### 2022-08-23 bayesian/bayesian.tex
Initial Overleaf Import

\newif\ifisarxiv
\isarxivtrue

% ready for submission
\ifisarxiv
  \documentclass[11pt]{article}
  \usepackage{fullpage}

  \usepackage{natbib}
\else
  \documentclass[twoside]{article}
  \usepackage{aistats2020}

  \usepackage{natbib}
  \renewcommand{\bibname}{References}
  \renewcommand{\bibsection}{\subsubsection*{\bibname}}
\fi

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{bm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{xfrac}
\usepackage{microtype}      % microtypography
\usepackage{color}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}

\input{shortdefs}
\newcommand{\mub}{\bm{\mu}}

\usepackage{grffile}

\begin{document}
\ifisarxiv
  \title{Bayesian experimental design using regularized \\
    determinantal point processes}
  \author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
  Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
    % examples of more authors
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \AND
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
  }
\else
  \twocolumn[
  \aistatstitle{Bayesian experimental design \\using regularized 
    determinantal point processes}
\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
}
  % \aistatsauthor{%
  %   Micha{\l } Derezi\'{n}ski
  %   \And
  %   Feynman Liang
  %   \And
  %   Michael W. Mahoney
  % }
  % \aistatsaddress {
  %   Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{mderezin@berkeley.edu}
  %   \And
  %   Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{feynman@berkeley.edu}
  %   \And
  %   ICSI and Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{mmahoney@stat.berkeley.edu}
  % } 
  ]
\fi

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\ifisarxiv
  \date{}
  \def\And{\and}
\fi

\ifisarxiv
  \maketitle
\fi

\begin{abstract}
We establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
 vectors to minimize a statistical optimality criterion.
 We show that a new regularized variant of
 DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. 
\end{abstract}

\section{Introduction}

Consider a collection of $n$ experiments parameterized by
$d$-dimensional vectors $\x_1,\dots,\x_n$, and let $\X$ denote the
$n\times d$ matrix with rows $\x_i^\top$. The outcome of the $i$th
experiment is a random variable $y_i = \x_i^\top\w + \xi_i$, where
$\w$ is the parameter vector of a linear model with prior distribution
$\Nc(\zero,\sigma^2\A^{-1})$, and $\xi_i\sim \Nc(0,\sigma^2)$ is
independent noise. In experimental design, we have access to the vectors 
$\x_i^\top$, for $i\in\{1,\ldots,n\} = [n]$, but we are allowed to
observe only a small number of outcomes $y_i$ for experiments we choose.
Suppose that we observe the outcomes from a subset
$S\subseteq [n]$ of $\lvert S \vert = k$ experiments. The
posterior distribution of $\w$ given $\y_S$ (the vector of outcomes in $S$) is:
\begin{align*}
  \w\mid \y_S \ &\sim\ \Nc(\mub, \Sigmab), \\
  \text{where } \mub &= (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
  \Sigmab &= \sigma^2(\X_S^\top\X_S+\A)^{-1}.
\end{align*}
Here $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
$i\in S$.

In Bayesian experimental design \citep{bayesian-design-review},
the prior precision matrix $\A$ is used to encode prior knowledge
and our goal is to choose $S$ so as to 
  minimize a function (a.k.a.~an optimality criterion) measuring the ``size''
  of the posterior covariance matrix
  $\Sigmab_{\w\mid\y_S} = \sigma^2(\X_S^\top\X_S+\A)^{-1}$.
Note that $\Sigmab_{\w\mid\y_S}$ is well defined even if $\A$ is not
invertible (i.e., an ``improper prior''). In particular, it includes classical
experimental design as the special case $\A = \zero$, as well as
the ridge-regularized case for $\A=\lambda\I$.
Denoting $\Sigmab$ as the subset covariance $\X_S^\top\X_S$, we will
use $f_{\A}(\Sigmab)$ to represent the following standard Bayesian 
optimality criteria \citep{bayesian-design-review,optimal-design-pukelsheim}:
\begin{enumerate}
\item A-optimality:\quad $f_{\A}(\Sigmab) = \tr\big((\Sigmab+\A)^{-1}\big)$;
\item C-optimality:\quad \!\!$f_{\A}(\Sigmab) = \cb^\top(\Sigmab+\A)^{-1}\cb$ for
$\cb \in \R^d$;
\item D-optimality:\quad $f_{\A}(\Sigmab) = \det(\Sigmab+\A)^{-1/d}$;
  \item V-optimality:\quad $f_{\A}(\Sigmab) =
    \frac1n\tr\big(\X(\Sigmab+\A)^{-1}\X^\top\big)$.
  \end{enumerate}
  Applications including clinical trials
  \citep{ryan2015fully,ding2008bayesian,spiegelhalter2004incorporating,berry2002adaptive,stangl1998bayesian,flournoy1993clinical},
  medical imaging \citep{owen2016optimisation},
  biological process models \citep{ryan2016optimal},
  and materials science
  \citep{frazier2016bayesian,ueno2016combo,terejanu2012bayesian}
  all use these optimality criteria and thus stand to benefit from
  our contributions.
  % Our algorithm can also be directly applied to improve
  % approximate design methods which leverage normal-based approximations
  % \citep{overstall2018approach}.

The general task we consider is the following combinatorial
optimization problem, where $[n]$
denotes $\{1,...,n\}$:

%\noindent
\textbf{Bayesian experimental design.}
Given an $n\times d$ matrix $\X$,
a criterion $f_{\A}(\cdot)$ and~$k\in[n]$,
efficiently compute or approximate
\begin{align*}
\argmin_{S \subseteq [n]} f_{\A}(\X_S^\top\X_S)\quad\text{subject
  to}\quad |S|=k.
\end{align*}
We denote the value at the optimal solution as $\opt$.
The prior work around this problem can be grouped into two research questions.
% categories, each asking a question about the optimum of the above
% problem, which we will denote $\opt =
% \min_{S:|S|=k}f_{\A}(\X_S^\top\X_S)$.
The first question asks when does there exist a
polynomial time algorithm for finding a $(1+\epsilon$)-approximation
for $\opt$.
The second question asks what we can
infer about $\opt$ just from the spectral
information about the problem, which is contained in the data covariance matrix
$\Sigmab_\X=\X^\top\X\in\R^{d\times d}$.

% We aim to address two research questions which have been outlined
% by the prior work on this topic:
%\noindent

\textbf{Question 1:} \quad
Given $\X$, $f_{\A}$ and $k$, can we efficiently find a $(1+\epsilon)$-approximation for
$\opt$?

\textbf{Question 2:} \quad
Given only $\Sigmab_{\X}$, $f_{\A}$ and $k$,
what is the upper bound on $\opt$?

A key aspect of both of these questions is how large the subset
size $k$ has to be for us to provide useful answers. As a baseline, we
should expect meaningful results when $k$ is at least $\Omega(d)$ \citep[see
discussion in][]{near-optimal-design}, and in fact,
for classical experimental design (i.e., when $\A=\zero$), the problem
becomes ill-defined 
when $k<d$. In the Bayesian setting we should be able to exploit the
additional prior knowledge 
to achieve strong results even for $k\ll d$. Intuitively, the larger
the prior precision matrix $\A$, the fewer degrees of freedom we have
in the problem. To measure this, we use the statistical notion of
\emph{effective dimension} \citep{ridge-leverage-scores}.
\begin{definition}
For $d\times d$ positive semi-definite (psd) matrices $\A$ and $\Sigmab$,
  let the $\A$-effective dimension of $\Sigmab$
be defined as $d_{\A}(\Sigmab) =
\tr\big(\Sigmab(\Sigmab+\A)^{-1}\big) \leq d$.
We will use the shorthand $d_{\A}$ when referring to $d_{\A}(\Sigmab_{\X})$.
\end{definition}
\cite{klivans-goel17} showed that $d_\A$ can be orders of
  magnitude smaller than the 
actual dimension $d$ when the eigenvalues of $\Sigmab_\X$ exhibit fast
decay, which is often the case in real datasets
\citep{revisiting-nystrom}. Recently,
\cite{regularized-volume-sampling} obtained bounds on Bayesian 
A/V-optimality criteria for $k\geq d_{\A}$, suggesting that $d_{\A}$
is the right notion of degrees of freedom for this problem.

\subsection{Main results}
Our main results provide new answers to Questions 1 and 2
by proposing a novel algorithm for Bayesian experimental design with
strong theoretical guarantees.

  \paragraph{Answer to Question 1.}
We propose an efficient $(1+\epsilon)$-approximation algorithm for
A/C/D/V-optimal Bayesian  experimental design:
\begin{theorem}\label{t:q2}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be $n\times d$. If
  $k=\Omega\big(\frac{d_{\A}}{\epsilon} +
  \frac{\log1/\epsilon}{\epsilon^2}\big)$ for some $\epsilon\in(0,1)$, then 
we can find in polynomial time a subset $S$ of size $k$ s.t.
  \begin{align*}
    f_{\A}\big(\X_S^\top\X_S\big) \leq (1+\epsilon)\cdot\opt.
  \end{align*}
\end{theorem}
\begin{remark}
  The algorithm referred to in Theorem~\ref{t:q2} first solves a convex
  relaxation of the task via a semi-definite program (SDP) to find a
  weight vector $p\in[0,1]^n$, then uses our new randomized algorithm 
  to round the weights to $\{0,1\}$, obtaining the subset
  $S$. The expected cost after SDP is $O(ndk+k^2d^2)$.
\end{remark}
% Note that, unlike in Theorem \ref{t:q1}, we use the unrescaled effective
% dimension instead of the rescaled one, $d_{\frac nk\!\A}$. The
% actual effective dimension that applies here (given in the proof in
% Section \ref{s:guarantees}) depends on the SDP solution. It is
% always upper bounded by $d_{\A}$, but it may be significantly
% smaller.

A number of recent works studied $(1+\epsilon)$-approximate SDP-based algorithms 
for classical and Bayesian experimental design (see Table
\ref{tab:comp} and Section \ref{s:related-work} for a comparison). Unlike \emph{all} prior work on
this topic, we are able to eliminate the
dependence of the subset size $k$ on the dimension $d$, replacing it
with the potentially much smaller effective dimension $d_\A$.
Our result also improves over the existing approaches in terms of the computational
cost of the rounding procedure that is performed after solving the SDP. A
number of different methods can be used to solve the SDP relaxation (see
Section \ref{s:experiments}). For
example, \cite{near-optimal-design} suggest using an iterative
optimizer called entropic mirror descent, which is known to exhibit
fast convergence and can run in $O(nd^2T)$ time, where $T$ is the
number of iterations.

\begin{figure*}[htbp]
  \vspace{-4mm}
  \centering
  \renewcommand{\arraystretch}{1.5}
\begin{tabular}{r||c|c|l|l}
  &Criteria&Bayesian& $k$ %& dep.~on $d$ & dep.~on $1/\epsilon$
  & Cost after SDP\\%of rounding\\
  \hline\hline
\small  \cite{tractable-experimental-design}
  &\small A,V&\xmark  &$d^2/\epsilon$
  % &\Red{quadratic}& \Orange{linear}
  &$n^2\cdot d$\\
\small  \cite{near-optimal-design}
  &\small A,C,D,E,G,V&\cmark
&$ d/\epsilon^2$
% &\Orange{linear}& \Red{quadratic}
  &$n\cdot kd^2$\\ 
\small  \cite{proportional-volume-sampling}
  &\small A,D&\xmark &$d/\epsilon$
  % & \Orange{linear}& \Orange{linear}
  & $n^4\cdot k^2d$\\ 
  \hline
  \small\textbf{this paper}&\small A,C,D,V&\cmark
    & $d_{\A}/\epsilon$
    % &\Green{none}& \Orange{linear}
  & $n\cdot kd + k^2d^2$
\end{tabular}
\vspace{2mm}
\captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
  classical and Bayesian experimental design (X-mark means that only
  the classical setting applies).
In the cost analysis, $n$ could be replaced by the number of
non-zero weights in the SDP solution. For simplicity we omit the log
terms and assume that $\epsilon = \Omega(\frac1{d_{\A}})$. Our
approach beats other methods both in terms of the
runtime and the dependence of $k$ on $d$ (when $d_\A=o(d)$).} 
\label{tab:comp}
\vspace{-3mm}
\end{figure*}

\paragraph{Answer to Question 2.}
%Formalizing the above intuitions,
By performing a careful theoretical analysis of the performance of our
algorithm, we are able to give an improved upper bound on $\opt$. In the below
result, we use a more refined notion of effective dimensionality for
Bayesian experimental design,
$d_{\frac nk\A}$ (where the precision matrix $\A$ is scaled by factor $\frac
nk$), which is smaller than $d_\A$ and therefore leads 
to a tighter bound.
% $d_{\frac nk\A}\leq d_{\A}$ as the notion of 
% effective dimensionality, which is an improvement over the prior work
% We now present our first main result, which provides an analytic bound
% on the four optimality criteria listed above (Question 1) for most
% practically relevant values of $k$.
\begin{theorem}\label{t:q1}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be
  $n\times d$. For any $k$ such that $k\geq 4d_{\frac
    nk\!\A}$, %we have
  \begin{align*}
\opt    \leq \bigg(\ 1+8\,\frac{d_{\frac
    nk\!\A}}{k} + 8\sqrtshort{\frac{\ln (k/d_{\frac nk\!\A})}{k}}\
    \bigg)\cdot f_{\A}\big(\tfrac kn\Sigmab_{\X}\big).
  \end{align*}
\end{theorem}
\begin{remark}
  We give a (randomized) algorithm which (with probability 1) finds
  the subset $S$ that certifies this bound and has expected time
  complexity $O(ndk+k^2d^2)$. 
  % and is $O(t (n d^2 + k d^2))$ time with
  % probability exponentially increasing in $t$.
  % Michal: please check this statement.
\end{remark}
  In particular, this means that if $k\geq 4d_{\frac nk\!\A}$
  then there is $S$ of size $k$
  which satisfies $f_{\A}(\X_S^\top\X_S) = O(1)\cdot f_{\A}(\frac kn\Sigmab_{\X})$.
  This not only improves on \cite{regularized-volume-sampling} in terms
  of the supported range of sizes $k$, but also in terms of the obtained bound (see
  Section \ref{s:related-work} for a~comparison). In Section
  \ref{s:experiments}, we we provide numerical evidence suggesting
  that for many real datasets the quantity $f_{\A}(\frac
  kn\Sigmab_{\X})$ provides a good estimate for $\opt$ to within a
  factor of 2.

Theorem \ref{t:q1} suggests that the right notion of degrees of
freedom for Bayesian experimental design can in fact be smaller
than $d_{\A}$.  
Intuitively, since $d_{\A}$ is computed using the full data covariance
$\Sigmab_{\X}$, it is not in the same scale as the smaller covariance
$\X_S^\top\X_S$ based on the subset $S$ of size $k\ll n$. In our
result this is corrected by increasing the 
regularization on $\Sigmab_\X$ from $\A$ to $\frac{n}{k}\A$ and using
$d_{\frac nk\A}=d_{\frac{n}{k}\A}(\Sigmab_\X)$ as the degrees of
freedom. 
Note that $d_{\frac nk\!\A}\leq d_{\A}$ and this gap can be very large for some
problems (see discussion in Appendix~\ref{a:deff}).

\subsection{Technical contributions}
  To establish Theorems \ref{t:q2} and \ref{t:q1}, 
  we develop a theoretical framework for a new sampling distribution
  % $\DPPreg{p}(\X,\A)$, where 
  % $p=(p_1,...,p_n)\in[0,1]^n$ is a vector of weights. 
  which can be seen as a \emph{regularized} variant of a determinantal point
  process (DPP). DPPs are a well-studied family of distributions
with numerous applications in sampling diverse subsets
  of negatively correlated elements \citep[see][]{dpp-ml}.

  Given a psd matrix $\A$ and a weight vector $p=(p_1,...,p_n)\in[0,1]^n$, we
define $\DPPreg{p}(\X,\A)$ as a distribution over subsets $S\subseteq[n]$ (of all
sizes) such that (see Definition~\ref{d:r-dpp}):
\begin{align*}
 \Pr(S)\propto \det(
  \X_S^\top\X_S+\A)\ \cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in
  S}(1-p_i).
\end{align*}
A number of regularized DPPs have been proposed recently
\citep{dpp-intermediate,regularized-volume-sampling}, mostly within the context
of Randomized Numerical Linear Algebra (RandNLA)~
\citep{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR}.
To our knowledge, ours is the first such definition that strictly falls
under the umbrella of traditional DPPs \citep{dpp-ml}.
We show this in Section \ref{s:r-dpp}, where we also prove that
regularized DPPs can be decomposed into a low-rank DPP plus
i.i.d. Bernoulli sampling (Theorem \ref{t:algorithm}). This decomposition reduces
the sampling cost from $O(n^3)$ to $O(nd^2)$,
and involves a more general result about DPPs defined via a
correlation kernel (Lemma~\ref{l:decomposition}), which is of
independent interest.

In Section \ref{s:guarantees} we demonstrate a fundamental connection between an
$\A$-regularized DPP and Bayesian experimental design with precision
matrix $\A$. For simplicity of exposition, let the weight vector $p$ be uniformly equal $(\frac kn,...,\frac kn)$. If
$S\sim\DPPreg{p}(\X,\A)$ and $f_{\A}$ is any one of the
A/C/D/V-optimality criteria, then:
 \begin{align*}
   \E\big[f_{\A}(\X_S^\top\X_S)\big]
   \leq f_{\A}\big(\tfrac kn\Sigmab_{\X}\big)
   \quad\text{and}\quad \E\big[|S|\big]
   \leq d_{\frac nk\!\A}+k. 
 \end{align*}
The proof of Theorem \ref{t:q1} relies on these two inequalities and a
concentration bound for the subset size $|S|$,
whereas to obtain Theorem
\ref{t:q2} we additionally use the SDP relaxation to find the optimal
weight vector~$p$. When $\A=\zero$, then $\DPPreg{p}(\X,\A)$ bears a
lot of similarity to 
\emph{proportional volume sampling} which is an (unregularized) determinantal
distribution proposed by \cite{proportional-volume-sampling}.
Our algorithm not only extends it to the Bayesian setting but also
offers a drastic time complexity improvement from 
the $O(n^4dk^2\log k)$ required by \cite{proportional-volume-sampling}
down to the $O(nd^2)$ required for sampling from $\DPPreg{p}(\X,\A)$,
and recent advances in RandNLA for DPP sampling
\citep{leveraged-volume-sampling,correcting-bias,dpp-intermediate} suggest that
$O(nd\log n + \poly(d))$ time is also possible.

\section{Related work}
\label{s:related-work}

A number of works proposed $(1+\epsilon)$-approximation algorithms
for experimental design which start with solving a convex
relaxation of the problem, and then use some rounding strategy to
obtain a discrete solution (see Table \ref{tab:comp} for comparison).
In this line of work we wish to find the smallest $k$ for which a
polynomial time approximation algorithm is possible. For example,
\cite{tractable-experimental-design} gave an approximation algorithm
for classical A/V-optimality with $k=\Omega(\frac{d^2}{\epsilon})$, where the
rounding is done in a greedy fashion, and some randomized rounding
strategies are also discussed. \cite{proportional-volume-sampling}
suggested \emph{proportional volume sampling} for the rounding step
and obtained approximation algorithms for classical A/D-optimality with
$k=\Omega(\frac d\epsilon+\frac {\log1/\epsilon}{\epsilon^2})$. Their
approach is particularly similar to ours (when $\A=\zero$). However,
as discussed earlier, while their algorithms run in polynomial time,
they scale very poorly with the number of experiments $n$ (see Table 
\ref{tab:comp}). \cite{near-optimal-design} proposed  
an efficient algorithm with a $(1+\epsilon)$-approximation guarantee for a wide
range of optimality criteria, including A/C/D/E/V/G-optimality, both classical and Bayesian, when
$k=\Omega(\frac d{\epsilon^2})$.  Our results (in Theorem \ref{t:q2}) improve on this work in
two important ways: 
\begin{itemize}
  \item In terms of the dependence on $\epsilon$ for A/C/D/V-optimality,
  \item In terms of the dependence on the dimension 
  (by replacing $d$ with $d_{\A}$) in the Bayesian setting.
\end{itemize}
A lower bound shown by \cite{proportional-volume-sampling} implies that our
Theorem \ref{t:q2} cannot be
directly extended to E-optimality, but a similar lower bound does not
exist for G-optimality. We remark that the approximation approaches
relying on a convex relaxation can generally be converted to
an upper bound on $\opt$ akin to our Theorem \ref{t:q1},
however, unlike our bound, none of them apply to the regime of $k\leq d$.

Non-trivial bounds for the \emph{classical} A-optimality criterion
(i.e., $\opt$
with $\A=\zero$) were first given by \cite{avron-boutsidis13}, where
they show that
for any $k\geq d$, $\opt \leq (1+ \frac{d-1}{k-d+1})\cdot f_{\zero}(\frac kn\Sigmab_\X)$
and the subset $S$ attaining the bound can be found in polynomial time.
The result was later extended \citep{unbiased-estimates,
regularized-volume-sampling,unbiased-estimates-journal} to the case
where $\A=\lambda\I$, proving that for any $k\geq d_{\lambda\I}$,
we have
$\opt\leq (1 +\frac{d_{\lambda\I}-1}{k-d_{\lambda\I}+1})\cdot f_{\frac
  kn\!\lambda\I}(\frac kn\Sigmab_\X)$,
and also a faster $O(nd^2)$ time algorithm was provided.
In comparison, our results (in Theorem \ref{t:q1}) offer the following
improvements for upper bounding $\opt$:
\begin{itemize}
  \item We cover a wider range of subset sizes, because
    $d_{\frac nk\!\lambda\I}\leq d_{\lambda\I}$,
  \item Our upper bound can be much tighter because
    $f_{\lambda\I}(\frac kn\Sigmab_\X)\leq f_{\frac kn\!\lambda\I}(\frac
    kn\Sigmab_\X)$. 
\end{itemize}
Additionally, \cite{minimax-experimental-design} propose a new notion of
\emph{minimax} experimental design, which is related to A/V-optimality.  They
also use a determinantal distribution for subset selection, however, due to
different assumptions, their bounds are incomparable.

Purely greedy approximation algorithms have been shown to provide
guarantees in a number of special cases for experimental design. One
example is classical D-optimality criterion, which can be converted to
a submodular function \citep{submodularity-optimal-design}. Also, greedy algorithms
for Bayesian A/V-optimality criteria have been considered by
\cite{greedy-supermodular} and \cite{greedy-graph-sampling}. These
methods can only provide a constant factor approximation guarantee
(as opposed to $1+\epsilon$), and the factor is generally problem
dependent (which means it could be arbitrarily large).
Finally, a number of
heuristics with good empirical performance have been proposed, such as Fedorov's
exchange method \citep{cook1980comparison}. However, in this work
we focus on methods that provide theoretical approximation guarantees.
% In Section \ref{s:experiments} we experimentally show that
% our regularized DPP sampling (without solving an SDP) performs
% better than iid sampling methods like uniform and predictive length
% sampling \cite{zhu2015optimal} and that by using the SDP solution
% we can improve it to match the performance of
% greedy algorithms for A-optimal design \cite{chamon2017approximate,greedy-supermodular}
% which empirically have been found to perform well on real world data.

\section{A new regularized determinantal point process}
\label{s:r-dpp}
  In this section we develop the theory for a novel regularized extension of
  determinantal point processes (DPP) which we use as the sampling distribution
  for obtaining guarantees in Bayesian experimental design.
  DPPs form a family of distributions which are used to model repulsion between
  elements in a random set, with many applications in machine learning
  \citep{dpp-ml}. Here, we focus on the setting where we are sampling out of all
  $2^n$ subsets $S\subseteq[n]$. Traditionally, a DPP is defined by a
  correlation kernel, which is an $n\times n$ psd matrix $\K$ with eigenvalues
  between 0 and 1, i.e., such that $\zero\preceq\K\preceq \I$. Given a
  correlation kernel $\K$, the corresponding DPP is
  defined as
  \begin{align*}
    S\sim\DPPcor(\K)\quad\text{iff}\quad
    \Pr(T\subseteq S) = \det(\K_{T,T})\ \  \forall_{T\in[n]},
  \end{align*}
where $\K_{T,T}$ is the submatrix of $\K$ with rows and columns
indexed by $T$. Another way of defining a DPP, popular in the machine
learning community, is via an ensemble kernel $\L$. Any psd matrix
$\L$ is an ensemble kernel of a DPP defined as:
  \begin{align*}
    S\sim\DPPens(\L)\quad\text{iff}\quad \Pr(S) \propto \det(\L_{S,S}).\quad~
  \end{align*}
Crucially, every $\DPPens$ is also a $\DPPcor$, but not the other way
around. Specifically, $\DPPens(\L)=\DPPcor(\K)$ when:
\begin{align*}
  &\text{(a)}\ \  \L=\K(\I-\K)^{-1},\qquad
  &\text{(b)}\ \  \K = \I -  (\I+\L)^{-1},
\end{align*}
but (a) requires that $\I-\K$ be invertible which is not true
for some DPPs.
(This will be important in our analysis.)
The classical algorithm for sampling from a DPP requires the
eigendecomposition of either matrix $\K$ or $\L$, which in general
costs $O(n^3)$, followed by a sampling procedure which costs
$O(n\,|S|^2)$ \cite{dpp-independence,dpp-ml}.

We now define our regularized DPP and describe its
connection with correlation and ensemble~DPPs.
\begin{definition}\label{d:r-dpp}
Given matrix $\X\in\R^{n\times d}$, a sequence  $p=(p_1,\dots,p_n)\in[0,1]^n$
and a psd matrix $\A\in\R^{d\times d}$ such that
$\sum_ip_i\x_i\x_i^\top\!+\A$ is
full rank, let
$\DPPreg{p}(\X,\A)$ be a distribution over $S\subseteq [n]$:
  \begin{align}
  \Pr(S) = \frac{\det(
  \X_S^\top\X_S+\A)}{\det\!\big(
 \sum_ip_i\x_i\x_i^\top+ \A\big)}
\cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in S}(1-p_i).\label{eq:poisson-prob}
\end{align}
%   \Pr(\pi)\ \propto\ \det\!\big(\A +
%   \X_{\pi}^\top\X_{\pi}\big)
%   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
% \end{align*}
\end{definition}
The fact that this is a proper distribution (i.e., that it sums to
one) can be restated as a determinantal expectation formula: if
$b_i\sim\mathrm{Bernoulli}(p_i)$ are independent Bernoulli random
variables, then
\begin{align*}
  &\sum_{S\subseteq [n]}\!\det(
  \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
  &=
  \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
  \det\!\Big(\sum_i\E[b_i]\x_i\x_i^\top\!+\A\Big),
\end{align*}
where $(*)$ was shown by \cite[Lemma
7]{determinantal-averaging} and $\E[b_i]=p_i$.
% follows from a recently shown determinantal expectation formula.
% \begin{lemma}[\cite{determinantal-averaging}]\label{l:cb}
% For $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$
% and a random set $S\subseteq [n]$ drawn with independent Bernoullis,
% \begin{align*}
%   \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)\Big]
%   = \det\!\Big( \E\big[\X_S^\top\X_S\big]+\A\Big).
% \end{align*}
% \end{lemma}

  The main theoretical contribution in this section is the
  following efficient algorithm for $\DPPreg{p}(\X,\A)$ which reduces it to
  sampling from a correlation DPP and unioning with i.i.d.~Bernoulli samples:
\begin{theorem}\label{t:algorithm}
For any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
full rank, let
\begin{align*}
  T&\sim\DPPcor\big(\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big),
  \quad\\
  &\quad\text{where}\quad\D_p=\diag(p).
\end{align*}
If $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random variables, then
$T\cup \{i:b_i\!=\!1\}\sim \DPPreg{p}(\X,\A)$.
\end{theorem}

\begin{remark}
    Figure~\ref{fig:algorithm-pseudocode} illustrates how to exploit this result
    to build an efficient sampling algorithm.
  Since the correlation kernel matrix has rank at most $d$, the preprocessing
  cost of eigendecomposition is $O(nd^2)$. Then, each sample costs only $O(n\,|T|^2)$.
\end{remark}

We prove the theorem in three steps. 
First, we express $\DPPreg{p}(\X,\A)$ as an ensemble DPP, which requires some
additional assumptions on $\A$ and $p$ to be possible.
Then, we convert the ensemble to a correlation kernel (eliminating the extra
assumptions),
and finally show that this kernel can be decomposed into a rank $d$ kernel plus
Bernoulli sampling.
In the process, we establish several novel theoretical properties
regarding the representation, decomposition, and closure properties
of regularized DPPs which may be of independent interest.

\renewcommand{\thealgorithm}{}
\floatname{algorithm}{}
\vspace{-3mm}
\begin{figure}[htbp]
\begin{center}
  \begin{algorithm}[H]
    {\fontsize{9}{9}\selectfont
      \caption{\bf \small Sampling \ $S\sim\DPPreg{p}(\X,\A)$}
      \begin{algorithmic}[0]
        \STATE \textbf{Input:} $\X\!\in\!\R^{n\times d}\!,\text{ psd } \A\!\in\!\R^{d\times
        d}\!, p\!\in\![0,1]^n$\\[1mm]
        \STATE Compute $\Z \leftarrow \A+\X^\top\D_p\X$\\[1mm]
        \STATE Compute \textsc{SVD} of  $\B=\D_p^{\sfrac12}\X\Z^{-\sfrac12}$\\[1mm]
        \STATE Sample $T\sim\DPPcor(\B\B^\top)$ \hfill \citep{dpp-independence}\\[1mm]
        \STATE Sample $b_i\sim\mathrm{Bernoulli}(p_i)$ for $i\in[n]$\\[1mm]
        \RETURN $S=T\cup\{i:b_i=1\}$
      \end{algorithmic}
    }
  \end{algorithm}
\end{center}
\vspace{-3mm}
\caption{Algorithm which exploits
  Theorem~\ref{t:algorithm} to sample $S \sim \DPPreg{p}(\X, \A)$
  in $O( nd^2 )$ time.}
\label{fig:algorithm-pseudocode}
\end{figure}

\vspace{2mm}
\begin{lemma}\label{t:reduction}
  Given $\X$, $\A$ and $\D_p$ as in Theorem \ref{t:algorithm}, assume that $\A$ and $\I-\D_p$ are
  invertible. Then,
  \begin{align*}
    \DPPreg{p}(\X,\A)&=\DPPens\big(\Dbt+
    \Dbt^{\sfrac12}\X\A^{-1}\X^\top
    \Dbt^{\sfrac12}\big),\quad\\
    \text{where}
    \quad\Dbt &= \D_p(\I-\D_p)^{-1}.
  \end{align*}
\end{lemma}
\begin{proof}
  Let $S\sim \DPPreg{p}(\X,\A)$. By Definition \ref{d:r-dpp} and
  the fact that $\det(\A\B+\I)=\det(\B\A+\I)$,
  \begin{align*}
    \Pr(S) &\propto \det(\X_S^\top\X_S+\A)\cdot
\prod_{i\in S}p_i\cdot\prod_{i\not\in S}(1-p_i) \\
&= \det(\X_S^\top\X_S+\A)\cdot
\prod_{i\in S}\frac{p_i}{1-p_i}\cdot\prod_{i=1}^n(1-p_i)
\\ &\propto\det\!\big(\A(\A^{-1}\X_S^\top\X_S+\I)\big) \det(\Dbt_{S,S}) \\
     &=\det(\A)\det(\A^{-1}\X_S^\top\X_S+\I) \det(\Dbt_{S,S})
\\ &\propto \det(\X_S \A^{-1}\X_S^\top+\I) \det(\Dbt_{S,S}) \\
&=\det\!\Big(\big[\Dbt^{\sfrac12}\X \A^{-1}\X^\top\Dbt^{\sfrac12}+\Dbt\big]_{S,S}\Big),
  \end{align*}
which matches the definition of the L-ensemble DPP.
\end{proof}
 At this point, to sample from  $\DPPreg{p}(\X,\A)$, we could simply
invoke any algorithm for sampling from
an ensemble DPP. However, this would only work for invertible
$\A$, which in particular excludes the important case of
$\A=\zero$ corresponding to classical experimental
design. Moreover, the standard algorithm would require computing the
eigendecomposition of the ensemble kernel, which (at
least if done na\"ively) costs $O(n^3)$. Even after this is done, the
sampling cost would still be $O(n\,|S|^2)$ which can be considerably
more than $O(nd^2)$. We first address the issue of invertibility of matrix
$\A$ by expressing our distribution via a correlation DPP.
\begin{lemma}\label{l:correlation}
  Given $\X$, $\A$, and $\D_p$ as in Theorem \ref{t:algorithm} (without
  any additional assumptions), we have
  \begin{align*}
    &\DPPreg{p}(\X,\A) 
    = \DPPcor\big(\D_p + \\
    &\ (\I\!-\!\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A\!+\!\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}(\I\!-\!\D_p)^{\sfrac12}\big).
    \end{align*}
  \end{lemma}
  When $\A$ and $\I-\D_p$ are invertible, then the proof (given in
  Appendix \ref{a:proofs}) is a straightforward calculation.
Then, we use a limit argument with $p_\epsilon=(1-\epsilon)p$ and
$\A_\epsilon=\A+\epsilon\I$, where $\epsilon\rightarrow 0$.

Finally, we show that the correlation DPP arrived at in Lemma
\ref{l:correlation} can be decomposed into a smaller DPP plus
Bernoulli sampling. In fact, in the following lemma we obtain a more
general recipe for combining DPPs with Bernoulli sampling, which may
be of independent interest. Note that if $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random
variables then $\{i:b_i\!=1\!\}\sim\DPPcor(\D_p)$.
\begin{lemma}\label{l:decomposition}
  Let $\K$ and $\D$ be $n\times n$ psd matrices with eigenvalues between
0 and 1, and assume that $\D$ is diagonal. If\, $T\!\sim\DPPcor(\K)$ and
$R\sim\DPPcor(\D)$, then
\begin{align*}T\cup R\sim \DPPcor\big(\D+(\I-\D)^{\sfrac12}\K
  (\I-\D)^{\sfrac12}\big).
  \end{align*}
\end{lemma}
The lemma is proven in Appendix \ref{a:proofs}. Theorem
\ref{t:algorithm} now follows by combining Lemmas \ref{l:correlation} and \ref{l:decomposition}.

\section{Guarantees for Bayesian experimental design}
\label{s:guarantees}
In this section we prove our main results regarding Bayesian
experimental design (Theorems \ref{t:q2} and \ref{t:q1}). First, we
establish certain properties of the regularized DPP distribution that
make it effective in this setting. Even though the size of the
sampled subset $S\sim\DPPreg{p}(\X,\A)$ is random and can be as large
as $n$, it is also highly concentrated around its expectation, which
can be bounded in terms of the $\A$-effective dimension. This is
crucial, since both of our main results require a subset of
deterministically bounded size. Recall that the effective dimension is
defined as a function $d_{\A}(\Sigmab) =
\tr\big(\Sigmab(\A+\Sigmab)^{-1}\big)$. The omitted proofs are in
Appendix \ref{a:proofs}.
\begin{lemma}\label{l:size}
  Given any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
full rank, let $S=T\cup \{i:b_i=1\}\sim \DPPreg{p}(\X,\A)$ be defined
as in Theorem \ref{t:algorithm}. Then
\begin{align*}
  \E\big[|S|\big] &\leq \E\big[|T|\big] + \E\Big[\sum_i b_i\Big]  \\
                  & = d_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big) + \sum_i p_i.
\end{align*}
\end{lemma}
Next, we show two expectation inequalities for the matrix inverse and
matrix determinant, which hold for the regularized DPP.
We use them to bound the Bayesian optimality criteria in expectation.
% These are
% immediately followed by a corollary applying them to bound
% Bayesian optimality criteria.
\begin{lemma}\label{t:expectations}
Whenever $S\sim \DPPreg{p}(\X,\A)$ is a well-defined distribution it holds that
\begin{align}
  \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
  &\preceq \Big(
    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1},\label{eq:sqinv}
  \\
\E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
  &\leq\det\!\Big(
    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1}.\label{eq:det}
\end{align}
\end{lemma}
\begin{corollary}\label{c:expectations}
Let $f_{\A}$ be A/C/D/V-optimality. Whenever $S\sim\DPPreg{p}(\X,\A)$ is
well-defined,
\begin{align*}
  \E\big[f_{\A}(\X_S^\top\X_S)\big] \leq f_{\A}\Big(\sum_i p_i\x_i\x_i^\top\Big).
\end{align*}
\end{corollary}
\begin{proof}
  In the case of A-, C-, and V-optimality, the function $f_{\A}$ is a
  linear transformation of the matrix $(\X_S^\top\X_S+\A)^{-1}$ so the
  bound follows from \eqref{eq:sqinv}. For D-optimality, we apply
  \eqref{eq:det} as follows:
  \begin{align*}
    \E\big[f_{\A}(\X_S^\top\X_S)\big]
    & = \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-\sfrac1d}\Big] \\
    &\leq
     \E\bigg[\Big(\det\!\big(\X_S^\top\X_S+\A\big)^{-\sfrac1d}\Big)^d\bigg]^{\sfrac1d} \\
    \\ &=  \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]^{\sfrac1d}\\
       &=
         \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{-\sfrac1d},
  \end{align*}
  which completes the proof.
\end{proof}
Finally, we present the key lemma that puts everything together. This
result is essentially a generalization of Theorem \ref{t:q1} from
which also follows Theorem \ref{t:q2}.
\begin{lemma}\label{l:guarantees}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be $n\times
  d$.  For some $w=(w_1,\dots,w_n)\in[0,1]^n$, let
  $\Sigmab_w=\sum_iw_i\x_i\x_i^\top$ and assume that
  $\sum_i w_i=k\in[n]$. If $k\geq 4\,d_{\A}(\Sigmab_w)$, then a
subset $S\subseteq[n]$ of size $k$ can be found in $O(ndk+k^2d^2)$ time that satisfies
  \begin{align*}
    &f_{\A}\big(\X_S^\top\X_S\big)  \\
    &\leq \bigg(\ 1+8\,\frac{d_{\A}(\Sigmab_w)}{k} + 8\sqrtshort{\frac{\ln (k/d_{\A}(\Sigmab_w))}{k}}\ \bigg)\cdot f_{\A}\big(\Sigmab_w\big).
  \end{align*}
  \end{lemma}
\begin{proof}
  Let $p=(p_1,\dots,p_n)$ be defined so that $p_i =
  \frac{w_i}{1+\epsilon}$, and suppose that $S\sim
  \DPPreg{p}(\X,\A)$. Then, using Corollary \ref{c:expectations}, we have
  \begin{align*}
\Pr&\big(|S|\leq k\big)\,\E\Big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\Big]\\
&\leq
\E\big[f_{\A}(\X_S^\top\X_S)\big] \\
&\leq
f_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big)
\\ &\leq (1+\epsilon)\cdot f_{\A}\Big(\sum_iw_i\x_i\x_i^\top\Big).
  \end{align*}
  Using Lemma \ref{l:size} we can bound the expected size of $S$ as
follows:
\begin{align*}
  \E\big[|S|\big]
  &\leq d_{\A}(\Sigmab_w) + \sum_i p_i \\
  &=d_{\A}(\Sigmab_w)+\frac{k}{1+\epsilon}\\
  &=k\cdot \Big(1+\frac {d_{\A}(\Sigmab_w) }k - \frac{\epsilon}{1+\epsilon}\Big).
\end{align*}
Let $d_w = d_{\A}(\Sigmab_w)$ and $\alpha=1+\frac {d_w} k -\frac{\epsilon}{1+\epsilon}$.
If $1\geq\epsilon\geq \frac{4d_w}{k}$, then $\alpha\leq 1 +
\frac\epsilon4-\frac\epsilon2=1-\frac\epsilon4$. Since $\DPPreg{p}(\X,\A)$
is a  determinantal point process, $|S|$ is a
Poisson binomial r.v.~so for $\epsilon\geq 6\sqrt{\frac{\ln(k/d_w)}{k}}$,
\begin{align*}
  \Pr(|S|>k) \leq \ee^{-\frac{(k-\alpha k)^2}{2k}} = \ee^{-\frac
  k2(1-\alpha)^2}\leq \ee^{-\frac{k\epsilon^2}{32}}\leq \frac{d_w}k.
\end{align*}
% If $k\geq\frac{4 d_{\A}(\Sigmab_w)}\epsilon$,
% then $\alpha \leq 1 + \frac\epsilon4 -\frac\epsilon2=1-\frac\epsilon4$.
%  Since $\Vol_p(\X,\A)$ is a determinantal point process, Lemma
% \ref{l:concent} implies that for $\epsilon'=\frac\epsilon4$ and $k\geq
%  48\frac{\ln 1/\epsilon'}{\epsilon'^2}$ we have
% \begin{align*}
% \Pr(|S|> k)\leq  \Pr\big(|S|\ \geq\ \alpha k + \epsilon' k\big)\leq
%   3\ee^{-\frac{\epsilon'^2k^2}{16(\epsilon'k+2\alpha k)}} \leq
%   3\ee^{-\frac{\epsilon'^2k}{48}}\leq 3\epsilon'.
% \end{align*}
For any $\epsilon\geq
4\,\frac{d_w}k + 6\sqrt{\frac{\ln(k/d_w)}{k}}$, we have
\begin{align*}
\E&\big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\big] \\
&\leq \frac{1+\epsilon}{1-\frac{d_w}k}\cdot f_{\A}(\Sigmab_w)\\
  &\leq
  \bigg(1+\frac{\epsilon+\frac{d_w}k}{1-\frac{d_w}k}\bigg)\cdot
f_{\A}(\Sigmab_w)
\\ &\leq \bigg(1+7\,\frac{d_w}{k} +
  8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\cdot f_{\A}(\Sigmab_w).
\end{align*}
Denoting $\E\big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\big]$ as $F_k$,
Markov's inequality implies that
\[\Pr\Big(f_{\A}(\X_S^\top\X_S) \geq
  (1+\delta) F_k\ \mid\ |S|\leq k\Big)\leq \frac1{1+\delta}.\]
Also, we showed
that $\Pr(|S|\leq k)\geq 1-\frac{d_w}{k}\geq \frac34$. Setting $\delta=\frac{d_w}{C
  k}$ for sufficiently large $C$ we obtain that with probability
$\Omega(\frac{d_w}{k})$, the random set $S$ has size at most $k$ and
\begin{align*}
f_{\A}&(\X_S^\top\X_S) \\
&\leq \bigg(1+\frac{d_w}{C k}\bigg)\!\cdot\! \bigg(1+7\,\frac{d_w}{k} +
  8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\!\cdot\! f_{\A}(\Sigmab_w)\\
&\leq
\bigg(1+8\,\frac{d_w}{k} +
  8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\cdot f_{\A}(\Sigmab_w).
\end{align*}
We can sample from $\DPPreg{p}(\X,\A)$ conditioned on
$|S|\leq k$ and $f_{\A}(\X_S^\top\X_S)$ bounded as above by rejection
sampling. When $|S|<k$,  the set is completed to
$k$ with arbitrary indices. On average, $O(\frac k{d_w})$ samples from
$\DPPreg{p}(\X,\A)$ are needed, so the cost is $O(nd^2)$ for the
eigendecomposition, $O(\frac{k}{d_w}\cdot nd_w^2)=O(nd_wk)$ for
sampling and $O(\frac k{d_w}\cdot kd^2)$ for recomputing $f_{\A}(\X_S^\top\X_S)$.
\end{proof}
To prove the main results, we use Lemma \ref{l:guarantees} with
appropriately chosen weights $w$.
\begin{proofof}{Theorem}{\ref{t:q2}}
  As discussed by \cite{near-optimal-design} and
  \cite{boyd2004convex}, the following convex 
relaxation of experimental design can be written as a semi-definite
program and solved using standard SDP solvers:
\begin{align}
  &w^* \ = \  \argmin_w\ \
  f_{\A}\Big(\sum_{i=1}^nw_i\x_i\x_i^\top\Big),\quad\\
  &\text{subject to}\quad
  \forall_i\ \ 0\leq w_i\leq 1,\quad \sum_i w_i=k.\label{eq:sdp}
\end{align}
The solution $w^*$ satisfies $f_{\A}\big(\Sigmab_{w^*}\big)\leq
\opt$. If we use $w^*$ in Lemma \ref{l:guarantees}, then observing
that $d_{\A}(\Sigmab_{w^*})\leq d_{\A}$, and setting $k\geq C(\frac
{d_{\A}}{\epsilon} + \frac{\log1/\epsilon}{\epsilon^2})$ for
sufficiently large $C$, the algorithm in the lemma finds subset $S$
such that
\[f_{\A}(\X_S^\top\X_S)\leq (1+\epsilon)\cdot
f_{\A}(\Sigmab_{w^*})\leq
(1+\epsilon)\cdot\opt.\]
% \cite{tractable-experimental-design} show that
% the support of $w^*$ has size at most $k+d^2$, which reduces the cost of
% our sampling to $O(k^2d^2)$.
Note that we did not need to
solve the SDP exactly, so approximate solvers could be used instead.
\end{proofof}
\begin{proofof}{Theorem}{\ref{t:q1}}
Let $w=(\frac kn,...,\frac kn)$ in Lemma~\ref{l:guarantees}. Then, we have
$\Sigmab_w=\frac kn\Sigmab_{\X}$ and also $d_{\A}(\Sigmab_w) =
d_{\frac nk\!\A}$. Since for any set $S$ of size $k$, we
have $\opt\leq f_{\A}(\X_S^\top\X_S)$, the result follows.
\end{proofof}

\input{bayesian_experiments.tex}

\section{Conclusions}

We proposed a new algorithm for finding
$(1+\epsilon)$-approximate Bayesian experimental designs by leveraging 
a fundamental connection with determinantal point processes. Compared
to the state-of-the-art approaches, our
method provides stronger theoretical guarantees in terms of the allowed
range of subset sizes, as well as
offering significantly better time complexity guarantees. At the same
time, our experiments show that on the task of A-optimal design the
proposed algorithm performs as well as or better than several methods that
are used in practice.

% \subsubsection*{Acknowledgements}

% MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
%   support of this work. 
%   Also, MWM and MD \Red{
%     thank the NSF TRIPODS program
%     and FL thanks the NPSC for funding.
%   }
%   The authors thank Uthaipon
%   T.~Tantipongpipat for valuable discussions.

\bibliographystyle{apalike}
\bibliography{pap}

\newpage
\appendix

\onecolumn

\section{Properties of regularized DPPs}
\label{a:proofs}
In this section we provide proofs omitted from Sections \ref{s:r-dpp}
and \ref{s:guarantees}. We start with showing the fact that the
regularized DPP distribution $\DPPreg{p}(\X,\A)$ is a correlation DPP.
\begin{lemma}[restated Lemma \ref{l:correlation}]
  Given $\X$, $\A$, and $\D_p$ as in Theorem \ref{t:algorithm}, we have
  \begin{align*}
    \DPPreg{p}(\X,\A)= \DPPcor\big(\D_p +
    (\I-\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}(\I-\D_p)^{\sfrac12}\big).
    \end{align*}
  \end{lemma}
  \begin{proof}
    First, we show this under the invertibility assumptions of Lemma
    \ref{t:reduction}, i.e., given that $\A$ and $\I-\D_p$ are
    invertible. In this case $\DPPreg{p}(\X,\A)=\DPPens(\L)$, where
    \begin{align}\L=\Dbt+
    \Dbt^{\sfrac12}\X\A^{-1}\X^\top
      \Dbt^{\sfrac12}\quad\text{ and }\quad\Dbt=\D_p(\I-\D_p)^{-1}.\label{eq:ensemble}
\end{align}
Converting this to
    a correlation kernel $\K$ and denoting $\Xt=\D_p^{\sfrac12}\X$, we obtain
    \begin{align*}
      \K&=\I-(\I+\L)^{-1}
\\      &=\I - \big(\I + (\I-\D_p)^{-1}\D_p +
(\I-\D_p)^{-\sfrac12}\Xt\A^{-1}\Xt^\top(\I-\D_p)^{-\sfrac12}\big)^{-1}
\\ &= \I - \big((\I-\D_p)^{-1} +
     (\I-\D_p)^{-\sfrac12}\Xt\A^{-1}\Xt^\top(\I-\D_p)^{-\sfrac12}\big)^{-1}
\\ &=\I -
     (\I-\D_p)^{\sfrac12}(\I+\Xt\A^{-1}\Xt^\top)^{-1}(\I-\D_p)^{\sfrac12}
      \\
      &\overset{(*)}{=}\I -
     (\I-\D_p)^{\sfrac12}\big(\I-\Xt\A^{-\sfrac12}
     (\I+\A^{-\sfrac12}\Xt^\top\Xt\A^{-\sfrac12})^{-1}
     \A^{-\sfrac12}\Xt^\top\big)(\I-\D_p)^{\sfrac12}
      \\
      &=\I - (\I-\D_p) +
     (\I-\D_p)^{\sfrac12}\Xt(\A+\Xt^\top\Xt)^{-1}\Xt^\top(\I-\D_p)^{\sfrac12}
\\ &=\D_p + (\I-\D_p)^{\sfrac12}\Xt(\A+\Xt^\top\Xt)^{-1}\Xt^\top(\I-\D_p)^{\sfrac12},
    \end{align*}
    where $(*)$ follows from Fact 2.16.19 in
    \cite{matrix-mathematics}. Note that converting from $\L$ to $\K$
    got rid of the inverses $\A^{-1}$  and $(\I-\D_p)^{-1}$ appearing
    in \eqref{eq:ensemble}. The intuition 
    is that when $\A$ or $\I-\D_p$ is non-invertible, then
    $\DPPreg{p}(\X,\A)$ is not an L-ensemble but it is still a
    correlation DPP. To show this, we use a limit argument. For
    $\epsilon\in[0,1]$, let $p_\epsilon=(1-\epsilon)p$ and
    $\A_\epsilon=\A+\epsilon\I$. Observe that if $\epsilon>0$ then $\A_\epsilon$ and
    $\I-\D_{p_{\epsilon}}$ are always invertible even if $\A$ and
    $\I-\D_p$ are not. Denote $\K_\epsilon$ as the
    above correlation kernel with $p$ replaced by $p_{\epsilon}$ and
    $\A$ replaced by $\A_{\epsilon}$. Note that all matrix operations
    defining kernel $\K_\epsilon$ are continuous w.r.t. $\epsilon\in[0,1]$, including the inverse, since
    $\A+\Xt^\top\Xt$ is assumed to be invertible. Therefore, the
    following equalities hold (with limits taken point-wise and $\epsilon>0$):
    \begin{align*}
      \DPPreg{p}(\X,\A) = \lim_{\epsilon\rightarrow
      0}\DPPreg{p_\epsilon}(\X,\A_{\epsilon}) =
      \lim_{\epsilon\rightarrow 0}\DPPcor(\K_\epsilon) = \DPPcor(\K),
    \end{align*}
where we did not have to assume invertibility of $\A$ or $\I-\D_p$.
\end{proof}
We now prove a lemma about combining a determinantal point process
with Bernoulli sampling, which itself is a DPP with a diagonal correlation
kernel.
\begin{lemma}[restated Lemma \ref{l:decomposition}]
  Let $\K$ and $\D$ be $n\times n$ psd matrices with eigenvalues between
0 and 1, and assume that $\D$ is diagonal. If\, $T\!\sim\DPPcor(\K)$ and
$R\sim\DPPcor(\D)$, then
\begin{align*}T\cup R\sim \DPPcor\big(\D+(\I-\D)^{\sfrac12}\K
  (\I-\D)^{\sfrac12}\big).
  \end{align*}
\end{lemma}
\begin{proof}
For this proof we will use the shorthand $\K_A$ for $\K_{A,A}$. If
$\D$ has no zeros on the diagonal then $\det(\D_A)>0$ for all
$A\subseteq[n]$ and
  \begin{align*}
    \Pr(A \subset T \cup R)
    &=\sum_{B\subset A}\Pr(R\cap A=A\setminus B) \ \Pr(B\subseteq T)\\
    &=\sum_{B\subset A}\det(\D_{A\setminus
      B})\det\!\big([\I-\D]_B\big)\,\det(\K_B) \\
    &=\sum_{B\subset A}\det(\D_{A\setminus
      B})\det\!\Big(\big[(\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\big]_B\Big) \\
    &=\det(\D_A) \sum_{B\subset A}\det\!\Big(\big[\D^{-\sfrac12}
      (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\D^{-\sfrac12}\big]_{B}\Big)\\[-1mm]
    &\overset{(*)}{=}\det(\D_A)\det\!\Big(\I + \big[\D^{-\sfrac12}
      (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\D^{-\sfrac12}\big]_A\Big)\\
    &=\det\!\Big(\big[\D+ (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\big]_A\Big),
  \end{align*}
  where $(*)$ follows from a standard determinantal identity used to
  compute the L-ensemble partition function
  \cite[Theorem~2.1]{dpp-ml}. If $\D$ has zeros on the diagonal, a
  similar limit argument as in Lemma \ref{l:correlation} with
  $\D_\epsilon=\D+\epsilon\,\I$ holds.
\end{proof}
Next, we give a bound on the expected size of a regularized DPP.
\begin{lemma}[restated Lemma \ref{l:size}]
  Given any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
  s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
  full rank, let $S=T\cup \{i:b_i=1\}\sim \DPPreg{p}(\X,\A)$ be defined
as in Theorem \ref{t:algorithm}. Then
\begin{align*}
\E\big[|S|\big]\leq \E\big[|T|\big] + \E\Big[\sum_i b_i\Big] =
  d_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big) + \sum_i  p_i.
\end{align*}
\end{lemma}
\begin{proof}
For correlation kernels it is known that the expected size of $\DPPcor(\K)$
is $\tr(\K)$. Thus, using $\D_p=\diag(p)$, we can invoke Lemma \ref{l:correlation} to obtain
\begin{align*}
  \E\big[|S|\big]
  &= \tr\big(\D_p +
  (\I-\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}
                    \X^\top\D_p^{\sfrac12}(\I-\D_p)^{\sfrac12}\big)
  \\ &\leq \tr(\D_p) +\tr\big(\D_p^{\sfrac12}
       \X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big)
\\ &=\tr(\D_p) +\tr\big(\X^\top\D_p\X (\A+\X^\top\D_p\X)^{-1}\big) =
     \tr(\D_p) + d_{\A}(\X^\top\D_p\X),
\end{align*}
from which the claim follows.
\end{proof}
Next, we show two expectation inequalities for the matrix inverse and
matrix determinant.
\begin{lemma}[restated Lemma \ref{t:expectations}]
Whenever $S\sim \DPPreg{p}(\X,\A)$ is a well-defined distribution it holds that
\begin{align}
  \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
  &\preceq \Big(
    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1},\label{eq:sqinv2}
  \\
\E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
  &\leq\det\!\Big(
    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1}.\label{eq:det2}
\end{align}
\end{lemma}
\begin{proof}
For a square matrix $\M$, define its adjugate, denoted $\adj(\M)$, as a matrix
whose $i,j$-th entry is $(-1)^{i+j}\det(\M_{-j,-i})$, where
$\M_{-j,-i}$ is the matrix $\M$ without $j$th row and $i$th column. If
$\M$ is invertible, then $\adj(\M) = \det(\M)\M^{-1}$. Now, let
$b_i\sim\mathrm{Bernoulli}(p_i)$ be independent random variables. As seen
in previous section, the identity
$\E[\det(\sum_ib_i\x_i\x_i^\top+\A)]=\det(\sum_ip_i\x_i\x_i^\top+\A)$ gives
us the normalization constant for $\DPPreg{p}(\X,\A)$. Moreover, as
noted in a different context by \cite{determinantal-averaging}, when
applied entrywise to the adjugate matrix, this identity implies that
$\E[\adj(\sum_ib_i\x_i\x_i^\top+\A)]=\adj(\sum_ip_i\x_i\x_i^\top+\A)$. Let
$\Ic$ denote the set of all subsets $S\subseteq [n]$ such that
$\X_S^\top\X_S+\A$ is invertible. We have
\begin{align*}
  \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
  &= \sum_{S\in\Ic}\big(\X_S^\top\X_S+\A\big)^{-1}
\frac{\det(\X_S^\top\X_S+\A)}{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
    S}p_i\prod_{i\not\in S}(1-p_i)
\\ &=\sum_{S\in\Ic}\frac{\adj(\X_S^\top\X_S+\A\big)}
{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
     S}p_i\prod_{i\not\in S}(1-p_i)
\\ &\preceq \sum_{S\subseteq[n]}\frac{\adj(\X_S^\top\X_S+\A\big)}
{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
     S}p_i\prod_{i\not\in S}(1-p_i)
\\ &=\frac{\E\big[\!\adj(\sum_i
     b_i\x_i\x_i^\top+\A)\big]}{\det(\sum_ip_i\x_i\x_i^\top+\A)}
  \\
  &=\frac{\adj(\sum_ip_i\x_i\x_i^\top+\A)}{\det(\sum_ip_i\x_i\x_i^\top+\A)}
    = \Big(\sum_ip_i\x_i\x_i^\top+\A\Big)^{-1}.
\end{align*}
Note that if $\Ic$ contains all subsets of $[n]$, for example when
$\A\succ\zero$, then the inequality turns into equality. Thus, we
showed \eqref{eq:sqinv2}, and \eqref{eq:det2} follows even more easily:
\begin{align*}
  \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
  &= \sum_{S\in\Ic}
\frac{1}{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
    S}p_i\prod_{i\not\in S}(1-p_i)
\leq \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{\!-1}\!,
\end{align*}
where the equality holds if $\Ic$ consists of all subsets of $[n]$.
\end{proof}

\section{Comparison of different effective dimensions}
\label{a:deff}
In this section we compare the two notions of effective dimension for
Bayesian experimental design considered in this work. Here, we let
$\X$ be the full $n\times d$ design matrix and use $k$ to denote the
desired subset size. Recall that the effective dimension is  defined
as a function of the data covariance matrix $\Sigmab_\X=\X^\top\X$ and
the prior precision matrix $\A$:
It is given by $d_{\A} =
\tr\big(\Sigmab_\X(\A+\Sigmab_\X)^{-1}\big)$. In
\cite{regularized-volume-sampling} it was suggested that $d_{\A}$
should also be used as the effective dimension for the experimental
design problem. Theorem \ref{t:q1} suggests it may not reflect the true degrees
of freedom of the problem because it does not scale with subset size
$k$. Instead we propose to use the \emph{scaled effective dimension}
$d_{\frac nk\!\A}$. Thus, the two definitions we are comparing can be
summarized as follows:
\begin{description}
  \item[Full effective dimension]\quad\ \ \,$d_{\A} =
    \tr\big(\Sigmab_\X(\A+\Sigmab_\X)^{-1}\big)$,
    \item[Scaled effective dimension] $d_{\frac nk\!\A} =
      \tr\big(\Sigmab_\X(\tfrac nk\A + \Sigmab_\X)^{-1}\big)$.
    \end{description}
    Here, we demonstrate that these two effective dimensions can be
    very different for some matrices and quite similar on others. For
    simplicity, we consider two diagonal data covariance matrices as
    our examples: \emph{identity covariance}, $\Sigmab_1 = \I$, and an
    \emph{approximately low-rank covariance}, $\Sigmab_2 =
    (1-\epsilon)\frac ds \I_S+\epsilon\I$, where $\I_S$ is the
    diagonal matrix with ones on the entries indexed by subset
    $S\subseteq [d]$ of
    size $s<d$ and zeros
    everywhere else. The second matrix is scaled in such way so that
    $\tr(\Sigmab_1)=\tr(\Sigmab_2)$. We use $d=100$, $s=10$ and
    $\epsilon=10^{-2}$. The prior precision matrix is
    $\A=10^{-2}\,\I$. Figure \ref{f:deff} plots the scaled effective
    dimension $d_{\frac nk\!\A}$ as a function of $k$, against the
    full effective dimension for both examples. Unsurprisingly, for
    the identity covariance the full effective dimension is almost
    $d$, and the scaled effective dimension goes up very quickly to
    match it. On the other hand, for the approximately low-rank
    covariance, $d_{\A}\approx 55$ is considerably less then
    $d=100$. Interestingly, the gap between the $d_{\frac nk\!\A}$ and 
    $d_{\A}$ for moderately small values of $k$ is even bigger. Our
    theory suggests that $d_{\frac nk\!\A}$ is a valid indicator of
    Bayesian degrees of freedom when $k\geq C\cdot d_{\frac nk\!\A}$
    for some small constant $C$ (Theorem \ref{t:q1} has $C=4$, but we
    believe this can be improved to $1$). While for the identity
    covariance the condition $k\geq
    d_{\frac nk\!\A}$ is almost equivalent to $k\geq d_{\A}$, in the
    approximately low-rank case, $k\geq d_{\frac nk\!\A}$ holds for
    $k$ as small as 20, much less than $d_{\A}$.
    \begin{figure}[htbp]
      \centering
      \includegraphics[width=0.5\textwidth]{figs/deff_identity}\nobreak\includegraphics[width=0.5\textwidth]{figs/deff_lowrank}
      \caption{Scaled effective dimension compared to the full effective
        dimension for two diagonal data covariance matrices, with
      $\A=10^{-2}\,\I$.}
      \label{f:deff}
    \end{figure}

\section{Additional details for the experiments}
\label{a:experiments}
\input{bayesian_experiments-appendix.tex}

\end{document}

### 2022-08-23 ftvi/ftvi.tex
Initial Overleaf Import

\documentclass{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage{amssymb,amsmath,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage[bold]{hhtensor}

\usepackage{natbib}
\bibliographystyle{apalike}

\usepackage{hyperref}
\usepackage{cleveref}

\usepackage{todonotes}

\usepackage{booktabs}
\usepackage{float}

\input{macros}

\title{Fat-Tailed Variational Inference}

\begin{document}
\maketitle

Given a probabilistic model $p(x,y)$, the goal of automatic variational
inference is to (1) construct a variational family of distributions
$\{q_\phi\}_{\phi \in \Phi}$ and (2) search for a good variational
approximation to the posterior $q_\phi \approx p(x \mid y)$.
We focus on (1) and consider variational families specifically constructed
to include fat-tailed members. Our contributions:
\begin{enumerate}
    \item We rectify the failure of ADVI against fat tailed target densities by
          extending it to utilize Student-T variational family, and extend its
          applicability by transforming the density through a normalizing flow
          analogous to \citep{webb2019improving}.
          Our results on a location-scale representation for a StudentT shows that (1)
          allowing users to specify a tail-index for variational approximators leads to
          better fits and (2) even without prior knowledge, estimating the tail
          coefficient during training also works.
    \item We provide alternatives for estimation of the variational
          approximation's tail index, adapting ideas from classical tail index
          estimation (\cite{hill1975simple,pickands1975statistical}) to the
          variational inference setting where posterior samples are expensive but
          joint log density evaluations are cheap
\end{enumerate}

\section{Background}

Probabilistic machine learning problems are coveniently formalized using
probabilistic programming languages (PPLs) which provide high level
primitives for probabilistic modelling and inference. In order to enable more
flexibile and precise control over variational inference methods, recent work
has explored a design space for variational inference APIs ranging from
limited but completely automated (ADVI, \cite{kucukelbir2017automatic}) to
manually specified guide programs (WebPPL, Pyro). Our work here focuses on
automated variational inference where we seek to automatically construct a
reasonable variational family automatically after tracing the execution of
the probabilistic program.

Fat-tailed distributions commonly arise in robust machine learning where a
standard approach is to replace light-tailed noise distributions fat-tailed
ones \cite{tipping2005variational}. They are also commonly used as weakly
informative prior distributions in Bayesian heirarchical models
\cite{gelman2006prior}.

\subsection{Related Work}

The Box-Cox transform \citep{box1964analysis} is a commonly used parametric
power transform to address fat tails and make data more similar to a normal
distribution. This is analogous to composing a power transform before
the normalizing flow. Whereas \cite{asar2017estimating} fit this transform
using goodness-of-fit tests against a normal, whereas our optimization signal
is back-propagated from an ELBO against the target posterior.

The bulk of related work focuses on fat-tails arising from relaxing priors.
Recent work in VAEs has extensively studied the impact of relaxing Gaussian
assumptions to heavier-tailed distributions. \citep{mathieu2019disentangling}
consider a StudentT prior distribution $p(z)$ over the latent code $z$ in a
VAE with Gaussian encoder $q(z \mid x)$
\footnote{\url{https://github.com/iffsid/disentangling-disentanglement/blob/3396d40f46c34dd928a1241f567a86276b0ff41b/src/main.py\#L52}},
showing that the anisotropy of a StudentT product distribution (compared to
the standard choice of Normal prior) leads to more disentangled
representations. \cite{chen2020use} perform a similar modification except in
a coupled VAE \cite{cao2019coupled} and showed improvements in the marginal
likelihoods of reconstructed images. In addition,
\cite{boenninghoff2020variational} consider a mixture of StudentTs for the
prior $p(z)$. Finally, \feynman{maybe remove this? it was rejected from
    ICLR}, \cite{abiri2020variational} considered both StudentT prior and
variational approximation family and showed improvements in SSIM (structural
similarity) score of between original and reconstructed images. To position
our work in context, note that the VAE's encoder $q(z \mid x)$ may be viewed
as a variational approximation to the posterior $p(z \mid x)$ defined by the
decoder model $p(x \mid z)$ and the prior $p(z)$. Our work differs from
\citep{mathieu2019disentangling,chen2020use,boenninghoff2020variational} in
that we consider heavy-tailed variational approximations $q(z \mid x)$ rather
than priors $p(z)$, and although \citep{abiri2020variational} also considers
a StudentT approximate posterior our work (1) considers a more general
variational family comprised of flow transforms of StudentTs and (2) conducts
a more thorough investigation across a broader range of models beyond a VAE
on FashionMNIST.

Relaxation of priors to heavy-tailed distributions has numerous
applications beyond VAEs. In \cite{silnova2018fast}, the authors perform
inference in heavy-tailed probabilistic linear discriminant analysis
using Gaussian mean-field variational inference and show improved
accuracy in speaker identification. Our work is complementary to these approaches;
whereas they consider heavy-tailed priors $p(z)$ we consider heavy-tailed
variational families $q(z \mid x)$.

More directly comparable recent work on fat-tailed variational families
\cite{ding2011t,futami2017expectation} studies the $t$-exponential family
variational approximation (which includes Student-Ts and other
heavier-tailed) includes heavy-tailed variational families, but critically do
not discuss selection of the parameter $t$ (which is deterministically to the
Student-T's DoF $v$). Other differences include their derivation of
expectation propagation update equations whereas we directly backprop a noisy
ELBO estimate, and our broader variational family which includes flow
transforms.

Most related to our work is \cite{jaini2020tails}, which shows similar
impossibility results with Gaussian tailed base distributions and consider
flow transforms of StudentT base distributions. However, our tail adaptive
flows (1) do not mandate equal tail index across all dimensions, (2) composes
techniques from \cite{kucukelbir2017automatic} to address target
distributions with constrained support, (3) are investigated within the
setting of density estimatino ($\propto KL(p,q)$, test set likelihood) as
variational inference ($KL(q,p)$, ELBO, since $p$ is a posterior hard to
sample), and (4) are automatically generated as part of a probabilistic
programming engine.

\section{Fat Tailed Variational Inference}

\subsection{Distribution classes}

While there is significant attention on parameterizing the transport map, there is comparatively little interest in considering the distribution $p$ of the latent variable $Z$. Because the vast majority of transport map parameterizations are Lipschitz continuous, we will find that this choice of $p$ can have a profound impact on the quality of the approximation of the tails of the target distribution. Here, we shall consider the following three main types of tail behaviour.
\begin{enumerate}
    \item The class $\mathcal{G}$ of subgaussian random vectors $X$ satisfying \[\limsup_{r\to\infty}r^{-2}\log\mathbb{P}(\|X\|>r)<0.\]
    \item The class $\mathcal{E}$ of subexponential random vectors $X$ satisfying \[\limsup_{r\to\infty}r^{-1}\log\mathbb{P}(\|X\|>r)<0.\]
    \item The class $\mathcal{P}_\nu$ of random vectors $X$ with tails no heavier than a $\nu$-power law, that is, \[\limsup_{r\to\infty}\log\mathbb{P}(\|X\|>r)/\log r<-\nu.\]
\end{enumerate}
It is more convenient to characterize these distribution classes according to the concentration function $\alpha_X$ defined for a random vector $X$ by
\[
    \alpha_{X}(r)=\sup_{A\,:\,\mathbb{P}(X\in A)\geq1/2}\mathbb{P}(\mbox{dist}(X,A)>r),
\]
where $\dist(x,A) = \inf_{y \in A}\|x - y\|$. Equivalent characterizations of $\mathcal{G},\mathcal{E},\mathcal{P}_\nu$ follow by replacing $\mathbb{P}(\|X\|>r)$ with $\alpha_X(r)$. For example, a random variable $X \in \mathcal{G}$ if and only if $\limsup_{r\to\infty} r^{-2} \alpha_X(r) < 0$. Under this observation, the following lemma is an immediate consequence of \cite[Proposition 1.3]{ledoux2001concentration}.
\begin{lemma}
    \label{lem:distn_class_closed}
    The classes $\mathcal{G}$, $\mathcal{E}$, and $\mathcal{P}_\nu$ for any $\nu > 0$, are closed under Lipschitz transformations. In other words, if $f$ is Lipschitz, then for any $X \in \mathcal{G}$ (resp. $\mathcal{E}$, $\mathcal{P}_\nu$), $f(X) \in \mathcal{G}$ (resp. $\mathcal{E}$, $\mathcal{P}_\nu$). 
\end{lemma}
In other words, using only transport map approximators based on Lipschitz transformations, to achieve an accurate approximation of the tails of the target distribution, one must ensure that the reference distribution exhibits the same tail behaviour. Density estimators using a base distribution of incorrect class will inevitably fail at approximating slower tail decay. Note that this is not in violation of universal approximation theory since $\mathcal{G}$ is dense in $L^{2}$ (and therefore in $\mathcal{E}$ and $\bigcup_{\nu>0}\mathcal{P}_{\nu}$).

Two noteworthy choices for $p$ in the literature are the standard Gaussian distribution, as seen in  \cite{kingma2016improved}, and the logistic distribution, as seen in \cite{dinh2014nice} (recommended because it ``tends to provide a better behaved gradient''). Indeed, in the former case, only subgaussian approximations are available. In the latter case, since the logistic distribution is subexponential and not subgaussian, subexponential approximations are available. However, neither of these choices are effective for distributions which exhibit power-law tails, such as the Student $t$-distribution.

\feynman{From 1/29/2021:

    New theory? Extend multivariate reuslts of \cite{jaini2020tails} to invalidate shared $\nu$, eg 
    product of two studentT with very different DoF.
    Interaction between ADVI transformed distributions and StudentTs.

}

\subsection{Fat tails and tail indices}
\feynman{Reconcile with distribution classes}

\begin{definition}[\cite{resnick2007heavy}]
    A random variable $X$ has \emph{fat tails} with
    tail index $\alpha > 0$
    if $\Pr[X > x] \sim x^{-\alpha}$ as $x \to \infty$.
\end{definition}

Variance undefined if $\alpha < 3$, generally all moments $> \alpha - 1$
are infinite. $\alpha$ is called the \emph{tail index}.

Can relax power law to regularly/slowly varying \footnote{\url{https://journals.aps.org/prresearch/abstract/10.1103/PhysRevResearch.1.033034}}, which permits deviations without
affecting tail exponent.

Pareto: $\Pr[X > x] = \left(\frac{x_m}{x}\right)^\alpha \sim x^{-\alpha}$

Student T \footnote{\url{https://math.stackexchange.com/questions/3092190/asymptotics-of-hypergeometric-2f-1abcz-for-large-z-to-infty}}: 
$\Pr[X > x] \sim x \left(\frac{x^2}{\nu}\right)^{-\frac{1}{2}} + \left(\frac{x^2}{\nu}\right)^{-\frac{\nu+1}{2}} \sim x^{-\nu}$.

Cauchy: $\Pr[X > x] \sim -\arctan\left(\frac{x - x_0}{\gamma}\right) \sim x^{-1}$,
which makes since because standard Cauchy is StudentT($\nu=1$).

Stable distributions? $X$ is stable if $\forall a, b > 0$, 
$a X_1 + b X_2 = c X + d$ for some $c > 0$, $d$ where $X_1, X_2$ independent copies of $X$
(closure under convolution for fixed $\alpha$).
This includes normal, Cauchy, and Levy. Four parameter family with stability parameter $\alpha$
controlling the asymptotic behavior (tail index?). $\alpha = 2$ gives a Guassian,
and $\alpha < 2$ gives $p(x) \sim \lvert x \rvert^{-(\alpha + 1)}$.

\feynman{Reconcile definitions with \cite{jaini2020tails}}

\subsection{Flows in variational inference}

Universal approximation theorems about flows are not violated; they are asymptotic
so finite realizations will still have support $\RR$ and Lipschitz continuity.

\textbf{IMPORTANT}: support $\RR$ makes them unsuitable for VI's $KL(q,p)$, need \cite{kucukelbir2017automatic}
transforms

\subsection{Failure modes on fat-tailed target densities}
\label{ssec:failure}

One major limitation of $p = N(0,I)$ is:
\begin{theorem}[Chapter 2 \cite{wainwright2019high}]
    Let $(X_i)_1^n$ be a vector of iid $N(0,1)$ RVs,
    $f : \RR^n \to \RR$ be $L$-Lipschitz.
    Then $f(X) - \EE f(X)$ is $L$-sub-Gaussian.
\end{theorem}
In particular, density estimators using a Gaussian base
distribution $p$ will inevitably fail at approximating slower
tail decay.

When the target density is a fat-tailed Cauchy distribution as in
\cref{fig:cauchy_normal_student}, variational inference using
flow-transformed Gaussian base distributions \cite{webb2019improving} (orange
in left) result in tails which decay inappropriately fast as measured
using a Kolmogorov-Smirnov goodness-of-fit test (right). In contrast,
the learned flow-transformed Student-T base distribution (green left)
provides a much better approximation of the tail behavior.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.49\textwidth]{Figures/cauchy_normal_student.png}
    \includegraphics[width=0.49\textwidth]{Figures/cauchy_ks.png}
    \caption{VI against a Cauchy target, PDFs (left) and $p$-value of Kolmogorov-Smirnov test statistics (right, $\leq 0.05$
        suggests poor approximation). 
    }
    \label{fig:cauchy_normal_student}
\end{figure}

\subsubsection{Analytical tail index algebra}

Restrict to StudentT or Pareto.

Sum, PDFs are convolved, max of coefficients?

Product,

Ratio

Max

\feynman{Method of steepest descent / saddle point approximations to approximate asymptotics of integrals, so given the PDF (eg from a PP) in theory can work out
    the power law exponent}

\subsection{A flexibile fat-tailed variational family}

We consider inverse autoregressive flow \cite{kingma2016improved} transforms
of heavy-tailed base distributions for variational inference, and investigate
the effects of directly learning the tail index versus applying asymptotic
approximations.

\begin{claim}
    Inverse autoregressive flows \cite{kingma2016improved} are Lipschitz continuous
    (because they are compositions of saturating non-linearities and matrix-vector products)
\end{claim}

To avoid the pitfalls of \cref{ssec:failure}, we see from
\cref{lem:distn_class_closed} that the root cause was closure of subgaussian
distributions under Lipschitz maps. Accordingly, the two modifications we can
consider relaxing are (1) Lipschitz-continuity of the transport map or (2)
the distribution class of the base distribution.

\subsection{Relaxing to non-Lipschitz transport maps}

Classical statistical practice is to apply a power transform in an effort to
make the data closer to a normal distribution. As power transforms are not
Lipschitz continuous on the real line, this corresponds to a relaxation of
assumption (1). We operationalize this idea by precomposing a ??? transform
onto the transport map and propose learning the transform parameter through
stochastic optimization of ELBO (c.f. estimating the Box-Cox parameter using
goodness-of-fit against normal in \cite{asar2017estimating}).

\feynman{Need to log transform a Normal, powers of normal are still sub-exponential}

\feynman{Experiment with this}

\subsection{Relaxing the base distribution}

Complementary to relaxing Lipschitz continuity, another workaround for
\cref{lem:distn_class_closed} is to change the base distribution class.
In our work, we restrict attention to fat-tailed distributions with
tails no heavier than a $\nu$-power law $\cP_\nu$.

Selection of the tail index parameter $\nu$ is a well studied problem with
classical solutions \citep{hill1975simple,pickands1975statistical} using
order statistics of samples. 
We consider:
\begin{itemize}
    \item Defining APIs to allow researchers to succinctly specify fat-tailness of latent
          variables and their automatic variational inference semantics
    \item Estimating the tail index from samples (either provided or obtained from MCMC)
    \item Estimating the tail index of the posterior $p(x \mid y)$ using the log joint density $p(x,y)$,
          which is easy to evaluate and (given samples) approximately marginalize
    \item Sample-free estimation by including the tail index as a variational
          parameter and optimizing ELBO
    \item \feynman{TODO: tail-coefficient algebra}
\end{itemize}

\subsubsection{Fine-grained control over tail indices of variational approximations}

% \verbatim{@random_variable(tail_index=2.0, hidden_size=64, num_layers=3)}
decorator to instruct mean-field guide with fixed tail index of $2.0$.

\subsubsection{Estimating tail index without samples}

% \verbatim{@random_variable(tail_index=2.0, hidden_size=64, num_layers=3, optim=True)}

Owing to the difficulty of generating posterior samples,
techniques for estimating the tail index from samples (e.g. the Hill estimator)
are not directly applicable to inference. As a result, it is of interest
to be able to estimate a tail index given only access to an unnormalized log density.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/hill.png}
    \caption{Hill estimator against $\nu=1.5$ StudentT}
    \label{fig:hill}
\end{figure}

Directly optimizing $\nu$ in the variational ELBO results in a trade-off between
accurate tail behaviour and matching the distribution around the mode.

For $x, y$ large:
\begin{align*}
    \Pr[X > x]             & \sim x^{-\alpha}                                       \\
    p(x)                   & \sim \alpha x^{-\alpha - 1}                            \\
    \log p(x)              & \sim \log \alpha - (\alpha + 1) \log x                 \\
    \log \frac{p(x)}{p(y)} & \sim (\alpha + 1) \log \frac{y}{x}                     \\
    \alpha                 & \sim \frac{\log p(x) - \log p(y)}{\log y - \log x} - 1
\end{align*}
Note that $p$ can be unnormalized as the partition function is cancelled out.
For example. setting $x=10$ and $y=20$ yields an estimate of $1.4799$
for a $\nu=1.5$ StudentT.

\feynman{Consider Richardson extrapolation to accelerate this limit. Rate of convergece in limit must be a power}

\subsubsection{Approximate marginalization}

\feynman{Is this feasible for PP?}

\textbf{Problem}: Joint distributions (e.g. location-scale mixtures) require
marginalization before we can get $p(x)$.

\textbf{Our Method}: Approximate with discrete mixture
\begin{align*}
    p(x) = \int p(x|z) p(z) dz \approx \frac{1}{N} \sum_{i=1}^N p(x \mid z_i)\quad\text{where}~z_i\simiid p(z)
\end{align*}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{Figures/our-estimator.png}
    \caption{Tail index estimation on location-scale mixture representation
        for StudentT, where the mixture is discretized to $n$ components.}
    \label{fig:approx_marg}
\end{figure}

\section{Experiments}

These experiments investigate the behavior of neural density estimators with
\emph{heavy-tailed base distribution}.
Specifically, we consider a masked autoregressive flow \cite{papamakarios2017masked}
transform of a generalized Student's t distribution as a density estimator $q_\theta(X)$
in a variational inference framework. To fit $q_\theta$ to a target distribution
$\pi$, the ELBO gradient is reparameterized and Monte-Carlo approximated
\begin{align*}
    \nabla_\theta \EE_{q_\theta} \log \frac{\pi(X)}{q_\theta(X)}
     & = \nabla_\theta \EE_p \log \frac{
        \pi(X)
    }{p_\theta(f_\theta^{-1}(X))
    \left\lvert \det \nabla f_\theta^{-1}(X) \right\rvert}    \\
     & = \EE_p \nabla_\theta \log \frac{
        \pi(X)
    }{p_\theta(f_\theta^{-1}(X))
    \left\lvert \det \nabla f_\theta^{-1}(X) \right\rvert}    \\
     & \approx \frac{1}{n} \sum_i^n \nabla_\theta \log \frac{
        \pi(x_i)
    }{p_\theta(f_\theta^{-1}(x_i))
        \left\lvert \det \nabla f_\theta^{-1}(x_i) \right\rvert}
\end{align*}

\feynman{Careful, $D_{KL}(N(0,1), \text{Cauchy}(0,1) \approx 0.2592 < \infty = D_{KL}(\text{Cauchy(0,1), N(0,1)})$}

\subsection{Applications in PPL inference}

The experiments in this section are conducted using the \texttt{beanmachine} PPL, where
inference is conducted following a Metropolis-within-Gibbs framework.

\subsection{Normal-normal location mixture}

We first consider a Normal-Normal conjugate inference problem where the posterior
is known to be a Normal distribution as well.
\Cref{fig:normal_normal} shows the resulting density approximation, which can
be seen to be reasonable for both a Normal base distribution (the ``correct'' one)
and a StudentT base distribution. This suggests that mis-specification (i.e. heavier
tails in the base distribution than the target) may not be too problematic.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/normal_normal_posterior.png}
    \caption{VI against a Normal posterior}
    \label{fig:normal_normal}
\end{figure}

\subsection{StudentT scale mixture representation}

We next consider a heavy-tailed posterior target density by using a scale-mixture
representation for Student's t. Specifically, if $v \sim \chi_2(\nu)$
and $y \mid v \sim N(0, v)$ then the marginal distribution of $y$ is StudentT
with $\nu$ degrees of freedom.
In this experiment, we also allow the degrees of freedom for the base StudentT
distribution to be optimized as well in \texttt{p\_student\_df\_vi} and set
the degrees of freedom equal to the true $\nu$ in \texttt{p\_student\_vi}.
While using the true $\nu$ does yield an almost exact fit, optimizing $\nu$ is
more practical.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/studentt_log_scale.png}
    \caption{VI against StudentT represented by a Normal scale mixture}
    \label{fig:studentt_loc_scale}
\end{figure}

\subsection{Bayesian Robust Regression}

$n = 100$, $d = 10$.

$X_{ij} \simiid N(0,1)$ for $i \in [n]$, $j \in [d]$.

$y_i \simiid \text{StudentT}(\text{loc}=X \beta, df=5)$

Improper ``flat'' prior on $\beta$ to ensure heavy-tailed posterior.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/blr.png}
    \caption{VI of Bayesian robust linear regression}
    \label{fig:blr}
\end{figure}

\feynman{Can work out the exponent with asymptotic approximations; compare?}

\subsection{Eight schools}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/eight_schools.png}
    \begin{tabular}{rcc}
        \toprule
                  & ELBO                & $\log P(y)$       \\
        \midrule
        ADVI      & -193.86 $\pm$ 33.50 & -70.11 $\pm$ 0.52 \\
        ADVI-HT   & -121.54 $\pm$ 19.59 & -70.29 $\pm$ 0.54 \\
        MAF       & -55.06 $\pm$ 5.46   & -59.14 $\pm$ 1.99 \\
        MAF-HT    & -59.63 $\pm$ 6.74   & -57.84 $\pm$ 4.97 \\
        MAF-2L    & -45.01 $\pm$ 11.02  & -52.19 $\pm$ 2.06 \\
        MAF-2L-HT & -51.67 $\pm$ 8.72   & -51.53 $\pm$ 4.23 \\
        IAF       & -77.67 $\pm$ 6.47   & -64.79 $\pm$ 0.80 \\
        IAF-HT    & -76.67 $\pm$ 3.71   & -64.91 $\pm$ 0.76 \\
        \bottomrule
    \end{tabular}
    \caption{Final ELBO and (MC estimate of) log marginal $\log P(y)$ after 5000 steps}
    \label{fig:blr}
\end{figure}

\subsection{Importance weights}

When the importance sampling density is more peaked than the target density.

\cite{wang2018variational} example 3.1: Let $p = N(0, \sigma_p^2)$, $q = N(0, \sigma_q^2)$,
and for $x \sim q$ let $w(x) = \frac{p(x)}{q(x)}$. If $\sigma_p > \sigma_q$,
then $w$ has tail index $\frac{\sigma_p^2}{\sigma_p^2 - \sigma_q^2}$.
Otherwise, $w$ is not fat-tailed.

\feynman{TODO}

\subsection{Some HTVAE example reconstructions and likelihoods}

\feynman{TODO. From \cite{sohn2015learning}, MNIST, Caltech-UCSD Birds, and Labeled Faces in the Wild.}

\section{Single-Site Normalizing Flow Variational Inference}

Let $\{q_\phi\}_{\phi \in \Phi}$ be a parameterized (variational) family.
Recall the variational ELBO
\begin{align*}
    \log p(y)
    = \log \int dx \frac{q_\phi(x)}{q_\phi(x)} p(x,y)
    \geq \int dx q_\phi(x) \log \frac{p(x,y)}{q_\phi(x)}
\end{align*}
The variational family's PDF $q_\phi(x)$ is assumed to be tractable, and
the joint PDF $p(x,y)$ may be obtained by running the probabilistic program forwards.
Finally, the Monte-Carlo method enables tractable approximation of the integral:
\begin{align*}
    \int dx q_\phi(x) \log \frac{p(x,y)}{q_\phi(x)}
     & \approx \log \frac{p(x,y)}{q_\phi(x)}\qquad x \sim q_\phi
\end{align*}
making the ELBO tractable.

Assume there are $d$ latent variables $x = (x_i)_{i=1}^d$.
Single-site variational inference (i.e.\ CaVI, coordinate-ascent variational
inference) sequentially iterates over $i \in [d]$, performing Monte-Carlo
\begin{align*}
    \phi \leftarrow
    \phi + \alpha_t \nabla_{\phi} \EE_{\setminus \{i\}} \EE_{i} q_\phi(x) \log \frac{p(x,y)}{q_\phi(x)}
\end{align*}

\todo{Finish description, vary the number of MC samples between two expectations}

\subsection{Cached Inverses}

The Monte-Carlo ELBO
\begin{align*}
    \sum_i \log p(x_i, y) - \log q(x_i) \qquad x_i \sim q
\end{align*}
requires both sampling and evaluating log densities with respect to $q$.
When $q$ is taken to be a distribution parameterized by a normalizing flow
$q(x) = \lvert \frac{df}{dx} \rvert p(f^{-1}(x))$, log density evaluations
require evaluating the inverse flow $f^{-1}$ whereas sampling involves
pushing forward samples $x \sim p$ along the flow's forward direction.

Modern normalizing flow architectures typically
are much more expensive to compute in one direction (c.f. IAF vs MAF), with
some flows (NAF, SoS, ResidualFlows) requiring an optimization to approximate
the inverse flows. To circumvent this problem, we identify a critical optimization
for VI that caching can be used to avoid flow inversion during Monte-Carlo training.
That is, by caching pairs $(x_i, y_i = f(x_i))$ sampled during MC ELBO,
the entropy term $\log q(y_i) = \log \lvert \frac{df}{dx}(x_i) \rvert + \log p(x_i)$
can be evaluated without explicit inversion.

\todo{Run-time comparisons for ELBO with/without caching}

\subsection{NaMI structured conditional flows}

The NaMI algorithm \cite{webb2017faithful} can be used for model inversion, enabling
single-site flow distributions which depend on the values of adjacent nodes (as well as
amortized variational inference where the variational families $q_{(\phi,y)}(x)$
depend on the conditioned observations $y$.

\section{Meeting Notes}

\begin{enumerate}
    \item We added a knob for heavy tail, how do we make fair comparison? Increase the flexibility of the flow for non-Gaussian.
    \item Neural network weights heavy tailed; can we learn VI with heavy-tailed distributions.
\end{enumerate}

\section{Discussion}

Limitations: only considered symmetric base distributions, could also consider skewness \citep{gupta2003multivariate}.

\bibliography{refs_ftvi}

\end{document}

### 2022-08-23 ftvi/ftvi_icml/ftvi_icml.tex
Initial Overleaf Import

%%%%%%%% ICML 2022 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass[nohyperref]{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2022} with \usepackage[nohyperref]{icml2022} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2022}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2022}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
    \usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Fat--Tailed VI with Anisotropic Tail Adaptive Flows}

% custom stuff
\usepackage[bold]{hhtensor}
\usepackage{mathtools}
\usepackage{float}
\usepackage{braket}
\usepackage{dsfont}
\usepackage{graphbox}
\usepackage{pgf}
\usepackage{comment}
\usepackage{subcaption}
\usepackage{enumitem}
\input{../macros}

\begin{document}

\twocolumn[
\icmltitle{Fat--Tailed Variational Inference with Anisotropic Tail Adaptive Flows}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2022
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Feynman Liang}{yyy,fb}
\icmlauthor{Liam Hodgkinson}{yyy,zzz}
\icmlauthor{Michael W. Mahoney}{yyy,zzz}
% \icmlauthor{Firstname4 Lastname4}{sch}
% \icmlauthor{Firstname5 Lastname5}{yyy}
% \icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
% \icmlauthor{Firstname7 Lastname7}{comp}
% %\icmlauthor{}{sch}
% \icmlauthor{Firstname8 Lastname8}{sch}
% \icmlauthor{Firstname8 Lastname8}{yyy,comp}
% %\icmlauthor{}{sch}
% %\icmlauthor{}{sch}
\end{icmlauthorlist}

\icmlaffiliation{fb}{Meta, Menlo Park, CA}
\icmlaffiliation{yyy}{Department of Statistics, University of California, Berkeley, CA}
\icmlaffiliation{zzz}{International Computer Science Institute, Berkeley, CA}
% \icmlaffiliation{comp}{Company Name, Location, Country}
% \icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}

\icmlcorrespondingauthor{Feynman Liang}{feynman@berkeley.edu}
%\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
    While fat-tailed densities commonly arise as posterior and marginal distributions in
    robust models and scale mixtures, they present challenges when Gaussian-based 
    variational inference fails to capture tail decay accurately. 
    We first improve previous theory on tails of Lipschitz flows 
    %\citep{jaini2020tails} 
    by quantifying how the tails affect the \emph{rate} of tail decay 
    and by expanding the theory to non-Lipschitz polynomial flows.
    We then develop an alternative theory for multivariate tail parameters which is 
    sensitive to tail-anisotropy. 
    In doing so, we unveil a fundamental problem which plagues many existing flow-based 
    methods: they can only model tail-isotropic distributions (i.e., distributions 
    having the same tail parameter in every direction).
    To mitigate this and enable modeling of tail-anisotropic targets, we propose 
    anisotropic tail-adaptive flows (ATAF).
    Experimental results on both synthetic and real-world targets confirm that ATAF 
    is competitive with prior work while also exhibiting appropriate tail-anisotropy.
\end{abstract}

\section{Introduction}
\label{sec:intro}

%\michael{Example}
%\liam{Example}

Flow-based methods 
\citep{papamakarios2021normalizing}
have proven to be effective techniques to model complex
probability densities. They compete with the state of the art on
density estimation \citep{huang2018neural,durkan2019neural,jaini2020tails},
generative modeling \citep{chen2019residual,kingma2018glow}, and variational inference \citep{kingma2016improved,agrawal2020advances} tasks.
These methods start with a random variable $X$ having a simple and tractable
distribution $\mu$, and then apply a learnable transport map $f_\theta$ to build
another random variable $Y = f_\theta(X)$ with a more expressive \emph{pushforward}
probability measure $(f_\theta)_\ast \mu$ \citep{papamakarios2021normalizing}.
In contrast to the implicit distributions \citep{huszar2017variational} produced by generative adversarial networks (GANs), flow-based methods restrict the transport map $f_\theta$ to be invertible and to have efficiently-computable Jacobian determinants.
As a result, probability density functions can be tractably computed
through direct application of a change of variables
\begin{align}
    \label{eq:change-of-variable}
    p_{Y}(y)
      = p_{X}(f_\theta^{-1}(y)) \left\lvert \det
        \left.\frac{d f_\theta^{-1}(z)}{dz} \right\vert_{z=y}
      \right\rvert .
\end{align}

While recent developments \citep{chen2019residual,huang2018neural,durkan2019neural} have focused primarily
on the transport map $f_\theta$, the base distribution $\mu$ has received comparatively less investigation. 
%PREVIOUS%We believe this asymmetric focus is detrimental to the research community because the sensible default choice of Gaussian base distribution $\mu = \cN(0,\mI)$ may result in significant limitations to the expressivity of the model (\Cref{thm:distn_class_closed}). 
%PREVIOUS%\michael{Clarify what does ``asymmetric'' mean in that sentence, does it mean that the focus is more on the transport map than the base distribution, or something else.}
The most common choice for the base distribution is standard Gaussian $\mu = \cN(0,\mI)$.
However, in \Cref{thm:distn_class_closed}, we show this choice results in significant
restrictions on the expressivity of the model, limiting its utility for data that
exhibits fat-tailed (or heavy-tailed) structure.
Prior work addressing heavy-tailed flows \citep{jaini2020tails}
are limited to tail-isotropic base distributions.
In \Cref{prop:isotropic-pushforward}, we prove flows built on these base distributions
are unable to model accurately multivariate anisotropic fat-tailed structure.
% \michael{I'm a little confused by this paragraph.  We cite two results of ours in this paper, but it sounds like prior work, since we then say that we will address those issues.  Are we saying that prior work did such-and-such, and one of our contributions is to make that explicit and then improve that.  Or was that known.  In either case, we should reword here.}
%PREVIOUS%Addressing this issue was an important aim for recent work \citep{jaini2020tails}, and our work here represents an additional advancement towards this goal.

\begin{figure*}[htbp]
  \centering
  \includegraphics{../Figures/pancake.pdf}
    \vspace{-5mm}
  \caption{
    Variational inference against a tail-anisotropic target distribution $\cN(0,1) \otimes \text{StudentT}(\nu=1)$ (top left).
    Only ATAF (bottom right) is able to correctly reproduce the tail-anisotropy (fat-tailed along $x$-axis,
    Gaussian along $y$-axis).
    In contrast, ADVI's (top right) Gaussian base distribution and TAF's (bottom left) tail-isotropic $\prod_{i=1}^2 \text{StudentT}(\nu)$
    base distribution  can only model tail-isotropic distributions (\Cref{prop:isotropic-pushforward}),
    which erroneously imposes power-law tails with the same rate of decay along both the $x$ and $y$ axes.
    \vspace{-5mm}
  }
  \label{fig:pancake}
\end{figure*}

Our work here aims to identify and address these deficiencies.
To understand the impact of the base distribution $\mu$ in flow-based models,
we develop and apply theory for fat-tailed random variables and their transformations under Lipschitz-continuous functions.
Our approach leverages the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration} to sharpen significantly and extend prior results
\citep[Theorem 4]{jaini2019sum} by describing precisely the tail parameters of the pushforward distribution $(f_\theta)_\ast \mu$ under both Lipschitz-continuous (\Cref{thm:distn_class_closed}) and polynomial (\Cref{corr:closure_polynomials}) transport maps.
% \michael{We say ``characterize'' but what do we mean.  Probably this is the place to say that it is particularly important to be flexible to have different tail indices in different directions.  Then some of the later sentences will make sense.}
%In the multivariate setting, we observe that it is important to be flexible and allow for a multivariate distribution
%to have different tail parameters in different directions. %Accordingly,
In the multivariate setting,
we develop a theory of direction-dependent tail parameters (\Cref{def:mv-tail-param}), and we show that tail-isotropic base distributions yield tail-isotropic pushforward measures (\Cref{prop:isotropic-pushforward}). 
As a consequence of \Cref{prop:isotropic-pushforward}, prior methods \citep{jaini2020tails} are limited in that
they are unable to capture \emph{tail-anisotropy}.
This motivates the construction of \emph{anisotropic tail adaptive flows} (ATAF, \Cref{def:ataf}) as a means to
alleviate this issue (\Cref{remark:anisotropic}) and to improve modeling of tail-anisotropic distributions.
% \michael{What limitation is precisely being referred to, since in the previous sentence we say that we get several results in this paper, and then we say how our results are a limitation.  Probably the best way to do this is to be more detailed, e.g., say that our \Cref{prop:isotropic-pushforward} means that the mapping can't introduce anisotropy.}
Our experiments show that ATAF exhibits correct tail behaviour in synthetic target distributions exhibiting fat-tails (\Cref{fig:cauchy_normal_student} of \Cref{sec:cauchy_normal_student}) and tail-anisotropy (\Cref{fig:pancake}).
On realistic targets,
% prior work \citep[Table 2]{jaini2020tails} showed tail-adaptive flows (TAF, a subset of ATAF)
% can improve density modeling and 
we find that ATAF can yield improvements in variational inference (VI) by capturing potential tail-anisotropy (\Cref{sec:experiments}).
% \michael{That sentence sounds like we are a footnote on them.  I would say that we get results for a ragne of cases, both for isotropic and anisotropic cases, thereby improving prior work that considered Gaussian things as well as prior work from \cite{jaini2020tails} that considered isotropic HT.}
\vspace{-3mm}
\subsection*{Related Work}
\vspace{-1mm}
\paragraph{Fat-Tails in Variational Inference.}

% The bulk of related work focuses on fat-tails arising from relaxing priors.
Recent work in variational autoencoders (VAEs) have considered relaxing Gaussian assumptions to heavier-tailed distributions \citep{mathieu2019disentangling,chen2019residual,boenninghoff2020variational,abiri2020variational}.
In \citet{mathieu2019disentangling}, a StudentT prior distribution $p(z)$ is considered over the latent code
$z$ in a VAE with Gaussian encoder $q(z \mid x)$. They argue
%\footnote{\url{https://github.com/iffsid/disentangling-disentanglement/blob/3396d40f46c34dd928a1241f567a86276b0ff41b/src/main.py\#L52}},
that the anisotropy of a StudentT product distribution leads to more disentangled representations, as compared to the standard choice of Normal distributions.
A similar modification is performed in \citet{chen2020use} for
a coupled VAE \citep{cao2019coupled}. 
This result showed improvements in the marginal
likelihoods of reconstructed images. In addition,
\citet{boenninghoff2020variational} consider a mixture of StudentTs for the
prior $p(z)$. To position
our work in context, note that the encoder $q(z \mid x)$ may be viewed
as a variational approximation to the posterior $p(z \mid x)$ defined by the
decoder model $p(x \mid z)$ and the prior $p(z)$. Our work differs from
\citet{mathieu2019disentangling,chen2020use,boenninghoff2020variational}, in
that we consider fat-tailed variational approximations $q(z \mid x)$ rather
than priors $p(z)$. Although \citet{abiri2020variational} also considers
a StudentT approximate posterior, our work involves a more general
variational family which uses normalizing flows.
% Relaxation of priors to heavy-tailed distributions has numerous
% applications beyond VAEs. In \cite{silnova2018fast}, the authors perform
% inference in heavy-tailed probabilistic linear discriminant analysis
% using Gaussian mean-field variational inference and show improved
% accuracy in speaker identification. Our work is complementary to these approaches;
% whereas they consider heavy-tailed priors $p(z)$ we consider heavy-tailed
% variational families $q(z \mid x)$.
Similarly, although \citet{wang2018variational} also deals with fat-tails in variational inference,
their goal is to improve $\alpha$-divergence VI by controlling the moments of importance
sampling ratios (which may be heavy-tailed). Our work here adopts
Kullback-Leibler divergence and is concerned with enriching the variational family
to include anisotropic fat-tailed distributions.
More directly comparable recent work \citep{ding2011t,futami2017expectation} studies the $t$-exponential family
variational approximation which includes StudentTs and other
heavier-tailed densities. Critically, the selection of their parameter $t$ (directly related to the
StudentT's degrees of freedom $\nu$), and the issue of tail anisotropy, are not discussed. %Other differences include their derivation of
%expectation-propagation update equations while 
% We directly backpropagate a noisy
% ELBO estimate, and introduce a richer variational family used by ATAF.

\vspace{-1mm}
\paragraph{Flow-Based Methods.}

Normalizing flows and other flow-based methods have a rich history within variational
inference \citep{kingma2016improved,rezende2015variational,agrawal2020advances,webb2019improving}.
Consistent with our experience (\Cref{fig:blr-anisotropic}), \citet{webb2019improving}
documents normalizing flows can offer improvements over ADVI and NUTS across thirteen different
Bayesian linear regression models from \citet{gelman2006data}.
\citet{agrawal2020advances} shows that normalizing flows compose nicely with other
advances in black-box VI (e.g., stick the landing, importance weighting).
However, none of these works treat the issue of fat-tailed targets and inappropriate tail
decay.
To our knowledge, only TAFs \citep{jaini2020tails} explicitly consider flows with tails
heavier than Gaussians. Our work here can be viewed as a direct improvement of \citet{jaini2020tails},
and we make extensive comparison to this work throughout the body of this paper. At
a high level, we provide a theory for fat-tails which is sensitive to the rate of
tail decay and develop a framework to characterize and address the tail-isotropic limitations plaguing
TAFs.

\vspace{-2mm}
\section{Flow-Based Methods for Fat-Tailed Variational Inference}

%Continuing from the notation introduced in \Cref{sec:intro}, in this section we
%finish establishing the VI setting and definitions/notation
%\michael{I'm not sure what ``Continuing ...'' means.  Can we say ``In this section we do such-and such ...".  Maybe something like:
%``In this section, we provide background on flow-vased methods for VI, and we describe some of the issues that arise when dealing with fat-tailed data and fat-tailed VI that will be important for understanding our main results. 
%}

\vspace{-2mm}
\subsection{Flow-Based VI Methods}

The objective of VI is to approximate a target distribution $\pi(x)$ by searching over
a \emph{variational family} $\cQ = \{q_\phi : \phi \in \Phi\}$ of probability distributions $q_\phi$.
While alternatives exist \citep{li2016variational,wang2018variational}, VI typically
seeks to find $q_\phi$ ``close'' to $\pi$, as measured by Kullback-Leibler divergence $\infdiv{q_\phi}{\pi}$.
To ensure tractability without sacrificing generality, in practice \citep{wingate2013automated,ranganath2014black}
a Monte-Carlo approximation of the evidence lower bound (ELBO) is maximized:
\begin{align*}
  %-\infdiv{q_\phi}{\pi}
  %\propto 
  \text{ELBO}(\phi)
  &= \int q_\phi(x) \log \frac{\bar\pi(x)}{q_\phi(x)} dx\\
  & \approx \frac{1}{n} \sum_{i=1}^n \log \frac{\bar\pi(x_i)}{q_\phi(x_i)},\;
  x_i \simiid q_\phi,\;
  \bar{\pi} \propto \pi. 
\end{align*}
To summarize, this procedure enables tractable black-box VI
by replacing $\pi$ with $\bar\pi \propto \pi$ and approximating expectations with respect to $q_\phi$ (which are tractable only in simple variational families) through Monte-Carlo approximation. In Bayesian inference and probabilistic programming applications, the target posterior
$\pi(x) = p(x \mid y) = \frac{p(x, y)}{p(y)}$ is typically intractable but
$\bar\pi(x) = p(x,y)$ is computable (i.e., represented by the probabilistic program's
generative / forward execution).

\begin{table*}
  \centering
  \begin{tabular}{ccc}
    \toprule
    Model                                  & Autoregressive transform                                                                              & Suff. cond. for Lipschitz-continuity        \\
    \midrule
    NICE\citep{dinh2014nice}               & $z_j + \mu_j \cdot \mathds{1}_{k \not \in [j]}$                                                       & $\mu_j$ Lipschitz                      \\
    MAF\citep{papamakarios2017masked}      & $\sigma_j z_j + (1 - \sigma_j) \mu_j$                                                                 & $\sigma_j$ bounded                     \\
    IAF\citep{kingma2016improved}          & $z_j \cdot \exp(\lambda_j) + \mu_j$                                                                   & $\lambda_j$ bounded, $\mu_j$ Lipschitz \\
    Real-NVP\citep{dinh2016density}        & $\exp(\lambda_j \cdot \mathds{1}_{k \not\in[j]}) \cdot z_j + \mu_j \cdot \mathds{1}_{k \not \in [j]}$ & $\lambda_j$ bounded, $\mu_j$ Lipschitz \\
    Glow\citep{kingma2018glow}             & $\sigma_j \cdot z_j + \mu_j \cdot \mathds{1}_{k \not\in [j]}$                                         & $\sigma_j$ bounded, $\mu_j$ Lipschitz  \\
    NAF\citep{huang2018neural}             & $\sigma^{-1}(w^\top \cdot \sigma(\sigma_j z_j + \mu_j))$                                              & Always \par (logistic mixture CDF)          \\
    NSF\citep{durkan2019neural}            & $z_j \mathds{1}_{z_j \not\in [-B,B]} + M_j(z_j;z_{<j}) \mathds{1}_{x_j \in [-B,B]}$         & Always \par (linear outside $[-B,B]$)     \\
    FFJORD\citep{grathwohl2018ffjord} & n/a (not autoregressive)                                                                              & Always \par (required for invertibility)    \\
    ResFlow\citep{chen2019residual} & n/a (not autoregressive)                                                                              & Always \par (required for invertibility)    \\
    \bottomrule
  \end{tabular}
  \caption{Some popular / recently developed flows, the autoregressive transform used in the flow (if applicable),
  and sufficient conditions conditions for Lipschitz-continuity. A subset of this table was first presented
  in \citet{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.
  \vspace{-5mm}
  }
  \label{tab:flows}
\end{table*}

While it is possible to construct a variational family $\cQ$ tailored to a specific task, we are interested in VI methods which are more broadly applicable and convenient to use: $\cQ$ should be automatically constructed from introspection of a given probabilistic model/program.
Automatic differentiation variational inference (ADVI, \citet{kucukelbir2017automatic}) is an early implementation of automatic VI and it is still the default in certain probabilistic programming languages \citep{carpenter2017stan}.
ADVI uses a Gaussian base distribution $\mu$ and a transport map $f_\theta = f \circ \Phi_\text{Affine}$ comprised of an invertible affine transform composed with a deterministic transformation $f$ from $\RR$ to the target distribution's support (e.g., $\exp : \RR \to \RR_{\geq 0}$, $\text{sigmoid} : \RR \to [0,1]$).
As Gaussians are closed under affine transformations, ADVI's representational capacity is limited to deterministic transformations of Gaussians. Hence it cannot represent complex multi-modal distributions.
To address this, more recent work \citep{kingma2016improved,webb2019improving} replaces the affine map $\Phi_\text{Affine}$ with a flow $\Phi_{\text{Flow}}$ typically parameterized by an invertible neural network:
\begin{definition}
    \label[definition]{def:advi}
  ADVI (with normalizing flows) comprise the variational family
  $\cQ_\text{ADVI}~\coloneqq~\{
    (f~\circ~\Phi_\text{Flow})_\ast \mu
    \}$, 
  where $\mu = \text{Normal}(0_d, I_d)$,
  $\Phi_\text{Flow}$ is an invertible flow transform (e.g., \Cref{tab:flows})
  and $f$ is a deterministic bijection between constrained supports \citep{kucukelbir2017automatic}.
\end{definition}

As first noted in \citet{jaini2020tails}, the pushforward of a light-tailed Gaussian base distribution under a Lipschitz-continuous flow will remain light-tailed and provide poor approximation to fat-tailed targets. 
Despite this, 
%at the time of writing this paper 
many major probabilistic programming packages still make a default choice of Gaussian base distribution (\texttt{AutoNormalizingFlow}/\texttt{AutoIAFNormal} in Pyro \citep{bingham2019pyro}, \texttt{method=variational} in Stan \citep{carpenter2017stan}, \texttt{NormalizingFlowGroup} in PyMC \citep{patil2010pymc}).
%\michael{What is the verb in the following sentence}
To address this issue, tail-adaptive flows \citep{jaini2020tails} use a
base distribution $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$,
where a single degrees-of-freedom $\nu \in \RR$ is used across all $d$ dimensions. Here is a more precise definition.
\begin{definition}
  \label[definition]{def:taf}
  Tail adaptive flows (TAF) comprise the variational family
  $\cQ_\text{TAF}
    \coloneqq \{
    (f \circ \Phi_\text{Flow})_\ast \mu_\nu
    \}$,
  where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$ with $\nu$ shared across all $d$ dimensions,
  $\Phi_{\text{Flow}}$ is an invertible flow,
  and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
  During training, the shared degrees of freedom $\nu$ is treated as an additional variational parameter.
\end{definition}

\vspace{-2mm}
\subsection{Fat-Tailed Variational Inference}

\vspace{-1mm}
Fat-tailed variational inference (FTVI)
% \michael{I'm inclined to say that we should have ``(FTVI)'' and then refer to that acronym everywhere; then we can say: VI is important, FTVI is very important, ADVI did a good job with FTVI, we do a great job with FTVI, and you the reader should follow up on us and do FTVI.  It's the sort of acronym that is nice to remember.}
% will do, and add this discussion to future directions in conclusion
considers the setting where the target $\pi(x)$ is fat-tailed. 
Such distributions commonly arise during a standard
``robustification'' approach where light-tailed noise distributions are
replaced with fat-tailed ones \citep{tipping2005variational}. They also
appear when weakly informative prior distributions are used in Bayesian
hierarchical models \citep{gelman2006prior}.

To formalize these notions of fat-tailed versus light-tailed distributions, a quantitative classification for tails is required.
While prior work classified distribution tails according to quantiles and the existence of moment generating functions
\citep[Section 3]{jaini2020tails}, here we propose a more natural and finer-grained classification based upon
the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration}, which is sensitive to
the rate of tail decay.

\begin{definition}[Classification of tails]
    \label[definition]{def:tail-classification}
    For each $\alpha,p > 0$, we let 
    \vspace{-1mm}
    \begin{itemize}[leftmargin=*]
        \item $\mathcal{E}_\alpha^p$ denote the set of \emph{exponential-type} random variables $X$ with $\mathbb{P}(|X| \geq x) = \Theta(e^{-\alpha x^p})$;
    \vspace{-1mm}
        \item $\mathcal{L}_\alpha^p$ denote the set of \emph{logarithmic-type} random variables $X$ with $\mathbb{P}(|X| \geq x) = \Theta(e^{-\alpha(\log x)^p})$.
    \end{itemize}
    \vspace{-1mm}
    % A random variable $X$ is of %\textbf{exponential-type}, denoted $X \in \cE_\alpha^p$, w
    % \begin{description}
    %     \item[exponential-type], denoted $X \in \cE_\alpha^p$,
    %   whenever $\PP(\lvert X \rvert \geq x) = \Theta(e^{-\alpha x^p})$, and
    %     \item[logarithmic-type], denoted $X \in \cL^p_\alpha$, whenever $\PP(\lvert X \rvert \geq x) = \Theta(e^{-\alpha (\log x)^p})$.
    % \end{description}
    In both cases, we call $p$ the \emph{class index} and $\alpha$ the \emph{tail parameter} for $X$.
    Note that every $\cE_\alpha^p$ and $\cL_\beta^q$ are disjoint, that is,
    $\cE_\alpha^p \cap \cL_\beta^q = \emptyset$ for all $\alpha,\beta,p,q > 0$.
    For brevity, we define the ascending families
    $\overline{\cE_\alpha^p}$ and $\overline{\cL_\alpha^p}$
    analogously as before except with $\Theta(\cdot)$ replaced
    by $\cO(\cdot)$. Similarly, we denote the class of distributions with exponential-type
    tails with class index at least $p$
    by $\overline{\cE^p} = \cup_{\alpha \in \RR_+} \overline{\cE_\alpha^p}$, and similarly for $\overline{\cL^p}$.
\end{definition}

For example, $\overline{\cE_\alpha^2}$ corresponds to $\alpha^{-1/2}$-sub-Gaussian random variables,
    $\overline{\cE_\alpha^1}$ corresponds to sub-exponentials, and (of particular relevance to this paper) $\cL^1_\alpha$ corresponds to the class of power-law distributions.

\vspace{-2mm}
\section{Tail Behavior of Lipschitz Flows}

\vspace{-1mm}
This section states our main theoretical contributions; proofs are deferred to \Cref{sec:proofs}.
We sharpen previous impossibility results approximating fat-tailed targets
using light-tailed base distributions \citep[Theorem 4]{jaini2020tails}
by characterizing the effects of Lipschitz-continuous transport maps on not only the tail class
but also the class index and tail parameter (\Cref{def:tail-classification}). Furthermore, we extend the theory
to include polynomial flows \citep{jaini2019sum}. For the multivariate setting,
we define the tail-parameter function (\Cref{def:mv-tail-param}) to help formalize the notion
of tail-isotropic distributions and prove a fundamental limitation that tail-isotropic
pushforwards remain tail-isotropic (\Cref{prop:isotropic-pushforward}).

Most of our results are developed within the context of Lipschitz-continuous transport maps $f_\theta$.
In practice, many flow-based methods exhibit Lipschitz-continuity in their transport map, either by design \citep{grathwohl2018ffjord,chen2019residual}, or as a consequence of choice of architecture and activation function (\Cref{tab:flows}). % or by being enforced stemming from the Banach fixed point theorem 
%
%or .
The following assumption encapsulates this premise.
\begin{assumption}\label[assumption]{assump:lipschitz}
    $f_\theta$ is invertible, and both $f_\theta$ and $f^{-1}_\theta$
    are $L$-Lipschitz continuous (e.g., sufficient conditions in \Cref{tab:flows} are satisfied).
\end{assumption}
It is worth noting that domains other than $\mathbb{R}^d$ may require an additional bijection between supports (e.g. $\exp : \mathbb{R} \to \mathbb{R}_+$) which
could violate \cref{assump:lipschitz}.

\vspace{-2mm}
\subsection{Closure of Tail Classes}
\label{ssec:failure}

\vspace{-1mm}
Our first set of results pertains to the closure of the tail classes in \Cref{def:tail-classification}
under Lipschitz-continuous transport maps. While earlier work \citep{jaini2020tails} demonstrated
closure of exponential-type distributions $\cup_{p > 0} \overline{\cE^p}$ under flows satisfying \Cref{assump:lipschitz}, our results in Theorem \ref{thm:distn_class_closed} and Corollaries \ref{corr:heavy_to_light} and \ref{corr:closure_polynomials} sharpen these observations, showing that: (1) Lipschitz transport maps cannot decrease the class index $p$ for exponential-type random variables, but they can alter the tail parameter $\alpha$; and
(2) under additional assumptions, they cannot change either class index $p$ or the tail parameter $\alpha$ for logarithmic-type random variables.
%         tail parameter $\alpha$; 
% \begin{itemize}[leftmargin=*]
%     \item cannot decrease the class index $p$ for exponential-type random variables, but can alter the
%         tail parameter $\alpha$; and
%     \item under additional assumptions, cannot change either class index $p$ or the tail
%         parameter $\alpha$ for logarithmic-type random variables.
% \end{itemize}

\begin{theorem}[Lipschitz maps of tail classes]
  \label{thm:distn_class_closed}
  Under \Cref{assump:lipschitz},
  the distribution classes $\overline{\cE^p}$
  and $\overline{\cL^p_\alpha}$ (with $p,\alpha > 0$) are closed
  under every flow transformation in \Cref{tab:flows}.
\end{theorem}

Informally, \Cref{thm:distn_class_closed} asserts that light-tailed base distributions cannot be transformed
via Lipschitz transport maps into fat-tailed target distributions.
Note this does not violate universality theorems for certain flows \citep{huang2018neural}
as these results only apply in the infinite-dimensional limit. Indeed, certain exponential-type families (such as Gaussian mixtures) are dense in the class of \emph{all} distributions, including those that are fat-tailed.

Note that $\overline{\cL^p_\alpha} \supset \cE^q_\beta$ for all $p,q,\alpha,\beta$, so \Cref{thm:distn_class_closed}
by itself does not preclude transformations of fat-tailed base distributions to light-tailed targets.
Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushfoward,
hence heavy-to-light transformations are impossible. Note here there is no ascending union over
tail parameters (i.e., $\cL^p_\alpha$ instead of $\overline{\cL^p_\alpha}$).

\begin{corollary}[Closure of $\cL^p_\alpha$]
  \label[corollary]{corr:heavy_to_light}
  If in addition $f_\theta$ is smooth
  with no critical points on the interior or boundary of
  its domain, then $\cL_\alpha^p$ is closed.
\end{corollary}

This implies that simply fixing a fat-tailed base
distribution \emph{a priori} is insufficient; the tail-parameter(s) of the base distribution must be explicitly optimized alongside
the other variational parameters during training.
While these additional assumptions may seem restrictive, note that many flow transforms
explicitly enforce smoothness and monotonicity \citep{wehenkel2019unconstrained,huang2018neural,durkan2019neural}
and hence satisfy the premises. In fact, we can show a version of \Cref{thm:distn_class_closed} ensuring closure of exponential-type
distributions under polynomial transport maps which do not satisfy \Cref{assump:lipschitz}.
This is significant because it extends the closure results to
include polynomial flows such as sum-of-squares flows \citep{jaini2019sum}.

\begin{corollary}[Closure under polynomial maps]
    \label[corollary]{corr:closure_polynomials}
  For any $\alpha, \beta, p, q \in \RR_+$, there does not exist a
  finite-degree polynomial map from $\cE_\alpha^p$ into $\cL_\beta^q$.
\end{corollary}

% Suffices to show for k \in \RR_{> 0}, this covers inverse powers

% \begin{remark}
%   There does not exist a inverse polynomial map (e.g., sqrt) from $\cL_\alpha$ to $\cE$.
% \end{remark}

\vspace{-2mm}
\subsection{Multivariate Fat-Tails and Anisotropic Tail Adaptive Flows}

\vspace{-1mm}
Next, we restrict attention to power-law tails $\cL^1_\alpha$, and we develop a multivariate fat-tailed theory and notions of isotropic/anisotropic tail indices. Using our theory, we prove that both ADVI and TAF are fundamentally limited because they
are only capable of fitting tail-isotropic target measures (\Cref{prop:isotropic-pushforward}).
We consider anisotropic tail adaptive flows (ATAF): a density
modeling method which can represent tail-anisotropic distributions (\Cref{remark:anisotropic}).

% Karamata's theorem enables recovery of a regularly varying random variable's
% index: $X \in \cL^1_\alpha$ means that $\lim_{x \to \infty} \frac{x \PP(X \geq
%     x)}{\int_x^\infty \PP(X \geq t) dt} = \alpha$ for $\alpha < -1$.
% Motivated by Karamata's theorem, we introduce
% the following definition to aid in describing multivariate fat-tailed random variables:

For example, consider the target distribution shown earlier in \Cref{fig:pancake} formed as the product of $\mathcal{N}(0,1)$ and $\text{StudentT}(\nu=1)$ distributions.
The marginal/conditional distribution along a horizontal slice (e.g., the distribution of $\braket{X,e_0}$)
is fat-tailed, while along a vertical slice (e.g., $\braket{X,e_1}$) it is Gaussian.
Another extreme example of tail-anisotropy where the tail parameter for
$\braket{X,v}$ is different in every direction $v \in \cS^{1}$
is given in \Cref{fig:radial-fat-tail}. Here $\mathcal{S}^{d-1}$ denotes the $(d-1)$-sphere in $d$ dimensions. 
Noting that the tail parameter depends on the choice of direction, we are motivated to consider
the following direction-dependent definition of multivariate tail parameters.

\begin{definition}
  \label[definition]{def:mv-tail-param}
  For a $d$-dimensional random vector $X$,
  its \emph{tail parameter function} $\alpha_X : \cS^{d-1} \to \bar{\RR}_+$
  is defined as
  % $\alpha_X(v) = \lim_{x \to \infty} \frac{x \PP(\braket{v,X} \geq x)}{\int_x^\infty \PP(\braket{v,X} \geq t) dt}$.
  $\alpha_X(v) = -\lim_{x \to \infty} \log \PP(\braket{v,X} \geq x) / \log x$ when the limit exists, and $\alpha_X(v) = +\infty$ otherwise.
  In other words, $\alpha_X(v)$ maps directions $v$ into the tail parameter of the corresponding one-dimensional projection $\braket{v,X}$. The random vector $X$ is \emph{tail-isotropic} if $\alpha_X(v) \equiv c$ is constant and
  \emph{tail-anisotropic} if $\alpha_X(v)$ is not constant but bounded.
\end{definition}

\begin{figure*}[htbp]
  \centering
  \includegraphics{../Figures/radial-fat-tail.pdf}%
%   \includegraphics[trim={0 0 7cm 0},clip]{../Figures/radial-fat-tail.pdf}\\%
%   \vspace{-8mm}
%   \includegraphics[trim={7.2cm 0 0 0},clip]{../Figures/radial-fat-tail.pdf}%
  %\includegraphics[width=\textwidth]{../Figures/radial-fat-tail.png}%
  %\input{../Figures/radial-fat-tail.pgf}
  \vspace{-7mm}
  \caption{
    Illustration of the direction-dependent tail-parameter function (right) on a tail-anisotropic distribution (left)
    with PDF $\dd P(r,\theta) = r^{-\alpha(\theta)} r \dd r \dd\theta$ and tail parameter $\alpha(\theta) = 2 + \cos(2\theta)$.
    While prior fat-tailed theory based on $\|X\|_2 = \sup_{\|v\|_2 = 1} \braket{X,v}$
    is only sensitive to the largest tail parameter $\max_{\theta \in [0, 2\pi]} \alpha(\theta) = 3.0$,
    our direction-dependent tail parameter function (bottom, red line)
    and its values along the standard basis axes ($\alpha(0)$ and $\alpha(\pi/2)$)
    capture \emph{tail-anisotropy}.
    \vspace{-5mm}
  }
  \label{fig:radial-fat-tail}
\end{figure*}
% \begin{definition}
%   \label[definition]{def:isotropic}
%   Multivarate random variable $X$ is tail-isotropic if $\alpha_X(v) \equiv c$ is constant.
%   Otherwise, it is tail-anisotropic.
% \end{definition}

% \begin{align*}
%   \PP[\braket{X,u} \geq x]
%   = \PP[\braket{\mu,u} + R \braket{AU,u} \geq x]
%   = \PP[R \geq \frac{x - \braket{\mu,u}}{\braket{u,AU}}]
%   \sim C_1 \left(\frac{x - \braket{\mu,u}}{\braket{u,AU}}\right)^{-\alpha}
%   \sim C_2 x^{-\alpha}
% \end{align*}
% and
% theory around tail parameters is developed for elliptically
% contoured multivariate distributions $X = \mu + R A U$
% for some heavy-tailed random variable $R$, fixed vectors $\mu \in \RR^d$
% and $U \in S^{d-1}$, and $A$ a (Cholesky factor)
% defining the ellipsoid axes.

% While elliptically contoured multivariate distributions admit a
% straightforward generalization from the scalar case, they are severely
% limited in practical applications.  One fundamental limitation of
% elliptically contoured $X$ is that the tail parameter is the same for every
% 1-dimension projection.  If $R$ has tail index $\alpha$ then for all for $u
%   \in S^{d-1}$
% \begin{align*}
%   \PP[\braket{X,u} \geq x]
%   = \PP[\braket{\mu,u} + R \braket{AU,u} \geq x]
%   = \PP[R \geq \frac{x - \braket{\mu,u}}{\braket{u,AU}}]
%   \sim C_1 \left(\frac{x - \braket{\mu,u}}{\braket{u,AU}}\right)^{-\alpha}
%   \sim C_2 x^{-\alpha}
% \end{align*}

% Furthermore, the assumption of elliptical distributions are
% easily violated in many common and useful applications.  For example, in
% probabilistic programming collections of random variables are oftentimes
% grouped together into a single multivariate random variable (e.g.,  blocked
% Gibbs, Hamiltonian Monte Carlo).  In fact, $Q_{TAF}$ (used in the experiments
% in \citep{jaini2020tails}) uses a StudentT product base distribution which
% is not elliptically symmetric.

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.4\textwidth]{../Figures/studentt-product.png}
%   \caption{
%     Illustrating multivariate fat-tails using a 2-dimensional StudentT.
%     The multivariate theory from \citep{jaini2020tails} is for elliptically
%     contoured distributions and not applicable to the StudentT product base distributions
%     used in their experiments.
%   }
%   \feynman{Consider just using words here to delineate from our contributions, this is not a super important fact}
%   \label{fig:studentt-product}
% \end{figure}

% Elliptical distributions are \emph{tail isotropic} i.e., $\alpha(v) \equiv c$ is constant.
% The base distribution for \citep{jaini2020tails}, $\prod_1^d \text{StudentT}(\nu_i)$ with $\nu_i \equiv \nu$,
% is also tail isotropic because
% $\braket{v,X}$ is a sum of StudentT and by tail index algebra
% has index $\max_i \nu_i = \nu$.
Of course, one can construct pathological densities where this definition is not effective
(see \Cref{eg:spiral}), but it will suffice for our purposes. 
It is illustrative to contrast with the theory presented for TAF \citep{jaini2020tails},
where only the tail exponent of $\|X\|_2$ is considered.
For $X = (X_1, \ldots, X_d)$ with $X_i \in \cL^1_{\alpha_i}$, 
by Fatou-Lebesgue and \Cref{lem:sum-rule}
\begin{align*}
  &\PP[\|X\|_2 \geq t]
  = \PP\left[\sup_{z \in \cS^{d-1}} \braket{X,z} \geq t\right]\\
  &\quad \geq \sup_{z \in \cS^{d-1}} \PP[\braket{X,z} \geq t]
  = \max_{1 \leq i \leq d} \nu_i
  = \max_{0 \leq i \leq d-1} \alpha_X(e_i) .
\end{align*}
Therefore, considering only the tail exponent of $\|X\|_2$ is equivalent to summarizing $\alpha_X(\cdot)$ by an upper bound.
Given the absence of the tail parameters for other directions (i.e., $\alpha_X(v) \neq \sup_{\|v\|=1} \alpha_X(v)$)
in the theory for TAF \citep{jaini2020tails}, it should be unsurprising that both their multivariate 
theory as well as their experiments only consider tail-isotropic distributions obtained either
as an elliptically-contoured distribution with fat-tailed radial distribution or 
$\prod_{i=1}^d \text{StudentT}(\nu)$ (tail-isotropic by \Cref{lem:sum-rule}). 
Our next proposition shows that this presents a significant limitation when the target distribution is
tail-anisotropic.

\begin{proposition}[Pushforwards of tail-isotropic distributions]
  \label[proposition]{prop:isotropic-pushforward}
  Let $\mu$ be tail isotropic with non-integer parameter $\nu$
  and suppose $f_\theta$ satisfies \Cref{assump:lipschitz}.
  Then $(f_\theta)_\ast \mu$ is tail isotropic with parameter $\nu$.
\end{proposition}

% However, $\alpha(v)$ is non-trivial to work with; it is an asymptotic quantity and is defined
% for uncountably many $v \in S^{d-1}$. In this work, we propose approximating
% $\alpha(v)$ using the standard basis vectors:

% \begin{definition}
%   The \emph{standard basis tail parameters} of a fat-tailed $X \in \RR^d$
%   is $\{\alpha(e_i) : i \in [d]\}$ where $\alpha$ is defined in
%   \Cref{def:mv-tail-param} and $e_i$ is the $i$th standard basis vector.
% \end{definition}

% The standard basis vectors provide a natural choice of projections for
% multivariate product distributions (as commonly encountered during blocking /
% grouping of random variables). Going back to our previous example,
% we still have that $\alpha_{\|X\|} = \max_i \alpha(e_i)$ but now the tail
% indices of the remaining $e_j$ need not be equal. Admittedly, the standard basis
% is less suitable for correlated multivariate distributions. For example,
% let $t_i \sim \text{StudentT}(\nu_i)$ and consider the rotated random variable
% $X = R_{\pi/4} [t_1; t_2]$. Its standard basis tail parameters are
% \begin{align*}
%   \Pr[\braket{X,e_1} > x]
%   = \Pr[\braket{X,e_2} > x]
%   = \Pr[0.5 t_1 + 0.5 t_2 > x]
%   \leq \max_i \nu_i
% \end{align*}
% This example illustrates that the standard basis tail parameters
% provide multivariate tail parameters which are no worse than
% previous work.

% \subsection{Limitations of our theory}

% \begin{itemize}
%   \item Standard basis parameters not general, only works for axis-aligned tails
%   i.e., independent product distributions
%   \item Still fails on the spiral, but it should do better than TAF
%   \item Theory not fully general, only considers rays from origin rather than
%   path integrals
% \end{itemize}

To work around this limitation without relaxing \Cref{assump:lipschitz}, it is evident
that tail-anisotropic base distributions $\mu$ must be considered. Perhaps the most straightforward modification to incorporate a tail-anisotropic base distribution replaces TAF's isotropic base distribution $\prod_{i=1}^d \text{StudentT}(\nu)$
with $\prod_{i=1}^d \text{StudentT}(\nu_i)$. Note that $\nu$ is no longer shared across dimensions,
enabling $d$ different tail parameters to be represented:

\begin{definition}\label[definition]{def:ataf}
  Anisotropic Tail-Adaptive Flows (ATAF) comprise the variational family
  $\cQ_\text{ATAF}~\coloneqq~\{
    (f \circ \Phi_\text{Flow})_\ast \mu_\nu
    \},$
  where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu_i)$, each $\nu_i$ is \emph{distinct}, and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
  Analogous to \citet{jaini2020tails}, ATAF's implementation
  treats $\nu_i$ identically to the other parameters in the flow and jointly optimizes over them.
\end{definition}

\begin{remark}\label[remark]{remark:anisotropic}
  Anisotropic tail-adaptive flows can represent tail-anisotropic distributions with up to $d$ different
  tail parameters while simultaneously satisfying \Cref{assump:lipschitz}.
  For example, if $\Phi_\text{Flow} = \text{Identity}$ and $\mu_\nu = \prod_{i=1}^d \text{StudentT}(i)$
  then the pushforward $(\Phi_\text{Flow})_\ast \mu_\nu = \mu_\nu$ is tail-anisotropic.
  % Moreover, its standard basis tail parameters are equal
  % (up to permutation) to those for $\mu$.
\end{remark}

Naturally, there are other parameterizations of the tail parameters $\nu_i$ that may be more effective depending on the application. For example, in high dimensions, one might prefer not to allow for $d$ unique indices, but perhaps only fewer. On the other hand, by using only $d$ tail parameters, an approximation error will necessarily be incurred when more than $d$ different tail parameters are present. \Cref{fig:radial-fat-tail} presents a worst-case scenario where the target distribution has a continuum of tail parameters. In theory, this density could itself be used as an underlying base distribution, although we have not found this to be a good option in practice. The key takeaway is that to capture several different tails in the target density, one must consider a base distribution that incorporates sufficiently many \emph{distinct} tail parameters.

Concerning the choice of StudentT families, we remark that since $\text{StudentT}(\nu) \Rightarrow \cN(0,1)$ 
as $\nu \to \infty$, ATAF should still provide reasonably good approximations to target 
distributions in $\overline{\cE^2}$ by taking $\nu$ sufficiently large. This can be seen in practice in \Cref{sec:normal-normal-location-mixture}.
%whereas the tail-parameter function may take on more than $d$ values. \Cref{fig:radial-fat-tail}
%presents a worst-case scenario.
%As ATAF is only capable of representing $d$ different tail parameters in its pushforward
%(\Cref{remark:anisotropic}), it will necessarily incur an approximation error when $> d$ different
%tail parameters are present.

% Compared to \citep{jaini2020tails}, the degrees of freedom $\nu$ is no longer
% shared across all $d$ dimensions so the variational approximations may exhibit
% varying degrees of fat-tailedness across different dimensions. As seen in
% \Cref{fig:pancake}, such anisotropy is particularly important in situations
% such as the ``heavy-tailed pancake.'' This situation is particularly relevant in
% probabilistic programming, where multiple latent variariables (potentially of different
% tail index) are blocked together for joint sampling / approximation.

%\vspace{-2mm}
%\subsection{Discussion of limitations}
%\label{ssec:limitations}

%\vspace{-1mm}
%By only considering projections $\braket{X,v}$ along rays through the origin
%$v \in \cS^{d-1}$, the limit in $\alpha_X(\cdot)$ may not be defined
%(see \cref{eg:spiral}).

%Furthermore, 
% We found that applying ATAF when the target is light tailed results in minimal
% error (\Cref{sec:normal-normal-location-mixture}); the result agrees with
% intuition because $\text{StudentT}(\nu) \to N(0,1)$ as $\nu \to \infty$ so
% it is reasonable to expect ATAF to learn reasonable approximations.

% \subsection{Non-parametric tail index estimation}
% \label{ssec:non-param-index}

% To perform variational inference with $\cQ_{ATAF}(\nu,\Phi)$, note that if
% $\Phi$ is Lipschitz then the tail index of any element of
% $\cQ_{ATAF}(\nu,\Phi)$ depends only on $\nu$. This motivates a two step
% procedure where (1) $\hat\nu$ estimates the degrees of freedom and
% (2) standard normalizing-flow variational inference $\max_\Phi \cQ_{ATAF}(\hat\nu,\Phi)$ is performed. The key advantage here is that both steps are standard,
% so existing techniques for tail index estimation may be applied.

% Let $X$ be a random variable with fat-tailed density $\pi(x)$
% with tail index $\alpha$ which we wish to estimate.
% When samples $X_i \simiid \pi$ are tractable
% (e.g., no \texttt{observe} or \texttt{condition} statements), traditional
% tail-index estimation techniques may be applied:

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=0.6\textwidth]{../Figures/hill.png}
%   \caption{Hill estimator against $\nu=1.5$ StudentT}
%   \label{fig:hill}
% \end{figure}

% Oftentimes $\pi$ is the posterior distribution of latent variable $X$
% after conditioning on observations $Y=y$, in which case
% $\pi(x) = p(X=x \mid Y=y) = \frac{p(x,y)}{p(y)}$
% and sampling from $\pi$ is non-trivial.
% To circumvent this, we propose the following asymptotically correct
% estimator. Note for $a, b$ large:
% \begin{align*}
%   \Pr[X > a]                 & \sim b^{-\alpha}                                           \\
%   \pi(a)                     & \sim \alpha a^{-\alpha - 1}                                \\
%   \log \pi(a)                & \sim \log \alpha - (\alpha + 1) \log a                     \\
%   \log \frac{p(a,y)}{p(b,y)} & = \log \frac{\pi(a)}{\pi(b)}
%   \sim (\alpha + 1) \log \frac{b}{a}                                                      \\
%   \alpha                     & \sim \frac{\log p(a,y) - \log p(b,y)}{\log b - \log a} - 1
% \end{align*}
% The joint density $p(x,y)$ is tractably computed by running the
% probabilistic program forwards.
% For example. setting $a=10$ and $b=20$ yields an estimate of $1.4799$ for a
% $\nu=1.5$ StudentT.

% \subsubsection{Approximating 1D-marginals}

% This is an example of a two-point log-linear extrapolation at points $a$ and $b$.
% \feynman{Consider Richardson extrapolation to accelerate this limit. Rate of convergece in limit must be a power}

% In general, PPs may include several latent variables.
% The 1-dimensional marginal $p(x, y) = \int_{\setminus x} p(x,z,y) dz$
% is required to apply the above interpolation method.
% We propose approximating this integral using discrete samples.
% If $z$ is a descendent of $y$, then
% we can compute $p(x,y)$ by executing the probabilistic program
% (possibly terminating early).
% If $z$ is an ancestor of $x$ (hence also an ancestor of $y$), then
% \begin{align*}
%   p(x,y) \approx \frac{1}{N} \sum_{i=1}^N p(y, x \mid z_i)
%   \qquad z_i \simiid p(z)
% \end{align*}
% In the last case where $x \rightarrow z \rightarrow y$,
% we have
% \begin{align*}
%   p(x,y) \approx p(x) \frac{1}{N} \sum_{i=1}^n p(y \mid x, z_i)
%   \qquad z_i \simiid p(z_i \mid x)
% \end{align*}

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=0.9\textwidth]{../Figures/our-estimator.png}
%   \caption{Tail index estimation on location-scale mixture representation
%     for StudentT, where the mixture is discretized to $n$ components.}
%   \label{fig:approx_marg}
% \end{figure}

\vspace{-2mm}
\section{Experiments}
\label{sec:experiments}

\vspace{-1mm}
Here we validate ATAF's ability to improve
a range of probabilistic modeling tasks.
Prior work \citep{jaini2020tails} demonstrated improved
density modelling when fat tails are considered, and
our experiments are complementary by evaluating TAFs and ATAFs for variational inference tasks as well as by demonstrating the effect of tail-anisotropy for modelling real-world financial returns and insurance claims datasets.
We implement using the 
\texttt{beanmachine} probabilistic programming language \citep{tehrani2020bean}
%\texttt{REDACTED FOR REVIEW} probabilistic programming language[REDACTED]
and the
\texttt{flowtorch} library for normalizing flows \citep{flowtorchai},
%\texttt{REDACTED FOR REVIEW} library for normalizing flows[REDACTED],
%and we have open-sourced the code\citep{ghlic}
and we have open-sourced code for reproducing experiments in Supplementary Materials. 
% TODO: add facebookresearch reference 
Additional details for the experiments are detailed in \Cref{sec:additional-exp-details}.

% These experiments investigate the behavior of neural density estimators with
% \emph{heavy-tailed base distribution}.
% Specifically, we consider a masked autoregressive flow \cite{papamakarios2017masked}
% transform of a generalized Student's t distribution as a density estimator $q_\theta(X)$
% in a variational inference framework. To fit $q_\theta$ to a target distribution
% $\pi$, the ELBO gradient is reparameterized and Monte-Carlo approximated
% \begin{align*}
%   \nabla_\theta \EE_{q_\theta} \log \frac{\pi(X)}{q_\theta(X)}
%   & = \nabla_\theta \EE_p \log \frac{
%     \pi(X)
%   }{p_\theta(f_\theta^{-1}(X))
%   \left\lvert \det \nabla f_\theta^{-1}(X) \right\rvert}    \\
%   & = \EE_p \nabla_\theta \log \frac{
%     \pi(X)
%   }{p_\theta(f_\theta^{-1}(X))
%   \left\lvert \det \nabla f_\theta^{-1}(X) \right\rvert}    \\
%   & \approx \frac{1}{n} \sum_i^n \nabla_\theta \log \frac{
%     \pi(x_i)
%   }{p_\theta(f_\theta^{-1}(x_i))
%     \left\lvert \det \nabla f_\theta^{-1}(x_i) \right\rvert}
% \end{align*}

\begin{figure*}[htbp]
  \centering
  \vspace{-0.2cm}
  \includegraphics{../Figures/blr_aniso.pdf}
  \vspace{-0.3cm}
  \caption{
    Bayesian linear regression's tail-anisotropic posterior
    (top left) exhibits a fat-tailed conditional in $\sigma$ (as evidenced by
    the convex power-law decay in the top middle panel) and a Gaussian conditional in $\beta$ (concave graph in top right panel).
    While all methods appear to provide a good approximation of the bulk (left column),
    \Cref{prop:isotropic-pushforward} implies
    Gaussian (Gaussian, second row) or isotropic StudentT product (TAF, third row) base distributions
    yield Gaussian or power-law tails, respectively, for \emph{both} $\sigma$ and $\beta$.
    In contrast, ATAF (bottom row) illustrates \Cref{remark:anisotropic} by
    modeling simultaneously a power-law tail on $\sigma$ and Gaussian tail on $\beta$.
  }
  \label{fig:blr-anisotropic}
\end{figure*}

\begin{table*}[htbp]
  \centering
%   \includegraphics[width=1.0\textwidth]{../Figures/eight_schools.png}
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \begin{tabular}{rcc}
            \toprule
                      & ELBO                & $\log p(y)$       \\
            \midrule
            ADVI      & $\mathbf{2873.90} \pm 6.95$    & $2969.73 \pm 1.73$ \\
            TAF       & $2839.64 \pm 9.10$    & $2973.85 \pm 0.87$ \\
            ATAF      & $2842.75 \pm 8.83$    & $\mathbf{2976.75} \pm 0.66$ \\
            \hline
            NUTS      & n/a                  & $3724.59 \pm 0.036$ \\
            \bottomrule
        \end{tabular}
        \caption{diamonds}
        \label{tab:diamonds}
    \end{subfigure}
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \begin{tabular}{rcc}
            \toprule
                      & ELBO                & $\log p(y)$       \\
            \midrule
            ADVI      & $-72.13 \pm 6.89$    & $-53.25 \pm 3.44$ \\
            TAF       & $-64.64 \pm 4.88$    & $-52.51 \pm 4.41$ \\
            ATAF      & $\mathbf{-58.63} \pm 4.75$    & $\mathbf{-51.01} \pm 3.71$ \\
            \hline
            NUTS      & n/a                  & $-47.78 \pm 0.093$ \\
            \bottomrule
        \end{tabular}
        \caption{Eight schools}
        \label{tab:eight_schools}
    \end{subfigure}
    \vspace{-2mm}
        \caption{Monte-Carlo ELBO and importance weighted Monte-Carlo marginal likelihood
        $p(y) = \EE_{x \sim q_\theta} \frac{p(x,y)}{q_\theta(x)}$ (higher is better, $\pm$ standard errors)
        estimates from VI on real-world datasets.
        To understand the variational approximation gap, we include marginal likelihoods based on ``golden samples'' from \texttt{posteriordb} \citep{ghposteriordb} computed using No-U-Turn-Sampling (NUTS, \citet{hoffman2014no,carpenter2017stan}).
    }
  \label{fig:eight_schools}
  \vspace{-4mm}
\end{table*}

\begin{table*}[htbp]
    \centering
    \begin{tabular}{rcc}
        \toprule
                  & Fama-French 5 Industry Daily & CMS 2008-2010 DE-SynPUF       \\
        \midrule
        ADVI      & $-5.018 \pm 0.056$    & $-1.883 \pm 0.012$ \\
        TAF       & $-4.703 \pm 0.023$    & $-1.659 \pm 0.004$ \\
        ATAF      & $\mathbf{-4.699} \pm 0.024$    & $\mathbf{-1.603} \pm 0.034$ \\
        \bottomrule
    \end{tabular}
    \vspace{-2mm}
    \caption{
        Log-likelihoods (higher is better, $\pm$ standard errors) achieved on
        density modeling tasks involving financial returns \citep{fama2015five} and insurance claims \citep{cms} data. 
        \vspace{-5mm}
    }
    \label{tab:density-estimation}
\end{table*}

\subsection{Bayesian Linear Regression}

Consider one-dimensional Bayesian linear regression (BLR)
with conjugate priors, defined by priors and likelihood
\begin{gather*}
    \sigma^2 \sim \text{Inv-Gamma}(a_0, b_0)\\
    \beta \mid \sigma^2 \sim \cN(0, \sigma^2),\qquad
    y \mid X, \beta, \sigma \sim \cN(X \beta, \sigma^2) ,
\end{gather*}
where $a_0$, $b_0$ are hyperparameters and the task is to approximate the posterior
distribution $p(\beta,\sigma^2 \mid X, y)$. Owing to conjugacy,
the posterior distribution can be explicitly computed. Indeed, $p(\beta,\sigma^2 \mid X, y) = \rho(\sigma^2)\rho(\beta \mid \sigma)$ where $\rho(\beta \mid \sigma) = \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 \Sigma_n)$, $\Sigma_n = (X^\top X + \sigma^{-2})^{-1}$, $\hat\beta = (X^\top X)^{-1} X^\top y$, and
%\begin{align*}
    %p(\beta,\sigma^2 \mid X, y) &= \rho(\sigma^2) \rho(\beta \mid \sigma) \\
    \[
    \rho(\sigma^2) = \text{Inv-Gamma}\bigg(
    %\underbrace{a_0 + \frac{n}{2}}_{\eqqcolon a_n}, 
    a_0 + \frac{n}{2}, 
    b_0 + \frac{1}{2}(y^\top y - \mu_n^\top \Sigma_n \mu_n)\bigg). % \\
    \]
    %\rho(\beta \mid \sigma) &= \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 (X^\top X + \sigma^{-2} I)^{-1})
%\end{align*}
This calculation reveals that the posterior distribution is tail-anisotropic:
for fixed $c$ we have that $p(\sigma^2, \beta=c \mid X, y) \propto \rho(\sigma^2) \in \cL^1_{\alpha_n}$
as a function of $\sigma$ (with $\alpha_n$ a function of $n$)
and $p(\sigma^2=c, \beta \mid X, y) \propto \rho(\beta \mid c) \in \overline{\cE^2}$
as a function of $\beta$.
As a result of \Cref{prop:isotropic-pushforward}, we expect ADVI and TAF to erroneously impose
Gaussian and power-law tails respectively for both $\beta$ and $\sigma^2$ as neither method
can produce a tail-anisotropic pushforward. This intuition is confirmed in \Cref{fig:blr-anisotropic},
where we see that only ATAF is the only method capable of modeling the tail-anisotropy present in the data.

Conducting Bayesian linear regression is among the standard tasks requested of a probabilistic programming language, yet it still displays tail-anisotropy. To accurately capture large quantiles, this tail-anisotropy  should not be ignored, necessitating a method such as ATAF.

\subsection{Diamond Price Prediction Using Non-Conjugate Bayesian Regression}
\label{ssec:diamonds}

Without conjugacy, the BLR posterior is intractable and there is no reason \emph{a priori} to expect tail-anisotropy.
Regardless, this presents a realistic and practical scenario for evaluating ATAF's ability to improve VI.
For this experiment, we consider BLR on the \texttt{diamonds} dataset \citep{wickham2011ggplot2} included in
\texttt{posteriordb} \citep{ghposteriordb}.
This dataset contains a covariate matrix $X \in \RR^{5000 \times 24}$ consisting of $5000$
diamonds each with $24$ features as well as an outcome variable $y \in \RR^{5000}$ representing each diamond's price.
The probabilistic model for this inference task is specified in Stan code provided by \citet{ghposteriordb} and is reproduced here
for convenience:
\begin{gather*}
    \alpha \sim \text{StudentT}(\nu=3, \text{loc}=8, \text{scale}=10)\\
    \sigma \sim \text{HalfStudentT}(\nu=3, \text{loc}=0, \text{scale}=10)\\
    \beta \sim \cN(0, \mI_{24}),\qquad
    y \sim \cN(\alpha + X \beta, \sigma) .
\end{gather*}

For each VI method, we performed 100 trials each consisting of 5000 descent steps
on the Monte-Carlo ELBO estimated using 1000 samples and report the results in
\Cref{tab:diamonds}. We report both the final Monte-Carlo ELBO
as well as a Monte-Carlo importance-weighted approximation to
the log marginal likelihood $\log p(y) = \log \EE_{x \sim q_\theta} \frac{p(x,y)}{q_\theta(y)}$
both estimated using 1000 samples.

% \subsection{Bayesian Robust Linear Regression}

% $n = 100$, $d = 10$.

% $X_{ij} \simiid N(0,1)$ for $i \in [n]$, $j \in [d]$.

% $y_i \simiid \text{StudentT}(\text{loc}=X \beta, df=5)$

% Improper ``flat'' prior on $\beta$ to ensure heavy-tailed posterior.

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=0.9\textwidth]{../Figures/blr.png}
%   \caption{VI of Bayesian robust linear regression}
%   \label{fig:blr}
% \end{figure}

\subsection{Eight Schools SAT Score Modelling with Fat-tailed Scale Mixtures}
\label{ssec:eight_schools}

The eight-schools model \citep{rubin1981estimation,gelman2013bayesian} is a classical
Bayesian hierarchical model used originally to consider the relationship between standardized
test scores and coaching programs in place at eight schools.
A variation using half Cauchy non-informative priors \citep{gelman2006prior} provides
a real-world inference problem involving fat-tailed distributions, and is formally specified
by the probabilistic model
\begin{gather*}
    \tau \sim \text{HalfCauchy}(\text{loc}=0, \text{scale}=5)\\
    \mu \sim \cN(0, 5),\qquad
    \theta \sim \cN(\mu, \tau),\qquad
    y \sim \cN(\theta, \sigma) .
\end{gather*}
Given test scores and standard errors $\{(y_i, \sigma_i)\}_{i=1}^8$, we are interested in the
posterior distribution over treatment effects $\theta_1,\ldots,\theta_d$. The experimental
parameters are identical to \Cref{ssec:diamonds}, and results are reported in \Cref{tab:eight_schools}.

\subsection{Financial and Actuarial Applications}

To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021 \citep{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010 \citep{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events. Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}. Both datasets exhibited superior fits after allowing for heavier tails, with a further improved fit using ATAF for the CMS claims dataset. 
%Using identical flow architectures and optimizers, the log likelihoods in \Cref{tab:density-estimation} suggest that models permitting tail-anisotropy (ATAF) yield significantly superior fits.

% \feynman{TODO}

% \subsection{Bayesian regression analysis on diamonds}
% \feynman{Remove? These are taking too long and don't look good}

% \begin{table}[htbp]
%   \caption{Final ELBO and (MC estimate of) log marginal $\log P(y)$ after 10000 steps on \texttt{diamonds}}
%   \label{fig:blr_diamonds}
%   \centering
%   \begin{tabular}{rcc}
%     \toprule
%               & ELBO   & $\log P(y)$ \\
%     \midrule
%     ADVI      & -374.4 & -6912.0           \\
%     ADVI-HT   & -246.3 & -6876.6     \\
%     MAF       & -211.1 & -6894.8     \\
%     MAF-HT    & -208.5 & -7442.0     \\
%     MAF-2L    & -197.9 & -6532.4     \\
%     MAF-2L-HT & -191.9 & -6839.9     \\
%     MAF-3L    & -194.1 & -7027.3     \\
%     MAF-3L-HT & -209.8 & -7128.1     \\
%     \bottomrule
%   \end{tabular}
% \end{table}

% \feynman{
%   % \subsection{Importance weights}

%   When the importance sampling density is more peaked than the target density.

%   \cite{wang2018variational} example 3.1: Let $p = N(0, \sigma_p^2)$, $q = N(0, \sigma_q^2)$,
%   and for $x \sim q$ let $w(x) = \frac{p(x)}{q(x)}$. If $\sigma_p > \sigma_q$,
%   then $w$ has tail index $\frac{\sigma_p^2}{\sigma_p^2 - \sigma_q^2}$.
%   Otherwise, $w$ is not fat-tailed.
% }

%\section{Related Work}

\section{Conclusion}
\label{sec:conclusion}

In this work, we have sharpened existing theory for approximating fat-tailed distributions with normalizing flows, and we formalized tail-(an)isotropy through a direction-dependent tail parameter. With this, we have shown that many prior flow-based methods are inherently limited by tail-isotropy. With this in mind, we proposed a simple flow-based method capable of modeling tail-anisotropic targets.
As we have seen, anisotropic FTVI is already applicable in fairly elementary examples such as Bayesian linear regression;
and ATAFs provide one of the first methods for using the representational capacity of flow-based methods,
while simultaneously producing tail-anisotropic distributions. A number of open problems still remain, including the study of other parameterizations of the tail behaviour of the base distribution. Even so, going forward, it seems prudent that density estimators, especially those used in black-box settings, consider accounting for tail-anisotropy using a method such as~ATAF.% expect the scientific impact of ATAFs to be
%fairly easily realized, as it can be applied wherever a parametric density estimator is used.
%While we do not expect ATAF to directly have any negative societal impacts,
%its broad applicability as a parametric density estimator which accommodates fat-tails means that it may be
%used to improve applications which do cause negative societal impact.

% VI is important, FTVI is very important, ADVI did a good job with FTVI, we do a great job with FTVI, and you the reader should follow up on us and do FTVI.

\paragraph{Acknowledgements}
L.H. and M.M's contributions were supported in part by DARPA, NSF, and ONR.
F.L.’s research while at the University of California, Berkeley was supported in part by a GFSD fellowship.

\bibliography{refs_ftvi}
\bibliographystyle{icml2022}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Checklist}

% %%% BEGIN INSTRUCTIONS %%%
% % The checklist follows the references.  Please
% % read the checklist guidelines carefully for information on how to answer these
% % questions.  For each question, change the default \answerTODO{} to \answerYes{},
% % \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
% %     justification to your answer}, either by referencing the appropriate section of
% % your paper or providing a brief inline description.  For example:
% % \begin{itemize}
% %   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
% %   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
% %   \item Did you include the license to the code and datasets? \answerNA{}
% % \end{itemize}
% % Please do not modify the questions and only use the provided macros for your
% % answers.  Note that the Checklist section does not count towards the page
% % limit.  In your paper, please delete this instructions block and only keep the
% % Checklist section heading above along with the questions/answers below.
% %%% END INSTRUCTIONS %%%

% \begin{enumerate}

%   \item For all authors...
%         \begin{enumerate}
%           \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
%                 \answerYes{}
%           \item Did you describe the limitations of your work?
%                 \answerYes{See \Cref{ssec:limitations}}
%           \item Did you discuss any potential negative societal impacts of your work?
%                 \answerYes{See \Cref{sec:conclusion}}
%           \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
%                 \answerYes{}
%         \end{enumerate}

%   \item If you are including theoretical results...
%         \begin{enumerate}
%           \item Did you state the full set of assumptions of all theoretical results?
%                 \answerYes{}
%           \item Did you include complete proofs of all theoretical results?
%                 \answerYes{Deferred to \Cref{sec:proofs}}
%         \end{enumerate}

%   \item If you ran experiments...
%         \begin{enumerate}
%           \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
%                 \answerYes{See \Cref{sec:experiments}}
%           \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
%                 \answerYes{See \Cref{sec:experiments}, \Cref{sec:additional-exp-details}, and open-sourced code}
%           \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
%                 \answerYes{Tables/figures report standard errors}
%           \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
%                 \answerYes{See \Cref{sec:additional-exp-details}}
%         \end{enumerate}

%   \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
%         \begin{enumerate}
%           \item If your work uses existing assets, did you cite the creators?
%                 \answerYes{}
%           \item Did you mention the license of the assets?
%                 \answerYes{See \texttt{LICENSE} and \texttt{README.md} file on Github repository.}
%           \item Did you include any new assets either in the supplemental material or as a URL?
%                 \answerYes{Experiments use newly developed code within \texttt{beanmachine} which is released in \Cref{sec:experiments}.}
%           \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
%                 \answerNo{Datasets used are either synthetic or released as part of MIT/BSD-3 licensed code.}
%           \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
%                 \answerNo{We did not screen datasets for PII / offensive content. All experiments perform probabilistic modeling on top of
%                 existing data, so we do not expect any PII / offensive content to appear in the body of the paper. Please notify us if something
%                 could potentially make a reader uncomfortable and we will be happy to fix it!
%                 }
%         \end{enumerate}

%   \item If you used crowdsourcing or conducted research with human subjects...
%         \begin{enumerate}
%           \item Did you include the full text of instructions given to participants and screenshots, if applicable?
%                 \answerNA{}
%           \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
%                 \answerNA{}
%           \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
%                 \answerNA{}
%         \end{enumerate}

% \end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\section{Experiments Performing VI Against a Fat-tailed Cauchy Target}
\label{sec:cauchy_normal_student}
The motivation for the fat-tailed variational families used in TAF/ATAF
is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.

\begin{figure*}[htbp]
  \centering
  \includegraphics{../Figures/fat_tail_ks.pdf}
  \vspace{-6mm}
  \caption{
    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
    the use of a Gaussian variational family (ADVI, solid blue line) can incur
    exponentially bad tail approximations (middle panel) compared to
    methods such as ATAF which permit heavier tails (orange dashed line).
    As a consequence, ADVI samples (blue, right panel) are rejected by
    the Kolmogorov-Smirnov test more often than ATAF samples (orange, right panel).
    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.  It might be good to show ATAF is much better but still not perfect.  Also, what is the X axis on each figure.  }
  }
  \label{fig:cauchy_normal_student}
\end{figure*}

\section{Proofs of Our Main Theoretical Results}
\label{sec:proofs}

\begin{proof}[Proof of \Cref{thm:distn_class_closed}]
  \label{proof:distn_class_closed}
  Let $X$ be a random variable from either $\cE_\alpha^p$
  or $\cL_\alpha^p$.
  Its concentration function
  (Equation 1.6 \citet{ledoux2001concentration}
  is given by
  \[
    \alpha_X(r)
    \coloneqq \sup \{ \mu\{x : d(x,A) \geq r\}; A \subset \text{supp}~X, \mu(A) \geq 1/2\}
    = \PP(\lvert X - m_X \rvert \geq r) .
  \]
  Under Assumption 1, $f_\theta$ is Lipschitz (say with Lipschitz
  constant $L$) so by Proposition 1.3 of \citet{ledoux2001concentration},
  \[
    \PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)
    \leq 2 \alpha_X(r/L)
    = \cO(\alpha_X(r/L)),
  \]
  where $m_{f_\theta(X)}$ is a median of $f_\theta(X)$.
  Furthermore, by the triangle inequality
  \begin{align}
    \PP(\lvert f_\theta(X) \rvert \geq r)
    &= \PP(\lvert f_\theta(X) - m_{f_\theta(X)} + m_{f_\theta(X)} \rvert \geq r) \nonumber\\
    &\leq \PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r - \lvert m_{f_\theta(X)}\rvert ) \nonumber\\
    &= \cO(\PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)) \nonumber\\
    &= \cO(\alpha_X(r/L)) ,
    \label{eq:pushforward-conc-fn}
  \end{align}
  where the asymptotic equivalence holds because $\lvert m_{f_\theta(X)} \rvert$ is independent of $r$.
  When $X \in \cE_\alpha^p$, \Cref{eq:pushforward-conc-fn} implies
  \[
    \PP(\lvert f_\theta(X) \rvert \geq r)
    = \cO(e^{-\frac{\alpha}{L} r^p}) \implies f_\theta(X) \in \overline{\cE}_{\alpha/L}^p,
  \]
  from whence we find that the Lipschitz transform of exponential-type
  tails continues to possess exponential-type tails with the same
  class index $p$, although the tail parameter may have changed. Hence,
  $\overline{\cE^p}$ is closed under Lipschitz maps for each $p \in \RR_{>0}$.
  On the other hand, when $X \in \cL_\alpha^p$, \Cref{eq:pushforward-conc-fn} also implies that
  \begin{align*}
    \PP(\lvert f_\theta(X) \rvert \geq r)
    &= \cO(e^{-\alpha (\log (r/L))^p})
    % &= \cO(e^{-\alpha (\log r)^p} e^{-\alpha (-\log L)^p}) \\
    = \cO(e^{-\alpha (\log r)^p}),
  \end{align*}
  and therefore, $f_\theta(X) \in \overline{\cL_\alpha^p}$.
  Unlike exponential-type tails, Lipschitz transforms of
  logarithmic-type tails not only remain logarithmic, but
  their tails decay no slower than a logarithmic-type tail
  of the same class index with the \emph{same} tail parameter $\alpha$.
  This upper bound suffices to show closure under Lipschitz maps for the
  ascending family $\overline{\cL_\alpha^p}$.
\end{proof}

\begin{proof}[Proof of \Cref{corr:heavy_to_light}]
    Let $f_\theta$ be as before with the additional assumptions.
    Since $f_\theta$ is a smooth continuous bijection, it is a diffeomorphism.
    Furthermore, by assumption $f_\theta$ has invertible Jacobian on the closure of its
    domain hence $\sup_{x \in \text{dom}~f_\theta} \lvert (f_\theta)'(x) \rvert \geq M > 0$.
    By the inverse function theorem, $(f_\theta)^{-1}$ exists and is
    a diffeomorphism with
    \[
    \frac{d}{dx}(f_\theta)^{-1}(x) = \frac{1}{(f_\theta)'((f_\theta)^{-1}(x))} \leq \frac{1}{M} .
    \]
    Therefore, $(f_\theta)^{-1}$ is $M^{-1}$-Lipschitz and we may apply
    \Cref{thm:distn_class_closed} to conclude the desired result.
    %\footnote{\url{https://math.stackexchange.com/questions/394908/diffeomorphism-from-inverse-function-theorem}}
\end{proof}

\begin{proof}[Proof of \Cref{corr:closure_polynomials}]
  Let $X \in \cE^p_\alpha$.
  By considering sufficiently large $X$ such that leading powers dominate, it suffices to consider monomials $Y = X^k$.
  Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$. 
\end{proof}

\begin{lemma}
    \label[lemma]{lem:sum-rule}
    Suppose $X \in \cL^1_\alpha$ and $Y \in \cL^1_\beta$.
    Then $X + Y \in \cL^1_{\min\{\alpha,\beta\}}$.
\end{lemma}

\begin{proof}
First, let $\gamma=\min\{\alpha,\beta\}$. It will suffice to show that (I) $\mathbb{P}(|X+Y|\geq r)=\mathcal{O}(r^{-\gamma})$, and (II) $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\gamma})$. Since $(X,Y)\mapsto|X+Y|$ is a 1-Lipschitz function on $\mathbb{R}^{2}$ and $\mathbb{P}(|X|\geq r)+\mathbb{P}(|Y|\geq r)=\mathcal{O}(r^{-\gamma})$, (I) follows directly from the hypotheses and Proposition 1.11 of \citet{ledoux2001concentration}. To show (II), note that for any $M>0$, conditioning on the event $|Y|\leq M$,\[
\mathbb{P}\left(\left|X\right|+|Y|\geq r\,\vert\,|Y|\leq M\right)\geq\mathbb{P}\left(\left|X\right|\geq r-M\right).
\]
Therefore, by taking $M$ to be sufficiently large so that $\mathbb{P}(|Y|\leq M)\geq\frac{1}{2}$,
\begin{align*}
\mathbb{P}\left(|X+Y|\geq r\right)&\geq\mathbb{P}\left(|X|+|Y|\geq r\right)\\
&\geq\mathbb{P}\left(\left|X\right|+|Y|\geq r\,\vert\,|Y|\leq M\right)\mathbb{P}\left(\left|Y\right|\leq M\right)\\
&\geq\frac{1}{2}\mathbb{P}\left(\left|X\right|\geq r-M\right)=\Theta(r^{-\alpha}).
\end{align*}
The same process with $X$ and $Y$ reversed implies $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\beta})$ as well. Both (II) and the claim follow.
\end{proof}

To show Proposition \ref{prop:isotropic-pushforward}, we will require a few extra assumptions to rule out pathological cases. The full content of Proposition \ref{prop:isotropic-pushforward} is contained in the following theorem.

\begin{theorem}
Suppose there exists $\nu > 0$ such that $C:\mathcal{S}^{d-1}\to(0,\infty)$ satisfies $C(v) \coloneqq \lim_{x \to \infty}x^{\nu}\mathbb{P}(|\langle v, X\rangle| > x)$ for all $v \in \mathcal{S}^{d-1}$. If $\nu$ is not an integer and $f$ is a bilipschitz function,
%such that $f(X)/\|f(X)\|$ has full support on $\mathcal{S}^{d-1}$
then $f(X)$ is tail-isotropic with tail index $\nu$. 
\end{theorem}
\begin{proof}
Since $x \mapsto \langle v, f(x)\rangle$ is Lipschitz continuous for any $v \in \mathcal{S}^{d-1}$, Theorem \ref{thm:distn_class_closed} implies $\langle v, f(X)\rangle \in \overline{\mathcal{L}_{\nu}^{1}}$. Let $\theta \in (0,\pi/2)$ (say, $\theta = \pi / 4$), and let $S_v = \{x\, : \, \cos^{-1}(\langle x/\|x\|, v\rangle)\leq\theta\}$ for each $v \in \mathcal{S}^{d-1}$. Then
\[
H_v \coloneqq \{x\,:\,\langle v, x \rangle > 1\} \supset \{x\,:\,\|x\| > (1-\cos\theta)^{-1}\} \cap S_v.
\]
From Theorem C.2.1 of \citet{buraczewski2016stochastic}, since $\nu \not \in \mathbb{Z}$, there exists a non-zero measure $\mu$ such that
\[
\mu(E) = \lim_{x \to \infty} \frac{\mathbb{P}(x^{-1}X \in E)}{\mathbb{P}(\|X\| > x)},
\]
for any Borel set $E$. Consequently, $\mu$ is regularly varying, and so 
by the spectral representation of regularly varying random vectors (see p. 281 \citet{buraczewski2016stochastic}), there exists a measure $P$ such that
\[
\lim_{x \to \infty}\frac{\mathbb{P}(\|X\|>tx, X/\|X\| \in E)}{\mathbb{P}(\|X\| > x)} = t^{-\nu} P(E),
\]
for any Borel set $E$ on $\mathcal{S}^{d-1}$ and any $t > 0$. Letting $F_v = \{y / \|y\|\,:\, f(y) \in S_v\} \subset \mathcal{S}^{d-1}$ (noting that $P(F_v) > 0$ by assumption), since $m\|x - y\| \leq \|f(x) - f(y)\| \leq M\|x - y\|$ for all $x,y$,
\begin{align*}
\liminf_{x \to \infty}
\frac{\mathbb{P}(f(X) \in x H_v)}{\mathbb{P}(\|f(X)\| > x)} 
&\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|f(X)\| > x(1-\cos\theta)^{-1}, f(X) \in S_v)}{\mathbb{P}(\|f(X)\| > x)} \\
&\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|X\| > x(m(1-\cos\theta))^{-1}, X/\|X\| \in F_v)}{\|X\| > x / M} \\
&\geq P(F_v) \left(\frac{M}{m(1-\cos\theta)}\right)^{-\nu} > 0,
yaB\end{align*}
where $P(F_v) > 0$ follows from the bilipschitz condition for $f$. Therefore, we have shown that $\mathbb{P}(\langle v, f(X)\rangle > x) = \Theta(\mathbb{P}(\|f(X)\| > x))$ for every $v \in \mathcal{S}^{d-1}$.
Since $\mathbb{P}(\|f(X)\| > x)$ obeys a power law with exponent $\nu$ by Corollary \ref{corr:heavy_to_light}, $f(X)$ is tail-isotropic with exponent $\nu$.
\end{proof}

%\begin{proof}[Proof of \Cref{prop:isotropic-pushforward}]
%  Since $f_\theta$ satisfies \cref{assump:lipschitz}, we have $\max(\|f_\theta\|_{\text{Lip}},\|f_\theta^{-1}\|_{\text{Lip}} \leq L < \infty$

%   Fix $z \in \cS^{d-1}$.
%   Note tail parameters are invariant under scalar multiplication as a consequence of
%   asymptotic notation, so combined with repeated applications of \Cref{lem:sum-rule} we
%   have that the linear combination $\braket{f_\theta(X), z}$
%   of $\{f_\theta(X)_i\}_{i \in [d]}$ is tail-isotropic provided
%   every $f_\theta(X)_i \in \cL_\nu^1$.
%   Let $i \in [d]$ be arbitrary.
%   Notice that the projection map $\pi_i = x \mapsto x_i$
%   is $1$-Lipschitz hence $\pi_i \circ f_\theta = x \mapsto f_\theta(x)_i$
%   is $L$-Lipschitz. In particular, $g_{x_{-i}} = x_i \mapsto (\pi \circ_i f_\theta)(x_{-i}, x_i)$ is $L$-Lipschitz for every $x_{-i} \in \mathbb{R}^{d-1}$.

%   Consider the $d$ times iterated conditional expectation
%   \begin{align*}
%       \mathbb{P}[f_\theta(X)_i \geq t]
%       &= \EE[\EE[\cdots\EE[1(f(X)_i \geq t) \mid X_1, \ldots, X_{i-1},X_{i+1},\ldots,X_d] \cdots \mid X_1 ]] \\
%       &= \EE[\EE[\cdots\mathbb{P}[g_{X_{-i}}(X_i) \geq t \mid X_{-i} ] \cdots \mid X_1 ]]
%   \end{align*}
%   Since $X \sim \mu$ is $\nu$-isotropic, we have that
%   $X_i = \braket{X, e_i} \in \cL_\nu^1$.
%   \feynman{To apply \cref{corr:heavy_to_light}, we need $x_i \mapsto f(x_{-i}, x_i)_i$ to be bi-Lipschitz for every $i \in [d]$ i.e. separately bilipschitz, see component Lipschitz constants in \url{http://www.optimization-online.org/DB_FILE/2014/12/4679.pdf}}
%\end{proof}

\section{Example of Non-existence of Tail Parameter Due to Oscillations}
\label{eg:spiral}

Consider $\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.

\begin{figure*}[htbp]
    \centering
    \vspace{-.5in}
    \includegraphics[scale=0.8]{Figures/spiral.pdf}
    % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\
    % \includegraphics[trim={5cm 0 4.9cm 0},clip]{Figures/spiral.pdf}\\
    % \includegraphics[trim={9.3cm 0 0cm 0},clip]{Figures/spiral.pdf}\\
    \caption{Taking a tail-anisotropic distribution (left) and ``spinning'' it (middle) results in
        one-dimensional projections which oscillate between tail parameters (as seen in
        $\log p(\braket{X,e_0})$ in right panel) and result in an ill-defined
        direction-dependent tail parameter function $\alpha_X(\cdot)$ due to a
        divergent limit.
    }
    \label{fig:spiral}
\end{figure*}

\section{Normal-normal Conjugate Model}
\label{sec:normal-normal-location-mixture}

We consider a Normal-Normal conjugate inference problem where the posterior
is known to be a Normal distribution as well. Here, we aim to show that ATAF
performs no worse than ADVI because $\text{StudentT}(\nu) \to N(0, 1)$ as $\nu \to \infty$.
\Cref{fig:normal_normal} shows the resulting density approximation, which can
be seen to be reasonable for both a Normal base distribution (the ``correct'' one)
and a StudentT base distribution. This suggests that mis-specification (i.e., heavier
tails in the base distribution than the target) may not be too problematic.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\textwidth]{../Figures/normal_normal_posterior.png}
  \caption{Variational inference against a light tailed Normal posterior. Both light and heavy tail
  variational families yield similar results.}
  \label{fig:normal_normal}
\end{figure}

\section{Additional Details For Experiments}
\label{sec:additional-exp-details}

All experiments were performed on an Intel i8700K with 32GB RAM and a NVIDIA GTX 1080
running PyTorch 1.9.0 / Python 3.8.5 / CUDA 11.2 / Ubuntu Linux 20.04 via Windows Subsystem for Linux.
For all flow-transforms $\Phi_{\text{Flow}}$, we used inverse autoregressive flows \citep{kingma2016improved} with a
dense autoregressive conditioner consisting of two layers of either 32 or 256 hidden units depending on problem (see code for details) and
ELU activation functions.
As described in \citet{jaini2020tails}, TAF is trained by including $\nu$ within the Adam optimizer alongside other flow parameters. For ATAF, we include all $\nu_i$ within the optimizer.
Models were trained using the Adam optimizer with $10^{-3}$ learning rate
for 10000 iterations, which we found empirically in all our experiments to result in negligible change in ELBO
at the end of training.

For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
is comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empkkirically observed
this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
the absence of outliers when using a Gaussian base distribution resulting in more stable ELBO gradients. This suggests
other techniques for handling outliers such as winsorization may also be helpful, and we leave further investigation
for future work.

For \cref{fig:blr-anisotropic}, the closed-form posterior was computed over a finite element grid to produce
the ``Target'' row. A similar progressive training scheme used for \cref{tab:diamonds} was also used here, with
the TAF flow transform $\Phi_{\text{Flow}}$ initialized from the result of ADVI and ATAF additionally initialized
all tail parameters $\nu_i$ based on the final shared tail parameter obtained from TAF training. Tails are computed
along the $\beta = 1$ or $\sigma = 1$ axes because the posterior is identically zero for $\sigma = 0$, hence it reveals
no information about the tails.

% \section{Electronic Submission}
% \label{submission}

% Submission to ICML 2022 will be entirely electronic, via a web site
% (not email). Information about the submission process and \LaTeX\ templates
% are available on the conference web site at:
% \begin{center}
% \textbf{\texttt{http://icml.cc/}}
% \end{center}

% The guidelines below will be enforced for initial submissions and
% camera-ready copies. Here is a brief summary:
% \begin{itemize}
% \item Submissions must be in PDF\@. 
% \item \textbf{New to this year}: If your paper has appendices, submit the appendix together with the main body and the references \textbf{as a single file}. Reviewers will not look for appendices as a separate PDF file. So if you submit such an extra file, reviewers will very likely miss it.
% \item Page limit: The main body of the paper has to be fitted to 8 pages, excluding references and appendices; the space for the latter two is not limited. For the final version of the paper, authors can add one extra page to the main body.
% \item \textbf{Do not include author information or acknowledgements} in your
%     initial submission.
% \item Your paper should be in \textbf{10 point Times font}.
% \item Make sure your PDF file only uses Type-1 fonts.
% \item Place figure captions \emph{under} the figure (and omit titles from inside
%     the graphic file itself). Place table captions \emph{over} the table.
% \item References must include page numbers whenever possible and be as complete
%     as possible. Place multiple citations in chronological order.
% \item Do not alter the style template; in particular, do not compress the paper
%     format by reducing the vertical spaces.
% \item Keep your abstract brief and self-contained, one paragraph and roughly
%     4--6 sentences. Gross violations will require correction at the
%     camera-ready phase. The title should have content words capitalized.
% \end{itemize}

% \subsection{Submitting Papers}

% \textbf{Paper Deadline:} The deadline for paper submission that is
% advertised on the conference website is strict. If your full,
% anonymized, submission does not reach us on time, it will not be
% considered for publication.

% \textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
% author information may appear on the title page or in the paper
% itself. \cref{author info} gives further details.

% \textbf{Simultaneous Submission:} ICML will not accept any paper which,
% at the time of submission, is under review for another conference or
% has already been published. This policy also applies to papers that
% overlap substantially in technical content with conference papers
% under review or previously published. ICML submissions must not be
% submitted to other conferences and journals during ICML's review
% period.
% %Authors may submit to ICML substantially different versions of journal papers
% %that are currently under review by the journal, but not yet accepted
% %at the time of submission.
% Informal publications, such as technical
% reports or papers in workshop proceedings which do not appear in
% print, do not fall under these restrictions.

% \medskip

% Authors must provide their manuscripts in \textbf{PDF} format.
% Furthermore, please make sure that files contain only embedded Type-1 fonts
% (e.g.,~using the program \texttt{pdffonts} in linux or using
% File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
% might come from graphics files imported into the document.

% Authors using \textbf{Word} must convert their document to PDF\@. Most
% of the latest versions of Word have the facility to do this
% automatically. Submissions will not be accepted in Word format or any
% format other than PDF\@. Really. We're not joking. Don't send Word.

% Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
% Those using \texttt{latex} and \texttt{dvips} may need the following
% two commands:

% {\footnotesize
% \begin{verbatim}
% dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
% ps2pdf paper.ps
% \end{verbatim}}
% It is a zero following the ``-G'', which tells dvips to use
% the config.pdf file. Newer \TeX\ distributions don't always need this
% option.

% Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
% results. This program avoids the Type-3 font problem, and supports more
% advanced features in the \texttt{microtype} package.

% \textbf{Graphics files} should be a reasonable size, and included from
% an appropriate format. Use vector formats (.eps/.pdf) for plots,
% lossless bitmap formats (.png) for raster graphics with sharp lines, and
% jpeg for photo-like images.

% The style file uses the \texttt{hyperref} package to make clickable
% links in documents. If this causes problems for you, add
% \texttt{nohyperref} as one of the options to the \texttt{icml2022}
% usepackage statement.

% \subsection{Submitting Final Camera-Ready Copy}

% The final versions of papers accepted for publication should follow the
% same format and naming convention as initial submissions, except that
% author information (names and affiliations) should be given. See
% \cref{final author} for formatting instructions.

% The footnote, ``Preliminary work. Under review by the International
% Conference on Machine Learning (ICML). Do not distribute.'' must be
% modified to ``\textit{Proceedings of the
% $\mathit{39}^{th}$ International Conference on Machine Learning},
% Baltimore, Maryland, USA, PMLR 162, 2022.
% Copyright 2022 by the author(s).''

% For those using the \textbf{\LaTeX} style file, this change (and others) is
% handled automatically by simply changing
% $\mathtt{\backslash usepackage\{icml2022\}}$ to
% $$\mathtt{\backslash usepackage[accepted]\{icml2022\}}$$
% Authors using \textbf{Word} must edit the
% footnote on the first page of the document themselves.

% Camera-ready copies should have the title of the paper as running head
% on each page except the first one. The running title consists of a
% single line centered above a horizontal rule which is $1$~point thick.
% The running head should be centered, bold and in $9$~point type. The
% rule should be $10$~points above the main text. For those using the
% \textbf{\LaTeX} style file, the original title is automatically set as running
% head using the \texttt{fancyhdr} package which is included in the ICML
% 2022 style file package. In case that the original title exceeds the
% size restrictions, a shorter form can be supplied by using

% \verb|\icmltitlerunning{...}|

% just before $\mathtt{\backslash begin\{document\}}$.
% Authors using \textbf{Word} must edit the header of the document themselves.

% \section{Format of the Paper}

% All submissions must follow the specified format.

% \subsection{Dimensions}

% The text of the paper should be formatted in two columns, with an
% overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
% between the columns. The left margin should be 0.75~inches and the top
% margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
% whether you print on US letter or A4 paper, but all final versions
% must be produced for US letter size.
% Do not write anything on the margins.

% The paper body should be set in 10~point type with a vertical spacing
% of 11~points. Please use Times typeface throughout the text.

% \subsection{Title}

% The paper title should be set in 14~point bold type and centered
% between two horizontal rules that are 1~point thick, with 1.0~inch
% between the top rule and the top edge of the page. Capitalize the
% first letter of content words and put the rest of the title in lower
% case.

% \subsection{Author Information for Submission}
% \label{author info}

% ICML uses double-blind review, so author information must not appear. If
% you are using \LaTeX\/ and the \texttt{icml2022.sty} file, use
% \verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
% will not be printed unless \texttt{accepted} is passed as an argument to the
% style file.
% Submissions that include the author information will not
% be reviewed.

% \subsubsection{Self-Citations}

% If you are citing published papers for which you are an author, refer
% to yourself in the third person. In particular, do not use phrases
% that reveal your identity (e.g., ``in previous work \cite{langley00}, we
% have shown \ldots'').

% Do not anonymize citations in the reference section. The only exception are manuscripts that are
% not yet published (e.g., under submission). If you choose to refer to
% such unpublished manuscripts \cite{anonymous}, anonymized copies have
% to be submitted
% as Supplementary Material via CMT\@. However, keep in mind that an ICML
% paper should be self contained and should contain sufficient detail
% for the reviewers to evaluate the work. In particular, reviewers are
% not required to look at the Supplementary Material when writing their
% review (they are not required to look at more than the first $8$ pages of the submitted document).

% \subsubsection{Camera-Ready Author Information}
% \label{final author}

% If a paper is accepted, a final camera-ready copy must be prepared.
% %
% For camera-ready papers, author information should start 0.3~inches below the
% bottom rule surrounding the title. The authors' names should appear in 10~point
% bold type, in a row, separated by white space, and centered. Author names should
% not be broken across lines. Unbolded superscripted numbers, starting 1, should
% be used to refer to affiliations.

% Affiliations should be numbered in the order of appearance. A single footnote
% block of text should be used to list all the affiliations. (Academic
% affiliations should list Department, University, City, State/Region, Country.
% Similarly for industrial affiliations.)

% Each distinct affiliations should be listed once. If an author has multiple
% affiliations, multiple superscripts should be placed after the name, separated
% by thin spaces. If the authors would like to highlight equal contribution by
% multiple first authors, those authors should have an asterisk placed after their
% name in superscript, and the term ``\textsuperscript{*}Equal contribution"
% should be placed in the footnote block ahead of the list of affiliations. A
% list of corresponding authors and their emails (in the format Full Name
% \textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
% Ideally only one or two names should be listed.

% A sample file with author names is included in the ICML2022 style file
% package. Turn on the \texttt{[accepted]} option to the stylefile to
% see the names rendered. All of the guidelines above are implemented
% by the \LaTeX\ style file.

% \subsection{Abstract}

% The paper abstract should begin in the left column, 0.4~inches below the final
% address. The heading `Abstract' should be centered, bold, and in 11~point type.
% The abstract body should use 10~point type, with a vertical spacing of
% 11~points, and should be indented 0.25~inches more than normal on left-hand and
% right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
% abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
% sentences. Gross violations will require correction at the camera-ready phase.

% \subsection{Partitioning the Text}

% You should organize your paper into sections and paragraphs to help
% readers place a structure on the material and understand its
% contributions.

% \subsubsection{Sections and Subsections}

% Section headings should be numbered, flush left, and set in 11~pt bold
% type with the content words capitalized. Leave 0.25~inches of space
% before the heading and 0.15~inches after the heading.

% Similarly, subsection headings should be numbered, flush left, and set
% in 10~pt bold type with the content words capitalized. Leave
% 0.2~inches of space before the heading and 0.13~inches afterward.

% Finally, subsubsection headings should be numbered, flush left, and
% set in 10~pt small caps with the content words capitalized. Leave
% 0.18~inches of space before the heading and 0.1~inches after the
% heading.

% Please use no more than three levels of headings.

% \subsubsection{Paragraphs and Footnotes}

% Within each section or subsection, you should further partition the
% paper into paragraphs. Do not indent the first line of a given
% paragraph, but insert a blank line between succeeding ones.

% You can use footnotes\footnote{Footnotes
% should be complete sentences.} to provide readers with additional
% information about a topic without interrupting the flow of the paper.
% Indicate footnotes with a number in the text where the point is most
% relevant. Place the footnote in 9~point type at the bottom of the
% column in which it appears. Precede the first footnote in a column
% with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
% appear in each column, in the same order as they appear in the text,
% but spread them across columns and pages if possible.}

% \begin{figure}[ht]
% \vskip 0.2in
% \begin{center}
% \centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
% \caption{Historical locations and number of accepted papers for International
% Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
% Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
% produced, the number of accepted papers for ICML 2008 was unknown and instead
% estimated.}
% \label{icml-historical}
% \end{center}
% \vskip -0.2in
% \end{figure}

% \subsection{Figures}

% You may want to include figures in the paper to illustrate
% your approach and results. Such artwork should be centered,
% legible, and separated from the text. Lines should be dark and at
% least 0.5~points thick for purposes of reproduction, and text should
% not appear on a gray background.

% Label all distinct components of each figure. If the figure takes the
% form of a graph, then give a name for each axis and include a legend
% that briefly describes each curve. Do not include a title inside the
% figure; instead, the caption should serve this function.

% Number figures sequentially, placing the figure number and caption
% \emph{after} the graphics, with at least 0.1~inches of space before
% the caption and 0.1~inches after it, as in
% \cref{icml-historical}. The figure caption should be set in
% 9~point type and centered unless it runs two or more lines, in which
% case it should be flush left. You may float figures to the top or
% bottom of a column, and you may set wide figures across both columns
% (use the environment \texttt{figure*} in \LaTeX). Always place
% two-column figures at the top or bottom of the page.

% \subsection{Algorithms}

% If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
% environments to format pseudocode. These require
% the corresponding stylefiles, algorithm.sty and
% algorithmic.sty, which are supplied with this package.
% \cref{alg:example} shows an example.

% \begin{algorithm}[tb]
%   \caption{Bubble Sort}
%   \label{alg:example}
% \begin{algorithmic}
%   \STATE {\bfseries Input:} data $x_i$, size $m$
%   \REPEAT
%   \STATE Initialize $noChange = true$.
%   \FOR{$i=1$ {\bfseries to} $m-1$}
%   \IF{$x_i > x_{i+1}$}
%   \STATE Swap $x_i$ and $x_{i+1}$
%   \STATE $noChange = false$
%   \ENDIF
%   \ENDFOR
%   \UNTIL{$noChange$ is $true$}
% \end{algorithmic}
% \end{algorithm}

% \subsection{Tables}

% You may also want to include tables that summarize material. Like
% figures, these should be centered, legible, and numbered consecutively.
% However, place the title \emph{above} the table with at least
% 0.1~inches of space before the title and the same after it, as in
% \cref{sample-table}. The table title should be set in 9~point
% type and centered unless it runs two or more lines, in which case it
% should be flush left.

% % Note use of \abovespace and \belowspace to get reasonable spacing
% % above and below tabular lines.

% \begin{table}[t]
% \caption{Classification accuracies for naive Bayes and flexible
% Bayes on various data sets.}
% \label{sample-table}
% \vskip 0.15in
% \begin{center}
% \begin{small}
% \begin{sc}
% \begin{tabular}{lcccr}
% \toprule
% Data set & Naive & Flexible & Better? \\
% \midrule
% Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
% Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
% Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
% Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
% Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
% Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
% Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
% Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
% \bottomrule
% \end{tabular}
% \end{sc}
% \end{small}
% \end{center}
% \vskip -0.1in
% \end{table}

% Tables contain textual material, whereas figures contain graphical material.
% Specify the contents of each row and column in the table's topmost
% row. Again, you may float tables to a column's top or bottom, and set
% wide tables across both columns. Place two-column tables at the
% top or bottom of the page.

% \subsection{Theorems and such}
% The preferred way is to number definitions, propositions, lemmas, etc. consecutively, within sections, as shown below.
% \begin{definition}
% \label{def:inj}
% A function $f:X \to Y$ is injective if for any $x,y\in X$ different, $f(x)\ne f(y)$.
% \end{definition}
% Using \cref{def:inj} we immediate get the following result:
% \begin{proposition}
% If $f$ is injective mapping a set $X$ to another set $Y$, 
% the cardinality of $Y$ is at least as large as that of $X$
% \end{proposition}
% \begin{proof} 
% Left as an exercise to the reader. 
% \end{proof}
% \cref{lem:usefullemma} stated next will prove to be useful.
% \begin{lemma}
% \label{lem:usefullemma}
% For any $f:X \to Y$ and $g:Y\to Z$ injective functions, $f \circ g$ is injective.
% \end{lemma}
% \begin{theorem}
% \label{thm:bigtheorem}
% If $f:X\to Y$ is bijective, the cardinality of $X$ and $Y$ are the same.
% \end{theorem}
% An easy corollary of \cref{thm:bigtheorem} is the following:
% \begin{corollary}
% If $f:X\to Y$ is bijective, 
% the cardinality of $X$ is at least as large as that of $Y$.
% \end{corollary}
% \begin{assumption}
% The set $X$ is finite.
% \label{ass:xfinite}
% \end{assumption}
% \begin{remark}
% According to some, it is only the finite case (cf. \cref{ass:xfinite}) that is interesting.
% \end{remark}
%restatable

% \subsection{Citations and References}

% Please use APA reference format regardless of your formatter
% or word processor. If you rely on the \LaTeX\/ bibliographic
% facility, use \texttt{natbib.sty} and \texttt{icml2022.bst}
% included in the style-file package to obtain this format.

% Citations within the text should include the authors' last names and
% year. If the authors' names are included in the sentence, place only
% the year in parentheses, for example when referencing Arthur Samuel's
% pioneering work \yrcite{Samuel59}. Otherwise place the entire
% reference in parentheses with the authors and year separated by a
% comma \cite{Samuel59}. List multiple references separated by
% semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
% construct only for citations with three or more authors or after
% listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

% Authors should cite their own work in the third person
% in the initial version of their paper submitted for blind review.
% Please refer to \cref{author info} for detailed instructions on how to
% cite your own papers.

% Use an unnumbered first-level section heading for the references, and use a
% hanging indent style, with the first line of the reference flush against the
% left margin and subsequent lines indented by 10 points. The references at the
% end of this document give examples for journal articles \cite{Samuel59},
% conference publications \cite{langley00}, book chapters \cite{Newell81}, books
% \cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
% \cite{mitchell80}, and dissertations \cite{kearns89}.

% Alphabetize references by the surnames of the first authors, with
% single author entries preceding multiple author entries. Order
% references for the same authors by year of publication, with the
% earliest first. Make sure that each reference includes all relevant
% information (e.g., page numbers).

% Please put some effort into making references complete, presentable, and
% consistent, e.g. use the actual current name of authors.
% If using bibtex, please protect capital letters of names and
% abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
% in your .bib file.

% \section*{Accessibility}
% Authors are kindly asked to make their submissions as accessible as possible for everyone including people with disabilities and sensory or neurological differences.
% Tips of how to achieve this and what to pay attention to will be provided on the conference website \url{http://icml.cc/}.

% \section*{Software and Data}

% If a paper is accepted, we strongly encourage the publication of software and data with the
% camera-ready version of the paper whenever appropriate. This can be
% done by including a URL in the camera-ready copy. However, \textbf{do not}
% include URLs that reveal your institution or identity in your
% submission for review. Instead, provide an anonymous URL or upload
% the material as ``Supplementary Material'' into the CMT reviewing
% system. Note that reviewers are not required to look at this material
% when writing their review.

% % Acknowledgements should only appear in the accepted version.
% \section*{Acknowledgements}

% \textbf{Do not} include acknowledgements in the initial version of
% the paper submitted for blind review.

% If a paper is accepted, the final camera-ready version can (and
% probably should) include acknowledgements. In this case, please
% place such acknowledgements in an unnumbered section at the
% end of the paper. Typically, this will include thanks to reviewers
% who gave useful comments, to colleagues who contributed to the ideas,
% and to funding agencies and corporate sponsors that provided financial
% support.

% % In the unusual situation where you want a paper to appear in the
% % references without citing it in the main text, use \nocite
% % \nocite{langley00}

% % \bibliography{example_paper}
% % \bibliographystyle{icml2022}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % APPENDIX
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
% \appendix
% \onecolumn
% \section{You \emph{can} have an appendix here.}

% You can have as much text here as you want. The main body must be at most $8$ pages long.
% For the final version, one more page can be added.
% If you want, you can use an appendix like this one, even using the one-column format.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% % This document was modified from the file originally made available by
% % Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% % by Iain Murray in 2018, and modified by Alexandre Bouchard in
% % 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022. 
% % Previous contributors include Dan Roy, Lise Getoor and Tobias
% % Scheffer, which was slightly modified from the 2010 version by
% % Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% % 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% % slightly modified from Prasad Tadepalli's 2007 version which is a
% % lightly changed version of the previous year's version by Andrew
% % Moore, which was in turn edited from those of Kristian Kersting and
% % Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2022-08-23 ftvi/rlmc.tex
Initial Overleaf Import

\documentclass{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage[bold]{hhtensor}

\usepackage{natbib}
\bibliographystyle{apalike}

\usepackage{hyperref}
\usepackage{cleveref}

\usepackage{todonotes}

\input{macros}

\title{Riemann Manifold Langevin Monte Carlo}

\begin{document}
\maketitle

XXX.  MWM TEST PUSH.

A number of samplers arise from discretizations of a stochastic differential equation (SDE) of the form
\[
    dX_t = \mu(X_t, t) dt + \sigma(X_t, t) d\eta_t
\]
where $\mu$ is a drift term and $\eta_t$ is a noise process.

To sample a target probability density $\pi(X)$, one can set $\mu(X_t) =
    \nabla \log \pi(\theta)$, $\sigma(X_t, t) \equiv 1$, and take $\eta_t = W_t$
a Wiener process. This corresponds to overdamped Langevin dynamics with potential $\log \pi(X)$,
which has $pi(X)$ as stationary solution. First-order Euler discretization of the dynamics
at scale $\eps$ yields
\[
    X_{t+1} \sim q(\cdot\mid X_t) = \cN\left(
    X_t + \frac{\eps^2}{2} \nabla \log \pi(X_t),
    \eps^2 \mI_k
    \right)
\]
When used in series with a Metropolis-Hastings accept/reject to correct for discretization error,
the result is known as the Metropolis Adjusted Langevin Algorithm (MALA, \cite{robert1996exponential}).
To accommodate  different variances between dimensions, \cite{roberts2002langevin} proposes using a
preconditioning ``mass'' matrix
\[
    q(\cdot\mid X_t) = \cN\left(
    X_t + \frac{\eps^2}{2} \nabla \log \pi(X_t),
    \eps^2 \mM
    \right)
\]
To choose $M$, \cite{roberts2002langevin} propose Riemannian-manifold
Langevin Monte Carlo (RLMC) via a Langevin diffusion on a Riemannian manifold
with metric tensor $\mM$ the of Fisher information matrix.
There are a number of choices:
\begin{description}
    \item[Expected Fisher Information] $\cov \log \nabla^2 p(\vy \mid \vx) = -E_{\vy \mid \vx} \nabla^2 \log p(\vy \mid \vx)$, which has the disadvantage that integration over the posterior is required
    \item[Empirical Fisher Information] $\widehat{\cov} \nabla^2 \log p(\vy \mid \vx)$, which only uses observed data
    \item[Observed Fisher Information] $\nabla^2 \log p(\vy \mid \vx) \mid_{\vx = \vx^{ML}}$, which requires only a single Hessian evaluation, is independent of the current particle position $\vx$, and is ensured to be positive semi-definite (hence PD after regularization).
\end{description}
Additionally, \cite{roberts2002langevin} suggest capturing prior informativeness in the metric tensor
by considering the Fisher information of the joint rather than the posterior, i.e.
\[
    -E_{\vy \mid \vx} \nabla^2 \log p(\vy, \vx)
    = \underbrace{-E_{\vy \mid \vx} \nabla^2 \log p(\vy \mid \vx)}_{\text{expected Fisher information}} - \underbrace{\nabla^2 \log p(\vx)}_{\text{Hessian of prior}}
\]

Newtonian Monte Carlo (NMC,\cite{arora2020newtonian}) is the default
inference method in a recently released open-source probabilistic programming
language (PPL) called \texttt{beanmachine} \citep{tehranibean}. Given
a 2-parameter proposal density $f(\cdot; \alpha, \beta)$, NMC performs a second-order approximation
by solving for $\alpha_t$ and $\beta_t$ in
\begin{alignat*}{1}
    \nabla \log \pi(X_t)   & = \nabla \log f(X_t; \alpha, \beta)   \\
    \nabla^2 \log \pi(X_t) & = \nabla^2 \log f(X_t; \alpha, \beta)
\end{alignat*}
and then setting $q(\cdot \mid X_t) = f(\cdot; \alpha_t, \beta_t)$. In the case of a Gaussian PDF $f$
with mean $\alpha$ and variance $\beta$, this becomes
\[
    q(\cdot \mid X_t)
    = \cN(X_t - (\nabla^2 \log \pi(X_t))^{-1} \nabla \log \pi(X_t), -(\nabla^2 \log \pi(X_t))^{-1})
\]
To relate this to RLMC, take $\pi(x) = p(x, y) \propto p(x \mid y)$ a
posterior distribution given (fixed) i.i.d. observations $y$. Then
\begin{alignat*}{1}
    - \nabla^2 \log \pi(x)
     & = - \nabla^2 \log p(x, y)                 \\
     & = - \sum_i^n \nabla^2 \log p(x, y_i)      \\
     & = -n E_{\hat{F}_n} \nabla^2 \log p(x, y)  \\
     & = -n \widehat{\cov} \nabla^2 \log p(x, y)
\end{alignat*}
from which we see that (after rescaling by $n$) NMC corresponds RLMC using the empirical
rather than expected Fisher information.

\todo{Heavy tailed $f$s? Can we show this discretizes a Levy-Flight vs Brownian Motion diffusion?}

There exists a major discrepancy between NMC as formualted and RLMC. From \cite{roberts2002langevin},
a diffusion which has invariant measure $\pi(X)$ on $\RR^d$ is given by the SDE
\begin{alignat}{1}
    dX_t & = \frac{1}{2} \mM(X_t)^{-1} (\nabla \log \pi(X)) dt + d\tilde{\vb}(t)
\end{alignat}
with the Brownian motion on the Riemannian manifold has as $i$th component
\begin{alignat}{1}
    d\tilde{\vb}_i(t)
     & = \underbrace{\lvert \mM(X_t) \rvert^{-1/2} \sum_{j=1}^d \frac{\pd}{\pd X_j} ( \mM(X_t)_{ij} \lvert \mM(X_t) \rvert^{1/2} )}_{A(X_t)} dt
    + (\mM(X_t)^{-1/2} d\vb(t))_i
\end{alignat}
where $\vb$ is standard Brownian motion.
Note that if $A(X_t) = 0$ (e.g. in the case $\mM$ is the observed Fisher information, which is constant),
then taking $\mM(X_t) = \nabla^2 \log \pi(X_t)$ and a first-order Euler discretization
at scale $1$ recovers NMC.

\begin{itemize}
    \item What effect does the missing drift term $A(X_t)$ have? Can we increase acceptance rates by including it?
    \item Is a scale of $1$ appropriate? Given that NMC uses $n$ times the empirical Fisher information, is there a ``better'' scaling involving $n$?
\end{itemize}

A diffusion which NMC does indeed represent the Euler discretization of is given by
\begin{alignat}{1}
    dX_t =\mM(X_t)^{-1} (\nabla \log \pi(X)) dt + \mM(X_t)^{-1/2} db_t
\end{alignat}

\begin{itemize}
    \item What is an (the?) invariant distribution of this distribution?
    \item Is there a bound on a divergence
          measure between this distribution and the target $\pi(X)$?
    \item Since this coincides with the previous diffusion in the case of
          $\mM$ the observed Fisher information, how does the performance of a MCMC
          sampling algorithm using that compare?
\end{itemize}

\bibliography{refs}

\end{document}

### 2022-08-23 hta/body.tex
Initial Overleaf Import

\maketitle

\begin{abstract}
	Despite the successes of probabilistic models based on passing noise through
	neural networks, recent work has identified that such methods often fail to
	capture tail behavior accurately unless the tails of the base distribution
	are appropriately calibrated. To overcome this deficiency, we propose a
	systematic approach for analyzing the tails of random variables during the
	static analysis (before drawing samples) pass of a probabilistic programming
	language (PPL) compiler. To characterize how the tails change under
	algebraic operations, we develop an algebra acting on a three-parameter
	family of tail asymptotics based on the generalized Gamma distribution. Our
	algebraic operations are closed under addition and multiplication, capable
	of distinguishing sub-Gaussians with differing scales, and handle ratios
	sufficiently well to reproduce the tails of most important statistical
	distributions directly from their definitions. Our experiments confirm that
	inference algorithms leveraging generalized Gamma algebra metadata attain
	superior performance across a number of density modeling and variational
	inference tasks.
\end{abstract}

\section{Introduction}

To facilitate efficient probabilistic modelling and inference, modern probabilistic programming languages (PPLs) draw upon recent developments in functional programming \citep{tolpin2016design},
programming languages \citep{bernstein2019static}, 
and deep variational inference \citep{bingham2019pyro}.
Despite their broadening appeal, common pitfalls such as mismatched
distribution supports \citep{lee2019towards} and non-integrable expectations
\citep{wang2018variational,vehtari2015pareto,yao2018yes} remain uncomfortably
commonplace and challenging to debug.
Recent innovations aiming to improve PPLs have automated verification of
distribution constraints \citep{lee2019towards}, tamed noisy gradient estimates
\citep{eslami2016attend} and unruly density ratios
\citep{vehtari2015pareto,wang2018variational}, and approximated high-dimensional
distributions with non-trivial bulks \citep{papamakarios2021normalizing} and non-Gaussian tails \cite{jaini2020tails}.

Continuing this line of work, here we 
we consider how to statically analyze a probabilistic program in order to automate the inference of 
tail behavior for any
random variables present. At present, correct inference of tail behaviour for target distributions remains an outstanding issue \citep{yao2018yes,wang2018variational}, which causes challenges for downstream Monte Carlo tasks. For example, importance sampling estimators can exhibit infinite variance if the tail of the approximating density is lighter than the target. Most prominant black-box variational inference methods are incapable of changing their tail behaviour from an initial proposal distribution \citep{jaini2020tails,ftvi}. MCMC algorithms may also lose ergodicity when the tail of the target density falls outside of a particular family \citep{roberts1996exponential}. All of these issues could be avoided if the tail of the target is known before runtime.

To classify tail asymptotics and define calibration, we propose a three-parameter
family based on the generalized Gamma distribution (\cref{eq:GenGammaDensity})
which interpolates between established asymptotics on sub-Gaussian \citep{ledoux2001concentration}
and regularly varying \citep{mikosch} random variables.
Algebraic operations on random variables
can be lifted to computations on the tail parameters resulting in what we call the \emph{generalized Gamma algebra (GGA)}.
Through analyzing operations like $X + Y$, $X^2$, and $X / Y$ at the level of densities (e.g.\ additive convolution $p_X \oplus p_Y$), the tail parameters of a target density can be estimated from the parameters of any input distributions using \Cref{tab:gga_operations}.

Operationalizing the GGA, we propose \emph{tail inferential} static analysis analogous to traditional \emph{type inference} and provide a reference implementation using the \texttt{beanmachine graph} \citep{tehrani2020bean}
PPL compiler. 
GGA tail metadata can be used to diagnose and address tail-related problems in downstream tasks, such as
employing Riemannian-manifold methods \citep{girolami2011riemann} to sample heavy tails
or pre-emptively detect unbounded expectations.
Here, we consider density estimation and variational inference where we use the GGA-computed tail of the target density to calibrate our density approximation.
%
%use it to ensure tail calibration and study its impact in density estimation and variational inference.
% Given a target random variable $X$ in a probabilistic program (PP) with density
% $p(x)$, an approximation with density $q(x)$ has calibrated tails if $p(x) \sim c q(x)$ as
% $\lvert x \rvert \to \infty$. Miscalibrated tails lead to pathological behaviors such as
% importance ratios $w(X) = \frac{p(X)}{q(X)}$ with unbounded variances.
%
%While GGA tail metadata may be more broadly useful, our experiments
%
When composed with a learnable Lipschitz pushforward map (\Cref{ssec:repr_dist}),
the resulting combination is a flexible density approximator with provably calibrated tails.
%n a range of elementary statistical computations and
%found that GGA yielded greater gains on targets with heavier tails.

\begin{figure}
	\includegraphics[width=\textwidth]{figures/schematic.pdf}
	\caption{Our overall approach for density approximations with calibrated tails.
		A generative model expressed in a PPL (1) and analyzed using the GGA (2) to
		compute the tail parameters of the target. A representative distribution
		with calibrated tails is chosen for the initial approximation (3) and a learnable
		Lipschitz pushforward (see \Cref{lem:lipschitz}) is optimized (4) to correct the
		bulk approximation.}
\end{figure}

% \begin{wrapfigure}[18]{r}{0.35\textwidth}
% 	\centering
% 	\includegraphics[width=.3\textwidth]{figures/pgm_vi.png}
% 	\caption{Graphical model and PPL API for VI with IID data $x_i$.}
% 	% 	Digraph fig1 {
% 	% rankdir = BT; //order things from left to right

% 	% { rank="same"; z; z_guide; }

% 	% //define alpha and beta as existing
% 	% z [shape=circle];
% 	% z_guide [shape=circle];

% 	% z->x_i; //quite literally z points at w

% 	% subgraph cluster_N
% 	% {
% 	%     label = "i=1,...,N"
% 	%     x_i [shape=circle, style=filled]
% 	% }
% 	% x_i -> params;
% 	% params -> z_guide

% 	% }
% \end{wrapfigure}

\textbf{Contributions}

\begin{itemize}[leftmargin=*]
	\item The GGA is introduced, generalizing prior work on classifying tail asymptotics
	      while including both sub-Gaussian / sub-exponentials \citep{ledoux2001concentration} as well as power-law / Pareto-based tail indices \citep{clauset2009power}. Composing operations outlined in \cref{tab:gga_operations}, one can compute the
	      GGA tail class for downstream random variables of interest.
	\item The GGA is implemented in the static analysis phase of a PPL compiler.
	      This unlocks the ability to leverage GGA metadata in order to
	      better tailor the emitted inference algorithm.
	\item Finally, we propose and evaluate a density estimator which combines GGA tails with normalizing flows
	      in order to simultaneously achieve good bulk approximation as well as correct tails.
\end{itemize}

\section{The Generalized Gamma Algebra}\label{sec:gga}

Here we formulate an algebra of random variables that is closed under most standard elementary operations (addition, multiplication, powers) which forms the foundation for our static analysis.

\begin{definition}\label{def:gg_tail}
A random variable $X$ is said to have a \emph{generalized Gamma tail} if the Lebesgue density of $|X|$ satisfies
\begin{equation}
\label{eq:GenGammaTails}
p_{|X|}(x) \sim c x^\nu e^{-\sigma x^\rho}, \qquad \text{as } x \to \infty,
\end{equation}
for some $c > 0$, $\nu \in \mathbb{R}$, $\sigma > 0$ and $\rho \in \mathbb{R}$. Denote the set of all such random variables by $\mathcal{G}$.
\end{definition}
Consider the following equivalence relation on $\mathcal{G}$: $X \equiv Y$ if and only if $0 < p_{|X|}(x) / p_{|Y|}(x) < +\infty$ for all sufficiently large $x$. The resulting equivalence classes can be represented by their corresponding parameters $\nu, \sigma, \rho$, and hence, we denote the class of random variables $X$ satisfying \cref{eq:GenGammaTails} by $(\nu,\sigma,\rho)$. In the special case where $\rho = 0$, for a fixed $\nu < -1$, each class $(\nu,\sigma,0)$ for $\sigma > 0$ is equivalent, and is denoted by $\mathcal{R}_{|\nu|}$, representing \emph{regularly varying} tails. Our algebra operates on these equivalence classes of $\mathcal{G}$, characterizing the change in tail behaviour under various operations.

\begin{table}[t]
    \centering
    \bgroup
\def\arraystretch{2.5}
\tolerance=1
\emergencystretch=\maxdimen
\hyphenpenalty=10000
\hbadness=10000
\begin{tabular}{|M{0.15\linewidth} | p{0.75\linewidth}|}
\hline
\centering \multirow{2}{*}{\textbf{Ordering}} & 
$\max\{(\nu_1,\sigma_1,\rho_1),(\nu_2,\sigma_2,\rho_2)\}$ \\ & $\qquad \equiv \begin{cases}
(\nu_1,\sigma_1,\rho_1) & \text{ if } \limsup_{x\to\infty} \frac{x^\nu_1 e^{-\sigma_1 x^{\rho_1}}}{x^\nu_2 e^{-\sigma_2 x^{\rho_2}}} < +\infty \\
(\nu_2,\sigma_2,\rho_2) & \text{ otherwise.}
\end{cases}$ \\ \hline
\centering \multirow{2}{*}{\textbf{Addition}} &  $(\nu_{1},\sigma_{1},\rho_{1})\oplus(\nu_{2},\sigma_{2},\rho_{2})$ \\
 & $\equiv \begin{cases}
\max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\} & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1\\
\left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right) & \text{ if }\rho_{1}=\rho_{2}=1\\
(\nu_{1}+\nu_{2}+\frac{2-\rho}{2},(\sigma_{1}^{-\frac{1}{\rho-1}}+\sigma_{2}^{-\frac{1}{\rho-1}})^{1-\rho},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}>1.
\end{cases}$ 
\\ \hline
\centering \textbf{Powers} & $(\nu,\sigma,\rho)^\beta \equiv (\frac{\nu+1}{\beta} - 1,\sigma,\frac{\rho}{\beta})$ for $\beta > 0$ \\\hline
\centering \textbf{Reciprocal*} & $(\nu,\sigma,\rho)^{-1} \equiv \begin{cases} (-\nu-2,\sigma,-\rho) & \text{ if } (\nu + 1)/\rho > 0 \text{ and } \rho \neq 0 \\
\mathcal{R}_2 & \text{ otherwise}
\end{cases}$ \\ \hline
\centering \textbf{Scalar \linebreak Multiplication} & $c(\nu,\sigma,\rho) \equiv (\nu,\sigma/|c|^\rho,\rho)$\\\hline
\multirow{3}{*}{\centering \textbf{Multiplication}} & $(\nu_{1},\sigma_{1},\rho_{1})\otimes(\nu_{2},\sigma_{2},\rho_{2})$ \\
&$\qquad \equiv\begin{cases}
\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0\\
\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}>0\\
\mathcal{R}_{|\nu_1|} & \mbox{ if }\rho_{1}\leq0,\rho_{2}>0 \\
\mathcal{R}_{\min\{|\nu_1|,|\nu_2|\}} & \mbox{ if }\rho_{1}=0,\rho_{2}=0
\end{cases}$ \\
&where $\mu=\frac{1}{|\rho_{1}|}+\frac{1}{|\rho_{2}|}=\frac{|\rho_{1}|+|\rho_{2}|}{|\rho_{1}\rho_{2}|}$, $\sigma=\mu(\sigma_{1}|\rho_{1}|)^{\frac{1}{\mu|\rho_{1}|}}(\sigma_{2}|\rho_{2}|)^{\frac{1}{\mu|\rho_{2}|}}$.\\ \hline
\centering \textbf{Product of Densities} & $(\nu_1,\sigma_1,\rho_1)\&(\nu_2,\sigma_2,\rho_2) 
\equiv \begin{cases}
(\nu_{1}+\nu_{2},\sigma_{1},\rho_{1}) & \text{ if }\rho_{1}<\rho_{2}\\
(\nu_{1}+\nu_{2},\sigma_{1}+\sigma_{2},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}\\
(\nu_{1}+\nu_{2},\sigma_{2},\rho_{2}) & \text{ otherwise.}
\end{cases}$ \\ \hline
\centering \textbf{Functions ($L$-Lipschitz)} & 
$f(X_1,\dots,X_n) \equiv L \max\{X_1,\dots,X_n\}$ \\ \hline
    \end{tabular}
    \vspace{.25cm}
    \caption{\label{tab:gga_operations}Operations on random variables (e.g.$X_1 + X_2$) are 
    viewed as actions on density functions
    (e.g. convolution $(\nu_1, \sigma_1, \rho_1)\oplus (\nu_2, \sigma_2, \rho_2)$) and the tail parameters of the result are analyzed and reported.}
\egroup
\end{table}

The form of \cref{eq:GenGammaTails} and the name of the algebra is derived from the generalized Gamma distribution. 
\begin{definition}
Let $\nu \in \mathbb{R}$, $\sigma > 0$, and $\rho \in \mathbb{R} \backslash \{0\}$ be such that $(\nu+1)/ \rho > 0$.
A non-negative random variable $X$ is \emph{generalized Gamma distributed} with parameters $\nu,\sigma,\rho$ if it has Lebesgue density
\begin{equation}
\label{eq:GenGammaDensity}
p_{\nu,\sigma,\rho}(x) = c_{\nu,\sigma,\rho} x^\nu e^{-\sigma x^\rho},\qquad x > 0,
\end{equation}
where $c_{\nu,\sigma,\rho} = \rho \sigma^{(\nu+1)/\rho} / \Gamma((\nu+1)/\rho)$ is the normalizing constant. %The generalized Gamma density incorporates many other well-known densities, including that of the Gamma distribution ($\rho = 1$), and the Frechet/Weibull distribution ($\nu = \rho - 1$). 
\end{definition}
The importance of the generalized Gamma form arises due to a combination of two factors:
\begin{enumerate}[label={(\roman*)},leftmargin=*]
    \item The majority of interesting continuous univariate distributions with infinite support satisfy \cref{eq:GenGammaTails}, including
    Gaussians ($\nu=0$, $\rho=2$),
    gamma/exponential/chi-squared ($\nu > -1$, $\rho=1$), Weibull/Frechet ($\rho = \nu + 1$), and
    Student $T$/Cauchy/Pareto ($\mathcal{R}_\nu$).
    However, some notable exceptions include the log-normal distributions. 
    \item The set $\mathcal{G}$ is known to be closed under additive convolution, positive powers, and Lipschitz functions --- we will show it is closed under multiplicative convolution as well. This covers the majority of elementary operations on independent random variables, with reciprocals, exponentials and logarithms the only exceptions. However, we will introduce a few ``tricks'' to handle these cases as well. 
\end{enumerate}
The full list of operations in GGA is compiled in \cref{tab:gga_operations}. All operations in the GGA can be proven to exhibit identical behaviour with their corresponding operations on random variables, with the sole exception of reciprocals (marked by asterisk), where additional assumptions are required. For further details, refer to the Supplementary Material.

%To incorporate tails which lie outside of $\mathcal{G}$, we let $\mathcal{R}_1$ incorporate \emph{super-heavy tails}, which denote random variables with tails heavier than any random variable in $\mathcal{G}$. All operations remain consistent with this notation. Likewise, we let $\mathcal{L}$ denote \emph{super-light tails}, which are treated in our algebra as a class where $\rho = +\infty$ (effectively constants).

% \paragraph{Exponential and Logarithm.} Tails of the generalized Gamma form are not closed under exponentiation or logarithms. Indeed, if both $X$ and $\exp X$ have generalized Gamma tails, then $X$ is exponentially distributed (and $\exp X$ has power law tails). As a workaround, we can consider an upper bound on the tail by projecting onto the nearest possible exponentially distributed / power law tail. If $\rho > 1$, then a change of variables shows the density of $\exp X$ satisfies
% \[
% p_{\exp X}(x) \sim \frac{c}{x}(\log x)^{\nu}\exp\left(-\sigma(\log x)^{\rho}\right)\leq\frac{\tilde{c}}{x}\exp\left(-\sigma(\log x)\right)=cx^{-\sigma-1},\,\mbox{ as }x \to \infty.
% \]
% The inverse of this operation sends $\mathcal{R}_{\sigma+1}$ to $(0,\sigma,1)$. With this in mind, we define the exponential and logarithmic operations according to the following: $\exp (\nu, \sigma, \rho) \equiv \mathcal{R}_{\sigma+1}$ if $\rho \geq 1$, otherwise $\mathcal{R}_1$; $\log (\nu, \sigma, \rho) \equiv (0, |\nu|-1, 1)$ if $\nu < -1$ and $\rho \leq 0$, otherwise $\mathcal{L}$.

\section{Implementation}\label{sec:impl}

\subsection{Compile-time static analysis}

To illustrate an implementation of GGA for static analysis, we sketch the operation of the PPL compiler at a high-level
and defer to the supplementary code for details.
A probabilistic program is first inspected using Python's built-in \texttt{ast} module
and transformed to static single assignment (SSA) form \citep{rosen1988global}.
Next, standard compiler optimizations (e.g. dead code elimination, constant propagation)
are applied and an execution of the optimized program is traced \citep{wingate2011lightweight,bingham2019pyro}
and accumulated in a directed acyclic graph representation. A breadth-first type checking pass, as seen in Algorithm~\ref{alg:bfs_typecheck}, completes in linear time, and GGA results may be applied to implement \texttt{computeGGA()} using the following steps:
\begin{itemize}[leftmargin=*]
	\item If a node has no parents, then it is an atomic distribution and its tail parameters are known (\Cref{tab:dist_list})
 	\item Otherwise, the node is an operation taking its potentially stochastic inputs (parents) to its output. Consult \Cref{tab:gga_operations} for the output GGA tails.
\end{itemize}

%% This declares a command \Comment
%% The argument will be surrounded by /* ... */
\SetKwComment{Comment}{/* }{ */}

\begin{algorithm}
	\caption{Pseudocode for a GGA tails static analysis pass}\label{alg:bfs_typecheck}
	\KwData{Abstract syntax tree for a PPL program}
	\KwResult{GGA parameter estimates for all random variables}
	frontier $\gets$ [rv : Parents(rv) = $\emptyset$]\;
	GGAs $\gets \{\}$\;
	\While{\text{frontier} $\neq \emptyset$}{
		next $\gets$ frontier.popLeft()\;
		GGAs[next] $\gets$ computeGGA(next.op, next.parent)\;
		frontier $\gets$ frontier + next.children()\;
	}
	\Return{GGAs}
\end{algorithm}

\subsection{Representative distributions}\label{ssec:repr_dist}

For each $(\nu,\sigma,\rho)$ we make a carefully defined choice of $p$ on $\mathbb{R}$ such that if $X \sim p$, then $X \equiv (\nu,\sigma,\rho)$. This way, any random variable $f(X)$, where $f$ is $1$-Lipschitz, will exhibit the correct tail, and so approximations of this form may be used for variational inference or density estimation. Let $X \equiv (\nu,\sigma,\rho)$ and $0 < \epsilon \ll 1$ denote a small parameter such that tails $e^{-x^\epsilon}$ are deemed to be ``very heavy'' (we chose $\epsilon = 0.1$). %We also let $X\equiv(\nu,\sigma,\rho)$ and $\zeta$ denote an independent Rademacher random variable with $\mathbb{P}(\zeta=1)=\mathbb{P}(\zeta=-1)=\frac12$.

\begin{itemize}
    \item[($\rho \leq 0$)] If $\rho \leq -1$, then $p_X(x) \sim c x^{-|\nu|}$. A prominent distribution on $\mathbb{R}$ with power law tails is the \emph{Student $t$ distribution}, in this case, with $|\nu|-1$ degrees of freedom if $\nu < -1$ (generate $X \sim t_{|\nu|-1})$.
    \item[($\rho > \epsilon$)] For moderately sized $\rho > 0$, we consider a symmetrized variant of the generalized Gamma density (\Cref{eq:GenGammaDensity}). 
    \item[($\rho \leq \epsilon$)] If $X \equiv (\nu,\sigma,\rho)$ where $\rho$ is small, then $X$ will exhibit much heavier tails, and the generalized Gamma distribution in Case 1 will become challenging to sample from. In these cases, we expect that the tail of $X$ should be well represented by a power law. 
    %Letting $\alpha = \sigma \rho - \nu - 1$, using an approximation to the logarithm, $p_{|X}(x) \approx c x^{-\alpha - 1}$ (see the Supplementary Material for more details). Therefore, we represent tails of this form by the Student $t$ distribution with $\alpha$ degrees  
    The generalized Gamma density (\Cref{eq:GenGammaDensity}) satisfies $\mathbb{E}X^r = \sigma^{-r/\rho} \Gamma(\frac{\nu+1+r}{\rho})/\Gamma(\frac{\nu+1}{\rho})$ for $r > 0$. Let $\alpha > 0$ be such that $\mathbb{E}X^\alpha = 2$. By Markov's inequality, the tail of $X$ satisfies $\mathbb{P}(X>x)\leq 2 x^{-\alpha}$. Therefore, we can represent tails of this form by the Student $t$ distribution with $\alpha+1$ degrees of freedom (generate $X \sim t_{\alpha}$).
\end{itemize}

\subsection{Bulk correction by Lipschitz mapping}

While a representative distribution will exhibit the desired tails,
the target distribution's bulk may be very different from a generalized Gamma and result in poor distributional approximation. To address this, we propose splicing together the tails from a generalized Gamma with a flexible density approximation for the bulk.
While many combinations are possible, in this work we rely on \Cref{lem:lipschitz} and post-compose neural
spline flows \citep{durkan2019neural} (which are identity functions outside of a bounded interval) after
properly initialized generalized Gamma distributions.
Optimizing the parameters of the flow results in good bulk approximation while simultaneously
preserving the tail correctness guarantees attained by the GGA.

\begin{example}
	Let $A \in \mathbb{R}^{k \times k}$, $x,y\in\mathbb{R}^k$, with $x_i,y_i,A_{ij} \overset{\text{iid}}{\sim} \mathcal{N}(-1,1)$.
	The distribution of $x^\top A y = \sum_{i,j} x_i A_{ij} y_j$ is convolution of normal-powers \citep{gupta2008analyzing} and has no convenient
	closed form expression.
	Using GGA's closure theorems (\cref{tab:gga_operations}), one can compute
	its tail parameters to be $(\frac{k}{2}-1,\frac{3}{2}, \frac{2}{3})$.

	\begin{figure}[h]
		\centering
		\input{experiments/normal-power-sum.pgf}
		\input{experiments/lip-nf.pgf}
		\caption{(Top) $5000$ samples of $\lvert x^\top A y \rvert$ vs the
		calibrated GGA density $q(x)$. While calibrated tails are provably guarantees,
		the target distribution's bulk differs from the assumed generalized Gamma representative distribution (\cref{ssec:repr_dist}) for all $k$.
		To fix the bulk approximation, a normalizing flow is composed with
		the GGA representative to form $\text{flow}(q)(x)$. The bulk approximation
		is improved (bottom) while the tails continue to exhibit the same
		behavior (bottom right).
		}
		\label{fig:power_normal}
	\end{figure}

	The GGA representative is a gamma distribution with the correct tails, but there is non-negligible error in the bulk where $x$ is small. To address this, a learnable bijector can be
	optimized as in \Cref{fig:power_normal} bottom left to correct the bulk approximation. Guaranteed by \Cref{lem:lipschitz} and
	visualized in \Cref{fig:power_normal} bottom right, the tails of the overall composition remain calibrated.
\end{example}

\section{Experiments}\label{sec:experiments}

In this section we demonstrate that GGA-based density estimation yields improvements across a variety of metrics.
We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushfowards of Normal distributions. To understand the individual
effect of using a GGA base distribution over standard normals versus more expressive pushforward maps \citep{durkan2019neural}, we also report ablation results where
normalizing flows are replaced by affine transforms as originally
proposed in \citep{kucukelbir2017automatic}. All experiments are repeated for 50 trials, trained to convergence using the Adam optimizer with manually tuned learning rate,
and conducted on i7-8700K CPU and GTX 1080 GPU hardware.

All target distributions in this section are expressed as generative PPL programs:
Cauchy using a reciprocal normal, Chi2 using a sum of squared normals, IG (Inverse Gamma)
using a reciprocal exponential, Normal using a sum of normals, and StudentT using a
normal and Cauchy ratio. Doing so tasks the static analyzer to infer the target's tails and makes the analysis non-trivial.
See supplementary for full details.

Our results in the following tables share a consistent narrative where a GGA base
distribution rarely hurts and can significantly help with heavy tailed targets.
Except for when targets are truly light tailed ($\alpha=\infty$ in Chi2 and Normal),
GGA-based approximations are the only ones to reproduce appropriate GPD tail index $\hat\alpha$ in density estimation and achieve a passing Pareto $\hat{k}$ diagnostic
\citep{yao2018yes} below $0.2$ in variational inference. When viewed through traditional
evaluation metrics such as negative cross-entropy $H(p,q) = E_p \log p(X)$,
ELBO $E_q \log \frac{q(X)}{p(X)}$,
and importance-weighted autoencoder bound \citep{burda2015importance} $E_q \log \sum_i^{1000} \frac{p(X)}{q(X)}$, GGA-based approximations remain favorable on almost all heavy-tailed
targets and have negligible difference for light tailed targets. Less surprising is the result that adding a flow improved approximation metrics, as we expect the additional representation flexibility to be beneficial.

%\feynman{explain the targets, generative and marginalized, reference appendix}

\textbf{Density Estimation} We minimize a Monte-Carlo estimate of the cross entropy $H(p,q) = -E_p[\log q(X)] \approx -\frac{1}{N} \sum_{i=1}^N \log q(x_i),\quad x_i \sim p$.
The results are shown in \Cref{tab:de} along with power-law tail index estimates \citep{clauset2009power} $\hat\alpha$.
Overall, we see that GGA performs better (lower NLL, $\hat\alpha$ closer to target) when the target
has heavier tails (lower $\hat\alpha$ target/theory) and that the difference is smaller but still non-negligible
for distributions such as Chi1 which possess tails heavier than Gaussian.

\begin{table}[h]
	\centering
	\caption{Density estimation metrics attained (mean, standard deviation in parenthesis)
	on targets of varying tail index (smaller $\alpha =$ heavier tails).
	Higher negative cross entropy $-H(p,q) = E_p \log q(X)$ implies a better overall approximation (row maxes bolded)
	while close agreement between the target Pareto tail index $\alpha$ \citep{clauset2009power}
	and its estimate $\hat\alpha$ in $q(x)$ suggest calibrated tails (closest in row bolded).
	}
	\label{tab:de}
	\input{experiments/de_table.tex}
\end{table}

\textbf{Variational Inference} The optimization objective is the ELBO $E_q \log \frac{p(X)}{q(X)} \approx \frac{1}{N} \sum_{i=1}^N \log \frac{p(x_i)}{q(x_i)},\quad x_i \sim q$. Here, the density $p$ must also be evaluated so for simplicity
experiments in \cref{tab:vi} use closed-form marginalized densities for targets.
The overall trends also show that GGA yields consistent improvements as measured by both ELBO and importance-weighted
estimates of marginal likelihood and that the difference was greater when the tails of $p(z)$ were heavier.
The $\hat{k}$ diagnostics \citep{yao2018yes} corroborate our findings that variational inference succeeds
($\hat{k} < -1.2$) when a GGA with appropriately matched tails is used and fails ($\hat{k} > 1$) when
Gaussian tails are erroneously imposed.

\begin{table}[h]
	\centering
	\caption{Variational inference metrics (mean, standard deviation in parenthesis)
	on targets of varying tail index (smaller $\alpha =$ heavier tails). Both the IWAE bound $E_q \log \sum_i^K \frac{p(X_i)}{q(X_i)}$ and the ELBO ($K=1$) measure (a lower bound) on the marginal likelihood where larger is better (row maxes bolded). In \citet{yao2018yes},
	a Pareto $\hat{k}$ diagnostic $> 0.2$ is interpreted as potentially problematic so only values below are bolded.}
	\label{tab:vi}
	\small\input{experiments/vi_table.tex}
\end{table}

\subsection{SGD for least-squares linear regression}

\begin{figure}[ht]
\centering
\includegraphics[width=0.32\textwidth]{figures/Kesten_0.40_1.5.png}
\includegraphics[width=0.32\textwidth]{figures/Kesten_0.50_1.5.png}
\includegraphics[width=0.32\textwidth]{figures/Kesten_0.50_2.png}
\caption{\label{fig:SGD}Density of iterates of SGD vs. predicted tail behaviour}
\end{figure}

For inputs $X$ and labels $Y$ from a dataset $\mathcal{D}$, the least squares estimator for linear regression satisfies $\beta = \min_\beta \tfrac12 \mathbb{E}_{X,Y\sim\mathcal{D}}(Y - X\beta)^2$. To solve for this estimator, one can apply stochastic gradient descent (SGD) sampling over independent $X_k,Y_k\sim \mathcal{D}$ to obtain the sequence of iterations
\[
\beta_{k+1} = (I - \delta X_k X_k^\top) \beta_k + \delta Y_k X_k
\]
for a step size $\delta > 0$. For large $\delta$, the iterates $\beta_k$ typically exhibit heavy-tailed fluctuations; in this regard, this sequence of iterates has been used as a simple model for more general stochastic optimization dynamics \citep{gurbuzbalaban2021heavy,hodgkinson2021multiplicative}. In particular, generalization performance has been tied to the heaviness of the tails in the iterates \cite{simsekli2019tail}. Here we use our algebra to predict the tail behaviour in a simple one-dimensional setting where $X_k \sim \mathcal{N}(0,\sigma^2)$ and $Y_k \sim \mathcal{N}(0,1)$. From classical theory \citep{buraczewski2016stochastic}, it is known that $X_k$ converges in distribution to a power law with tail exponent $\alpha > 0$ satisfying $\mathbb{E}|1 - \delta X_k^2|^\alpha = 1$. In \cref{fig:SGD}, we plot the density of the representative obtained using our algebra after $10^4$ iterations against a kernel density estimate of the first $10^6$ iterates when $\sigma \in \{0.4,0.5\}$ and $\delta \in \{1.5,2.0\}$. In all cases, the density obtained from the algebra provides a surprisingly close fit.

\section{Related Work}

\textbf{Heavy tails and probabilistic machine learning}

For studying heavy tails, methods based on subexponential distributions
\citep{goldie1998subexponential} and generalized Pareto distributions (GPD) or
equivalently regularly varying distributions \citep{tajvidi2003confidence}
have recieved attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
Heavy tails can impact probabilistic machine learning methods in a number of ways.
The observation that density ratios $\frac{p(x)}{q(x)}$ tend to be heavy tailed
has resulted in new methods for smoothing importance sampling \citep{vehtari2015pareto},
adaptively modifying divergences \citep{wang2018variational}, and
diagnosing variational inference through the Pareto $\hat{k}$ diagnostic \citep{yao2018yes}.
These works are complementary to our paper and our reported results include $\hat{k}$
diagnostics for VI and $\hat\alpha$ tail index estimates based on GPD.

Our work considers heavy-tailed targets $p(x)$ which is the same setting as
\citet{jaini2020tails,ftvi}. Whereas those respective works
lump the tail parameter in as another variational parameter and may be more generally applicable, the GGA may be applied before samples are drawn and leads to perfectly calibrated tails when applicable.

\textbf{Probabilistic programming}

PPLs can be characterized by the primary use case optimized for, whether that's
Gibbs sampling over Bayes nets \citep{spiegelhalter1996bugs,de2017programming},
stochastic control flow \citep{goodman2012church,wingate2011lightweight},
deep stochastic variational inference \citep{tran2018simple,bingham2019pyro}, or
Hamiltonian Monte-Carlo \citep{carpenter2017stan,xu2020advancedhmc}. Our implementation target
\texttt{beanmachine} \citep{tehrani2020bean} is a declarative PPL selected
due to availability of a PPL compiler and support for static analysis plugins.
Similar to \citet{bingham2019pyro,siddharth2017learning}, it uses PyTorch \citep{paszke2019pytorch} for GPU tensors and automatic differentiation.
Synthesizing an approximating distribution during PPL compilation (\Cref{sec:impl}) is
also performed in the Stan language by \citet{kucukelbir2017automatic} and normalizing
flow extensions in \citet{webb2019aml}. We compare directly against these related density approximators in \Cref{sec:experiments}.

\textbf{Static analysis}

There is a long history of formal methods and probabilistic programming
\cite{kozen1979semantics,jones1989probabilistic}. While much of the research
\cite{claret2013bayesian}
is concerned with defining formal semantics
and establishing invariants \cite{wang2018pmaf}
See \cite{bernstein2019static} for a recent review.
Static analysis utilizes the abstract syntax tree (AST)
representation of a program in order to compute invariants
(e.g. the return type of a function, the number of classes implementing a trait)
without executing the underlying program.
As dynamic analysis in PPs is less reliable due to non-determinism, static analysis
methods for PPs become increasingly important.

Within PPLs, static analysis has traditionally been applied in the context of formalizing
semantics \citep{kozen1979semantics} and has been used to verify probabilistic
programs by ensuring termination, bounding random values values
\citep{sankaranarayanan2013static}. \cite{lee2019towards} proposes a static analyzer for the
Pyro PPL \citep{bingham2019pyro} to verify
distribution supports and avoid $-\texttt{Inf}$ log probabilities.

More relevant to our work are applications of static analysis to improve inference. \citet{nori2014r2} statically analyzes a probabilistic program and computes pre-images of observations in order to better adapt MCMC proposal distributions.
While we also perform static analysis over
abstract syntax tree (AST) representations of a probabilistic program,
applying GGA yields an upper bound on the tails of all random variables
so that calibrated tails can be imposed on distribution estimates.

\section{Conclusion}\label{sec:conclusion}

In this work, we have proposed a novel systematic approach for conducting tail inferential static analysis by implementing a three-parameter generalized Gamma algebra into a PPL compiler. Initial results are promising, showing that improved inference with simpler approximation families is possible when combined with tail metadata.
While already useful, the generalized Gamma algebra and its implementation currently has
some notable limitations:
\begin{itemize}[leftmargin=*]
\item Since the algebra assumes independence, handling of dependencies between defined random variables must be conducted externally. This will inevitably require interoperability with a symbolic package to decompose complex expressions into operations on independent random variables.
\item The GGA is formulated for univariate distributions only. Suitably defining multivariate tails is an open problem with interesting alternatives \citep{jaini2020tails,ftvi} all of which
could extend GGA to higher dimensions. 
\item Conditioning is arguably the most important feature of a PPL and what distinguishes it from a glorified simulator. Exact marginalization in general is NP-hard \citep{koller2009probabilistic}, so treatment of conditional distributions using symbolic manipulations is a significant open problem, with some basic developments \cite{shan2017exact,cho2019disintegration}. Since only the tails are required in our setup, it may be possible to construct a dual algebra for operations under conditioning; this is left for future work.
\item Compile-time static analysis only applicable to fixed model structure. While out of scope for our current work, open-universe models \citep{milch2010extending}
and PPLs to support them \citep{bingham2019pyro} are an important research direction.
\item The most significant omission to the algebra itself is classification of log-normal tails; while addition may be treated using \cite{gulisashvili2016tail} for example, multiplicative convolution with log-normal tails remains elusive. 
\item At present, reciprocals are approximated by assuming behaviour near zero. Reciprocals may be better treated by covering near-zero asymptotics separately. 
\end{itemize}

The GGA provides a necessary first step into the static analysis of tails in
a probabilistic program.
As the above limitations are improved in future work and GGA becomes more broadly applicable,
we are excited to see how improved tail modelling will improve downstream PPL applications
as well as
other researchers will utilize GGA metadata to develop
novel PPL applications.

\begin{ack}
	Use unnumbered first level headings for the acknowledgments. All acknowledgments
	go at the end of the paper before the list of references. Moreover, you are required to declare
	funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
	More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2021/PaperInformation/FundingDisclosure}.

	Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
\end{ack}

% \subsection{Limitations}

% \todo{Liam: Log-Normal insensitivity, edge case (see example)}

% The GGA is formulated for univariate distributions only.

{
\small
\bibliographystyle{unsrtnat}
\bibliography{refs}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Checklist}

% %%% BEGIN INSTRUCTIONS %%%
% The checklist follows the references.  Please
% read the checklist guidelines carefully for information on how to answer these
% questions.  For each question, change the default \answerTODO{} to \answerYes{},
% \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
% justification to your answer}, either by referencing the appropriate section of
% your paper or providing a brief inline description.  For example:
% \begin{itemize}
%   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
%   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
%   \item Did you include the license to the code and datasets? \answerNA{}
% \end{itemize}
% Please do not modify the questions and only use the provided macros for your
% answers.  Note that the Checklist section does not count towards the page
% limit.  In your paper, please delete this instructions block and only keep the
% Checklist section heading above along with the questions/answers below.
% %%% END INSTRUCTIONS %%%

\begin{enumerate}

\item For all authors...
\begin{enumerate}
  \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
    \answerYes{}
  \item Did you describe the limitations of your work?
    \answerYes{See \cref{sec:conclusion}}
  \item Did you discuss any potential negative societal impacts of your work?
    \answerYes{See \cref{sec:conclusion}}
  \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
    \answerYes{}
\end{enumerate}

\item If you are including theoretical results...
\begin{enumerate}
  \item Did you state the full set of assumptions of all theoretical results?
    \answerYes{Stated at start of \cref{sec:gga}}
	\item Did you include complete proofs of all theoretical results?
    \answerYes{Most are deferred to supplementary materials}
\end{enumerate}

\item If you ran experiments...
\begin{enumerate}
  \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
    \answerYes{See supplemental, a URL for code will be provided witn camera ready}
  \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
    \answerYes{}
	\item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
    \answerYes{Standard errors reported in parenthesis on \cref{tab:de} and \cref{tab:vi}}
	\item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
    \answerYes{See \cref{sec:experiments}}
\end{enumerate}

\item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
\begin{enumerate}
  \item If your work uses existing assets, did you cite the creators?
    \answerYes{See citations of \citep{tehrani2020bean}}
  \item Did you mention the license of the assets?
    \answerYes{Licence included in code}
  \item Did you include any new assets either in the supplemental material or as a URL?
    \answerYes{See supplemental material}
  \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
    \answerNA{}
  \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
    \answerNA{}
\end{enumerate}

\item If you used crowdsourcing or conducted research with human subjects...
\begin{enumerate}
  \item Did you include the full text of instructions given to participants and screenshots, if applicable?
    \answerNA{}
  \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
    \answerNA{}
  \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
    \answerNA{}
\end{enumerate}

\end{enumerate}

### 2022-08-23 hta/neurips_2022.tex
Initial Overleaf Import

--- 2022-08-23 hta/body.tex
@@
+\documentclass{article}
+\csname remappendix\endcsname
+
+% if you need to pass options to natbib, use, e.g.:
+%     \PassOptionsToPackage{numbers, compress}{natbib}
+% before loading neurips_2022
+
+% ready for submission
+\usepackage{neurips_2022}
+
+% to compile a preprint version, e.g., for submission to arXiv, add add the
+% [preprint] option:
+%     \usepackage[preprint]{neurips_2022}
+
+% to compile a camera-ready version, add the [final] option, e.g.:
+%     \usepackage[final]{neurips_2022}
+
+% to avoid loading the natbib package, add option nonatbib:
+%    \usepackage[nonatbib]{neurips_2022}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{xcolor}         % colors
+\usepackage{longtable}
+\usepackage{wrapfig}
+\usepackage{array}
+\usepackage{enumerate}
+\usepackage{enumitem}
+\usepackage{hhline}
+\usepackage{multirow}
+
+\usepackage{amsmath}
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}{Proposition}
+\newtheorem{lemma}[theorem]{Lemma}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}
+\newtheorem{example}{Example}
+\usepackage{todonotes}
+\title{Static Analysis of Tail Behaviour with a \\ Generalized Gamma Algebra}
+
+\newcommand{\feynman}[1]{\todo{(Feynman): #1}}
+\newcommand{\dd}{\mathrm{d}}
+\usepackage{tikz}
+\usetikzlibrary{bayesnet}
+\usepackage{cleveref}
+\usepackage{listings}
+\usepackage{multirow}
+\usepackage{siunitx}
+\usepackage{etoolbox}
+\robustify\bfseries
+\usepackage[ruled]{algorithm2e}
+\usepackage{float}
+
+\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
+\newcolumntype{M}[1]{>{\centering\arraybackslash}m{#1}}
+
+% The \author macro works with any number of authors. There are two commands
+% used to separate the names and addresses of multiple authors: \And and \AND.
+%
+% Using \And between authors leaves it to LaTeX to determine where to break the
+% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
+% authors names on the first line, and the last on the second line, try using
+% \AND instead of \And before the third author name.
+
+\author{%
+  David S.~Hippocampus\thanks{Use footnote for providing further information
+    about author (webpage, alternative address)---\emph{not} for acknowledging
+    funding agencies.} \\
+  Department of Computer Science\\
+  Cranberry-Lemon University\\
+  Pittsburgh, PA 15213 \\
+  \texttt{hippo@cs.cranberry-lemon.edu} \\
+  % examples of more authors
+  % \And
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+  % \AND
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+  % \And
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+  % \And
+  % Coauthor \\
+  % Affiliation \\
+  % Address \\
+  % \texttt{email} \\
+}
+
+\sisetup{detect-all = true}
+
+\begin{document}
+
 \maketitle
@@
 \end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\ifdefined\remappendix
+\newpage
+
+\appendix
+
+\section{Additional examples}\label{sec:addtl_eg}
+
+To further illustrate the GGA through example, in this section we work
+out explicit GGA computations using distributions
+from \cref{tab:dist_list} and operations in \cref{tab:gga_operations}
+and recover some common probability identities.
+
+\begin{example}[Chi-squared random variables]
+	Let $X_1,\dots,X_k$ be $k$ independent standard normal random variables.
+The variable $Z = \sum_{i=1}^k X_i^2$ is \emph{chi-squared distributed} with $k$ degrees of freedom.
+Using the generalized Gamma algebra, we can accurately determine the tail behaviour of this random variable directly from its construction.
+Recall that each $X_i \equiv (0,1/2,2)$, and by the power operation, $X_i^2 \equiv (-1/2,1/2,1)$.
+Applying the addition operation $k$ times reveals that $Z \equiv (k/2-1,1/2,1)$ and implies that the density of $Z$ is asymptotically $c x^{k/2-1} e^{-x / 2}$ as $x \to \infty$.
+In fact, the density of $Z$ is exactly $p_Z(x) = c_k x^{k/2-1} e^{-x/2}$ where $c_k = 2^{-k/2} / \Gamma(k/2)$.
+\end{example}
+
+\begin{example}[Products of random variables]
+	To demonstrate the efficacy of the multiplication operation in our algebra, we consider the product of two exponential, Gaussian, and reciprocal Gaussian random variables.
+In \cref{sec:proofs}, we manually prove the following.
+	\begin{lemma}
+	\label{lem:Products}
+		Let $X_1,X_2 \sim \mbox{Exp}(\lambda)$ and $Z_1,Z_2 \sim \mathcal{N}(0,1)$ be independent. The densities of $X_1 X_2$, $Z_1 Z_2$ and $Z = 1/Z_1 \cdot 1/Z_2$ satisfy as $x \to \infty$,
+		\[
+			p_{X_1 X_2}(x) \sim \frac{\lambda^{3/2}\sqrt{\pi}}{x^{1/4}} e^{-2\lambda \sqrt{x}},\quad p_{Z_1 Z_2}(x) \sim \frac{1}{\sqrt{2\pi x}} e^{-x},\quad
+			p_{Z}(x) \sim \frac{1}{\sqrt{2\pi}|z|^{3/2}}e^{-1/|z|}.
+		\]
+	\end{lemma}
+	With ease, our algebra correctly determines that $X_1 X_2 \equiv (-\frac14,2\lambda,\frac12)$, $Z_1 Z_2 \equiv (-\frac12, 1, 1)$ and $Z \equiv (-\frac32,1,-1)$.
+	% Recall that the Bessel $K$-function of zeroth order, denoted $K_0$, is given by the Fourier transform $K_0(x) = \frac12 \int_{-\infty}^{\infty} e^{ixt} (1+t^2)^{-1/2} \dd t$.
+This function behaves as $x \to \infty$ as $K_{0}(x)\sim\sqrt{\frac{\pi}{2x}}e^{-x}$.
+Let $X_1,X_2 \sim \mathrm{Exp}(\lambda)$ be two independent exponentially distributed variables.
+The density of the product $Z = X_1 X_2$ is given by
+	% \[
+	% p_Z(x) = 2\lambda^{2}K_{0}(2\lambda\sqrt{z})\sim\lambda^{3/2}\sqrt{\pi}z^{-1/4}e^{-2\lambda\sqrt{z}}.
+	% \]
+	% This tail behaviour can be obtained using our algebra as follows: 
+	% The density of
+	% \[
+	% p_Z(x) = \frac1\pi K_0(|x|) \sim \frac{1}{\sqrt{2\pi x}} e^{-x},\qquad \mbox{as }x \to \infty.
+	% \]
+\end{example}
+
+\begin{example}[Reciprocal distributions]
+	Perhaps the most significant challenge with a tail algebra is correctly identifying the tail behaviour of reciprocal distributions.
+Here, we test the efficacy of our formulation with known reciprocal distributions.
+	\begin{itemize}
+		\item \emph{Reciprocal normal:} $X \sim \mathcal{N}(0,1) \equiv (0,1/2,2)$, and $X^{-1} \equiv (-2,1/2,-2)$.
+		\item \emph{Inverse exponential:} $X \sim \text{Exp}(\lambda) \equiv (0,\lambda,1)$, and $X^{-1} \equiv (-2,\lambda,-1)$.
+		\item \emph{Inverse $t$-distribution:} $X \equiv \mathcal{R}_\nu$, and $X^{-1} \equiv \mathcal{R}_2$.
+		\item \emph{Inverse Cauchy:} $X \equiv \mathcal{R}_2$, it is known $X^{-1}$
+		      has the same distribution and our theory predicts $X^{-1} \equiv \mathcal{R}_2$.
+	\end{itemize}
+\end{example}
+
+\begin{example}[Cauchy distribution]
+	A simple special case of the Student $T$ distribution is the Cauchy distribution, which arises as the ratio of two standard normal random variables.
+For $X\sim\mathcal{N}(0,1)$, $X \equiv (0,1/2,2)$ and $X^{-1} \equiv (-2,1/2,-2)$.
+%The reciprocal of a standard normal random variable $X \equiv (0,1/2,2)$ satisfies $X^{-1} \equiv (-2,1/2,-2)$.
+	Hence, the multiplication operation correctly predicts that the ratio of two standard normal random variables is in $\mathcal{R}_2$.
+\end{example}
+
+\begin{example}[Student $T$ distribution]
+	Let $X$ be a standard normal random variable, and $V$ a chi-squared random variable with $\nu$ degrees of freedom.
+The random variable $T = X / \sqrt{V / \nu}$ is \emph{$t$-distributed} with $\nu$ degrees of freedom.
+Since $V \equiv (\nu/2-1,1/2,1)$, multiplying by the constant $1/\nu$ reveals $V / \nu \equiv (\nu/2-1,1/(2\nu),1)$.
+Applying the square root operation, $\sqrt{V / \nu} \equiv (\nu-1,1/(2\nu),2)$.
+To compute the division operation, we first take the reciprocal to find $(V/\nu)^{-1/2} \equiv (-\nu-1,1/(2\nu),-2)$.
+Finally, since $\rho = -2 < 1$ for this random variable, the multiplication operation with $X \equiv (0,1/2,2)$ yields $T \equiv \mathcal{R}_{\nu+1}$, and so the density of $T$ is asymptotically $c x^{-\nu-1}$ as $x \to \infty$.
+	Indeed, the density of $T$ satisfies $p_T(x) = c_\nu (1 + x^2 / \nu)^{-(\nu+1)/2}$ where $c_\nu = \Gamma(\frac{\nu+1}{2})/\Gamma(\frac{\nu}{2}) (\nu\pi)^{-1/2}$, which exhibits the predicted tail behaviour.
+\end{example}
+
+\begin{example}{Log-normal distribution}
+	Although the log-normal distribution does not lie in $\mathcal{G}$, the existence of log-normal tails arising from the multiplicative central limit theorem is suggested by our algebra.
+Let $X_1,X_2,\dots$ be independent standard normal random variables and let $Z_k = X_1\cdots X_{2^k}$ for each $k=1,2,\dots$.
+By the multiplicative central limit theorem, letting $\tau = \exp(\mathbb{E}\log |X_i|) \approx 1.13$, $\left(\frac{X_{1}\cdots X_{n}}{\tau}\right)^{1/\sqrt{n}}$ converges in distribution as $n\to\infty$ to a log-normal random variable $Z$ with density
+	\[
+		p_Z(x) = \frac{1}{x\sqrt{2\pi}} \exp(-\tfrac12 (\log x)^2).
+	\]
+	Therefore, the same is true for $V_k = (Z_k / \tau)^{2^{-k/2}}$.
+Using our algebra, we will attempt to reproduce the tail of this density.
+Letting $\tilde{Z}_k = X_{2^k} \cdots X_{2^{k+1}}$, we see that $Z_{k+1} = Z_k \tilde{Z}_k$, and $Z_k,\tilde{Z}_k$ are iid.
+Let $Z_k \equiv (\nu_k, \sigma_k, \rho_k)$, by induction using the multiplication operation, we find that $\nu_{k+1} ~=~\frac{1}{\mu}\left(\frac{2\nu_{k}}{\rho_{k}}-\frac{1}{2}\right)=\nu_{k}-\frac{\rho_{k}}{4}$, $\sigma_{k+1}=\mu\left(\sigma_{k}\rho_{k}\right)^{\frac{2}{\mu\rho_{k}}}=\frac{2}{\rho_{k}}\left(\sigma_{k}\rho_{k}\right)=2\sigma_{k}$, and $\rho_{k+1}=\frac{1}{\mu}=\frac{\rho_{k}}{2}$.
+	Since $\rho_0 = 2$, $\sigma_0 = 1/2$, and $\nu_0 = 0$, we find that $\rho_k = 2^{1-k}$ and $\sigma_k = 2^{k-1}$.
+Furthermore, $\nu_{k+1} = \nu_k - 2^{-k-1}$ and so $\nu_k = -1 + 2^{-k}$.
+Therefore $Z_k \equiv (-1+2^{-k},2^{k-1},2^{1-k})$, and
+	\[
+		V_k \equiv (-1+2^{-k/2},2^{k-1}\tau^{-2^{1-k}},2^{1-k/2}),
+	\]
+	and letting $\epsilon_k = 2^{-k/2}$, the tail behaviour of the density of $V_k$ satisfies
+	\begin{align*}
+		p_k(x) & \sim c_k x^{-1+\epsilon_{k}}\exp\left(-\frac{\epsilon_{k}^{-2}}{2\tau^{-2\epsilon_{k}^{2}}}x^{2\epsilon_k}\right) \\& \sim c_k x^{-1+\epsilon_{k}}\exp\left(-\frac{1}{2\tau^{-2\epsilon_{k}^{2}}}\left(\frac{x^{\epsilon_{k}}-1}{\epsilon_{k}}\right)^{2}\right) \approx c_k x^{-1}\exp\left(-\frac{1}{2}(\log x)^2\right),
+	\end{align*}
+	as $x \to \infty$, where the approximation improves as $k$ gets larger. The quality of this approximation is shown in \cref{fig:lognormal}.
+\end{example}
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.48\textwidth]{figures/LogNormal_Bulk.png}
+	\includegraphics[width=0.48\textwidth]{figures/LogNormal_Tail.png}
+	\caption{Estimation of the log-normal density by tail algebra applied to $V_k$.}
+	\label{fig:lognormal}
+\end{figure}
+
+\section{Additional details for experiments}
+
+The targets in \Cref{tab:de} and \Cref{tab:vi} are analyzed using
+the GGA in \Cref{sec:addtl_eg}. Note that Inverse Gamma (``IG'') corresponds
+to the inverse exponential. We selected closed form targets so that
+the Pareto tail index $\alpha$ is known analytically and the quality
+of theoretical predictions as well as empirical results can be evaluated
+against. All experiments are repeated for $100$ trials and $1,000$ samples
+from the model (as well as the approximation in VI) were used to compute
+each gradient estimate. Losses were trained until convergence, which 
+all occurred in under $10,000$ iterations at a $0.05$ learning rate and
+the Adam \citep{kingma2014adam} optimizer.
+
+\subsection{Normal target}
+
+Consider the toy example of a Normal target. This case is trivial for
+Gaussian based methods and is oftentimes the initialization.
+This lack of approximation gap in ADVI is seen in \Cref{fig:normal},
+where we also see that GGA achieves similar approximation quality.
+This is unsurprising as the GGA approximation in \cref{tab:dist_list}
+is also a Normal distribution.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.6\textwidth]{figures/de_normal_sum.png}
+	\includegraphics[width=.6\textwidth]{figures/vi_normal_sum.png}
+	\caption{Density estimation and VI against a known normal target}
+	\label{fig:normal}
+\end{figure}
+
+\subsection{Chi-square}
+
+Now let $X_{ij} \sim N(0,1)$ and consider $tr X^\top X$.
+Such quantities arise in the analysis of random projections.
+It is important here to recognize that the power operation $X \mapsto X^2$ is not equivalent to the multiplication operation $X \mapsto X \otimes X$, as multiplication assumes independence.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\textwidth]{figures/jl-chi2.png}
+	\caption{$5000$ samples of JL matrix trace (blue) vs GGA prediction (yellow)}
+	\label{fig:jl_chi2}
+\end{figure}
+
+\section{Operations in the Generalized Gamma Algebra}
+
+Whereas \cref{tab:gga_operations} provides a summary of our theory useful
+for referencing, in this section we provide additional explanation and references
+for how operations on random variables affect their GGA tails.
+
+\paragraph{Ordering.} A total ordering is imposed on the equivalence classes of $\mathcal{G}$ according to the heaviness of tails.
+In particular, we say that $(\nu_1,\sigma_1,\rho_1) \leq (\nu_2,\sigma_2,\rho_2)$ if $(x^{\nu_1} e^{-\sigma_1 x^{\rho_1}}) / (x^{\nu_2} e^{-\sigma_2 x^{\rho_2}})$ is bounded as $x \to \infty$.
+As usual, we say $(\nu_1,\sigma_1,\rho_1) < (\nu_2,\sigma_2,\rho_2)$ if $(\nu_1,\sigma_1,\rho_1) \leq (\nu_2,\sigma_2,\rho_2)$ but $(\nu_1,\sigma_1,\rho_1) \not\equiv (\nu_2,\sigma_2,\rho_2)$.
+
+%\paragraph{Multiplication by a Constant.}
+
+\paragraph{Addition.} Tails of this form are  closed under addition. Combining subexponentiality for $\rho < 1$
+%Any random variable $X$ in $(\nu,\sigma,\rho)$ for $\rho < 1$ is subexponential 
+\cite[Chapter X.1]{asmussen2010ruin}, 
+%that is, any sum of iid copies of $X$ asymptotically saturates the union bound: $\mathbb{P}\left(\sum_{i=1}^{n}X_{i}>x\right)\sim\sum_{i=1}^{n}\mathbb{P}(X_{i}>x)$ as $x \to \infty$. Combining this
+with \cite[Thm 3.1 \& eqn. (8.3)]{asmussen2017tail},
+\begin{proposition}
+\label{prop:gga_add}
+Denoting the addition of random variables (additive convolution of densities) by $\oplus$,
+\begin{multline}
+(\nu_{1},\sigma_{1},\rho_{1})\oplus(\nu_{2},\sigma_{2},\rho_{2})
+\\\equiv \begin{cases}
+\max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\} & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1\\
+\left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right) & \text{ if }\rho_{1}=\rho_{2}=1\\
+(\nu_{1}+\nu_{2}+1-\frac{\rho}{2},(\sigma_{1}^{-\frac{1}{\rho-1}}+\sigma_{2}^{-\frac{1}{\rho-1}})^{1-\rho},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}>1.
+\end{cases}
+\end{multline}
+\end{proposition}
+
+\paragraph{Powers.} For all exponents $\beta > 0$, by invoking a change of variables $x \mapsto x^\beta$, it is easy to show that
+%\begin{equation}
+%\label{eq:PowerEqn}
+$(\nu, \sigma, \rho)^\beta \equiv  \left(\frac{\nu+1}{\beta}-1, \sigma, \frac{\rho}\beta\right).$
+%\end{equation}
+%\paragraph{Reciprocals.}
+We \emph{define} negative powers and reciprocals equivalently to positive powers in the case $\beta < 0$.
+This equivalence cannot be proven to hold in general since we cannot determine tail asymptotics of the reciprocal without knowledge of its behaviour around zero.
+Therefore, we implicitly assume that the behaviour around zero mimics the tail behaviour, that is, \cref{eq:GenGammaTails} holds as $x \to 0^+$.
+However, this can only hold provided $(\nu + 1)/\rho > 0$ and $\rho \neq 0$.
+In all other cases, including $\mathcal{R}_\nu$, we assume that the density of $X$ approaches a nonzero value near zero, and define the reciprocal to be $\mathcal{R}_{2}$.
+%By assuming $p_X(x) \to \tilde{c} x^\nu e^{-\sigma x^\rho}$ as $x \to 0^+$, then \cref{eq:PowerEqn} also holds for $\beta < 0$.
+
+\paragraph{Multiplication.}
+For any $c \in \mathbb{R} \backslash \{0\}$, it can be readily seen from a change of variables $x \mapsto c x$ that $c (\nu, \sigma, \rho) = (\nu, \sigma / |c|^\rho, \rho)$.
+The class $\mathcal{G}$ is also closed under multiplication (assuming independence of random variables), as we show in the following result --- the proof is delayed to Appendix C.
+\begin{proposition}
+\label{prop:Mult}
+Denoting the multiplication of independent random variables (multiplicative convolution) by $\otimes$,
+\[
+(\nu_{1},\sigma_{1},\rho_{1})\otimes(\nu_{2},\sigma_{2},\rho_{2})
+\equiv\begin{cases}
+\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0\\
+\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}>0\\
+\mathcal{R}_{|\nu_1|} & \mbox{ if }\rho_{1}\leq0,\rho_{2}>0 \\
+\mathcal{R}_{\min\{|\nu_1|,|\nu_2|\}} & \mbox{ if }\rho_{1}=0,\rho_{2}=0
+\end{cases}
+\]
+where $\mu=\frac{1}{|\rho_{1}|}+\frac{1}{|\rho_{2}|}=\frac{|\rho_{1}|+|\rho_{2}|}{|\rho_{1}\rho_{2}|}$ and $\sigma=\mu(\sigma_{1}|\rho_{1}|)^{\frac{1}{\mu|\rho_{1}|}}(\sigma_{2}|\rho_{2}|)^{\frac{1}{\mu|\rho_{2}|}}$.
+\end{proposition}
+% \begin{proof}
+% Appendix C.
+% \end{proof}
+
+%\feynman{I added this back; we need it for the StudentT example but this result currently doesn't give what we need}
+%The next binary operation of interest is \emph{multiplication}. Using Cauchy's inequality, we can adopt the naive
+%\[
+%X \cdot Y \equiv \max\{X^2, Y^2\}.
+%\]
+%In the case where $X,Y \in \mathcal{E}$, this turns out to be reasonable.
+However, for any other use case, Breiman's lemma suggests that $X \cdot Y \equiv \max\{X, Y\}$.
+\todo{Perform Tauberian analysis to determine sharpness of these operations}.
+
+\paragraph{Product of Densities.}
+
+We can also consider a product of densities operation acting on two random variables $X,Y$, denoted $X \& Y$, by
+$p_{X \& Y}(x) = c p_X(x) p_Y(x)$,
+where $c > 0$ is an appropriate normalizing constant and $p_X,p_Y,p_{X\& Y}$ are the densities of $X$, $Y$, and $X \& Y$, respectively. In terms of the equivalence classes:
+\[
+(\nu_1,\sigma_1,\rho_1)\&(\nu_2,\sigma_2,\rho_2) 
+\equiv \begin{cases}
+(\nu_{1}+\nu_{2},\sigma_{1},\rho_{1}) & \text{ if }\rho_{1}<\rho_{2}\\
+(\nu_{1}+\nu_{2},\sigma_{1}+\sigma_{2},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}\\
+(\nu_{1}+\nu_{2},\sigma_{2},\rho_{2}) & \text{ otherwise.}
+\end{cases}
+\]
+Note that this particular operation does not require either $p_X$ or $p_Y$ to be normalized --- only the tail behaviour is needed.
+We may also use this to work out the tail behaviour of a posterior density, provided the tail behaviour of the likelihood in the parameters is known.
+
+\paragraph{Lipschitz Functions.}
+
+There are many multivariate functions that cannot be readily represented in terms of the operations covered thus far.
+For these, it is important to specify the tail behaviour of pushforward measures under Lipschitz-continuous functions.
+Fortunately, this is covered by \cref{lem:lipschitz} below, presented in \cite[Proposition 1.3]{ledoux2001concentration}.
+H\"{o}lder-continuous functions can also be represented as a composition of a power operation and a Lipschitz-continuous function. 
+\begin{lemma}\label{lem:lipschitz}
+For any Lipschitz continuous function $f:\mathbb{R}^d \to \mathbb{R}$ satisfying $\|f(x)-f(y)\|\leq L\|x -y\|$ for $x,y \in \mathbb{R}^d$, there is $f(X_1,\dots,X_d) \equiv L \max\{X_1,\dots,X_d\}.$
+\end{lemma}
+
+\paragraph{Power Law Approximation.} Note that as $x \to \infty$, $p_{|X|}(x) \sim c x^\nu e^{-\sigma x^\rho} = \tilde{c}x^\nu e^{-\sigma\rho\frac{x^\rho - 1}{\rho}} \approx \tilde{c} x^\nu e^{-\sigma\rho \log x}$
+    \[
+    p_{|X|}(x) \sim c x^\nu e^{-\sigma x^\rho} = \tilde{c} x^\nu e^{-\sigma(x^\rho - 0)} = \tilde{c} x^\nu e^{-\sigma\rho\frac{x^\rho - 1}{\rho}} \approx \tilde{c} x^\nu e^{-\sigma\rho \log x} = \tilde{c} x^{\nu-\sigma \rho},
+    \]
+    where we have used the approximation $\log x = \rho^{-2}(x^\rho - 1) + \mathcal{O}(\rho^2)$.
+Consequently, we can represent tails of this form by the Student $t$ distribution with $|\nu-\sigma\rho|-1$ degrees of freedom.
+In practice, we find this approximation tends to \emph{overestimate} the heaviness of the tail.
+Alternatively, the generalized Gamma density (\ref{eq:GenGammaDensity}) satisfies $\mathbb{E}X^r = \sigma^{-r/\rho} \Gamma(\frac{\nu+1+r}{\rho})/\Gamma(\frac{\nu+1}{\rho})$ for $r > 0$.
+Let $\alpha > 0$ be such that $\mathbb{E}X^\alpha = 2$.
+By Markov's inequality, the tail of $X$ satisfies $\mathbb{P}(X>x)\leq 2 x^{-\alpha}$.
+Therefore, we can represent tails of this form by the Student $t$ distribution with $\alpha+1$ degrees of freedom (generate $X \sim t_{\alpha}$).
+In practice, we find this approximation to be more accurate, and is hence used in Section 4.1.
+
+\section{List of univariate distributions}
+
+\bgroup
+\def\arraystretch{2}
+\begin{longtable}{|c|c|c|c|}
+\caption{List of univariate distributions}\label{tab:dist_list}\\
+\hline 
+Name & Support & Density $p(x)$ & Class\tabularnewline
+\hline 
+\hline 
+Benktander Type II & $(0,\infty)$ & $e^{\frac{a}{b}(1-x^{b})}x^{b-2}(ax^{b}-b+1)$ & $(2b-2,\frac{a}{b},b)$\tabularnewline
+\hline 
+Beta prime distribution & $(0,\infty)$ & $\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}x^{\alpha-1}(1+x)^{-\alpha-\beta}$ & $\mathcal{R}_{\beta+1}$\tabularnewline
+\hline 
+Burr distribution & $(0,\infty)$ & $ckx^{c-1} (1+x^{c})^{-k-1}$ & $\mathcal{R}_{ck+1}$\tabularnewline
+\hline 
+Cauchy distribution & $(-\infty,\infty)$ & $(\pi\gamma)^{-1}\left[1+\left(\frac{x-x_{0}}{\gamma}\right)^{2}\right]^{-1}$ & $\mathcal{R}_{2}$\tabularnewline
+\hline 
+Chi distribution & $(0,\infty)$ & $\frac{1}{2^{k/2-1}\Gamma(k/2)}x^{k-1}e^{-x^{2}/2}$ & $(k-1,\frac{1}{2},2)$\tabularnewline
+\hline 
+Chi-squared distribution & $(0,\infty)$ & $\frac{1}{2^{k/2}\Gamma(k/2)}x^{\frac{k}{2}-1}e^{-x/2}$ & $(\frac{k}{2}-1,\frac{1}{2},1)$\tabularnewline
+\hline 
+Dagum distribution & $(0,\infty)$ & $\frac{ap}{x}\left(\frac{x}{b}\right)^{ap}\left(\left(\frac{x}{b}\right)^{a}+1\right)^{-p-1}$ & $\mathcal{R}_{a+1}$\tabularnewline
+\hline 
+Davis distribution & $(0,\infty)$ & $\propto(x-\mu)^{-1-n}/\left(e^{\frac{b}{x-\mu}}-1\right)$ & $(-1-n,b,-1)$\tabularnewline
+\hline 
+Exponential distribution & $(0,\infty)$ & $\lambda e^{-\lambda x}$ & $(0,\lambda,1)$\tabularnewline
+\hline 
+$F$ distribution & $(0,\infty)$ & $\propto x^{d_{1}/2-1}(d_{1}x+d_{2})^{-(d_{1}+d_{2})/2}$ & $\mathcal{R}_{d_{2}/2+1}$\tabularnewline
+\hline 
+Fisher $z$-distribution & $(-\infty,\infty)$ & $\propto\frac{e^{d_{1}x}}{(d_{1}e^{2x}+d_{2})^{(d_{1}+d_{2})/2}}$ & $(0,d_{2},1)$\tabularnewline
+\hline 
+Frechet distribution & $(0,\infty)$ & $\frac{\alpha}{\lambda}\left(\frac{x-m}{\lambda}\right)^{-1-\alpha}e^{-\left(\frac{x-m}{\lambda}\right)^{-\alpha}}$ & $(-1-\alpha,\lambda^{\alpha},-\alpha)$\tabularnewline
+\hline 
+Gamma distribution & $(0,\infty)$ & $\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta x}$ & $(\alpha-1,\beta,1)$\tabularnewline
+\hline 
+Gamma/Gompertz distribution & $(0,\infty)$ & $bse^{bx}\beta^{s}/(\beta-1+e^{bx})^{s+1}$ & $(0,bs,1)$\tabularnewline
+\hline 
+Generalized hyperbolic distribution & $(-\infty,\infty)$ & $\propto e^{\beta(x-\mu)}\frac{K_{\lambda-1/2}(\alpha\sqrt{\delta^{2}+(x-\mu)^{2}})}{(\delta^{2}+(x-\mu)^{2})^{1/4-\lambda/2}}$ & $(\lambda-1,\alpha-\beta,1)$\tabularnewline
+\hline 
+Generalized Normal distribution & $(-\infty,\infty)$ & $\frac{\beta}{2\alpha\Gamma(1/\beta)}\exp\left(-\left(\frac{|x-\mu|}{\alpha}\right)^{\beta}\right)$ & $(0,\alpha^{-\beta},\beta)$\tabularnewline
+\hline 
+Geometric stable distribution & $(-\infty,\infty)$ & no closed form & $\mathcal{R}_{\alpha+1}$\tabularnewline
+\hline 
+Gompertz distribution & $(0,\infty)$ & $\sigma \eta \exp(\eta + \sigma x - \eta e^{\sigma x})$ & $\mathcal{L}$\tabularnewline
+\hline
+Gumbel distribution & $(0,\infty)$ & $\beta^{-1} e^{-(\beta^{-1}(x-\mu)+e^{-\beta^{-1}(x-\mu)})}$ & $(0,\frac{1}{\beta},1)$\tabularnewline
+\hline
+Gumbel Type II distribution & $(0,\infty)$ & $\alpha\beta x^{-\alpha-1}e^{-\beta x^{-\alpha}}$ & $(-\alpha-1,\beta,-\alpha)$\tabularnewline
+\hline 
+Holtsmark distribution & $(-\infty,\infty)$ & no closed form & $\mathcal{R}_{5/2}$\tabularnewline
+\hline 
+Hyperbolic secant distribution & $(-\infty,\infty)$ & $\frac{1}{2}\text{sech}\left(\frac{\pi x}{2}\right)$ & $(0,\frac{\pi}{2},1)$\tabularnewline
+\hline 
+Inverse chi-squared distribution & $(0,\infty)$ & $\frac{2^{-k/2}}{\Gamma(k/2)}x^{-k/2-1}e^{-1/(2x)}$ & $(-\frac{k}{2}-1,\frac{1}{2},-1)$\tabularnewline
+\hline 
+Inverse gamma distribution & $(0,\infty)$ & $\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{-\alpha-1}e^{-\beta/x}$ & $(-\alpha-1,\beta,-1)$\tabularnewline
+\hline 
+Levy distribution & $(0,\infty)$ & $\sqrt{\frac{c}{2\pi}}(x-\mu)^{-3/2}e^{-\frac{c}{2(x-\mu)}}$ & $(-\frac{3}{2},\frac{c}{2},-1)$\tabularnewline
+\hline 
+Laplace distribution & $(-\infty,\infty)$ & $\frac{1}{2\lambda}\exp\left(-\frac{|x-\mu|}{\lambda}\right)$ & $(0,\frac{1}{\lambda},1)$\tabularnewline
+\hline 
+Logistic distribution & $(-\infty,\infty)$ & $\frac{e^{-(x-\mu)/\lambda}}{\lambda(1+e^{-(x-\mu)/\lambda})^{2}}$ & $(0,\frac{1}{\lambda},1)$\tabularnewline
+\hline 
+Log-Cauchy distribution &
+$(0,\infty)$ & $\frac{\sigma}{x\pi}((\log x - \mu)^2 + \sigma^2)^{-1}$ & $\mathcal{R}_1$\tabularnewline
+\hline
+Log-Laplace distribution & $(0,\infty)$ & $\frac{1}{2\lambda x}\exp\left(-\frac{\left|\log x-\mu\right|}{\lambda}\right)$ & $\mathcal{R}_{1/\lambda+1}$\tabularnewline
+\hline 
+Log-logistic distribution & $(0,\infty)$ & $\frac{\beta}{\alpha}\left(\frac{x}{\alpha}\right)^{\beta-1}\left(1+\left(\frac{x}{\alpha}\right)^{\beta}\right)^{-2}$ & $\mathcal{R}_{\beta+1}$\tabularnewline
+\hline
+Log-$t$ distribution &
+$(0,\infty)$ &
+$\propto x^{-1} (1 + \frac1{\nu}(\log x - \mu)^2)^{-\frac{\nu+1}{2}}$ & $\mathcal{R}_1$\tabularnewline
+\hline 
+Lomax distribution & $(0,\infty)$ & $\frac{\alpha}{\lambda}\left(1+\frac{x}{\lambda}\right)^{-\alpha-1}$ & $\mathcal{R}_{\alpha+1}$\tabularnewline
+\hline 
+Maxwell-Boltzmann distribution & $(0,\infty)$ & $\sqrt{\frac{2}{\pi}}\frac{x^{2}e^{-x^{2}/(2\sigma^{2})}}{\sigma^{3}}$ & $(2,\frac{1}{2\sigma^{2}},2)$\tabularnewline
+\hline 
+Normal distribution & $(-\infty,\infty)$ & $\frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^{2}}$ & $(0,\frac{1}{2\sigma^{2}},2$)\tabularnewline
+\hline 
+Pareto distribution & $(x_{0},\infty)$ & $\alpha x_{0}^{\alpha}x^{-\alpha-1}$ & $\mathcal{R}_{\alpha+1}$\tabularnewline
+\hline 
+Rayleigh distribution & $(0,\infty)$ & $\frac{x}{\sigma^{2}}e^{-x^{2}/(2\sigma^{2})}$ & $(1,\frac{1}{2\sigma^{2}},2)$\tabularnewline
+\hline 
+Rice distribution & $(0,\infty)$ & $\frac{x}{\sigma^{2}}\exp\left(-\frac{(x^{2}+\nu^{2})}{2\sigma^{2}}\right)I_{0}\left(\frac{x\nu}{\sigma^{2}}\right)$ & $(\frac{1}{2},\frac{1}{2\sigma^{2}},2)$\tabularnewline
+\hline 
+Skew normal distribution & $(-\infty,\infty)$ & no closed form & $(0,\frac{1}{2\sigma^{2}},2)$\tabularnewline
+\hline 
+Slash distribution & $(-\infty,\infty)$ & $\frac{1-e^{-\frac{1}{2}x^{2}}}{\sqrt{2\pi}x^{2}}$ & $(-2,\frac{1}{2},2)$\tabularnewline
+\hline 
+Stable distribution & $(-\infty,\infty)$ & no closed form & $\mathcal{R}_{\alpha+1}$\tabularnewline
+\hline 
+Student's $t$-distribution & $(-\infty,\infty)$ & $\frac{\Gamma(\frac{\nu+1}{2})}{\sqrt{\nu\pi}\Gamma(\frac{\nu}{2})}\left(1+\frac{x^{2}}{\nu}\right)^{-\frac{\nu+1}{2}}$ & $\mathcal{R}_{\nu+1}$\tabularnewline
+\hline 
+Tracy-Widom distribution & $(-\infty,\infty)$ & no closed form & $(-\frac{3\beta}{4}-1,\frac{2\beta}{3},\frac{3}{2})$\tabularnewline
+\hline 
+Voigt distribution & $(-\infty,\infty)$ & no closed form & $\mathcal{R}_{2}$\tabularnewline
+\hline 
+Weibull distribution & $(0,\infty)$ & $\frac{\rho}{\lambda}\left(\frac{x}{\lambda}\right)^{\rho-1}e^{-(x/\lambda)^{\rho}}$ & $(\rho-1,\lambda^{-\rho},\rho)$\tabularnewline
+\hline 
+\end{longtable}
+\egroup
+
+The following densities are not supported by our algebra: Benini distribution; Benktander Type I distribution; Johnson's $S_U$-distribution; and the log-normal distribution.
+All of these densities exhibit log-normal tails.
+
+\section{Proofs}\label{sec:proofs}
+
+\begin{proof}[Proof of Lemma \ref{lem:Products}]
+The proof relies on the following integral definition \cite[pg.
+183]{watson1995treatise} and asymptotic relation as $z \to \infty$ \cite[pg.
+202]{watson1995treatise} of the modified Bessel function $K_\nu(z)$ for $z > 0$ and $\nu \geq 0$,
+\begin{equation}
+\label{eq:BesselInt}
+K_\nu(z) = \frac12 \left(\frac{z}{2}\right)^\nu \int_0^\infty u^{-\nu-1}\exp\left(-u - \frac{z^2}{4u}\right) \dd u \sim \sqrt{\frac{\pi}{2z}}e^{-z}. 
+\end{equation}
+We also make use of the known density for the product of two independent continuous random variables: if $X$ and $Y$ have densities $p_X$ and $p_Y$ respectively, then $Z = XY$ has density
+\[
+p_Z(z) = \int_\mathbb{R} p_X(x) p_Y(z/x) |x|^{-1} \dd x.
+\]
+\paragraph{Exponentials.} Recalling that the density of $X \sim \mathrm{Exp}(\lambda)$ is $p_X(x) = \lambda e^{-\lambda x}$ for $x \geq 0$, for $Z = XY$ where $X \sim \mathrm{Exp}(\lambda_1)$ and $Y \sim \mathrm{Exp}(\lambda_2)$ are independent,
+\[
+p_Z(z) = \int_{0}^{\infty}x^{-1}\lambda_{1}e^{-\lambda_{1}x}\lambda_{2}e^{-\lambda_{2}z/x}\dd x = \lambda_{1}\lambda_{2}\int_{0}^{\infty}x^{-1}e^{-\lambda_{1}x-\lambda_{2}z/x}\dd x.
+\]
+Since $2K_{0}(2\sqrt{z})=\int_{0}^{\infty}u^{-1}\exp(-u-\frac{z}{u})\dd u$, let $u = \lambda_1 v$, so that $\dd u = \lambda_1 \dd v$,
+\[
+2K_{0}(2\sqrt{\lambda_{1}\lambda_{2}z})=\int_{0}^{\infty}u^{-1}\exp\left(-\lambda_{1}v-\lambda_{2}\frac{z}{v}\right)\dd v.
+\]
+Therefore, letting $\lambda = \sqrt{\lambda_1\lambda_2}$, 
+\[
+p_Z(z) = 2\lambda^2 K_0(2\lambda\sqrt{z}) \sim \sqrt{\pi} \lambda^{3/2} z^{-1/4} e^{-2\lambda z^{1/2}}.
+\]
+\paragraph{Normals.} Recalling that the density of $X \sim \mathcal{N}(0,1)$ is $p_X(x) = (2\pi)^{-1/2} \exp(-\frac12 x^2)$, for $Z = XY$ where $X,Y \sim \mathcal{N}(0,1)$ are independent, 
+\begin{align*}
+p_Z(z) &= \frac{1}{2\pi}\int_{\mathbb{R}}\left|x\right|^{-1}e^{-\frac{1}{2}x^{2}}e^{-\frac{1}{2}z^{2}/x^{2}}\dd x\\
+&=\frac{1}{\pi}\int_{0}^{\infty}x^{-1}e^{-\frac{1}{2}x^{2}-\frac{1}{2}z^{2}/x^{2}}\dd x\\
+&=\frac{1}{\pi}\int_{0}^{\infty}x^{-1}e^{-\frac{1}{2}x^{2}-\frac{1}{2}z^{2}/x^{2}}\dd x.
+\end{align*}
+Let $u=\frac{1}{2}x^{2}$ so that $\dd u=x \dd x$ and
+\[
+K_{\nu}(z)=z^{\nu}\int_{0}^{\infty}x^{-2\nu-1}\exp\left(-\frac{1}{2}x^{2}-\frac{z^{2}}{2x^{2}}\right)\dd x.
+\]
+In particular, for any $z \in \mathbb{R}$,
+\begin{equation}
+\label{eq:BesselGauss}
+K_{0}(|z|)=\int_{0}^{\infty}x^{-1}\exp\left(-\frac{1}{2}x^{2}-\frac{z^{2}}{2x^{2}}\right)\dd x,
+\end{equation}
+and so
+\[
+p_Z(z) = \frac{1}{\pi} K_0(|z|) \sim \frac{1}{\sqrt{2\pi|z|}} e^{-|z|}.
+\]
+\paragraph{Reciprocal Normals.} Finally, by a change of variables, we note that the density of $X^{-1}$ where $X \sim \mathcal{N}(0,1)$ is $p_{X^{-1}}(x) = (2\pi)^{-1/2} x^{-2}\exp(-\frac{1}{2x^2})$.
+Therefore, the density of $Z = 1/(XY)$ where $X,Y \sim \mathcal{N}(0,1)$ are independent is given by
+\begin{align*}
+p_{Z}(z)&=\int_{\mathbb{R}}\frac{1}{\sqrt{2\pi}x^{2}}e^{-\frac{1}{2x^{2}}}\frac{x^{2}}{\sqrt{2\pi}z^{2}}e^{-\frac{x^{2}}{2z^{2}}}\frac{1}{\left|x\right|}\dd x\\
+&=\frac{1}{2\pi z^{2}}\int_{\mathbb{R}}e^{-\frac{1}{2x^{2}}-\frac{x^{2}}{2z^{2}}}\frac{1}{\left|x\right|}\dd x\\
+&=\frac{1}{\pi z^{2}}\int_{0}^{\infty}e^{-\frac{1}{2x^{2}}-\frac{x^{2}}{2z^{2}}}\frac{1}{x}\dd x\\
+&=\frac{1}{\pi z^{2}}K_{0}(|z|^{-1}) \sim \sqrt{\frac{1}{2\pi}}|z|^{-3/2}e^{-|z|^{-1}},
+\end{align*}
+where we have once again used (\ref{eq:BesselGauss}). 
+\end{proof}
+
+% \begin{lemma}
+% Suppose that $X$ is generalized Gamma distributed with density $p_X(x)$. Then $\int_z^\infty x^{-1} p_X(x) \dd x = o(\bar{p}_X(x))$ as $x \to \infty$.
+% \end{lemma}
+% \begin{proof}
+% Suppose that $X$ is generalized Gamma distributed with parameters $\nu,\sigma,\rho$. Then
+% \[
+% \int_z^\infty x^{-1} p_X(x) \dd x = \frac{\Gamma(\frac{\nu}{\rho}, \sigma z^\rho)}{\Gamma(\frac{\nu}{\rho})}.
+% \]
+% where $\Gamma(a,z)$ is the (upper) incomplete Gamma function. From \cite[8.11.2]{NIST:DLMF}, 
+% \[
+% \frac{\Gamma(\frac{\nu}{\rho}, \sigma z^\rho)}{\Gamma(\frac{\nu}{\rho})} \sim c\sigma^{\frac{\nu}{\rho}-1}z^{\nu-\rho}e^{-\sigma z^{\rho}}=o(\bar{p}_{X}(z)),
+% \]
+% for an appropriate $c > 0$.
+% \end{proof}
+% \begin{lemma}
+% Let $\bar{X},\bar{Y}$ be independent generalized Gamma random variables with densities $\bar{p}_X$ and $\bar{p}_Y$ respectively.
+If $X,Y$ are independent random variables with densities $p_X(x) \sim c_X \bar{p}_X(x)$ and $p_Y(x) \sim c_Y \bar{p}_Y(x)$ as $x\to\infty$, then the densities $p_{XY}$ and $\bar{p}_{XY}$ of $XY$ and $\bar{X}\bar{Y}$, respectively, satisfy $p_{XY}(x) \sim \bar{p}_{XY}(x)$ as $x \to \infty$.
+% \end{lemma}
+% \begin{proof}
+% Let $C_{X}=\sup_{x}|p_{X}(x)-\bar{p}_{X}(x)|$ and $C_{Y}=\sup_{y}|p_{Y}(y)-\bar{p}_{Y}(y)|$, and for any $\epsilon > 0$, let $M_\epsilon$ be such that
+% \[
+% \max_{x\geq M_{\epsilon}}\left\{ \left|\frac{p_{X}(x)}{\bar{p}_{X}(x)}-1\right|,\left|\frac{p_{Y}(x)}{\bar{p}_{Y}(x)}-1\right|\right\} <\epsilon.
+% \]
+% \end{proof}
+
+Recall that the Mellin transform of a function $f$ on $(0,\infty)$ is given by
+\[
+\mathcal{M}_s[f] = \int_0^\infty x^{s-1} f(x) \dd x.
+\]
+Letting $p_{XY}$ denote the density of the product of independent random variables $X,Y$ with respective densities $p_X$ and $p_Y$, $\mathcal{M}_s[p_{XY}] = \mathcal{M}_s[p_X] \mathcal{M}_s[p_Y]$.
+There is
+\[
+\mathcal{M}_s[c x^\nu e^{-\sigma x^\rho}] = \frac{c \sigma^{-\nu/\rho}}{\rho} \sigma^{-s/\rho} \Gamma\left(\frac{\nu}{\rho} + \frac{s}{\rho}\right).
+\]
+To facilitate the proof of Proposition \ref{prop:Mult}, we define the Fox $H$-function
+\[
+H_{p,q}^{m,n}\left[z\left|\substack{(a_{1},A_{1}),\dots,(a_{p},A_{p})\\
+(b_{1},B_{1}),\dots,(b_{q},B_{q})
+}
+\right.\right]
+\]
+as the inverse Mellin transform of
+\[
+\Theta(s) = z^{-s} \frac{\prod_{j=1}^m \Gamma(b_j + B_j s) \cdots \prod_{j=1}^n \Gamma(1-a_j - A_j s)}{\prod_{j=m+1}^q \Gamma(1-b_j - B_j s) \prod_{j=n+1}^p \Gamma(a_j + A_j s)}.
+\]
+An important property of the Fox $H$-function is its asymptotic behaviour as $z \to \infty$. From \cite[Theorem 1.3]{mathai2009h},
+\[
+H_{p,q}^{q,0}\left[z\left|\substack{(a_{1},A_{1}),\dots,(a_{p},A_{p})\\
+(b_{1},B_{1}),\dots,(b_{q},B_{q})}\right.\right] \sim c x^{(\delta+\frac12)/\mu} \exp(-\mu \beta^{-1/\mu} x^{1/\mu}),\qquad \mbox{as }x\to\infty,
+\]
+for some constant $c > 0$, where $\beta = \prod_{j=1}^p (A_j)^{-A_j} \prod_{j=1}^q B_j^{B_j}$, $\mu = \sum_{j=1}^q B_j - \sum_{j=1}^p A_j$, and $\delta = \sum_{j=1}^q b_j - \sum_{j=1}^p a_j + \frac{p-q}{2}$.
+
+\begin{proof}[Proof of Proposition \ref{prop:Mult}]
+The $\rho_1 \leq 0, \rho_2 > 0$ and $\rho_1 = \rho_2 = 0$ cases follow from Breiman's lemma \cite[Lemma B.5.1]{buraczewski2016stochastic}.
+Our argument proceeds similar to \cite{asmussen2017tail}.
+Assume that $\rho_1,\rho_2 > 0$ and let $0 < \epsilon < 1$ be such that $0 < a_- < a_+ < 1$, where
+\[
+a_+ = \frac{(1+\epsilon)\rho_2}{\rho_1+\rho_2},\qquad a_{-} = 1 - \frac{(1+\epsilon)\rho_1}{\rho_1+\rho_2}.
+\]
+Then for $\rho = \frac{\rho_1\rho_2}{\rho_1 + \rho_2}$, if $X \equiv (\nu_1,\sigma_1,\rho_1)$ and $Y \equiv (\nu_2,\sigma_2,\rho_2)$, then
+\begin{align*}
+\mathbb{P}(XY>x,X\notin[x^{a_{-}},x^{a_{+}}])  & \leq\mathbb{P}(X>x^{a_{+}})+\mathbb{P}(Y>x^{1-a_{-}}) \\
+& \sim c_{1}x^{\nu_{1}a_{+}}e^{-\sigma_{1}x^{\rho_{1}a_{+}}}+c_{2}x^{\nu_{2}(1-a_{-})}e^{-\sigma_{2}x^{\rho_{2}(1-a_{-})}}\\
+& \leq\left(c_{1}x^{\nu_{1}a_{+}}+c_{2}x^{\nu_{2}(1-a_{-})}\right)e^{-\min\{\sigma_{1},\sigma_{2}\}x^{(1+\epsilon)\rho}} = o(x^\nu e^{-\sigma x^\rho}),
+\end{align*}
+for any $\nu,\sigma > 0$.
+Hence, it will suffice to show the claimed tail asymptotics for the generalized Gamma distribution.
+In this case, since $a_- > 0$ and $a_+ < 1$, the tail of the distribution for the product of $X,Y$ depends only on the tail of the distributions for $X$ and $Y$.
+
+Therefore, assume without loss of generality that $p_X(x) = c_X x^{\nu_1} e^{-\sigma_1 x^{\rho_1}}$ and $p_Y(x) = c_Y x^{\nu_2} e^{-\sigma_2 x^{\rho_2}}$. Then
+\[
+\mathcal{M}_s[p_{XY}] = c_X c_Y \frac{\sigma_{1}^{-\nu_{1}/\rho_{1}}}{\rho_{1}}\frac{\sigma_{2}^{-\nu_{2}/\rho_{2}}}{\rho_{2}}\left(\sigma_{1}^{1/\rho_{1}}\sigma_{2}^{1/\rho_{2}}\right)^{-s}\Gamma\left(\frac{\nu_{1}}{\rho_{1}}+\frac{s}{\rho_{1}}\right)\Gamma\left(\frac{\nu_{2}}{\rho_{2}}+\frac{s}{\rho_{2}}\right).
+\]
+Consequently,
+\[
+p_{XY}(z) = c_X c_Y \frac{\sigma_{1}^{-\nu_{1}/\rho_{1}}}{\rho_{1}}\frac{\sigma_{2}^{-\nu_{2}/\rho_{2}}}{\rho_{2}}H_{p,q}^{m,n}\left[\sigma_{1}^{1/\rho_{1}}\sigma_{2}^{1/\rho_{2}}z\left|\substack{-\\
+(\frac{\nu_{1}}{\rho_{1}},\frac{1}{\rho_{1}}),(\frac{\nu_{2}}{\rho_{2}},\frac{1}{\rho_{2}})
+}
+\right.\right]
+\]
+Computing the corresponding $\beta,\delta,\mu$ for the asymptotic expansion, we find that
+\[
+\mu=\frac{1}{\rho_{1}}+\frac{1}{\rho_{2}},\qquad\delta=\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-1,\qquad\beta=\rho_{1}^{-1/\rho_{1}}\rho_{2}^{-1/\rho_{2}}.
+\]
+Consequently, for some $c > 0$,
+\[
+p_{XY}(z) \sim c z^{\frac{1}{\mu}(\frac{1}{2}+\delta)}\exp\left(-\mu\beta^{-\frac{1}{\mu}}(\sigma_{1}^{1/\rho_{1}}\sigma_{2}^{1/\rho_{2}})^{\frac{1}{\mu}}z^{\frac{1}{\mu}}\right),
+\]
+which completes the $\rho_1,\rho_2 > 0$ case. The final case follows by composing the multiplication and reciprocal operations. Note that
+\begin{align*}
+(\nu_{1},\sigma_{1},-\rho_{1})^{-1}\otimes(\nu_{2},\sigma_{2},-\rho_{2})^{-1}	&\equiv\left(-\nu_{1}-2,\sigma_{1},\rho_{1}\right)\otimes\left(-\nu_{2}-2,\sigma_{2},\rho_{2}\right)\\
+&\equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}-2}{\rho_{1}}+\frac{-\nu_{2}-2}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)\\
+&\equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}}{\rho_{1}}+\frac{-\nu_{2}}{\rho_{2}}-2\mu-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)\\
+&\equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}}{\rho_{1}}+\frac{-\nu_{2}}{\rho_{2}}-\frac{1}{2}\right)-2,\sigma,\frac{1}{\mu}\right),
+\end{align*}
+and therefore
+\[
+(\nu_{1},\sigma_{1},-\rho_{1})\otimes(\nu_{2},\sigma_{2},-\rho_{2})\equiv\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right).
+\]
+\end{proof}
+
+\fi
+\end{document}

### 2022-08-23 hta/working_doc.tex
Initial Overleaf Import

\documentclass{article}
\usepackage[utf8]{inputenc}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{enumerate}
\usepackage{color}

\usepackage{float}
\usepackage{graphicx}

\usepackage{hyperref}
\usepackage{cleveref}

\usepackage{todonotes}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{bayesnet}

\newcommand{\dd}{\mathrm{d}}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\title{An Algebra on Tail Behaviour for Informing Inference Algorithm Design}

\begin{document}

\maketitle

\section{Introduction}
Our objective is to construct an algebra on the tail behaviour of one-dimensional distributions. For multi-dimensional distributions, we are primarily focused on tail behaviour of the marginals. The algebra will include exponential, Gaussian, log-normals, and power law tails, and form a commutative semiring under addition and multiplication (no inverses), while supporting powers, and some exponential and logarithmic operations. The complexities of the exact algebra for random variables (e.g. convolution) necessitate some approximations that will be suitable for most purposes.

\subsection{Riemannian-Manifold MCMC}
\label{ssec:mcmc}

Among the most commonly used MCMC algorithms is the Metropolis-adjusted Langevin algorithm (MALA), based on discretizations of Langevin diffusion:
\[
\dd X_t = -\tfrac12 \nabla \log(X_t) \dd t + \dd B_t,
\]
where $B_t$ is Brownian motion.

\begin{theorem}\label{thm:nmc-ergodic}
    From \cite{roberts1996exponential} theorem 4.3, MALA is not geometrically ergodic
    in the presence of heavy-tails ($\nabla  \log \pi(x) \to 0$).
    However, theorem 2.3 implies the Newtonian Monte-Carlo algorithm's use of the Hessian to define a metric tensor ensures the continuous
    time Langevin diffusion is exponentially ergodic. 
\end{theorem}
Our algebra may be used
to statically analyze a PPL program (representing a probabilistic model) and
recommend usage of potentially more expensive Riemannian methods.

\subsection{Proposal distributions}

Proposal distributions $q(x)$ are utilized in importance sampling, Metropolis-Hastings,
and variational inference, where the importance ratio $w(x) = \frac{p(x)}{q(x)}$ plays a
central role in their analysis. In particular, after establishing bounded variance of $w$ one
can apply Cauchy-Schwarz to yield efficient estimation procedures. For bounded variance, notice
\begin{align*}
    \text{Var}_q w
    &= \int_x \frac{p^2(x)}{q(x)} dx - 1
\end{align*}
Restricting comparison to polynomial orders, for $\int_1^\infty \frac{p^2(x)}{q(x)} dx$
to converge we must have $\frac{p^2(x)}{q(x)} = O(x^{-(1+\epsilon)})$ (with
upper bound $\frac{1+\epsilon}{\epsilon}$).

\begin{theorem}\label{thm:bdd-importance-ratio}
    If $p \in \mathcal{L}_\beta^2$ then $q \in \mathcal{L}_\alpha^2$ for $\alpha < 2 \beta$.
\end{theorem}

% Prior work (\cite{jaini2020tails}, TODO: cite NeurIPS submission) investigated
% the use of heavy-tailed variational families for variational inference, where
% tail parameters are jointly optimized as part of training.

Our algebra provides upper bounds on tail-coefficients which can be used as initializations
for the variational approximations. In doing so, we guarantee that proposal distributions $q(x)$ are at least as heavy as targets $p(x)$.

\section{Classes of Tail Behaviour}

\begin{definition}
The set of \emph{exponential-type densities} with scale $\alpha$ and exponent $\rho$, denoted $\mathcal{E}_\alpha^\rho$ is comprised of random variables $X$ satisfying
\[
\mathbb{P}(|X| \geq x) = \mathcal{O}(e^{-\alpha x^\rho}).
\]
Similarly, the set of \emph{logarithmic-type densities} with scale $\alpha$ and exponent $\rho$, denoted $\mathcal{L}_\alpha^\rho$, is comprised of random variables $X$ satisfying
\[
\mathbb{P}(|X| \geq x) = \mathcal{O}(e^{-\alpha (\log x)^\rho}).
\]
For brevity, we let $\mathcal{E}^\rho = \bigcup_{\alpha > 0} \mathcal{E}_\alpha^\rho$, $\mathcal{L}^\rho = \bigcup_{\alpha > 0} \mathcal{L}_\alpha^\rho$,  $\mathcal{E}=\bigcup_{\rho>0} \mathcal{E}^\rho$ and $\mathcal{L}=\bigcup_{\rho>0} \mathcal{L}^\rho$. We let $\mathcal{P} = \mathcal{E} \cup \mathcal{L}$ denote the set of viable random variables under this framework. 
\end{definition}

The tail behaviour is encoded by three properties: whether $X \in \mathcal{E}$ or $X \in \mathcal{L}$; the scale $\alpha$; and the exponent $\rho$. The collection of admissible random variables encompasses the following four classes:
\begin{enumerate}[(I)]
\item \textbf{Exponential:} $X \in \mathcal{E}_\alpha^1$ implies $\mathbb{P}(|X| \geq x) = \mathcal{O}(e^{-\alpha x})$
\item \textbf{Gaussian:}  $X\in\mathcal{E}_{\alpha}^{2}$ implies $\mathbb{P}(X\geq x)=\mathcal{O}(e^{-\alpha x^{2}})$.
\item \textbf{Power laws:} $X\in\mathcal{L}_{\alpha}^{1}$, implies $\mathbb{P}(X\geq x)=\mathcal{O}(x^{-\alpha})$.
\item \textbf{Log-normal:} $X\in\mathcal{L}_{\alpha}^{2}$, implies $\mathbb{P}(X\geq x)=\mathcal{O}(e^{-\alpha(\log x)^{2}})$.
\end{enumerate}

\section{Operations}

\paragraph{Ordering.}
First, we can assign a total ordering to random variables in the sets $\mathcal{E}$ and $\mathcal{L}$ over the equivalence classes specified by the parameters $\alpha$ and $\rho$. This ordering is defined so that $X > Y$ if $X$ has heavier tails than $Y$. In tail behaviour, $X \equiv Y$ if $X,Y$ belong to the same class with equivalent scales $\alpha$ and exponents $\rho$. Furthermore, $X > Y$ if one of the following holds:
\begin{enumerate}[(i)]
\item $X \in \mathcal{L}, Y \in \mathcal{E}$.
\item Both $X,Y \in \mathcal{L}$, but $X \in \mathcal{L}^\rho, Y \in \mathcal{L}^\lambda$,  and $\rho < \lambda$.
\item Both $X,Y \in \mathcal{E}$, but $X \in \mathcal{E}^\rho, Y \in \mathcal{E}^\lambda$, and $\rho < \lambda$.
\item Both $X,Y \in \mathcal{L}^\rho$, but $X \in \mathcal{L}_\alpha^\rho, Y \in \mathcal{L}_\beta^\rho$, and $\alpha < \beta$.
\item Both $X,Y \in \mathcal{E}^\rho$, but $X \in \mathcal{E}_\alpha^\rho, Y \in \mathcal{E}_\beta^\rho$, and $\alpha < \beta$.
\end{enumerate}

\paragraph{Exponential and Logarithm.}
By design, $X \in \mathcal{L}_\alpha^\rho$ if and only if $\log X \in \mathcal{E}_\alpha^\rho$ --- this is easy to show by a change of variables. Similarly, if $Y \in \mathcal{E}_\alpha^\rho$, then $\exp Y \in \mathcal{L}_\alpha^\rho$. If $Y \in \mathcal{E}_\alpha^\rho$, then $\log Y$ is lighter than any exponential density, so we are free to take $\log Y \in \mathcal{E}_\alpha^{\tilde{\infty}}$, where $\tilde{\infty}$ is a suitably large number. On the other hand, if $X \in \mathcal{L}_\alpha^\rho$, then $\exp X$ is heavier than any power law, falling into the family of \emph{super-heavy tails}. Our analysis is not capable of treating these variables, so we leave $\exp X$ to be undefined.

\paragraph{Powers.} If $X \in \mathcal{E}_\alpha^\rho$ (resp. $\mathcal{L}_\alpha^\rho$) then $X^\beta \in \mathcal{E}_\alpha^{\rho / \beta}$ (resp. $\mathcal{L}_\alpha^{\rho / \beta}$). If $\beta < 0$, then $X^{\beta} = (X^{-1})^{|\beta|}$, which will be covered by the reciprocal case later.

\paragraph{Multiplication by a Constant.}
It can be readily seen from the definitions that if $X \in \mathcal{E}_\alpha^\rho$, then for any $c \in \mathbb{R}$, $c X \in \mathcal{E}_{\alpha / |c|^\rho}^\rho$. On the other hand, the tails of logarithmic-type distributions are unaffected by constant factors, so if $X \in \mathcal{L}_\alpha^\rho$, then $c X \in \mathcal{L}_\alpha^\rho$.

\paragraph{Reciprocal.}
Unfortunately, determining the tail behaviour of the reciprocal $X^{-1}$ from the tail behaviour of $X$ is technically ill-posed, as it depends on the probability density of $X$ about zero. For example, a $t_\nu$-distributed random variable is defined as a ratio $\sqrt{\nu} X / Y$ where $X,Y \in \mathcal{E}_{1/2}^2$. This is true for any $\nu > 1$. Therefore, no rule depending only on the scale and exponent is capable of predicting the degrees of freedom $\nu$ in this setting. 
However, since the majority of heavy tails seem to arise in ratio distributions, it is critical to treat reciprocal operations. 
Fortunately, most cases will be covered by the following general rule shown in Lemma \ref{lem:Reciprocal}.

\begin{lemma}
\label{lem:Reciprocal}
Suppose that $X$ is a random variable with bounded density $p_X$. Then $X^{-1} \in \mathcal{L}_1^1$. 
\end{lemma}
\begin{proof}
By a change of variables, the density of $X^{-1}$ is
\[
p_{X^{-1}}(x) = x^{-2}p_X(x^{-1}).
\]
Since $p_X$ is bounded, $p_{X^{-1}}(x) = \mathcal{O}(x^{-2})$ as $x \to \infty$, which implies the result.
\end{proof}

\paragraph{Addition.}
Addition is the first operation not covered by a simple change of variables. Of course, at complete generality, one can use the naive bound
\[
X + Y \equiv \max\{2 X, 2 Y\}.
\]
This turns out to be effective in most cases when the scales do not matter. The single exception occurs when $X,Y \in \mathcal{E}^\rho$ for $\rho \geq 1 $, in which case, the scale does matter. This case is covered by the following lemma, which assumes independence.
\begin{lemma}
\label{lem:sum}
Suppose that $X, Y \in \mathcal{E}^\rho$ are independent random variables for $\rho \geq 1$. If $X \in \mathcal{E}_\alpha^\rho$ and $Y \in \mathcal{E}_\beta^\rho$, then $X + Y \in \mathcal{E}_\sigma^\rho$ where
\[
\sigma = \begin{cases}
\min\{\alpha,\beta\}&\mbox{ if } \rho = 1 \\
(\alpha^{-\frac{1}{\rho-1}}+\beta^{-\frac{1}{\rho-1}})^{1-\rho} &\mbox{ if } \rho > 1.
\end{cases}
\]
\end{lemma}
\begin{proof}
We focus on the $\rho > 1$ case, since the $\rho = 1$ case follows by taking $\rho \to 1^+$. It also suffices to consider the case where $\limsup_{x \to \infty} \mathbb{P}(|X| > x^{1/\rho}) = -\alpha$ and $\limsup_{x\to\infty} \mathbb{P}(|Y| > x^{1/\rho}) = -\beta$, as the tail behaviour will be bounded above by this case. By Kasahara's Tauberian Theorem, 
\begin{align*}
\limsup_{\lambda \to \infty} \lambda^{-\frac{\rho}{\rho-1}}\log \mathbb{E}e^{\lambda |X|} &= \frac{\rho - 1}{\rho} (\alpha \rho)^{-\frac{1}{\rho-1}} \\
\limsup_{\lambda \to \infty} \lambda^{-\frac{\rho}{\rho-1}}\log \mathbb{E}e^{\lambda |Y|} &= \frac{\rho - 1}{\rho} (\beta \rho)^{-\frac{1}{\rho-1}},
\end{align*}
and since $X$ and $Y$ are independent,
\[
\limsup_{\lambda \to \infty} \lambda^{-\frac{\rho}{\rho-1}}\log \mathbb{E}e^{\lambda (|X|+|Y|)} = \frac{\rho - 1}{\rho} (\sigma \rho)^{-\frac{1}{\rho-1}},
\]
which, by the same theorem, implies that $\mathbb{P}(|X|+|Y|>x) = \mathcal{O}(e^{-\sigma x^\rho})$.
Since $|X+Y| \leq |X| + |Y|$, the result follows.
\end{proof}

\paragraph{Multiplication by a Random Variable.}
The next binary operation of interest is \emph{multiplication}. Using Cauchy's inequality, we can adopt the naive
\[
X \cdot Y \equiv \max\{X^2, Y^2\}.
\]
In the case where $X,Y \in \mathcal{E}$, this turns out to be reasonable. However, for any other use case, Breiman's lemma suggests that $X \cdot Y \equiv \max\{X, Y\}$. \todo{Perform Tauberian analysis to determine sharpness of these operations}.

\paragraph{Product of Densities.}

To handle the application of priors, it is necessary to support a product of densities operation acting on two random variables $X,Y$, denoted $X \& Y$, by
\[
p_{X \& Y}(x) = \frac{1}{\mathcal{Z}} p_X(x) p_Y(x),
\]
where $\mathcal{Z}$ is an appropriate normalizing constant and $p_X,p_Y,p_{X\& Y}$ are the densities of $X$, $Y$, and $X \& Y$, respectively.
Like addition of random variables, a simple naive bound
\[
X \& Y \equiv \min\{X, Y\},
\]
often suffices when $X,Y$ have different forms of their tails. If $X,Y \in \mathcal{E}^\rho$, $X \in \mathcal{E}_\alpha^\rho$, $Y \in \mathcal{E}_\beta^\rho$, then $X\& Y \in \mathcal{E}_{\alpha+\beta}^\rho$. Similarly, if $\rho > 1$, $X,Y \in \mathcal{L}^\rho$, $X \in \mathcal{L}_\alpha^\rho$, $Y \in \mathcal{L}_\beta^\rho$, then $X\& Y \in \mathcal{L}_{\alpha+\beta}^\rho$. The final exception lies in the $\rho \leq 1$ case, where the densities necessarily exhibit different tail behaviour to their survival functions $\mathbb{P}(|X| > x)$. Using Pareto densities as a model, we say that if $X \in \mathcal{L}_\alpha^\rho$ and $Y \in \mathcal{L}_\beta^\rho$ for $\rho \leq 1$, then $X \& Y \in \mathcal{L}_{\alpha+\beta-1}^\rho$.

\paragraph{Lipschitz Functions.}

There are many multivariate functions that cannot be readily represented in terms of the operations covered thus far. For these, it is important to specify the tail behaviour of pushforward measures under Lipschitz-continuous functions. Fortunately, this is covered by \cite[Proposition 1.3]{ledoux2001concentration}.
\begin{lemma}
For any Lipschitz continuous function $f:\mathbb{R}^d \to \mathbb{R}$ satisfying $\|f(x)-f(y)\|\leq L\|x -y\|$ for $x,y \in \mathbb{R}^d$, there is
\[
f(X_1,\dots,X_d) \equiv L \max\{X_1,\dots,X_d\}.
\]
\end{lemma}
H\"{o}lder-continuous functions can be represented as a composition of a power operation and a Lipschitz-continuous function.

\paragraph{Conditioning.}
A primary application of sampling techniques is for Bayesian inference. To cover this use case, it is necessary to prescribe a procedure to deal with \emph{conditional distributions}. For this, we introduce a very simple approach involving inverses.

Suppose that a random variable $X$ is dependent on a parameter $\theta$ and a latent random element $Z$ through a function $f$ by
\[
X = f(\theta; Z).
\]
Assuming that $f$ is invertible with respect to $\theta$, with inverse $f^{-1}$, it follows that
\[
\theta = f^{-1}(X; Z).
\]
It follows from the Doob-Dynkin lemma that the distribution of $\theta$ conditioned on $X = x$ is precisely that of $f^{-1}(x;Z)$.

\begin{example}
As a simple example, if $X \sim \mathcal{N}(\mu,\sigma^2)$, then since $X = \mu + \sigma Z$ for $Z \sim \mathcal{N}(0,1)$, $\mu = X - \sigma Z$ and so
\[
\mu \vert X = x \sim \mathcal{N}(x, \sigma^2).
\]
\end{example}

\section{New Algebra}

Consider a tail algebra for random variables with tails of the generalized gamma form:
\[
f_X(x) \sim c x^\nu e^{-\sigma x^\rho}
\]
The tail of a random variable $X$ of this form can be represented as a triple $(\nu,\sigma,\rho)$. Additionally, we require
\[
    \nu \in \mathbb{R}, \sigma > 0, \rho \in \mathbb{R}
\]
while noting that $\sigma = 0$ corresponds to the theory of regularly varying distributions.

\todo{Compare against Lemma 1.3.4 of \cite{mikosch}}
Tails of this form are  closed under addition: denoting the addition of random variables (additive convolution) by $\oplus$,
\[
(\nu_{1},\sigma_{1},\rho_{1})\oplus(\nu_{2},\sigma_{2},\rho_{2})
=\begin{cases}
\max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\} & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1\\
\left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right) & \text{ if }\rho_{1}=\rho_{2}=1\\
(\nu_{1}+\nu_{2}+1-\frac{\rho}{2},(\sigma_{1}^{-\frac{1}{\rho-1}}+\sigma_{2}^{-\frac{1}{\rho-1}})^{1-\rho},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}>1
\end{cases}
\]
\todo{Chi squared distribution has both $\nu,\rho > 0$}
\todo{Student-T tail is exactly recovered }
where $\max$ selects the heavier tail. Similarly, powers and scalar products satisfy
\[
(\nu,\sigma,\rho)^m = \bigg(\frac{\nu+1}{m} - 1, \sigma, \frac{\rho}{m}\bigg),
\]
and $c(\nu,\sigma,\rho) = (\nu,|c|^{-1} \sigma,\rho)$. 
Tails are also (almost) closed under multiplication: denoting the multiplication of random variables (multiplicative convolution) by $\otimes$,
\[
(\nu_{1},\sigma_{1},\rho_{1})\otimes(\nu_{2},\sigma_{2},\rho_{2})
=\begin{cases}
\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}-\frac{3}{2}\right),\sigma,\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0\\
\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}>0\\
(\nu_{1},0,0) & \mbox{ if }\rho_{1}\leq0,\rho_{2}>0 \\
(\max\{\nu_1,\nu_2\},0,0) & \mbox{ if }\rho_{1}=0,\rho_{2}=0
\end{cases}
\]
where $\mu=\frac{1}{|\rho_{1}|}+\frac{1}{|\rho_{2}|}=\frac{|\rho_{1}|+|\rho_{2}|}{|\rho_{1}\rho_{2}|}$ and $\sigma=\sigma(\sigma_{1}|\rho_{1}|)^{\frac{1}{\mu\rho_{1}}}(\sigma_{2}|\rho_{2}|)^{\frac{1}{\mu\rho_{2}}}$.

\subsection{Relation with regular variation}

Both our theory as well as the theory of regularly varying functions represent functions modulo
asymptotic behavior.
\begin{definition}
    A positive measurable function $f$ is \emph{regularly varying at infinity} with index $\alpha \in \mathbb{R}$ if
    \[
        \lim_{x \to \infty} \frac{f(tx)}{f(x)} = t^{\alpha}\qquad\forall t > 0
    \]
    If $\alpha = 0$, then $f$ is \emph{slowly varying}
\end{definition}
However, our theory is strictly finer in that it may interpolate between
the regularly varying theory and sub-exponential theory which offers finer control. In particular, the
generalized gamma density $e^{-\sigma x^\rho}$ is neglected by the regularly varying theory.

\begin{lemma}
    $F$ with density $f$ is a regularly varying distribution with
    index $\alpha$ iff $f \sim (\alpha, \cdot, \cdot)$
\end{lemma}

\subsection{Application: importance sampling ratio}

\cite{vehtari2015pareto} propose stabilizing importance weights using a Pareto approximation to the importance weights. Can we use our algebra to avoid the empirical estimation procedure \cite{zhang2009new} into a symbolic analysis one?

Given probability densities $p$ and $q$ over a common support, we are interested in determining the tail index of
\[
    \frac{p(X)}{q(x)},\qquad X \sim q
\]
The elementary case is when we are interested in sampling $X \sim p$ over $\mathcal{X}$ and while $p$ may be intractable a readily sample-able proposal $q$ is available. 
So long as $p \ll q$, we need not worry about division by zero.
Note $p(X)$ and $q(X)$ are $\mathbb{R}_+$-valued random variables with high correlation if $q$ is chosen well.

\textbf{Problem}: Heavy-tailed PDFs are usually not Lipschitz!

Even without heavy tails, consider $p = Exp(1)$ and $q = Exp(\theta)$. Then
$Y = p(X) = e^{-X}$ is the exponential transform of $X$ and has density 
\[
    \lvert \frac{1}{Y} \rvert e^{-\theta \log Y}
    = Y^{-\theta - 1}
\]
because $Y \geq 0$. Similarly, the density of $Z = q(X) = \theta e^{-\theta X}$ is given by
\[
\lvert \frac{1}{Z \theta} \rvert e^{(\log Z - \log \theta )}
= (\theta Z)^{-1-1}
\]
Cast into our algebra, we have that $Y = (-\theta-1,0,0)$ and $Z = (-2,0,0)$
so the reciprocal
\[
    Z^{-1} = (0, 0, 0)
\]
(TODO: improper distribution? or computation error?)

\section{Static analysis of probabilistic programs}

To conveniently express stochastic processes and automatically perform inference,
researchers oftentimes employ probabilistic programming languages. For closed-world
probabilistic programs (i.e. representing static graphical models), static analysis
may be performed prior to inference in order to inform and improve
inference through methods such as automatically marginalizing \cite{murray2018automated} 
or exploiting conjugacy relations \cite{hoffman2018autoconj}. In this section, we describe
how our heavy-tailed algebra can similarly be leveraged during static analysis of a
probabilistic program in order to detect and bound the tail behavior of heavy-tailed
random variables. Combined with \cref{thm:nmc-ergodic} and \cref{{thm:bdd-importance-ratio},
this leads to more informed inference algorithms which achieve strong guarantees even in the presence of heavy tails.

Consider the following PP:
\begin{lstlisting}
    X1 ~ Normal(0, 1)
    X2 = 2*X1
    X3 ~ Normal(0, 2)
    Y = X2 + X3
    Z1 = exp Y
    Z2 ~ StudentT(3)
    W = Z1 * Z2
\end{lstlisting}

Note that the random variables present in the program is fixed; only the values are random.
As a result, it is possible to construct an equivalent non-plated graph representation for the program
(e.g. by transforming the AST, or alternatively by tracing program execution)

\begin{tikzpicture}
  \node[latent]                               (W) {$W$};
  \factor[above=of W] {mult-W} {$\times$} {} {};
  \node[latent, above=of mult-W, xshift=-1.2cm] (Z1) {$Z1$};
  \node[latent, above=of mult-W, xshift=1.2cm]  (Z2) {$Z2$};
  \factor[above=of Z1] {exp-Z1} {$\exp$} {} {};
  \node[latent, above=of exp-Z1] (Y) {$Y$};
  \factor[above=of Y] {plus-Y} {$+$} {} {};
  \node[latent, above=of plus-Y, xshift=-1.2cm] (X2) {$X2$};
  \node[latent, above=of plus-Y, xshift=1.2cm] (X3) {$X3$};
  \factor[above=of X2] {t2-X3} {$2\times$} {} {};
  \node[latent, above=of t2-X3] (X1) {$X1$};

  \factoredge {Z1, Z2} {mult-W} {W};
  \factoredge {Y} {exp-Z1} {Z1};
  \factoredge {X2, X3} {plus-Y} {Y};
  \factoredge {X1} {t2-X3} {X2};
\end{tikzpicture}

To statically analyze the program with our algebra, the first step is to annotate all ``primitive'' distributions
(i.e. nodes given by \texttt{sample} statements ``~'', no ancestors in the PGM) with their tail classes

\begin{tikzpicture}
  \node[latent]                               (W) {$W$};
  \factor[above=of W] {mult-W} {$\times$} {} {};
  \node[latent, above=of mult-W, xshift=-1.2cm] (Z1) {$Z1$};
  \node[latent, above=of mult-W, xshift=1.2cm]  (Z2) {$Z2 \in \mathcal{L}^1_3$};
  \factor[above=of Z1] {exp-Z1} {$\exp$} {} {};
  \node[latent, above=of exp-Z1] (Y) {$Y$};
  \factor[above=of Y] {plus-Y} {$+$} {} {};
  \node[latent, above=of plus-Y, xshift=-1.2cm] (X2) {$X2$};
  \node[latent, above=of plus-Y, xshift=1.2cm] (X3) {$X3 \in \mathcal{E}^2_2$};
  \factor[above=of X2] {t2-X3} {$2\times$} {} {};
  \node[latent, above=of t2-X3] (X1) {$X1 \in \mathcal{E}^2_1$};

  % Connect the nodes
  \factoredge {Z1, Z2} {mult-W} {W};
  \factoredge {Y} {exp-Z1} {Z1};
  \factoredge {X2, X3} {plus-Y} {Y};
  \factoredge {X1} {t2-X3} {X2};
\end{tikzpicture}

Next, for each child node whose parents are completely annotated, a matching algebra rule
is applied to derive its tail class. If no matching rule exists, then our algebra is unable to
bound the rate of tail decay of that node and any of its ancestors. In this example,
``multiplication by a constant'' is applied for $X2$, ``addition'' for $Y$, ``exponential and logarithm'' for $Z1$,
and finally ``multiplication by a random variable'' for $W$.

\begin{tikzpicture}
  \node[latent]                               (W) {$W \in \mathcal{L}^1_1$};
  \factor[above=of W] {mult-W} {$\times$} {} {};
  \node[latent, above=of mult-W, xshift=-1.2cm] (Z1) {$Z1 \in \mathcal{L}^1_1$};
  \node[latent, above=of mult-W, xshift=1.2cm]  (Z2) {$Z2 \in \mathcal{L}^1_3$};
  \factor[above=of Z1] {exp-Z1} {$\exp$} {} {};
  \node[latent, above=of exp-Z1] (Y) {$Y \in \mathcal{E}^2_1$};
  \factor[above=of Y] {plus-Y} {$+$} {} {};
  \node[latent, above=of plus-Y, xshift=-1.2cm] (X2) {$X2 \in \mathcal{E}^2_1$};
  \node[latent, above=of plus-Y, xshift=1.2cm] (X3) {$X3 \in \mathcal{E}^2_2$};
  \factor[above=of X2] {t2-X3} {$2\times$} {} {};
  \node[latent, above=of t2-X3] (X1) {$X1 \in \mathcal{E}^2_1$};

  % Connect the nodes
  \factoredge {Z1, Z2} {mult-W} {W};
  \factoredge {Y} {exp-Z1} {Z1};
  \factoredge {X2, X3} {plus-Y} {Y};
  \factoredge {X1} {t2-X3} {X2};
\end{tikzpicture}

\section{Experiments}

\subsection{TODO: new toy experiments}

\begin{itemize}
    \item Chi square from squared Normal
    \item Cauchy from ratio distribution
    \item Standardized residual is Student-t distributed
    \item Kestern's law and iterated add/mult with iid random variables (?)
\end{itemize}

\subsection{Informing MCMC algorithm selection in Neal's heavy funnel}

Neal's funnel is a bivariate probability density commonly used to benchmark MCMC inference algorithm's ability to explore posterior distributions. Here, we consider a variant with anisotropic heavy tails where
\[
    z \sim t_\nu(\text{loc}=0, \text{scale}=10),
    x \mid z \sim N(\text{loc}=0, \text{scale}=e^{z/2})
\]
See \Cref{fig:neals-funnel}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.3\textwidth]{figures/nf-1.png}
    \includegraphics[width=0.3\linewidth]{figures/nf-2.png}
    \includegraphics[width=0.3\linewidth]{figures/nf-3.png}
    \caption{Samples from Neal's heavy funnel (left), which both random walk Metropolis-Hastings (middle), NUTS (TODO) and Newtonian Monte Carlo (right) aim to approximate. The presence of a heavy-tailed $z$ necessitates a more sophisticated algorithm than MALA, and our algebra is able to detect this during static analysis to inform the selection of inference routine.}
    \label{fig:neals-funnel}
\end{figure}

\subsection{MCMC kernel selection for simulating an option price}

Let $X$ be the price of an underlying asset (e.g. a stock price). Paying $P$ to purchase a put option on $X$ with strike prise $S$ has payoff

$$V = 1(X \geq S) (X - S) - P$$

Consider MCMC simulations of this put option's value $V$. See \Cref{fig:option}.
Using the algebra, we may compute $V < X$ so by theory on importance sampling ratios
the proposal distribution for $V$ can be at least as heavy as $X$ while retaining bounded
variances. This suggests a ``informed heavy tailed proposer'' of $t_{\nu=\alpha(X)}$ which
exploits the result of our algebra to ensure bounded variances.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.3\textwidth]{figures/option-1.png}
    \includegraphics[width=0.3\textwidth]{figures/option-2.png}
    \includegraphics[width=0.3\textwidth]{figures/option-3.png}
    \includegraphics[width=0.3\textwidth]{figures/option-4.png}
    \includegraphics[width=0.3\textwidth]{figures/option-5.png}
    \caption{Empirical distributions of options prices from 
    (left to right, top to bottom) ground truth, RWMH, NMC, NUTS, and
    an informed heavy tailed proposer. The informed proposer has an accurate estimate of the
    $P(V=0)$ probability compared to RWMH and NUTS and generates samples comparable to the ground truth/NMC
    all without requiring computation of the Hessian like NMC.}
    \label{fig:option}
\end{figure}

\subsubsection{Additive time series}

Other applications may involve a time series with stochastic shocks at each time-step.
If these shocks are heavy-tailed, then repeated applications of \cref{lem:sum} allow
for controlling the tail behavior of forecasts.

\todo{MC (or VI?) for distribution of $x(T)$}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.4\textwidth]{figures/ar2.png}
    \caption{Empirical distribution of samples from an AR(2) heavy-tailed additive time series.
    The predictive distribution over future states $x(T)$ has tails controlled by \cref{lem:sum}
    allowing for construction of informed proposers.
    }
\end{figure}

\subsection{Informed initializations of variational approximations}

Consider Bayesian robust linear regression $y = X \beta + \epsilon$
with $\epsilon \sim t_\nu$ and $\beta \sim N(0,\sigma)$. Then the distribution
of responses $y$ satisfies (under tail behavior ordering) $y < \min(\epsilon, X)$ so
when performing VI against $y$ a reasonable initialization is $\alpha(\min(\epsilon, X))$.
See \Cref{fig:vi-lin}.

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{figures/vi-lin.png}
    \caption{Traces of the log likelihood (left) and values of the tail parameters of
    the variational approximations (true $\nu=2$). Informed initialization at $\nu=2$ leads
    to fast training and an optimal value attaining a higher log likelihood}
    \label{fig:vi-lin}
\end{figure}

\appendix

\section{External results}
The following theorem is \cite[Corollary 1]{kasahara1978tauberian}.
\begin{theorem}[Kasahara's Tauberian Theorem]
For any random variable $X$,
\[
\limsup_{x\to\infty} \frac{1}{x} \log \mathbb{P}(|X| > x^{1/\rho}) = -\alpha,
\]
if and only if
\[
\limsup_{\lambda\to\infty} \lambda^{-\frac{\rho}{\rho - 1}} \log \mathbb{E}e^{\lambda |X|} = \frac{\rho - 1}{\rho}(\alpha \rho)^{-\frac{1}{\rho - 1}}.
\]
\end{theorem}

\bibliographystyle{plain}
\bibliography{main}

\end{document}

### 2022-08-23 ic/build.sh

@@ -1,0 +1,2 @@
+#!/bin/sh
+"/mnt/c/Program Files/MiKTeX/miktex/bin/x64/latexmk.exe" -synctex=1 -interaction=nonstopmode -pdf

### 2022-08-23 design/equivalents_nips/response.tex
Commit original repos

\documentclass{article}

\usepackage{../sty/neurips2020/neurips_2020_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{verbatim}
\usepackage{lipsum}

\begin{document}

Two reviewers rated this paper very highly.
The other reviewer was anomalously low and provided a review that
lacked substance and was completely inappropriate.  
We thus ask the senior PC member to discount this review and make a
decision based on the other two thoughtful reviews, or to find
another reviewer to provide a more appropriate review.
(We will, however, address to the extent possible the comments of this reviewer below.)

\vspace{-3mm}

\paragraph{To Reviewer \#2} Thanks for the clear accept and the
helpful suggestions. In the final version, we will clarify the
discussion of $\gamma$ as a function of $k$ in Section~1.2, and we
will also expand Section~1.3 (note that a detailed discussion of
randomized iterative optimization is provided in Appendix B).

\vspace{-3mm}

\paragraph{To Reviewer \#3} Thanks for the clear accept and the helpful
suggestions. It is true that our proofs define a high probability
event, however this is merely for the analysis. The final
statements of Theorems 1 and 2 hold absolutely, rather than with high
probability. In the final version, we will expand our discussion of
low-rank approximation metrics.

\vspace{-3mm}

\paragraph{Addressing Reviewer \#4}
Below, we demonstrate that the review is inappropriate and should be
discounted, because:\vspace{-2mm}
\begin{enumerate}
\item The reviewer completely misrepresents and fails to understand
  the stated aims of the paper.
  \vspace{-1mm}
\item The reviewer makes sweeping claims which are technically wrong
  and unsupported.
\end{enumerate}\vspace{-2mm}

First of all, while we agree that the criticisms voiced by the
reviewer apply to many NeurIPS papers, this paper is \emph{not one of
  them}. We are well aware of the TCS work on low-rank
approximation and sketching. If we were to cite all of those TCS papers, it would
include hundreds of references (we choose to cite a few reviews, as is
common when an area gets to a certain level of maturity). However, more importantly,
unlike the TCS papers, in this work we are \emph{not} interested in obtaining
worst-case approximation or concentration bounds (see, e.g., lines 4
and 27). Yet, the reviewer appears to be
confused about this and states that ``the authors
show that, given a matrix A and a sketch S, the difference of the
singular values of $A^TA$  and $A^TS^TSA$ is concentrated around the
expectation.'' This is \emph{not at all} accurate in describing our
results. Instead, our goal is to provide a precise characterization,
which goes \emph{beyond worst-case bounds}, for the expected residual
projection matrix, $\mathbb{E}[I-(SA)^\dagger SA]$ (i.e., approximating an
analytically intractable deterministic quantity with a simpler
analytically tractable expression). Thus, in the context of low-rank
approximation, our goal is not
to improve on a TCS-style approximation objective (e.g., by showing a
$1+\epsilon$ error bound relative to the best rank $k$ subspace),
but rather to express the error in a simple form as a function of the
spectrum of the data matrix.  Also, unlike standard worst-case analysis, our analysis
does \emph{not} rely on satisfying some notion of the \emph{subspace
  embedding} property, which
significantly differentiates our work from that cited by the
reviewer. Note that a subspace embedding is
neither sufficient nor necessary for many numerical implementations of
sketching \citep{blendenpik,lsrn},
or statistical results
\citep{GarveshMahoney_JMLR,dobriban2019asymptotics,yang2020reduce},
as well as in the context of iterative optimization and
implicit regularization (see Section~1.3), which are discussed in
detail in the paper.

Finally, we point out the false and unsupported claims made by the
reviewer when comparing our paper to prior work.  This
likely arises from the reviewer's confusion regarding the nature of our results. (Meanwhile, the other two reviewers describe the paper as
``very well-written'' and ``clearly written''.)
Regarding \cite{optimal-matrix-product}, Reviewer \#4 states
that ``Main result here is a generalization of theorem 1'', and then
later, regarding the submission, the reviewer claims that ``the results presented here are, in
one form or another, either known results from TCS literature, or easy
corollaries''. The latter statement is incredibly broad, completely
unsupported and simply false, so we focus on the former. Regarding the former claim, Cohen et al. do provide a
low-rank approximation guarantee for sub-Gaussian sketches. However,
this result differs from ours in several respects. First of
all, instead of analyzing $\tilde A=A(SA)^\dagger SA$ directly as a
low-rank approximation with a sketch of size
$k$ (as we do), they use a larger sketch
(e.g., of size $C k/\epsilon^2$, where $C>1$ and $\epsilon< 1$) and
consider the matrix
$\tilde A _k$, defined as the best rank $k$ approximation of $\tilde
A$. This distinction is crucial for their analysis, which relies on
showing that a sketch of size sufficiently larger than $k$ ensures
a rank $k$ subspace embedding condition. This condition is not known to hold
(at least in the worst case) if the sketch is of size $k$. Our
novel analysis completely avoids subspace embeddings (which, as the reviewer points out, are central to TCS-style analysis).  This is why we can still provide
upper/lower bounds for the low-rank approximation error in this
important case. Also, the form of our bounds is completely different than that
of Cohen et al., in that we compare the error with a certain implicit
function of the singular values of $A$, which is different from the
error of the best rank $k$ approximation (used by Cohen et al.), and
so the role of $\epsilon$ in our paper is different than in
theirs. All in all, different methods are being analyzed, different types of
bounds are obtained, and completely different analysis is
used. Thus, by all indications, the reviewer is wrong, and our result is \emph{not} a corollary or a special case
of the results of Cohen et al.

\vspace{-.7\baselineskip}
\renewcommand\refname{\normalsize References}
{\scriptsize
\setlength{\bibsep}{0pt plus 0.2ex}
\bibliographystyle{abbrvnat}
\bibliography{../pap}
}

\end{document}

### 2022-08-23 design/nystrom-neurips2020-rebuttal.tex
Commit original repos

\documentclass{article}

\usepackage{sty/neurips2020/neurips_2020_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{verbatim}
\usepackage{lipsum}
\usepackage{amsmath}
\begin{document}
We thank all the reviewers for the positive comments.

\paragraph{R1: Correctness of sr(A)-1 < k < sr(A)} 
Yes, this statement is correct. Our goal in Remark 1 is to illustrate a sharp transition in the approximation factor as $k$ approaches the stable rank from below. Remark 1(2) is an existential result i.e. there exists an $\mathbf{A}$ for which the lower bound on the approximation factor holds. 
\vspace{-2mm}
\paragraph{R1: L92 ``stable rank of A is 20" should be 40}
Here, we are referring to $\text{sr}_0(\mathbf{A})$, which is 20 and in Figure
1 it coincides with the first peak in the approximation factor.
\vspace{-2mm}
\paragraph{R1: Minor typos/comments}
Thanks for the careful read. We will make the suggested changes in the final version.
\vspace{-2mm}
\paragraph{R1: Fig 2.  Is the min $\Phi_s(k)$ line taken over $s\in\{10, 15, 25\}$ or over all values of $s\in [0,\text{rank}(\mathbf{A})]$ } The minimum is taken over all values of $s$. 
\vspace{-2mm}
\paragraph{R1: Proof of Thm 3. Importance of orthogonality of simplices}
For our setup, because of multiple simplices, the symmetry that holds for~\citet{pca-volume-sampling} breaks, and thus different subsets of the same size have different possible errors. Orthogonality, along with the choice of $\alpha_i, \beta_i$, ensures that we can pinpoint the best possible set of size $k$ and its corresponding error amongst the different choices (see proof of Lemma 4 in Appendix F). 
\vspace{-2mm}
\paragraph{R1: Other works on beyond worst case analysis}
We feel that any thorough discussion of works on ``beyond worst-case
analysis'' would have to span a very wide range of topics, which is
beyond the scope of this paper. In the final version, we will
reference \cite{roughgarden2019beyond} which provides a broader discussion on the area.
% In interest of keeping the paper precise, we did not feel that discussion of works for beyond worst case analysis on unrelated problems added anything of value to the exposition.
If the reviewer has any other works in mind which are particularly
relevant, we will definitely consider adding them.

\vspace{5mm}

\paragraph{R2: Suggested additional related works on matrix approximation}
The list of published papers on algorithms for matrix
approximation is indeed extensive. We thank the reviewer for the additional
pointers, as they are relevant to our work. In particular, we
believe that extending our results to the $\ell_p$ low-rank approximation
setting is an interesting direction for future work.

\vspace{5mm}

\paragraph{R3: Connections with overparametrization}
While we are not aware of a formal connection between the
multiple-descent phenomenon in column subset selection and
overparameterization in machine learning, both settings have to do
with obtaining a stable approximation of the data, and this stability can often be
captured by the spectral properties of the problem.
% We have established that the multiple descent phenomenon we observe is
% dependent on the spectrum of the matrix. Beyond that, there is no
% visible relationship of our work to overparameterization.

\vspace{5mm}

\paragraph{R4: Difference between k-DPP and greedy on enuite2001}
We also found the different behaviors of the approximation factor for
the two methods intriguing. One explanation for this is that the
expected approximation factor for a k-DPP is fully determined by the
eigenvalues, whereas the performance of the greedy method depends on
other characteristics of the data. However, it is not clear to us why
this is most pronounced on the eunite2001 dataset.
\vspace{-2mm}
\paragraph{R4: Correcting the effect of noisy eigenvalues}
We have not considered this, but it is indeed an
interesting question. One simple observation is that if one chooses the subset
size $k$ small enough so that the noisy tail of the spectrum consists of
only the eigenvalues with index sufficiently larger than $k$, then the effect of noise
on the approximation factor should be~minimal.

{
\setlength{\bibsep}{0pt plus 0.2ex}
\bibliographystyle{abbrvnat}
\bibliography{pap}
}
\end{document}

### 2022-08-23 design/save/colt19/rebuttal.txt
Commit original repos

### Review 1 ###
Title: Thanks for the good suggestions

>> " 'minimax optimal experimental design'  is really 'minimax optimal among all unbiased estimators', right?"
Good point! We will clarify the section and make it clear that
we are looking for minimax optimal 'unbiased' estimators

+++MW: Not done
You only changed the title of the section.
Change title seems to be too much now?
I stressed unbiasedness in abstract

>> "...why you restrict to unbiased estimators here"
The key advantage of using unbiased estimators for the
minimax definition is that we can say that the least
squares estimator w_LS is the minimum variance estimator
among them (Proposition 7). If we include all biased
estimators, then it is not clear what w_LS
should be replaced with to even define the minimax value. Also, unbiased estimators 
are closed under averaging and this immediately lets us
reduce the variance/loss. We will add more discussion to clarify this.

+++MW: not done
Added a minimal remark before Prop. 7 in Magenta

>> "how does this compare to the pseudo-sequential setting"
This setting gives more freedom to the design, so we would expect that the bounds should be at least as good. Certainly, this is a great direction for future work!

>> "I found it confusing that \hat{\bf w}({\bf y}_S ) is not written conditional on {\bf X}"
Good point! Of course the estimator function has to depend on X. Since X is fixed beforehand, and then w_hat is chosen afterwards, we treat the dependence on X as implicit in the choice of w_hat. We will clarify this point in the final version!

+++MW: not done
Sometimes with condition on X and sometimes we dont
Indeed confusing. See for example statement of Thm 4
Would be hard to fix though

hat w (y_S)   but  w_LS(y|X)

### Review 2 ###
Title: The minimax part is not a distraction

Thanks for all the comments and suggestions!
Yes - the paper is a bit overbearing and we understand that
you want us to simplify it. However Prop. 9 is not a minor
result. We will highlight this proposition more clearly.

>> "The authors define the minimax-optimal value in Definition 8 and provide some minor results in Prop 9."
>> "Such a ``minimax'' discussion acts as a distraction"
>> "make the paper less daunting (in terms of the number of results/lemmas)."
Reasons for keeping Prop. 9:
Almost all of the claims made in the paper are needed to obtain Prop. 9
and it provides BOTH upper and lower bounds for the minimax value Rk*. 
The bounds are tight up to within a constant factor. 
We will make this point clearer and the whole section more approachable in the final version.

+++MW: done

### Review 3 ###
Title: We have upper and lower bounds for the minimax value

>> "My major concern is with Sec. 1.2, and the so-called “minimax optimality” "
>> "For a typical paper, the next step would then be establishing matching upper and lower bounds of Rk*(X),"
BOTH upper and lower bounds for Rk* are given in Prop. 9 (part 2 and 3). 
The bounds are tight up to within a constant factor. 
In fact, the upper bound of C \phi/k (part 2) is a consequence of our main result, Theorem 4. 
Thus, while our main proposed estimator does not obtain exact minimax optimality 
(this appears very hard to attain), it does lead to a very strong guarantee for Rk*.

>> "The guarantees ... are additive, which are very different from the approximation
>> guarantees from Allen-Zhu et al. (2017), Wang et al. (2017) and Nikolov et al
>> (2018), deriving multiplicative guarantees."
>> "I feel this difference should be highlighted and discussed."
We completely agree. While we already point this out in the related work, we will highlight it further in the final version.
+++MW: not done
Dont know how to do this

>> "I suggest avoiding the usage of the term “minimax”, while replacing it with “minimum variance unbiased” "
>> "I suggest deriving some example scalings of \phi assuming X is randomly generated from some underlying distribution"
Good suggestions, thank you! We will address them in the final version.

### 2022-08-23 design/save/icml20/response.txt
Commit original repos

We thank all the reviewers for their comments and feedback.

## To Reviewer #2

Thanks for the (updated) positive overall evaluation.

Motivation.
The CSSP is a well-established task in the literature, where the goal is *not* to replace SVD by a faster procedure, but rather to find a subset of columns that provides a good approximation of the matrix (e.g. for interpretability). SVD does not yield such a subset of columns. Many celebrated CSSP algorithms have been slower than the SVD, e.g., Boutsidis et al. (2011), Guruswami and Sinop (2012).

Runtime.
Our subset selection procedure is a standard k-DPP, which can be significantly faster than the cost of SVD. There are several almost-linear time k-DPP sampling algorithms, including by Deshpande and Rademacher (2010), Anari et al. (2016) and Derezinski et al. (2019).
 > "Alpha is defined through opt_k, which seems to require solving SVD "
Parameter alpha is only used in the analysis. It is *not* required for the algorithm.
 > "Even with carefully tuned concentration, aren’t a total number of $sqrt{k}$ trials needed before a size $k$ subset is returned?"
There are algorithms that sample directly from a k-DPP, e.g., Anari et al (2016), without repeated trials.

Metrics.
We agree that some approximation metrics for the CSSP and Nystrom may not exhibit multiple-descent peaks. Nevertheless, the approximation ratio we use is the most commonly used metric in the CSSP literature, and therefore precise understanding of its behavior is of great relevance to the community. We also believe that the multiple-descent curve is not unique to this metric.

Matern kernel.
The polynomial eigenvalue decay rate for the Matern kernel is discussed in Section 4.2 of Rasmussen and Williams (2006), see equation (4.15). These kernels are commonly studied, see, e.g., Burt et al (2019) and Mutny et al (2019).

## To Reviewer #4

Thanks for the thorough review, and the positive feedback on our technical contributions.

Connections to double descent.
We agree that the multiple descent curve in the CSSP exhibits important differences from the double descent phenomenon in the generalization error of over-parameterized ML models. We will clarify this in the final version to avoid any confusion.
We too are suspicious of papers that try to make a tenuous connection with a recently-popular topic, but in this case the connections are quite strong, and so we believe that our paper would be incomplete without mentioning double descent. We will make sure to clarify these connections (summarized below) in the final version.
 1. The peaks exhibited by the approximation ratio of the CSSP occur when the subset size k crosses a sharp drop in the spectrum of the data. Analogously, the double descent peak in linear regression occurs when the number of model features crosses the point where the spectrum of the design matrix drops to zero (i.e., the rank of that matrix; see Section 2 of Belkin et al, 2019b).
 2. Bartlett et al (2019) show that double descent can occur not only at the point where the spectrum drops to zero, but also when it exhibits a significant drop to a small non-zero value. Moreover, the notion of stable rank used in their analysis matches our notion (see line 109 of our submission). In other words, the peaks exhibited in the CSSP closely align with the peaks of the linear regression generalization error.
 3. Derezinski et al (2019b) provide an analysis of double descent for linear regression under DPP-based data distribution, and their analysis of the bias of the minimum norm solution leads to expressions that are closely related to our formulas for the CSSP error under DPP sampling (see line 258 of our submission).

 > "I would be interested to see more methods"
In Appendix F, we provide experimental results for the greedy selection method (instead of a k-DPP), which also exhibits the error peaks predicted by our analysis.

 > "The first regime of Remark 1 seems odd."
It is true that, in this regime, a non-negligible portion of the spectrum will not be recovered even by the best rank-k approximation. Nevertheless, this regime is still relevant for certain tasks, such as in applications to clustering, where k can be chosen independently of the data spectrum, for the sake of visualization or interpretability.

 > "In intro, ... would be worth citing the many results which give 1+epsilon bicriteria approximation."
We will address this in the final version.

 > "I think it should be justified a bit more why the focus is on the case of |S| = k."
Good point. Our results can also be applied to the setting where |S| > k, and we will address this in the final version.

 > "The definition of stable rank is a bit odd"
Our definition of stable rank (which is the same as in Definition 3 of Bartlett et al, 2019), sr_s(A), reduces to a commonly used definition of stable rank for s = 0. When s > 0, then this can be viewed as the stable rank of the matrix after removing top s singular vectors.

### 2022-08-23 design/aistats/bayesian.tex
Commit original repos

--- 2022-08-23 bayesian/bayesian.tex
@@
 \newif\ifisarxiv
-\isarxivtrue
+\isarxivfalse
 
@@
   \usepackage{fullpage}
-
-  \usepackage{natbib}
 \else
   \documentclass[twoside]{article}
-  \usepackage{aistats2020}
+  \usepackage[accepted]{aistats2020}
+
+  % \special{papersize = 8.5in, 11in}
+  % \setlength{\pdfpageheight}{11in}
+  % \setlength{\pdfpagewidth}{8.5in}
 
@@
 \usepackage{caption}
-
-\input{shortdefs}
+\hypersetup{
+	colorlinks,
+	linkcolor={red!40!gray},
+	citecolor={blue!40!gray},
+	urlcolor={blue!70!gray}
+}
+
+\input{../shortdefs}
 \newcommand{\mub}{\bm{\mu}}
@@
   \else
+
+  \runningtitle{Bayesian experimental design using regularized DPPs}
   \twocolumn[
@@
     determinantal point processes}
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
+  \aistatsauthor{%
+    Micha{\l } Derezi\'{n}ski
+    \And
+    Feynman Liang
+    \And
+    Michael W. Mahoney
   }
-  % \aistatsauthor{%
-  %   Micha{\l } Derezi\'{n}ski
-  %   \And
-  %   Feynman Liang
-  %   \And
-  %   Michael W. Mahoney
-  % }
-  % \aistatsaddress {
-  %   Department of Statistics\\
-  %   University of California, Berkeley\\
-  %   \texttt{mderezin@berkeley.edu}
-  %   \And
-  %   Department of Statistics\\
-  %   University of California, Berkeley\\
-  %   \texttt{feynman@berkeley.edu}
-  %   \And
-  %   ICSI and Department of Statistics\\
-  %   University of California, Berkeley\\
-  %   \texttt{mmahoney@stat.berkeley.edu}
-  % } 
+  \aistatsaddress {
+    Department of Statistics\\
+    University of California, Berkeley\\
+    \texttt{mderezin@berkeley.edu}
+    \And
+    Department of Statistics\\
+    University of California, Berkeley\\
+    \texttt{feynman@berkeley.edu}
+    \And
+    ICSI and Department of Statistics\\
+    University of California, Berkeley\\
+    \texttt{mmahoney@stat.berkeley.edu}
+  } 
+% \aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+% \aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }
   ]
@@
 \end{align*}
-Here $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
+Here, $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
 $i\in S$.
@@
   medical imaging \citep{owen2016optimisation},
-  biological process models \citep{ryan2016optimal},
-  and materials science
-  \citep{frazier2016bayesian,ueno2016combo,terejanu2012bayesian}
+  materials science
+  \citep{frazier2016bayesian,ueno2016combo,terejanu2012bayesian},
+ and biological process models \citep{ryan2016optimal}
   all use these optimality criteria and thus stand to benefit from
@@
 infer about $\opt$ just from the spectral
-information about the problem, which is contained in the data covariance matrix
-$\Sigmab_\X=\X^\top\X\in\R^{d\times d}$.
+information about the problem, which is contained in the
+data covariance matrix~$\Sigmab_\X=\X^\top\X$.
 
@@
 \begin{figure*}[htbp]
-  \vspace{-4mm}
+  \vspace{-2mm}
   \centering
@@
 \end{tabular}
-\vspace{2mm}
+%\vspace{2mm}
 \captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
@@
   \Pr(S) = \frac{\det(
-  \X_S^\top\X_S+\A)}{\det\!\big(
- \sum_ip_i\x_i\x_i^\top+ \A\big)}
-\cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in S}(1-p_i).\label{eq:poisson-prob}
+  \X_S^\top\X_S+\A)}{\det\!\big(\!
+ \sum_ip_i\x_i\x_i^\top\!+ \A\big)}
+\cdot \!\prod_{i\in S}p_{i}\cdot\!\prod_{i\not\in S}(1\!-\!p_i).\label{eq:poisson-prob}
 \end{align}
@@
   \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
-  &=
+  &\!=
   \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
@@
 \end{align*}
-where $(*)$ was shown by \cite[Lemma
-7]{determinantal-averaging} and $\E[b_i]=p_i$.
+where $(*)$ follows  from Lemma 7 of \citet{determinantal-averaging}.
 % follows from a recently shown determinantal expectation formula.
@@
   \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &\preceq \Big(
-    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1},\label{eq:sqinv}
+  &\preceq \Big(\!
+    \sum_ip_i\x_i\x_i^\top \!+\A \Big)^{-1}\!\!,\label{eq:sqinv}
   \\
-\E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &\leq\det\!\Big(
-    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1}.\label{eq:det}
+\E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
+  &\leq\det\!\Big(\!
+    \sum_ip_i\x_i\x_i^\top\!+\A \Big)^{\!-1}\!\!.\!\!\label{eq:det}
 \end{align}
@@
     \\ &=  \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]^{\sfrac1d}\\
-       &=
+       &\leq
          \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{-\sfrac1d},
@@
 
-% \subsubsection*{Acknowledgements}
-
-% MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
-%   support of this work. 
-%   Also, MWM and MD \Red{
-%     thank the NSF TRIPODS program
-%     and FL thanks the NPSC for funding.
-%   }
-%   The authors thank Uthaipon
-%   T.~Tantipongpipat for valuable discussions.
+\subsubsection*{Acknowledgements}
+
+MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
+  support of this work. Also, MWM and MD thank the NSF TRIPODS program
+    and FL thanks the NPSC for funding.  The authors thank Uthaipon
+  T.~Tantipongpipat for valuable discussions.
 
 \bibliographystyle{apalike}
-\bibliography{pap}
+\bibliography{../pap}
 
@@
       \centering
-      \includegraphics[width=0.5\textwidth]{figs/deff_identity}\nobreak\includegraphics[width=0.5\textwidth]{figs/deff_lowrank}
+      \includegraphics[width=0.5\textwidth]{../figs/deff_identity}\nobreak\includegraphics[width=0.5\textwidth]{../figs/deff_lowrank}
       \caption{Scaled effective dimension compared to the full effective

### 2022-08-23 design/aistats/bayesian_experiments-appendix.tex
Commit original repos

This section presents additional details and experimental results omitted from
the main body of the paper. In addition to the \texttt{mg\_scale} dataset presented in
Section~\ref{s:experiments}, we also benchmarked on three other data sets
described in Table~\ref{tab:libsvm-datasets}.
\begin{table}[ht]
    \centering
    \caption{Datasets used in the experiments \citep{libsvm}.}
    \label{tab:libsvm-datasets}
    \begin{tabular}[t]{lcccc}
        \toprule
        & \texttt{mg\_scale} & \texttt{bodyfat\_scale} & \texttt{mpg\_scale} & \texttt{housing\_scale} \\
        \midrule
        $n$ & 1385 & 252 & 392 & 506 \\
        $d$ & 6 & 14 & 7 & 13\\
        \bottomrule
    \end{tabular}
\end{table}

The A-optimality values obtained are illustrated in Figure~\ref{f:obj-grid}.
The general trend observed in Section~\ref{s:experiments} of our method
(without SDP) outperforming independent sampling methods (uniform and
predictive length) and our method (with SDP) matching the performance of the
greedy bottom up method continues to hold across the additional datasets considered.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{../bayesian_figures/obj_grid.pdf}
    \caption{A-optimality values achieved by the methods compared. In all cases
        considered, we found our method (without SDP) to be superior to independent
        sampling methods like uniform and predictive length sampling. After paying the price
        to solve an SDP, our method (with SDP) is able to consistently match the performance
        of a greedy method which has been noted
        \citep{chamon2017approximate} to work well empirically.}
    \label{f:obj-grid}
\end{figure}

The relative ranking and overall order of magnitude differences
between runtimes (Figure~\ref{f:runtimes}) are also similar across the various
datasets. An exception to the rule is on $\texttt{mg\_scale}$, where we see
that our method (without SDP) costs more than the greedy method
(whereas everywhere else it costs~less).

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{../bayesian_figures/runtime_grid.pdf}
    \caption{Runtimes of the methods compared. Our method (without SDP) is
        within an order of magnitude of greedy bottom up and faster in 3 out of
        4 cases. The gap between our method with and without SDP is
        attributable to the SDP solver, making investigation of more efficient
        solvers and approximate solutions an interesting direction for future
        work.
    }
    \label{f:runtimes}
\end{figure}

The claim that $f_{\A}(\frac{k}{n} \Sigmab_\X)$ is an appropriate
quantity to summarize the contribution of problem-dependent factors
on the performance of Bayesian A-optimal designs is further evidenced in Figure~\ref{f:ratios}.
Here, we see that after normalizing the A-optimality values by this
quantity, the remaining quantities are all on the same scale and close to $1$.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{../bayesian_figures/ratios_grid.pdf}
    \caption{The ratio controlled by Lemma~\ref{l:guarantees}. This ratio converges
        to $1$ as $k \to n$ and is close to $1$ across all
        real world datasets,
        suggesting that $f_{\A}(\frac{k}{n} \Sigmab_\X)$
        is an appropriate problem-dependent scale for Bayesian A-optimal
        experimental design.
    }
    \label{f:ratios}
\end{figure}

% The choice of $\lambda=43$ is motivated by keeping 95\% of the spectral
% mass. Figure~\ref{fig:scree} shows a scree plot for \texttt{mpg\_scale}
% after applying a polynomial kernel, where we see that the majority
% of the spectral mass is contained in the top few eigenvalues.

% \begin{figure}[htbp]
%     \centering
%     % \hspace{-0.55cm}
%     \includegraphics[width=0.5\textwidth]{Figures/screeplot.pdf}
%     \caption{
%         To determine the amount of regularization $\lambda$, a scree
%         plot of the eigenvalues of the polynomially expanded
%         covariance matrix $\Phi(X)^\top \Phi(X)$.
%     }
%     \label{fig:scree}
% \end{figure}

### 2022-08-23 design/aistats/bayesian_experiments.tex
Commit original repos

\begin{figure*}
    \centering
    \hspace{-0.35cm}
    \includegraphics[width=\textwidth]{Figures/mg_combined.pdf}
     % \includegraphics[width=0.51\textwidth]{bayesian_figures/objectives-mg_scale-full.pdf}\hspace{-2mm}\nobreak\includegraphics[width=0.51\linewidth]{bayesian_figures/ratios-mg_scale.pdf}
    % \includegraphics[width=0.49\textwidth]{bayesian_figures/objectives-mg_scale-zoom.pdf}
    \caption{(left) A-optimality value obtained by the various methods on
        the \texttt{mg\_scale} dataset \citep{libsvm} with
        prior precision $\A = 10^{-5}\, \I$,\quad (right)
A-optimality value for our method (with and without SDP) divided by
$f_{\A}(\frac kn\Sigmab_{\X})$, the baseline estimate suggested by Theorem \ref{t:q1}.}
    \label{f:experiments}
\end{figure*}

\section{Experiments}\label{s:experiments}
We confirm our theoretical
results with experiments on real world data from \texttt{libsvm} datasets
\citep{libsvm} (more details in Appendix~\ref{a:experiments}).
For all our experiments, the prior precision matrix is set to $\A = n^{-1} \I$
and we consider sample sizes $k \in [d, 5d]$. Each experiment is
averaged over 25 trials and bootstrap 95\% confidence intervals are shown.
% We provide open source implementations for efficient RDPP sampling
% and reproducing our results at \Red{TODO: fill in after blind review}.
The quality of our method, as measured by the A-optimality
criterion,
\[f_{\A}(\X_S^\top \X_S) = \tr \left((\X_S^\top \X_S + \A)^{-1}\right),\]
is compared against several baselines and recently proposed methods
for A-optimal design that have been shown to perform well in
practice. Note that none of these algorithms come with theoretical
guarantees as strong as those offered by our approach. The list of
implemented methods is as follows: 
\begin{description}
    \item[Our method (with SDP)] uses the efficient algorithms
        developed in proving Theorem~\ref{t:q2} to sample
        $\DPPreg{p}(\X,\A)$ constrained to subset size $k$
        with $p = w^*$, see \eqref{eq:sdp},
        obtained using a         recently developed first order convex cone solver called Splitting
        Conical Solver \citep[SCS, see][]{o2016conic}.
        We chose SCS because it can handle the SDP constraints in
        \eqref{eq:sdp} and has provable termination guarantees, while
        also finding solutions faster \citep{o2016conic} than alternative
        off-the-shelf optimization software libraries such as SDPT3
        and Sedumi.

    \item[Our method (without SDP)] samples $\DPPreg{p}(\X,\A)$ with uniform
      probabilities $p \equiv \frac{k}{n}$.
      % Compared to uniform and predictive length
      %   sampling, there is an additional DPP sampling overhead which is independent
      %   of sample size $k$ (but does depend on $d$).

          \item[Greedy bottom-up] adds an index $i \in [n]$ to the sample $S$
        maximizing the increase in A-optimality criterion
        \citep{greedy-supermodular,chamon2017approximate}.
       %  This method posesses well-understood theoretical bounds
       %  (however they
       % require additional assumptions on the matrix $\X$) and has strong
       %  empirical performance \cite{tractable-experimental-design}.

    \item[Uniform] samples every size $k$ subset $S \subseteq [n]$
        with equal probability.

    \item[Predictive length] sampling \citep{zhu2015optimal} samples
        each row $\x_i$ of $\X$ with probability $\propto\|\x_i\|$.
\end{description}

Figure~\ref{f:experiments} reveals that our method (without SDP) is superior
to both uniform and predictive length sampling, producing designs which
achieve lower $A$-optimality criteria values for all sample sizes.
As Theorem~\ref{t:algorithm} shows that our method (without SDP) only differs
from uniform sampling by an additional DPP sample with controlled
expected size (see Lemma~\ref{l:size}), we may conclude
that adding even a small DPP sample can improve a uniformly sampled design.

Consistent with prior observations
\citep{tractable-experimental-design,chamon2017approximate}, the greedy bottom up
method achieves surprisingly good performance, despite the limited
theoretical guarantees it offers. However, if our method is used
in conjunction with an SDP solution, then we are able to match and
even slightly exceed the performance of the greedy bottom up
method. Furthermore, the overall run-time costs (see Appendix~\ref{a:experiments})
between the two are comparable. As the majority of the runtime of our
method (with SDP) is occupied by solving the SDP, an interesting future direction
is to investigate alternative solvers such as interior point methods as well
as terminating the solvers early once an approximate solution is reached.

Figure~\ref{f:experiments} (right) numerically evaluates the tightness of the
bound obtained in Theorem \ref{t:q1} by plotting the ratio:
\[\frac{f_{\A}(\X_S^\top \X_S)}{ f_{\A}(\frac{k}{n}\Sigmab_\X)}\]
for subsets returned by our method (with and without SDP). Note that the
line for our method with SDP on
Figure~\ref{f:experiments} (right) shows that the ratio never goes
below 0.5, and we saw similar behavior across all examined datasets
(see Appendix~\ref{a:experiments}). This evidence suggests that for
    many real datasets $\opt$ is within only a small constant factor
    away from $f_{\A}(\frac{k}{n}\Sigmab_\X)$, matching the upper bound of
    Theorem~\ref{t:q1}. 
% As sample size $k \to n$, this should converge towards $1$ because $\X_S^\top
% \X_S \to \X^\top \X = \Sigma_X$.
% Across all datasets we examined, we see that this ratio is close to
% its asymptotic value of $1$. Similar to how condition
% numbers are the proper problem dependent constants in matrix inversion
% \cite{nocedal2006numerical}
% and $\tr (\frac{k}{n} \X^\top \X)^{-1}$
% is the problem-dependent constant which appear in results on non-Bayesian
% experimental design \feynman{Citation for this COLT paper?}, our results here
% suggest that $f_{\A}\left(\frac{k}{n} \X^\top \X\right)$
% is the proper problem-dependent constant for Bayesian optimal experimental
% design with a $k$ cardinality constraint.
% \feynman{Does this make sense?}

### 2022-08-23 design/aistats/bayesian_experiments2.tex
Commit original repos

\section{Experiments}\label{s:experiments}

In this section we demonstrate that our method is capable
of finding better experimental designs $S \subset [n]$
on data matrices $\X$ occuring in real-world applications.
We compare our method against the following references and recently proposed
methods for A-optimal design:
\begin{description}
    \item[Greedy bottom-up] adds an index $i \in [n]$ to the sample $S$
        maximizing the increase in A-optimality criterion
        \cite{greedy-supermodular,chamon2017approximate}.
       %  This method posesses well-understood theoretical bounds
       %  (however they
       % require additional assumptions on the matrix $\X$) and has strong
       %  empirical performance \cite{tractable-experimental-design}.

    \item[Our method (with SDP)] uses the efficient algorithms
        developed in proving Theorem~\ref{t:q2} to sample
        $\DPPreg{p}(\X,\A)$ constrained to subset size $k$
        with $p = w^*$, see \eqref{eq:sdp},
        obtained using a         recently developed first order convex cone solver called Splitting
        Conical Solver (SCS) \cite{o2016conic}.
        We chose SCS because it can handle the SDP constraints in
        \eqref{eq:sdp} and has provable termination guarantees, while
        also finding solutions faster \cite{o2016conic} than alternative
        off-the-shelf optimization software libraries such as SDPT3 and Sedumi.

    \item[Our method (without SDP)] samples $\DPPreg{p}(\X,\A)$ with uniform
      probabilities $p \equiv \frac{k}{n}$.
      % Compared to uniform and predictive length
      %   sampling, there is an additional DPP sampling overhead which is independent
      %   of sample size $k$ (but does depend on $d$).

    \item[Uniform] samples every size $k$ subset $S \subseteq [n]$
        with equal probability.

    \item[Predictive length] sampling \cite{zhu2015optimal} samples
        each row $\x_i$ of $\X$ with probability $\propto\|\x_i\|$.
\end{description}
Our experiments use data matrices $\X$ obtained from \texttt{libsvm} datasets
\cite{libsvm} (more details in Appendix~\ref{a:experiments}).

\subsection{Bayesian A-optimal design}

\Red{
    We first verify empirically that our method produces experimental
    designs which achieve good (lower) optimality criteria values, confirming
    the $(1+\epsilon)$-approximation guarantee provided by Theorem~\ref{t:q2}.
    Consistent with prior work in optimal Bayesian experimental design,
    we use the commonly adopted \citep{mariet2017elementary,chaloner1984optimal}
    A-optimality criterion
    $f_{\A}(\X_S^\top \X_S) = \tr \left((\X_S^\top \X_S + \A)^{-1}\right)$
}
The prior precision matrix is set to $\A = n^{-1} \I$,
and we consider sample sizes $k \in [d, 5d]$.
Each experiment is averaged over 25 trials and bootstrap 95\% confidence
intervals are shown.
% We provide open source implementations for efficient RDPP sampling
% and reproducing our results at \Red{TODO: fill in after blind review}.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.5\textwidth,trim={0 0 10.2cm 0},clip]{../bayesian_figures/mg_combined.pdf}
     % \includegraphics[width=0.51\textwidth]{../bayesian_figures/objectives-mg_scale-full.pdf}\hspace{-2mm}\nobreak\includegraphics[width=0.51\linewidth]{../bayesian_figures/ratios-mg_scale.pdf}
    % \includegraphics[width=0.49\textwidth]{../bayesian_figures/objectives-mg_scale-zoom.pdf}
    \caption{(left) A-optimality value obtained by the various methods on
        the \texttt{mg\_scale} dataset \cite{libsvm} with
        prior precision $\A = 10^{-5}\, \I$
    }
%         ,\quad (right)
% A-optimality value for our method (with and without SDP) divided by
% $f_{\A}(\frac kn\Sigmab_{\X})$, the baseline estimate suggested by Theorem \ref{t:q1}.
    \label{f:experiments}
\end{figure}

Figure~\ref{f:experiments} (left) reveals that our method (without SDP) is superior
to both uniform and predictive length sampling, producing designs which
achieve lower $A$-optimality criteria values for all sample sizes.
As Theorem~\ref{t:algorithm} shows that our method (without SDP) only differs
from uniform sampling by an additional DPP sample with controlled
expected size (see Lemma~\ref{l:size}), we may conclude
that adding even a small DPP sample can improve a uniformly sampled design.

Consistent with prior observations
\cite{chamon2017approximate,tractable-experimental-design}, the greedy bottom up
method achieves surprisingly good performance. However, if our method is used
in conjunction with an SDP solution, then we are able to match and
even slightly exceed the performance of the greedy bottom up
method. Furthermore, the overall run-time costs (see Appendix \ref{a:experiments})
between the two are comparable. As the majority of the runtime of our
method (with SDP) is occupied by solving the SDP, an interesting future direction
is to investigate alternative solvers such as interior point methods as well
as terminating the solvers early once an approximate solution is reached.

% Figure~\ref{f:experiments} (right) displays the ratio
% $f_{\A}(\X_S^\top \X_S)\, /\, f_{\A}(\frac{k}{n}\Sigmab_\X)$ for
% subsets returned by our method (with and without SDP). Note that the
% line for our method with SDP on
% Figure~\ref{f:experiments} (right) shows that the ratio never goes
% below 0.5, and we saw similar behavior across all examined datasets
% (see Appendix~\ref{a:experiments}). This evidence suggests that for
%     many real datasets $\opt$ is within a small constant factor of
%     $f_{\A}(\frac{k}{n}\Sigmab_\X)$, matching the upper bound of Theorem~\ref{t:q1}.
% As sample size $k \to n$, this should converge towards $1$ because $\X_S^\top
% \X_S \to \X^\top \X = \Sigma_X$.
% Across all datasets we examined, we see that this ratio is close to
% its asymptotic value of $1$. Similar to how condition
% numbers are the proper problem dependent constants in matrix inversion
% \cite{nocedal2006numerical}
% and $\tr (\frac{k}{n} \X^\top \X)^{-1}$
% is the problem-dependent constant which appear in results on non-Bayesian
% experimental design \feynman{Citation for this COLT paper?}, our results here
% suggest that $f_{\A}\left(\frac{k}{n} \X^\top \X\right)$
% is the proper problem-dependent constant for Bayesian optimal experimental
% design with a $k$ cardinality constraint.
% \feynman{Does this make sense?}

\subsection{Polynomial regression with low effective dimension}

\begin{figure}[htbp]
    \centering
    % \hspace{-0.55cm}
    \includegraphics[width=0.5\textwidth]{Figures/screeplot.pdf}
    \caption{
        To determine the amount of regularization $\lambda$, a scree
        plot of the eigenvalues of the polynomially expanded
        covariance matrix $\Phi(X)^\top \Phi(X)$.
    }
    \label{fig:scree}
\end{figure}

\begin{figure}[htbp]
    \centering
    % \hspace{-0.85cm}
    \includegraphics[width=0.5\textwidth]{Figures/mse_low_d_eff.pdf}
    \caption{Mean-squared error of subsampled least squares estimates
        for a degree 2 polynomial regression on \texttt{mpg\_scale} with
        $\lambda = 43$ (so the dimension $d = 36$ is much larger than the
        effective dimension $d_\A \approx 9.36$).
    }
    \label{fig:low_d_eff}
\end{figure}

\Red{
    As experimental design is oftentimes employed as a preprocessing
    step within a predictive modelling pipeline,
    we validate the practical utility of our algorithm by demonstrating
    superior performance for row subset selection in a polynomial
    regression task.  The \texttt{mpg\_scale} dataset \cite{uci-repository},
    which originally has $7$ features, is expanded using a degree 2 polynomial
    feature expansion to yield $d=36$ features.
    We choose $\A = \lambda \I$ where $\lambda = 43$ \feynman{Should we show
    scree plot Figure~\ref{fig:scree}?}, resulting in an effective dimension $d_\A \approx 9.35 \ll
    36 = d$.  Because $d_\A$ is more than three times smaller than $d$, the new
    theoretical results in this paper (Theroem~\cite{t:q1}) provide tighter
    guarantees than previous bounds which depended only on $d$.

    To perform polynomial regression, we run our algorithm on the
    data matrix with polynomial features to produce a $S \subset [n]$
    and subsequently form the subsampled least squares estimator
    $\hat{w}_S = \X_S^\dag y_S$. The quality of $\hat{w}_S$ is evaluated
    using mean-squared error from the full least squares estimator
    \begin{align}
        MSE &= \mathbb{E}_S\left[
            \|\hat{w}_S - \hat{w}_{[n]} \|^2
        \right]
    \end{align}
    Figure~\ref{fig:low_d_eff} illustrates our results over
    $100$ trials with $95\%$ confidence intervals. Our method outperforms all
    of the independent sampling (uniform and predictive length) methods,
    demonstrating the gains resulting from joint sampling.
    Also, contrary to Figure~\ref{f:experiments} we see that in terms
    of MSE our method is superior to the greedy bottom up method in
    almost all cases. This difference may be due to
    smaller effective dimension or because of the difference between
    A-optimality and MSE, and we plan to further investigate this in
    upcoming work.
}

### 2022-08-23 design/bayesian_experiments-appendix.tex
Commit original repos

--- 2022-08-23 design/aistats/bayesian_experiments-appendix.tex
@@
+\section*{Additional Experiments}
+
 This section presents additional details and experimental results omitted from
@@
     \centering
-    \caption{Datasets used in the experiments \citep{libsvm}.}
+    \caption{\cite{libsvm} datasets used in experiments}
     \label{tab:libsvm-datasets}
@@
     \centering
-    \includegraphics[width=\textwidth]{../bayesian_figures/obj_grid.pdf}
+    \includegraphics[width=\textwidth]{bayesian_figures/obj_grid.pdf}
     \caption{A-optimality values achieved by the methods compared. In all cases
@@
         of a greedy method which has been noted
-        \citep{chamon2017approximate} to work well empirically.}
+        \cite{chamon2017approximate} to work well empirically.}
     \label{f:obj-grid}
@@
     \centering
-    \includegraphics[width=\textwidth]{../bayesian_figures/runtime_grid.pdf}
+    \includegraphics[width=\textwidth]{bayesian_figures/runtime_grid.pdf}
     \caption{Runtimes of the methods compared. Our method (without SDP) is
@@
     \centering
-    \includegraphics[width=\textwidth]{../bayesian_figures/ratios_grid.pdf}
+    \includegraphics[width=\textwidth]{bayesian_figures/ratios_grid.pdf}
     \caption{The ratio controlled by Lemma~\ref{l:guarantees}. This ratio converges
@@
 % \end{figure}
+
+In addition to \texttt{mg\_scale}, the results for the low effective dimension
+($d_\lambda$) ridge regression experiments on the remaining \texttt{libsvm}
+datasets are shown in Figure~\ref{f:small_deff_ridge_and_a_opt}.
+
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=\textwidth]{bayesian_figures/small_deff_ridge_and_a_opt.pdf}
+    \caption{Ridge regression in low effective dimension on \texttt{libsvm} datasets.}
+    \label{f:small_deff_ridge_and_a_opt}
+\end{figure}

### 2022-08-23 design/bayesian_experiments.tex
Commit original repos

--- 2022-08-23 design/aistats/bayesian_experiments.tex
@@
     \hspace{-0.35cm}
-    \includegraphics[width=\textwidth]{Figures/mg_combined.pdf}
+    \includegraphics[width=\textwidth]{aistats/Figures/mg_combined.pdf}
      % \includegraphics[width=0.51\textwidth]{bayesian_figures/objectives-mg_scale-full.pdf}\hspace{-2mm}\nobreak\includegraphics[width=0.51\linewidth]{bayesian_figures/ratios-mg_scale.pdf}
@@
     \caption{(left) A-optimality value obtained by the various methods on
-        the \texttt{mg\_scale} dataset \citep{libsvm} with
+        the \texttt{mg\_scale} dataset \cite{libsvm} with
         prior precision $\A = 10^{-5}\, \I$,\quad (right)
@@
 \section{Experiments}\label{s:experiments}
-We confirm our theoretical
-results with experiments on real world data from \texttt{libsvm} datasets
-\citep{libsvm} (more details in Appendix~\ref{a:experiments}).
-For all our experiments, the prior precision matrix is set to $\A = n^{-1} \I$
-and we consider sample sizes $k \in [d, 5d]$. Each experiment is
-averaged over 25 trials and bootstrap 95\% confidence intervals are shown.
-% We provide open source implementations for efficient RDPP sampling
-% and reproducing our results at \Red{TODO: fill in after blind review}.
-The quality of our method, as measured by the A-optimality
-criterion,
-\[f_{\A}(\X_S^\top \X_S) = \tr \left((\X_S^\top \X_S + \A)^{-1}\right),\]
-is compared against several baselines and recently proposed methods
+
+In this section we demonstrate that our method is capable of finding better
+experimental designs $S \subset [n]$ on data matrices $\X$ occuring in on real
+world data from \texttt{libsvm} datasets \citep{libsvm} (more details in
+Appendix~\ref{a:experiments}).
+We compare our method against the following baselines and recently proposed methods
 for A-optimal design that have been shown to perform well in
-practice. Note that none of these algorithms come with theoretical
-guarantees as strong as those offered by our approach. The list of
-implemented methods is as follows: 
+practice:
 \begin{description}
@@
         obtained using a         recently developed first order convex cone solver called Splitting
-        Conical Solver \citep[SCS, see][]{o2016conic}.
+        Conical Solver (SCS) \citep{o2016conic}.
         We chose SCS because it can handle the SDP constraints in
@@
         also finding solutions faster \citep{o2016conic} than alternative
-        off-the-shelf optimization software libraries such as SDPT3
-        and Sedumi.
+        off-the-shelf optimization software libraries such as SDPT3 and Sedumi.
 
@@
 
+\subsection{Bayesian A-optimal design}
+
+Consistent with prior work \citep{mariet2017elementary,chaloner1984optimal}
+in optimal Bayesian experimental design,
+we consider the task of A-optimal design:
+given a data matrix $\X \in \R^{n \times d}$, select
+$S \subset [n]$ of size $\lvert S \rvert = k$ such that
+the A-optimality
+criterion,
+\[f_{\A}(\X_S^\top \X_S) = \tr \left((\X_S^\top \X_S + \A)^{-1}\right),\]
+is small.
+All our experiments set the prior precision matrix to $\A = n^{-1} \I$
+and we consider sample sizes $k \in [d, 5d]$.
+Each experiment is averaged over 25 trials and bootstrap 95\% confidence
+intervals are shown.
+
 Figure~\ref{f:experiments} reveals that our method (without SDP) is superior
@@
 Consistent with prior observations
-\citep{tractable-experimental-design,chamon2017approximate}, the greedy bottom up
+\citep{chamon2017approximate,tractable-experimental-design}, the greedy bottom up
 method achieves surprisingly good performance, despite the limited
@@
 % \feynman{Does this make sense?}
+
+\subsection{Subsampled ridge regression}
+
+\begin{figure}[htbp]
+    \centering
+    % \hspace{-0.85cm}
+    \includegraphics[width=0.8\textwidth]{bayesian_figures/mg_scale_ridge.pdf}
+    \caption{Mean-squared error for subsampled ridge estimates
+        on \texttt{mg\_scale}
+    }
+    \label{fig:low_d_eff}
+\end{figure}
+
+As experimental design is oftentimes employed as a data reduction step within a
+predictive modelling pipeline, we validate the practical utility of our
+algorithm by demonstrating superior performance for row subset selection in a
+ridge regression task on $\texttt{mg\_scale}$.
+We chose $\A = \lambda \I$ such that $\approx 95\%$ of the spectral mass was
+contained in eigenvalues exceeding $\lambda$, resulting in an effective
+dimension $d_\A \approx 3.97 \ll 6 = d$.  Because $d_\A$ is smaller than $d$,
+the new theoretical results in this paper (Theroem~\ref{t:q1}) provide tighter
+guarantees than previous bounds which depended only on $d$.
+
+We first use the various methods to produce a design $S \subset [n]$,
+form the subsampled ridge estimator $\hat{\w}_S = (\X_S^\top \X_S + \A)^{-1}
+\X_S^\top \y_S$ (which is also the posterior mean of $\w$ given $\y_S$),
+and plots its $\ell_2$ distance from the full unsubsampled
+estimator $\hat{\w}_{[n]}$ (i.e. the posterior mean of $\w$ given $\y_{[n]}$)
+in Figure~\ref{fig:low_d_eff} ($100$ trials with $95\%$ confidence intervals).
+Our method outperforms all of the other methods compared, achieveing a lower
+relative approximation error as the sample size increases.

### 2022-08-23 design/colt.tex
Commit original repos

\newif\ifisarxiv
\isarxivfalse

\ifisarxiv
\documentclass[12pt]{sty/colt2019/colt2018-arxiv}
\else
\documentclass[final,12pt]{sty/colt2019/colt2019} % Anonymized submission
% \documentclass[12pt]{colt2019} % Include author names
\fi

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

% \title[Short Title]{Full Title of Article}
\title[Minimax experimental design]{Minimax experimental design:
  Bridging the gap between statistical and worst-case approaches to
  least squares regression}
\usepackage{times}

\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{xfrac}
\usepackage{algorithmic}
\input{shortdefs}
% Use \Name{Author Name} to specify the name.
% If the surname contains spaces, enclose the surname
% in braces, e.g. \Name{John {Smith Jones}} similarly
% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
% If the first letter in the forenames is a diacritic
% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

% Two authors with the same address
% \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
%  \Name{Author Name2} \Email{xyz@sample.com}\\
%  \addr Address}

% Three or more authors with the same address:
% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
%  \Name{Author Name2} \Email{an2@sample.com}\\
%  \Name{Author Name3} \Email{an3@sample.com}\\
%  \addr Address}

%  Authors with different addresses:
\coltauthor{\Name{Micha{\l } Derezi\'{n}ski}
  \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics, UC Berkeley 
  \AND
  \Name{Kenneth L. Clarkson}
  \Email{klclarks@us.ibm.com}\\
  \addr IBM Research -- Almaden
  \AND
  \Name{Michael W. Mahoney}
  \Email{mmahoney@stat.berkeley.edu}\\
  \addr ICSI and Department of Statistics, UC Berkeley
  \AND
\Name{Manfred K. Warmuth} \Email{manfred@ucsc.edu}\\
\addr Google Inc. Z\"urich \& UC Santa Cruz
}

\begin{document}

\maketitle

\begin{abstract}%
In experimental design, we are given a large collection of vectors,
each with a hidden response value that we assume derives from an
underlying linear model, and we wish to pick a small subset of the
vectors such that querying the corresponding responses will lead to a
good estimator of the model.  
A classical approach in statistics is to assume the responses are
linear, plus zero-mean i.i.d.~Gaussian noise, in which case the goal
is to provide an unbiased estimator with smallest mean squared error
(A-optimal design).  
A related approach, more common in computer science, is to assume the
responses are arbitrary but fixed, in which case the goal is to
estimate the least squares solution using few responses, as quickly as
possible, for worst-case inputs.   
Despite many attempts, characterizing the relationship between these
two approaches has proven elusive.  
We address this by proposing a framework for experimental 
design where the responses are produced by
an arbitrary unknown distribution.  
We show that there is an efficient randomized experimental design
procedure that achieves strong variance bounds for an
unbiased estimator using few responses  in this general model.
Nearly tight bounds for the classical A-optimality criterion, as well
as improved bounds for worst-case responses, emerge as special cases
of this result.  
In the process, we develop a new algorithm for a joint sampling
distribution called volume sampling, and we propose a new
i.i.d.~importance sampling method: inverse score sampling.  
A key novelty of our analysis is in developing new expected error
bounds for worst-case regression by controlling the tail behavior of
i.i.d.~sampling via the jointness of volume sampling.  
Our result motivates a new minimax-optimality criterion for
experimental design with unbiased estimators, 
which can be viewed as an extension of both
A-optimal design and sampling for worst-case regression. 
\end{abstract}

\begin{keywords}%
  A-optimality, 
  worst-case, 
  volume sampling,  
  minimax, 
  linear regression, 
  least squares.%
\end{keywords}

\section{Introduction}
Consider fixed design regression in $d$ dimensions, with $n\gg d$
experiments parameterized by vectors $\x_1,\dots,\x_n\in\R^d$ and the
associated real random response variables $y_1,\dots,y_n$. Suppose
that each response variable is modeled as a linear function of the
parameters plus i.i.d.~Gaussian noise:
$y_i=\x_i^\top\w^*+\xi_i$, where $\xi_i\sim\Nc(0,\sigma^2)$. 
Let $\X$ be the $n\times d$ matrix whose rows are $\x_i^\top$ 
(assumed to be full rank)
and let $\y$ be the vector of the $n$ random responses
$y_i$. Under the above standard statistical assumptions the \textit{least squares estimator}
$\wols(\y|\X)=\X^\dagger \y$ (where $\X^\dagger=(\X^\top\X)^{-1}\X^\top$ is
the Moore-Penrose pseudo-inverse) is known to be the minimum
variance unbiased estimator for $\w^*\in\R^d$.
This implies that it satisfies $\E_\y[\wols(\y|\X)]=\w^*$, while achieving the smallest possible \textit{mean squared error}:
$\text{MSE}[\wols(\y|\X)]=\E_\y\big[\|\wols(\y|\X)-\w^*\|^2\big]=\sigma^2\phi$, where
$\sigma^2$ is the magnitude of the noise and $\phi=\tr((\X^\top\X)^{-1})$ captures the relevant spectral structure of $\X$. 
To compute this estimator exactly, we have to observe all $n$ responses.

In the realm of experimental design \citep{optimal-design-book}, one asks: what if we are given all
$n$ vectors $\x_i$ but are allowed to query only $k\ll n$ of the
responses? An unbiased estimator produced under this additional
restriction will certainly be no better than $\wols(\y|\X)$ (in terms of its MSE). 
There are many experimental design criteria that have been considered.
For example, we may wish to find a weight vector that minimizes the excess mean squared error resulting from the restricted access to the responses. 
This criterion is known as an A-optimal design. 
In this model, the problem reduces to finding a
subset $S\subseteq [n]$ of $k$ experiments for which the mean squared
error of the least squares estimator is minimized. Its MSE then becomes
$\min_S\sigma^2\tr((\X_S^\top\X_S)^{-1})$, where $\X_S$ is a submatrix
with $k$ rows selected by $S$.
Other optimality criteria have been studied for selecting subset $S$,
e.g., V-optimality (which we discuss below), as well as D- and
E-optimality (which are not based on the variance of the estimator,
therefore they are not as relevant to this discussion).   
%A number of recent papers 
% such as \cite{efficient-experimental-design} gave efficient algorithms
% for finding a subset $S$ that is approximately A-optimal.

How good (in terms of the MSE) can the A-optimal subset be in general?
Not surprisingly, this will depend on the total noise of the responses,
i.e.~$\E\big[\|\xib\|^2\big]=n\sigma^2$, where $\xib\in\R^n$ is the vector of
noise variables $\xi_1,\dots,\xi_n$, as well as the structure of $\X$ described
by $\phi=\tr((\X^\top\X)^{-1})$.
The following result from numerical linear algebra %Numerical Linear Algebra (RandNLA)
shows the existence of a subset $S$ with a
good A-optimality bound as a function of its size $k$ 
which is known to be asymptotically tight for some
matrices. The resulting experimental design given in
the corollary can be computed efficiently. 
\begin{theorem}[\citeauthor{avron-boutsidis13}, \citeyear{avron-boutsidis13}]
\label{thm:avron-boutsidis}
For any full rank $\X\in\R^{n\times d}$ and $d\leq k\leq n$, there is a
subset $S\subseteq[n]$ of size $k$ s.t.~$\tr((\X_S^\top\X_S)^{-1})\leq
\frac{n-d+1}{k-d+1}\tr((\X^\top\X)^{-1})$. 
\end{theorem}
Although Theorem \ref{thm:avron-boutsidis}
% By combining Theorem \ref{thm:avron-boutsidis} with the formula for
% the MSE of the least squares estimator under the statistical model
% with i.i.d.~zero-mean noise, we immediately obtain
was originally stated as a worst-case linear algebra statement, it
easily leads to the following corollary regarding the statistical MSE.
Here $\y_S$ denotes the vector of the selected random responses.
\begin{corollary}\label{c:classic}
Given $\X\in\R^{n\times d}$ such that 
$\tr((\X^\top\X)^{-1})\!=\!\phi$ and $\epsilon>0$, there is an
experimental design $S\subseteq [n]$
of size $k\leq d+\phi/\epsilon$ s.t.~for any $\y=\X\w^*+\xib$, where $\E[\xib] = \zero$
and $\Var[\xib]=\sigma^2\I$,
\begin{align*}
  \E_\y\big[\wols(\y_S|\X_S)\big]=\w^*\quad\text{and}\quad
  \underbrace{\mathrm{MSE}\big[\wols(\y_S|\X_S)\big]}_{\leq\frac{n-d+1}{k-d+1}\sigma^2\phi} -
    \underbrace{\mathrm{MSE}\big[\wols(\y|\X)\big]}_{\sigma^2\phi} \leq \epsilon\cdot
  \underbrace{\E_\y\big[\|\xib\|^2\big]}_{n\sigma^2}.%\quad\text{and}\quad k\leq d + \phi/\epsilon.
\end{align*}
\end{corollary}
Note that the bound in Corollary \ref{c:classic} holds even without
subtracting $\MSE{\wols(\y|\X)}$, however we include it here for the sake of
consistency with the later discussion.

\subsection{Experimental design with arbitrary random responses}
% All of the above discussion strongly relies on the very restrictive
% but standard response model where the noise $\xib$ is independent, homoscedastic
% (same variance) and mean zero.
% We show that it is possible to motivate
% experimental design in a way that is agnostic to the response
% model. 
While noise $\xib$ need not be i.i.d.~Gaussian to show
Corollary~\ref{c:classic}, it still has to be
zero-mean, homoscedastic (same variances) and uncorrelated.
In this section we show that there are experimental designs for
which the MSE bound from Corollary \ref{c:classic} holds
for any (even adversarial) noise. This will allow us to propose a new
``minimax-optimality'' criterion for experimental design (in Section
\ref{s:minimax}) which can be viewed as a generalization of
A-optimality to arbitrary random responses. From now on, the only
assumption we make on the random variables $y_i$ is that they have a
finite second moment. We next redefine the optimal linear predictor $\w^*$
and the vector of noise variables $\xib$ as:\footnote{Using the fact
  that $\E_\y \big[\norm{\X\w-\y}^2\big]=\E_\y
  \big[\norm{\X\w-\E[\y]}^2\big] + \E\big[\norm{\y - \E[\y]}^2\big]$.} 
\begin{align*}
  \w^*\defeq  \argmin_\w \E_\y \big[\|\X\w-\y\|^2\big]
  	=\X^\dagger\E[\y],\qquad\xib_{\y|\X}\defeq \X\w^*-\y.
\end{align*}
Note that when the noise
  happens to have mean zero, i.e.~$\E[\xib_{\y|\X}]=\zero$, then this definition
  of $\w^*$ is consistent with the statistical setting. Having no
  knowledge of the response model means that we cannot commit to a
  particular fixed subset $S$ because those responses could be
  adversarially noisy. To avoid this, we allow randomization in the design
  procedure.
  \begin{definition}
      \label{d:rand}
A ``random experimental design'' $(S,\wbh)$ of
size $k$ consists of a \textbf{random} set $S\subseteq [n]$ of size at most $k$
and a \textbf{random} function $\wbh:\R^{|S|}\rightarrow \R^d$, which
returns an estimator $\wbh(\y_S)$.
\end{definition}
The mean squared error in this context is defined as:
  $\text{MSE}\big[\wbh(\y_S)\big] =
  \E_{S,\wbh,\y}\big[\|\wbh(\y_S)-\w^*\|^2\big]$,
so it is exactly the standard MSE, except with the expectation taken
over the randomness of both the responses and the design. 
Our main result shows that when we allow the experimental design procedure to be randomized,
the mean squared error bound
given in Corollary \ref{c:classic} for homoscedastic noise can be
recovered almost exactly for arbitrary random response vectors $\y$ 
(which includes deterministically chosen response vectors as a special case).
\begin{theorem}\label{t:mse}
Given $\X\in\R^{n\times d}$ such that 
$\tr((\X^\top\X)^{-1})\!=\!\phi$ and $\epsilon>0$, there is a
random experimental design $(S,\wbh)$
of size $k=O(d\log n+\phi/\epsilon)$ s.t.~for \textbf{any} random
response vector $\y$, 
\begin{align*}
\E_{S,\wbh,\y}\big[\wbh(\y_S)\big]=\w^*\quad\text{and}\quad
  \mathrm{MSE}\big[\wbh(\y_S)\big] - \mathrm{MSE}\big[\wols(\y|\X)\big]
  \leq \epsilon\cdot 
  \E_\y\big[\|\xib_{\y|\X}\|^2\big].%\quad\text{and}\quad k= O(d\log n + \phi/\epsilon).
\end{align*}
\end{theorem}
To put this result in context we consider several different response
models to which it applies:
\begin{enumerate}
\item \textit{A-optimal experimental design}. If we assume
independent homoscedastic zero-mean 
noise, then
  our model matches the classical A-optimal
  experimental design, except for allowing the design procedure to be
  randomized. Despite the broadness of Theorem~\ref{t:mse} it still
  offers sample complexity that is only a log factor away from that of
  Corollary~\ref{c:classic}. 
  \item \textit{Heteroscedastic regression}. We let each
    response have zero-mean noise with some unknown variance
    $\Var[\xi_i]=\sigma_i^2$. In this case, unlike existing work such as
    \cite{regularized-volume-sampling}, we bound the MSE in terms of
    $\sum_i\sigma_i^2$ rather than $n\cdot\max_i\sigma_i^2$. Our
    design achieves this without having to adaptively estimate the
    variances as done by \cite{v-optimal}.
  \item \textit{Bayesian regression}. Suppose that
    $\y=\X\w+\z$, where $\w\sim D_\w$ is a random vector 
with a prior $D_\w$ and mean $\w^*$, whereas $\z$ is a zero-mean
random noise. In this case we may wish to minimize MSE w.r.t.~$\w$ (and not
$\w^*$),
i.e., $\E_{S,\wbh,\w,\z}\big[\|\wbh(\y_S)-\w\|^2\big]$.
For this purpose we can still apply Theorem \ref{t:mse} to the response
vector $\y$ conditioned on $\w$, obtaining:
% Since $\wbh(\y_S)$ is unbiased
%     w.r.t.~any response model it is also conditionally unbiased given
%     $\w$, so we have:
    \begin{align*}
      \underbrace{\E\big[\|\wbh(\y_S)-\w\|^2\big]
      \ -\ \E\big[\|\wols(\y|\X)-\w\|^2\big]}_{
      \E\big[\E\big[\|\wbh(\y_S)-\w\|^2-\|\wols(\y|\X)-\w\|^2\ |\ \w \big]\big]} 
      \ \leq\hspace{-2mm}\underbrace{\epsilon\cdot\tr\big(\Var[\z]\big)}_{\E\big[\epsilon\,\cdot\,\E[\|\X\w-\y\|^2\,|\w]\big]}\hspace{-4mm}.
    \end{align*}
    While traditional Bayesian experimental design
    \citep[see][]{bayesian-design-review} focuses on 
    i.i.d.~Gaussian noise, our results apply to arbitrary
    zero-mean noise. A natural future direction is to extend Theorem \ref{t:mse} to
    biased estimators that take advantage of the prior information.
  \item \textit{Worst-case regression}. We let $\y$ be some arbitrary
    fixed vector $\y\in\R^n$, i.e., $\Var[\xib_{\y|\X}]=\zero$ \citep[a well-studied
    problem; see, e.g.,][]{drineas2006sampling}. Then $\wols(\y|\X)=\w^*$
    and we get: 
    \begin{align*}
      \E_{S,\wbh}\big[\|\wbh(\y_S)-\w^*\|^2\big] \leq \epsilon\cdot \|\X\w^*-\y\|^2,
    \end{align*}
the first such bound that holds: (a) ``in
    expectation'' (rather than ``with constant probability''), (b) for an unbiased
    estimator, (c) for sample size $O(\phi/\epsilon)$ (when $\epsilon$
    is sufficiently small).
\end{enumerate}
As a corollary to Theorem \ref{t:mse}, we give an additional result
which bounds the \textit{mean squared prediction error} (MSPE) instead
of MSE, defined as   $\text{MSPE}\big[\wbh(\y_S)\big] =
\E_{S,\wbh,\y}\big[\|\X(\wbh(\y_S)-\w^*)\|^2\big]$.
In many tasks, the 
performance of an estimator is evaluated 
in terms of the prediction accuracy, in which case MSPE may be a
natural metric. Note that here the sample complexity no longer depends
on the spectral parameter $\phi$ (which is replaced by $d$), just as
it happens when bounding MSPE in the classical homoscedastic setting.
\begin{theorem}\label{t:mspe}
Given a full rank $\X\in\R^{n\times d}$ and $\epsilon>0$, there is a
random experimental design $(S,\wbh)$
of size $k=O(d\log n+d/\epsilon)$ such that for \textbf{any} random
response vector $\y$, 
  \begin{align*}
    \E_{S,\wbh,\y}\big[\wbh(\y_S)\big] = \w^*\quad\text{and}\quad
    \mathrm{MSPE}\big[\wbh(\y_S)\big]-\mathrm{MSPE}\big[\wols(\y|\X)\big]\leq
    \epsilon\cdot\E_\y\big[\|\xib_{\y|\X}\|^2\big].
  \end{align*}
\end{theorem}
In the statistical setting, minimizing the MSPE is often referred to
as V-optimal design \citep[see][]{v-optimal}.
On the other hand, in worst-case regression analysis (when responses
form a fixed vector $\y\in\R^n$), the mean squared prediction error is
often replaced by the ``square loss'': $L(\w) =
\|\X\w-\y\|^2$. Theorem~\ref{t:mspe} implies a bound on the expected
square loss of the estimator $\wbh(\y_S)$:
\begin{align}
    \E_{S,\wbh}\big[L(\wbh(\y_S))\big] \overset{(*)}{=} \MSPE{\wbh(\y_S)}+L(\w^*) \leq
(1+\epsilon)\cdot L(\w^*).
    \label{e:regret}
\end{align}
where $(*)$ follows from the unbiasedness of $\wbh(\y_S)$ via the
bias-variance decomposition of the expected square loss.
The only
\textit{expected} loss bound of this kind known prior to this result
required sample size $k=O(d^2/\epsilon)$ \citep{unbiased-estimates}.

Since our experimental design is randomized, each evaluation may
produce a different result. In fact this can go to our advantage:
instead of using one design with a larger $k$ we can choose to produce
multiple independent designs with a small $k$, say
$(S_1,\wbh_1),\dots,(S_m,\wbh_m)$, and then average them. This
strategy may be preferrable in distributed settings and when data
privacy is a concern. Since all the
designs are unbiased for the random responses $\y$, it follows that:
\begin{align*}
  \mathrm{MSE}\bigg[\frac1m\sum_{t=1}^m\wbh_t(\y_{S_t})\bigg] -
  \mathrm{MSE}\big[\wols(\y|\X)\big] = 
\frac1m\Big(\mathrm{MSE}\big[\wbh_1(\y_{S_1})\big] - \mathrm{MSE}\big[\wols(\y|\X)\big]\Big),
\end{align*}
with an analogous formula also holding for the MSPE.

\subsection{Minimax-optimal experimental design with unbiased estimators}
%(and related open problems)}
\label{s:minimax}
% impervious to scaling the unknown distribution of $\y$ by any factor:
If we divide both sides of the inequality in Theorem~\ref{t:mse} by
the right-hand-side $\E_\y[\|\xib_{\y|\X}\|^2]$, we see a ratio bounded above by $\epsilon$
for all $\y$. This ratio, or rather its maximum over all $\y$, can be considered a
quality criterion for experimental designs,  to be minimized instead of only
bounded. We will call the optimum a
\emph{minimax-optimal design}. Similarly as in the standard setup, we
minimize over all \emph{unbiased} estimators.
The key difference
is that we allow the design to be randomized.
Let $\Fc$ denote the family
of \textit{all} random vectors in $\R^n$ with finite second moment.
% s.t.~the total noise $\E_\y\big[\|\xib_{\y|\X}\|^2\big]$ is
% non-zero, where recall that 
% $\xib_{\y|\X}=\X\X^\dagger\E[\y]-\y$.
\begin{definition}[unbiased estimators]
Given matrix $\X\in\R^{n\times d}$ and budget $k\in\{d,\dots,n\}$, let
$\Wc_k(\X)$ be the family of all random experimental designs $(S,\wbh)$ of
size $k$ such that: %that are unbiased w.r.t.~any response vector $\y\in\Fc$:
\begin{align*}
\E_{S,\wbh,\y}\big[\wbh(\y_S)\big]
    = \argmin_\w\E\big[\|\X\w-\y\|^2\big]
    = \X^\dagger\E[\y]\quad\text{for
  all }\ \y\in\Fc.
\end{align*}
\end{definition}
In Appendix \ref{a:minimax} we show that the least squares estimator
$\wols(\y|\X)=\X^\dagger\y$ is 
the \textit{minimum variance unbiased estimator} (MVUE) among all
such estimators with an unrestricted budget, i.e., $\Wc_n(\X)$. 
\begin{proposition}\label{p:mvue}
Given any full rank matrix $\X\in\R^{n\times d}$ and any random function
$\wbh:\R^n\rightarrow\R^d$,
\begin{align*}
\text{if}\quad
  \E_{\y,\wbh}[\wbh(\y)]=\X^\dagger\E[\y]\quad \forall_{\y\in\Fc},
  \quad\text{then}\quad
  \Var\big[\wbh(\y)\big]\succeq\Var\big[\wols(\y|\X)\big]\quad \forall_{\y\in\Fc}.
\end{align*}
\end{proposition}
It is thus natural to minimize the excess mean squared error incurred by an
unbiased estimator with a restricted budget compared to that of
$\wols(\y|\X)$. Note that if we did not introduce the unbiasedness constraint it
would be unclear what comparator should be used in place of $\wols(\y|\X)$,
which makes that constraint particularly important here.
Since we take a maximum over all response vectors in $\Fc$, we
normalize the error by the noise
$\E_\y\big[\|\xib_{\y|\X}\|^2\big]$ (equal to $n\sigma^2$ in the
classical setting).
To avoid division by zero, we exclude all fixed vectors in the
  column span of $\X$, denoted $\mathrm{Sp}(\X)\subseteq\R^n$.
\begin{definition}[minimax-optimal value]\label{d:minimax}
For a full-rank $\X\in\R^{n\times d}$ and  $d\leq k\leq n$,
define:
\begin{align*}
R_k^*(\X)&\defeq
  \min_{(S,\wbh)\in\Wc_k(\X)}\ \max_{\y\in\Fc\backslash\mathrm{Sp}(\X)}\,\frac{\mathrm{MSE}\big[\wbh(\y_S)\big]-
     \mathrm{MSE}\big[\wols(\y|\X)\big]
  }{\E_\y\big[\|\xib_{\y|\X}\|^2\big]},%\\
  % \text{where}\quad\MSE{\wbh}\ &\text{ denotes }\
  % \E\big[\|\wbh-\E[\wbh]\|^2\big]\quad\text{for any unbiased
  %                               estimator $\wbh$.}
\end{align*}
where $\MSE{\wbh}$ for any unbiased estimator $\wbh$ denotes $\E\big[\|\wbh-\E[\wbh]\|^2\big]$.
\end{definition}

\begin{proposition}[matching upper and lower bounds]\label{p:minimax-bounded}
Let $\X$ denote a full rank $n\times d$ matrix and
$\phi=\tr((\X^\top\X)^{-1})$. Then, for any $d\leq k\leq n$, we have $0\leq
R_k^*(\X)<\infty$. Furthermore:

% 1. \  For any $\X$ and $d\leq k\leq n$,  we have $0\leq
% R_k^*(\X)<\infty$;

\textnormal{1. (upper)} \ There is $C>0$ s.t.~for any $\X$ and
$k\geq C\cdot d\log n$, we have $R_k^*(\X)\leq C\cdot \phi/k$;

\textnormal{2. (lower)} \ For any $n$, $d$ and $\epsilon\!\in\!(0,1)$, there is $\X$
    s.t.~if $k^2\!<\!\epsilon nd/3$ then $R_k^*(\X)\geq
    (1-\epsilon)\cdot\phi/k$.
  %For any $\X$ and $k$ as above, $0\leq R_k^*(\X)<\infty$ and there is
% an absolute constant $C>0$ such that if $k\geq C\cdot d\log n$ then
% $R_k^*(\X)\leq C\cdot \phi/k$, where $\phi=\tr\big((\X^\top\X)^{-1}\big)$.
  \end{proposition}

\vspace{-1mm}
Part 1 of Proposition \ref{p:minimax-bounded} follows from our main result (Theorem \ref{t:mse}), whereas part 2 is an application of a
matrix inequality of \cite{avron-boutsidis13}, see details in Appendix \ref{a:minimax}. Note that if we defined $\Fc$ as the family of all
random vectors $\y$ such that the noise  $\xib_{\y|\X}$ is
i.i.d.~centered Gaussian (with any variance), then in this case the
least squares estimator would also be the MVUE, and the above
definition would become equivalent to the classical A-optimality criterion.
Even in this special case, finding an exactly optimal design is hard
(to our knowledge, NP-hardness has not been established),
although efficient approximation algorithms exist for A-optimality
(see Section~\ref{s:related-work}). Similar questions can be asked
about minimax-optimal design, however without any
restrictions on the design procedure, this task appears daunting. In
Section \ref{s:vol} we present one such restriction based on
``volume sampling'' which leads to a family of efficient
unbiased estimators that we used in Theorems \ref{t:mse} and
\ref{t:mspe}.

% A number of open questions arise for minimax-optimal experimental
% design:
% \begin{enumerate}
%   \item For which families $\Fc$ is it tractable to compute (or
%     approximate) and for which is it NP-hard?
%   \item Does the unbiasedness constraint ever change the minimax value for
%     the maximal family $\Fc_{\max}$ (all random vectors with bounded second
%     moment), or for the Gaussian family $\Fc_N(\X)$?
%   \item In Lemma \ref{l:reduction} we show that for all $\X$ and
%    designs $(S,\wbh)$ that are unbiased in $\Fc_{\max}$ we have:
%     \begin{align*}
%       \max_{\y\in\Fc_{\max}}R_k(S,\wbh\,|\,\X,\y)\ \leq\ 2\cdot \max_{\y\in\R^n}R_k(S,\wbh\,|\,\X,\y).
%     \end{align*}
%     Can the constant $2$ be improved to $1$, i.e.,~is it sufficient to
%     consider deterministic reponse vectors $\y$ when finding the
%     minimax-optimal value w.r.t.~$\Fc_{\max}$?
% \end{enumerate}

\subsection{Construction and efficiency of random experimental designs}\label{ss:RED}
The random experimental design used in Theorems
\ref{t:mse} and \ref{t:mspe} consists of two primary components:
	\vspace{-1mm}
\begin{enumerate}
  \item \textit{volume sampling:} the initial few experiments are drawn
    from a joint sampling distribution over 
    sets $S\subseteq [n]$ of size $d$ such that $\Pr(S)\propto
    \det(\X_S)^2$;
	\vspace{-1mm}
  \item \textit{i.i.d.~sampling:} the remaining $k-d$ experiments are
    sampled independently from a carefully chosen distribution
    $q=(q_1,\dots,q_n)$. 
  \end{enumerate}
	\vspace{-1mm}
  While it is mainly the i.i.d.~sampling that is responsible for 
bounding the sample size $k$, volume sampling is
  necessary for establishing both the unbiasedness and the expected
  bounds. The key novelty of our analysis is using volume sampling to
  control the MSE in the \textit{tail} of the
  distribution, and using the concentration properties of
  i.i.d.~sampling to bound it in the \textit{bulk} of the
  distribution (see Section \ref{s:proofs}). The i.i.d.~sampling
  distribution $q$ used in the proof is a mixture of uniform
  distribution with two importance sampling techniques:
  \begin{enumerate}
  \item  \textit{Leverage score sampling}: $\Pr(i)=p_i^{\mathrm{lev}}
\defeq \frac1d \x_i^\top(\X^\top\X)^{-1}\x_i$ for
  $i\in[n]$. This is a standard sampling method which has been used in
  obtaining bounds for worst-case linear regression.
  \item \textit{Inverse score sampling}: $\Pr(i)= p_i^{\mathrm{inv}}\defeq
    \frac1\phi\x_i^\top(\X^\top\X)^{-2}\x_i$ for
    $i\in[n]$. This is a novel sampling technique which is essential
    for achieving $O(\phi/\epsilon)$ sample size for small $\epsilon$.
  \end{enumerate}
As discussed earlier, having chosen a design, we may wish to
produce multiple independent samples of it, for example to construct
an averaged estimator. Thus, we break down the computational cost into
the preprocessing cost (incurred once per given matrix $\X$) and
sampling/estimation cost (incurred every time a new estimator is produced).
% The total cost of such a procedure is:
% \begin{align*}
%   \text{preprocessing}\quad+\quad m\ \cdot\ \big(\,\text{sampling}\ +\ \text{estimation}\,\big),
% \end{align*}
% where $m$ is the number of independent designs we wish to produce. 
The estimation step simply requires computing a least squares estimator
from $k$ samples, which costs $O(kd^2)$. The
preprocessing involves all the calculations necessary to construct the
sampling distributions. Both of the above importance
sampling distributions can be computed exactly in time $O(nd^2)$ or
approximately in time $O(nd\log n +d^3\log d)$ using standard sketching
techniques \citep[see][]{DMMW12_JMLR}. Once they are obtained, the sampling cost is
negligible. On the other hand, for volume sampling both preprocessing
and sampling cost can be significant.
\cite{leveraged-volume-sampling} showed that a volume sampled set of size
$d$ can be generated in time $O(d^4)$ by selecting it from a sequence
of $O(d^2)$ i.i.d.~samples from the leverage score distribution
$p^{\mathrm{lev}}$ (see Theorem 6 there). Below, we improve on that result.
% We improve on % this in the following result.
\begin{theorem}\label{t:bottom-up}
  For any $\X$ and $q$ such that $q_i\geq
  \frac12 p_i^{\mathrm{lev}}$, there is an algorithm which,
  given matrix $\X^\top\X$ and a stream of i.i.d~samples from
  $q$, returns a set $S$
  s.t.~$\Pr(S)\propto \det(\X_S)^2$, and w.p.~at least $1-\delta$
  it runs in time $O\big(d^3\log d \log\frac1\delta\big)$ using
  $O(d\log d\log\frac1\delta)$ i.i.d.~samples. 
\end{theorem}
Our algorithm improves on the best known sampling time for volume
sampling from $O(d^4)$ to $O(d^3\log d)$, which has important
implications for other applications of this distribution such as
determinantal point processes (see Section \ref{s:vol} for the proof
and further discussion). To establish correctness of the sampling,
this algorithm requires the exact computation of matrix $\X^\top\X$,
which typically costs $O(nd^2)$. The algorithm of
\cite{leveraged-volume-sampling} only requires an approximation of
this matrix, which can be computed in time $O(nd\log n + d^4\log d)$.
% Thus, the overall
% time complexity of producing $m$ random designs of size $k=O(d\log n)$
% (the smallest size for which we have bounds) is:
% \begin{align*}
% &\text{either}\qquad &O\big( nd\log n +d^4\log d\big)& &&+&& m\,\cdot\, O(d^4 + d^3\log n),&&\\
%  & \text{or}\qquad &O(nd^2)& &&+&& m\,\cdot\, O(d^3\log n).&&
% \end{align*}
Similar improvements in the preprocessing cost for our algorithm may be possible. We leave this as an open question.

\section{Related work}
\label{s:related-work}

There is a large body of related work, and we describe only that which
most informed our approach.

\paragraph{Classical experimental design.}

Many optimality criteria have been considered as functions $F(S)$ of a
subset $S\subseteq [n]$ \citep{optimal-design-pukelsheim}, assuming
that $y_i=\x_i^\top\w^*+\xi_i$ where $\xi_i\sim\Nc(0,\sigma^2)$, and
these typically have natural interpretations for the least squares
estimator.  
% Typically, those functions reflect some property of the
% covariance matrix $\X_S^\top\X_S$ of the selected input points, which have
Recent work has studied the tractability of finding an approximately
optimal subset $\widehat{S}$ of size $k$, i.e., such that
$F(\widehat{S}) \leq (1+\epsilon) \min_{S:\,|S|=k}F(S)$.  
For example, 
\cite{near-optimal-design} showed that polynomial time algorithms are
possible for many classical optimality criteria, such as A-optimality,
$F_A(S)=\tr((\X_S^\top\X_S)^{-1})$, D-optimality,
$F_D(S)=\det(\X_S^\top\X_S)^{-1}$, V-optimality,
$F_V(S)=\tr(\X(\X_S^\top\X_S)^{-1}\X^\top)$ and others, as long as
$k=\Omega(d/\epsilon^2)$;  
\cite{tractable-experimental-design} showed tractable approximability of A/V-optimality for $k=\Omega(d^2/\epsilon)$; and 
this was later improved by \cite{proportional-volume-sampling} to $k=\Omega(d/\epsilon + (\log\epsilon^{-1})/\epsilon^2)$. 
%
Robust variants of experimental design have been considered to address more general response models. 
In particular, 
\cite{robust-design} assume that the covariance matrix of the noise is
known only approximately and defines a minimax-type criterion where
the maximization goes over a neighborhood of that covariance; and  
\cite{v-optimal} use an active learning procedure to estimate the 
individual noise variances before constructing the design.  
None of these procedures, however, are truly agnostic to the response
model.

\paragraph{Subset selection for worst-case regression.}

Subset selection has been studied extensively for both statistical and worst-case regression models.
Perhaps most relevant is the work of \cite{coresets-regression}, which
showed a lower bound for any deterministically chosen subset $S$ and
function $\wbh$, when the hidden response vector $\y$ is arbitrary but
fixed.
This implies that random sampling is necessary in this setting.
% This implies that (unlike in the statistical setting) an effective approach to subset selection has to be randomized (when it has to be independent of the responses). 
% \michael{Michal, that sentence is hard to parse; rewrite it.}
% \Brown{What is the nature of this lower bound so that you can
% draw this conclusion?}
% <<<<<<< HEAD
% In the context of RandNLA, %\emph{randomized numerical linear algebra}
% %\citep[RandNLA; see][]{woodruff2014sketching,DM16_CACM},
% %it was shown by
% \cite{drineas2006sampling} showed that a random sampling algorithm
% based on the statistical leverage scores 
% constructs an estimator $\wbh(\y_S)$ which, with constant probability,
% achieves $\|\wbh(\y_S)-\w^*\|^2\leq\epsilon\cdot \|\X\w^*-\y\|^2$ by
% using $k=O(d\log d + \lambda_{\max}((\X^\top\X)^{-1})\cdot
% d/\epsilon)$ samples. 
In the context of \emph{randomized numerical linear algebra}
\citep[RandNLA; see][]{woodruff2014sketching, DM16_CACM}, it was shown by
\cite{drineas2006sampling} that a random sampling algorithm based on the statistical leverage scores
constructs an estimator $\wbh(\y_S)$ which, with constant probability, achieves $\|\wbh(\y_S)-\w^*\|^2\leq\epsilon\cdot \|\X\w^*-\y\|^2$ by using $k=O(d\log d + \lambda_{\max}((\X^\top\X)^{-1})\cdot d/\epsilon)$ samples.
%>>>>>>> 4d450329e1fb19384bb1593d44a0d9a6703beb81
The estimator we propose in Theorem \ref{t:mse} achieves the same
bound for $k=O(d\log d + \phi/\epsilon)$. 
Since 
$ %%\begin{align*}
\phi=\tr\big((\X^\top\X)^{-1}\big) \leq
  \lambda_{\max}\big((\X^\top\X)^{-1}\big)\cdot d \leq   \phi\cdot d  ,
$ %%\end{align*}
our result is better by up to a factor of $d$.

% As we
% mentioned in discussing Theorem \ref{t:mspe}, in the worst-case
% regression it is often desirable to bound the squared prediction loss of the
% obtained estimator, i.e., $L(\wbh)=\|\X\wbh-\y\|^2$. In this case,

\paragraph{Statistical versus algorithmic approaches.}

% Since it considers algorithms that involve random
% sampling~\citep{Mah-mat-rev_JRNL, DM16_CACM, RandNLA_PCMIchapter_TR},
% RandNLA provides a good venue to address the statistical versus
% worst-case approaches we consider. 
 \cite{ping-ma2014} and \cite{garvesh2015} were the first to 
 consider statistical guarantees (such as A-optimality) that can be
 obtained by sampling methods developed for RandNLA (primarily leverage
 score sampling), contrasting them with some common worst-case guarantees.  
However, these works treat those two settings separately (in particular, the statistical setting is limited to i.i.d.~Gaussian noise), rather than putting them under one umbrella of minimax experimental design, as we do.
Subsequently, \cite{chen2017condition} showed loss bounds for
worst-case regression which extend to a randomized response model
that is comparable to ours. 
They give a randomized estimator 
({\em not unbiased}) that with constant probability
achieves the following bound on the square loss:
$L(\wbh(\y_S))\leq(1+\epsilon)L(\w^*)$, for sample size
$k=O(d/\epsilon)$, where $L(\w)=\|\X\w-\y\|^2$. % and they also show
% that this is optimal.
%However, their estimator is not
%unbiased and the bound is not in expectation, in contrast to what we
%we obtained in \eqref{e:regret} as a corolllary of Theorem \ref{t:mspe}.
%(and so their result does not imply a bound on the MSPE). 
In contrast we obtain an {\em unbiased} estimator achieving
the same bound \emph{in expectation} with only slightly larger
sample size of $k=O(d\log n+d/\epsilon)$.

\paragraph{Constant probability versus unbiased expectations.}
Unlike our Theorems \ref{t:mse} and \ref{t:mspe}, most results in
RandNLA are stated to hold with high or constant probability
\citep{woodruff2014sketching, DM16_CACM, RandNLA_PCMIchapter_TR} as opposed to in 
expectation, and they do not provide unbiased estimators, which often
makes them incomparable to statistical approaches.
In fact, expected bounds are often impossible for these
techniques \citep[e.g., for leverage score sampling; see][]{unbiased-estimates-journal}.
% Contrast this with our Theorem~\ref{t:mse}, which claims $d\log n$ instead of $d\log d$ only to achieve the bound in expectation as opposed to with constant probability (which is not possible for leverage score sampling).
% \michael{Michal, that sentence is hard to parse; rewrite it.}
Unbiased estimators were first introduced to worst-case regression by
\cite{unbiased-estimates}, who gave the first \textit{expected}
square loss bound for sample size of $k=O(d^2/\epsilon)$ via volume
sampling.  
Subsequently, \cite{leveraged-volume-sampling} demonstrated an
unbiased estimator with a \textit{constant probability} loss bound for
sample size $k=O(d\log d+d/\epsilon)$. Our result builds on the latter
by obtaining an unbiased estimator with an \textit{expected} loss
bound for $k=O(d\log n+d/\epsilon)$.

\section{Rescaled volume sampling}\label{s:vol}

We now discuss the sampling distribution introduced by
\cite{leveraged-volume-sampling}, based on earlier work by
\cite{avron-boutsidis13}, that allows for constructing
unbiased least squares estimators. 
\begin{definition}%[\cite{leveraged-volume-sampling}]
  \label{d:vs}
  Given full rank matrix $\X\in\R^{n\times d}$ and a distribution
  $q=(q_1,\dots,q_n)$ s.t.~$q_i> 0$ for all $i\in[n]$, we define
$q$-rescaled volume sampling of size $k\geq d$, written
$\Vol_q^k(\X)$, as a distribution 
over index sequences $\pi=(\pi_1,\dots,\pi_k)\in[n]^k$ such that:
\vspace{-4mm}
  \begin{align*}
    \Pr(\pi) =
    \frac{\det\!\big(\X^\top\S_\pi^\top\S_\pi\X\big)}{\frac{d!}{k^d}{k\choose
    d}\det(\X^\top\X)}\prod_{i=1}^kq_{\pi_i},\qquad\quad\text{where}\qquad\quad
    \S_\pi = \begin{bmatrix}\frac1{\sqrt{k q_{\pi_1}}}\e_{\pi_1}^\top\\
      \vdots\\\frac1{\sqrt{k q_{\pi_k}}}\e_{\pi_k}^\top\end{bmatrix}
    \in\R^{k\times n}. 
  \end{align*}
\end{definition}
\vspace{-1mm}
It is easy to see that for $k=d$, the matrix $\S_\pi\X$ is square 
and thus
\begin{align*}
  \det(\X^\top\S_\pi^\top\S_\pi\X)
    =\det(\S_\pi\X)^2=\frac{\det(\X_\pi)^2}{d^d\prod_i q_{\pi_i}},
\end{align*}
where $\X_\pi$ selects the rows indexed by $\pi$ from $\X$. So the
distribution $\Vol_q^d(\X)$ is the same for every $q$. For this
reason, we will write it simply as $\Vol^d(\X)$.
We mention the following
recently shown results regarding rescaled volume sampling which we use
later in the proofs. 
\begin{lemma}[\citeauthor{leveraged-volume-sampling}, \citeyear{leveraged-volume-sampling}]\label{l:expectations}
For any $\X$, $q$ and $k$ as in Definition \ref{d:vs}, if
$\pi\sim\Vol_q^k(\X)$, then  
\begin{align}
  \E\big[(\S_\pi\X)^\dagger \S_\pi\big] &= \X^\dagger,\label{eq:unbiased}\\
  \E\big[(\X^\top\S_\pi^\top\S_\pi\X)^{-1}\big]
  &\preceq \frac k{k\!-\!d\!+\!1}(\X^\top\X)^{-1}.\label{eq:square-inverse}
\end{align}
\end{lemma}
The following random experimental design emerges as a natural
candidate for proving Theorem \ref{t:mse}:
\begin{align}
  S = \{\pi_1\}\cup\dots\cup\{\pi_k\},\quad 
    \wbh(\y_S) = (\S_\pi\X)^ \dagger\S_\pi\y
    ,\quad\text{where }\pi\sim\Vol_q^k(\X).\label{eq:design}
\end{align}
Note that since sequence $\pi$ may include repetitions whereas set $S$
may not (it is not a multi-set), the function 
$\wbh(\y_S)$ has to depend not only on $\y_S$ but also on the multiplicities of each
response in sequence $\pi$. Thus both set $S$ and function $\wbh(\cdot)$ 
in this design are in fact randomized and satisfy Definition \ref{d:rand}. 
%
Lemma \ref{l:expectations} shows that this design 
is unbiased for any $\y$, leading to a restricted notion of 
minimax-optimality which provides an upper-bound on $R_k^*(\X)$ (proof in
Appendix \ref{a:minimax}):
\begin{lemma}\label{l:reduction}
  Let $\Vc_k(\X)$ consist of all random experimental designs
  based on $q$-rescaled volume sampling as in \eqref{eq:design},
  parameterized by  distribution $q$, and let $\mathrm{Sp}(\X)$ be the
  column span of $\X$. Then: 
\begin{align*}
R_k^*(\X)\leq \min_{(S,\wbh)\in\Vc_k(\X)}\ 
  \max_{\y\in\R^n\backslash \mathrm{Sp}(\X)}\frac{\E_{S,\wbh}\big[\|\wbh(\y_S)-\wols(\y|\X)\|^2\big]}{\|\X\,\wols(\y|\X)-\y\|^2}.
  % =\min_q \big\|\Var\big[(\S_\pi\X)^\dagger\S_\pi\big]\big\|,
\end{align*}
%where $\Var[\M]=\E[(\M-\E[\M])^\top(\M-\E[\M])]$.
\end{lemma}
Even for the restricted minimax-optimality, finding the exact or even
approximate optimum $q^*$ is open. However,
in Section \ref{s:proofs} we bound the restricted minimax value by
selecting a particular distribution $q$ and utilizing the following
decomposition of $q-$rescaled volume sampling in the analysis. % of size $k$ into a
% volume sample of size $d$ and an i.i.d.~sample of size $k-d$ from $q$. 
\begin{lemma}[\citeauthor{correcting-bias}, \citeyear{correcting-bias}]\label{l:augmenting}
For any $\X$, $q$ and $k$ as in Definition \ref{d:vs}, let
$\pi\sim\Vol^d(\X)$ and $\tilde{\pi}_1,\dots,\tilde{\pi}_{k-d}\simiid
q$. Finally let $\sigma$  
be a permutation of $(1,\dots,k)$ drawn uniformly at random. Then:
\begin{align*}
  \sigma\big(\pi_1,\dots,\pi_d,\tilde{\pi}_1,\dots,\tilde{\pi}_{k-d}\big)\sim
  \Vol_q^k(\X). 
\end{align*}
\end{lemma}

\begin{wrapfigure}{r}{0.45\textwidth}
  \vspace{-6mm}
  \centering
\begin{minipage}{0.45\textwidth}
\floatname{algorithm}{\small Algorithm}
\begin{algorithm}[H] 
    \caption{\small (Bottom-up) volume sampling}\label{alg:bottom-up}
  \begin{algorithmic}[0]
    \STATE \textbf{input:} $\X\in\R^{n\times d}$, \ $q$, \ 
    $\A_1=(\X^\top\X)^{-1}$ 
    \STATE \textbf{output:} $\pi\sim\Vol^d(\X)$
    \STATE \textbf{for }$i=1..d$
    \STATE \quad\textbf{repeat}
    \STATE  \quad\quad Sample $\pi_i \sim q$
    \STATE \quad\quad Sample $a\sim
    \text{Bernoulli}\Big(\frac{\x_{\pi_i}^\top\A_i\x_{\pi_i}}{2d \,q_{\pi_i}}\Big)$
    \STATE \quad\textbf{until} $a=1$
    \STATE \quad$\A_{i+1} \leftarrow \A_i -
    \frac{\A_i\x_{\pi_i}\x_{\pi_i}^\top\A_i}{\x_{\pi_i}^\top\A_i\x_{\pi_i}}$
    \STATE \textbf{end for}
    \RETURN $\pi_1,\dots,\pi_d$
  \end{algorithmic}
\end{algorithm}
\end{minipage}
\vspace{-4mm}
\end{wrapfigure}
% Lemma \ref{l:augmenting} shows that we can decompose a
% $q-$rescaled volume sample of size $k$ into a volume sample of size $d$ and an
% i.i.d.~sample of size $k-d$ from $q$.
If distribution $q$
is sufficiently close to the leverage score sampling distribution
$p^{\mathrm{lev}}$, then even the initial volume sample of size $d$ can
be \textit{selected out of} an i.i.d.~sample of size $O(d\log d)$ as
shown in Algorithm \ref{alg:bottom-up}. This algorithm is a new
implementation of a classical method for sampling from a so-called
elementary determinantal point process, due to
\cite{dpp-independence}. To our knowledge, the best previously known
runtime for this method was $O(nd^2)$ for each produced volume sample
(see \cite{dpp-coreset}), whereas the runtime of this implementation is
$O(d^3\log d)$ (in addition to a preprocessing step which involves
computing distribution $q$ and matrix $(\X^\top\X)^{-1}$). Note that
for some applications of volume sampling, such as determinantal point
process sampling, one can often assume that $\X^\top\X=\I$ 
\citep[see][]{dpp-intermediate}, in which case the preprocessing becomes much
cheaper than $O(nd^2)$. We now prove Theorem
\ref{t:bottom-up} by establishing correctness and runtime of Algorithm
\ref{alg:bottom-up}.

\begin{proofof}{Theorem}{\ref{t:bottom-up}}
Since $\A_i\preceq (\X^\top\X)^{-1}$, then $\x_{\pi_i}^\top\A_i\x_{\pi_i}/(2d\, q_{\pi_i})
\leq p_{\pi_i}^{\mathrm{lev}}/(2q_{\pi_i})\leq 1$ is a valid
  Bernoulli probability. We start with the proof of correctness, which
  is an adaptation of the one given by \cite{dpp-independence}.
% Let $\U=\X(\X^\top\X)^{-\sfrac12}$. For any set $S$ of size $d$ we
% have $\det(\U_S)^2=\det(\X_S)^2\det\!\big((\X^\top\X)^{-1}\big)$ so
% $\Vol^d(\U)=\Vol^d(\X)$. Since $\U^\top\U=\I$, w.l.o.g.~we can assume
% that $\A_1=(\X^\top\X)^{-1}=\I$.
For any $i,j$ define
$\u_j^{(i)}=\A_{i}^{\sfrac12}\x_{j}$. The
marginal probability of sampling $\pi_{i+1}$ conditioned on previous steps
is proportional to $\|\u_{\pi_{i+1}}^{(i+1)}\|^2$, which can be written as:
\begin{align*}
  \|\u_{\pi_{i+1}}^{(i+1)}\|^2&=  \x_{\pi_{i+1}}^\top\A_{i+1}\x_{\pi_{i+1}}
  =
  \x_{\pi_{i+1}}^\top\A_{i}^{\sfrac12}\bigg(\I -
    \frac{\A_{i}^{\sfrac12}\x_{\pi_i}\x_{\pi_i}^\top\A_{i}^{\sfrac12}}
    {\x_{\pi_i}^\top\A_{i}\x_{\pi_i}}\bigg)\A_{i}^{\sfrac12}\x_{\pi_{i+1}}\\
  &=\u_{\pi_{i+1}}^{(i)\top}\bigg(\I-\frac{\u_{\pi_i}^{(i)}\u_{\pi_i}^{(i)\top}}
    {\|\u_{\pi_i}^{(i)}\|^2}\bigg)\u_{\pi_{i+1}}^{(i)}
    =\big\|\P_i\u_{\pi_{i+1}}^{(i)}\big\|^2,
    \quad\text{where }\ \P_i=\I-\frac{\u_{\pi_i}^{(i)}\u_{\pi_i}^{(i)\top}}
    {\|\u_{\pi_i}^{(i)}\|^2}
\end{align*}
is a projection onto the $(d-1)$-dimensional
subspace of $\R^d$ orthogonal to $\u_{\pi_i}^{(i)}$. We conclude that
vectors $\u_{j}^{(i)}$ are obtained from $\u_{j}^{(1)}=(\X^\top\X)^{-\sfrac12}\x_j$ by
repeatedly \textit{projecting away} the points that were already
sampled. This means that since $\U=\X (\X^\top\X)^{-\sfrac12}$
satisfies $\U^\top\U=\I$, we have: 
\begin{align*}
  \sum_{j=1}^n\|\u_{j}^{(i+1)}\|^2
  =\tr\bigg(\sum_{j=1}^n\u_j^{(1)}\u_j^{(1)\top}\cdot\prod_{t=1}^{i}\P_t\bigg)
  = \tr\bigg(\U^\top\U\cdot\prod_{t=1}^{i}\P_t\bigg)= d-i.
\end{align*}
We can now write the probability of sampling a sequence
$\pi_1,\dots,\pi_d$ as:
\begin{align*}
  \Pr(\pi) = \prod_{i=1}^d\frac{\|\u_{\pi_i}^{(i)}\|^2}{d-i+1} =
  \frac{\det(\U_\pi)^2}{d!} = \frac{\det(\X_\pi)^2}{d!\det(\X^\top\X)},
\end{align*}
which follows because $\det(\U_\pi)^2$ is the squared volume spanned
by the vectors $\u^{(1)}_{\pi_1},\dots,\u^{(1)}_{\pi_d}$ and it is obtained as a
series of applications of the ``base $\times$ height'' formula. To
bound the runtime, we note that the expected acceptance probability in
the $i$th step of Algorithm \ref{alg:bottom-up} is:
\begin{align*}
  \sum_{j=1}^nq_j\cdot \frac{\x_{j}^\top\A_i\x_j}{2d
  \,q_j}=\frac1{2d}\sum_{j=1}^n\|\u_{j}^{(i)}\|^2 = \frac{d-i+1}{2d}.
\end{align*}
Thus, the expected total number of trials of rejection sampling
throughout the algorithm is:
\begin{align*}
  \sum_{i=1}^d\frac{2d}{d-i+1} = 2d\sum_{i=1}^d\frac1i\leq 2d\big(\ln(d)+1\big).
\end{align*}
Standard tail bounds for a sum of geometric random variables show that
with probability at least $1-\delta$ the number of rejection sampling
trials is $O(d\log d\log\frac1\delta)$. Each trial costs
$O(d^2)$, as does updating the matrix $\A_i$, which concludes the proof.
\end{proofof}

\section{Proof of Theorem \ref{t:mse}}
\label{s:proofs}
In this section we use $\wbh$ and $\wols$ as shorthands for
$\wbh(\y_S)$ and $\wols(\y|\X)$. % Also, unless otherwise specified, the
% expectations are taken over all random variables (such as $S$ and $\y$).
% First, we point out that the random response vector $\y$ can be reduced
% to a fixed vector $\y$ in the analysis while losing only a constant factor in
% the error bound.
To prove the error bound in Theorem \ref{t:mse} we will invoke Lemma
\ref{l:reduction},
thereby restricting ourselves to a fixed response vector $\y\in\R^n$,
in which case $\w^*=\wols$, and a volume sampled random design
as discussed in the previous section.
% W.l.o.g.~we can prove the error
% bound in Theorem \ref{t:mse}  assuming that $\y$ is equal to some
% fixed vector $\y\in\R^n$ (in which case $\w^*=\wols$), and then apply
% Lemma \ref{l:reduction}, losing only a factor of 2 in the bound
% (which goes into the big-O notation). 
The construction in our proof uses leverage scores and inverse scores, as
discussed in Subsection \ref{ss:RED}.
\begin{definition}
Given full rank matrix $\X$, its $i$th \textbf{ leverage score} is
defined as $l_i(\X)\defeq \x_i^\top(\X^\top\X)^{-1}\x_i$,
and its $i$th \textbf{inverse score} as  $v_i(\X) \defeq \x_i^\top(\X^\top\X)^{-2}\x_i$.
\end{definition}
The key challenge in obtaining the result
is that standard techniques developed for
i.i.d.~sampling \citep[see, e.g., ][]{drineas2006sampling} only show the
least squares error bounds with constant probability. 
Such bounds do not suffice to show an expected bound because
  we do not have  control over what happens in the \textit{failure 
    event} (where the expectation may be unbounded). In fact, an
  expected bound of this type is not possible 
  for any i.i.d.~sampling 
  \citep[see Proposition 11 in][]{unbiased-estimates-journal}. Our key contribution is to
  define an event $A$ s.t.:
  \begin{enumerate}
    \item if $A$ occurs, then we can show a strong expected
      bound relying on i.i.d.~sampling techniques,
    \item if $A$ fails to occur, a weaker bound still holds
      because of the jointness of volume sampling.
    \end{enumerate}
    Crucially, the probability of failure will be exponentially small,
    thus allowing us to obtain the desired result. This technique is
    described in the proof  of the following key lemma.
  \begin{lemma}\label{l:key}
  %   For any $\alpha\in[0.5,0.75]$, define distribution  $q(\alpha) =
  % \alpha \,\big(0.5\cdot p^{\mathrm{uni}}+0.5\cdot p^{\mathrm{inv}}\big) 
  % \,+\,(1-\alpha)\,p^{\mathrm{lev}}$, where $p_i^{\mathrm{uni}}= 1/n$,
  % $p_i^{\mathrm{inv}}= v_i(\X)/\phi$ and $p_i^{\mathrm{lev}} =
  % l_i(\X)/d$.
  There is $C>0$ s.t.~for any full rank
matrix  $\X\in\R^{n\times d}$, if
  $\pi\sim\Vol_{q(\alpha)}^k(\X)$ where
  \begin{align*}
 q(\alpha) &=
  \alpha \,\big(0.5\cdot p^{\mathrm{uni}}+0.5\cdot p^{\mathrm{inv}}\big) 
  \,+\,(1-\alpha)\,p^{\mathrm{lev}}  %\frac{l_i(\X)}{d} 
  \quad\text{for}\quad \alpha\in\big[0.5,0.75\big],\\
  \text{with}\quad p_i^{\mathrm{uni}}
     &= 1/n,\quad
 p_i^{\mathrm{inv}}= v_i(\X)/\phi,
\quad	p_i^{\mathrm{lev}} = l_i(\X)/d,\ \text{ and }\ \phi=\tr\big((\X^\top\X)^{-1}\big),
  \end{align*}
then for any $k\geq d+ C\max\{ d\log n , \phi/\epsilon\big\}$ and an arbitrary
vector $\xib\in\R^n$ we have
%  $k\geq d+ C\max\{ d\log n , \phi/\epsilon\big\}$, then
%   Suppose that $\pi\sim\Vol_{q(\alpha)}^k(\X)$, with sample size
%   $k\geq d+ C\max\{ d\log n , 
% \phi/\epsilon\big\}$ and
% \begin{align*}
%  q(\alpha) &=
%   \alpha \,\big(0.5\cdot p^{\mathrm{uni}}+0.5\cdot p^{\mathrm{inv}}\big) 
%   \,+\,(1-\alpha)\,p^{\mathrm{lev}}  %\frac{l_i(\X)}{d} 
%   \quad\text{for}\quad \alpha\in\big[0.5,0.75\big],\\
%   \text{where}\quad p_i^{\mathrm{uni}}
%      &= 1/n,\quad
%  p_i^{\mathrm{inv}}= v_i(\X)/\phi,
% 	\text{ and }
% 	p_i^{\mathrm{lev}} = l_i(\X)/d.
% \end{align*}
% Also, let $\S_\pi$ be defined as in Definition \ref{d:vs}. Then, for
% any vector $\xib\in\R^n$, we have
\begin{align*}
 \E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\big] \leq \frac\epsilon 8
  \|\xib\|^2 +4\, \big\|\X^\dagger\E[\S_\pi^\top\S_\pi]\xib\big\|^2.
\end{align*}
\end{lemma}
\begin{proof}
%   Let $\pi\sim\Vol_q^k(\X)$ be a volume sampled sequence with sample size $k\geq d+ C\max\{ d\log n ,
% \phi/\epsilon\big\}$ and
% \begin{align*}
%  q &=
%   \alpha \,\big(0.5\cdot p^{\mathrm{uni}}+0.5\cdot p^{\mathrm{inv}}\big) 
%   \,+\,(1-\alpha)\,p^{\mathrm{lev}}  %\frac{l_i(\X)}{d} 
%   \quad\text{for}\quad \alpha\in\big[0.5,0.75\big],\\
%   \text{where}\quad p_i^{\mathrm{uni}}
%      &= 1/n,\quad
%  p_i^{\mathrm{inv}}= v_i(\X)/\phi,
% 	\text{ and }
% 	p_i^{\mathrm{lev}} = l_i(\X)/d
% \end{align*}
% %\ken{put in definition of lev and inv. scores}
% Recall that $\phi=\tr((\X^\top\X)^{-1})=\|\X^\dagger\|_F^2$. 
% We will
% use the unbiased estimator defined in \eqref{eq:design}, i.e.,
% $\wbh=(\S_\pi\X)^\dagger\S_\pi\y$.
Observe that we chose $q(\alpha)$ as a
mixture of three distributions in such a way that each of them
has at least 0.25 weight in the mixture. Lemma
    \ref{l:augmenting} allows us to decompose sample $\pi$ into
    the volume part, $\pi_{[d]}=(\pi_1,\dots,\pi_d)\sim \Vol^d(\X)$,
    and the i.i.d.~part, 
    $\tilde{\pi}=(\pi_{d+1},\dots,\pi_k)\sim q $ (technically, this
    requires reordering the sequence $\pi$). We now define an event $A$ as a variant of the so-called
    \textit{subspace  embedding} condition:
    \begin{align*}
      \text{event $A$ holds iff }\quad
\frac1k\sum_{i=d+1}^k\frac1{q_{\pi_i}}\x_{\pi_i}\x_{\pi_i}^\top\
      \succeq\ \frac12\,\X^\top\X.  
    \end{align*}
    Note that $A$ is  defined only over the i.i.d.~samples
    $\tilde{\pi}$ and is therefore completely independent of the volume sample
    $\pi_{[d]}$.
% Recall that by Lemma \ref{l:reduction} w.l.o.g.~we can assume that $\y$
% is equal to some fixed 
% $\y\in\R^n$, so $\w^*=\X^\dagger\y$ and we need to
% bound the mean squared error in terms of the noise residual, i.e.,
% $\xib=\X\w^*-\y$. For the sake of the proof, we define a ``rescaled
% model residual'' vector, a function of some $p=(p_1,\dots,p_n)$:
% \begin{align*}
%   \Delta_\pi^{p}\defeq \overbrace{(\X^\top\S_\pi^\top\S_\pi\X)^{-1}
%         \X^\top\S_\pi^\top\S_\pi}^{(\S_\pi\X)^\dagger\S_\pi}\D\overbrace{(\X\w^*-\y)}^{\xib}
% ,\qquad\text{where}
%   \quad\D=\text{diag}\Big(\frac{q_i}{p_i}\Big).
% \end{align*}
% Note that for $p=q$ (i.e.~when $\D=\I$) the square norm of the residual is
% $\|\Delta_\pi^q\|^2 = \|\w^*-\wbh\|^2$. Ultimately, this is what
% we wish to bound, however for now we will bound the expected squared
% norm of any rescaled residual.
We start by decomposing the expectation
into two terms: 
    \begin{align}
      \E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\big] =
      \Pr(A)\,\E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\ |\,A\big] + \Pr(\neg
      A)\,\E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\ |\,\neg A\big].\label{eq:terms}
    \end{align}
    % We bound the first term by shifting to
    % an orthogonalized matrix
    % $\U=\X(\X^\top\X)^{-\sfrac12}$. 
To bound the first term we decompose the squared norm into
two factors 
    \begin{align*}
\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2
      &= \big\|(\X^\top\S_\pi^\top\S_\pi\X)^{-1}
        \X^\top\S_\pi^\top\S_\pi\xib\big\|^2
    \\ & \leq \big\|(\X^\top\S_\pi^\top\S_\pi\X)^{-1}\X^\top\X\big\|^2
        \cdot\big\|(\X^\top\X)^{-1}\X^\top\S_\pi^\top\S_\pi\xib\big\|^2.
    \end{align*}
   When event $A$ occurs, then the first factor can be easily bounded by 4, because
   \begin{align*}
(\X^\top\S_\pi^\top\S_\pi\X)^{-1}\X^\top\X\preceq
     \bigg(\frac1k\sum_{i=d+1}^k\frac1{q_{\pi_i}}\x_{\pi_i}\x_{\pi_i}^\top\bigg)^{\!-1}\X^\top\X\preceq 2\,\I.
   \end{align*}
Thus, it remains to bound the second factor in expectation,
i.e.
$\E\big[\|\X^\dagger\S_\pi^\top\S_\pi\xib\|^2\,|A\big]$. For this,
we 
   need an extension of a result by \cite{leveraged-volume-sampling}
(see proof in Appendix \ref{a:mult}).
   \begin{lemma}\label{l:mult}
  Given $\A\in\R^{n\times d}$,
  $\b\in\R^n$ and $\beta>0$, let $\pi\sim\Vol_q^k(\A)$ where $q_i\geq 
  \beta\max\!\big\{\frac{\|\a_i\|^2}{\|\A\|_F^2},\frac{l_i(\A)}{d}\big\}$
  for all $i\in [n]$ and $k\geq d$. Then
  $\v_\pi=\A^\top\S_\pi^\top\S_\pi\b$ satisfies 
  \begin{align*}
\E\big[\|\v_\pi-\E[\v_\pi]\|^2\big]\leq
    \frac1{\beta^2 k}\|\A\|_F^2\|\b\|^2.
  \end{align*}
\end{lemma}
We use Lemma \ref{l:mult} with %Noting that $\|\X^\dagger\|_F^2=b$,
$\A=\X^{\dagger\top}$ and $\b=\xib$. In this case
$\|\a_i\|^2=v_i(\X)$ and $l_i(\A)=l_i(\X)$. Also let
$\beta=0.25$ and observe that $\phi=\|\X^\dagger\|_F^2$.
Now for $k\ge C\cdot\phi/\epsilon$ and $C\geq 16\cdot4/\beta^2$ we have:
 \begin{align}
   \Pr(A) \,\E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\ |\,A\big]
   &\leq 4\ \Pr(A)\ \E\Big[\big\| \X^\dagger\S_\pi^\top\S_\pi\xib\big\|^2\ |\,A\Big] 
\nonumber \\
   &\leq 4\, \E\big[\|\v_\pi\|^2\big]\ = \ 4\,\E\Big[\big\|\v_\pi-\E[\v_\pi]\big\|^2\Big] + 4\,\big\|\E[\v_\pi]\big\|^2\nonumber  \\
   &\leq \frac{4\epsilon}{C\beta^2\phi}\|\X^\dagger\|_F^2\|\xib\|^2 +
     4\,\big\|
     \X^\dagger\E[\S_\pi^\top\S_\pi]\xib\big\|^2\nonumber \\
     &\leq \frac{\epsilon}{16}\|\xib\|^2+4\,\big\|
     \X^\dagger\E[\S_\pi^\top\S_\pi]\xib\big\|^2. \label{eq term 1}
 \end{align}
 Next, we bound the second term in \eqref{eq:terms} by using a
 different decomposition of $\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2$. We will
 use the fact that since $q_i(\alpha)\geq \frac14
 p_i^{\mathrm{uni}}=\frac1{4n}$ for all $i\in[n]$, then
 $\S_\pi^\top\S_\pi\preceq 4n\,\I$ for all $\pi\in[n]^k$. It is only
 here that we use the $p^{\mathrm{uni}}$ term in $q(\alpha)$. It
 follows that
%  \begin{align*}
% \|(\S_\pi\X)^\dagger\S_\pi\xib\|^2
%    &= \big\|(\X^\top\S_\pi^\top\S_\pi\X)^{-1}
%      \X^\top\S_\pi^\top\S_\pi\xib\big\|^2\\
%    &\leq
%      \|\xib\|^2\|\S_\pi^\top\S_\pi\X(\X^\top\S_\pi^\top\S_\pi\X)^{-2}\X^\top\S_\pi^\top\S_\pi\|\\
%    &\leq
%      \|\xib\|^2\tr\big(\S_\pi^\top\S_\pi\X(\X^\top\S_\pi^\top\S_\pi\X)^{-2}\X^\top\S_\pi^\top\S_\pi\big)\\
%    &=
% \|\xib\|^2\tr\big(\X^\top(\S_\pi^\top\S_\pi)^2\X(\X^\top\S_\pi^\top\S_\pi\X)^{-2}\big)\\ 
%    &\leq \|\xib\|^24n\,\tr\big((\X^\top\S_\pi^\top\S_\pi\X)^{-1}\big)
%      \leq 4n \|\xib\|^2\,\tr\big((\X^\top\S_{\pi}^\top\I_{[d]}\S_{\pi}\X)^{-1}\big),
%  \end{align*}
 % \ken{Alternatively,
\begin{align*}
 \norm{(\S_\pi\X)^\dagger\S_\pi}^2
 	   & \le \norm{(\S_\pi\X)^\dagger\S_\pi}_F^2
\le  \norm{(\S_\pi\X)^\dagger}_F^2 \norm{\S_\pi}^2
\le  \norm{(\S_\pi\X)^\dagger}_F^2 4n
	\\ & = 4n\,\tr \big((\X^\top\S_\pi^\top \S_\pi \X)^{-1}\big)
\le 4n\, \tr\big((\X^\top\S_{\pi}^\top\I_{[d]}\S_{\pi}\X)^{-1}\big),
\end{align*}
 % }
 %\ken{last two inequalities use von neumann trace inequality and that matrices arePSD, or is there a more elementary argument?}
 where $\I_{[d]}=\sum_{i=1}^d\e_i\e_i^\top$ selects the
 first $d$ rows from $\S_\pi$.
 Since $\pi_{[d]}\sim \Vol^d(\X)$ and it is independent of the event
 $A$, from inequality \eqref{eq:square-inverse} in Lemma \ref{l:expectations} it
 follows that:
 \begin{align*}
   \E\big[\tr\big((\X^\top\S_{\pi}^\top\I_{[d]}\S_{\pi}\X)^{-1}\big)\
   |\,\neg A\big]
   &=   \frac{k}{d}\cdot\tr\big(
     \E\big[(\X^\top\S_{\pi_{[d]}}^\top\S_{\pi_{[d]}}\X)^{-1}\big]\big)
\leq \frac kd\cdot d\,\tr\big((\X^\top\X)^{-1}\big) = k\phi.
 \end{align*}
 Here the $k/d$ term arises from the different multipliers used for $\S_\pi$ and
 $\S_{\pi_{[d]}}$.
 %\ken{probably there's a better way to say where the $k/d$ comes from}
 Thus, we obtain that $\E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\,|\,\neg A\big]\leq
 4nk\phi\, \|\xib\|^2$. It remains to show that $\Pr(\neg A)$ is sufficiently
 small to obtain the desired bound. For this, we refer to a standard
matrix concentration result for obtaining subspace embeddings, which
follows from \cite{matrix-tail-bounds}.
\begin{lemma}
Given a full rank $\X\in\R^{n\times d}$, if distribution $q$ is such that
$q_i\geq \beta\,\frac{l_i(\X)}{d}$ for all $i\in[n]$, then an i.i.d.~sample
$\pi_1,\dots,\pi_s\sim q$ for $s\geq C'\frac d\beta\log
\frac d\delta$ with probability at least $1-\delta$ satisfies
\begin{align*}
\frac12\,\X^\top\X\preceq
  \frac1s\sum_{i=1}^s\frac1{q_{\pi_i}}\x_{\pi_i}\x_{\pi_i}^\top\preceq
  \frac32\,\X^\top\X. 
\end{align*}
\end{lemma}
Setting $\beta=\frac14$, $\delta = \frac1{6nk^2}$, $s=k-d$ and $C\geq
16C'$, we conclude that
since $k-d\geq 16C' d\log n\geq C'\frac d\beta\log \frac d\delta$, we have $\Pr(\neg A)
\E\big[\|(\S_\pi\X)^\dagger\S_\pi\xib\|^2\,|\,\neg A\big] \leq \frac \phi k \|\xib\|^2 \leq \frac\epsilon
{16} \|\xib\|^2$. Combining this with \eqref{eq:terms} and \eqref{eq term 1}, 
we complete the proof of the bound of Lemma \ref{l:key}.
% desigrthat for any
% $\alpha\in[0.5,0.75]$ and any distribution $p$, if $\D=\diag(q_i/p_i)$ then:
% \begin{align}
%   \E\big[\|\Delta_\pi^p\|^2\big]\leq \frac{\epsilon}{8}\,\|\D\xib\|^2 + 4\,\big\| \X^\dagger\big(\E[\S_\pi^\top\S_\pi]\D-\I\big)\xib\big\|^2.\label{eq:delta}
% \end{align}
\end{proof}
The last term in the bound of Lemma \ref{l:key} can be
controlled when the vector $\xib$ is orthogonal to the columns of
$\X$. The following bound is shown in Appendix \ref{a:mult}.
\begin{lemma}\label{l:bias}
  For $\X$ and $\xib$ s.t.~$\X^\top\xib=\zero$, if
  $\pi\sim\Vol_{q(\alpha)}^k(\X)$ (as in Lemma \ref{l:key}) with
  $\alpha=0.5$, then 
  \begin{align*}
    \big\|\X^\dagger\E[\S_\pi^\top\S_\pi]\xib\big\|^2\leq
    \frac\epsilon 8\|\xib\|^2.
  \end{align*}
\end{lemma}
We put the two lemmas together to complete the proof of our main result.
\begin{proofof}{Theorem}{\ref{t:mse}}
% We use inequality \eqref{eq:delta} twice, with different values of
% $\alpha$ and $p$. The key idea is to first choose distribution $p$
% so that $\D = \big(\E[\S_\pi^\top\S_\pi]\big)^{-1}$, in which case the
% second term becomes zero. This leads to the following bound proven in
% Appendix \ref{a:mult}.
% \begin{lemma}
%   For any $\X$ and $\xib$ such that $\X^\top\xib=\zero$ we have
%   \[\bigg\|\X^\dagger\mathrm{diag}\bigg(\frac{q_i((1-\frac dk)\cdot
%       0.5)}{q_i(0.5)}\bigg)\xib\bigg\|^2\leq \frac\epsilon 8 \|\xib\|^2.\]
% \end{lemma}
Setting $\wbh=(\S_\pi\X)^\dagger\S_\pi\y$ for a fixed vector
$\y\in\R^n$ with $\pi$ as in Lemma
\ref{l:key} and $\alpha=0.5$, using Lemmas \ref{l:key} and
\ref{l:bias} with $\xib=\y-\X\wols$ we obtain that:
\begin{align*}
  \E\big[\|\wbh-\wols\|^2 \big] \overset{(*)}{=}
  \E\big[\|(\S_\pi\X)^\dagger\S_\pi (\y- \X\wols)\|^2\big]
  \leq \frac\epsilon 8\|\xib\|^2 + 4\cdot\frac\epsilon 8\|\xib\|^2\leq
  \epsilon\cdot\|\xib\|^2,
\end{align*}
where $(*)$ follows because volume sampling ensures that
$\rank(\S_\pi\X)=d$, so $(\S_\pi\X)^\dagger\S_\pi\X=\I$.
%which concludes the proof via the reduction of Lemma \ref{l:reduction}.
\end{proofof}

In Appendix \ref{a:mspe} we prove Theorem \ref{t:mspe} as a reduction from Theorem
\ref{t:mse} by transforming the matrix $\X$ into matrix
$\U=\X(\X^\top\X)^{-\sfrac12}$. This transformation makes MSE equal to
MSPE, while preserving most key properties of least squares estimators.
The random experimental design obtained via this reduction is different
than the one used to bound the mean squared error. While the leverage
scores are preserved during the
transformation from $\X$ to $\U$, the inverse scores change, and in
fact they become equal to the leverage scores, i.e.,
$v_i(\U)=l_i(\U)$, so distribution $q$ is somewhat simpler in this
case. However, it can be shown that the exact experimental design
used for Theorem \ref{t:mse} also satisfies the guarantee from Theorem
\ref{t:mspe}, albeit with slightly different constants.

The logarithmic dependence on $n$ in the sample size $k$ for Theorems
\ref{t:mse} and \ref{t:mspe} comes from our analysis of the expected
error in the tail of the sampling distribution. It is possible that
the dependence on $n$ can be eliminated altogether, even when using
the same distribution. We leave this as an open question for future work.

% Acknowledgments---Will not appear in anonymized version
\acks{MWM would like
  to acknowledge ARO, DARPA, NSF and ONR for providing partial
  support of this work. Also, MWM and MD thank the NSF for
  funding via the NSF TRIPODS program. Part of this work
  was done 
  while MD, KLC and MWM were visiting the Simons Institute for the
  Theory of Computing and while MKW was at UC Santa Cruz, supported by
  NSF grant IIS-1619271.}

\bibliography{pap}

\appendix
\newpage

\section{Basic properties of the minimax-optimal experimental design}
\label{a:minimax}
We start by formally showing that
the least squares estimator $\wols(\y|\X)=\X^\dagger\y$ is 
the minimum variance unbiased estimator (MVUE) for
$\w^*=\X^\dagger\E[\y]$ w.r.t.~the family $\Fc$ consisting of all
random response vectors $\y$ with finite second moment.
\begin{proofof}{Proposition}{\ref{p:mvue}}
  Since all fixed vectors $\y\in\R^n$ belong to $\Fc$, it follows
  that $\E\big[\wbh(\y)\,|\,\y\big] = \wols(\y|\X)$ and using
  shorthands $\wbh=\wbh(\y)$, $\wols=\wols(\y|\X)$ and
  $\w^*=\X^\dagger\E[\y]$ we have
  \begin{align*}
    \Var\big[\wbh\big]
    &= \E\big[\wbh\wbh^\top\big] - \w^*\w^{*\top}\\
    &=\E_\y \Big[\E_{\wbh}\big[\wbh\wbh^\top-\wols\wols^\top\,|\,\y\big]
      +\wols\wols^\top\Big] - \w^*\w^{*\top}\\
    &=\E\big[(\wbh-\wols)(\wbh-\wols)^\top\big] +
      \E\big[\wols\wols^\top\big] - \w^*\w^{*\top}\succeq \Var\big[\wols\big],
  \end{align*}
  because the first term is a positive semi-definite matrix.
\end{proofof}
In the next lemma, we observe that it suffices to consider fixed
vectors $\y\in\R^n$ to bound $R_k^*(\X)$.
\begin{lemma}\label{l:fixed}
  Given a matrix $\X\in\R^{n\times d}$, suppose that a random design
  $(S,\wbh)$ for all fixed response vectors $\y\in\R^n$ satisfies 
  \begin{align*}
    \E\big[\wbh(\y_S)\big]=\wols\quad\text{and}\quad
    \E\big[\|\wbh(\y_S)-\wols\|^2\big]\leq
    \epsilon\cdot\|\X\wols-\y\|^2,
  \end{align*}
where $\wols=\wols(\y|\X)$. Then, for all random response vectors
$\y\in \Fc$, we have
  \begin{align*}
    \E\big[\wbh(\y_S)\big] = \w^*\quad\text{and}\quad
    \E\big[\|\wbh(\y_S)-\w^*\|^2\big]\leq \E\big[\|\wols-\w^*\|^2\big]+ \epsilon\cdot \E\big[\|\X\w^*-\y\|^2\big].
  \end{align*}
\end{lemma}
%\ken{all \emph{finite} random response vectors, yes?}
\begin{proof}
We first decompose the mean squared error using the unbiasedness of
$\wbh(\y_S)$ as follows:  
\begin{align*}
  \E\big[\|\wbh(\y_S)-\w^*\|^2\big]
  % & = \E\Big[\E\big[\|\wbh-\E[\wbh|\y]\|^2\,|\,\y\big]
  %   + \E\big[\|\E[\wbh|\y]-\w^*\|^2\,|\,\y\big]\Big]\\
  &=\E\big[\|\wbh(\y_S)-\wols\|^2\big] +
    \E\big[\|\wols-\w^*\|^2\big]\\
  &\leq \epsilon\cdot \E\big[\|\X\wols-\y\|^2\big] + \E\big[\|\wols-\w^*\|^2\big].
\end{align*}
Also note that $\E[\wbh(\y_S)] = \E\big[\E[\wbh(\y_S)\,|\,\y]\big]=\E[\wols]=\w^*$.
It remains to bound $\E\big[\|\X\wols-\y\|^2\big]$.
Note that $\wols = \argmin_\w \norm{\X\w - \y}^2$,
so that in particular $\norm{\X\wols - \y}^2 \le \norm{\X\w^* - \y}^2$,
and therefore $\E\big[\|\X\wols-\y\|^2\big] \le \E\big[\|\X\w^*-\y\|^2\big]$,
%Noting
%that $\wols=\X^\dagger \y$ and $\w^*=\X^\dagger\E[\y]$, we once
%again do a bias-variance decomposition:
%\begin{align*}
%  \E\big[\|\X\X^\dagger \y - \y\|^2\big]
%  &= \E\big[\|\X\X^\dagger (\y-\E[\y])\|^2\big]
%    +\E\big[\|\X\X^\dagger \E[\y]-\y\|^2\big]\\
%  &\leq \E\big[\|\y-\E[\y]\|^2\big] + \E\big[\|\X\w^*-\y\|^2\big]\ \leq\
%    2 \cdot\E\big[\|\X\w^*-\y\|^2\big], 
%\end{align*}
%\ken{$\X\wols$ is the closest vector to $\y$ in the column space of $\X$,
%	$\X\w^*$ is in the column space of $\X$.}
which concludes the proof.
\end{proof}
\begin{proofof}{Lemma}{\ref{l:reduction}}
Follows immediately from Lemma \ref{l:fixed} and the fact that $\Vc_k(\X)\subseteq\Wc_k(\X)$.
\end{proofof}
We next prove Proposition \ref{p:minimax-bounded}, showing that
the minimax-optimal value $R_k^*(\X)$ given in Definition
\ref{d:minimax} is well-defined for all $d\leq k\leq n$ and
for most $k$ it has matching upper and lower bounds of $\Theta(\phi/\epsilon)$.
\begin{proofof}{Proposition}{\ref{p:minimax-bounded}} 
  %\textit{Part 1.} \
  Since $\mathrm{MSE}[\wbh]=\tr\big(\Var[\wbh]\big)$
for any unbiased estimator, Proposition
\ref{p:mvue} immediately implies that $R_k^*(\X)\geq 0$. Next,
let $\pi\sim\Vol_q^k(\X)$ with $q$ chosen as in the proof of Theorem
\ref{t:mse} and consider the volume sampled estimator defined as in
\eqref{eq:design}, i.e. $\wbh=(\S_\pi\X)^\dagger\S_\pi\y$. First, we
can use Lemma \ref{l:reduction} to assume w.l.o.g.~that $\y$ is a
fixed vector in $\R^n$ so that $\w^*=\X^\dagger\y$. Then, as in
the proof of Theorem \ref{t:mse} we use equation \eqref{eq:square-inverse} in Lemma
\ref{l:expectations}:
\begin{align*}
  \E\big[\|\wbh-\w^*\|^2\big]&\leq
  4n\|\X\w^*-\y\|^2\,\E\big[\tr\big((\X^\top\S_\pi^\top\S_\pi\X)^{-1}\big)\big]\\
  &\leq 4n\|\X\w^*-\y\|^2\frac{k}{k-d+1}\tr\big((\X^\top\X)^{-1}\big).
\end{align*}
This implies that $R_k^*(\X)\leq 4n\frac{k}{k-d+1}\,\phi<\infty$, 
where unlike in the proof of Theorem \ref{t:mse} we did not need to
assume any lower bound on $k$ other than that $k\geq d$.

\textit{Part 1.} \
Theorem \ref{t:mse} implies that there is a constant $C>0$ such that
for any $\X$, $\epsilon>0$ and $k\geq C\cdot (d\log n+\phi/\epsilon)$ there
is a random experimental design $(S,\wbh)$ of size at most $k$ which
demonstrates an upper bound on the minimax-optimal value, i.e., that
$R_k^*(\X)\leq \epsilon$. Now, suppose that $k\geq 2C\cdot d\log n$ and
let $\epsilon=2C\cdot\phi/k$. Then,
$
  k\geq 2C\cdot \max\{d\log n,\,\phi/\epsilon\}\geq C\cdot(d\log n + \phi/\epsilon),
$
which means that for any $\X$ we have $R_k^*(\X)\leq \epsilon=2C\cdot\phi/k$.

\textit{Part 2.} \ This result is based on the following lower bound for
classical $A$-optimal design.
\begin{theorem}[\citeauthor{avron-boutsidis13},
  \citeyear{avron-boutsidis13}, Theorem 4.5]\label{t:lower}
  For any $\alpha>0$, $n$, $d$ such that $n>2d$ and
  $\mathrm{mod}(n,d)=0$ there is a full rank $n\times d$ matrix $\X$
  such that for any subset $S\subseteq[n]$
  with $\mathrm{rank}(\X_S)=d$ and $|S|=k$, we have
  \begin{align*}
    \tr\big((\X_S^\top\X_S)^{-1}\big)\geq
    \bigg(\frac{n-k}{k+\alpha^2}+1-\frac kd\bigg)\cdot\tr\big((\X^\top\X)^{-1}\big).
  \end{align*}
\end{theorem}
We remark that $\X$ in the theorem depends on $\alpha$.
%\ken{role of $\alpha$ is unclear: arbitrariliy small? nothing depends on it?}
Note that the condition $\mathrm{mod}(n,d)=0$ can be eliminated by
padding matrix $\X$ with appropriate number of $\zero$ rows and
replacing $n$ in the bound with $n-d$. Let $\X$ be the
matrix from Theorem \ref{t:lower} (padded if necessary, with $\alpha$
chosen later) and let
  $\Fc_N(\X)$ be the family of random response vectors $\y=\X\w^*+\xib$
  such that $\xib\sim\Nc(\zero,\I)$ and $\w^*\in\R^d$. 
Also let $\Fc_B(\X)$ be the Bayesian counterpart, where $\y=\X\w+\xib$
for independent Gaussian random vectors
$\w\sim\Nc(\zero,\sigma_\w^2\I)$ and $\xib\sim\Nc(\zero,\I)$, with any
$\sigma_\w^2>0$. Given any $\y\in\Fc_B(\X)$ with prior variance
$\sigma_\w^2$, the following bound is known for any (possibly biased)
estimator $\wbh(\y)$ of $\w$, called the minimum mean squared error
bound \citep[MMSE; see][]{gray-davisson-stats-book}: 
\begin{align*}
  \E\big[\|\wbh(\y)-\w\|^2\big] \geq \tr\big((\X^\top\X+(1/\sigma_\w^2)\I)^{-1}\big).
\end{align*}
An analogous lower bound holds when applied to the same regression model
restricted to a fixed subset $S_\circ\subseteq[n]$,
i.e.~such that $\y_{S_\circ}=\X_{S_\circ}\w+\xib_{S_\circ}$, and any estimator
$\wbh(\y_{S_\circ})$: 
\begin{align}
  \E\big[\|\wbh(\y_{S_\circ})-\w\|^2\big]\geq \tr\big((\X_{S_\circ}^\top\X_{S_\circ}+(1/\sigma_\w^2)\I)^{-1}\big).\label{eq:bayes}
\end{align}
We use this to give a lower bound for the MSE of any random experimental
design $(S,\wbh)\in\Wc_k(\X)$ following Definition \ref{d:rand} for
the (non-Bayesian) response model $\Fc_N(\X)$:
\begin{align*}
  \max_{\y\in\Fc_N(\X)}\MSE{\wbh(\y_S)}
  &= \max_{\underset{\y=\X\w+\xib}{\y\in\Fc_B(\X)}}
\max_{\w^*\in\R^d}\E\big[\|\wbh(\y_S)-\w\|^2\,|\,\w\!=\!\w^*\big]\\[-2mm]
    &\overset{(a)}{\geq} \max_{\underset{\y=\X\w+\xib}{\y\in\Fc_B(\X)}}\E_\w\Big[\E\big[\|\wbh(\y_S)-\w\|^2\,|\,\w\big]\Big]\\
    &= \max_{\underset{\y=\X\w+\xib}{\y\in\Fc_B(\X)}} \E\big[\|\wbh(\y_S)-\w\|^2\big]\\
  &=\max_{\underset{\y=\X\w+\xib}{\y\in\Fc_B(\X)}}\E_S\Big[\E\big[\|\wbh(\y_S)-\w\|^2\,|\,S\big]\Big]\\[-2mm]
  &\overset{(b)}{\geq} \lim_{\sigma_\w\rightarrow\infty}
    \E_S\Big[\tr\big((\X_S^\top\X_S+(1/\sigma_\w^2)\I)^{-1}\big)\Big]\\
  &%= \E_S\Big[ \tr\big((\X_S^\top\X_S)^{-1}\big)\Big]
 \geq \min_{\underset{\rank(\X_S)=d}{S:|S|\leq k}}\tr\big((\X_S^\top\X_S)^{-1}\big),
\end{align*}
where $(a)$ follows because expectation $\E_\w$ is always upper
bounded by the maximum over all possible values of $\w$, and $(b)$
follows from \eqref{eq:bayes} applied to the conditional expectation
for a fixed $S$ and the fact that the trace is monotonically increasing with $\sigma_\w$.
We now lower bound $R_k^*(\X)$ by the minimax value based on the
response family $\Fc_N(\X)$: 
  \begin{align*}
    R_k^*(\X)
    &\geq
\min_{(S,\wbh)\in\Wc_k(\X)}\
      \max_{\y\in\Fc_N(\X)}\frac{\MSE{\wbh(\y_S)}-
      \MSE{\wols(\y|\X)}}{\E_\y\big[\|\xib_{\y|\X}\|^2\big]}\\
   &\geq\min_{\underset{\rank(\X_S)=d}{S:|S|\leq k}}\frac1n\Big(\tr\big((\X_S^\top\X_S)^{-1}\big)-\tr\big((\X^\top\X)^{-1}\big)\Big)\\ 
    \text{(Theorem \ref{t:lower})}\quad&\geq
\frac1n\bigg(\frac{n-d-k}{k+\alpha^2}-\frac kd\bigg)\cdot
                                         \tr\big((\X^\top\X)^{-1}\big)\\
&    \geq  \bigg(\frac k{k+\alpha^2} - \frac{3k^2}{nd}\bigg)\cdot
      \frac\phi k\ \geq\  (1-\epsilon)\cdot\frac\phi k,
  \end{align*}
because $k^2<\epsilon nd/3$ and we can choose $\alpha$ small enough
so that $\frac{\alpha^2}{k+\alpha^2}+\frac{3k^2}{nd} \leq \epsilon$.
\end{proofof}

\section{Omitted proofs from Section \ref{s:proofs}}
\label{a:mult}

\begin{proofof}{Lemma}{\ref{l:mult}}
Recall that $\v_\pi = \A^\top\S_\pi^\top\S_\pi\b$.  We rewrite the expectation as follows:
\begin{align*}
  % \E\bigg[\Big\|\A^\top\!\Big(\frac1k\S_{\pi}^\top\S_{\pi}\!-\!\I\Big)\b\|^2\big]
  \E\big[\|\v_\pi-\E[\v_\pi]\|^2\big]
  &=\b^\top\E\Big[
    \big(\S_{\pi}^\top\S_{\pi}\!-\!\E[\S_{\pi}^\top\S_{\pi}]\big)\A\A^\top\!
    \big(\S_{\pi}^\top\S_{\pi}\!-\! \E[\S_{\pi}^\top\S_{\pi}]\big)\Big]\ \b\\
  &\le \bigg\|\underbrace{
    \begin{bmatrix}\cov\big[\frac{s_i}{kq_i},\frac{s_j}{kq_j}\big]\a_i^\top\a_j\end{bmatrix}_{n\times n}}_{\M}\bigg\|\cdot\|\b\|^2,
\end{align*}
where $s_i=|\{t\,:\,\pi_t=i\}|$.
Note that $\M$ is the Hadamard product of two PSD matrices, and therefore also PSD
by the Schur product theorem.
Next, we use two formulas
shown by \cite{leveraged-volume-sampling} for rescaled volume sampling:
\begin{align}
  \E[s_i] &=(k-d)\,q_i+l_i,\label{eq:marginal}\\
  \cov(s_i,s_j)&=\one_{i=j}\E[s_i] - (k-d)q_iq_j - l_{ij}^2,\nonumber
\end{align}
where $l_i=\a_i^\top(\A^\top\A)^{-1}\a_i$ is the $i$th leverage score
of $\A$ and $l_{ij}=\a_i^\top(\A^\top\A)^{-1}\a_j$ is the $(i,j)$th 
cross-leverage score. Using the above we get:
\begin{align*}
  \M = \text{diag}\bigg(\begin{bmatrix}\frac{\|\a_i\|^2}{kq_i}\frac{(k-d)q_i+l_i}{kq_i}\end{bmatrix}_{n\times 1}\bigg) -
  \frac{k-d}{k^2}\A\A^\top
      - \begin{bmatrix}\frac{l_{ij}^2}{k^2q_iq_j}\a_i^\top\a_j\end{bmatrix}_{n\times n}.
\end{align*}
%\ken{I think I understand what this notation means, but it's a bit odd: the last term seems to indicate an $ij$'th entry, the first
%term an implicit vector inside the diag}
Note that the second term is a PSD matrix being subtracted from
$\M$.
% This matrix is PSD because it can be written as a Hadamard
% product of $\A\A^\top$ (which is PSD) with a rank-one PSD matrix, so we
% can invoke the Schur product theorem.
Similarly, the last term is also
subtracting a PSD matrix. To see this, note that the matrix
formed by the cross-leverage scores is $[l_{ij}]_{n\times n}=\A(\A^\top\A)^{-1}\A^\top$,
so it is PSD.
%\ken{what does $[l_{ij}]_{ij}$ mean?}
Therefore, we can write the last term as a Hadamard product of
PSD matrices and apply Schur product theorem. Thus it remains to bound
the diagonal term: 
\begin{align*}
  \frac{\|\a_i\|^2}{kq_i}\,\frac{(k-d)q_i+l_i}{kq_i}\leq
  \frac{\|\A\|_F^2}{\beta k} \, \bigg(\frac{k-d}{k} +
  \frac{d}{k\beta}\bigg)\leq  \frac{\|\A\|_F^2}{\beta^2 k}.
\end{align*}
Since the spectral norm of the diagonal term is bounded by $\frac{\|\A\|_F^2}{\beta^2 k}$,
and subtracting the two PSD terms leaves $\M$ a PSD matrix, we have
$\norm{\M} \le \frac{\|\A\|_F^2}{\beta^2 k}$, and the result follows.
%which completes the proof.
% \end{proof}
\end{proofof}

\begin{proofof}{Lemma}{\ref{l:bias}}
% If we set $\D = \big(\E[\S_\pi^\top\S_\pi]\big)^{-1}$ then the
% second term becomes zero. 
Using the marginal expectation formula \eqref{eq:marginal} for
volume sampling we have:
\begin{align*}
  \E[\S_\pi^\top\S_\pi] =
  \text{diag}\bigg(\begin{bmatrix}\frac{(k-d)q_i+l_i(\X)}{kq_i}
  \end{bmatrix}_{n\times
  1}\bigg)
  =   \text{diag}\bigg(\begin{bmatrix}\frac{q_i(\frac
  {k-d}k\alpha)}{q_i(\alpha)}\end{bmatrix}_{n\times 1}\bigg)\defeq \D_\alpha,
\end{align*}
where 
  $q(\alpha)= \alpha
  \big(0.5\cdot p^{\mathrm{uni}}+0.5\cdot p^{\mathrm{inv}}\big)
  +(1\!-\!\alpha)p^{\mathrm{lev}}$. Now, let $\beta =
  \alpha\cdot\frac{k}{k-d}$ and $\pi'\sim \Vol_{q(\beta)}^k(\X)$. 
%   so to zero out the second term in \eqref{eq:delta} we
%   need to use $p=q(\tilde{\alpha})$. Therefore, in this case we have
%   $\|\D\|\leq \alpha/\tilde{\alpha}$ and 
% moreover, u
Using expectation formula \eqref{eq:unbiased} of Lemma
\ref{l:expectations} and the fact that $\X^\dagger\xib=\zero$, we have:
\begin{align*}
  \E\big[(\S_{\pi'}\X)^\dagger\S_{\pi'}\,\D_\beta^{-1}\xib]
  &= \X^\dagger \D_{\beta}^{-1}\xib \\
&=\X^\dagger(\D_{\beta}^{-1}-\I)\xib\\
  &=\X^\dagger \text{diag}\bigg(\frac{q_i(\beta) -
    q_i(\alpha)}{q_i(\alpha)}\bigg)\xib\\
  &=-\frac{k}{k-d}\X^\dagger(\D_{\alpha}-\I)\xib\\
  &=-\frac{k}{k-d}\X^\dagger\D_\alpha\xib.
%\frac{\alpha}{\tilde{\alpha}}\X^\dagger\Q_{\tilde{\alpha}}\xib,\qquad
  % \text{where}\quad  \Q_{\tilde{\alpha}}
  % = \frac{d\tilde{\alpha}}{k}\,\text{diag} \bigg(\frac{
  %   \frac12p_i^{\mathrm{uni}}+\frac12p_i^{\mathrm{inv}}-p_i^{\mathrm{lev}}}
  %   {q_i(\tilde{\alpha})}\bigg).
\end{align*}
We can assume $k\geq 3d$, adjusting $C$ of Lemma~\ref{l:key}.
With $\alpha=0.5$, this implies $\beta\in[0.5,0.75]$, so
applying Lemma \ref{l:key} to sequence $\pi'$ and vector
$\D_\beta^{-1}\xib$ combined with Jensen's inequality we obtain:
\begin{align*}
  \|\X^\dagger\D_\alpha\xib\|^2
  &=\Big(\frac{k-d}{k}\Big)^2\cdot
\big\|\E[(\S_{\pi'}\X)^\dagger\S_{\pi'}\D_\beta^{-1}\xib]\big\|^2\\ 
  \text{(Jensen's inequality)}\quad
  &\leq
    \Big(\frac{k-d}{k}\Big)^2\cdot\E\big[\|(\S_{\pi'}\X)^\dagger\S_{\pi'}\D_\beta^{-1}\xib\|^2\big]\\
  \text{(Lemma \ref{l:key})}\quad
  &\leq \Big(\frac{k-d}{k}\Big)^2\cdot \Big(\frac\epsilon8\,\|\D_\beta^{-1}\xib\|^2 +
    4\underbrace{\|\X^\dagger\D_\beta\D_\beta^{-1}\xib\|^2}_{0}\Big)\\[-3mm]
  &\leq \frac\epsilon8\cdot\|\xib\|^2,
%   \frac{\tilde{\alpha}^2}{\alpha^2}\big\|\E[\Delta_\pi^p]\big\|^2\leq
% \frac{\tilde{\alpha}^2}{\alpha^2}  \E\big[\|\Delta_\pi^p\|^2\big]\leq
% \frac{\tilde{\alpha}^2}{\alpha^2}  \frac{\epsilon}{8} \,\|\D\xib\|^2\leq
%   \frac{\epsilon}{8}\,\|\xib\|^2.
%  \label{eq:q-bound}
\end{align*}
because $\|\D_\beta^{-1}\|\leq \frac{k}{k-d}$, which concludes the proof.
% Note that the above bound holds for any $\Q_{\tilde{\alpha}}$ such that
% $\alpha\in[0.5,0.75]$, so in particular as long as $\frac dk\leq
% \frac13$, then $\Q_{\alpha}$ satisfies \eqref{eq:q-bound} for $\alpha=0.5$.
% We now choose this $\alpha$ and set $p=q=q(\alpha)$
% (i.e.,~$\D=\I$).
% \
\end{proofof}

\section{Proof of Theorem \ref{t:mspe}}
\label{a:mspe}
  The key idea in the proof is a standard transformation of the
  data matrix $\X$ which has the property that it preserves the
  predictions of the least squares estimator, while transforming the
  actual estimator in such a way that the mean squared error becomes
  equal to the mean squared prediction error. Specifically, consider
  matrix $\U=\X(\X^\top\X)^{-\sfrac12}$. This matrix has the property
  that $\U^\top\U=\I$ so $\tr((\U^\top\U)^{-1})=d$ and $\U\U^\dagger=\U\U^\top=\X\X^\dagger$. Replacing all least squares
  estimators for this new matrix we have $\v^* = \U^\top\E[\y]$,
  $\v_{\mathrm{LS}}=\U^\top \y$ and $\vbh=(\S_\pi\U)^\dagger\S_\pi
  \y$. Note that we have $\U\v^*=\X\X^\dagger\E[\y]=\X\w^*$ and
  similarly $\U\v_{\mathrm{LS}}=\X\wols$. A simple calculation also
  reveals that $\U\vbh=\X\wbh$ for
  $\wbh=(\S_\pi\X)^\dagger\S_\pi\y$.
  %(Alternatively, note that $\U$ and $\X$ have the same column space,
  %so that $\U\cdot \argmin_\z\norm{\S_\pi (\U\z - \y)} = \X\cdot \argmin_\z\norm{\S_\pi (\X\z - \y)}$.
  %\ken{probably this added sentence is too much or too little})
   Suppose that $\pi\sim\Vol_q^k(\U)$
  is produced as in
  the proof of Theorem \ref{t:mse} when applied 
  to matrix $\U$. Then: 
  \begin{align*}
    \E\big[\|\X(\wbh-\w^*)\|^2\big]
    &= \E\big[\|\U(\vbh-\v^*)\|^2\big]\\
&= \E\big[\|\vbh-\v^*\|^2\big]\\
    \text{(Theorem \ref{t:mse})}\quad
    &\leq \E\big[\|\v_{\mathrm{LS}}-\v^*\|^2\big] + \epsilon\cdot \E\big[\|\U\v^*-\y\|^2\big]\\
      &= \E\big[\|\X(\wols-\w^*)\|^2\big] + \epsilon\cdot \E\big[\|\X\w^*-\y\|^2\big].
  \end{align*}

\end{document}

### 2022-08-23 design/continuous-empirical.tex
Commit original repos

\subsection{Empirical analysis of asymptotic consistency}
\label{s:empirical}

Theorem \ref{t:asymptotic} states that our surrogate expressions for the MSE under certain Gaussian designs are asymptotically consistent with the multiplicative error rate of $O(1/d)$. 
In this section, we show strong empirical evidence that this fact extends to
the setting not covered by the theorem: under-determined regime (i.e.,
$n<d$) with a non-isotropic Gaussian distribution (i.e., $\Sigmab\neq
\I$). We break down our analysis into verifying two
conjectures which are of independence interest to multivariate
Gaussian analysis. The first conjecture addresses the variance term in
the MSE and postulates an asymptotically
consistent formula for the expected Moore-Penrose pseudo-inverse
of the singular Wishart distribution. Recall that matrix
$\W\sim\Wc(\Sigmab,n)$ is distributed according to the Wishart
distribution with $n$ degrees of freedom if it can be decomposed as
$\W=\X^\top\X$, where $\X\sim\Nc(\zero,\Sigmab)^n$.
\begin{conjecture}[Pseudo-inverse of singular Wishart]\label{c:wishart}
  Fix $n/d<1$ and let $\W\sim\Wc(\Sigmab,n )$, where
$\Sigmab$ is $d\times d$ positive definite with condition number bounded by a constant.
  Then:
\begin{align}
\bigg|\frac{\E\big[\tr(\W^\dagger)\big]}{\Vc(\Sigmab,n)} -1\bigg|=
  O(1/d)\qquad\text{for}\quad\Vc(\Sigmab,n)=\tr\big((\Sigmab+\lambda_n\I)^{-1}\big)\frac{1-\alpha_n}{d-n},\label{eq:wishart}
\end{align}
where $\lambda_n\geq 0$ satisfies $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$ and
$\alpha_n=\det(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$.
\end{conjecture}
Our second conjecture involves the projection onto the
orthogonal complement of a Gaussian sample $\X$, i.e., the matrix
$\I-\X^\dagger\X$, and addresses the bias term in the MSE.
\begin{conjecture}[Gaussian orthogonal projection]\label{c:projection}
  Fix $n/d<1$ and let $\X\sim \Nc(\zero,\Sigmab)^n$, where
$\Sigmab$ is $d\times d$ positive definite with condition number bounded by a constant. Then:
\begin{align}
\sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top
  \Bc(\Sigmab,n)\w} - 1\bigg| = O(1/d)\qquad\text{for}\quad
  \Bc(\Sigmab,n) =
  (\Sigmab+\lambda_n\I)^{-1}\frac{d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})},\label{eq:projection}  
\end{align}
where $\lambda_n\geq 0$ satisfies $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$.
\end{conjecture}
Note that the surrogate expression for the mean squared
error can be written as:
\[
\Mc(\Sigmab_\mu,\w^*,\sigma^2,n)=\sigma^2 \Vc(\Sigmab_\mu,n)
+ \w^{*\top}\Bc(\Sigmab_\mu,n)\w^*.
\]
So, if the conjectures are true,
this would immediately imply that the asymptotic consistency claim
given in Theorem \ref{t:asymptotic} for
the surrogate Gaussian MSE extends to
the under-determined setting with arbitrary covariance $\Sigmab$.

Furthermore, both of these conjectures are related to open problems
which have been extensively studied in the literature. 
With respect to Conjecture~\ref{c:wishart}, \cite{srivastava2003} first derived the probability
density function of a singular Wishart
distribution, and \cite{cook2011} computed the first and second moments of
generalized inverses of a singular Wishart distribution. However, for the
Moore-Penrose pseudo-inverse and arbitrary covariance $\Sigmab$,
\cite{cook2011} claims that the quantities required
to express the mean ``do not have tractable closed-form representation.''
%
Conjecture~\ref{c:projection} has connections to directional statistics.
Using the SVD, we have the equivalent representation $\X^\dagger \X = \V \V^\top$
where $\V$ is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal
$n$-frames in $\R^d$).
The distribution of $\V$ is known as the matrix angular central
Gaussian (MACG) distribution \cite{chikuse1990matrix}. While prior work
has considered high dimensional limit theorems \cite{CHIKUSE1991145}
as well as density estimation and hypothesis testing \cite{CHIKUSE1998188}
on $V_{n,d}$, they only analyzed the invariant measure
(which corresponds in our setting to $\Sigmab = \I$),
and to our knowledge a closed form expression of $\E[\V\V^\top]$ where
$\V$ is distributed according to MACG with
arbitrary $\Sigmab$ remains an open question.

For verifying these two conjectures, it suffices to only consider
diagonal covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D
\Q^\top$ is its eigendecomposition and $\X\sim\Nc(\zero, \Q\D\Q^\top)^n$, then
we have for $\W \sim \Wc(\Sigmab, n)$ that $\W
\overset{d}{=} \X^\top \X$ and hence, defining
$\Xt\sim\Nc(\zero,\D)^n$, by linearity and unitary invariance of trace,
\begin{align*}
  \E[\tr(\W^\dagger)]
  &= \tr\big( \E[(\X^\top\X)^\dagger] \big)
  = \tr\Big( \Q\E\big[(\Xt^\top\Xt)^\dagger\big]\Q^\top \Big)
  = \tr\Big( \E\big[(\Xt^\top\Xt)^\dagger\big] \Big)
  = \E\left[\tr \big((\Xt^\top\Xt)^\dagger\big) \right].
\end{align*}
Similarly, we have that
$\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$, and a simple
calculation shows that the expression in Conjecture~\ref{c:projection}
is also independent of the choice of matrix $\Q$.
Thus, we empirically validate our conjectures for diagonal matrices $\Sigmab$
with several different eigenvalue decay profiles. Denoting
$\lambda_1,...,\lambda_d$ as the eigenvalues of $\Sigmab$, we consider
the following decays:
\begin{itemize}
  \item \texttt{diag\_linear}: linear decay, $\lambda_{i} = b-a i$;
  \item \texttt{diag\_exp}: exponential decay, $\lambda_{i} = b\,10^{-
      a i} $;
  \item \texttt{diag\_poly}: fixed-degree polynomial decay, $\lambda_{i} = (b-a i)^2$;
  \item \texttt{diag\_poly\_2}: variable-degree polynomial decay, $\lambda_i = b i^{-a}$.
\end{itemize}
The constants $a$ and $b$ are chosen to ensure $\lambda_{\text{max}}(\Sigmab) = 1$ and
$\lambda_{\text{min}}(\Sigmab) = 10^{-4}$ (i.e., the condition number
$\kappa(\Sigmab) = 10^{4}$ remains constant).
Figure~\ref{fig:eig-decays} illustrates an example of these decay profiles
for $d=100$.

\begin{figure}[H]
    \includegraphics[width=\textwidth]{continuous_figures/decays.pdf}
  \caption{Scree-plots of $\Sigmab$ for the eigenvalue decays examined
    in our empirical valuations. Here $d=100$ for visualization, whereas
    our experiments increase $d$ while preserving the ratio $n/d$ and
    the decay profile,
    with $\lambda_{\text{max}}(\Sigmab) = 1$ to
    $\lambda_{\text{min}}(\Sigmab) = 10^{-4}$.}
  \label{fig:eig-decays}
\end{figure}

We verify our conjectures by incrasing $d$ while keeping the aspect ratio
$n/d$ fixed and examining the rate of decay of the quantities asserted
in the conjectures. As no closed form expressions are available for
the expectations in the conjectures, we estimate $\E\big[\tr(\W^\dagger)\big]$ (for
Conjecture~\ref{c:wishart}) and $\E[\I-\X^\dagger\X]$ (for
Conjecture~\ref{c:projection}) through Monte Carlo sampling. To ensure that
estimation noise is sufficiently small, we continually increase the number
of Monte Carlo samples until the bootstrap confidence intervals are
within $\pm 12.5\%$ of the quantities in \eqref{eq:wishart} and
\eqref{eq:projection}. We found that while
Conjecture~\ref{c:wishart} required a relatively small number of
trials (up to one thousand),
estimation noise was much larger in Conjecture~\ref{c:projection} and
necessitated over two million trials to obtain good estimates
near $d=100$.

% For Conjecture~\ref{c:wishart}, sampling $\W\sim\Wc(\Sigmab,n)$ is
% elementary and the only nontrivial aspect of computing $\Vc(\Sigmab, n)$
% is solving for $\lambda_n$ such that
% $n = \tr(\Sigmab(\Sigmab + \lambda_n \I)^{-1})$. But notice that
% if we let $\tau_i > 0$ be the eigenvalues of $\Sigmab \succ 0$,
% then
% \begin{align*}
%   \tr(\Sigmab(\Sigmab + \lambda \I)^{-1})
%   &= \sum_{i=1}^d \frac{\tau_i}{\tau_i + \lambda}
% \end{align*}
% which is equal to $d > n$ when $\lambda = 0$ and monotonically decreasing in
% $\lambda$. Hence, $\lambda_n$ is unique and can be found using bisection.

\begin{figure}[h]
    \includegraphics[width=\textwidth]{continuous_figures/wishart_var.pdf}
  \caption{
    Empirical verification of Conjecture~\ref{c:wishart}. We show
    the quantity $\big|\E[\tr(\W^\dagger)]\,\Vc(\Sigmab,n)^{-1} -1\big|$
    as $d$ increases for various aspect ratios $n/d$. Consistent with our conjecture,
    a $O(1/d)$ decay (linear with slope $-1$ on a log-log plot) is
    exhibited across all eigenvalue decay profiles and
    aspect ratios investigated.
  }
  \label{f:conj-wishart}
\end{figure}

The results of empirically validating Conjecture~\ref{c:wishart} are
illustrated in Figure~\ref{f:conj-wishart}, where we performed
Monte Carlo estimation of $\E\big[\tr(\W^\dagger)\big]$ and
plot $\big|\E[\tr(\W^\dagger)]\,\Vc(\Sigmab,n)^{-1} -1\big|$
as $d$ increases from $10$ to $1000$, across a range of aspect ratios $n/d$ and eigenvalue
decay profiles for $\Sigmab$. Confidence intervals are estimated by
bootstrapping. We observe that on log-log axes all of the plots are decreasing with
a linear $-1$ slope, consistent with the $O(1/d)$ rate predicted
by Conjecture~\ref{c:wishart}.

Conjecture \ref{c:projection} is handled similarly, by sampling
$\X\sim\mu^n$ where $\mu=\Nc(\zero,\Sigmab )$ to obtain a Monte Carlo estimate of
$\E[\I-\X^\dagger\X]$. To handle the supremum over $\w$, notice that
we can rewrite \eqref{eq:projection} as a spectral norm
\begin{align}
  \sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top
  \Bc(\Sigmab,n)\w} - 1\bigg|
\  =\ \big\|\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-1}
  - \I\big\|.
  \label{eq:proj-to-operator-norm}
\end{align}
Confidence intervals can now be constructed
using existing methods for constructing operator norm
confidence intervals, and our results use the bootstrapping
method described in \cite{lopes2019bootstrapping}.

Figure~\ref{f:conj-bias} shows
how $\big\|\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-1} - \I\big\|$ decays as we
hold the aspect ratio $n/d$ fixed and increase $d$ between $10$ and
$100$ across the listed
eigenvalue decay profiles and aspect ratios. Again, we observe on
log-log axes a linear decay with slope $-1$ consistent with the $O(1/d)$
rate posed by Conjecture~\ref{c:projection}. Note that the range of
$d$ is smaller than in Figure \ref{f:conj-wishart} because the large
number of Monte Carlo samples (up to two million) required for this
experiment made the computations much more expensive.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=\textwidth]{continuous_figures/proj_bias.pdf}
  \end{center}
  \caption{
    Empirical results validating Conjecture~\ref{c:projection}.
We show how
    $\big\|\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-1} - \I\big\|$
    (which by Equation~\ref{eq:proj-to-operator-norm}
    is equal to the quantity controlled by Conjecture~\ref{c:projection})
    decays as $d$ increases and observe a linear slope consistent with the conjectured $O(1/d)$ rate.
  }
  \label{f:conj-bias}
\end{figure}

### 2022-08-23 design/continuous.tex
Commit original repos

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}
\usepackage{cleveref}
\input{shortdefs}

\title{Exact expressions for double descent
and implicit regularization
\\
via surrogate random design}
  \author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \and
  Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \and
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
 }

% \author{
%       Micha{\l } Derezi\'{n}ski \\
% Department of Statistics\\
% University of California, Berkeley\\
% \texttt{mderezin@berkeley.edu}
% }

% \setlength{\parindent}{0cm}
% \setlength{\parskip}{0mm}

\begin{document}
\maketitle

\begin{abstract}
  Double descent refers to the phase transition that is exhibited by
  the generalization error of unregularized learning models when varying the ratio
  between the number of parameters and the number of training
  samples. The recent success of highly over-parameterized machine learning
  models such as deep neural networks has motivated a theoretical analysis of
  the double descent phenomenon in classical models such as linear
  regression which can also generalize well in the over-parameterized
  regime. We build on recent advances in Randomized Numerical Linear
  Algebra (RandNLA) to provide the first exact non-asymptotic
  expressions for double descent of the minimum norm linear
  estimator. Our approach involves constructing
  what we call a surrogate random design to replace the standard
  i.i.d.~design of the training sample. This surrogate design admits
  exact expressions for the mean squared error of the estimator while
  preserving the key properties of the standard design. 
  We also establish an exact implicit regularization result for
over-parameterized training samples. In particular, we show that, for 
the surrogate design, the implicit bias of the unregularized minimum
norm estimator precisely corresponds to solving a ridge-regularized
least squares problem on the population distribution. 
\end{abstract}

\section{Introduction}

Classical statistical learning theory asserts that to achieve generalization one must use training sample size that sufficiently exceeds the complexity of the learning model, where the latter is typically represented by the number of parameters (or some related structural parameter) \cite{HFT09}. 
In particular, this seems to suggest the conventional wisdom that one should not use models that fit the training data exactly. 
However, modern machine learning
practice often seems to go against this intuition, using models with so many
parameters that the training data can be perfectly interpolated, in which case
the training error vanishes. It has been shown that 
models such as deep neural networks, as well as certain so-called interpolating kernels and decision
trees, can generalize well in this regime. In particular, recent work \cite{BHMM19}
empirically demonstrated a phase transition in generalization
performance of learning models which occurs at an
\emph{interpolation thershold}, i.e., a point where training error
goes to zero (as one varies the ratio between the model complexity and
the sample size). Moving away from this threshold in 
either direction tends to reduce the generalization error, leading to
the so-called \emph{double descent} curve.

To understand this
surprising phenomenon, in perhaps the simplest possible setting, we
study it in the context of linear or least squares regression.
Consider a full rank $n\times d$ data matrix $\X$ and a vector $\y$ of
responses corresponding to each of the $n$ data points (the rows of $\X$), where we wish to
find the best linear model $\X\w\approx \y$, parameterized by a
$d$-dimensional vector $\w$.
The simplest example of an estimator that has been shown to exhibit
the double descent phenomenon \cite{belkin2019two} is the
Moore-Penrose estimator, $\wbh=\X^\dagger\y$: 
in the so-called over-determined regime, i.e., when $n>d$, it corresponds to the
least squares solution, i.e., $\argmin_{\w} \|\X\w-\y\|^2$; and in the
under-determined regime (also known as
over-parameterized or interpolating), i.e., when $n<d$, it
corresponds to the minimum norm solution to the linear system $\X\w=\y$.  
Given the ubiquity of linear regression and the Moore-Penrose
solution, e.g., in kernel-based machine learning, studying the
performance of this estimator can shed some light on the effects of
over-parameterization/interpolation in machine learning more generally. 
Of particular interest are results that are exact (i.e., not upper/lower bounds) and
non-asymptotic (i.e., for large but still finite $n$ and $d$).

We build on methods from Randomized Numerical
Linear Algebra (RandNLA) in order to obtain \emph{exact non-asymptotic
expressions} for the mean squared error (MSE) of the
Moore-Penrose estimator (see Theorem~\ref{t:mse}). 
This provides a precise characterization of the double descent
phenomenon for perhaps the simplest and most ubiquitous regression
problem. 
In obtaining these results, we are able to provide precise
formulas for the \emph{implicit regularization} induced by
minimum norm solutions of
under-determined training samples, relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}). 
This result has been observed empirically for RandNLA
methods~\cite{Mah-mat-rev_JRNL}, but it has also been shown 
in deep learning~\cite{Ney17_TR} and machine
learning~\cite{Mah12} more generally. 
To obtain our precise results, we use a somewhat non-standard random
design, which we term surrogate random design (see
Section~\ref{s:determinantal} for a detailed discussion), and which we
expect will be of more general interest. 
Informally, the goal of a surrogate random design is to modify an
original design to capture its main properties
while being ``nicer'' in some useful way. In
Theorem~\ref{t:asymptotic} and Section \ref{s:asymptotic} we show,
both theoretically and empirically, that our surrogate design
accurately preserves the key properties of the original design when
the data distribution is a multivariate Gaussian.

\subsection{Main results: double descent and implicit regularization}

As the performance metric in our analysis, we use the \emph{mean
  squared error} (MSE), defined as
$\mathrm{MSE}[\wbh]=\E\big[\|\wbh-\w^*\|^2\big]$, where $\w^*$ is a fixed
underlying linear model of the responses.  
In analyzing the MSE, we make the following standard assumption on the response noise.
\begin{assumption}[Homoscedastic noise]\label{a:linear}
  Responses are $y(\x) = \x^\top\w^*+\xi$ where
$\xi\sim\Nc(0,\sigma^2)$.
\end{assumption}

\noindent
Our main result provides an exact expression for the MSE of the
Moore-Penrose estimator under our surrogate design denoted $S_\mu^n$, where
$\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
size (details in Section~\ref{s:determinantal}). This surrogate is
used in place of the standard $n\times d$ random design $\X\sim\mu^n$, where $n$
data points (the rows of $\X$) are sampled independently from
$\mu$. Unlike for the standard design, our MSE formula is
fully expressible as a function of the covariance
matrix $\Sigmab_\mu=\E_\mu[\x\x^\top]$. To state our main result, we
need an additional minor assumption on $\mu$ which is satisfied by
most standard continuous distributions such as any multivariate
Gaussian with positive definite covariance matrix. 
\begin{assumption}[General position]\label{a:general-position}
For $1\leq n \leq d$, if $\X\sim\mu^n$, then $\rank(\X)=n$ almost surely.
\end{assumption}

\noindent
Under Assumptions~\ref{a:linear} and~\ref{a:general-position}, we can establish our first main result, stated as the following theorem.

\begin{theorem}[Exact non-asymptotic MSE]
\label{t:mse}
  If the response noise is homoscedastic with variance~$\sigma^2$
  (Assumption~\ref{a:linear}) and $\mu$ is in
  general position (Assumption~\ref{a:general-position}), then for
  $\Xb\sim S_\mu^n$ (Definition~\ref{d:main}) and
  $\yb_i=y(\xbb_i)$, we have
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} =
    \begin{cases}
    \sigma^2\,\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\cdot
    \frac{1-\alpha_n}{d-n}\ +\
\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
{\tr((\Sigmab_\mu+\lambda_n\I)^{-1})}\cdot (d-n),
&\text{for }n<d,\\
\sigma^2\, \tr(\Sigmab_\mu^{-1}),& \text{for }n=d,\\
\sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d},&\text{for
}n>d,
\end{cases}
  \end{align*}
  where $\lambda_n\geq 0$ satisfies
  $\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})=n$, while
  $\alpha_n=\det(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$
  and $\beta_n=\ee^{d-n}$.
\end{theorem}
\begin{definition}
  We will use $\Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
  for $\MSE{\Xb^\dagger\ybb}$.
\end{definition}
For illustration, we plot these MSE expressions in Figure \ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
and Section \ref{s:asymptotic}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error 
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/descent-intro}
%  \end{center}
  \caption{
Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
numerical estimates even for non-isotropic
features. Eigenvalue decay leads to a steeper 
descent curve in the under-determined regime ($n<d$).}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/descent-shrinkage}
%  \end{center}
  \caption{
    The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.}
\end{subfigure}
\caption{Illustration of the main results for $d=100$ and
$\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
eigenvalues decaying exponentially and scaled so that
$\tr(\Sigmab^{-1})=d$. We use our surrogate
formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
\ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
lines), accompanied by the empirical estimates based on the standard
i.i.d.~design (error bars are three times the standard error of the
mean). We consider three different condition numbers $\kappa$ of
$\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
i.e., $\Sigmab=\I$. In all cases, we use $\sigma^2=\|\w^*\|=1$ with
$\w^*=\frac1{\sqrt{d}}\one$.}
\label{f:intro}
\end{figure}

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$. 
This expression is precisely the classical notion of \emph{effective dimension} for ridge regression regularized with $\lambda_n$~\cite{Alaoui15}, and it arises here even though there is no explicit ridge regularization in the problem being considered in Theorem~\ref{t:mse}. 
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as: 
\begin{align*}
\argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
    + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\qquad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us, 
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in Assumption~\ref{a:linear}.  
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
\label{t:unbiased}
For $\mu$ satisfying\footnote{The proof of Theorem \ref{t:unbiased}
  can be easily extended to probability measures $\mu$ that do not
  satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and any
  $y(\cdot)$ such that   $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
  $\Xb\sim S_\mu^n$ and $\yb_i=y(\xbb_i)$, then
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] = 
    \begin{cases}
       (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} &\text{for }n<d,\\
        \Sigmab_\mu^{-1}\v_{\mu,y} &\text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is chosen so that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

\noindent
That is, when $n < d$, the Moore-Penrose estimator (which itself is
not regularized), computed on the
random training sample, in expectation equals the global ridge-regularized least
squares solution of the underlying regression
problem. Moreover, $\lambda_n$, i.e., the amount
of implicit $\ell_2$-regularization, is controlled by the degree of
over-parameterization in such a way as to ensure that $n$ becomes the ridge effective dimension
(a.k.a.~the effective degrees of freedom).

We illustrate this result in Figure
\ref{f:intro}b, plotting the norm of the expectation of the
Moore-Penrose estimator. As for the MSE, our surrogate theory aligns
well with the empirical estimates for i.i.d.~Gaussian designs, showing
that the shrinkage of the unregularized estimator in the
under-determined regime matches the implicit
ridge-regularization characterized by Theorem \ref{t:unbiased}. While the shrinkage
is a linear function of the sample size $n$ for isotropic features
(i.e., $\Sigmab=\I$), it 
exhibits a non-linear behavior for other spectral decays.
Such \emph{implicit regularization} has been studied
previously~\cite{MO11-implementing, PM11, GM14_ICML, Mah12}; it has
been observed empirically for RandNLA sampling
algorithms~\cite{Mah-mat-rev_JRNL, MMY15}; and it has also received attention more
generally within the context of neural networks~\cite{Ney17_TR}. While our implicit regularization result
is limited to the Moore-Penrose estimator, this new connection (and
others, described below) between the minimum norm solution of an unregularized
under-determined system and a ridge-regularized least squares solution
offers a simple interpretation for the implicit regularization
observed in modern machine learning architectures.

Our exact non-asymptotic expressions in Theorem~\ref{t:mse} and 
our exact implicit regularization results in Theorem~\ref{t:unbiased}
are derived for the surrogate design, but Figure \ref{f:intro}
suggests that they accurately describe the MSE (up to lower order
terms) also under the standard i.i.d.~design $\X\sim\mu^n$,
particularly when $\mu$ is a multivariate Gaussian.  
As a third result, we can verify this in the cases where there exist
known expressions for the MSE under the i.i.d.~design (standard
Gaussian for the under-determined setting, and 
arbitrary Gaussian for the over-determined one).
\begin{theorem}[Asymptotic consistency of surrogate design]
\label{t:asymptotic}
Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
c_\rho=\frac3{|1-\rho|}$ and
\begin{align*}
  \mu &= \begin{cases}
    \Nc(\zero,\I)&\text{when }n<d-1,\\
    \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
  \end{cases}
\end{align*}
then the absolute difference between surrogate
expressions and the true MSE is bounded as follows:
% &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
%     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
%     \Sigmab\succ\zero,\\
\begin{align*}
\Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
\frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
\end{align*}
\end{theorem}
\begin{remark}
For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
random design can be infinite, whereas the
surrogate MSE is finite and has a closed form expression.
\end{remark}

\noindent
Empirical estimates given in Figure \ref{f:intro} suggest that the
consistency of surrogate expressions holds much more generally than
it is stated above. Based on a detailed empirical analysis described in
Section~\ref{s:empirical}, we conjecture that an asymptotic consistency
result of the form similar to the statement of Theorem~\ref{t:asymptotic}
holds true in the under-determined regime without the assumption that
$\Sigmab=\I$. In this case, no formula is known for
$\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
Gaussian design naturally extend.

\subsection{Key techniques: surrogate designs and
  determinant preserving matrices}

The standard random design model for linear regression assumes that
each pair $(\x_i^\top,y_i)$ is drawn independently, where the row vector
$\x_i^\top$ comes from some $d$-variate distribution $\mu$ and $y_i=y(\x_i)$ is a random response variable drawn conditionaly on $\x_i$.
Precise theoretical analysis of under-determined regression in this
setting poses significant challenges, even in such special cases as
the Moore-Penrose estimator and a Gaussian data distribution $\mu$.  
Rather than trying to directly analyze the usual i.i.d.~random design
$\X\sim\mu^n$ described above, we modify it slightly by introducing
the notion of a \emph{surrogate random design}, $\Xb\sim S_\mu^n$.  
Informally, the goal of a surrogate random design is to modify an
original design to capture the main properties of the original design,
while being ``nicer'' for theoretical or empirical analysis. 
In particular, here, we will modify the distribution of matrix $\X$ so
as to: 
\begin{enumerate}
  \item closely preserve the behavior of the Moore-Penrose estimator
from the i.i.d.~design; and
  \item obtain exact expressions for double descent in terms of the
    mean squared error.
\end{enumerate}
A key element in the construction of our surrogate designs involves rescaling the measure $\X\sim\mu^n$ by the pseudo-determinant $\pdet(\X\X^\top)$, i.e., a product of the non-zero eigenvalues. 
A similar type of determinantal design was suggested in prior work
\cite{correcting-bias-journal}, but it was restricted there only to
$n\geq d$.  
We broaden this definition by not only allowing the sample size to be
less than $d$, but also allowing it to be randomized.  
Our definition of a determinantal design matrix $\Xb$ follows by
expressing $\E[F(\Xb)]$ for any real-valued function $F(\cdot)$ as
(see Definition~\ref{d:det}):  
\begin{align*}
  \E[F(\Xb)]\ \propto\ \E[\pdet(\X\X^\top)F(\X)],
  \end{align*}
where $\X\sim\mu^K$ and $K$ is a random variable.
Then, we define (in Definition~\ref{d:surrogate}) our surrogate design $S_\mu^n$ for each $n>0$ as a determinantal design with a carefully chosen random variable $K$, so that the expected sample size is equal to $n$ and so that it is possible to derive closed form expressions for the MSE. 
We achieve this by using modifications of the Poisson distribution to construct the variable $K$.

The key technical contribution that allows us to derive the MSE for determinantal designs is the concept of \emph{determinant preserving random matrices}, a notion that we expect to be useful more generally. 
Specifically, in Section~\ref{s:dp} we define a class of $d\times d$ random
matrices $\A$ for which taking the determinant commutes with taking
the expectation, for the matrix itself and any of its square submatrices (see Definition~\ref{d:main}):
\begin{align*}
  \E\big[\!\det(\A_{\Ic,\Jc})\big] =
  \det\!\big(\E[\A_{\Ic,\Jc}]\big)\quad \text{for all }\Ic,\Jc\subseteq
  [d]\text{ s.t. }|\Ic|=|\Jc|.
\end{align*}
  Not all random matrices satisfy this property, however many
interesting and non-trivial examples can be found. Constructing
these examples is facilitated by the closure properties that this
class enjoys. In particular, if $\A$ and $\B$ are determinant
preserving and independent, then $\A+\B$ and $\A\B$ are also
determinant preserving (see Lemma \ref{t:ring}). We use these
techniques to prove a number of determinantal expectation
formulas. For example, we show that if $\X\sim\mu^K$, where $K$ is a
Poisson random variable, then:
\begin{align*}
  \text{(Lemma \ref{l:poisson})}\quad  \E\big[\det(\X^\top\X)\big]
  &= \det\!\big(\E[\X^\top\X]\big),\\
\text{(Lemma \ref{l:normalization})}\quad  \E\big[\det(\X\X^\top)\big]
&= \ee^{-\E[K]}\det\!\big(\I + \E[\X^\top\X]\big). 
\end{align*}
These formulas are used to derive the normalization constants for our
surrogate design distribution, which are later used in proving
Theorems~\ref{t:mse} and~\ref{t:unbiased}.

\subsection{Related work}
\label{s:related-work}

There is a large body of related work, which for simplicity we cluster into three groups.

\textbf{Double descent.}
The double descent phenomenon (a term introduced by \cite{BHMM19})
corresponds to the phase transition in the generalization error that
occurs when the ratio between the model complexity and the
sample size crosses the so-called interpolation threshold. It has been observed empirically
in a number of learning models, including neural networks
\cite{BHMM19,GJSx19_TR}, kernel methods \cite{BMM18_TR,BRT18_TR},
nearest neighbor models \cite{BHM18_TR}, and decision trees \cite{BHMM19}. The
theoretical analysis of double descent, and more broadly the generalization
properties of interpolating estimators, have primarily focused on various forms of
linear regression. The most comparable
to our work are \cite{BLLT19_TR,LR18_TR} and \cite{HMRT19_TR}, who provide
non-asymptotic upper/lower bounds and asymptotic formulas,
respectively, for the generalization error of the Moore-Penrose 
estimator under essentially the same i.i.d.~random design setting as
ours. On the other hand, \cite{MVSS19_TR} provide bounds for the error
of the ideal linear interpolator (instead of the minimum norm
one). Note that while we analyze the classical mean squared error, many
works focus on the squared prediction error instead (some of them still
refer to it as the MSE).
Another line of literature deals with linear regression in the
so-called \emph{misspecified} setting, where the set of observed
features does not match the feature space in
which the response model is linear
\cite{belkin2019two,HMRT19_TR,Mit19_TR,MM19_TR}, e.g., when the
learner observes a random subset of $d$ features from a larger
population. This is an important distinction, 
because it allows varying the model complexity by changing the number
of observed features while keeping the linear model fixed (see further
discussion in Section \ref{s:conclusions}). We believe
that our results can be extended to this important setting, and we leave
this as a direction for future work.

% Here are things to cite about double descent and related things.
% Original double descent papers:
% \cite{BHMM19}
% \cite{BMM18_TR}
% \cite{BRT18_TR}
% \cite{belkin2019two}
% Other double descent papers:
% \cite{AS17_TR}
% \cite{BLLT19_TR}
% \cite{BHM18_TR}
% \cite{BFF19_TR}
% \cite{COB18_TR}
% \cite{GJSx19_TR}
% \cite{HMRT19_TR}
% \cite{LR18_TR}
% \cite{MBB17_TR}
% \cite{MM19_TR}
% \cite{Mit19_TR}
% \cite{MVSS19_TR}
% \cite{NMBx18_TR}
% \cite{SKS18_TR}
% \cite{SGAx18_TR}
% \cite{WGSx19_TR}
% \michael{Say something about how we use MSE but others use one of a wide range of slightly different generalization metrics. Would you put in a draft, and then I can make a pass.}

\textbf{RandNLA.}
Randomized numerical linear algebra
\cite{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR} has
traditionally focused on obtaining algorithmic improvements for tasks
such as least squares and low-rank approximation via techniques that include
sketching \cite{sarlos-sketching} and i.i.d.~leverage score sampling
\cite{Drineas2006sampling}. However, there has been growing
interest in understanding the statistical properties of these
randomized methods \cite{MMY15,GarveshMahoney_JMLR}, for example
looking at the mean squared error of the least squares estimator
obtained via i.i.d.~subsampling under the standard linear response
model. Determinantal sampling methods (a.k.a.~volume sampling, or
determinantal point processes), which first found their way
into RandNLA in the context of low-rank approximation
\cite{pca-volume-sampling}, have been recently shown to combine
strong worst-case guarantees with elegant statistical properties.
In particular, \cite{unbiased-estimates} showed that the
least-squares estimator subsampled via the so-called size $d$ volume
sampling (loosely corresponding to the special case of our surrogate design
$S_\mu^n$ where $n=d$) is an unbiased estimator that admits exact
formulas for both the expected square loss (a worst-case metric)
and the mean squared error (a statistical metric). These results were
developed further by
\cite{leveraged-volume-sampling,correcting-bias,minimax-experimental-design},
however they were still limited to the over-determined setting (with
the exception of
\cite{regularized-volume-sampling,bayesian-experimental-design} who
gave upper bounds on the mean squared error of the ridge estimator
under different determinantal samplings). Also in the over-determined setting,
\cite{correcting-bias-journal} provided evidence for the fact that
determinantal rescaling can be used to modify the original data
distribution (particularly, a multivariate Gaussian) without a
significant distortion to the estimator, while making certain
statistical quantities expressible analytically. We take this
direction further by analyzing the unregularized least squares
estimator in the under-determined setting which is less well
understood, partly due to the presence of implicit regularization.

\textbf{Implicit regularization.}
The term implicit regularization typically refers to the notion that approximate
computation (e.g., rather than exactly minimizing a function $f$,
instead running an approximation algorithm to get an approximately
optimal solution) can implicitly lead to
statistical regularization (e.g., exactly minimizing an objective of
the form $f + \lambda g$, for some well-specified $\lambda$ and $g$).  
See~\cite{MO11-implementing, PM11, GM14_ICML} and references therein
for early work on the topic; and see~\cite{Mah12} for an overview. 
More recently, often motivated by neural networks, there has been work
on implicit regularization that typically considered SGD-based
optimization algorithms. 
See, e.g., theoretical results on simplified
models~\cite{NTS14_TR,Ney17_TR,SHNx17_TR,GWBNx17_TR,ACHL19_TR,KBMM19_TR}
as well as extensive empirical and phenomenological results on state-of-the-art neural
network models~\cite{MM18_TR,MM19_HTSR_ICML}.
The implicit regularization observed by us is different in that it is
not caused by an inexact approximation algorithm (such as SGD) but rather by the
selection of one out of many exact solutions (e.g., the minimum norm
solution). In this context, most relevant are the 
asymptotic results of~\cite{LJB19_TR} (which used the asymptotic risk
results for ridge regression of~\cite{DW15_TR}) and~\cite{KLS18_TR}. 
Our non-asymptotic results are also related to recent work in
RandNLA on the expectation of the
inverse~\cite{determinantal-averaging} and generalized
inverse~\cite{MDK19_TR} of a subsampled matrix.

\section{Surrogate random designs}
\label{s:determinantal}

In this section, we provide the definition of our surrogate random
design $S_\mu^n$, where $\mu$ is a $d$-variate probability measure and
$n$ is the sample size. This distribution is used in place
of the standard random design $\mu^n$ consisting of $n$ row vectors drawn
independently from $\mu$. Our surrogate design uses determinantal
rescaling to alter the joint distribution of the vectors so that
certain expected quantities (such as the mean squared error of the
Moore-Penrose estimator) can be expressed in a closed form.
We start by introducing notation.
%to be used throughout the rest of the paper.

%\paragraph{Preliminaries}
\textbf{Preliminaries.}
The set $\{1,...,n\}$ will be denoted by $[n]$. 
For an $n\times n$ matrix $\A$, we use $\pdet(\A)$ to denote the pseudo-determinant of $\A$,
which is the product of non-zero eigenvalues. For index subsets $\Ic$
and $\Jc$, we use $\A_{\Ic,\Jc}$ to denote the submatrix of $\A$ with
rows indexed by $\Ic$ and columns indexed by $\Jc$. We may write
$\A_{\Ic,*}$ to indicate that we take a subset of rows. We use $\adj(\A)$ to
denote the adjugate of $\A$, defined as follows: the
$(i,j)$th entry of $\adj(\A)$ is
$(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$. We will
use two useful identities related to the adjugate: (1)
$\adj(\A)=\det(\A)\A^{-1}$ for invertible $\A$, and (2)
$\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$.
% A formula called Sylvester's theorem
% relates the adjugate and the determinant: $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$.
For a probability measure $\mu$ over $\R^d$, we use $\x^\top\!\sim\mu$
to denote a random row vector $\x^\top$ sampled according to this distibution.
We let $\X\sim\mu^k$ denote a $k\times d$ random matrix with rows
drawn i.i.d.~according to $\mu$, and the $i$th row is denoted as $\x_i^\top$.  
We also let $\Sigmab_\mu=\E_{\mu}[\x\x^\top]$, where $\E_{\mu}$ refers to
the expectation with respect to $\x^\top\!\sim\mu$, assuming throughout that
$\Sigmab_\mu$ is well-defined and positive definite.
We use $\Poisson(\gamma)_{\leq a}$ as the Poisson distribution restricted to values less than or equal to $a$, and a similar convention is used for the restriction to values greater or equal $a$. 
Finally, we use $\#(\X)$ to denote the number of rows of $\X$.

We now define a family of determinantal distributions over random matrices $\Xb$,
where not only the entries but also the number of rows is
randomized. 
This randomized sample size is a crucial property of our
designs that enables our analysis. 
Our definition follows by
expressing $\E[F(\Xb)]$ for real-valued functions $F:\bigcup_{k\geq 0}\R^{k\times
  d}\rightarrow \R$ (the expectation may be undefined for some functions).  
\begin{definition}\label{d:det}
  Let $\mu$ satisfy Assumption~\ref{a:general-position} and let $K$ be
  a random variable over non-negative integers. A determinantal design
    $\Xb\sim \Det(\mu,K)$ is a 
distribution such that~for any $F(\cdot)$ as above,
\begin{align*}
  \E\big[F(\Xb)\big]\  \propto\ \E[\pdet(\X\X^\top)F(\X)]
  \quad\text{for}\quad\X\sim\mu^K.
\end{align*}
\end{definition}

\noindent
Setting $F(\cdot)$ to 1, observe that the proportionality constant
must be $1/\E[\pdet(\X\X^\top)]$. The above definition can be
interpreted as rescaling the density function of $\mu^K$ by the
pseudo-determinant, and then renormalizing it.

We now construct our surrogate design $S_\mu^n$ by appropriately
selecting the random variable $K$. 
One might be tempted to use the obvious choice of $K=n$, but this does \emph{not} result in simple closed form expressions for the MSE in the under-determined regime (i.e., $n<d$), which is the regime of primary interest to us. 
Instead, we derive our random variables $K$ from the Poisson distribution.

\begin{definition}\label{d:surrogate}
For $\mu$ satisfying Assumption~\ref{a:general-position},
define surrogate design $S_\mu^n$ as $\Det(\mu,K)$ where: 
    \begin{enumerate}
\item if $n<d$, then $K\sim \Poisson(\gamma_n)_{\leq d}$ with
 $\gamma_n$ being the solution of $n=\tr(\Sigmab_\mu(\Sigmab_\mu+\frac1{\gamma_n}\I)^{-1})$, 
\item if $n=d$, then we simply let $K=d$,
\item if $n>d$, then $K\sim\Poisson(\gamma_n)_{\geq d}$ with $\gamma_n=n-d$.
\end{enumerate}
\end{definition}

\noindent
Note that the under-determined case, i.e., $n<d$, is restricted to $K\leq d$ so that, under Assumption~\ref{a:general-position}, $\pdet(\X\X^\top)=\det(\X\X^\top)$ with probability 1. 
On the other hand in the over-determined case, i.e., $n>d$, we have
$K\geq d$ so that $\pdet(\X\X^\top)=\det(\X^\top\X)$. In the special case
of $n=d=K$ both of these equations are satisfied: $\pdet(\X\X^\top)=\det(\X^\top\X)=\det(\X\X^\top)=\det(\X)^2$.

The first non-trivial property of the surrogate design $S_\mu^n$ is
that the expected sample size is in fact always equal to $n$, which we
prove at the end of this section.
\begin{lemma} \label{l:size}
Let $\Xb\sim S_\mu^n$ for any $n>0$. 
 Then, we have $\E[\#(\Xb)] = n$.
\end{lemma}

 Our general template for computing expectations under
 a surrogate design $\Xb\sim\S_\mu^n$ is to use the following expressions based on the
i.i.d.~random design $\X\sim\mu^K$:
\begin{align*}
  \E[F(\Xb)] &=\begin{cases}
    \frac{\E[\det(\X\X^\top)F(\X)]}{\E[\det(\X\X^\top)]}
   \quad K\sim\Poisson(\gamma_n)&\text{for }n<d,\\[2mm]
    \frac{\E[\det(\X)^2F(\X)]}{\E[\det(\X)^2]}\hspace{3mm}
    \quad K=d&\text{for }n=d,\\[2mm]
    \frac{\E[\det(\X^\top\X)F(\X)]}{\E[\det(\X^\top\X)]}
   \quad K\sim\Poisson(\gamma_n)&\text{for }n>d.    
  \end{cases}
\end{align*}
These formulas follow from Definitions \ref{d:det} and
\ref{d:surrogate} because the determinants $\det(\X\X^\top)$ and
$\det(\X^\top\X)$ are non-zero precisely in the regimes $n\leq d$ and
$n\geq d$, respectively, which is why we can drop the restrictions on the
range of the Poisson distribution.
Crucially, the normalization constants for computing the expectations
can be obtained using the following formulas: if $\X\sim\mu^K$ then 
\begin{align*}
\text{(Lemma~\ref{l:normalization})}&&
\E\big[\det(\X\X^\top)\big]  &= \ee^{-\gamma_n}\det(\I +
\gamma_n\Sigmab_\mu)&&\text{for } K\sim\Poisson(\gamma_n),\ n<d,\\
  \text{(Lemma~\ref{l:cb})}&& \E\big[\det(\X)^2\big] &=
  d!\det(\Sigmab_\mu),&&\text{for } K=n=d,\\
  \text{(Lemma~\ref{l:poisson})}&& \E\big[\det(\X^\top\X)\big]
&=\det(\gamma_n\Sigmab_\mu),&&  \text{for }K\sim\Poisson(\gamma_n),\ n>d.
\end{align*}
\begin{remark}
We will use $Z_\mu^n$ as a shorthand for the above normalization constants.
\end{remark}
% For the case of $n=d$, the normalization constant can be found in the
% literature \cite{expected-generalized-variance,correcting-bias}:
We prove Lemmas \ref{l:poisson} and \ref{l:normalization} in Section
\ref{s:dp} by introducing the
concept of determinant preserving random matrices. The lemmas play a 
crucial role in deriving a number of new expectation formulas for the
under- and over-determined surrogate designs that we use to prove Theorems~\ref{t:mse}
and~\ref{t:unbiased} in Section~\ref{s:expectations}.
On the other hand, Lemma  \ref{l:cb} and the design $S_\mu^d$ can be found in the
literature \cite{expected-generalized-variance,correcting-bias-journal},
and we will rely on those known results in this special case. 
Importantly, the $n=d$ case offers a continuous transition between the under-
and over-determined regimes because the distribution $S_\mu^n$
converges to $S_\mu^d$ when $n$ approaches $d$ from above and
below. Another important property of the design $S_\mu^d$ is that it
can be used to construct an over-determined design for any $n>d$. A
similar version of this result was also previously shown by
\cite{correcting-bias-journal} for a different determinantal design.
  \begin{lemma}\label{l:decomposition}
Let $\Xb\sim S_\mu^d$ and $\X\sim \mu^K$, where
$K\sim\Poisson(\gamma)$. Then the matrix composed of a random
permutation of the rows from $\Xb$ and $\X$ is distributed according to
$S_\mu^{d+\gamma}$.
\end{lemma}
\begin{proof}
Let $\Xt$ denote the matrix constructed from the permuted rows of
$\Xb$ and $\X$.  Letting $\Z\sim\mu^{K+d}$, we derive the expectation
$\E[F(\Xt)]$ by summing over the possible index subsets  $S\subseteq
[K+d]$ that correspond to the rows coming from $\Xb$:
\begin{align*}
\E[F(\Xt)] &= \E\bigg[\frac1{{K+d\choose
  d}}\sum_{S:\,|S|=d}\frac{\E[\det(\Z_{S,*})^2F(\Z)\mid
  K]}{d!\det(\Sigmab_\mu)}\bigg]\\
  &=\sum_{k=0}^\infty
    \frac{\gamma^k\ee^{-\gamma}}{k!}\,\frac{\gamma^dk!}{(k+d)!}\,
    \frac{\E\big[\sum_{S:\,|S|=d}\det(\Z_{S,*})^2F(\Z)\mid
    K=k\big]}{\det(\gamma\Sigmab_\mu)}\\
  &\overset{(*)}{=}\sum_{k=0}^\infty
    \frac{\gamma^{k+d}\ee^{-\gamma}}{(k+d)!}\,\frac{\E[\det(\Z^\top\Z)F(\Z)\mid K=k]}{\det(\gamma\Sigmab_\mu)},
\end{align*}
where $(*)$ uses the Cauchy-Binet formula to sum over all subsets $S$
of size $d$. Finally, since the sum shifts from $k$
to $k+d$, the last expression can be rewritten as
$\E[\det(\X^\top\X)F(\X)]/\det(\gamma\Sigmab_\mu)$, where recall that
$\X\sim\mu^K$ and $K\sim\Poisson(\gamma)$, matching the definition of $S_\mu^{d+\gamma}$.
\end{proof}
We now return to the proof of Lemma \ref{l:size}, where we establish
that the expected sample size of $S_\mu^n$ is indeed $n$.
\begin{proof} (of Lemma \ref{l:size}) \
  The result is obvious when $n=d$, whereas
  for $n>d$ it is an immediate consequence 
  of Lemma \ref{l:decomposition}.
  Finally, for $n<d$ the expected sample
  size follows as a corollary of a more general expectation formula
  proven in Section \ref{s:expectations}, which states that
  \begin{align*}
\text{(Lemma \ref{l:proj})} \qquad\E\big[\I - \Xb^\dagger\Xb\big] =
    (\gamma_n\Sigmab_\mu + \I)^{-1},
  \end{align*}
  where $\Xb^\dagger\Xb$ is the orthogonal projection onto
  the subspace spanned by the rows of $\Xb$. Since the rank of this
  subspace is equal to the number of the rows, we have
  $\#(\Xb)=\tr(\Xb^\dagger\Xb)$, so
  \begin{align*}
    \E\big[\#(\Xb)\big] = d - \tr\big((\gamma_n\Sigmab_\mu +
    \I)^{-1}\big) =
    \tr\big(\gamma_n\Sigmab_\mu(\gamma_n\Sigmab_\mu+\I)^{-1}\big) = n,
  \end{align*}
  which completes the proof.
\end{proof}

\section{Determinant preserving random matrices}
\label{s:dp}

In this section, we introduce the key tool for computing expectation formulas of matrix determinants.
It is used in our analysis of the surrogate design, and it should be of independent interest.

The key question motivating the following definition is: when does taking expectation commute with computing a determinant for a square random matrix? 
\begin{definition}\label{d:main}
A random $d\times d$ matrix $\A$ is called \textbf{determinant
  preserving} (d.p.), if
\begin{align*}
  \E\big[\!\det(\A_{\Ic,\Jc})\big] =
  \det\!\big(\E[\A_{\Ic,\Jc}]\big)\quad \text{for all }\Ic,\Jc\subseteq
  [d]\text{ s.t. }|\Ic|=|\Jc|.
\end{align*}
\end{definition}
Note that from the definition of an adjugate matrix (see Section
\ref{s:determinantal}) it immediately follows that if $\A$ is
determinant preserving then adjugate commutes with expectation for this matrix:
\begin{align}
  \E\big[\big(\!\adj(\A)\big)_{i,j}\big] &=
  \E\big[(-1)^{i+j}\det(\A_{[d]\backslash\{j\},[d]\backslash\{i\}})\big]\nonumber
  \\
&=(-1)^{i+j}\det\!\big(\E[\A_{[d]\backslash\{j\},[d]\backslash\{i\}}]\big)
  = \big(\!\adj(\E[\A])\big)_{i,j}.\label{eq:adj}
\end{align}
We next give a few simple examples to provide some intuition. First, note
that every $1\times 1$ random matrix is determinant preserving simply
because taking a determinant is an identity transfomation in one
dimension. Similarly, every fixed matrix is determinant preserving because
in this case taking the expectation is an identity
transformation. In all other cases, however, Definition \ref{d:main}
has to be verified more carefully. Further examples (positive and
negative) follow.
\begin{example}
If $\A$ has i.i.d. Gaussian entries $a_{ij}\sim\Nc(0,1)$, then
$\A$ is d.p.~because $\E[\det(\A)]=0$.
\end{example}
In fact, it can be shown that all random matrices with independent entries
are determinant preserving. However, this is not a necessary condition.
\begin{example}\label{e:rank-1}
Let $\A = s\,\Z$, where $\Z$ is fixed with $\rank(\Z) = r$, and $s$
is a scalar random variable. Then for $|\Ic|=|\Jc|=r$ we have
\begin{align*}
  \E\big[\det(s\,\Z_{\Ic,\Jc})\big] &= \E[s^r]\det(\Z_{\Ic,\Jc})
                                      =\det\Big(\big(\E[s^r]\big)^{\frac1r}\,\Z_{\Ic,\Jc}\Big), 
\end{align*}
  so if $r=1$ then $\A$ is determinant preserving, whereas if $r>1$
  and $\Var[s]>0$ then it is not.
\end{example}
To construct more complex examples, we show that determinant preserving random matrices are
closed under addition and multiplication. The proof of this result is an extension of an
argument given by \cite{determinantal-averaging} (Lemma~7) for
computing the expected determinant of the sum of rank-1 random matrices.
\begin{lemma}\label{t:ring}
  If $\A,\B$ are independent and d.p.~then
  $\A+\B$ and $\A\B$ are also determinant preserving.
\end{lemma}
\begin{proof}
 First, we show that $\A+\u\v^\top$ is d.p.~for any fixed
 $\u,\v\in\R^d$. Below, we use a standard identity for the rank one
 update of a determinant:
 $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$. It follows that
 for any $\Ic$ and $\Jc$ of the same size,
  \begin{align*}
\E\big[\!\det(\A_{\Ic,\Jc}\!+\u_{\Ic}\v_{\Jc}^\top)\big] &=
    \E\big[\!\det(\A_{\Ic,\Jc}) +
    \v_{\Jc}^\top\adj(\A_{\Ic,\Jc}) \u_{\Ic}\big]\\
    &\overset{(*)}{=}\det\!\big(\E[\A_{\Ic,\Jc}]\big) +
      \v_{\Jc}^\top\adj\!\big(\E[\A_{\Ic,\Jc}]\big) \u_{\Ic}\\
    &=\det\!\big(\E[\A_{\Ic,\Jc} \!+ \u_{\Ic}\v_{\Jc}^\top]\big),
  \end{align*}
  where $(*)$ used \eqref{eq:adj}, i.e., the fact that for d.p.~matrices, adjugate commutes
  with expectation. Crucially, through the definition of an adjugate
  this step implicitly relies on the assumption that all the square
  submatrices of $\A_{\Ic,\Jc}$ are also  determinant preserving.
  Iterating this, we get that $\A+\Z$ is d.p.~for any fixed
  $\Z$. We now show the same for $\A+\B$:
  \begin{align*}
\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\big]
    &=
      \E\Big[\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\mid\B\big]\Big]\\
    &\overset{(*)}{=}\E\Big[\!\det\!\big(\E[\A_{\Ic,\Jc}]\!+\B_{\Ic,\Jc}\big)\Big]\\
      &= \det\!\big(\E[\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc}]\big),
  \end{align*}
  where $(*)$  uses the fact that after conditioning on $\B$ we can
  treat it as a fixed matrix. Next, we show that $\A\B$ is determinant preserving via the Cauchy-Binet formula:
  \begin{align*}
    \E\big[\!\det\!\big((\A\B)_{\Ic,\Jc}\big)\big]
    &= \E\big[\!\det(\A_{\Ic,*}\B_{*,\Jc})\big]\\
    &=\E\bigg[\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\A_{\Ic,S}\big)
      \det\!\big(\B_{S,\Jc}\big)\bigg]\\
&=\!\!\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\E[\A]_{\Ic,S}\big)
                                                \det\!\big(\E[\B]_{S,\Jc}\big)\\
    &=\det\!\big(\E[\A]_{\Ic,*}\, \E[\B]_{*,\Jc}\big)\\
      &= \det\!\big(\E[\A\B]_{\Ic,\Jc}\big),
  \end{align*}
  where recall that $\A_{\Ic,*}$ denotes the submatrix of $\A$
  consisting of its (entire) rows indexed by $\Ic$.
  \end{proof}
Finally, we introduce another important class of d.p.~matrices:
a sum of i.i.d.~rank-1 random matrices with the number of
i.i.d.~samples being a Poisson random variable. Our use of the Poisson
distribution is crucial for the below result to hold. It is an
extension of an expectation formula given by \cite{dpp-intermediate}
for sampling from discrete distributions.
\begin{lemma}\label{l:poisson}
If $K$ is a Poisson random variable and $\A,\B$ are random $K\times d$
matrices whose rows  are sampled as an i.i.d.~sequence of joint pairs of
random vectors, then $\A^\top\B$ is d.p., and in particular,
\begin{align*}
  \E\big[\det(\A^\top\B)\big] &= \det\!\big(\E[\A^\top\B]\big).
  \end{align*}
\end{lemma}

To prove the above result, we will use the following
lemma, many variants of which appeared in the literature
(e.g., \cite{expected-generalized-variance}). We use the one given by
\cite{correcting-bias}.
\begin{lemma}[\cite{correcting-bias}]\label{l:cb}
If the rows of random matrices $\A,\B\in\R^{k\times d}$
  are sampled as an i.i.d.~sequence of $k\geq d$ pairs of joint random vectors, then
\begin{align}
  k^d\,\E \big[\det(\A^\top\B)\big]
  &= \ktd\,\det\!\big(\E[\A^\top\B]\big).
     \end{align}
   \end{lemma}
Here, we use the following standard shorthand: $\ktd =
\frac{k!}{(k-d)!} = k\,(k-1)\dotsm(k-d+1)$. Note that the above result
almost looks like we are claiming that the matrix $\A^\top\B$ is d.p.,
but in fact it is not because $k^d\neq \ktd$. The difference
in those factors is precisely what we are going to correct with the
Poisson random variable. We now present the proof of Lemma
\ref{l:poisson}.
\begin{proof} (of Lemma \ref{l:poisson}) \
Without loss of generality, it suffices to check Definition \ref{d:main} with both $\Ic$ and
$\Jc$ equal $[d]$. We first expand the expectation by
conditioning on the value of $K$ and letting $\gamma=\E[K]$:
    \begin{align*}
      \E\big[\!\det(\A^\top\B)\big]
      &= \sum_{k=0}^\infty
\E\big[\det(\A^\top\B)\mid K\!=\!k\big]\
\Pr(K\!=\!k)\\
      \text{(Lemma \ref{l:cb})}
      \quad&=
        \sum_{k=d}^\infty\frac{k! k^{-d}}{(k-d)!}\det\!\big(\E[\A^\top\B\mid
        K\!=\!k]\big)
        \frac{\gamma^k\ee^{-\gamma}}{k!}\\
      &=\sum_{k=d}^\infty
\Big(\frac\gamma k\Big)^d\det\!\big(\E[\A^\top\B\mid K\!=\!k]\big)
        \frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}.
     %  \\
     %  &=\sum_{k=0}^\infty \det\!\big(\E[\A^\top\B]\big)\,\Pr(K\!=\!k)
     % \ =\ \det\!\big(\E[\A^\top\B]\big).
    \end{align*}
    Note that $\frac\gamma k\,\E[\A^\top\B\mid K\!=\!k]=\E[\A^\top\B]$,
    which is independent of $k$. Thus we can rewrite the above
    expression as:
    \begin{align*}
\det\!\big(\E[\A^\top\B]\big)\sum_{k=d}^\infty\frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}
      =
      \det\!\big(\E[\A^\top\B]\big)\sum_{k=0}^\infty
      \frac{\gamma^{k}\ee^{-\gamma}}{k!}=\det\!\big(\E[\A^\top\B]\big),
    \end{align*}
    which concludes the proof.
\end{proof}
Finally we use Lemma \ref{l:poisson} combined with Lemma
\ref{t:ring} to show the expectation formula needed for obtaining the
normalization constant of the under-determined surrogate
design (proven by setting $\A=\B=\X\sim\mu^K$). Note that the below result is
more general than the normalization constant requires, because it allows the matrices
$\A$ and $\B$ to be different. In fact, we will use this more general
statement later on in our analysis.
\begin{lemma}\label{l:normalization}
If $K$ is a Poisson random variable and $\A$, $\B$ are random $K\times d$
matrices whose rows  are sampled as an i.i.d.~sequence of joint pairs of
random vectors, then
\begin{align*}
  \E\big[\det(\A\B^\top)\big] &= \ee^{-\E[K]}\det\!\big(\I + \E[\B^\top\A]\big).
  \end{align*}
\end{lemma}
\begin{proof}
By Lemma \ref{l:poisson}, the matrix $\B^\top\A$ is determinant
preserving. Applying Lemma \ref{t:ring} we conclude that
$\I+\B^\top\A$ is also d.p., so
\begin{align*}
  \det\!\big(\I+\E[\B^\top\A]\big) = \E\big[\det(\I+\B^\top\A)\big] =\E\big[\det(\I+\A\B^\top)\big],
\end{align*}
where the second equality is known as Sylvester's Theorem.
We now use the following standard determinantal formula.
\begin{lemma}[\cite{dpp-ml}]
  For any $k\times d$ matrices $\A,\B$ we have
  $\det(\I+\A\B^\top)=\sum_{S\subseteq[k]}\det(\A_{S,*}\B_{S,*}^\top)$.
\end{lemma}
We rewrite the expectation of $\det(\I+\A\B^\top)$ by applying the
lemma.  Letting $\gamma=\E[K]$, we obtain:
\begin{align*}
\E\big[\det(\I+\A\B^\top)\big]  &=\E\bigg[\sum_{S\subseteq [K]}\E\big[\det(\A_{S,*}\B_{S,*}^\top)\mid
    K\big]\bigg]\\
  &\overset{(*)}{=}\sum_{k=0}^\infty\frac{\gamma^k\ee^{-\gamma}}{k!} \sum_{i=0}^k{k\choose
    i}\E\big[\det(\A\B^\top)\mid K=i\big]\\
  &=\sum_{i=0}^\infty \E\big[\det(\A\B^\top)\mid K=i\big]\sum_{k\geq
    i}^\infty{k\choose i}\frac{\gamma^k\ee^{-\gamma}}{k!}\\
  &=\sum_{i=0}^\infty
    \frac{\gamma^i\ee^{-\gamma}}{i!}\E\big[\det(\A\B^\top)\mid K=i\big]
    \sum_{k\geq i}^\infty\frac{\gamma^{k-i}}{(k-i)!} \\
  &= \E\big[\det(\A\B^\top)\big]\cdot\ee^\gamma,
\end{align*}
where $(*)$ follows from the exchangeability of the rows of $\A$ and
$\B$, which implies that the distribution of $\A_{S,*}\B_{S,*}^\top$ is the
same for all subsets $S$ of a fixed size $k$.
\end{proof}

\section{Expectation formulas for surrogate designs}
\label{s:expectations}
In this section we prove a number of expectation formulas for
determinantal surrogate designs, which we then use to prove Theorems \ref{t:mse}
and \ref{t:unbiased}. In the process, we derive closed form
expressions for  $\E[\Xb^\dagger\Xb]$, i.e., the expectation of the
orthogonal projection onto the subspace spanned by the the rows of
$\Xb$, and for $\E[\tr((\Xb^\top\Xb)^\dagger)]$, the trace of the
pseudo-inverse of the sample covariance matrix. To our knowledge,
neither of these quantities admit closed form expressions for standard
i.i.d.~random designs such as Gaussian with general covariance (except
for the isotropic case).

\subsection{Proof of Theorem \ref{t:mse}}
\label{s:mse-proof}
Let $y(\x)$ follow the homoscedastic noise model with variance $\sigma^2$
(Assumption \ref{a:linear}). Recall that we have
$\Xb\sim\S_\mu^n$ and $\yb_i=y(\xbb_i)=\xbb_i^\top\w^*+\xi_i$, where
$\xi_i\sim\Nc(0,\sigma^2)$. A standard decomposition of
the MSE of the Moore-Penrose estimator proceeds as follows:
\begin{align*}
  \MSE{\Xb^\dagger\ybb}
  &= \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\\
  &=\E\big[\|\Xb^\dagger\xib\|^2\big] +
    \E\big[\|(\Xb^\dagger\Xb-\I)\w^*\|^2\big]\\
  &=\sigma^2\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big] +
    \w^{*\top}\E\big[\I-\Xb^\dagger\Xb\big]\w^*.
\end{align*}
Thus, our task is to find closed form expressions for the two
expectations above. If $n\geq d$, then the latter goes away because
when $\Xb$ has full column rank then $\I-\Xb^\dagger\Xb=\zero$. When
$n<d$, this expectation is given in the following result. %can be obtained as a consequence of Lemma \ref{l:ridge-under}.
\begin{lemma}\label{l:proj}
If  $\Xb\sim S_\mu^n$ and $n<d$, then we have
\[\E\big[\I-\Xb^\dagger\Xb\big] = (\gamma_n\Sigmab_\mu+\I)^{-1}.\]
\end{lemma}
The proof of Lemma \ref{l:proj} is deferred to Section
\ref{s:unbiased-proof} because it follows as a corollary of a more
general result (Lemma \ref{l:ridge-under}). We next derive the second
expectation needed to compute the MSE. The under- and over-determined
cases are proven separately, starting with the former.
% Note that those
% results require $\mu$ to satisfy general position (Assumption \ref{a:general-position}),
% which implies that if $\X\sim\mu^k$ for $k\leq d$ then $\rank(\X)=k$.
\begin{lemma}\label{l:sqinv-under}
If  $\Xb\sim S_\mu^n$ for $n<d$, then we have
\begin{align*}
    \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
    &={\gamma_n}\big(1- \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big).
\end{align*}
\end{lemma}
\begin{proof}
Let $\X\sim\mu^K$ for $K\sim\Poisson({\gamma_n})$. Note that if
$\det(\X\X^\top)>0$ then using the fact that
$\det(\A)\A^{-1}=\adj(\A)$ for any invertible matrix $\A$, we can write:
  \begin{align*}
    \det(\X\X^\top)\tr\big((\X^\top\X)^{\dagger}\big)
    &= \det(\X\X^\top)\tr\big((\X\X^\top)^{-1}\big) \\
    &= \tr(\adj(\X\X^\top)) \\[-1mm]
    &= \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top),
  \end{align*}
  where $\X_{-i}$ is a shorthand for $\X_{[K]\backslash\{i\},*}$.
Assumption \ref{a:general-position} ensures that
$\Pr\big\{\det(\X\X^\top)>0\big\}=1$, which allows us to write:
  \begin{align*}
Z_\mu^n\cdot \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
    &=\E\bigg[
    \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top)\ \big|\ 
    \det(\X\X^\top)>0\bigg]\cdot\overbrace{\Pr\big\{\det(\X\X^\top)>0\big\}}^{1}\\
    &=\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\E\Big[
      \sum_{i=1}^k\det(\X_{-i}\X_{-i}^\top)\ \big|\  K=k\Big]\\
    &=\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\, k\
      \E\big[\det(\X\X^\top)\mid K=k-1\big]\\
    &=\gamma_n\sum_{k=0}^{d-1}\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}
      \E\big[\det(\X\X^\top)\mid K=k\big]\\
    &=\gamma_n\Big(\E\big[\det(\X\X^\top)\big]\ -\
      \frac{\gamma_n^{d}\ee^{-\gamma_n}}{d!}\E\big[\det(\X)^2\mid K=d\big]
      \Big) \\
    &\overset{(*)}{=}\gamma_n\big(\ee^{-\gamma_n}\det(\I +\gamma_n\Sigmab_\mu) -
      \ee^{-\gamma_n}\det(\gamma_n\Sigmab_\mu)\big),
  \end{align*}
  where $(*)$ uses Lemma \ref{l:normalization} for the first term and
  Lemma \ref{l:cb} for the second term. We obtain the desired result by
  dividing both sides by
  $Z_\mu^n=\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)$. 
\end{proof}
In the over-determined regime, a more general matrix expectation
formula can be shown (omitting the trace). The following result is
related to an expectation formula derived by
\cite{correcting-bias-journal}, however they use a slightly
different determinantal design so the results are incomparable.
\begin{lemma}\label{l:sqinv-over}
If $\Xb\sim S_\mu^n$ and $n>d$, then we
have
\begin{align*}
  \E\big[ (\Xb^\top\Xb)^{\dagger}\big] =
  \Sigmab_\mu^{-1}\cdot \frac{1-\ee^{-\gamma_n}}{\gamma_n}.
\end{align*}
\end{lemma}
\begin{proof}
Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$. Assumption
\ref{a:general-position} implies that for $K\neq d-1$ we have
\begin{align}
  \det(\X^\top\X)(\X^\top\X)^\dagger=\adj(\X^\top\X),\label{eq:adj-over}
  \end{align}
however when $k=d-1$ then \eqref{eq:adj-over} does not hold because
$\det(\X^\top\X)=0$ while $\adj(\X^\top\X)$ may be non-zero. It 
follows that: 
  \begin{align*}
Z_\mu^n\cdot
    \E\big[ (\Xb^\top\Xb)^{\dagger}\big]
    &=\E\big[\det(\X^\top\X)(\X^\top\X)^\dagger\big]\\
    &=\E\big[\adj(\X^\top\X)\big]-
\frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)!}
      \E\big[\adj(\X^\top\X)\mid K=d-1\big]\\
    &\overset{(*)}{=}\adj\!\big(\E[\X^\top\X]\big) -
      \frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)^{d-1}}
      \adj\!\big(\E[\X^\top\X\mid K=d-1]\big)\\
    &=\adj(\gamma_n\Sigmab_\mu) - \ee^{-\gamma_n}\adj(\gamma_n\Sigmab_\mu)\\
    &=\det(\gamma_n\Sigmab_\mu)\,(\gamma_n\Sigmab_\mu)^{-1}(1-\ee^{-\gamma_n})\\
    &=\det(\gamma_n\Sigmab_\mu)\,\Sigmab_\mu^{-1}\cdot\frac{1-\ee^{-\gamma_n}}{\gamma_n},
  \end{align*}
  where the first term in $(*)$ follows from Lemma
  \ref{l:normalization} and \eqref{eq:adj}, whereas the second term comes
  from Lemma 2.3 of \cite{correcting-bias-journal}.
Dividing both sides by $Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ completes the proof.
\end{proof}
Finally, applying the closed form expressions from Lemmas
\ref{l:proj}, \ref{l:sqinv-under} and \ref{l:sqinv-over}, we derive
the formula for the MSE and prove Theorem \ref{t:mse}.
\begin{proof} (of Theorem \ref{t:mse}) \
  First, assume that $n<d$, in which case we have
  $\gamma_n=\frac1{\lambda_n}$ and moreover
  \begin{align*}
    n &= \tr\big(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1}\big)\\
      &=\tr\big((\Sigmab_\mu+\lambda_n\I-\lambda_n\I)(\Sigmab_\mu+\lambda_n\I)^{-1}\big)\\
    &=d - \lambda_n\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big),    
    %\frac{\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)}{d-n},
  \end{align*}
so we can write $\lambda_n$ as $(d-n)/\tr((\Sigmab_\mu+\lambda_n\I)^{-1})$.
  From this and Lemmas \ref{l:proj} and \ref{l:sqinv-under}, we
obtain the desired expression, where recall
  that $\alpha_n = \det\!\big(\Sigmab_\mu (\Sigmab_\mu+\frac1{\gamma_n})^{-1}\big)$:
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} &= \sigma^2\,\gamma_n(1-\alpha_n) +
    \tfrac1{\gamma_n} \,\w^{*\top}(\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\w^*
    \\
    &\overset{(a)}{=}\sigma^2\,\frac{1-\alpha_n}{\lambda_n} +
    \lambda_n\,\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*\\
    &\overset{(b)}{=}\sigma^2\tr\big((\Sigmab_\mu+\lambda_n)^{-1}\big)\frac{1-\alpha_n}{d-n}
      +
      (d-n)\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
      {\tr\big((\Sigmab_\mu+\lambda_n)^{-1}\big)}.
  \end{align*}
  While the expression given after $(a)$ is simpler than the one
after $(b)$, the latter better illustrates how the MSE depends on
the sample size $n$ and the dimension $d$.
  Now, assume that $n>d$. In this case, we have $\gamma_n=n-d$ and apply Lemma
  \ref{l:sqinv-over}:
  \begin{align*}
    \MSE{\Xb^\dagger\ybb}
    &= \sigma^2\,\tr(\Sigmab_\mu^{-1})\,
\frac{1-\ee^{-\gamma_n}}{\gamma_n}
=\sigma^2\,\tr(\Sigmab_\mu^{-1})
\,\frac{1-\beta_n}{n-d}.
  \end{align*}
The case of $n=d$ was shown by \cite{correcting-bias-journal}
(Theorem~2.12). This concludes the proof. 
\end{proof}
\subsection{Proof of Theorem \ref{t:unbiased}}
\label{s:unbiased-proof}
Recall that our goal is to compute the expected value of
$\Xb^\dagger\ybb$ under the surrogate design $S_\mu^n$. Similarly as for Theorem
\ref{t:mse}, the case of $n=d$ was shown by
\cite{correcting-bias-journal} (Theorem 2.10). We break the rest down into the
under-determined case $(n<d)$ and the over-determined case ($n>d$),
starting with the former. Recall that we do \emph{not} require any
modeling assumptions on the responses.
\begin{lemma}\label{l:ridge-under}
If $\Xb\sim S_\mu^n$ and $n<d$, then for any $y(\cdot)$
such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
denoting $\yb_i$ as $y(\xbb_i)$, we have
\begin{align*}
  \E\big[\Xb^\dagger \ybb\big]
  &=
    \big(\Sigmab_\mu+\tfrac1{\gamma_n}\I\big)^{-1}\E_{\mu,y}[y(\x)\,\x].
\end{align*}
\end{lemma}
\begin{proof}
   Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
   $y(\x_i)$ as $y_i$.
  Note that when $\det(\X\X^\top)>0$, then
  the $j$th entry of $\X^\dagger\y$ equals
  $\f_j^\top(\X\X^\top)^{-1}\y$, where $\f_j$ is the $j$th
  column of $\X$, so:
\begin{align*}
  \det(\X\X^\top)\,(\X^\dagger\y)_j
  &= \det(\X\X^\top)\, \f_j^\top(\X\X^\top)^{-1}\y \\
  &=
  \det(\X\X^\top+\y\f_j^\top) - \det(\X\X^\top).
\end{align*}
If $\det(\X\X^\top)=0$, then also
$\det(\X\X^\top+\y\f_j^\top)=0$, so we can write:
\begin{align*}
Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
  &=  \E\big[\det(\X\X^\top)(\X^\dagger\y)_j\big] \\
  &= \E\big[\det(\X\X^\top+\y\f_j^\top)-\det(\X\X^\top)\big]  \\
  &=\E\big[\det\!\big([\X,\y][\X,\f_j]^\top\big)\big] - \E\big[\det(\X\X^\top)\big]\\
  &\overset{(1)}{=}\ee^{-\gamma_n}\det\!\bigg(\I +
    \gamma_n\,\E_{\mu,y}\bigg[\begin{pmatrix}\x\x^\top&
      \!\!\x\, y(\x)\\ x_j\,\x^\top&\!\!
      x_j\,y(\x)\end{pmatrix}\bigg]\bigg)
                          -\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)\\
  &\overset{(2)}{=}\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)\Big(\E_{\mu,y}\big[\gamma_n
    x_j\,y(\x)\big] - \E_{\mu}\big[\gamma_n
    x_j\,\x^\top\big](\I+\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[\gamma_n\x\,
    y(\x)\big]\Big),
\end{align*}
where $(1)$ uses Lemma \ref{l:normalization} twice, with the first
application involving two different matrices $\A=[\X,\y]$ and
$\B=[\X,\f_j]$, whereas $(2)$ is a standard determinantal identity
(see Fact 2.14.2 in \cite{matrix-mathematics}).
  Dividing both sides by $Z_\mu^n$ and letting $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$, we obtain that:
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big]
    &= \gamma_n\v_{\mu,y} - \gamma_n^2\Sigmab_\mu(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y}\\
&=\gamma_n\big(\I - \gamma_n\Sigmab_\mu (\I+\gamma_n\Sigmab_\mu)^{-1}\big)\v_{\mu,y}
=\gamma_n(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y},
  \end{align*}
  which completes the proof.
\end{proof}
We return to Lemma \ref{l:proj} from Section \ref{s:mse-proof}, regarding the expected orthogonal
projection onto the complement of the row-span of $\Xb$, i.e.,
$\E[\I-\Xb^\dagger\Xb]$, which follows as a corollary of Lemma~\ref{l:ridge-under}.
\begin{proof} (of Lemma \ref{l:proj}) \
  We let $y(\x)=x_j$ for each $j\in[d]$ and apply Lemma
  \ref{l:ridge-under} $d$ times, obtaining:
  \begin{align*}
    \I - \E\big[\Xb^\dagger\Xb] = \I -
    (\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\Sigmab_\mu,
  \end{align*}
  from which the result follows by simple algebraic manipulation.
\end{proof}
We move on to the over-determined case, where the ridge regularization
of adding the identity to $\Sigmab_\mu$ vanishes. Recall that we
assume throughout the paper that $\Sigmab_\mu$ is invertible.
\begin{lemma}\label{l:ridge-over}
  If $\Xb\sim S_\mu^n$ and $n>d$, then for any real-valued random function $y(\cdot)$
  such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
denoting $\yb_i$ as $y(\xbb_i)$, we have
 \begin{align*}
  \E\big[\Xb^\dagger \ybb\big]
  &=\Sigmab_\mu^{-1}\E_{\mu,y}\big[y(\x)\,\x\big].
\end{align*}
\end{lemma}
\begin{proof}
   Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
   $y_i=y(\x_i)$. Similarly as in the proof of
   Lemma~\ref{l:ridge-under}, we note that when $\det(\X^\top\X)>0$,
   then 
  the $j$th entry of $\X^\dagger\y$ equals
  $\e_j^\top(\X^\top\X)^{-1}\X^\top\y$, where $\e_j$ is the $j$th
standard basis vector, so:
\begin{align*}
  \det(\X^\top\X)\,(\X^\dagger\y)_j =
  \det(\X^\top\X)\, \e_j^\top(\X^\top\X)^{-1}\X^\top\y =
  \det(\X^\top\X+\X^\top\y\e_j^\top) - \det(\X^\top\X).
\end{align*}
If $\det(\X^\top\X)=0$, then also
$\det(\X^\top\X+\X^\top\y\e_j^\top)=0$. We proceed to compute the
expectation:
\begin{align*}
Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
  &=  \E\big[\det(\X^\top\X)(\X^\dagger\y)_j\big] \\
  &= \E\big[\det(\X^\top\X+\X^\top\y\e_j^\top)-\det(\X^\top\X)\big]  \\
  &=\E\big[\det\!\big(\X^\top(\X+\y\e_j^\top)\big)\big] - \E\big[\det(\X^\top\X)\big]\\
  &\overset{(*)}{=}\det\!\Big(
    \gamma_n\,\E_{\mu,y}\big[\x(\x+ y(\x)\e_j)^\top\big]\Big)
    -\det(\gamma_n\Sigmab_\mu)\\
  &=\det\!\big(\gamma_n\Sigmab_\mu + \gamma_n\E_{\mu,y}[\x\,y(\x)]\e_j^\top\big)
    -\det(\gamma_n\Sigmab_\mu)\\
  &=\det(\gamma_n\Sigmab_\mu)\cdot
    \gamma_n\e_j^\top(\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[y(\x)\,\x\big],
\end{align*}
where $(*)$ uses Lemma \ref{l:poisson} twice (the first time, with
$\A=\X$ and $\B=\X+\y\e_j^\top$). Dividing both sides by
$Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ concludes the proof. 
\end{proof}

We combine Lemmas \ref{l:ridge-under} and \ref{l:ridge-over} to obtain
the proof of Theorem \ref{t:unbiased}.
\begin{proof} (of Theorem \ref{t:unbiased}) \
The case of $n=d$ follows directly from the result of
\cite{correcting-bias} (Theorem~2.10). 
Assume that $n<d$. Then we have
$\gamma_n=\frac1{\lambda_n}$, so the result follows
from Lemma \ref{l:ridge-under}.
% , noting that unlike in the lemma,
% Theorem \ref{t:unbiased} allows for the response model $y(\cdot)$ to be randomized:
% \begin{align*}
%   \E[\Xb^\dagger\ybb] = \E\big[\E[\Xb^\dagger\ybb\mid\ybb]\big] = 
% % \w_{\lambda_n}^* = \argmin_\w\E_\mu\big[(\x^\top\w-y(\x))^2\big] +
% %   \lambda_n\|\w\|^2 =
% %   \big(\Sigmab_\mu+\lambda_n\I\big)^{-1}
% %   \E_{\mu}\big[y(\x)\,\x\big].
% \end{align*}
If $n>d$, then the result follows from Lemma \ref{l:ridge-over}.
\end{proof}

\section{Asymptotic consistency of surrogate Gaussian designs}
\label{s:asymptotic}
In this section we compare the MSE expressions obtained for the
surrogate design with the true MSE of the standard i.i.d.~design for
the case where $\mu$ is a centered Gaussian distribution
$\Nc(\zero,\Sigmab)$. First, we theoretically compare the analytical
expressions available in certain cases for the i.i.d.~design with their
surrogate equivalents, showing that the surrogate expressions are
asymptotically consistent (Theorem \ref{t:asymptotic}). In fact, we
show even more: that the surrogate expressions provide a
multiplicative approximation for the true MSE with the error decaying
at the rate of $O(1/d)$. However, this result
does not cover the under-determined setting for a non-isotropic
Gaussian, because to the best of our knowledge, analytical expressions
for the i.i.d.~design are not available there. To address this case,
which is of primary interest to us, we perform thorough empirical
analysis showing that our surrogate expressions are asymptotically
consistent, with the error decaying as $O(1/d)$, in all of the
non-isotropic Gaussian test-cases that we evaluated. Based on 
this, we formulate Conjectures \ref{c:wishart} and \ref{c:projection},
regarding the asymptotic behavior of the pseudo-inverse of the
singular Wishart distribution and of a random Gaussian projection, that
are of independent interest to multivariate statistics.

\subsection{Proof of Theorem \ref{t:asymptotic}}
Once again, we break down the proof into under- and
over-determined cases, starting with the former. Note that in this
case we require that the covariance be equal to identity.
\begin{lemma}\label{l:asymptotic-under}
Let $\rho=n/d$, $\X\sim \Nc(\zero,\I)^n$ and $y_i=y(\x_i)$ under Assumption
  \ref{a:linear}. If $d>n+1$ then
  \begin{align*}
    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\I,
\w^*,\sigma^2,n)}{\Mc(\I,\w^*,\sigma^2,n)}\ \leq\
\frac1d\cdot \frac{1}{1-\rho-\frac1d} + 3\rho^d.
  \end{align*}
\end{lemma}

\begin{proof} 
  We first recall the standard decomposition of $\MSE{\X^\dagger \y}$:
  \begin{align*}
    \MSE{\X^\dagger \y}
    &=\sigma^2 \E\big[\tr\big((\X^\top \X)^\dagger \big)\big]+
      \w^{*\top}\!\big( \I - \E[\X^\dagger \X]\big) \w^*. 
  \end{align*}
  Since the rows of $\X$ are standard normal random variables,
  $\X\X^\top$ is an $n\times n$ Wishart random matrix with $d>n+1$ degrees
  of freedom. Using the formula for the mean of the Inverse-Wishart
  distribution, it follows that
  \begin{align*}
\E\big[\tr((\X^\top\X)^\dagger)\big] = \E\big[\tr((\X\X^\top)^{-1})\big]= \frac{n}{d - n - 1}.
  \end{align*}
  % and therefore
  % \begin{align*}
  %   \MSE{\X^\dagger \y}
  %   &= \|\w\|_2^2 - \w^\top \E[\X^\dagger \X] \w
  %   + \frac{\sigma^2 n}{d - n - 1}
  % \end{align*}
  % Let $\X = \U \S \V^\top$ be its full singular value decomposition. Notice
  % \begin{align*}
  %   \X^\dagger \X
  %   &= \V \S^\dagger \U^\top \U \S \V^\top \\
  %   &= \V \S^\dagger \S \V^\top \\
  %   &= \V \begin{bmatrix}
  %     \I_n & 0 \\
  %     0 & 0
  %   \end{bmatrix} \V^\top \\
  %   &= \V_n \V_n^\top \\
  %   \w^\top \E[\X^\dagger \X] \w
  %   &= \E[\|\V_n^\top \w\|_2^2]
  % \end{align*}
  % where $\V_n$ denotes the $d \times n$ matrix with columns equal to the
  % $n$ eigenvectors of $\X^\top \X$ with non-zero eigenvalue.
  % But since $\Sigmab = \I$, by rotation invariance
  % $\V_n^\top$ is a change of basis onto a uniformly random
Note that $\X^\dagger\X$ is a uniformly random projection matrix so by
rotational symmetry it follows that
\begin{align*}
  \w^{*\top}\E\big[\X^\dagger\X]\w^* =
  \E\big[\|\X^\dagger\X\w^*\|^2\big] = \frac nd \,\|\w^*\|^2.
\end{align*}
Putting this together we obtain that
  \begin{align*}
    \MSE{\X^\dagger \y}
    &=\frac{\sigma^2 n}{d - n - 1}+ \|\w^*\|^2\,\frac{d - n}{d}.
  \end{align*}
  On the other hand, the surrogate MSE expression can be derived by
observing that for $\Sigmab=\I$ we have
$\tr((\Sigmab+\lambda_n\I)^{-1})=d/(1+\lambda_n)=n$ (see definition of
$\lambda_n$ in Theorem \ref{t:mse}): 
  \begin{align*}
    \Mc(\I, \w^*,\sigma^2,n) =
    \sigma^2n\cdot\frac{1-\alpha_n}{d-n} + \|\w^*\|^2\,\frac{d-n}{d}.
  \end{align*}
  Note that the second term is the same in both cases, even though
  this may not be true for non-isotropic Gaussians. We now compute the
  normalized difference between the 
  expressions, 
  \begin{align*}
 \frac{\MSE{\X^\dagger\y}-\Mc(\I, \w^*,\sigma^2,n)}{\Mc(\I,
    \w^*,\sigma^2,n)}
    &= \frac{\sigma^2n\cdot(\frac1{d-n-1} -
    \frac{1-\alpha_n}{d-n})}{\sigma^2n\cdot\frac{1-\alpha_n}{d-n} +
    \|\w^*\|^2\,\frac{d-n}{d}}\\
&\leq \frac {d-n}{1-\alpha_n}
\Big(\frac1{d-n-1}-\frac{1-\alpha_n}{d-n}\Big)\\
    &=\frac{d-n}{d-n-1}\,\frac{1}{1-\alpha_n}- 1\\
    &=\frac1{d-n-1} + \frac{\alpha_n}{1-\alpha_n}\Big(1+\frac1{d-n-1}\Big).
  \end{align*}
Let $\rho=n/d$. Then $\frac1{d-n-1} = \frac1d\cdot
\frac1{1-\rho-\frac1d}$ and moreover
$\alpha_n = (\frac1{1+\lambda_n})^d=(\frac
nd)^d=\rho^d$. From the assumption that $d>n+1$, we
conclude that $\alpha_n \leq (\frac{d-2}{d})^d\leq \ee^{-2}$
so that $\frac{\alpha_n}{1-\alpha_n}(1+\frac1{d-n-1})\leq
  \frac{2 \alpha_n}{1-\ee^{-2}}\leq 3\rho^d$. This shows the right-hand-side inequality of the
  theorem. That fact that $\MSE{\X^\dagger\y}\geq\Mc(\I,
  \w^*,\sigma^2,n)$ follows easily.
\end{proof}
\begin{lemma}\label{l:asymptotic-over}
Let $\rho=n/d$, $\X\sim \Nc(\zero,\Sigmab)^n$ and $y_i=y(\x_i)$ under Assumption
  \ref{a:linear}. If $n>d+1$ then
  \begin{align*}
    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab,
\w^*,\sigma^2,n)}{\Mc(\Sigmab,\w^*,\sigma^2,n)}\ \leq\
\frac1d\cdot \frac{1}{\rho-1-\frac1d} + 3(\ee^{1-\rho})^d.
  \end{align*}
\end{lemma}
\begin{proof}
The MSE for the over-determined Gaussian design can be derived by
using the formula for the mean of the Inverse-Wishart distribution:
\begin{align*}
  \MSE{\X^\dagger\y}=\sigma^2\tr\big(\E[(\X^\top\X)^{-1}]\big) =
  \frac{\sigma^2\tr(\Sigmab^{-1}) }{n-d-1}.
\end{align*}
To compute the normalized difference we follow similar derivations as
in the proof of Lemma \ref{l:asymptotic-under}:
\begin{align*}
 \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)}{\Mc(\Sigmab,
    \w^*,\sigma^2,n)}
    &=  \frac {n-d}{1-\beta_n}
      \Big(\frac1{n-d-1}-\frac{1-\beta_n}{n-d}\Big)\\
  &\leq\frac1d\cdot\frac1{\rho-1-\frac1d} + \frac{2\beta_n}{1-\beta_n}.
\end{align*}
Recall that $\beta_n = \ee^{d-n} = (\ee^{1-\rho})^d$ and for $n-d\geq
2$ we have $\frac2{1-\beta_n}\leq 3$. The desired inequalities follow immediately.
\end{proof}
Theorem \ref{t:asymptotic} now follows as a consequence of
Lemmas \ref{l:asymptotic-under} and \ref{l:asymptotic-over}.
\begin{proof} (of Theorem \ref{t:asymptotic}) \
  Since $\frac1d\leq \frac13|1-\rho|$ and $\ee^{-|1-\rho|d}\leq
  \frac1{2d|1-\rho|}$, it follows that
  \begin{align*}
    \frac1d\cdot \frac{1}{|1-\rho|-\frac1d} + 3(\ee^{-|1-\rho|})^d
    &\leq \frac1d\cdot \frac{1}{|1-\rho|-\frac13|1-\rho|} +
      \frac1d\cdot\frac{3}{2|1-\rho|}=\frac {c_\rho}{d}.
  \end{align*}
  The case of $n>d+1$ now follows from Lemma
  \ref{l:asymptotic-over}. Also, since $\rho\leq \ee^{\rho-1}$, for
$n<d-1$ we have $3\rho^d\leq 3(\ee^{-|1-\rho|})^{d}$, so the same bound follows
from Lemma \ref{l:asymptotic-under}.
\end{proof}

\input{continuous-empirical.tex}

\section{Conclusions}
\label{s:conclusions}
\begin{wrapfigure}{r}{0.45\textwidth}
 \vspace{-.8cm}
 \includegraphics[width=.45\textwidth]{figs/descent-model-log}
 \vspace{-6mm}
  \caption{
Surrogate MSE as a function of $d/n$, with $n$
fixed to $100$ and varying $d$.}
  \label{f:model}
  \vspace{-3mm}
\end{wrapfigure}

We derived exact non-asymptotic expressions for the MSE of the
Moore-Penrose estimator in the standard regression task, reproducing
the double descent phenomenon as the sample size crosses between the
under- and over-determined regime. To achieve this, we modified the
standard i.i.d.~random design distribution using a determinantal
rescaling to obtain a surrogate design which admits exact MSE expressions,
while capturing the key properties of the i.i.d.~design. We
also provided a result that relates the expected value of the
Moore-Penrose estimator of a training sample in the under-determined regime (i.e., the
minimum norm solution) to the ridge-regularized least squares solution
for the population distribution, thereby providing an interpretation for the
implicit regularization resulting from over-parameterization.

An important technical issue is that, in this work, we focus on the classical \emph{well-specified}
linear regression task, where the underlying response model is linear
with respect to the observed feature space. A significant effort in
the related literature (see Section \ref{s:related-work}) has been
directed towards a number of \emph{misspecified} linear regression
tasks, where the set of $d$ observed features is different than the
set of $D$ features which define the linear model (typically, $d\ll
D$). Crucially, unlike in the well-specified task, here it is possible
to vary the number of observed features without changing the
underlying linear model. 
Recent work \cite{HMRT19_TR} has compared how
varying the feature dimension affects the (asymptotic) generalization
error for both well-specified and misspecified tasks, however their
analysis was limited to certain special settings such as an
isotropic data distribution. As an additional 
point of comparison, in Figure~\ref{f:model} we plot the MSE
expressions of Theorem \ref{t:mse} for our well-specified setting when varying the
feature dimension~$d$. The model is chosen just like in
Figure~\ref{f:intro}, where the covariances $\Sigmab_\mu$ are diagonal
with condition number $\kappa$ and exponentially decaying spectrum
scaled so that $\tr(\Sigmab_\mu^{-1})=d$. We also use $\sigma^2=1$ and
$\w^*=\frac1{\sqrt{d}}\one_d$. Qualitatively, our plots follow the
trends outlined by \cite{HMRT19_TR} for the isotropic case (i.e., $\kappa=1$), 
but the spectral decay of the covariance matrix (captured by our new
MSE expressions) does have a significant effect on the descent curve
in the under-determined regime. Note that the plots achieve their
minimum as $d$ goes to zero because in the well-specified task as
the complexity of the prediction model decreases, so does the complexity of the
true response model. Nevertheless, even in the well-specified setting
we achieve non-trivial generalization in the under-determined regime,
as seen by the fact that the MSE curve goes below the error of the
null estimator, $\mathrm{MSE}[\zero]=\|\w^*\|^2=1$.

Our work opens up a number of new directions for future research.
Most obviously, this includes extending our surrogate analysis to the misspecified linear regression discussed above. 
In addition, in Section \ref{s:asymptotic}, we provided two conjectures, the goal of which is to bound the difference between the results obtained for our surrogate designs and the true values corresponding to the i.i.d.~design, in the case of multivariate Gaussians. 
We believe that addressing these conjectures will have a broader impact on our understanding of non-isotropic Gaussian designs. 
Finally, it remains open whether the analysis we provided for the mean squared error can be reproduced in the context of mean squared \emph{prediction} error, which is relevant in many machine learning tasks.

%\noingent
\paragraph{Acknowledgements.}
We would like to acknowledge ARO, DARPA, NSF, and ONR for providing partial support of this work.

%\bibliographystyle{plain}
\bibliographystyle{alpha}
  \bibliography{pap}

  \iffalse

  \appendix

  \section{Literature review of double descent}
  \begin{itemize}
\item \cite{BHMM19} Introduction of double descent. Empirical evidence and
experiments for: random fourier features (interpreted as a class of
two-layer neural network), decision trees and ensemble
methods. Uses the term \emph{inductive bias} to describe choosing a
minimum norm solution.

\item\cite{BMM18_TR}
Empirical evaluation of the performance of Gaussian and Laplacian
kernels for classification in the interpolation regime. Draws a distinction between
\emph{implicit regularization} (e.g., early stopping) and
\emph{inductive bias} (e.g., minimum norm solutions).

\item\cite{BRT18_TR}
An error bound for a special interpolating kernel
method, showing that you can generalize despite zero training error.

\item\cite{belkin2019two} Two theoretical models of double
  descent where features are chosen as a uniformly random subset of a
  larger design matrix. Case 1: design matrix is a random i.i.d
  isotropic Gaussian. Case 2: discrete fourier transform matrix. They
  obtain exact expressions in those special cases.

\item \cite{BLLT19_TR} Non-asymptotic upper- and lower-bounds for
  least squares in infinite dimensional Hilbert spaces. Input data is
  Gaussian with arbitrary covariance. Two notions of
 effective rank, one of them is an extension of the stable rank.

\item \cite{BHM18_TR} Analysis of nearest neighbor in the
   interpolating regime.

 \item \cite{BFF19_TR} Analysis of least squares
   under the Predictive Normalized Maximum Likelihood (pNML) model,
   showing the capacity for generalization in the over-parameterized
   regime.
 \item \cite{COB18_TR} Analysis of lazy training, i.e., when the
   optimization path for a non-convex loss function follows
   that of a linearized model. The conclusion is that lazy training
   does not explain the success of over-parameterized neural networks.

 \item \cite{GJSx19_TR} Bias and variance analysis of deep neural
   networks via a Neural Tangent Kernel model. Suggested an ensemble
   approach of averaging  several over-parameterized networks (for us,
   the ensemble approximates the ridge solution).

 \item \cite{HMRT19_TR} Asymptotic analysis of over-parameterized
   least squares (and ridge), with exact expressions for isotropic
   features (not necessarily Gaussian), and then for general
   covariances (correlated features) the asymptotic expressions depend on the Stieltjes
   transform, which does not in general have a closed form. They show
   that the performance of over-parameterized least squares depends on
   the signal-to-noise ratio (SNR). They also show that optimally
   tuned ridge is always better than least squares. Finally, they
   differentiate between the well-specified model and the
   \emph{misspecified model}, where only a subset of features is
   observed by the learner.

 \item \cite{MM19_TR} Asymptotic analysis of the random features
   model with a nonlinear activation (i.e., a two-layer neural network
   with fixed first layer). The random features are drawn uniformly
   from the unit sphere. The expressions are fairly complicated but
   also quite accurate. This is not directly comparable to us because
   our model is well-specified (i.e., we get all the featuers). They
   point out that for random features the peak in double descent
   occurs both for the bias and the variance, unlike for
   well-specified models where only variance exhibits a peak.

 \item \cite{LR18_TR}   Upper bounds for the generalization error of
   Kernel ``Ridgeless''   Regression, i.e, least squares in the
   over-parameterized regime over an infinite dimensional Hilbert
   space. Arbitrary covariance matrices are allowed, and an inner
   product kernel with a non-linear smooth activation is
   assumed. Additionally, they
   perform analysis for different rates of exponentially decaying
   eigenvalues.

\item   \cite{MVSS19_TR} They derive the test error of the “ideal interpolator”
   for well-specified linear regression,
   showing that it can be significantly better than the minimum norm
   one. Also, they analyze sparsity seeking interpolators and they
   look at several random feature models (but always well-specified).

\item  \cite{Mit19_TR} They introduce a model for misspecified sparse
  regression and provide exact asymptotic expressions for the
  generalization error of both the minimum norm and the regularized
  estimators, where the norm can be either $\ell_2$ or $\ell_1$. 
\end{itemize}
\fi

\end{document}

### 2022-08-23 design/continuous_colt/appendix.tex
Commit original repos

\section{Proof of Lemma~\ref{l:size}}
\label{appx: proof-of-l-size}

We first record an important property of the design $S_\mu^d$
which can be used to construct an over-determined design for any $n>d$. A similar
version of this result was also previously shown by
\cite{correcting-bias-journal} for a different determinantal design.

\begin{lemma}\label{l:decomposition}
  Let $\Xb\sim S_\mu^d$ and $\X\sim \mu^K$, where
  $K\sim\Poisson(\gamma)$. Then the matrix composed of a random
  permutation of the rows from $\Xb$ and $\X$ is distributed according to
  $S_\mu^{d+\gamma}$.
\end{lemma}

\begin{proof}
Let $\Xt$ denote the matrix constructed from the permuted rows of
$\Xb$ and $\X$.  Letting $\Z\sim\mu^{K+d}$, we derive the probability
$\Pr\big\{\Xt\!\in\! E\big\}$ by summing over the possible index subsets  $S\subseteq
[K+d]$ that correspond to the rows coming from $\Xb$:
\begin{align*}
  \Pr\big\{\Xt\in E\big\} &= \E\bigg[\frac{1}{\binom{K+d}{d}}
  \sum_{S:\,|S|=d}\frac{\E[\det(\Z_{S,*})^2\one_{[\Z\in E]}\mid
  K]}{d!\det(\Sigmab_\mu)}\bigg]\\
  &=\sum_{k=0}^\infty
    \frac{\gamma^k\ee^{-\gamma}}{k!}\,\frac{\gamma^dk!}{(k+d)!}\,
    \frac{\E\big[\sum_{S:\,|S|=d}\det(\Z_{S,*})^2\one_{[\Z\in E]}\mid
    K=k\big]}{\det(\gamma\Sigmab_\mu)}\\
  &\overset{(*)}{=} \sum_{k=0}^\infty
    \frac{\gamma^{k+d}\ee^{-\gamma}}{(k+d)!}
    \,\frac{\E[\det(\Z^\top\Z)\one_{[\Z\in E]}\mid K=k]}{\det(\gamma\Sigmab_\mu)},
\end{align*}
where $(*)$ uses the Cauchy-Binet formula to sum over all subsets $S$
of size $d$. Finally, since the sum shifts from $k$
to $k+d$, the last expression can be rewritten as
$\E[\det(\X^\top\X)\one_{[\X\in E]}]/\det(\gamma\Sigmab_\mu)$, where recall that
$\X\sim\mu^K$ and $K\sim\Poisson(\gamma)$, matching the definition of $S_\mu^{d+\gamma}$.
\end{proof}

We now proceed with the proof of Lemma \ref{l:size}, where we establish
that the expected sample size of $S_\mu^n$ is indeed $n$.

\begin{proofof}{Lemma}{\ref{l:size}}
  The result is obvious when $n=d$, whereas
  for $n>d$ it is an immediate consequence
  of Lemma \ref{l:decomposition}.
  Finally, for $n<d$ the expected sample
  size follows as a corollary of Lemma \ref{l:proj}, which states that
  \begin{align*}
\text{(Lemma \ref{l:proj})} \qquad\E\big[\I - \Xb^\dagger\Xb\big] =
    (\gamma_n\Sigmab_\mu + \I)^{-1},
  \end{align*}
  where $\Xb^\dagger\Xb$ is the orthogonal projection onto
  the subspace spanned by the rows of $\Xb$. Since the rank of this
  subspace is equal to the number of the rows, we have
  $\#(\Xb)=\tr(\Xb^\dagger\Xb)$, so
  \begin{align*}
    \E\big[\#(\Xb)\big] = d - \tr\big((\gamma_n\Sigmab_\mu +
    \I)^{-1}\big) =
    \tr\big(\gamma_n\Sigmab_\mu(\gamma_n\Sigmab_\mu+\I)^{-1}\big) = n,
  \end{align*}
  which completes the proof.
\end{proofof}

\section{Proofs for Section \ref{s:dp}}
\label{a:dp}

\begin{proofof}{Lemma}{\ref{t:ring}} \
 First, we show that $\A+\u\v^\top$ is d.p.~for any fixed
 $\u,\v\in\R^d$. Below, we use a standard identity for the rank one
 update of a determinant:
 $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$. It follows that
 for any $\Ic$ and $\Jc$ of the same size,
  \begin{align*}
\E\big[\!\det(\A_{\Ic,\Jc}\!+\u_{\Ic}\v_{\Jc}^\top)\big] &=
    \E\big[\!\det(\A_{\Ic,\Jc}) +
    \v_{\Jc}^\top\adj(\A_{\Ic,\Jc}) \u_{\Ic}\big]\\
    &\overset{(*)}{=}\det\!\big(\E[\A_{\Ic,\Jc}]\big) +
      \v_{\Jc}^\top\adj\!\big(\E[\A_{\Ic,\Jc}]\big) \u_{\Ic}\\
    &=\det\!\big(\E[\A_{\Ic,\Jc} \!+ \u_{\Ic}\v_{\Jc}^\top]\big),
  \end{align*}
  where $(*)$ used \eqref{eq:adj}, i.e., the fact that for d.p.~matrices, adjugate commutes
  with expectation. Crucially, through the definition of an adjugate
  this step implicitly relies on the assumption that all the square
  submatrices of $\A_{\Ic,\Jc}$ are also  determinant preserving.
  Iterating this, we get that $\A+\Z$ is d.p.~for any fixed
  $\Z$. We now show the same for $\A+\B$:
  \begin{align*}
\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\big]
    &=
      \E\Big[\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\mid\B\big]\Big]\\
    &\overset{(*)}{=}\E\Big[\!\det\!\big(\E[\A_{\Ic,\Jc}]\!+\B_{\Ic,\Jc}\big)\Big]\\
      &= \det\!\big(\E[\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc}]\big),
  \end{align*}
  where $(*)$  uses the fact that after conditioning on $\B$ we can
  treat it as a fixed matrix. Next, we show that $\A\B$ is determinant preserving via the Cauchy-Binet formula:
  \begin{align*}
    \E\big[\!\det\!\big((\A\B)_{\Ic,\Jc}\big)\big]
    &= \E\big[\!\det(\A_{\Ic,*}\B_{*,\Jc})\big]\\
    &=\E\bigg[\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\A_{\Ic,S}\big)
      \det\!\big(\B_{S,\Jc}\big)\bigg]\\
&=\!\!\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\E[\A]_{\Ic,S}\big)
                                                \det\!\big(\E[\B]_{S,\Jc}\big)\\
    &=\det\!\big(\E[\A]_{\Ic,*}\, \E[\B]_{*,\Jc}\big)\\
      &= \det\!\big(\E[\A\B]_{\Ic,\Jc}\big),
  \end{align*}
  where recall that $\A_{\Ic,*}$ denotes the submatrix of $\A$
  consisting of its (entire) rows indexed by $\Ic$.
  \end{proofof}

To prove Lemma \ref{l:poisson}, we will use the following
lemma, many variants of which appeared in the literature
\cite[e.g.,][]{expected-generalized-variance}. We use the one given by
\cite{correcting-bias}.
\begin{lemma}[\citeauthor{correcting-bias}, \citeyear{correcting-bias}]\label{l:cb}
If the rows of random $k\times d$ matrices $\A,\B$
  are sampled as an i.i.d.~sequence of $k\geq d$ pairs of joint random vectors, then
\begin{align}
  k^d\,\E \big[\det(\A^\top\B)\big]
  &= \ktd\,\det\!\big(\E[\A^\top\B]\big).
     \end{align}
 \end{lemma}

\noindent
Here, we use the following standard shorthand: $\ktd =
\frac{k!}{(k-d)!} = k\,(k-1)\dotsm(k-d+1)$. Note that the above result
almost looks like we are claiming that the matrix $\A^\top\B$ is d.p.,
but in fact it is not because $k^d\neq \ktd$. The difference
in those factors is precisely what we are going to correct with the
Poisson random variable. We now present the proof of Lemma
\ref{l:poisson}.
\begin{proofof}{Lemma}{\ref{l:poisson}}
Without loss of generality, it suffices to check Definition \ref{d:main} with both $\Ic$ and
$\Jc$ equal $[d]$. We first expand the expectation by
conditioning on the value of $K$ and letting $\gamma=\E[K]$:
    \begin{align*}
      \E\big[\!\det(\A^\top\B)\big]
      &= \sum_{k=0}^\infty
\E\big[\det(\A^\top\B)\mid K\!=\!k\big]\
\Pr(K\!=\!k)\\
      \text{(Lemma \ref{l:cb})}
      \quad&=
        \sum_{k=d}^\infty\frac{k! k^{-d}}{(k-d)!}\det\!\big(\E[\A^\top\B\mid
        K\!=\!k]\big)
        \frac{\gamma^k\ee^{-\gamma}}{k!}\\
      &=\sum_{k=d}^\infty
\Big(\frac\gamma k\Big)^d\det\!\big(\E[\A^\top\B\mid K\!=\!k]\big)
        \frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}.
     %  \\
     %  &=\sum_{k=0}^\infty \det\!\big(\E[\A^\top\B]\big)\,\Pr(K\!=\!k)
     % \ =\ \det\!\big(\E[\A^\top\B]\big).
    \end{align*}
    Note that $\frac\gamma k\,\E[\A^\top\B\mid K\!=\!k]=\E[\A^\top\B]$,
    which is independent of $k$. Thus we can rewrite the above
    expression as:
    \begin{align*}
\det\!\big(\E[\A^\top\B]\big)\sum_{k=d}^\infty\frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}
      =
      \det\!\big(\E[\A^\top\B]\big)\sum_{k=0}^\infty
      \frac{\gamma^{k}\ee^{-\gamma}}{k!}=\det\!\big(\E[\A^\top\B]\big),
    \end{align*}
    which concludes the proof.
  \end{proofof}

To prove Lemma \ref{l:normalization}, we use the following standard
determinantal formula which is used to derive the normalization
constant of a discrete determinantal point process.
\begin{lemma}[\citeauthor{dpp-ml}, \citeyear{dpp-ml}]\label{l:det-standard}
  For any $k\times d$ matrices $\A,\B$ we have
  \[\det(\I+\A\B^\top)=\sum_{S\subseteq[k]}\det(\A_{S,*}\B_{S,*}^\top).\]
\end{lemma}

\begin{proofof}{Lemma}{\ref{l:normalization}}
By Lemma \ref{l:poisson}, the matrix $\B^\top\A$ is determinant
preserving. Applying Lemma \ref{t:ring} we conclude that
$\I+\B^\top\A$ is also d.p., so
\begin{align*}
  \det\!\big(\I+\E[\B^\top\A]\big) = \E\big[\det(\I+\B^\top\A)\big] =\E\big[\det(\I+\A\B^\top)\big],
\end{align*}
where the second equality is known as Sylvester's Theorem.
We rewrite the expectation of $\det(\I+\A\B^\top)$ by applying Lemma
\ref{l:det-standard}.  Letting $\gamma=\E[K]$, we obtain:
\begin{align*}
\E\big[\det(\I+\A\B^\top)\big]  &=\E\bigg[\sum_{S\subseteq [K]}\E\big[\det(\A_{S,*}\B_{S,*}^\top)\mid
    K\big]\bigg]\\
  &\overset{(*)}{=}\sum_{k=0}^\infty\frac{\gamma^k\ee^{-\gamma}}{k!}
  \sum_{i=0}^k\binom{k}{i} \E\big[\det(\A\B^\top)\mid K=i\big]\\
  &=\sum_{i=0}^\infty \E\big[\det(\A\B^\top)\mid K=i\big]
  \sum_{k\geq i}^\infty \binom{k}{i}
  \frac{\gamma^k\ee^{-\gamma}}{k!}\\
  &=\sum_{i=0}^\infty
    \frac{\gamma^i\ee^{-\gamma}}{i!}\E\big[\det(\A\B^\top)\mid K=i\big]
    \sum_{k\geq i}^\infty\frac{\gamma^{k-i}}{(k-i)!} = \E\big[\det(\A\B^\top)\big]\cdot\ee^\gamma,
\end{align*}
where $(*)$ follows from the exchangeability of the rows of $\A$ and
$\B$, which implies that the distribution of $\A_{S,*}\B_{S,*}^\top$ is the
same for all subsets $S$ of a fixed size $k$.
\end{proofof}

\section{Proof of Theorem \ref{t:mse}}
\label{a:mse-proof}
In this section we use $Z_\mu^n$ to denote the normalization
constant that appears in \eqref{eq:cases} when computing an expectation for surrogate design
$S_\mu^n$.
We first prove Lemma \ref{l:sqinv-all} by splitting the under- and
over-determined cases and start with proving the former.
% Note that those
% results require $\mu$ to satisfy general position (Assumption \ref{a:general-position}),
% which implies that if $\X\sim\mu^k$ for $k\leq d$ then $\rank(\X)=k$.
\begin{lemma}\label{l:sqinv-under}
If  $\Xb\sim S_\mu^n$ for $n<d$, then we have
\begin{align*}
    \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
    &={\gamma_n}\big(1- \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big).
\end{align*}
\end{lemma}
\begin{proof}
Let $\X\sim\mu^K$ for $K\sim\Poisson({\gamma_n})$. Note that if
$\det(\X\X^\top)>0$ then using the fact that
$\det(\A)\A^{-1}=\adj(\A)$ for any invertible matrix $\A$, we can write:
  \begin{align*}
    \det(\X\X^\top)\tr\big((\X^\top\X)^{\dagger}\big)
    &= \det(\X\X^\top)\tr\big((\X\X^\top)^{-1}\big) \\
    &= \tr(\adj(\X\X^\top)) \\[-1mm]
    &= \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top),
  \end{align*}
  where $\X_{-i}$ is a shorthand for $\X_{[K]\backslash\{i\},*}$.
Assumption \ref{a:general-position} ensures that
$\Pr\big\{\det(\X\X^\top)>0\big\}=1$, which allows us to write:
  \begin{align*}
Z_\mu^n\cdot \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
    &=\E\bigg[
    \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top)\ \big|\
    \det(\X\X^\top)>0\bigg]\cdot\overbrace{\Pr\big\{\det(\X\X^\top)>0\big\}}^{1}\\
    &=\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\E\Big[
      \sum_{i=1}^k\det(\X_{-i}\X_{-i}^\top)\ \big|\  K=k\Big]\\
    &=\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\, k\
      \E\big[\det(\X\X^\top)\mid K=k-1\big]\\
    &=\gamma_n\sum_{k=0}^{d-1}\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}
      \E\big[\det(\X\X^\top)\mid K=k\big]\\
    &=\gamma_n\Big(\E\big[\det(\X\X^\top)\big]\ -\
      \frac{\gamma_n^{d}\ee^{-\gamma_n}}{d!}\E\big[\det(\X)^2\mid K=d\big]
      \Big) \\
    &\overset{(*)}{=}\gamma_n\big(\ee^{-\gamma_n}\det(\I +\gamma_n\Sigmab_\mu) -
      \ee^{-\gamma_n}\det(\gamma_n\Sigmab_\mu)\big),
  \end{align*}
  where $(*)$ uses Lemma \ref{l:normalization} for the first term and
  Lemma \ref{l:cb} for the second term. We obtain the desired result by
  dividing both sides by
  $Z_\mu^n=\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)$.
\end{proof}
In the over-determined regime, a more general matrix expectation
formula can be shown (omitting the trace). The following result is
related to an expectation formula derived by
\cite{correcting-bias-journal}, however they use a slightly
different determinantal design so the results are incomparable.
\begin{lemma}\label{l:sqinv-over}
If $\Xb\sim S_\mu^n$ and $n>d$, then we
have
\begin{align*}
  \E\big[ (\Xb^\top\Xb)^{\dagger}\big] =
  \Sigmab_\mu^{-1}\cdot \frac{1-\ee^{-\gamma_n}}{\gamma_n}.
\end{align*}
\end{lemma}
\begin{proof}
Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$. Assumption
\ref{a:general-position} implies that for $K\neq d-1$ we have
\begin{align}
  \det(\X^\top\X)(\X^\top\X)^\dagger=\adj(\X^\top\X),\label{eq:adj-over}
  \end{align}
however when $k=d-1$ then \eqref{eq:adj-over} does not hold because
$\det(\X^\top\X)=0$ while $\adj(\X^\top\X)$ may be non-zero. It
follows that:
  \begin{align*}
Z_\mu^n\cdot
    \E\big[ (\Xb^\top\Xb)^{\dagger}\big]
    &=\E\big[\det(\X^\top\X)(\X^\top\X)^\dagger\big]\\
    &=\E\big[\adj(\X^\top\X)\big]-
\frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)!}
      \E\big[\adj(\X^\top\X)\mid K=d-1\big]\\
    &\overset{(*)}{=}\adj\!\big(\E[\X^\top\X]\big) -
      \frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)^{d-1}}
      \adj\!\big(\E[\X^\top\X\mid K=d-1]\big)\\
    &=\adj(\gamma_n\Sigmab_\mu) - \ee^{-\gamma_n}\adj(\gamma_n\Sigmab_\mu)\\
    &=\det(\gamma_n\Sigmab_\mu)\,(\gamma_n\Sigmab_\mu)^{-1}(1-\ee^{-\gamma_n})\\
    &=\det(\gamma_n\Sigmab_\mu)\,\Sigmab_\mu^{-1}\cdot\frac{1-\ee^{-\gamma_n}}{\gamma_n},
  \end{align*}
  where the first term in $(*)$ follows from Lemma
  \ref{l:normalization} and \eqref{eq:adj}, whereas the second term comes
  from Lemma 2.3 of \cite{correcting-bias-journal}.
Dividing both sides by $Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ completes the proof.
\end{proof}

Applying the closed form expressions from Lemmas
\ref{l:proj} and \ref{l:sqinv-all}, we derive
the formula for the MSE and prove Theorem \ref{t:mse} (we defer the
proof of Lemma \ref{l:proj} to Appendix \ref{s:unbiased-proof}).
\begin{proofof}{Theorem}{\ref{t:mse}}
  First, assume that $n<d$, in which case we have
  $\gamma_n=\frac1{\lambda_n}$ and moreover
  \begin{align*}
    n &= \tr\big(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1}\big)\\
      &=\tr\big((\Sigmab_\mu+\lambda_n\I-\lambda_n\I)(\Sigmab_\mu+\lambda_n\I)^{-1}\big)\\
    &=d - \lambda_n\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big),
    %\frac{\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)}{d-n},
  \end{align*}
so we can write $\lambda_n$ as $(d-n)/\tr((\Sigmab_\mu+\lambda_n\I)^{-1})$.
  From this and Lemmas \ref{l:proj} and \ref{l:sqinv-under}, we
obtain the desired expression, where recall
  that $\alpha_n = \det\!\big(\Sigmab_\mu (\Sigmab_\mu+\frac1{\gamma_n})^{-1}\big)$:
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} &= \sigma^2\,\gamma_n(1-\alpha_n) +
    \tfrac1{\gamma_n} \,\w^{*\top}(\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\w^*
    \\
    &\overset{(a)}{=}\sigma^2\,\frac{1-\alpha_n}{\lambda_n} +
    \lambda_n\,\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*\\
    &\overset{(b)}{=}\sigma^2\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\frac{1-\alpha_n}{d-n}
      +
      (d-n)\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
      {\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)}.
  \end{align*}
  While the expression given after $(a)$ is simpler than the one
after $(b)$, the latter better illustrates how the MSE depends on
the sample size $n$ and the dimension $d$.
  Now, assume that $n>d$. In this case, we have $\gamma_n=n-d$ and apply Lemma
  \ref{l:sqinv-over}:
  \begin{align*}
    \MSE{\Xb^\dagger\ybb}
    &= \sigma^2\,\tr(\Sigmab_\mu^{-1})\,
\frac{1-\ee^{-\gamma_n}}{\gamma_n}
=\sigma^2\,\tr(\Sigmab_\mu^{-1})
\,\frac{1-\beta_n}{n-d}.
  \end{align*}
The case of $n=d$ was shown in Theorem~2.12 of \cite{correcting-bias-journal}.
This concludes the proof.
\end{proofof}

\section{Proof of Theorem \ref{t:unbiased}}
\label{s:unbiased-proof}

As in the previous section, we use $Z_\mu^n$ to denote the normalization
constant that appears in \eqref{eq:cases} when computing an expectation
for surrogate design $S_\mu^n$.
Recall that our goal is to compute the expected value of
$\Xb^\dagger\ybb$ under the surrogate design $S_\mu^n$. Similarly as for Theorem
\ref{t:mse}, the case of $n=d$ was shown in Theorem 2.10 of
\cite{correcting-bias-journal}. We break the rest down into the
under-determined case $(n<d)$ and the over-determined case ($n>d$),
starting with the former. Recall that we do \emph{not} require any
modeling assumptions on the responses.
\begin{lemma}\label{l:ridge-under}
If $\Xb\sim S_\mu^n$ and $n<d$, then for any $y(\cdot)$
such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
denoting $\yb_i$ as $y(\xbb_i)$, we have
\begin{align*}
  \E\big[\Xb^\dagger \ybb\big]
  &=
    \big(\Sigmab_\mu+\tfrac1{\gamma_n}\I\big)^{-1}\E_{\mu,y}[y(\x)\,\x].
\end{align*}
\end{lemma}
\begin{proof}
   Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
   $y(\x_i)$ as $y_i$.
  Note that when $\det(\X\X^\top)>0$, then
  the $j$th entry of $\X^\dagger\y$ equals
  $\f_j^\top(\X\X^\top)^{-1}\y$, where $\f_j$ is the $j$th
  column of $\X$, so:
\begin{align*}
  \det(\X\X^\top)\,(\X^\dagger\y)_j
  &= \det(\X\X^\top)\, \f_j^\top(\X\X^\top)^{-1}\y \\
  &=
  \det(\X\X^\top+\y\f_j^\top) - \det(\X\X^\top).
\end{align*}
If $\det(\X\X^\top)=0$, then also
$\det(\X\X^\top+\y\f_j^\top)=0$, so we can write:
\begin{align*}
Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
  &=  \E\big[\det(\X\X^\top)(\X^\dagger\y)_j\big] \\
  &= \E\big[\det(\X\X^\top+\y\f_j^\top)-\det(\X\X^\top)\big]  \\
  &=\E\big[\det\!\big([\X,\y][\X,\f_j]^\top\big)\big] - \E\big[\det(\X\X^\top)\big]\\
  &\overset{(a)}{=}\ee^{-\gamma_n}\det\!\bigg(\I +
    \gamma_n\,\E_{\mu,y}\bigg[\begin{pmatrix}\x\x^\top&
      \!\!\x\, y(\x)\\ x_j\,\x^\top&\!\!
      x_j\,y(\x)\end{pmatrix}\bigg]\bigg)
                          -\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)\\
  &\overset{(b)}{=}\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu) \\
    &\qquad \times \Big(\E_{\mu,y}\big[\gamma_n
    x_j\,y(\x)\big] - \E_{\mu}\big[\gamma_n
    x_j\,\x^\top\big](\I+\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[\gamma_n\x\,
    y(\x)\big]\Big),
\end{align*}
where $(a)$ uses Lemma \ref{l:normalization} twice, with the first
application involving two different matrices $\A=[\X,\y]$ and
$\B=[\X,\f_j]$, whereas $(b)$ is a standard determinantal identity
\cite[see Fact 2.14.2 in][]{matrix-mathematics}.
  Dividing both sides by $Z_\mu^n$ and letting $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$, we obtain that:
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big]
    &= \gamma_n\v_{\mu,y} - \gamma_n^2\Sigmab_\mu(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y}\\
&=\gamma_n\big(\I - \gamma_n\Sigmab_\mu (\I+\gamma_n\Sigmab_\mu)^{-1}\big)\v_{\mu,y}
=\gamma_n(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y},
  \end{align*}
  which completes the proof.
\end{proof}
We return to Lemma \ref{l:proj}, regarding the expected orthogonal
projection onto the complement of the row-span of $\Xb$, i.e.,
$\E[\I-\Xb^\dagger\Xb]$, which follows as a corollary of Lemma~\ref{l:ridge-under}.
\begin{proofof}{Lemma}{\ref{l:proj}}
  We let $y(\x)=x_j$ where $j\in[d]$ and apply Lemma
  \ref{l:ridge-under} for each $j$, obtaining:
  \begin{align*}
    \I - \E\big[\Xb^\dagger\Xb] = \I -
    (\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\Sigmab_\mu,
  \end{align*}
  from which the result follows by simple algebraic manipulation.
\end{proofof}

We move on to the over-determined case, where the ridge regularization
of adding the identity to $\Sigmab_\mu$ vanishes. Recall that we
assume throughout the paper that $\Sigmab_\mu$ is invertible.
\begin{lemma}\label{l:ridge-over}
  If $\Xb\sim S_\mu^n$ and $n>d$, then for any real-valued random function $y(\cdot)$
  such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
denoting $\yb_i$ as $y(\xbb_i)$, we have
 \begin{align*}
  \E\big[\Xb^\dagger \ybb\big]
  &=\Sigmab_\mu^{-1}\E_{\mu,y}\big[y(\x)\,\x\big].
\end{align*}
\end{lemma}
\begin{proof}
   Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
   $y_i=y(\x_i)$. Similarly as in the proof of
   Lemma~\ref{l:ridge-under}, we note that when $\det(\X^\top\X)>0$,
   then
  the $j$th entry of $\X^\dagger\y$ equals
  $\e_j^\top(\X^\top\X)^{-1}\X^\top\y$, where $\e_j$ is the $j$th
standard basis vector, so:
\begin{align*}
  \det(\X^\top\X)\,(\X^\dagger\y)_j =
  \det(\X^\top\X)\, \e_j^\top(\X^\top\X)^{-1}\X^\top\y =
  \det(\X^\top\X+\X^\top\y\e_j^\top) - \det(\X^\top\X).
\end{align*}
If $\det(\X^\top\X)=0$, then also
$\det(\X^\top\X+\X^\top\y\e_j^\top)=0$. We proceed to compute the
expectation:
\begin{align*}
Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
  &=  \E\big[\det(\X^\top\X)(\X^\dagger\y)_j\big] \\
  &= \E\big[\det(\X^\top\X+\X^\top\y\e_j^\top)-\det(\X^\top\X)\big]  \\
  &=\E\big[\det\!\big(\X^\top(\X+\y\e_j^\top)\big)\big] - \E\big[\det(\X^\top\X)\big]\\
  &\overset{(*)}{=}\det\!\Big(
    \gamma_n\,\E_{\mu,y}\big[\x(\x+ y(\x)\e_j)^\top\big]\Big)
    -\det(\gamma_n\Sigmab_\mu)\\
  &=\det\!\big(\gamma_n\Sigmab_\mu + \gamma_n\E_{\mu,y}[\x\,y(\x)]\e_j^\top\big)
    -\det(\gamma_n\Sigmab_\mu)\\
  &=\det(\gamma_n\Sigmab_\mu)\cdot
    \gamma_n\e_j^\top(\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[y(\x)\,\x\big],
\end{align*}
where $(*)$ uses Lemma \ref{l:poisson} twice (the first time, with
$\A=\X$ and $\B=\X+\y\e_j^\top$). Dividing both sides by
$Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ concludes the proof.
\end{proof}

\noindent
We combine Lemmas \ref{l:ridge-under} and \ref{l:ridge-over} to obtain
the proof of Theorem \ref{t:unbiased}.
\begin{proofof}{Theorem}{\ref{t:unbiased}}
The case of $n=d$ follows directly from Theorem~2.10 of
\cite{correcting-bias}.
Assume that $n<d$. Then we have
$\gamma_n=\frac1{\lambda_n}$, so the result follows
from Lemma \ref{l:ridge-under}.
% , noting that unlike in the lemma,
% Theorem \ref{t:unbiased} allows for the response model $y(\cdot)$ to be randomized:
% \begin{align*}
%   \E[\Xb^\dagger\ybb] = \E\big[\E[\Xb^\dagger\ybb\mid\ybb]\big] =
% % \w_{\lambda_n}^* = \argmin_\w\E_\mu\big[(\x^\top\w-y(\x))^2\big] +
% %   \lambda_n\|\w\|^2 =
% %   \big(\Sigmab_\mu+\lambda_n\I\big)^{-1}
% %   \E_{\mu}\big[y(\x)\,\x\big].
% \end{align*}
If $n>d$, then the result follows from Lemma \ref{l:ridge-over}.
\end{proofof}

\section{Proof of Theorem \ref{t:asymptotic}}
\label{sec:proof-of-t-asymptotic}
Once again, we break down the proof into under- and
over-determined cases, starting with the former. Note that in this
case we require that the covariance be equal to identity.
\begin{lemma}\label{l:asymptotic-under}
  Let $\rho=n/d$, $\X\sim \Nc_{n,d}(\zero,\I_n \otimes \I_d)$ and $y_i=y(\x_i)$ under Assumption
  \ref{a:linear}. If $d>n+1$ then
  \begin{align*}
    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\I,
\w^*,\sigma^2,n)}{\Mc(\I,\w^*,\sigma^2,n)}\ \leq\
\frac1d\cdot \frac{1}{1-\rho-\frac1d} + 3\rho^d.
  \end{align*}
\end{lemma}

\begin{proof}
  We first recall the standard decomposition of $\MSE{\X^\dagger \y}$:
  \begin{align*}
    \MSE{\X^\dagger \y}
    &=\sigma^2 \E\big[\tr\big((\X^\top \X)^\dagger \big)\big]+
      \w^{*\top}\!\big( \I - \E[\X^\dagger \X]\big) \w^*.
  \end{align*}
  Since the rows of $\X$ are standard normal random variables,
  $\X\X^\top$ is an $n\times n$ Wishart random matrix with $d>n+1$ degrees
  of freedom. Using the formula for the mean of the Inverse-Wishart
  distribution, it follows that
  \begin{align*}
\E\big[\tr((\X^\top\X)^\dagger)\big] = \E\big[\tr((\X\X^\top)^{-1})\big]= \frac{n}{d - n - 1}.
  \end{align*}
  % and therefore
  % \begin{align*}
  %   \MSE{\X^\dagger \y}
  %   &= \|\w\|_2^2 - \w^\top \E[\X^\dagger \X] \w
  %   + \frac{\sigma^2 n}{d - n - 1}
  % \end{align*}
  % Let $\X = \U \S \V^\top$ be its full singular value decomposition. Notice
  % \begin{align*}
  %   \X^\dagger \X
  %   &= \V \S^\dagger \U^\top \U \S \V^\top \\
  %   &= \V \S^\dagger \S \V^\top \\
  %   &= \V \begin{bmatrix}
  %     \I_n & 0 \\
  %     0 & 0
  %   \end{bmatrix} \V^\top \\
  %   &= \V_n \V_n^\top \\
  %   \w^\top \E[\X^\dagger \X] \w
  %   &= \E[\|\V_n^\top \w\|_2^2]
  % \end{align*}
  % where $\V_n$ denotes the $d \times n$ matrix with columns equal to the
  % $n$ eigenvectors of $\X^\top \X$ with non-zero eigenvalue.
  % But since $\Sigmab = \I$, by rotation invariance
  % $\V_n^\top$ is a change of basis onto a uniformly random
Note that $\X^\dagger\X$ is a uniformly random projection matrix so by
rotational symmetry it follows that
\begin{align*}
  \w^{*\top}\E\big[\X^\dagger\X]\w^* =
  \E\big[\|\X^\dagger\X\w^*\|^2\big] = \frac nd \,\|\w^*\|^2.
\end{align*}
Putting this together we obtain that
  \begin{align*}
    \MSE{\X^\dagger \y}
    &=\frac{\sigma^2 n}{d - n - 1}+ \|\w^*\|^2\,\frac{d - n}{d}.
  \end{align*}
  On the other hand, the surrogate MSE expression can be derived by
observing that for $\Sigmab=\I$ we have
$\tr((\Sigmab+\lambda_n\I)^{-1})=d/(1+\lambda_n)=n$ (see definition of
$\lambda_n$ in Theorem \ref{t:mse}):
  \begin{align*}
    \Mc(\I, \w^*,\sigma^2,n) =
    \sigma^2n\cdot\frac{1-\alpha_n}{d-n} + \|\w^*\|^2\,\frac{d-n}{d}.
  \end{align*}
  Note that the second term is the same in both cases, even though
  this may not be true for non-isotropic Gaussians. We now compute the
  normalized difference between the
  expressions,
  \begin{align*}
 \frac{\MSE{\X^\dagger\y}-\Mc(\I, \w^*,\sigma^2,n)}{\Mc(\I,
    \w^*,\sigma^2,n)}
    &= \frac{\sigma^2n\cdot(\frac1{d-n-1} -
    \frac{1-\alpha_n}{d-n})}{\sigma^2n\cdot\frac{1-\alpha_n}{d-n} +
    \|\w^*\|^2\,\frac{d-n}{d}}\\
&\leq \frac {d-n}{1-\alpha_n}
\Big(\frac1{d-n-1}-\frac{1-\alpha_n}{d-n}\Big)\\
    &=\frac{d-n}{d-n-1}\,\frac{1}{1-\alpha_n}- 1\\
    &=\frac1{d-n-1} + \frac{\alpha_n}{1-\alpha_n}\Big(1+\frac1{d-n-1}\Big).
  \end{align*}
Let $\rho=n/d$. Then $\frac1{d-n-1} = \frac1d\cdot
\frac1{1-\rho-\frac1d}$ and moreover
$\alpha_n = (\frac1{1+\lambda_n})^d=(\frac
nd)^d=\rho^d$. From the assumption that $d>n+1$, we
conclude that $\alpha_n \leq (\frac{d-2}{d})^d\leq \ee^{-2}$
so that $\frac{\alpha_n}{1-\alpha_n}(1+\frac1{d-n-1})\leq
  \frac{2 \alpha_n}{1-\ee^{-2}}\leq 3\rho^d$. This shows the right-hand-side inequality of the
  theorem. That fact that $\MSE{\X^\dagger\y}\geq\Mc(\I,
  \w^*,\sigma^2,n)$ follows easily.
\end{proof}
\begin{lemma}\label{l:asymptotic-over}
  Let $\rho=n/d$, $\X\sim \Nc_{n,d}(\zero,\I_n \otimes \Sigmab)$ and $y_i=y(\x_i)$ under Assumption
  \ref{a:linear}. If $n>d+1$ then
  \begin{align*}
    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab,
\w^*,\sigma^2,n)}{\Mc(\Sigmab,\w^*,\sigma^2,n)}\ \leq\
\frac1d\cdot \frac{1}{\rho-1-\frac1d} + 3(\ee^{1-\rho})^d.
  \end{align*}
\end{lemma}
\begin{proof}
The MSE for the over-determined Gaussian design can be derived by
using the formula for the mean of the Inverse-Wishart distribution:
\begin{align*}
  \MSE{\X^\dagger\y}=\sigma^2\tr\big(\E[(\X^\top\X)^{-1}]\big) =
  \frac{\sigma^2\tr(\Sigmab^{-1}) }{n-d-1}.
\end{align*}
To compute the normalized difference we follow similar derivations as
in the proof of Lemma \ref{l:asymptotic-under}:
\begin{align*}
 \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)}{\Mc(\Sigmab,
    \w^*,\sigma^2,n)}
    &=  \frac {n-d}{1-\beta_n}
      \Big(\frac1{n-d-1}-\frac{1-\beta_n}{n-d}\Big)\\
  &\leq\frac1d\cdot\frac1{\rho-1-\frac1d} + \frac{2\beta_n}{1-\beta_n}.
\end{align*}
Recall that $\beta_n = \ee^{d-n} = (\ee^{1-\rho})^d$ and for $n-d\geq
2$ we have $\frac2{1-\beta_n}\leq 3$. The desired inequalities follow immediately.
\end{proof}
Theorem \ref{t:asymptotic} now follows as a consequence of
Lemmas \ref{l:asymptotic-under} and \ref{l:asymptotic-over}.
\begin{proofof}{Theorem}{\ref{t:asymptotic}}
  Since $\frac1d\leq \frac13|1-\rho|$ and $\ee^{-|1-\rho|d}\leq
  \frac1{2d|1-\rho|}$, it follows that
  \begin{align*}
    \frac1d\cdot \frac{1}{|1-\rho|-\frac1d} + 3(\ee^{-|1-\rho|})^d
    &\leq \frac1d\cdot \frac{1}{|1-\rho|-\frac13|1-\rho|} +
      \frac1d\cdot\frac{3}{2|1-\rho|}=\frac {c_\rho}{d}.
  \end{align*}
  The case of $n>d+1$ now follows from Lemma
  \ref{l:asymptotic-over}. Also, since $\rho\leq \ee^{\rho-1}$, for
$n<d-1$ we have $3\rho^d\leq 3(\ee^{-|1-\rho|})^{d}$, so the same bound follows
from Lemma \ref{l:asymptotic-under}.
\end{proofof}

\section{Empirical analysis of the asymptotic consistency conjectures}
\label{sec:asymp-conj-details}

Conjectures \ref{c:wishart} and \ref{c:projection} are related to open problems
which have been extensively studied in the literature.
With respect to Conjecture~\ref{c:wishart}, \cite{srivastava2003} first derived the probability
density function of the pseudo-Wishart
distribution (also called the singular Wishart), and \cite{cook2011}
computed the first and second moments of 
generalized inverses for the distribution. However, for the
Moore-Penrose inverse and arbitrary covariance $\Sigmab$,
\cite{cook2011} claims that the quantities required
to express the mean ``do not have tractable closed-form representation.''
%
Conjecture~\ref{c:projection} has connections to directional statistics.
Using the SVD, we have the equivalent representation $\X^\dagger \X = \V \V^\top$
where $\V$ is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal
$n$-frames in $\R^d$).
The distribution of $\V$ is known as the matrix angular central
Gaussian (MACG) distribution \citep{chikuse1990matrix}. While prior work
has considered high dimensional limit theorems \citep{CHIKUSE1991145}
as well as density estimation and hypothesis testing \citep{CHIKUSE1998188}
on $V_{n,d}$, they only analyzed the invariant measure
(which corresponds in our setting to $\Sigmab = \I$),
and to our knowledge a closed form expression of $\E[\V\V^\top]$ where
$\V$ is distributed according to MACG with
arbitrary $\Sigmab$ remains an open question.

For verifying these two conjectures, it suffices to only consider
diagonal covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D
\Q^\top$ is its eigendecomposition and $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Q\D\Q^\top)$, then
we have for $\W \sim \Pc\Wc(\Sigmab, n)$ that $\W
\overset{d}{=} \X^\top \X$ and hence, defining
$\Xt\sim\Nc_{n,d}(\zero,\I_n \otimes \D)$, by linearity and unitary invariance of trace,
\begin{align*}
  \E[\tr(\W^\dagger)]
  &= \tr\big( \E[(\X^\top\X)^\dagger] \big)
  = \tr\Big( \Q\E\big[(\Xt^\top\Xt)^\dagger\big]\Q^\top \Big)
  = \tr\Big( \E\big[(\Xt^\top\Xt)^\dagger\big] \Big)
  = \E\left[\tr \big((\Xt^\top\Xt)^\dagger\big) \right].
\end{align*}
Similarly, we have that
$\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$, and a simple
calculation shows that the expression in Conjecture~\ref{c:projection}
is also independent of the choice of matrix $\Q$.

\input{continuous-empirical.tex}

### 2022-08-23 design/continuous_colt/colt2020-sample.tex
Commit original repos

\documentclass[anon]{colt2020} % Anonymized submission
% \documentclass{colt2020} % Include author names

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

\title[Short Title]{Full Title of Article}
\usepackage{times}
% Use \Name{Author Name} to specify the name.
% If the surname contains spaces, enclose the surname
% in braces, e.g. \Name{John {Smith Jones}} similarly
% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
% If the first letter in the forenames is a diacritic
% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

% Two authors with the same address
% \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
%  \Name{Author Name2} \Email{xyz@sample.com}\\
%  \addr Address}

% Three or more authors with the same address:
% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
%  \Name{Author Name2} \Email{an2@sample.com}\\
%  \Name{Author Name3} \Email{an3@sample.com}\\
%  \addr Address}

% Authors with different addresses:
\coltauthor{%
 \Name{Author Name1} \Email{abc@sample.com}\\
 \addr Address 1
 \AND
 \Name{Author Name2} \Email{xyz@sample.com}\\
 \addr Address 2%
}

\begin{document}

\maketitle

\begin{abstract}%
  This is a great paper and it has a concise abstract.%
\end{abstract}

\begin{keywords}%
  List of keywords%
\end{keywords}

\section{Introduction}

This is where the content of your paper goes.
\begin{itemize}
  \item Limit the main text (not counting references and appendices) to 12 PMLR-formatted pages, using this template.
  \item Include, either in the main text or the appendices, all details, proofs and derivations required to substantiate the results.
  \item Include {\em in the main text} enough details, including proof details, to convince the reviewers of the contribution, novelty and significance of the submissions.
  \item Use the \textbackslash documentclass[anon]{colt2020} option during submission process -- 
this automatically hides the author names listed under \textbackslash coltauthor. 
Do not include author names in the remainder of the text, and to the extent possible, avoid directly identifying the authors. 
You should still include all relevant references, including your own, and any other relevant discussion, 
even if this might allow a reviewer to infer the author identities. 
Use \textbackslash documentclass[final]{colt2020} only during camera-ready submission.
\end{itemize}

% Acknowledgments---Will not appear in anonymized version
\acks{We thank a bunch of people.}

\bibliography{yourbibfile}

\appendix

\section{My Proof of Theorem 1}

This is a boring technical proof.

\section{My Proof of Theorem 2}

This is a complete version of a proof sketched in the main text.

\end{document}

### 2022-08-23 design/continuous_colt/continuous-empirical.tex
Commit original repos

--- 2022-08-23 design/continuous-empirical.tex
@@
-\subsection{Empirical analysis of asymptotic consistency}
-\label{s:empirical}
-
-Theorem \ref{t:asymptotic} states that our surrogate expressions for the MSE under certain Gaussian designs are asymptotically consistent with the multiplicative error rate of $O(1/d)$. 
-In this section, we show strong empirical evidence that this fact extends to
-the setting not covered by the theorem: under-determined regime (i.e.,
-$n<d$) with a non-isotropic Gaussian distribution (i.e., $\Sigmab\neq
-\I$). We break down our analysis into verifying two
-conjectures which are of independence interest to multivariate
-Gaussian analysis. The first conjecture addresses the variance term in
-the MSE and postulates an asymptotically
-consistent formula for the expected Moore-Penrose pseudo-inverse
-of the singular Wishart distribution. Recall that matrix
-$\W\sim\Wc(\Sigmab,n)$ is distributed according to the Wishart
-distribution with $n$ degrees of freedom if it can be decomposed as
-$\W=\X^\top\X$, where $\X\sim\Nc(\zero,\Sigmab)^n$.
-\begin{conjecture}[Pseudo-inverse of singular Wishart]\label{c:wishart}
-  Fix $n/d<1$ and let $\W\sim\Wc(\Sigmab,n )$, where
-$\Sigmab$ is $d\times d$ positive definite with condition number bounded by a constant.
-  Then:
-\begin{align}
-\bigg|\frac{\E\big[\tr(\W^\dagger)\big]}{\Vc(\Sigmab,n)} -1\bigg|=
-  O(1/d)\qquad\text{for}\quad\Vc(\Sigmab,n)=\tr\big((\Sigmab+\lambda_n\I)^{-1}\big)\frac{1-\alpha_n}{d-n},\label{eq:wishart}
-\end{align}
-where $\lambda_n\geq 0$ satisfies $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$ and
-$\alpha_n=\det(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$.
-\end{conjecture}
-Our second conjecture involves the projection onto the
-orthogonal complement of a Gaussian sample $\X$, i.e., the matrix
-$\I-\X^\dagger\X$, and addresses the bias term in the MSE.
-\begin{conjecture}[Gaussian orthogonal projection]\label{c:projection}
-  Fix $n/d<1$ and let $\X\sim \Nc(\zero,\Sigmab)^n$, where
-$\Sigmab$ is $d\times d$ positive definite with condition number bounded by a constant. Then:
-\begin{align}
-\sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top
-  \Bc(\Sigmab,n)\w} - 1\bigg| = O(1/d)\qquad\text{for}\quad
-  \Bc(\Sigmab,n) =
-  (\Sigmab+\lambda_n\I)^{-1}\frac{d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})},\label{eq:projection}  
-\end{align}
-where $\lambda_n\geq 0$ satisfies $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$.
-\end{conjecture}
-Note that the surrogate expression for the mean squared
-error can be written as:
-\[
-\Mc(\Sigmab_\mu,\w^*,\sigma^2,n)=\sigma^2 \Vc(\Sigmab_\mu,n)
-+ \w^{*\top}\Bc(\Sigmab_\mu,n)\w^*.
-\]
-So, if the conjectures are true,
-this would immediately imply that the asymptotic consistency claim
-given in Theorem \ref{t:asymptotic} for
-the surrogate Gaussian MSE extends to
-the under-determined setting with arbitrary covariance $\Sigmab$.
-
-Furthermore, both of these conjectures are related to open problems
-which have been extensively studied in the literature. 
-With respect to Conjecture~\ref{c:wishart}, \cite{srivastava2003} first derived the probability
-density function of a singular Wishart
-distribution, and \cite{cook2011} computed the first and second moments of
-generalized inverses of a singular Wishart distribution. However, for the
-Moore-Penrose pseudo-inverse and arbitrary covariance $\Sigmab$,
-\cite{cook2011} claims that the quantities required
-to express the mean ``do not have tractable closed-form representation.''
-%
-Conjecture~\ref{c:projection} has connections to directional statistics.
-Using the SVD, we have the equivalent representation $\X^\dagger \X = \V \V^\top$
-where $\V$ is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal
-$n$-frames in $\R^d$).
-The distribution of $\V$ is known as the matrix angular central
-Gaussian (MACG) distribution \cite{chikuse1990matrix}. While prior work
-has considered high dimensional limit theorems \cite{CHIKUSE1991145}
-as well as density estimation and hypothesis testing \cite{CHIKUSE1998188}
-on $V_{n,d}$, they only analyzed the invariant measure
-(which corresponds in our setting to $\Sigmab = \I$),
-and to our knowledge a closed form expression of $\E[\V\V^\top]$ where
-$\V$ is distributed according to MACG with
-arbitrary $\Sigmab$ remains an open question.
-
-For verifying these two conjectures, it suffices to only consider
-diagonal covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D
-\Q^\top$ is its eigendecomposition and $\X\sim\Nc(\zero, \Q\D\Q^\top)^n$, then
-we have for $\W \sim \Wc(\Sigmab, n)$ that $\W
-\overset{d}{=} \X^\top \X$ and hence, defining
-$\Xt\sim\Nc(\zero,\D)^n$, by linearity and unitary invariance of trace,
-\begin{align*}
-  \E[\tr(\W^\dagger)]
-  &= \tr\big( \E[(\X^\top\X)^\dagger] \big)
-  = \tr\Big( \Q\E\big[(\Xt^\top\Xt)^\dagger\big]\Q^\top \Big)
-  = \tr\Big( \E\big[(\Xt^\top\Xt)^\dagger\big] \Big)
-  = \E\left[\tr \big((\Xt^\top\Xt)^\dagger\big) \right].
-\end{align*}
-Similarly, we have that
-$\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$, and a simple
-calculation shows that the expression in Conjecture~\ref{c:projection}
-is also independent of the choice of matrix $\Q$.
-Thus, we empirically validate our conjectures for diagonal matrices $\Sigmab$
-with several different eigenvalue decay profiles. Denoting
-$\lambda_1,...,\lambda_d$ as the eigenvalues of $\Sigmab$, we consider
-the following decays:
-\begin{itemize}
-  \item \texttt{diag\_linear}: linear decay, $\lambda_{i} = b-a i$;
-  \item \texttt{diag\_exp}: exponential decay, $\lambda_{i} = b\,10^{-
-      a i} $;
-  \item \texttt{diag\_poly}: fixed-degree polynomial decay, $\lambda_{i} = (b-a i)^2$;
-  \item \texttt{diag\_poly\_2}: variable-degree polynomial decay, $\lambda_i = b i^{-a}$.
-\end{itemize}
-The constants $a$ and $b$ are chosen to ensure $\lambda_{\text{max}}(\Sigmab) = 1$ and
-$\lambda_{\text{min}}(\Sigmab) = 10^{-4}$ (i.e., the condition number
-$\kappa(\Sigmab) = 10^{4}$ remains constant).
-Figure~\ref{fig:eig-decays} illustrates an example of these decay profiles
-for $d=100$.
-
-\begin{figure}[H]
+\begin{figure}[t]
   \includegraphics[width=\textwidth]{continuous_figures/decays.pdf}
+  \vspace{-1cm}
   \caption{Scree-plots of $\Sigmab$ for the eigenvalue decays examined
@@
 
-We verify our conjectures by incrasing $d$ while keeping the aspect ratio
+% Denoting
+% $\lambda_1,...,\lambda_d$ as the eigenvalues of $\Sigmab$, w
+Letting $\lambda_i(\Sigmab)$ be the $i$th largest eigenvalue of
+$\Sigmab$, we consider the
+following eigenvalue profiles:
+\begin{itemize}
+  \item \texttt{diag\_linear}: linear decay, $\lambda_i(\Sigmab)= b-a i$;
+  \item \texttt{diag\_exp}: exponential decay, $\lambda_i(\Sigmab) = b\,10^{- a i} $;
+  \item \texttt{diag\_poly}: fixed-degree polynomial decay, $\lambda_i(\Sigmab) = (b-a i)^2$;
+  \item \texttt{diag\_poly\_2}: variable-degree polynomial decay, $\lambda_i(\Sigmab) = b i^{-a}$.
+\end{itemize}
+The constants $a$ and $b$ are chosen to ensure $\lambda_{\text{max}}(\Sigmab) = 1$ and
+$\lambda_{\text{min}}(\Sigmab) = 10^{-4}$ (i.e., the condition number
+$\kappa(\Sigmab) = 10^{4}$ remains constant).
+
+We verify our conjectures by increasing $d$ while keeping the aspect ratio
 $n/d$ fixed and examining the rate of decay of the quantities asserted
@@
 
-% For Conjecture~\ref{c:wishart}, sampling $\W\sim\Wc(\Sigmab,n)$ is
+% For Conjecture~\ref{c:wishart}, sampling $\W\sim\Pc\Wc(\Sigmab,n)$ is
 % elementary and the only nontrivial aspect of computing $\Vc(\Sigmab, n)$
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \includegraphics[width=\textwidth]{continuous_figures/wishart_var.pdf}
+  \includegraphics[width=\textwidth]{continuous_figures/proj_bias.pdf}
+  \vspace{-1cm}
   \caption{
-    Empirical verification of Conjecture~\ref{c:wishart}. We show
-    the quantity $\big|\E[\tr(\W^\dagger)]\,\Vc(\Sigmab,n)^{-1} -1\big|$
-    as $d$ increases for various aspect ratios $n/d$. Consistent with our conjecture,
+    Empirical verification of Conjectures~\ref{c:wishart} (top) and
+    \ref{c:projection} (bottom).
+    We show the quantities
+    $|\E[\tr(\W^\dagger)]\,\Vc(\Sigmab,n)^{-1} -1|$ and
+    $\|\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-1} - \I\|$, respectively,
+    as $d$ increases for various aspect ratios $n/d$.
+    Consistent with our conjectures,
     a $O(1/d)$ decay (linear with slope $-1$ on a log-log plot) is
@@
 The results of empirically validating Conjecture~\ref{c:wishart} are
-illustrated in Figure~\ref{f:conj-wishart}, where we performed
+illustrated in Figure~\ref{f:conj-wishart} (top), where we performed
 Monte Carlo estimation of $\E\big[\tr(\W^\dagger)\big]$ and
@@
 confidence intervals, and our results use the bootstrapping
-method described in \cite{lopes2019bootstrapping}.
+method described by \cite{lopes2019bootstrapping}.
 
-Figure~\ref{f:conj-bias} shows
+Figure~\ref{f:conj-wishart} (bottom) shows
 how $\big\|\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-1} - \I\big\|$ decays as we
@@
 rate posed by Conjecture~\ref{c:projection}. Note that the range of
-$d$ is smaller than in Figure \ref{f:conj-wishart} because the large
+$d$ is smaller than in Figure \ref{f:conj-wishart} (top) because the large
 number of Monte Carlo samples (up to two million) required for this
 experiment made the computations much more expensive.
-
-\begin{figure}[H]
-  \begin{center}
-    \includegraphics[width=\textwidth]{continuous_figures/proj_bias.pdf}
-  \end{center}
-  \caption{
-    Empirical results validating Conjecture~\ref{c:projection}.
-We show how
-    $\big\|\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-1} - \I\big\|$
-    (which by Equation~\ref{eq:proj-to-operator-norm}
-    is equal to the quantity controlled by Conjecture~\ref{c:projection})
-    decays as $d$ increases and observe a linear slope consistent with the conjectured $O(1/d)$ rate.
-  }
-  \label{f:conj-bias}
-\end{figure}

### 2022-08-23 design/continuous_colt/continuous.tex
Commit original repos

--- 2022-08-23 design/continuous.tex
@@
+\newif\ifisarxiv
+\isarxivtrue
+
+\ifisarxiv
 \documentclass[11pt]{article}
+\usepackage[round]{natbib}
 \usepackage{fullpage}
@@
 \usepackage{wrapfig}
-\usepackage{subcaption}
+\usepackage{subfigure}
 \usepackage{caption}
@@
 \usepackage{cleveref}
-\input{shortdefs}
 
@@
 
-% \author{
-%       Micha{\l } Derezi\'{n}ski \\
-% Department of Statistics\\
-% University of California, Berkeley\\
-% \texttt{mderezin@berkeley.edu}
-% }
-
-% \setlength{\parindent}{0cm}
-% \setlength{\parskip}{0mm}
+%  \newcommand{\jmlrBlackBox}{\rule{1.5ex}{1.5ex}}
+  \providecommand{\BlackBox}{\rule{1.5ex}{1.5ex}}
+%  \newcommand{\jmlrQED}{\hfill\jmlrBlackBox\par\bigskip}
+  \renewenvironment{proof}%
+  {%
+   \par\noindent{\bfseries\upshape Proof\ }%
+  }%
+  {\hfill\BlackBox\par\bigskip}
+
+\else
+\documentclass[anon]{colt2020}
+
+\title[Exact expressions for double descent and implicit
+regularization]{Exact expressions for double descent and implicit
+  regularization\\
+  via surrogate random design}
+\usepackage{times}
+
+\coltauthor{%
+  \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
+  \addr Department of Statistics\\
+  University of California, Berkeley\\
+  \AND
+  \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
+  \addr Department of Statistics\\
+  University of California, Berkeley\\
+  \AND
+  \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
+  \addr ICSI and Department of Statistics\\
+  University of California, Berkeley
+}
+
+\usepackage{float}
+\usepackage{wrapfig}
+
+\fi
+
+\input{../shortdefs}
 
@@
 
+\ifisarxiv\else
+\begin{keywords}
+  Least squares, determinantal point process,
+  double descent, implicit regularization
+\end{keywords}
+\fi
+
 \section{Introduction}
 
-Classical statistical learning theory asserts that to achieve generalization one must use training sample size that sufficiently exceeds the complexity of the learning model, where the latter is typically represented by the number of parameters (or some related structural parameter) \cite{HFT09}.
-In particular, this seems to suggest the conventional wisdom that one should not use models that fit the training data exactly. 
-However, modern machine learning
-practice often seems to go against this intuition, using models with so many
-parameters that the training data can be perfectly interpolated, in which case
-the training error vanishes. It has been shown that 
-models such as deep neural networks, as well as certain so-called interpolating kernels and decision
-trees, can generalize well in this regime. In particular, recent work \cite{BHMM19}
-empirically demonstrated a phase transition in generalization
-performance of learning models which occurs at an
-\emph{interpolation thershold}, i.e., a point where training error
-goes to zero (as one varies the ratio between the model complexity and
-the sample size). Moving away from this threshold in 
-either direction tends to reduce the generalization error, leading to
-the so-called \emph{double descent} curve.
+Classical statistical learning theory asserts that to achieve generalization
+one must use training sample size that sufficiently exceeds the complexity of
+the learning model, where the latter is typically represented by the number of
+parameters \citep[or some related structural parameter; see][]{HFT09}.  In particular,
+this seems to suggest the conventional wisdom that one should not use models
+that fit the training data exactly.  However, modern machine learning practice
+often seems to go against this intuition, using models with so many parameters
+that the training data can be perfectly interpolated, in which case the
+training error vanishes. It has been shown that models such as deep neural
+networks, as well as certain so-called interpolating kernels and decision
+trees, can generalize well in this regime. In particular,
+\cite{BHMM19} empirically demonstrated a phase transition in generalization
+performance of learning models which occurs at an \emph{interpolation
+thershold}, i.e., a point where training error goes to zero (as one varies the
+ratio between the model complexity and the sample size). Moving away from this
+threshold in either direction tends to reduce the generalization error, leading
+to the so-called \emph{double descent} curve.
 
@@
 The simplest example of an estimator that has been shown to exhibit
-the double descent phenomenon \cite{belkin2019two} is the
+the double descent phenomenon \citep{belkin2019two} is the
 Moore-Penrose estimator, $\wbh=\X^\dagger\y$:
@@
 Of particular interest are results that are exact (i.e., not upper/lower bounds) and
-non-asymptotic (i.e., for large but still finite $n$ and $d$).
-
-We build on methods from Randomized Numerical
-Linear Algebra (RandNLA) in order to obtain \emph{exact non-asymptotic
-expressions} for the mean squared error (MSE) of the
-Moore-Penrose estimator (see Theorem~\ref{t:mse}). 
-This provides a precise characterization of the double descent
-phenomenon for perhaps the simplest and most ubiquitous regression
-problem. 
-In obtaining these results, we are able to provide precise
-formulas for the \emph{implicit regularization} induced by
-minimum norm solutions of
-under-determined training samples, relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}). 
+non-asymptotic (i.e., for large but still finite $n$ and~$d$).
+
+We build on methods from Randomized Numerical Linear Algebra (RandNLA) in order
+to obtain \emph{exact non-asymptotic expressions} for the mean squared error
+(MSE) of the Moore-Penrose estimator (see Theorem~\ref{t:mse}).  This provides
+a precise characterization of the double descent phenomenon for perhaps the
+simplest and most ubiquitous regression problem.  In obtaining these results,
+we are able to provide precise formulas for the \emph{implicit regularization}
+induced by minimum norm solutions of under-determined training samples,
+relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}).
 This result has been observed empirically for RandNLA
-methods~\cite{Mah-mat-rev_JRNL}, but it has also been shown 
-in deep learning~\cite{Ney17_TR} and machine
-learning~\cite{Mah12} more generally. 
-To obtain our precise results, we use a somewhat non-standard random
-design, which we term surrogate random design (see
-Section~\ref{s:determinantal} for a detailed discussion), and which we
-expect will be of more general interest. 
-Informally, the goal of a surrogate random design is to modify an
-original design to capture its main properties
-while being ``nicer'' in some useful way. In
-Theorem~\ref{t:asymptotic} and Section \ref{s:asymptotic} we show,
-both theoretically and empirically, that our surrogate design
-accurately preserves the key properties of the original design when
-the data distribution is a multivariate Gaussian.
+methods~\citep{Mah-mat-rev_JRNL}, but it has also been shown in deep
+learning~\citep{Ney17_TR} and machine learning~\citep{Mah12} more generally.  To
+obtain our precise results, we use a somewhat non-standard random design, which
+we term surrogate random design (see Section~\ref{s:determinantal} for a
+detailed discussion), and which we expect will be of more general interest.
+Informally, the goal of a surrogate random design is to modify an original
+design to capture its main properties while being ``nicer'' in some useful way.
+In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
+theoretically and empirically, that our surrogate design accurately preserves
+the key properties of the original design when the data distribution is a
+multivariate Gaussian.
 
@@
 underlying linear model of the responses.
-In analyzing the MSE, we make the following standard assumption on the response noise.
+In analyzing the MSE, we make the following standard assumption that
+the response noise is homoscedastic with variance $\sigma^2$.
 \begin{assumption}[Homoscedastic noise]\label{a:linear}
@@
 Our main result provides an exact expression for the MSE of the
-Moore-Penrose estimator under our surrogate design denoted $S_\mu^n$, where
+Moore-Penrose estimator under our surrogate design denoted $\Xb\sim S_\mu^n$, where
 $\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
@@
 \noindent
-Under Assumptions~\ref{a:linear} and~\ref{a:general-position}, we can establish our first main result, stated as the following theorem.
+Under Assumptions~\ref{a:linear} and~\ref{a:general-position}, we can
+establish our first main result, stated as the following theorem, where
+we use $\X^\dagger$ to denote the Moore-Penrose inverse of $\X$.
+% , which gives an exact non-asymptotic
+% expression for the MSE of the Moore-Penrose estimator  the
+% surrogate design $S_\mu^n$.
 
@@
 \label{t:mse}
-  If the response noise is homoscedastic with variance~$\sigma^2$
+  If the response noise is homoscedastic % with variance~$\sigma^2$
   (Assumption~\ref{a:linear}) and $\mu$ is in
   general position (Assumption~\ref{a:general-position}), then for
-  $\Xb\sim S_\mu^n$ (Definition~\ref{d:main}) and
-  $\yb_i=y(\xbb_i)$, we have
+  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and
+  $\yb_i=y(\xbb_i)$, %we have
   \begin{align*}
@@
   \end{align*}
-  where $\lambda_n\geq 0$ satisfies
-  $\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})=n$, while
+with $\lambda_n\geq 0$ defined by
+  $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$, \
   $\alpha_n=\det(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$
@@
 \end{definition}
-For illustration, we plot these MSE expressions in Figure \ref{f:intro}a,
-comparing them with empirical estimates of the true MSE under the
-i.i.d.~design for a multivariate Gaussian distribution
-$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
-$100$ and vary the number of samples $n$, observing a double descent
-peak at $n=d$. We observe that our theory aligns well with
-the empirical estimates, whereas
-previously, no such theory was available except for special
-cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
-and Section \ref{s:asymptotic}). The plots
-show that varying the spectral decay of $\Sigmab$ has a significant effect on the
-shape of the curve in the under-determined regime. We use the
-horizontal line to denote the MSE of the null estimator
-$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
-decay rapidly, then the Moore-Penrose estimator suffers less error 
-than the null estimator for some values of $n<d$, and the curve
-exhibits a local optimum in this regime.
-
-\begin{figure}[h]
+
+\begin{figure}[t]
 \centering
-\begin{subfigure}[t]{0.48\textwidth}
-%  \begin{center}
-    \includegraphics[width=\textwidth]{figs/descent-intro}
-%  \end{center}
-  \caption{
+\subfigure[%
   Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
@@
   features. Eigenvalue decay leads to a steeper
-descent curve in the under-determined regime ($n<d$).}
-\end{subfigure}
+  descent curve in the under-determined regime ($n<d$).]{%
+    \includegraphics[width=0.48\textwidth]{figs/descent-intro}
+  }
 \hfill
-\begin{subfigure}[t]{0.48\textwidth}
-%  \begin{center}
-    \includegraphics[width=\textwidth]{figs/descent-shrinkage}
-%  \end{center}
-  \caption{
+\subfigure[%
     The mean of the estimator $\X^\dagger\y$ exhibits
@@
     ridge-regularized least squares optimum (theory lines), as characterized by
-    Theorem \ref{t:unbiased}.}
-\end{subfigure}
+    Theorem \ref{t:unbiased}.]{%
+      \includegraphics[width=0.48\textwidth]{figs/descent-shrinkage}
+    }
 \caption{Illustration of the main results for $d=100$ and
@@
 $\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
-i.e., $\Sigmab=\I$. In all cases, we use $\sigma^2=\|\w^*\|=1$ with
+i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
 $\w^*=\frac1{\sqrt{d}}\one$.}
+\vspace{-5mm}
 \label{f:intro}
@@
 
+\noindent
+Proof of Theorem \ref{t:mse} is given in Appendix \ref{a:mse-proof}.
+For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
+comparing them with empirical estimates of the true MSE under the
+i.i.d.~design for a multivariate Gaussian distribution
+$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
+$100$ and vary the number of samples $n$, observing a double descent
+peak at $n=d$. We observe that our theory aligns well with
+the empirical estimates, whereas
+previously, no such theory was available except for special
+cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
+and Appendices \ref{sec:proof-of-t-asymptotic} and \ref{sec:asymp-conj-details}). The plots
+show that varying the spectral decay of $\Sigmab$ has a significant effect on the
+shape of the curve in the under-determined regime. We use the
+horizontal line to denote the MSE of the null estimator
+$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
+decay rapidly, then the Moore-Penrose estimator suffers less error
+than the null estimator for some values of $n<d$, and the curve
+exhibits a local optimum in this regime.
+
 One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$.
-This expression is precisely the classical notion of \emph{effective dimension} for ridge regression regularized with $\lambda_n$~\cite{Alaoui15}, and it arises here even though there is no explicit ridge regularization in the problem being considered in Theorem~\ref{t:mse}.
+This expression is precisely the classical notion of \emph{effective
+  dimension} for ridge regression regularized with
+$\lambda_n$~\citep{ridge-leverage-scores}, and it arises here even though there is
+no explicit ridge regularization in the problem being considered in
+Theorem~\ref{t:mse}. 
 The global solution to the ridge regression task (i.e., $\ell_2$-regularized
@@
     + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
-  \lambda\I)^{-1}\v_{\mu,y},\qquad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
+  \lambda\I)^{-1}\v_{\mu,y},\quad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
 \end{align*}
@@
 theorem applies more generally than Theorem~\ref{t:mse}, in that it
-does \emph{not} require the responses to follow any linear model as in Assumption~\ref{a:linear}.  
+does \emph{not} require the responses to follow any linear model as in
+Assumption~\ref{a:linear} (proof in Appendix~\ref{s:unbiased-proof}). 
 \begin{theorem}[Implicit regularization of Moore-Penrose estimator]
@@
   satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
-  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and any
-  $y(\cdot)$ such that   $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
-  $\Xb\sim S_\mu^n$ and $\yb_i=y(\xbb_i)$, then
+  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and
+  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
+  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$, then
   \begin{align*}
@@
   \end{align*}
-  where, as in Theorem \ref{t:mse}, $\lambda_n$ is chosen so that the effective dimension
+  where, as in Theorem \ref{t:mse}, $\lambda_n$ is such that the effective dimension
   $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
@@
 Such \emph{implicit regularization} has been studied
-previously~\cite{MO11-implementing, PM11, GM14_ICML, Mah12}; it has
+previously~\citep[see, e.g.,][]{MO11-implementing, %PM11, GM14_ICML,
+Mah12}; it has
 been observed empirically for RandNLA sampling
-algorithms~\cite{Mah-mat-rev_JRNL, MMY15}; and it has also received attention more
-generally within the context of neural networks~\cite{Ney17_TR}. While our implicit regularization result
+algorithms~\citep{MMY15}; and it has also received attention more
+generally within the context of neural networks~\citep{Ney17_TR}. While our implicit regularization result
 is limited to the Moore-Penrose estimator, this new connection (and
@@
 arbitrary Gaussian for the over-determined one).
+We give the proof in Appendix~\ref{sec:proof-of-t-asymptotic}.
 \begin{theorem}[Asymptotic consistency of surrogate design]
@@
 \noindent
-Empirical estimates given in Figure \ref{f:intro} suggest that the
-consistency of surrogate expressions holds much more generally than
-it is stated above. Based on a detailed empirical analysis described in
-Section~\ref{s:empirical}, we conjecture that an asymptotic consistency
+Empirical estimates given in Figure
+\ref{f:intro} suggest that the consistency of surrogate expressions
+holds much more generally than it is stated above. Based on a detailed
+empirical analysis described in Appendix \ref{sec:asymp-conj-details},
+we conjecture that an asymptotic consistency 
 result of the form similar to the statement of Theorem~\ref{t:asymptotic}
 holds true in the under-determined regime without the assumption that
-$\Sigmab=\I$. In this case, no formula is known for
+$\Sigmab=\I$ (see Conjectures \ref{c:wishart} and
+\ref{c:projection}). In this case, no formula is known for 
 $\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
@@
 A key element in the construction of our surrogate designs involves rescaling the measure $\X\sim\mu^n$ by the pseudo-determinant $\pdet(\X\X^\top)$, i.e., a product of the non-zero eigenvalues.
-A similar type of determinantal design was suggested in prior work
+A similar type of determinantal design was suggested by
 \cite{correcting-bias-journal}, but it was restricted there only to
@@
 less than $d$, but also allowing it to be randomized.
-Our definition of a determinantal design matrix $\Xb$ follows by
-expressing $\E[F(\Xb)]$ for any real-valued function $F(\cdot)$ as
-(see Definition~\ref{d:det}):  
+A key property of our determinantal design matrix $\Xb$ is that
+the expectation of a function $F(\Xb)$ can be, up to normalization,
+expressed as:
 \begin{align*}
@@
   \end{align*}
-where $\X\sim\mu^K$ and $K$ is a random variable.
+where $\X\sim\mu^K$ and $K$ is a random variable (see details in Definition~\ref{d:det}).
 Then, we define (in Definition~\ref{d:surrogate}) our surrogate design $S_\mu^n$ for each $n>0$ as a determinantal design with a carefully chosen random variable $K$, so that the expected sample size is equal to $n$ and so that it is possible to derive closed form expressions for the MSE.
-We achieve this by using modifications of the Poisson distribution to construct the variable $K$.
+We achieve this by using the Poisson distribution to construct the variable~$K$.
 
@@
 techniques to prove a number of determinantal expectation
-formulas. For example, we show that if $\X\sim\mu^K$, where $K$ is a
-Poisson random variable, then:
+formulas needed in obtaining Theorems~\ref{t:mse}
+and~\ref{t:unbiased}. For example, we show that if $\X\sim\mu^K$,
+where $K$ is a Poisson random variable, then:
 \begin{align*}
@@
 \end{align*}
-These formulas are used to derive the normalization constants for our
-surrogate design distribution, which are later used in proving
-Theorems~\ref{t:mse} and~\ref{t:unbiased}.
-
-\subsection{Related work}
+
+\section{Related work}
 \label{s:related-work}
@@
 \textbf{Double descent.}
-The double descent phenomenon (a term introduced by \cite{BHMM19})
+The double descent phenomenon \cite[a term introduced by][]{BHMM19}
 corresponds to the phase transition in the generalization error that
@@
 in a number of learning models, including neural networks
-\cite{BHMM19,GJSx19_TR}, kernel methods \cite{BMM18_TR,BRT18_TR},
-nearest neighbor models \cite{BHM18_TR}, and decision trees \cite{BHMM19}. The
+\citep{BHMM19,GJSx19_TR}, kernel methods \citep{BMM18_TR,BRT18_TR},
+nearest neighbor models \citep{BHM18_TR}, and decision trees \citep{BHMM19}. The
 theoretical analysis of double descent, and more broadly the generalization
@@
 which the response model is linear
-\cite{belkin2019two,HMRT19_TR,Mit19_TR,MM19_TR}, e.g., when the
+\citep{belkin2019two,HMRT19_TR,Mit19_TR,MM19_TR}, e.g., when the
 learner observes a random subset of $d$ features from a larger
@@
 Randomized numerical linear algebra
-\cite{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR} has
+\citep{DM16_CACM,RandNLA_PCMIchapter_TR} has
 traditionally focused on obtaining algorithmic improvements for tasks
 such as least squares and low-rank approximation via techniques that include
-sketching \cite{sarlos-sketching} and i.i.d.~leverage score sampling
-\cite{Drineas2006sampling}. However, there has been growing
+sketching \citep{sarlos-sketching} and i.i.d.~leverage score sampling
+\citep{Drineas2006sampling}. However, there has been growing
 interest in understanding the statistical properties of these
-randomized methods \cite{MMY15,GarveshMahoney_JMLR}, for example
+randomized methods \citep{MMY15,GarveshMahoney_JMLR}, for example
 looking at the mean squared error of the least squares estimator
@@
 into RandNLA in the context of low-rank approximation
-\cite{pca-volume-sampling}, have been recently shown to combine
+\cite[][]{pca-volume-sampling}, have been recently shown to combine
 strong worst-case guarantees with elegant statistical properties.
@@
 \cite{leveraged-volume-sampling,correcting-bias,minimax-experimental-design},
-however they were still limited to the over-determined setting (with
-the exception of
-\cite{regularized-volume-sampling,bayesian-experimental-design} who
-gave upper bounds on the mean squared error of the ridge estimator
-under different determinantal samplings). Also in the over-determined setting,
+however they were still limited to the over-determined setting 
+\cite[with the exception
+of][who gave upper bounds on the mean squared error of the ridge estimator
+under different determinantal
+samplings]{regularized-volume-sampling,bayesian-experimental-design}.
+Also in the over-determined setting, 
 \cite{correcting-bias-journal} provided evidence for the fact that
@@
 See, e.g., theoretical results on simplified
-models~\cite{NTS14_TR,Ney17_TR,SHNx17_TR,GWBNx17_TR,ACHL19_TR,KBMM19_TR}
+models~\citep{NTS14_TR,Ney17_TR,SHNx17_TR,GWBNx17,ACHL19,KBMM19_TR}
 as well as extensive empirical and phenomenological results on state-of-the-art neural
-network models~\cite{MM18_TR,MM19_HTSR_ICML}.
+network models~\citep{MM18_TR,MM19_HTSR_ICML}.
 The implicit regularization observed by us is different in that it is
@@
 solution). In this context, most relevant are the
-asymptotic results of~\cite{LJB19_TR} (which used the asymptotic risk
-results for ridge regression of~\cite{DW15_TR}) and~\cite{KLS18_TR}. 
+asymptotic results of~\cite{LJB19_TR} \cite[which used the asymptotic risk
+results for ridge regression of][]{DW15_TR} and~\cite{KLS18_TR}.
 Our non-asymptotic results are also related to recent work in
 RandNLA on the expectation of the 
-inverse~\cite{determinantal-averaging} and generalized
-inverse~\cite{MDK19_TR} of a subsampled matrix.
+inverse~\citep{determinantal-averaging} and generalized
+inverse~\citep{MDK19_TR} of a subsampled matrix.
 
@@
 $(i,j)$th entry of $\adj(\A)$ is
-$(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$. We will
-use two useful identities related to the adjugate: (1)
-$\adj(\A)=\det(\A)\A^{-1}$ for invertible $\A$, and (2)
-$\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$.
+$(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$.
+% We will
+% use two useful identities related to the adjugate: (1)
+% $\adj(\A)=\det(\A)\A^{-1}$ for invertible $\A$, and (2)
+% $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$.
 % A formula called Sylvester's theorem
@@
 $\Sigmab_\mu$ is well-defined and positive definite.
-We use $\Poisson(\gamma)_{\leq a}$ as the Poisson distribution restricted to values less than or equal to $a$, and a similar convention is used for the restriction to values greater or equal $a$. 
-Finally, we use $\#(\X)$ to denote the number of rows of $\X$.
+We use $\Poisson(\gamma)_{\leq a}$ as the Poisson distribution
+restricted to $[0,a]$, whereas $\Poisson(\gamma)_{\geq a}$ is restricted
+to $[a,\infty)$.
+We also let $\#(\X)$ denote the number of rows of $\X$.
 
@@
 designs that enables our analysis.
-Our definition follows by
-expressing $\E[F(\Xb)]$ for real-valued functions $F:\bigcup_{k\geq 0}\R^{k\times
-  d}\rightarrow \R$ (the expectation may be undefined for some functions).  
+% Our definition follows by
+% expressing $\E[F(\Xb)]$ for real-valued functions $F:\bigcup_{k\geq 0}\R^{k\times
+%   d}\rightarrow \R$ (the expectation may be undefined for some functions).
 \begin{definition}\label{d:det}
@@
     $\Xb\sim \Det(\mu,K)$ is a
-distribution such that~for any $F(\cdot)$ as above,
-\begin{align*}
-  \E\big[F(\Xb)\big]\  \propto\ \E[\pdet(\X\X^\top)F(\X)]
-  \quad\text{for}\quad\X\sim\mu^K.
+distribution with the same domain as $\X\sim\mu^K$ such that~for any
+event $E$ measurable w.r.t.~$\X$, we have
+\begin{align*}
+\Pr\big\{\Xb\in E\big\}\ = \frac{\E[\pdet(\X\X^\top)\one_{[\X\in E]}]}{\E[\pdet(\X\X^\top)]}.
 \end{align*}
@@
 \noindent
-Setting $F(\cdot)$ to 1, observe that the proportionality constant
-must be $1/\E[\pdet(\X\X^\top)]$. The above definition can be
+% Setting $F(\cdot)$ to 1, observe that the proportionality constant
+% must be $1/\E[\pdet(\X\X^\top)]$.
+The above definition can be
 interpreted as rescaling the density function of $\mu^K$ by the
 pseudo-determinant, and then renormalizing it.
-
 We now construct our surrogate design $S_\mu^n$ by appropriately
 selecting the random variable $K$.
-One might be tempted to use the obvious choice of $K=n$, but this does \emph{not} result in simple closed form expressions for the MSE in the under-determined regime (i.e., $n<d$), which is the regime of primary interest to us.
+The obvious choice of $K=n$ does \emph{not} result in simple closed form expressions for the MSE in the under-determined regime (i.e., $n<d$), which is the regime of primary interest to us.
 Instead, we derive our random variables $K$ from the Poisson distribution.
-
 \begin{definition}\label{d:surrogate}
@@
 define surrogate design $S_\mu^n$ as $\Det(\mu,K)$ where:
+\vspace{-2mm}
     \begin{enumerate}
 \item if $n<d$, then $K\sim \Poisson(\gamma_n)_{\leq d}$ with
- $\gamma_n$ being the solution of $n=\tr(\Sigmab_\mu(\Sigmab_\mu+\frac1{\gamma_n}\I)^{-1})$, 
+ $\gamma_n$ being the solution of
+ $n=\tr(\Sigmab_\mu(\Sigmab_\mu+\frac1{\gamma_n}\I)^{-1})$,
+ \vspace{-2mm}
 \item if $n=d$, then we simply let $K=d$,
+  \vspace{-2mm}
 \item if $n>d$, then $K\sim\Poisson(\gamma_n)_{\geq d}$ with $\gamma_n=n-d$.
@@
 that the expected sample size is in fact always equal to $n$, which we
-prove at the end of this section.
+prove in Appendix \ref{appx: proof-of-l-size}.
 \begin{lemma} \label{l:size}
@@
 
+\noindent
  Our general template for computing expectations under
@@
 i.i.d.~random design $\X\sim\mu^K$:
-\begin{align*}
+\begin{align}
   \E[F(\Xb)] &=\begin{cases}
@@
    \quad K\sim\Poisson(\gamma_n)&\text{for }n>d.
-  \end{cases}
-\end{align*}
+  \end{cases}\label{eq:cases}
+\end{align}
 These formulas follow from Definitions \ref{d:det} and
@@
 range of the Poisson distribution.
-Crucially, the normalization constants for computing the expectations
+The normalization constants for computing the expectations
 can be obtained using the following formulas: if $\X\sim\mu^K$ then
@@
 \end{align*}
-\begin{remark}
-We will use $Z_\mu^n$ as a shorthand for the above normalization constants.
-\end{remark}
+% \begin{remark}
+% We will use $Z_\mu^n$ as a shorthand for the above normalization constants.
+% \end{remark}
 % For the case of $n=d$, the normalization constant can be found in the
@@
 \ref{s:dp} by introducing the
-concept of determinant preserving random matrices. The lemmas play a 
-crucial role in deriving a number of new expectation formulas for the
-under- and over-determined surrogate designs that we use to prove Theorems~\ref{t:mse}
-and~\ref{t:unbiased} in Section~\ref{s:expectations}.
+concept of determinant preserving random matrices.
+% The lemmas play a
+% crucial role in deriving a number of new expectation formulas for the
+% under- and over-determined surrogate designs that we use to prove Theorems~\ref{t:mse}
+% and~\ref{t:unbiased} in Section~\ref{s:expectations}.
 On the other hand, Lemma  \ref{l:cb} and the design $S_\mu^d$ can be found in the
-literature \cite{expected-generalized-variance,correcting-bias-journal},
-and we will rely on those known results in this special case. 
-Importantly, the $n=d$ case offers a continuous transition between the under-
-and over-determined regimes because the distribution $S_\mu^n$
-converges to $S_\mu^d$ when $n$ approaches $d$ from above and
-below. Another important property of the design $S_\mu^d$ is that it
-can be used to construct an over-determined design for any $n>d$. A
-similar version of this result was also previously shown by
-\cite{correcting-bias-journal} for a different determinantal design.
-  \begin{lemma}\label{l:decomposition}
-Let $\Xb\sim S_\mu^d$ and $\X\sim \mu^K$, where
-$K\sim\Poisson(\gamma)$. Then the matrix composed of a random
-permutation of the rows from $\Xb$ and $\X$ is distributed according to
-$S_\mu^{d+\gamma}$.
+literature \citep{expected-generalized-variance,correcting-bias-journal},
+and we will rely on those known results in this case.
+% Importantly, the $n=d$ case offers a continuous transition between the under-
+% and over-determined regimes because the distribution $S_\mu^n$
+% converges to $S_\mu^d$ when $n$ approaches $d$ from above and
+% below.
+
+We now highlight some key expectation formulas for surrogate designs, which are used to
+derive the MSE expressions from Theorem \ref{t:mse}. A standard decomposition of
+the MSE yields:% for the Moore-Penrose estimator yields:
+\begin{align}
+  \MSE{\Xb^\dagger\ybb}
+  &= \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\big]
+  =\sigma^2\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big] +
+    \w^{*\top}\E\big[\I-\Xb^\dagger\Xb\big]\w^*.\label{eq:mse-derivation}
+\end{align}
+Thus, our task is to find closed form expressions for the two
+expectations above. If $n\geq d$, then the latter goes away because
+$\Xb^\dagger\Xb$ is the projection onto the row-span of $\Xb$ so
+when $\Xb$ has rank $d$ then $\I-\Xb^\dagger\Xb=\zero$. When
+$n<d$, it is given in the following result (proof in
+Appendix \ref{s:unbiased-proof}).
+% can be obtained as a consequence of Lemma \ref{l:ridge-under}.
+\begin{lemma}\label{l:proj}
+If  $\Xb\sim S_\mu^n$ and $n<d$, then we have:
+$\E\big[\I-\Xb^\dagger\Xb\big] = (\gamma_n\Sigmab_\mu+\I)^{-1}$.
 \end{lemma}
-\begin{proof}
-Let $\Xt$ denote the matrix constructed from the permuted rows of
-$\Xb$ and $\X$.  Letting $\Z\sim\mu^{K+d}$, we derive the expectation
-$\E[F(\Xt)]$ by summing over the possible index subsets  $S\subseteq
-[K+d]$ that correspond to the rows coming from $\Xb$:
-\begin{align*}
-\E[F(\Xt)] &= \E\bigg[\frac1{{K+d\choose
-  d}}\sum_{S:\,|S|=d}\frac{\E[\det(\Z_{S,*})^2F(\Z)\mid
-  K]}{d!\det(\Sigmab_\mu)}\bigg]\\
-  &=\sum_{k=0}^\infty
-    \frac{\gamma^k\ee^{-\gamma}}{k!}\,\frac{\gamma^dk!}{(k+d)!}\,
-    \frac{\E\big[\sum_{S:\,|S|=d}\det(\Z_{S,*})^2F(\Z)\mid
-    K=k\big]}{\det(\gamma\Sigmab_\mu)}\\
-  &\overset{(*)}{=}\sum_{k=0}^\infty
-    \frac{\gamma^{k+d}\ee^{-\gamma}}{(k+d)!}\,\frac{\E[\det(\Z^\top\Z)F(\Z)\mid K=k]}{\det(\gamma\Sigmab_\mu)},
-\end{align*}
-where $(*)$ uses the Cauchy-Binet formula to sum over all subsets $S$
-of size $d$. Finally, since the sum shifts from $k$
-to $k+d$, the last expression can be rewritten as
-$\E[\det(\X^\top\X)F(\X)]/\det(\gamma\Sigmab_\mu)$, where recall that
-$\X\sim\mu^K$ and $K\sim\Poisson(\gamma)$, matching the definition of $S_\mu^{d+\gamma}$.
-\end{proof}
-We now return to the proof of Lemma \ref{l:size}, where we establish
-that the expected sample size of $S_\mu^n$ is indeed $n$.
-\begin{proof} (of Lemma \ref{l:size}) \
-  The result is obvious when $n=d$, whereas
-  for $n>d$ it is an immediate consequence 
-  of Lemma \ref{l:decomposition}.
-  Finally, for $n<d$ the expected sample
-  size follows as a corollary of a more general expectation formula
-  proven in Section \ref{s:expectations}, which states that
-  \begin{align*}
-\text{(Lemma \ref{l:proj})} \qquad\E\big[\I - \Xb^\dagger\Xb\big] =
-    (\gamma_n\Sigmab_\mu + \I)^{-1},
-  \end{align*}
-  where $\Xb^\dagger\Xb$ is the orthogonal projection onto
-  the subspace spanned by the rows of $\Xb$. Since the rank of this
-  subspace is equal to the number of the rows, we have
-  $\#(\Xb)=\tr(\Xb^\dagger\Xb)$, so
-  \begin{align*}
-    \E\big[\#(\Xb)\big] = d - \tr\big((\gamma_n\Sigmab_\mu +
-    \I)^{-1}\big) =
-    \tr\big(\gamma_n\Sigmab_\mu(\gamma_n\Sigmab_\mu+\I)^{-1}\big) = n,
-  \end{align*}
-  which completes the proof.
-\end{proof}
+
+\noindent
+No such expectation formula is known for i.i.d.~designs, except when
+$\mu$ is an isotropic Gaussian. In Appendix \ref{s:unbiased-proof}, we
+also prove a generalization of Lemma \ref{l:proj} which is then used
+to establish our implicit regularization result
+(Theorem~\ref{t:unbiased}). We next give an expectation formula 
+for the trace of the Moore-Penrose inverse of the covariance
+matrix for a surrogate design (proof in Appendix \ref{a:mse-proof}).
+\begin{lemma}\label{l:sqinv-all}
+If  $\Xb\sim S_\mu^n$, then:\vspace{-2mm}
+\begin{align*}
+    \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
+  &=
+    \begin{cases}
+\gamma_n\big(1-
+\det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big),
+&\text{for }n<d,\\
+\tr(\Sigmab_\mu^{-1}),
+&\text{for }n=d,\\
+\tr(\Sigmab_\mu^{-1})\,\frac{1-\ee^{-\gamma_n}}{\gamma_n},
+&\text{for }n>d.
+\end{cases}
+\end{align*}
+\end{lemma}
+
+\noindent
+% The case of $n=d$ in the above lemma was shown by
+% \cite{correcting-bias-journal} in Theorem~2.12, whereas the case of $n>d$
+% is closely related to their Theorem 2.9.
+Our key contribution is the under-determined
+regime (i.e., $n<d$), where we observe implicit regularization
+given by $\lambda_n=\frac1{\gamma_n}$. Since $n =
+\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})=d-\lambda_n\tr((\Sigmab_\mu+\lambda_n\I)^{-1})$,
+it follows that $\lambda_n=(d-n)/\tr((\Sigmab_\mu+\lambda_n\I)^{-1})$.
+Combining this with Lemmas \ref{l:proj} and \ref{l:sqinv-all}, we
+recover the surrogate MSE expression in Theorem~\ref{t:mse} (see details
+in Appendix \ref{a:mse-proof}).
 
@@
 
-The key question motivating the following definition is: when does taking expectation commute with computing a determinant for a square random matrix? 
+The key question motivating the following definition is: When does taking expectation commute with computing a determinant for a square random matrix?
 \begin{definition}\label{d:main}
-A random $d\times d$ matrix $\A$ is called \textbf{determinant
-  preserving} (d.p.), if
+A random $d\times d$ matrix $\A$ is called determinant
+  preserving (d.p.), if
 \begin{align*}
@@
 \end{definition}
-Note that from the definition of an adjugate matrix (see Section
-\ref{s:determinantal}) it immediately follows that if $\A$ is
+
+\noindent
+Note that from the definition of an adjugate matrix (see preliminaries
+in Section \ref{s:determinantal}) it immediately follows that if $\A$ is
 determinant preserving then adjugate commutes with expectation for this matrix:
@@
 \end{align}
+The adjugate is useful in our analysis because it connects the
+determinant and the inverse via the formula
+$\adj(\A)=\det(\A)\A^{-1}$, which holds for any invertible $\A$.
 We next give a few simple examples to provide some intuition. First, note
@@
 \end{example}
+
+\noindent
 In fact, it can be shown that all random matrices with independent entries
@@
 \end{example}
+
+\noindent
 To construct more complex examples, we show that determinant preserving random matrices are
-closed under addition and multiplication. The proof of this result is an extension of an
-argument given by \cite{determinantal-averaging} (Lemma~7) for
-computing the expected determinant of the sum of rank-1 random matrices.
-\begin{lemma}\label{t:ring}
-  If $\A,\B$ are independent and d.p.~then
-  $\A+\B$ and $\A\B$ are also determinant preserving.
+closed under addition and multiplication. The proof of this result is
+an extension of an existing argument, given by
+\cite{determinantal-averaging} in the proof of Lemma~7, for computing
+the expected determinant of the sum of rank-1 random matrices (proof in Appendix \ref{a:dp}). 
+\begin{lemma}[Closure properties]\label{t:ring}
+  If $\A$ and $\B$ are independent and determinant preserving, then:
+  \begin{enumerate}
+  \item $\A+\B$ is determinant preserving,
+  \item $\A\B$ is determinant preserving.
+    \end{enumerate}
 \end{lemma}
-\begin{proof}
- First, we show that $\A+\u\v^\top$ is d.p.~for any fixed
- $\u,\v\in\R^d$. Below, we use a standard identity for the rank one
- update of a determinant:
- $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$. It follows that
- for any $\Ic$ and $\Jc$ of the same size,
-  \begin{align*}
-\E\big[\!\det(\A_{\Ic,\Jc}\!+\u_{\Ic}\v_{\Jc}^\top)\big] &=
-    \E\big[\!\det(\A_{\Ic,\Jc}) +
-    \v_{\Jc}^\top\adj(\A_{\Ic,\Jc}) \u_{\Ic}\big]\\
-    &\overset{(*)}{=}\det\!\big(\E[\A_{\Ic,\Jc}]\big) +
-      \v_{\Jc}^\top\adj\!\big(\E[\A_{\Ic,\Jc}]\big) \u_{\Ic}\\
-    &=\det\!\big(\E[\A_{\Ic,\Jc} \!+ \u_{\Ic}\v_{\Jc}^\top]\big),
-  \end{align*}
-  where $(*)$ used \eqref{eq:adj}, i.e., the fact that for d.p.~matrices, adjugate commutes
-  with expectation. Crucially, through the definition of an adjugate
-  this step implicitly relies on the assumption that all the square
-  submatrices of $\A_{\Ic,\Jc}$ are also  determinant preserving.
-  Iterating this, we get that $\A+\Z$ is d.p.~for any fixed
-  $\Z$. We now show the same for $\A+\B$:
-  \begin{align*}
-\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\big]
-    &=
-      \E\Big[\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\mid\B\big]\Big]\\
-    &\overset{(*)}{=}\E\Big[\!\det\!\big(\E[\A_{\Ic,\Jc}]\!+\B_{\Ic,\Jc}\big)\Big]\\
-      &= \det\!\big(\E[\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc}]\big),
-  \end{align*}
-  where $(*)$  uses the fact that after conditioning on $\B$ we can
-  treat it as a fixed matrix. Next, we show that $\A\B$ is determinant preserving via the Cauchy-Binet formula:
-  \begin{align*}
-    \E\big[\!\det\!\big((\A\B)_{\Ic,\Jc}\big)\big]
-    &= \E\big[\!\det(\A_{\Ic,*}\B_{*,\Jc})\big]\\
-    &=\E\bigg[\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\A_{\Ic,S}\big)
-      \det\!\big(\B_{S,\Jc}\big)\bigg]\\
-&=\!\!\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\E[\A]_{\Ic,S}\big)
-                                                \det\!\big(\E[\B]_{S,\Jc}\big)\\
-    &=\det\!\big(\E[\A]_{\Ic,*}\, \E[\B]_{*,\Jc}\big)\\
-      &= \det\!\big(\E[\A\B]_{\Ic,\Jc}\big),
-  \end{align*}
-  where recall that $\A_{\Ic,*}$ denotes the submatrix of $\A$
-  consisting of its (entire) rows indexed by $\Ic$.
-  \end{proof}
-Finally, we introduce another important class of d.p.~matrices:
+% \begin{lemma}\label{t:ring}
+%   If $\A,\B$ are independent and d.p.~then
+%   $\A+\B$ and $\A\B$ are also determinant preserving.
+% \end{lemma}
+
+\noindent
+Next, we introduce another important class of d.p.~matrices:
 a sum of i.i.d.~rank-1 random matrices with the number of
@@
 extension of an expectation formula given by \cite{dpp-intermediate}
-for sampling from discrete distributions.
+for sampling from discrete distributions (proof in Appendix \ref{a:dp}).
 \begin{lemma}\label{l:poisson}
@@
 matrices whose rows  are sampled as an i.i.d.~sequence of joint pairs of
-random vectors, then $\A^\top\B$ is d.p., and in particular,
+random vectors, then $\A^\top\B$ is d.p., and so:
 \begin{align*}
@@
 
-To prove the above result, we will use the following
-lemma, many variants of which appeared in the literature
-(e.g., \cite{expected-generalized-variance}). We use the one given by
-\cite{correcting-bias}.
-\begin{lemma}[\cite{correcting-bias}]\label{l:cb}
-If the rows of random matrices $\A,\B\in\R^{k\times d}$
-  are sampled as an i.i.d.~sequence of $k\geq d$ pairs of joint random vectors, then
-\begin{align}
-  k^d\,\E \big[\det(\A^\top\B)\big]
-  &= \ktd\,\det\!\big(\E[\A^\top\B]\big).
-     \end{align}
-   \end{lemma}
-Here, we use the following standard shorthand: $\ktd =
-\frac{k!}{(k-d)!} = k\,(k-1)\dotsm(k-d+1)$. Note that the above result
-almost looks like we are claiming that the matrix $\A^\top\B$ is d.p.,
-but in fact it is not because $k^d\neq \ktd$. The difference
-in those factors is precisely what we are going to correct with the
-Poisson random variable. We now present the proof of Lemma
-\ref{l:poisson}.
-\begin{proof} (of Lemma \ref{l:poisson}) \
-Without loss of generality, it suffices to check Definition \ref{d:main} with both $\Ic$ and
-$\Jc$ equal $[d]$. We first expand the expectation by
-conditioning on the value of $K$ and letting $\gamma=\E[K]$:
-    \begin{align*}
-      \E\big[\!\det(\A^\top\B)\big]
-      &= \sum_{k=0}^\infty
-\E\big[\det(\A^\top\B)\mid K\!=\!k\big]\
-\Pr(K\!=\!k)\\
-      \text{(Lemma \ref{l:cb})}
-      \quad&=
-        \sum_{k=d}^\infty\frac{k! k^{-d}}{(k-d)!}\det\!\big(\E[\A^\top\B\mid
-        K\!=\!k]\big)
-        \frac{\gamma^k\ee^{-\gamma}}{k!}\\
-      &=\sum_{k=d}^\infty
-\Big(\frac\gamma k\Big)^d\det\!\big(\E[\A^\top\B\mid K\!=\!k]\big)
-        \frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}.
-     %  \\
-     %  &=\sum_{k=0}^\infty \det\!\big(\E[\A^\top\B]\big)\,\Pr(K\!=\!k)
-     % \ =\ \det\!\big(\E[\A^\top\B]\big).
-    \end{align*}
-    Note that $\frac\gamma k\,\E[\A^\top\B\mid K\!=\!k]=\E[\A^\top\B]$,
-    which is independent of $k$. Thus we can rewrite the above
-    expression as:
-    \begin{align*}
-\det\!\big(\E[\A^\top\B]\big)\sum_{k=d}^\infty\frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}
-      =
-      \det\!\big(\E[\A^\top\B]\big)\sum_{k=0}^\infty
-      \frac{\gamma^{k}\ee^{-\gamma}}{k!}=\det\!\big(\E[\A^\top\B]\big),
-    \end{align*}
-    which concludes the proof.
-\end{proof}
-Finally we use Lemma \ref{l:poisson} combined with Lemma
-\ref{t:ring} to show the expectation formula needed for obtaining the
-normalization constant of the under-determined surrogate
-design (proven by setting $\A=\B=\X\sim\mu^K$). Note that the below result is
-more general than the normalization constant requires, because it allows the matrices
-$\A$ and $\B$ to be different. In fact, we will use this more general
-statement later on in our analysis.
+\noindent
+Finally, we show the expectation formula needed for obtaining the
+normalization constant of the under-determined surrogate design, given
+in \eqref{eq:cases}.
+The below result is more general than the normalization constant
+requires, because it allows the matrices $\A$ and $\B$ to be different
+(the constant is obtained by setting $\A=\B=\X\sim\mu^K$).
+In fact, we use this more general statement to show Theorems
+\ref{t:mse} and~\ref{t:unbiased}. The proof uses
+Lemmas \ref{t:ring} and \ref{l:poisson} (see Appendix \ref{a:dp}).
 \begin{lemma}\label{l:normalization}
@@
 \end{lemma}
-\begin{proof}
-By Lemma \ref{l:poisson}, the matrix $\B^\top\A$ is determinant
-preserving. Applying Lemma \ref{t:ring} we conclude that
-$\I+\B^\top\A$ is also d.p., so
-\begin{align*}
-  \det\!\big(\I+\E[\B^\top\A]\big) = \E\big[\det(\I+\B^\top\A)\big] =\E\big[\det(\I+\A\B^\top)\big],
-\end{align*}
-where the second equality is known as Sylvester's Theorem.
-We now use the following standard determinantal formula.
-\begin{lemma}[\cite{dpp-ml}]
-  For any $k\times d$ matrices $\A,\B$ we have
-  $\det(\I+\A\B^\top)=\sum_{S\subseteq[k]}\det(\A_{S,*}\B_{S,*}^\top)$.
-\end{lemma}
-We rewrite the expectation of $\det(\I+\A\B^\top)$ by applying the
-lemma.  Letting $\gamma=\E[K]$, we obtain:
-\begin{align*}
-\E\big[\det(\I+\A\B^\top)\big]  &=\E\bigg[\sum_{S\subseteq [K]}\E\big[\det(\A_{S,*}\B_{S,*}^\top)\mid
-    K\big]\bigg]\\
-  &\overset{(*)}{=}\sum_{k=0}^\infty\frac{\gamma^k\ee^{-\gamma}}{k!} \sum_{i=0}^k{k\choose
-    i}\E\big[\det(\A\B^\top)\mid K=i\big]\\
-  &=\sum_{i=0}^\infty \E\big[\det(\A\B^\top)\mid K=i\big]\sum_{k\geq
-    i}^\infty{k\choose i}\frac{\gamma^k\ee^{-\gamma}}{k!}\\
-  &=\sum_{i=0}^\infty
-    \frac{\gamma^i\ee^{-\gamma}}{i!}\E\big[\det(\A\B^\top)\mid K=i\big]
-    \sum_{k\geq i}^\infty\frac{\gamma^{k-i}}{(k-i)!} \\
-  &= \E\big[\det(\A\B^\top)\big]\cdot\ee^\gamma,
-\end{align*}
-where $(*)$ follows from the exchangeability of the rows of $\A$ and
-$\B$, which implies that the distribution of $\A_{S,*}\B_{S,*}^\top$ is the
-same for all subsets $S$ of a fixed size $k$.
-\end{proof}
-
-\section{Expectation formulas for surrogate designs}
-\label{s:expectations}
-In this section we prove a number of expectation formulas for
-determinantal surrogate designs, which we then use to prove Theorems \ref{t:mse}
-and \ref{t:unbiased}. In the process, we derive closed form
-expressions for  $\E[\Xb^\dagger\Xb]$, i.e., the expectation of the
-orthogonal projection onto the subspace spanned by the the rows of
-$\Xb$, and for $\E[\tr((\Xb^\top\Xb)^\dagger)]$, the trace of the
-pseudo-inverse of the sample covariance matrix. To our knowledge,
-neither of these quantities admit closed form expressions for standard
-i.i.d.~random designs such as Gaussian with general covariance (except
-for the isotropic case).
-
-\subsection{Proof of Theorem \ref{t:mse}}
-\label{s:mse-proof}
-Let $y(\x)$ follow the homoscedastic noise model with variance $\sigma^2$
-(Assumption \ref{a:linear}). Recall that we have
-$\Xb\sim\S_\mu^n$ and $\yb_i=y(\xbb_i)=\xbb_i^\top\w^*+\xi_i$, where
-$\xi_i\sim\Nc(0,\sigma^2)$. A standard decomposition of
-the MSE of the Moore-Penrose estimator proceeds as follows:
-\begin{align*}
-  \MSE{\Xb^\dagger\ybb}
-  &= \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\\
-  &=\E\big[\|\Xb^\dagger\xib\|^2\big] +
-    \E\big[\|(\Xb^\dagger\Xb-\I)\w^*\|^2\big]\\
-  &=\sigma^2\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big] +
-    \w^{*\top}\E\big[\I-\Xb^\dagger\Xb\big]\w^*.
-\end{align*}
-Thus, our task is to find closed form expressions for the two
-expectations above. If $n\geq d$, then the latter goes away because
-when $\Xb$ has full column rank then $\I-\Xb^\dagger\Xb=\zero$. When
-$n<d$, this expectation is given in the following result. %can be obtained as a consequence of Lemma \ref{l:ridge-under}.
-\begin{lemma}\label{l:proj}
-If  $\Xb\sim S_\mu^n$ and $n<d$, then we have
-\[\E\big[\I-\Xb^\dagger\Xb\big] = (\gamma_n\Sigmab_\mu+\I)^{-1}.\]
-\end{lemma}
-The proof of Lemma \ref{l:proj} is deferred to Section
-\ref{s:unbiased-proof} because it follows as a corollary of a more
-general result (Lemma \ref{l:ridge-under}). We next derive the second
-expectation needed to compute the MSE. The under- and over-determined
-cases are proven separately, starting with the former.
-% Note that those
-% results require $\mu$ to satisfy general position (Assumption \ref{a:general-position}),
-% which implies that if $\X\sim\mu^k$ for $k\leq d$ then $\rank(\X)=k$.
-\begin{lemma}\label{l:sqinv-under}
-If  $\Xb\sim S_\mu^n$ for $n<d$, then we have
-\begin{align*}
-    \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
-    &={\gamma_n}\big(1- \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big).
-\end{align*}
-\end{lemma}
-\begin{proof}
-Let $\X\sim\mu^K$ for $K\sim\Poisson({\gamma_n})$. Note that if
-$\det(\X\X^\top)>0$ then using the fact that
-$\det(\A)\A^{-1}=\adj(\A)$ for any invertible matrix $\A$, we can write:
-  \begin{align*}
-    \det(\X\X^\top)\tr\big((\X^\top\X)^{\dagger}\big)
-    &= \det(\X\X^\top)\tr\big((\X\X^\top)^{-1}\big) \\
-    &= \tr(\adj(\X\X^\top)) \\[-1mm]
-    &= \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top),
-  \end{align*}
-  where $\X_{-i}$ is a shorthand for $\X_{[K]\backslash\{i\},*}$.
-Assumption \ref{a:general-position} ensures that
-$\Pr\big\{\det(\X\X^\top)>0\big\}=1$, which allows us to write:
-  \begin{align*}
-Z_\mu^n\cdot \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
-    &=\E\bigg[
-    \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top)\ \big|\ 
-    \det(\X\X^\top)>0\bigg]\cdot\overbrace{\Pr\big\{\det(\X\X^\top)>0\big\}}^{1}\\
-    &=\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\E\Big[
-      \sum_{i=1}^k\det(\X_{-i}\X_{-i}^\top)\ \big|\  K=k\Big]\\
-    &=\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\, k\
-      \E\big[\det(\X\X^\top)\mid K=k-1\big]\\
-    &=\gamma_n\sum_{k=0}^{d-1}\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}
-      \E\big[\det(\X\X^\top)\mid K=k\big]\\
-    &=\gamma_n\Big(\E\big[\det(\X\X^\top)\big]\ -\
-      \frac{\gamma_n^{d}\ee^{-\gamma_n}}{d!}\E\big[\det(\X)^2\mid K=d\big]
-      \Big) \\
-    &\overset{(*)}{=}\gamma_n\big(\ee^{-\gamma_n}\det(\I +\gamma_n\Sigmab_\mu) -
-      \ee^{-\gamma_n}\det(\gamma_n\Sigmab_\mu)\big),
-  \end{align*}
-  where $(*)$ uses Lemma \ref{l:normalization} for the first term and
-  Lemma \ref{l:cb} for the second term. We obtain the desired result by
-  dividing both sides by
-  $Z_\mu^n=\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)$. 
-\end{proof}
-In the over-determined regime, a more general matrix expectation
-formula can be shown (omitting the trace). The following result is
-related to an expectation formula derived by
-\cite{correcting-bias-journal}, however they use a slightly
-different determinantal design so the results are incomparable.
-\begin{lemma}\label{l:sqinv-over}
-If $\Xb\sim S_\mu^n$ and $n>d$, then we
-have
-\begin{align*}
-  \E\big[ (\Xb^\top\Xb)^{\dagger}\big] =
-  \Sigmab_\mu^{-1}\cdot \frac{1-\ee^{-\gamma_n}}{\gamma_n}.
-\end{align*}
-\end{lemma}
-\begin{proof}
-Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$. Assumption
-\ref{a:general-position} implies that for $K\neq d-1$ we have
-\begin{align}
-  \det(\X^\top\X)(\X^\top\X)^\dagger=\adj(\X^\top\X),\label{eq:adj-over}
-  \end{align}
-however when $k=d-1$ then \eqref{eq:adj-over} does not hold because
-$\det(\X^\top\X)=0$ while $\adj(\X^\top\X)$ may be non-zero. It 
-follows that: 
-  \begin{align*}
-Z_\mu^n\cdot
-    \E\big[ (\Xb^\top\Xb)^{\dagger}\big]
-    &=\E\big[\det(\X^\top\X)(\X^\top\X)^\dagger\big]\\
-    &=\E\big[\adj(\X^\top\X)\big]-
-\frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)!}
-      \E\big[\adj(\X^\top\X)\mid K=d-1\big]\\
-    &\overset{(*)}{=}\adj\!\big(\E[\X^\top\X]\big) -
-      \frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)^{d-1}}
-      \adj\!\big(\E[\X^\top\X\mid K=d-1]\big)\\
-    &=\adj(\gamma_n\Sigmab_\mu) - \ee^{-\gamma_n}\adj(\gamma_n\Sigmab_\mu)\\
-    &=\det(\gamma_n\Sigmab_\mu)\,(\gamma_n\Sigmab_\mu)^{-1}(1-\ee^{-\gamma_n})\\
-    &=\det(\gamma_n\Sigmab_\mu)\,\Sigmab_\mu^{-1}\cdot\frac{1-\ee^{-\gamma_n}}{\gamma_n},
-  \end{align*}
-  where the first term in $(*)$ follows from Lemma
-  \ref{l:normalization} and \eqref{eq:adj}, whereas the second term comes
-  from Lemma 2.3 of \cite{correcting-bias-journal}.
-Dividing both sides by $Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ completes the proof.
-\end{proof}
-Finally, applying the closed form expressions from Lemmas
-\ref{l:proj}, \ref{l:sqinv-under} and \ref{l:sqinv-over}, we derive
-the formula for the MSE and prove Theorem \ref{t:mse}.
-\begin{proof} (of Theorem \ref{t:mse}) \
-  First, assume that $n<d$, in which case we have
-  $\gamma_n=\frac1{\lambda_n}$ and moreover
-  \begin{align*}
-    n &= \tr\big(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1}\big)\\
-      &=\tr\big((\Sigmab_\mu+\lambda_n\I-\lambda_n\I)(\Sigmab_\mu+\lambda_n\I)^{-1}\big)\\
-    &=d - \lambda_n\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big),    
-    %\frac{\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)}{d-n},
-  \end{align*}
-so we can write $\lambda_n$ as $(d-n)/\tr((\Sigmab_\mu+\lambda_n\I)^{-1})$.
-  From this and Lemmas \ref{l:proj} and \ref{l:sqinv-under}, we
-obtain the desired expression, where recall
-  that $\alpha_n = \det\!\big(\Sigmab_\mu (\Sigmab_\mu+\frac1{\gamma_n})^{-1}\big)$:
-  \begin{align*}
-    \MSE{\Xb^\dagger\ybb} &= \sigma^2\,\gamma_n(1-\alpha_n) +
-    \tfrac1{\gamma_n} \,\w^{*\top}(\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\w^*
-    \\
-    &\overset{(a)}{=}\sigma^2\,\frac{1-\alpha_n}{\lambda_n} +
-    \lambda_n\,\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*\\
-    &\overset{(b)}{=}\sigma^2\tr\big((\Sigmab_\mu+\lambda_n)^{-1}\big)\frac{1-\alpha_n}{d-n}
-      +
-      (d-n)\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
-      {\tr\big((\Sigmab_\mu+\lambda_n)^{-1}\big)}.
-  \end{align*}
-  While the expression given after $(a)$ is simpler than the one
-after $(b)$, the latter better illustrates how the MSE depends on
-the sample size $n$ and the dimension $d$.
-  Now, assume that $n>d$. In this case, we have $\gamma_n=n-d$ and apply Lemma
-  \ref{l:sqinv-over}:
-  \begin{align*}
-    \MSE{\Xb^\dagger\ybb}
-    &= \sigma^2\,\tr(\Sigmab_\mu^{-1})\,
-\frac{1-\ee^{-\gamma_n}}{\gamma_n}
-=\sigma^2\,\tr(\Sigmab_\mu^{-1})
-\,\frac{1-\beta_n}{n-d}.
-  \end{align*}
-The case of $n=d$ was shown by \cite{correcting-bias-journal}
-(Theorem~2.12). This concludes the proof. 
-\end{proof}
-\subsection{Proof of Theorem \ref{t:unbiased}}
-\label{s:unbiased-proof}
-Recall that our goal is to compute the expected value of
-$\Xb^\dagger\ybb$ under the surrogate design $S_\mu^n$. Similarly as for Theorem
-\ref{t:mse}, the case of $n=d$ was shown by
-\cite{correcting-bias-journal} (Theorem 2.10). We break the rest down into the
-under-determined case $(n<d)$ and the over-determined case ($n>d$),
-starting with the former. Recall that we do \emph{not} require any
-modeling assumptions on the responses.
-\begin{lemma}\label{l:ridge-under}
-If $\Xb\sim S_\mu^n$ and $n<d$, then for any $y(\cdot)$
-such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
-denoting $\yb_i$ as $y(\xbb_i)$, we have
-\begin{align*}
-  \E\big[\Xb^\dagger \ybb\big]
-  &=
-    \big(\Sigmab_\mu+\tfrac1{\gamma_n}\I\big)^{-1}\E_{\mu,y}[y(\x)\,\x].
-\end{align*}
-\end{lemma}
-\begin{proof}
-   Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
-   $y(\x_i)$ as $y_i$.
-  Note that when $\det(\X\X^\top)>0$, then
-  the $j$th entry of $\X^\dagger\y$ equals
-  $\f_j^\top(\X\X^\top)^{-1}\y$, where $\f_j$ is the $j$th
-  column of $\X$, so:
-\begin{align*}
-  \det(\X\X^\top)\,(\X^\dagger\y)_j
-  &= \det(\X\X^\top)\, \f_j^\top(\X\X^\top)^{-1}\y \\
-  &=
-  \det(\X\X^\top+\y\f_j^\top) - \det(\X\X^\top).
-\end{align*}
-If $\det(\X\X^\top)=0$, then also
-$\det(\X\X^\top+\y\f_j^\top)=0$, so we can write:
-\begin{align*}
-Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
-  &=  \E\big[\det(\X\X^\top)(\X^\dagger\y)_j\big] \\
-  &= \E\big[\det(\X\X^\top+\y\f_j^\top)-\det(\X\X^\top)\big]  \\
-  &=\E\big[\det\!\big([\X,\y][\X,\f_j]^\top\big)\big] - \E\big[\det(\X\X^\top)\big]\\
-  &\overset{(1)}{=}\ee^{-\gamma_n}\det\!\bigg(\I +
-    \gamma_n\,\E_{\mu,y}\bigg[\begin{pmatrix}\x\x^\top&
-      \!\!\x\, y(\x)\\ x_j\,\x^\top&\!\!
-      x_j\,y(\x)\end{pmatrix}\bigg]\bigg)
-                          -\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)\\
-  &\overset{(2)}{=}\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)\Big(\E_{\mu,y}\big[\gamma_n
-    x_j\,y(\x)\big] - \E_{\mu}\big[\gamma_n
-    x_j\,\x^\top\big](\I+\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[\gamma_n\x\,
-    y(\x)\big]\Big),
-\end{align*}
-where $(1)$ uses Lemma \ref{l:normalization} twice, with the first
-application involving two different matrices $\A=[\X,\y]$ and
-$\B=[\X,\f_j]$, whereas $(2)$ is a standard determinantal identity
-(see Fact 2.14.2 in \cite{matrix-mathematics}).
-  Dividing both sides by $Z_\mu^n$ and letting $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$, we obtain that:
-  \begin{align*}
-    \E\big[\Xb^\dagger\ybb\big]
-    &= \gamma_n\v_{\mu,y} - \gamma_n^2\Sigmab_\mu(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y}\\
-&=\gamma_n\big(\I - \gamma_n\Sigmab_\mu (\I+\gamma_n\Sigmab_\mu)^{-1}\big)\v_{\mu,y}
-=\gamma_n(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y},
-  \end{align*}
-  which completes the proof.
-\end{proof}
-We return to Lemma \ref{l:proj} from Section \ref{s:mse-proof}, regarding the expected orthogonal
-projection onto the complement of the row-span of $\Xb$, i.e.,
-$\E[\I-\Xb^\dagger\Xb]$, which follows as a corollary of Lemma~\ref{l:ridge-under}.
-\begin{proof} (of Lemma \ref{l:proj}) \
-  We let $y(\x)=x_j$ for each $j\in[d]$ and apply Lemma
-  \ref{l:ridge-under} $d$ times, obtaining:
-  \begin{align*}
-    \I - \E\big[\Xb^\dagger\Xb] = \I -
-    (\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\Sigmab_\mu,
-  \end{align*}
-  from which the result follows by simple algebraic manipulation.
-\end{proof}
-We move on to the over-determined case, where the ridge regularization
-of adding the identity to $\Sigmab_\mu$ vanishes. Recall that we
-assume throughout the paper that $\Sigmab_\mu$ is invertible.
-\begin{lemma}\label{l:ridge-over}
-  If $\Xb\sim S_\mu^n$ and $n>d$, then for any real-valued random function $y(\cdot)$
-  such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
-denoting $\yb_i$ as $y(\xbb_i)$, we have
- \begin{align*}
-  \E\big[\Xb^\dagger \ybb\big]
-  &=\Sigmab_\mu^{-1}\E_{\mu,y}\big[y(\x)\,\x\big].
-\end{align*}
-\end{lemma}
-\begin{proof}
-   Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
-   $y_i=y(\x_i)$. Similarly as in the proof of
-   Lemma~\ref{l:ridge-under}, we note that when $\det(\X^\top\X)>0$,
-   then 
-  the $j$th entry of $\X^\dagger\y$ equals
-  $\e_j^\top(\X^\top\X)^{-1}\X^\top\y$, where $\e_j$ is the $j$th
-standard basis vector, so:
-\begin{align*}
-  \det(\X^\top\X)\,(\X^\dagger\y)_j =
-  \det(\X^\top\X)\, \e_j^\top(\X^\top\X)^{-1}\X^\top\y =
-  \det(\X^\top\X+\X^\top\y\e_j^\top) - \det(\X^\top\X).
-\end{align*}
-If $\det(\X^\top\X)=0$, then also
-$\det(\X^\top\X+\X^\top\y\e_j^\top)=0$. We proceed to compute the
-expectation:
-\begin{align*}
-Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
-  &=  \E\big[\det(\X^\top\X)(\X^\dagger\y)_j\big] \\
-  &= \E\big[\det(\X^\top\X+\X^\top\y\e_j^\top)-\det(\X^\top\X)\big]  \\
-  &=\E\big[\det\!\big(\X^\top(\X+\y\e_j^\top)\big)\big] - \E\big[\det(\X^\top\X)\big]\\
-  &\overset{(*)}{=}\det\!\Big(
-    \gamma_n\,\E_{\mu,y}\big[\x(\x+ y(\x)\e_j)^\top\big]\Big)
-    -\det(\gamma_n\Sigmab_\mu)\\
-  &=\det\!\big(\gamma_n\Sigmab_\mu + \gamma_n\E_{\mu,y}[\x\,y(\x)]\e_j^\top\big)
-    -\det(\gamma_n\Sigmab_\mu)\\
-  &=\det(\gamma_n\Sigmab_\mu)\cdot
-    \gamma_n\e_j^\top(\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[y(\x)\,\x\big],
-\end{align*}
-where $(*)$ uses Lemma \ref{l:poisson} twice (the first time, with
-$\A=\X$ and $\B=\X+\y\e_j^\top$). Dividing both sides by
-$Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ concludes the proof. 
-\end{proof}
-
-We combine Lemmas \ref{l:ridge-under} and \ref{l:ridge-over} to obtain
-the proof of Theorem \ref{t:unbiased}.
-\begin{proof} (of Theorem \ref{t:unbiased}) \
-The case of $n=d$ follows directly from the result of
-\cite{correcting-bias} (Theorem~2.10). 
-Assume that $n<d$. Then we have
-$\gamma_n=\frac1{\lambda_n}$, so the result follows
-from Lemma \ref{l:ridge-under}.
-% , noting that unlike in the lemma,
-% Theorem \ref{t:unbiased} allows for the response model $y(\cdot)$ to be randomized:
-% \begin{align*}
-%   \E[\Xb^\dagger\ybb] = \E\big[\E[\Xb^\dagger\ybb\mid\ybb]\big] = 
-% % \w_{\lambda_n}^* = \argmin_\w\E_\mu\big[(\x^\top\w-y(\x))^2\big] +
-% %   \lambda_n\|\w\|^2 =
-% %   \big(\Sigmab_\mu+\lambda_n\I\big)^{-1}
-% %   \E_{\mu}\big[y(\x)\,\x\big].
-% \end{align*}
-If $n>d$, then the result follows from Lemma \ref{l:ridge-over}.
-\end{proof}
-
-\section{Asymptotic consistency of surrogate Gaussian designs}
-\label{s:asymptotic}
-In this section we compare the MSE expressions obtained for the
-surrogate design with the true MSE of the standard i.i.d.~design for
-the case where $\mu$ is a centered Gaussian distribution
-$\Nc(\zero,\Sigmab)$. First, we theoretically compare the analytical
-expressions available in certain cases for the i.i.d.~design with their
-surrogate equivalents, showing that the surrogate expressions are
-asymptotically consistent (Theorem \ref{t:asymptotic}). In fact, we
-show even more: that the surrogate expressions provide a
-multiplicative approximation for the true MSE with the error decaying
-at the rate of $O(1/d)$. However, this result
-does not cover the under-determined setting for a non-isotropic
-Gaussian, because to the best of our knowledge, analytical expressions
-for the i.i.d.~design are not available there. To address this case,
-which is of primary interest to us, we perform thorough empirical
-analysis showing that our surrogate expressions are asymptotically
-consistent, with the error decaying as $O(1/d)$, in all of the
-non-isotropic Gaussian test-cases that we evaluated. Based on 
-this, we formulate Conjectures \ref{c:wishart} and \ref{c:projection},
-regarding the asymptotic behavior of the pseudo-inverse of the
-singular Wishart distribution and of a random Gaussian projection, that
-are of independent interest to multivariate statistics.
-
-\subsection{Proof of Theorem \ref{t:asymptotic}}
-Once again, we break down the proof into under- and
-over-determined cases, starting with the former. Note that in this
-case we require that the covariance be equal to identity.
-\begin{lemma}\label{l:asymptotic-under}
-Let $\rho=n/d$, $\X\sim \Nc(\zero,\I)^n$ and $y_i=y(\x_i)$ under Assumption
-  \ref{a:linear}. If $d>n+1$ then
-  \begin{align*}
-    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\I,
-\w^*,\sigma^2,n)}{\Mc(\I,\w^*,\sigma^2,n)}\ \leq\
-\frac1d\cdot \frac{1}{1-\rho-\frac1d} + 3\rho^d.
-  \end{align*}
-\end{lemma}
-
-\begin{proof} 
-  We first recall the standard decomposition of $\MSE{\X^\dagger \y}$:
-  \begin{align*}
-    \MSE{\X^\dagger \y}
-    &=\sigma^2 \E\big[\tr\big((\X^\top \X)^\dagger \big)\big]+
-      \w^{*\top}\!\big( \I - \E[\X^\dagger \X]\big) \w^*. 
-  \end{align*}
-  Since the rows of $\X$ are standard normal random variables,
-  $\X\X^\top$ is an $n\times n$ Wishart random matrix with $d>n+1$ degrees
-  of freedom. Using the formula for the mean of the Inverse-Wishart
-  distribution, it follows that
-  \begin{align*}
-\E\big[\tr((\X^\top\X)^\dagger)\big] = \E\big[\tr((\X\X^\top)^{-1})\big]= \frac{n}{d - n - 1}.
-  \end{align*}
-  % and therefore
-  % \begin{align*}
-  %   \MSE{\X^\dagger \y}
-  %   &= \|\w\|_2^2 - \w^\top \E[\X^\dagger \X] \w
-  %   + \frac{\sigma^2 n}{d - n - 1}
-  % \end{align*}
-  % Let $\X = \U \S \V^\top$ be its full singular value decomposition. Notice
-  % \begin{align*}
-  %   \X^\dagger \X
-  %   &= \V \S^\dagger \U^\top \U \S \V^\top \\
-  %   &= \V \S^\dagger \S \V^\top \\
-  %   &= \V \begin{bmatrix}
-  %     \I_n & 0 \\
-  %     0 & 0
-  %   \end{bmatrix} \V^\top \\
-  %   &= \V_n \V_n^\top \\
-  %   \w^\top \E[\X^\dagger \X] \w
-  %   &= \E[\|\V_n^\top \w\|_2^2]
-  % \end{align*}
-  % where $\V_n$ denotes the $d \times n$ matrix with columns equal to the
-  % $n$ eigenvectors of $\X^\top \X$ with non-zero eigenvalue.
-  % But since $\Sigmab = \I$, by rotation invariance
-  % $\V_n^\top$ is a change of basis onto a uniformly random
-Note that $\X^\dagger\X$ is a uniformly random projection matrix so by
-rotational symmetry it follows that
-\begin{align*}
-  \w^{*\top}\E\big[\X^\dagger\X]\w^* =
-  \E\big[\|\X^\dagger\X\w^*\|^2\big] = \frac nd \,\|\w^*\|^2.
-\end{align*}
-Putting this together we obtain that
-  \begin{align*}
-    \MSE{\X^\dagger \y}
-    &=\frac{\sigma^2 n}{d - n - 1}+ \|\w^*\|^2\,\frac{d - n}{d}.
-  \end{align*}
-  On the other hand, the surrogate MSE expression can be derived by
-observing that for $\Sigmab=\I$ we have
-$\tr((\Sigmab+\lambda_n\I)^{-1})=d/(1+\lambda_n)=n$ (see definition of
-$\lambda_n$ in Theorem \ref{t:mse}): 
-  \begin{align*}
-    \Mc(\I, \w^*,\sigma^2,n) =
-    \sigma^2n\cdot\frac{1-\alpha_n}{d-n} + \|\w^*\|^2\,\frac{d-n}{d}.
-  \end{align*}
-  Note that the second term is the same in both cases, even though
-  this may not be true for non-isotropic Gaussians. We now compute the
-  normalized difference between the 
-  expressions, 
-  \begin{align*}
- \frac{\MSE{\X^\dagger\y}-\Mc(\I, \w^*,\sigma^2,n)}{\Mc(\I,
-    \w^*,\sigma^2,n)}
-    &= \frac{\sigma^2n\cdot(\frac1{d-n-1} -
-    \frac{1-\alpha_n}{d-n})}{\sigma^2n\cdot\frac{1-\alpha_n}{d-n} +
-    \|\w^*\|^2\,\frac{d-n}{d}}\\
-&\leq \frac {d-n}{1-\alpha_n}
-\Big(\frac1{d-n-1}-\frac{1-\alpha_n}{d-n}\Big)\\
-    &=\frac{d-n}{d-n-1}\,\frac{1}{1-\alpha_n}- 1\\
-    &=\frac1{d-n-1} + \frac{\alpha_n}{1-\alpha_n}\Big(1+\frac1{d-n-1}\Big).
-  \end{align*}
-Let $\rho=n/d$. Then $\frac1{d-n-1} = \frac1d\cdot
-\frac1{1-\rho-\frac1d}$ and moreover
-$\alpha_n = (\frac1{1+\lambda_n})^d=(\frac
-nd)^d=\rho^d$. From the assumption that $d>n+1$, we
-conclude that $\alpha_n \leq (\frac{d-2}{d})^d\leq \ee^{-2}$
-so that $\frac{\alpha_n}{1-\alpha_n}(1+\frac1{d-n-1})\leq
-  \frac{2 \alpha_n}{1-\ee^{-2}}\leq 3\rho^d$. This shows the right-hand-side inequality of the
-  theorem. That fact that $\MSE{\X^\dagger\y}\geq\Mc(\I,
-  \w^*,\sigma^2,n)$ follows easily.
-\end{proof}
-\begin{lemma}\label{l:asymptotic-over}
-Let $\rho=n/d$, $\X\sim \Nc(\zero,\Sigmab)^n$ and $y_i=y(\x_i)$ under Assumption
-  \ref{a:linear}. If $n>d+1$ then
-  \begin{align*}
-    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab,
-\w^*,\sigma^2,n)}{\Mc(\Sigmab,\w^*,\sigma^2,n)}\ \leq\
-\frac1d\cdot \frac{1}{\rho-1-\frac1d} + 3(\ee^{1-\rho})^d.
-  \end{align*}
-\end{lemma}
-\begin{proof}
-The MSE for the over-determined Gaussian design can be derived by
-using the formula for the mean of the Inverse-Wishart distribution:
-\begin{align*}
-  \MSE{\X^\dagger\y}=\sigma^2\tr\big(\E[(\X^\top\X)^{-1}]\big) =
-  \frac{\sigma^2\tr(\Sigmab^{-1}) }{n-d-1}.
-\end{align*}
-To compute the normalized difference we follow similar derivations as
-in the proof of Lemma \ref{l:asymptotic-under}:
-\begin{align*}
- \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)}{\Mc(\Sigmab,
-    \w^*,\sigma^2,n)}
-    &=  \frac {n-d}{1-\beta_n}
-      \Big(\frac1{n-d-1}-\frac{1-\beta_n}{n-d}\Big)\\
-  &\leq\frac1d\cdot\frac1{\rho-1-\frac1d} + \frac{2\beta_n}{1-\beta_n}.
-\end{align*}
-Recall that $\beta_n = \ee^{d-n} = (\ee^{1-\rho})^d$ and for $n-d\geq
-2$ we have $\frac2{1-\beta_n}\leq 3$. The desired inequalities follow immediately.
-\end{proof}
-Theorem \ref{t:asymptotic} now follows as a consequence of
-Lemmas \ref{l:asymptotic-under} and \ref{l:asymptotic-over}.
-\begin{proof} (of Theorem \ref{t:asymptotic}) \
-  Since $\frac1d\leq \frac13|1-\rho|$ and $\ee^{-|1-\rho|d}\leq
-  \frac1{2d|1-\rho|}$, it follows that
-  \begin{align*}
-    \frac1d\cdot \frac{1}{|1-\rho|-\frac1d} + 3(\ee^{-|1-\rho|})^d
-    &\leq \frac1d\cdot \frac{1}{|1-\rho|-\frac13|1-\rho|} +
-      \frac1d\cdot\frac{3}{2|1-\rho|}=\frac {c_\rho}{d}.
-  \end{align*}
-  The case of $n>d+1$ now follows from Lemma
-  \ref{l:asymptotic-over}. Also, since $\rho\leq \ee^{\rho-1}$, for
-$n<d-1$ we have $3\rho^d\leq 3(\ee^{-|1-\rho|})^{d}$, so the same bound follows
-from Lemma \ref{l:asymptotic-under}.
-\end{proof}
-
-\input{continuous-empirical.tex}
-
-\section{Conclusions}
+
+\section{Conclusions and open problems}
 \label{s:conclusions}
-\begin{wrapfigure}{r}{0.45\textwidth}
- \vspace{-.8cm}
- \includegraphics[width=.45\textwidth]{figs/descent-model-log}
- \vspace{-6mm}
-  \caption{
-Surrogate MSE as a function of $d/n$, with $n$
-fixed to $100$ and varying $d$.}
-  \label{f:model}
-  \vspace{-3mm}
-\end{wrapfigure}
 
@@
 
+\begin{wrapfigure}{r}{\ifisarxiv 0.5\else 0.47\fi\textwidth}
+ \ifisarxiv\else \vspace{-.3cm}\fi
+  \centering
+ \includegraphics[width=.46\textwidth]{figs/descent-model-log}
+\ifisarxiv\else \vspace{-2mm}\fi
+  \caption{
+Surrogate MSE as a function of $d/n$, with $n$
+fixed to $100$ and varying $d$.}
+  \label{f:model}
+\ifisarxiv\else  \vspace{-.5cm}\fi
+\end{wrapfigure}
+
 An important technical issue is that, in this work, we focus on the classical \emph{well-specified}
@@
 underlying linear model.
-Recent work \cite{HMRT19_TR} has compared how
+Recent work \citep{HMRT19_TR} has compared how
 varying the feature dimension affects the (asymptotic) generalization
@@
 point of comparison, in Figure~\ref{f:model} we plot the MSE
-expressions of Theorem \ref{t:mse} for our well-specified setting when varying the
+expressions of Theorem~\ref{t:mse} for our well-specified setting when varying the
 feature dimension~$d$. The model is chosen just like in
@@
 scaled so that $\tr(\Sigmab_\mu^{-1})=d$. We also use $\sigma^2=1$ and
-$\w^*=\frac1{\sqrt{d}}\one_d$. Qualitatively, our plots follow the
+$\w^*=\frac1{\sqrt{d}}\one$. Qualitatively, our plots follow the
 trends outlined by \cite{HMRT19_TR} for the isotropic case (i.e., $\kappa=1$),
@@
 the complexity of the prediction model decreases, so does the complexity of the
-true response model. Nevertheless, even in the well-specified setting
-we achieve non-trivial generalization in the under-determined regime,
+true response model. Nevertheless,
+we observe generalization in the under-determined regime,
 as seen by the fact that the MSE curve goes below the error of the
-null estimator, $\mathrm{MSE}[\zero]=\|\w^*\|^2=1$.
+null estimator, $\mathrm{MSE}[\zero]=1$.
 
 Our work opens up a number of new directions for future research.
-Most obviously, this includes extending our surrogate analysis to the misspecified linear regression discussed above. 
-In addition, in Section \ref{s:asymptotic}, we provided two conjectures, the goal of which is to bound the difference between the results obtained for our surrogate designs and the true values corresponding to the i.i.d.~design, in the case of multivariate Gaussians.
-We believe that addressing these conjectures will have a broader impact on our understanding of non-isotropic Gaussian designs. 
-Finally, it remains open whether the analysis we provided for the mean squared error can be reproduced in the context of mean squared \emph{prediction} error, which is relevant in many machine learning tasks.
-
-%\noingent
+This includes extending our surrogate analysis to the
+misspecified linear regression discussed above.
+Also, it remains open whether the analysis we provided for the mean
+squared error can be reproduced in the context of mean squared
+\emph{prediction} error, which is relevant in many machine learning
+tasks. Finally, while Theorem \ref{t:asymptotic} states that our surrogate
+expressions for the MSE under certain Gaussian designs are
+asymptotically consistent with the multiplicative error rate of
+$O(1/d)$, we believe that this fact extends to
+the setting not covered by the theorem: under-determined regime (i.e.,
+$n<d$) with a non-isotropic Gaussian distribution (i.e., $\Sigmab\neq
+\I$). We break down our analysis into verifying two
+conjectures which are of independent interest to multivariate
+Gaussian analysis. The first conjecture addresses the first term in
+the MSE derivation \eqref{eq:mse-derivation} and postulates an asymptotically
+consistent formula for the expected Moore-Penrose inverse
+of the pseudo-Wishart distribution. Matrix
+$\W\sim\Pc\Wc(\Sigmab,n)$ is distributed according to the pseudo-Wishart
+distribution with $n<d$ degrees of freedom \citep{srivastava1979introduction} if it can be written as
+$\W=\X^\top\X$, where $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Sigmab)$
+is the matrix-variate Normal \citep{gupta2018matrix}, i.e., an
+i.i.d.~Gaussian design.
+% we call it a pseudo-Wishart 
+% and write $\W \sim \Pc\Wc(\Sigmab, n)$
+% if $n<d$ but still
+\ifisarxiv\pagebreak\fi
+\begin{conjecture}[Moore-Penrose inverse of pseudo-Wishart]\label{c:wishart}
+  Fix $n/d<1$ and let $\W\sim\Pc\Wc(\Sigmab,n )$, where
+$\Sigmab$ is $d\times d$ positive definite with condition number bounded by a constant.
+Then:
+\begin{align}
+\bigg|\frac{\E\big[\tr(\W^\dagger)\big]}{\Vc(\Sigmab,n)} -1\bigg|=
+  O(1/d)\qquad\text{for}\quad\Vc(\Sigmab,n)=\frac{1-\alpha_n}{\lambda_n},\label{eq:wishart}
+\end{align}
+where $\lambda_n\geq 0$ satisfies $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$ and
+$\alpha_n=\det(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$.
+\end{conjecture}
+
+\noindent
+Our second conjecture involves the projection onto the
+orthogonal complement of a Gaussian sample $\X$, i.e., the matrix
+$\I-\X^\dagger\X$, and addresses
+the second term in the MSE derivation \eqref{eq:mse-derivation}.
+\begin{conjecture}[Gaussian orthogonal projection]\label{c:projection}
+  Fix $n/d<1$ and let $\X\sim \Nc_{n,d}(\zero,\I_n \otimes \Sigmab)$, where
+$\Sigmab$ is $d\times d$ positive definite with condition number
+bounded by a constant. Then:
+\begin{align}
+\sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top
+  \Bc(\Sigmab,n)\w} - 1\bigg| = O(1/d)\qquad\text{for}\quad
+  \Bc(\Sigmab,n) =
+  \lambda_n(\Sigmab+\lambda_n\I)^{-1}.\label{eq:projection}
+\end{align}
+\end{conjecture}
+% Note that the surrogate expression for the mean squared
+% error can be written as:
+% \[
+% \Mc(\Sigmab_\mu,\w^*,\sigma^2,n)=\sigma^2 \Vc(\Sigmab_\mu,n)
+% + \w^{*\top}\Bc(\Sigmab_\mu,n)\w^*.
+% \]
+% So, if the conjectures are true,
+% this would immediately imply that the asymptotic consistency claim
+% given in Theorem \ref{t:asymptotic} for
+% the surrogate Gaussian MSE extends to
+% the under-determined setting with arbitrary covariance $\Sigmab$.
+
+\noindent
+Recall that $\lambda_n=\frac {d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})}$,
+so our surrogate MSE is recovered as
+$\sigma^2\Vc(\Sigmab,n)+\w^{*\top}\Bc(\Sigmab,n)\w^*$. 
+% The expressions $\Vc(\Sigmab,n)$ and $\Bc(\Sigmab,n)$ and recover
+% surrogate MSE from Theorem \ref{t:mse} since $\lambda_n=\frac{\tr((\Sigmab+\lambda_n)^{-1})}{d-n}$.
+Conjectures \ref{c:wishart} and \ref{c:projection} provide new insights into classical matrix-variate
+distributions with extensive literature dedicated to them \citep[see,
+e.g.,][]{chikuse1990matrix,cook2011}. We discuss 
+this further in Appendix \ref{sec:asymp-conj-details},
+where we also provide detailed empirical evidence supporting these claims.
+
+% In Appendix \ref{sec:asymp-conj-details} we provide further discussion
+% and detailed empirical evidence for these conejctures.
+% We believe that addressing these conjecture will have a broader
+% impact on our understanding of non-isotropic Gaussian designs.
+
+% In addition, in Section \ref{s:asymptotic}, we provided two conjectures, the goal of which is to bound the difference between the results obtained for our surrogate designs and the true values corresponding to the i.i.d.~design, in the case of multivariate Gaussians.
+
 \paragraph{Acknowledgements.}
-We would like to acknowledge ARO, DARPA, NSF, and ONR for providing partial support of this work.
-
-%\bibliographystyle{plain}
-\bibliographystyle{alpha}
-  \bibliography{pap}
-
-  \iffalse
+We would like to acknowledge ARO, DARPA, NSF, ONR, and GFSD for providing
+partial support of this work.
+
+\ifisarxiv\bibliographystyle{plainnat}\fi
+\bibliography{../pap}
 
@@
 
-  \section{Literature review of double descent}
-  \begin{itemize}
-\item \cite{BHMM19} Introduction of double descent. Empirical evidence and
-experiments for: random fourier features (interpreted as a class of
-two-layer neural network), decision trees and ensemble
-methods. Uses the term \emph{inductive bias} to describe choosing a
-minimum norm solution.
-
-\item\cite{BMM18_TR}
-Empirical evaluation of the performance of Gaussian and Laplacian
-kernels for classification in the interpolation regime. Draws a distinction between
-\emph{implicit regularization} (e.g., early stopping) and
-\emph{inductive bias} (e.g., minimum norm solutions).
-
-\item\cite{BRT18_TR}
-An error bound for a special interpolating kernel
-method, showing that you can generalize despite zero training error.
-
-\item\cite{belkin2019two} Two theoretical models of double
-  descent where features are chosen as a uniformly random subset of a
-  larger design matrix. Case 1: design matrix is a random i.i.d
-  isotropic Gaussian. Case 2: discrete fourier transform matrix. They
-  obtain exact expressions in those special cases.
-
-\item \cite{BLLT19_TR} Non-asymptotic upper- and lower-bounds for
-  least squares in infinite dimensional Hilbert spaces. Input data is
-  Gaussian with arbitrary covariance. Two notions of
- effective rank, one of them is an extension of the stable rank.
-
-\item \cite{BHM18_TR} Analysis of nearest neighbor in the
-   interpolating regime.
-
- \item \cite{BFF19_TR} Analysis of least squares
-   under the Predictive Normalized Maximum Likelihood (pNML) model,
-   showing the capacity for generalization in the over-parameterized
-   regime.
- \item \cite{COB18_TR} Analysis of lazy training, i.e., when the
-   optimization path for a non-convex loss function follows
-   that of a linearized model. The conclusion is that lazy training
-   does not explain the success of over-parameterized neural networks.
-
- \item \cite{GJSx19_TR} Bias and variance analysis of deep neural
-   networks via a Neural Tangent Kernel model. Suggested an ensemble
-   approach of averaging  several over-parameterized networks (for us,
-   the ensemble approximates the ridge solution).
-
- \item \cite{HMRT19_TR} Asymptotic analysis of over-parameterized
-   least squares (and ridge), with exact expressions for isotropic
-   features (not necessarily Gaussian), and then for general
-   covariances (correlated features) the asymptotic expressions depend on the Stieltjes
-   transform, which does not in general have a closed form. They show
-   that the performance of over-parameterized least squares depends on
-   the signal-to-noise ratio (SNR). They also show that optimally
-   tuned ridge is always better than least squares. Finally, they
-   differentiate between the well-specified model and the
-   \emph{misspecified model}, where only a subset of features is
-   observed by the learner.
-
- \item \cite{MM19_TR} Asymptotic analysis of the random features
-   model with a nonlinear activation (i.e., a two-layer neural network
-   with fixed first layer). The random features are drawn uniformly
-   from the unit sphere. The expressions are fairly complicated but
-   also quite accurate. This is not directly comparable to us because
-   our model is well-specified (i.e., we get all the featuers). They
-   point out that for random features the peak in double descent
-   occurs both for the bias and the variance, unlike for
-   well-specified models where only variance exhibits a peak.
-
- \item \cite{LR18_TR}   Upper bounds for the generalization error of
-   Kernel ``Ridgeless''   Regression, i.e, least squares in the
-   over-parameterized regime over an infinite dimensional Hilbert
-   space. Arbitrary covariance matrices are allowed, and an inner
-   product kernel with a non-linear smooth activation is
-   assumed. Additionally, they
-   perform analysis for different rates of exponentially decaying
-   eigenvalues.
-
-\item   \cite{MVSS19_TR} They derive the test error of the “ideal interpolator”
-   for well-specified linear regression,
-   showing that it can be significantly better than the minimum norm
-   one. Also, they analyze sparsity seeking interpolators and they
-   look at several random feature models (but always well-specified).
-
-\item  \cite{Mit19_TR} They introduce a model for misspecified sparse
-  regression and provide exact asymptotic expressions for the
-  generalization error of both the minimum norm and the regularized
-  estimators, where the norm can be either $\ell_2$ or $\ell_1$. 
-\end{itemize}
-\fi
+\input{appendix}
 

### 2022-08-23 design/continuous_nips/continuous-empirical.tex
Commit original repos

\begin{figure}[t]
  \includegraphics[width=\textwidth]{continuous_figures/decays.pdf}
  \vspace{-1cm}
  \caption{Scree-plots of $\Sigmab$ for the eigenvalue decays examined
    in our empirical valuations.
    % Here $d=100$ for visualization, whereas
    % our experiments increase $d$ while preserving the ratio $n/d$ and
    % the decay profile,
    % with $\lambda_{\text{max}}(\Sigmab) = 1$ to
    % $\lambda_{\text{min}}(\Sigmab) = 10^{-4}$.
  }
  \label{fig:eig-decays}
\end{figure}

\section{Empirical evaluation of asymptotic consistency}
\label{sec:asymp-conj-details}

% While Theorem~\ref{t:asymptotic} is a result on asymptotic consistency of the
% MSE, the proof (Appendix~\ref{sec:proof-of-t-asymptotic}) follows the standard
% decomposition of MSE in Equation~\ref{eq:mse-derivation} in the process
% establish consistency on the bias and variance terms independently.

% Recall that $\lambda_n=\frac {d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})}$,
% so our surrogate MSE is recovered as
% $\sigma^2\Vc(\Sigmab,n)+\w^{*\top}\Bc(\Sigmab,n)\w^*$.
% % The expressions $\Vc(\Sigmab,n)$ and $\Bc(\Sigmab,n)$ and recover
% % surrogate MSE from Theorem \ref{t:mse} since $\lambda_n=\frac{\tr((\Sigmab+\lambda_n)^{-1})}{d-n}$.
% Lemmas \ref{c:wishart} and \ref{c:projection} provide new insights into classical matrix-variate
% distributions with extensive literature dedicated to them \citep[see,
% e.g.,][]{chikuse1990matrix,cook2011}.
%
%
%

%\section{Empirical evaluation of asymptotic consistency}

% Theorem \ref{t:asymptotic} shows that the surrogate MSE expressions
% derived in Theorem \ref{t:mse} are asymptotically consistent with the
% true MSE for the i.i.d.~design under a class of sub-gaussian data
% distributions.
In this section, we empirically quantify the convergence rates for
the asymptotic result of Theorem~\ref{t:asymptotic}.
% when $\mu$ is a centered multivariate% Gaussian $\Nc(\zero,\Sigmab).$
We focus on the under-determined regime (i.e.,
$n<d$) and separate the evaluation into the bias and
variance terms, following the MSE decomposition given
in \eqref{eq:mse-derivation}. Consider  $\X = \Z\Sigmab^{1/2} $, where the entries of $\Z$ are
i.i.d. standard Gaussian, and define:\vspace{-1mm}
\begin{enumerate}
  \item Variance discrepancy:\quad
    $\big|\frac{\E[\tr((\X^\top\X)^\dagger)]}{\Vc(\Sigmab,n)}-1\big|$ where
    $\Vc(\Sigmab,n)=\frac{1-\alpha_n}{\lambda_n}$.
  \item Bias discrepancy:\quad
     $\sup_{\w\in\R^d\backslash\{\zero\}}\big|\frac{\w^\top\E[\I-\X^\dagger\X]\w}
     {\w^\top\Bc(\Sigmab,n)\w} - 1\big|$
    where $\Bc(\Sigmab,n) = \lambda_n(\Sigmab+\lambda_n\I)^{-1}$.
  \end{enumerate}\vspace{-1mm}
   Recall that $\lambda_n=\frac {d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})}$,
so our surrogate MSE can be written as
$\Mc=\sigma^2\Vc(\Sigmab,n)+\w^{*\top}\Bc(\Sigmab,n)\w^*$, and when both
discrepancies are bounded by $\epsilon$, then $(1-2\epsilon)\Mc\leq\MSE{\X^\dagger\y}\leq (1+2\epsilon)\Mc$.
In our experiments, we consider four standard eigenvalue decay profiles
for $\Sigmab$, including polynomial and exponential decay (see
 \Cref{fig:eig-decays} and \Cref{sec:eig-decay-details}).
\begin{figure} %[ht]
  \includegraphics[width=\textwidth]{continuous_figures/wishart_var.pdf}
  \includegraphics[width=\textwidth]{continuous_figures/proj_bias.pdf}
  \vspace{-.8cm}
  \caption{
    Empirical verification of the asymptotic consistency of surrogate MSE.
    We show the discrepancies for the variance (top) and bias
    (bottom),  with bootstrapped $95\%$ confidence intervals, as $d$
    increases and $n/d$ is fixed. We observe
     $O(1/d)$ decay (linear with slope $-1$ on a log-log plot).
  }
  \label{f:conj-wishart}
\end{figure}

Figure~\ref{f:conj-wishart} (top) plots the variance discrepancy (with
$\E[\tr((\X^\top\X)^\dagger)]$ estimated via Monte Carlo
sampling and bootstrapped confidence intervals) as $d$ increases from $10$ to
$1000$, across a range of aspect ratios $n/d$. In all cases, we observe that
the discrepancy decays to zero at a rate of $O(1/d)$.
Figure~\ref{f:conj-wishart} (bottom) plots the bias discrepancy, with the same
rate of decay observed throughout.  Note that the range of $d$ is smaller than
in Figure \ref{f:conj-wishart} (top) because the large number of Monte Carlo
samples (up to two million) required for this experiment made the computations
much more expensive (more details in Appendix \ref{a:empirical}). Based on the
above empirical results, we conclude with a conjecture.
\begin{conjecture}
  \label{c:1-over-d-rate}
  When $\mu$ is a centered multivariate Gaussian and its covariance
  has a constant condition
  number, then, for $n/d$ fixed, the surrogate MSE satisfies:
  $\big|\frac{\textnormal{MSE}[\X^\dagger\y]}{\Mc}-1\big|= O(1/d)$.
\end{conjecture}

% The results of empirically validating \Cref{c:wishart} are illustrated in
% Figure~\ref{f:conj-wishart} (top), where we performed Monte Carlo estimation of
% $\E\big[\tr((\X^\top\X)^\dagger)\big]$ and plot
% $\big|\E[\tr((\X^\top\X)^\dagger)]\,\Vc(\Sigmab,n)^{-1} -1\big|$ as $d$ increases from
% $10$ to $1000$, across a range of aspect ratios $n/d$ and eigenvalue decay
% profiles for $\Sigmab$.
% We observe that on log-log axes all of the plots are
% decreasing with a linear $-1$ slope, consistent with the $O(1/d)$ rate
% predicted by \Cref{c:wishart}. \Cref{c:projection} is handled similarly, by
% sampling $\X\sim\mu^n$ where $\mu=\Nc(\zero,\Sigmab )$ to obtain a Monte Carlo
% estimate of $\E[\I-\X^\dagger\X]$.

% Figure~\ref{f:conj-wishart} (bottom) shows how \eqref{eq:projection} decays as
% we hold the aspect ratio $n/d$ fixed and increase $d$ between $10$ and $100$
% across the listed eigenvalue decay profiles and aspect ratios. Again, we
% observe on log-log axes a linear decay with slope $-1$ consistent with the
% $O(1/d)$ rate posed by \Cref{c:projection}. Note that the range of $d$ is
% smaller than in Figure \ref{f:conj-wishart} (top) because the large number of
% Monte Carlo samples (up to two million) required for this experiment made the
% computations much more expensive.

### 2022-08-23 design/continuous_nips/continuous.tex
Commit original repos

--- 2022-08-23 design/continuous_colt/continuous.tex
@@
  \newif\ifisarxiv
-\isarxivtrue
+ \isarxivfalse
 
@@
    \author{%
-          Micha{\l } Derezi\'{n}ski \\
+           \textbf{Micha{\l } Derezi\'{n}ski} \\
    Department of Statistics\\
@@
    \and
-  Feynman Liang \\
+   \textbf{Feynman Liang} \\
    Department of Statistics\\
@@
    \and
-   Michael W. Mahoney\\
+    \textbf{Michael W. Mahoney}\\
    ICSI and Department of Statistics\\
@@
 \else
-\documentclass[anon]{colt2020}
-
-\title[Exact expressions for double descent and implicit
-regularization]{Exact expressions for double descent and implicit
-  regularization\\
-  via surrogate random design}
-\usepackage{times}
-
-\coltauthor{%
-  \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
-  \addr Department of Statistics\\
+
+\documentclass{article}
+\usepackage[final]{neurips_2020}
+
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+
+\usepackage{amsmath}
+\usepackage{cancel}
+\usepackage{cleveref}
+\usepackage{subfigure}
+\usepackage{todonotes}
+ \usepackage{wrapfig}
+\usepackage{xcolor}
+\usepackage{xfrac}
+\hypersetup{
+	colorlinks,
+	linkcolor={red!40!gray},
+	citecolor={blue!40!gray},
+	urlcolor={blue!70!gray}
+}
+
+\title{Exact expressions for double descent and implicit regularization via
+surrogate random design}
+  \author{%
+Micha{\l } Derezi\'{n}ski \\
+  Department of Statistics\\
   University of California, Berkeley\\
-  \AND
-  \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
-  \addr Department of Statistics\\
+  \texttt{mderezin@berkeley.edu}\\
+  \And
+Feynman Liang \\
+  Department of Statistics\\
   University of California, Berkeley\\
-  \AND
-  \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
-  \addr ICSI and Department of Statistics\\
-  University of California, Berkeley
+  \texttt{feynman@berkeley.edu}
+  \And
+Michael W. Mahoney\\
+  ICSI and Department of Statistics\\
+  University of California, Berkeley\\
+  \texttt{mmahoney@stat.berkeley.edu}
 }
-
-\usepackage{float}
-\usepackage{wrapfig}
+% \usepackage{times}
+
+% \coltauthor{%
+%   \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
+%   \addr Department of Statistics\\
+%   University of California, Berkeley\\
+%   \AND
+%   \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
+%   \addr Department of Statistics\\
+%   University of California, Berkeley\\
+%   \AND
+%   \Name{Zhenyu Liao} \Email{TODO}\\
+%   \addr Department of Statistics\\
+%   University of California, Berkeley\\
+%   \AND
+%   \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
+%   \addr ICSI and Department of Statistics\\
+%   University of California, Berkeley
+% }
+
+% \usepackage{float}
+% \usepackage{wrapfig}
 
@@
 \input{../shortdefs}
+
+\newcommand{\RED}{\color[rgb]{0.70,0,0}}
 
@@
   regression which can also generalize well in the over-parameterized
-  regime. We build on recent advances in Randomized Numerical Linear
-  Algebra (RandNLA) to provide the first exact non-asymptotic
+  regime. We provide the first exact non-asymptotic
   expressions for double descent of the minimum norm linear
-  estimator. Our approach involves constructing
-  what we call a surrogate random design to replace the standard
-  i.i.d.~design of the training sample. This surrogate design admits
-  exact expressions for the mean squared error of the estimator while
-  preserving the key properties of the standard design.
-  We also establish an exact implicit regularization result for
-  over-parameterized training samples. In particular, we show that, for
-  the surrogate design, the implicit bias of the unregularized minimum
-  norm estimator precisely corresponds to solving a ridge-regularized
-  least squares problem on the population distribution.
+  estimator. Our approach involves constructing a special
+  determinantal point process  which we call surrogate random
+  design, to replace the standard i.i.d.~design of the training
+  sample. This surrogate design admits exact expressions for the mean
+  squared error of the estimator while preserving the key properties
+  of the standard design. We also establish an exact implicit
+  regularization result for over-parameterized training samples. In
+  particular, we show that, for the surrogate design, the implicit bias
+  of the unregularized minimum norm estimator precisely corresponds to
+  solving a ridge-regularized least squares problem on the population
+  distribution. In our analysis we introduce a new mathematical tool of
+  independent interest: the class of random matrices for which
+  determinant commutes with expectation.
 \end{abstract}
-
-\ifisarxiv\else
-\begin{keywords}
-  Least squares, determinantal point process,
-  double descent, implicit regularization
-\end{keywords}
-\fi
 
@@
 (MSE) of the Moore-Penrose estimator (see Theorem~\ref{t:mse}).  This provides
-a precise characterization of the double descent phenomenon for perhaps the
-simplest and most ubiquitous regression problem.  In obtaining these results,
+a precise characterization of the double descent phenomenon for the
+linear regression problem. In obtaining these results,
 we are able to provide precise formulas for the \emph{implicit regularization}
@@
 relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}).
-This result has been observed empirically for RandNLA
-methods~\citep{Mah-mat-rev_JRNL}, but it has also been shown in deep
-learning~\citep{Ney17_TR} and machine learning~\citep{Mah12} more generally.  To
-obtain our precise results, we use a somewhat non-standard random design, which
-we term surrogate random design (see Section~\ref{s:determinantal} for a
-detailed discussion), and which we expect will be of more general interest.
-Informally, the goal of a surrogate random design is to modify an original
-design to capture its main properties while being ``nicer'' in some useful way.
-In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
-theoretically and empirically, that our surrogate design accurately preserves
-the key properties of the original design when the data distribution is a
-multivariate Gaussian.
+% This result has been observed empirically for RandNLA
+% methods~\citep{Mah-mat-rev_JRNL}, but it has also been shown in deep
+% learning~\citep{Ney17_TR} and machine learning~\citep{Mah12} more generally.
+To obtain our precise results, we use a somewhat non-standard random
+design, based on a specially chosen determinantal point process (DPP), which we
+term surrogate random design. DPPs are a family of non-i.i.d.~sampling
+distributions which are typically used to induce diversity in the
+produced samples \citep{dpp-ml}. Our aim in using a DPP as a surrogate
+design is very different: namely, to make certain quantities (such as the MSE)
+analytically tractable, while accurately \emph{preserving} the underlying
+properties of the original data distribution. This strategy might seem
+counter-intuitive since DPPs are typically found most useful when they
+\emph{differ} from the data distribution. However, we show both theoretically
+(Theorem~\ref{t:asymptotic}) and empirically
+(Section~\ref{sec:asymp-conj-details}), that for many commonly studied data
+distributions, such as multivariate Gaussians, our DPP-based surrogate
+design accurately preserves the key properties of the standard i.i.d.~design
+(such as the MSE), and even matches it exactly in the high-dimensional
+asymptotic limit. In our analysis of the surrogate design, we
+introduce the concept of \emph{determinant preserving
+  random matrices} (Section \ref{s:dp}), a class of random matrices for which determinant
+commutes with expectation, which should be of independent interest.
+
+%  To
+% obtain our precise results, we use a somewhat non-standard random design, which
+% we term surrogate random design (see Section~\ref{s:determinantal} for a
+% detailed discussion), and which we expect will be of more general interest.
+% Informally, the goal of a surrogate random design is to modify an original
+% design to capture its main properties while being ``nicer'' in some useful way.
+% In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
+% theoretically and empirically, that our surrogate design accurately preserves
+% the key properties of the original design when the data distribution is a
+% multivariate Gaussian.
 
@@
 In analyzing the MSE, we make the following standard assumption that
-the response noise is homoscedastic with variance $\sigma^2$.
+the response noise is homoscedastic.
 \begin{assumption}[Homoscedastic noise]\label{a:linear}
-  Responses are $y(\x) = \x^\top\w^*+\xi$ where
-$\xi\sim\Nc(0,\sigma^2)$.
+  The noise $\xi =y(\x)- \x^\top\w^*$ has mean $0$ and variance $\sigma^2$.
 \end{assumption}
@@
 $\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
-size (details in Section~\ref{s:determinantal}). This surrogate is
+size. This surrogate is
 used in place of the standard $n\times d$ random design $\X\sim\mu^n$, where $n$
 data points (the rows of $\X$) are sampled independently from
-$\mu$. Unlike for the standard design, our MSE formula is
-fully expressible as a function of the covariance
+$\mu$. We form the surrogate by constructing a determinantal point
+process with $\mu$ as the background measure, so that $S_\mu^n(\X)\propto
+  \pdet(\X\X^\top)\mu(\X)$, where $\pdet(\cdot)$ denotes the pseudo-determinant
+(details in Section~\ref{s:determinantal}).  Unlike for the standard
+design, our MSE formula is fully expressible as a function of the covariance
 matrix $\Sigmab_\mu=\E_\mu[\x\x^\top]$. To state our main result, we
 need an additional minor assumption on $\mu$ which is satisfied by
-most standard continuous distributions such as any multivariate
-Gaussian with positive definite covariance matrix.
+most standard continuous distributions (e.g., multivariate Gaussians).
 \begin{assumption}[General position]\label{a:general-position}
@@
   general position (Assumption~\ref{a:general-position}), then for
-  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and
+  any $\w \in \mathbb{R}^d$,
+  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}), and
   $\yb_i=y(\xbb_i)$, %we have
+  \vspace{-2mm}
   \begin{align*}
@@
       {\tr((\Sigmab_\mu+\lambda_n\I)^{-1})}\cdot (d-n),
-&\text{for }n<d,\\
-\sigma^2\, \tr(\Sigmab_\mu^{-1}),& \text{for }n=d,\\
-\sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d},&\text{for
+                                                                 & \text{for }n<d, \\
+      \sigma^2\, \tr(\Sigmab_\mu^{-1}),                          & \text{for }n=d, \\
+      \sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d}, & \text{for
       }n>d,
@@
 \end{theorem}
-\begin{definition}
-  We will use $\Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
+\begin{definition}\label{d:expression}
+  We will use $\Mc = \Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
   for $\MSE{\Xb^\dagger\ybb}$.
@@
     $\w^*=\frac1{\sqrt{d}}\one$.}
-\vspace{-5mm}
+  %\vspace{-5mm}
   \label{f:intro}
@@
 previously, no such theory was available except for special
-cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
-and Appendices \ref{sec:proof-of-t-asymptotic} and \ref{sec:asymp-conj-details}). The plots
+cases such as $\Sigmab=\I$ (more details in Theorem~\ref{t:asymptotic}
+and Section~\ref{sec:asymp-conj-details}). The plots
 show that varying the spectral decay of $\Sigmab$ has a significant effect on the
@@
   \label{t:unbiased}
-For $\mu$ satisfying\footnote{The proof of Theorem \ref{t:unbiased}
-  can be easily extended to probability measures $\mu$ that do not
-  satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
-  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and
-  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
-  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$, then
+  For $\mu$ satisfying
+  % \footnote{The proof of Theorem \ref{t:unbiased}
+  %   can be easily extended to probability measures $\mu$ that do not
+  %   satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
+  %   assumption here to simplify the presentation.}
+  Assumption~\ref{a:general-position} and
+  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
+  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$,
   \begin{align*}
@@
     \begin{cases}
-       (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} &\text{for }n<d,\\
-        \Sigmab_\mu^{-1}\v_{\mu,y} &\text{for }n \ge d,
+      (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} & \text{for }n<d,     \\
+      \Sigmab_\mu^{-1}\v_{\mu,y}                 & \text{for }n \ge d,
     \end{cases}
@@
 our exact implicit regularization results in Theorem~\ref{t:unbiased}
-are derived for the surrogate design, but Figure \ref{f:intro}
-suggests that they accurately describe the MSE (up to lower order
-terms) also under the standard i.i.d.~design $\X\sim\mu^n$,
-particularly when $\mu$ is a multivariate Gaussian.
-As a third result, we can verify this in the cases where there exist
-known expressions for the MSE under the i.i.d.~design (standard
-Gaussian for the under-determined setting, and
-arbitrary Gaussian for the over-determined one).
-We give the proof in Appendix~\ref{sec:proof-of-t-asymptotic}.
+are derived for the surrogate design, which is a
+non-i.i.d.~distribution based on a determinantal point process. However, Figure \ref{f:intro}
+suggests that those expressions accurately describe the MSE (up to lower order
+terms) also under the standard i.i.d.~design $\X\sim\mu^n$ when $\mu$ is a multivariate Gaussian.
+As a third result, we verify that the surrogate expressions for the
+MSE are asymptotically consistent with the MSE of an i.i.d.~design,
+for a wide class of distributions which include multivariate Gaussians.
+
 \begin{theorem}[Asymptotic consistency of surrogate design]
   \label{t:asymptotic}
-Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
-and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
-c_\rho=\frac3{|1-\rho|}$ and
+  Let $\X \in \mathbb{R}^{n \times d}$ have i.i.d.~rows $\x_i^\top = \z_i^\top\Sigmab^{\frac12}$
+  where $\z_i$ has independent zero mean and unit variance
+  sub-Gaussian entries, and suppose that Assumptions~\ref{a:linear}
+  and \ref{a:general-position} are satisfied. Furthermore, suppose
+  that there
+  exist $c, C, C^* \in \mathbb{R}_{> 0}$ such that
+  $C \I \succeq \Sigmab \succeq c \I \succ 0$ and $\| \w^* \| \le
+    C^*$. Then
+  %and the limiting spectral measure (i.e., normalized counting measure of the eigenvalues) of $\Sigmab \in \mathbb R^{d \times d}$ converges weakly to some measure $\nu$ as $d \to \infty$. Then
   \begin{align*}
-  \mu &= \begin{cases}
-    \Nc(\zero,\I)&\text{when }n<d-1,\\
-    \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
-  \end{cases}
+    \MSE{\X^\dagger\y} - \Mc(\Sigmab, \w^*,\sigma^2,n) \to 0
   \end{align*}
-then the absolute difference between surrogate
-expressions and the true MSE is bounded as follows:
-% &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
-%     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
-%     \Sigmab\succ\zero,\\
-\begin{align*}
-\Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
-\frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
-\end{align*}
+  with probability one as $d,n \to \infty$ with $n/d \to \bar c \in (0,\infty) \setminus \{1\}$.
+  % Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
+  % and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
+  % c_\rho=\frac3{|1-\rho|}$ and
+  % \begin{align*}
+  %   \mu &= \begin{cases}
+  %     \Nc(\zero,\I)&\text{when }n<d-1,\\
+  %     \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
+  %   \end{cases}
+  % \end{align*}
+  % then the absolute difference between surrogate
+  % expressions and the true MSE is bounded as follows:
+  % % &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
+  % %     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
+  % %     \Sigmab\succ\zero,\\
+  % \begin{align*}
+  % \Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
+  % \frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
+  % \end{align*}
 \end{theorem}
-\begin{remark}
-For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
-random design can be infinite, whereas the
-surrogate MSE is finite and has a closed form expression.
-\end{remark}
-
-\noindent
-Empirical estimates given in Figure
-\ref{f:intro} suggest that the consistency of surrogate expressions
-holds much more generally than it is stated above. Based on a detailed
-empirical analysis described in Appendix \ref{sec:asymp-conj-details},
-we conjecture that an asymptotic consistency 
-result of the form similar to the statement of Theorem~\ref{t:asymptotic}
-holds true in the under-determined regime without the assumption that
-$\Sigmab=\I$ (see Conjectures \ref{c:wishart} and
-\ref{c:projection}). In this case, no formula is known for 
-$\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
-Gaussian design naturally extend.
-
-\subsection{Key techniques: surrogate designs and
-  determinant preserving matrices}
-
-The standard random design model for linear regression assumes that
-each pair $(\x_i^\top,y_i)$ is drawn independently, where the row vector
-$\x_i^\top$ comes from some $d$-variate distribution $\mu$ and $y_i=y(\x_i)$ is a random response variable drawn conditionaly on $\x_i$.
-Precise theoretical analysis of under-determined regression in this
-setting poses significant challenges, even in such special cases as
-the Moore-Penrose estimator and a Gaussian data distribution $\mu$.
-Rather than trying to directly analyze the usual i.i.d.~random design
-$\X\sim\mu^n$ described above, we modify it slightly by introducing
-the notion of a \emph{surrogate random design}, $\Xb\sim S_\mu^n$.
-Informally, the goal of a surrogate random design is to modify an
-original design to capture the main properties of the original design,
-while being ``nicer'' for theoretical or empirical analysis.
-In particular, here, we will modify the distribution of matrix $\X$ so
-as to:
-\begin{enumerate}
-  \item closely preserve the behavior of the Moore-Penrose estimator
-from the i.i.d.~design; and
-  \item obtain exact expressions for double descent in terms of the
-    mean squared error.
-\end{enumerate}
-A key element in the construction of our surrogate designs involves rescaling the measure $\X\sim\mu^n$ by the pseudo-determinant $\pdet(\X\X^\top)$, i.e., a product of the non-zero eigenvalues.
-A similar type of determinantal design was suggested by
-\cite{correcting-bias-journal}, but it was restricted there only to
-$n\geq d$.
-We broaden this definition by not only allowing the sample size to be
-less than $d$, but also allowing it to be randomized.
-A key property of our determinantal design matrix $\Xb$ is that
-the expectation of a function $F(\Xb)$ can be, up to normalization,
-expressed as:
-\begin{align*}
-  \E[F(\Xb)]\ \propto\ \E[\pdet(\X\X^\top)F(\X)],
-  \end{align*}
-where $\X\sim\mu^K$ and $K$ is a random variable (see details in Definition~\ref{d:det}).
-Then, we define (in Definition~\ref{d:surrogate}) our surrogate design $S_\mu^n$ for each $n>0$ as a determinantal design with a carefully chosen random variable $K$, so that the expected sample size is equal to $n$ and so that it is possible to derive closed form expressions for the MSE.
-We achieve this by using the Poisson distribution to construct the variable~$K$.
-
-The key technical contribution that allows us to derive the MSE for determinantal designs is the concept of \emph{determinant preserving random matrices}, a notion that we expect to be useful more generally.
-Specifically, in Section~\ref{s:dp} we define a class of $d\times d$ random
-matrices $\A$ for which taking the determinant commutes with taking
-the expectation, for the matrix itself and any of its square submatrices (see Definition~\ref{d:main}):
-\begin{align*}
-  \E\big[\!\det(\A_{\Ic,\Jc})\big] =
-  \det\!\big(\E[\A_{\Ic,\Jc}]\big)\quad \text{for all }\Ic,\Jc\subseteq
-  [d]\text{ s.t. }|\Ic|=|\Jc|.
-\end{align*}
-  Not all random matrices satisfy this property, however many
-interesting and non-trivial examples can be found. Constructing
-these examples is facilitated by the closure properties that this
-class enjoys. In particular, if $\A$ and $\B$ are determinant
-preserving and independent, then $\A+\B$ and $\A\B$ are also
-determinant preserving (see Lemma \ref{t:ring}). We use these
-techniques to prove a number of determinantal expectation
-formulas needed in obtaining Theorems~\ref{t:mse}
-and~\ref{t:unbiased}. For example, we show that if $\X\sim\mu^K$,
-where $K$ is a Poisson random variable, then:
-\begin{align*}
-  \text{(Lemma \ref{l:poisson})}\quad  \E\big[\det(\X^\top\X)\big]
-  &= \det\!\big(\E[\X^\top\X]\big),\\
-\text{(Lemma \ref{l:normalization})}\quad  \E\big[\det(\X\X^\top)\big]
-&= \ee^{-\E[K]}\det\!\big(\I + \E[\X^\top\X]\big).
-\end{align*}
+
+% \begin{remark}
+% For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
+% random design can be infinite, whereas the
+% surrogate MSE is finite and has a closed form expression.
+% \end{remark}
+
+\noindent
+The above result is particularly remarkable since our surrogate design
+is a determinantal point process. DPPs are commonly used in ML to
+ensure that the data points in a sample are well spread-out. However,
+if the data distribution is sufficiently regular (e.g., a multivariate
+Gaussian), then the i.i.d.~samples are already spread-out reasonably
+well, so rescaling the distribution by a determinant has a negligible
+effect that vanishes in the high-dimensional regime.
+Furthermore, our empirical estimates (Figure~\ref{f:intro}) suggest that the surrogate expressions
+are accurate not only in the asymptotic limit, but even for moderately large
+dimensions. Based on a detailed empirical analysis described in
+Section \ref{sec:asymp-conj-details},
+we conjecture that the convergence described in
+Theorem~\ref{t:asymptotic} has the rate of~$O(1/d)$.
+% holds true in the under-determined regime without the assumption that
+% $\Sigmab=\I$ (see Conjectures \ref{c:wishart} and
+% \ref{c:projection}). In this case, no formula is known for 
+% $\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
+% Gaussian design naturally extend.
+
+% \subsection{Key techniques: surrogate designs and
+%   determinant preserving matrices}
+
+% The standard random design model for linear regression assumes that
+% each pair $(\x_i^\top,y_i)$ is drawn independently, where the row vector
+% $\x_i^\top$ comes from some $d$-variate distribution $\mu$ and $y_i=y(\x_i)$ is a random response variable drawn conditionaly on $\x_i$.
+% Precise theoretical analysis of under-determined regression in this
+% setting poses significant challenges, even in such special cases as
+% the Moore-Penrose estimator and a Gaussian data distribution $\mu$.
+% Rather than trying to directly analyze the usual i.i.d.~random design
+% $\X\sim\mu^n$ described above, we modify it slightly by introducing
+% the notion of a \emph{surrogate random design}, $\Xb\sim S_\mu^n$.
+% Informally, the goal of a surrogate random design is to modify an
+% original design to capture the main properties of the original design,
+% while being ``nicer'' for theoretical or empirical analysis.
+% In particular, here, we will modify the distribution of matrix $\X$ so
+% as to:
+% \begin{enumerate}
+%   \item closely preserve the behavior of the Moore-Penrose estimator
+% from the i.i.d.~design; and
+%   \item obtain exact expressions for double descent in terms of the
+%     mean squared error.
+% \end{enumerate}
+% A key element in the construction of our surrogate designs involves rescaling the measure $\X\sim\mu^n$ by the pseudo-determinant $\pdet(\X\X^\top)$, i.e., a product of the non-zero eigenvalues.
+% A similar type of determinantal design was suggested by
+% \cite{correcting-bias-journal}, but it was restricted there only to
+% $n\geq d$.
+% We broaden this definition by not only allowing the sample size to be
+% less than $d$, but also allowing it to be randomized.
+% A key property of our determinantal design matrix $\Xb$ is that
+% the expectation of a function $F(\Xb)$ can be, up to normalization,
+% expressed as:
+% \begin{align*}
+%   \E[F(\Xb)]\ \propto\ \E[\pdet(\X\X^\top)F(\X)],
+%   \end{align*}
+% where $\X\sim\mu^K$ and $K$ is a random variable (see details in Definition~\ref{d:det}).
+% Then, we define (in Definition~\ref{d:surrogate}) our surrogate design $S_\mu^n$ for each $n>0$ as a determinantal design with a carefully chosen random variable $K$, so that the expected sample size is equal to $n$ and so that it is possible to derive closed form expressions for the MSE.
+% We achieve this by using the Poisson distribution to construct the variable~$K$.
+
+% The key technical contribution that allows us to derive the MSE for determinantal designs is the concept of \emph{determinant preserving random matrices}, a notion that we expect to be useful more generally.
+% Specifically, in Section~\ref{s:dp} we define a class of $d\times d$ random
+% matrices $\A$ for which taking the determinant commutes with taking
+% the expectation, for the matrix itself and any of its square submatrices (see Definition~\ref{d:main}):
+% \begin{align*}
+%   \E\big[\!\det(\A_{\Ic,\Jc})\big] =
+%   \det\!\big(\E[\A_{\Ic,\Jc}]\big)\quad \text{for all }\Ic,\Jc\subseteq
+%   [d]\text{ s.t. }|\Ic|=|\Jc|.
+% \end{align*}
+%   Not all random matrices satisfy this property, however many
+% interesting and non-trivial examples can be found. Constructing
+% these examples is facilitated by the closure properties that this
+% class enjoys. In particular, if $\A$ and $\B$ are determinant
+% preserving and independent, then $\A+\B$ and $\A\B$ are also
+% determinant preserving (see Lemma \ref{t:ring}). We use these
+% techniques to prove a number of determinantal expectation
+% formulas needed in obtaining Theorems~\ref{t:mse}
+% and~\ref{t:unbiased}. For example, we show that if $\X\sim\mu^K$,
+% where $K$ is a Poisson random variable, then:
+% \begin{align*}
+%   \text{(Lemma \ref{l:poisson})}\quad  \E\big[\det(\X^\top\X)\big]
+%   &= \det\!\big(\E[\X^\top\X]\big),\\
+% \text{(Lemma \ref{l:normalization})}\quad  \E\big[\det(\X\X^\top)\big]
+% &= \ee^{-\E[K]}\det\!\big(\I + \E[\X^\top\X]\big).
+% \end{align*}
 
@@
 \textbf{Double descent.}
-The double descent phenomenon \cite[a term introduced by][]{BHMM19}
-corresponds to the phase transition in the generalization error that
-occurs when the ratio between the model complexity and the
-sample size crosses the so-called interpolation threshold. It has been observed empirically
+The double descent phenomenon has been observed empirically
 in a number of learning models, including neural networks
@@
 properties of interpolating estimators, have primarily focused on various forms of
-linear regression. The most comparable
-to our work are \cite{BLLT19_TR,LR18_TR} and \cite{HMRT19_TR}, who provide
-non-asymptotic upper/lower bounds and asymptotic formulas,
-respectively, for the generalization error of the Moore-Penrose
-estimator under essentially the same i.i.d.~random design setting as
-ours. On the other hand, \cite{MVSS19_TR} provide bounds for the error
-of the ideal linear interpolator (instead of the minimum norm
-one). Note that while we analyze the classical mean squared error, many
-works focus on the squared prediction error instead (some of them still
-refer to it as the MSE).
-Another line of literature deals with linear regression in the
+linear regression \citep{BLLT19_TR,LR18_TR,HMRT19_TR,MVSS19_TR}. Note
+that while we analyze the classical mean squared error, many
+works focus on the squared prediction error.
+Also, unlike in our work, some of the literature on double descent deals with linear regression in the
 so-called \emph{misspecified} setting, where the set of observed
@@
 learner observes a random subset of $d$ features from a larger
-population. This is an important distinction,
-because it allows varying the model complexity by changing the number
-of observed features while keeping the linear model fixed (see further
-discussion in Section \ref{s:conclusions}). We believe
-that our results can be extended to this important setting, and we leave
-this as a direction for future work.
-
-% Here are things to cite about double descent and related things.
-% Original double descent papers:
-% \cite{BHMM19}
-% \cite{BMM18_TR}
-% \cite{BRT18_TR}
-% \cite{belkin2019two}
-% Other double descent papers:
-% \cite{AS17_TR}
-% \cite{BLLT19_TR}
-% \cite{BHM18_TR}
-% \cite{BFF19_TR}
-% \cite{COB18_TR}
-% \cite{GJSx19_TR}
-% \cite{HMRT19_TR}
-% \cite{LR18_TR}
-% \cite{MBB17_TR}
-% \cite{MM19_TR}
-% \cite{Mit19_TR}
-% \cite{MVSS19_TR}
-% \cite{NMBx18_TR}
-% \cite{SKS18_TR}
-% \cite{SGAx18_TR}
-% \cite{WGSx19_TR}
-% \michael{Say something about how we use MSE but others use one of a wide range of slightly different generalization metrics. Would you put in a draft, and then I can make a pass.}
-
-\textbf{RandNLA.}
-Randomized numerical linear algebra
+population.
+
+\begin{wrapfigure}{r}{\ifisarxiv 0.5\else 0.44\fi\textwidth}
+  \ifisarxiv\else \vspace{-.2cm}\fi
+  \centering
+  \includegraphics[width=.43\textwidth]{figs/descent-model-log}
+  \ifisarxiv\else \vspace{-3mm}\fi
+  \caption{
+    Surrogate MSE as a function of $d/n$, with $n$
+    fixed to $100$ and varying dimension $d$ and condition number
+    $\kappa$, for signal-to-noise ratio $\mathrm{SNR}=1$.}
+  \label{f:model}
+  \ifisarxiv\else  \vspace{-.5cm}\fi
+\end{wrapfigure}
+
+The most directly comparable to our setting is the recent work of
+\cite{HMRT19_TR}. They study how
+varying the feature dimension affects the (asymptotic) generalization
+error for linear regression, % for both well-specified and misspecified tasks
+however their analysis is limited to certain special settings such as an
+isotropic data distribution. As an additional
+point of comparison, in Figure~\ref{f:model} we plot the MSE
+expressions of Theorem~\ref{t:mse} when varying the
+feature dimension~$d$ (the setup is the same as in
+Figure~\ref{f:intro}). Our plots follow the
+trends outlined by \cite{HMRT19_TR} for the isotropic case (see their Figure~2),
+but the spectral decay of the covariance (captured by our new
+MSE expressions) has a significant effect on the descent curve.
+This leads to generalization in the under-determined regime even when
+the signal-to-noise ratio ($\mathrm{SNR}=\|\w^*\|^2/\sigma^2$) is 1,
+unlike suggested by \cite{HMRT19_TR}.
+
+\textbf{RandNLA and DPPs.}
+Randomized Numerical Linear Algebra
 \citep{DM16_CACM,RandNLA_PCMIchapter_TR} has
-traditionally focused on obtaining algorithmic improvements for tasks
-such as least squares and low-rank approximation via techniques that include
-sketching \citep{sarlos-sketching} and i.i.d.~leverage score sampling
-\citep{Drineas2006sampling}. However, there has been growing
+traditionally focused on obtaining purely algorithmic improvements for tasks
+such as least squares regression, %  and low-rank approximation via techniques that include
+% sketching \citep{sarlos-sketching} and i.i.d.~leverage score sampling
+% \citep{Drineas2006sampling}. However,
+but there has been growing
 interest in understanding the statistical properties of these
-randomized methods \citep{MMY15,GarveshMahoney_JMLR}, for example
-looking at the mean squared error of the least squares estimator
-obtained via i.i.d.~subsampling under the standard linear response
-model. Determinantal sampling methods (a.k.a.~volume sampling, or
-determinantal point processes), which first found their way
-into RandNLA in the context of low-rank approximation
-\cite[][]{pca-volume-sampling}, have been recently shown to combine
-strong worst-case guarantees with elegant statistical properties.
-In particular, \cite{unbiased-estimates} showed that the
-least-squares estimator subsampled via the so-called size $d$ volume
-sampling (loosely corresponding to the special case of our surrogate design
-$S_\mu^n$ where $n=d$) is an unbiased estimator that admits exact
-formulas for both the expected square loss (a worst-case metric)
-and the mean squared error (a statistical metric). These results were
-developed further by
-\cite{leveraged-volume-sampling,correcting-bias,minimax-experimental-design},
-however they were still limited to the over-determined setting 
-\cite[with the exception
-of][who gave upper bounds on the mean squared error of the ridge estimator
-under different determinantal
-samplings]{regularized-volume-sampling,bayesian-experimental-design}.
-Also in the over-determined setting, 
-\cite{correcting-bias-journal} provided evidence for the fact that
-determinantal rescaling can be used to modify the original data
-distribution (particularly, a multivariate Gaussian) without a
-significant distortion to the estimator, while making certain
-statistical quantities expressible analytically. We take this
-direction further by analyzing the unregularized least squares
-estimator in the under-determined setting which is less well
-understood, partly due to the presence of implicit regularization.
+randomized methods \citep{MMY15,GarveshMahoney_JMLR} and
+a beyond worst-case analysis \cite{precise-expressions}.
+Determinantal point processes \citep{dpps-in-randnla,dpp-ml} have been recently shown to combine
+strong worst-case regression guarantees with elegant statistical
+properties \citep{unbiased-estimates}.
+However, these results are limited to the over-determined setting \citep{leveraged-volume-sampling,correcting-bias,minimax-experimental-design} and
+ridge regression
+\citep{regularized-volume-sampling,bayesian-experimental-design}.
+Our results are also related to recent work on using DPPs
+to analyze the expectation of the
+inverse~\citep{determinantal-averaging,debiasing-second-order} and generalized
+inverse~\citep{randomized-newton,nystrom-multiple-descent} of a subsampled matrix.
+
+% Also in the over-determined setting, 
+% \cite{correcting-bias-journal} provided evidence for the fact that
+% determinantal rescaling can be used to modify the original data
+% distribution (particularly, a multivariate Gaussian) without a
+% significant distortion to the estimator, while making certain
+% statistical quantities expressible analytically. We take this
+% direction further by analyzing the unregularized least squares
+% estimator in the under-determined setting which is less well
+% understood, partly due to the presence of implicit regularization.
 
@@
 The term implicit regularization typically refers to the notion that approximate
-computation (e.g., rather than exactly minimizing a function $f$,
-instead running an approximation algorithm to get an approximately
-optimal solution) can implicitly lead to
-statistical regularization (e.g., exactly minimizing an objective of
-the form $f + \lambda g$, for some well-specified $\lambda$ and $g$).
+computation % (e.g., rather than exactly minimizing a function $f$,
+% instead running an approximation algorithm to get an approximately
+% optimal solution)
+can implicitly lead to
+statistical regularization. % (e.g., exactly minimizing an objective of
+% the form $f + \lambda g$, for some well-specified $\lambda$ and $g$)
 See~\cite{MO11-implementing, PM11, GM14_ICML} and references therein
@@
 optimization algorithms.
-See, e.g., theoretical results on simplified
-models~\citep{NTS14_TR,Ney17_TR,SHNx17_TR,GWBNx17,ACHL19,KBMM19_TR}
-as well as extensive empirical and phenomenological results on state-of-the-art neural
-network models~\citep{MM18_TR,MM19_HTSR_ICML}.
+See, e.g., theoretical results~\citep{NTS14_TR,Ney17_TR,SHNx17_TR,GWBNx17,ACHL19,KBMM19_TR}
+as well as extensive empirical studies~\citep{MM18_TR,MM19_HTSR_ICML}.
 The implicit regularization observed by us is different in that it is
@@
 solution). In this context, most relevant are the
-asymptotic results of~\cite{LJB19_TR} \cite[which used the asymptotic risk
-results for ridge regression of][]{DW15_TR} and~\cite{KLS18_TR}.
-Our non-asymptotic results are also related to recent work in
-RandNLA on the expectation of the 
-inverse~\citep{determinantal-averaging} and generalized
-inverse~\citep{MDK19_TR} of a subsampled matrix.
+asymptotic results of \cite{KLS18_TR} and \cite{LJB19_TR}.
 
@@
 of the standard random design $\mu^n$ consisting of $n$ row vectors drawn
-independently from $\mu$. Our surrogate design uses determinantal
-rescaling to alter the joint distribution of the vectors so that
-certain expected quantities (such as the mean squared error of the
-Moore-Penrose estimator) can be expressed in a closed form.
-We start by introducing notation.
+independently from $\mu$.
+% Our surrogate design uses determinantal
+% rescaling to alter the joint distribution of the vectors so that
+% certain expected quantities (such as the mean squared error of the
+% Moore-Penrose estimator) can be expressed in a closed form.
+% We start by introducing notation.
 %to be used throughout the rest of the paper.
@@
 \textbf{Preliminaries.}
-The set $\{1,...,n\}$ will be denoted by $[n]$.
+%The set $\{1,...,n\}$ will be denoted by $[n]$.
 For an $n\times n$ matrix $\A$, we use $\pdet(\A)$ to denote the pseudo-determinant of $\A$,
-which is the product of non-zero eigenvalues. For index subsets $\Ic$
+which is the product of non-zero eigenvalues (repeated eigenvalues are taken to the power of their
+algebraic multiplicity).
+For index subsets $\Ic$
 and $\Jc$, we use $\A_{\Ic,\Jc}$ to denote the submatrix of $\A$ with
 rows indexed by $\Ic$ and columns indexed by $\Jc$. We may write
-$\A_{\Ic,*}$ to indicate that we take a subset of rows. We use $\adj(\A)$ to
-denote the adjugate of $\A$, defined as follows: the
-$(i,j)$th entry of $\adj(\A)$ is
-$(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$.
+$\A_{\Ic,*}$ to indicate that we take a subset of rows.
+% We use $\adj(\A)$ to denote the adjugate of $\A$, defined as follows: the
+% $(i,j)$th entry of $\adj(\A)$ is
+% $(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$.
 % We will
@@
 % relates the adjugate and the determinant: $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$.
-For a probability measure $\mu$ over $\R^d$, we use $\x^\top\!\sim\mu$
-to denote a random row vector $\x^\top$ sampled according to this distibution.
+% For a probability measure $\mu$ over $\R^d$, we use $\x^\top\!\sim\mu$
+% to denote a random row vector $\x^\top$ sampled according to this distibution.
 We let $\X\sim\mu^k$ denote a $k\times d$ random matrix with rows
@@
 
-We now define a family of determinantal distributions over random matrices $\Xb$,
-where not only the entries but also the number of rows is
-randomized.
-This randomized sample size is a crucial property of our
-designs that enables our analysis.
+% We now define a family of determinantal distributions over random matrices $\Xb$,
+% where not only the entries but also the number of rows is
+% randomized.
+% This randomized sample size is a crucial property of our
+% designs that enables our analysis.
 % Our definition follows by
@@
   Let $\mu$ satisfy Assumption~\ref{a:general-position} and let $K$ be
-  a random variable over non-negative integers. A determinantal design
+  a random variable over $\mathbb{Z}_{\geq 0}$. A determinantal design
   $\Xb\sim \Det(\mu,K)$ is a
   distribution with the same domain as $\X\sim\mu^K$ such that~for any
-event $E$ measurable w.r.t.~$\X$, we have
+  event $E$ measurable w.r.t.~$\X$, we have\vspace{-2mm}
   \begin{align*}
@@
   \end{align*}
-\end{definition}
+\end{definition}\vspace{-2mm}
 
@@
 selecting the random variable $K$.
-The obvious choice of $K=n$ does \emph{not} result in simple closed form expressions for the MSE in the under-determined regime (i.e., $n<d$), which is the regime of primary interest to us.
-Instead, we derive our random variables $K$ from the Poisson distribution.
+The obvious choice of $K=n$ does \emph{not} result in simple closed
+form expressions for the MSE in the under-determined regime (i.e.,
+$n<d$), which is the regime of primary interest to us.
+Instead, we derive our random variables $K$ from the Poisson
+distribution.
 \begin{definition}\label{d:surrogate}
@@
     \item if $n<d$, then $K\sim \Poisson(\gamma_n)_{\leq d}$ with
- $\gamma_n$ being the solution of
+          $\gamma_n$ as the solution of
           $n=\tr(\Sigmab_\mu(\Sigmab_\mu+\frac1{\gamma_n}\I)^{-1})$,
@@
 \begin{align}
-  \E[F(\Xb)] &=\begin{cases}
+  \E[F(\Xb)] & =\begin{cases}
     \frac{\E[\det(\X\X^\top)F(\X)]}{\E[\det(\X\X^\top)]}
-   \quad K\sim\Poisson(\gamma_n)&\text{for }n<d,\\[2mm]
+    \quad K\sim\Poisson(\gamma_n) & \text{for }n<d, \\[2mm]
     \frac{\E[\det(\X)^2F(\X)]}{\E[\det(\X)^2]}\hspace{3mm}
-    \quad K=d&\text{for }n=d,\\[2mm]
+    \quad K=d                     & \text{for }n=d, \\[2mm]
     \frac{\E[\det(\X^\top\X)F(\X)]}{\E[\det(\X^\top\X)]}
-   \quad K\sim\Poisson(\gamma_n)&\text{for }n>d.
+    \quad K\sim\Poisson(\gamma_n) & \text{for }n>d.
   \end{cases}\label{eq:cases}
@@
 range of the Poisson distribution.
-The normalization constants for computing the expectations
-can be obtained using the following formulas: if $\X\sim\mu^K$ then
-\begin{align*}
-\text{(Lemma~\ref{l:normalization})}&&
-\E\big[\det(\X\X^\top)\big]  &= \ee^{-\gamma_n}\det(\I +
-\gamma_n\Sigmab_\mu)&&\text{for } K\sim\Poisson(\gamma_n),\ n<d,\\
-  \text{(Lemma~\ref{l:cb})}&& \E\big[\det(\X)^2\big] &=
-  d!\det(\Sigmab_\mu),&&\text{for } K=n=d,\\
-  \text{(Lemma~\ref{l:poisson})}&& \E\big[\det(\X^\top\X)\big]
-&=\det(\gamma_n\Sigmab_\mu),&&  \text{for }K\sim\Poisson(\gamma_n),\ n>d.
-\end{align*}
+We compute the normalization constants
+by introducing the concept of determinant preserving random matrices,
+discussed in Section \ref{s:dp}.
+% can be obtained using the following formulas: if $\X\sim\mu^K$ then
+% \begin{align*}
+% \text{(Lemma~\ref{l:normalization})}&&
+% \E\big[\det(\X\X^\top)\big]  &= \ee^{-\gamma_n}\det(\I +
+% \gamma_n\Sigmab_\mu)&&\text{for } K\sim\Poisson(\gamma_n),\ n<d,\\
+%   \text{(Lemma~\ref{l:cb})}&& \E\big[\det(\X)^2\big] &=
+%   d!\det(\Sigmab_\mu),&&\text{for } K=n=d,\\
+%   \text{(Lemma~\ref{l:poisson})}&& \E\big[\det(\X^\top\X)\big]
+% &=\det(\gamma_n\Sigmab_\mu),&&  \text{for }K\sim\Poisson(\gamma_n),\ n>d.
+% \end{align*}
 % \begin{remark}
@@
 % literature \cite{expected-generalized-variance,correcting-bias}:
-We prove Lemmas \ref{l:poisson} and \ref{l:normalization} in Section
-\ref{s:dp} by introducing the
-concept of determinant preserving random matrices.
+% We prove Lemmas \ref{l:poisson} and \ref{l:normalization} in Section
+% \ref{s:dp} by introducing the
+% concept of determinant preserving random matrices.
 % The lemmas play a
@@
 % and~\ref{t:unbiased} in Section~\ref{s:expectations}.
-On the other hand, Lemma  \ref{l:cb} and the design $S_\mu^d$ can be found in the
-literature \citep{expected-generalized-variance,correcting-bias-journal},
-and we will rely on those known results in this case.
+% On the other hand, Lemma  \ref{l:cb} and the design $S_\mu^d$ can be found in the
+% literature \citep{expected-generalized-variance,correcting-bias-journal},
+% and we will rely on those known results in this case.
 % Importantly, the $n=d$ case offers a continuous transition between the under-
@@
 
-We now highlight some key expectation formulas for surrogate designs, which are used to
-derive the MSE expressions from Theorem \ref{t:mse}. A standard decomposition of
-the MSE yields:% for the Moore-Penrose estimator yields:
+\paragraph{Proof sketch of Theorem \ref{t:mse}}
+% We now highlight some key expectation formulas for surrogate designs, which are used to
+% derive the MSE expressions from Theorem \ref{t:mse}. 
+We focus here on the under-determined regime (i.e., $n<d$),
+highlighting the key new expectation formulas we develop to derive the
+MSE expressions for surrogate designs. A standard decomposition of the MSE yields:
 \begin{align}
   \MSE{\Xb^\dagger\ybb}
-  &= \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\big]
+   & = \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\big]
   =\sigma^2\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big] +
@@
 Thus, our task is to find closed form expressions for the two
-expectations above. If $n\geq d$, then the latter goes away because
-$\Xb^\dagger\Xb$ is the projection onto the row-span of $\Xb$ so
-when $\Xb$ has rank $d$ then $\I-\Xb^\dagger\Xb=\zero$. When
-$n<d$, it is given in the following result (proof in
-Appendix \ref{s:unbiased-proof}).
-% can be obtained as a consequence of Lemma \ref{l:ridge-under}.
+expectations above. The latter, which is the expected projection onto
+the complement of the row-span of $\Xb$, is proven in
+Appendix \ref{s:unbiased-proof}.
 \begin{lemma}\label{l:proj}
@@
 \begin{lemma}\label{l:sqinv-all}
-If  $\Xb\sim S_\mu^n$, then:\vspace{-2mm}
-\begin{align*}
-    \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
-  &=
-    \begin{cases}
-\gamma_n\big(1-
-\det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big),
-&\text{for }n<d,\\
-\tr(\Sigmab_\mu^{-1}),
-&\text{for }n=d,\\
-\tr(\Sigmab_\mu^{-1})\,\frac{1-\ee^{-\gamma_n}}{\gamma_n},
-&\text{for }n>d.
-\end{cases}
-\end{align*}
+  If  $\Xb\sim S_\mu^n$ and $n<d$, then:
+  $\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
+    =\gamma_n\big(1-
+    \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big)$.
 \end{lemma}
@@
 % is closely related to their Theorem 2.9.
-Our key contribution is the under-determined
-regime (i.e., $n<d$), where we observe implicit regularization
+Note the implicit regularization term which appears in both formulas,
 given by $\lambda_n=\frac1{\gamma_n}$. Since $n =
@@
 Combining this with Lemmas \ref{l:proj} and \ref{l:sqinv-all}, we
-recover the surrogate MSE expression in Theorem~\ref{t:mse} (see details
-in Appendix \ref{a:mse-proof}).
+recover the surrogate MSE expression in Theorem~\ref{t:mse}.
 
@@
 
-The key question motivating the following definition is: When does taking expectation commute with computing a determinant for a square random matrix?
+The key question motivating the following definition is: \textit{When does taking expectation commute with computing a determinant for a square random matrix?}
 \begin{definition}\label{d:main}
@@
 \noindent
-Note that from the definition of an adjugate matrix (see preliminaries
-in Section \ref{s:determinantal}) it immediately follows that if $\A$ is
-determinant preserving then adjugate commutes with expectation for this matrix:
-\begin{align}
-  \E\big[\big(\!\adj(\A)\big)_{i,j}\big] &=
-  \E\big[(-1)^{i+j}\det(\A_{[d]\backslash\{j\},[d]\backslash\{i\}})\big]\nonumber
-  \\
-&=(-1)^{i+j}\det\!\big(\E[\A_{[d]\backslash\{j\},[d]\backslash\{i\}}]\big)
-  = \big(\!\adj(\E[\A])\big)_{i,j}.\label{eq:adj}
-\end{align}
-The adjugate is useful in our analysis because it connects the
-determinant and the inverse via the formula
-$\adj(\A)=\det(\A)\A^{-1}$, which holds for any invertible $\A$.
 We next give a few simple examples to provide some intuition. First, note
@@
   \begin{align*}
-  \E\big[\det(s\,\Z_{\Ic,\Jc})\big] &= \E[s^r]\det(\Z_{\Ic,\Jc})
+    \E\big[\det(s\,\Z_{\Ic,\Jc})\big] & = \E[s^r]\det(\Z_{\Ic,\Jc})
     =\det\Big(\big(\E[s^r]\big)^{\frac1r}\,\Z_{\Ic,\Jc}\Big),
@@
   If $\A$ and $\B$ are independent and determinant preserving, then:
+  \vspace{-1mm}
   \begin{enumerate}
     \item $\A+\B$ is determinant preserving,
+          \vspace{-2mm}
     \item $\A\B$ is determinant preserving.
@@
   \begin{align*}
-  \E\big[\det(\A^\top\B)\big] &= \det\!\big(\E[\A^\top\B]\big).
+    \E\big[\det(\A^\top\B)\big] & = \det\!\big(\E[\A^\top\B]\big).
   \end{align*}
@@
   \begin{align*}
-  \E\big[\det(\A\B^\top)\big] &= \ee^{-\E[K]}\det\!\big(\I + \E[\B^\top\A]\big).
+    \E\big[\det(\A\B^\top)\big] & = \ee^{-\E[K]}\det\!\big(\I + \E[\B^\top\A]\big).
   \end{align*}
@@
 
-\section{Conclusions and open problems}
+\input{continuous-empirical.tex}
+
+\section{Conclusions}
 \label{s:conclusions}
@@
 We derived exact non-asymptotic expressions for the MSE of the
-Moore-Penrose estimator in the standard regression task, reproducing
+Moore-Penrose estimator in the linear regression task, reproducing
 the double descent phenomenon as the sample size crosses between the
@@
 standard i.i.d.~random design distribution using a determinantal
-rescaling to obtain a surrogate design which admits exact MSE expressions,
+point process to obtain a surrogate design which admits exact MSE expressions,
 while capturing the key properties of the i.i.d.~design. We
@@
 
-\begin{wrapfigure}{r}{\ifisarxiv 0.5\else 0.47\fi\textwidth}
- \ifisarxiv\else \vspace{-.3cm}\fi
-  \centering
- \includegraphics[width=.46\textwidth]{figs/descent-model-log}
-\ifisarxiv\else \vspace{-2mm}\fi
-  \caption{
-Surrogate MSE as a function of $d/n$, with $n$
-fixed to $100$ and varying $d$.}
-  \label{f:model}
-\ifisarxiv\else  \vspace{-.5cm}\fi
-\end{wrapfigure}
-
-An important technical issue is that, in this work, we focus on the classical \emph{well-specified}
-linear regression task, where the underlying response model is linear
-with respect to the observed feature space. A significant effort in
-the related literature (see Section \ref{s:related-work}) has been
-directed towards a number of \emph{misspecified} linear regression
-tasks, where the set of $d$ observed features is different than the
-set of $D$ features which define the linear model (typically, $d\ll
-D$). Crucially, unlike in the well-specified task, here it is possible
-to vary the number of observed features without changing the
-underlying linear model.
-Recent work \citep{HMRT19_TR} has compared how
-varying the feature dimension affects the (asymptotic) generalization
-error for both well-specified and misspecified tasks, however their
-analysis was limited to certain special settings such as an
-isotropic data distribution. As an additional
-point of comparison, in Figure~\ref{f:model} we plot the MSE
-expressions of Theorem~\ref{t:mse} for our well-specified setting when varying the
-feature dimension~$d$. The model is chosen just like in
-Figure~\ref{f:intro}, where the covariances $\Sigmab_\mu$ are diagonal
-with condition number $\kappa$ and exponentially decaying spectrum
-scaled so that $\tr(\Sigmab_\mu^{-1})=d$. We also use $\sigma^2=1$ and
-$\w^*=\frac1{\sqrt{d}}\one$. Qualitatively, our plots follow the
-trends outlined by \cite{HMRT19_TR} for the isotropic case (i.e., $\kappa=1$),
-but the spectral decay of the covariance matrix (captured by our new
-MSE expressions) does have a significant effect on the descent curve
-in the under-determined regime. Note that the plots achieve their
-minimum as $d$ goes to zero because in the well-specified task as
-the complexity of the prediction model decreases, so does the complexity of the
-true response model. Nevertheless,
-we observe generalization in the under-determined regime,
-as seen by the fact that the MSE curve goes below the error of the
-null estimator, $\mathrm{MSE}[\zero]=1$.
-
-Our work opens up a number of new directions for future research.
-This includes extending our surrogate analysis to the
-misspecified linear regression discussed above.
-Also, it remains open whether the analysis we provided for the mean
-squared error can be reproduced in the context of mean squared
-\emph{prediction} error, which is relevant in many machine learning
-tasks. Finally, while Theorem \ref{t:asymptotic} states that our surrogate
-expressions for the MSE under certain Gaussian designs are
-asymptotically consistent with the multiplicative error rate of
-$O(1/d)$, we believe that this fact extends to
-the setting not covered by the theorem: under-determined regime (i.e.,
-$n<d$) with a non-isotropic Gaussian distribution (i.e., $\Sigmab\neq
-\I$). We break down our analysis into verifying two
-conjectures which are of independent interest to multivariate
-Gaussian analysis. The first conjecture addresses the first term in
-the MSE derivation \eqref{eq:mse-derivation} and postulates an asymptotically
-consistent formula for the expected Moore-Penrose inverse
-of the pseudo-Wishart distribution. Matrix
-$\W\sim\Pc\Wc(\Sigmab,n)$ is distributed according to the pseudo-Wishart
-distribution with $n<d$ degrees of freedom \citep{srivastava1979introduction} if it can be written as
-$\W=\X^\top\X$, where $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Sigmab)$
-is the matrix-variate Normal \citep{gupta2018matrix}, i.e., an
-i.i.d.~Gaussian design.
-% we call it a pseudo-Wishart 
-% and write $\W \sim \Pc\Wc(\Sigmab, n)$
-% if $n<d$ but still
-\ifisarxiv\pagebreak\fi
-\begin{conjecture}[Moore-Penrose inverse of pseudo-Wishart]\label{c:wishart}
-  Fix $n/d<1$ and let $\W\sim\Pc\Wc(\Sigmab,n )$, where
-$\Sigmab$ is $d\times d$ positive definite with condition number bounded by a constant.
-Then:
-\begin{align}
-\bigg|\frac{\E\big[\tr(\W^\dagger)\big]}{\Vc(\Sigmab,n)} -1\bigg|=
-  O(1/d)\qquad\text{for}\quad\Vc(\Sigmab,n)=\frac{1-\alpha_n}{\lambda_n},\label{eq:wishart}
-\end{align}
-where $\lambda_n\geq 0$ satisfies $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$ and
-$\alpha_n=\det(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$.
-\end{conjecture}
-
-\noindent
-Our second conjecture involves the projection onto the
-orthogonal complement of a Gaussian sample $\X$, i.e., the matrix
-$\I-\X^\dagger\X$, and addresses
-the second term in the MSE derivation \eqref{eq:mse-derivation}.
-\begin{conjecture}[Gaussian orthogonal projection]\label{c:projection}
-  Fix $n/d<1$ and let $\X\sim \Nc_{n,d}(\zero,\I_n \otimes \Sigmab)$, where
-$\Sigmab$ is $d\times d$ positive definite with condition number
-bounded by a constant. Then:
-\begin{align}
-\sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top
-  \Bc(\Sigmab,n)\w} - 1\bigg| = O(1/d)\qquad\text{for}\quad
-  \Bc(\Sigmab,n) =
-  \lambda_n(\Sigmab+\lambda_n\I)^{-1}.\label{eq:projection}
-\end{align}
-\end{conjecture}
-% Note that the surrogate expression for the mean squared
-% error can be written as:
-% \[
-% \Mc(\Sigmab_\mu,\w^*,\sigma^2,n)=\sigma^2 \Vc(\Sigmab_\mu,n)
-% + \w^{*\top}\Bc(\Sigmab_\mu,n)\w^*.
-% \]
-% So, if the conjectures are true,
-% this would immediately imply that the asymptotic consistency claim
-% given in Theorem \ref{t:asymptotic} for
-% the surrogate Gaussian MSE extends to
-% the under-determined setting with arbitrary covariance $\Sigmab$.
-
-\noindent
-Recall that $\lambda_n=\frac {d-n}{\tr((\Sigmab+\lambda_n\I)^{-1})}$,
-so our surrogate MSE is recovered as
-$\sigma^2\Vc(\Sigmab,n)+\w^{*\top}\Bc(\Sigmab,n)\w^*$. 
-% The expressions $\Vc(\Sigmab,n)$ and $\Bc(\Sigmab,n)$ and recover
-% surrogate MSE from Theorem \ref{t:mse} since $\lambda_n=\frac{\tr((\Sigmab+\lambda_n)^{-1})}{d-n}$.
-Conjectures \ref{c:wishart} and \ref{c:projection} provide new insights into classical matrix-variate
-distributions with extensive literature dedicated to them \citep[see,
-e.g.,][]{chikuse1990matrix,cook2011}. We discuss 
-this further in Appendix \ref{sec:asymp-conj-details},
-where we also provide detailed empirical evidence supporting these claims.
-
-% In Appendix \ref{sec:asymp-conj-details} we provide further discussion
-% and detailed empirical evidence for these conejctures.
-% We believe that addressing these conjecture will have a broader
-% impact on our understanding of non-isotropic Gaussian designs.
-
-% In addition, in Section \ref{s:asymptotic}, we provided two conjectures, the goal of which is to bound the difference between the results obtained for our surrogate designs and the true values corresponding to the i.i.d.~design, in the case of multivariate Gaussians.
+% Our work opens up a number of new directions for future research.
+% This includes extending our surrogate analysis to the
+% misspecified linear regression discussed above.
+% Also, it remains open whether the analysis we provided for the mean
+% squared error can be reproduced in the context of mean squared
+% \emph{prediction} error, which is relevant in many machine learning
+% tasks. Finally, while Theorem \ref{t:asymptotic} states that our surrogate
+% expressions for the MSE under certain Gaussian designs are
+% asymptotically consistent with the multiplicative error rate of
+% $O(1/d)$, we believe that this fact extends to
+% the setting not covered by the theorem: under-determined regime (i.e.,
+% $n<d$) with a non-isotropic Gaussian distribution (i.e., $\Sigmab\neq
+% \I$). We break down our analysis into verifying two
+% conjectures which are of independent interest to multivariate
+% Gaussian analysis.
+\ifisarxiv\else
+  \section*{Broader Impact}
+
+  While the double descent phenomenon has been empirically observed in a variety
+  of applications, mathematical descriptions of it are oftentimes complex and
+  inaccesible to non-experts. In contrast, our surrogate design and the
+  accompanying closed-form expressions provide an easily computable rule of thumb
+  for estimating the generalization error. One important application is
+  the high-dimensional (i.e. underdetermined, $n < d$) regime
+  experienced by modern machine learning systems where the number of
+  parameters vastly exceeds the quantity of
+  available data. Our research can be applied here to provide a theoretical
+  understanding of the surprising phenomenon where without proper
+  regularization \citep{nakkiran2020optimal} more data (i.e. increasing $n$)
+  may lead to worse generalization performance.
+
+  Better theoretical understanding of generalization error in the small data
+  regime has important societal impact. Through theoretically modeling a system's
+  performance, we can build safer systems by better understanding how badly an
+  estimator fails and accounting for these failure modes in the system's design.
+  Furthermore, the improved understanding of minimum norm solutions performing
+  worse with more data offers an appealing trade-off where certain systems can
+  both improve their performance and respect the privacy of its users by
+  collecting less data.
+\fi
 
@@
 We would like to acknowledge ARO, DARPA, NSF, ONR, and GFSD for providing
-partial support of this work.
-
-\ifisarxiv\bibliographystyle{plainnat}\fi
+partial support of this work. We also thank Zhenyu Liao for pointing
+out fruitful connections between our results and the asymptotic
+analysis of random matrix resolvents.
+
+\ifisarxiv
+  \bibliographystyle{plainnat}
+\else
+  \bibliographystyle{alpha}
+\fi
+
 \bibliography{../pap}
 
+\ifisarxiv\else\newpage\fi
+
 \appendix

### 2022-08-23 design/continuous_topml/appendix.tex
Commit original repos

--- 2022-08-23 design/continuous_colt/appendix.tex
@@
 
+We use $\adj(\A)$ to denote the adjugate of $\A$, defined as follows: the
+$(i,j)$th entry of $\adj(\A)$ is
+$(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$.
+We will use two useful identities related to the adjugate: (1)
+$\adj(\A)=\det(\A)\A^{-1}$ for invertible $\A$, and (2)
+$\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$
+\citep[see Fact 2.14.2 in][]{matrix-mathematics}.
+
+First, note that from the definition of an adjugate matrix it immediately follows that if $\A$ is
+determinant preserving then adjugate commutes with expectation for this matrix:
+\begin{align}
+  \E\big[\big(\!\adj(\A)\big)_{i,j}\big] &=
+  \E\big[(-1)^{i+j}\det(\A_{[d]\backslash\{j\},[d]\backslash\{i\}})\big]\nonumber
+  \\
+&=(-1)^{i+j}\det\!\big(\E[\A_{[d]\backslash\{j\},[d]\backslash\{i\}}]\big)
+  \\
+  &= \big(\!\adj(\E[\A])\big)_{i,j}.\label{eq:adj}
+\end{align}
+% The adjugate is useful in our analysis because it connects the
+% determinant and the inverse via the formula
+% $\adj(\A)=\det(\A)\A^{-1}$, which holds for any invertible $\A$.
+
 \begin{proofof}{Lemma}{\ref{t:ring}} \
  First, we show that $\A+\u\v^\top$ is d.p.~for any fixed
- $\u,\v\in\R^d$. Below, we use a standard identity for the rank one
+ $\u,\v\in\R^d$. Below, we use the identity for a rank one
  update of a determinant:
@@
 \cite{correcting-bias}.
-\begin{lemma}[\citeauthor{correcting-bias}, \citeyear{correcting-bias}]\label{l:cb}
+\begin{lemma}[\cite{correcting-bias}]\label{l:cb}
 If the rows of random $k\times d$ matrices $\A,\B$
@@
 constant of a discrete determinantal point process.
-\begin{lemma}[\citeauthor{dpp-ml}, \citeyear{dpp-ml}]\label{l:det-standard}
+\begin{lemma}[\cite{dpp-ml}]\label{l:det-standard}
   For any $k\times d$ matrices $\A,\B$ we have
@@
 $S_\mu^n$.
-We first prove Lemma \ref{l:sqinv-all} by splitting the under- and
-over-determined cases and start with proving the former.
+We first prove Lemma \ref{l:sqinv-all}. % by splitting the under- and
+% over-determined cases and start with proving the former.
 % Note that those
@@
 % which implies that if $\X\sim\mu^k$ for $k\leq d$ then $\rank(\X)=k$.
-\begin{lemma}\label{l:sqinv-under}
+\begin{lemma}[restated Lemma \ref{l:sqinv-all}]\label{l:sqinv-under}
 If  $\Xb\sim S_\mu^n$ for $n<d$, then we have
@@
 Applying the closed form expressions from Lemmas
-\ref{l:proj} and \ref{l:sqinv-all}, we derive
+\ref{l:proj}, \ref{l:sqinv-all} and \ref{l:sqinv-over}, we derive
 the formula for the MSE and prove Theorem \ref{t:mse} (we defer the
@@
 \label{sec:proof-of-t-asymptotic}
-Once again, we break down the proof into under- and
-over-determined cases, starting with the former. Note that in this
-case we require that the covariance be equal to identity.
-\begin{lemma}\label{l:asymptotic-under}
-  Let $\rho=n/d$, $\X\sim \Nc_{n,d}(\zero,\I_n \otimes \I_d)$ and $y_i=y(\x_i)$ under Assumption
-  \ref{a:linear}. If $d>n+1$ then
-  \begin{align*}
-    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\I,
-\w^*,\sigma^2,n)}{\Mc(\I,\w^*,\sigma^2,n)}\ \leq\
-\frac1d\cdot \frac{1}{1-\rho-\frac1d} + 3\rho^d.
-  \end{align*}
+
+The proof of Theorem \ref{t:asymptotic} follows the standard decomposition of
+MSE in Equation~\ref{eq:mse-derivation}, and in the process,
+establishes consistency of the variance and bias terms
+independently. To this end, we introduce the following two useful
+lemmas that capture the limiting behavior of the 
+variance and bias terms, respectively.
+
+\begin{lemma}\label{c:wishart}
+  Under the setting of Theorem~\ref{t:asymptotic}, we have, as $n,d \to \infty$
+with $n/d \to \bar c \in (0, \infty) \setminus \{ 1 \}$ that
+\begin{align}
+%\bigg|\frac{\E\big[\tr((\X^\top \X)^\dagger)\big]}{\Vc(\Sigmab,n)} -1\bigg| \to 0 \qquad\text{for}\quad\Vc(\Sigmab,n)= \frac{1-\alpha_n}{\lambda_n}
+\begin{cases}
+  \E\big[\tr((\X^\top \X)^\dagger)\big] - (1-\alpha_n) \lambda_n^{-1} \to 0, & \text{for }\bar c < 1, \\
+  \E\big[\tr((\X^\top \X)^\dagger)\big] - \frac{1-\beta_n}{n - d} \cdot \tr \Sigmab^{-1} \to 0, & \text{for }\bar c > 1
+\end{cases}
+\label{eq:wishart}
+\end{align}
+where $\lambda_n\geq 0$ is the unique solution to
+$n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$,
+$\alpha_n=\det(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$,
+and $\beta_n = e^{d-n}$.
 \end{lemma}
 
-\begin{proof}
-  We first recall the standard decomposition of $\MSE{\X^\dagger \y}$:
+\noindent
+
+The second term in the MSE derivation \eqref{eq:mse-derivation}, $\E[\I -
+\X^\dag \X]$, involves the expectation of a projection onto the orthogonal
+complement of a sub-Gaussian general position sample $\X$.
+This term is zero when $n > d$, and for $n < d$
+we prove in \cref{sec:proof-of-c-projection} that the surrogate design's
+bias $\mathcal{B}(\Sigmab, n)$ provides an asymptotically consistent approximation
+to all of the eigenvectors and eigenvalues:
+
+\begin{lemma}\label{c:projection}
+Under the setting of Theorem~\ref{t:asymptotic}, for $\w \in \R^d$ of bounded Euclidean norm (i.e., $\| \w \| \le C'$ for all $d$), we have, as $n,d \to \infty$ with $n/d \to \bar c \in (0,1)$ that
+\begin{align}
+%\sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top \Bc(\Sigmab,n)\w} - 1\bigg| \to 0 \qquad\text{for}\quad\Bc(\Sigmab,n) = \lambda_n(\Sigmab+\lambda_n\I)^{-1}
+  \w^\top\E[\I-\X^\dagger\X]\w - \lambda_n \w^\top (\Sigmab + \lambda_n \I)^{-1}\w \to 0
+  \label{eq:projection}
+\end{align}
+while $\I - \X^\dagger \X = 0$ for $\bar c > 1$.
+\end{lemma}
+
+\subsection{Proof of \cref{c:wishart}}
+\label{sec:proof-of-c-wishart}
+
+\subsubsection{The $\bar c \in (0,1)$ case}
+
+For $n < d$, we first establish (1) $\liminf_n \lambda_n > 0$ and (2) $\alpha_n \to 0$.
+To prove (1), by hypothesis $\Sigmab \succeq c \I$ for all $d$. Since $\frac{n}{d} < 1$,
+we have (by definition of $\lambda_n$) for some $\delta > 0$
 \begin{align*}
-    \MSE{\X^\dagger \y}
-    &=\sigma^2 \E\big[\tr\big((\X^\top \X)^\dagger \big)\big]+
-      \w^{*\top}\!\big( \I - \E[\X^\dagger \X]\big) \w^*.
+  1 - \delta
+  > \frac{n}{d}
+  = \frac{1}{d} \tr(\Sigmab(\Sigmab + \lambda_n \I)^{-1})
+  > \frac{c}{c + \lambda_n}
 \end{align*}
-  Since the rows of $\X$ are standard normal random variables,
-  $\X\X^\top$ is an $n\times n$ Wishart random matrix with $d>n+1$ degrees
-  of freedom. Using the formula for the mean of the Inverse-Wishart
-  distribution, it follows that
+Rearranging, we have $\lambda_n > \frac{\delta c}{1 - \delta} > 0$.
+For (2), let $(\tau_i)_{i \in [d]}$ denote the eigenvalues of $\Sigmab$.
+Since $1 - x \leq e^{-x}$ and $C \I \succeq \Sigmab \succeq c \I$ for all $d$,
 \begin{align*}
-\E\big[\tr((\X^\top\X)^\dagger)\big] = \E\big[\tr((\X\X^\top)^{-1})\big]= \frac{n}{d - n - 1}.
+  \alpha_n
+  = \prod_{i=1}^d \frac{\tau_i}{\tau_i + \lambda_n}
+  \leq \left(\frac{C}{C + \lambda_n}\right)^d
+  = \left( 1 - \frac{\lambda_n}{C + \lambda_n}\right)^d
+  \leq \exp\left(-d \frac{\lambda_n}{C + \lambda_n}\right)
 \end{align*}
-  % and therefore
+and since $\lambda_n > 0$ eventually as $d \to \infty$ we have
+$\alpha_n \to 0$ so that $(1-\alpha_n) \lambda_n^{-1} - \lambda_n^{-1} \to 0$.
+
+% (1) and (2) together imply $\mathcal{V}(\Sigmab, n) \to \lambda_n^{-1}$, and
+% by Slutsky's theorem
 % \begin{align*}
-  %   \MSE{\X^\dagger \y}
-  %   &= \|\w\|_2^2 - \w^\top \E[\X^\dagger \X] \w
-  %   + \frac{\sigma^2 n}{d - n - 1}
+%   \lim_{n,d} \left\lvert
+%   \frac{\E[\tr((\X^\top \X)^\dag)] - \mathcal{V}(\Sigmab, n)}{\mathcal{V}(\Sigmab, n)}
+%   \right\rvert
+%   = \lim_{n,d} \left\lvert
+%   \frac{\E[\tr((\X^\top \X)^\dag)] - \lambda_n^{-1}}{\lambda_n^{-1}}
+%   \right\rvert
 % \end{align*}
-  % Let $\X = \U \S \V^\top$ be its full singular value decomposition. Notice
+% Since $\liminf_n \lambda_n > 0$ was shown in (1),
+As a consequence of (2) and Slutsky's theorem, it suffices to show
+$\tr(\X^\top \X)^\dag - \lambda_n^{-1} \overset{d}{\to} 0$ as $n,d \to \infty$.
+To do this, we consider the limiting behavior of $\tr
+(\X^\top \X)^\dagger/n = \tr (\X \X^\top)^\dagger/n$ as $n/d \to \bar c \in
+(0,1)$, for $\X = \Z \Sigmab^{\frac12}$ with $\Z \in \mathbb R^{n \times d}$
+having i.i.d.~zero mean, unit variance sub-Gaussian entries, i.e., the behavior
+of
+\begin{equation}\label{eq:limit1}
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left( \frac1n \X \X^\top + z \I_n \right)^{-1}
+\end{equation}
+by definition of the pseudo-inverse.
+
+The proof comes in three steps: (i) for fixed $z > 0$, consider the limiting behavior of $ \delta(z) \equiv \tr (\X \X^\top/n + z \I_n)^{-1}/n$ as $n,d \to \infty$ and state
+\begin{equation}\label{eq:limit2}
+  \lim_{n,d \to \infty} \delta(z) - m(z) \to 0
+\end{equation}
+almost surely for some $m(z)$ to be defined; (ii) show that both $\delta(z)$ and its derivate $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take $z \to 0^+$ in \eqref{eq:limit2} and state
+\begin{equation}\label{eq:limit3}
+  \lim_{z \to 0^+} \lim_{n,d \to \infty} \delta(z) - \lim_{z \to 0^+} m(z) \to 0
+\end{equation}
+almost surely, given that the limit $\lim_{z \to 0^+} m(z) \equiv m(0)$ exists and eventually (iii) exchange the two limits in \eqref{eq:limit3} with Moore-Osgood theorem, to reach
+\[
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left(\frac1n \X \X^\top + z \I_n \right)^{-1} - m(0) \to 0.
+\]
+
+Step (i) follows from \cite{silverstein1995empirical} that, we have, for $z > 0$ that
+\[
+  \delta(z) \equiv \frac1n \tr \left( \frac1n \X \X^\top  + z \I_n \right)^{-1}  - m(z) \to 0
+\]
+almost surely as $n,d \to \infty$, for $m(z)$ the unique positive solution to
+\begin{equation}\label{eq:def-m}
+  m(z) = \left( z + \frac1n \tr \Sigmab (\I + m(z) \Sigmab)^{-1} \right)^{-1}.
+\end{equation}
+
+For the above step (ii), we use the assumption $\Sigmab \succeq c \I \succ 0$
+for all $d$ large, so that with $\X = \Z \Sigmab^{\frac12}$, we have for large enough $n,d$ that
+\[
+  \lambda_{\min} (\X \X^\top/n) \ge \lambda_{\min} (\Z \Z^\top/n) \lambda_{\min} (\Sigmab) \ge \frac{c}2 (\sqrt{\bar c} - 1 )^2
+\]
+almost surely, where we used Bai-Yin theorem \cite{bai1993limit}, which states that the minimum eigenvalue of $\Z \Z^\top/n$ is almost surely larger than $( \sqrt{\bar c} - 1 )^2/2$ for $n<d$ sufficiently large.
+Note that here the case $\bar c = 1$ is excluded.
+
+Observe that
+\[
+  |\delta(z)| = \left| \frac1n \tr \left( \frac1n \X \X^\top  + z \I_n \right)^{-1} \right| \le \frac1{ \lambda_{\min} (\X \X^\top/n) }
+\]
+and similarly for its derivative, so that we are allowed to take the $z \to 0^+$ limit.
+Note that the existence of the $\lim_{z \to 0^+} m(z)$ for $m(z)$ defined in \eqref{eq:def-m} is well known, see for example \cite{ledoit2011eigenvectors}.
+Then, by Moore-Osgood theorem we finish step (iii) and by concluding that
+\[
+  \tr (\X^\top \X)^\dagger - m(0) \to 0
+\]
+for $m(0) = \lambda_n^{-1}$ the unique solution to $\lambda_n^{-1} = \left( \frac1n \tr \Sigmab (\I + \lambda_n^{-1} \Sigmab)^{-1} \right)^{-1}$, or equivalently, to
+\[
+  n = \tr \Sigmab (\Sigmab + \lambda_n \I)^{-1}
+\]
+as desired.
+
+\subsubsection{The $\bar c \in (1, \infty)$ case}
+
+First note that as $n, d \to \infty$ with $n > d$, we have $\beta_n = e^{d-n} \to 0$ and it
+it suffices to show
+\[
+  \tr (\X^\top \X)^\dagger - \frac1{n-d} \tr \Sigmab^{-1} \to 0
+\]
+almost surely to conclude the proof.
+
+In the $\bar c \in (1, \infty)$ case, it is more convenient to work on the following co-resolvent
+\[
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left( \frac1n \X^\top \X + z \I_d \right)^{-1}
+\]
+where we recall $\X^\top \X = \Sigmab^{\frac12} \Z^\top \Z \Sigmab^{\frac12} \in \R^{d \times d}$ and following the same three-step procedure as in the $\bar c < 1$ case above.
+The only difference is in step (i) we need to assess the asymptotic behavior of $\delta \equiv \tr (\X^\top \X/n + z \I_d)^{-1}/n$.
+This was established in \cite{bai1998no} where it was shown that, for $z > 0$ we have
+\[
+  \frac1n \tr (\X^\top \X/n + z \I_d)^{-1} - \frac{d}n m(z) \to 0
+\]
+almost surely as $n,d \to \infty$, for $m(z)$ the unique solution to
+\[
+  m(z) = \frac1d \tr \left( \left(1- \frac{d}n - \frac{d}n z m(z) \right) \Sigmab - z \I_d \right)^{-1}
+\]
+so that for $d < n$ by taking $z = 0$ we have
+\[
+  m(0) = \frac{n}d \frac1{n-d} \tr \Sigmab^{-1}.
+\]
+The steps (ii) and (iii) follow exactly the same line of arguments as the $\bar c < 1$ case and are thus omitted.
+
+% {\RED TODO: proof for $n>d$ sub-Gaussians. We can't appeal
+% to inverse Wishart mean anymore}
+% Now consider $n > d$ with $n/d \to \bar c \in (1, \infty)$.  Firstly notice
+% $\beta_n = e^{d - n} \to 0$, so $\lim \mathcal{V}(\Sigmab, n) = \lim
+% \frac{1}{n-d} \tr(\Sigmab^{-1})$. Noting
+% $\lim \frac{1}{n-d} \tr(\Sigmab^{-1}) = \int \tau^{-1} dH(\tau)$ (where $H$ is the
+% limiting spectral distribution of $\Sigmab$) is a constant, by Slutsky's theorem
+% it suffices to show
+% \[
+%   \tr (\X^\top \X)^\dagger - \frac1{n-d} \tr \Sigmab^{-1} \to 0
+% \]
+% Recognizing $(\X^\top \X)^\dag = (\X^\top\X)^{-1}$ as an inverse-Wishart
+% matrix when $n > d$, we have that the left-hand side is identically zero.
+
+\subsection{Proof of \cref{c:projection}}
+\label{sec:proof-of-c-projection}
+
+Since $\X^\dagger \X = \X^\top (\X \X^\top)^\dagger \X$, to prove
+\cref{c:projection}, we are interested in the limiting behavior of the
+following quadratic form
+\[
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w
+\]
+for deterministic $\w \in \mathbb R^{d}$ of bounded Euclidean norm (i.e., $\| \w \| \le C'$ as $n,d \to \infty$), as $n,d \to \infty$ with $n/d \to \bar c \in (0,1)$.
+The limiting behavior of the above quadratic form, or more generally, bilinear form of the type $\frac1n \w_1^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w_2$ for $\w_1, \w_2 \in \mathbb R^{d}$ of bounded Euclidean norm are widely studied in random matrix literature, see for example \cite{hachem2013bilinear}.
+
+For the proof of \Cref{c:projection} we follow the same protocol as that of
+\Cref{c:wishart}, namely: (i) we consider, for fixed $z > 0$, the limiting
+behavior of $\frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n
+\right)^{-1} \X \w$. Note that
+\begin{align*}
+  \delta(z) &\equiv \frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w = \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \frac1n \X^\top \X \w \\
+  &= \| \w \|^2 - z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w
+\end{align*}
+and it remains to work on the second $z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w$ term.
+It follows from \cite{hachem2013bilinear} that
+\[
+   %\delta(z) = \| \w \|^2 - z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w  \to \| \w \|^2 - \w^\top (\I_d + m(z) \Sigmab)^{-1} \w^\top
+   z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w - \w^\top (\I_d + m(z) \Sigmab)^{-1} \w^\top \to 0
+\]
+almost surely as $n,d \to \infty$, where we recall $m(z)$ is the unique solution to \eqref{eq:def-m}.
+
+We move on to step (ii), under the assumption that $c \le \lambda_{\min} (\Sigmab) \le \lambda_{\max} (\Sigmab) \le C$ and $\| \w \| \le C'$, we have
+\begin{align*}
+  \lambda_{\max} \left( \frac1n \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \right) &\le \frac{ \lambda_{\max} (\X \X^\top/n) }{ \lambda_{\min} (\X \X^\top/n) + z } \le \frac{ \lambda_{\max} (\Z \Z^\top/n) \lambda_{\max} (\Sigmab) }{ \lambda_{\min} (\Z \Z^\top/n) \lambda_{\min} (\Sigmab) } \\
+  &\le 4 \frac{ (\sqrt{\bar c} + 1)^2 C }{ (\sqrt{\bar c} -1)^2 c }
+\end{align*}
+so that $\delta(z)$ remains bounded and similarly for its derivative
+$\delta'(z)$, which, by Arzela-Ascoli theorem, yields uniform
+convergence and we are allowed to take the $z \to 0^+$
+limit. Ultimately, in step (iii) we exchange the two limits with
+Moore-Osgood theorem, concluding the proof.
+
+\subsection{Finishing the proof of Theorem~\ref{t:asymptotic}}
+
+To finish the proof of Theorem~\ref{t:asymptotic}, it remains to write
+\[
+  \MSE{\X^\dagger\y} = \sigma^2\E\big[\tr\big((\X^\top\X)^{\dagger}\big)\big] +
+    \w^{*\top}\E\big[\I-\X^\dagger\X\big]\w^*
+\]
+Since $\lambda_n = \frac{d-n}{ \tr (\Sigmab + \lambda_n \I)^{-1} }$,
+by \Cref{c:wishart} and \Cref{c:projection} we have
+$\MSE{\X^\dagger\y} - \Mc(\Sigmab, \w^*, \sigma^2, n) \to 0$ as $n,d
+\to \infty$ with $n/d \to \bar c \in (0,\infty) \setminus \{ 1 \}$,
+which concludes the proof of Theorem~\ref{t:asymptotic}.
 % \begin{align*}
-  %   \X^\dagger \X
-  %   &= \V \S^\dagger \U^\top \U \S \V^\top \\
-  %   &= \V \S^\dagger \S \V^\top \\
-  %   &= \V \begin{bmatrix}
-  %     \I_n & 0 \\
-  %     0 & 0
-  %   \end{bmatrix} \V^\top \\
-  %   &= \V_n \V_n^\top \\
-  %   \w^\top \E[\X^\dagger \X] \w
-  %   &= \E[\|\V_n^\top \w\|_2^2]
+%   &\MSE{\X^\dagger\y} - \Mc(\Sigmab, \w^*, \sigma^2, n) \\
+%   &=
+%     \underbrace{\left(
+%       \sigma^2\E\big[\tr\big((\X^\top\X)^{\dagger}\big)\big]
+%       - \frac{\sigma^2(1 - \alpha_n)}{\lambda_n}
+%     \right)}_{\to 0} + \underbrace{\left(
+%       \w^{*\top}\E\big[\I-\X^\dagger\X\big]\w^*
+%       - \lambda_n \w^{*\top} (\Sigmab + \lambda_n \I)^{-1} \w^*
+%     \right)}_{\to 0}
 % \end{align*}
-  % where $\V_n$ denotes the $d \times n$ matrix with columns equal to the
-  % $n$ eigenvectors of $\X^\top \X$ with non-zero eigenvalue.
-  % But since $\Sigmab = \I$, by rotation invariance
-  % $\V_n^\top$ is a change of basis onto a uniformly random
-Note that $\X^\dagger\X$ is a uniformly random projection matrix so by
-rotational symmetry it follows that
-\begin{align*}
-  \w^{*\top}\E\big[\X^\dagger\X]\w^* =
-  \E\big[\|\X^\dagger\X\w^*\|^2\big] = \frac nd \,\|\w^*\|^2.
-\end{align*}
-Putting this together we obtain that
-  \begin{align*}
-    \MSE{\X^\dagger \y}
-    &=\frac{\sigma^2 n}{d - n - 1}+ \|\w^*\|^2\,\frac{d - n}{d}.
-  \end{align*}
-  On the other hand, the surrogate MSE expression can be derived by
-observing that for $\Sigmab=\I$ we have
-$\tr((\Sigmab+\lambda_n\I)^{-1})=d/(1+\lambda_n)=n$ (see definition of
-$\lambda_n$ in Theorem \ref{t:mse}):
-  \begin{align*}
-    \Mc(\I, \w^*,\sigma^2,n) =
-    \sigma^2n\cdot\frac{1-\alpha_n}{d-n} + \|\w^*\|^2\,\frac{d-n}{d}.
-  \end{align*}
-  Note that the second term is the same in both cases, even though
-  this may not be true for non-isotropic Gaussians. We now compute the
-  normalized difference between the
-  expressions,
-  \begin{align*}
- \frac{\MSE{\X^\dagger\y}-\Mc(\I, \w^*,\sigma^2,n)}{\Mc(\I,
-    \w^*,\sigma^2,n)}
-    &= \frac{\sigma^2n\cdot(\frac1{d-n-1} -
-    \frac{1-\alpha_n}{d-n})}{\sigma^2n\cdot\frac{1-\alpha_n}{d-n} +
-    \|\w^*\|^2\,\frac{d-n}{d}}\\
-&\leq \frac {d-n}{1-\alpha_n}
-\Big(\frac1{d-n-1}-\frac{1-\alpha_n}{d-n}\Big)\\
-    &=\frac{d-n}{d-n-1}\,\frac{1}{1-\alpha_n}- 1\\
-    &=\frac1{d-n-1} + \frac{\alpha_n}{1-\alpha_n}\Big(1+\frac1{d-n-1}\Big).
-  \end{align*}
-Let $\rho=n/d$. Then $\frac1{d-n-1} = \frac1d\cdot
-\frac1{1-\rho-\frac1d}$ and moreover
-$\alpha_n = (\frac1{1+\lambda_n})^d=(\frac
-nd)^d=\rho^d$. From the assumption that $d>n+1$, we
-conclude that $\alpha_n \leq (\frac{d-2}{d})^d\leq \ee^{-2}$
-so that $\frac{\alpha_n}{1-\alpha_n}(1+\frac1{d-n-1})\leq
-  \frac{2 \alpha_n}{1-\ee^{-2}}\leq 3\rho^d$. This shows the right-hand-side inequality of the
-  theorem. That fact that $\MSE{\X^\dagger\y}\geq\Mc(\I,
-  \w^*,\sigma^2,n)$ follows easily.
-\end{proof}
-\begin{lemma}\label{l:asymptotic-over}
-  Let $\rho=n/d$, $\X\sim \Nc_{n,d}(\zero,\I_n \otimes \Sigmab)$ and $y_i=y(\x_i)$ under Assumption
-  \ref{a:linear}. If $n>d+1$ then
-  \begin{align*}
-    &0\ \leq\ \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab,
-\w^*,\sigma^2,n)}{\Mc(\Sigmab,\w^*,\sigma^2,n)}\ \leq\
-\frac1d\cdot \frac{1}{\rho-1-\frac1d} + 3(\ee^{1-\rho})^d.
-  \end{align*}
-\end{lemma}
-\begin{proof}
-The MSE for the over-determined Gaussian design can be derived by
-using the formula for the mean of the Inverse-Wishart distribution:
-\begin{align*}
-  \MSE{\X^\dagger\y}=\sigma^2\tr\big(\E[(\X^\top\X)^{-1}]\big) =
-  \frac{\sigma^2\tr(\Sigmab^{-1}) }{n-d-1}.
-\end{align*}
-To compute the normalized difference we follow similar derivations as
-in the proof of Lemma \ref{l:asymptotic-under}:
-\begin{align*}
- \frac{\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)}{\Mc(\Sigmab,
-    \w^*,\sigma^2,n)}
-    &=  \frac {n-d}{1-\beta_n}
-      \Big(\frac1{n-d-1}-\frac{1-\beta_n}{n-d}\Big)\\
-  &\leq\frac1d\cdot\frac1{\rho-1-\frac1d} + \frac{2\beta_n}{1-\beta_n}.
-\end{align*}
-Recall that $\beta_n = \ee^{d-n} = (\ee^{1-\rho})^d$ and for $n-d\geq
-2$ we have $\frac2{1-\beta_n}\leq 3$. The desired inequalities follow immediately.
-\end{proof}
-Theorem \ref{t:asymptotic} now follows as a consequence of
-Lemmas \ref{l:asymptotic-under} and \ref{l:asymptotic-over}.
-\begin{proofof}{Theorem}{\ref{t:asymptotic}}
-  Since $\frac1d\leq \frac13|1-\rho|$ and $\ee^{-|1-\rho|d}\leq
-  \frac1{2d|1-\rho|}$, it follows that
-  \begin{align*}
-    \frac1d\cdot \frac{1}{|1-\rho|-\frac1d} + 3(\ee^{-|1-\rho|})^d
-    &\leq \frac1d\cdot \frac{1}{|1-\rho|-\frac13|1-\rho|} +
-      \frac1d\cdot\frac{3}{2|1-\rho|}=\frac {c_\rho}{d}.
-  \end{align*}
-  The case of $n>d+1$ now follows from Lemma
-  \ref{l:asymptotic-over}. Also, since $\rho\leq \ee^{\rho-1}$, for
-$n<d-1$ we have $3\rho^d\leq 3(\ee^{-|1-\rho|})^{d}$, so the same bound follows
-from Lemma \ref{l:asymptotic-under}.
-\end{proofof}
+%as $n,d \to \infty$ with $n/d \to \bar c \in (0,1)$.
+%This allows us to conclude the proof of
 
-\section{Empirical analysis of the asymptotic consistency conjectures}
-\label{sec:asymp-conj-details}
+% For
+% the case $n <d$, note that $\X^\dagger \X = \I$ and
+% \[
+%   \MSE{\X^\dagger\y} - \frac1{n-d} \tr (\Sigmab^{-1}) \to 0.
+% \]
 
-Conjectures \ref{c:wishart} and \ref{c:projection} are related to open problems
-which have been extensively studied in the literature.
-With respect to Conjecture~\ref{c:wishart}, \cite{srivastava2003} first derived the probability
-density function of the pseudo-Wishart
-distribution (also called the singular Wishart), and \cite{cook2011}
-computed the first and second moments of 
-generalized inverses for the distribution. However, for the
-Moore-Penrose inverse and arbitrary covariance $\Sigmab$,
-\cite{cook2011} claims that the quantities required
-to express the mean ``do not have tractable closed-form representation.''
-%
-Conjecture~\ref{c:projection} has connections to directional statistics.
-Using the SVD, we have the equivalent representation $\X^\dagger \X = \V \V^\top$
-where $\V$ is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal
-$n$-frames in $\R^d$).
-The distribution of $\V$ is known as the matrix angular central
-Gaussian (MACG) distribution \citep{chikuse1990matrix}. While prior work
-has considered high dimensional limit theorems \citep{CHIKUSE1991145}
-as well as density estimation and hypothesis testing \citep{CHIKUSE1998188}
-on $V_{n,d}$, they only analyzed the invariant measure
-(which corresponds in our setting to $\Sigmab = \I$),
-and to our knowledge a closed form expression of $\E[\V\V^\top]$ where
-$\V$ is distributed according to MACG with
-arbitrary $\Sigmab$ remains an open question.
+\section{Additional details for empirical evaluation}
+\label{a:empirical}
 
-For verifying these two conjectures, it suffices to only consider
-diagonal covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D
-\Q^\top$ is its eigendecomposition and $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Q\D\Q^\top)$, then
-we have for $\W \sim \Pc\Wc(\Sigmab, n)$ that $\W
-\overset{d}{=} \X^\top \X$ and hence, defining
-$\Xt\sim\Nc_{n,d}(\zero,\I_n \otimes \D)$, by linearity and unitary invariance of trace,
+Our empirical investigation of the rate of asymptotic convergence in \Cref{t:asymptotic} (and, more
+specifically, the variance and bias discrepancies defined in
+Section~\ref{sec:asymp-conj-details}), in the context of
+Gaussian random matrices, is related to open problems which have been
+extensively studied in the literature. Note that when
+$\X=\Z\Sigmab^{1/2}$ were $\Z$ has i.i.d.~Gaussian entries (as in
+Section \ref{sec:asymp-conj-details}), then $\W=\X^\top\X$ is known as the
+pseudo-Wishart distribution (also called the singular Wishart),
+denoted as $\W \sim \Pc\Wc(\Sigmab, n)$, and the variance
+term from the MSE can be written as $\sigma^2\E[\tr(\W^\dagger)]$.
+\cite{srivastava2003} first derived the probability density function of the
+pseudo-Wishart distribution, and
+\cite{cook2011} computed the first and second moments of generalized
+inverses. However, for the Moore-Penrose inverse and arbitrary 
+covariance $\Sigmab$, \cite{cook2011} claims that the quantities required to
+express the mean ``do not have tractable closed-form representation.''
+The bias term, $\w^{*\top}\E[\I-\X^\dagger\X]\w^*$, has connections to
+directional statistics.  Using the SVD, 
+we have the equivalent representation $\X^\dagger \X = \V \V^\top$ where $\V$
+is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal $n$-frames
+in $\R^d$).  The distribution of $\V$ is known as the matrix angular central
+Gaussian (MACG) distribution \citep{chikuse1990matrix}. While prior work has
+considered high dimensional limit theorems \citep{CHIKUSE1991145} as well as
+density estimation and hypothesis testing \citep{CHIKUSE1998188} on $V_{n,d}$,
+they only analyzed the invariant measure (which corresponds in our setting to
+$\Sigmab = \I$), and to our knowledge a closed form expression of
+$\E[\V\V^\top]$ where $\V$ is distributed according to MACG with arbitrary
+$\Sigmab$ remains an open question.
+
+For analyzing the rate of decay of variance and bias discrepancies (as
+defined in Section \ref{sec:asymp-conj-details}), it suffices to only consider diagonal
+covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D \Q^\top$ is
+its eigendecomposition and $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Q\D\Q^\top)$,
+then we have for $\W \sim \Pc\Wc(\Sigmab, n)$ that $\W \overset{d}{=} \X^\top
+\X$ and hence, defining $\Xt\sim\Nc_{n,d}(\zero,\I_n \otimes \D)$, by linearity
+and unitary invariance of trace,
 \begin{align*}
@@
 \end{align*}
-Similarly, we have that
-$\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$, and a simple
-calculation shows that the expression in Conjecture~\ref{c:projection}
-is also independent of the choice of matrix $\Q$.
+Similarly, we have that $\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$,
+and a simple calculation shows that the bias discrepancy is
+also independent of the choice of matrix $\Q$.
 
-\input{continuous-empirical.tex}
+In our experiments, we increase $d$ while keeping the aspect ratio $n/d$
+fixed and examining the rate of decay of the discrepancies.
+We estimate $\E\big[\tr(\W^\dagger)\big]$ (for the variance) and
+$\E[\I-\X^\dagger\X]$ (for the bias) through Monte Carlo sampling.
+Confidence intervals are constructed using ordinary bootstrapping for
+the variance. We rewrite the supremum over $\w$ in bias discrepancy as
+a spectral norm: 
+\[\big\|\Bc(\Sigmab,n)^{-\frac12}\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-\frac12} -
+  \I\big\|,\]
+and apply existing methods for constructing bootstrapped operator
+norm confidence intervals described in \cite{lopes2019bootstrapping}.  To
+ensure that estimation noise is sufficiently small, we continually increase the
+number of Monte Carlo samples until the bootstrap confidence intervals are
+within $\pm 12.5\%$ of the measured discrepancies.  We found that while
+variance discrepancy required a relatively small number of trials (up
+to one thousand), estimation noise was much larger for the bias
+discrepancy, and it necessitated over two million trials to obtain
+good estimates near $d=100$.
+
+\subsection{Eigenvalue decay profiles}
+\label{sec:eig-decay-details}
+
+Letting $\lambda_i(\Sigmab)$ be the $i$th largest eigenvalue of
+$\Sigmab$, we consider the
+following eigenvalue profiles (visualized in Figure~\ref{fig:eig-decays}):
+\begin{itemize}
+  \item \texttt{diag\_linear}: linear decay, $\lambda_i(\Sigmab)= b-a i$;
+  \item \texttt{diag\_exp}: exponential decay, $\lambda_i(\Sigmab) = b\,10^{- a i} $;
+  \item \texttt{diag\_poly}: fixed-degree polynomial decay, $\lambda_i(\Sigmab) = (b-a i)^2$;
+  \item \texttt{diag\_poly\_2}: variable-degree polynomial decay, $\lambda_i(\Sigmab) = b i^{-a}$.
+\end{itemize}
+The constants $a$ and $b$ are chosen to ensure $\lambda_{\text{max}}(\Sigmab) = 1$ and
+$\lambda_{\text{min}}(\Sigmab) = 10^{-4}$ (i.e., the condition number
+$\kappa(\Sigmab) = 10^{4}$ remains constant).

### 2022-08-23 design/double-descent.tex
Commit original repos

\documentclass[12pt]{sty/colt2019/colt2018-arxiv}
% \else
% \documentclass[anon,12pt]{sty/colt2019/colt2019} % Anonymized submission
% \documentclass[12pt]{colt2019} % Include author names
% \fi

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

% \title[Short Title]{Full Title of Article}
\title{Implicit regularization in DPP subset selection}
\usepackage{times}

\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{xfrac}
\usepackage{algorithmic}
\input{shortdefs}

\coltauthor{\Name{Micha{\l } Derezi\'{n}ski}
  \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics, UC Berkeley 
}

\begin{document}

\maketitle

\section{Literature review for double descent}

\begin{enumerate}
\item \citet*{BHMM19} Introduction of double descent. Empirical evidence and
experiments for: random fourier features (interpreted as a class of
two-layer neural network), decision trees and ensemble
methods. Uses the term \emph{inductive bias} to describe choosing a
minimum norm solution.

\item\citet*{BMM18_TR}
Empirical evaluation of the performance of Gaussian and Laplacian
kernels for classification in the interpolation regime. Draws a distinction between
\emph{implicit regularization} (e.g., early stopping) and
\emph{inductive bias} (e.g., minimum norm solutions).

\item\citet*{BRT18_TR}
An error bound for a special interpolating kernel
method, showing that you can generalize despite zero training error.

\item\citet*{belkin2019two} Two theoretical models of double
  descent where features are chosen as a uniformly random subset of a
  larger design matrix. Case 1: design matrix is a random i.i.d
  isotropic Gaussian. Case 2: discrete fourier transform matrix. They
  obtain exact expressions in those special cases.

\item \citet*{BLLT19_TR} Non-asymptotic upper- and lower-bounds for
  least squares in infinity dimensional Hilbert spaces. Input data is
  Gaussian with arbitrary covariance. Two notions of
 effective rank, one of them is an extension of the stable rank.

\item \citet*{BHM18_TR} Analysis of nearest neighbor in the
   interpolating regime.

 \item \citet*{BFF19_TR} Analysis of least squares
   under the Predictive Normalized Maximum Likelihood (pNML) model,
   showing the capacity for generalization in the over-parameterized
   regime.
 \item \citet*{COB18_TR} Analysis of lazy training, i.e., when the
   optimization path for a non-convex loss function follows
   that of a linearized model. The conclusion is that lazy training
   does not explain the success of over-parameterized neural networks.

 \item \citet*{GJSx19_TR} Bias and variance analysis of deep neural
   networks via a Neural Tangent Kernel model. Suggested an ensemble
   approach of averaging  several over-parameterized networks (for us,
   the ensemble approximates the ridge solution).

 \item \citet*{HMRT19_TR} Asymptotic analysis of over-parameterized
   least squares (and ridge), with exact expressions for isotropic
   features (not necessarily Gaussian), and then for general
   covariances (correlated features) the asymptotic expressions depend on the Stieltjes
   transform, which does not in general have a closed form. They show
   that the performance of over-parameterized least squares depends on
   the signal-to-noise ratio (SNR). They also show that optimally
   tuned ridge is always better than least squares. Finally, they
   differentiate between the well-specified model and the
   \emph{misspecified model}, where only a subset of features is
   observed by the learner.

 \item \citet*{MM19_TR} Asymptotic analysis of the random features
   model with a nonlinear activation (i.e., a two-layer neural network
   with fixed first layer). The random features are drawn uniformly
   from the unit sphere. The expressions are fairly complicated but
   also quite accurate. This is not directly comparable to us because
   our model is well-specified (i.e., we get all the featuers). They
   point out that for random features the peak in double descent
   occurs both for the bias and the variance, unlike for
   well-specified models where only variance exhibits a peak.

 \item \citet*{LR18_TR}   Upper bounds for the generalization error of
   Kernel ``Ridgeless''   Regression, i.e, least squares in the
   over-parameterized regime over an infinite dimensional Hilbert
   space. Arbitrary covariance matrices are allowed, and an inner
   product kernel with a non-linear smooth activation is
   assumed. Additionally, they
   perform analysis for different rates of exponentially decaying eigenvalues. 
\end{enumerate}

\appendix
\section{Sampling instances}
Let $(\X,\y)$ be an $n\times d$ linear regression problem. For a
randomly selected subset of data $S\subseteq [n]$, we predict with the
minimum norm (when $|S|\leq d$) or least squares (when $|S|>d$)
estimator: $\wbh_S = \X_S^\dagger\y_S$, where $^\dagger$ denotes 
the Moore-Penrose pseudoinverse. 
\begin{theorem}\label{t:unbiased}
  If $S$ is sampled out of all subsets of $[n]$ so that $\Pr(S)\propto 
  \det(\frac1\lambda\X_S\X_S^\top)$, then $\X_S^\dagger\y_S$ is an unbiased
  estimator of the full ridge estimator with parameter $\lambda$, i.e.:
  \begin{align*}
    \E\big[\X_S^\dagger\y_S\big] = (\X^\top\X+\lambda\I)^{-1}\X^\top\y.
  \end{align*}
\end{theorem}
\begin{remark}
Furthermore, the expected subset size is equal to the
$\lambda$-statistical dimension:
\begin{align*}
  \E\big[|S|\big] =
    \tr\big(\X^\top\X(\X^\top\X+\lambda\I)^{-1}\big) = d_\lambda\leq d.
\end{align*}
\end{remark}
Theorem \ref{t:unbiased} means that subsampling points with a DPP
makes the minimum norm estimator $\X_S^\dagger\y_S$ behave (in expectation) as if it
was ridge regularized, and we can control the amount of regularization
with the sample size. Since regularized estimators often perform
better than unregularized ones, this suggests that sampling fewer
points may some times actually improve the performance of the
estimator. We observe this in the result below, where we get an exact
closed form expression for the mean squared error of this estimator.
\begin{definition}
  Let $\X$ by a full rank $n\times d$ matrix. For any $\lambda\neq 0$
  define $\Vol_\lambda(\X)$ as a distribution over all subsets
  $S\subseteq[n]$ such that
  \begin{align*}
    \Pr(S) \propto \begin{cases}
      \det(\X_S\X_S^\top)\lambda^{-|S|}&\text{for }\lambda > 0,\\
      \det(\X_S^\top\X_S)(-\lambda)^{|S|}&\text{for }\lambda<0.
      \end{cases}
  \end{align*}
\end{definition}
\begin{assumption}\label{a:general-position}
  $\X$ is an $n\times d$ matrix with rows in general position.
\end{assumption}
\begin{assumption}\label{a:linear}
  Response vector is modeled as $\y = \X\w+\xib$ where
$\xib\sim\Nc(\zero,\sigma^2\I)$.
\end{assumption}
\begin{theorem}\label{t:mse}
Under Assumptions \ref{a:general-position} and \ref{a:linear}, if $S\sim\Vol_\lambda(\X)$ and $k_{\lambda}=\E[|S|]$, then
\begin{align*}
  \MSE{\X_S^\dagger\y_S} =
\begin{cases}
\sigma^2\tr(\Sigmab_\lambda)\cdot\big(1 +
\frac{n-d}{d-k_\lambda} (1-p_\lambda)\big)+
    (d-k_\lambda)\,\frac{\|\w\|_{\Sigmab_\lambda}^2}{\tr(\Sigmab_\lambda)},&\text{for
    }\lambda>0,\\
\sigma^2\tr(\Sigmab_0)\cdot \frac{n-d}{k_\lambda-d}(1-p_\lambda),&\text{for
}\lambda<0.
\end{cases}
\end{align*}
where $\Sigmab_\lambda=(\X^\top\X+\lambda\I)^{-1}$ and $p_\lambda=\Pr(|S|=d)$.
% Moreover, we have $k=\tr\big(\Sigmab_0\Sigmab_\lambda^{-1})$ and
% $\Pr(|S|\!=\!d) = \det(\Sigmab_0\Sigmab_\lambda^{-1})$. 
\end{theorem}
\begin{remark}\label{r:sqinv}
  If we sample subset $S$ of fixed size $k\geq d$ so that $\Pr(S)\propto
  \det(\X_S^\top\X_S)$, i.e., volume sampling, then \cite{unbiased-estimates-journal} showed that:
  \begin{align*}
    \MSE{\X_S^\dagger\y_S} = \sigma^2\frac{n-d+1}{k-d+1}\tr\big((\X^\top\X)^{-1}\big).
  \end{align*}
If we take $\lambda\rightarrow 0$ in Theorem \ref{t:mse} then the DPP
distribution converges to volume sampling of size $k=d$. Also,
the MSE expression from Theorem \ref{t:mse} converges to the above
expression, because:
\begin{align*}
  \frac{\det(\X^\top\X+\lambda\I)-\det(\X^\top\X)}{\lambda\,\det(\X^\top\X+\lambda\I)}
\ \ \underset{\lambda\rightarrow 0}\longrightarrow \ \ \tr\big((\X^\top\X)^{-1}\big).
\end{align*}
\end{remark}
\begin{proof} (of Theorem \ref{t:mse})
  Let $\alpha = \frac{\det(\X^\top\X)}{\det(\X^\top\X+\lambda\I)}$. We have
  \begin{align*}
    \E\big[\|\X_S^\dagger\y_S - \w\|^2\big]
    &= \E\big[\|\X_S^\dagger(\X_S\w+\xib_S)-\w\|^2\big]\\
    & = \E\big[\|\X_S^\dagger\xib_S\|^2\big] +
      \E\big[\|(\X_S^\dagger\X_S-\I)\w\|^2\big]\\
    &=\sigma^2\E\big[\tr(\X_S^\dagger\X_S^{\dagger\top})\big] +
      \E\big[\tr\big((\I- \X_S^\dagger\X_S)\w\w^\top\big)\big]\\
    &=\sigma^2\E\big[\tr\big((\I_S\X\X^\top\I_S)^\dagger\big)\big] +
      \tr\big(\E[\I- \X_S^\dagger\X_S]\w\w^\top\big)\\
    &= \sigma^2\bigg[\tr\big((\X\X^\top+\lambda\I)^{-1}\big) -
      \frac{\tr(\I-\X^\dagger\X)\alpha}{\lambda}\bigg]
      +
      \tr\big((\tfrac1\lambda\X^\top\X+\I)^{-1}\w\w^\top\big)\\
    & =
      \tfrac{\sigma^2}{\lambda}\bigg[\tr\big(\I-\X(\X^\top\X+\lambda\I)^{-1}\X^\top\big)
   -         (n-d)\alpha\bigg]
      + \lambda\tr\big((\X^\top\X+\lambda\I)^{-1}\w\w^\top\big)\\
    &=\frac{\sigma^2}\lambda\bigg[(1-\alpha)n -
      \Big(d-\lambda\tr\big((\X^\top\X+\lambda\I)^{-1}\big)\Big) +
      \alpha d\bigg] +
      \lambda\tr\big((\X^\top\X+\lambda\I)^{-1}\w\w^\top\big)\\
    &=\frac{\sigma^2}\lambda\bigg[(1-\alpha)(n-d) +\lambda\tr\big((\X^\top\X+\lambda\I)^{-1}\bigg]
     + \lambda\tr\big((\X^\top\X+\lambda\I)^{-1}\w\w^\top\big).
  \end{align*}
\end{proof}
We can use the following inequalities to approximate the determinantal
term in Theorem \ref{t:mse}:
\begin{align*}
\exp\Big(\lambda\tr\big((\X^\top\X)^{-1}\big)\Big)
\leq  \frac{\det\!\big(\X^\top\X)}{\det(\X^\top\X+\lambda\I)}
  \leq \exp\Big(\tr\big(\X^\top\X(\X^\top\X+\lambda\I)^{-1} -
  \I\big)\Big) = \exp(d_\lambda-d).
\end{align*}
Let $\Sigmab_{\lambda} = \X^\top\X+\lambda\I$. Using the right hand
side inequality and the fact that $\lambda = 
\frac{d-d_\lambda}{\tr(\Sigmab_\lambda^{-1})}$, we can
approximately express the MSE from Theorem \ref{t:mse} as
follows:
\begin{align*}
  \MSE{\X_S^\dagger\y_S} \geq \sigma^2\tr(\Sigmab_\lambda^{-1})\bigg[1 +
  \frac{n-d}{d-d_\lambda}\Big(1-\ee^{d_\lambda-d}\Big)\bigg] + (d-d_\lambda)\frac{\w^\top\Sigmab_\lambda^{-1}\w}{\tr(\Sigmab_\lambda^{-1})}.
\end{align*}
While the above expression depends both on the subset size $d_\lambda$ and
on $\Sigmab_\lambda$, it seems that the dependence on the subset size
is much more significant.
\begin{theorem}
Suppose that $\y = \X\w+\xib$ where
$\xib\sim\Nc(\zero,\sigma^2\I)$. If
$\Pr(S)\propto\det(\frac1\lambda\X_S\X_S^\top)$, then:
\begin{align*}
 \MSPE{\X_S^\dagger\y_S}= \E\big[\|\X\X_S^\dagger\y_S - \X\w\|^2\big]
  = ???.
\end{align*}
\end{theorem}
\begin{proof}
  \begin{align*}
    \E\big[\|\X\X_S^\dagger\y_S - \X\w\|^2\big]
    &= \E\big[\|\X\X_S^\dagger(\X_S\w+\xib_S)-\X\w\|^2\big]\\
        & = \E\big[\|\X\X_S^\dagger\xib_S\|^2\big] +
      \E\big[\|\X(\X_S^\dagger\X_S-\I)\w\|^2\big]\\
    &=???
  \end{align*}
\end{proof}
\begin{theorem}
  Let $\X$ be an $n\times d$ matrix of rank $r$ and suppose that
  $\y=\X\w+\xib$ where $\xib\sim\Nc(\zero,\sigma^2\I)$. Let
  $\B\in\R^{n\times r}$ be such that $\B\B^\top=\X\X^\top$. Denoting
  $\tilde\w=\B^\dagger\X\w$, for any $S$ we have
  \begin{align*}
\MSE{\X_S^\dagger\y_S}=\E\big[\|\X_S^\dagger\y_S-\w\|^2\big]= \E\big[\|\B_S^\dagger\y_S -
    \tilde\w\|^2\big] + \|(\I-\X^\dagger\X)\w\|^2.
  \end{align*}
\end{theorem}
\begin{proof}
  We first note that the transformation from $(\X,\w)$ to $(\B,\tilde\w)$
  preserves the response model:
  \[\B\tilde\w =
    \B\B^\dagger\X\w=\B\B^\top(\B\B^\top)^\dagger\X\w=\X\X^\top(\X\X^\top)^\dagger\X\w=\X\X^\dagger\X\w=\X\w.\]
  It follows that:
  \begin{align*}
    \E\big[\|\X_S^\dagger\y_S-\w\|^2\big] &=
    \sigma^2\tr\big((\X_S\X_S^{\top})^\dagger\big) +\big\|(\I-
    \X_S^\dagger\X_S)\w\big\|^2\\
&= \sigma^2\tr\big((\B_S\B_S^\top)^\dagger\big) +
\big\|(\I-\X_S^\dagger\X_S)\X^\dagger\X\w\big\|^2
                                    + \|(\I-\X^\dagger\X)\w\|^2\\
&=\sigma^2\tr\big((\B_S\B_S^\top)^\dagger\big) +
\big\|(\I-\B_S^\dagger\B_S)\B^\dagger\X\w\big\|^2
                                    + \|(\I-\X^\dagger\X)\w\|^2\\
&=\E\big[\|\B_S^\dagger\y_S-\tilde\w\|^2\big] + \|(\I-\X^\dagger\X)\w\|^2.
  \end{align*}
\end{proof}

We next plotting the two MSE expressions of Theorem \ref{t:mse} and
Remark \ref{r:sqinv} against the expected subset size
(for DPP that is $d_\lambda$, and for volume sampling that is $k$), we
observe a ``spike'' at $d$. This behavior is a variant of the double
descent curve discussed by \cite{belkin2019two}.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.49\textwidth]{figs/double-descent-primal}
    \includegraphics[width=0.49\textwidth]{figs/double-descent-primal-peak}
  \caption{Plot of the expressions from Theorem \ref{t:mse} and Remark
    \ref{r:sqinv} (sampling instances) against
    the expected subset size for a random $200\times 100$ matrix with
    exponentially decaying eigenvalues (left zooms in on the
    Y axis, right zooms in on the X axis).} 
\end{figure}
\paragraph{Next} Run experiments on real high-dimensional data by
averaging over multiple DPP samples.
\section{Sampling features}
Let $(\X,\y)$ be an $n\times d$ linear regression problem. We now
randomly select a subset of features $T\subseteq [d]$ and we predict
with the sparse
minimum-norm (when $|T|\geq n$) or least-squares (when $|T|< n$)
estimator: $\wbh_T = (\X\I_S)^\dagger\y$, where $^\dagger$ denotes 
the Moore-Penrose pseudoinverse and $\I_S=\sum_{i\in S}\e_i\e_i^\top$. 
% For the sake of consistency with our
% notation above, we let $\Z = \X^\top$ and use
\begin{theorem}
  If $T$ is sampled out of all subsets of $[d]$ so that $\Pr(T)\propto 
  \det(\frac1\lambda\X_{:,T}^\top\X_{:,T})$, then $\wbh_T$ is an unbiased
  estimator of the full ridge estimator with parameter $\lambda$, i.e.:
  \begin{align*}
    \E\big[\wbh_T\big] = (\X^\top\X+\lambda\I)^{-1}\X^\top\y.
  \end{align*}
\end{theorem}
\begin{theorem}\label{t:mspe-dual}
  Let $\X$ be an $n\times d$ matrix with columns in general
  position. Suppose that $\y=\X\w+\xib$, where
  $\xib\sim\Nc(\zero,\sigma^2\I)$. If $T\sim\Vol_\lambda(\X^\top)$, $k_\lambda=\E[|T|]$ and
  $\Sigmab_\lambda=(\X\X^\top+\lambda\I)^{-1}$, then:
  \begin{align*}
    \MSPE{\wbh_T}
    =\begin{cases}
    \sigma^2k_\lambda +
    (n-k_\lambda)\frac{\|\X\w\|^2_{\Sigmab_\lambda}}{\tr(\Sigmab_\lambda)},
    &\text{for }\lambda>0,\\
    \sigma^2n    &\text{for }\lambda\leq 0.
    \end{cases}
  \end{align*}
\end{theorem}
\begin{proof}
Let $\Z = \X^\top$. We will use the fact
that
\[\Z^\top(\Z^\top\I_T)^\dagger = \Z^\top\I_T\Z(\Z^\top\I_T\Z)^\dagger=
  \Z_T^\top\Z_T\Z_T^\dagger\Z_T^{\dagger\top}=\Z_T^\top\Z_T^{\dagger\top}= \Z_T^\dagger\Z_T,\]
which is a
projection matrix, and for $\lambda\leq 0$ we have $\Z_T^\dagger\Z_T=\I$.
\begin{align*}
  \MSPE{\wbh_T} &= \E\big[\|\Z^\top(\Z^\top\I_T)^\dagger\y - \Z^\top\w\|^2\big]\\
               &= \E\big[\|\Z_T^\dagger\Z_T(\Z^\top\w+\xib)-\Z^\top\w\|^2\big]\\
               &= \E\big[\|\Z_T^\dagger\Z_T\xib\|^2\big] +
                 \E\big[\|(\I-\Z_T^\dagger\Z_T)\Z^\top\w\|^2\big]\\
  &=\sigma^2\E\big[\tr(\Z_T^\dagger\Z_T)\big] +
    \E\big[\tr\big((\I-\Z_T^\dagger\Z_T)\Z^\top\w\w^\top\Z\big)\big]\\
(\text{for }\lambda>0)\quad  &=\sigma^2k_\lambda +
    \tr\big((\I+\tfrac1\lambda\Z^\top\Z)^{-1}\Z^\top\w\w^\top\Z\big)\\
  &=\sigma^2k_\lambda + \lambda\,\|\X\w\|^2_{(\X\X^\top+\lambda\I)^{-1}}.
\end{align*}
Note that $k = n-\lambda\, \tr\big((\X\X^\top+\lambda\I)^{-1}\big)$, so
the result follows.
\end{proof}

\begin{theorem}\label{t:mse-dual}
  Let $\X$ be an $n\times d$ matrix with columns in general
  position. Suppose that $\y=\X\w+\xib$, where
  $\xib\sim\Nc(\zero,\sigma^2\I)$. If $\Pr(T)\propto
  \det(\frac1\lambda\X_{:,T}^\top\X_{:,T})$, then:
  \begin{align*}
    \MSE{\wbh_T} &=
    \sigma^2\tr\big((\X\X^\top+\lambda\I)^{-1}\big)\bigg[1 +
    \frac{d-n}{n-d_\lambda}\Big(1-\frac{\det(\X\X^\top)}{\det(\X\X^\top+\lambda\I)}\Big)\bigg] &&\textnormal{variance}\\
    &\quad+\quad ???&&\textnormal{bias}^2
  \end{align*}
\end{theorem}
\Red{\bf It is not clear how to compute the bias. Previous proof had a mistake.}
\begin{proof}
Let $\Z = \X^\top$. Then we are sampling $T$ so that $\Pr(T)\propto\det(\frac1\lambda\Z_T\Z_T^\top)$.
\begin{align*}
  \MSE{\wbh_T} &= \E\big[\|(\Z^\top\I_T)^\dagger\y - \w\|^2\big]\\
               &= \E\big[\|(\Z^\top\I_T)^\dagger(\Z^\top\w+\xib)-\w\|^2\big]\\
               &= \E\big[\|(\Z^\top\I_T)^\dagger\xib\|^2\big] +
                 \E\big[\|(\I-(\Z^\top\I_T)^\dagger\Z^\top)\w\|^2\big].
\end{align*}
Note that $(\Z^\top\I_T)^\dagger = \I_T\Z(\Z_T^\top\Z_T)^\dagger$, so the variance part can be written as follows:
\begin{align*}
  \E\big[\|(\Z^\top\I_T)^\dagger\xib\|^2\big]
  &= \sigma^2\E\Big[\tr\big((\Z^\top\I_T)^\dagger (\Z^\top\I_T)^{\dagger\top}\big)\Big]\\
  &=
    \sigma^2\E\Big[\tr\big((\Z_T^\top\Z_T)^\dagger(\Z_T^\top\Z_T)^\dagger\Z_T^\top\Z_T\big)\Big]\\
  &=\sigma^2\E\Big[\tr\big((\Z_T^\top\Z_T)^\dagger\big)\Big]\\
  &=\sigma^2\bigg[\tr\big((\Z^\top\Z+\lambda\I)^{-1}\big) +
    \frac{(d-n)}{\lambda}\Big(1-\frac{\det(\Z^\top\Z)}{\det(\Z^\top\Z+\lambda\I)}\Big)\bigg]
\end{align*}
Next, we compute the bias term, letting $\bar T = [d]\backslash T$:
\begin{align*}
  \E\big[\|(\I-(\Z^\top\I_T)^\dagger\Z^\top)\w\|^2\big]
  &= \E\Big[\w^\top\big(\I - \Z(\I_T\Z)^\dagger\big)\big(\I-(\Z^\top\I_T)^\dagger\Z^\top\big)\w\Big]\\
  &= \|\w\|^2 - \E\big[2\,\w^\top\Z(\Z_T^\top\Z_T)^\dagger\Z^\top\I_T\w +
    \w^\top\Z(\Z_T^\top\Z_T)^\dagger\Z^\top(\I_T+\I_{\bar T})\w\big]\\
  &= \|\w\|^2 -
    \w^\top\Z\E\big[(\Z_T^\top\Z_T)^\dagger\Z^\top\I_T\big]\w +
    \w^\top\Z\E\big[(\Z_T^\top\Z_T)^\dagger\Z^\top\I_{\bar T}\big]\w\\
  &=\|\w\|^2 - \w^\top\Z\Z^\top(\lambda\I+\Z\Z^\top)^{-1}\w +
    \w^\top\Z\E\big[(\Z_T^\top\Z_T)^\dagger\Z^\top\I_{\bar T}\big]\w\\
  &=\lambda\,\w^\top(\X^\top\X+\lambda\I)^{-1}\w +
    \w^\top\Z\E\big[(\Z_T^\top\Z_T)^\dagger\Z^\top\I_{\bar T}\big]\w.
\end{align*}
% If $|T|< n$ and the rows of $\Z$ are in general position, then
% $(\Z_T^\top\Z_T)^\dagger\z_i=\zero$ for any $i\not\in T$. So
% \begin{align*}
%   \E\big[(\Z_T^\top\Z_T)^\dagger\Z^\top\I_{\bar T}\big]
%   &=
%   \Pr(|T|=n)\,\E\big[(\Z_T^\top\Z_T)^\dagger\Z^\top\I_{\bar T}\mid
%     |T|=n\big]\\
%   &= \frac{\det(\frac1\lambda\Z^\top\Z)}{\det(\frac1\lambda\Z^\top\Z+\I)}\,\E\big[(\Z_T^\top\Z_T)^{-1}\Z^\top -
%     (\Z_T^\top\Z_T)^{-1}\Z^\top\I_T\mid |T|=n\big]\\
%   &= \frac{\det(\Z^\top\Z)}{\det(\Z^\top\Z+\lambda\I)}\,\Big[(d-n+1)(\Z^\top\Z)^{-1}\Z^\top -
%     (\Z^\top\Z)^{-1}\Z^\top\Big]\\
%   &=\frac{\det(\Z^\top\Z) (d-n)}{\det(\Z^\top\Z+\lambda\I)} \,\Z^\dagger.
% \end{align*}
\end{proof}
\begin{theorem}\label{t:mse-dual-tail}
If we sample subset $T$ of fixed size $k\geq n$ so that $\Pr(T)\propto
  \det(\X_{:,T}\X_{:,T}^\top)$, i.e., volume sampling, then:
  \begin{align*}
    \MSE{\X_S^\dagger\y_S}
    &= \sigma^2\frac{d-n+1}{k-n+1}\tr\big((\X\X^\top)^{-1}\big)&&\textnormal{variance}\\
    &\quad+ \w^\top(\I-\X^\dagger\X)\w + \frac{d-k}{k-n+1}\w^\top\X^\dagger\X\w.&&\textnormal{bias}^2
  \end{align*}
\end{theorem}
% \begin{figure}[H]
%   \centering
%   \includegraphics[width=0.49\textwidth]{figs/double-descent-dual}
%     \includegraphics[width=0.49\textwidth]{figs/double-descent-dual-peak}
%   \caption{Plot of the expressions from Theorems \ref{t:mse-dual} and
%     \ref{t:mse-dual-tail} (sampling features)  against
%     the expected subset size for a random $100\times 200$ matrix with
%     exponentially decaying eigenvalues (left zooms in on the
%     Y axis, right zooms in on the X axis).} 
% \end{figure}

\section{Motivating DPPs as a model for feature selection}
Suppose that $\Z\in\R^{d\times n}$ is a population of feature vectors
($d$ rows of $\Z$). We may imagine that $d$ could be infinite, in
which case we would consider a distribution over features.
The task of a feature engineer is to select a set $S$ of $k$
features that will represent the population well. We propose the following model of a
feature engineer:
\begin{quote}
The \emph{oblivious feature engineer} does not have knowledge of the
population matrix $\Z$, however they 
can indicate their preference by defining a score function $f$ such that
$f(\Z_S)$ is a non-negative weight representing the believed quality
of a feature set $S$. The feature selection is realized by sampling
the set so that $\Pr(S)\propto f(\Z_S)$.
\end{quote}
Note that this idealized feature selection can be motivated by
considering a Markov chain procedure where we start with some subset
$S$ and then randomly pick two points $i\in S$ and $j\not\in S$ to
swap. We then accept the swap with probability
$\min\big\{1,\frac{f(\Z_{S-i+j})}{f(\Z_S)}\big\}$. The Markov chain
only requires a uniformly random access to features from $\Z$ and
converges to the distribution $\Pr(S)\propto
f(\Z_S)$. Note that this scheme requires us to fix the size
of the subset.

\textbf{Question:} What is the optimal score function for the
oblivious feature engineer? For now, we focus on the case of $k <
\min\{n,d\}$. One way to determine its quality is to ask how much 
 information do we lose by applying the orthogonal projection onto the
 subspace spanned by the features, $\P_{\!S}=\Z_S^\dagger\Z_S$, to the
 whole population matrix $\Z$. 
 We can evaluate this by looking at the sum of squares of the
 residuals: $\|\Z-\P_{\!S}\Z\|_F^2$.
The natural gold standard is top-$k$ PCA, which is the
best rank $k$ approximation of $\Z$, denoted $\Z_k$, and also a simple
example of an auto-encoder. Therefore, the \emph{value} of a score
function $f$ can be defined as:
\begin{align*}
  v_k(f) =  \max_{\Z}\frac{\E_{S\sim
  f\,:\,|S|=k}\big[\|\Z-\P_{\!S}\Z\|_F^2\big]}{\|\Z-\Z_k\|_F^2}. 
\end{align*}
\cite{pca-volume-sampling} showed that if
$f_{\DPP}(\Z_S)=\det(\Z_S\Z_S^\top)$, then $v_k(f)=k+1$. Also, they showed a
lower bound which says that for any $f$  we have $v_k(f)\geq
k+1$. This means that a cardinality constrained DPP is minimax optimal
in terms of $v_k$. Here are some further questions:
\begin{enumerate}
  \item Is the minimax optimal score function unique? In other words,
    can we show that for any $f\neq f_{\DPP}$ we have $v_k(f)>k+1$?
Simple first step, show that for $f$ uniformly equal $1$ we have
$v_k(f)=\infty$.
 \item The above requires that we fix the subset size, but we are
   actually analyzing a DPP with unconstrained size. Can we redefine
   the minimax value by replacing the cardinality constraint with an
   expected cardinality constraint? What is the minimax optimal score
   function then?
 \item The projection error is only well-defined for
   $k<\min\{n,d\}$. What is the natural way of extending this
   motivation to $k\geq \min\{n,d\}$? 
\end{enumerate}

\section{Tight analysis of the inverse formula}
Let $\pdet(\A)$ denote the pseudo-determinant of $\A$, i.e.,
the product of its non-zero eigenvalues.
\begin{theorem}
  Let $\X$ be an $n\times d$ matrix of rank $r$ whose rows are in
  general position within an $r$-dimensional subspace.
    If $S$ is sampled out of all subsets of $[n]$ so that $\Pr(S)\propto 
    \det(\X_S\X_S^\top)$, then
    \begin{align*}
    \E\big[(\I_S\X\X^\top\I_S)^\dagger\big] = (\I+\X\X^\top)^{-1} - \frac{\pdet(\X\X^\top)}{\det(\I+\X\X^\top)}\,(\I-\X^\dagger\X).  
    \end{align*}
  \end{theorem}
  \begin{remark}
    If $r= n$ then $\X^\dagger\X=\I$ and we get the
     formula: $\E\big[(\I_S\X\X^\top\I_S)^\dagger\big] = (\I+\X\X^\top)^{-1}$.
   \end{remark}
   \begin{proof}
     W.l.o.g.~assume that $r=d$.
     Let $\y\in\R^n$ be any vector. We will first compute
     \begin{align*}
       \det(\I+\X^\top\X)\,\E\big[\y^\top(\I_S\X\X^\top\I_S)^\dagger\y\big] =
       \sum_{S:|S|\leq d}\y_S^\top\adj(\X_S\X_S^\top)\y_S
     \end{align*}
     Each of those terms can be further simplified to:
  \begin{align*}
   \y_S^\top\adj(\X_S\X_S^\top)\y_S =
    \det(\X_S\X_S^\top+\y_S\y_S^\top) - \det(\X_S\X_S^\top)
    =\det\!\big([\X,\y]_S[\X,\y]_S^\top\big) - \det(\X_S\X_S^\top).
  \end{align*}
  Summing up over $S$ we obtain:
  \begin{align*}
    \sum_{S:|S|\leq d}\y_S^\top\adj(\X_S\X_S^\top)\y_S
    & =
    \sum_{S:|S|\leq d} \Big(\det\!\big([\X,\y]_S[\X,\y]_S^\top\big) -
\det(\X_S\X_S^\top)\Big) \\
&= \det\!\big(\I+[\X,\y][\X,\y]^\top\big) -\det(\I + \X\X^\top) -
\sum_{S:|S|=d+1}\det\!\big([\X,\y]_S[\X,\y]_S^\top\big) \\
&= \y^\top\!\adj(\I + \X\X^\top)\y -
\det\!\big([\X,\y]^\top[\X,\y]\big)\\
&=\det(\I+\X^\top\X)\,\y^\top(\I+\X\X^\top)^{-1}\y -
                                            \det(\X^\top\X)\,\y^\top(\I-\X^\dagger\X)\y.
  \end{align*}
  Since the formula holds for all $\y$'s, the result follows.
   \end{proof}

   \section{Square inverse formula for volume plus binomial sampling}
   \begin{theorem}\label{t:sqinv-binom}
     Let $\X$ be an $n\times d$ matrix whose rows are in general
     position. If $T\sim\Vol(\X)$ and $R\sim\mathrm{Bin}(n,p)$, then
     $S=T\cup R$ satisfies:
     \begin{align*}
       \E\big[(\X_S^\top\X_S)^{-1}\big] = (p\X^\top\X)^{-1}\cdot \big(1-(1-p)^{n-d+1}\big).
     \end{align*}
   \end{theorem}
   \begin{remark}
     If we set $p = \frac{k-d}{n-d}$ then $\E\big[|S|\big]=k$ and we get
\[     \E\big[(\X_S^\top\X_S)^{-1}\big] =
  \frac{n-d}{k-d}\,(\X^\top\X)^{-1}\cdot\bigg(1-\Big(\frac{n-k}{n-d}\Big)^{n-d+1}\bigg).\]
On the other hand, if we take $p\rightarrow 0$, then we recover
$\E\big[(\X^\top\X)^{-1}\big] = (n-d+1)\,(\X^\top\X)^{-1}$. 
\end{remark}
     \begin{proof}
       \begin{align*}
         \det(p\X^\top\X)&\,\E\big[(\X_S^\top\X_S)^{-1}\big]
         = \sum_{S}\det(\X_S^\top\X_S)(\X_S^\top\X_S)^{-1}p^{|S|}(1-p)^{n-|S|}\\
          &=
            \sum_{k=0}^n\sum_{S:|S|=k}\adj(\X_S^\top\X_S)p^k(1-p)^{n-k}
            \ - \sum_{S:|S|=d-1}\!\!\adj(\X_S^\top\X_S)p^{d-1}(1-p)^{n-d+1}\\
         &=\adj(p\X^\top\X) - \adj(\X^\top\X)p^{d-1}(1-p)^{n-d+1}\\
         &=\det(p\X^\top\X)(p\X^\top\X)^{-1}\big(1-(1-p)^{n-d+1}\big).
       \end{align*}
     \end{proof}
     \begin{corollary}
  Let $\X$ be an $n\times d$ matrix of rank $r$ whose rows are in
  general position within an $r$-dimensional subspace. If $T\sim\DPP(\X\X^\dagger)$ and $R\sim\mathrm{Bin}(n,p)$, then
  $S=T\cup R$ satisfies:
  \begin{align*}
    \E\big[\tr\big((\X_S^\top\X_S)^\dagger\big)\big] =
    \tr\big((p\X^\top\X)^\dagger\big)\cdot \big(1-(1-p)^{n-r+1}\big).
  \end{align*}
     \end{corollary}
     \begin{proof}
       Let $\L=\X\X^\top$. Since $\rank(\L)=r$, there is a matrix
       $\B\in\R^{n\times r}$ such that $\L=\B\B^\top$. It follows that
       $\DPP(\X\X^\dagger)=\Vol(\B)$. Moreover, since
       $\X_S\X_S^\top=\L_{S,S}=\B_S\B_S^\top$ for any $S\subseteq [n]$, we have
       \begin{align*}
         \tr\big((\X_S^\top\X_S)^\dagger\big) =
         \tr\big((\X_S\X_S^\top)^\dagger\big) =
         \tr\big((\B_S\B_S^\top)^\dagger\big) = \tr\big((\B_S^\top\B_S)^{-1}\big).
       \end{align*}
       In particular, we also have $\tr\big((\X^\top\X)^\dagger\big) =
       \tr\big((\B^\top\B)^{-1}\big)$. The result now follows by
       applying Theorem \ref{t:sqinv-binom}.
     \end{proof}

\section{Proof of Theorem \ref{t:unbiased}}
   \begin{proofof}{Theorem}{\ref{t:unbiased}}
Assume that $S\sim\DPP(\X\X^\top)$ (we can drop the $\lambda$
w.l.o.g.).  Let $\z\in\R^{d}$ be any vector. We will compute 
     \begin{align*}
       \det(\I+\X^\top\X)\,\z^\top\E\big[\X_S^\dagger\y_S\big] =
       \sum_{S:|S|\leq
       d}\det(\X_S\X_S^\top)\z^\top\X_S^\top(\X_S\X_S^\top)^{-1}\y_S
       =
\sum_{S:|S|\leq
       d}\z^\top\X_S^\top\adj(\X_S\X_S^\top)\y_S
     \end{align*}
     Each of those terms can be further simplified to:
  \begin{align*}
\z^\top\X_S^\top\adj(\X_S\X_S^\top)\y_S =
    \det(\X_S\X_S^\top+\y_S\z^\top\X_S^\top) - \det(\X_S\X_S^\top)
    =\det\!\big([\X,\y]_S[\X,\X\z]_S^\top\big) - \det(\X_S\X_S^\top).
  \end{align*}
  Summing up over $S$ we obtain:
  \begin{align*}
    \sum_{S:|S|\leq d}\z^\top\X^\top\adj(\X_S\X_S^\top)\y_S
    & =
    \sum_{S:|S|\leq d} \Big(\det\!\big([\X,\y]_S[\X, \X\z]_S^\top\big) -
\det(\X_S\X_S^\top)\Big) \\
&= \det\!\big(\I+[\X,\y][\X, \X\z]^\top\big) -\det(\I + \X\X^\top)\\
&= \z^\top\X^\top\!\adj(\I + \X\X^\top)\y \\
    &=\det(\I+\X\X^\top)\,\z^\top\X^\top(\I+\X\X^\top)^{-1}\y\\
    &=\det(\I+\X^\top\X)\,\z^\top(\I+\X^\top\X)^{-1}\X^\top\y.
  \end{align*}
  Since the formula holds for all $\z$'s, the result follows.
   \end{proofof}

\bibliography{pap}

\end{document}

### 2022-08-23 design/equivalents_nips/neurips_2020.tex
Commit original repos

\newif\ifisarxiv
\isarxivtrue

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2020

% ready for submission
% \usepackage{neurips_2020}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2020}

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2020}

% to avoid loading the natbib package, add option nonatbib:
\ifisarxiv
\usepackage{fullpage}
\usepackage{natbib}
\else
\usepackage[final]{neurips_2020}
\fi
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
 \usepackage{subfigure}
\usepackage{amsmath,amsthm}
\usepackage{cancel}
\usepackage{cleveref}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{todonotes}
\usepackage{xcolor}
\usepackage{xfrac}
\usepackage{caption}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}
\newcommand{\BLUE}{\color[rgb]{0,0,0.69}}
\input{../shortdefs}

\title{Precise expressions for random projections: \\
  Low-rank approximation and randomized Newton}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\ifisarxiv
\author{
          \textbf{Micha{\l } Derezi\'{n}ski} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \and
  \textbf{Feynman Liang} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \and
   \textbf{Zhenyu Liao} \\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{zhenyu.liao@berkeley.edu}
  \and
   \textbf{Michael W. Mahoney}\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
}
\else
\author{%
 \textbf{Micha{\l } Derezi\'{n}ski} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
  \textbf{Feynman Liang} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
   \textbf{Zhenyu Liao} \\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{zhenyu.liao@berkeley.edu}
  \And
   \textbf{Michael W. Mahoney}\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
  % David S.~Hippocampus\thanks{Use footnote for providing further information
  %   about author (webpage, alternative address)---\emph{not} for acknowledging
  %   funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}
\fi

\begin{document}
\maketitle

\begin{abstract}
It is often desirable to reduce the dimensionality of a large dataset
by projecting it onto a low-dimensional subspace.  Matrix sketching
has emerged as a powerful technique for performing such dimensionality
reduction very efficiently.  Even though there is an extensive
literature on the worst-case performance of sketching, existing
guarantees are typically very different from what is observed in
practice.  We exploit recent developments in the spectral analysis of
random matrices to develop novel techniques that provide provably
accurate expressions for the expected value of random projection
matrices obtained via sketching.  These expressions can be used to
characterize the performance of dimensionality reduction in a variety
of common machine learning tasks, ranging from low-rank approximation
to iterative stochastic optimization.  Our results apply to several
popular sketching methods, including Gaussian and Rademacher sketches,
and they enable precise analysis of these methods in terms of spectral
properties of the data.  Empirical results show that the expressions
we derive reflect the practical performance of these sketching
methods, down to lower-order effects and even constant factors. 
\end{abstract}

\section{Introduction}
Many settings in modern machine learning, optimization and scientific
computing require us to work with data matrices that are so large that
some form of dimensionality reduction is a necessary component of the
process. One of the most popular families of methods for dimensionality reduction,
coming from the literature on Randomized Numerical Linear Algebra
(RandNLA), consists of data-oblivious sketches \cite{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}. 
Consider a large $m\times n$
matrix $\A$. A \emph{data-oblivious sketch} of size $k$ is the matrix $\S\A$,
where $\S$ is a $k\times m$ random matrix such that
$\E[\frac1k\S^\top\S]=\I$, whose distribution does not
depend on $\A$. This sketch reduces the first dimension of $\A$ from
$m$ to a much smaller $k$ (we assume without loss of generality that
$k \ll n \le m$), and an analogous procedure can be defined for
reducing the second 
dimension as well. This approximate representation of $\A$ is central
to many algorithms in areas such as linear regression, low-rank approximation, kernel methods,
and iterative second-order optimization. While there is a long line of research
aimed at bounding the worst-case approximation error of such
representations, these bounds are often too loose to reflect accurately
the practical performance of these methods. In this paper, we develop new theory which
enables more precise analysis of the accuracy of sketched data
representations.

A common way to measure the accuracy of the sketch $\S\A$ is by
considering the $k$-dimensional subspace spanned by its rows. The goal
of the sketch is to choose a subspace that best aligns with the
distribution of all of the $m$ rows of $\A$ in $\R^n$. Intuitively, our goal
is to minimize the (norm of the) residual when projecting a vector $\a\in\R^n$ onto that
subspace, i.e., $\a - \P\a=(\I-\P)\a$, where
$\P = (\S\A)^\dagger\S\A$ is the orthogonal projection matrix onto the subspace
spanned by the rows of $\S\A$ (and $(\cdot)^\dagger$ denotes the
Moore-Penrose pseudoinverse). For this reason, the quantity that has appeared 
ubiquitously in the error analysis of RandNLA sketching is what we call the residual
projection matrix:
\begin{align*}
  \textbf{(residual projection matrix)}\quad \P_{\!\perp}\ :=\ \I - \P\ =\ \I
  - (\S\A)^\dagger\S\A.
\end{align*}
Since $\P_{\!\perp}$ is random, the average performance of the sketch
can often be characterized by its expectation, $\E[\P_{\!\perp}]$.
For example, the low-rank approximation error of the
sketch can be expressed as
$\E[\|\A - \A\P\|_F^2]= \tr\,\A^\top\A\,\E[\P_{\!\perp}]$, where
$\|\cdot\|_F$ denotes the Frobenius norm.  A similar formula follows for
the trace norm error of a sketched Nystr\"om approximation
\cite{Williams01Nystrom,revisiting-nystrom}.
Among others, this approximation error
appears in the analysis of sketched kernel 
ridge regression \cite{fanuel2020diversity} and Gaussian process
regression \cite{sparse-variational-gp}. Furthermore, a variety of
iterative algorithms, such as randomized second-order methods for
convex optimization \cite{Qu2015Feb,Qu2016,Gower2019,jacsketch}
and linear system solvers based on 
the generalized Kaczmarz method \cite{generalized-kaczmarz},
have convergence guarantees which depend on the extreme eigenvalues of
$\E[\P_{\!\perp}]$. Finally, a generalized form of the expected
residual projection has been recently used to model the implicit
regularization of the interpolating solutions in over-parameterized
linear models \cite{surrogate-design,BLLT19_TR}.

\subsection{Main result}
Despite its prevalence in the literature, the 
expected residual projection is not well understood, even in such
simple cases as when $\S$ is a Gaussian sketch (i.e., with
i.i.d.~standard normal entries). We address this by providing a
surrogate expression, i.e., a simple analytically
tractable approximation, for this matrix quantity:
\begin{align}
  \E[\P_{\!\perp}]\ \overset\epsilon\simeq \ \bar\P_{\!\perp}:=(\gamma\A^\top\A +
  \I)^{-1},\quad\text{with \ $\gamma>0$ \ s.t. \ }\tr\,\bar\P_{\!\perp} = n-k.\label{eq:surrogate}
\end{align}
Here, $\overset{\epsilon}{\simeq}$ means that while the surrogate expression is
not exact, it approximates the true quantity up to some $\epsilon$
accuracy. Our main result provides a rigorous approximation guarantee
for this surrogate expression with respect to a
range of sketching matrices $\S$, including the standard Gaussian and
Rademacher sketches. We state the result using the
positive semi-definite ordering denoted by $\preceq$. 
\begin{theorem}\label{t:main}
Let $\S$ be a sketch of size $k$ with i.i.d.~mean-zero sub-gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.  If
  we let $\rho = r/k$ be a fixed constant larger than $1$, then
  \begin{align*}
    (1-\epsilon)\,\bar\P_{\!\perp}\preceq\E[\P_{\!\perp}]\preceq
    (1+\epsilon)\,\bar\P_{\!\perp}\quad\text{for}\quad \epsilon =
    O(\tfrac1{\sqrt r}).
  \end{align*}
\end{theorem}
In other words, when the sketch size $k$ is smaller than the stable rank $r$
of $\A$, then the discrepancy between our surrogate expression
$\bar\P_{\!\perp}$ and $\E[\P_{\!\perp}]$ is of the order
$1/\sqrt r$, where the big-O notation hides only the dependence
on $\rho$ and on the sub-gaussian constant (see Theorem \ref{t:main-tech} for
more details). Our proof of Theorem \ref{t:main} is inspired by the techniques from
random matrix theory which have been used to analyze the asymptotic
spectral distribution of large random matrices by focusing on the
associated matrix resolvents and Stieltjes transforms
\cite{hachem2007deterministic,bai2010spectral}. However, our analysis
is novel in several respects:
\begin{enumerate}
  \item The residual projection matrix can be obtained
    from the appropriately scaled resolvent matrix $z(\A^\top\S^\top\S\A+z\I)^{-1}$ by
    taking $z\rightarrow 0$. Prior work (e.g.,
      \cite{HMRT19_TR}) combined this with an exchange-of-limits argument to
      analyze the asymptotic behavior of the residual projection. 
      This approach, however, does not allow for a precise control in
      finite-dimensional problems. We are able to provide a more
      fine-grained, non-asymptotic analysis by working directly with
      the residual projection itself, instead of the resolvent.
  \item We require no assumptions on the largest and smallest singular
    value of $\A$. Instead, we derive our bounds in terms of the
    stable rank of $\A$ (as opposed to its actual rank), which
    implicitly compensates for ill-conditioned data matrices.
  \item We obtain upper/lower bounds for $\E[\P_{\!\perp}]$ in terms
    of the positive semi-definite ordering~$\preceq$, which can be
    directly converted to guarantees for the precise
    expressions of expected low-rank approximation error derived in
    the following section. 
    % are stronger than the standard error bounds
    % $\|\E[\P_{\!\perp}]-\bar\P_{\!\perp}\|$ because they lead to relative error
    % approximations for \emph{all} of the eigenvalues of
    % $\E[\P_{\!\perp}]$ (with respect to those of $\bar \P_{\!\perp}$).
  \end{enumerate}

It is worth mentioning that the proposed analysis is significantly
different from the sketching literature based on subspace
embeddings (e.g.,
\cite{sarlos-sketching,cw-sparse,nn-sparse,projection-cost-preserving,optimal-matrix-product}),
in the sense that here our object of interest is not to obtain
a worst-case approximation with high probability, but rather, our
analysis provides \emph{precise} characterization on the
\emph{expected} residual projection matrix that goes \emph{beyond
  worst-case bounds}. From an application perspective, the subspace
embedding property is neither sufficient nor necessary for many
numerical implementations of sketching \citep{blendenpik,lsrn}, 
or statistical results \citep{GarveshMahoney_JMLR,dobriban2019asymptotics,yang2020reduce}, as well as in the context of iterative optimization and implicit regularization (see Sections~\ref{s:newton}~and~\ref{s:implicit} below), which are discussed in detail as concrete applications of the proposed analysis.

\subsection{Low-rank approximation}
\label{s:low-rank}
We next provide some immediate corollaries of Theorem~\ref{t:main},
where we use $x\overset\epsilon\simeq y$ to denote a 
multiplicative approximation $|x-y|\leq \epsilon y$.
Note that our analysis is new even for the classical Gaussian sketch where
the entries of $\S$ are i.i.d.~standard normal. However the results
apply more broadly, including a standard class of data-base friendly
Rademacher sketches where each entry $s_{ij}$ is a $\pm1$ Rademacher
random variable \cite{achlioptas2003database}. We start by analyzing the Frobenius
norm error $\|\A-\A\P\|_F^2=\tr\,\A^\top\A\,\P_{\!\perp}$ of
sketched low-rank approximations. Note that by the definition of $\gamma$ in
\eqref{eq:surrogate}, we have $k =
\tr\,(\I-\bar\P_{\!\perp})=\tr\,\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1}$,
so the surrogate expression we obtain for the expected error is
remarkably simple.
\begin{corollary} %[low-rank approximation]
  \label{c:low-rank}
Let $\sigma_i$ be the singular values of $\A$. Under the assumptions of Theorem \ref{t:main}, we have:\vspace{-1mm}
  \begin{align*}
    \E\big[\|\A-\A\P\|_F^2\big] \ \overset\epsilon\simeq\ 
    k/\gamma
\quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1} = k.
%\quad \text{with \ $\gamma>0$ \ s.t. \ }\tr\,\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1} = k.
    % \tr\A^\top\A(\gamma\A^\top\A+\I)^{-1},
   % \quad \text{with \ $\gamma>0$ \ s.t. \ }\tr\,(\gamma\A^\top\A+\I)^{-1} = n-k.
  \end{align*}
\end{corollary}\vspace{-4mm}
\begin{remark}
  The parameter $\gamma=\gamma(k)$ increases at least linearly
  as a function of $k$, which is why the expected error will always
  decrease with increasing $k$. For example, when the singular values
  of $\A$ exhibit exponential decay, i.e., $\sigma_i^2=C\cdot\alpha^{i-1}$ for $\alpha\in(0,1)$, then the error also decreases
  exponentially, at the rate of $k/(\alpha^{-k}-1)$.
We discuss this further in Section \ref{s:explicit}, giving explicit
formulas for the error as a function of $k$ under both exponential and polynomial spectral
decay profiles.
\end{remark}
The above result is important for many RandNLA methods, and it is also relevant in the context of kernel methods,
where the data is represented via a positive semi-definite $m\times m$
kernel matrix $\K$ which corresponds to the matrix of dot-products of
the data vectors in some reproducible kernel Hilbert space. In this
context, sketching can be applied directly to the matrix $\K$ via an
extended variant of the
Nystr\"om method \cite{revisiting-nystrom}. A Nystr\"om approximation
constructed from a sketching matrix $\S$ is defined as $\tilde\K =
\C^\top\W^\dagger\C$, where $\C=\S\K$ and $\W=\S\K\S^\top$, and it is
applicable to a variety of settings, including Gaussian Process
regression, kernel machines and Independent
Component Analysis \cite{sparse-variational-gp,Williams01Nystrom,Bach2003}. By setting
$\A=\K^{\frac12}$, it is easy to see \cite{nystrom-multiple-descent} that
the trace norm error $\|\K-\tilde\K\|_*$ is identical to the squared
Frobenius norm error of the low-rank 
sketch $\S\A$, so Corollary \ref{c:low-rank} implies that
\begin{align}
  \E\big[\|\K-\tilde\K\|_*\big] \ \overset\epsilon\simeq\  k/\gamma \quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\lambda_i}{\gamma\lambda_i+1} = k,\label{eq:nystrom}
\end{align}
with any sub-gaussian sketch, where $\lambda_i$ denote the
eigenvalues of $\K$. Our error analysis 
given in Section \ref{s:explicit} is particularly relevant here, since
commonly used kernels such as the Radial 
Basis Function (RBF) or the Mat\'ern kernel induce a well-understood eigenvalue
decay \cite{Santa97Gaussianregression,RasmussenWilliams06}.

Metrics other than the aforementioned Frobenius norm error,
such as the spectral norm error \cite{tropp2011structure}, are also of significant
interest in the low-rank approximation literature. We leave these
directions for future investigation.

\subsection{Randomized iterative optimization}
\label{s:newton}
We next turn to a class of iterative methods which take
advantage of sketching to reduce the per iteration cost of
optimization. These methods have been developed in a variety of
settings, from solving linear systems to convex optimization and
empirical risk minimization, and in many cases
the residual projection matrix appears as a black box quantity whose spectral
properties determine the convergence behavior of the algorithms
\cite{generalized-kaczmarz}. With
our new results, we can precisely characterize not only the rate of
convergence, but also, in some cases, the complete evolution of
the parameter vector, for the following algorithms:
% The following algorithms
% benefit from our improved convergence analysis:
\begin{enumerate}
  \item \emph{Generalized Kaczmarz method}
    \cite{generalized-kaczmarz} for approximately solving a linear system $\A\x=\b$;
  \item \emph{Randomized Subspace Newton} \cite{Gower2019}, a second order
    method, where we sketch the Hessian matrix.
  \item \emph{Jacobian Sketching} \cite{jacsketch}, a
    class of first order methods which use additional information via a
    weight matrix $\W$ that is sketched at every iteration.
    % \item \emph{Iterative Adaptive Sketching} \cite{lacotte2019high},
    %   a second order method where the sketching matrices are
    %   adaptively rescaled to improve the convergence rate.
      % \footnote{While
      %   \cite{lacotte2019high} also use the residual projection in
      %   their analysis, adapting our techniques is less
      %   straightforward in this case.}
\end{enumerate}
 We believe that
extensions of our techniques will apply to other algorithms,
such as that of \cite{lacotte2019high}.

We next 
%Next, we 
give a result in the context of linear systems
for the generalized Kaczmarz method \cite{generalized-kaczmarz}, but a similar
convergence analysis is given for the methods of
\cite{Gower2019,jacsketch} in Appendix~\ref{a:newton}.
\begin{corollary}\label{c:kaczmarz}
Let $\x^*$ be the unique solution of $\A\x^*=\b$ and consider
  the iterative algorithm:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S\A\x=\S\b.
  \end{align*}
Under the assumptions of Theorem \ref{t:main}, with $\gamma$ defined in
\eqref{eq:surrogate} and $r=\|\A\|_F^2/\|\A||^2$, we have:
  \begin{align*}
    \E\big[\x^{t+1}-\x^*\big] \overset\epsilon\simeq 
    (\gamma\A^\top\A+\I)^{-1}\,\E\big[\x^t-\x^*\big]
    \quad\text{for}\quad\epsilon=O(\tfrac1{\sqrt r}). 
  \end{align*}
\end{corollary}
The corollary follows from Theorem \ref{t:main} combined with Theorem
4.1 in \cite{generalized-kaczmarz}. 
Note that when $\A^\top\A$ is positive definite then
$(\gamma\A^\top\A+\I)^{-1}\prec\I$, so the algorithm will
converge from any starting point, and the worst-case convergence rate of the above method can be
obtained by evaluating the largest eigenvalue of
$(\gamma\A^\top\A+\I)^{-1}$. However the result itself is much
stronger, in that it can be used to describe the (expected) trajectory of the
iterates for any starting point $\x^0$. Moreover, when the spectral decay
profile of $\A$ is known, then the explicit expressions for $\gamma$
as a function of $k$ derived in Section \ref{s:explicit} can be used
to characterize the convergence properties of generalized Kaczmarz as
well as other methods discussed above.

\subsection{Implicit regularization}\label{s:implicit}
%\paragraph{Implicit regularization.}

Setting $\x^t=\zero$, we can view one step of the iterative method in Corollary~\ref{c:kaczmarz} as
  finding a minimum norm interpolating solution of an under-determined linear
  system $(\S\A,\S\b)$. Recent interest in the generalization
capacity of over-parameterized machine learning models has
  motivated extensive research on the statistical properties of such interpolating
  solutions \cite[e.g.,][]{BLLT19_TR,HMRT19_TR,surrogate-design}. In this
  context, Theorem \ref{t:main} provides new evidence for the implicit
  regularization conjecture posed by \cite{surrogate-design} (see
  their Theorem 2 and associated discussion), with the amount of
  regularization equal $\frac1\gamma$, where $\gamma$ is implicitly
  defined in \eqref{eq:surrogate}: 
  \begin{align*}
  \underbrace{\E\Big[\argmin_\x\|\x\|^2\ \ \textnormal{s.t.}\ \
    \S\A\x=\S\b\Big] - \x^*}_{\text{Bias of sketched minimum norm solution}}
    \  \ \overset\epsilon\simeq\ \
    \underbrace{\argmin_\x\Big\{\|\A\x-\b\|^2+\tfrac1\gamma\|\x\|^2\Big\}
    - \x^*}_{\text{Bias of $l_2$-regularized solution}}.
  \end{align*}
While implicit regularization has received attention recently in the context of SGD algorithms for overparameterized machine learning models, it was originally discussed in the context of approximation algorithms more generally~\cite{Mah12}.
Recent work has made precise this notion in the context of RandNLA \cite{surrogate-design}, and our results here can be viewed in terms of implicit regularization of scalable RandNLA methods.

%\section{Related work}
\subsection{Related work}

A significant body of research has been dedicated to understanding the
guarantees for low-rank approximation via sketching,
particularly in the context of 
RandNLA \cite{DM16_CACM,RandNLA_PCMIchapter_chapter}. This line of work includes
i.i.d.~row sampling methods \cite{BoutsidisMD08,ridge-leverage-scores} which 
preserve the structure of the data, and data-oblivious methods such as
Gaussian and Rademacher sketches \cite{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}. However,
all of these results focus on worst-case 
upper bounds on the approximation error. One exception is a recent
line of works on non-i.i.d.~row sampling with Determinantal Point Processes
(DPP, \cite{dpps-in-randnla}). In this case, exact analysis of the low-rank approximation
error \cite{nystrom-multiple-descent}, as well as precise convergence analysis
of stochastic second order methods \cite{randomized-newton}, have been
obtained. Remarkably, the expressions they obtain are analogous to
\eqref{eq:surrogate}, despite using completely different techniques. However, their analysis
is limited only to DPP-based sketches, which are considerably more
expensive to construct and thus much less widely used. The connection
between DPPs and Gaussian sketches was recently explored by 
\cite{surrogate-design} in the context of analyzing the implicit
regularization effect of choosing a minimum norm solution in
under-determined linear regression. They conjectured that the
expectation formulas obtained for DPPs are a good proxy for the
corresponding quantities obtained under a Gaussian
distribution. Similar observations were made by
\cite{debiasing-second-order} in the context of sketching for
regularized least squares and second order optimization. While
both of these works only provide empirical evidence for this
particular claim, our Theorem \ref{t:main} can be 
viewed as the first theoretical non-asymptotic justification of that
conjecture.

The effectiveness of sketching has also been extensively studied in
the context of second order optimization. These methods differ
depending on how the sketch is applied to the Hessian matrix, and
whether or not it is applied to the gradient as well. The class of
methods discussed in Section \ref{s:newton}, including Randomized
Subspace Newton and the Generalized Kaczmarz method, relies on
projecting the Hessian downto a low-dimensional subspace, which makes our
results directly applicable. A related family of methods uses
the so-called Iterative Hessian Sketch (IHS) approach \cite{pilanci2016iterative,lacotte2019faster}. The
similarities between IHS and the Subspace Newton-type methods (see
\cite{Qu2015Feb} for a comparison) suggest that our techniques could be
extended to provide precise convergence guarantees also to the IHS.
Finally, yet another family of Hessian sketching methods has been studied by
\cite{roosta2019sub,sketched-ridge-regression,XRM17_theory_TR,YXRM18_TR,fred_newtonMR_TR,distributed-newton,determinantal-averaging}.
These methods  preserve the  rank of the 
 Hessian, and so their convergence
guarantees do not rely on the residual projection.

\section{Precise analysis of the residual projection}

In this section, we give a detailed statement of our main technical
result, along with a sketch of the proof. First, recall the definition of
sub-gaussian random variables and vectors.
\begin{definition}
We say that $x$ is a
$K$-sub-gaussian random variable if its sub-gaussian Orlicz norm $\| x
\|_{\psi_2} \le K$, where $\| x \|_{\psi_2} := \inf\{ t > 0:~\E[
\exp(x^2/t^2) ] \le 2 \}$. Similarly, we say that a random vector $\x$ is
$K$-sub-gaussian if for all $\|\a\| \leq 1$ we have % that $\x^\top
% \a$ is a sub-gaussian random variable with
$\|\x^\top \a \|_{\psi_2} \leq K$.
\end{definition}

For convenience, we state the main result in a slightly different form than Theorem
\ref{t:main}. Namely, we replace the $m\times n$ matrix $\A$ with a  positive semi-definite
$n\times n$ matrix
$\Sigmab^{\frac12}$. Furthermore, instead of a sketch $\S$ with
i.i.d.~sub-gaussian entries, we use a random matrix $\Z$ with
i.i.d.~sub-gaussian \emph{rows}, which is a strictly weaker condition
because it allows for the entries of each row to be correlated. Since
the rows of $\Z$ are also assumed to have mean zero and identity covariance,
each row of $\Z\Sigmab^{\frac12}$ has covariance $\Sigmab$. In
Section \ref{s:reduction} we show how to 
convert this statement back to the form of Theorem~\ref{t:main}.

\begin{theorem}\label{t:main-tech}
Let $\P_\perp = \I - \X^\dagger \X$ for $\X = \Z\Sigmab^{\frac12}$,
where $\Z \in \mathbb R^{k\times n}$ has i.i.d.~$K$-sub-gaussian rows with zero mean and identity covariance, and
$\Sigmab$ is an $n\times n$ positive semi-definite matrix. Define:
\ifisarxiv\vspace{-1.25mm}\fi
\begin{align*}
\bar\P_\perp= (\gamma\Sigmab + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P_\perp=n-k.
\end{align*}
\ifisarxiv\vspace{-4.75mm}

\noindent
\fi
Let $r =\tr(\Sigmab)/\|\Sigmab\|$
be the stable rank of
$\Sigmab^{\frac12}$ and fix $\rho=r/k > 1$. There exists a constant
$C_{\rho}>0$, depending only on $\rho$ and $K$, such that if $r\geq
C_\rho$, then
\ifisarxiv\vspace{-1.5mm}\fi
\begin{align}
\Big(1-\frac {C_\rho}{\sqrt
  r}\Big)\cdot\bar\P_\perp\preceq\E[\P_\perp]\preceq
  \Big(1+\frac {C_\rho}{\sqrt r}\Big)\cdot\bar\P_\perp.
\end{align}
\end{theorem}

We first provide the following informal derivation of the expression for $\bar \P_\perp$
given in Theorem~\ref{t:main-tech}. 
% Let us use $\P$ to denote the matrix $\X^\dagger \X=\I-\P_{\perp}$. Using a 
% rank-one update formula for the Moore-Penrose pseudoinverse (see
% Lemma~\ref{l:rank-one-update} in the appendix) we have $\E[\P] = \E\big[(\X^\top \X)^\dagger \X^\top \X\big]= \sum_{i=1}^k \E[(\X^\top \X)^\dagger \x_i \x_i^\top]  = k\, \E\! \left[\frac{ (\I - \P_{-k}) \x_k \x_k^\top }{ \x_k^\top (\I - \P_{-k}) \x_k } \right]$, where we use $\x_i^\top$ to denote the $i$-th row of $\X$, and
% $\P_{-k} = \X_{-k}^\dagger \X_{-k} $, where $\X_{-i} $ is the matrix
% $\X$ without its $i$-th row. Due to the 
% sub-gaussianity of $\x_k$, the quadratic form $\x_k^\top (\I -
% \P_{-k}) \x_k$ in the denominator is expected to concentrate around
% its expectation (with respect to $\x_k$) and thus $\E[\X^\dagger \X] \simeq k\, \E\! \left[\frac{ (\I - \P_{-k}) \x_k \x_k^\top }{ \tr \Sigmab (\I - \P_{-k}) } \right] = k\E \left[\frac{ (\I - \P_{-k}) \Sigmab }{ \tr \Sigmab (\I - \P_{-k}) } \right]$, where we use $\E[\x_k \x_k^\top] = \Sigmab$. Further note that, with
% $\P_{-k} \simeq \P$ for large $k$ and $\frac1k \tr \Sigmab (\I -
% \P_{-k}) \simeq \frac1k \tr \Sigmab \E[\P_\perp]$ from a concentration
% argument, we conclude that $\E[\X^\dagger \X] = \I - \E[\P_\perp] \simeq \frac{ k \E[\P_\perp] \Sigmab }{ \tr \Sigmab \E[\P_\perp] }$, from which it follows that $\E[\P_\perp] \simeq  \big( \frac{ k\Sigmab}{ \tr \Sigmab \E[\P_\perp] } + \I \big)^{-1}$, and thus $\E[\P_\perp] \simeq \bar \P_\perp$ for $\bar \P_\perp =
% (\gamma \Sigmab + \I)^{-1}$ and $\gamma^{-1} = \frac1k \tr \Sigmab
% \bar \P_\perp$. This leads to the (implicit) expression for $\bar
% \P_\perp$ and $ \gamma$ given in Theorem \ref{t:main-tech}.  
Let us use $\P$ to denote the matrix $\X^\dagger \X=\I-\P_{\perp}$. Using a 
rank-one update formula for the Moore-Penrose pseudoinverse (see
Lemma~\ref{l:rank-one-update} in the appendix) we have
\[\I-\E[\P_{\perp}]=\E[\P] = \E\big[(\X^\top \X)^\dagger \X^\top \X\big]= \sum_{i=1}^k
  \E[(\X^\top \X)^\dagger \x_i \x_i^\top]  = k\, \E\! \left[\frac{ (\I
      - \P_{-k}) \x_k \x_k^\top }{ \x_k^\top (\I - \P_{-k}) \x_k }
  \right],\]
where we use $\x_i^\top$ to denote the $i$-th row of $\X$, and
$\P_{-k} = \X_{-k}^\dagger \X_{-k} $, where $\X_{-i} $ is the matrix
$\X$ without its $i$-th row. Due to the 
sub-gaussianity of $\x_k$, the quadratic form $\x_k^\top (\I -
\P_{-k}) \x_k$ in the denominator concentrates around
its expectation (with respect to $\x_k$), i.e., $\tr\Sigmab(\I-\P_{-k})$,
% and thus $\E[\P] \simeq k\, \E\! \left[\frac{ (\I - \P_{-k}) \x_k \x_k^\top }{ \tr \Sigmab (\I - \P_{-k}) } \right] = k\E \left[\frac{ (\I - \P_{-k}) \Sigmab }{ \tr \Sigmab (\I - \P_{-k}) } \right]$,
where we use $\E[\x_k \x_k^\top] = \Sigmab$. Further note that, with
$\P_{-k} \simeq \P$ for large $k$ and $\frac1k \tr \Sigmab (\I -
\P_{-k}) \simeq \frac1k \tr \Sigmab \E[\P_\perp]$ from a concentration
argument, we conclude~that
\begin{align*}
\I - \E[\P_\perp] \simeq \frac{ k \E[\P_\perp] \Sigmab }{
  \tr \Sigmab \E[\P_\perp] }\qquad\Longrightarrow\qquad \E[\P_\perp] \simeq  \Big( \frac{ k\Sigmab}{ \tr \Sigmab \E[\P_\perp] } + \I \Big)^{-1},
\end{align*}
  % from which it follows that $\E[\P_\perp] \simeq  \big( \frac{ k\Sigmab}{ \tr \Sigmab \E[\P_\perp] } + \I \big)^{-1}$,
  and thus $\E[\P_\perp] \simeq \bar \P_\perp$ for $\bar \P_\perp =
(\gamma \Sigmab + \I)^{-1}$ and $\gamma^{-1} = \frac1k \tr \Sigmab
\bar \P_\perp$. This leads to the (implicit) expression for $\bar
\P_\perp$ and $ \gamma$ given in Theorem \ref{t:main-tech}.

\medskip

\subsection{Proof sketch of Theorem~\ref{t:main-tech}}
\label{subsec:proof-sketch}

To make the above intuition rigorous, we next present a proof sketch
for Theorem~\ref{t:main-tech}, with the detailed proof deferred to
Appendix~\ref{sec:proof-of-theo-main-tech}.  The proof can be divided
into the following three steps.

\paragraph{Step 1.} First note that, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the spectral
norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}$, so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ for $\rho = r/k > 1$ from the definition of $\bar \P_\perp$, we have
\[
  \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
    \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P^{-1})\bar\P_\perp^{\frac12}\|\leq
  \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}} =: \epsilon.
\]
This means that all eigenvalues 
  of the p.s.d.\@ matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
  in the interval $[1-\epsilon,1+\epsilon]$, so $(1-\epsilon)\I\preceq
    \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.$ Multiplying by $\bar\P_{\perp}^{\frac12}$ from both sides, we obtain the
  desired bound.
  % Note that for p.s.d.~matrices $\A,\B,\C$, if $\A\preceq \B$ then
  % $\C\A\C\preceq \C\B\C$, so letting $\C=\bar\P_\perp^{\frac12}$ in the above,
  % we get:
  % \begin{align*}
  %   (1-\epsilon)\bar\P_\perp\preceq
  %   \E[\P_\perp]\preceq (1+\epsilon)\bar\P_\perp
  % \end{align*}
  % as desired.
 \paragraph{Step 2.} Then, we carefully design an event $E$ that (i)
 is provable to occur with high probability and (ii) ensures that the
 denominators in the following decomposition are bounded away from
 zero:
 % based on (the discussion of) which the object of interest $\I - \E[\P_\perp ] \bar \P_\perp^{-1}$ can be decomposed as
  \begin{align*}
  \I-\E[\P_\perp]\bar\P_\perp^{-1}
    &= \E[\P] - \gamma\E[\P_\perp]\Sigmab = \E[\P \cdot \one_E] +
    \E[\P  \cdot \one_{\neg E}]
    -\gamma\E[\P_\perp]\Sigmab\\
  %&=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg] -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
  &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
    s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
    - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2} \\ 
    &\quad+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
    %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
\end{align*}
where we let $\hat s = \x_k^\top(\I - \P_{-k})\x_k$ and $\bar s=k/\gamma$. 
\paragraph{Step 3.}
It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3
,\T_4$ respectively to reach the conclusion. More precisely, the terms
$\| \T_2 \| $ and $\| \T_4 \|$ are proportional to $\Pr (\neg E)$,
while the term $\| \T_3 \|$ can be bounded using the rank-one update
formula for the pseudoinverse (Lemma~\ref{l:rank-one-update} in the
appendix). The remaining term $\| \T_1 \|$ is more subtle and can be
bounded with a careful application of  the Hanson-Wright type
\cite{rudelson2013hanson} sub-gaussian concentration 
inequalities (Lemmas~\ref{l:concentration-sub-gaussian}~and~\ref{l:trace} in the
appendix). This allows for a bound on the operator norm $\| \I -
\E[\P_\perp ] \bar \P_\perp^{-1}\| $ and hence the conclusion.

\subsection{Proof of Theorem \ref{t:main}}\label{s:reduction}
We now discuss how Theorem \ref{t:main} can be obtained from Theorem
\ref{t:main-tech}. The crucial difference between the statements is
that in Theorem \ref{t:main} we let $\A$ be an arbitrary rectangular
matrix, whereas in Theorem \ref{t:main-tech} we instead use a square,
symmetric and positive semi-definite matrix $\Sigmab$. To convert between the
two notations, consider the SVD decomposition
$\A=\U\D\V^\top$ of $\A \in \R^{m \times n}$ (recall that we assume $m
\ge n$), where $\U\in\R^{m\times n}$ and $\V\in\R^{n\times 
  n}$ have orthonormal columns and $\D$ is a diagonal matrix. Now, let
$\Z = \S\U$, $\Sigmab=\D^2$ and  
$\X=\Z\Sigmab^{\frac12}=\S\U\D$. Using the fact that
$\V^\top\V=\V\V^\top=\I$, it follows that:
\begin{align*}
  \I - (\S\A)^\dagger\S\A = \V(\I-\X^\dagger\X)\V^\top\quad\text{and}\quad
  (\gamma\A^\top\A+\I)^{-1} = \V(\gamma\Sigmab + \I)^{-1}\V^\top.
\end{align*}
Note that since $\|\U\v\|=\|\v\|$, the rows of $\Z$ are
sub-gaussian with the same constant as the rows of $\S$. Moreover,
using the fact that $\B\preceq\C$ implies
$\V\B\V^\top\preceq\V\C\V^\top$ for any p.s.d.~matrices $\B$ and $\C$,
Theorem \ref{t:main} follows as a corollary of Theorem \ref{t:main-tech}.

% We can also assume wlog $\Sigmab$ is $d$ by $d$ and full rank, 
%   otherwise consider the SVD $\Sigmab = \U \S \V^\top$.
%   Conjugating \cref{eq:thm1-psd} by $\V^\top$ and using
%   orthogonality of the columns of $\V$,
%   it suffices to show the result
%   (after conjugating $\bar{\P}$ by $\V^\top$ as well)
%   for $\X \V$.
%   We have
%   \begin{align*}
%     \X \V
%     = \Z \Sigmab \V
%     = \underbrace{\Z \U}_{\eqqcolon \tilde{\Z}} \S \V^\top \V
%     =\tilde{\Z} \S
%   \end{align*}
%   where $\tilde{\Z}$ is $K$-sub-gaussian and $\S$ is diagonal, full rank,
%   and has $\tr(\S) = \tr((\Sigmab^\top \Sigmab)^{1/2})$.
%   Hence, it suffices to prove the result for
%   $\Sigmab$ diagonal and full rank and the corresponding
%   bounds hold without modification.

\section{Explicit formulas under known spectral decay}
\label{s:explicit}
The expression we give for the expected residual projection,
$\E[\P_{\perp}]\simeq (\gamma\A^\top\A+\I)^{-1}$, is implicit
in that it depends on the parameter $\gamma$ which is the solution of
the following equation:
\begin{align}
  \sum_{i\geq 1}\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1} =
  k,\qquad\text{where $\sigma_i$ are the singular values of $\A$.}\label{eq:equation}
\end{align}
In general, it is impossible to solve this equation analytically, i.e., to write
$\gamma$ as an explicit formula of $n$, $k$ and
the singular values of $\A$. However, we show that when the 
singular values exhibit a known rate of decay, then it is
possible to obtain explicit formulas for $\gamma$. In particular, this
allows us to provide precise and easily interpretable rates of decay for the low-rank
approximation error of a sub-gaussian sketch.

Matrices that have known spectral decay, most commonly with either
exponential or polynomial rate, arise in many machine
learning problems \cite{randomized-newton}. Such behavior can be naturally
occurring in data, or it can be induced by feature expansion using, say, the
RBF kernel (for exponential decay) \cite{Santa97Gaussianregression} or the
Mat\'ern kernel (for polynomial decay) \cite{RasmussenWilliams06}. Understanding these
two classes of decay plays an important role in distinguishing the
properties of light-tailed and heavy-tailed data distributions.
Note that in the kernel setting we may often represent our data via
the $m\times m$ kernel matrix $\K$, instead of the $m\times n$ data
matrix $\A$, and study the sketched Nystr\"om method
\cite{revisiting-nystrom} for low-rank approximation. To handle 
the kernel setting in our analysis, it suffices to replace the squared singular
values $\sigma_i^2$ of $\A$ with the eigenvalues of $\K$.
\begin{figure}[t]
\centering
\subfigure[%
Singular values are given by $\sigma_i^2=C\cdot \alpha^{i-1}$.]{%
    \includegraphics[width=0.48\textwidth]{explicit_exp}
  }
\hfill
\subfigure[%
Singular values are given by $\sigma_i^2=C\cdot i^{-\beta}$.
]{%
      \includegraphics[width=0.48\textwidth]{explicit_poly}
    }
\caption{
Theoretical predictions of low-rank approximation error of a Gaussian
sketch under known spectral decays, compared to the empirical results.
The constant $C$ is scaled so that $\|\A\|_F^2=1$ and we let
$n=m=1000$. For the theory, we plot the explicit formulas
\eqref{eq:explicit-exp} and \eqref{eq:explicit-poly} (dashed lines),
as well as the implicit expression from Corollary \ref{c:low-rank}
(thin solid lines) obtained by numerically solving
\eqref{eq:equation}. Observe that the explicit and implicit
predictions are nearly (but not exactly) identical.
}
%\vspace{-5mm}
\label{f:explicit}
\end{figure}

\subsection{Exponential spectral decay}
Suppose that the squared singular values of $\A$ exhibit exponential
decay, i.e. $\sigma_i^2 = C\cdot \alpha^{i-1}$, where $C$ is a constant
and $\alpha\in(0,1)$. For simplicity of presentation, we will let
$m,n\rightarrow\infty$. Under this spectral decay, we can approximate
the sum in \eqref{eq:equation} by the analytically computable integral
$\int_y^\infty\!\frac1{1+(C\gamma)^{-1}\alpha^{-x}} dx$,
% \begin{align*}
%   \underbrace{\int_{0}^\infty\frac1{1+(C\gamma)^{-1}\alpha^{-x}}
%   dx}_{\frac{\ln(1/(C\gamma+1))}{\ln(\alpha)}}\ \leq\ 
%   \underbrace{\sum_{i\geq
%   1}\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1}}_{k}
%  \ \leq\
%   \underbrace{\int_{-1}^\infty\frac1{1+(C\gamma)^{-1}\alpha^{-x}}
%   dx}_{\frac{\ln(\alpha/(C\gamma+\alpha))}{\ln(\alpha)}}.
% % =\frac1{2\ln(\alpha)}\bigg(\ln\Big(\frac{\alpha}{C\gamma+\alpha}\Big)
% %     +\ln\Big(\frac{1}{C\gamma+1}\Big)\bigg). 
% % 
%   \end{align*}
% Using each integral as a proxy for the sum, solving for $\gamma$, and
% taking a geometric mean of the two results,
obtaining $\gamma\approx(\alpha^{-k}-1)\sqrt\alpha/C$. Applying this to the
formula from Corollary~\ref{c:low-rank}, we can express the low-rank
approximation error  
for a sketch of size $k$ as follows:
\begin{align}
  \E\big[\|\A-\A\P\|_F^2\big] \approx
\frac C{\sqrt\alpha}\cdot
  \frac{k}{\alpha^{-k}-1},\quad\text{when}\quad\sigma_i^2=C\cdot\alpha^{i-1}\
  \ \text{for all $i$.}\label{eq:explicit-exp}
\end{align}
In Figure \ref{f:explicit}a, we plot the above formula against the
numerically obtained implicit expression from
Corollary~\ref{c:low-rank}, as well as empirical results for a Gaussian 
sketch. First, we observe that the theoretical predictions closely
align with empirical values even after the sketch size crosses the
stable rank $r\approx\frac1{1-\alpha}$, suggesting that Theorem \ref{t:main} can
be extended to this regime. Second, while it is not surprising that
the error decays at a similar 
rate as the singular values, our predictions offer a much more precise
description, down to lower order effects and
even constant factors. For instance, we observe that the error
(normalized by $\|\A\|_F^2$, as in the figure) only
starts decaying exponentially after $k$ crosses the stable rank, and
until that point it decreases at a linear rate with slope
$-\frac{1-\alpha}{2\sqrt\alpha}$.

\subsection{Polynomial spectral decay}
We now turn to polynomial spectral decay, which is a natural model for
analyzing heavy-tailed data distributions. Let $\A$ have squared
singular values $\sigma_i^2=C\cdot i^{-\beta}$ for some $\beta\geq 2$, and let
$m,n\rightarrow\infty$. As in the case of exponential decay,
we use the integral $\int_y^\infty\!\frac1{1+(C\gamma)^{-1}x^{-\beta}}dx$
to approximate the sum in \eqref{eq:equation}, and 
solve for $\gamma$, obtaining $\gamma \approx
\big((k+\frac12)\frac\beta\pi\sin(\frac\pi\beta)\big)^\beta$. Combining
this with Corollary \ref{c:low-rank} we get:
\begin{align}
  \E\big[\|\A-\A\P\|_F^2\big] \approx
  C\cdot\frac{k}{(k+\frac12)^\beta}\bigg(\frac{\pi/\beta}{\sin(\pi/\beta)}\bigg)^\beta,
  \quad\text{when}\quad\sigma_i^2=C\cdot
  i^{-\beta}\ \ \text{for all $i$.}\label{eq:explicit-poly}
\end{align}

Figure \ref{f:explicit}b compares our predictions to the empirical
results for several values of $\beta$. In all of these cases, the
stable rank is close to 1, and yet the theoretical predictions align
very well with the empirical results. Overall, the asymptotic rate of decay of the error is
$k^{1-\beta}$.  However it is easy to verify that the lower order
effect of $(k+\frac12)^\beta$ appearing instead of $k^\beta$
in \eqref{eq:explicit-poly} significantly changes the trajectory for
small values of $k$. Also, note that as $\beta$ grows large, the
constant $\big(\frac{\pi/\beta}{\sin(\pi/\beta)}\big)^\beta$ goes to
$1$, but it plays a significant role for $\beta=2$ or $3$
(roughly, scaling the expression by a factor of $2$). Finally, we
remark that for $\beta\in(1,2)$, our integral approximation of
\eqref{eq:equation} becomes less accurate. We expect that a corrected
expression is possible, but likely more complicated and less interpretable.
\vspace{3mm}

\ifisarxiv
\else
\begin{minipage}{.48\textwidth}
\hspace{-4.5mm}\includegraphics[width=.57\textwidth]{abalone-nystrom}\nobreak\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{eunite-nystrom}\vspace{-1mm}
  \captionof{figure}{Theoretical predictions versus approximation error for the
    sketched Nystr\"om with the RBF kernel
    (spectral decay shown at the bottom).}\label{f:nystrom}
\end{minipage}\hspace{4mm}
\begin{minipage}{.48\textwidth}
\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{covtype-linear}\nobreak\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{splice-linear}\vspace{-1mm}
  \captionof{figure}{Theoretical predictions versus approximation error for
the Gaussian and Rademacher sketches
  (spectral decay shown at the bottom).}\label{f:linear}
\end{minipage}
\fi

\section{Empirical results}\label{s:experiments}

In this section, we numerically verify the accuracy of our theoretical
predictions for the low-rank approximation error of sketching on
benchmark datasets from the libsvm repository 
\cite{libsvm} (further numerical results are in
Appendix~\ref{a:experiments}). We repeated every experiment 10 times,
and plot 
both the average and standard deviation of the results. We use the
following $k\times m$ sketching matrices $\S$:\vspace{-1mm}
\begin{enumerate}
\item \emph{Gaussian sketch:} with i.i.d.\@ standard normal entries;
\item \emph{Rademacher sketch:} with
  i.i.d.\@ entries equal $1$ with probability 0.5 and $-1$ otherwise.
\end{enumerate}
\vspace{-2mm}

\ifisarxiv
\begin{figure}
  \centering
\includegraphics[width=.43\textwidth]{abalone-nystrom}\nobreak\includegraphics[width=.43\textwidth]{eunite-nystrom}
  \caption{Theoretical predictions versus approximation error for the
    sketched Nystr\"om with the RBF kernel
    (spectral decay shown at the bottom).}\label{f:nystrom}
\end{figure}
\fi

\ifisarxiv
\begin{figure}
  \centering
\includegraphics[width=.43\textwidth]{covtype-linear}\nobreak\includegraphics[width=.43\textwidth]{splice-linear}
  \caption{Theoretical predictions versus approximation error for
the Gaussian and Rademacher sketches
  (spectral decay shown at the bottom).}\label{f:linear}
\end{figure}
\fi

\paragraph{Varying spectral decay.}  To demonstrate the role of spectral
decay and the stable rank on the approximation error, we performed
feature expansion using the radial basis function (RBF) kernel
$k(\a_i,\a_j)=\exp(-\|\a_i-\a_j\|^2/(2\sigma^2))$, obtaining an $m\times m$
kernel matrix $\K$. We used the sketched Nystr\"om method to construct
a low-rank approximation $\tilde
\K=\K\S^\top(\S\K\S^\top)^\dagger\S\K$, and computed the normalized trace norm 
error $\|\K-\tilde\K\|_*/\|\K\|_*$. The theoretical predictions are coming from
\eqref{eq:nystrom}, which in turn uses Theorem \ref{t:main}.
% The scale parameter
% $\sigma>0$ significantly affects the spectral properties of the kernel,
% so varying it allows us to observe qualitatively different behavior of
% the approximation error. Note that t
% The choice of $\sigma$ used in
% practice depends on the down-stream task, and is often determined by
% cross-validation.
Following \cite{revisiting-nystrom}, we use the RBF kernel because
varying the scale parameter $\sigma$ 
allows us to observe the approximation error under qualitatively
different spectral decay profiles of the kernel. In Figure \ref{f:nystrom},
we present the results for the Gaussian sketch on two datasets, with three values
of $\sigma$, and in all cases our theory aligns with the empirical
results. Furthermore, as smaller $\sigma$ leads to slower spectral decay and 
larger stable rank, it also makes the approximation error decay more
linearly for small sketch sizes. This behavior is predicted by our explicit
expressions \eqref{eq:explicit-exp} for the error under exponential spectral decay from
Section \ref{s:explicit}. Once the sketch sizes are sufficiently
larger than the stable rank of $\K^{\frac12}$, the error starts decaying at an
exponential rate. Note that Theorem \ref{t:main} only guarantees
accuracy of our expressions for sketch sizes below the stable rank,
however the predictions are accurate regardless of this constraint.
\vspace{-2mm}
\paragraph{Varying sketch type.} In the next set of
empirical results, we compare the performance of Gaussian and Rademacher
sketches, and also verify the theory when sketching the data matrix $\A$
without kernel expansion, plotting
$\|\A-\A(\S\A)^\dagger\S\A\|_F^2/\|\A\|_F^2$.
Since both of the sketching methods have 
sub-gaussian entries, Corollary~\ref{c:low-rank} predicts that they
should have comparable performance in this task and match our
expressions. This is exactly what we observe in Figure \ref{f:linear}
for two datasets and a range of sketching sizes, as well as in other
empirical results shown in Appendix \ref{a:experiments}.

\section{Conclusions}

We derived the first theoretically supported precise expressions for the
expected residual projection matrix, which is a central component in
the analysis of RandNLA dimensionality reduction via sketching. Our
analysis provides a new understanding of low-rank approximation, the Nystr\"om method, and the
convergence properties of many randomized iterative algorithms. As a direction for future work, we conjecture that our
main result can be extended to sketch sizes larger than the stable
rank of the data matrix.

\ifisarxiv\else
\section*{Broader Impact}
In this paper, we investigate the spectral properties of residual
(random) projection matrices, commonly appearing in various
sketching-based methods. The precise theoretical description given in
this paper provides performance guarantees for popular algorithms
such as low-rank approximation and many randomized (iterative)
optimization methods, and contributes to the development of more
robust and reliable large-scale learning systems. The
theoretical framework developed in this work presents no foreseeable
negative societal consequence. 
\fi

\paragraph{Acknowledgments.}
We would like to acknowledge DARPA, NSF, and ONR for providing partial support of this work.

\bibliographystyle{alpha}
\bibliography{../pap}

% In this section, we report experimental evidence illustrating our results
% provide meaningful approximations on a variety of conditions and real-world
% datasets.

% In \Cref{f:main} we use synthetic data to investigate the size of the error
% ($\epsilon$) as well as its rate of decay. $\A$ is taken to be diagonal with
% eigenvalues decaying exponentially from $1$ to $0.1$. As a result, we are both
% able to let $r$ grow arbitrarily large as well as choose a sketch size $k$ such
% that the aspect ratio $\rho$ remains constant. As the constants hidden
% in $O(1/\sqrt{r})$ depend on $\rho$, such a scaling enables isolation
% of the relation between approximation quality $\epsilon$ and stable rank $r$.
% We construct bootstrapped operator norm confidence intervals using
% \cite{lopes2019bootstrapping}.

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=1.0\linewidth]{../equivalents_code/theorem1.png}
%   \caption{Visualization of the $\epsilon$ in \Cref{t:main} for two different
%     sub-gaussian sketching matrices $\mathbf{S}$.
%   }
%   \label{f:main}
% \end{figure}

% The results in \Cref{f:main} show that the approximation quality becomes
% small relatively quickly, and that the reciprocal-like rate of decay holds
% both across a range of aspect ratios $\rho$ as well as over different
% sub-gaussian sketching matrices $\S$.

% Beyond synthetic data, we interrogate the practical applicability of our
% results by performing experiments on real-world datasets to empirically verify
% that \Cref{c:low-rank} provides useful predictions. We chose to present
% datasets representing a diversity of conditions here
% (see \Cref{tab:datasets}), though we conducted a more extensive investigation
% and report full results in Appendix \todo{feynman}.

% \begin{table}[htpb]
%   \centering
%   \caption{Summary statistics for \texttt{libsvm} \citep{libsvm} datasets used}
%   \label{tab:datasets}
%   \begin{tabular}{ccccc}
%     \toprule
%     Name & Rows ($n$) & Columns ($d$) & Rows / Columns & Stable rank ($r$) \\
%     \midrule
%     %abalone\_scale & 4177 & 8 & 5.2e+02 & 1.57 \\
%     %housing\_scale & 506 & 13 & 39 & 1.75 \\
%     %mg\_scale & 1385 & 6 & 2.3e+02 & 2.32 \\
%     %mpg\_scale & 392 & 7 & 56 & 1.85 \\
%     %pyrim\_scale & 74 & 27 & 2.7 & 1.50 \\
%     %triazines\_scale & 186 & 60 & 3.1 & 1.40 \\
%     cpusmall\_scale & 8192 & 12 & \textbf{6.8e+02} & 1.06 \\
%     bodyfat\_scale & 252 & 14 & 18 & 1.32 \\
%     splice\_scale & 1000 & 60 & 17 & \textbf{18.65} \\
%     real-sim & 72309 & \textbf{20958} & 3.45 & 7.46 \\
%     cifar10 & 50000 & 3072 & 16.28 & 1.15 \\
%     SVHN & \textbf{73257} & 3072 & 23.85 & 1.06 \\
%     \bottomrule
%   \end{tabular}
% \end{table}

% % \begin{figure}[htbp]
% %   \centering
% %   \includegraphics[width=1.0\linewidth]{../equivalents_code/corollary1.png}
% %   \caption{
% %     \Cref{c:low-rank} approximations of the reconstruction error (blue) 
% %     closely follows the mean across many real-world datasets.
% %   }
% %   \label{f:low-rank}
% % \end{figure}

% In \Cref{f:low-rank}, we display the predicted reconstruction error using our
% theory as well as a box plot summarizing the empirical distribution of samples
% of $\|\A - \A\P\|_F^2$. Notice that our theoretical predictions are very
% close to the empirical means even for extremely small sketch sizes, and
% that the results are applicable on datasets as small as
% $252 \times 14$ (bodyfat\_scale) to as large as $72309 \times 20958$
% (real-sim). As expected, cifar10 and SVHN results appear similar because
% they are both image datasets with the same number of dimensions
% and similar stable ranks.
% Also, when the stable rank is large (splice\_scale and real-sim)
% we see that both the empirical and predicted reconstruction errors decay
% at a significantly slower rate than when it is small. This further supports
% the view that stable rank is the ``correct'' notion to consider in the context
% of low rank approximation.

% \appendix
% \input{appendix}

\ifisarxiv\else\newpage\fi
\appendix

\section{Proof of Theorem~\ref{t:main-tech}}
\label{sec:proof-of-theo-main-tech}

We first introduce the following technical lemmas.

\begin{lemma}\label{l:rank-one-update}
    For $\X \in \mathbb R^{k \times n}$ with $k<n$, denote $\P = \X^\dagger \X$ and $\P_{-k} = \X_{-k}^\dagger \X_{-k}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its i-th row $\x_i \in \mathbb R^n$. Then, conditioned on the event $E_k: \left\{ \left| \frac{\tr \Sigmab (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} - 1 \right| \le \frac12 \right\}$:
    \begin{align*}
      (\X^\top\X)^\dagger\x_k = \frac{(\I - \P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}, \quad \P - \P_{-k} = \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top (\I - \P_{-k}) \x_k}.
    \end{align*}
\end{lemma}

\begin{proof}
% Recall $\P = \I - \X^\dagger \X = \I - (\X^\top \X)^\dagger \X^\top \X = \I - (\X^\top \X)^\dagger (\A + \x_k \x_k^\top)$ and $\P_{-k} = \I - \X_{-k}^\dagger \X_{-k} = \I - \A^\dagger \A$ 
Since conditioned on $E_k$ we have $\x_k^\top (\I - \P_{-k}) \x_k \neq 0$, from \cite[Theorem 1]{10.2307/2099767} we deduce
\begin{align*}
  (\X^\top \X)^\dag = \left( \A + \x_k \x_k^\top \right)^\dagger
  &= \A^\dag
  - \frac{\A^\dag \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top ( \I - \P_{-k}) \x_k}
  - \frac{( \I -\P_{-k}) \x_k \x_k^\top \A^\dag}{\x_k^\top ( \I - \P_{-k}) \x_k} \\ 
  &+ (1 + \x_k^\top \A^\dag \x_k) \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k})}{(\x_k^\top (\I - \P_{-k}) \x_k)^2}
\end{align*}
for $\A = \X_{-k}^\top \X_{-k}$ so that $\I - \P_{-k} = \I - \A^\dagger \A$, where we used the fact that $\I - \P_{-k}$ is a projection matrix so that $(\I - \P_{-k})^2 = \I - \P_{-k}$. As a consequence, multiplying by $\x_k$ and simplifying we get
\[
  (\X^\top \X)^\dag \x_k = \frac{ ( \I -\P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}.
\]
By definition of the pseudoinverse, $\P = \X^\dagger \X = (\X^\top \X)^\dagger \X^\top \X$ so that 
\[
  \P - \P_{-k} = \X^\dagger \X - \X_{-k}^\dagger \X_{-k} = \frac{( \I - \P_{-k}) \x_k \x_k^\top ( \I - \P_{-k})}{\x_k^\top ( \I - \P_{-k}) \x_k}
\]
where we used $\A (\I - \P_{-k}) = \A - \A \A^\dagger \A = 0$ and thus the conclusion.
\end{proof}

% \begin{lemma}[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Corollary 2.6]{zajkowski2018bounds}}] \label{l:concentration-sub-gaussian}
% For a $K$-sub-gaussian random vector $\x \in \mathbb R^d$ (i.e., $\| \x \|_{\psi_2} \leq K$) and $\A \in \mathbb R^{d \times d}$, we have
% \begin{align*}
%   \Pr\left[
%     \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert
%     \geq t
%     \right] \leq 2 \exp\left(
%     -\min\left\{
%       \frac{t^2}{16 C^2  K^4 (\tr \A)^2},
%       \frac{t}{4 C K^2 (\tr \A)},
%     \right\}
%   \right)
% \end{align*}
% and 
% \begin{align*}
%   \E\left[ \left( \x^\top \A \x - \E[\x^\top \A \x] \right)^2 \right]
%   &\leq C~K^4~\tr(\A)^2
% \end{align*}
% for some universal constant $C$.
% \end{lemma}

% \begin{lemma}
% %[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}]
% \label{l:concentration-sub-gaussian}
% For a $K$-sub-gaussian random vector $\x \in \mathbb R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and positive semi-definite matrix $\A \in \mathbb R^{n \times n}$ such thats $\| \A \| \le 1 $, we have
% \[
%   \Pr\left[
%     \lvert \x^\top \A \x - \tr \A \rvert
%     \geq \frac12 \tr \A
%     \right] \leq 2 \exp\left( - \min \left\{ \frac{r_\A}{4 C^2 K^4}, \frac{\sqrt r_\A}{2 C K^2}  \right\} \right)
% \]
% with $r_\A = \frac{\tr \A}{\| \A \|}$ the nuclear rank of $\A$, and 
% \begin{align*}
%   \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
%   &\leq cK^4\tr\A^2
% \end{align*}
% for some $C,c > 0$ independent of $k$.
% \end{lemma}

% \begin{proof}
% % From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \Sigmab$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that 
% % \[
% %   \Pr \left\{ \left| \x^\top \A \x - \tr \A \Sigmab \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
% % \]
% % for some universal constant $C > 0$. Taking $t = \frac12 \tr \A \Sigmab$ we obtain
% From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that 
% \[
%   \Pr \left\{ \left| \x^\top \A \x - \tr \A  \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
% \]
% for some universal constant $C > 0$. Taking $t = \frac12 \tr \A$ we have
% \[
%   \frac{t^2}{C^2 K^4 \tr\A^2 } = \frac{ (\tr \A)^2 }{4 C^2 K^4 \tr\A^2 } \ge \frac{ \tr \A }{4 C^2 K^4 \| \A \| } = \frac{r_\A}{4 C^2 K^4}, \quad \frac{t}{C K^2 \sqrt{\tr\A^2}  } \ge \frac{\sqrt r_\A}{2 C K^2}
% \]
% where we use the fact that $\tr \A^2 \le \| \A \| \tr \A$ and denote $r_\A = \frac{\A}{\| \A \|}$ the nuclear rank of $\A$.

% Integrating this bound yields, we get
% \[
%   \E\left[ (\x^\top \A \x - \tr \A )^2 \right] \le c~K^4~\tr \A^2
% \]
% and thus the conclusion.
% \end{proof}
%
%
%
\begin{lemma}
%[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}]
\label{l:concentration-sub-gaussian}
For a $K$-sub-gaussian random vector $\x \in \mathbb R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and positive semi-definite matrix $\A \in \mathbb R^{n \times n}$, we have
\[
  \Pr\left[
    \lvert \x^\top \A \x - \tr \A \rvert
    \geq \frac13 \tr \A
    \right] \leq 2 \exp\left( - \min \left\{ \frac{r_{\!\A}}{9 C^2 K^4}, \frac{\sqrt{r_{\!\A}}}{3 C K^2}  \right\} \right)
\]
with $r_{\!\A} = \tr \A/\| \A \|$ the stable rank of $\A$, and 
\begin{align*}
  \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
  &\leq c~K^4~\tr \A^2
\end{align*}
for some $C,c > 0$ independent of $K$.
\end{lemma}

\begin{proof}
% From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \Sigmab$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that 
% \[
%   \Pr \left\{ \left| \x^\top \A \x - \tr \A \Sigmab \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
% \]
% for some universal constant $C > 0$. Taking $t = \frac12 \tr \A \Sigmab$ we obtain
This follows from a Hanson-Wright type \cite{rudelson2013hanson} sub-gaussian concentration inequality. More precisely,
from \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that 
\[
  \Pr \left\{ \left| \x^\top \A \x - \tr \A  \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
\]
for some universal constant $C > 0$. Taking $t = \frac13 \tr \A$ we have
\[
  \frac{t^2}{C^2 K^4 \tr\A^2 } = \frac{ (\tr \A)^2 }{9 C^2 K^4 \tr\A^2 } \ge \frac{ \tr \A }{9 C^2 K^4 \| \A \| } = \frac{r_{\!\A}}{9 C^2 K^4}, \quad \frac{t}{C K^2 \sqrt{\tr\A^2}  } \ge \frac{\sqrt{r_{\!\A}}}{3 C K^2}
\]
where we use the fact that $\tr \A^2 \le \| \A \| \tr \A$.

Integrating this bound yields:
\[
  \E\left[ (\x^\top \A \x - \tr \A )^2 \right] \le c~K^4~\tr \A^2
\]
and thus the conclusion.
\end{proof}

% \begin{lemma}\label{l:rank-one-projection}
%   For any $d\times d$ p.s.d.~matrix $\A$ and
%   any random vector $\z$ with jointly
%   $K$-sub-gaussian (sub-gaussian Orlicz norm
%   $\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$),
%   \begin{align*}
%     \Bigg\|\E\bigg[\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\bigg]\Bigg\|
% \leq
%     \sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\v\Big)^2\bigg]}
%     \leq C\cdot K^2 \cdot \frac{\|\A\|}{\tr\,\A}.
%   \end{align*}
%   for some universal constant $C$.
% \end{lemma}
%\textbf{** Lemma 3 seems no longer necessary?**}

\begin{lemma}\label{l:trace}
With the notations of Lemma~\ref{l:rank-one-update}, for $X=\tr\,\Sigmab(\P_{-k}-\E[\P_{-k}])$
and $\|\Sigmab\| = 1$, we have
\begin{align*}
  \E[X^2]\leq
  Ck\quad\text{and}\quad\Pr\{|X|\geq t\}\leq 2 \ee^{-\frac{t^2}{ck}}.
  \end{align*}
  for some universal constant $C,c > 0$.
\end{lemma}
\begin{proof}
To simplify notations, we work on $\P$ instead of $\P_{-k}$, the same line of argument applies to $\P_{-k}$ by changing the sample size $k$ to $k-1$.

First note that
\begin{align*}
  X
  &= \tr \Sigmab (\P - \E \P)
  = \E_k [\tr \Sigmab \P]  - \E_0[ \tr \Sigmab \P ]\\
  %%%
  &= \sum_{i=1}^k \left( \E_i[\tr \Sigmab \P] - \E_{i-1}[\tr \Sigmab \P]\right)
  = \sum_{i=1}^k (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})
\end{align*}
where we used the fact that $\E_i [\tr \Sigmab \P_{-i}] = \E_{i-1} [\tr \Sigmab \P_{-i}]$, for $\E_i[\cdot]$ the conditional expectation with respect to $\mathcal F_i$ the $\sigma$-field generating the rows $\x_1 \ldots, \x_i$ of $\X$. This forms a martingale difference sequence (it is a difference sequence of
the Doob martingale for $\tr \Sigmab (\P - \P_{-i})$ with respect to filtration $\mathcal{F}_i$)
hence it falls within the scope of the Burkholder inequality \cite{burkholder1973distribution}, recalled as follows.

\begin{lemma}\label{l:burkholder}
For $\{ x_i \}_{i=1}^k$ a real martingale difference sequence with respect to the increasing $\sigma$ field $\mathcal F_i$, we have, for $L > 1$, there exists $C_L > 0$ such that
\[
  \E \bigg[
    \Big| \sum_{i=1}^k x_i \Big|^L
  \bigg]
  \le C_L \E \bigg[
    \Big( \sum_{i=1}^k |x_i|^2 \Big)^{L/2}
  \bigg].
\]
\end{lemma}

From Lemma~\ref{l:rank-one-update}, $\P - \P_{-i} = \frac{(\I - \P_{-i}) \x_i \x_i^\top (\I - \P_{-i}) }{ \x_i^\top (\I - \P_{-i}) \x_i}$ is positive semi-definite, we have $\tr \Sigmab (\P - \P_{-i}) \le \| \Sigmab \| = 1$ so that with Lemma~\ref{l:burkholder} we obtain with $x_i = (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})$ that, for $L > 1$
\[
  \E |X|^L \le C_L k^{L/2}.
\]
In particular, for $L=2$, we obtain $\E |X|^2 \le C k$.
%for $c = (36 \sqrt{2})^2$ (details in proof of Theorem~3.2 in \cite{burkholder1973distribution}).

For the second result, since we have almost surely bounded martingale differences
($\lvert x_i \rvert \leq 2$), by the Azuma-Hoeffding inequality
\begin{align*}
  \Pr\{
    \lvert X \rvert
    \geq t
  \}
  \leq 2 \ee^{\frac{-t^2}{8 k}}
\end{align*}
as desired.
\end{proof}

% \subsection{Complete proof of Theorem~\ref{t:main-tech}}

% Equipped with the lemmas above, we are ready to prove Theorem~\ref{t:main-tech}. First note that:
% \begin{enumerate}[leftmargin=*]
%   \item  Since $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, we can assume without loss of generality (after rescaling $\bar\P_\perp$ correspondingly) that $\|\Sigmab\| = 1$.
%   \item According to the definition of $\bar \P_\perp$ and $\gamma$, the following bounds hold
%   \begin{equation}\label{eq:bound-bar-gamma-P}
%     \frac1{\gamma + 1} \I \preceq \bar \P_\perp \preceq \I, \quad \gamma \le \frac{k}{r-k} = \frac1{\rho - 1}
%   \end{equation}
%   for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}k > 1$, where we used the fact that
%   \[
%     k = n - \tr\,\bar\P_\perp = \tr\,\bar\P_\perp (\gamma\Sigmab + \I) - \tr\,\bar\P_\perp
%   = \gamma \tr\,\bar\P_\perp \Sigmab\geq \frac{\gamma}{\gamma+1}\tr\,\Sigmab,
%   \]
%   so that $r = \tr \Sigmab \le k \cdot \frac{\gamma + 1}{\gamma}$.%= k + \tr \bar \P_\perp \Sigmab
%   \item As already discussed in Section~\ref{subsec:proof-sketch}, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the following spectral norm
%   \begin{equation}\label{eq:proof-1-SM}
%     \| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}
%   \end{equation}
%   so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ from \eqref{eq:bound-bar-gamma-P}, we have
%   \begin{align*}
%     \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
%     \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp^{\frac12}\|\leq
%   \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
%   \end{align*}
%   Defining $\epsilon =\frac{C_\rho}{\sqrt r}
%   \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
%   of the p.s.d.~matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
%   in the interval $[1-\epsilon,1+\epsilon]$, and
%   \begin{align*}
%     (1-\epsilon)\I\preceq
%     \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.
%   \end{align*}
%   so that by multiplying $\bar \P_{\perp}^{\frac12}$ on both sides, we obtain the
%   desired bound.
% \end{enumerate}

% \bigskip

% As a consequence of the above observations, we only need to prove \eqref{eq:proof-1-SM} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:

% \begin{enumerate}[leftmargin=*]
%   \item For $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its $i$-th row, we define, for $i \in \{ 1, \ldots, k \}$, the following events
%   \begin{equation}\label{eq:def-events-E-F}
%     E_i: \left\{ \left| \frac{\tr (\I - \P_{-i}) \Sigmab}{\x_i^\top ( \I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\}, \quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\}. %F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
%   \end{equation}
%   where we recall $\x_i \in \R^n$ is the $i$-th row of $\X$ so that $\E[\x_i] = 0$ and $\E[\x_i \x_i^\top] = \Sigmab$. With Lemma~\ref{l:concentration-sub-gaussian}, we can bound the probability of $\neg E_i$ and $\neg F_i$, and consequently that of $\neg E$ for $E = \bigwedge_{i=1}^k(E_i \wedge F_i)$;
%   \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \|$. More precisely, since
% %   \begin{align*}
% %   \I&-\E[\P]\bar\P_\perp^{-1}
% %   = \E[\X^\dagger\X] - \gamma\E[\P]\Sigmab\\
% %   &= \E[\X^\dagger\X\one_E] +
% %     \E[\X^\dagger\X \one_{\neg E}]
% %     -\gamma\E[\P]\Sigmab\\
% %   &=n\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top}{\x_k^\top\P_{-k}\x_k}\one_E\bigg]
% %     -\gamma\E[\P]\Sigmab \ +\     \E[\X^\dagger\X\one_{\neg E}]\\
% %   &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
% %     s)\cdot\frac{\P_{-k}\x_k\x_k^\top}{\x_k^\top\P_{-k}\x_k}\one_E\bigg]}_{\T_1}
% %     + \gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2} + \gamma \underbrace{\E[\P_{-k}-\P]\Sigmab}_{\T_3} + \underbrace{\E[\X^\dagger\X\one_{\neg E}]}_{\T_4}
% % \end{align*}
% \begin{align*}
%   \I-\E[\P_\perp]\bar\P_\perp^{-1}
%     &= \E[\P] - \gamma\E[\P_\perp]\Sigmab
%     \\
%     &= \E[\P \cdot \one_E] +
%     \E[\P  \cdot \one_{\neg E}]
%     -\gamma\E[\P_\perp]\Sigmab\\
%   &=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg]
%     -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
%   &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
%     s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
%     - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2} \\ 
%     &+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
%     %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
% \end{align*}
% where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_k^\top ( \I - \P_{-k}) \x_k$ as well as $\bar s = \tr \bar \P_\perp \Sigmab =k/\gamma$. It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ to reach the conclusion.
% \end{enumerate}

% Another important relation that will be constantly used throughout the proof is
% \begin{equation}\label{eq:bound-trace-P-Sigma}
%     \tr (\I - \P_{-k}) \Sigmab = \tr \Sigmab^{\frac12} (\I - \P_{-k})^2 \Sigmab^{\frac12} =
%     \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-k}^\dagger\X_{-k}\|_F^2\geq
% \sum_{i\geq k}\lambda_i (\Sigmab)\geq r-k
% \end{equation}
% where we used the fact that $\rank(\X_{-k}^\dagger \X_{-k}) \le \rank(\X_{-k}) \le k-1$ and arranged the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_n(\Sigmab)$ in a non-increasing order. As a consequence, we also have
% \begin{equation}\label{eq:bound-norm-P-Sigma}
%   \frac{\tr (\I - \P_{-k}) \Sigmab}{ \|  (\I - \P_{-k}) \Sigmab \| } \ge \tr (\I - \P_{-k}) \Sigmab \geq r-k.
% \end{equation}

% \bigskip

% For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian}~and~\eqref{eq:bound-norm-P-Sigma} that
% \begin{align*}
%   \Pr (\neg E_i) &\le \Pr \left\{ \lvert \x_i^\top (\I - \P_{-i}) \x_i - \tr \Sigmab (\I - \P_{-i})  \rvert
%     \geq \frac13 \tr \Sigmab (\I - \P_{-i}) \right\} \\ 
%     &\le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }. % \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
% \end{align*}
% %for some constant $c_\rho$ depending on fixed $\rho \equiv \frac{r}n > 1$. 
% Similarly we have
% \[
%   \Pr (\neg F_i) \le 2 \ee^{ - \min \left\{ \frac{ r }{9 C^2 K^4}, \frac{\sqrt{r} }{3 C K^2} \right\} } \le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }
% \]
% and with the union bound we obtain
% \begin{equation}
%   \Pr (\neg E) \le 4 k \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{k}{(r-k)^2} \cdot 4 (r-k)^2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{C_\rho}{r - k}
% \end{equation}
% where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ and  $x^4 \ee^{-\alpha x} \le \frac{256 \ee^{-4}}{\alpha^4}$ on $x > 0$. Also, denote $c_\rho = \frac{r - k}r = \frac{\rho - 1}{\rho} > 0$, we have
% \begin{equation}\label{eq:proba-E}
%   \Pr(\neg E) \le \frac{C_\rho}{r - k} = \frac{C_\rho}{c_\rho r} = \frac{C'_\rho}{r} 
% \end{equation}
% for some $C'_\rho > 0$ that depends on $\rho = r/k > 1$ and the sub-gaussian norm $K$.

% \medskip

% At this point, note that, conditioned on the event $E$, we have for $i \in \{1, \ldots, k \}$
% \begin{equation}
%     \frac12 \frac1{\tr (\I - \P_{-i}) \Sigmab} \le \frac1{\x_i^\top (\I - \P_{-i}) \x_i} \le \frac32 \frac1{\tr (\I - \P_{-i}) \Sigmab}, \quad  \frac1{2r} \le \frac1{\x_i^\top \x_i} \le \frac3{2r}. %\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
% \end{equation}
% Also, with \eqref{eq:proba-E} and the fact that $\| \I - \P_{-k} \|
% \le 1$ and $\| \P \| \le 1$, we have $\| \T_2 \| + \| \T_4 \| \le
% \frac{C_\rho}{r}$ for some $\C_\rho > 0$ that depends on $\rho$ and $K$. And it thus remains to handle the terms $\T_1$ and $\T_3$ to obtain a bound on $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| $.

% To bound $\T_3$, with $\P - \P_{-k} = \frac{(\I - \P_{-k})\x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}$ in Lemma~\ref{l:rank-one-update}, we have
% \begin{align*}
% \|\T_3\| &\leq  \bigg\|
%   \E\bigg[\frac{ (\I - \P_{-k}) \x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}
%   \cdot \one_E \bigg]\bigg\|  + \| \E[ (\P - \P_{-k}) \cdot \one_{\neg E} ] \| \\ 
%   %%%
%   &\le \frac32 \E \left[ \frac1{ \tr (\I - \P_{-k} ) \Sigmab } \right] + \frac{c_\rho}{r - k} \leq \frac{C_\rho}{r-k} = \frac{C'_\rho}{r}
% \end{align*}
% where we used the fact that $\tr\,(\I - \P_{-k}) \Sigmab \ge r -k $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv r/k > 1$.

% For $\T_1$ we write
% \begin{align*}
%   \|\T_1\|    &\leq\E\bigg[\|\I - \P_{-k}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
%     \frac{\x_k\x_k^\top}{\x_k^\top( \I - \P_{-k}) \x_k} \cdot \one_E\mid\P_{-k}\Big]
%     \Big\|\bigg]
%   \\
%           &\leq \frac32 \frac1{r-k} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
%             |\bar s-\hat s|\cdot \v^\top\x_k\x_k^\top\v\cdot\one_E\mid
%             \P_{-k}\Big]\bigg]
%   \\
%   &\leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
%     s)^2\cdot \one_E \mid\P_{-k}\big]}}_{ T_{1,1} }\cdot
%     \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_k)^4\big]}}_{ T_{1,2} }\bigg]
% \end{align*}
% where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} for the second inequality, and Cauchy–Schwarz for the third inequality.

% We first bound $T_{1,2}$ by definition of sub-gaussian random
% vectors. We have for $\x_k$ a $K$-sub-gaussian and $\| \v \| = 1$
% that, $\v^\top \x_k$ is a sub-gaussian random variable with $\|\v^\top
% \a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute
% constant $C > 0$, see for example
% \cite[Section~2.5.2]{vershynin2018high}.

% For $T_{1,1}$ we have
% \[
%   \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-k}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
% \]
% where we denote $s=\E[\hat s]=\tr\,\E[\I - \P_{-k}]\Sigmab$. Note that
% % \begin{align*}
% %   \E\big[(s - \hat s)^2 \cdot\one_E \big] &=
% % \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2 \cdot\one_E \big] +
% %   \E\big[(\tr\,\P_{-k}\Sigmab-\x_k^\top\P_{-k}\x_k)^2 \cdot\one_E \big]\\
% %   &\leq C\Big(n + \E\big[\tr\,(\P_{-k}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
% %     C\big(n + \tr\,\E[\P_{-k}]\Sigmab \cdot\one_E\big)\\
% %   &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
% % \end{align*}
% \begin{align*}
%   \E\big[(s - \hat s)^2 \big] &=
% \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2  \big] +
%   \E\big[(\tr\,(\I - \P_{-k})\Sigmab-\x_k^\top(\I - \P_{-k})\x_k)^2  \big]\\
%   &\leq C_1 k +  C_2 \E\big[\tr\,(\Sigmab - \P_{-k}\Sigmab)^2 \big] \le C(k+s)\leq C\big(k+\bar s + |s- \bar s| \big)
% \end{align*}
% where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}. Recall that $\bar s = \tr \bar \P_\perp \Sigmab \le \tr \Sigmab = r$ and $k < r$, we have
% \begin{equation}\label{eq:bound-T11}
%   T_{1,1} \le \sqrt{ (\bar s - s)^2 + C (|\bar s - s| + 2 r) }
% \end{equation}

% % Moreover, using the fact that
% % $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$, we obtain that:
% % \begin{align*}
% %   |\bar s - s| \cdot\one_E &= |\tr (\bar \P_\perp - \E[\P_{-k}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P_\perp - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-k}] \Sigmab| \\
% %   &= \big|\tr\,(\I-\E[\P]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-k}\x_k\x_k^\top\P_{-k}}{\x_k^\top \P_{-k} \x_k} \right] \Sigmab \\
% %   %&= \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P_\perp\Sigmab\big| \cdot\one_E + 1 \\
% %   &\leq \gamma\,\E\bigg[|\bar s-\hat s|\cdot
% %     \frac{\tr\,\P_{-k}\x_k\x_k^\top\bar\P_\perp\Sigmab}{\tr\,\P_{-k}\x_k\x_k^\top} \cdot\one_E\bigg]
% %     + \gamma\,
% %     \E\bigg[\frac{\tr\,\P_{-k}\x_k\x_k^\top\P_{-k}\Sigmab\bar\P_\perp\Sigmab}
% %     {\tr\,\P_{-k}\x_k\x_k^\top}\bigg] + 1\\
% %   &\leq \frac{\gamma}{\gamma+1}\Big(\E\left[|\bar s-\hat
% %     s| \cdot \frac{\x_k^\top \x_k}{\x_k^\top \P_{-k} \x_k} \cdot\one_E \right] + 1\Big)\leq \frac{\gamma}{\gamma+1}\Big(|\bar
% %     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
% %     s|\big] + 1\Big)\\
% %   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
% %      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
% % \end{align*}

% It remains to bound $|\bar s - s|$. Note that $\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X
% (\X^\top\X)^\dagger$ and is symmetric, so
% \begin{align*}
%   &\I - \E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar\P_\perp^{-1} \E[\P_\perp]
%   = 2\E[\P] - \E[\gamma\P_\perp\Sigmab] - \E[\gamma\Sigmab\P_\perp]\\
%   &= \sum_{i=1}^k\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp])\\
%   %&=k\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top + \x_k\x_k^\top \P_{-k}}{\x_k^\top\P_{-k}\x_k}\bigg] - \gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
%   &=\gamma\,
%     \E\bigg[\bar s\cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})} {\x_k^\top(\I - \P_{-k}) \x_k}\bigg]
%     -\gamma\,\E\bigg[\hat s
%     \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg] \\ 
%   & +\gamma\,(\E[(\I - \P_{-k})\Sigmab] + \E[\Sigmab ( \I -\P_{-k}) ])- \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp]) \\
%   &=\gamma\,\E\bigg[(\bar s - \hat s)
%     \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]
% +    \gamma (\E[\P-\P_{-k}]\Sigmab + \Sigmab \E[\P-\P_{-k}]). 
% \end{align*}
% %where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
% Moreover, using the fact that $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$ and $\bar \P_\perp \Sigmab = \Sigmab \bar \P_\perp$, we obtain that
% \begin{align*}
%   &|\bar s - s|  = |\tr (\bar \P_\perp - \E[\I - \P_{-k}] )\Sigmab|   \le |\tr (\bar \P_\perp - \E[\P_\perp] )\Sigmab|   + |\tr \E[\P - \P_{-k}] \Sigmab| \\
%   &= \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab + \tr\,\bar\P_\perp (\I - \bar \P_\perp^{-1} \E[\P_\perp]) \Sigmab \big|   + \tr\,\E \left[ \frac{(\I - \P_{-k})\x_k\x_k^\top(\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} \right] \Sigmab \\
%   &\leq \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar \P_\perp^{-1} \E[\P_\perp])\bar\P_\perp\Sigmab \big|   + 1 \\
%   &\leq \frac{\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,( (\I -\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I -\P_{-k})) \bar\P_\perp\Sigmab}{\tr\,(\I -\P_{-k})\x_k\x_k^\top}  \bigg] \\
%   &+ \gamma\, \E\bigg[\frac{\tr\,(\I-\P_{-k})\x_k\x_k^\top(\I-\P_{-k})\bar\P_\perp\Sigmab}
%     {\tr\,(\I-\P_{-k})\x_k\x_k^\top}\bigg] + 1\\
%   &\leq \frac{\gamma}{\gamma+1}\Bigg(\E\bigg[|\bar s-\hat
%     s| \cdot \frac{\x_k^\top (\I -\P_{-k}) \x_k}{\x_k^\top (\I -\P_{-k}) \x_k}  \bigg] + 1\Bigg) + 1\leq \frac{\gamma}{\gamma+1}\Big(|\bar
%     s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1 \\
%   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
%      C\sqrt{|\bar s- s|} + C\sqrt{2 r} + 1\Big) + 1.
% \end{align*}

% Solving for $|\bar s-s|$, we deduce that
% \begin{align*}
%   |\bar s-s| \leq C_1 \sqrt{r} + C_2,
%   %\gamma(c\sqrt{k+\bar s}+2) + c^2\gamma^2\leq \gamma (c+1)\sqrt{k+\bar s} + c^2\gamma^2.
% \end{align*}
% so plugging back to \eqref{eq:bound-T11} we get $T_{1,1} \le C
% \sqrt r$ and $\| \T_1\| \le \frac{C_\rho}{\sqrt r} $, thus completing
% the proof.

\subsection{Complete proof of Theorem~\ref{t:main-tech}}

Equipped with the lemmas above, we are ready to prove Theorem~\ref{t:main-tech}. First note that:
\begin{enumerate}[leftmargin=*]
  \item  Since $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, we can assume without loss of generality (after rescaling $\bar\P_\perp$ correspondingly) that $\|\Sigmab\| = 1$.
  \item According to the definition of $\bar \P_\perp$ and $\gamma$, the following bounds hold
  \begin{equation}\label{eq:bound-bar-gamma-P}
    \frac1{\gamma + 1} \I \preceq \bar \P_\perp \preceq \I, \quad \gamma \le \frac{k}{r-k} = \frac1{\rho - 1}
  \end{equation}
  for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}k > 1$, where we used the fact that
  \[
    k = n - \tr\,\bar\P_\perp = \tr\,\bar\P_\perp (\gamma\Sigmab + \I) - \tr\,\bar\P_\perp
  = \gamma \tr\,\bar\P_\perp \Sigmab\geq \frac{\gamma}{\gamma+1}\tr\,\Sigmab,
  \]
  so that $r = \tr \Sigmab \le k \cdot \frac{\gamma + 1}{\gamma}$.%= k + \tr \bar \P_\perp \Sigmab
  \item As already discussed in Section~\ref{subsec:proof-sketch}, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the following spectral norm
  \begin{equation}\label{eq:proof-1-SM}
    \| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}, %\quad~\text{or}~\| \I - \bar \P_\perp^{-1} \E[\P_\perp] \| \le \frac{C_\rho}{\sqrt r}
  \end{equation}
  so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ from \eqref{eq:bound-bar-gamma-P}, we have
  \begin{align*}
    \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
    %\|\bar\P_\perp^{\frac12}(\I-\bar\P_\perp^{-1}\E[\P_\perp])\bar\P_\perp^{-\frac12}\| =
    \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp^{\frac12}\|\leq
  \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
  \end{align*}
  Defining $\epsilon =\frac{C_\rho}{\sqrt r}
  \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
  of the p.s.d.~matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
  in the interval $[1-\epsilon,1+\epsilon]$, and
  \begin{align*}
    (1-\epsilon)\I\preceq
    \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.
  \end{align*}
  so that by multiplying $\bar \P_{\perp}^{\frac12}$ on both sides, we obtain the
  desired bound.
\end{enumerate}

\bigskip

As a consequence of the above observations, we only need to prove \eqref{eq:proof-1-SM} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:

\begin{enumerate}[leftmargin=*]
  \item For $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its $i$-th row, we define, for $i \in \{ 1, \ldots, k \}$, the following events
  \begin{equation}\label{eq:def-events-E-F}
    E_i: \left\{ \left| \frac{\tr (\I - \P_{-i}) \Sigmab}{\x_i^\top ( \I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\}, 
    %\quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\}. %F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
  \end{equation}
  where we recall $\x_i \in \R^n$ is the $i$-th row of $\X$ so that $\E[\x_i] = 0$ and $\E[\x_i \x_i^\top] = \Sigmab$. With Lemma~\ref{l:concentration-sub-gaussian}, we can bound the probability of $\neg E_i$, and consequently that of $\neg E$ for $E = \bigwedge_{i=1}^k E_i$;
  \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \|$. More precisely, since
\begin{align*}
  \I-\E[\P_\perp]\bar\P_\perp^{-1}
    &= \E[\P] - \gamma\E[\P_\perp]\Sigmab
    \\
    &= \E[\P \cdot \one_E] +
    \E[\P  \cdot \one_{\neg E}]
    -\gamma\E[\P_\perp]\Sigmab\\
  &=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg]
    -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
  &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
    s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
    - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2} \\ 
    &+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
    %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
\end{align*}
where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_k^\top ( \I - \P_{-k}) \x_k$ as well as $\bar s = \tr \bar \P_\perp \Sigmab =k/\gamma$. It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ to reach the conclusion.
\end{enumerate}

% Note that, $\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I- \bar\P_\perp^{-1}\E[\P_\perp]$ is symmetric and 
% \begin{equation}
%   \I-\E[\P_\perp]\bar\P_\perp^{-1} + \I- \bar\P_\perp^{-1}\E[\P_\perp] = \gamma (\T_1 + \T_1^\top) - \gamma (\T_2 + \T_2^\top) + 2 \gamma \T_3 + 2 \T_4
% \end{equation}

Another important relation that will be constantly used throughout the proof is
\begin{equation}\label{eq:bound-trace-P-Sigma}
    \tr (\I - \P_{-k}) \Sigmab = \tr \Sigmab^{\frac12} (\I - \P_{-k})^2 \Sigmab^{\frac12} =
    \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-k}^\dagger\X_{-k}\|_F^2\geq
\sum_{i\geq k}\lambda_i (\Sigmab)\geq r-k
\end{equation}
where we used the fact that $\rank(\X_{-k}^\dagger \X_{-k}) \le \rank(\X_{-k}) \le k-1$ and arranged the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_n(\Sigmab)$ in a non-increasing order. As a consequence, we also have
\begin{equation}\label{eq:bound-norm-P-Sigma}
  \frac{\tr (\I - \P_{-k}) \Sigmab}{ \|  (\I - \P_{-k}) \Sigmab \| } \ge \tr (\I - \P_{-k}) \Sigmab \geq r-k.
\end{equation}

\bigskip

For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian}~and~\eqref{eq:bound-norm-P-Sigma} that
\begin{align*}
  \Pr (\neg E_i) &\le \Pr \left\{ \lvert \x_i^\top (\I - \P_{-i}) \x_i - \tr \Sigmab (\I - \P_{-i})  \rvert
    \geq \frac13 \tr \Sigmab (\I - \P_{-i}) \right\} \\ 
    &\le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }. % \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
\end{align*}
so that with the union bound we obtain
\begin{equation}\label{eq:union-bound-neg-E}
  \Pr (\neg E) \le 2k \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{k}{(r-k)^2} \cdot 2 (r-k)^2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{C_\rho}{r - k}
\end{equation}
where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ and  $x^4 \ee^{-\alpha x} \le \frac{256 \ee^{-4}}{\alpha^4}$ on $x > 0$. Also, denote $c_\rho = \frac{r - k}r = \frac{\rho - 1}{\rho} > 0$, we have
\begin{equation}\label{eq:proba-E}
  \Pr(\neg E) \le \frac{C_\rho}{r - k} = \frac{C_\rho}{c_\rho r} = \frac{C'_\rho}{r} 
\end{equation}
for some $C'_\rho > 0$ that depends on $\rho = r/k > 1$ and the sub-gaussian norm $K$.

\medskip

At this point, note that, conditioned on the event $E$, we have for $i \in \{1, \ldots, k \}$
\begin{equation}
    \frac12 \frac1{\tr (\I - \P_{-i}) \Sigmab} \le \frac1{\x_i^\top (\I - \P_{-i}) \x_i} \le \frac32 \frac1{\tr (\I - \P_{-i}) \Sigmab}, %\quad  \frac1{2r} \le \frac1{\x_i^\top \x_i} \le \frac3{2r}. %\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
\end{equation}
Also, with \eqref{eq:proba-E} and the fact that $\| \P \| \le 1$, we have $ \| \T_4 \| \le
\frac{C_\rho}{r}$ for some $C_\rho > 0$ that depends on $\rho$ and $K$. To handle non-symmetric matrix $\T_2$, note that $\T_2 + \T_2^\top$ is symmetric and
\begin{equation}
  - \E[ (\I - \P_{-k}) \cdot \one_{\neg E}] - \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}] \preceq \T_2 + \T_2^\top \preceq \E[ (\I - \P_{-k}) \cdot \one_{\neg E}] + \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}]
\end{equation}
with $-(\A \A^\top + \B \B^\top) \preceq \A \B^\top + \B \A^\top \preceq \A \A^\top + \B \B^\top$. To obtain an upper bound for operator norm of $\E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}]$, note that 
\begin{align*}
  \| \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}] \| &\le \E[ (\x_k^\top \x_k)^2 \cdot \one_{\neg E}] = \int_0^\infty \Pr ( \x^\top \x \cdot \one_{\neg E} \ge \sqrt t ) dt \\ 
  &\le \int_0^X \Pr ( \x^\top \x \cdot \one_{\neg E} \ge \sqrt t ) dt + \int_X^\infty \Pr ( \x^\top \x \ge \sqrt t ) dt  \\ 
  &\le X \cdot \Pr(\neg E) + \int_X^\infty \ee^{- \min \left\{ \frac{t}{C^2 K^4 r}, \frac{\sqrt t}{CK^2 \sqrt r} \right\} } dt \le \frac{c_\rho}r
\end{align*}
where we recall $\E[\x^\top \x] = \tr \Sigmab = r$ and take $X \ge C^2 K^4 r$, the third line follows from the proof of Lemma~\ref{l:concentration-sub-gaussian} and the forth line from the same argument as in \eqref{eq:union-bound-neg-E}. Moreover, since $\| \T_2 \| \le \| \T_2 + \T_2^\top \|$ (see for example \cite[Proposition~5.11]{serre2010matrices}), we conclude that $\| \T_2 \| \le \frac{C_\rho}r$.

And it thus remains to handle the terms $\T_1$ and $\T_3$ to obtain a bound on $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| $.

To bound $\T_3$, with $\P - \P_{-k} = \frac{(\I - \P_{-k})\x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}$ in Lemma~\ref{l:rank-one-update}, we have
\begin{align*}
\|\T_3\| &\leq  \bigg\|
  \E\bigg[\frac{ (\I - \P_{-k}) \x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}
  \cdot \one_E \bigg]\bigg\|  + \| \E[ (\P - \P_{-k}) \cdot \one_{\neg E} ] \| \\ 
  %%%
  &\le \frac32 \E \left[ \frac1{ \tr (\I - \P_{-k} ) \Sigmab } \right] + \frac{c_\rho}{r - k} \leq \frac{C_\rho}{r-k} = \frac{C'_\rho}{r}
\end{align*}
where we used the fact that $\tr\,(\I - \P_{-k}) \Sigmab \ge r -k $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv r/k > 1$.

For $\T_1$ we write
\begin{align*}
  \|\T_1\|    &\leq\E\bigg[\|\I - \P_{-k}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
    \frac{\x_k\x_k^\top}{\x_k^\top( \I - \P_{-k}) \x_k} \cdot \one_E\mid\P_{-k}\Big]
    \Big\|\bigg]
  \\
          &\leq \frac32 \frac1{r-k} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
            |\bar s-\hat s|\cdot \v^\top\x_k\x_k^\top\v\cdot\one_E\mid
            \P_{-k}\Big]\bigg]
  \\
  &\leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
    s)^2\cdot \one_E \mid\P_{-k}\big]}}_{ T_{1,1} }\cdot
    \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_k)^4\big]}}_{ T_{1,2} }\bigg]
\end{align*}
where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} for the second inequality, and Cauchy–Schwarz for the third inequality.

We first bound $T_{1,2}$ by definition of sub-gaussian random
vectors. We have for $\x_k$ a $K$-sub-gaussian and $\| \v \| = 1$
that, $\v^\top \x_k$ is a sub-gaussian random variable with $\|\v^\top
\a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute
constant $C > 0$, see for example
\cite[Section~2.5.2]{vershynin2018high}.

For $T_{1,1}$ we have
\[
  \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-k}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
\]
where we denote $s=\E[\hat s]=\tr\,\E[\I - \P_{-k}]\Sigmab$. Note that
% \begin{align*}
%   \E\big[(s - \hat s)^2 \cdot\one_E \big] &=
% \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2 \cdot\one_E \big] +
%   \E\big[(\tr\,\P_{-k}\Sigmab-\x_k^\top\P_{-k}\x_k)^2 \cdot\one_E \big]\\
%   &\leq C\Big(n + \E\big[\tr\,(\P_{-k}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
%     C\big(n + \tr\,\E[\P_{-k}]\Sigmab \cdot\one_E\big)\\
%   &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
% \end{align*}
\begin{align*}
  \E\big[(s - \hat s)^2 \big] &=
\E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2  \big] +
  \E\big[(\tr\,(\I - \P_{-k})\Sigmab-\x_k^\top(\I - \P_{-k})\x_k)^2  \big]\\
  &\leq C_1 k +  C_2 \E\big[\tr\,(\Sigmab - \P_{-k}\Sigmab)^2 \big] \le C(k+s)\leq C\big(k+\bar s + |s- \bar s| \big)
\end{align*}
where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}. Recall that $\bar s = \tr \bar \P_\perp \Sigmab \le \tr \Sigmab = r$ and $k < r$, we have
\begin{equation}\label{eq:bound-T11}
  T_{1,1} \le \sqrt{ (\bar s - s)^2 + C (|\bar s - s| + 2 r) }
\end{equation}

% Moreover, using the fact that
% $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$, we obtain that:
% \begin{align*}
%   |\bar s - s| \cdot\one_E &= |\tr (\bar \P_\perp - \E[\P_{-k}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P_\perp - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-k}] \Sigmab| \\
%   &= \big|\tr\,(\I-\E[\P]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-k}\x_k\x_k^\top\P_{-k}}{\x_k^\top \P_{-k} \x_k} \right] \Sigmab \\
%   %&= \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P_\perp\Sigmab\big| \cdot\one_E + 1 \\
%   &\leq \gamma\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,\P_{-k}\x_k\x_k^\top\bar\P_\perp\Sigmab}{\tr\,\P_{-k}\x_k\x_k^\top} \cdot\one_E\bigg]
%     + \gamma\,
%     \E\bigg[\frac{\tr\,\P_{-k}\x_k\x_k^\top\P_{-k}\Sigmab\bar\P_\perp\Sigmab}
%     {\tr\,\P_{-k}\x_k\x_k^\top}\bigg] + 1\\
%   &\leq \frac{\gamma}{\gamma+1}\Big(\E\left[|\bar s-\hat
%     s| \cdot \frac{\x_k^\top \x_k}{\x_k^\top \P_{-k} \x_k} \cdot\one_E \right] + 1\Big)\leq \frac{\gamma}{\gamma+1}\Big(|\bar
%     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
%     s|\big] + 1\Big)\\
%   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
% \end{align*}

It remains to bound $|\bar s - s|$. Note that $\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X
(\X^\top\X)^\dagger$ and is symmetric, so
\begin{align*}
  &\I - \E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar\P_\perp^{-1} \E[\P_\perp]
  = 2\E[\P] - \E[\gamma\P_\perp\Sigmab] - \E[\gamma\Sigmab\P_\perp]\\
  &= \sum_{i=1}^k\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp])\\
  %&=k\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top + \x_k\x_k^\top \P_{-k}}{\x_k^\top\P_{-k}\x_k}\bigg] - \gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
  &=\gamma\,
    \E\bigg[\bar s\cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})} {\x_k^\top(\I - \P_{-k}) \x_k}\bigg]
    -\gamma\,\E\bigg[\hat s
    \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg] \\ 
  & +\gamma\,(\E[(\I - \P_{-k})\Sigmab] + \E[\Sigmab ( \I -\P_{-k}) ])- \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp]) \\
  &=\gamma\,\E\bigg[(\bar s - \hat s)
    \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]
+    \gamma (\E[\P-\P_{-k}]\Sigmab + \Sigmab \E[\P-\P_{-k}]). 
\end{align*}
%where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
Moreover, using the fact that $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$ and $\bar \P_\perp \Sigmab = \Sigmab \bar \P_\perp$, we obtain that
\begin{align*}
  &|\bar s - s|  = |\tr (\bar \P_\perp - \E[\I - \P_{-k}] )\Sigmab|   \le |\tr (\bar \P_\perp - \E[\P_\perp] )\Sigmab|   + |\tr \E[\P - \P_{-k}] \Sigmab| \\
  &= \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab + \tr\,\bar\P_\perp (\I - \bar \P_\perp^{-1} \E[\P_\perp]) \Sigmab \big|   + \tr\,\E \left[ \frac{(\I - \P_{-k})\x_k\x_k^\top(\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} \right] \Sigmab \\
  &\leq \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar \P_\perp^{-1} \E[\P_\perp])\bar\P_\perp\Sigmab \big|   + 1 \\
  &\leq \frac{\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
    \frac{\tr\,( (\I -\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I -\P_{-k})) \bar\P_\perp\Sigmab}{\tr\,(\I -\P_{-k})\x_k\x_k^\top}  \bigg] \\
  &+ \gamma\, \E\bigg[\frac{\tr\,(\I-\P_{-k})\x_k\x_k^\top(\I-\P_{-k})\bar\P_\perp\Sigmab}
    {\tr\,(\I-\P_{-k})\x_k\x_k^\top}\bigg] + 1\\
  &\leq \frac{\gamma}{\gamma+1}\Bigg(\E\bigg[|\bar s-\hat
    s| \cdot \frac{\x_k^\top (\I -\P_{-k}) \x_k}{\x_k^\top (\I -\P_{-k}) \x_k}  \bigg] + 1\Bigg) + 1\leq \frac{\gamma}{\gamma+1}\Big(|\bar
    s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1 \\
  &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
     C\sqrt{|\bar s- s|} + C\sqrt{2 r} + 1\Big) + 1.
\end{align*}

Solving for $|\bar s-s|$, we deduce that
\begin{align*}
  |\bar s-s| \leq C_1 \sqrt{r} + C_2,
  %\gamma(c\sqrt{k+\bar s}+2) + c^2\gamma^2\leq \gamma (c+1)\sqrt{k+\bar s} + c^2\gamma^2.
\end{align*}
so plugging back to \eqref{eq:bound-T11} we get $T_{1,1} \le C
\sqrt r$ and $\| \T_1\| \le \frac{C_\rho}{\sqrt r} $, thus completing
the proof.
\section{Convergence analysis of randomized iterative methods}
\label{a:newton}

Here, we discuss how our surrogate expressions for the expected
residual projection can be used to perform convergence analysis for
several randomized iterative optimization methods discussed in
Section~\ref{s:newton}.

\subsection{Generalized Kaczmarz method}
Generalized Kaczmarz \cite{generalized-kaczmarz} is an iterative
method for solving an $m\times n$ linear system
$\A\x=\b$, which uses a $k\times m$ sketching matrix $\S_t$ to reduce
the linear system and update an iterate $\x^t$ as follows:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S_t\A\x=\S_t\b.
  \end{align*}
Assume that $\x^*$ is the unique solution to the linear system $\A\x=\b$. In Theorems 4.1 and 4.6, \cite{generalized-kaczmarz} show that the
expected trajectory of the generalized Kaczmarz iterates, as
they converge to $\x^*$, is controlled
by the projection matrix $\P=(\S_t\A)^\dagger\S_t\A$ as follows:
\begin{align*}
  \text{(\cite{generalized-kaczmarz}, Theorem 4.1)}\qquad\quad \ \ \,
  \E[\x^{t+1}-\x^*]
  &= \big(\I-\E[\P]\big)\,\E[\x^t-\x^*],\\
  \text{(\cite{generalized-kaczmarz}, Theorem 4.6)}\qquad
  \E\big[\|\x^{t+1}-\x^*\|^2\big]
&\leq (1-\kappa)\,\E\big[\|\x^t-\x^*\|^2\big],
\text{ where }\kappa=\lambda_{\min}\big(\E[\P]\big).
\end{align*}
Both of these results depend on the expected projection
$\E[\P]$. The first one describes the expected trajectory of the
iterate, whereas the second one gives the worst-case convergence
rate in terms of the so-called \emph{stochastic condition number}
$\kappa$. We next demonstrate how Theorem \ref{t:main} can be used  in 
combination with the above results to
obtain convergence analysis for generalized Kaczmarz which is
formulated in terms of the spectral properties of $\A$. This includes
precise expressions for both the expected trajectory and $\kappa$. The following
result is a more detailed version of Corollary \ref{c:kaczmarz} from
Section \ref{s:newton}.
\begin{corollary}\label{c:kaczmarz2}
Let $\sigma_i$ denote the singular values of $\A$, and let $k$ denote
the size of sketch $\S_t$. Define:
\begin{align*}
\Delta_t=\x^t-\x^*\quad\text{and}\quad  \bar\Delta_{t+1} =
  (\gamma\A^\top\A+\I)^{-1}\E[\Delta_t]\quad\text{s.t.}\quad
  \sum_i\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1}=k.
\end{align*}
  Suppose that $\S_t$ has i.i.d.~mean-zero sub-gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$. 
Assume that $\rho = r/k$ is a constant larger than $1$. Then, the
expected trajectory satisfies:
\begin{align}
  \big\|\E[\Delta_{t+1}] - \bar\Delta_{t+1}\big\|
  \leq \epsilon\cdot \|\bar\Delta_{t+1}\|,
  \quad\text{for}\quad\epsilon=O\big(\tfrac1{\sqrt
  r}\big).\label{eq:trajectory}
\end{align}
Moreover, we obtain the following worst-case convergence guarantee:
\begin{align}
  \E\big[\|\Delta_{t+1}\|^2\big]
  \leq \big(1-
  (\bar\kappa-\epsilon)\big)\,\E\big[\|\Delta_t\|^2\big],
  \quad\text{where}\quad
  \bar\kappa = \frac{\sigma_{\min}^2}{\sigma_{\min}^2+1/\gamma}.\label{eq:worst-case}
\end{align}
\end{corollary}
\begin{remark}
Our worst-case convergence guarantee \eqref{eq:worst-case}
  requires the matrix $\A$ to be sufficiently well-conditioned so
  that $\bar\kappa-\epsilon>0$. However, we believe that our surrogate expression
  $\bar\kappa$ for the stochastic condition number is far more
  accurate than suggested by the current analysis. 
\end{remark}

\subsection{Randomized Subspace Newton}
Randomized Subspace Newton (RSN, \cite{Gower2019})  is a randomized
Newton-type method  for minimizing a
smooth, convex and twice differentiable function
$f:\R^d\times \R$. The iterative 
update for this algorithm is defined as follows:
\begin{align*}
  \x^{t+1} = \x^t - \frac1L\S_t^\top(\S_t\H(\x^t)\S_t^\top)^\dagger\S_t\g(\x^t),
\end{align*}
where $\H(\x^t)$ and $\g(\x^t)$ are the Hessian and gradient of $f$ at
$\x^t$, respectively, whereas $\S_t$ is a $k\times d$ sketching
matrix (with $k\ll d$) which is refreshed at every iteration. Here,
$L$ denotes the \emph{relative smoothness} constant defined by
\cite{Gower2019} in Assumption 1, which also defines relative strong
convexity, denoted by $\mu$. In Theorem 2, they prove the
following convergence guarantee for RSN:
\begin{align*}
  \E[f(\x^t)] - f(\x^*) \leq \Big(1-\kappa\frac{\mu}{L}\Big)^t(f(\x^0)-f(\x^*)),
\end{align*}
where $\kappa =\min_{\x}\kappa(\x)$ and
$\kappa(\x)=\lambda_{\min}^+(\E[\P(\x)])$ is the smallest positive
eigenvalue of the expectation of the projection matrix
$\P(\x)=
\H^{\frac12}(\x)\S_t^\top(\S_t\H(\x)\S_t^\top)^\dagger\S_t\H^{\frac12}(\x)$. Our
results lead to the following surrogate expression for this expected
projection when the sketch is sub-gaussian:
\begin{align*}
  \E[\P(\x)] \simeq \H(\x)\big(\H(\x) +
  \tfrac1{\gamma(\x)}\I\big)^{-1}\quad\text{for}\quad\gamma(\x)>0
  \quad\text{s.t.}\quad\tr\,\H(\x)\big(\H(\x) +
  \tfrac1{\gamma(\x)}\I\big)^{-1} = k.
\end{align*}
Thus, the condition number $\kappa$ of RSN can be estimated using the
following surrogate expression:
\begin{align*}
\kappa\simeq \bar\kappa :=
  \min_\x\frac{\lambda_{\min}^+(\H(\x))}{\lambda_{\min}^+(\H(\x)) +
  1/\gamma(\x)}. 
\end{align*}
Just as in Corollary \ref{c:kaczmarz2}, an approximation
of the form $|\bar\kappa-\kappa|\leq\epsilon$ can be shown
from Theorem \ref{t:main}.
\begin{corollary}\label{c:rsn}
  Suppose that sketch $\S_t$ has size $k$ and i.i.d.~mean-zero sub-gaussian entries. Let
  $r=\min_\x\tr\,\H(\x)/\|\H(\x)\|$ be the (minimum) stable rank of
  the (square root) Hessian and assume that $\rho = r/k$ is a constant larger than
  $1$. Then,
  \begin{align*}
    |\kappa-\bar\kappa| \leq O\big(\tfrac1{\sqrt r}\big).
  \end{align*}
\end{corollary}

\subsection{Jacobian Sketching}
Jacobian Sketching (JacSketch, \cite{jacsketch}) defines an $n\times n$ positive semi-definite weight matrix
$\W$, and combines it with an $k\times n$ sketching matrix $\S$
(which is refreshed at every iteration of the algorithm), to
implicitly construct the following projection matrix:
\begin{align*}
  \Pi_\S = \S^\top(\S\W\S^\top)^\dagger\S\W,
\end{align*}
which is used to sketch the Jacobian at the current iterate (for the
complete method, we refer to their Algorithm 1).
The convergence rate guarantee given in their Theorem 3.6 for
JacSketch is given in terms of the Lyapunov function:
\begin{align*}
 \Psi^t = \|\x^t-\x^*\|^2 + \frac{\alpha}{2\mathcal{L}_2}\|\mathbf{J}^t-\nabla F(\x^*)\|_{\W^{-1}}^2,
\end{align*}
where $\alpha$ is the step size used by the algorithm. Under
appropriate choice of the step-size, Theorem~3.6 states that:
\begin{align*}
\E[\Psi^t] \leq \bigg(1 - \mu\,\min\Big\{\frac{1}{4\mathcal{L}_1},\frac{\kappa}{4\mathcal{L}_2\rho/n^2+\mu}\Big\}\bigg)^t\cdot\Psi^0,
\end{align*}
where $\kappa=\lambda_{\min}(\E[\Pi_\S])$ is the 
\emph{stochastic condition number} analogous to the one defined for
the Generalized Kaczmarz method, $n$ is the data size and
parameters $\rho$, $\mathcal{L}_1$, $\mathcal{L}_2$ and $\mu$ are
problem dependent constants defined in Theorem~3.6. Similarly as
before, we can use our surrogate expressions 
for the expected residual projection to obtain a precise estimate for
the stochastic condition number 
$\kappa$ under sub-gaussian sketching:
\begin{align*}
  \kappa\simeq\bar\kappa :=\frac{\lambda_{\min}(\W)}{\lambda_{\min}(\W) +
  1/\gamma}
  \quad\text{for}\quad\gamma>0\quad\text{s.t.}\quad\tr\,\W(\W +
  \tfrac1\gamma\I)^{-1} = k.
\end{align*}
\begin{corollary}\label{c:jacsketch}
  Suppose $\S_t$ has size $k$ and i.i.d.~mean-zero sub-gaussian entries. Let
  $r=\tr\,\W/\|\W\|$ be the stable rank of $\W^{\frac12}$ and assume
  that $\rho = r/k$ is a constant larger than 
  $1$. Then,
  \begin{align*}
    |\kappa-\bar\kappa| \leq O\big(\tfrac1{\sqrt r}\big).
  \end{align*}
\end{corollary}

\subsection{Omitted proofs}
\begin{proofof}{Corollary}{\ref{c:kaczmarz2}}
Using Theorem \ref{t:main}, for $\bar\P_{\perp}$ as defined in \eqref{eq:surrogate}, we have
  \begin{align*}
    (1-\epsilon)\bar\P_{\perp}\preceq \I-\E[\P] =
    \E[\P_{\perp}]\preceq(1+\epsilon)\bar\P_{\perp},
    \quad\text{where}\quad\epsilon=O\big(\tfrac 1{\sqrt r}\big).
  \end{align*}
In particular, this implies that
$\|\bar\P_{\perp}^{-\frac12}(\E[\P_{\perp}]-\bar\P_{\perp})\bar\P_{\perp}^{-\frac12}\|\leq\epsilon$.
Moreover, in the proof of Theorem \ref{t:main-tech} we showed that
  $\frac{\rho-1}\rho\I\preceq\bar\P_{\perp}\preceq\I$, see
  \eqref{eq:bound-bar-gamma-P}, so it follows that:
  \begin{align*}
    \bar\P_{\perp}^{-1}(\E[\P_{\perp}]-\bar\P_{\perp})^2\bar\P_{\perp}^{-1}\preceq
    \frac{\rho}{\rho-1}\big(\bar\P_{\perp}^{-\frac12}(\E[\P_{\perp}]-\bar\P_{\perp})\bar\P_{\perp}^{-\frac12}\big)^2
    \preceq     \frac{\rho}{\rho-1}\,\epsilon^2\cdot \I,
  \end{align*}
  where note that  $\frac{\rho}{\rho-1}\,\epsilon^2=O(1/r)$, since
  $\rho$ is treated as a constant. Thus we conclude that:
  \begin{align*}
    \|\E[\Delta_{t+1}]-\bar\Delta_{t+1}\|^2
    &=\E[\Delta_t]^\top(\E[\P_{\perp}]-\bar\P_{\perp})^2\E[\Delta_t]
    \\
    &\leq O(1/r)\cdot \E[\Delta_t]^\top\bar\P_{\perp}^2\E[\Delta_t] =
      O(1/r)\cdot \|\bar\Delta_{t+1}\|^2,
  \end{align*}
  which completes the proof of \eqref{eq:trajectory}. To show
  \eqref{eq:worst-case}, it suffices to observe that
  \begin{align*}
    \lambda_{\min}(\E[\P])
    = 1- \lambda_{\max}(\E[\P_{\perp}])
    \geq 1-(1+\epsilon)\lambda_{\max}(\bar\P_{\perp})
    \geq \lambda_{\min}(\I-\bar\P_{\perp}) - \epsilon,
  \end{align*}
  which completes the proof since
  $\I-\bar\P_{\perp}=\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1}$. 
\end{proofof}

Corollaries \ref{c:rsn} and \ref{c:jacsketch} follow analogously from
Theorem \ref{t:main}.

\begin{figure}
  \centering
  \includegraphics[width=.47\textwidth]{bodyfat-supp}\nobreak\includegraphics[width=.47\textwidth]{housing-supp}
  \includegraphics[width=.47\textwidth]{mg-supp}\nobreak\includegraphics[width=.47\textwidth]{triazines-supp} 
  \caption{Theoretical predictions versus approximation error for the
    sketched Nystr\"om with the RBF kernel, using Gaussian and
    Rademacher sketches (spectral decay shown at the bottom).}\label{f:nystrom2}
\end{figure}

\section{Additional empirical results}
\label{a:experiments}

We complement the results of Section \ref{s:experiments} with
empirical results on four additional libsvm datasets \cite{libsvm} (bringing
the total number of benchmark datasets to eight), which further
establish the accuracy of our surrogate expressions for the low-rank approximation
error. Similary as in Figure \ref{f:nystrom}, we use the sketched
Nystr\"om method \cite{revisiting-nystrom} with the RBF kernel
$k(\a_i,\a_j)=\exp(-\|\a_i-\a_j\|^2/(2\sigma^2))$, for several values
of the parameter $\sigma$. The values of $\sigma$ were chosen so as to
demonstrate the effectiveness of our theoretical predictions both when the
stable rank is moderately large and when it is very small.

In Figure \ref{f:nystrom2} we show the results for both Gaussian and
Rademacher sketches. These results reinforce the conclusions we made in
Section \ref{s:experiments}: our theoretical estimates are very
accurate in all cases, for both sketching methods, and even when the
stable rank is close to 1 (a regime that is not supported by the
current theory).

\end{document}

% \begin{document}

% \maketitle

% \begin{abstract}
%   The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
%   both the left- and right-hand margins. Use 10~point type, with a vertical
%   spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
%   bold, and in point size 12. Two line spaces precede the abstract. The abstract
%   must be limited to one paragraph.
% \end{abstract}

% \section{Submission of papers to NeurIPS 2020}

% NeurIPS requires electronic submissions.  The electronic submission site is
% \begin{center}
%   \url{https://cmt3.research.microsoft.com/NeurIPS2020/}
% \end{center}

% Please read the instructions below carefully and follow them faithfully.

% \subsection{Style}

% Papers to be submitted to NeurIPS 2020 must be prepared according to the
% instructions presented here. Papers may only be up to eight pages long,
% including figures. Additional pages \emph{containing only a section on the broader impact, acknowledgments and/or cited references} are allowed. Papers that exceed eight pages of content will not be reviewed, or in any other way considered for
% presentation at the conference.

% The margins in 2020 are the same as those in 2007, which allow for $\sim$$15\%$
% more words in the paper compared to earlier years.

% Authors are required to use the NeurIPS \LaTeX{} style files obtainable at the
% NeurIPS website as indicated below. Please make sure you use the current files
% and not previous versions. Tweaking the style files may be grounds for
% rejection.

% \subsection{Retrieval of style files}

% The style files for NeurIPS and other conference information are available on
% the World Wide Web at
% \begin{center}
%   \url{http://www.neurips.cc/}
% \end{center}
% The file \verb+neurips_2020.pdf+ contains these instructions and illustrates the
% various formatting requirements your NeurIPS paper must satisfy.

% The only supported style file for NeurIPS 2020 is \verb+neurips_2020.sty+,
% rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
%   Microsoft Word, and RTF are no longer supported!}

% The \LaTeX{} style file contains three optional arguments: \verb+final+, which
% creates a camera-ready copy, \verb+preprint+, which creates a preprint for
% submission to, e.g., arXiv, and \verb+nonatbib+, which will not load the
% \verb+natbib+ package for you in case of package clash.

% \paragraph{Preprint option}
% If you wish to post a preprint of your work online, e.g., on arXiv, using the
% NeurIPS style, please use the \verb+preprint+ option. This will create a
% nonanonymized version of your work with the text ``Preprint. Work in progress.''
% in the footer. This version may be distributed as you see fit. Please \textbf{do
%   not} use the \verb+final+ option, which should \textbf{only} be used for
% papers accepted to NeurIPS.

% At submission time, please omit the \verb+final+ and \verb+preprint+
% options. This will anonymize your submission and add line numbers to aid
% review. Please do \emph{not} refer to these line numbers in your paper as they
% will be removed during generation of camera-ready copies.

% The file \verb+neurips_2020.tex+ may be used as a ``shell'' for writing your
% paper. All you have to do is replace the author, title, abstract, and text of
% the paper with your own.

% The formatting instructions contained in these style files are summarized in
% Sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.

% \section{General formatting instructions}
% \label{gen_inst}

% The text must be confined within a rectangle 5.5~inches (33~picas) wide and
% 9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).  Use 10~point
% type with a vertical spacing (leading) of 11~points.  Times New Roman is the
% preferred typeface throughout, and will be selected for you by default.
% Paragraphs are separated by \nicefrac{1}{2}~line space (5.5 points), with no
% indentation.

% The paper title should be 17~point, initial caps/lower case, bold, centered
% between two horizontal rules. The top rule should be 4~points thick and the
% bottom rule should be 1~point thick. Allow \nicefrac{1}{4}~inch space above and
% below the title to rules. All pages should start at 1~inch (6~picas) from the
% top of the page.

% For the final version, authors' names are set in boldface, and each name is
% centered above the corresponding address. The lead author's name is to be listed
% first (left-most), and the co-authors' names (if different address) are set to
% follow. If there is only one co-author, list both author and co-author side by
% side.

% Please pay special attention to the instructions in Section \ref{others}
% regarding figures, tables, acknowledgments, and references.

% \section{Headings: first level}
% \label{headings}

% All headings should be lower case (except for first word and proper nouns),
% flush left, and bold.

% First-level headings should be in 12-point type.

% \subsection{Headings: second level}

% Second-level headings should be in 10-point type.

% \subsubsection{Headings: third level}

% Third-level headings should be in 10-point type.

% \paragraph{Paragraphs}

% There is also a \verb+\paragraph+ command available, which sets the heading in
% bold, flush left, and inline with the text, with the heading followed by 1\,em
% of space.

% \section{Citations, figures, tables, references}
% \label{others}

% These instructions apply to everyone.

% \subsection{Citations within the text}

% The \verb+natbib+ package will be loaded for you by default.  Citations may be
% author/year or numeric, as long as you maintain internal consistency.  As to the
% format of the references themselves, any style is acceptable as long as it is
% used consistently.

% The documentation for \verb+natbib+ may be found at
% \begin{center}
%   \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
% \end{center}
% Of note is the command \verb+\citet+, which produces citations appropriate for
% use in inline text.  For example,
% \begin{verbatim}
%    \citet{hasselmo} investigated\dots
% \end{verbatim}
% produces
% \begin{quote}
%   Hasselmo, et al.\ (1995) investigated\dots
% \end{quote}

% If you wish to load the \verb+natbib+ package with options, you may add the
% following before loading the \verb+neurips_2020+ package:
% \begin{verbatim}
%    \PassOptionsToPackage{options}{natbib}
% \end{verbatim}

% If \verb+natbib+ clashes with another package you load, you can add the optional
% argument \verb+nonatbib+ when loading the style file:
% \begin{verbatim}
%    \usepackage[nonatbib]{neurips_2020}
% \end{verbatim}

% As submission is double blind, refer to your own published work in the third
% person. That is, use ``In the previous work of Jones et al.\ [4],'' not ``In our
% previous work [4].'' If you cite your other papers that are not widely available
% (e.g., a journal paper under review), use anonymous author names in the
% citation, e.g., an author of the form ``A.\ Anonymous.''

% \subsection{Footnotes}

% Footnotes should be used sparingly.  If you do require a footnote, indicate
% footnotes with a number\footnote{Sample of the first footnote.} in the
% text. Place the footnotes at the bottom of the page on which they appear.
% Precede the footnote with a horizontal rule of 2~inches (12~picas).

% Note that footnotes are properly typeset \emph{after} punctuation
% marks.\footnote{As in this example.}

% \subsection{Figures}

% \begin{figure}
%   \centering
%   \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
%   \caption{Sample figure caption.}
% \end{figure}

% All artwork must be neat, clean, and legible. Lines should be dark enough for
% purposes of reproduction. The figure number and caption always appear after the
% figure. Place one line space before the figure caption and one line space after
% the figure. The figure caption should be lower case (except for first word and
% proper nouns); figures are numbered consecutively.

% You may use color figures.  However, it is best for the figure captions and the
% paper body to be legible if the paper is printed in either black/white or in
% color.

% \subsection{Tables}

% All tables must be centered, neat, clean and legible.  The table number and
% title always appear before the table.  See Table~\ref{sample-table}.

% Place one line space before the table title, one line space after the
% table title, and one line space after the table. The table title must
% be lower case (except for first word and proper nouns); tables are
% numbered consecutively.

% Note that publication-quality tables \emph{do not contain vertical rules.} We
% strongly suggest the use of the \verb+booktabs+ package, which allows for
% typesetting high-quality, professional tables:
% \begin{center}
%   \url{https://www.ctan.org/pkg/booktabs}
% \end{center}
% This package was used to typeset Table~\ref{sample-table}.
%
% \begin{table}
%   \caption{Sample table title}
%   \label{sample-table}
%   \centering
%   \begin{tabular}{lll}
%     \toprule
%     \multicolumn{2}{c}{Part}                   \\
%     \cmidrule(r){1-2}
%     Name     & Description     & Size ($\mu$m) \\
%     \midrule
%     Dendrite & Input terminal  & $\sim$100     \\
%     Axon     & Output terminal & $\sim$10      \\
%     Soma     & Cell body       & up to $10^6$  \\
%     \bottomrule
%   \end{tabular}
% \end{table}
%
% \section{Final instructions}
%
% Do not change any aspects of the formatting parameters in the style files.  In
% particular, do not modify the width or length of the rectangle the text should
% fit into, and do not change font sizes (except perhaps in the
% \textbf{References} section; see below). Please note that pages should be
% numbered.
%
% \section{Preparing PDF files}
%
% Please prepare submission files with paper size ``US Letter,'' and not, for
% example, ``A4.''
%
% Fonts were the main cause of problems in the past years. Your PDF file must only
% contain Type 1 or Embedded TrueType fonts. Here are a few instructions to
% achieve this.
%
% \begin{itemize}
%
% \item You should directly generate PDF files using \verb+pdflatex+.
%
% \item You can check which fonts a PDF files uses.  In Acrobat Reader, select the
%   menu Files$>$Document Properties$>$Fonts and select Show All Fonts. You can
%   also use the program \verb+pdffonts+ which comes with \verb+xpdf+ and is
%   available out-of-the-box on most Linux machines.
%
% \item The IEEE has recommendations for generating PDF files whose fonts are also
%   acceptable for NeurIPS. Please see
%   \url{http://www.emfield.org/icuwb2010/downloads/IEEE-PDF-SpecV32.pdf}
%
% \item \verb+xfig+ "patterned" shapes are implemented with bitmap fonts.  Use
%   "solid" shapes instead.
%
% \item The \verb+\bbold+ package almost always uses bitmap fonts.  You should use
%   the equivalent AMS Fonts:
% \begin{verbatim}
%    \usepackage{amsfonts}
% \end{verbatim}
% followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or \verb+\mathbb{C}+
% for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You can also use the following
% workaround for reals, natural and complex:
% \begin{verbatim}
%    \newcommand{\RR}{I\!\!R} %real numbers
%    \newcommand{\Nat}{I\!\!N} %natural numbers
%    \newcommand{\CC}{I\!\!\!\!C} %complex numbers
% \end{verbatim}
% Note that \verb+amsfonts+ is automatically loaded by the \verb+amssymb+ package.
%
% \end{itemize}
%
% If your file contains type 3 fonts or non embedded TrueType fonts, we will ask
% you to fix it.
%
% \subsection{Margins in \LaTeX{}}
%
% Most of the margin problems come from figures positioned by hand using
% \verb+\special+ or other commands. We suggest using the command
% \verb+\includegraphics+ from the \verb+graphicx+ package. Always specify the
% figure width as a multiple of the line width as in the example below:
% \begin{verbatim}
%    \usepackage[pdftex]{graphicx} ...
%    \includegraphics[width=0.8\linewidth]{myfile.pdf}
% \end{verbatim}
% See Section 4.4 in the graphics bundle documentation
% (\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})
%
% A number of width problems arise when \LaTeX{} cannot properly hyphenate a
% line. Please give LaTeX hyphenation hints using the \verb+\-+ command when
% necessary.
%
%
% \section*{Broader Impact}
%
% Authors are required to include a statement of the broader impact of their work, including its ethical aspects and future societal consequences.
% Authors should discuss both positive and negative outcomes, if any. For instance, authors should discuss a)
% who may benefit from this research, b) who may be put at disadvantage from this research, c) what are the consequences of failure of the system, and d) whether the task/method leverages
% biases in the data. If authors believe this is not applicable to them, authors can simply state this.
%
% Use unnumbered first level headings for this section, which should go at the end of the paper. {\bf Note that this section does not count towards the eight pages of content that are allowed.}
%
% \begin{ack}
% Use unnumbered first level headings for the acknowledgments. All acknowledgments
% go at the end of the paper before the list of references. Moreover, you are required to declare
% funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
% More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2020/PaperInformation/FundingDisclosure}.
%
%
% Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
% \end{ack}
%
% \section*{References}
%
% References follow the acknowledgments. Use unnumbered first-level heading for
% the references. Any choice of citation style is acceptable as long as you are
% consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
% when listing the references.
% {\bf Note that the Reference section does not count towards the eight pages of content that are allowed.}
% \medskip
%
% \small
%
% [1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
% connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
% (eds.), {\it Advances in Neural Information Processing Systems 7},
% pp.\ 609--616. Cambridge, MA: MIT Press.
%
% [2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
%   Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
% TELOS/Springer--Verlag.
%
% [3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
% recall at excitatory recurrent synapses and cholinergic modulation in rat
% hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
%
% \end{document}

### 2022-08-23 design/implicit.tex
Commit original repos

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}
\usepackage{cleveref}
\input{shortdefs}

\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\bPhi}{\boldsymbol{\Phi}}

\DeclareMathOperator{\ReLU}{ReLU}
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\erf}{erf}

\title{Implicit Regularization of Interpolating Models}
  \author{%
    Micha{\l } Derezi\'{n}ski
 }
\date{}

% \setlength{\parindent}{0cm}
% \setlength{\parskip}{0mm}

\begin{document}
\maketitle

\section{Introduction}

Consider a set $\Dc=\{(\x_i,y_i)\}_{i=1}^n$ with $\x_i \in \R^d$, $y_i \in \R$ of $n$ data points drawn i.i.d.~from
some distribution $\mu$, denoted as $\Dc\sim\mu^n$. The learner uses a
hypothesis class $\{f_\w:\w\in \mathcal W\}$ of functions
parameterized by $\w$. We consider the highly over-parameterized
regime where there exist many $\w$ such that $f_\w(\x_i)=y_i$
for all $(\x_i,y_i)\in\Dc$, i.e., $f_\w$ interpolates the data. The
learner chooses one of those functions by minimizing 
some capacity metric defined by $g(\w)$, as follows:
\begin{align*}
  \hat \w_\Dc(g) \triangleq \argmin_{\w}\Big\{g(\w)\ \text{ s.t. }f_\w(\x_i)=y_i\ \,
  \forall_{(\x_i,y_i)\in\Dc}\Big\}.
\end{align*}
Our goal is to characterize the implicit regularization that is
induced by the choice of metric $g(\cdot)$. We do this by comparing the
bias of $\hat \w_\Dc(g)$ to the bias introduced by explicitly
regularizing the minimizer of a loss function
$\E_{(\x,y)\sim\mu}\,\ell\big(f_\w(\x),y\big)$ over the 
population distribution $\mu$:
\begin{align*}
\bar \w_\mu(\ell,\lambda g) \triangleq \argmin_{\w}\Big\{ 
\E_{(\x,y)\sim\mu}\,\ell\big(f_\w(\x),y\big) + \lambda
\,g(\w)\Big\}.
\end{align*}  
As a guiding 
principle of our analysis, we formulate the following working
hypothesis, which we aim to prove (or disprove) under particular settings: 
\begin{align}
  \E\big[\hat \w_\Dc(g)\big]\quad \simeq\quad \bar \w_\mu(\ell,\lambda g),\label{eq:hypothesis}
\end{align}
for an appropriate choice of loss function $\ell(\yh,y)$ and the parameter $\lambda$
determining the amount of implicit regularization. Our setting
encompasses a range of statistical learning models such as:
\begin{enumerate}
\item \emph{Generalized linear model:} $f_\w(\x) =
  \sigma(\w^\top \x)$, with $\x,\w\in\R^d$ and a link function $\sigma$.
\item \emph{Random features model:} $f_\w(\x) =
  \w^\top \phi(\x)$, where $\phi(\x)=\langle\sigma(\varphi_1^\top
  \x),...,\sigma(\varphi_p^\top \x)\rangle$ and $\w\in\R^p$.
\item \emph{Neural network model:} $f_\w(\x)$ is a neural
  network parameterized by $\w$ for a fixed architecture.
\end{enumerate}
In addition to the learning model, we can vary the choice of the
capacity function $g(\cdot)$, which will have a significant effect on
the implicit regularization. Common choices for $g(\cdot)$ include:
\begin{enumerate}
  \item \emph{Euclidean norm:} $g(\w) = \|\w\|_2^2$, or more
    generally, $g(\w)=\w^\top \A \w$ for some positive definite matrix $\A$.
    \item \emph{Sparsifying norm:} $g(\w)=\|\w\|_1$ or other sparsity-inducing norm.
\end{enumerate}

\section{Euclidean norm and general Euclidean norm in linear model}

In this section, we consider the linear model $f_\w(\x) = \w^\top \x$ and noisy linear output $y = \w_*^\top \x + \xi$ with independent noise of zero mean and finite variance. We focus on the set of $\w$ that interpolates the training set, i.e., $f(\x_i) = y_i$ for $i \in \{0, \ldots, n\}$ and choose the minimum (general) Euclidean norm from this set, i.e., $g(\w) = \| \w \|_\A^2 = \w^\top \A \w$ for some positive definite $\A$ as
\[
  \hat \w \equiv \argmin_\w g(\w) = \argmin_\w \w^\top \A \w, \quad \textmd{such that $f(\x_i) = y_i$ for all $i$}.
\]
Compared to following population minimizer of (regularized) quadratic loss
\[
  \bar \w_\mu \equiv \argmin_\w \left\{ \E_{(\x,y) \sim \mu} (f_\w(\x) - y)^2 + \lambda \| \w \|_\A^2 \right\}
\]
we have the following theorem.

\begin{theorem}\label{t:main-Mahalanobis}
Let $\x\sim\Nc(0,\Sigmab)$ and assume $\Sigmab \A^{-1}$ has nuclear rank
  $r>2n$, we have:
  \begin{align*}
    \E [\hat \w]
   \ \overset{\epsilon}{\simeq}\ \bar \w_\mu
\quad    \text{with}\quad\epsilon = O\big(\tfrac1{\sqrt r}\big),
  \end{align*}
  where $\w'\overset\epsilon\simeq\w$ iff
  $\|\w'-\w\|_\A \leq \epsilon\,\|\w\|_\A$ for $\| \w \|_\A^2 = \w^\top \A \w $ for $\lambda$ so that $n=\tr\,\Sigmab(\Sigmab+\lambda\A)^{-1}$.
\end{theorem}
\begin{proof}
For $f_\w(\x) = \w^\top \x$, this is equivalent to find $\w_\A \equiv \A^{\frac12} \w$ such that $f_{\w_\A} (\x) = \w_\A^\top \A^{-\frac12} \x$ with $g(\w_\A) = \| \w_\A \|^2$ minimized, or, to find $\w_\A$ such that $f_{\w_\A} (\tilde \x) = \w_\A^\top \tilde \x$ for $\tilde \x \equiv \A^{-\frac12} \x \sim \Nc (\zero, \A^{-\frac12} \Sigmab \A^{-\frac12} )$ with minimal Euclidean norm $\| \w_\A \|^2$. The desired solution can then be retrieved as $\w = \A^{-\frac12} \w_\A$.

We thus focus on $f_{\w_\A} (\tilde \x) = \w_\A^\top \tilde \x$ for $\tilde \x \equiv \A^{-\frac12} \x \sim \Nc (\zero, \Sigmab_\A )$, $\Sigmab_\A = \A^{-\frac12} \Sigmab \A^{-\frac12}$ with minimal Euclidean norm $\| \w_\A \|^2$. In this case, we have $\hat \w_\A$ the minimum norm solution explicitly given by $\hat \w_\A = \tilde \X^\dagger \y$ with $y = \tilde \x^\top \A^{\frac12} \w_* + \xi$, so that
\[
  \E [\hat \w_\A]= \E[\tilde \X^\dagger \y] = \E[\tilde \X^\dagger \tilde \X] \A^{\frac12} \w_*
\]
as well as
\[
  \bar{\w}_{\A,\mu} = \argmin_{\w_\A} \left\{ \E_{(\x,y) \sim \mu} (y - \w_\A^\top \A^{-\frac12} \x)^2 + \lambda \| \w_\A \|^2 \right\} = (\Sigmab_\A + \lambda \I)^{-1} \Sigmab_\A \cdot \A^{\frac12} \w_*.
\]

Recall $\Sigmab_\A \equiv \A^{-\frac12} \Sigmab \A^{-\frac12}$, it then follows from Theorem~\ref{t:projection} that, for $\lambda$ the solution to $n = \tr \Sigmab_\A (\Sigmab_\A + \lambda \I)^{-1}$ we have $ (\Sigmab_\A + \lambda \I)^{-1} \Sigmab_\A = \I - \bar \P $ with
\[
  \bar \P = (\gamma \Sigmab_\A + \I)^{-1}, \quad \lambda = \gamma^{-1}
\]
so that
\begin{align*}
  \E[\tilde \X^\dagger \tilde \X] \A^{\frac12} \w_* - (\Sigmab_\A + \lambda \I)^{-1} \Sigmab_\A \A^{\frac12} \w_* &= \left( \E[\tilde \X^\dagger \tilde \X] - (\I - \bar \P) \right) (\I - \bar \P)^{-1} \cdot (\Sigmab_\A + \lambda \I)^{-1} \Sigmab_\A \A^{\frac12} \w_* \\
  %%%
  &= \left( \E[\tilde \X^\dagger \tilde \X] (\I - \bar \P)^{-1} - \I  \right) \cdot \bar \w_{\A,\mu}.
\end{align*}
Since $\w = \A^{-\frac12} \w_\A$, we have
\begin{align*}
  \A^{\frac12} ( \E[\hat \w] - \w_\mu ) = \E[\hat \w_\A] - \bar \w_{\A,\mu} = \left( \E[\tilde \X^\dagger \tilde \X] (\I - \bar \P)^{-1} - \I  \right) \cdot \A^{\frac12} \bar \w_\mu
\end{align*}
and thus
\[
  \| \E[\hat \w] - \bar \w_\mu \|_\A \le \left\| \left( \E[\tilde \X^\dagger \tilde \X] (\I - \bar \P)^{-1} - \I  \right) \right\| \cdot \| \bar \w_\mu \|_\A.
\]
\end{proof}

\section{Euclidean norm in generalized linear response model}

In this section, we consider the underlying model to be a noisy generalized linear model $y = \sigma(\w_*^\top \x) + \xi$ for some $\sigma(\cdot): \R \to \R$ and independent $\xi$ such that $\E[\xi] = 0$ and $\E[\xi^2]$ is finite. When the nonlinear linking function is unknown, one may simply use a linear model $f_\w(\x) = \w^\top \x$ to fit the response $y$. Again, we focus on the minimum Euclidean norm solution
\[
  \hat \w \equiv \argmin_\w \| \w \|^2, \quad \textmd{such that $f_\w(\x_i) = \w^\top \x_i = y_i$ for all $i$},
\]
that is explicitly given by $\hat \w = \X^\dagger \y$ so that 
\begin{align*}
  \E[\hat \w] = \E[\X^\dagger \y] &= \E[\X^\dagger \sigma(\X \w_*)] = \E[(\X^\top \X)^\dagger \X^\top \sigma(\X \w_*)] = \E \left[ \sum_{i=1}^n (\X^\top \X)^\dagger \x_i \sigma(\x_i^\top \w_*) \right] \\ 
  %%%
  &= n \E\left[\frac{(\I - \X_{-i}^\dagger \X_{-i}) \x_i \sigma(\x_i^\top \w_*)}{\x_i^\top (\I - \X_{-i}^\dagger \X_{-i}) \x_i} \right]
\end{align*}
with Lemma~\ref{lem:rank-one-update}.

Compared to following population minimizer of (regularized) quadratic loss
\[
  \bar \w_\mu \equiv \argmin_\w \left\{ \E_{(\x,y) \sim \mu} (f_\w(\x) - y)^2 + \lambda \| \w \|^2 \right\}.
\]
Since 
\begin{align*}
  \E_{(\x,y) \sim \mu} (f_\w(\x) - y)^2 + \lambda \| \w \|^2 &= \E[\w^\top \x - \sigma(\w_*^\top \x)]^2 + \E[\xi^2]+ \lambda \| \w \|^2 \\ 
  & = \w^\top (\Sigmab + \lambda \I) \w - 2 \w^\top \E[\sigma(\w_*^\top \x) \x] + \E[\sigma^2(\w_*^\top \x)] + \E[\xi^2]
\end{align*}
so that 
\[
  \bar \w_\mu \equiv \argmin_\w \left\{ \E_{(\x,y) \sim \mu} (f_\w(\x) - y)^2 + \lambda \| \w \|^2 \right\} = (\Sigmab + \lambda \I)^{-1} \E[\sigma(\w_*^\top \x) \x].
\]

\begin{lemma}\label{lem:rank-one-update}
    For $\X \in \mathbb R^{n \times d}$ with $n<d$, denote $\P = \X^\dagger \X$ and $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(n-1) \times d}$ the matrix $\X$ without its i-th row $\x_i \in \mathbb R^d$. Then, conditioned on the event $E_i: \left\{ \left| \frac{\tr \Sigmab (\I - \P_{-i})}{\x_i^\top (\I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\}$:
    \begin{align*}
      (\X^\top\X)^\dagger\x_i = \frac{(\I - \P_{-i}) \x_i}{\x_i^\top (\I - \P_{-i}) \x_i}, \quad \P - \P_{-i} = \frac{(\I - \P_{-i}) \x_i \x_i^\top (\I - \P_{-i}) }{\x_i^\top (\I - \P_{-i}) \x_i}.
    \end{align*}
\end{lemma}

\subsection{Computation of the expectation}

In this subsection, we evaluate the expectation $\E[\sigma(\w_*^\top \x) \x]$ with the following lemma.

\begin{lemma}[Stein's lemma, from \cite{stein1981estimation}]\label{lem:stein}
Let $x \sim \Nc(0,1)$ and $y: \mathbb R \to \mathbb R$ a continuously differentiable function such that $\E[y'(x)] < \infty$. Then,
\begin{equation}
  \E[x y(x)] = \E[y'(x)].
\end{equation}
Moreover, for $\x \sim \Nc(\zero, \Sigmab)$ with $\Sigmab \in \mathbb R^{d \times d}$ and $y: \mathbb R^d \mapsto \mathbb R$ a continuously differentiable function with derivatives having at most polynomial growth with respect to $d$,
\begin{equation}
  \E[[\x]_i y(\x)] = \sum_{j=1}^d \Sigmab_{ij} \E \left[ \frac{\partial y(\x)}{\partial [\x]_j} \right]
\end{equation}
where $\partial/\partial [\x]_i$ indicates differentiation with respect to the $i$-th entry of $\x$. Or in vector form
\[
  \E[\x y(\x)] = \Sigmab \E[\nabla y(\x)].
\]
\end{lemma}

As a consequence of Lemma~\ref{lem:stein}, for $\x \sim \Nc(0, \Sigmab)$ and continuously differentiable $\sigma (\cdot)$, we have
\[
  \E[\sigma(\w_*^\top \x) \x] = \Sigmab \w_* \cdot \E[\sigma'(\w_*^\top \x)] = \Sigmab \w_* \cdot \E[\sigma'(\sqrt{\w_*^\top \Sigmab \w_*} \eta)] = \Sigmab \w_* \cdot \frac1{\sqrt{2\pi}} \int \sigma'(x\sqrt{\w_*^\top \Sigmab \w_*}) e^{-x^2/2} dx.
\]
for $\eta \sim \Nc(0,1)$.

\bigskip

For possibly non-smooth function $\sigma(\cdot)$ of interest, we denote $\x = \Sigmab^{\frac12} \z$ with $\z \sim \Nc(0, \I_d)$ standard normal. We can decompose $\z$ as follows
\[
  \z = \eta \frac{\Sigmab^{\frac12} \w_*}{ \sqrt{\w_*^\top \Sigmab \w_*} } + \z_\perp, \quad \eta = \frac{\w_*^\top \Sigmab^{\frac12} \z}{ \sqrt{\w_*^\top \Sigmab \w_*} }
\]
with $\Sigmab^{\frac12} \w_*/\sqrt{\w_*^\top \Sigmab \w_*}$ the unit vector in the direction of $\Sigmab^{\frac12} \w_*$ and $\z_\perp$ lying on the $(d-1)$-dimensional subspace orthogonal to $\Sigmab^{\frac12} \w_*$. As a consequence, using the orthogonal invariance of the standard Gaussian we have:
\begin{enumerate}
  \item $\eta \sim \mathcal N(0,1)$ and $\z_\perp \sim \mathcal N(0,\I_{d-1})$ and they are \emph{independent}; as well as 
  \item $\w_*^\top \Sigmab^{\frac12} \z_\perp = 0$.
\end{enumerate}

As such, we have
\[
  \E[ \sigma(\w_*^\top \x) \x ] = \E \left[ \sigma(\eta \sqrt{\w_*^\top \Sigmab \w_*} ) \left( \eta \frac{\Sigmab \w_*}{ \sqrt{\w_*^\top \Sigmab \w_*} } + \Sigmab^{\frac12} \z_\perp \right) \right] = \E \left[ \eta \sigma(\eta \sqrt{\w_*^\top \Sigmab \w_*} ) \right] \frac{\Sigmab \w_*}{ \sqrt{\w_*^\top \Sigmab \w_*} }.
\]
with
\[
  \E \left[ \eta \sigma(\eta \sqrt{\w_*^\top \Sigmab \w_*} ) \right] = \frac1{\sqrt{2 \pi}} \int x \sigma(x \sqrt{\w_*^\top \Sigmab \w_*}) e^{-x^2/2} dx.
\]

\subsection{Simulations}

\begin{figure}[!htb]
\centering
\includegraphics[width=1.1\textwidth]{figs/implicit_reg_nonlinear.png}
%%%
\caption{ $\E[\hat \w]$ and versus $\bar \w_\mu$ for different nonlinearities, $d = 50$, $n = 20$, expectation estimated from $50000$ independent runs.}
\end{figure}

\section{Implicit regularization in single-hidden-layer neural network model}

Consider the random feature model of the type $f_{\bbeta}(\x) = \bbeta^\top \phi(\W^\top \x)$ for some nonlinear function $\phi(\cdot): \R \to \R$, random input $\x \in \R^d$, first layer weight matrix $\W \in \R^{d \times p}$, second layer weight $\bbeta \in \R^p$ and noisy output $y = \bbeta_*^\top \phi(\W^\top \x) + \xi$ with independent noise such that $\E[\xi] = 0$ and $\E[\xi^2]$ is finite. Denote
\begin{equation}
  \bphi_i \equiv \phi(\W^\top \x_i) \in \R^p, \quad \bPhi \equiv \phi(\X \W) \in \R^{n \times p}
\end{equation}
and focus on the set of $\bbeta \in \R^p$ that interpolates the training set $\{\x_i, y_i\}_{i=1}^n$ so that $f_{\bbeta}(\x_i) = y_i$ and select the minimum Euclidean norm solution from this set that is explicitly given by
\[
  \hat \bbeta \equiv \argmin_{ \bbeta^\top \bphi_i =y_i} \| \bbeta\|^2 = \bPhi^\dagger \y.
\]
As a consequence, we have 
\[
  \E[\hat \bbeta] = \E[\bPhi^\dagger \y] = \E[\bPhi^\dagger \bPhi] \bbeta_*
\]
and
\[
  \bar \bbeta_\mu \equiv \argmin_{\bbeta} \left\{ \E_{(\x,y) \sim \mu} (f_{\bbeta}(\x) - y)^2 + \lambda \| \bbeta \|^2 \right\} = (\K + \lambda \I_p)^{-1} \K \bbeta_*
\]
where we denote 
\begin{equation}\label{eq:def-K}
  \K \equiv \E[\bphi \bphi^\top] = \E[\phi(\W^\top \x) \phi(\x^\top \W)] \in \R^{p \times p}.
\end{equation}

It thus remains to show
\[
  \E_\X[\bPhi^\dagger \bPhi] \equiv \E_\X[\phi(\X \W)^\dagger \phi(\X \W)] \simeq (\K + \lambda \I)^{-1} \K
\]
in an operator norm sense, for $\lambda$ the solution to $n = \tr \K (\K + \lambda \I)^{-1}$. \textbf{Important here! Do not \emph{explicitly} depend on the data dimension $d$!}

\begin{lemma}[Concentration of quadratic forms, {\cite[Lemma~1]{louart2018random}}]
For $\x \in \R^d$ having i.i.d.\@ zero mean, unit variance and $K$-sub-Gaussian entries with sub-Gaussian norm $\| [\x]_i \|_{\psi_2} \le K$ and $\A \in \R^{p \times p}$ such that $\| \A \| \le 1$, then, define the random vector $\bphi \equiv \phi(\W^\top \x) \in \R^p$ for $\W \in \R^{d \times p}$ and Lipschitz continuous $\phi(\cdot)$ with parameter $\gamma_\phi$, we have
\[
  \Pr \left\{ \left| \frac1p \bphi^\top \A \bphi - \frac1p \tr \A \K \right| \ge t \right\} \le C e^{- \frac{c p}{ \| \W \|^2 K^2 \gamma_\phi^2 } \min \left( \frac{t^2}{t_0^2}, t \right) }
\]
for $t_0 \equiv |\phi(0)| + \gamma_\rho K \| \W \| \sqrt{ \frac{d}{p} }$ and $\K \in \R^{p \times p}$ defined in \eqref{eq:def-K}.
\end{lemma}

\subsection{Explicit form of \texorpdfstring{$\K$}{K} for Gaussian data}

For standard Gaussian $\x \sim \mathcal N(0,\I_d)$, the matrix $\K(\W)$ defined in \eqref{eq:def-K} involves the $d$-dimensional integration of
\begin{align*}
  \K_{ij} &= \E_\x[\phi(\w_i^\top \x) \phi(\x^\top \w_j)] = (2\pi)^{-\frac{d}2} \int_{\mathbb{R}^d} \phi(\x^\top \w_i) \phi(\x^\top \w_j) e^{-\frac12 \|\x\|^2} d\,\x \\ 
  %%
  &= \frac1{2\pi} \int_{\mathbb{R}} \int_{\mathbb{R}} \sigma(\tilde x_1 \tilde w_{i,1}) \sigma(\tilde x_1 \tilde w_{j,1} + \tilde x_2 \tilde w_{j,2}) e^{-\frac12 (\tilde x_1^2 + \tilde x_2^2) } d\,\tilde x_1 d\,\tilde x_2 \\
  %%%
  &= \frac1{2\pi} \int_{\mathbb{R}^2} \sigma(\tilde \x^\top \tilde \w_i) \sigma(\tilde \x^\top \tilde \w_j) e^{-\frac12 \|\tilde \x\|^2} d\,\tilde \x
\end{align*}
where we applied the Gram-Schmidt process (by first assuming that $\w_i$ and $\w_j$ are linearly independent) and denote $\tilde w_{i,1} = \|\w_i\|$, $\tilde w_{j,1} = \frac{\w_i^\top \w_j}{\|\w_i\|}$, $\tilde w_{j,2} = \|\w_j\| \sqrt{ 1 - \frac{(\w_i^\top \w_j)^2}{ \|\w_i\|^2 \|\w_j\|^2 } }$ as well as $\tilde \x = [\tilde x_1, \tilde x_2]^\top$, $\tilde \w_i = [\tilde w_{i,1},0]^\top$ and $\tilde \w_j = [\tilde w_{j,1},\tilde w_{j,2}]^\top$. As an example, for $\phi(t) = \max(t,0) \equiv \ReLU(t)$, this further simplifies as
\begin{align*}
  \K_{i,j} = \frac1{2\pi} \int_{ \min(\tilde \x^\top \tilde \w_i, \tilde \x^\top \tilde \w_j) \ge 0} \tilde \x^\top \tilde \w_i \cdot \tilde \x^\top \tilde \w_j \cdot e^{-\frac12 \|\tilde \x\|^2} d\,\tilde \x
\end{align*}

With a simple geometric representation we observe 
\[
  \{ \tilde \x~|~\min(\tilde \x^\top \tilde \w_i, \tilde \x^\top \tilde \w_j) \ge 0 \} = \left\{ r \cos(\theta) + r \sin (\theta)~|~r\ge0,~\theta \in \left[ \theta_0 - \frac{\pi}2, \frac{\pi}2 \right] \right\}
\]
with $\theta_0 \equiv \arccos \left( \frac{ \tilde w_{j,1} }{ \|\w_j\| } \right) = \frac{\pi}2 - \arcsin \left( \frac{ \tilde w_{j,1} }{ \|\w_j\| } \right)$. Therefore with a polar coordinate change of variable we deduce, for $\phi(t) = \ReLU(t)$ that
\begin{align*}
  \K_{ij} &= \tilde w_{i,1} \frac1{2\pi} \int_{\theta_0 - \frac{\pi}2}^{\frac{\pi}2} \cos(\theta) \left( \tilde w_{j,1} \cos(\theta) + \tilde w_{j,2} \sin(\theta) \right) d\,\theta \ \int_0^\infty r^3 e^{-\frac12 r^2} d\,r \\
  %%%
  &=\frac1{2\pi} \|\w_i\| \|\w_j\| \left( \sqrt{ 1-\angle(\w_i,\w_j)^2 } + \angle(\w_i,\w_j) \arccos\left( - \angle(\w_i,\w_j) \right) \right)
\end{align*}
with $\angle(\w_i,\w_j) \equiv \frac{\w_i^\top \w_j}{ \|\w_i\| \|\w_j\| }$. More results for various commonly used activations function are given in Table~\ref{tab:K-phi}.

\begin{table*}[t]
\caption{$\K_{ij}$ for different $\phi(\cdot)$, $\angle(\w_i,\w_j) \equiv \frac{\w_i^\top \w_j}{\|\w_i\| \|\w_j\|}$.}
\label{tab:K-phi}
\begin{center}
\begin{tabular}{l|c}
\hline
$\sigma(t)$     & $\K_{ij}$ \\
\hline
$t$    & $\w_i^\top \w_j$\\

$\max(t,0) \equiv \ReLU(t)$     & $\frac1{2\pi} \|\w_i\| \|\w_j\| \left(\angle(\w_i,\w_j)\arccos\left(-\angle(\w_i,\w_j)\right)+\sqrt{1-\angle(\w_i,\w_j)^2}\right)$   \\

$|t|$           & $\frac2\pi \|\w_i\| \|\w_j\| \left(\angle(\w_i,\w_j)\arcsin\left(\angle(\w_i,\w_j)\right)+\sqrt{1-\angle(\w_i,\w_j)^2}\right)$   \\

%$\varsigma_+ \max(t,0)+ \varsigma_- \max(-t,0)$ & $\frac{\|\w_i\| \|\w_j\|}{2\pi} \left[ (\varsigma_+ + \varsigma_-)^2 \sqrt{1 - \angle(\w_i,\w_j)^2 } + \angle(\w_i,\w_j) \left( (\varsigma_+^2 + \varsigma_-^2) \arccos(-\angle(\w_i,\w_j)) - 2\varsigma_+ \varsigma_- \arccos(\angle(\w_i,\w_j)) \right) \right]$\\
$\varsigma_+ \max(t,0)+ \varsigma_- \max(-t,0)$ & $ \frac12 (\varsigma_+^2 + \varsigma_-^2) \w_i^\top\w_j + \frac{\|\w_i\| \|\w_j\|}{2\pi} (\varsigma_+ + \varsigma_-)^2 \left( \sqrt{1 - \angle(\w_i,\w_j)^2 } - \angle(\w_i,\w_j) \arccos(\angle(\w_i,\w_j)) \right) $\\

$1_{t>0}$       & $\frac12 -\frac1{2\pi} \arccos\left(\angle(\w_i,\w_j)\right)$    \\

$\sign(t)$ & $\frac2\pi\arcsin\left(\angle(\w_i,\w_j)\right)$    \\

$\varsigma_2 t^2 + \varsigma_1 t + \varsigma_0$ & $\varsigma_2^2 \left( 2\left( \w_i^\top \w_j \right)^2 + \|\w_i\|^2 \|\w_j\|^2 \right) + \varsigma_1^2 \w_i^\top \w_j + \varsigma_2 \varsigma_0 \left( \|\w_i\|^2 + \|\w_j\|^2 \right) + \varsigma_0^2$\\

$\cos(t)$       & $\exp\left(-\frac12 \left(\|\w_i\|^2 + \|\w_j\|^2\right)\right) \cosh(\w_i^\top \w_j)$  \\

$\sin(t)$       & $\exp\left(-\frac12 \left(\|\w_i\|^2 + \|\w_j\|^2\right)\right) \sinh(\w_i^\top \w_j)$  \\

$\erf(t)$  & $\frac2\pi \arcsin\left(\frac{2\w_i^\top \w_j}{\sqrt{(1+2\|\w_i\|^2)(1+2\|\w_j\|^2)}}\right)$    \\
$\exp(-\frac{t^2}2)$ & $\frac1{ \sqrt{ (1+\|\w_i\|^2) (1+\|\w_j\|^2) - (\w_i^\top \w_j)^2 } }$\\
\hline
\end{tabular}
\end{center}
\end{table*}

\subsection{Simulations}

Questions: 
\begin{enumerate}
  \item condition on the dimensions $n,p,d$?
  \item condition on the weight matrix $\W \in \R^{n \times d}$, should depend on the operator norm of $\W$, is there any dependence on the minimum singular value?
  \item dependence on the nonlinear function $\phi(\cdot)$? should be assumed to be at least Lipschitz (and the results should depend on the Lipschitz constant of $\phi$), but what's the impact of different nonlinearity? Note that the expression of $\K(\W)$ can be given explicitly for some commonly used activation function $\phi$ such as ReLU, etc.
  \item can we go beyond the case of i.i.d.~Gaussian $\X$? should be able to extend to i.i.d.~sub-Gaussian entries, in that case however, the expression of $\K(\W)$ may no longer be explicit.
\end{enumerate}

\clearpage

Consider the random feature model of the type $f_{\bbeta}(\x) = \bbeta^\top \phi(\W^\top \x)$ for some nonlinear function $\phi(\cdot): \R \to \R$, \emph{deterministic} $\x \in \R^d$, $\W \in \R^{d \times p}$ having i.i.d.~Gaussian entries and $\bbeta \in \R^p$ and noisy output $y =  \bbeta_*^\top \phi(\W^\top \x) + \xi$ with independent noise such that $\E[\xi] = 0$ and $\E[\xi^2]$ is finite. Denote
\begin{equation}
  \bphi_i \equiv \phi(\W^\top \x_i) \in \R^p, \quad \bPhi \equiv \phi(\X \W) \in \R^{n \times p}
\end{equation}
and focus on the set of $\bbeta \in \R^p$ that interpolates the training set $\{\x_i, y_i\}_{i=1}^n$ so that $f_{\bbeta}(\x_i) = y_i$ and select the minimum Euclidean norm solution from this set that is explicitly given by
\[
  \hat \bbeta \equiv \argmin_{f_{\bbeta}(\x_i) =y_i} \| \bbeta\|^2 = \bPhi^\dagger \y.
\]
As a consequence, we have 
\[
  \E[\hat \bbeta] = \E[\bPhi^\dagger \y] = \E[\bPhi^\dagger \bPhi] \bbeta_*
\]
and
\[
  \bar \bbeta_\mu \equiv \argmin_{\bbeta} \left\{ \E_{(\x,y) \sim \mu} (f_{\bbeta}(\x) - y)^2 + \lambda \| \bbeta \|^2 \right\} = (\K + \lambda \I_p)^{-1} \K \bbeta_*
\]
where we denote 
\begin{equation}
  \K \equiv \E[\bphi \bphi^\top] = \E_\W[\phi(\W^\top \x) \phi(\x^\top \W)] \in \R^{p \times p}.
\end{equation}

It thus remains to show
\[
  \E_\W[\bPhi^\dagger \bPhi] \equiv \E_\W[\phi(\X \W)^\dagger \phi(\X \W)] \simeq (\K + \lambda \I)^{-1} \K
\]
in an operator norm sense, for $\lambda$ the solution to $n = \tr \K (\K + \lambda \I)^{-1}$.

Consider the particular case of logistic regression, we have 
\[
  \Pr(y_i = 1 \mid \x_i) = \sigma(\w_*^\top \x_i) = 1 - \Pr(y_i = -1 \mid \x_i), \quad \sigma(t) = \frac1{1 + e^{-t}}.
\]
The associated maximum likelihood estimate is given by 
\[
  \argmin_\w \ell(y_i \x_i^\top \w) = , \quad \ell(t) = \ln(1 + e^{-t}).
\]

The idea is to evaluate the implicit ``regularization'' effect. For the minimum norm solution, it thus remains to assess
\begin{equation}\label{eq:implicit-regularization}
  \E[ \X^\dagger \y] = \begin{cases} (\Sigmab + \lambda \I_d)^{-1} \E[y(\x) \x], \quad n < d \\ \Sigmab^{-1} \E[y(\x) \x], \quad n \ge d \end{cases}
\end{equation}
with $\lambda$ the unique solution to $n = \tr \Sigmab (\Sigmab+ \lambda \I_d)^{-1}$.

Consider the nonlinear model $\y = \sigma(\X \w) + \xi$, with $\y \in \mathbb R^n$, $\w \in \mathbb R^d$, $\X \in \mathbb R^{n \times d}$ and independent noise $\xi$ of zero mean. Here $\sigma: \mathbb R \mapsto \mathbb R$ is the \emph{link function}, popular choices of $\sigma(\cdot)$ are
\begin{enumerate}
  \item \emph{logit link function}: $\sigma(t) = \frac1{ 1 + e^{-t} }$;
  \item \emph{prohibit link function}: $\sigma(t) = \Phi(t)$ with $\Phi$ the cumulative distribution function of standard normal;
  \item \emph{cloglog link function}: $\sigma(t) = 1 - e^{-e^t}$;
  \item possible non-differentiable function such as the ReLU $\sigma(t) = \max(t,0)$ in a neural network context.
\end{enumerate}

\begin{theorem}\label{t:main}
Let $f_\theta(x)=x^\top\theta$ and $(x,y)\sim\mu$
  be s.t.~$x\sim\Nc(0,\Sigmab)$ and $y=x^\top\theta^*+\xi$,
  with $\E[\xi]=0$. Assume that $\Sigmab$ has nuclear rank
  $r>2n$. Then, for $g(x)=\|x\|_2^2$ 
  and $\ell(\yh,y)=(\yh-y)^2$, we have:
  \begin{align*}
    \E_{\Dc\sim\mu^n}\big[\theta_\Dc^*(g)\big]
   \ \overset{\epsilon}{\simeq}\ \theta_\mu^*(\ell, \lambda g)
\quad    \text{with}\quad\epsilon = O\big(\tfrac{\sqrt n}r\big),
  \end{align*}
  where $\theta'\overset\epsilon\simeq\theta$ iff
  $\|\theta'-\theta\|_2\leq \epsilon\,\|\theta\|_2$, and we define
  $\lambda$ so that $n=\tr\,\Sigmab(\Sigmab+\lambda\I)^{-1}$.
\end{theorem}
Possible extensions of Theorem \ref{t:main} which we can expect to
obtain:
\begin{enumerate}
\item General Euclidean norms of the form $g(x) = x^\top\A x$,
\item General distribution of $y$, i.e., without the underlying
  $x^\top\theta^*+\xi$ model, 
\item Non-Gaussian distribution of $x$, which might include the random
  features model. 
\end{enumerate}

% \begin{align}
%   \E\Big[\argmin_{\w:\,\X\w=\y}g(\w)\Big]\ \simeq\ \argmin_{\w\in\R^d}\Big\{
%   \E_\mu\big[l(\x^\top\w,y)\big] + \lambda \,g(\w)\Big\},\label{eq:hypothesis}
% \end{align}
% We are motivated by the
% theoretical and empirical results of \cite{surrogate-design} in one
% particular case where the input distribution $\mu_\x$ is Gaussian,
% function $g(\w)$ is the squared $\ell_2$-norm $\|\w\|_2^2$ and
% $l(\yh,y)$ is the square loss $(\yh-y)^2$, which suggest that
% \eqref{eq:hypothesis} is an accurate characterization of implicit
% regularization for a minimum $\ell_2$-norm estimator.  In this case,
% the estimator is given via the Moore-Penrose pseudoinverse,
% $\wbh=\X^\dagger\y$, whereas the right hand side of
% \eqref{eq:hypothesis} is the ridge-regularized least squares
% solution. The characterization is only exact when the random design
% distribution $\X\sim\mu^n$ of the data is appropriately \emph{rescaled}, obtaining a
% \emph{surrogate design} distribution $\Xb\sim S_\mu^n$. However,
% the theoretical characterization derived under the
% surrogate design remains remarkably accurate under the unrescaled
% Gaussian random design. Is the squared $\ell_2$-norm the only choice
% for function $g(\cdot)$ that  yields the explicit
% characterization of implicit regularization? In our first result, we show that the result
% of \cite{surrogate-design} can be extended to all Mahalanobis norms of
% the form $\|\w\|_\A^2=\w^\top\A\w$ (where the $\ell_2$-norm is
% recovered by setting $\A$ to identity), suggesting that the observed phenomenon is more general.
% To establish the result, we have to extend the definition of a
% surrogate design, taking into account the matrix $\A$.
\begin{figure}[H]
\centering
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/implicit/implicit_bias_uniform}
%  \end{center}
  \caption{}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/implicit/implicit_bias_decaying}
%  \end{center}
  \caption{}
\end{subfigure}
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/implicit/implicit_mse_uniform}
%  \end{center}
  \caption{}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/implicit/implicit_mse_decaying}
%  \end{center}
  \caption{}
\end{subfigure}
\caption{Illustration of using minimum norm estimators for three
  different Mahalanobis norms $\|\cdot\|_A$, with $d=100$ and
  $\mu=\Nc(\zero,\I)$. We consider three matrices $\A$ which are diagonal with
eigenvalues decaying exponentially with condition numbers
$\kappa$. Plots (a) and (b) show the bias, i.e., $\|\E[\wbh]-\w^*\|$,
while plots (c) and (d) show the MSE, i.e.,
$\E\big[\|\wbh-\w^*\|^2\big]$. All plots assume a linear model
$y(\x)=\x^\top\w^*+\xi$, where $\xi\sim \Nc(0,0.3)$. The plots on the
left use a  uniformly linear model, i.e., $\w^*=\frac1{\sqrt{d}}\one$,
whereas the plots on the right use weights increasing exponentially
between  1 and 100, then scaled so that $\|\w^*\|=1$.}
\label{f:intro}
\end{figure}

% \begin{theorem}
%   For any positive definite $\A$,  if $\Xb \sim S_{\!\A,\mu}^n$ then:
%   \begin{align*}
%     \E\Big[\argmin_{\w:\,\Xb\w=\ybb}\|\w\|_\A^2\Big] \ = \
%     \argmin_{\w\in\R^d}\Big\{
%   \E_\mu\big[(\x^\top\w-y)^2\big] + \lambda_n\|\w\|_\A^2\Big\},
%   \end{align*}
%   where $\lambda_n$ is such that
%   $n=\tr\big(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\A)^{-1}\big)$ and
%   $\Sigmab_\mu=\E_\mu[\x\x^\top]$.
% \end{theorem}
% Observe that our surrogate design $S_{\!\A,\mu}^n$ depends on the
% choice of matrix $\A$ even though the original design $\mu^n$ does
% not. Nevertheless, our empirical results (Figure \ref{f:intro}) suggest that under very mild
% assumptions on $\A$, any surrogate design $S_{\!\A,\mu}^n$ closely
% preserves the properties of the original design.

\begin{enumerate}
  \item Can we empirically show \eqref{eq:hypothesis} for
    non-Euclidean norms, where the minimum-norm estimator does not
    have a closed form expression, such as the $\ell_1$-norm?
  \item If the data distribution is not Gaussian, can we replace the
    square loss  $\l(\yh,y)=(\yh-y)^2$ with some other loss to improve
    the accuracy of the characterization?
\end{enumerate}

\section*{Connections to Fenchel duality}

For a convex function $g : \mathbb{R}^d \to \mathbb{R}$, denote its convex
conjugate $g^*(z) = \sup_x z^\top x - f(x) = -\inf f(x) - z^\top x$. We have
\begin{align*}
  \min_{\w : \X \w = \y} g(\w)
  &= \min_{\w \in \mathbb{R}^d} \max_{\v \in \mathbb{R}^n} g(\w) + \v^\top (\y - \X \w) \\
  &\geq \max_{\v \in \mathbb{R}^n} \min_{\w \in \mathbb{R}^d} g(\w) + \v^\top (\y - \X \w) \\
  &= \max_{\v \in \mathbb{R}^n} \v^\top \y - g^*(\X^\top \v)
\end{align*}
with the inequality an equality under certain conditions (Sion minimax theorem).
Assume equality for what remains.

From first order conditions, the optimal $\v$ satisfies
\begin{align*}
  \y &= \X (\nabla g^*)(\X^\top \v)
\end{align*}
For example, when $g(\z) = \frac{1}{2} \z^\top \z$ then $g = g^*$, $\nabla g^*(\z) = \z$,
and we have $\y = \X \X^\top \v$ hence $\v = (\X \X^\top)^{-1} \y$ and
\begin{align*}
  \max_{\v \in \mathbb{R}^n} \min_{\w \in \mathbb{R}^d} g(\w) + \v^\top (\y - \X \w)
  &= \min_{\w \in \mathbb{R}^d} g(\w) + \y^\top (\X \X^\top)^{-1} (\y - \X \w) \\
  &= \min_{\w \in \mathbb{R}^d} \|\y - \X \w\|_2^2 + g(\w) + \left(\y^\top (\X \X^\top)^{-1} - (\y - \X \w)^\top\right) (\y - \X \w) \\
  &= \min_{\w \in \mathbb{R}^d} \|\y - \X \w\|_2^2 + g(\w) + \left(
    \y^\top ((\X \X^\top)^{-1} - \I) + \w^\top \X^\top
  \right) (\y - \X \w)
\end{align*}
Optimizing over $\w$ only involves retaining terms containing $\w$ so
ignoring constants and focusing on those terms shows
\begin{align*}
  \min_{\w : \X \w = \y} g(\w)
  &= \min_{\w \in \mathbb{R}^d} \|\y - \X \w\|_2^2 + g(\w)
    - \y^\top ((\X \X^\top)^{-1} - \I) \X \w + \w^\top \X^\top \y - \w^\top \X^\top \X \w \\
  &= \min_{\w \in \mathbb{R}^d} \|\y - \X \w\|_2^2 + g(\w)
    + \y^\top (2 \I - (\X \X^\top)^{-1}) \X \w
    - \w^\top \X^\top \X \w
\end{align*}
Assume a linear $y(\x) = \x^\top \w^* + \xi$ model and take expectation,
using Fatou's lemma to move the expectation into the minimization
\begin{align*}
  \E \min_{\w \in \mathbb{R}^d} (\cdot)
  &\leq \min_{\w \in \mathbb{R}^d} \E\left[
    \|\y - \X \w\|_2^2 + g(\w)
    + \y^\top (2 \I - (\X \X^\top)^{-1}) \X \w
    - \w^\top \X^\top \X \w
  \right] \\
  &= \min_{\w \in \mathbb{R}^d} \E\left[
    \|\y - \X \w\|_2^2
    + \w^\top \X^\top (\I - (\X \X^\top)^{-1}) \X \w
  \right]
  + g(\w) \\
  &= \min_{\w \in \mathbb{R}^d} \E\left[
    \|\y - \X \w\|_2^2
    + \w^\top (\X^\top \X - \X^\dag \X) \w
  \right]
  + g(\w)
\end{align*}
Approximating $\E[\X^\top \X - \X^\dag \X] \approx \lambda \I$ and
recalling $g(\w) = \frac{1}{2} \w^\top \w$, we have
\begin{align*}
  \E \min_{\w \in \mathbb{R}^d} (\cdot)
  &\leq \min_{\w \in \mathbb{R}^d} \E\left[
    \|\y - \X \w\|_2^2
  \right]
  + \left(\lambda + 1\right)g(\w)
\end{align*}

\bibliographystyle{alpha}
\bibliography{pap}

\end{document}

### 2022-08-23 design/nips/neurips_2019_author_response.tex
Commit original repos

\documentclass{article}

\usepackage{neurips_2019_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{mwe}

\bibliographystyle{unsrtnat}

\begin{document}

We would like to first stress that the main contribution of this
paper is theoretical (Theorems 1 and 3). For a long time, the area of
experimental design has been dominated by heuristic approaches (such
as greedy bottom-up and Fedorov's exchange method), and obtaining efficient
algorithms which guarantee a $(1+\epsilon)$-approximate solution is
considered a challenging task. For this task, our results
significantly improve on previous works that were published at top
machine learning and computational theory venues such as ICML'17
(\cite{pmlr-v70-allen-zhu17e}), SODA'19 (\cite{Nikolov:2019}) and COLT'19
(\cite{pmlr-v99-madan19a}). Those improvements are not limited to
approximation guarantees but also involve significantly faster
runtimes and a broader scope of applicable optimality criteria. While we do also
present numerical experiments comparing our algorithm to heuristic
approaches (Figure 1 in the submission), our primary point of comparison is with other methods that
provide theoretical guarantees for this task (Table 1 in the submission).

We also note that in both Bayesian and classical experimental design, A/C/D/V-optimality criteria have
historically been the standard evaluation metric (see \cite{chaloner1984optimal}) and
are still used in recent works published at NeurIPS
(\cite{mariet2017elementary}, Figure 1).  Improvements in algorithms for
optimal Bayesian design, such as our paper, immediately imply improved
performance guarantees for practical experimental design applications
ranging from clinical trials \cite{flournoy1993clinical,berry2002adaptive}
to materials science \cite{terejanu2012bayesian}.

In the final version, we will add a conclusion section which
consolidates our results for ease of reference and differentiation
from prior work (as requested by Reviewers 1 and 4).

\subsubsection*{Comments of Reviewer 3}
\begin{quote}
  \emph{``There are many Theorems/Lemmas and it is not always clear
    which are new results, and which are previous work.''} 
\end{quote}
As discussed above, our primary contribution is theoretical, and in
fact \emph{all} the lemmas and theorems presented in this paper are
novel results. We will clarify this in the final version.

\begin{quote}
\emph{``It would be much more convincing to add another figure that
  shows mean/sd of test error in k-fold CV.''}
\end{quote}
Typically, in experimental design literature we use the optimality
criteria as the performance metric. Nevertheless, we acknowledge that evaluating
mean-squared error under cross-validation may be more meaningful for NeurIPS
readers and we are happy to include this in the final version.%Figure~\ref{fig:mse} in our paper.

\subsubsection*{Comments of Reviewer 4}
\begin{quote}
  \emph{``The main theorems (Theorem 1-2) rely on sampling from DPPs,
    but the statement does not involve probabilistic modifications (in
    expectation, with high probability, etc.).''}
\end{quote}
Even though the output of our algorithms is randomized, the
approximation guarantees in Theorems 1 and 3 are in fact
deterministic. We achieve this by running the randomized procedure
repeatedly until the guarantee is satisfied. Because of this, the time
complexity guarantee holds with high probability, as explained in the
proof of Lemma 13 (page 7). We will clarify this further in the final version.

\begin{quote}
  \emph{``It would be helpful if the algorithms are summarized as
  pseudo-code and pointed from Introduction.''}
\end{quote}
We provide pseudo-code for our algorithm on page 5, document the
SDP being solved in equation (5), and describe the algorithm used to solve
the SDP in the experiments section (Section 5). We will try to
consolidate this information into one pseudo-code in the final version.

% \begin{figure}[h]
%   \centering
%   \includegraphics[width=0.30\textwidth]{example-image-a}
%   \caption{Regularized mean-squared error loss computed on remaining $n-k$
%     points.}
%   \label{fig:mse}
% \end{figure}

% While reviewer 4 is correct in observing that the greedy bottom-up method
% performs similarly to our method, we do not believe this is a significant
% concern because (1) most previous work in Bayesian experimental design
% (\cite{singh2018approximate,mariet2017elementary}) also do not significantly
% outperform greeedy bottom up / Fedorov exchange algorithms in experiments and
% (2) the theoretical guarantees for greedy methods (which rely on weak
% submodularity) are much looser and only applicable to D-optimality
% \cite{bouhtou2010submodularity}) whereas our theoretical results apply to all
% of the A/C/D/V-optimality criteria.

% Finally, we do not believe reviewer 4's concerns on pseudo-code are valid
% because we provide pseudo-code for our algorithm on page 5, document the
% SDP being solved in equation 5, and describe the algorithm used to solve
% the SDP in the experiments section (section 5).

\renewcommand\refname{\normalsize References}
\setlength{\bibsep}{0pt plus 0.2ex}
\tiny
\bibliography{refs}

\end{document}

### 2022-08-23 design/nystrom-appendix.tex
Commit original repos

\documentclass{article}

% Recommended, but optional, packages for figures and better
% typesetting:
\usepackage{fullpage}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{enumitem}
%\usepackage{hyperref}
\usepackage{cleveref}
\input{shortdefs}

\title{\textbf{Appendix for:} A multiple-descent curve in the Column Subset Selection
  Problem
  and the Nystr\"om method}
\date{}

  \begin{document}
  \maketitle

  \appendix

  \section{Expectation formula for the random projection matrix}
  \begin{theorem}\label{l:expectation_proj}
  For any $\A$ and $S\subseteq [n]$, let $\P_S$ denote the  
projection onto the $\mathrm{span}\{\a_i:i\in S\}$. If $S\sim \DPP(\A^\top\A)$, then
\begin{align*}
  \E[\P_S] & = \A(\I+\A^\top\A)^{-1}\A^\top.
\end{align*}
\end{theorem}
\begin{proof}
Fix $m$ as the column dimension of $\A$ and let $\A_S$ denote the
submatrix of $\A$ consisting of the columns 
indexed by $S$. We have $\P_S=\A_S(\K_{S,S})^+\A_S$, where $^+$
denotes Moore-Penrose pseudo-inverse and $\K=\A^\top\A$. Let
$\v\in\R^m$ be an arbitrary vector. When $\K_{S,S}$ is
invertible, then a standard determinantal 
identity states that:
  \begin{align*}
\det(\K_{S,S})\v^\top\P_S\v
    =\det(\K_{S,S})\v^\top\A_S\K_{S,S}^{-1}\A_S^\top\v = \det(\K_{S,S}+\A_S^\top\v\v^\top\A_S)-\det(\K_{S,S}).
  \end{align*}
When $\K_{S,S}$ is not invertible then
$\det(\K_{S,S})=\det(\K_{S,S}+\A_S^\top\v\v^\top\A_S)=0$, because the
rank of $\K_{S,S}+\A_S^\top\v\v^\top\A_S=\A_S^\top(\I+\v\v^\top)\A_S$
cannot be higher than the rank of $\K_{S,S}=\A_S^\top\A_S$. Thus,
\begin{align*}
  \det(\I+\K)\v^\top\E[\P_S]\v
  &= \sum_{S\subseteq[n]:\,\det(\K_{S,S})>0}\det(\K_{S,S})
\v^\top\A_S\K_{S,S}^{-1}\A_S^\top\v\\
  &=\sum_{S\subseteq[n]}\det(\K_{S,S}+\A_S^\top\v\v^\top\A_S)-\det(\K_{S,S})\\
  &=\sum_{S\subseteq[n]}\det\!\big([\K+\A^\top\v\v^\top\A]_{S,S}\big)-\det(\K_{S,S})\\
  &\overset{(*)}{=}\det(\I + \K + \A^\top\v\v^\top\A) - \det(\I+\K)\\
  &=\det(\I+\K)\v^\top\A(\I+\K)^{-1}\A^\top\v,
\end{align*}
where $(*)$ involves two applications of the determinantal summation identity (Theorem
2.1, \cite{Kulesza2012book}) that corresponds to computing the
normalization constant $\det(\I+\K)$ for a DPP. Since the above
calculation holds for arbitrary vector $\v$, the claim follows.
\end{proof}

  \section{Non-trivial results for size less than stable rank}
  We first point out the trivial upper bound.
  \begin{theorem}
    For any $\A$ and 
    $r=\sr(\A)$, for all subsets $S$ of size $k<r$ we have
    \begin{align*}
      \Er_\A(S)\leq \Big(\,1+\tfrac{k}{r-k}\,\Big)\,\|\A-\A_k\|_F^2.
    \end{align*}
  \end{theorem}
  \begin{proof}
    First note that $\|\A-\A_k\|_F^2\geq \lambda_1(r-k)$, where
    $r=\sr(\A)$ and $\lambda_1$ is the largest eigenvalue of
    $\A^\top\A$. It follows that:
    \begin{align*}
      \Er_\A(S) = \|\A-\P_S\A\|_F^2 \leq \|\A\|_F^2= r\cdot \lambda_1
      \leq \frac{r}{r-k}\,\|\A-\A_k\|_F^2,
    \end{align*}
    which concludes the proof.
  \end{proof}

  Next, we point out a non-trivial bound which improves the factor
  when $k>r/2$.
  \begin{theorem}
Let $r=\sr(\A)$ and $k\in (\frac r2, r)$. There is $\alpha$ such that for
    $S\sim\DPP(\frac1\alpha\A^\top\A)$,
    \begin{align*}
      \E[|S|]\leq k,\quad\text{and}\quad\E\big[\Er_\A(S)\big] \leq
     \Big(\,1 +  \sqrt{\!\tfrac {k}{r-k}}\,\Big)\cdot\|\A-\A_k\|_F^2.
    \end{align*}
  \end{theorem}

  \begin{proof}
    Without loss of generality, assume that $\lambda_1=1$. Set
    $\alpha=E_k/(\sqrt{r-k}(\sqrt k-\sqrt{r-k}))$, where $E_k=\sum_{j>k}\lambda_j=\|\A-\A_k\|_F^2\geq r-k$. Let
    $\lambda_1'=...=\lambda_k'=1$, and for $i>k$ set
    $\lambda_i'=\beta\lambda_i$, where $\beta = \frac{r-k}{E_k}\leq 1$. Now, let $\alpha'=\beta\alpha$. We have:
  \begin{align*}
    \E[|S|] &= \sum_{i}\frac{\lambda_i}{\lambda_i+\alpha} 
   \leq\sum_{i=1}^k\frac{\lambda_i}{\lambda_i+\beta\alpha}
    + \sum_{i>k}\frac{\beta\lambda_i}{\beta\lambda_i+\beta\alpha}\\
&   \leq \sum_i \frac{\lambda_i'}{\lambda_i'+\alpha'}
=\frac{k}{1+\alpha'} + \sum_{i> k}\frac{\lambda_i'}{\lambda_i'+\alpha'}\\
    &\leq
k - \frac k{1+\frac1{\alpha'}}+  \frac1{\alpha'} \sum_{i>k} \lambda_i' \\
    &\overset{(*)}{=} k - \frac k{1 + \frac{\sqrt{k(r-k)} - (r-k)}{r-k}}
      +\frac{r-k}{\alpha'}\\
    &=k - \sqrt{k(r-k)} + \sqrt{k(r-k)} - (r-k)\leq k - (r-k).
  \end{align*}
    where $(*)$ follows because $\sum_{i>
      k}\lambda_i'=(r-k)E_k/E_k=r-k$. We bound the error using as follows:
    \begin{align*}
      \E\big[\Er_\A(S)\big] = \alpha\,\E[|S|]\leq
      \frac{k-(r-k)}{\sqrt{r-k}(\sqrt k-\sqrt{r-k})}\,E_k
      =\frac{\sqrt k +\sqrt{r-k}}{\sqrt{r-k}}\,E_k,
    \end{align*}
which concludes the proof.
  \end{proof}

    \begin{theorem}
    	\label{t:upperbound_withs}
      For any $s\in[n]$, let $r=\sr(\A-\A_s)$ and suppose that
      $s+\frac7{\epsilon^4}\ln^2\frac1\epsilon\leq k\leq r-1$ for $\epsilon\in(0,\frac12)$.
%      and $k\geq (\frac6{\epsilon^2}\ln\frac1\epsilon)^2$ for $\epsilon\in(0,\frac12)$.
      There is $\alpha$ such that for 
    $S\sim\DPP(\frac1\alpha\A^\top\A)$,
    \begin{align*}
\E\big[\Er_\A(S) \mid |S|\leq s+k\big] \leq
      (1+2\epsilon)^2\Big(\,1+\frac{s}{k}\,\Big)\sqrt{1+\tfrac{2k}{r-k}\,}\cdot\|\A-\A_{s+k}\|_F^2.
    \end{align*}
  \end{theorem}
  \begin{proof}
      Without loss of generality, assume that $\lambda_{s+1}=1$. Set
    $\alpha=\sqrt{\frac{r+k}{r-k}}\frac{E_{s+k}}{(1-\epsilon)k}$, where
    $E_{s+k}=\sum_{j>s+k}\lambda_j=\|\A-\A_{s+k}\|_F^2\geq r-k$. Let 
    $\lambda_{s+1}'=...=\lambda_{s+k}'=1$, and for $i>s+k$ set
    $\lambda_i'=\beta\lambda_i$, where $\beta =
    \frac{r-k}{E_{s+k}}\leq 1$. Now, let
    $\alpha_{i}=\beta\alpha=\frac{\sqrt{r^2-k^2}}{(1-\epsilon)k}$
for $i>s+k$    and for $s+1\leq i\leq s+k$, let
    \[\alpha_{i}'=(1-\epsilon)\frac{\sqrt{r+k}+\sqrt{r-k}}{2\sqrt{r+k}}\beta\alpha
    =\frac{\sqrt{r-k}(\sqrt{r+k}+\sqrt{r-k})}{r+k - (r-k)} = \frac{\sqrt{r-k}}{\sqrt{r+k}-\sqrt{r-k}}.\]
Note that $\alpha_i'\leq \alpha$ for all $i$, so we have:
  \begin{align*}
    \E[|S|] &= \sum_{i}\frac{\lambda_i}{\lambda_i+\alpha} 
   \leq s + \sum_{i=s+1}^{s+k}\frac{\lambda_i}{\lambda_i+\alpha_i'}
    + \sum_{i>s+k}\frac{\beta\lambda_i}{\beta\lambda_i+\beta\alpha}\\
&   \leq s + \sum_{i>s} \frac{\lambda_i'}{\lambda_i'+\alpha_i'}
=s + \frac{k}{1+\alpha'} + \sum_{i>s+k}\frac{\lambda_i'}{\lambda_i'+\alpha'}\\
    &\leq s+
k - \frac {k}{1+\frac1{\alpha'}}+  \frac1{\alpha'} \sum_{i>s+k} \lambda_i' \\
    &\overset{(*)}{=} s + k - \frac{k}{1+\frac{\sqrt{r+k}-\sqrt{r-k}}{\sqrt{r-k}}}
      +\frac{r-k}{\alpha'}\\
                &=s + k - \frac{k\sqrt{r-k}}{\sqrt{r+k}}+
                  (1-\epsilon)\frac{k\sqrt{r-k}}{\sqrt{r+k}}\\
            &= s + k - \frac{\epsilon k\sqrt{r-k}}{\sqrt{r+k}}
              \leq s+k,
  \end{align*}
    where $(*)$ follows because
    $\sum_{i>s+ k}\lambda_i'=(r-k)E_{s+k}/E_{s+k}=r-k$.
      Now, let $p_i=\frac{\lambda_i'}{\lambda_i'+\alpha_i'}$ be the
      Bernoulli probabilities for
  $b_i\sim\mathrm{Bernoulli}(p_i)$
  and $X =\sum_{i>s} b_i$. From the above, we have:
  \begin{align*}
k - \E[X] \geq \frac{\epsilon k\sqrt{r-k}}{\sqrt{r+k}},
  \end{align*}
  and furthermore:
  \begin{align*}
    \Var[X]&\leq \sum_{i=s+1}^{s+k}(1-p_i) + \sum_{i>s+k}p_i
    \leq \frac{k\sqrt{r-k}}{\sqrt{r+k}}+
      (1-\epsilon)\frac{k\sqrt{r-k}}{\sqrt{r+k}} = (2-\epsilon) \frac{k\sqrt{r-k}}{\sqrt{r+k}} .
  \end{align*}
  Using
  Theorem 2.6 from \cite{ChungLu2006book} with $\lambda=\frac{\epsilon
    k\sqrt{r-k}}{\sqrt{r+k}}$, we have:
  \begin{align*}
    \Pr(|S|>s+k)&\leq \Pr(X>k)\leq \Pr(X > \E[X]+\lambda)
    \leq \exp\Big(-\frac{\lambda^2}{2(\Var[X]+\lambda/3)}\Big)\\
    &\leq \exp\Big(-\frac{\lambda^2}{2(\frac{2-\epsilon}\epsilon
      \lambda+\lambda/3)}\Big)\leq\exp(-\epsilon\lambda /4)
      =\exp\Big(-\frac{\epsilon^2k\sqrt{r-k}}{4\sqrt{r+k}}\Big). 
  \end{align*}
  Note that if $7\leq k\leq r-1$, then $k\frac{\sqrt{r-k}}{\sqrt{r+k}}\geq
\frac{k}{\sqrt{2k+1}}\geq \frac{7}{16}\sqrt k$, so by simple algebra
it follows that for $k\geq \frac 7{\epsilon^4}\ln^2\frac1\epsilon$,
we have $k\frac{\sqrt{r-k}}{\sqrt{r+k}}\geq
\frac4{\epsilon^2}\ln\frac1\epsilon$ and therefore
$\Pr(|S|>s+k)\leq \epsilon$. Finally, we get:
    \begin{align*}
      \frac{\E\big[\Er_\A(S)\mid |S|\leq s+k\big]}{\|\A-\A_{s+k}\|_F^2}
      &=
        \frac{\alpha\,\E[|S|]}{E_{s+k}}
        \leq(1-\epsilon)^{-2}\frac{s+k}{k}\sqrt{\tfrac{r+k}{r-k}\,},
        =(1+2\epsilon)^2\Big(1+\frac sk\Big)\sqrt{1 + \tfrac {2k}{r-k}\,},        
    \end{align*}
    which completes the proof.
\end{proof}

  \begin{theorem}
For any $s\in[n]$, let $r=\sr(\A-\A_s)$ and suppose that $k\in(\frac
r2,r-1)$ and $k\ge0q (\frac6{\epsilon^2}\ln\frac1\epsilon)^2$ for
$\epsilon\in(0,\frac12)$. There is $\alpha$ such that for 
    $S\sim\DPP(\frac1\alpha\A^\top\A)$,
    \begin{align*}
\E\big[\Er_\A(S)\mid |S|\leq s+k\big] \leq
(1+2\epsilon)^2\Big(\,1+\frac{s}{2k-r}\,\Big)\Big(\,1 +  \sqrt{\!\tfrac {k}{r-k}}\,\Big)\cdot\|\A-\A_{s+k}\|_F^2.
    \end{align*}
  \end{theorem}

  \begin{proof}
    Without loss of generality, assume that $\lambda_{s+1}=1$. Set
    $\alpha=E_{s+k}/((1-\epsilon)\sqrt{r-k}(\sqrt k-\sqrt{r-k}))$, where
    $E_{s+k}=\sum_{j>s+k}\lambda_j=\|\A-\A_{s+k}\|_F^2\geq r-k$. Let 
    $\lambda_{s+1}'=...=\lambda_{s+k}'=1$, and for $i>s+k$ set
    $\lambda_i'=\beta\lambda_i$, where $\beta = \frac{r-k}{E_{s+k}}\leq 1$. Now, let $\alpha'=\beta\alpha$. We have:
  \begin{align*}
    \E[|S|] &= \sum_{i}\frac{\lambda_i}{\lambda_i+\alpha} 
   \leq s + \sum_{i=s+1}^{s+k}\frac{\lambda_i}{\lambda_i+\beta\alpha}
    + \sum_{i>s+k}\frac{\beta\lambda_i}{\beta\lambda_i+\beta\alpha}\\
&   \leq s + \sum_{i>s} \frac{\lambda_i'}{\lambda_i'+\alpha'}
=s + \frac{k}{1+\alpha'} + \sum_{i>s+k}\frac{\lambda_i'}{\lambda_i'+\alpha'}\\
    &\leq s+
k - \frac k{1+\frac1{\alpha'}}+  \frac1{\alpha'} \sum_{i>s+k} \lambda_i' \\
    &\overset{(*)}{\leq} s + k - \frac k{1 + \frac{\sqrt{k(r-k)} - (r-k)}{r-k}}
      +\frac{r-k}{\alpha'}\\
    &=s + k - \sqrt{k(r-k)} + (1-\epsilon)\big(\sqrt{k(r-k)} -
      (r-k)\big)\\
            &\leq s+k -\epsilon \sqrt{k(r-k)} - (1-\epsilon)(r-k)
              \leq s+k-(r-k).
  \end{align*}
  where $(*)$ follows because
  $\sum_{i>s+ k}\lambda_i'=(r-k)E_{s+k}/E_{s+k}=r-k$.
  Now, let
  $b_i\sim\mathrm{Bernoulli}(\frac{\lambda_i'}{\lambda_i'+\alpha'})$. Using
  Theorem 2.6 from \cite{ChungLu2006book}, we have:
  \begin{align*}
    \Pr(|S|>s+k) \leq \Pr\Big(\sum_{i>s}b_i>k\Big)\leq
    \exp\Big(-\frac{\epsilon^2k(r-k)}{2(\sqrt{k(r-k)}(\frac1{1-\epsilon}+(1-\epsilon)+\frac\epsilon
   3))}\Big)\leq \ee^{-\epsilon^2\sqrt{k}/6}\leq \epsilon.
  \end{align*}
  We bound the error as follows: 
    \begin{align*}
      \frac{\E\big[\Er_\A(S)\mid |S|\leq
      s+k\big]}{\|\A-\A_{s+k}\|_F^2}
      &= \frac{\alpha\,\E[|S|]}{\Pr(|S|\leq s+k)E_{s+k}}\leq
      \frac1{(1-\epsilon)^2}\frac{s+k-(r-k)}{\sqrt{r-k}(\sqrt k-\sqrt{r-k})}\\
      &\leq(1+2\epsilon)^2\bigg(\frac{s(\sqrt{k}+\sqrt{r-k})}{\sqrt{r-k}(k-(r-k))}+\frac{\sqrt k
        +\sqrt{r-k}}{\sqrt{r-k}}\bigg) \\
      &=(1+2\epsilon)^2\bigg(1 +
        \frac{s}{k-(r-k)}\bigg) \frac{\sqrt k +\sqrt{r-k}}{\sqrt{r-k}},
    \end{align*}
which concludes the proof.
  \end{proof}
  \begin{theorem}\label{t:generalupperbound}
    For any $s\in[n]$, let $r=\sr(\A-\A_s)$ and suppose that
    $\frac2{\epsilon^2}\ln\frac1\epsilon\leq k<r$ for $\epsilon\in(0,\frac12)$.
There is $\alpha$ such that if $S\sim\DPP(\frac1\alpha\A^\top\A)$ then
\begin{align*}
  \E\big[\Er_\A(S)\mid |S|\leq s+k\big]\leq (1+2\epsilon)^2 \Big(1+\frac
          sk\Big)\Big(1 + \frac{k}{r-k}\Big)\,\|\A -
  \A_{s+k}\|_F^2.%\quad\text{and}\quad \E[|S|]\leq s+k.
\end{align*}
  \end{theorem}
\begin{proof}
	We have
	\begin{align*}
	E_{s+k}=\|\A-\A_{s+k}\|_F^2\geq 
	\frac{\lambda_{s+1}}{\lambda_{s+1}}\sum_{i>s+k}\lambda_i
          =\lambda_{s+1} \Big( r -
          \frac{\sum_{s< i\leq s+k}\lambda_i}{\lambda_{s+1}}\Big)\geq
\lambda_{s+1}(r - k).
	\end{align*}

Let $\alpha = \frac{r}{r-k}\frac{E_{s+k}}{(1-\epsilon)k}$. This makes sure that,

\begin{align*}
  \E[|S|] = \sum_i\frac{\lambda_i}{\lambda_i+\frac{r}{r-k} \frac{E_{s+k}}{(1-\epsilon)k}}
  \leq \sum_i\frac{\lambda_i}{\lambda_i+\frac{\lambda_{s+1}}{1-\epsilon}\frac{r}{k}}
  \leq
  s+(1-\epsilon)\sum_{i>s}\frac{\lambda_i}{\lambda_{s+1}\frac{r}k}
  =s + (1-\epsilon)k.
\end{align*}
Since $|S|=\sum_i b_i$, where
$b_i\sim\mathrm{Bernoulli}(\frac{\lambda_i}{\lambda_i+\alpha})$, and
$\E\big[\sum_{i>s}b_i\big]\leq (1-\epsilon)k$, we have:
\begin{align*}
  \Pr(|S|>s+k)\leq \Pr\Big(\sum_{i>s}b_i>k\Big)\leq \ee^{-\epsilon^2 k/2}.
\end{align*}
Note that if $\epsilon\in(0,\frac12)$ and $k\geq
\frac2{\epsilon^2}\ln\frac1\epsilon$, then $\ee^{-\epsilon^2 k/2}\leq
\epsilon$, so we get:
	\begin{align*}
          \frac{\E[\Er_{\A}(S)\mid |S|\leq s+k]}{\|\A-\A_{s+k}\|_F^2}
          &= \frac{\alpha \E[|S|]}{\Pr(|S|\leq s+k) E_{s+k}} 
          \leq
            \frac{1}{(1-\epsilon)^2}\frac{r}{r-k}\frac{s+k}{k} \\
          &\leq (1+2\epsilon)^2 \Big(1+\frac
          sk\Big)\Big(1 + \frac{k}{r-k}\Big),
	\end{align*}
        which concludes the proof.
\end{proof}

  \begin{theorem}\label{t:upper_conditionnum}
Given $\A$ and $r<k\leq n$, let $\kappa_r$ denote the condition
    number of $\A-\A_{r}$. There is $\alpha$ such
    that if $S\sim \DPP(\frac1\alpha\A^\top\A)$, then
    \begin{align*}
\E[|S|]\leq k\quad\text{and}\quad \E[\Er_\A(S)] \leq
      \kappa_r^2\Big(1+\frac r{k-r}\Big)\,\|\A-\A_k\|_F^2.
      \end{align*}
    \end{theorem}
    \begin{proof}
      Let $\alpha=\frac{\kappa_r^2}{k-r}E_k$, where
      $E_k=\|\A-\A_k\|_F^2$. Note that
      $\kappa_r^2=\frac{\lambda_{r+1}}{\lambda_n}$, where $\lambda_i$ is
      the $i$th largest eigenvalue of $\A^\top\A$. Note that $E_k\geq
      (n-k)\lambda_n$, so we have
      \begin{align*}
        \E[|S|]&\leq \sum_i\frac{\lambda_i}{\lambda_i+\alpha}\leq r +
        \frac{(n-r)\lambda_{r+1}}{\lambda_{r+1}+
        \frac{\kappa_r^2}{k-r}(n-k)\lambda_n}\\
               &=r + \frac{1}{\frac1{n-r} + \frac1{k-r}\,\frac{n-k}{n-r}} \\
        &=r + \frac{1}{\frac1{n-r} + \frac1{k-r}\,(1 - \frac{k-r}{n-r})} \\
& =  r + (k-r)= k.
\end{align*}
Finally, same as before, we have $\E[\Er_\A(S)]=\alpha\E[|S|]\leq
\kappa_r^2(1+\frac r{k-r})\,E_k$.
    \end{proof}

\begin{theorem}
	\label{t:upperbound} 	For any $\A \in \R^{m\times n}$, any $s\in[n]$, let $r=\sr(\A-\A_s)$. For any $\epsilon > 0$ s.t. $\epsilon^2 \geq 4s(1+\epsilon)/r$, let

	\begin{align*}
	 &\Phi_l = \frac{r}{2(1+\epsilon)} \left(  \epsilon - \sqrt{\epsilon^2 - \frac{4s(1+\epsilon)}{r} }  \right), \; & \Phi_u = \frac{r}{2(1+\epsilon)} \left(  \epsilon + \sqrt{\epsilon^2 - \frac{4s(1+\epsilon)}{r}  }  \right) \\
	 &\Psi_l =  \frac{2}{\epsilon^2} \log (2+ \frac{1}{\epsilon}) - s
	 \end{align*}

	If k satisfies: $ \max(\Phi_l, \Psi_l) \leq k \leq  \Phi_u$, we can find a set $S$ of size $|S|\leq s+k$ with $\Er_\A(S)  \leq (1+3\epsilon)  \| \A - \A_{s+k}\|_F^2$ with $O(\frac1{\epsilon})$ rounds of DPP rejection sampling on average. 
\end{theorem}
\begin{proof}
	We make use of concentration arguments in conjunction with Theorem~\ref{t:generalupperbound}.

	To re-write the RHS in the result of Theorem~\ref{t:generalupperbound} in the form of $(1+\epsilon)$ we need (with $p=\frac{1+\epsilon}{r}$),

	\begin{align*}
	 & (1 + \frac{s}{k}) (\frac{r}{r-k}) \leq 1+ \epsilon  \iff \left( \frac{s+k}{k} \right) \frac{1}{r-k} \leq p \\ 
	 & \iff rkp - k^2p \geq s + k  \iff pk^2 - \epsilon k + s \leq 0,
	\end{align*}

	which holds for $k \in \frac{1}{2p}[ \epsilon - (\epsilon^2 - 4ps )^{\frac{1}{2}}, \epsilon + (\epsilon^2 - 4ps )^{\frac{1}{2}}]$.

	 In other words, if $k$ satisfies the above condition, from Theorem~\ref{t:generalupperbound}, we have $\E[\Er_{\A}(S)] = \Big(1+\frac
	 sk\Big)\Big(1 + \frac{k}{r-k}\Big)\,\|\A-\A_{s+k}\|_F^2 \leq (1+\epsilon) \|\A-\A_{s+k}\|_F^2$.

	For concentration, we make use of the fact that the size of the selected set through DPP sampling follows Poisson Binomial, which has the following concentration bound~\cite{ChungLu2006book}:

	\begin{align*}
	\Pr(|S| > t) \leq \ee^{-\frac{(t-(1-\epsilon)t))^2}{2t}} =
	\ee^{-\epsilon^2 t/2},
	\end{align*}

	where $t=k+s$ for this result. Thus, using Theorem~\ref{t:generalupperbound}, we get	
	\begin{align*}
	& \E\big[\Er_{\A}(S)\mid |S|\leq t\big] \leq \frac{\E[\Er_{\A}(S)]}{\Pr(|S|\leq t)} \leq \frac {(1+\epsilon)\| \A - \A_{t}\|_F^2}{(1-\ee^{-\delta^2t/2}) }.  
	\end{align*}

	With $\delta \leq \epsilon$, and $t \geq \frac{2}{\epsilon^2} \ln \left(2 + \frac1{\epsilon} \right)$, which ensures $\ee^{-\delta^2 t/2} \leq \frac{\epsilon}{1+2\epsilon}$, we get, $\E\big[\Er_{\A}(S)\mid |S|\leq t\big] \leq (1+2\epsilon) \| \A - \A_{t} \|_F^2$.

	Note that for $t \geq \frac{2}{\epsilon^2} \ln \left(2 + \frac1{\epsilon} \right)$, we have $|S| \leq t$ with probability $\frac{1+\epsilon}{1+2\epsilon}$. Using shorthand $E_t=\E\big[\Er_{\A}(S)\mid |S|\leq t\big]$ in the Markov inequality, we have,

	\[ P\big[\Er_{\A}(S) \geq \eta E_t \mid |S| \leq t \big] \leq \frac{1}{\eta}. \]

	Thus, with probability $\geq  \frac{1+\epsilon}{1+2\epsilon} (1 - \frac1{\eta})$, we have $|S| \leq t$ and  $\Er_{\A}(S) \leq \eta E_k$. Choosing $\eta = (1+\frac \epsilon {1+2\epsilon})$ ensures $\Er_{\A}(S) \leq (1+3\epsilon)E^*_k$ with probability $\geq  \frac{(1+\epsilon)\epsilon}{(1+2\epsilon)(1+3\epsilon)}$, which means on average $\frac{(1+2\epsilon)(1+3\epsilon)}{(1+\epsilon)\epsilon} = O(1/\epsilon)$ rounds of rejection sampling suffice.

\end{proof}

\begin{theorem}
  For any $\A\in\R^{m\times n}$, $k\in[n]$ and $\alpha>0$, if
  $S\sim\DPP(\frac1\alpha\A^\top\A)$, then
  \begin{align*}
    \E\big[\Er_\A(S)\mid |S|=k\big] \leq \E\big[\Er_\A(S)\mid |S|\leq k\big].
  \end{align*}
\end{theorem}
\begin{proof}
Let $e_k$ denote the $k$th elementary symmetric polynomial defined
as:
\begin{align*}
e_k = \sum_{T:|T|=k}\det(\A_T^\top\A_T) = \sum_{T:|T|=k}\prod_{i\in T}\lambda_i.
\end{align*}
Also let $\bar{e}_k=e_k/{n\choose k}$ denote the $k$th
elementary symmetric mean. Newton's inequalities imply that:
\begin{align*}
1\geq \frac{\bar e_{k-1}\bar e_{k+1}}{\bar e_k^2} =
  \frac{e_{k-1}e_{k+1}}{e_k^2}\,\frac{{n\choose k}}{{n\choose
  k-1}}\frac{{n\choose k}} {{n\choose k+1}} =
  \frac{e_{k-1}e_{k+1}}{e_k^2}\,\frac{n+1-k}{k}\,\frac{k+1}{n-k}. 
\end{align*}
The results of
\cite{pca-volume-sampling,more-efficient-volume-sampling} establish
that $\E[\Er_\A(S)\mid |S|=k] = (k+1)\frac{e_{k+1}}{e_k}$, so it
follows that:
\begin{align}
  \frac{\E[\Er_\A(S)\mid |S|=k]}{\E[\Er_\A(S)\mid|S|=k-1]}
  = \frac{k+1}{k}\,\frac{e_{k+1}e_{k-1}}{e_k^2}\leq
  \frac{n-k}{n+1-k}\leq 1.\label{eq:decreasing-error}
\end{align}
Finally, note that $\E[\Er_\A(S)\mid|S|\leq k]$ is a weighted average
of components $\E[\Er_\A(S)\mid|S|= s]$ for $s\in[k]$, and
\eqref{eq:decreasing-error} implies that the smallest of those components
is associated with $s=k$. Since the weighted average is lower bounded by the
smallest component, this completes the proof.
\end{proof}
\begin{remark}
If we show that the function $f(k)=\E[\Er_\A(S)\mid |S|=k]$ is
convex then we avoid the concentration argument, because then if
$S\sim\DPP(\frac1{\alpha}\A^\top\A)$ satisfies
$\E[|S|]=k$, Jensen's inequality implies:
\begin{align*}
  \E\big[\Er_\A(S)\mid |S|\!=\!k\big]=f\Big(\sum_{i\geq 0}\Pr(|S|\!=\!i)\,i\Big) \leq
  \sum_{i\geq 0}\Pr(|S|\!=\!i)f(i)=\E\big[\Er_\A(S)\big].
\end{align*}
\end{remark}
% \begin{theorem}
% If $S\sim\DPP(\frac1\alpha\A^\top\A)$, then
% $f(k)=\E[\Er_\A(S)\mid|S|=k]$ is a convex function.
% \end{theorem}
% \begin{proof}
% % It suffices to show that $f(k)$ is log-convex (a stronger statement),
% % i.e., that $f(k)^2\leq f(k-1)f(k+1)$ for all $k$. Note that
% We can write $f(k)$ in terms of the ratios of elementary symmetric
% means as:
% \begin{align*}
%   f(k) = (k+1)\frac{e_{k+1}}{e_k} = (n-k)\frac{\bar e_{k+1}}{\bar e_k}.
% \end{align*}
% We will first show that $g(k) = \frac{\bar e_{k+1}}{\bar e_k}$ is
% log-convex. We use Theorem 77 p.~64 from \cite{hlp-inequalities} with
% \begin{align*}
%   \alpha_i'=\begin{cases}
%     2 &\text{if }i=k-1,\\
%     4&\text{if }i=k+1,\\
%     1&\text{otherwise},
%   \end{cases}\qquad
%        \alpha_i=\begin{cases}
%          4&\text{if }i=k,\\
%          2&\text{if }i=k+2,\\
%          1&\text{otherwise}.
%          \end{cases}
% \end{align*}
% Verifying the condition of Theorem 77, for any $m$ we have:
% \begin{align*}
%   \alpha_{k-1}'(m-1)+\alpha_k'm + \alpha_{k+1}'(m+1)+\alpha_{k+2}'(m+2)
%   &= 2(m-1)+m+4(m+1)+(m+2)\\
%   &=8m+4\\
%   &=(m-1)+4m+(m+1)+2(m+2)\\
%   &=\alpha_{k-1}(m-1)+\alpha_km+\alpha_{k+1}(m+1)+\alpha_{k+2}(m+2).
% \end{align*}
% Therefore, Theorem 77 implies that $\prod_i \bar
% e_i^{\alpha_i'}\leq\prod_i\bar e_i^{\alpha_i}$, which simplifies to
% $\bar e_{k-1}\bar e_{k+1}^3\leq \bar e_k^3\bar e_{k+2}$. From this we
% obtain that:
% \begin{align*}
%   g(k)^2 = \frac{\bar e_{k+1}^2}{\bar e_k^2}\leq \frac{\bar e_k}{\bar
%   e_{k-1}}\frac{\bar e_{k+2}}{\bar e_{k+1}} = g(k-1)g(k+1),
% \end{align*}
% so be definition, $g(k)$ is log-convex, which also implies
% convexity. Finally, we establish convexity of $f(k)$:
% \begin{align*}
%   f(k-1) + f(k+1) - 2f(k)
%   &= (n-k+1)g(k-1) + (n-k-1)g(k+1) - 2(n-k)g(k)\\
%   &= (n-k)\big(g(k-1) + g(k+1) - 2g(k)\big) + \big(g(k-1) -
%     g(k+1)\big)\\
%   &\overset{(*)}{\geq} 0,
% \end{align*}
% where in $(*)$ we used that $g(k)$ is convex and non-increasing.
% \end{proof}
%
%\begin{theorem}\label{t:upperbounds_withdecay} For any $\A$ and let $\lambda_1, \ldots, \lambda_n$ be the eigenvalues of $\A^\top \A$. 
%	
%
%	Suppose that $s+ \frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$,
%	where $\epsilon\in(0,\frac12)$. Say $S$ is a volume sampled set of size
%	$k$. We have the following approximation guarantees based on the decay of eigenvalues of $\K$.
%	 \begin{enumerate}
%	 		\item Exponential decay: $\lambda_i = \lambda_1(1-\delta)^i$ for $0<\delta<1$. Then,
%	 		\[\frac{\E[\Er(S)]}{\opt} \leq (1+2\epsilon)^2 \Phi_s(k), \]
%	 		
%	 		where,
%	 		
%	 		\begin{itemize}
%	 		\item $\Phi_s(k) \leq \frac{1}{1-2\delta k}$, if $k \leq \frac{1}{4\delta}$ 	
%	 		\item $\Phi_s(k) \leq 8\delta k$, if $ \frac{1}{4\delta} <k \leq n-\frac{\ln 2}{\delta}$
%	 		\item  $\Phi_s(k) \leq \frac{\delta k}{4\ln 2}$ if $k>n-\frac{\ln 2}{\delta}$
%	 		\end{itemize}
%	 		
%	 		
%	 	\item Polynomial decay: $\lambda_i = \lambda_1/i^p, p >1$. We can bound the approximation guarantee as 
%	 	 
%	 	 	\[\tfrac{\E[\Er(S)]}{\opt} \leq (1+2\epsilon)^2\Big(\,1+\frac{s}{k}\,\Big)\sqrt{1+\tfrac{2pk}{s-p -pk }}\]
%	 	\end{enumerate}
%
%\end{theorem}

\begin{theorem}\label{t:upperbounds_withdecay}
	Let $p\geq 1$, $\delta\in(0,1)$ and $\lambda_1\!\geq\!\lambda_2\!\geq\!...$ be the eigenvalues of
	$\A^\top\A$.
	% If for $i\in[n]$ and $p\geq 1$ we have $\lambda_i=\Omega(\lambda_1/i^p)$,
	% then for $k\in[n]$,
	% A volume sampled set $S$ of size $k$ satisfies:
	If $S$ is  a volume sampled set of size $k$, then
	\begin{align*}
	\frac{\E\big[\Er_\A(S)\big]}{\opt} =
	\begin{cases}
	1+O(p),&\text{if
	}\lambda_i=\Omega(\lambda_1/i^p),\\
	1+O(\delta k), &\text{if }\lambda_i=\Omega\big(\lambda_1\,(1-\delta)^i\big).
	\end{cases}
	\end{align*}
\end{theorem}

\begin{proof}
	(1) Exponential decay.

	 The stable rank of the matrix residual after removing top $s$ eigenvalues:
	\[ \sr_s(\A) = \sum_{j>s} \lambda_j/\lambda_{s+1} =  \frac{\lambda_{s+1}(1-(1-\delta)^{n-s})/\delta }{\lambda_{s+1}} =  \frac{1-(1-\delta)^{n-s}}{\delta}.
	\]

		We present the proof by considering two subcases separately : when $n \geq k +  \frac{\ln 2}{\delta}$ and $n < k+  \frac{\ln 2}{\delta}$.

	\textbf{Case 1(a):}  $n \geq k +  \frac{\ln 2}{\delta}$. From the assumption,

	\begin{align*}
	& k \leq n -  \frac{\ln 2}{\delta}  \\
	\implies & s \leq n - \frac{\ln 2}{\delta}  \\
	\implies  & s \leq n - \frac{\ln 2}{\ln \frac{1}{1-\delta}} \\
	\iff & 1 - (1-\delta)^{n-s} \geq \frac{1}{2}   \\
	\implies & \sr_s(\K) \geq \frac{1}{2\delta},
	\end{align*}

	where the third inequality follows because $1-\frac{1}{x} \leq \ln x$ with $x = \delta$.

	We'll use $u=k-s$ for simplicity. From Theorem~\ref{t:upper}, using $\sr_s \geq \frac{1}{2\delta}$ we have the following upper bound:
%	\[ \frac{\E[\Er(S)]}{\opt} \leq (1+2\epsilon)^2\Big(\,1+\frac{s}{k}\,\Big)\sqrt{1+\tfrac{2\delta k}{1-\delta k}\,} \leq (1+2\epsilon)^2\Big(\,1+\frac{s}{k}\,\Big) \left( 1+ \frac{2\delta k}{1-2\delta k}\right)
%	\].

	\[ \Phi_s(k) \leq \frac{k}{u} \left(1+ \frac{2\delta u}{1 - 2\delta u}\right)  = \frac{k}{u}\cdot \frac{1}{1-2\delta u}\].

	Since $\Phi_s$ holds for any $u$, we can optimize the RHS over $u$. RHS is minimized for $\hat{u}=\frac{1}{4\delta}$. Thus,  $k  \leq \hat{u}$, increasing $u$ only tightens the upper bound. So, in this case we can set $s=0 \implies u = k = \hat{u}$ to get $\Phi_s(k) \leq \frac{1}{1- 2\delta k}\leq 2$ . On the other hand, if $k > \hat{u}$, we can set $u=\hat{u}$, which implies $\Phi_s(k) \leq 8\delta k$.

	\bigskip
	\textbf{Case 1(b)}:$k > n - \frac{\ln 2}{\delta}$. We make use of Theorem~\ref{t:upper_conditionnum} for the case when $k$ becomes too close to $n$. The approximation guarantee has:

	\[\Phi_s(k) = \frac{\lambda_{s+1}}{\lambda_n} \frac{k}{k-s}\],

	which holds for all $s <k$. For our bound, we choose $s = k- \frac{\ln 2}{\delta}$. This implies $n-s < \frac{2\ln 2}{\delta}$. Consider,

	\begin{align*}
	& \frac{\lambda_{s+1}}{\lambda_n}  = \frac{1}{(1-\delta)^{n-s}} \leq \frac{1}{(1-\delta)^{(\ln 4)/\delta}} = \left[(1-\delta)^{-\frac{1}{\delta}}\right] ^ {\ln 4} \leq e^{- \ln 4} = \frac{1}{4}. 
	\end{align*}

	Finally, we have,

	\[\frac{k}{k-s} = \frac{\delta k}{\ln 2}\].

	Plugging in we get,

	\[\Phi_s(k) \leq \frac{\delta k}{4 \ln 2} \]

	(2) Polynomial decay. We provide the proof by splitting in two cases.

	\textbf{Case 2(a)}: $\left(\frac{k+1}{n}\right)^{p-1} \leq \frac{1}{2}$

	 We know that $(s+1)^{1-p} \leq (p-1)\sum_{i\geq s}\frac{1}{i^p} \leq s^{1-p} $ [CITE].

	For calculating the stable rank:

	\[ \sr_s(\A) = \frac{\sum_{i=s+1}^n \lambda_i}{\lambda_{s+1}} =\frac1{\lambda_{s+1}} \left( \sum_{i > s} \lambda_i - \sum_{i > n} \lambda_i \right).\]

	Consider,

	\[ \frac{\sum_{i>s} \lambda_i}{\lambda_{s+1}} \geq \frac{(s+2)^{1-p}}{(p-1) (s+1)^{-p}} = \frac{s+2}{p-1} \left(\frac{s+1}{s+2}\right)^p =  \frac{s+2}{p-1} \left(1 - \frac1{s+2}\right)^p  \geq \frac{s+2}{p-1}(1 - \frac{p}{s+2}) \geq \frac{s+1}{p-1} -1 .\]

	Similarly for the second part,

	\[\frac{\sum_{i>n} \lambda_i}{\lambda_{s+1}} \leq \frac{(n+1)^{1-p}}{(p-1) (s+1)^{-p}} = \frac{s+1}{p-1}\left(\frac{s+1}{n+1}\right)^{p-1} =\frac{s+1}{p-1}\left(\frac{s+1}{n+1}\right)^{p}\frac{n+1}{s+1} \leq \frac{1}{2} \frac{s+1}{p-1} . \]

	Combining, we can get the following lower bound on the stable rank:

	\[\sr_s(\A) \geq \frac{1}{2} \frac{s+1}{p-1} -1  \]

	Recall we use $u = k-s$. Proceeding as before, we can call upon Theorem~\ref{t:upper} to get, 
	\[ \Phi_s(k) \leq \frac{k}{u}\sqrt{1+\frac{2u}{ \sr_s-u }\,} \leq  \frac{k}{u} + \frac{k}{  \frac{1}{2} \frac{s+1}{p-1} - 1 - u } = \frac{k}{u} + \frac{(2p-2)k}{   s+1 -2 p +2 - (2p-2)u } \leq \frac{k}{u} + \frac{(2p-1)k}{   k+3 -2 p  - (2p-1)u } \]

	Optimizing over $u$, we see that the minima is reached for $u = \hat{u} = \frac{k+3-2p}{2(2p-1)}$ which achieves the value $ \frac{4(2p-1)k}{   k+3 -2 p}$ which is upper bounded by $\frac{8pk}{k-2p}$.

	\bigskip 
	\textbf{Case 2(b)}: $\left(\frac{k+1}{n}\right)^{p-1} > \frac{1}{2}$

	From Theorem~\ref{t:upper_conditionnum}, we know that the approximation ratio is upper bounded by $\frac{\lambda_{s+1}}{\lambda_n}.\frac{k}{k-s} = \frac{n^p}{(s+1)^p}\frac{k}{k-s} $. Consider,

	\begin{align*}
	&\frac{n^p}{(s+1)^p}\frac{k}{k-s} = \frac{n^p}{(k+1)^p} \frac{{(k+1)^p}}{{(s+1)^p}} \frac{k}{k-s} = \left(\frac{n}{k+1}\right)^{p-1} \frac{k+1}{n}\frac{{(k+1)^p}}{{(s+1)^p}} \frac{k}{k-s}\\ 
	& \leq 2 \left(\frac{k+1}{s+1}\right)^p \frac{k}{k-s},
	\end{align*}

	which holds true for all $s \leq k$. We set $s$ so that $\left(\frac{k+1}{s+1}\right)^p = 2p$, which gives the upper bound of $4p\frac{p^{1/p}2^{1/p}}{p^{1/p}2^{1/p}-1} $. Using the fact that $1 < p^{1/p}<2$, we can further refine the upper bound as $8p\frac{2^{1/p}}{2^{1/p}-1}$ which is $O(p)$.

	\emph{Note:} In proofs of both the polynomial decay and eponential decay, we have ignored the importance of $\epsilon$. In principle, our chosen $s$ may get too close to $k$ that may require adjusting $\epsilon$, but this will only change the bounds by a constant factor.

 \end{proof}

Theorem~\ref{t:upperbounds_withdecay} is directly applicable for many well-known kernels. For example, $t+\frac{1}{2}$-Mat\'ern kernel has polynomially decaying eigenvalues with $p=2t+2$~\cite{Ritter1995Mattern}, and thus our obtained bounds are applicable for Mat\'ern kernels. Similarly, the squared exponential kernel has exponentially decaying eigenvalues~\cite{Zhu1997GaussianRA} with the decay factor $\delta$ governed by the hyperparameters.

\section{General lower bound}

\begin{lemma}\label{l:lowerbounddouble}
	Fix $\epsilon\in(0,1)$.  Consider any $k+1$ vectors $\x_i$ ,
        with $\|x_i\|^2=a \,\forall i,$ spanning a 
	$k$-dimensional subspace with their centroid at $\zero$. Similarly, consider another set of $s+1$ vectors $\y_j$, with $\|y_j\|^2=b
        \,\forall j,$ lying in the orthogonal complement of the span
        of $\x_i$ and spanning $s$-dimensional subspace, again with
        centroid at $\zero$. 
	Let $\v_1$ and $\v_2$
	be unit vectors that are orthogonal to each other and to all $\x_i$ and $\y_j$.  We construct
	matrix $\A$ with $n=k+s+2$ columns by stacking
	$\tilde\x_i:=\x_i + \alpha \v_1$ and $\tilde \y_j
	:=\y_j+\alpha \v_2$. For sufficiently small
	$\alpha>0$ and if $b < \min\{a, \frac{ \frac{\epsilon}{2} (k+1) \alpha^2}{(s+1) (1-\epsilon/2)} \} $, for any subset $S\subseteq[n]$ of size $k$,
	\begin{align*}
	\Er_{\A}(S)\geq (1-\epsilon)(k+1) \| \A -\A_{k}\|^2_F,
	\end{align*}
	and for any subset $S_2\subseteq[n]$ of size $k+s+1$,
		\begin{align*}
	\Er_{\A}(S_2)\geq (1-\epsilon)(s+1) \| \A -\A_{k+s+1}\|^2_F,
	\end{align*}
\end{lemma}
\begin{proof}
(a)	Say $\P^\star$ is the span of $\{\x_1, \x_2, \ldots \x_{k+1}\}$. Say $\Abh$ is the matrix obtained by stacking all the $\x_i$ and $\y_j$. Note that $\one$ is an eigenvector of $\Abh^\top\Abh$ with eigenvalue $0$. Further,

\begin{align*}
\A^\top \A =	\Abh^\top\Abh  + \alpha^2 \one \one^\top.
\end{align*}

Thus, $\A^\top\A$ has the same eigenvectors and eigenvalues as $\Abh^\top\Abh$ except the eigenvalue $0$ has been replaced with $\alpha^2$. Hence, if $\alpha^2 < \lambda_k$, where $\lambda_k$ is the second smallest eigenvalue of $\Abh^\top\Abh$ restricted to the set of vectors $\x_i$, and since $b<a$, then
	\begin{align*}
	E_k = \|\A- \A_k\|_F^2  = \| \A - \P^\star \A \|_F^2 = (k+1)\alpha^2 + (s+1) b
		\end{align*}

	Since the centroid of $\{\tilde\x_1,\ldots, \tilde\x_{k+1}\}$ is $\alpha \v_1$, we can write 
	$\tilde\x_{k+1} = (k+1)\alpha\v_1 - \sum_{i\leq k} \tilde\x_i$
	For selecting the set $S\subset [n]$ of size $k$, since $a>b$, we can assume without loss of generality that $S = \{\x_1,\ldots,\x_k\}$.
	\begin{align*}
	\Er_{\A}(S) &= \| \tilde\x_{k+1}- \P_S \tilde\x_{k+1}  \|_F^2
                      +  \sum_{i=1}^{s+1}\| \tilde\y_i - \P_S
                      \tilde\y_i \|_F^2\\ 
	&=\|(k+1)\alpha\v_1 - \sum_{i\leq k}\tilde\x_i- \P_S
   ((k+1) \alpha\v_1 - \sum_{i\leq k}\tilde\x_i )\|_F^2 + (s+1)b
          \\ 
	&= (k+1)^2\alpha^2 \|\v_1 - \P_S\v_1\|^2_F + (k_2+1)b \\
	& \geq (k+1) (E_k - (s+1)b) \|\v_1 - \P_S\v_1\|^2_F + (s+1)b \\
	& = (k+1) E_k \|\v_1 - \P_S\v_1\|^2_F + (s+1)b (1 - (k+1)\|\v_1 - \P_S\v_1\|^2_F ) \\
	& \geq (k+1) E_k \|\v_1 - \P_S\v_1\|^2_F - (s+1)kb 
	\end{align*}	
	Note that $\lim_{\alpha\rightarrow 0}\P_S\v_1 = \zero$ because $\v_1$
	is orthogonal to the subspace spanned by $\{\x_i:i\in S\}$, so there is
	$\alpha_S>0$ such that for any $\alpha\leq\alpha_S$ we have
        $\|\v-\P_S\v\|^2\geq 1-\frac\epsilon2$.

Further, given $\epsilon$ and $\alpha$ selected above, the assumption on $b$ ensures that the following is satisfied: 
\begin{align*}
	&(s+1)b (1-\epsilon/2) \leq \frac{\epsilon}{2}(k+1) \alpha^2 \\
	&\implies (s+1)b \leq \epsilon/2 E_k \\
	& \implies (s+1)bk \leq \epsilon (k+1)/2 E_k,
\end{align*}

which completes the desired result.

%        \michal{Now, we need to ensure that $(s+1)kb\leq
%          \frac\epsilon2(k+1)E_k$. This will require an additional
%          assumption upper-bounding $b$ and also a lower bound on $E_k$.}

 (b) Let $ t = k+s+1$. Proceeding as in the proof of part (a), let $\P^\star$ be the span of $\{\x_1,\ldots,\x_{k+1}, \y_1, \ldots,\y_s\}$
 \begin{align*}
 E_t \leq (s+1)\alpha^2
 \end{align*}

 Further, we can write $\tilde\y_{s+1} = (s + 1)\v_2 -\sum_{j\leq s } \tilde \y_j$.

 	\begin{align*}
 \Er_{\A}(S_2) &= \| \tilde\y_{k+1}- \P_{S_2} \tilde\y_{k+1}  \|_F^2 \\
 & = (s+1)\alpha^2\| \v_2 - \P_{S_2} \v_2  \|_F^2 \\
 & \geq (s + 1) E_k\| \v_2 - \P_{S_2} \v_2  \|_F^2,
 \end{align*}

 and the result follows as in part(a).

\end{proof}

\begin{lemma}\label{l:lowerbounddouble2}
	Fix $\epsilon\in(0,1)$.  Consider $r$ sets of vectors. For $i\in [r]$ $i^\text{th}$ set $\Gamma_i$ has $k_i+1$ vectors $\x^{(i)}_j$ where $j\in[k_i+1]$
	with $\|x_j^{(i)}\|^2=a_i$ spanning a 
	$k_i$-dimensional subspace with their centroid at $\zero$. 
	Let $\v_i, i \in [r]$ 
	be unit vectors that are orthogonal to each other and to all $\x^{(i)}_j$. We construct
	matrix $\A$ with $n= r + \sum_i k_i$ columns by stacking
	$\tilde\x^{(i)}_j:=\x^{(i)}_j + \alpha \v_i$. For sufficiently small
	$\alpha>0$ and if $\forall j,\; a_{j+1} \leq \min\left\{ a_j, \frac{\frac{\epsilon}{2} (k_j+1) \alpha^2 }{(1-\frac{\epsilon}{2}) (k_{j+1}+1)r} \right\}$ for a given $c \in [r]$, any subset $S\subseteq[n]$ of size $k = \sum_{i\leq c}k_i +c-1$,
	\begin{align*}
	\Er_{\A}(S)\geq (1-\epsilon)(k_c+1) \| \A -\A_{k}\|^2_F.
	\end{align*}
\end{lemma}
\begin{proof}
	(a) Say $\P^\star$ is the span of $\{\x^{(1)}_1, \x^{(2)}_2, \ldots \x^{(1)}_{k_1+1}, \x^{(2)}_{1}, \ldots, \x^{(c)}_{k_c} \}$. Say $\Abh$ is the matrix obtained by stacking all the $\x^{(i)}_j$. Note that $\one$ is an eigenvector of $\Abh^\top\Abh$ with eigenvalue $0$. Further,

	\begin{align*}
	\A^\top \A =	\Abh^\top\Abh  + \alpha^2 \one \one^\top.
	\end{align*}

	Thus, $\A^\top\A$ has the same eigenvectors and eigenvalues as $\Abh^\top\Abh$ except the eigenvalue $0$ has been replaced with $\alpha^2$. Hence, if $\alpha^2 < \lambda$, where $\lambda$ is the smallest non-zero eigenvalue of $\Abh^\top\Abh$, and since $a_{i+1}<a_i \;\forall i$, then
	\begin{align*}
	E_k = \|\A- \A_k\|_F^2  = \| \A - \P^\star \A \|_F^2 =  (k_c+1)\alpha^2 + \sum_{i>c}(k_i+1) a_i =  (k_c+1)\alpha^2 + \phi_c, 
	\end{align*}

	where $\phi_c =\sum_{i>c}(k_i+1) a_i $ for ease of notation.

	Since the centroid of $\{\tilde\x^{(c)}_1,\ldots, \tilde\x^{(c)}_{k_c+1}\}$ is $\alpha \v_c$, we can write 
	$\tilde\x^{(c)}_{k_c+1} = (k_c+1)\alpha\v_c - \sum_{j\leq k_c} \tilde\x^{(c)}_j$.
	For selecting the set $S\subset [n]$ of size $k$, since $a_i>a_{i+1}$, we can assume without loss of generality that $S = \cup_{i<c} \Gamma_i \cup \{\x^{(1)}_1,\ldots,\x^{(c)}_{k_c}\}$.
	\begin{align*}
	\Er_{\A}(S) &= \| \tilde\x^{(c)}_{k_c+1}- \P_S \tilde\x^{(c)}_{k_c+1}  \|_F^2
	+  \sum_{i>c}\sum_{j=1}^{k_i+1}\| \tilde\x^{(i)}_j - \P_S
	\tilde\x^{(i)}_j \|_F^2\\ 
	&=\|(k_c+1)\alpha\v_c - \sum_{j\leq k_c}\tilde\x^{(c)}_j- \P_S
	((k+1) \alpha\v_c - \sum_{j\leq k_c}\tilde\x^{(c)}_j )\|_F^2 + \sum_{i>c}(k_i+1)a_i
	\\ 
	&= (k_c+1)^2\alpha^2 \|\v_c - \P_S\v_c\|^2_F + \sum_{i>c}(k_i+1)a_i \\
	& = (k_c+1) (E_k - \phi_c) \|\v_c - \P_S\v_c\|^2_F + \phi_c \\
	& = (k_c+1) E_k \|\v_c - \P_S\v_c\|^2_F + \phi_c (1 - (k_c+1)\|\v_c - \P_S\v_c\|^2_F ) \\
	& \geq (k_c+1) E_k \|\v_c - \P_S\v_c\|^2_F - k_c \phi_c 
	\end{align*}	
	Note that $\lim_{\alpha\rightarrow 0}\P_S\v_1 = \zero$ because $\v_1$
	is orthogonal to the subspace spanned by $\{\x_i:i\in S\}$, so there is
	$\alpha_S>0$ such that for any $\alpha\leq\alpha_S$ we have
	$\|\v-\P_S\v\|^2\geq 1-\frac\epsilon2$.

	Further, given $\epsilon$ and $\alpha$ selected above, the assumption on decreasing lengths $a_j$ for an arbitrary $j \in [r]$ is: 
	\begin{align*}
	&  a_{j+1} \leq \frac{\frac{\epsilon}{2} (k_j+1) \alpha^2 }{(1-\frac{\epsilon}{2}) (k_{j+1}+1)r}\\
	\implies & (1-\frac{\epsilon}{2}) (k_{j+1}+1) a_{j+1} \leq \frac{\frac{\epsilon}{2} (k_j+1) \alpha^2 }{r}\\
	\implies & \sum_{i>c}(1-\frac{\epsilon}{2}) (k_{i}+1) a_{i} \leq (r-c)\frac{\frac{\epsilon}{2} (k_c+1) \alpha^2 }{r} \leq \frac{\epsilon}{2} (k_c+1) \alpha^2\\
	\implies & (1-\frac{\epsilon}{2})\phi_c  \leq  \frac{\epsilon}{2} (k_c+1) \alpha^2 \\
	\implies & \phi_c \leq \frac{\epsilon}{2} E_k \\
	\implies & k_c\phi_c \leq \frac{\epsilon}{2} (k_c+1) E_k,
	\end{align*}

	which completes the desired result.

	%        \michal{Now, we need to ensure that $(s+1)kb\leq
	%          \frac\epsilon2(k+1)E_k$. This will require an additional
	%          assumption upper-bounding $b$ and also a lower bound on $E_k$.}

\end{proof}

\section{Old Lower bounds}

 We discuss lower bounds of two forms -- algorithmic lower bounds specific to the DPPs and algorithm indepedent lower bounds. For the former case, we obtain lower bounds for $k<r$ which are tight upto a factor of $\sqrt{\epsilon}$. For the latter case, we discuss lower bounds in two regimes $k\leq r$ and $k>r$. We show that for the case $k>r$, the known upper bound of multiplicative factor of $(k+1)$ are tight. On the other hand, for the case of $k\leq r$, however, we obtain lower bounds that are loose compared to what our newly obtained upper bounds entail. Removing this gap between the upper and lower bounds should be an interesting future direction for research.

 \subsection{Algorithmic lower bounds for DPPs}

  \begin{lemma}
  	\label{l:algolowerbound}
  	For any $\epsilon>0$, $\exists$ a matrix $\A$ with stable rank $r$ such that using DPP sampling with $E(|S|) = k$ with $k \leq \frac{\epsilon}{1+\epsilon} r$, we can lower bound $\E(\Er_{\A}(S)) \geq \left(\frac1{2} + \sqrt{\epsilon}\right) \|\A -\A_k \|_F^2 $.
  \end{lemma}
  \begin{proof}
  	Consider a matrix $\A \in \R^{m\times n}$ so that $\K = \A^\top \A$ has $k$ eigenvalues as $1$, while the remaining eigenvalues are all $\delta \in (0,1)$.

  	 Say, $n$ is large enough so that $r=(1+\rho) k$ for some $\rho > 0$.  To calculate $r$,

  	\begin{align*}
  	r = \sum_i \frac {\lambda_i}{ \lambda_1}= k + \delta(d-k).
  	\end{align*}

  	Since  $r=(1+\rho) k$, we get $\delta(d-k) = \rho k$. Furthermore, say we want $\beta$ multiplicative bound i.e. $\frac{\E(\Er_{\A}(S))}{\| \A - \A_k\|_F^2} = \beta$, which gives,

  	\begin{align*}
  	&\frac{\alpha k }{\delta(d-k) } = \beta \\ & \implies \alpha k = \beta \rho k \implies \alpha = \beta \rho.
  	\end{align*}

  	For $k$,

  	\begin{align*}
  	& k = \sum_i \frac{\lambda_i}{\lambda_i + \alpha} 
  	= \frac{k}{1+\alpha} + \frac{\delta(d-k)}{\delta+\alpha} \\
  	& \implies k =  \frac{k}{1+\alpha} + \frac{\rho k}{\delta+\alpha}\\
  	&\implies 1 =  \frac{1}{1+\alpha} + \frac{\rho}{\delta+\alpha} \\
  	&\implies \delta = \rho + \frac{\rho}{\alpha} - \alpha = \rho + \frac1 {\beta} - \beta \rho.
  	\end{align*}

  	For the bound, we need $0 < \delta < 1$. While $\delta < 1$ is trivially satisfied, for $\delta > 0$, we need $\rho < \frac{1}{\beta (\beta - 1)}$.

  	With $\rho = \frac1 {\epsilon}$, we have $k = \frac{\epsilon}{1+ \epsilon}r$. And with $\beta =1+\frac{\epsilon}{1+ \epsilon} $, the condition $\rho < \frac{1}{\beta (\beta - 1)}$ is satisfied. Hence, we get for any $\epsilon>0$, with $k = \frac{\epsilon}{1+ \epsilon}r $, we are ensured that $\frac{E(\Er_{\A}(S))}{ \| \A - \A_k \|_F^2} \geq \left( 1 + \frac{\epsilon}{1+\epsilon} \right)$.

  	We can also write the condition as $\beta (\beta -1) -\epsilon < 0$, which implies
  	\[\beta \in \left(\frac{( 1- \sqrt(1+ 4 \epsilon)}{2}, \frac{( 1+ \sqrt(1+ 4 \epsilon)}{2} \right)\]. $\beta = (1/2 + \sqrt{\epsilon})$ satisfies this condition.

  \end{proof}

 \subsection{Algorithm independent lower bounds} 
  \subsubsection{Case $k\leq r$}
\begin{theorem}\label{t:lower-small-k}
  For any $\epsilon\in(0,k^2)$ and $n,m,k,r\in\N$ such
  that $k\leq\frac{\epsilon}{1+\epsilon}\,r\leq\min\{n,m\}$, there is
  an $m\times n$ matrix $\X$ with stable rank  
  $r$ such that for any subset $S$ of size $k$,
  \begin{align*}
    \Er_\X(S) \geq \bigg(1+\epsilon\cdot \frac{k-1}{2k^2}\bigg)\cdot \|\X-\X_k\|_F^2.
  \end{align*}
\end{theorem}

\begin{proof}
	Let $\x_1,...,\x_{k+1}$ be vectors forming a regular simplex centered
	at $\zero$ which lies inside of the subspace spanned by the first $k$
	coordinates. Denote $a=\|\x_i\|$ as the length of each vector. We
	shift those vectors in the $(k+1)$-st coordinate by $\alpha$, getting
	$\xbt_i=\x_i+\alpha\e_{k+1}$. Then, we add further vectors
	$\x_{k+2},...,\x_n$ for some $n\gg k$ so that $\x_i=\delta\e_{i}$ for
	$i>k+1$. Denote the entire $n\times d$ matrix as $\X$. The optimum
	rank $k$ error is upper bounded as follows: 
	\begin{align*}
	\|\X - \X_k\|_F^2 \leq (k+1)\alpha^2 + (n-k-1)\delta^2.
	\end{align*}
	The top $k$ eigenvalues of $\X^\top\X$ associated with the simplex are all
	$\lambda_i=\frac{k+1}{k}\,a^2$. This means that,
	assuming that $\delta^2\leq(k+1)\alpha^2\leq \frac{k+1}{k}\,a^2$ (to
	achieve the correct ordering of the eigenvalues), the
	stable rank of $\X$ is given by:
	\begin{align*}
	r = \frac1{\lambda_1}\sum_{i=1}^d\lambda_i = k +
	k\,\frac{\alpha^2}{a^2} + \frac{(n-k-1)\delta^2}{\frac{k+1}k\,a^2}.
	\end{align*}
	Finally, for any subset $S\subseteq[k+1]$, the loss is:
	\begin{align*}
	L(S) &= (k+1)^2\frac{(\frac ak)^2\alpha^2}{(\frac ak)^2+\alpha^2} +
	(n-k-1)\delta^2\\
	&=(k+1)^2\frac{a^2\alpha^2}{a^2+k^2\alpha^2} +
	(n-k-1)\delta^2.
	\end{align*}
	Suppose that $\alpha^2=a^2/k^2$ and $(n-k-1)\delta^2 = (k+1)(\rho a^2-\alpha^2)$ for some
	$\rho\in[\frac1{k^2},1]$. It follows that $r=(1+\rho)k$ and
	$\|\X - \X_k\|_F^2=\rho(k+1)a^2$. Also we get:
	\begin{align*}
	\Er_{\X}(S)&=  (k+1)^2 \frac{a^2}{2k^2} + (k+1)\Big(\rho-\frac1{k^2}\Big) a^2\\
	&=\rho(k+1)a^2 + a^2\frac{k+1}{k^2}\Big(\frac{k+1}{2} - 1\Big)\\
	&=\|\X - \X_k\|_F^2\cdot\Big(1 + \frac{k-1}{2\rho k^2}\Big).
	\end{align*}
	By design, note that $\epsilon = \frac{1}{\rho}$, then we obtain the desired result. 
	\end{proof}

\subsubsection{Case $k >r$}
To obtain the lower bounds for the regime where the selected subset is larger than the stable rank, we first extend the lower bound result of
\cite{pca-volume-sampling}. Their construction uses a set of vectors that
form a regular simplex centered at $\zero$ and than slightly
shifts it. We show that in fact the regularity of the simplex is not
necessary for their proof to work.
\begin{lemma}\label{l:simplex}
	Fix $\epsilon\in(0,1)$.  Consider any $k+1$ vectors $\x_i$ spanning a
	$k$-dimensional subspace with their centroid at $\zero$ and let $\v$
	be a unit vector orthogonal to that subspace. For any sufficiently small
	$\alpha>0$, if we construct
	matrix $\X$ with $k+1$ columns so that the $i$th column vector is $\x_i^\top+\alpha\v^\top$, 
	then for any subset $S\subseteq[k+1]$ of size $k$,
	\begin{align*}
	\Er_{\X}(S)\geq (1-\epsilon)(k+1) \| \X -\X_k\|^2_F.
	\end{align*}
\end{lemma}
\begin{proof}
	Let $\P^*$ be the projection onto the span of
	$\{\x_1,\dots,\x_{k+1}\}$. Then we have
	\begin{align*}
	 \| \X -\X_k\|^2_F \leq \|\X - \P^*\X\|_F^2 = (k+1)\cdot\alpha^2.
	\end{align*}
	We will use the shorthand $\xbt_i=\x_i+\alpha\v$. Consider any subset
	of $k$ columns from $\X$. W.l.o.g. suppose 
	$S=\{1,\dots,k\}$. Since the centroid of all the columns is $\alpha\v$,
	we can rewrite the last vector as 
	\begin{align*}
	\xbt_{k+1} = (k+1)\alpha\v - \sum_{i\in S}\xbt_i.
	\end{align*}
	Since for any $i\in S$ we have $\P_S\xbt_i=\xbt_i$, it follows that:
	\begin{align*}
	\Er_{\X} &= \|\xbt_{k+1} - \P_S\xbt_{k+1}\|^2 \\
	&=\bigg\| (k+1)\alpha\v - \sum_{i\in S}\xbt_i - \Big(\P_S (k+1)\alpha\v -
	\sum_{i\in S}\xbt_i\Big)\bigg\|^2\\
	&=(k+1)^2\alpha^2\|\v-\P_S\v\|^2.
	\end{align*}
	Note that $\lim_{\alpha\rightarrow 0}\P_S\v = \zero$ because $\v$
	is orthogonal to the subspace spanned by $\{\x_i:i\in S\}$, so there is
	$\alpha_S>0$ such that for any $\alpha\leq\alpha_S$ we have $\|\v-\P_S\v\|^2\geq 1-\epsilon$. This shows that
	$L(S)\geq (1-\epsilon)(k+1)E_k^*$ for $\alpha=\alpha_S$. Taking
	$\alpha = \min_{S:|S|=k}\alpha_S$ completes the proof.
\end{proof}
We next obtain our main hardness result which states that the lower bound
factor $k+1$ of \cite{pca-volume-sampling} cannot be improved by constraining to matrices of a
particular stable rank $r$ as long as $k>r$. This stands in contrast
to our upper bound which showed that when $k<r$, then the factor $k+1$
can be significantly improved.
\begin{theorem}
	\label{t:lowerboundrleqk}
	For any $\epsilon\in(0,1)$ and $n,d,k,r\in\N$ such that  $0<r< k<\min\{n,d\}$, there is a matrix
	$\X\in\R^{n\times d}$ with stable rank in $[r,r+1]$ such that for any subset
	$S$ of size $k$
	\begin{align*}
	\Er_{\X}(S)\geq (1-\epsilon)(k+1) \|\X - \X_k \|_F^2.
	\end{align*}
\end{theorem}
\begin{proof}
	We will construct a set of vectors by gradually modifying and augmenting it while
	preserving two invariants: (1) that the stable rank is in $[r,r+\frac23]$
	and (2) the centroid of the vectors is  $\zero$.

	First, construct matrix $\X_{(r+1)}$  from $r+1$ column vectors
	$\x_1^\top,...,\x_{r+1}^\top$ that form 
	a regular simplex centered at $\zero$, scaled so that the $r$ non-zero
	singular values of $\X_{(r+1)}$ are all equal 1. Since all the non-zero
	singular values of the matrix are equal, the stable rank
	is $r$. Next, we iteratively add vectors one by one (and modify the
	existing ones) until there is $k+1$ in the set. Let $\x_{1},...,\x_{t-1}$ and
	$\xbt_1,...,\xbt_t$ denote the 
	sets of vectors \emph{before} and \emph{after} adding the $t$th vector,
	with corresponding matrices $\X_{(t-1)}$ and $\X_{(t)}$. Also, let
	$\tilde r_{t-1}$ denote the stable rank of $\X_{(t-1)}$.
	Adding the $i$th vector proceeds as follows:
	\begin{enumerate}
		\item shift all $t-1$ vectors along a normal direction $\v_t$
		(orthogonal to their span) by some $\delta_t$,
		\item let the new vector be $\xbt_{t}=- (t-1)\,\delta_t\v_t$.
	\end{enumerate}
	After the transformation and adding the $i$th vector, the centroid
	stays at $\zero$, and moreover:
	\begin{align*}
	\X_{(t)}^\top\X_{(t)} &=
	\sum_{i=1}^{t-1}(\x_i+\delta_t\v_t)(\x_i+\delta_t\v_t)^\top +
	(t-1)^2\delta_t^2\v_t\v_t^\top\\
	&=\sum_{i=1}^{t-1}\x_i\x_i^\top +
	\delta_t\v_t\Big(\sum_{i=1}^{t-1}\x_i^\top\Big) +
	\Big(\sum_{i=1}^{t-1}\x_i\Big)\delta_t\v_t^\top +
	(t-1)\delta_t^2\v_t\v_t^\top+        (t-1)^2\delta_t^2\v_t\v_t^\top\\
	&=\X_{(t-1)}^\top\X_{(t-1)}+((t-1)^2+t-1)\delta_t^2\v_t\v_t^\top.
	\end{align*}

q

	Therefore, since $\v_i$ is orthogonal to the span of $\{\x_{1},...,\x_{t-1}\}$,
	it becomes a new eigenvector of $\X_{(t)}^\top\X_{(t)}$, while the old
	eigenvectors from $\X_{(t-1)}^\top\X_{(t-1)}$ remain the same in the
	new matrix. Thus, as long as $\delta_t$ is sufficiently small (so that
	the new eigenvalue does not become the largest), we have:
	\begin{align*}
	\tilde r_{t} = \tilde r_{t-1} + \|\X_{(t)}\|_F^2
	-\|\X_{(t-1)}\|_F^2=\tilde r_{t-1} + \delta_t^2((t-1)^2+t-1) = \tilde r_{t-1} + \delta_t^2((t-1)^2+t-1).
	\end{align*}
	To make sure that stable rank remains less than $r+\frac23$, we
	assume that $\delta_t\leq\frac1{t2^{t}}$ and bound the stable rank of the final set:
	\begin{align*}
	\tilde r_{k+1} = r + \sum_{t=r+2}^{k+1} \delta_t^2 (t^2-t) \leq r +
	\sum_{t=1}^{k+1}\frac2{4^t}\leq r + \frac23.
	\end{align*}
	Now, it remains to apply Lemma \ref{l:simplex} to the set
	of $k+1$ vectors, obtaining matrix $\Xt_{(k+1)}$ that satisfies the
	desired lower bound. When choosing
	$\alpha$, we have to again ensure that it 
	is sufficiently small so that stable rank does not change by
	more than $\frac13$. Naturally, this can be done just as we did it
	in the construction of the set. We construct the final matrix $\X$
	from $\Xt_{(k+1)}$ by padding with zeros to get the shape right,
	which does not affect the loss $\Er_{\X}(\cdot)$ or the stable rank. 
\end{proof}

  \bibliographystyle{sty/icml2020/icml2020}
  \bibliography{pap}

  \end{document}

### 2022-08-23 design/nystrom-main-ijcai.tex
Commit original repos

%%%% ijcai21.tex
%
\typeout{Improved guarantees and a multiple-descent curve for 
  Column Subset Selection 
  and the Nystr\"om method$^*$}

% These are the instructions for authors for IJCAI-21.

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in
% The file ijcai21.sty is NOT the same than previous years'
\usepackage{sty/ijcai2021/ijcai21}

% Use the postscript times font!
\usepackage{natbib}
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{xcolor}
\usepackage{xfrac}
\usepackage{amssymb}

\urlstyle{same}

% the following package is optional:
%\usepackage{latexsym}

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}

% Following comment is from ijcai97-submit.tex:
% The preparation of these files was supported by Schlumberger Palo Alto
% Research, AT\&T Bell Laboratories, and Morgan Kaufmann Publishers.
% Shirley Jowell, of Morgan Kaufmann Publishers, and Peter F.
% Patel-Schneider, of AT\&T Bell Laboratories collaborated on their
% preparation.

% These instructions can be modified and used in other conferences as long
% as credit to the authors and supporting agencies is retained, this notice
% is not changed, and further modification or reuse is not restricted.
% Neither Shirley Jowell nor Peter F. Patel-Schneider can be listed as
% contacts for providing assistance without their prior permission.

% To use for other conferences, change references to files and the
% conference appropriate and use other authors, contacts, publishers, and
% organizations.
% Also change the deadline and address for returning papers and the length and
% page charge instructions.
% Put where the files are available in the appropriate places.

%PDF Info Is REQUIRED.
\pdfinfo{
/TemplateVersion (IJCAI.2021.0)
}

\title{Improved Guarantees and a Multiple-descent Curve\\
  for Column Subset Selection 
  and the Nystr\"om Method (Extended Abstract)$^*$}

% Single author syntax
%\author{
%    Zhi-Hua Zhou
%    \affiliations
%    Nanjing University
%    \emails
%    pcchair@ijcai-21.org
%}

% Multiple author syntax (remove the single-author syntax above and the \iffalse ... \fi here)
% Check the ijcai21-multiauthor.tex file for detailed instructions
%\iffalse
\author{
Micha{\l } Derezi\'{n}ski$^{1\dagger}$
\and
{Rajiv Khanna}$^1$\and
Michael W. Mahoney$^{1,2}$
\affiliations
$^1$Department of Statistics,
  University of California at Berkeley\\
$^2$International Computer Science Institute, Berkeley \\
\emails
\{mderezin, rajivak\}@berkeley.edu,
mmahoney@stat.berkeley.edu
}
%\fi
\input{shortdefs2}

\begin{document}

\maketitle

\begin{abstract}
 The Column Subset Selection Problem (CSSP) and the Nystr\"om method
are among the leading tools for constructing interpretable low-rank
approximations of large datasets by selecting a small but
representative set of features or instances. 
A fundamental question in this area is: what is the cost of this
interpretability, i.e., how well can a data subset of
size $k$ compete with the best rank $k$ approximation?
We develop techniques which exploit spectral properties of the data
matrix to obtain improved approximation guarantees which go beyond the
standard worst-case analysis.
Our approach leads to significantly better bounds for datasets with
known rates of singular value decay, e.g., polynomial or exponential
decay.
Our analysis also reveals an intriguing phenomenon: the cost of
interpretability as a function of $k$ may exhibit multiple peaks and
valleys, which we call a multiple-descent curve.
A lower bound we establish shows that this behavior is not an artifact
of our analysis, but rather it is an inherent property of the CSSP and
Nystr\"om tasks.
Finally, using the example of a radial basis function (RBF)
kernel, we show that both our improved bounds and the multiple-descent
curve can be observed on real datasets simply by varying the
RBF~parameter.
\end{abstract}
\section{Introduction}
\label{s:intro}

{\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{This is an abridged version invited to IJCAI 2021 of a
  longer paper with the same title that appeared in NeurIPS 2020 and
  received a Best Paper Award.}
  \footnotetext[2]{Corresponding author.}}

We consider the task of selecting a small but representative sample of
column vectors from a large matrix. Known as the \emph{Column Subset Selection
Problem} (CSSP), this is a well-studied combinatorial optimization task with
many applications in machine learning. %  
%\citep[e.g., feature selection, see][]{Guyon03FeatureSelection,BoutsidisMD08},
%scientific computing \citep[e.g.,][]{Chan92RankRevealingQR,Drineas08CUR} and signal processing
%\citep[e.g.,][]{Balzano10MatchedSubspace}. 
In a 
commonly studied variant of this task, we aim to minimize the
squared error of projecting all columns of the matrix onto the
subspace spanned by the chosen column subset.
\begin{definition}[CSSP]
  Given an $m\times n$ matrix $\A$, pick a set $S\subseteq\{1,...,n\}$ of 
  $k$ column indices, to minimize
  \begin{align*}
   \Er_\A(S) := \|\A - \P_S\A\|_F^2,
  \end{align*}
  where $\|\cdot\|_F$ is the Frobenius norm, $\P_S$ is the
  projection onto $\mathrm{span}\{\a_i:i\in S\}$ and $\a_i$ denotes the $i$th
  column of~$\A$.
\end{definition}
Another variant of the CSSP emerges in the kernel
setting under the name \emph{Nystr\"om method}
\citep{Williams01Nystrom,dm_kernel_JRNL,revisiting-nystrom}. We also
discuss this variant, showing how our analysis applies in this context.
Both the CSSP and the Nystr\"om method are ways of
constructing accurate low-rank approximations by using submatrices of
the target matrix. Therefore, it is natural to ask how
close we can get to the best possible rank~$k$ approximation error:
\begin{align*}
  \opt := \!\!\min_{\B:\,\rank(\B)=k} \!\!\|\A-\B\|_F^2\leq \min_{S:\,|S|=k}\Er_\A(S).
\end{align*}

While the best possible rank $k$ approximation has the lowest
approximation error, the approximated matrix $\B$ is not
{input-sparse}. As such, it is not \emph{interpretable} since a
practitioner is unable to attribute the quality of approximation to
specific physical quantities represented by the columns of the
matrix~$\A$ \citep{mahoney2009cur}. Such \emph{interpretable}
dimensionality reduction is desirable 
in many machine learning applications. 
 Our goal is to find a subset $S$ of size $k$ for which the cost of
 interpretability is small, as measured by what we call the
 \emph{approximation factor}: the ratio
between $\Er_\A(S)$ and $\opt$. Furthermore, a brute force
search requires iterating over all 
${n \choose k}$ subsets, which is prohibitively expensive, so we
would like to find our subset more efficiently.

\begin{figure}
%\begin{wrapfigure}{r}{0.45\textwidth}
 % \vspace{-5mm}
  %\fi
  \centering  
  \includegraphics[width=0.47\textwidth]{figs/nystrom/nystrom-bounds}
  \caption{Empirical study of the
    expected approximation factor $\E[\Er_\A(S)]/\opt$ for
a $k$-DPP with
    different subset sizes $|S|=k$, compared to our theory. We use a
    data matrix $\A$ 
    whose
    spectrum exhibits two sharp drops, demonstrating
multiple-descent. The lower bounds are based on Theorem
    \ref{t:lower}, whereas, as our upper bound, we plot the
    minimum over all $\Phi_s(k)$ from
    Theorem~\ref{t:upper}. Note that multiple-descent vanishes under
    smooth spectral decay, resulting in improved guarantees (see
    Theorem \ref{t:decay}).
  }
  \label{f:intro}
  \end{figure}
%  \else
%  \end{wrapfigure}\fi

In terms of worst-case analysis, 
\citet{pca-volume-sampling} gave a randomized
method 
which returns a
set $S$ of size $k$ such that: %\vspace{-4mm}
\begin{align}
  \frac{\E[\Er_\A(S)]}{\opt}\leq k+1.\label{eq:old-bound}
\end{align}
While the original algorithm was slow, 
efficient implementations have been provided since then
\citep[e.g., see][]{efficient-volume-sampling,dpp-intermediate}.
The method belongs to the family of cardinality constrained
Determinantal Point Processes \citep[DPPs, see][]{dpps-in-randnla}, and will
be denoted as $S\sim k\text{-}\DPP(\A^\top\A)$. 
The approximation  factor  $k+1$ is optimal in the worst-case,
since for any $0<k<n\leq m$ and $0<\delta<1$, an $m\times n$
matrix~$\A$ can be constructed for which $\frac{\Er_{\A}(S)}{\opt}\geq
(1-\delta)(k+1)$ for all subsets $S$ of size $k$. 
Yet it is known that, in practice, CSSP algorithms perform better
than worst-case, so the question we consider is: how can we go beyond
the usual worst-case analysis to accurately reflect what is possible
in the CSSP?

\textbf{Contributions.} We provide improved guarantees for the CSSP
approximation factor, which go beyond the worst-case analysis and 
which lead to surprising conclusions.%\vspace{-3mm}
\begin{enumerate}
\item\underline{New upper bounds}:
We develop a family of upper bounds on the CSSP approximation factor
(Theorem~\ref{t:upper}), which we call the Master Theorem
as they can be used to derive a number 
of new guarantees. In particular, we show that when the data matrix $\A$ exhibits a
known spectral decay, then \eqref{eq:old-bound} can often be
drastically improved (Theorem \ref{t:decay}).
\item \underline{New lower bound}:
Even though the worst-case upper bound in \eqref{eq:old-bound} can often
be loose, there are cases when it cannot be improved.
We give a new lower bound construction (Theorem \ref{t:lower}) showing
that there are matrices $\A$ for which multiple different subset sizes exhibit
worst-case behavior.
\item \underline{Multiple-descent curve}: 
Our upper and lower bounds
reveal that for some matrices the CSSP approximation factor can exhibit
peaks and valleys as a function of the subset size $k$ (see Figure
\ref{f:intro}).  We show that this phenomenon is an inherent property
of the CSSP  (Corollary \ref{c:multiple-descent}).
\end{enumerate}

\section{Determinantal Point Processes}
\label{s:dpp}

Since our main results rely on randomized subset selection via
determinantal point processes (DPPs), we provide a brief overview of
the relevant aspects of this class of distributions.
First introduced by \citet{Macchi1975}, a determinantal point process is a
probability distribution over subsets $S \subseteq[n]$, where we use $[n]$
to denote the set $\{1,...,n\}$. The relative probability of a subset being drawn is 
governed by a positive semidefinite (p.s.d.) matrix $\K \in \R^{n\times n}$,
as stated in the definition below, where we use $\K_{S,S}$ to denote
the $|S|\times |S|$ submatrix of $\K$ with rows and columns indexed
by $S$.
\begin{definition}\label{d:dpp}
  For an $n\times n$ p.s.d.~matrix $\K$, define $S\sim
  \DPP(\K)$ as a distribution 
  over all subsets $S\subseteq [n]$ so
  that
  \[\Pr(S)=\frac{\det(\K_{S,S})}{\det(\I+\K)}.\]
A restriction to subsets of size $k$ is denoted as $k$-$\DPP(\K)$.
\end{definition}
DPPs can be used to introduce diversity in the selected set or to
model the preference for selecting dissimilar items, where the
similarity is stated by the kernel matrix $\K$. DPPs are
commonly used in many machine learning applications where these
properties are desired, e.g.,~recommender systems~\citep{Warlop2019}, model
interpretation~\citep{kim:2016MMD}, text
and video summarization~\citep{dpp-video}, and others
\citep{dpp-ml}. For a recent survey, see \cite{dpps-in-randnla}.

Given a p.s.d.~matrix $\K \in \R^{n \times n}$ with eigenvalues
$\lambda_1,...\, \lambda_n$, the size of the set $S \sim \DPP(\K)$
is distributed as a Poisson binomial random variable, namely, the
number of successes in $n$ Bernoulli random trials where the
probability of success in the $i$th trial is given by 
$\frac{\lambda_i}{\lambda_i+1}$. This leads to a simple
expression for the expected subset size:
\begin{align}
\E[|S|] = \sum_i \frac{\lambda_i}{\lambda_i + 1} = \tr(\K(\I + \K)^{-1}).\label{eq:size}
\end{align}
Note that if $S\sim\DPP(\frac1\alpha\K)$, where $\alpha>0$, then
$\Pr(S)$ is proportional to $\alpha^{-|S|}\det(\K_{S,S})$, so rescaling the kernel
by a scalar only affects the distribution of the subset sizes, giving
us a way to set the expected size to a desired value (larger $\alpha$
means smaller expected size).
Nevertheless, it is still often preferrable to restrict the size of
$S$ to a fixed $k$, obtaining a $k$-$\DPP(\K)$ \citep{k-dpp}.

Both DPPs and $k$-DPPs can be sampled efficiently, with some of the
first algorithms provided by
\citet{dpp-independence},
\citet{efficient-volume-sampling}, \citet{k-dpp} and others. These
approaches rely on an eigendecomposition of the kernel $\K$, at the cost of $O(n^3)$. When
$\K=\A^\top\A$, as in the CSSP, and the dimensions satisfy $m\ll n$, then this can be improved
to $O(nm^2)$. More recently, algorithms that avoid computing the
eigendecomposition have been proposed
\citep{dpp-intermediate,dpp-sublinear,alpha-dpp,rayleigh-mcmc},
resulting in running times of $\Ot(n)$ when given matrix $\K$ and
$\Ot(nm)$ for matrix $\A$, assuming small desired subset size.
See \citet{dppy} for an efficient Python implementation of DPP sampling.

The key property of DPPs that enables our analysis is a
formula for the expected value of the random matrix that is the orthogonal projection onto the
subspace spanned by vectors selected by $\DPP(\A^\top\A)$.
In the special case when $\A$ is a square full rank matrix, the
following result can be derived as a corollary of Theorem 1 by
\citet{randomized-newton}, and a variant for DPPs over
continuous domains can be found as Lemma~8 of
\citet{surrogate-design}. 
\begin{lemma}\label{l:expectation_proj}
  For any $\A$ and $S\subseteq [n]$, let $\P_S$ be the  
projection onto the $\mathrm{span}\{\a_i:i\in S\}$. If $S\sim \DPP(\A^\top\A)$, then
\begin{align*}
  \E[\P_S] & = \A(\I+\A^\top\A)^{-1}\A^\top.
\end{align*}
\end{lemma}
Lemma~\ref{l:expectation_proj} implies a simple closed form expression for
the expected error in the CSSP presented next. Here, we use a
rescaling parameter $\alpha>0$ for controlling
the distribution of the subset sizes. Note that it is crucial that we
are using a DPP with random subset size, because the corresponding
expression for the expected error of the fixed size $k$-DPP is
combinatorial, and therefore much harder to work with.
\begin{lemma}\label{l:expected-error}
  For any $\alpha>0$, if $S\sim
  \DPP(\frac1\alpha\A^\top\A)$, then
  \begin{align*}
    \E\big[\Er_\A(S)\big] =
    \tr\big(\A\A^\top(\I+\tfrac1\alpha\A\A^\top)^{-1}\big) = \E[|S|]\cdot\alpha.
  \end{align*}
\end{lemma}

\section{Main Results}
Our upper bounds rely on the notion of effective dimensionality called
stable rank~\citep{ridge-leverage-scores}. Here, we use an extended version of this concept, as
defined by \citet{BLLT19_TR}.
\begin{definition}[Stable rank]
Let $\lambda_1\geq \lambda_2\geq ...$ denote the eigenvalues of the
matrix $\A^\top\A$. For  $0\leq s<\rank(\A)$, we define the stable
rank of order $s$ as $\sr_s(\A) = \lambda_{s+1}^{-1}\sum_{i>s}\lambda_i$.
\end{definition}
In the following result, we define a family of functions $\Phi_s(k)$ which
bound the approximation factor $\Er_\A(S)/\opt$ in the range of $k$ between $s$ and
$s+\sr_s(\A)$. We call this the Master Theorem because we use it to derive
a number of more specific upper bounds.
\begin{theorem}[Master Theorem]\label{t:upper}
Given $0\leq s<\rank(\A) $, let  $t_s = s+\sr_s(\A)$,
and suppose that $s+ \frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$,
where $0<\epsilon\leq\frac12$. If $S\sim k$-$\DPP(\A^\top\A)$, then
\begin{align*}
  &\frac{\E[\Er_\A(S)]}{\opt}\leq (1+2\epsilon)^2\,\Phi_{s}(k),\\
  \text{where }& \Phi_s(k)=\big(1+\tfrac{s}{k-s}\big)\sqrt{1 + \tfrac{2(k-s)}{t_s-k}\,}.
\end{align*}
\end{theorem}
Note that we separated out the dependence on $\epsilon$ from the function
$\Phi_s(k)$, because the term $(1+2\epsilon)^2$ is an artifact of a
concentration of measure analysis that is unlikely to be of practical
significance. We conjecture that the dependence on $\epsilon$
can be eliminated from the statement entirely.% (see Conjecture \ref{c:convex}).

We next examine the consequences of
the Master Theorem, starting with a sharp transition that occurs as $k$
approaches the stable rank of $\A$. 
\begin{remark}[Sharp transition]\label{r:phase-transition}
For any $k$ it is true that:
%  \vspace{-3mm}
  \begin{enumerate}
\item  For all $\A$, if $k\leq \sr_0(\A)\!-\!1$, then there is a subset $S$ of size $k$ such that
$\frac{\Er_\A(S)}{\opt} = O(\sqrt  k\,)$.
\item There is $\A$ such that $\sr_0(\A)\!-\!1<k<\sr_0(\A)$ and for
  every size $k$ subset $S$, $\frac{\Er_\A(S)}{\opt}
  \geq 0.9\,k$.
  \end{enumerate}
\end{remark}
Part 1 of Remark \ref{r:phase-transition} follows from the Master Theorem by setting
$s=0$, whereas part 2 follows from the lower bound of
\citet{more-efficient-volume-sampling}. Observe how the worst-case
approximation factor jumps from $O(\sqrt k\,)$ to $\Omega(k)$, as $k$
approaches $\sr_0(\A)$. An example of this sharp
transition is shown in Figure \ref{f:intro}, where the stable
rank of $\A$ is around $20$.

While certain matrices directly exhibit the sharp transition from
Remark \ref{r:phase-transition}, many do not. In particular, for
matrices with a known rate of spectral decay, the Master Theorem can
be used to provide improved guarantees on the CSSP approximation
factor over \emph{all} subset sizes.

To illustrate this, we give novel bounds
for the two most commonly studied decay rates:  polynomial and exponential.
\begin{theorem}[Examples without sharp transition]\label{t:decay}
Let $\lambda_1\!\geq\!\lambda_2\!\geq\!...$ be the
eigenvalues of $\A^\top\A$. There is an absolute constant $c$ such that
for any $0\!<\!c_1\!\leq\!c_2$, with $\gamma=c_2/c_1$, if:\\[2mm]
\textnormal{1.} (\textbf{polynomial spectral decay}) $c_1i^{-p}\!\leq\!
  \lambda_i\!\leq\! c_2i^{-p}$ for all $i$, with $p>1$, then $S\sim
k$-$\DPP(\A^\top\A)$ satisfies
  \begin{align*}
    \frac{\E[\Er_\A(S)]}{\opt}\leq c \gamma p. 
  \end{align*}
\textnormal{2.} (\textbf{exponential spectral decay}) $c_1(1\!-\!\delta)^{i}\leq \lambda_i\leq
  c_2(1\!-\!\delta)^{i}$ for all $i$, with $\delta\in(0,1)$, then $S\sim
k$-$\DPP(\A^\top\A)$ satisfies %\vspace{-4mm}
  \begin{align*}
    \frac{\E[\Er_\A(S)]}{\opt}\leq c\gamma(1+ \delta k).
  \end{align*}
\end{theorem}
Note that for polynomial decay, unlike in \eqref{eq:old-bound}, the
approximation factor is constant, i.e., it does not depend on $k$. For
exponential decay, our bound provides an improvement over
\eqref{eq:old-bound} when $\delta=o(1)$.
To illustrate how these types of bounds can be obtained from the
Master Theorem, consider the function $\Phi_s(k)$ for some
$s>0$. The first term in the function, $1+\frac{s}{k-s}$, decreases with $k$, whereas
the second term (the square root) increases, albeit at a slower rate. This creates a
U-shaped curve which, if sufficiently wide, has a valley where the
approximation factor can get arbitrarily close to 1. This will occur
when $\sr_s(\A)$ is large, i.e., when the spectrum of $\A^\top\A$ has
a relatively flat region after the $s$th eigenvalue (Figure \ref{f:intro}
for $k$ between 20 and 40). Note that a peak value of some function
$\Phi_{s_1}$ may coincide with a valley of some $\Phi_{s_2}$, so only
taking a minimum over all functions reveals the true approximation
landscape predicted by the Master Theorem.

The peaks and valleys of the CSSP approximation
factor suggested by Theorem \ref{t:upper} are in fact an
inherent property of the problem, rather than an artifact of our
analysis or the result of using a particular algorithm. We prove this by
constructing a family of matrices $\A$ for which the best possible approximation
factor is large, i.e., close to the worst-case upper bound of
\citet{pca-volume-sampling}, not just for one size $k$, but for a
sequence of increasing sizes.
 \begin{theorem}[Lower bound]\label{t:lower}
For any $\delta\in(0,1)$ and
$0\!=\!k_0\!<\!k_1\!<\!...\!<\!k_t\!<\!n\leq m$, there is a matrix 
$\A\in\R^{m\times n}$ such that for any subset $S$ of size $k_i$,
where $i\in\{1,...,t\}$,
\begin{align*}
\frac{\Er_{\A}(S)}{\textsc{OPT}_{k_i}}\geq (1-\delta)(k_i-k_{i-1}).
  \end{align*}
\end{theorem}
Combining the Master Theorem with the lower
bound of Theorem \ref{t:lower} we can easily provide an example matrix
for which the optimal solution to the CSSP problem exhibits multiple
peaks and valleys. We refer to this phenomenon as the multiple-descent curve.
\begin{corollary}[Multiple-descent curve]\label{c:multiple-descent}
  For $t\in\N$ and $\delta\in(0,1)$, there is a
  sequence $0<k_1^l<k_1^u<k_2^l<k_2^u<...<k_t^l<k_t^u$ and
  $\A\in\R^{m\times n}$ such that for any $i\in\{1,...,t\}$:
  \begin{align*}
    \min_{S:|S|=k_i^l}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^l}}\leq 1+\delta,
                                 \qquad\text{and } \\
    \min_{S:|S|=k_i^u}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^u}} \geq
    (1-\delta)(k_i^u+1).
  \end{align*}
\end{corollary}

  \paragraph{The Nystr\"om method.} 
  We briefly discuss how our results translate to guarantees for the Nystr\"om
  mehod, a variant of the CSSP in the kernel setting which has
  gained considerable interest in the machine learning literature~\citep{dm_kernel_JRNL,revisiting-nystrom}.
In this context, rather than being given the column vectors
explicitly, we consider the 
$n\times n$ matrix $\K$ whose entry $(i,j)$ is the dot product
between the $i$th and $j$th vector in the kernel space,
$\langle\a_i,\a_j\rangle_\text{K}$. A Nystr\"om approximation of $\K$ based on subset $S$
is defined as $\Kbh(S) = \C\B^{\dagger}\C^\top$, where $\B$ is the
$|S|\times |S|$ submatrix of $\K$ indexed by $S$, whereas $\C$ is the
$n\times |S|$  submatrix with columns indexed by $S$. 
\begin{remark}
  If $\K=\A^\top\A$ and $\|\cdot\|_*$ is the trace norm, then
    $\big\|\K-\Kbh(S)\big\|_* = \Er_{\A}(S)$ for all $S\subseteq\{1,...,n\}$.
Moreover, the trace norm error of the best rank $k$ approximation of
$\K$, 
is equal to the squared Frobenius norm error of the 
best rank $k$ approximation of $\A$, i.e.,
\[\min_{\Kbh:\,\rank(\K)=k}\|\K-\Kbh\|_* = \opt.\]
\end{remark}

\section{Empirical Evaluation}
\label{s:experiments}

In this section, we provide an empirical evaluation designed to demonstrate how our improved guarantees for the CSSP and Nystr\"om method, as well as the multiple-descent phenomenon, can be easily observed on real datasets. 
We use a standard experimental setup for data subset selection using
the Nystr\"om method \citep{revisiting-nystrom}, where an $n\times n$
kernel matrix $\K$ for a dataset of size $n$ is defined so that the
entry $(i,j)$ is computed using the Gaussian Radial Basis Function  (RBF)
kernel:
$\langle\a_i,\a_j\rangle_\text{K}=\exp(-\|\a_i\!-\!\a_j\|^2/\sigma^2)$,
where $\sigma$ is a free parameter. 
We are particularly interested in the effect of varying $\sigma$.
Nystr\"om subset selection is performed using $S\sim k$-$\DPP(\K)$
(Definition \ref{d:dpp}), and we plot the expected approximation
factor $\E[\|\K-\Kbh(S)\|_*]/\opt$ (averaged over 1000 runs), where
$\Kbh(S)$ is the Nystr\"om approximation of $\K$ based on the subset
$S$, $\|\cdot\|_*$ is the trace norm,
and $\opt$ is the trace norm error of the best rank $k$
approximation. 
This task is
equivalent to the CSSP task defined on the matrix $\A$ such that
$\K=\A^\top\A$.

\begin{figure}[t]
   \centering
  \includegraphics[width=0.45\textwidth]{figs/nystrom/rbf-eunite2001-double}

  \caption{Top plot shows the Nystr\"om approximation factor $\E[\|\K-\Kbh(S)\|_*]/\opt$,
    where $S\sim k$-$\DPP(\K)$ for the eunite2001 Libsvm dataset ($\sigma$ is the RBF parameter). Error bars show three times the standard
    error  of the mean over 1000 trials. Bottom plot shows the spectral decay for  the
  top $40$ eigenvalues of the kernel $\K$. Note that the
  peaks in the approximation factor align with the drops in
  the spectrum.}
  \label{f:rbf}
\end{figure}

The aim of our empirical evaluation is to verify the following two claims motivated by our theory (and to illustrate that doing so is as easy as varying the RBF parameter $\sigma$):
\begin{enumerate}
  \item When the spectral decay is sufficiently slow/smooth, the
    approximation factor for CSSP/Nystr\"om is much better than
    suggested by previous worst-case bounds.
  \item  A drop in spectrum around the $k$th eigenvalue results in
    a peak in the approximation factor near
    subset size~$k$. Several drops result in the
    multiple-descent curve.
  \end{enumerate}
In Figure \ref{f:rbf} (top), we plot the approximation factor against
the subset size $k$ (in the range of 1 to 40) for a benchmark regression dataset from the Libsvm repository
\citep[\emph{eunite2001}, see][]{libsvm}. 
In Figure~\ref{f:rbf} (bottom), we also show the top
40 eigenvalues of the RBF kernel $\K$ in decreasing order, for
three different values of the parameter~$\sigma$.

The dataset \emph{eunite2001} (Figure~\ref{f:rbf}) exhibits a full
multiple-descent curve with up to three peaks for large values of
$\sigma$ (see top plot), and the peaks are once again aligned with the spectrum
drops (see bottom plot). Decreasing $\sigma$ gradually eliminates the peaks, 
resulting in a uniformly small approximation factor. Thus, both of our theoretical
claims can easily be verified on this dataset simply by adjusting the RBF parameter.

While the right choice of the parameter $\sigma$ ultimately depends on
the downstream machine learning task, it has been observed that
varying $\sigma$ has a pronounced effect on the spectral properties of
the kernel matrix, \citep[see,
e.g.,][]{revisiting-nystrom}.  
The main takeaway from our results here is that, depending on the
structure of the problem, we may end up in the regime where the
Nystr\"om approximation factor exhibits a multiple-descent curve
(e.g., due to a hierarchical nature of the data) or in the regime
where it is relatively flat.

\bibliographystyle{sty/ijcai2021/named}
\bibliography{pap}

\end{document}

### 2022-08-23 design/nystrom-main-neurips2020.tex
Commit original repos

\newif\ifisarxiv
\isarxivfalse

\documentclass{article}

\usepackage{multibib}
% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{cleveref}
%\usepackage{natbib}
\usepackage{xcolor}

\hypersetup{%draft, % TODO: remove "draft" option before final version after making sure no hyperlinks are broken due to pagebreaks.
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}

% repeating theorem statement in the supplement 
\newtheorem*{rep@theorem}{\rep@title}
\newcommand{\newreptheorem}[2]{%
	\newenvironment{rep#1}[1]{%
		\def\rep@title{#2 \ref{##1}}%
		\begin{rep@theorem}}%
		{\end{rep@theorem}}}
\newreptheorem{lemma}{Lemma'}
\newreptheorem{definition}{Definition'}
\newreptheorem{proposition}{Proposition'}
\newreptheorem{theorem}{Theorem'}
\newreptheorem{corollary}{Corollary'}

\input{shortdefs}

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2020} with \usepackage[nohyperref]{icml2020} above.

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

\ifisarxiv
\usepackage{fullpage}
\usepackage{natbib}
\else
\usepackage[final]{sty/neurips2020/neurips_2020}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[final]{neurips_2020}

\title{Improved guarantees and a multiple-descent curve for %the CSSP
  Column Subset Selection %Problem
  and the Nystr\"om method}
\author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
{Rajiv Khanna} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{rajivak@berkeley.edu}
  \And
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
  }
\fi
\begin{document}

\ifisarxiv
\title{Improved guarantees and a multiple-descent curve for\\ %the CSSP
  Column Subset Selection %Problem
  and the Nystr\"om method}
  \author{%
          \textbf{Micha{\l } Derezi\'{n}ski} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \and
\textbf{Rajiv Khanna} \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{rajivak@berkeley.edu}
  \and
   \textbf{Michael W. Mahoney}\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
  }

\else\fi

\ifisarxiv
  \date{}
\fi
  \maketitle

\begin{abstract}
The Column Subset Selection Problem (CSSP) and the Nystr\"om method
are among the leading tools for constructing small low-rank
approximations of large datasets in machine learning and scientific
computing.  
A fundamental question in this area is: how well can a data subset of
size $k$ compete with the best rank $k$ approximation?
We develop techniques which exploit spectral properties of the data
matrix to obtain improved approximation guarantees which go beyond the
standard worst-case analysis. 
Our approach leads to significantly better bounds for datasets with
known rates of singular value decay, e.g., polynomial or exponential
decay.  
Our analysis also reveals an intriguing phenomenon: the approximation
factor as a function of $k$ may exhibit multiple peaks and valleys,
which we call a multiple-descent curve.  
A lower bound we establish shows that this behavior is not an artifact
of our analysis, but rather it is an inherent property of the CSSP and
Nystr\"om tasks.
Finally, using the example of a radial basis function (RBF)
kernel, we show that both our improved bounds and the multiple-descent
curve can be observed on real datasets simply by varying the 
RBF parameter.  
\end{abstract}

\section{Introduction}
\label{s:intro}
We consider the task of selecting a small but representative sample of
column vectors from a large matrix. Known as the Column Subset Selection
Problem (CSSP), this is a well-studied combinatorial optimization task with
many applications in machine learning  
\citep[e.g., feature selection, see][]{Guyon03FeatureSelection,BoutsidisMD08},
scientific computing \citep[e.g.,][]{Chan92RankRevealingQR,Drineas08CUR} and signal processing
\citep[e.g.,][]{Balzano10MatchedSubspace}. In a 
commonly studied variant of this task, we aim to minimize the
squared error of projecting all columns of the matrix onto the
subspace spanned by the chosen column subset.
\begin{definition}[CSSP]
  Given an $m\times n$ matrix $\A$, pick a set $S\subseteq\{1,...,n\}$ of 
  $k$ column indices, to minimize
  \begin{align*}
   \Er_\A(S) := \|\A - \P_S\A\|_F^2,
  \end{align*}
  where $\|\cdot\|_F$ is the Frobenius norm, $\P_S$ is the
  projection onto $\mathrm{span}\{\a_i:i\in S\}$ and $\a_i$ denotes the $i$th
  column of~$\A$.
\end{definition}
Another variant of the CSSP emerges in the kernel
setting under the name \emph{Nystr\"om method}
\citep{Williams01Nystrom,dm_kernel_JRNL,revisiting-nystrom}. We also
discuss this variant, showing how our analysis applies in this context.
Both the CSSP and the Nystr\"om method are ways of
constructing accurate low-rank approximations by using submatrices of
the target matrix. Therefore, it is natural to ask how
close we can get to the best possible rank~$k$ approximation error:
\begin{align*}
  \opt := \!\!\min_{\B:\,\rank(\B)=k} \!\!\|\A-\B\|_F^2\leq \min_{S:\,|S|=k}\Er_\A(S).
\end{align*}
Our goal is to find a subset $S$ of size $k$ for which the ratio
between $\Er_\A(S)$ and $\opt$ is small. Furthermore, a brute force
search requires iterating over all 
${n \choose k}$ subsets, which is prohibitively expensive, so we
would like to find our subset more efficiently.

\ifisarxiv\begin{figure}\else
\begin{wrapfigure}{r}{0.45\textwidth}
  \vspace{-5mm}
  \fi
  \centering  
  \includegraphics[width=\ifisarxiv 0.6\else 0.47\fi\textwidth]{figs/nystrom/nystrom-bounds}
  \caption{Empirical study of the
    expected approximation factor $\E[\Er_\A(S)]/\opt$ for
a $k$-DPP with
    different subset sizes $|S|=k$, compared to our theory. We use a
    data matrix $\A$ 
    whose
    spectrum exhibits two sharp drops, demonstrating
multiple-descent. The lower bounds are based on Theorem
    \ref{t:lower}, whereas, as our upper bound, we plot the
    minimum over all $\Phi_s(k)$ from
    Theorem~\ref{t:upper}. Note that multiple-descent vanishes under
    smooth spectral decay, resulting in improved guarantees (see
    Theorem \ref{t:decay} and Figure \ref{f:sliding}).
  }
  \label{f:intro}
  \ifisarxiv\end{figure}\else
  \end{wrapfigure}\fi

Extensive literature has been dedicated to developing algorithms for
the CSSP (an in depth discussion of the related work can be found in
Appendix \ref{s:related-work}).
In terms of worst-case analysis, 
\citet{pca-volume-sampling} gave a randomized
method 
which returns a
set $S$ of size $k$ such that: %\vspace{-4mm}
\begin{align}
  \frac{\E[\Er_\A(S)]}{\opt}\leq k+1.\label{eq:old-bound}
\end{align}
While the original algorithm was slow, 
efficient implementations have been provided since then
\citep[e.g., see][]{efficient-volume-sampling,dpp-intermediate}.
The method belongs to the family of cardinality constrained
Determinantal Point Processes \citep[DPPs, see][]{dpps-in-randnla}, and will
be denoted as $S\sim k\text{-}\DPP(\A^\top\A)$. 
The approximation  factor  $k+1$ is optimal in the worst-case,
since for any $0<k<n\leq m$ and $0<\delta<1$, an $m\times n$
matrix~$\A$ can be constructed for which $\frac{\Er_{\A}(S)}{\opt}\geq
(1-\delta)(k+1)$ for all subsets $S$ of size $k$. 
Yet it is known that, in practice, CSSP algorithms perform better
than worst-case, so the question we consider is: how can we go beyond
the usual worst-case analysis to accurately reflect what is possible
in the CSSP?

\textbf{Contributions.} We provide improved guarantees for the CSSP
approximation factor, which go beyond the worst-case analysis and 
which lead to surprising conclusions.%\vspace{-3mm}
\begin{enumerate}
\item\underline{New upper bounds}:
We develop a family of upper bounds on the CSSP approximation factor
(Theorem~\ref{t:upper}), which we call the Master Theorem
as they can be used to derive a number 
of new guarantees. In particular, we show that when the data matrix $\A$ exhibits a
known spectral decay, then \eqref{eq:old-bound} can often be
drastically improved (Theorem \ref{t:decay}).
\item \underline{New lower bound}:
Even though the worst-case upper bound in \eqref{eq:old-bound} can often
be loose, there are cases when it cannot be improved.
We give a new lower bound construction (Theorem \ref{t:lower}) showing
that there are matrices $\A$ for which multiple different subset sizes exhibit
worst-case behavior.
\item \underline{Multiple-descent curve}: 
Our upper and lower bounds
reveal that for some matrices the CSSP approximation factor can exhibit
peaks and valleys as a function of the subset size $k$ (see Figure
\ref{f:intro}).  We show that this phenomenon is an inherent property
of the CSSP  (Corollary \ref{c:multiple-descent}).
\end{enumerate}

\subsection{Main results}
Our upper bounds rely on the notion of effective dimensionality called
stable rank~\citep{ridge-leverage-scores}. Here, we use an extended version of this concept, as
defined by \citet{BLLT19_TR}.
\begin{definition}[Stable rank]
Let $\lambda_1\geq \lambda_2\geq ...$ denote the eigenvalues of the
matrix $\A^\top\A$. For  $0\leq s<\rank(\A)$, we define the stable
rank of order $s$ as $\sr_s(\A) = \lambda_{s+1}^{-1}\sum_{i>s}\lambda_i$.
\end{definition}
In the following result, we define a family of functions $\Phi_s(k)$ which
bound the approximation factor $\Er_\A(S)/\opt$ in the range of $k$ between $s$ and
$s+\sr_s(\A)$. We call this the Master Theorem because we use it to derive
a number of more specific upper bounds.
\begin{theorem}[Master Theorem]\label{t:upper}
Given $0\leq s<\rank(\A) $, let  $t_s = s+\sr_s(\A)$,
and suppose that $s+ \frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$,
where $0<\epsilon\leq\frac12$. If $S\sim k$-$\DPP(\A^\top\A)$, then
\begin{align*}
  \frac{\E[\Er_\A(S)]}{\opt}\leq (1+2\epsilon)^2\,\Phi_{s}(k),
  \qquad\text{where}\quad \Phi_s(k)=\big(1+\tfrac{s}{k-s}\big)\sqrt{1 + \tfrac{2(k-s)}{t_s-k}\,}.
\end{align*}
\end{theorem}
Note that we separated out the dependence on $\epsilon$ from the function
$\Phi_s(k)$, because the term $(1+2\epsilon)^2$ is an artifact of a
concentration of measure analysis that is unlikely to be of practical
significance. In fact, we believe that the dependence on $\epsilon$
can be eliminated from the statement entirely (see
Conjecture \ref{c:convex}).

We next examine the consequences of
the Master Theorem, starting with a sharp transition that occurs as $k$
approaches the stable rank of $\A$. 
\begin{remark}[Sharp transition]\label{r:phase-transition}
For any $k$ it is true that:
%  \vspace{-3mm}
  \begin{enumerate}
\item  For all $\A$, if $k\leq \sr_0(\A)\!-\!1$, then there is a subset $S$ of size $k$ such that
$\frac{\Er_\A(S)}{\opt} = O(\sqrt  k\,)$.
\item There is $\A$ such that $\sr_0(\A)\!-\!1<k<\sr_0(\A)$ and for
  every size $k$ subset $S$, $\frac{\Er_\A(S)}{\opt}
  \geq 0.9\,k$.
  \end{enumerate}
\end{remark}
Part 1 of the remark follows from the Master Theorem by setting
$s=0$, whereas part 2 follows from the lower bound of
\citet{more-efficient-volume-sampling}. Observe how the worst-case
approximation factor jumps from $O(\sqrt k\,)$ to $\Omega(k)$, as $k$
approaches $\sr_0(\A)$. An example of this sharp
transition is shown in Figure \ref{f:intro}, where the stable
rank of $\A$ is around $20$.

While certain matrices directly exhibit the sharp transition from
Remark \ref{r:phase-transition}, many do not. In particular, for
matrices with a known rate of spectral decay, the Master Theorem can
be used to provide improved guarantees on the CSSP approximation
factor over \emph{all} subset sizes.

To illustrate this, we give novel bounds
for the two most commonly studied decay rates:  polynomial and exponential.
\begin{theorem}[Examples without sharp transition]\label{t:decay}
Let $\lambda_1\!\geq\!\lambda_2\!\geq\!...$ be the
eigenvalues of $\A^\top\A$. There is an absolute constant $c$ such that
for any $0\!<\!c_1\!\leq\!c_2$, with $\gamma=c_2/c_1$, if:\\[2mm]
\textnormal{1.} (\textbf{polynomial spectral decay}) $c_1i^{-p}\!\leq\!
  \lambda_i\!\leq\! c_2i^{-p}$ $\forall_i$, with $p>1$, then $S\sim
k$-$\DPP(\A^\top\A)$ satisfies
  \begin{align*}
    \frac{\E[\Er_\A(S)]}{\opt}\leq c \gamma p. 
  \end{align*}
\textnormal{2.} (\textbf{exponential spectral decay}) $c_1(1\!-\!\delta)^{i}\leq \lambda_i\leq
  c_2(1\!-\!\delta)^{i}$ $\forall_i$, $\delta\in(0,1)$, then $S\sim
k$-$\DPP(\A^\top\A)$ satisfies %\vspace{-4mm}
  \begin{align*}
    \frac{\E[\Er_\A(S)]}{\opt}\leq c\gamma(1+ \delta k).
  \end{align*}
\end{theorem}
Note that for polynomial decay, unlike in \eqref{eq:old-bound}, the
approximation factor is constant, i.e., it does not depend on $k$. For
exponential decay, our bound provides an improvement over
\eqref{eq:old-bound} when $\delta=o(1)$.
To illustrate how these types of bounds can be obtained from the
Master Theorem, consider the function $\Phi_s(k)$ for some
$s>0$. The first term in the function, $1+\frac{s}{k-s}$, decreases with $k$, whereas
the second term (the square root) increases, albeit at a slower rate. This creates a
U-shaped curve which, if sufficiently wide, has a valley where the
approximation factor can get arbitrarily close to 1. This will occur
when $\sr_s(\A)$ is large, i.e., when the spectrum of $\A^\top\A$ has
a relatively flat region after the $s$th eigenvalue (Figure \ref{f:intro}
for $k$ between 20 and 40). Note that a peak value of some function
$\Phi_{s_1}$ may coincide with a valley of some $\Phi_{s_2}$, so only
taking a minimum over all functions reveals the true approximation
landscape predicted by the Master Theorem. To prove Theorem
\ref{t:decay}, we show that the stable ranks $\sr_s(\A)$ are
sufficiently large so that any $k$ lies in  the valley of some
function $\Phi_s(k)$ (see Section \ref{s:upper}).

The peaks and valleys of the CSSP approximation
factor suggested by Theorem \ref{t:upper} are in fact an
inherent property of the problem, rather than an artifact of our
analysis or the result of using a particular algorithm. We prove this by
constructing a family of matrices $\A$ for which the best possible approximation
factor is large, i.e., close to the worst-case upper bound of
\citet{pca-volume-sampling}, not just for one size $k$, but for a
sequence of increasing sizes.
 \begin{theorem}[Lower bound]\label{t:lower}
For any $\delta\in(0,1)$ and
$0\!=\!k_0\!<\!k_1\!<\!...\!<\!k_t\!<\!n\leq m$, there is a matrix 
$\A\in\R^{m\times n}$ such that for any subset $S$ of size $k_i$,
where $i\in\{1,...,t\}$,
\begin{align*}
\frac{\Er_{\A}(S)}{\textsc{OPT}_{k_i}}\geq (1-\delta)(k_i-k_{i-1}).
  \end{align*}
\end{theorem}
Combining the Master Theorem with the lower
bound of Theorem \ref{t:lower} we can easily provide an example matrix
for which the optimal solution to the CSSP problem exhibits multiple
peaks and valleys. We refer to this phenomenon as the multiple-descent curve.
\begin{corollary}[Multiple-descent curve]\label{c:multiple-descent}
  For $t\in\N$ and $\delta\in(0,1)$, there is a
  sequence $0<k_1^l<k_1^u<k_2^l<k_2^u<...<k_t^l<k_t^u$ and
  $\A\in\R^{m\times n}$ such that for any $i\in\{1,...,t\}$:
  \begin{align*}
    \min_{S:|S|=k_i^l}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^l}}\leq 1+\delta
                                 \qquad\text{and }\qquad
    \min_{S:|S|=k_i^u}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^u}} \geq
    (1-\delta)(k_i^u+1).
  \end{align*}
\end{corollary}
\paragraph{Connection to double descent.}
A number of phase transitions have been recently observed in the
machine learning literature which are commonly dubbed \emph{double
  descent}. The term was introduced by \citet{BHMM19} in the context of generalization
error of statistical learning, however
\citet{double-descent-condition} observed that the behavior of the
generalization error, at least for linear models, can be explained by
a more fundamental double descent phenomenon observed in the condition
number of random matrices. Also, \citet{zhenyu2020double} showed that
double descent is merely a manifestation of the
spectral phase transitions in high-dimensional random kernel
matrices. Further, \citet{liang2020multiple} observed that the spectral phase transitions
of certain kernels can lead to multiple double descent peaks. The
multiple-descent phenomenon in the CSSP that 
emerges from our analysis is
related to those works in that the spectral properties of the data
matrix (and, in particular, the condition number) determine the peaks
(i.e., phase transitions) discussed in
Corollary~\ref{c:multiple-descent}. Those parallels manifest
themselves most clearly when comparing our analysis to the work of
\citet{BLLT19_TR}, which uses the same notion of stable rank as we do,
and \citet{surrogate-design}, where determinantal sampling plays a
central role in the analysis of double descent. However, we stress that there are
  also important  differences in our setting: (1) the CSSP is a
  \emph{deterministic} optimization task, and (2) we study the
  \emph{approximation factor}, rather than the generalization error (further discussion in
  Appendix~\ref{s:related-work}).

  \subsection{The Nystr\"om method}\label{s:nystrom}
  We briefly discuss how our results translate to guarantees for the Nystr\"om
  mehod, a variant of the CSSP in the kernel setting which has
  gained considerable interest in the machine learning literature~\citep{dm_kernel_JRNL,revisiting-nystrom}.
In this context, rather than being given the column vectors
explicitly, we consider the 
$n\times n$ matrix $\K$ whose entry $(i,j)$ is the dot product
between the $i$th and $j$th vector in the kernel space,
$\langle\a_i,\a_j\rangle_\text{K}$. A Nystr\"om approximation of $\K$ based on subset $S$
is defined as $\Kbh(S) = \C\B^{\dagger}\C^\top$, where $\B$ is the
$|S|\times |S|$ submatrix of $\K$ indexed by $S$, whereas $\C$ is the
$n\times |S|$  submatrix with columns indexed by $S$. The Nystr\"om
method has many applications in machine learning, including for
kernel machines \citep{Williams01Nystrom}, Gaussian Process regression 
\citep{sparse-variational-gp} and Independent Component Analysis
\citep{Bach2003}. 
\begin{remark}
  If $\K=\A^\top\A$ and $\|\cdot\|_*$ is the trace norm, then
    $\big\|\K-\Kbh(S)\big\|_* = \Er_{\A}(S)$ for all $S\subseteq\{1,...,n\}$.
Moreover, the trace norm error of the best rank $k$ approximation of
$\K$, 
is equal to the squared Frobenius norm error of the 
best rank $k$ approximation of $\A$, i.e.,
\[\min_{\Kbh:\,\rank(\K)=k}\|\K-\Kbh\|_* = \opt.\]
\end{remark}
%\vspace{-2mm}
This connection was used by~\citet{belabbas-wolfe09} to adapt the
$k+1$ approximation factor bound 
of~\citet{pca-volume-sampling} to the Nystr\"om method.   
Similarly, all of our results for the CSSP, including the
multiple-descent curve that we have observed, can be translated into analogous
statements for the trace norm approximation error in the Nystr\"om
method. Of particular interest 
are the improved bounds for kernel matrices with known eigenvalue
decay rates. Such matrices arise naturally in machine learning when
using standard kernel functions such as the Gaussian Radial Basis Function
(RBF) kernel (a.k.a.~the squared exponential kernel) and the
Mat\'ern kernel \citep{sparse-variational-gp}.

\underline{RBF kernel}: If
$\langle\a_i,\a_j\rangle_\text{K}=\exp(-\|\a_i\!-\!\a_j\|^2/\sigma^2)$ 
and the data comes from $\Nc(0,\eta^2)$, then,
  for large $n$,
  $\lambda_i\!\asymp\!\lambda_1(\frac{b}{a+b+c})^i$, where
  $a=\frac1{4\eta^2}$, $b=\frac1{\sigma^2}$ and $c=\sqrt{a^2\!+\!2ab}$
  \citep{Santa97gaussianregression},
  so Theorem~\ref{t:decay} yields an approximation factor of
  $O(1\!+\!\frac{a+c}{a+b+c}k)$, better than $k\!+\!1$ when
  $\sigma^2\ll\eta^2$. Note that the parameter $\sigma$ defines the
  size of a neighborhood around which the data points are deemed
  similar by the RBF kernel. Therefore, smaller $\sigma$ means that
  each data point has fewer similar neighbors.

  \underline{Mat\'ern kernel}: If $\K$ is the Mat\'ern kernel with
  parameters $\nu$ and $\ell$ and the data is distributed according to a
  uniform measure in one dimension,
  then $\lambda_i\asymp \lambda_1 i^{-2\nu-1}$
  \citep{RasmussenWilliams06}, so Theorem~\ref{t:decay} yields a
  Nystr\"om approximation factor of $O(1+\nu)$ for any subset size
  $k$.

In Section \ref{s:experiments}, we also empirically
demonstrate our improved guarantees and the multiple-descent curve for the Nystr\"om method with the
RBF kernel.

\section{Upper bounds}\label{s:upper}\vspace{-2mm}
In this section, we derive the upper bound given in
Theorem~\ref{t:upper} by using a novel expectation formula for the squared
projection error of a DPP. We then show
how this result can be used to 
obtain improved guarantees for matrices with known eigenvalue decays,
i.e., Theorem \ref{t:decay}. Our analysis heavily relies on the theory
of DPPs \citep{dpps-in-randnla}, so for completeness, in Appendix~\ref{s:dpp} we provide a
brief summary of DPPs and the relevant results.

Let $S\sim \DPP(\frac1\alpha\A^\top\A)$ denote a
distribution over all subsets $S\subseteq [n]$ so that
$\Pr(S)\propto\det(\frac1\alpha\A_S^\top\A_S)$, where $\alpha>0$. Then,
$k$-$\DPP(\A^\top\A)$ is simply a restriction of
$\DPP(\frac1\alpha\A^\top\A)$ to the subsets of size $k$ (regardless
of the choice of $\alpha$).
However, the expected subset size for $\DPP(\frac1\alpha\A^\top\A)$ does
depend on~$\alpha$. Our analysis relies on a careful selection of
this parameter. In Lemma \ref{l:expected-error}  (Appendix \ref{s:dpp}), we show the
following expectation formula for the CSSP approximation error: 
\[\E[\Er_\A(S)]=\E[|S|]\cdot\alpha,\quad\text{for}\quad S\sim \DPP(\tfrac1\alpha\A^\top\A).\]
If we set $\alpha =\opt = \sum_{i=k+1}^n\lambda_i$,
where $\lambda_i$ are the eigenvalues of
$\A^\top\A$ in decreasing order, then:
\begin{align*}
 \E[|S|]& = \sum_{i=1}^n\frac{\lambda_i}{\alpha+\lambda_i}
\leq\sum_{i=1}^k\frac{\lambda_i}{\alpha+\lambda_i} + 1\leq k+1.
\end{align*}
This recovers the upper bound of \citet{pca-volume-sampling}, i.e., $\E[\Er_\A(S)]\leq(k+1)\opt$,
except that the
subset size is randomized with expectation bounded by 
$k+1$, instead of a fixed subset size equal $k$. However, a more refined
choice of the parameter $\alpha$
allows us to significantly improve on the above error bound in certain
regimes, as shown below.
\begin{lemma}\label{l:upper}
For any $\A$, $0\leq\epsilon<1$ and $s<k<t_s$, where $t_s=s+\sr_s(\A)$,
say $S\sim\DPP(\frac1\alpha\A^\top\A)$ for
$\alpha=\frac{\gamma_s(k)\opt}{(1-\epsilon)(k-s)}$ and
$\gamma_s(k):=\sqrt{1+\frac{2(k-s)}{t_s-k}\,}$. Then, defining $\Phi_s(k)
:=\big(1+\frac{s}{k-s}\big)\,\gamma_s(k)$,
\begin{align*}
  \frac{\E\big[\Er_\A(S)\big]}{\opt}\leq \frac{\Phi_s(k)}{1-\epsilon}
  \quad\text{and}\quad \E[|S|]\leq k-\epsilon\,\frac{k-s}{\gamma_s(k)}.
\end{align*}
  \end{lemma}
Note that, setting $\epsilon=0$, the above lemma implies that we can
achieve approximation factor $\Phi_s(k)$ with a DPP whose expected
size is bounded by $k$. We introduce $\epsilon$ so that we can
convert the bound from DPP to the fixed size k-DPP via a
concentration argument. Intuitively, our strategy is to show that the
randomized subset size of a DPP is sufficiently concentrated around
its expectation that with high probability it will be bounded by $k$,
and for this we need the expectation to be strictly below $k$. A
careful application of the Chernoff bound for a Poisson binomial
random variable yields the following concentration bound.
\begin{lemma}\label{l:concentration}
Let $S$ be as in Lemma \ref{l:upper} with $\epsilon\leq \frac12$.
If $s+\frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$, then
$\Pr(|S|>k) \leq \epsilon$.
\end{lemma}
Finally, any expected bound for random size DPPs can be converted to
an expected bound for a fixed size k-DPP via the following
result.
\begin{lemma}\label{l:monotonic}
  For any $\A\in\R^{m\times n}$, $k\in[n]$ and $\alpha>0$, if
  $S\sim\DPP(\frac1\alpha\A^\top\A)$ and $S'\sim k$-$\DPP(\A^\top\A)$,
  then %\vspace{-4mm}
  \begin{align*}
    \E\big[\Er_\A(S')\big] \leq \E\big[\Er_\A(S)\mid |S|\leq k\big].
  \end{align*}
\end{lemma}
The above inequality may seem intuitively obvious since adding more columns
to a set $S$ to complete it to size $k$ always reduces the
error. However, a priori, it could happen that going from subsets 
of size $k-1$ to subsets of size $k$ results in a redistribution of
probabilities to the subsets with larger error. To show that this will
not happen, our proof relies on classic but non-trivial combinatorial
bounds called Newton's inequalities. Putting together
Lemmas~\ref{l:upper}, \ref{l:concentration} and
\ref{l:monotonic}, we obtain our Master Theorem.
\begin{proofof}{Theorem}{\ref{t:upper}}
Let $S\sim\DPP(\frac1\alpha\A^\top\A)$ be sampled as in Lemma
\ref{l:upper}, and let $S'\sim k$-$\DPP(\A^\top\A)$. We have:\vspace{-2mm}
\begin{align*}
  \E\big[\Er_\A(S')\big]
\overset{(a)}{\leq} \E\big[\Er_\A(S)\mid|S|\leq k\big]
  \leq \frac{\E\big[\Er_\A(S)\big]}{\Pr(|S|\leq k)}
  \overset{(b)}{\leq} \frac{\Phi_s(k)}{(1-\epsilon)^2}\cdot\opt,
\end{align*}
where $(a)$ follows from Lemma \ref{l:monotonic} and $(b)$ follows
from Lemmas \ref{l:upper} and \ref{l:concentration}. Since
$0<\epsilon\leq\frac12$, we have
$\frac1{(1-\epsilon)^2}\leq(1+2\epsilon)^2$, which completes the proof.
\end{proofof}

\begin{figure}
  \centering
  \ifisarxiv
  \includegraphics[width=0.6\textwidth]{figs/nystrom/sliding}
\includegraphics[width=0.6\textwidth]{figs/nystrom/sliding-peak}
\else  
%\centering
\hspace{-8mm}
\includegraphics[width=0.54\textwidth]{figs/nystrom/sliding}\hspace{-5mm}
\includegraphics[width=0.54\textwidth]{figs/nystrom/sliding-peak}
%  \vspace{-6mm}
  \fi
  \caption{Illustration of the upper bound functions $\Phi_s(k)$ for
    different values of $s$, with a $200\times 200$ matrix $\A$ such
    that the $i$th eigenvalue of $\A^\top\A$ is set to:
    \ifisarxiv(top)\else(left)\fi\
    $1/i$;
    \ifisarxiv(bottom)\else(right)\fi\
    $1$ for $i<30$ and $0.01$ for $i\geq 30$. For each
    function, we marked the window of applicable $k$'s with a
    horizontal line. For polynomial spectral decay
    \ifisarxiv(top)\else(left)\fi
    , the stable rank
    $\sr_s(\A)$ (i.e., the width of the window starting at $s$)
    increases, while for the sharp spectrum drop
    \ifisarxiv(bottom)\else(right)\fi\
    the stable
    rank shrinks as the window approaches the drop, causing a peak in
    the upper~bound.}\vspace{-2mm}
  \label{f:sliding}
\end{figure}

We now demonstrate how Theorem \ref{t:upper} can be used as the Master Theorem to derive new
bounds on the CSSP approximation factor under additional assumptions
on the singular value decay of matrix $\A$. Rather than a single upper
bound, Theorem \ref{t:upper} provides a family of upper bounds
$\Phi_s$, each with a range of applicable values $k$. Since each
$\Phi_s(k)$ forms a U-shaped curve, its smallest point falls near the
middle of that range. In Figure \ref{f:sliding} we visualize these bounds as a sliding
window that sweeps across the axis representing possible subset
sizes. The width of the window varies: when it starts at $s$ 
then its width is the stable rank $\sr_s(\A)$. The wider the window,
the lower is the valley of the corresponding U-curve. Thus, when
bounding the approximation factor for a given $k$, we should choose
the widest window such that $k$ falls near the bottom of its
U-curve. Showing a guarantee that holds for all $k$ requires lower-bounding the
stable ranks $\sr_s(\A)$ for each $s$. This is straightforward for
both polynomial and exponential decay. Specifically, using the
notation from Theorem \ref{t:decay}, in Appendix \ref{a:decay} we
prove that: %\vspace{-3mm}
\begin{align*}
  \sr_s(\A) =
  \begin{cases}
    \Omega(s/p), &\text{for polynomial rate  $\lambda_i\asymp1/i^p$},\\
    \Omega(1/\delta),&\text{for exponential rate $\lambda_i\asymp(1-\delta)^i$}.
    \end{cases}
  \end{align*}
As an example, Figure \ref{f:sliding} (left) shows that the stable rank $\sr_s(\A)$, i.e.,
the width of the window starting at $s$, grows linearly with $s$ for eigenvalues
decaying polynomially with $p=1$. As a result, the bottom of each
U-shaped curve remains at roughly the same level, making the CSSP
approximation factor independent of $k$, as in Theorem
\ref{t:decay}. In contrast, Figure \ref{f:sliding} (right) provides
the same plot for a different matrix $\A$ with a sharp drop in the spectrum. The U-shaped
curves cannot slide smoothly across that drop because of the
shrinking stable ranks, which results in a peak similar to the ones
observed in Figure \ref{f:intro}.

 \section{Lower bound}\label{s:lower}
%\vspace{-2mm}
 As discussed in the previous section, our upper bounds for the CSSP
 approximation factor exhibit a peak (a high point,
 with the bound decreasing on either side) around a
 subset size $k$ when there is a sharp drop in the spectrum of $\A$
 around the $k$th singular value. It is natural to ask whether this
 peak is an artifact of our analysis, or a property of the k-DPP
 distribution, or whether even optimal CSSP subsets exhibit this 
 phenomenon. In this section, we extend a lower
 bound construction of \citet{pca-volume-sampling} and use it to show
 that for certain matrices the approximation factor of the optimal CSSP subset, i.e.,
 $\min_{|S|=k}\Er_\A(S)/\opt$, can exhibit not just one but any
 number of peaks as a function of $k$, showing that the
 multiple-descent curve in Figure \ref{f:intro} describes a
real phenomenon in the CSSP.

The lower bound construction of \citet{pca-volume-sampling} relies on
arranging the column vectors of a $(k+1)\times (k+1)$ matrix $\A$ into a
centered symmetric $k$-dimensional simplex. This way, the $k+1$
columns are spanning a $k$ dimensional subspace which contains the $k$
leading singular vectors of $\A$. They then proceed to shift the
columns slightly in the direction orthogonal to that subspace so that
the $(k+1)$st singular value of $\A$ becomes non-zero. This results in
an instance of the CSSP with a sharp drop in the spectrum. Due
to the symmetry in this construction, all subsets of size $k$ have an identical squared
projection error. It is easy to show that this error satisfies $\Er_\A(S)\geq (1-\delta)(k+1)\opt$,
where $\delta$ is a parameter which depends on the condition number of
matrix $\A$ and it can be driven arbitrarily close to $0$. Another
variant of this construction was also provided 
by \citet{more-efficient-volume-sampling}. The key limitation of both
of these constructions is that they only provide a lower bound for a
single subset size $k$ in a given matrix, whereas our goal is to show
that the CSSP can exhibit the multiple-descent curve, which requires
lower bounds for multiple different values of $k$ holding with respect
to the same matrix $\A$.

Our strategy for constructing the lower bound matrix is to concatenate
together multiple sets of columns, each of which
represents a simplex spanning some subspace of $\R^m$. The key
challenge that we face in this approach is that, unlike in
the construction of \citet{pca-volume-sampling}, different subsets of
the same size will have different projection errors. 
\begin{lemma}\label{l:lower}
Fix $\delta\in(0,1)$ and consider unit vectors
$\a_{i,j}\in\R^m$ in general position, where $i\in[t]$, $j\in[l_i]$, such that
$\sum_j\a_{i,j}=0$ for each $i$, and for any $i,j,i',j'$, if $i\ne i'$
then $\a_{i,j}$ is orthogonal to $\a_{i'\!,j'}$. Also, let unit vectors
$\{\v_i\}_{i\in[t]}$ be orthogonal to  each other and to all
$\a_{i,j}$. There are positive scalars $\alpha_{i},\beta_i$ for $ i \in [t]$
such that matrix $\A$ with columns 
$\alpha_{i}\a_{i,j}+\beta_i\v_i$ over all $i$ and $j$ satisfies:
%\ifisarxiv\else\vspace{-1mm}\fi
\begin{align*}
  \min_{|S|=k_i}\frac{\Er_\A(S)}{\mathrm{OPT}_{k_i}}\geq
  (1-\delta)l_i,\quad\text{for }k_i=l_1+...+l_i-1.
\end{align*}
\end{lemma}
%\ifisarxiv\else \vspace{-5mm}\fi
% Theorem \ref{t:lower} now follows easily from the lemma.
\begin{proofof}{Theorem}{\ref{t:lower}}
We let $l_1 = k_1+1$ and then for $i>1$ we set $l_i=k_i-k_{i-1}$. We
then construct the vectors $\a_{i,j}$ that satisfy Lemma \ref{l:lower} 
by letting each set $\{\a_{i,j}\}_{j}$ be the corners of a centered
$(l_i-1)$-dimensional regular simplex. We ensure that each
simplex is orthogonal to every other simplex by placing them in
orthogonal subspaces. 
\end{proofof}

\section{Empirical evaluation}
\label{s:experiments}
\vspace{-2mm}

In this section, we provide an empirical evaluation designed to demonstrate how our improved guarantees for the CSSP and Nystr\"om method, as well as the multiple-descent phenomenon, can be easily observed on real datasets. 
We use a standard experimental setup for data subset selection using
the Nystr\"om method \citep{revisiting-nystrom}, where an $n\times n$
kernel matrix $\K$ for a dataset of size $n$ is defined so that the
entry $(i,j)$ is computed using the Gaussian Radial Basis Function  (RBF)
kernel:
$\langle\a_i,\a_j\rangle_\text{K}=\exp(-\|\a_i\!-\!\a_j\|^2/\sigma^2)$,
where $\sigma$ is a free parameter. 
We are particularly interested in the effect of varying $\sigma$.
Nystr\"om subset selection is performed using $S\sim k$-$\DPP(\K)$
(Definition \ref{d:dpp}), and we plot the expected approximation
factor $\E[\|\K-\Kbh(S)\|_*]/\opt$ (averaged over 1000 runs), where
$\Kbh(S)$ is the Nystr\"om approximation of $\K$ based on the subset
$S$ (see Section~\ref{s:nystrom}), $\|\cdot\|_*$ is the trace norm,
and $\opt$ is the trace norm error of the best rank $k$
approximation. Additional experiments, using greedy selection instead 
of a k-DPP, are in Appendix~\ref{a:greedy}.
As discussed in Section \ref{s:nystrom}, this task is
equivalent to the CSSP task defined on the matrix $\A$ such that
$\K=\A^\top\A$.

The aim of our empirical evaluation is to verify the following two claims motivated by our theory (and to illustrate that doing so is as easy as varying the RBF parameter $\sigma$):
%\ifisarxiv\else\vspace{-2mm}\fi
\begin{enumerate}
  \item When the spectral decay is sufficiently slow/smooth, the
    approximation factor for CSSP/Nystr\"om is much better than
    suggested by previous worst-case bounds.
%    \vspace{-1mm}
  \item  A drop in spectrum around the $k$th eigenvalue results in
    a peak in the approximation factor near
    subset size~$k$. Several drops result in the
    multiple-descent curve.
  \end{enumerate}
%\ifisarxiv\else  \vspace{-2mm}\fi
In Figure \ref{f:rbf} (top), we plot the approximation factor against
the subset size $k$ (in the range of 1 to 40) for an artificial toy
dataset and 
for two benchmark regression datasets from the Libsvm repository
\citep[\emph{bodyfat} and \emph{eunite2001}, see][]{libsvm}. 
The toy dataset is constructed by scaling the eigenvalues of a random
$50\times 50$ Gaussian matrix so that the spectrum is flat with a
single drop at the $21$-st eigenvalue. 
For each dataset, in Figure~\ref{f:rbf} (bottom), we also show the top
40 eigenvalues of the kernel $\K$ in decreasing order. 
For the toy dataset, to maintain full control over the spectrum we use
the linear kernel $\langle\a_i,\a_j\rangle_\text{K}=\a_i^\top\a_j$, and we show results
for three different values of the condition number $\kappa$ of kernel
$\K$.  
For the benchmark datasets, we show results on the RBF kernel with 
three different values of the parameter~$\sigma$.

\begin{figure*}[t]
   \centering
  \ifisarxiv
  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-toy-double}
   \hspace{-6mm}
  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-bodyfat-double}
  \hspace{-6mm}
  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-eunite2001-double}
  \vspace{-5mm}
  \else
    \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-toy-double}
   \hspace{-6mm}
  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-bodyfat-double}
  \hspace{-6mm}
  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-eunite2001-double}
  \vspace{-4mm}
  \fi
  \caption{Top three plots show the Nystr\"om approximation factor $\E[\|\K-\Kbh(S)\|_*]/\opt$,
    where $S\sim k$-$\DPP(\K)$ (experiments using greedy selection
    instead of a k-DPP are in Appendix \ref{a:greedy}), for a toy dataset
    ($\kappa$ is the condition number) and two
    Libsvm datasets ($\sigma$ is the RBF parameter). Error bars show three times the standard
    error  of the mean over 1000 trials. Bottom three plots show the spectral decay for  the
  top $40$ eigenvalues  of each kernel $\K$. Note that the
  peaks in the approximation factor align with the drops in
  the spectrum.}
\vspace{-2mm}
  \label{f:rbf}
\end{figure*}

Examining the toy dataset (Figure~\ref{f:rbf}, left), it is apparent
that a larger drop in spectrum leads to a sharper peak in the
approximation factor as a function of the subset size~$k$, whereas a flat spectrum results in  
the approximation factor being close to $1$. A similar trend is observed for
dataset \emph{bodyfat} (Figure~\ref{f:rbf}, center), where large
parameter $\sigma$ results in a peak that is aligned with a spectrum
drop, while 
decreasing $\sigma$ makes the spectrum flatter and the factor closer to
$1$. Finally, dataset \emph{eunite2001} (Figure~\ref{f:rbf}, right) exhibits a full
multiple-descent curve with up to three peaks for large values of
$\sigma$, and the peaks are once again aligned with the spectrum
drops. Decreasing $\sigma$ gradually eliminates the peaks, 
resulting in a uniformly small approximation factor. Thus, both of our theoretical
claims can easily be verified on this dataset simply by adjusting the RBF parameter.

While the right choice of the parameter $\sigma$ ultimately depends on
the downstream machine learning task, it has been observed that
varying $\sigma$ has a pronounced effect on the spectral properties of
the kernel matrix, \citep[see,
e.g.,][]{revisiting-nystrom,LBM16_TR,wang2019block}.  
The main takeaway from our results here is that, depending on the
structure of the problem, we may end up in the regime where the
Nystr\"om approximation factor exhibits a multiple-descent curve
(e.g., due to a hierarchical nature of the data) or in the regime
where it is relatively flat.

\section{Conclusions and open problems}
\label{s:conclusions}
%\vspace{-3mm}
We derived new guarantees for the Column Subset Selection Problem (CSSP) and
the Nystr\"om method, going beyond worst-case analysis by exploiting the
structural properties of a dataset, e.g., when the spectrum exhibits a
known rate of decay. Our upper and lower bounds
for the CSSP/Nystr\"om approximation factor reveal an intruiguing phenomenon we
call the multiple-descent curve: the approximation factor can exhibit a highly
non-monotonic behavior as a function of $k$, with multiple
peaks and valleys. These observations suggest a possible connection to the
double descent curve exhibited by
the generalization error of many machine learning models (see
Appendix~\ref{s:related-work} for a more in depth discussion of
similarities and differences between the two phenomena).

Our analysis technique relies on converting an error bound
from random-size DPPs to fixed-size k-DPPs, which results in an
additional constant factor of $(1+2\epsilon)^2$ in Theorem~\ref{t:upper}.
We put forward a
conjecture which would eliminate this factor from
Theorem~\ref{t:upper} and is of independent interest to the study of 
elementary symmetric polynomials, a classical topic in combinatorics
\citep{HLP-Inequalities}. 
\begin{conjecture}\label{c:convex}
  The following function is \underline{convex} with respect to $k\in[n]$ for any
$\lambda_1,...,\lambda_n> 0$:\vspace{-.5mm}
 \[f(k)=(k+1)\frac{\sum_{S:|S|=k+1}\prod_{i\in S}\lambda_i}{\sum_{S:|S|=k}\prod_{i\in S}\lambda_i}.\]
\end{conjecture} %\vspace{-1.5mm}

\citet{pca-volume-sampling} showed that if $S\sim
k$-$\DPP(\A^\top\A)$ and $\lambda_i$ are the eigenvalues of
$\A^\top\A$, then $\E[\Er_\A(S)]=f(k)$. If $f(k)$ is convex then
Jensen's inequality implies:
\begin{align*}
  \E[\Er_\A(S)]\leq \E[\Er_\A(S')]\quad\text{for }S'\!\sim\DPP(\tfrac1{\alpha_k}\A^\top\A),
\end{align*}
where $\alpha_k$ is chosen so that $\E[|S'|]=k$. This would allow us to use
the bound from Lemma \ref{l:upper} directly on a k-DPP
without relying on the concentration argument of
Lemma~\ref{l:concentration}, thereby improving the bounds in
Theorems \ref{t:upper} and \ref{t:decay}.

\subsection*{Acknowledgments}
We would like to acknowledge DARPA, IARPA (contract W911NF20C0035),
NSF, and ONR via its BRC on RandNLA for providing partial support of
this work.  Our conclusions do not necessarily reflect the position or
the policy of our sponsors, and no official endorsement should be
inferred.

\ifisarxiv\else
\section*{Broader Impact}
Our study offers a deeper theoretical understanding of the Column
Subset Selection Problem. The nature of our work is primarily
theoretical, with direct applications to feature selection and kernel
approximation, as we noted in Section \ref{s:intro}. The primary reason for
feature selection as a method for approximating a given matrix, as opposed to a low rank
approximation using an SVD, is interpretability, which is crucial in
many scientific disciplines. Our analysis shows that in many practical settings, feature
selection performs almost as well as SVD at approximating a matrix. As
such, our work makes a stronger case for feature selection, wherever
applicable, for the sake of interpretability. We also hope our work
motivates further research into a fine grained analysis to quantify if
machine learning problems are really as hard as worst-case bounds
suggest them to be.
\fi

\bibliographystyle{sty/icml2020/icml2020}
\bibliography{pap}

  \newpage
  \appendix
  \onecolumn

\section{Additional related works}
\label{s:related-work}
\underline{The Column Subset Selection Problem} is one of the most classical tasks
in matrix approximation~\citep{BoutsidisMD08}. The original version of the problem compares the
projection error of a subset of size $k$ to the best rank $k$
approximation error. The techniques used for finding good subsets have
included many randomized methods
\citep{pca-volume-sampling,BoutsidisMD08,dpp-for-cssp,boutsidis2014optimal}, as well as deterministic
methods \citep{GuE96}. Variants of these algorithms have also been extended to more general losses~\citep{chierichetti17algorithms, khanna2017scalable, elenberg2016restricted}. Later on, most works have relaxed the problem
formulation by allowing the number of selected columns $|S|$ to exceed
the rank $k$. These approaches include deterministic sparsification
based algorithms \citep{near-optimal-columns}, greedy selection
\citep[e.g.,][]{Bhaskara2016GreedyCSS} and randomized methods
\citep[e.g.,][]{Drineas08CUR,more-efficient-volume-sampling,Paul2015CSS}. Note that
we study the \emph{original} version of the CSSP (i.e., without the
relaxation), where the number of columns $|S|$ must be equal to the
rank $k$.

\underline{The Nystr\"om method} has been given significant attention
independently of the CSSP. The guarantees most comparable to our
setting are due to \citet{belabbas-wolfe09}, who show the approximation
factor $k+1$ for the trace norm error. Many recent works allow the subset size $|S|$ to exceed
the target rank $k$, which enables the use of i.i.d.~sampling
techniques such as leverage scores \citep{revisiting-nystrom}
and ridge leverage scores \citep{ridge-leverage-scores,Musco17Nystrom}.
In addition to the trace norm error, these works consider other types
of guarantees, e.g., based on spectral and Frobenius norms, which are not 
as readily comparable to the CSSP error bounds.

\underline{The double descent curve} was introduced by \citet{BHMM19} to explain
the remarkable success of machine learning models which generalize
well despite having more parameters than training data. This research 
has been primarily motivated by the success of deep neural networks,
but double descent has also been observed in linear regression
\citep{belkin2019two,BLLT19_TR,surrogate-design} and other learning models.
Double descent is typically presented by plotting
the absolute generalization error as a function of the number of parameters used in the
learning model, although \citet{double-descent-condition} and
\citet{zhenyu2020double} showed that the 
behavior of generalization error is merely an artifact of the phase
transitions in the spectral properties of random
matrices. Importantly, although the descent curves we obtain are 
reminiscent of the above works, our setting is different in that it is
a \emph{deterministic} combinatorial 
optimization problem for \emph{relative} error. In particular,
Corollary~\ref{c:multiple-descent} shows that our multiple-descent
curve can 
occur as a purely deterministic property of the optimal CSSP
solution. Despite the differences, there are certain similarities
between the two settings, namely (a) the notion of stable rank we use
matches the one used by~\citet{BLLT19_TR}, (b) the peaks in both the
settings are closely aligned -- these peaks coincide with the size $k$
crossing the corresponding sharp drops in the respective spectra, (c)
the analysis of bias of the minimum norm solution for double descent
for linear regression under DPP sampling obtained
by~\citet{surrogate-design} leads to expressions very similar to ours
for the CSSP error for DPP sampling.

\underline{Determinantal point processes} have been shown to provide
near-optimal guarantees not only for the CSSP but also other tasks in
numerical linear algebra, such as least squares regression
\citep[e.g.,][]{avron-boutsidis13,unbiased-estimates-journal,minimax-experimental-design}.
They are also used in recommender systems, stochastic optimization and other
tasks in machine learning \citep[for a review,
see][]{dpps-in-randnla,dpp-ml}. Efficient algorithms for sampling from these
distributions have been proposed both in the CSSP setting \citep[i.e.,
given matrix $\A$; see, e.g.,][]{efficient-volume-sampling,dpp-intermediate}
and in the Nystr\"om setting \citep[i.e., given kernel $\K$;
see, e.g.,][]{rayleigh-mcmc,dpp-sublinear}. The term ``cardinality constrained
DPP'' (also known as a ``k-DPP'' or ``volume sampling'') was introduced by \citet{k-dpp} to
differentiate from standard DPPs which have random cardinality.
Our proofs rely in part on converting DPP bounds to
k-DPP bounds via a refinement of the
concentration of measure argument used by
\citet{bayesian-experimental-design}.

\underline{Beyond worst-case analysis} of algorithms is crucial to understanding the often-noticed gap between practical performance and theoretical guarantees of these algorithms. However, there have been limited number of works in machine learning that undertake finer-grained studies for beyond worst-case analyses. We refer to~\citet{roughgarden2019beyond} for a recent survey of such studies for a few problems in machine learning.~\citet{Mah12} takes an alternative view and studies implicit statistical properties of worst case algorithms.

\section{Determinantal point processes}
\label{s:dpp}

Since our main results rely on randomized subset selection via
determinantal point processes (DPPs), we provide a brief overview of
the relevant aspects of this class of distributions.
First introduced by \citet{Macchi1975}, a determinantal point process is a
probability distribution over subsets $S \subseteq[n]$, where we use $[n]$
to denote the set $\{1,...,n\}$. The relative probability of a subset being drawn is 
governed by a positive semidefinite (p.s.d.) matrix $\K \in \R^{n\times n}$,
as stated in the definition below, where we use $\K_{S,S}$ to denote
the $|S|\times |S|$ submatrix of $\K$ with rows and columns indexed
by $S$.
\begin{definition}\label{d:dpp}
  For an $n\times n$ p.s.d.~matrix $\K$, define $S\sim
  \DPP(\K)$ as a distribution 
  over all subsets $S\subseteq [n]$ so
  that
  \[\Pr(S)=\frac{\det(\K_{S,S})}{\det(\I+\K)}.\]
A restriction to subsets of size $k$ is denoted as $k$-$\DPP(\K)$.
\end{definition}
DPPs can be used to introduce diversity in the selected set or to
model the preference for selecting dissimilar items, where the
similarity is stated by the kernel matrix $\K$. DPPs are
commonly used in many machine learning applications where these
properties are desired, e.g.,~recommender systems~\citep{Warlop2019}, model
interpretation~\citep{kim:2016MMD}, text
and video summarization~\citep{dpp-video}, and others
\citep{dpp-ml}. They have also played an important role in randomized
numerical linear algebra \citep{dpps-in-randnla}.

Given a p.s.d.~matrix $\K \in \R^{n \times n}$ with eigenvalues
$\lambda_1,...\, \lambda_n$, the size of the set $S \sim \DPP(\K)$
is distributed as a Poisson binomial random variable, namely, the
number of successes in $n$ Bernoulli random trials where the
probability of success in the $i$th trial is given by 
$\frac{\lambda_i}{\lambda_i+1}$. This leads to a simple
expression for the expected subset size:
\begin{align}
\E[|S|] = \sum_i \frac{\lambda_i}{\lambda_i + 1} = \tr(\K(\I + \K)^{-1}).\label{eq:size}
\end{align}
Note that if $S\sim\DPP(\frac1\alpha\K)$, where $\alpha>0$, then
$\Pr(S)$ is proportional to $\alpha^{-|S|}\det(\K_{S,S})$, so rescaling the kernel
by a scalar only affects the distribution of the subset sizes, giving
us a way to set the expected size to a desired value (larger $\alpha$
means smaller expected size).
Nevertheless, it is still often preferrable to restrict the size of
$S$ to a fixed $k$, obtaining a $k$-$\DPP(\K)$ \citep{k-dpp}.

Both DPPs and k-DPPs can be sampled efficiently, with some of the
first algorithms provided by
\citet{dpp-independence},
\citet{efficient-volume-sampling}, \citet{k-dpp} and others. These
approaches rely on an eigendecomposition of the kernel $\K$, at the cost of $O(n^3)$. When
$\K=\A^\top\A$, as in the CSSP, and the dimensions satisfy $m\ll n$, then this can be improved
to $O(nm^2)$. More recently, algorithms that avoid computing the
eigendecomposition have been proposed
\citep{dpp-intermediate,dpp-sublinear,alpha-dpp,rayleigh-mcmc},
resulting in running times of $\Ot(n)$ when given matrix $\K$ and
$\Ot(nm)$ for matrix $\A$, assuming small desired subset size.
See \citet{dppy} for an efficient Python implementation of DPP sampling.

The key property of DPPs that enables our analysis is a
formula for the expected value of the random matrix that is the orthogonal projection onto the
subspace spanned by vectors selected by $\DPP(\A^\top\A)$.
In the special case when $\A$ is a square full rank matrix, the
following result can be derived as a corollary of Theorem 1 by
\citet{randomized-newton}, and a variant for DPPs over
continuous domains can be found as Lemma~8 of
\citet{surrogate-design}. For completeness, we also provide a proof in
Appendix~\ref{a:expectration-proj}.
\begin{lemma}\label{l:expectation_proj}
  For any $\A$ and $S\subseteq [n]$, let $\P_S$ be the  
projection onto the $\mathrm{span}\{\a_i:i\in S\}$. If $S\sim \DPP(\A^\top\A)$, then
\begin{align*}
  \E[\P_S] & = \A(\I+\A^\top\A)^{-1}\A^\top.
\end{align*}
\end{lemma}
Lemma~\ref{l:expectation_proj} implies a simple closed form expression for
the expected error in the CSSP. Here, we use a
rescaling parameter $\alpha>0$ for controlling
the distribution of the subset sizes. Note that it is crucial that we
are using a DPP with random subset size, because the corresponding
expression for the expected error of the fixed size k-DPP is
combinatorial, and therefore much harder to work with.
\begin{lemma}\label{l:expected-error}
  For any $\alpha>0$, if $S\sim
  \DPP(\frac1\alpha\A^\top\A)$, then
  \begin{align*}
    \E\big[\Er_\A(S)\big] =
    \tr\big(\A\A^\top(\I+\tfrac1\alpha\A\A^\top)^{-1}\big) = \E[|S|]\cdot\alpha.
  \end{align*}
\end{lemma}
\begin{proof}
Using Lemma~\ref{l:expectation_proj}, the expected loss is given by:
	\begin{align*}
          \E\big[\Er_\A(S)\big]
          &= \E\big[ \|(\I-\P_S)\A\|_F^2\big]
            = \tr(\A\A^\top\E[\I-\P_S]) \\
          &=\tr\big(\A\A^\top
            (\I-\tfrac1\alpha\A(\I+\tfrac1\alpha\A^\top\A)^{-1}\A^\top)\big)\\
          &\overset{(*)}{=}\tr\big(\A\A^\top(\I+\tfrac1\alpha\A\A^\top)^{-1}\big),
	\end{align*}
where $(*)$ follows from the matrix identity
$(\I+\A\A^\top)^{-1}=\I-\A(\I+\A^\top\A)^{-1}\A^\top$. 
\end{proof}

\section{Proof of Lemma \ref{l:expectation_proj}}
\label{a:expectration-proj}
  We will use the following standard determinantal summation identity
  \citep[see Theorem 2.1 in][]{dpp-ml} which corresponds to computing the
normalization constant $\det(\I+\K)$ for a DPP.
  \begin{lemma}\label{l:normalization}
    For any $n\times n$ matrix $\K$, we have
    \begin{align*}
      \det(\I+\K) = \sum_{S\subseteq[n]}\det(\K_{S,S}).
      \end{align*}
    \end{lemma}
We now proceed with the proof of Lemma \ref{l:expectation_proj}
(restated below for convenience).
  \begin{replemma}{l:expectation_proj}
  For any $\A$ and $S\subseteq [n]$, let $\P_S$ denote the  
projection onto the $\mathrm{span}\{\a_i:i\in S\}$. If $S\sim \DPP(\A^\top\A)$, then
\begin{align*}
  \E[\P_S] & = \A(\I+\A^\top\A)^{-1}\A^\top.
\end{align*}
\end{replemma}
\begin{proof}
Fix $m$ as the column dimension of $\A$ and let $\A_S$ denote the
submatrix of $\A$ consisting of the columns 
indexed by $S$. We have $\P_S=\A_S(\K_{S,S})^\dagger\A_S$, where $^\dagger$
denotes the Moore-Penrose inverse and $\K=\A^\top\A$. Let
$\v\in\R^m$ be an arbitrary vector. When $\K_{S,S}$ is
invertible, then a standard determinantal 
identity states that:
  \begin{align*}
\det(\K_{S,S})\v^\top\P_S\v
    =\det(\K_{S,S})\v^\top\A_S\K_{S,S}^{-1}\A_S^\top\v =
    \det(\K_{S,S}+\A_S^\top\v\v^\top\A_S)-\det(\K_{S,S}). 
  \end{align*}
When $\K_{S,S}$ is not invertible then
$\det(\K_{S,S})=\det(\K_{S,S}+\A_S^\top\v\v^\top\A_S)=0$, because the
rank of $\K_{S,S}+\A_S^\top\v\v^\top\A_S=\A_S^\top(\I+\v\v^\top)\A_S$
cannot be higher than the rank of $\K_{S,S}=\A_S^\top\A_S$. Thus,
\begin{align*}
  \det(\I+\K)\v^\top\E[\P_S]\v
  &= \sum_{S\subseteq[n]:\,\det(\K_{S,S})>0}\det(\K_{S,S})
\v^\top\A_S\K_{S,S}^{-1}\A_S^\top\v\\
  &=\sum_{S\subseteq[n]}\det(\K_{S,S}+\A_S^\top\v\v^\top\A_S)-\det(\K_{S,S})\\
  &=\sum_{S\subseteq[n]}\det\!\big([\K+\A^\top\v\v^\top\A]_{S,S}\big)-\sum_{S\subseteq[n]}\det(\K_{S,S})\\
  &\overset{(*)}{=}\det(\I + \K + \A^\top\v\v^\top\A) - \det(\I+\K)\\
  &=\det(\I+\K)\v^\top\A(\I+\K)^{-1}\A^\top\v,
\end{align*}
where $(*)$ involves two applications of Lemma \ref{l:normalization}. Since the above
calculation holds for arbitrary vector $\v$, the claim follows.
\end{proof}

\section{Proofs omitted from Section \ref{s:upper}}
\begin{replemma}{l:upper}
For any $\A$, $0\leq\epsilon<1$ and $s<k<t_s$, where $t_s=s+\sr_s(\A)$,
suppose that $S\sim\DPP(\frac1\alpha\A^\top\A)$ for
$\alpha=\frac{\gamma_s(k)\opt}{(1-\epsilon)(k-s)}$ and
$\gamma_s(k)=\sqrt{1+\frac{2(k-s)}{t_s-k}\,}$. Then: 
\begin{align*}
  \frac{\E\big[\Er_\A(S)\big]}{\opt}\leq \frac{\Phi_s(k)}{1-\epsilon}
  \quad\text{and}\quad \E[|S|]\leq k-\epsilon\,\frac{k-s}{\gamma_s(k)},
\end{align*}
where $\Phi_s(k) =\big(1+\frac{s}{k-s}\big)\,\gamma_s(k)$.
 \end{replemma}
\begin{proof}
  Let $\lambda_1\geq \lambda_2\geq ...$ be the eigenvalues of
  $\A^\top\A$. Note that scaling the matrix $\A$ by any constant $c$ and
  scaling $\alpha$ by $c^2$ preserves the distribution of $S$ as well
  as the approximation ratio, so without loss of generality, assume
  that $\lambda_{s+1}=1$. Furthermore, using the shorthands
    $l=k-s$ and $r=\sr_s(\A)$, we have $t_s-k=r-l$ and so
    $\gamma_s(k)=\sqrt{\tfrac{r+l}{r-l}}$. We now lower bound the optimum as follows:
  \begin{align*}
\opt = \sum_{j>k}\lambda_j = \sr_s(\A) - \!\sum_{j=s+1}^k\!\lambda_j\
    \geq\  r-l.
  \end{align*}
  We will next define an alternate sequence of eigenvalues which is in
  some sense ``worst-case'', by shifting the spectral mass away from
  the tail. 
  Let $\lambda_{s+1}'=...=\lambda_{k}'=1$, and for $i>k$ set
    $\lambda_i'=\beta\lambda_i$, where $\beta =
    \frac{r-l}{\opt}\leq 1$. Additionally, define:
\begin{align}\alpha'&=\beta\alpha=\frac{\gamma_s(k)(r-l)}{(1-\epsilon)l} =
  \frac{\sqrt{r^2-l^2}}{(1-\epsilon)l},\nonumber\\
  \alpha''&=(1-\epsilon)\frac{\sqrt{r+l}+\sqrt{r-l}}{2\sqrt{r+l}}\alpha'
=\frac{(\sqrt{r+l}+\sqrt{r-l})\sqrt{r-l}}{r+l-(r-l)}=\frac{\sqrt{r-l}}{\sqrt{r+l}-\sqrt{r-l}}.
            \label{eq:alpha}
  \end{align}
and note that $\alpha''\leq\alpha'\leq \alpha$. Moreover, for $s+1\leq i\leq k$,
we let $\alpha_i'=\alpha''$, while for $i>k$ we set $\alpha_i'=\alpha'$.
We proceed to bound the expected subset size $\E[|S|]$ by converting
all the eigenvalues from $\lambda_i$ to $\lambda_i'$ and $\alpha$ to
$\alpha_i'$, which will allow us to easily bound the entire expression:
  \begin{align}
    \E[|S|] = \sum_{i}\frac{\lambda_i}{\lambda_i+\alpha} 
\leq s + \sum_{i=s+1}^{k}\frac{\lambda_i}{\lambda_i+\alpha_i'}
    + \sum_{i>k}\frac{\beta\lambda_i}{\beta\lambda_i+\beta\alpha}
    \leq s + \sum_{i=s+1}^k  \frac{\lambda_i'}{\lambda_i'+\alpha''} +
    \sum_{i>k}\frac{\lambda_i'}{\lambda_i'+\alpha'}.\label{eq:converting}
  \end{align}
  We bound each of the two sums separately starting with the first one:
  \begin{align}
    \sum_{i=s+1}^k  \frac{\lambda_i'}{\lambda_i'+\alpha''}
    =\frac l{1+\alpha''} = l- \frac {l}{1+\frac1{\alpha''}} =
    l-\frac{l}{1+\frac{\sqrt{r+l}-\sqrt{r-l}}{\sqrt{r-l}}} = l - \frac{l\sqrt{r-l}}{\sqrt{r+l}}.\label{eq:sum1}
  \end{align}
To bound the second sum, we use  the fact that
$\sum_{i>k}\lambda_i'=\beta\,\opt=r-l$, and obtain:
  \begin{align}
    \sum_{i>k}\frac{\lambda_i'}{\lambda_i'+\alpha'}
    \le \frac1{\alpha'} \sum_{i>k} \lambda_i' 
 = \frac{r-l}{\alpha'}
=                  (1-\epsilon)\frac{l\sqrt{r-l}}{\sqrt{r+l}}.\label{eq:sum2}
  \end{align}
Combining the two sums, we conclude that $\E[|S|]\leq s+l - \epsilon\,l
\sqrt{\tfrac{r-l}{r+l}}=k-\frac{\epsilon\,l}{\gamma_s(k)}$. Finally, Lemma \ref{l:expected-error} yields:
    \begin{align*}
      \frac{\E\big[\Er_\A(S)\big]}{\opt} = \frac{\E[|S|]\cdot\alpha}{\opt}\leq
\frac{k}{k-s}\,\frac{\gamma_s(k)}{1-\epsilon} = \frac{\Phi_s(k)}{1-\epsilon},
    \end{align*}
    which concludes the proof.
  \end{proof}

\begin{replemma}{l:concentration}
Let $S$ be sampled as in Lemma \ref{l:upper} with $\epsilon\leq \frac12$.
If $s+\frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$, then
$\Pr(|S|>k) \leq \epsilon$.
\end{replemma}
\begin{proof}
 Let $p_i=\frac{\lambda_i'}{\lambda_i'+\alpha_i'}$ be the 
      Bernoulli probabilities for
  $b_i\sim\mathrm{Bernoulli}(p_i)$
  and $X =\sum_{i>s} b_i$, where $\lambda_i'$ and $\alpha_i'$ are as
  defined in the proof of Lemma \ref{l:upper}. Note that $|S|$ is
  distributed as a Poisson binomial random variable such that the success
  probability associated with the $i$th eigenvalue is upper-bounded by
  $p_i$ for each $i>s$. It follows that $\Pr(|S|\!>\!k)\leq \Pr(X\!>\!l)$, where $l=k-s$.
Moreover, letting $r=\sr_s(\A)$, in the proof of Lemma \ref{l:upper} we showed that:
  \begin{align*}
k - \E[X] \geq \epsilon\,\frac{ l\sqrt{r-l}}{\sqrt{r+l}},
  \end{align*}
  and furthermore, using the derivations in \eqref{eq:sum1} and
  \eqref{eq:sum2} together with the formula $\Var[b_i]=p_i(1-p_i)$, we obtain that:
  \begin{align*}
    \Var[X]&\leq \sum_{i=s+1}^{k}(1-p_i) + \sum_{i>k}p_i
    \leq \frac{l\sqrt{r-l}}{\sqrt{r+l}}+
      (1-\epsilon)\frac{l\sqrt{r-l}}{\sqrt{r+l}} = (2-\epsilon) \frac{l\sqrt{r-l}}{\sqrt{r+l}}.
  \end{align*}
    Using Theorem 2.6 from \citet{ChungLu2006book} with $\lambda=\epsilon\,\frac{
    l\sqrt{r-l}}{\sqrt{r+l}}$, we have:
  \begin{align*}
    \Pr(|S|>k)&\leq \Pr(X>l)\leq \Pr(X > \E[X]+\lambda)
    \leq \exp\Big(-\frac{\lambda^2}{2(\Var[X]+\lambda/3)}\Big)\\
    &\leq \exp\Big(-\frac{\lambda^2}{2(\frac{2-\epsilon}\epsilon
      \lambda+\lambda/3)}\Big)\leq\exp(-\epsilon\lambda /4)
      =\exp\Big(-\frac{\epsilon^2l\sqrt{r-l}}{4\sqrt{r+l}}\Big). 
  \end{align*}
  Note that since $7\leq l\leq r-1$, we have $l\frac{\sqrt{r-l}}{\sqrt{r+l}}\geq
\frac{l}{\sqrt{2l+1}}\geq \frac{7}{16}\sqrt l$, so by simple algebra
it follows that for $l\geq \frac 7{\epsilon^4}\ln^2\frac1\epsilon$,
we have $l\frac{\sqrt{r-l}}{\sqrt{r+l}}\geq
\frac4{\epsilon^2}\ln\frac1\epsilon$ and therefore
$\Pr(|S|>k)\leq \epsilon$. 
\end{proof}
\begin{replemma}{l:monotonic}
  For any $\A\in\R^{m\times n}$, $k\in[n]$ and $\alpha>0$, if
  $S\sim\DPP(\frac1\alpha\A^\top\A)$ and $S'\sim k$-$\DPP(\A^\top\A)$, then
  \begin{align*}
    \E\big[\Er_\A(S')\big] \leq \E\big[\Er_\A(S)\mid |S|\leq k\big].
  \end{align*}
\end{replemma}
\begin{proof}
Let $\lambda_1\geq\lambda_2\geq ...$ denote the eigenvalues of
$\A^\top\A$ and let $e_k$ be the $k$th elementary symmetric polynomial
of $\A$:
\begin{align*}
e_k = \sum_{T:|T|=k}\det(\A_T^\top\A_T) = \sum_{T:|T|=k}\prod_{i\in T}\lambda_i.
\end{align*}
Also let $\bar{e}_k=e_k/{n\choose k}$ denote the $k$th
elementary symmetric mean. Newton's inequalities imply that:
\begin{align*}
1\geq \frac{\bar e_{k-1}\bar e_{k+1}}{\bar e_k^2} =
  \frac{e_{k-1}e_{k+1}}{e_k^2}\,\frac{{n\choose k}}{{n\choose
  k-1}}\frac{{n\choose k}} {{n\choose k+1}} =
  \frac{e_{k-1}e_{k+1}}{e_k^2}\,\frac{n+1-k}{k}\,\frac{k+1}{n-k}. 
\end{align*}
The results of
\citet{pca-volume-sampling} and \citet{more-efficient-volume-sampling} establish
that $\E[\Er_\A(S)\mid |S|=k] = (k+1)\frac{e_{k+1}}{e_k}$, so it
follows that:
\begin{align}
  \frac{\E[\Er_\A(S)\mid |S|=k]}{\E[\Er_\A(S)\mid|S|=k-1]}
  = \frac{k+1}{k}\,\frac{e_{k+1}e_{k-1}}{e_k^2}\leq
  \frac{n-k}{n+1-k}\leq 1.\label{eq:decreasing-error}
\end{align}
Finally, note that $\E[\Er_\A(S)\mid|S|\leq k]$ is a weighted average
of components $\E[\Er_\A(S)\mid|S|= s]$ for $s\in[k]$, and
\eqref{eq:decreasing-error} implies that the smallest of those components
is associated with $s=k$. Since the weighted average is lower bounded by the
smallest component, this completes the proof.
  \end{proof}

\section{Proof of Theorem~\ref{t:decay}}\label{a:decay}
Before showing Theorem \ref{t:decay}, we give an additional lemma
which covers the corner case of the theorem when $k$ is close to $n$.
  \begin{lemma}\label{l:upper_conditionnum}
Given $\A\in\R^{m\times n}$ and $s< k< n$, let
$\lambda_1\geq...\geq \lambda_n>0$ be the eigenvalues of $\A^\top\A$.
If $S\sim k$-$\DPP(\A^\top\A)$ and we let $b=\min\{k-s,n-k\}$, then
for any $0<\epsilon\leq \frac12$ we have
    \begin{align*}
\frac{\E[\Er_\A(S)]}{\opt} \leq \big(1-\ee^{-\frac{\epsilon^2b}{10}}\big)^{-1}(1-\epsilon)^{-1}\Psi_{s}(k),
    \end{align*}
    where $\Psi_s(k) = \frac{\lambda_{s+1}}{\lambda_n}\,\big(1+\frac s{k-s}\big)$.
    \end{lemma}
    \begin{proof}
      Let $\alpha=\frac{\lambda_{s+1}}{(1-\epsilon)\lambda_n}\frac{\opt}{k-s}$.
Note that $\opt=\sum_{i>k}\lambda_i\geq (n-k)\lambda_n$. Define
$b_i\sim\mathrm{Bernoulli}(\frac{\lambda_i}{\lambda_i+\alpha})$ and
let $X=\sum_{i>s}b_i$. We have: 
\begin{align*}
          \E[X]&= \sum_{i>s}\frac{\lambda_i}{\lambda_i+\alpha}\\
          &\leq 
            \frac{(n-s)\lambda_{s+1}}{\lambda_{s+1}+
            \frac{\lambda_{s+1}}{\lambda_n}\frac{(n-k)\lambda_n}{(1-\epsilon)(k-s)}}\\
          &= \frac{1}{\frac1{n-s} +
            \frac1{(1-\epsilon)(k-s)}\,\frac{n-k}{n-s}} \\
          &= \frac{1}{\frac1{n-s} +
            \frac{1}{(1-\epsilon)(k-s)}(1-\frac{k-s}{n-s})}\\
          &=\frac{1}{\frac{1}{(1-\epsilon)(k-s)} -
            \frac{\epsilon}{1-\epsilon}\frac1{n-s}}\\
          &=\frac{1-\epsilon}{\frac1{k-s}-\frac\epsilon{n-s}}.
\end{align*}
Let $S'\sim\DPP(\frac1\alpha\A^\top\A)$. It follows that
\begin{align*}
  k-\E[|S'|]&\geq k -(s+\E[X])\\
  & \geq (k-s) -
  \frac{1-\epsilon}{\frac1{k-s}-\frac\epsilon{n-s}}\\
&=(k-s)\bigg(1-\frac{1-\epsilon}{1-\epsilon\,\frac{k-s}{n-s}}\bigg)\\
&=(k-s)\,\frac{\epsilon - \epsilon\,
\frac{k-s}{n-s}}{1-\epsilon\,\frac{k-s}{n-s}}\\
             &\geq \epsilon\,(k-s)\bigg(1-\frac{k-s}{n-s}\bigg)\\
             &=\epsilon\cdot\frac{(k-s)(n-k)}{n-s}\\
  &\geq \frac\epsilon2\cdot\min\{k-s,n-k\}.
\end{align*}
From this, it follows that:
\begin{align*}
  \frac{\E[\Er_\A(S')]}{\opt}=\frac{\E[|S|]\cdot\alpha}{\opt}\leq
(1-\epsilon)^{-1}  \frac k{k-s}\frac{\lambda_{s+1}}{\lambda_n} =
(1-\epsilon)^{-1}\Big(1+\frac s{k-s}\Big)\frac{\lambda_{s+1}}{\lambda_n}.
\end{align*}
We now give an upper bound on $\Pr(|S'|>k)$ by considering two
cases.

\textbf{Case 1}: $k-s\leq n-k$. Then, using $\lambda=\epsilon(k-s)/2$,
we have $(k-s)-\E[X]\geq \lambda$, so using Theorem 2.4
from \citet{ChungLu2006book}, we get:
\begin{align*}
  \Pr(|S'|>k)\leq \Pr(X>k-s)\leq \Pr(X>\E[X]+\lambda)\leq
  \ee^{-\frac{\lambda^2}{2(k-s)}}
  =\ee^{-\epsilon^2(k-s)/8}.
\end{align*}
\textbf{Case 2}: $k-s>n-k$. Then, using Theorem 2.4
from \citet{ChungLu2006book} with $\lambda=k-\E[|S'|]=\frac{\epsilon(n-k)}2
+ \Delta$, where $\Delta>0$, we get:
\begin{align*}
  \Pr(|S'|>k)
  &= \Pr(n-|S'|<n-k)\\
 &\leq \exp\Big(-\frac{\lambda^2}{2\,\E[n-|S'|]}\Big) \\
  &=\exp\Big(-\frac\lambda2\,\frac{\frac\epsilon2(n-k)
    + \Delta}{n-k+\frac\epsilon2(n-k)+\Delta}\Big)\\
  &\leq\exp\Big(-\frac\lambda2\,\frac{\frac\epsilon2(n-k)}
    {n-k+\frac\epsilon2(n-k)}\Big)\\
  &=\exp\Big(-\frac{\epsilon^2(n-k)}{8(1+\epsilon/2)}\Big)\\
  &\overset{(*)}{\leq} \exp\Big(-\frac{\epsilon^2(n-k)}{10}\Big),
\end{align*}
where in $(*)$ we used the fact that $\epsilon\in(0,\frac12)$. Now,
the result follows easily by invoking Lemma \ref{l:monotonic}:
\begin{align*}
    \E\big[\Er_\A(S)\big] &\leq \E\big[\Er_\A(S')\mid|S'|\leq k\big]
  \leq \frac{\E\big[\Er_\A(S')\big]}{\Pr(|S'|\leq k)}
  \\
  &\leq \big(1-\ee^{-\frac{\epsilon^2b}{10}}\big)^{-1}(1-\epsilon)^{-1}
\frac{\lambda_{s+1}}{\lambda_n}\Big(1+\frac s{k-s}\Big)\cdot\opt,
\end{align*}
which completes the proof.
\end{proof}
Note that since $b\geq 1$, setting $\epsilon=\frac12$ in Lemma
\ref{l:upper_conditionnum} yields the following simpler (but usually
much weaker) bound:
\begin{align*}
  \frac{\E[\Er_\A(S)]}{\opt}
  \leq
  2\big(1-\ee^{-\frac1{40}}\big)^{-1}\Psi_s(k)
  \leq 82\, \Psi_s(k).
\end{align*}

\begin{reptheorem}{t:decay}
Let $\lambda_1\!\geq\!\lambda_2\!\geq\!...$ be the
eigenvalues of $\A^\top\A$. There is an absolute constant $c$ such that
for any $0\!<\!c_1\!\leq\!c_2$, with $\gamma=c_2/c_1$, if:\\[2mm]
\textnormal{1.} (\textbf{polynomial spectral decay}) $c_1i^{-p}\!\leq\!
  \lambda_i\!\leq\! c_2i^{-p}$ $\forall_i$, with $p>1$, then $S\sim
k$-$\DPP(\A^\top\A)$ satisfies
  \begin{align*}
    \frac{\E[\Er_\A(S)]}{\opt}\leq c \gamma p. 
  \end{align*}
\textnormal{2.} (\textbf{exponential spectral decay}) $c_1(1\!-\!\delta)^{i}\leq \lambda_i\leq
  c_2(1\!-\!\delta)^{i}$ $\forall_i$, with $\delta\in(0,1)$, then $S\sim
k$-$\DPP(\A^\top\A)$ satisfies
  \begin{align*}
    \frac{\E[\Er_\A(S)]}{\opt}\leq c\gamma(1+ \delta k).
  \end{align*}
 \end{reptheorem}
\begin{proof}

\textbf{  (1) Polynomial decay.} We provide the proof by splitting it into two cases.

  \textbf{Case 1(a)}: $\left(\frac{k+1}{n}\right)^{p-1} \leq \frac{1}{2} $

  We can use upper and lower integrals to bound the sum $\sum_{i\geq s}\frac{1}{i^p}$ as:
  \[ \int_{x\geq (s+1)}\frac{1}{i^p}dx \leq \sum_{i\geq
      s}\frac{1}{i^p} \leq \int_{x\geq s}\frac{1}{i^p}dx
    \implies \sum_{i=s+1}^n\frac{1}{i^p}
    \geq \frac{(s+2)^{1-p}}{p-1} -  \frac{(n+1)^{1-p}}{p-1}.\] 
  We lower bound the stable rank for $s\leq k$ using the upper/lower bounds on
  the eigenvalues and the condition for Case 1(a):
  \begin{align*}
    \sr_s(\A)&=\frac{\sum_{i=s+1}^n\lambda_i}{\lambda_{s+1}}\\
    &\geq
    \frac{c_1}{c_2}\bigg(\frac{(s+2)^{1-p}}{(p-1)(s+1)^{-p}}-
    \frac{(n+1)^{1-p}}{(p-1)(s+1)^{-p}}\bigg) \\
    &=\frac1\gamma\bigg(\frac{s+2}{p-1} \Big(1 - \frac1{s+2}\Big)^p
      -  \frac{s+1}{p-1}\Big(\frac{s+1}{n+1}\Big)^{p-1}\bigg)\\
    &\geq \frac1\gamma\bigg(\frac{s+2}{p-1} - 1
      -  \frac{s+1}{p-1}\cdot\frac12\bigg) = \frac1{2\gamma}\,\frac{s+1}{p-1}-\frac1\gamma.
  \end{align*}

  Further using $u =
  k-s$, we can call upon Theorem~\ref{t:upper} to get,  
  \begin{align*}
    \Phi_s(k) &\leq \frac{k}{u}\sqrt{1+\frac{2u}{ \sr_s-u }\,} \leq
    \frac{k}{u} + \frac{k}{  \frac{1}{2\gamma} \frac{s+1}{p-1} -
      \gamma^{-1} - u } = \frac{k}{u} + \frac{(2p-2)k}{
                \gamma^{-1}(s+1 - 2 p +2) - (2p-2)u } \\
    &\leq \frac{k}{u} +
    \frac{(2p-2 + \gamma^{-1})k}{   \gamma^{-1}(k+3 -2 p)  - (2p-2 +
    \gamma^{-1})u }
    \end{align*}

  Optimizing over $u$, we see that the minimum is reached for $u =
  \hat{u} = \frac{k+3-2p}{2\gamma(2p-2+\gamma^{-1})}$ which achieves the value
  $ \frac{4(\gamma(2p-2)+1)k}{k+3 -2 p}$ which is upper bounded by
  $\frac{12\gamma pk}{(k-2p)}$.

  We assume $k \geq \hat{u} > 60p > 60$. If not,~\citet{pca-volume-sampling} ensure an upper bound of $(k+1) \leq 60p+1 < 61p$. With $p < k/60$, we get:

  \[ \frac{12\gamma pk}{k-2p} \leq \frac{12\gamma pk}{k-k/30} =
    \frac{12\gamma p}{1-1/30} \leq \frac{360}{29}\gamma  p.  \]

  Since we assumed that $\hat u> 60$, then $k-s> \frac{7}{\epsilon^4}
  \ln^2\frac{1}{\epsilon}$ for $\epsilon = 0.5$ which means
  $(1+2\epsilon)^2 \leq 4$, which makes the approximation ratio upper
  bounded by $ \frac{1440}{29}\gamma p  $. The overall bound
  thus becomes $61 \gamma p$.

  \bigskip 
  \textbf{Case 1(b)}: $\left(\frac{k+1}{n}\right)^{p-1} >\frac12$

  From Lemma~\ref{l:upper_conditionnum}, we know that the
  approximation ratio is upper bounded by constant factor times
  $\Psi_s(k)=\frac{\lambda_{s+1}}{\lambda_n}\,\frac{k}{k-s}$. Consider,

  \begin{align*}
    \Psi_s(k) &= \frac{\lambda_{s+1}}{\lambda_n}\frac{k}{k-s}\leq
      \gamma\frac{n^p}{(s+1)^p}\frac{k}{k-s} =\gamma
      \left(\frac{n}{k+1}\right)^{p-1}
      \frac{k+1}{n}\frac{{(k+1)^p}}{{(s+1)^p}} \frac{k}{k-s}
    \leq 2\gamma\left(\frac{k+1}{s+1}\right)^p \frac{k}{k-s},
  \end{align*}

  which holds true for all $s \leq k$, and is optimized for $s=
  \hat{s}=\frac{pk-1}{p+1}$. We get that the approximation
  ratio is bounded as:

  \[\Psi_s(k) \leq
    \gamma\frac{k(p+1)}{k+1}\left(\frac{p+1}{p}\right)^p \leq
    e\gamma(p+1)\leq 2e\gamma p. \]

  Combining in the factor based on $\epsilon$ in
  Lemma~\ref{l:upper_conditionnum}, we get an upper bound of
  $164e\gamma p$ that is larger than the bound obtained in the case
  1(a) above and hence covers all the subcases.

\textbf{  (2) Exponential decay.}

We  first lower bound the stable rank of $\A$ of order $s$:
  \[ \sr_s(\A) = \sum_{j>s} \lambda_j/\lambda_{s+1} \geq
    \frac{c_1(1-(1-\delta)^{n-s})/\delta }{c_2} =
    \frac{1-(1-\delta)^{n-s}}{\gamma\delta}. 
  \]

  We present the proof by considering two subcases separately : when
  $k\leq n-  \frac{\ln 2}{\delta}$ and $k>n-  \frac{\ln 2}{\delta}$.

  \textbf{Case 2(a):}  $k\leq n-  \frac{\ln 2}{\delta}$. From the
  assumption, letting $s\leq k$  we have
  \begin{align*}
& s \leq n - \frac{\ln 2}{\delta}  \\
    \implies  & s \leq n - \frac{\ln 2}{\ln \frac{1}{1-\delta}} \\
    \implies  & (n-s)\ln \frac{1}{1-\delta} \geq \ln 2\\
    \iff & 1 - (1-\delta)^{n-s} \geq \frac{1}{2}   \\
    \implies & \sr_s(\K) \geq \frac{1}{2\gamma \delta},
  \end{align*}

  where the second inequality follows because $\frac{x}{1+x} \leq \ln (1+x)$ with $x = \delta/(1-\delta)$.

  We will use $u=k-s$. From Theorem~\ref{t:upper}, using $\sr_s \geq
  \frac{1}{2\gamma\delta}$ we have the following upper bound:   
  \[ \Phi_s(k) \leq \frac{k}{u} \left(1+ \frac{2\gamma\delta u}{1 -
        2\gamma\delta u}\right)  = \frac{k}{u}\cdot
    \frac{1}{1-2\gamma\delta u}.\]
  RHS is minimized for $\hat{u}=\frac{1}{4\gamma\delta}$. We let
  $\epsilon=0.5$ and assume that
  $\hat{u}\geq 60$ which is bigger than $\frac{7}{\epsilon^4} \ln^2 \frac{1}{\epsilon}$.
If not, then $\delta\geq\frac4{60\gamma}>\frac1\gamma$ and
the  worst-case bound of \citet{pca-volume-sampling} ensures that the
approximation factor is no more than $k+1 \leq \gamma(1+\frac1\gamma
k)\leq \gamma(1+\delta k)$. By a similar argument we can assume that
$k\geq 60$.

If  $k  \leq \hat{u}$, in this case we can set $s=0$, i.e., $u = k$, obtaining $\Phi_s(k) \leq \frac{1}{1- 2\gamma\delta k}\leq 2$. And so the
approximation ratio is bounded by $(1+2\epsilon)^2 \cdot 2 \leq 8$. On the other
hand, if $k > \hat{u}$, we can set $u=\hat{u}$, which implies
$\Phi_s(k) \leq 8\gamma\delta k$, and so the approximation ratio is
bounded by $ 32
\gamma\delta k$. The overall bound is thus $61\gamma(1+\delta k)$
covering all possible subcases.

  \bigskip
  \textbf{Case 2(b)}: $k > n - \frac{\ln 2}{\delta}$. We make use of
  Lemma~\ref{l:upper_conditionnum} for the case when $k$ is close to
  $n$. The approximation guarantee uses:

  \[\Psi_s(k) = \frac{\lambda_{s+1}}{\lambda_n} \frac{k}{k-s},\]

where $s <k$. For our bound, we choose $s = \lfloor k- \frac{\ln
  2}{\delta}\rfloor$. This implies that $n-s < \frac{2\ln
  2}{\delta}+1=\frac{\delta+\ln4}{\delta}$. It
follows that
  \begin{align*}
    & \frac{\lambda_{s+1}}{\lambda_n}  \leq
     \frac{ \gamma}{(1-\delta)^{n-s}} \leq
      \frac{\gamma}{(1-\delta)^{(\delta+\ln 4)/\delta}} =
      \gamma\left[(1-\delta)^{-\frac{1}{\delta}}\right] ^ {\delta+\ln 4} \leq
      \gamma e^{\frac{\delta+\ln 4}{1-\delta}}. 
  \end{align*}
  If $\delta\geq \frac1{20}$, then the worst-case result of
  \citet{pca-volume-sampling} suffices to show that the approximation ratio is
  bounded by $k+1\leq 20(1+\delta k)$, so assume that
  $\delta<\frac1{20}$. Then we have $\ee^{\frac{\delta+\ln 4}{1-\delta}}<5.$
  Combining this with the fact that $\frac{k}{k-s} \leq \frac{\delta k}{\ln 2}$,
we obtain:
  \[\Phi_s(k) \leq \frac{5\gamma\delta k}{\ln 2}. \]

  Combining with factor based on $\epsilon$ in
  Lemma~\ref{l:upper_conditionnum}, we get $82\cdot
  \frac{5\gamma\delta k}{\ln 2}$. Thus, the bound of $\frac{82\cdot
    5}{\ln 2}\gamma(1+\delta k)$ holds in all cases, completing the proof.
\end{proof}

\section{Proof of Lemma \ref{l:lower}}
\begin{replemma}{l:lower}
Fix $\delta\in(0,1)$ and consider unit vectors
$\a_{i,j}\in\R^m$ in general position, where $i\in[t]$, $j\in[l_i]$, such that
$\sum_j\a_{i,j}=0$ for each $i$, and for any $i,j,i',j'$, if $i\ne i'$
then $\a_{i,j}$ is orthogonal to $\a_{i'\!,j'}$. Also, let unit vectors
$\{\v_i\}_{i\in[t]}$ be orthogonal to  each other and to all
$\a_{i,j}$. There are positive scalars $\alpha_{i},\beta_i$ for $ i \in [t]$
such that matrix $\A$ with columns 
$\alpha_{i}\a_{i,j}+\beta_i\v_i$ over all $i$ and $j$ satisfies: 
\begin{align*}
  \min_{|S|=k_i}\frac{\Er_\A(S)}{\mathrm{OPT}_{k_i}}\geq
  (1-\delta)l_i,\quad\text{for }k_i=l_1+...+l_i-1.
\end{align*}
 \end{replemma}

\begin{proof}
Say $\Abh_i$ is the matrix obtained
  by stacking all the $\a_{i,j}$ and let $\lambda_{i,1}\geq
  \lambda_{i,2}\geq ...\geq \lambda_{i,l_i-1}$ denote the non-zero eigenvalues of
  $\Abh_i^\top\Abh_i$. We write $\tilde\a_{i,j} = 
  \alpha_i\a_{i,j}+\beta_i \v_i$ and note that for each $i$,
  $\one_{l_i}$ is an eigenvector of $\Abh_i^\top\Abh_i$ with
  eigenvalue $0$. Further, $\A^\top\A$ is a block-diagonal matrix
with blocks $\B_i=\alpha_i^2\Abh_i^\top\Abh_i  + \beta_i^2 \one_{l_i} \one_{l_i}^\top$: 
  \begin{align*}
    \A^\top \A = \begin{bmatrix}
\B_1&\zero
      &\zero\\
      \zero &\ddots &\zero\\
      \zero &\zero &\B_t
      \end{bmatrix}
  \end{align*}
  Therefore, the eigenvalues of $\A^\top\A$ are
  $\alpha_1^2\lambda_{1,1},...,\alpha_1^2\lambda_{1,l_1-1},\beta_1^2l_1,
  ...,
  \alpha_t^2\lambda_{t,1},...,\alpha_t^2\lambda_{t,l_t-1},\beta_t^2l_t$,
  and so we can always choose the parameters so that
  $\alpha_i\gg\beta_i\gg\alpha_{i+1}$ for each $i$, ensuring that
  these eigenvalues are in decreasing order.
Let us fix an arbitrary $c\in[t]$. From the above, it follows that for
$k_c=\big(\sum_{i\leq c}l_i\big)-1$ we have:
  \begin{align*}
  \mathrm{OPT}_{k_c} =l_c\beta_c^2 + \sum_{i>c}\tr(\B_i)=
l_c\beta_c^2 + \phi_c,  
\end{align*}
where we use $\phi_c =\sum_{i>c}\tr(\B_i) $ as a shorthand. 
Since the centroid of $\{\tilde\a_{c,1}, \ldots, \tilde\a_{c,l_c} \}$ is $\beta \v_c$, we can write 
$\tilde\a_{c, l_c} = l_c\beta\v_c - \sum_{j< l_c} \tilde\a_{c,j}$.
For selecting the set $S\subset [n]$ of size $k_c$, since
$\alpha_i\gg\alpha_{i+1}$, we can assume without loss of
generality that $S$ does not select any vectors $\tilde\a_{i,j}$ such
that $i>c$ and does not drop any such that $i<c$, and so for some
$j'\in[l_c]$ we let $S_{j'}$ be the index set such that $\P_{S_{j'}}$ is the
projection onto the span of $\Big(\bigcup_{i<c} \bigcup_j \{\tilde\a_{i,j}\}\Big) \cup
\{\tilde\a_{c,1},\ldots,\tilde\a_{c,l_c}\}\backslash\{\tilde\a_{c,j'}\}$.  We now lower bound the
squared projection error of that set:
\begin{align*}
  \Er_{\A}(S_{j'})
  &= \| \tilde\a_{c,j'}- \P_{S_{j'}} \tilde\a_{c,j'}  \|^2
    +  \sum_{i>c}\sum_{j=1}^{l_i}\| \tilde\a_{i,j} - \P_{S_{j'}}
    \tilde\a_{i,j} \|^2\\ 
  &=\bigg\|l_c\beta\v_c - \sum_{j<l_c}\tilde\a_{c,j}- \P_{S_{j'}}
    \Big(l_c \beta\v_c - \sum_{j< l_c}\tilde\a_{c,j} \Big)\bigg\|^2
    +
    \sum_{i>c}\sum_{j=1}^{l_i}\| \tilde\a_{i,j} \|^2
  \\ 
  &= l_c^2\beta^2 \|\v_c - \P_{S_{j'}}\v_c\|^2 + \phi_c \\
  & = l_c (\mathrm{OPT}_{k_c}- \phi_c) \|\v_c - \P_{S_{j'}}\v_c\|^2 + \phi_c \\
  & \geq l_c\mathrm{OPT}_{k_c} \|\v_c - \P_{S_{j'}}\v_c\|^2 - l_c \phi_c. 
\end{align*}	
Note that $\lim_{\beta\rightarrow 0}\P_{S_{j'}}\v_c = \zero$ because $\v_c$
is orthogonal to the subspace spanned by $S_{j'}$, so we can choose
$\beta_c$ small enough so that
$\|\v-\P_{S_{j'}}\v\|^2\geq 1-\frac\delta2$ for each $j'\in[l_c]$.
Furthermore, we have
\begin{align*}
  \phi_c = \sum_{i>c}\tr(\B_i) =
  \sum_{i>c}\alpha_i^2l_i+\beta_i^2l_i\leq 2\alpha_{c+1}^2\sum_{i>c}l_i,
\end{align*}
So, if we ensure that $\alpha_{c+1}^2\leq \frac{\delta}{4}
l_c\beta_c^2/(\sum_{i>c}l_i)$, then:
\begin{align*}
  l_c\phi_c \leq 2l_c\alpha_{c+1}^2\sum_{i>c}l_i\leq \frac\delta
  2\cdot l_c^2\beta^2\leq \frac\delta 2 l_c\cdot\textsc{OPT}_{k_c},
\end{align*}
which implies that $\Er_\A(S_{j'})\geq
(1-\delta)l_c\textsc{OPT}_{k_c}$. Note that all the conditions
we required on $\alpha_i$ and $\beta_i$ can be satisfied by a
sufficiently quickly decreasing sequence
$\alpha_1\gg\beta_1\gg\alpha_2\gg\beta_2\gg...\gg\alpha_t\gg\beta_t>0$,
which completes the proof.
\end{proof}

\section{Proof of Corollary~\ref{c:multiple-descent}}
\label{s:proofmultidescent}
\begin{repcorollary}{c:multiple-descent}
  For $t\in\N$ and $\delta\in(0,1)$, there is a
  sequence $k_1^l<k_1^u<k_2^l<k_2^u<...<k_t^l<k_t^u$ and
  $\A\in\R^{m\times n}$ such that for any $i\in[t]$:
  \begin{align*}
    \min_{S:|S|=k_i^l}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^l}}&\leq 1+\delta
                                 \quad\text{and }\\
    \min_{S:|S|=k_i^u}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^u}} &\geq
    (1-\delta)(k_i^u+1).
  \end{align*}
\end{repcorollary}
\begin{proof}
We will use Theorem~\ref{t:lower} to construct the matrix $\A$ using
the sequence we build below to make sure the upper and lower bounds
are satisfied. Theorem~\ref{t:lower} uses Lemma~\ref{l:lower} to
construct the matrix $\A$ which has a ``step" eigenvalue profile
i.e. there are multiple groups of eigenvalues and in each group the
eigenvalue is constant (each group corresponds to a regular simplex, see
Section \ref{s:lower}). Below we consider a single such group that
starts at $s = k^u_i$ and ends at $w=k^u_{i+1}$, and we let $k = k^l_{i+1}$,
for any $i \in \{0,\ldots,t-1\}$, with $k^u_0=0$.

Theorem~\ref{t:upper} implies that there is a set $S$ with an upper bound on the
approximation factor $\Er_\A(S)/\text{OPT}_k$ of
$(1+2\epsilon)^2\big(1+ \frac{s}{k-s}\big)\big( 1+ \frac{k-s}{t_s-
    k } \big)$. Consider the following three conditions to ensure that each
of the three terms in the above approximation factor is less than $(1+\delta_1)$ where $\delta_1 =
\delta/7$:

\begin{enumerate}
\item $\epsilon \leq \frac{ (1+\delta_1 )^{1/2} - 1}{2} \implies
  (1+2\epsilon)^2 \leq (1+\delta_1) $. Let $\tau_\epsilon =
  \frac{7}{\epsilon^4} \ln^2 \frac{1}{\epsilon}$, where $\epsilon$ is
  chosen so as to satisfy the above condition. 
\item  $k \geq \frac{s}{\delta_1} + s + \tau_\epsilon$ ensures that $(1+\frac{s}{k-s}) \leq (1+\delta_1)$ and that $k-s \geq \tau_\epsilon$.
\item $w \geq k(1+\frac{1}{\delta_1}) +1$.
  \end{enumerate}
  To see the usefulness of condition 3, note that each group of
  vectors in column set of  $\A$ constructed from Theorem \ref{t:lower} form a
  shifted regular simplex. A regular simplex has the smallest
  eigenvalue $0$ and the rest of the eigenvalues are all $(w-s)
  \alpha^2/(w-s-1)$, where $\alpha$ is the length of each of the
  $(w-s)$ vectors in the simplex. Thus, we can lower bound the stable
  rank of the shifted simplex as $\sr_s(\A) \geq
  \frac{(w-s)\alpha^2}{(w-s) \alpha^2} (w-s-1) = (w-s-1)$. From
  condition 3: 
  \[  w \geq k(1+\frac{1}{\delta_1}) +1  \implies s + \sr_s(\A) \geq k(1+\frac{1}{\delta_1}) \implies t_s \geq k(1+\frac{1}{\delta_1}) -\frac{s}{\delta_1} \implies 1+ \frac{k-s}{t_s-k}  \leq (1+\delta_1).\]

Thus if all the above three conditions are satisfied, the approximation ratio can be upper bounded by $(1+\delta_1)^3 \leq (1+\delta)$, since $\delta_1 = \delta/7$.

Similarly for the lower bound, we will need condition 4 below.
\begin{enumerate}
  \setcounter{enumi}{3}

	\item $w \geq \frac{ 2s}{\delta} +\frac{2}{\delta}$.
\end{enumerate}
 Now, we apply Theorem~\ref{t:lower} using $k_i=w$ and $k_{i-1}=s$ to get the following lower bound with $\delta_2 = \delta/2$:

\[\min_{S:|S|=w}\frac{\Er_\A(S)}{\textsc{OPT}_w}\geq (1-\delta_2) (w- s) \geq (w + 1) - \frac{\delta}{2} (w+ 1 + \frac{2s}{\delta} + \frac{2}{\delta}) \geq (1-\delta) (w+1),   \]

where the last inequality follows from condition 4. Also, observe that
we can replace conditions 3 and 4 with a single stronger condition: $w \geq k(1 + \frac{7}{\delta}) + 1+ \frac{2}{\delta}$.

We now iteratively construct the sequence that satisfies all of the
above conditions:
\begin{enumerate}
\item $k^u_0 = 0 $
\item For $1 \leq i \leq t$ 
  \begin{enumerate}
  \item $k^l_i =  \big\lceil\frac{7k^u_{i-1}}{\delta}  + k^u_{i-1} + \tau_\epsilon \big\rceil$.
  \item $k^u_i = \lceil k^l_i (1 + 7/\delta) + \frac{2}{\delta} + 1\rceil$. 
  \end{enumerate}
\end{enumerate}                
We can now use Theorem~\ref{t:lower} with subsequence
$\{k^u_i\}$ which also constructs the matrix $\A$ through
Lemma~\ref{l:lower}, to ensure that the lower bound of
$(1+\delta)(k^u_i+1)$ is satisfied for $\A$ for all $i$. We
can also use Theorem~\ref{t:upper} for the same matrix $\A$
and $k=k^l_i$ for any $i$ to ensure that the upper bound of
$(1+\delta)$ is also satisfied for any $i$. 
 \end{proof}

 \section{Empirical evaluation with greedy subset selection}
 \label{a:greedy}
 In this section, we provide a more detailed empirical evaluation to
 complement what we presented in Section~\ref{s:experiments}. Our aim here is to demonstrate
that our improved analysis of the CSSP/Nystr\"om approximation factor
can be useful in understanding the performance of not only the k-DPP
method, but also of greedy subset selection. Note that our theory
does not strictly apply to the greedy algorithm. Nevertheless, we show
that, similar to the k-DPP method, greedy selection also exhibits the
improved guarantees and the multiple-descent curve predicted by our
analysis.

The most standard version of the greedy algorithm
\citep[see, e.g.,][]{Bhaskara2016GreedyCSS} 
starts with an empty set and then
iteratively adds columns that minimize the 
approximation error at every step, until we reach a set of size
$k$. 
The pseudo-code is given below.
\renewcommand{\thealgorithm}{}
\floatname{algorithm}{}
\begin{algorithm}[H]
  \caption{Greedy subset selection algorithm for CSSP/Nystr\"om}
  \begin{algorithmic}[0]
    \STATE \textbf{Input:} $k\in[n]$\quad and \quad an $m\times n$
    matrix $\A$ (CSSP), \quad or\quad an $n\times 
    n$ p.s.d.~matrix $\K=\A^\top\A$ (Nystr\"om)\vspace{2mm}
    \STATE $S\leftarrow \emptyset$
    \STATE \textbf{for} $i$ $=$ $1$ \textbf{to} $k$ \textbf{do}
    \STATE \quad Pick $i\in[n]\backslash S$ that minimizes $\Er_\A(S\cup\{i\})$,\quad or
    equivalently, $\|\K-\Kbh(S\cup\{i\})\|_*$
    \STATE \quad$S\leftarrow S\cup\{i\}$
    \STATE \textbf{end for}\vspace{2mm}
    \RETURN $S$
  \end{algorithmic}
\end{algorithm}
In our empirical evaluation we use the same experimental setup as in
Section \ref{s:experiments}, by running greedy on a toy dataset with
the linear kernel $\langle\a_i,\a_j\rangle_\text{K}=\a_i^\top\a_j$
that has one sharp spectrum drop (controlled by the condition number $\kappa$),
and two Libsvm datasets with the RBF kernel
$\langle\a_i,\a_j\rangle_\text{K}=\exp(-\|\a_i\!-\!\a_j\|^2/\sigma^2)$
for three values of the RBF parameter $\sigma$. The main question
motivating these experiments is: does the approximation factor of the
greedy algorithm exhibit the multiple-descent curve that is predicted in
our analysis, and are the peaks in this curve aligned with the sharp drops
in the spectrum of the data?

 \begin{figure*}[t]
  \centering
  \ifisarxiv
    \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-toy-greedy}
   \hspace{-6mm}
  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-bodyfat-greedy}
  \hspace{-6mm}
  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-eunite2001-greedy}
  \else
  ~\hspace{-5mm}
  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-toy-greedy}\hspace{-5mm}\nolinebreak\includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-bodyfat-greedy}\hspace{-5mm}\nolinebreak\includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-eunite2001-greedy}
  \fi
  \caption{Top plots show the Nystr\"om approximation factor
    $\|\K-\Kbh(S)\|_*/\opt$,
    where $S$ is constructed using greedy subset selection,
    against the subset size $k$, for a toy dataset
    ($\kappa$ is the condition number) and two
    Libsvm datasets ($\sigma$ is the RBF parameter). Bottom plots show
    the spectral decay for  the top $40$ eigenvalues  of each kernel
    $\K$,  demonstrating how the peaks in the Nystr\"om approximation
    factor align with the drops in the spectrum.}
\vspace{-2mm}
  \label{f:greedy}
\end{figure*}

The plots in Figure \ref{f:greedy} confirm that the
Nystr\"om approximation factor of greedy subset selection exhibits
similar peaks and valleys as those indicated by our theoretical and
empirical analysis of the k-DPP method. This is most clearly observed
for the toy dataset (Figure \ref{f:greedy} left), where the peak grows with the condition number
$\kappa$, and for the \emph{bodyfat} dataset (Figure \ref{f:greedy} center), where the size of the peak is
proportional to the RBF parameter $\sigma$. Moreover, we observe that
when the spectral decay is slow/smooth, which corresponds to smaller values
of $\sigma$, then the approximation factor of the greedy algorithm
stays relatively close to 1. For the \emph{eunite2001}
dataset (Figure \ref{f:greedy} right), the behavior of the approximation factor is very
non-linear, with several peaks occurring for large values of
$\sigma$. Interestingly, while the peaks do align with some of the
drops in the spectrum, not all of the spectrum drops result in a peak for the
greedy algorithm. This goes in line with our analysis, in the sense
that a sharp drop in the spectrum following the $k$th eigenvalue is a
\emph{necessary but not sufficient} condition for the approximation
factor of the optimal subset $S$ of size $k$ to exhibit a peak.

Our empirical evaluation leads to an overall conclusion that the
multiple-descent curve of the CSSP/Nystr\"om approximation factor is a
phenomenon exhibited by both \emph{randomized}
methods, such as the k-DPP, and \emph{deterministic}
algorithms, such as greedy subset selection. While the exact behavior
of this curve is algorithm-dependent, significant insight can be
gained about it by studying the spectral properties of the data. Our
results suggest that performing a theoretical analysis of
the multiple-descent phenomenon for greedy methods is a promising direction for future work.

  \end{document}

### 2022-08-23 design/nystrom-main.tex
Commit original repos

--- 2022-08-23 design/nystrom-main-neurips2020.tex
@@
 \newif\ifisarxiv
-\isarxivfalse
+\isarxivtrue
 
@@
 
-\usepackage{multibib}
 % Recommended, but optional, packages for figures and better typesetting:
@@
 \usepackage{cleveref}
-%\usepackage{natbib}
+\usepackage{natbib}
 \usepackage{xcolor}
-
-\hypersetup{%draft, % TODO: remove "draft" option before final version after making sure no hyperlinks are broken due to pagebreaks.
+\hypersetup{
 	colorlinks,
@@
 
+% Use the following line for the initial blind version submitted for
+% review:
+
 \ifisarxiv
 \usepackage{fullpage}
-\usepackage{natbib}
 \else
-\usepackage[final]{sty/neurips2020/neurips_2020}
+\usepackage[nohyperref]{sty/icml2020/icml2020}
 
 % If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[final]{neurips_2020}
-
-\title{Improved guarantees and a multiple-descent curve for %the CSSP
-  Column Subset Selection %Problem
+%\usepackage[accepted]{icml2020}
+
+% The \icmltitle you define below is probably too long as a header.
+% Therefore, a short form for the running title is supplied here:
+\icmltitlerunning{Improved guarantees and a multiple-descent curve for
+  the CSSP and the Nystr\"om method}
+\fi
+\begin{document}
+
+\ifisarxiv
+\title{Improved guarantees and a multiple-descent curve
+  for the\\ Column Subset Selection Problem
   and the Nystr\"om method}
@@
   \texttt{mderezin@berkeley.edu}\\
-  \And
-{Rajiv Khanna} \\
+  \and
+Rajiv Khanna \\
   Department of Statistics\\
@@
   \texttt{rajivak@berkeley.edu}
-  \And
+  \and
    Michael W. Mahoney\\
@@
   }
+
+\else
+\twocolumn[
+\icmltitle{Improved guarantees and a multiple-descent curve
+  for the\\ Column Subset Selection Problem
+  and the Nystr\"om method}
+
+% It is OKAY to include author information, even for blind
+% submissions: the style file will automatically remove it for you
+% unless you've provided the [accepted] option to the icml2020
+% package.
+
+% List of affiliations: The first argument should be a (short)
+% identifier you will use later to specify author affiliations
+% Academic affiliations should list Department, University, City, Region, Country
+% Industry affiliations should list Company, City, Region, Country
+
+% You can specify symbols, otherwise they are numbered in order.
+% Ideally, you should not use this facility. Affiliations will be numbered
+% in order of appearance and this is the preferred way.
+\icmlsetsymbol{equal}{*}
+
+\begin{icmlauthorlist}
+\icmlauthor{Aeiau Zzzz}{equal,to}
+\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
+\icmlauthor{Cieua Vvvvv}{goo}
+\icmlauthor{Iaesut Saoeu}{ed}
+\icmlauthor{Fiuea Rrrr}{to}
+\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
+\icmlauthor{Aaoeu Iasoh}{goo}
+\icmlauthor{Buiui Eueu}{ed}
+\icmlauthor{Aeuia Zzzz}{ed}
+\icmlauthor{Bieea C.~Yyyy}{to,goo}
+\icmlauthor{Teoau Xxxx}{ed}
+\icmlauthor{Eee Pppp}{ed}
+\end{icmlauthorlist}
+
+\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
+\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
+\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
+
+\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
+\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
+
+% You may provide any keywords that you
+% find helpful for describing your paper; these are used to populate
+% the "keywords" metadata in the PDF but will not be shown in the document
+\icmlkeywords{Machine Learning, ICML}
+
+\vskip 0.3in
+]
+
+% this must go after the closing bracket ] following \twocolumn[ ...
+
+% This command actually creates the footnote in the first column
+% listing the affiliations and the copyright notice.
+% The command takes one argument, which is text to display at the start of the footnote.
+% The \icmlEqualContribution command is standard text for equal contribution.
+% Remove it (just {}) if you do not need this facility.
+
+%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+
 \fi
-\begin{document}
-
-\ifisarxiv
-\title{Improved guarantees and a multiple-descent curve for\\ %the CSSP
-  Column Subset Selection %Problem
-  and the Nystr\"om method}
-  \author{%
-          \textbf{Micha{\l } Derezi\'{n}ski} \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mderezin@berkeley.edu}\\
-  \and
-\textbf{Rajiv Khanna} \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{rajivak@berkeley.edu}
-  \and
-   \textbf{Michael W. Mahoney}\\
-  ICSI and Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mmahoney@stat.berkeley.edu}
-  }
-
-\else\fi
 
@@
   \date{}
+  \maketitle
 \fi
-  \maketitle
 
 \begin{abstract}
+% The Column Subset Selection Problem (CSSP) and the Nystr\"om method
+% are among the leading tools for constructing small low-rank
+% approximations of large datasets in machine learning and scientific
+% computing. A fundamental question in this area is: how well
+% can a data subset of size $k$ compete with the best rank $k$
+% approximation? To address this question, we develop improved
+% approximation guarantees which go beyond the standard worst-case
+% analysis by relying on the spectral properties of the data. Our approach leads
+% to significantly better bounds for datasets with known rates of singular value
+% decay, e.g., polynomial or exponential decay. Our analysis also
+% reveals an intriguing phenomenon: the approximation factor as a
+% function of $k$ may exhibit multiple peaks and valleys, which we call
+% a multiple-descent curve. Moreover, our new lower bounds show that
+% this behavior is not an artifact of our analysis, but rather an
+% inherent property of the CSSP and Nystr\"om tasks. Finally, we verify
+% our theoretical analysis on real datasets.
 The Column Subset Selection Problem (CSSP) and the Nystr\"om method
@@
 \section{Introduction}
-\label{s:intro}
+
+\begin{figure}
+  \centering  
+  \includegraphics[width=\ifisarxiv 0.6\else 0.5\fi\textwidth]{figs/nystrom/nystrom-bounds}
+  \caption{An empirical study showing the
+    expected approximation factor $\E[\Er_\A(S)]/\opt$ for
+$S\sim k$-$\DPP(\A^\top\A)$, with
+    different subset sizes $|S|=k$, compared to our theory. We use a
+    data matrix $\A$ 
+    whose
+    spectrum exhibits two sharp drops, demonstrating
+    the multiple-descent phenomenon. The lower bounds are based on Theorem
+    \ref{t:lower}, whereas, as our upper bound, we plot the
+    minimum over all functions $\Phi_s(k)$ from Theorem
+    \ref{t:upper}. Note that when the spectrum of
+    matrix $\A$ exhibits sufficiently smooth decay (instead of the sharp drops), then the
+    multiple-descent curve vanishes. In such cases, our upper bounds
+    significantly improve on the worst-case analysis of \citet{pca-volume-sampling}
+    for all subset sizes $k$ (see Theorem \ref{t:decay}).
+  }
+  \label{f:intro}
+\end{figure}
 We consider the task of selecting a small but representative sample of
@@
 \end{definition}
-Another variant of the CSSP emerges in the kernel
+Another variant of the CSSP, of particular interest in machine learning, emerges in the kernel
 setting under the name \emph{Nystr\"om method}
@@
 ${n \choose k}$ subsets, which is prohibitively expensive, so we
-would like to find our subset more efficiently.
-
-\ifisarxiv\begin{figure}\else
-\begin{wrapfigure}{r}{0.45\textwidth}
-  \vspace{-5mm}
-  \fi
-  \centering  
-  \includegraphics[width=\ifisarxiv 0.6\else 0.47\fi\textwidth]{figs/nystrom/nystrom-bounds}
-  \caption{Empirical study of the
-    expected approximation factor $\E[\Er_\A(S)]/\opt$ for
-a $k$-DPP with
-    different subset sizes $|S|=k$, compared to our theory. We use a
-    data matrix $\A$ 
-    whose
-    spectrum exhibits two sharp drops, demonstrating
-multiple-descent. The lower bounds are based on Theorem
-    \ref{t:lower}, whereas, as our upper bound, we plot the
-    minimum over all $\Phi_s(k)$ from
-    Theorem~\ref{t:upper}. Note that multiple-descent vanishes under
-    smooth spectral decay, resulting in improved guarantees (see
-    Theorem \ref{t:decay} and Figure \ref{f:sliding}).
-  }
-  \label{f:intro}
-  \ifisarxiv\end{figure}\else
-  \end{wrapfigure}\fi
-
-Extensive literature has been dedicated to developing algorithms for
-the CSSP (an in depth discussion of the related work can be found in
-Appendix \ref{s:related-work}).
+would like to find our subset more efficiently. %  We formalize these
+% goals as follows.\\
+% \textbf{Problem:} Given $\A$, $k$ and $\epsilon$, is there a subset $S$ of
+% size $k$ such that $\Er_\A(S)\leq (1+\epsilon)\cdot \|\A-\A_k\|_F^2$
+% and can we find such a set in time polynomial in $n$, $m$, $k$ and $1/\epsilon$?
+
 In terms of worst-case analysis, 
+%the task may be considered solved, since 
 \citet{pca-volume-sampling} gave a randomized
-method 
+method %randomized procedure 
 which returns a
-set $S$ of size $k$ such that: %\vspace{-4mm}
+set $S$ of size $k$ such that:
 \begin{align}
@@
 efficient implementations have been provided since then
-\citep[e.g., see][]{efficient-volume-sampling,dpp-intermediate}.
+\citep[e.g., see][]{efficient-volume-sampling}.
 The method belongs to the family of cardinality constrained
-Determinantal Point Processes \citep[DPPs, see][]{dpps-in-randnla}, and will
+determinantal point processes (see Definition \ref{d:dpp}), and will
 be denoted as $S\sim k\text{-}\DPP(\A^\top\A)$. 
@@
 \ref{f:intro}).  We show that this phenomenon is an inherent property
-of the CSSP  (Corollary \ref{c:multiple-descent}).
+of the CSSP  (Corollary \ref{c:multiple-descent}), which
+leads us to a new connection with the recently discovered double
+descent curve \citep{BHMM19,surrogate-design}.
 \end{enumerate}
@@
 \end{definition}
+% We show that the approximation factor achievable by a
+% subset of size $k$ in the
+% CSSP can be much better than $k+1$. This is done
 In the following result, we define a family of functions $\Phi_s(k)$ which
@@
   \frac{\E[\Er_\A(S)]}{\opt}\leq (1+2\epsilon)^2\,\Phi_{s}(k),
-  \qquad\text{where}\quad \Phi_s(k)=\big(1+\tfrac{s}{k-s}\big)\sqrt{1 + \tfrac{2(k-s)}{t_s-k}\,}.
-\end{align*}
+\end{align*}
+where  $\Phi_s(k)=\big(1+\frac{s}{k-s}\big)\sqrt{1 + \frac{2(k-s)}{t_s-k}\,}$.
 \end{theorem}
@@
   \begin{enumerate}
-\item  For all $\A$, if $k\leq \sr_0(\A)\!-\!1$, then there is a subset $S$ of size $k$ such that
+\item  For all $\A$, if $k\leq \sr_0(\A)\!-\!1$, then there exists a subset $S$ of size $k$ such that
 $\frac{\Er_\A(S)}{\opt} = O(\sqrt  k\,)$.
-\item There is $\A$ such that $\sr_0(\A)\!-\!1<k<\sr_0(\A)$ and for
-  every size $k$ subset $S$, $\frac{\Er_\A(S)}{\opt}
+\item There is $\A$ such that $\sr_0(\A)\!-\!1<k<\sr_0(\A)$ and for every subset $S$ of
+  size $k$ we have $\frac{\Er_\A(S)}{\opt}
   \geq 0.9\,k$.
@@
 \citet{more-efficient-volume-sampling}. Observe how the worst-case
-approximation factor jumps from $O(\sqrt k\,)$ to $\Omega(k)$, as $k$
+approximation factor jumps from $O(\sqrt k\,)$ to $O(k)$, as $k$
 approaches $\sr_0(\A)$. An example of this sharp
@@
 \textnormal{2.} (\textbf{exponential spectral decay}) $c_1(1\!-\!\delta)^{i}\leq \lambda_i\leq
-  c_2(1\!-\!\delta)^{i}$ $\forall_i$, $\delta\in(0,1)$, then $S\sim
-k$-$\DPP(\A^\top\A)$ satisfies %\vspace{-4mm}
+  c_2(1\!-\!\delta)^{i}$ $\forall_i$, with $\delta\in(0,1)$, then $S\sim
+k$-$\DPP(\A^\top\A)$ satisfies
   \begin{align*}
@@
 a relatively flat region after the $s$th eigenvalue (Figure \ref{f:intro}
-for $k$ between 20 and 40). Note that a peak value of some function
+for $k$ between 20 and 50). Note that a peak value of some function
 $\Phi_{s_1}$ may coincide with a valley of some $\Phi_{s_2}$, so only
@@
   \begin{align*}
-    \min_{S:|S|=k_i^l}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^l}}\leq 1+\delta
-                                 \qquad\text{and }\qquad
-    \min_{S:|S|=k_i^u}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^u}} \geq
+    \min_{S:|S|=k_i^l}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^l}}&\leq 1+\delta
+                                 \quad\text{and }\\
+    \min_{S:|S|=k_i^u}\frac{\Er_\A(S)}{\textsc{OPT}_{k_i^u}} &\geq
     (1-\delta)(k_i^u+1).
@@
 \end{corollary}
-\paragraph{Connection to double descent.}
-A number of phase transitions have been recently observed in the
-machine learning literature which are commonly dubbed \emph{double
-  descent}. The term was introduced by \citet{BHMM19} in the context of generalization
-error of statistical learning, however
-\citet{double-descent-condition} observed that the behavior of the
-generalization error, at least for linear models, can be explained by
-a more fundamental double descent phenomenon observed in the condition
-number of random matrices. Also, \citet{zhenyu2020double} showed that
-double descent is merely a manifestation of the
-spectral phase transitions in high-dimensional random kernel
-matrices. Further, \citet{liang2020multiple} observed that the spectral phase transitions
-of certain kernels can lead to multiple double descent peaks. The
-multiple-descent phenomenon in the CSSP that 
-emerges from our analysis is
-related to those works in that the spectral properties of the data
-matrix (and, in particular, the condition number) determine the peaks
-(i.e., phase transitions) discussed in
-Corollary~\ref{c:multiple-descent}. Those parallels manifest
-themselves most clearly when comparing our analysis to the work of
-\citet{BLLT19_TR}, which uses the same notion of stable rank as we do,
-and \citet{surrogate-design}, where determinantal sampling plays a
-central role in the analysis of double descent. However, we stress that there are
-  also important  differences in our setting: (1) the CSSP is a
-  \emph{deterministic} optimization task, and (2) we study the
-  \emph{approximation factor}, rather than the generalization error (further discussion in
-  Appendix~\ref{s:related-work}).
+The multiple-descent phenomenon that emerges from our analysis bears
+similarity to the double descent curve 
+described by \citet{BHMM19}. This curve illustrates the sharp transition
+between the generalization error of under- and over-parameterized machine
+learning models as we change the ratio between the number of parameters 
+and the number of samples. We further discuss this connection in
+Section \ref{s:related-work}.
 
@@
 $n\times |S|$  submatrix with columns indexed by $S$. The Nystr\"om
-method has many applications in machine learning, including for
+method has numerous applications in machine learning, including for
 kernel machines \citep{Williams01Nystrom}, Gaussian Process regression 
@@
   If $\K=\A^\top\A$ and $\|\cdot\|_*$ is the trace norm, then
-    $\big\|\K-\Kbh(S)\big\|_* = \Er_{\A}(S)$ for all $S\subseteq\{1,...,n\}$.
+  \begin{align*}
+    \big\|\K-\Kbh(S)\big\|_* = \Er_{\A}(S) \quad \text{for all}\quad
+    S\subseteq\{1,...,n\}.
+  \end{align*}
 Moreover, the trace norm error of the best rank $k$ approximation of
@@
 \end{remark}
-%\vspace{-2mm}
+\vspace{-2mm}
 This connection was used by~\citet{belabbas-wolfe09} to adapt the
@@
 
-\section{Upper bounds}\label{s:upper}\vspace{-2mm}
-In this section, we derive the upper bound given in
-Theorem~\ref{t:upper} by using a novel expectation formula for the squared
-projection error of a DPP. We then show
-how this result can be used to 
-obtain improved guarantees for matrices with known eigenvalue decays,
-i.e., Theorem \ref{t:decay}. Our analysis heavily relies on the theory
-of DPPs \citep{dpps-in-randnla}, so for completeness, in Appendix~\ref{s:dpp} we provide a
-brief summary of DPPs and the relevant results.
-
-Let $S\sim \DPP(\frac1\alpha\A^\top\A)$ denote a
-distribution over all subsets $S\subseteq [n]$ so that
-$\Pr(S)\propto\det(\frac1\alpha\A_S^\top\A_S)$, where $\alpha>0$. Then,
-$k$-$\DPP(\A^\top\A)$ is simply a restriction of
-$\DPP(\frac1\alpha\A^\top\A)$ to the subsets of size $k$ (regardless
-of the choice of $\alpha$).
-However, the expected subset size for $\DPP(\frac1\alpha\A^\top\A)$ does
-depend on~$\alpha$. Our analysis relies on a careful selection of
-this parameter. In Lemma \ref{l:expected-error}  (Appendix \ref{s:dpp}), we show the
-following expectation formula for the CSSP approximation error: 
-\[\E[\Er_\A(S)]=\E[|S|]\cdot\alpha,\quad\text{for}\quad S\sim \DPP(\tfrac1\alpha\A^\top\A).\]
-If we set $\alpha =\opt = \sum_{i=k+1}^n\lambda_i$,
-where $\lambda_i$ are the eigenvalues of
-$\A^\top\A$ in decreasing order, then:
-\begin{align*}
- \E[|S|]& = \sum_{i=1}^n\frac{\lambda_i}{\alpha+\lambda_i}
-\leq\sum_{i=1}^k\frac{\lambda_i}{\alpha+\lambda_i} + 1\leq k+1.
-\end{align*}
-This recovers the upper bound of \citet{pca-volume-sampling}, i.e., $\E[\Er_\A(S)]\leq(k+1)\opt$,
-except that the
-subset size is randomized with expectation bounded by 
-$k+1$, instead of a fixed subset size equal $k$. However, a more refined
-choice of the parameter $\alpha$
-allows us to significantly improve on the above error bound in certain
-regimes, as shown below.
-\begin{lemma}\label{l:upper}
-For any $\A$, $0\leq\epsilon<1$ and $s<k<t_s$, where $t_s=s+\sr_s(\A)$,
-say $S\sim\DPP(\frac1\alpha\A^\top\A)$ for
-$\alpha=\frac{\gamma_s(k)\opt}{(1-\epsilon)(k-s)}$ and
-$\gamma_s(k):=\sqrt{1+\frac{2(k-s)}{t_s-k}\,}$. Then, defining $\Phi_s(k)
-:=\big(1+\frac{s}{k-s}\big)\,\gamma_s(k)$,
-\begin{align*}
-  \frac{\E\big[\Er_\A(S)\big]}{\opt}\leq \frac{\Phi_s(k)}{1-\epsilon}
-  \quad\text{and}\quad \E[|S|]\leq k-\epsilon\,\frac{k-s}{\gamma_s(k)}.
-\end{align*}
-  \end{lemma}
-Note that, setting $\epsilon=0$, the above lemma implies that we can
-achieve approximation factor $\Phi_s(k)$ with a DPP whose expected
-size is bounded by $k$. We introduce $\epsilon$ so that we can
-convert the bound from DPP to the fixed size k-DPP via a
-concentration argument. Intuitively, our strategy is to show that the
-randomized subset size of a DPP is sufficiently concentrated around
-its expectation that with high probability it will be bounded by $k$,
-and for this we need the expectation to be strictly below $k$. A
-careful application of the Chernoff bound for a Poisson binomial
-random variable yields the following concentration bound.
-\begin{lemma}\label{l:concentration}
-Let $S$ be as in Lemma \ref{l:upper} with $\epsilon\leq \frac12$.
-If $s+\frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$, then
-$\Pr(|S|>k) \leq \epsilon$.
-\end{lemma}
-Finally, any expected bound for random size DPPs can be converted to
-an expected bound for a fixed size k-DPP via the following
-result.
-\begin{lemma}\label{l:monotonic}
-  For any $\A\in\R^{m\times n}$, $k\in[n]$ and $\alpha>0$, if
-  $S\sim\DPP(\frac1\alpha\A^\top\A)$ and $S'\sim k$-$\DPP(\A^\top\A)$,
-  then %\vspace{-4mm}
-  \begin{align*}
-    \E\big[\Er_\A(S')\big] \leq \E\big[\Er_\A(S)\mid |S|\leq k\big].
-  \end{align*}
-\end{lemma}
-The above inequality may seem intuitively obvious since adding more columns
-to a set $S$ to complete it to size $k$ always reduces the
-error. However, a priori, it could happen that going from subsets 
-of size $k-1$ to subsets of size $k$ results in a redistribution of
-probabilities to the subsets with larger error. To show that this will
-not happen, our proof relies on classic but non-trivial combinatorial
-bounds called Newton's inequalities. Putting together
-Lemmas~\ref{l:upper}, \ref{l:concentration} and
-\ref{l:monotonic}, we obtain our Master Theorem.
-\begin{proofof}{Theorem}{\ref{t:upper}}
-Let $S\sim\DPP(\frac1\alpha\A^\top\A)$ be sampled as in Lemma
-\ref{l:upper}, and let $S'\sim k$-$\DPP(\A^\top\A)$. We have:\vspace{-2mm}
-\begin{align*}
-  \E\big[\Er_\A(S')\big]
-\overset{(a)}{\leq} \E\big[\Er_\A(S)\mid|S|\leq k\big]
-  \leq \frac{\E\big[\Er_\A(S)\big]}{\Pr(|S|\leq k)}
-  \overset{(b)}{\leq} \frac{\Phi_s(k)}{(1-\epsilon)^2}\cdot\opt,
-\end{align*}
-where $(a)$ follows from Lemma \ref{l:monotonic} and $(b)$ follows
-from Lemmas \ref{l:upper} and \ref{l:concentration}. Since
-$0<\epsilon\leq\frac12$, we have
-$\frac1{(1-\epsilon)^2}\leq(1+2\epsilon)^2$, which completes the proof.
-\end{proofof}
-
-\begin{figure}
-  \centering
-  \ifisarxiv
-  \includegraphics[width=0.6\textwidth]{figs/nystrom/sliding}
-\includegraphics[width=0.6\textwidth]{figs/nystrom/sliding-peak}
-\else  
-%\centering
-\hspace{-8mm}
-\includegraphics[width=0.54\textwidth]{figs/nystrom/sliding}\hspace{-5mm}
-\includegraphics[width=0.54\textwidth]{figs/nystrom/sliding-peak}
-%  \vspace{-6mm}
-  \fi
-  \caption{Illustration of the upper bound functions $\Phi_s(k)$ for
-    different values of $s$, with a $200\times 200$ matrix $\A$ such
-    that the $i$th eigenvalue of $\A^\top\A$ is set to:
-    \ifisarxiv(top)\else(left)\fi\
-    $1/i$;
-    \ifisarxiv(bottom)\else(right)\fi\
-    $1$ for $i<30$ and $0.01$ for $i\geq 30$. For each
-    function, we marked the window of applicable $k$'s with a
-    horizontal line. For polynomial spectral decay
-    \ifisarxiv(top)\else(left)\fi
-    , the stable rank
-    $\sr_s(\A)$ (i.e., the width of the window starting at $s$)
-    increases, while for the sharp spectrum drop
-    \ifisarxiv(bottom)\else(right)\fi\
-    the stable
-    rank shrinks as the window approaches the drop, causing a peak in
-    the upper~bound.}\vspace{-2mm}
-  \label{f:sliding}
-\end{figure}
-
-We now demonstrate how Theorem \ref{t:upper} can be used as the Master Theorem to derive new
-bounds on the CSSP approximation factor under additional assumptions
-on the singular value decay of matrix $\A$. Rather than a single upper
-bound, Theorem \ref{t:upper} provides a family of upper bounds
-$\Phi_s$, each with a range of applicable values $k$. Since each
-$\Phi_s(k)$ forms a U-shaped curve, its smallest point falls near the
-middle of that range. In Figure \ref{f:sliding} we visualize these bounds as a sliding
-window that sweeps across the axis representing possible subset
-sizes. The width of the window varies: when it starts at $s$ 
-then its width is the stable rank $\sr_s(\A)$. The wider the window,
-the lower is the valley of the corresponding U-curve. Thus, when
-bounding the approximation factor for a given $k$, we should choose
-the widest window such that $k$ falls near the bottom of its
-U-curve. Showing a guarantee that holds for all $k$ requires lower-bounding the
-stable ranks $\sr_s(\A)$ for each $s$. This is straightforward for
-both polynomial and exponential decay. Specifically, using the
-notation from Theorem \ref{t:decay}, in Appendix \ref{a:decay} we
-prove that: %\vspace{-3mm}
-\begin{align*}
-  \sr_s(\A) =
-  \begin{cases}
-    \Omega(s/p), &\text{for polynomial rate  $\lambda_i\asymp1/i^p$},\\
-    \Omega(1/\delta),&\text{for exponential rate $\lambda_i\asymp(1-\delta)^i$}.
-    \end{cases}
-  \end{align*}
-As an example, Figure \ref{f:sliding} (left) shows that the stable rank $\sr_s(\A)$, i.e.,
-the width of the window starting at $s$, grows linearly with $s$ for eigenvalues
-decaying polynomially with $p=1$. As a result, the bottom of each
-U-shaped curve remains at roughly the same level, making the CSSP
-approximation factor independent of $k$, as in Theorem
-\ref{t:decay}. In contrast, Figure \ref{f:sliding} (right) provides
-the same plot for a different matrix $\A$ with a sharp drop in the spectrum. The U-shaped
-curves cannot slide smoothly across that drop because of the
-shrinking stable ranks, which results in a peak similar to the ones
-observed in Figure \ref{f:intro}.
-
- \section{Lower bound}\label{s:lower}
-%\vspace{-2mm}
- As discussed in the previous section, our upper bounds for the CSSP
- approximation factor exhibit a peak (a high point,
- with the bound decreasing on either side) around a
- subset size $k$ when there is a sharp drop in the spectrum of $\A$
- around the $k$th singular value. It is natural to ask whether this
- peak is an artifact of our analysis, or a property of the k-DPP
- distribution, or whether even optimal CSSP subsets exhibit this 
- phenomenon. In this section, we extend a lower
- bound construction of \citet{pca-volume-sampling} and use it to show
- that for certain matrices the approximation factor of the optimal CSSP subset, i.e.,
- $\min_{|S|=k}\Er_\A(S)/\opt$, can exhibit not just one but any
- number of peaks as a function of $k$, showing that the
- multiple-descent curve in Figure \ref{f:intro} describes a
-real phenomenon in the CSSP.
-
-The lower bound construction of \citet{pca-volume-sampling} relies on
-arranging the column vectors of a $(k+1)\times (k+1)$ matrix $\A$ into a
-centered symmetric $k$-dimensional simplex. This way, the $k+1$
-columns are spanning a $k$ dimensional subspace which contains the $k$
-leading singular vectors of $\A$. They then proceed to shift the
-columns slightly in the direction orthogonal to that subspace so that
-the $(k+1)$st singular value of $\A$ becomes non-zero. This results in
-an instance of the CSSP with a sharp drop in the spectrum. Due
-to the symmetry in this construction, all subsets of size $k$ have an identical squared
-projection error. It is easy to show that this error satisfies $\Er_\A(S)\geq (1-\delta)(k+1)\opt$,
-where $\delta$ is a parameter which depends on the condition number of
-matrix $\A$ and it can be driven arbitrarily close to $0$. Another
-variant of this construction was also provided 
-by \citet{more-efficient-volume-sampling}. The key limitation of both
-of these constructions is that they only provide a lower bound for a
-single subset size $k$ in a given matrix, whereas our goal is to show
-that the CSSP can exhibit the multiple-descent curve, which requires
-lower bounds for multiple different values of $k$ holding with respect
-to the same matrix $\A$.
-
-Our strategy for constructing the lower bound matrix is to concatenate
-together multiple sets of columns, each of which
-represents a simplex spanning some subspace of $\R^m$. The key
-challenge that we face in this approach is that, unlike in
-the construction of \citet{pca-volume-sampling}, different subsets of
-the same size will have different projection errors. 
-\begin{lemma}\label{l:lower}
-Fix $\delta\in(0,1)$ and consider unit vectors
-$\a_{i,j}\in\R^m$ in general position, where $i\in[t]$, $j\in[l_i]$, such that
-$\sum_j\a_{i,j}=0$ for each $i$, and for any $i,j,i',j'$, if $i\ne i'$
-then $\a_{i,j}$ is orthogonal to $\a_{i'\!,j'}$. Also, let unit vectors
-$\{\v_i\}_{i\in[t]}$ be orthogonal to  each other and to all
-$\a_{i,j}$. There are positive scalars $\alpha_{i},\beta_i$ for $ i \in [t]$
-such that matrix $\A$ with columns 
-$\alpha_{i}\a_{i,j}+\beta_i\v_i$ over all $i$ and $j$ satisfies:
-%\ifisarxiv\else\vspace{-1mm}\fi
-\begin{align*}
-  \min_{|S|=k_i}\frac{\Er_\A(S)}{\mathrm{OPT}_{k_i}}\geq
-  (1-\delta)l_i,\quad\text{for }k_i=l_1+...+l_i-1.
-\end{align*}
-\end{lemma}
-%\ifisarxiv\else \vspace{-5mm}\fi
-% Theorem \ref{t:lower} now follows easily from the lemma.
-\begin{proofof}{Theorem}{\ref{t:lower}}
-We let $l_1 = k_1+1$ and then for $i>1$ we set $l_i=k_i-k_{i-1}$. We
-then construct the vectors $\a_{i,j}$ that satisfy Lemma \ref{l:lower} 
-by letting each set $\{\a_{i,j}\}_{j}$ be the corners of a centered
-$(l_i-1)$-dimensional regular simplex. We ensure that each
-simplex is orthogonal to every other simplex by placing them in
-orthogonal subspaces. 
-\end{proofof}
-
-\section{Empirical evaluation}
-\label{s:experiments}
-\vspace{-2mm}
-
-In this section, we provide an empirical evaluation designed to demonstrate how our improved guarantees for the CSSP and Nystr\"om method, as well as the multiple-descent phenomenon, can be easily observed on real datasets.
-We use a standard experimental setup for data subset selection using
-the Nystr\"om method \citep{revisiting-nystrom}, where an $n\times n$
-kernel matrix $\K$ for a dataset of size $n$ is defined so that the
-entry $(i,j)$ is computed using the Gaussian Radial Basis Function  (RBF)
-kernel:
-$\langle\a_i,\a_j\rangle_\text{K}=\exp(-\|\a_i\!-\!\a_j\|^2/\sigma^2)$,
-where $\sigma$ is a free parameter. 
-We are particularly interested in the effect of varying $\sigma$.
-Nystr\"om subset selection is performed using $S\sim k$-$\DPP(\K)$
-(Definition \ref{d:dpp}), and we plot the expected approximation
-factor $\E[\|\K-\Kbh(S)\|_*]/\opt$ (averaged over 1000 runs), where
-$\Kbh(S)$ is the Nystr\"om approximation of $\K$ based on the subset
-$S$ (see Section~\ref{s:nystrom}), $\|\cdot\|_*$ is the trace norm,
-and $\opt$ is the trace norm error of the best rank $k$
-approximation. Additional experiments, using greedy selection instead 
-of a k-DPP, are in Appendix~\ref{a:greedy}.
-As discussed in Section \ref{s:nystrom}, this task is
-equivalent to the CSSP task defined on the matrix $\A$ such that
-$\K=\A^\top\A$.
-
-The aim of our empirical evaluation is to verify the following two claims motivated by our theory (and to illustrate that doing so is as easy as varying the RBF parameter $\sigma$):
-%\ifisarxiv\else\vspace{-2mm}\fi
-\begin{enumerate}
-  \item When the spectral decay is sufficiently slow/smooth, the
-    approximation factor for CSSP/Nystr\"om is much better than
-    suggested by previous worst-case bounds.
-%    \vspace{-1mm}
-  \item  A drop in spectrum around the $k$th eigenvalue results in
-    a peak in the approximation factor near
-    subset size~$k$. Several drops result in the
-    multiple-descent curve.
-  \end{enumerate}
-%\ifisarxiv\else  \vspace{-2mm}\fi
-In Figure \ref{f:rbf} (top), we plot the approximation factor against
-the subset size $k$ (in the range of 1 to 40) for an artificial toy
-dataset and 
-for two benchmark regression datasets from the Libsvm repository
-\citep[\emph{bodyfat} and \emph{eunite2001}, see][]{libsvm}. 
-The toy dataset is constructed by scaling the eigenvalues of a random
-$50\times 50$ Gaussian matrix so that the spectrum is flat with a
-single drop at the $21$-st eigenvalue. 
-For each dataset, in Figure~\ref{f:rbf} (bottom), we also show the top
-40 eigenvalues of the kernel $\K$ in decreasing order. 
-For the toy dataset, to maintain full control over the spectrum we use
-the linear kernel $\langle\a_i,\a_j\rangle_\text{K}=\a_i^\top\a_j$, and we show results
-for three different values of the condition number $\kappa$ of kernel
-$\K$.  
-For the benchmark datasets, we show results on the RBF kernel with 
-three different values of the parameter~$\sigma$.
-
-\begin{figure*}[t]
-   \centering
-  \ifisarxiv
-  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-toy-double}
-   \hspace{-6mm}
-  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-bodyfat-double}
-  \hspace{-6mm}
-  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-eunite2001-double}
-  \vspace{-5mm}
-  \else
-    \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-toy-double}
-   \hspace{-6mm}
-  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-bodyfat-double}
-  \hspace{-6mm}
-  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-eunite2001-double}
-  \vspace{-4mm}
-  \fi
-  \caption{Top three plots show the Nystr\"om approximation factor $\E[\|\K-\Kbh(S)\|_*]/\opt$,
-    where $S\sim k$-$\DPP(\K)$ (experiments using greedy selection
-    instead of a k-DPP are in Appendix \ref{a:greedy}), for a toy dataset
-    ($\kappa$ is the condition number) and two
-    Libsvm datasets ($\sigma$ is the RBF parameter). Error bars show three times the standard
-    error  of the mean over 1000 trials. Bottom three plots show the spectral decay for  the
-  top $40$ eigenvalues  of each kernel $\K$. Note that the
-  peaks in the approximation factor align with the drops in
-  the spectrum.}
-\vspace{-2mm}
-  \label{f:rbf}
-\end{figure*}
-
-Examining the toy dataset (Figure~\ref{f:rbf}, left), it is apparent
-that a larger drop in spectrum leads to a sharper peak in the
-approximation factor as a function of the subset size~$k$, whereas a flat spectrum results in  
-the approximation factor being close to $1$. A similar trend is observed for
-dataset \emph{bodyfat} (Figure~\ref{f:rbf}, center), where large
-parameter $\sigma$ results in a peak that is aligned with a spectrum
-drop, while 
-decreasing $\sigma$ makes the spectrum flatter and the factor closer to
-$1$. Finally, dataset \emph{eunite2001} (Figure~\ref{f:rbf}, right) exhibits a full
-multiple-descent curve with up to three peaks for large values of
-$\sigma$, and the peaks are once again aligned with the spectrum
-drops. Decreasing $\sigma$ gradually eliminates the peaks, 
-resulting in a uniformly small approximation factor. Thus, both of our theoretical
-claims can easily be verified on this dataset simply by adjusting the RBF parameter.
-
-While the right choice of the parameter $\sigma$ ultimately depends on
-the downstream machine learning task, it has been observed that
-varying $\sigma$ has a pronounced effect on the spectral properties of
-the kernel matrix, \citep[see,
-e.g.,][]{revisiting-nystrom,LBM16_TR,wang2019block}.  
-The main takeaway from our results here is that, depending on the
-structure of the problem, we may end up in the regime where the
-Nystr\"om approximation factor exhibits a multiple-descent curve
-(e.g., due to a hierarchical nature of the data) or in the regime
-where it is relatively flat.
-
-\section{Conclusions and open problems}
-\label{s:conclusions}
-%\vspace{-3mm}
-We derived new guarantees for the Column Subset Selection Problem (CSSP) and
-the Nystr\"om method, going beyond worst-case analysis by exploiting the
-structural properties of a dataset, e.g., when the spectrum exhibits a
-known rate of decay. Our upper and lower bounds
-for the CSSP/Nystr\"om approximation factor reveal an intruiguing phenomenon we
-call the multiple-descent curve: the approximation factor can exhibit a highly
-non-monotonic behavior as a function of $k$, with multiple
-peaks and valleys. These observations suggest a possible connection to the
-double descent curve exhibited by
-the generalization error of many machine learning models (see
-Appendix~\ref{s:related-work} for a more in depth discussion of
-similarities and differences between the two phenomena).
-
-Our analysis technique relies on converting an error bound
-from random-size DPPs to fixed-size k-DPPs, which results in an
-additional constant factor of $(1+2\epsilon)^2$ in Theorem~\ref{t:upper}.
-We put forward a
-conjecture which would eliminate this factor from
-Theorem~\ref{t:upper} and is of independent interest to the study of 
-elementary symmetric polynomials, a classical topic in combinatorics
-\citep{HLP-Inequalities}. 
-\begin{conjecture}\label{c:convex}
-  The following function is \underline{convex} with respect to $k\in[n]$ for any
-$\lambda_1,...,\lambda_n> 0$:\vspace{-.5mm}
- \[f(k)=(k+1)\frac{\sum_{S:|S|=k+1}\prod_{i\in S}\lambda_i}{\sum_{S:|S|=k}\prod_{i\in S}\lambda_i}.\]
-\end{conjecture} %\vspace{-1.5mm}
-
-\citet{pca-volume-sampling} showed that if $S\sim
-k$-$\DPP(\A^\top\A)$ and $\lambda_i$ are the eigenvalues of
-$\A^\top\A$, then $\E[\Er_\A(S)]=f(k)$. If $f(k)$ is convex then
-Jensen's inequality implies:
-\begin{align*}
-  \E[\Er_\A(S)]\leq \E[\Er_\A(S')]\quad\text{for }S'\!\sim\DPP(\tfrac1{\alpha_k}\A^\top\A),
-\end{align*}
-where $\alpha_k$ is chosen so that $\E[|S'|]=k$. This would allow us to use
-the bound from Lemma \ref{l:upper} directly on a k-DPP
-without relying on the concentration argument of
-Lemma~\ref{l:concentration}, thereby improving the bounds in
-Theorems \ref{t:upper} and \ref{t:decay}.
-
-\subsection*{Acknowledgments}
-We would like to acknowledge DARPA, IARPA (contract W911NF20C0035),
-NSF, and ONR via its BRC on RandNLA for providing partial support of
-this work.  Our conclusions do not necessarily reflect the position or
-the policy of our sponsors, and no official endorsement should be
-inferred.
-
-\ifisarxiv\else
-\section*{Broader Impact}
-Our study offers a deeper theoretical understanding of the Column
-Subset Selection Problem. The nature of our work is primarily
-theoretical, with direct applications to feature selection and kernel
-approximation, as we noted in Section \ref{s:intro}. The primary reason for
-feature selection as a method for approximating a given matrix, as opposed to a low rank
-approximation using an SVD, is interpretability, which is crucial in
-many scientific disciplines. Our analysis shows that in many practical settings, feature
-selection performs almost as well as SVD at approximating a matrix. As
-such, our work makes a stronger case for feature selection, wherever
-applicable, for the sake of interpretability. We also hope our work
-motivates further research into a fine grained analysis to quantify if
-machine learning problems are really as hard as worst-case bounds
-suggest them to be.
-\fi
-
-\bibliographystyle{sty/icml2020/icml2020}
-\bibliography{pap}
-
-  \newpage
-  \appendix
-  \onecolumn
-
-\section{Additional related works}
+\section{Related work}
 \label{s:related-work}
@@
 included many randomized methods
-\citep{pca-volume-sampling,BoutsidisMD08,dpp-for-cssp,boutsidis2014optimal}, as well as deterministic
-methods \citep{GuE96}.
-Variants of these algorithms have also been extended to more general losses~\citep{chierichetti17algorithms, khanna2017scalable, elenberg2016restricted}.
-Later on, most works have relaxed the problem
+\citep{pca-volume-sampling,BoutsidisMD08,dpp-for-cssp}, as well as deterministic
+methods \citep{GuE96}. Later on, most works have relaxed the problem
 formulation by allowing the number of selected columns $|S|$ to exceed
@@
 \citep{belkin2019two,BLLT19_TR,surrogate-design} and other learning models.
-Double descent is typically presented by plotting
-the absolute generalization error as a function of the number of parameters used in the
-learning model, although \citet{double-descent-condition} and
-\citet{zhenyu2020double} showed that the 
-behavior of generalization error is merely an artifact of the phase
-transitions in the spectral properties of random
-matrices. Importantly, although the descent curves we obtain are 
-reminiscent of the above works, our setting is different in that it is
-a \emph{deterministic} combinatorial 
-optimization problem for \emph{relative} error. In particular,
-Corollary~\ref{c:multiple-descent} shows that our multiple-descent
-curve can 
-occur as a purely deterministic property of the optimal CSSP
-solution. Despite the differences, there are certain similarities
-between the two settings, namely (a) the notion of stable rank we use
-matches the one used by~\citet{BLLT19_TR}, (b) the peaks in both the
-settings are closely aligned -- these peaks coincide with the size $k$
-crossing the corresponding sharp drops in the respective spectra, (c)
-the analysis of bias of the minimum norm solution for double descent
-for linear regression under DPP sampling obtained
-by~\citet{surrogate-design} leads to expressions very similar to ours
-for the CSSP error for DPP sampling.
+Double descent occurs when we plot the
+generalization error as a function of the number of parameters used in the
+learning model. The definition of generalization error
+relies on assuming a probabilistic generative model of
+the data. Importantly, our setting is different in that it is a \emph{deterministic} combinatorial
+optimization problem. In particular, Corollary
+\ref{c:multiple-descent} shows that our multiple-descent curve can
+occur as a purely deterministic property of the optimal CSSP solution.
 
@@
 tasks in machine learning \citep[for a review,
-see][]{dpps-in-randnla,dpp-ml}. Efficient algorithms for sampling from these
+see][]{dpp-ml}. Efficient algorithms for sampling from these
 distributions have been proposed both in the CSSP setting \citep[i.e.,
@@
 
-\underline{Beyond worst-case analysis} of algorithms is crucial to understanding the often-noticed gap between practical performance and theoretical guarantees of these algorithms.
-However, there have been limited number of works in machine learning that undertake finer-grained studies for beyond worst-case analyses.
-We refer to~\citet{roughgarden2019beyond} for a recent survey of such studies for a few problems in machine learning.~\citet{Mah12} takes an alternative view and studies implicit statistical properties of worst case algorithms.
-
 \section{Determinantal point processes}
 \label{s:dpp}
+% Determinantal point processes \cite[DPPs,]{Macchi1975} form a family of
+% non-uniform distributions over sets.
+% The key novelty in our analysis of the Column Subset Selection Problem
+% involves exploiting a close connection between 
+% determinantal point processes with cardinality constrained to some $k$
+% (i.e., k-DPPs) and the classical DPPs for which the subset size is a
+% random variable. This random subset size may seem counter-productive in the
+% context of the CSSP, but it leads to a \emph{simple closed form expression}
+% for the expected approximation error $\E[\Er_\A(S)]$ of a column
+% subset $S$ sampled with a DPP.
 
@@
 and video summarization~\citep{dpp-video}, and others
-\citep{dpp-ml}. They have also played an important role in randomized
-numerical linear algebra \citep{dpps-in-randnla}.
+\citep{dpp-ml}.
 
@@
 eigendecomposition have been proposed
-\citep{dpp-intermediate,dpp-sublinear,alpha-dpp,rayleigh-mcmc},
+\citep{rayleigh-mcmc,dpp-sublinear,dpp-intermediate},
 resulting in running times of $\Ot(n)$ when given matrix $\K$ and
@@
 \citet{surrogate-design}. For completeness, we also provide a proof in
-Appendix~\ref{a:expectration-proj}.
+Appendix~A.
 \begin{lemma}\label{l:expectation_proj}
@@
 \end{proof}
-
+The challenge in using the above formula for the expected error is
+that it applies to a DPP with a randomized subset size rather than a
+fixed size k-DPP. Therefore, the random subset
+$S$ with some positive probability has cardinality much greater than $k$.
+Our strategy in addressing this is to choose
+the rescaling parameter $\alpha$ so that the subset size is bounded by
+$k$ with sufficiently high probability, via a concentration of measure
+argument, as discussed in the following section.
+% $\Pr(|S|\leq k)$ is
+% guaranteed to be sufficiently large (via concentration of measure),
+% and then using the bound:
+% \[\E[\Er_\A(S)\mid |S|\leq k]\leq
+%   \frac{\E[\Er_\A(S)]}{\Pr(|S|\leq k)}.\]
+
+\section{Upper bounds}\label{s:upper}
+In this section, we derive the upper bound given in
+Theorem~\ref{t:upper} by using the expectation formula for the squared
+projection error of a DPP (Lemma \ref{l:expected-error}). We then show
+how this result can be used to 
+obtain improved guarantees for matrices with known eigenvalue decays,
+i.e., Theorem \ref{t:decay}.
+
+Recall that both the expected error formula and the expected subset
+size of $S\sim\DPP(\frac1\alpha\A^\top\A)$ depend on the rescaling
+parameter $\alpha$, and our analysis relies on a careful selection of
+this parameter. To illustrate this, consider setting it to $\alpha =\opt = \sum_{i=k+1}^n\lambda_i$,
+where $\lambda_i$ are the eigenvalues of
+$\A^\top\A$ in decreasing order. Now, \eqref{eq:size} implies that:
+\begin{align*}
+ \E[|S|]& = \sum_{i=1}^n\frac{\lambda_i}{\alpha+\lambda_i}
+\leq\sum_{i=1}^k\frac{\lambda_i}{\alpha+\lambda_i} + 1\leq k+1.
+\end{align*}
+Together with Lemma \ref{l:expected-error}, this recovers the
+upper bound of \citet{pca-volume-sampling} since
+$\E[\Er_\A(S)]=\E[|S|]\cdot\alpha \leq (k+1)\cdot\opt$, except that the
+subset size is randomized with expectation bounded by 
+$k+1$, instead of a fixed subset size equal $k$. However, a more refined
+choice of the parameter $\alpha$
+allows us to significantly improve on the above error bound in certain
+regimes, as shown below.
+\begin{lemma}\label{l:upper}
+For any $\A$, $0\leq\epsilon<1$ and $s<k<t_s$, where $t_s=s+\sr_s(\A)$,
+suppose that $S\sim\DPP(\frac1\alpha\A^\top\A)$ for
+$\alpha=\frac{\gamma_s(k)\opt}{(1-\epsilon)(k-s)}$ and
+$\gamma_s(k)=\sqrt{1+\frac{2(k-s)}{t_s-k}\,}$. Then: 
+\begin{align*}
+  \frac{\E\big[\Er_\A(S)\big]}{\opt}\leq \frac{\Phi_s(k)}{1-\epsilon}
+  \quad\text{and}\quad \E[|S|]\leq k-\epsilon\,\frac{k-s}{\gamma_s(k)},
+\end{align*}
+where $\Phi_s(k) =\big(1+\frac{s}{k-s}\big)\,\gamma_s(k)$.
+  \end{lemma}
+Note that, setting $\epsilon=0$, the above lemma implies that we can
+achieve approximation factor $\Phi_s(k)$ with a DPP whose expected
+size is bounded by $k$. We introduce $\epsilon$ so that we can
+convert the bound from DPP to the fixed size k-DPP via a
+concentration argument. Intuitively, our strategy is to show that the
+randomized subset size of a DPP is sufficiently concentrated around
+its expectation that with high probability it will be bounded by $k$,
+and for this we need the expectation to be strictly below $k$. A
+careful application of the Chernoff bound for a Poisson binomial
+random variable yields the following concentration bound.
+\begin{lemma}\label{l:concentration}
+Let $S$ be sampled as in Lemma \ref{l:upper} with $\epsilon\leq \frac12$.
+If $s+\frac7{\epsilon^4}\ln^2\!\frac1\epsilon \leq k\leq t_s-1$, then
+$\Pr(|S|>k) \leq \epsilon$.
+\end{lemma}
+Finally, any expected bound for random size DPPs can be converted to
+an expected bound for a fixed size k-DPP via the following
+result.
+\begin{lemma}\label{l:monotonic}
+  For any $\A\in\R^{m\times n}$, $k\in[n]$ and $\alpha>0$, if
+  $S\sim\DPP(\frac1\alpha\A^\top\A)$ and $S'\sim k$-$\DPP(\A^\top\A)$, then
+  \begin{align*}
+    \E\big[\Er_\A(S')\big] \leq \E\big[\Er_\A(S)\mid |S|\leq k\big].
+  \end{align*}
+\end{lemma}
+The above inequality may seem intuitively obvious since adding more columns
+to a set $S$ to complete it to size $k$ always reduces the
+error. However, a priori, it could happen that going from subsets 
+of size $k-1$ to subsets of size $k$ results in a redistribution of
+probabilities to the subsets with larger error. To show that this will
+not happen, our proof relies on classic but non-trivial combinatorial
+bounds called Newton's inequalities. Putting together
+Lemmas~\ref{l:upper}, \ref{l:concentration} and
+\ref{l:monotonic}, we obtain our Master Theorem.
+\begin{proofof}{Theorem}{\ref{t:upper}}
+Let $S\sim\DPP(\frac1\alpha\A^\top\A)$ be sampled as in Lemma
+\ref{l:upper}, and let $S'\sim k$-$\DPP(\A^\top\A)$. We have:\vspace{-2mm}
+\begin{align*}
+  \E\big[\Er_\A(S')\big]
+  &\overset{(a)}{\leq} \E\big[\Er_\A(S)\mid|S|\leq k\big]\\
+  &\leq \frac{\E\big[\Er_\A(S)\big]}{\Pr(|S|\leq k)}
+  \overset{(b)}{\leq} \frac{\Phi_s(k)}{(1-\epsilon)^2}\cdot\opt,
+\end{align*}
+where $(a)$ follows from Lemma \ref{l:monotonic} and $(b)$ follows
+from Lemmas \ref{l:upper} and \ref{l:concentration}. Since
+$0<\epsilon\leq\frac12$, we have
+$\frac1{(1-\epsilon)^2}\leq(1+2\epsilon)^2$, which completes the proof.
+\end{proofof}
+
+\begin{figure}
+  \centering
+  \ifisarxiv
+  \includegraphics[width=0.6\textwidth]{figs/nystrom/sliding}
+  \includegraphics[width=0.6\textwidth]{figs/nystrom/sliding-peak}
+\else  
+  \includegraphics[width=0.5\textwidth]{figs/nystrom/sliding}
+  \includegraphics[width=0.5\textwidth]{figs/nystrom/sliding-peak}
+  \vspace{-6mm}
+  \fi
+  \caption{Illustration of the upper bound functions $\Phi_s(k)$ for
+    different values of $s$, with a $200\times 200$ matrix $\A$ such
+    that the $i$th eigenvalue of $\A^\top\A$ is set to:
+  (top)  $1/i$; (bottom) $1$ for $i<30$ and $0.01$ for $i\geq 30$. For each
+    function, we marked the window of applicable $k$'s with a
+    horizontal line. For polynomial spectral decay (top), the stable rank
+    $\sr_s(\A)$ (i.e., the width of the window starting at $s$)
+    increases, while for the sharp spectrum drop (bottom) the stable
+    rank shrinks as the window approaches the drop, causing a peak in
+    the upper bound.}\vspace{-2mm}
+  \label{f:sliding}
+\end{figure}
+
+We now demonstrate how Theorem \ref{t:upper} can be used as the Master Theorem to derive new
+bounds on the CSSP approximation factor under additional assumptions
+on the singular value decay of matrix $\A$. Rather than a single upper
+bound, Theorem \ref{t:upper} provides a family of upper bounds
+$\Phi_s$, each with a range of applicable values $k$. Since each
+$\Phi_s(k)$ forms a U-shaped curve, its smallest point falls near the
+middle of that range. In Figure \ref{f:sliding} we visualize these bounds as a sliding
+window that sweeps accross the axis representing possible subset
+sizes. The width of the window varies: when it starts at $s$ 
+then its width is the stable rank $\sr_s(\A)$. The wider the window,
+the lower is the valley of the corresponding U-curve. Thus, when
+bounding the approximation factor for a given $k$, we should choose
+the widest window such that $k$ falls near the bottom of its
+U-curve. Showing a guarantee that holds for all $k$ requires lower-bounding the
+stable ranks $\sr_s(\A)$ for each $s$. This is straightforward for
+both polynomial and exponential decay. Specifically, using the
+notation from Theorem \ref{t:decay}, in Appendix \ref{a:decay} we
+prove that:
+\begin{align*}
+  \sr_s(\A) =
+  \begin{cases}
+    \Omega(s/p), &\text{for polynomial rate  $\lambda_i\asymp1/i^p$},\\
+    \Omega(1/\delta),&\text{for exponential rate $\lambda_i\asymp(1-\delta)^i$}.
+    \end{cases}
+  \end{align*}
+As an example, Figure \ref{f:sliding} (top) shows that the stable rank $\sr_s(\A)$, i.e.,
+the width of the window starting at $s$, grows linearly with $s$ for eigenvalues
+decaying polynomially with $p=1$. As a result, the bottom of each
+U-shaped curve remains at roughly the same level, making the CSSP
+approximation factor independent of $k$, as in Theorem
+\ref{t:decay}. In contrast, Figure \ref{f:sliding} (bottom) provides
+the same plot for a different matrix $\A$ with a sharp drop in the spectrum. The U-shaped
+curves cannot slide smoothly accross that drop because of the
+shrinking stable ranks, which results in a peak similar to the ones
+observed in Figure \ref{f:intro}.
+
+ \section{Lower bound}\label{s:lower}
+
+ As discussed in the previous section, our upper bounds for the CSSP
+ approximation factor exhibit a peak (a high point,
+ with the bound decreasing on either side) around a
+ subset size $k$ when there is a sharp drop in the spectrum of $\A$
+ around the $k$th singular value. It is natural to ask whether this
+ peak is an artifact of our analysis, or a property of the k-DPP
+ distribution, or whether even optimal CSSP subsets exhibit this 
+ phenomenon. In this section, we extend a lower
+ bound construction of \citet{pca-volume-sampling} and use it to show
+ that for certain matrices the approximation factor of the optimal CSSP subset, i.e.,
+ $\min_{|S|=k}\Er_\A(S)/\opt$, can exhibit not just one but any
+ number of peaks as a function of $k$, showing that the
+ multiple-descent curve from Figure \ref{f:intro} describes an
+inherent phenomenon in the CSSP.
+
+The lower bound construction of \citet{pca-volume-sampling} relies on
+arranging the column vectors of a $(k+1)\times (k+1)$ matrix $\A$ into a
+centered symmetric $k$-dimensional simplex. This way, the $k+1$
+columns are spanning a $k$ dimensional subspace which contains the $k$
+leading singular vectors of $\A$. They then proceed to shift the
+columns slightly in the direction orthogonal to that subspace so that
+the $(k+1)$st singular value of $\A$ becomes non-zero. This results in
+an instance of the CSSP with a sharp drop in the spectrum. Due
+to the symmetry in this construction, all subsets of size $k$ have an identical squared
+projection error. It is easy to show that this error satisfies $\Er_\A(S)\geq (1-\delta)(k+1)\opt$,
+where $\delta$ is a parameter which depends on the condition number of
+matrix $\A$ and it can be driven arbitrarily close to $0$. Another
+variant of this construction was also provided 
+by \citet{more-efficient-volume-sampling}. The key limitation of both
+of these constructions is that they only provide a lower bound for a
+single subset size $k$ in a given matrix, whereas our goal is to show
+that the CSSP can exhibit the multiple-descent curve, which requires
+lower bounds for multiple different values of $k$ holding with respect
+to the same matrix $\A$.
+
+Our strategy for constructing the lower bound matrix is to concatenate
+together multiple sets of columns, each of which
+represents a simplex spanning some subspace of $\R^m$. The key
+challenge that we face in this approach is that, unlike in
+the construction of \citet{pca-volume-sampling}, different subsets of
+the same size will have different projection errors. Nevertheless,
+we are able to lower bound these errors.
+%, as shown in the following result.
+\begin{lemma}\label{l:lower}
+Fix $\delta\in(0,1)$ and consider unit vectors
+$\a_{i,j}\in\R^m$ in general position, where $i\in[t]$, $j\in[l_i]$, such that
+$\sum_j\a_{i,j}=0$ for each $i$, and for any $i,j,i',j'$, if $i\ne i'$
+then $\a_{i,j}$ is orthogonal to $\a_{i'\!,j'}$. Also, let unit vectors
+$\{\v_i\}_{i\in[t]}$ be orthogonal to  each other and to all
+$\a_{i,j}$. There are positive scalars $\alpha_{i},\beta_i$ for $ i \in [t]$
+such that matrix $\A$ with columns 
+$\alpha_{i}\a_{i,j}+\beta_i\v_i$ over all $i$ and $j$ satisfies:
+\ifisarxiv\else\vspace{-1mm}\fi
+\begin{align*}
+  \min_{|S|=k_i}\frac{\Er_\A(S)}{\mathrm{OPT}_{k_i}}\geq
+  (1-\delta)l_i,\quad\text{for }k_i=l_1+...+l_i-1.
+\end{align*}
+\end{lemma}
+\ifisarxiv\else \vspace{-5mm}\fi
+% Theorem \ref{t:lower} now follows easily from the lemma.
+\begin{proofof}{Theorem}{\ref{t:lower}}
+We let $l_1 = k_1+1$ and then for $i>1$ we set $l_i=k_i-k_{i-1}$. We
+then construct the vectors $\a_{i,j}$ that satisfy Lemma \ref{l:lower} 
+by letting each set $\{\a_{i,j}\}_{j}$ be the corners of a centered
+$(l_i-1)$-dimensional regular simplex. We ensure that each
+simplex is orthogonal to every other simplex by placing them in
+orthogonal subspaces. 
+\end{proofof}
+
+We also use Lemma~\ref{l:lower} in Appendix \ref{s:proofmultidescent}
+to construct a matrix which exhibits the multiple-descent curve
+(Corollary~\ref{c:multiple-descent}).
+
+\section{Empirical evaluation}
+\label{s:experiments}
+
+\begin{figure*}[t]
+  \centering
+  \ifisarxiv
+  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-toy-double}
+   \hspace{-6mm}
+  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-bodyfat-double}
+  \hspace{-6mm}
+  \includegraphics[width=0.345\textwidth]{figs/nystrom/rbf-eunite2001-double}
+  \vspace{-5mm}
+  \else
+    \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-toy-double}
+   \hspace{-6mm}
+  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-bodyfat-double}
+  \hspace{-6mm}
+  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-eunite2001-double}
+  \vspace{-10mm}
+  \fi
+  \caption{Top three plots show the Nystr\"om approximation factor $\E[\|\K-\Kbh(S)\|_*]/\opt$,
+    where $S\sim k$-$\DPP(\K)$ (experiments using greedy selection
+    instead of a k-DPP are in Appendix \ref{a:greedy}), for a toy dataset
+    ($\kappa$ is the condition number) and two
+    Libsvm datasets ($\sigma$ is the RBF parameter). Error bars show three times the standard
+    error  of the mean over 1000 trials. Bottom three plots show the spectral decay for  the
+  top $40$ eigenvalues  of each kernel $\K$. Note that the
+  peaks in the approximation factor align with the drops in
+  the spectrum.}
+\vspace{-2mm}
+  \label{f:rbf}
+\end{figure*}
+
+In this section, we provide an empirical evaluation designed to demonstrate how our improved guarantees for the CSSP and Nystr\"om method, as well as the multiple-descent phenomenon, can be easily observed on real datasets.
+We use a standard experimental setup for data subset selection using
+the Nystr\"om method \citep{revisiting-nystrom}, where an $n\times n$
+kernel matrix $\K$ for a dataset of size $n$ is defined so that the
+entry $(i,j)$ is computed using the Gaussian Radial Basis Function  (RBF)
+kernel:
+$\langle\a_i,\a_j\rangle_\text{K}=\exp(-\|\a_i\!-\!\a_j\|^2/\sigma^2)$,
+where $\sigma$ is a free parameter. 
+We are particularly interested in the effect of varying $\sigma$.
+Nystr\"om subset selection is performed using $S\sim k$-$\DPP(\K)$
+(Definition \ref{d:dpp}), and we plot the expected approximation
+factor $\E[\|\K-\Kbh(S)\|_*]/\opt$ (averaged over 1000 runs), where
+$\Kbh(S)$ is the Nystr\"om approximation of $\K$ based on the subset
+$S$ (see Section~\ref{s:nystrom}), $\|\cdot\|_*$ is the trace norm,
+and $\opt$ is the trace norm error of the best rank $k$
+approximation. Additional experiments, using greedy selection instead 
+of a k-DPP, are in Appendix~\ref{a:greedy}.
+As discussed in Section \ref{s:nystrom}, this task is
+equivalent to the CSSP task defined on the matrix $\A$ such that
+$\K=\A^\top\A$.
+
+The aim of our empirical evaluation is to verify the following two claims motivated by our theory (and to illustrate that doing so is as easy as varying the RBF parameter $\sigma$):
+\ifisarxiv\else\vspace{-3mm}\fi
+\begin{enumerate}
+  \item When the spectral decay is sufficiently slow/smooth, the
+    approximation factor for CSSP/Nystr\"om is much better than
+    suggested by previous worst-case bounds.
+    \vspace{-1mm}
+  \item  A drop in spectrum around the $k$th eigenvalue results in
+    a peak in the approximation factor near
+    subset size~$k$. Several drops result in the
+    multiple-descent curve.
+  \end{enumerate}
+\ifisarxiv\else  \vspace{-3mm}\fi
+In Figure \ref{f:rbf} (top), we plot the approximation factor against
+the subset size $k$ (in the range of 1 to 40) for an artificial toy
+dataset and %\citep[following][]{revisiting-nystrom}
+for two benchmark regression datasets from the Libsvm repository
+\citep[\emph{bodyfat} and \emph{eunite2001}, see][]{libsvm}. 
+The toy dataset is constructed by scaling the eigenvalues of a random
+$50\times 50$ Gaussian matrix so that the spectrum is flat with a
+single drop at the $21$-st eigenvalue. 
+For each dataset, in Figure~\ref{f:rbf} (bottom), we also show the top
+40 eigenvalues of the kernel $\K$ in decreasing order. 
+For the toy dataset, to maintain full control over the spectrum we use
+the linear kernel $\langle\a_i,\a_j\rangle_\text{K}=\a_i^\top\a_j$, and we show results
+for three different values of the condition number $\kappa$ of kernel
+$\K$.  
+For the benchmark datasets, we show results on the RBF kernel with 
+three different values of the parameter~$\sigma$.
+
+Examining the toy dataset (Figure~\ref{f:rbf}, left), it is apparent
+that a larger drop in spectrum leads to a sharper peak in the
+approximation factor as a function of the subset size~$k$, whereas a flat spectrum results in  
+the approximation factor being close to $1$. A similar trend is observed for
+dataset \emph{bodyfat} (Figure~\ref{f:rbf}, center), where large
+parameter $\sigma$ results in a peak that is aligned with a spectrum
+drop, while 
+decreasing $\sigma$ makes the spectrum flatter and the factor closer to
+$1$. Finally, dataset \emph{eunite2001} (Figure~\ref{f:rbf}, right) exhibits a full
+multiple-descent curve with up to three peaks for large values of
+$\sigma$, and the peaks are once again aligned with the spectrum
+drops. Decreasing $\sigma$ gradually eliminates the peaks, 
+resulting in a uniformly small approximation factor. Thus, both of our theoretical
+claims can easily be verified on this dataset simply by adjusting the RBF parameter.
+
+While the right choice of the parameter $\sigma$ ultimately depends on
+the downstream machine learning task, it has been observed that
+varying $\sigma$ has a pronounced effect on the spectral properties of
+the kernel matrix, \citep[see,
+e.g.,][]{revisiting-nystrom,LBM16_TR,wang2019block}.  
+The main takeaway from our results here is that, depending on the
+structure of the problem, we may end up in the regime where the
+Nystr\"om approximation factor exhibits a multiple-descent curve
+(e.g., due to a hierarchical nature of the data) or in the regime
+where it is relatively flat.
+
+\section{Conclusions and open problems}
+\label{s:conclusions}
+We derived new guarantees for the Column Subset Selection Problem (CSSP) and
+the Nystr\"om method, going beyond worst-case analysis by exploiting the
+structural properties of a dataset, e.g., when the spectrum exhibits a
+known rate of decay. Our upper and lower bounds
+for the CSSP/Nystr\"om approximation factor reveal an intruiguing phenomenon we
+call the multiple-descent curve: the approximation factor can exhibit a highly
+non-monotonic behavior as a function of $k$, with multiple
+peaks and valleys. These observations suggest a connection to the
+double descent curve exhibited by
+the generalization error of many machine learning models (see
+Section~\ref{s:related-work}). This new 
+connection is remarkable, since, unlike generalization error, the CSSP
+approximation factor is a deterministic objective in a combinatorial
+optimization problem without any underlying statistical model. 
+% However, whereas double descent is characterized by a single sharp
+% transition, we have observed that multiple CSSP sharp transitions can
+% be present in one dataset. Thus, a natural question arises: Can the
+% multiple-descent behavior be observed also for the generalization
+% error of learning models?
+
+Our analysis technique relies on converting an error bound
+from random-size DPPs to fixed-size k-DPPs, which results in an
+additional constant factor of $(1+2\epsilon)^2$ in Theorem~\ref{t:upper}.
+% We
+% believe that this factor (and any dependence on $\epsilon$) can be
+% eliminated from this bound. In fact, w
+We put forward a
+conjecture which would eliminate this factor from
+Theorem~\ref{t:upper} and is of independent interest to the study of 
+elementary symmetric polynomials, a classical topic in combinatorics
+\citep{HLP-Inequalities}. 
+\begin{conjecture}\label{c:convex}
+  The following function is \underline{convex} with respect to $k\in[n]$ for any
+  positive sequence $\lambda_1,...,\lambda_n$:
+% For any sequence $\lambda_1,..,\lambda_n>0$, the function
+ \[f(k)=(k+1)\frac{\sum_{S:|S|=k+1}\prod_{i\in S}\lambda_i}{\sum_{S:|S|=k}\prod_{i\in S}\lambda_i}.\]
+   %\frac{e_{k+1}(\lambda)}{e_k(\lambda)}\]
+\end{conjecture}
+\citet{pca-volume-sampling} showed that if $S\sim
+k$-$\DPP(\A^\top\A)$ and $\lambda_i$ are the eigenvalues of
+$\A^\top\A$, then $\E[\Er_\A(S)]=f(k)$. If $f(k)$ is convex then
+Jensen's inequality implies:
+\begin{align*}
+  \E[\Er_\A(S)]\leq \E[\Er_\A(S')]\quad\text{for }S'\!\sim\DPP(\tfrac1{\alpha_k}\A^\top\A),
+\end{align*}
+where $\alpha_k$ is chosen so that $\E[|S'|]=k$. This would allow us to use
+the bound from Lemma \ref{l:upper} directly on a k-DPP
+without relying on the concentration argument of
+Lemma~\ref{l:concentration}, thereby improving the bounds in
+Theorems \ref{t:upper} and \ref{t:decay}.
+% We leave
+% Conjecture \ref{c:convex} as a new direction for future work.
+
+\subsubsection*{Acknowledgements}
+
+MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
+  support of this work. Also, the authors thank the NSF for funding
+  via the NSF TRIPODS program.
+
+\bibliographystyle{sty/icml2020/icml2020}
+\bibliography{pap}
+
+  \newpage
+  \appendix
+  \onecolumn
   \section{Proof of Lemma \ref{l:expectation_proj}}
-\label{a:expectration-proj}
   We will use the following standard determinantal summation identity
@@
   \leq \frac{\E\big[\Er_\A(S')\big]}{\Pr(|S'|\leq k)}
-  \\
-  &\leq \big(1-\ee^{-\frac{\epsilon^2b}{10}}\big)^{-1}(1-\epsilon)^{-1}
+  \leq \big(1-\ee^{-\frac{\epsilon^2b}{10}}\big)^{-1}(1-\epsilon)^{-1}
 \frac{\lambda_{s+1}}{\lambda_n}\Big(1+\frac s{k-s}\Big)\cdot\opt,
@@
 
+  % To lower bound the stable rank, we assume that the series
+  % $\{\lambda_i\}$ goes beyond $n$ by using $c_1  i^{-p} \leq
+  % \lambda_i \leq c_2 i^{-p} \;\;\forall i\in\N$. Using this
+  % extended series, we can write the stable rank as:
+
+  % \[ \sr_s(\A) = \frac{\sum_{i=s+1}^n \lambda_i}{\lambda_{s+1}} =\frac1{\lambda_{s+1}} \left( \sum_{i > s} \lambda_i - \sum_{i>n} \lambda_i \right).\]
+
+  % From the lower bound obtained from the lower integral:
+
+  % \[ \frac{\sum_{i>s} \lambda_i}{\lambda_{s+1}} \geq \frac{c_1}{c_2}\frac{(s+2)^{1-p}}{(p-1) (s+1)^{-p}} = \frac{c_1}{c_2}\frac{s+2}{p-1} \left(\frac{s+1}{s+2}\right)^p =  \frac{c_1}{c_2}\frac{s+2}{p-1} \left(1 - \frac1{s+2}\right)^p  \geq \frac{c_1}{c_2}\frac{s+2}{p-1}(1 - \frac{p}{s+2}) \geq \frac{c_1}{c_2}\frac{s+1}{p-1} - \frac{c_1}{c_2}.\]
+
+  % Similarly for the second part, for any $s\leq k$, using the
+  % condition of Case 1(a) we have
+
+  % \[\frac{\sum_{i>n} \lambda_i}{\lambda_{s+1}} \leq
+  %   \frac{c_2}{c_1}\frac{(n+1)^{1-p}}{(p-1) (s+1)^{-p}} =
+  %   \frac{c_2}{c_1} \frac{s+1}{p-1}\left(\frac{s+1}{n+1}\right)^{p-1}
+  %   \leq \frac{c_2}{c_1}\frac{c^2_1}{2c^2_2} \frac{s+1}{p-1} =
+  %   \frac{c_1}{2c_2} \frac{s+1}{p-1} . \]
+
+  % Combining, we can get the following lower bound on the stable rank:
+
+  % \[\sr_s(\A) \geq \frac{c_1}{2c_2} \frac{s+1}{p-1} - \frac{c_1}{c_2}
+  %   = \frac{1}{2\gamma} \, \frac{s+1}{p-1} - \frac1\gamma,  \]
+
+  % where recall that  $\gamma = \frac{c_2}{c_1}$.
   Further using $u =
@@
   $\hat{u}\geq 60$ which is bigger than $\frac{7}{\epsilon^4} \ln^2 \frac{1}{\epsilon}$.
+  % Further, $(1+2\epsilon)^2 \leq 4$
 If not, then $\delta\geq\frac4{60\gamma}>\frac1\gamma$ and
@@
 
+% \section{Proof of Lemma~\ref{l:lower}}
+
+% \begin{proof}
+% 	 Say $\Abh$ is the matrix obtained by stacking all the $\alpha_i\a_{i,j}$.
+We write $\tilde\a_{i,j} = \alpha_i\a_{i,j}+\beta \v_i$.
+Let $\w_i$, $i\in[t]$ be vectors of size $\sum_i l_i$, with $1s$ at positions $\sum_{i<c}l_i +1$ to $l_c$, and $0$s elsewhere.
+Note that $\w_i$ is an (unnormalized) eigenvector of $\Abh^\top\Abh$ with eigenvalue $0$.
+Further,
+
+% %\michal{How does this equation not depend on $\alpha_i$?}
+% 	\begin{align*}
+% 	\A^\top \A =	\Abh^\top\Abh  + \sum_i \beta_i^2 \w_i
+%           \w_i^\top.
+% 	\end{align*}
+
+% 	Thus, $\A^\top\A$ has the same eigenvectors and eigenvalues as $\Abh^\top\Abh$ except the repeated eigenvalue $0$ has been replaced with $l_i\beta_i^2$.
+
+% 	 We prove the result for an arbitrary $i=c$, so that $|S|= k_c$.
+Say $\P^\star$ is the span of $\{\a_{1,1}, \a_{1,2}, \ldots \a_{1,l_1}, \a_{2,1}, \ldots, \a_{c,l_c} \}$.
+If $\beta_i^2 < \lambda/l_i$, where $\lambda$ is the smallest non-zero eigenvalue of $\Abh^\top\Abh$, and since we can choose  $\beta_{i+1} < \beta_{i}$ as well as  $\alpha_{i+1}<\alpha_i \;\forall i$, then
+% 	\begin{align*}
+% 	\mathrm{OPT}_{k_c} = \| \A - \P^\star \A \|_F^2 =
+%           l_c\beta^2 + \phi_c,  
+% 	\end{align*}
+
+% 	where $\phi_c =\sum_{i>c} l_i (\alpha_i^2 + \beta_i^2) $ for ease of notation.
+
+% 	The centroid of $\{\tilde\a_{c,1}, \ldots, \tilde\a_{c,l_c} \}$ is $\beta_c \v_c$. We can write 
+% 	$\tilde\a_{c, l_c} = l_c \beta_c\v - \sum_{j < l_c} \tilde\a_{c,j}$.
+% 	For selecting the set $S\subset [n]$ of size $k_c$, since
+%         $\alpha_i>\alpha_{i+1}$, we can assume without loss of
+%         generality that $\P_S$ is a projection onto the span of
+%         $\Big(\bigcup_{i<c} \bigcup_j \{\tilde\a_{i,j}\}\Big) \cup
+%         \{\tilde\a_{c,1},\ldots,\tilde\a_{c,l_{c-1}}\}$.
+
+%       %  \michal{The size of set $S$ is not $k_c$}	
+% 	\begin{align*}
+% 	\Er_{\A}(S) &= \| \tilde\a_{c,l_c}- \P_S \tilde\a_{c,l_c}  \|_2^2
+% 	+  \sum_{i>c}\sum_{j=1}^{l_i}\| \tilde\a_{i,j} - \P_S
+% 	\tilde\a_{i,j} \|_2^2\\ 
+% 	&=\bigg\|  l_c\beta_c\v_c - \sum_{j < l_c} \tilde\a_{c,j} - \P_S
+% 	\Big( l_c\beta_c\v_c- \sum_{j < l_c} \tilde\a_{c,j} \Big)\bigg\|_2^2 +  \phi_c
+% 	\\ 
+% 	&= l_c^2\beta_c^2 \|\v_c - \P_S\v_c\|^2_2 + \phi_c \\
+% 	& = l_c(\mathrm{OPT}_{k_c} - \phi_c) \|\v_c - \P_S\v_c\|^2_2 + \phi_c \\
+% 	& = l_c \mathrm{OPT}_{k_c} \|\v_c - \P_S\v_c\|^2_2 + \phi_c (1 - l_c\|\v_c - \P_S\v_c\|^2_2 ) \\
+% 	& \geq l_c \mathrm{OPT}_{k_c} \|\v_c - \P_S\v_c\|^2_2 - (l_c-1) \phi_c 
+% 	\end{align*}	
+% 	Note that $\lim_{\beta\rightarrow 0}\P_S\v = \zero$, so there is
+% 	$\beta_S>0$ such that for any $\beta\leq\beta_S$ we have
+% 	$\|\v-\P_S\v\|^2\geq 1-\frac\epsilon2$.
+We choose $\beta^2_{c}  = 0.9* \min\left\{ \beta^2_{c-1}, \frac{\frac{\epsilon}{4} l_j \beta_{c-1}^2 }{(1-\frac{\epsilon}{2}) l_{j+1}t}, \min_S \beta^2_S \right\} $.
+
+% 	Further, given $\epsilon$, we select $\alpha_1=1$ and $\beta_1 = 0.9*\lambda/l_1$, and $ \forall j\geq 1,\,\, \alpha^2_{j+1}  = 0.9* \min\left\{ \alpha^2_j, \frac{\frac{\epsilon}{4} l_j \beta_j^2 }{(1-\frac{\epsilon}{2}) l_{j+1}t} \right\} $ and $\beta_{j+1}$ as above so that:
+% 	\begin{align*}
+% %	&  \alpha_{j+1} \leq \frac{\frac{\epsilon}{2} (l_j+1) \beta^2 }{(1-\frac{\epsilon}{2}) (l_{j+1}+1)t}\\
+% 	& (1-\frac{\epsilon}{2}) l_{j+1} (\alpha^2_{j+1}+\beta_{j+1}^2) \leq \frac{\frac{\epsilon}{2} l_j \beta^2 }{t}\\
+% 	\implies & \sum_{i>c}(1-\frac{\epsilon}{2}) l_{j+1} (\alpha^2_{j+1}+\beta_{j+1}^2)\leq (t-c)\frac{\frac{\epsilon}{2} l_c \beta^2 }{t} \leq \frac{\epsilon}{2} l_c \beta^2\\
+% 	\implies & (1-\frac{\epsilon}{2})\phi_c  \leq  \frac{\epsilon}{2} l_c \beta^2 \\
+% 	\implies & \phi_c \leq \frac{\epsilon}{2} \mathrm{OPT}_{k_c} \\
+% 	\implies & (l_c-1)\phi_c \leq \frac{\epsilon}{2} l_c \mathrm{OPT}_{k_c},
+% 	\end{align*}
+% which gives the desired result.
+% \end{proof}
+
 \section{Proof of Corollary~\ref{c:multiple-descent}}
@@
   To see the usefulness of condition 3, note that each group of
-  vectors in column set of  $\A$ constructed from Theorem \ref{t:lower} form a
+  vectors in column set of  $\A$ constructed from Theorem 3 form a
   shifted regular simplex. A regular simplex has the smallest
@@
  In this section, we provide a more detailed empirical evaluation to
- complement what we presented in Section~\ref{s:experiments}. Our aim here is to demonstrate
+ complement what we presented in Section~6. Our aim here is to demonstrate
 that our improved analysis of the CSSP/Nystr\"om approximation factor
@@
 \end{algorithm}
+\vspace{-4mm}
 In our empirical evaluation we use the same experimental setup as in
@@
   \else
-  ~\hspace{-5mm}
-  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-toy-greedy}\hspace{-5mm}\nolinebreak\includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-bodyfat-greedy}\hspace{-5mm}\nolinebreak\includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-eunite2001-greedy}
+  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-toy-greedy}
+   \hspace{-6mm}
+  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-bodyfat-greedy}
+  \hspace{-6mm}
+  \includegraphics[width=0.35\textwidth]{figs/nystrom/rbf-eunite2001-greedy}
+  \vspace{-10mm}
   \fi

### 2022-08-23 design/nystrom.tex
Commit original repos

\documentclass[12pt]{sty/colt2019/colt2018-arxiv}
% \else
% \documentclass[anon,12pt]{sty/colt2019/colt2019} % Anonymized submission
% \documentclass[12pt]{colt2019} % Include author names
% \fi

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

% \title[Short Title]{Full Title of Article}
\title{Row subset selection and the Nystr\"om method}
\usepackage{times}

\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{xfrac}
\usepackage{algorithmic}
\input{shortdefs}

\coltauthor{\Name{Micha{\l } Derezi\'{n}ski}
  \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics, UC Berkeley
}

\begin{document}

\maketitle

\section{Introduction}

\subsection{Nystr\"om}
Let $\L$ be a p.s.d.~$n\times n$ matrix with eigendecomposition
$\U\D\U^\top$, where the columns of $\U$ are the eigenvectors and
$\D=\diag(\lambda_1,...,\lambda_n)$ contains the eigenvalues. For any
unitarily invariant norm $\|\cdot\|$, the optimal rank $k$
approximation to $\L$ is
\begin{align*}
  \L_k =\U\diag(\lambda_1,...,\lambda_k,0,...,0)\U^\top,
\end{align*}
in the sense of minimizing $\|\L-\L_k\|$ among all rank $k$
matrices. To avoid the full eigendecomposition, the Nystr\"om method
was introduced, which constructs a rank $k$ approximation of $\L$
by only using the decomposition of a small $k\times k$ submatrix
$\L_{S,S}$ for some index subset $S$. 
\begin{definition}
 We define the Nystr\"om approximation of $\L$ based on a subset $S$
 as the $n \times n$ matrix $\widehat\L(S)=\L_{\cdot,S}\L_{S,S}^\dagger\L_{S,\cdot}$.
\end{definition}
An example application is a variational Gaussian Process approximation
$\widehat\L(S)$ for which \cite{sparse-variational-gp} showed that the
Kullback-Leibler divergence measuring the quality of the 
approximation can be represented (up to a factor of 2) by the nuclear (trace)
norm of the difference: $\|\L-\widehat\L(S)\|_*$.

\subsection{Connecting projections to Nystr\"om}
Given an $n\times d$ matrix $\X$ and subset
$S\subseteq [n]$ of size $k\leq d$, let $\P_S =
\X_S^\top(\X_S\X_S^\top)^{-1}\X_S$ denote the projection onto
the span of $\{\x_i:i\in S\}$.
The goal in row subset selection is often to minimize $L(S) =
\|\X-\X\P_S\|_F^2$ subject to $|S|=k$ (other norms are also
considered). We show that this problem is 
equivalent to minimizing the nuclear norm for the Nystr\"om
approximation of $\L=\X\X^\top$.
\begin{lemma}
For any $\X$ and $S\subseteq [n]$, we have $L(S) = \|\L -
\widehat\L(S)\|_*$ where $\L=\X\X^\top$.
\end{lemma}
\begin{proof}
\begin{align*}
L(S)&=\|\X(\I-\P_S)\|_F^2 = \tr((\I - \P_S)\X^\top\X(\I-\P_S))=
  \tr(\X(\I-\P_S) \X^\top)\\
  & = \tr(\X\X^\top - \X\X_S^\top(\X_S\X_S^\top)^{-1}\X_S\X^\top) =
    \tr(\L-\widehat\L(S)) = \|\L-\widehat\L(S)\|_*.
\end{align*}  
\end{proof}
\subsection{Optimal bounds using DPPs}
%From now on, w.l.o.g., we use the $\X$ notation instead of the $\L$ notation.
Denoting $\lambda_1,\dots,\lambda_d$ as the decreasing eigenvalues of
$\L$, we define the optimal rank-$k$ projection loss (not based on a subset) as
follows:
\begin{align*}
  E_k^*\defeq \min_{\text{proj. }\P:\,\rank(\P)=k}\|\X-\X\P\|_F^2 = \|\L-\L_k\|_* =
  \sum_{i=k+1}^d\lambda_i.
\end{align*}
\begin{definition} Define $S\sim \DPP(\L)$
  over all subsets $S\subseteq [n]$ so
  that \[\Pr(S)=\frac{\det(\L_{S,S})}{\det(\L+\I)}.\]
  \end{definition}
Prior work showed the following result when sampling from a
cardinality constrained DPP. The below result  was first shown by
\cite{pca-volume-sampling}, then later adapted to the context of
Nystr\"om approximations by \cite{belabbas-wolfe09}. 
\begin{theorem}[\cite{pca-volume-sampling}]\label{t:simple}
If $S\sim \DPP(\L)$, then
  \begin{align*}
    \E\big[L(S)\mid |S|=k\big]\leq (k+1)\cdot E_k^*.
  \end{align*}
\end{theorem}
Moreover, \cite{pca-volume-sampling} showed that  the factor $k+1$
cannot in general be improved when comparing to $E_k^*$. 
\begin{theorem}[\cite{pca-volume-sampling}]\label{t:lower}
For any $\epsilon>0$, there is a $k+1\times k+1$  matrix $\L$ such
that for every subset $S$ of size $k$ we have
\begin{align*}
  L(S)\geq (1-\epsilon)(k+1)\cdot E_k^*.
\end{align*}
\end{theorem}
We show a result closely related to Theorem \ref{t:simple}, where DPP
is not cardinality constrained but rather rescaled by a parameter
$\alpha$ which controls the \emph{expected} subset size.
\begin{theorem}\label{t:main}
For any $\X\in\R^{n\times d}$ and $\alpha>0$, if $S\sim
\DPP(\frac1\alpha\L)$, then
\begin{align*}
  \E\big[L(S)\big] =
  \alpha\cdot\E\big[|S|\big],\quad\E\big[|S|\big]=\tr\big(\L(\alpha\I+\L)^{-1}\big). 
\end{align*}
For $\alpha=E_k^*$ we have $\E\big[|S|\big]\leq k+1$ and the bound
becomes: $\E\big[L(S)\big]\leq (k+1)\cdot E_k^*$.
\end{theorem}

\subsection{Related work for selecting $r>k$ rows}

A result related to Theorem \ref{t:simple} comparing the expected
error of subsets of size $|S|=r>k$ to $E_k^*$ was given by
\cite{more-efficient-volume-sampling}. A more comprehensive analysis
of deterministic methods of row subset selection with $r>k$ for a variety of
norms was given by \cite{near-optimal-columns}. From the perspective
of the Nystr\"om method, this problem was also studied by
\cite{revisiting-nystrom} who used leverage score sampling as one of
the approaches.

\subsection{Open problem}
Denote the loss of the best subset of size $k$ as $L_k^* \defeq
\min_{S:|S|=k} L(S)$. Naturally we have $E_k^*\leq L_k^*$, so the
existing results show that:
\begin{align*}
  \E\big[L(S)\big] \leq (k+1)\cdot L_k^*
\end{align*}
However, the lower bound of Theorem \ref{t:lower} does not apply
here. So  the following remains open.

\textbf{Question:} When is it possible to efficiently find $S$ of size
$k$ such that $L(S)\leq (1+\epsilon)\cdot L_k^*$?

\section{Application of Theorem \ref{t:main}}

\subsection{Identical eigenvalues}

Suppose that all $d$ non-zero eigenvalues  of $\L=\X\X^\top$ are 1. In
this case we have $E_k^* = d-k$. Now, for some $\epsilon>0$ we choose
$\alpha>0$ such that
\begin{align*}
(1-\epsilon) k = \E\big[|S|\big] = \frac{d}{1+\alpha}.
\end{align*}
So, we have $\alpha = \frac d{(1-\epsilon)k}-1$. By Theorem
\ref{t:main},
\begin{align*}
  \E\big[L(S)\big] = \alpha\, \E\big[|S|\big] =
\Big(\frac d{(1-\epsilon)k}-1\Big)\,(1-\epsilon) k = d - (1-\epsilon)k
  = E_k^*(1 + \epsilon\tfrac k{d-k}).
\end{align*}
Note that for $\epsilon=0$, we have $\E[|S|]=k$ and $\E[L(S)]=E_k^*$,
however we introduce $\epsilon$ to ensure that $|S|\leq k$ with
sufficient probability. From the concentration bound for a Poisson binomial r.v.~we get
\begin{align*}
  \Pr(|S| > k) \leq \ee^{-\frac{(k-(1-\epsilon)k))^2}{2k}} =
  \ee^{-\epsilon^2 k/2}
\end{align*}
Now, it follows that:
\begin{align*}
\E\big[L(S)\mid |S|\leq k\big] \leq \frac{\E[L(S)]}{\Pr(|S|\leq k)}
  \leq \frac{1+\epsilon\frac k{d-k}}{1-\ee^{-\epsilon^2k/2}}\cdot E_k^*.
\end{align*}
In particular, this implies that for any $k$ we can find $S$ of size
$k$ such that $L(S)-E_k^* = O(1) \cdot E_k^*$.

\subsection{Hard example}
Suppose that $d-1$ eigenvalues are 1 and last eigenvalue is $\delta\ll
1$. Then $E_k^*=\delta+d-k-1$. We proceed similarly (omitting
$\epsilon$ for simplicity): 
\begin{align*}
k = \E\big[|S|\big] = \frac{d-1}{1+\alpha} +
  \frac{\delta}{\delta+\alpha} = \frac{d-1+\delta}{1+\alpha} +
  \frac{\delta(1-\delta)}{(\delta+\alpha)(1+\alpha)}.
\end{align*}
From this it follows that $\alpha=\frac{E_k^*}{k} +
\frac{\delta(1-\delta)}{k(\delta+\alpha)}$ and so we obtain that:
\begin{align*}
  \E\big[L(S)] = E_k^* +  \frac{\delta(1-\delta)}{\delta+\alpha}
  \leq E_k^* + 1 = \delta +d - k.
\end{align*}
So if $k\leq d/2$ then $\E[L(S)] = O(E_k^*)$. However, let us
consider $k=d-1$. Then $E_k^*=\delta$ and 
\begin{align*}
  \E[L(S)] = E_k^*(1 +  \tfrac{1-\delta}{\delta+\alpha}).
\end{align*}
Note that $\delta$ and $\alpha$ are defined in an entangled way. We
use this to get:
\begin{align*}
  \frac{\delta}{\delta+\alpha} = k - \frac{d-1}{1+\alpha} = (d-1)\frac{\alpha}{1+\alpha}
\end{align*}
and further transformation shows that:
\begin{align*}
\frac1{\delta+\alpha} = \frac1\alpha\Big(1 -
  (d-1)\frac\alpha{1+\alpha}\Big) = \frac1\alpha -
  \frac{d-1}{1+\alpha}.
  \end{align*}
  Putting the above two equations together we obtain:
  \begin{align*}
    \frac{1-\delta}{\delta+\alpha} = \frac1\alpha -
    \frac{d-1}{1+\alpha} - (d-1)\frac\alpha{1+\alpha} = \frac1\alpha
   - (d-1).
  \end{align*}
By making $\delta$ arbitrarily small we can also force $\alpha$ to be
arbitrarily small so:
\begin{align*}
  \frac{\E[L(S)]}{E_k^*}= 1+  \big(\tfrac1\alpha - (d-1)\big)
  \underset{\alpha\rightarrow 0}{\longrightarrow} \infty.
\end{align*}
This eigenvalue structure matches the construction used to obtain the
lower bound from Theorem \ref{t:lower}. So as we can see, our bound is
weaker than that of Theorem \ref{t:simple} in this very minor corner case.

\subsection{Improvement for exponentially decaying eigenvalues}
Suppose that the eigenvalues of $\L=\X\X^\top$
exhibit exponential decay: $\lambda_i=\gamma^{i}$, and for
simplicity assume that $d\rightarrow\infty$ so that the stable rank is 
$r=\sum_i\frac{\lambda_i}{\lambda_1} = \frac1{1-\gamma}$. Setting $\alpha =
\gamma^t$, if $S\sim\DPP(\frac1\alpha\L)$ then
\begin{align*}
  \E\big[|S|\big] \leq t + \gamma^{-t}\sum_{i>t}\gamma^i =
  t + \frac\gamma{1-\gamma} = t +\gamma\cdot r,
\end{align*}
Letting $k=\E[|S|\big]\geq r$ and noting that
$E_k^*=\gamma^k\cdot\gamma r$, Theorem \ref{t:main} obtains the following bound: 
\begin{align*}
  \E\big[L(S)\big] = k\cdot \gamma^{k-\frac\gamma{1-\gamma}}\leq \ee
  k\gamma^k = \frac{\ee k}{\gamma r}E_k^*,
\end{align*}
better than the bound obtained by \cite{pca-volume-sampling}
when $r\gg 1$ (i.e., decay is slow). In particular, we
achieve that if $k = O(r)$ then $\E[L(S)] = O(E_k^*)$.

 Concentration argument seems straight forward. We take $\E(|S|) = (1-\epsilon)k$, so that $t \geq (1-\epsilon)k - \gamma r$. We have, 
 for the expected loss,

 \begin{align*}
 \E[L(S)] = \alpha \E[|S|] =  (1- \epsilon) k \gamma^t \leq  (1- \epsilon)k\cdot \gamma^{(1-\epsilon)k-\frac\gamma{1-\gamma}}\leq \ee
  (1- \epsilon)k\gamma^{(1-\epsilon)k} = \frac{\ee  (1- \epsilon)k}{\gamma^(\epsilon k) r}E_k^* .
 \end{align*}

 Then, 
 \begin{align*}
 \E\big[L(S)\mid |S|\leq k\big] \leq \frac{\E[L(S)]}{\Pr(|S|\leq k)}
 \leq \frac{\ee  (1- \epsilon)k}{\gamma^{k\epsilon}\gamma r(1-\ee^{-\epsilon^2k/2})}\cdot E_k^*.
 \end{align*}

\subsection{General result for stable rank} 
Based on the above examples, we formulate the following result.

\begin{theorem}
For any $\lambda_1\geq...\geq\lambda_d\geq 0$, $\alpha>0$ and $k\in[d]$, define
\begin{align*}
  k_\alpha=\sum_i\frac{\lambda_i}{\lambda_i+\alpha},\quad
  r =\sum_i\frac{\lambda_i}{\lambda_1},\quad\text{and}\quad E_{k}^* =
  \sum_{i>k}\lambda_i.
  \end{align*}
For any $\epsilon>0$, if $k\leq \frac{\epsilon}{1+\epsilon}\cdot r$, then there is $\alpha$ such that
    $k_\alpha\leq k$ and  $\alpha k_\alpha\leq (1+\epsilon) E_k^*$.
 \end{theorem}
\begin{proof}
We have
  \begin{align*}
E_{k}^*=
    \frac{\lambda_1}{\lambda_1}\sum_{i>k}\lambda_i =\lambda_1\Big( r -
 \frac{\sum_{i\leq k}\lambda_i}{\lambda_1}\Big)\geq \lambda_1\Big( r -  \sum_{i\leq k}\frac{\lambda_1}{\lambda_1}\Big) =\lambda_1( r -
    k).
  \end{align*}
  Now we let $\alpha = (1+\epsilon) E_k^*/k$ so that $\alpha k
  =(1+\epsilon)E_k^*$ and moreover: 
  \begin{align*}
    k_\alpha=\sum_i\frac{\lambda_i}{\lambda_i+(1+\epsilon)E_{k}^*/k}
    \leq \sum_i\frac{\lambda_i}{\lambda_i+(1+\epsilon)\lambda_1(\frac rk-1)}
    \leq\sum_i\frac{\lambda_i}{(1+\epsilon)\lambda_1\frac{r-k}k}=\frac{r}{r-k}\cdot
    \frac k{1+\epsilon}.
  \end{align*}
  Note that $k$ is chosen so that $\frac
  r{r-k}\,\frac1{1+\epsilon}\leq 1$.
\end{proof}

\subsubsection{Concentration}

\begin{lemma} With $r$ as the stable rank and any $\epsilon > 0$, for $ \frac 2{\epsilon^2} \ln \frac1{\epsilon} \leq k \leq  \frac {\epsilon }{(1+\epsilon)}  r $, we can find a set $S$ of size $|S|\leq k$ with $L(S)  \leq (1+3\epsilon)  E^*_k$ with $O(\frac1{\epsilon})$ rounds of rejection sampling on average. 
\end{lemma}

Use $\alpha = \frac{(1+\epsilon)}{1-\delta} \frac{E^*_k}{k}$. This makes sure that,

\begin{align*}
\E(|S|) = \sum_i\frac{\lambda_i}{\lambda_i+(1+\epsilon)E_{k}^*/k(1-\delta)}
\leq \sum_i\frac{\lambda_i}{\lambda_i+\frac{(1+\epsilon)}{1-\delta}\lambda_1(\frac rk-1)}
\leq\sum_i\frac{\lambda_i}{\frac{(1+\epsilon)}{1-\delta}\lambda_1\frac{r-k}k}=\frac{(1-\delta)r}{r-k}\cdot
\frac k{1+\epsilon},
\end{align*}

which implies $\E(|S|) \leq (1-\delta)k$, since $k$ is chosen to ensure that $\frac{r}{r-k}.\frac{1}{1+\epsilon} \leq 1$. This gives $\E(L(S)) = \alpha \E(|S|) = \frac{(1+\epsilon)}{1-\delta} \frac{E^*_k}{k} \E(|S|) \leq \frac{(1+\epsilon)}{1-\delta} \frac{E^*_k}{k} (1-\delta) k = (1+\epsilon) E^*_k$.

 For concentration ,

\begin{align*}
	& \E\big[L(S)\mid |S|\leq k\big] \leq \frac{\E[L(S)]}{\Pr(|S|\leq k)} \leq \frac {(1+\epsilon)E^*_k}{(1-\ee^{-\delta^2k/2}) }.  
\end{align*}

With $\delta \leq \epsilon$, and $k \geq \frac{2}{\epsilon^2} \ln \left(2 + \frac1{\epsilon} \right)$, which ensures $\ee^{-\delta^2 k/2} \leq \frac{\epsilon}{1+2\epsilon}$, we get, $\E\big[L(S)\mid |S|\leq k\big] \leq (1+2\epsilon) E^*_k$.

Note that for $k \geq \frac{2}{\epsilon^2} \ln \left(2 + \frac1{\epsilon} \right)$, we have $|S| \leq k$ with probability $\frac{1+\epsilon}{1+2\epsilon}$. Using shorthand $E_k=\E\big[L(S)\mid |S|\leq k\big]$ in the Markov inequality, we have,

\[ P\big[L(S) \geq \eta E_k \mid |S| \leq k \big] \leq \frac{1}{\eta}. \]

Thus, with probability $\geq  \frac{1+\epsilon}{1+2\epsilon} (1 - \frac1{\eta})$, we have $|S| \leq k$ and  $L(S) \leq \eta E_k$. Choosing $\eta = (1+\frac \epsilon {1+2\epsilon})$ ensures $L(S) \leq (1+3\epsilon)E^*_k$ with probability $\geq  \frac{(1+\epsilon)\epsilon}{(1+2\epsilon)(1+3\epsilon)}$, which means on average $\frac{(1+2\epsilon)(1+3\epsilon)}{(1+\epsilon)\epsilon} = O(1/\epsilon)$ rounds of rejection sampling suffice.

\subsubsection{Longer tail}
Consider the case when $\frac{d}{2}$ eigenvalues are 1 and the remaining $\frac{d}{2}$ are all equal to $\delta \ll 1$. In this case, for $k>\frac{d}{2}$ we have,

\begin{align*}
\frac{\E(L(S)) } {E^*_k} &= \frac{\alpha k}{(d-k)\delta}.
\end{align*}

With $\alpha = \delta$, we get $k = \sum \frac{\lambda_i}{\lambda_i + alpha}= \frac{d}{2}(\frac{1}{1+\alpha} + \frac{\delta}{\delta + \alpha}) = \frac{d}{2} (\frac{1}{1+\alpha} + \frac{1}{2}) \approx \frac{3d}{4}$ for $\delta = \alpha \rightarrow 0$.

So, for $k \approx 3d/4$ in the above example, we see $\E(L(S))$ is constant factor approximation of $E^*_k$.

\subsubsection{General tail} 
Say $m$ eigenvalues are $1$ and $(d-m)$ eigenvalues are $\delta$. We want to consider the case when $k>m$. And, we would like to have a $\beta$-constant factor approximation for $\E(L(S))$ i.e. $\frac{\E(L(S))}{E^*_k} = \beta \implies \frac{\alpha k}{(d-k)\delta} = \beta \implies \frac{\alpha}{\delta} = \frac{\beta(d-k)}{k}$.

\begin{align*}
k &= \frac{m}{1+ \alpha} + \frac{\delta(d-m)}{\alpha + \delta} \\ 
& = \frac{m}{1+\alpha} + \frac{(d-m)}{1 + \frac{\alpha}{\delta}} \\
& = \frac{m}{1+\alpha} + \frac{(d-m)}{1 + \beta (\frac{d}{k} - 1)}.
% & = m \left( \frac{1}{1+ \alpha} - \frac{1}{\delta + \alpha} \right) + \frac{d}{\alpha + \delta}, \\ 
\end{align*}

Assuming $\frac{m}{d} = c$ (constant), and $\alpha, \delta \rightarrow 0$,we get

\begin{align*}
\frac{k}{d} &= c + \frac{1-c}{1 + \beta (\frac{d}{k} - 1)}.
\end{align*}

The above expression is quadratic in $\frac{k}{d}$. Let $x = \frac{d}{k}$. We can write,

\begin{align*}
\frac{1}{x} &= c+ \frac{1-c}{1 + \beta x - \beta}\\
 & = \frac{c + \beta c x - \beta c + 1 -c }{1 + \beta x - \beta}\\      
 \implies & \beta c x^2 + x (1-\beta - \beta c) +(\beta -1) = 0.
\end{align*}

For specific values, say $c=1/2$ and $\beta=2$, we get the quadratic $x^2 - 2x +1 = 0 \implies x =1$.

\subsection{Lower bound}

 \begin{lemma}
	For any $\epsilon>0$, $\exists$ a matrix $\mathbf{X}$ with stable rank $r$ such that using DPP sampling with $E(|S|) = k$ with $k \leq \frac{\epsilon}{1+\epsilon} r$, we can lower bound $E(L(S)) \geq \left(\frac1{2} + \sqrt{\epsilon}\right)E^*_k $.
\end{lemma}
\begin{proof}

Using the same setup as above, with $m=k$, and say we want $r=(1+\rho) k$.  To calculate $r$,

 \begin{align*}
r = \sum_i \frac {\lambda_i}{ \lambda_1} = m + \delta (d-m) = k + \delta(d-k).
 \end{align*}

Since  $r=(1+\rho) k$, we get $\delta(d-k) = \rho k$.

Furthermore, since we want $\frac{\E(L(S))}{E^*_k} = \beta$,

\begin{align*}
&\frac{\alpha k }{\delta(d-k) } = \beta \\ & \implies \alpha k = \beta \rho k \implies \alpha = \beta \rho.
\end{align*}

For $k$,

\begin{align*}
& k = \sum_i \frac{\lambda_i}{\lambda_i + \alpha} 
 = \frac{k}{1+\alpha} + \frac{\delta(d-k)}{\delta+\alpha} \\
& \implies k =  \frac{k}{1+\alpha} + \frac{\rho k}{\delta+\alpha}\\
&\implies 1 =  \frac{1}{1+\alpha} + \frac{\rho}{\delta+\alpha} \\
&\implies \delta = \rho + \frac{\rho}{\alpha} - \alpha = \rho + \frac1 {\beta} - \beta \rho.
\end{align*}

For the bound, we need $0 < \delta < 1$. While $\delta < 1$ is trivially satisfied, for $\delta > 0$, we need $\rho < \frac{1}{\beta (\beta - 1)}$.

With $\rho = \frac1 {\epsilon}$, we have $k = \frac{\epsilon}{1+ \epsilon}r$. And with $\beta =1+\frac{\epsilon}{1+ \epsilon} $, the condition $\rho < \frac{1}{\beta (\beta - 1)}$ is satisfied. Hence, we get for any $\epsilon>0$, with $k = \frac{\epsilon}{1+ \epsilon}r $, we are ensured that $\frac{E(L(S))}{E^*_k} \geq \left( 1 + \frac{\epsilon}{1+\epsilon} \right)$.

We can also write the condition as $\beta (\beta -1) -\epsilon < 0$, which implies
 \[\beta \in \left(\frac{( 1- \sqrt(1+ 4 \epsilon)}{2}, \frac{( 1+ \sqrt(1+ 4 \epsilon)}{2} \right)\]. $\beta = (1/2 + \sqrt{\epsilon})$ satisfies this condition.

\end{proof}

\section{Lower bound for subsets larger than stable rank}
We first extend the lower bound result of
\cite{pca-volume-sampling}. Their construction uses a set of vectors that
form a regular simplex centered at $\zero$ and than slightly
shifts it. We show that in fact the regularity of the simplex is not
necessary for their proof to work.
\begin{lemma}\label{l:simplex}
Fix $\epsilon\in(0,1)$.  Consider any $k+1$ vectors $\x_i$ spanning a
$k$-dimensional subspace with their centroid at $\zero$ and let $\v$
be a unit vector orthogonal to that subspace. For any sufficiently small
$\alpha>0$, if we construct
matrix $\X$ with $k+1$ rows so that the $i$th row vector is $\x_i^\top+\alpha\v^\top$, 
then for any subset $S\subseteq[k+1]$ of size $k$,
\begin{align*}
  L(S)\geq (1-\epsilon)(k+1) E_k^*.
  \end{align*}
\end{lemma}
\begin{proof}
Let $\P^*$ be the projection onto the span of
$\{\x_1,\dots,\x_{k+1}\}$. Then we have
\begin{align*}
  E_k^* \leq \|\X - \X\P^*\|_F^2 = (k+1)\cdot\alpha^2.
\end{align*}
We will use the shorthand $\xbt_i=\x_i+\alpha\v$. Consider any subset
of $k$ rows from $\X$. W.l.o.g. suppose 
$S=\{1,\dots,k\}$. Since the centroid of all the rows is $\alpha\v$,
we can rewrite the last vector as 
\begin{align*}
\xbt_{k+1} = (k+1)\alpha\v - \sum_{i\in S}\xbt_i.
\end{align*}
Since for any $i\in S$ we have $\P_S\xbt_i=\xbt_i$, it follows that:
\begin{align*}
  L(S) &= \|\xbt_{k+1} - \P_S\xbt_{k+1}\|^2 \\
  &=\bigg\| (k+1)\alpha\v - \sum_{i\in S}\xbt_i - \Big(\P_S (k+1)\alpha\v -
    \sum_{i\in S}\xbt_i\Big)\bigg\|^2\\
       &=(k+1)^2\alpha^2\|\v-\P_S\v\|^2.
\end{align*}
Note that $\lim_{\alpha\rightarrow 0}\P_S\v = \zero$ because $\v$
is orthogonal to the subspace spanned by $\{\x_i:i\in S\}$, so there is
$\alpha_S>0$ such that for any $\alpha\leq\alpha_S$ we have $\|\v-\P_S\v\|^2\geq 1-\epsilon$. This shows that
$L(S)\geq (1-\epsilon)(k+1)E_k^*$ for $\alpha=\alpha_S$. Taking
$\alpha = \min_{S:|S|=k}\alpha_S$ completes the proof.
\end{proof}
We next obtain our main hardness result which states that the lower bound
factor $k+1$ of \cite{pca-volume-sampling} cannot be improved by constraining to matrices of a
particular stable rank $r$ as long as $k>r$. This stands in contrast
to our upper bound which showed that when $k<r$, then the factor $k+1$
can be significantly improved.
\begin{theorem}
For any $\epsilon\in(0,1)$ and $n,d,k,r\in\N$ such that  $0<r< k<\min\{n,d\}$, there is a matrix
$\X\in\R^{n\times d}$ with stable rank in $[r,r+1]$ such that for any subset
$S$ of size $k$
\begin{align*}
  L(S)\geq (1-\epsilon)(k+1) E_k^*.
  \end{align*}
\end{theorem}
\begin{proof}
  We will construct a set of vectors by gradually modifying and augmenting it while
  preserving two invariants: (1) that the stable rank is in $[r,r+\frac23]$
  and (2) the centroid of the vectors is  $\zero$.

First, construct matrix $\X_{(r+1)}$  from $r+1$ row vectors
$\x_1^\top,...,\x_{r+1}^\top$ that form 
  a regular simplex centered at $\zero$, scaled so that the $r$ non-zero
  singular values of $\X_{(r+1)}$ are all equal 1. Since all the non-zero
  singular values of the matrix are equal, the stable rank
is $r$. Next, we iteratively add vectors one by one (and modify the
existing ones) until there is $k+1$ in the set. Let $\x_{1},...,\x_{t-1}$ and
$\xbt_1,...,\xbt_t$ denote the 
sets of vectors \emph{before} and \emph{after} adding the $t$th vector,
with corresponding matrices $\X_{(t-1)}$ and $\X_{(t)}$. Also, let
$\tilde r_{t-1}$ denote the stable rank of $\X_{(t-1)}$.
Adding the $i$th vector proceeds as follows:
  \begin{enumerate}
    \item shift all $t-1$ vectors along a normal direction $\v_t$
      (orthogonal to their span) by some $\delta_t$,
    \item let the new vector be $\xbt_{t}=-t\,\delta_t\v_t$.
    \end{enumerate}
    After the transformation and adding the $i$th vector, the centroid
    stays at $\zero$, and moreover:
    \begin{align*}
      \X_{(t)}^\top\X_{(t)} &=
      \sum_{i=1}^{t-1}(\x_i+\delta_t\v_t)(\x_i+\delta_t\v_t)^\top +
      t^2\delta_t^2\v_t\v_t^\top\\
      &=\sum_{i=1}^{t-1}\x_i\x_i^\top +
        \delta_t\v_t\Big(\sum_{i=1}^{t-1}\x_i^\top\Big) +
        \Big(\sum_{i=1}^{t-1}\x_i\Big)\delta_t\v_t^\top +
 (t-1)\delta_t^2\v_t\v_t^\top+        t^2\delta_t^2\v_t\v_t^\top\\
      &=\X_{(t-1)}^\top\X_{(t-1)}+(t^2+t-1)\delta_t^2\v_t\v_t^\top.
    \end{align*}
Therefore, since $\v_i$ is orthogonal to the span of $\{\x_{1},...,\x_{t-1}\}$,
it becomes a new eigenvector of $\X_{(t)}^\top\X_{(t)}$, while the old
eigenvectors from $\X_{(t-1)}^\top\X_{(t-1)}$ remain the same in the
new matrix. Thus, as long as $\delta_t$ is sufficiently small (so that
the new eigenvalue does not become the largest), we have:
    \begin{align*}
\tilde r_{t} = \tilde r_{t-1} + \|\X_{(t)}\|_F^2
      -\|\X_{(t-1)}\|_F^2=\tilde r_{t-1} + \delta_t^2(t^2+t-1).
    \end{align*}
    To make sure that stable rank remains less than $r+\frac23$, we
    assume that $\delta_t\leq\frac1{t2^{t}}$ and bound the stable rank of the final set:
    \begin{align*}
 \tilde r_{k+1} = r + \sum_{t=r+2}^{k+1} \delta_t^2 (t^2+t-1) \leq r +
      \sum_{t=1}^{k+1}\frac2{4^t}\leq r + \frac23.
    \end{align*}
  Now, it remains to apply Lemma \ref{l:simplex} to the set
  of $k+1$ vectors, obtaining matrix $\Xt_{(k+1)}$ that satisfies the
desired lower bound. When choosing
  $\alpha$, we have to again ensure that it 
  is sufficiently small so that stable rank does not change by
  more than $\frac13$. Naturally, this can be done just as we did it
  in the construction of the set. We construct the final matrix $\X$
from $\Xt_{(k+1)}$ by padding with zeros to get the shape right,
which does not affect the loss $L(\cdot)$ or the stable rank. 
\end{proof}

\section{Lower bound for subsets smaller than stable rank}
Let $\x_1,...,\x_{k+1}$ be vectors forming a regular simplex centered
at $\zero$ which lies inside of the subspace spanned by the first $k$
coordinates. Denote $a=\|\x_i\|$ as the length of each vector. We
shift those vectors in the $(k+1)$-st coordinate by $\alpha$, getting
$\xbt_i=\x_i+\alpha\e_{k+1}$. Then, we add further vectors
$\x_{k+2},...,\x_n$ for some $n\gg k$ so that $\x_i=\delta\e_{i}$ for
$i>k+1$. Denote the entire $n\times d$ matrix as $\X$. The optimum
rank $k$ error is upper bounded as follows: 
\begin{align*}
  E_k^* \leq (k+1)\alpha^2 + (n-k-1)\delta^2.
\end{align*}
The top $k$ eigenvalues of $\X^\top\X$ associated with the simplex are all
$\lambda_i=\frac{k+1}{k}\,a^2$. This means that,
assuming that $\delta^2\leq(k+1)\alpha^2\leq \frac{k+1}{k}\,a^2$ (to
achieve the correct ordering of the eigenvalues), the
stable rank of $\X$ is given by:
\begin{align*}
  r = \frac1{\lambda_1}\sum_{i=1}^d\lambda_i = k +
  k\,\frac{\alpha^2}{a^2} + \frac{(n-k-1)\delta^2}{\frac{k+1}k\,a^2}.
\end{align*}
Finally, for any subset $S\subseteq[k+1]$, the loss is:
\begin{align*}
  L(S) &= (k+1)^2\frac{(\frac ak)^2\alpha^2}{(\frac ak)^2+\alpha^2} +
  (n-k-1)\delta^2\\
  &=(k+1)^2\frac{a^2\alpha^2}{a^2+k^2\alpha^2} +
  (n-k-1)\delta^2.
\end{align*}
Suppose that $\alpha^2=a^2/k^2$ and $(n-k-1)\delta^2 = (k+1)(\rho a^2-\alpha^2)$ for some
$\rho\in[\frac1{k^2},1]$. It follows that $r=(1+\rho)k$ and
$E_k^*=\rho(k+1)a^2$. Also we get:
\begin{align*}
  L(S)&=  (k+1)^2 \frac{a^2}{2k^2} + (k+1)\Big(\rho-\frac1{k^2}\Big) a^2\\
      &=\rho(k+1)a^2 + a^2\frac{k+1}{k^2}\Big(\frac{k+1}{2} - 1\Big)\\
  &=E_k^*\cdot\Big(1 + \frac{k-1}{2\rho k^2}\Big).
\end{align*}
So if we let $\epsilon = \frac{k-1}{2\rho k^2}$ then we can turn
this into the following lower bound. 
\begin{theorem}
  For any $\epsilon\in(0,k^2)$ and $n,d,k,r\in\N$ such
  that $k\leq\frac{\epsilon}{1+\epsilon}\,r\leq\min\{n,d\}$, there is
  an $n\times d$ matrix $\X$ with stable rank  
  $r$ such that for any subset $S$ of size $k$,
  \begin{align*}
    L(S) \geq \bigg(1+\epsilon\cdot \frac{k-1}{2k^2}\bigg)\cdot E_k^*.
  \end{align*}
\end{theorem}
Note that this lower bound is worse than our upper bound by a factor
of about $\frac1{2k}$. The key question is which bound is loose: the
upper or the lower bound? Also, not sure how to cleanly
write the lower bound
in a way that concludes with $L(S)\geq (1+\epsilon) E_k^*$.

\subsection{Const lower bound}
Choose $\rho \in [\frac1 {k^2}, \frac{(k-1)}{2k^2\epsilon} ]$. This still means $r=(1+\rho)k$ and
$E_k^*=\rho(k+1)a^2$. Following the steps for $L(S)$, we get,
\begin{align*}
L(S) \geq E_k^*\cdot\Big(1 + \frac{k-1}{2\rho k^2}\Big) \geq (1 + \epsilon) E_k^*.
\end{align*}

Note that we need $\delta^2 \leq (k+1) \alpha^2$ to hold, but  $(n-k-1)\delta^2 = (k+1)(\rho a^2-\alpha^2)$ can still hold simultaneously by choosing large enough $n$ for  $\rho \in [\frac1 {k^2},\frac{(k-1)}{2k^2\epsilon} ]$.

Now we need to find the condition for $k$ wrt $r$. By $r = (1+\rho)k$, we have $r \geq  (1+\frac1{k^2})k \iff k + \frac1 k \leq r \impliedby k < r$.

Finally the range of $\epsilon$ can only be such that the lower bound of $1/k^2$ is not violated.

\begin{theorem}(Informal) For any $\epsilon \in (0,(k-1)/2 ]$, $\exists$ a matrix $\mathbf{X} $ of stable rank $r$, such that for selecting $k$ rows with $k < r$, ensures 
	\begin{align*}
	L(S) \geq (1+\epsilon) E^*_k
	\end{align*}

\end{theorem}
\section{Proof of Theorem \ref{t:main}}
Our result is based on the following matrix expectation formula.
\begin{lemma}
  For any $\X\in\R^{n\times d}$ and non-negative weights $p_1,...,p_n$, let $\Xt$ denote
  $\X$ with $i$th row rescaled by $\sqrt{p_i}$. If $S\sim\DPP(\Xt\Xt^\top)$, then
  \begin{align*}
    \E\big[\I-\X_S^\top(\X_S\X_S^\top)^{-1}\X_S\big] & = (\I +
    \Xt^\top\Xt)^{-1} = \Big(\I + \sum p_i\x_i\x_i^\top\Big)^{-1}.
  \end{align*}
\end{lemma}
\begin{proofof}{Theorem}{\ref{t:main}}
  Let $S\sim\DPP(\frac1\alpha\X\X^\top)$. The expected subset size is given by:
\begin{align*}
  \E\big[|S|\big] =
  \tr\big(\tfrac1\alpha\X\X^\top(\I+\tfrac1\alpha\X\X^\top)^{-1}\big).
\end{align*}
On the other hand the expected loss is given by:
\begin{align*}
\E\big[L(S)\big]=  
  \tr(\X^\top\X \,\E[\I-\P_S]) =
  \tr\big(\X^\top\X(\I+\tfrac1\alpha\X^\top\X)^{-1}\big) =
  \E[|S|]\cdot \alpha.
\end{align*}
Now, setting $\alpha = E_k^*$, it follows that:
\begin{align*}
  \E\big[|S|\big] = \sum_{i=1}^d\frac{\lambda_i}{\alpha+\lambda_i}\leq
  \sum_{i=1}^k\frac{\lambda_i}{E_k^*+\lambda_i} +
  \frac1{E_k^*}\sum_{i=k+1}^d\lambda_i\leq k+1.
\end{align*}
\end{proofof}

Note that the scaling weights $p_i$ we use to prove Theorem
\ref{t:main} are all set to $\frac1\alpha$. Can we improve the result
by finding optimal values of the weights?

\section{Algorithm using a weighted DPP}

\begin{definition}
Define the $\alpha$-regularized Nystr\"om approximation of $\L$ based on a subset $S$
 as the $n \times n$ matrix $\widehat\L_\alpha(S)=\L_{\cdot,S}(\alpha\I+\L_{S,S})^\dagger\L_{S,\cdot}$.
\end{definition}

Given some value $\alpha>0$, consider the following optimization problem:
\begin{align*}
  \min_p\quad
  &
  \tr\bigg[\X^\top\X\Big(\I+\frac1\alpha\sum_ip_i\x_i\x_i^\top\Big)^{-1}\bigg]\\
  \text{subject to}\quad
  &\sum_i p_i = k,
  % \sum_j p_j\,\x_j^\top\Big(\alpha\I+\sum_ip_i\x_i\x_i^\top\Big)^{-1}\!\x_j\leq k,
\quad 0\leq p_i\leq 1 \ \ \forall_i.
\end{align*}
Let $p^*(\alpha)$ denote the solution.%  and let $S\sim
% \DPP(\frac1\alpha\Xt\Xt^\top)$. Note that 
% \begin{align*}
% \E\big[|S|\big] =
%   \tr\big(\Xt^\top\Xt(\alpha\I+\Xt^\top\Xt)^{-1}\big)\leq \tr\big(\X^\top\X(\alpha\I+\X^\top\X)^{-1}\big)=d_\alpha.
%   \end{align*}
%Moreover, define the following
% secondary optimization:
% \begin{align*}
%   \alpha^* = \min_{\alpha>0} \ \ \alpha \quad\text{subject to}\quad
%   \tr\bigg[\Big(\sum_ip_i^*(\alpha)\x_i\x_i^\top\Big)\Big(\alpha\I + \sum_i
%   p_i^*(\alpha)\x_i\x_i^\top\Big)^{-1}\bigg]\leq k.
% \end{align*}

\begin{theorem}\label{t:weighted}
If $S\sim\DPP(\frac1{\alpha}\Xt\Xt^\top)$, where $i$th
row of $\Xt$ is rescaled by $\sqrt{p_i^*(\alpha)}$, we obtain:
\begin{align*}
  \E\,\big[\|\L-\widehat\L(S)\|_*\big] \leq \min_{S:|S|=k} \|\L -
  \widehat\L_{\alpha}(S)\|_*.
\end{align*}
\end{theorem}
\textbf{Problem:} It is not clear how to choose $\alpha$ to ensure
that $\E[|S|]\leq k$. However, a simple doubling strategy might work in
practice.

Another strategy would be to put the expected subset size as a
constraint in the optimization:
\begin{align*}
  \min_p\quad
  &
  \tr\bigg[\X^\top\X\Big(\I+\frac1\alpha\sum_ip_i\x_i\x_i^\top\Big)^{-1}\bigg]\\
  \text{subject to}\quad
  & \tr\bigg[\Big(\sum_ip_i\x_i\x_i^\top\Big)\Big(\alpha\I + \sum_i
 p_i\x_i\x_i^\top\Big)^{-1}\bigg]\leq k,
\quad 0\leq p_i\leq 1 \ \ \forall_i.
\end{align*}
Here, the problem is that we can no longer write the bound from
Theorem \ref{t:weighted}, and also the constraint becomes non-linear.

% \section{A limit argument for Theorem \ref{t:main}}\label{s:proof}
% \begin{definition}\label{d:r-dpp}
% Given matrix $\X\in\R^{n\times d}$, a sequence of Bernoulli probabilities $p=(p_1,\dots,p_n)\in[0,1]^n$
% and a p.s.d.~matrix $\A\in\R^{d\times d}$, let
% $\Vol_p(\X,\A)$ be a distribution over all sets $S\subseteq [n]$ s.t.:
%   \begin{align}
%   \Pr(S) = \frac{\det(
%   \X_S^\top\X_S+\A)}{\det\!\big(
%  \sum_ip_i\x_i\x_i^\top+ \A\big)}\
% \cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in S}(1-p_i).\label{eq:poisson-prob}
% \end{align}
% %   \Pr(\pi)\ \propto\ \det\!\big(\A +
% %   \X_{\pi}^\top\X_{\pi}\big)
% %   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
% % \end{align*}
% \end{definition}
% Whenever all $p_i=\frac sn$, we will write $\Vol^s(\X,\A)$.
% The below theorem was shown in prior work.
% \begin{theorem}[\cite{bayesian-experimental-design}]\label{t:expectations}
% Given $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$ and $p$,
% if $S\sim \Vol_p(\X,\A)$, then
% \begin{align}
%   \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
%   &\preceq \Big(
%   \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1}.\label{eq:sqinv}
% \end{align}
% \end{theorem}

% We will use the following limit argument.
% \begin{lemma}
% For any $p\in[0,1]^n$, let $\Xt$ denote $\X$ with $i$th
%   row rescaled by $\sqrt{p_i}$ for all $i$. We have
%   \[\lim_{\epsilon\rightarrow 0}\Vol_{\epsilon
%     p}(\tfrac1{\sqrt\epsilon}\,\X,\I) =\DPP(\Xt\Xt^\top).\]
% \end{lemma}
% % \begin{proof}
% %   Denoting $p=\frac sn$ we have
% %   \begin{align*}
% %     \Pr(S)
% % &= \frac{\det(\frac1p\X_S^\top\X_S+\I)}{\det(\X^\top\X+\I)}\cdot
% %     p^{|S|}\cdot (1-p)^{n-|S|}\\
% % &=    \frac{\det(\X_S\X_S^\top+p\I)}{\det(\X\X^\top+\I)}\cdot
% %                                    (1-p)^{n-|S|}\\
% % &\rightarrow \frac{\det(\X_S\X_S^\top)}{\det(\I+\X\X^\top)}.
% %   \end{align*}
% % \end{proof}

% % \begin{theorem}
% %   For any $\X\in\R^{n\times d}$ and $p\in[0,1]^n$, if $S\sim\DPP(\Xt\Xt^\top)$, then
% %   \begin{align*}
% %     \E\big[\X_S^\top(\X_S\X_S^\top)^{-1}\X_S\big] &=
% %     \Xt^\top(\I+\Xt\Xt^\top)^{-1}\Xt,\\
% %     \E\big[\I-\X_S^\top(\X_S\X_S^\top)^{-1}\X_S\big] &= (\I +
% %     \Xt^\top\Xt)^{-1} = \Big(\I + \sum_ip_i\x_i\x_i^\top\Big)^{-1},
% %   \end{align*}
% %   where $\Xt$ here denotes $\X$ with $i$th row rescaled by $\sqrt{p_i}$.
% % \end{theorem}
% We use these two facts to show our new expectation inequality.
% \begin{proofof}{Theorem}{\ref{t:main}}
%   If $S \sim \Vol_{\epsilon p}(\frac1{\sqrt{\epsilon}}\,\X,\I)$ then
%   \begin{align*}
%     \E\big[(\I + \tfrac1\epsilon\X_S^\top\X_S)^{-1}\big] \preceq
%     (\I+\Xt^\top\Xt)^{-1}.
%   \end{align*}
%   On the other hand, we have
%   \begin{align*}
%     (\I + \tfrac 1\epsilon\X_S^\top\X_S)^{-1} = \I - \tfrac
%     1\epsilon\X_S^\top(\I+\tfrac 1\epsilon\X_S\X_S^\top)^{-1}\X_S = \I -
%     \X_S^\top(\epsilon\,\I+\X_S\X_S^\top)^{-1}\X_S.
%   \end{align*}
%   Taking the limit $\epsilon\rightarrow 0$ we get the result.
% \end{proofof}

\section{Backup}

\subsubsection{Concentration}

\begin{lemma} With $r$ as the stable rank and $\epsilon \in (0,1)$, for $ \frac 8{\epsilon^2} \ln \frac1{\epsilon} \leq k \leq  \frac {\epsilon (1-\epsilon)}{2(1+\epsilon)}  r $, there exists an algorithm to select a set of rows $S$ with $|S| \leq k$, so that $L(S)  \leq \frac{(1+\epsilon)}{(1- \epsilon) } E^*_k$. 
\end{lemma}
It suffices to show that $\E\big[L(S)\mid |S|\leq k\big]  \leq \frac{(1+\epsilon)}{(1- \epsilon) } E^*_k$ if $S$ is selected using DPP sampling.

For concentration, we would need $k_\alpha \leq k$, which requires using similar arguments as above:

\begin{align*}
	& \frac r {r-k}\,\frac1{1+\epsilon} \leq 1 - \delta \iff \frac r {r-k} \leq (1-\delta) (1+\epsilon)   \iff 1 - \frac k{r} \geq \frac1{(1-\delta) (1+\epsilon)} \\
	& \iff \frac k{r} \leq \frac{\epsilon(1-\delta)- \delta}{(1+\epsilon)(1-\delta)}.
\end{align*}

For $\delta \leq \epsilon/2$, we have $\frac \delta{1-\delta} \leq \frac {\epsilon/2}{1- \epsilon/2}$. To satisfy the above condition on $\frac k{r}$,

\begin{align*}
	&	\frac k{r} \leq \frac{\epsilon(1-\delta)- \delta}{(1+\epsilon)(1-\delta)} \iff 	\frac  k{r} \leq \frac{\epsilon}{1+\epsilon} - \frac{\delta}{(1+\epsilon)(1-\delta)} \\
	&\impliedby 	\frac  k{r} \leq \frac{\epsilon}{1+\epsilon} - \frac{\epsilon/2}{(1+\epsilon)(1-\epsilon/2)} \iff 	\frac k{r} \leq \frac{\epsilon}{1+\epsilon} \left( 1- \frac{1/2}{1-\epsilon/2}\right)  \\
	&\iff \frac k{r} \leq \frac{\epsilon}{1+\epsilon} \left(  \frac{1/2 - \epsilon/2}{1-\epsilon/2}\right) \impliedby \frac kr \leq \frac {\epsilon (1-\epsilon)}{2(1+\epsilon)} 
\end{align*}

For concentration,

\begin{align*}
	& \E\big[L(S)\mid |S|\leq k\big] \leq \frac{\E[L(S)]}{\Pr(|S|\leq k)} \leq \frac {(1+\epsilon)E^*_k}{(1-\ee^{-\delta^2k/2}) } \leq \frac{(1+\epsilon)}{(1- \epsilon) } E^*_k,  
\end{align*}

for $\ee^{-\delta^2k/2} \leq \epsilon $, which gives $k \geq \frac 2{\delta^2} \ln \frac1{\epsilon} $ which is simplified to  $k \geq \frac 8{\epsilon^2} \ln \frac1{\epsilon} $ for $\delta \leq \epsilon/2$.
\bibliography{pap}

\end{document}

### 2022-08-23 design/open/open-minimax.tex
Commit original repos

\documentclass[12pt]{../sty/colt2019/colt2018-arxiv}
% \else
% \documentclass[anon,12pt]{sty/colt2019/colt2019} % Anonymized submission
% \documentclass[12pt]{colt2019} % Include author names
% \fi

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

% \title[Short Title]{Full Title of Article}
\title{Minimax experimental design: Open problems}
\usepackage{times}

\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{xfrac}
\usepackage{algorithmic}
\input{../shortdefs}
% Use \Name{Author Name} to specify the name.
% If the surname contains spaces, enclose the surname
% in braces, e.g. \Name{John {Smith Jones}} similarly
% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
% If the first letter in the forenames is a diacritic
% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

% Two authors with the same address
% \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
%  \Name{Author Name2} \Email{xyz@sample.com}\\
%  \addr Address}

% Three or more authors with the same address:
% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
%  \Name{Author Name2} \Email{an2@sample.com}\\
%  \Name{Author Name3} \Email{an3@sample.com}\\
%  \addr Address}

%  Authors with different addresses:
\coltauthor{\Name{Micha{\l } Derezi\'{n}ski}
  \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics, UC Berkeley 
}

\begin{document}

\maketitle

\section{A list of open problems and conjectures}

\subsection{Optimal sample complexity for the MSE}
\begin{conjecture}
  Given $\X\in\R^{n\times d}$ such that 
$\tr((\X^\top\X)^{-1})\!=\!\phi$ and $\epsilon>0$, there is a
random experimental design $(S,\wbh)$
of size $k=O(d+\phi/\epsilon)$ s.t.~for \textbf{any} random
response vector $\y$, 
\begin{align*}
\E_{S,\wbh,\y}\big[\wbh(\y_S)\big]=\w^*\quad\text{and}\quad
  \mathrm{MSE}\big[\wbh(\y_S)\big] - \mathrm{MSE}\big[\wols(\y|\X)\big]
  \leq \epsilon\cdot 
  \E_\y\big[\|\xib_{\y|\X}\|^2\big].%\quad\text{and}\quad k= O(d\log n + \phi/\epsilon).
\end{align*}
\end{conjecture}
\textbf{Best known:} \ $k=O(d\log n + \phi/\epsilon)$ via rescaled volume sampling.

We believe that rescaled volume sampling achieves sample complexity $\Theta(d\log d+\phi/\epsilon)$ (but not the conjectured $d+\phi/\epsilon$). Showing this would make the bound independent of the data size $n$, which has important implications for experimental design over continuous distributions (instead of finite $n\times d$ matrices).

\subsection{The no-bias conjecture}
\begin{conjecture}
  Given any $\X$, for any (possibly biased) random experimental design $(S,\wbh)$ of size $k$, there is an unbiased design $(\St,\wbt)\in\Wc_k(\X)$ such that
  \begin{align*}
    \max_{\y\in\Fc\backslash\mathrm{Sp}(\X)}\,\frac{\mathrm{MSE}\big[\wbt(\y_{\St})\big]-
    \mathrm{MSE}\big[\wols(\y|\X)\big]
    }{\E_\y\big[\|\xib_{\y|\X}\|^2\big]}\leq     \max_{\y\in\Fc\backslash\mathrm{Sp}(\X)}\,\frac{\mathrm{MSE}\big[\wbh(\y_S)\big]-
    \mathrm{MSE}\big[\wols(\y|\X)\big]
    }{\E_\y\big[\|\xib_{\y|\X}\|^2\big]}.
  \end{align*}
\end{conjecture}
The conjecture suggests that due to the worst-case nature of the criterion, no biased estimator should perform better than every unbiased design. If true, this would imply that dropping the unbiasedness constraint from the definition of the minimax value $R_k^*(\X)$ would not alter it.

\subsection{Bayesian minimax experimental design}
In Bayesian experimental design we assume that $\y=\X\w+\z$, where $\w\sim D_\w$  is a random vector with a known prior $D_\w$ and $\z$ has mean zero. Since wish to minimize $\E[\|\wbh(\y_S)-\w\|^2]$, knowledge of the prior should favor a biased estimator.

\textbf{Open:} Define a criterion that takes maximum over noise vector $\z$ given a prior $D_\w$. Show that this value is smaller than $R_k^*$ and that there is a biased estimator better than any unbiased estimator.

\subsection{Finding optimal volume-rescaled i.i.d.~sampling design}
We define the following minimization problem over distributions $q\in\Delta_n$ with support on $\{1..n\}$:
\begin{align*}
  \hat{R}_k^*(\X) = \argmin_{q\in\Delta_n}\max_{\y\in\R^n\backslash\mathrm{Sp}(\x)}
  \E_{\pi\sim\Vol_q^k(\X)}\bigg[\frac{\|(\S_\pi\X)^\dagger\S_\pi\y - \X^\dagger\y\|^2}{\|\X\X^\dagger\y-\y\|^2}\bigg].
\end{align*}
We know that $R_k^*(\X)\leq \hat{R}_k^*(\X)\leq C\cdot \phi/k$ for $k\geq C\cdot d\log n$.\\
\textbf{Open:} Find an efficient optimization algorithm that converges to the optimum distribution $q$.
\subsection{Minimax experimental design over distributions}
Consider a distribution $(\x,y)\sim D$ jointly drawing a vector $\x\in\R^d$ and a real response value $y$. We define the optimum solution  and noise variable as:
\begin{align*}
  \w^*=\argmin_\w\E\big[(\x^\top\w-y)^2\big] = \big(\E[\x\x^\top]\big)^{-1}\E[\x\,y],\qquad \xi = \x^\top\w^*-y.
\end{align*}
Random experimental design in this setting samples a set of points from the support of the input distribution $D_{\cal X}$ and then queries the labels from $D_{\cal Y|\x}$. Minimax experimental design extends naturally to this setting for both MSE and MSPE.\\
\textbf{Known:} The minimax value for MSPE is bounded by $d^2/k$ (unpublished)\\
\textbf{Open:} Bound the MSE minimax value, get closer to the conjectured $\phi/k$ bound

\subsection{Minimax experimental design for any loss function}
Consider a loss function $L_\y(\w)=\sum_{i=1}^n\ell(\x_i^\top\w,y_i)$. For $\X$ and random vector $\y\in\Fc$
define:
\begin{align*}
\w^* \defeq \argmin_\w\E\big[L_\y(\w)\big],\qquad
  \wols(\y|\X)\defeq \argmin_\w L_\y(\w),
\end{align*}
\begin{align*}
R_k^*(\X)&\defeq
  \min_{(S,\wbh)}\ \max_{\y\in\Fc\backslash\mathrm{Sp}(\X)}\,\frac{\mathrm{MSE}\big[\wbh(\y_S)\big]-
     \mathrm{MSE}\big[\wols(\y|\X)\big]
  }{\E\big[L_\y(\w^*)\big]}.
\end{align*}
\subsection{Kernel ridge experimental design}
\begin{conjecture}
Suppose that $\w\sim \Nc(\zero,\Sigmab)$. Let $\y=\X\w+\z$,
where $\E[\z]=\zero$. For $\C_\lambda=(\X^\top\X+\lambda\Sigmab^{-1})^{-1}$,
we define
$\phi_\lambda=\tr\big(\C_\lambda\big)$ and 
$\w_\lambda(\y)=\C_\lambda\X^\top\y$. 
\begin{align*}
  \E\Big[\E\big[\|\wbh(\y_S) - \w\|^2
  -\|\w_\lambda(\y)-\w\|^2\mid\w\big]\Big]
  &\leq
O\big(\phi_\lambda/k\big)\cdot\E\Big[\E\big[\|\X\w-\y\|^2\mid \w\big] +
    \lambda\|\Sigmab^{-\sfrac12}\w\|^2\Big],
  \\
  \MSE{\wbh(\y_S)} - \MSE{\w_\lambda(\y)}
&\leq O\big(\phi_\lambda/k\big) \cdot\Big(\tr\big(\Var[\z]\big)+ \lambda\Big)
\end{align*}

\begin{align*}
\min_{(S,\wbh)}\ \max_{\underset{\tr\Var[\z]\geq \sigma^2}{\z:\,\E[\z]=\zero}}\,\frac{  \MSE{\wbh(\y_S)} - \MSE{\w_\lambda(\y)}
  }{\tr\,\Var[\z]}\leq O\big(1/k\big)\cdot \min_\lambda\big\{ \phi_\lambda(1+\lambda/\sigma^2)\big\}.
\end{align*}
\end{conjecture}

\subsection{Extending the minimax framework to D,E,...-optimality}
Most classical optimality criteria for experimental design have an interpretation in terms of the response vector $\y$ being linear plus i.i.d.~noise. Thus, they should have corresponding minimax variants.\\
\textbf{Known:} A-optimality, $F_A(S)=\tr((\X_S^\top\X_S)^{-1})$, is the MSE,\\
and $V$-optimality, $F_V(S)=\tr(\X(\X_S^\top\X_S)^{-1}\X^\top)$,  is the MSPE\\ 
\textbf{Open:} Any other one, such as D-optimality, $F_D(S)=\det(\X_S^\top\X_S)^{-1}$.

\subsection{Active minimax experimental design}
Is it possible to achieve a better minimax value if the experimental design procedure is allowed to sample and query online?
\subsection{$\y$ independent ratio and optimize $\q$}
Write the ratio in the minimax w.o. $\y$ i.t.o. some norms
of matrices. Then try to optimize $\q$.
\Red{This would be a key additional piece for a journal paper.}
\end{document}

### 2022-08-23 design/open/open-projections.tex
Commit original repos

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
  colorlinks,
  linkcolor={red!40!gray},
  citecolor={blue!40!gray},
  urlcolor={blue!70!gray}
}
\usepackage{cleveref}
\input{../shortdefs}

\title{Exact expressions for random projections: Open problems}
\author{Micha{\l} Derezi\'nski, UC Bekeley}
% \setlength{\parindent}{0cm}
% \setlength{\parskip}{0mm}

\begin{document}
\maketitle

\section{Main results}
Consider a large $m\times n$
matrix $\A$. A data-oblivious sketch of size $k$ is the matrix $\S\A$,
where $\S$ is a $k\times m$ random matrix such that
$\E[\frac1k\S^\top\S]=\I$, and whose distribution does not
depend on $\A$.
\begin{theorem}\label{t:main}
Let $\S$ be a sketch of size $k$ with i.i.d.~mean-zero sub-Gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.  If
  we let $\rho = r/k$ be a fixed constant larger than $1$, then
  \begin{align*}
    (1-\epsilon)\,\bar\P_{\!\perp}\preceq\E[\P_{\!\perp}]\preceq
    (1+\epsilon)\,\bar\P_{\!\perp}\quad\text{for}\quad \epsilon =
    O(\tfrac1{\sqrt r}),
  \end{align*}
  where $\P_{\!\perp} :=\I - (\S\A)^\dagger\S\A,$ and $\bar\P_{\!\perp}:=(\gamma\A^\top\A +
  \I)^{-1}$, for $\gamma>0$  s.t. $\tr\,\bar\P_{\!\perp} = n-k$.
\end{theorem}
% \begin{theorem}\label{t:main-tech}
% Let $\P_\perp = \I - \X^\dagger \X$ for $\X = \Z\Sigmab^{\frac12}$,
% where $\Z \in \mathbb R^{k\times n}$ has i.i.d.~$K$-sub-Gaussian rows with zero mean and identity covariance, and
% $\Sigmab$ is an $n\times n$ positive semi-definite matrix. Define: 
% \begin{align*}
% \bar\P_\perp= (\gamma\Sigmab + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P_\perp=n-k.
% \end{align*}
% Let $r =\tr(\Sigmab)/\|\Sigmab\|$
% be the stable rank of
% $\Sigmab^{\frac12}$ and fix $\rho=r/k > 1$. There exists a constant
% $C_{\rho}>0$, depending only on $\rho$ and $K$, such that if $r\geq
% C_\rho$, then
% \begin{align}
% \Big(1-\frac {C_\rho}{\sqrt
%   r}\Big)\cdot\bar\P_\perp\preceq\E[\P_\perp]\preceq
%   \Big(1+\frac {C_\rho}{\sqrt r}\Big)\cdot\bar\P_\perp.
% \end{align}
% \end{theorem}
\begin{corollary} %[low-rank approximation]
  \label{c:low-rank}
Let $\sigma_i$ be the singular values of $\A$. Under the assumptions of Theorem \ref{t:main}, we have:\vspace{-1mm}
  \begin{align*}
    \E\big[\|\A-\A\P\|_F^2\big] \ \overset\epsilon\simeq\ 
    k/\gamma
\quad \text{for \ $\gamma>0$ \ s.t. \ } \sum_{i}\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1} = k.
  \end{align*}
\end{corollary}
\begin{corollary}\label{c:kaczmarz}
Let $\x^*$ be the unique solution of $\A\x^*=\b$ and consider
  the iterative algorithm:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S_t\A\x=\S_t\b.
  \end{align*}
Under the assumptions of Theorem \ref{t:main}, with $r=\|\A\|_F^2/\|\A||^2$, we have:
  \begin{align*}
    \E\big[\x^{t+1}-\x^*\big] \overset\epsilon\simeq 
    (\gamma\A^\top\A+\I)^{-1}\,\E\big[\x^t-\x^*\big]
    \quad\text{for}\quad\epsilon=O(\tfrac1{\sqrt r}). 
  \end{align*}
Moreover, we obtain the following worst-case convergence guarantee:
\begin{align}
  \E\big[\|\x^{t+1}-\x^*\|^2\big]
  \leq \big(1-
  (\bar\kappa-\epsilon)\big)\,\E\big[\|\x^{t}-\x^*\|^2\big],
  \quad\text{where}\quad
  \bar\kappa = \frac{\sigma_{\min}^2(\A)}{\sigma_{\min}^2(\A)+1/\gamma}.\label{eq:worst-case}
\end{align}
\end{corollary}

\section{Open questions}
\subsection{Extend the result beyond the stable rank}
\begin{figure}[h]
  \centering
\includegraphics[width=.4\textwidth]{../equivalents_nips/abalone-nystrom}\nobreak\includegraphics[width=.4\textwidth]{../equivalents_nips/eunite-nystrom}
\caption{
Theoretical predictions of low-rank approximation error of a Gaussian
sketch along with spectral decay profiles and stable rank of the
matrices. Note that we use a kernel matrix notation $\K$, which can be
converted to the data matrix $\A=\K^{\frac12}$. 
}
%\vspace{-5mm}
\label{f:explicit}
\end{figure}

In Figure \ref{f:explicit}, we observe that, for standard spectral
decay profiles, the surrogate expression
of low-rank approximation error (see Corollary \ref{c:low-rank}) is
accurate even when $k$ is larger than the stable rank of the data
matrix. In particular, it seems that as long as the spectral decay
does not have a sudden drastic drop around $k$, the surrogate
expressions are accurate. This, to a certain degree, can also be
observed for the expected residual projection, i.e. $\E[\P_{\perp}]$.
Thus, we raise a number of questions:
\begin{enumerate}
  \item Under what conditions on the spectral decay can we let $k>r$
    and still have the surrogate expressions be asymptotically
    consistent (i.e., error decays to zero)?
  \item Under what conditions on the spectral decay can we ensure a
    $\frac12$-approximation for surrogate expressions, i.e., that
    $\frac12\bar\P_{\perp}\preceq\E[\P_{\perp}]\preceq\frac32\bar\P_{\perp}$?
  \item Under what conditions on the spectral decay can we ensure a
    $\frac12$-approximation for the low-rank approximation error,
    i.e., as in Corollary \ref{c:low-rank}, with $\epsilon=\frac12$ (a
    strictly weaker guarantee)?
\end{enumerate}
%\subsection{Extend the result to other sketches, e.g., SRHT}

\subsection{Expectation of non-residual projection, $\E[\P]$, where
  $\P=(\S\A)^\dagger\S\A$}\label{s:non-residual}
Our surrogate expression for $\E[\P]$ has the form
$\bar\P:=\gamma\Sigmab(\gamma\Sigmab+\I)^{-1}$. Using the guarantees
for the residual projection, we can immediately obtain the following:
\begin{align*}
(1-\epsilon)\bar\P - \epsilon \I\preceq \E[\P]\preceq (1+\epsilon)\bar\P + \epsilon\I.
\end{align*}
This is an additive approximation gurantee, which is very
weak for small eigenvalues of $\E[\P]$. Note that unlike the surrogate
expression for the \emph{residual} projection, $\bar\P$ can have very small
eigenvalues, which presents additional challenges in the
analysis. Because of this, we expect some dependence of the guarantees
on the ambient dimension $n$, and possibly on the condition number.

\begin{conjecture}\label{h:non-residual}
Let $\A$ be an $m\times n$ matrix with stable rank
  $r=\|\A\|_F^2/\|\A\|^2$, such that $n/r\leq\tau$ and $r/k\geq\rho$ for
 fixed constants  $\tau,\rho>1$. Given any sketch $\S$ of size $k$ with
  i.i.d.~mean-zero sub-Gaussian entries, there is an event $E$ such
  that:
\begin{align*}
  \Pr(E)\geq 1-\ee^{-\Theta(r)}\quad\text{and}\quad
  (1-\epsilon)\,\bar\P\preceq\E[\,\P\mid E\,]
  \preceq(1+\epsilon)\,\bar\P\quad\text{for}\quad
  \epsilon=O(\tfrac1{\sqrt r}),
\end{align*}
where $\P:= (\S\A)^\dagger\S\A,$ and
$\bar\P:=\gamma\A^\top\A(\gamma\A^\top\A + \I)^{-1}$,
for $\gamma>0$  such that $\tr\,\bar\P= k$.
\end{conjecture}

\noindent
The event $E$ is used to avoid dependence on the
condition number of $\A$. This seems to be necessary for
discrete-valued sketches (e.g., Rademacher), but it is likely avoidable for the Gaussian sketch.

\subsection{Expectation of incomplete projection,
  $\E[\H]$, where $\H=\S^\top(\S\A\A^\top\S^\top)^\dagger\S$} 
Note that the incomplete projection is related to the projection
matrix as follows:
\begin{align*}
  \P = (\S\A)^\dagger\S\A =
  \A^\top\S^\top(\S\A\A^\top\S^\top)^\dagger\S\A = \A^\top\H\A.
\end{align*}
This quantity appears in certain optimization methods such as Sketched
Newton-Raphson \cite{sketched-newton-raphson}.
Our surrogate expression for $\E[\H]$ should have the form
\begin{align*}
  \E[\H]\ \simeq\ \bar\H:=(\A\A^\top+\tfrac1\gamma\I)^{-1}.
\end{align*}

\begin{conjecture}
  Let $\S$ be a sketch of size $k$ with i.i.d.~mean-zero sub-Gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.  If
$\rho = r/k>1$ is a constant, then for
  any $z\neq 0$
  \begin{align*}
    (1-\epsilon)\,\bar\H\preceq\E[\H]\preceq
    (1+\epsilon)\,\bar\H\quad\text{with}\quad \epsilon =
    O(\tfrac1{\sqrt r}),
  \end{align*}
  where $\H := \S^\top(\S\A\A^\top\S^\top)^\dagger\S$
  and $\bar\H:=(\A\A^\top+\frac1\gamma\I)^{-1}$, for $\gamma>0$ such
  that $\tr\,\A^\top\bar\H\A = k$.
\end{conjecture}

\subsection{Higher-order corrections for ill-conditioned matrices}

\begin{figure}[h]
  \centering
\includegraphics[width=.4\textwidth]{higher}\nobreak\includegraphics[width=.4\textwidth]{discrepancy}
\caption{}
%\vspace{-5mm}
\label{f:higher-svd}
\end{figure}

The surrogate expression proposed in Theorem \ref{t:main} does not appear
to be correct when the spectrum of $\A$ contains a sharp drop. To
analyze this, consider the following simple example, where $\A^\top\A
= \diag(1,...,1,\kappa^{-1},...,\kappa^{-1})$ for some $\kappa\gg
1$. We can compare the surrogate expression $\bar\P$ with the expected
projection $\E[\P]$ for $\P=(\S\A)^\dagger\S\A$ (similar
observations follow for $\P_{\perp}$), when using a Gaussian
sketch. Note that for the Gaussian sketch, if $\A^\top\A$ is diagonal,
then so are $\E[\P]$ and $\bar\P$, so we can focus on comparing just the
spectrum of both matrices. This is shown in Figure~\ref{f:higher-svd}
(left), where we let $n=40$ and $r\approx 20$, with $\kappa=100$. We
observe that when $k\approx r$, then there is a discrepancy between
the surrogate and the expectation, particularly noticable for the
small eigenvalues. The relative error of the surrogate estimate for
the small eigenvalues as a function of $k$ is shown in Figure
\ref{f:higher-svd} (right), showing that the peak of the error occurs
as $k$ gets close to $r$ (the location of the peak is less than $r$,
but it approaches $r$ as $\kappa$ grows).

Note that this phenomenon presents itself already in the simple case
where $k=1$ and we let
$\A^\top\A=\diag(1,\kappa^{-1},...,\kappa^{-1})$. In this case, our
goal is to find a surrogate expression for a rank-one projection, so
we can formulate the following basic open problem:

\paragraph{Open problem:} Let
$\A=\diag(1,\kappa^{-\frac12},...,\kappa^{-\frac12})$ be an $n\times
n$ matrix, and define $\z\sim\Nc(\zero,\I)$. Find a surrogate
expression $\bar\P_{\kappa,n}$ such that:
\begin{align*}
  \lambda_i\Bigg(\E\bigg[\frac{\A^\top\z\z^\top\A}{\z^\top\A^\top\A\z}\bigg]\Bigg)
  \simeq \lambda_i\big(\bar\P_{\kappa,n}\big)\quad\text{for all $i$}\quad\text{as}\quad \kappa,n\rightarrow\infty.
\end{align*}

\subsection{Upper/lower bounds for the extreme eigenvalues}
\begin{figure}[h]
  \centering
\includegraphics[width=.4\textwidth]{lambda_min}\nobreak\includegraphics[width=.4\textwidth]{lambda_max}
\caption{}
%\vspace{-5mm}
\label{f:extreme}
\end{figure}

As observed in Figure \ref{f:higher-svd}, even when there is a
significant discrepancy between the eigenvalues of $\E[\P]$ and the
corresponding surrogate expressions, the extreme eigenvalues of
$\bar\P$ provide upper/lower bounds for the extreme eigenvalues of
$\E[\P]$. This has been consistently observed empirically for the Gaussian sketch
regardless of the spectral decay (see Figure \ref{f:extreme}).

\begin{conjecture}\label{h:extreme}
Let $\A$ be an $m\times n$ matrix. For a sketch $\S$ of any size $k$ with
  i.i.d.~standard normal entries we have:
\begin{align*}
  \lambda_{\min}(\bar\P)\leq \lambda_{\min}(\E[\P])\leq
  \lambda_{\max}(\E[\P])\leq \lambda_{\max}(\bar\P),
\end{align*}
where $\P= (\S\A)^\dagger\S\A,$ and
$\bar\P=\gamma\A^\top\A(\gamma\A^\top\A + \I)^{-1}$,
for $\gamma>0$  such that $\tr\,\bar\P= k$.
\end{conjecture}

\noindent
Note that the conjecture is very broad in that it does not require any
assumptions on $\A$ or on the sketch size $k$, because no
example has been found where the statement would be violated. It may
be worthwhile to look for such counter-example.
Further, it is possible that the conjecture also extends to other
sub-Gaussian sketches, or even beyond, although this has not been
verified empirically.

\subsection{Further convergence analysis of Gaussian Kaczmarz method}

The convergence guarantee given in Corollary \ref{c:kaczmarz} invites
the following directions for future work:
\begin{enumerate}
\item  The worst-case convergence rate depends on the smallest
  eigenvalue of $\E[\P]$ so we only have a weak additive
  approximation, which is vacuous when that eigenvalue is small
  enough.
  Can we avoid replace the additive approximation with multiplicative
  one, or eliminate it altogether?
\item The expected trajectory depends on the expected direction of
  $\E[\x^t-\x^*]$. A priori, this direction could match the smallest
  eigenvalue of $\E[\P]$. Is there a way to precondition the
  problem in such a way as to avoid this?
\end{enumerate}
Improving the worst-case convergence  is closely related to proving
Conjectures \ref{h:non-residual} and \ref{h:extreme}. In particular, if Conjecture \ref{h:extreme} is
true, then this eliminates \emph{any} dependence on the approximation
accuracy of the surrogate expressions and leads to a clean statement.
\begin{conjecture}[Corollary of Conjecture \ref{h:extreme}]
Let $\x^*$ be the unique solution of $\A\x^*=\b$ and consider
  the iterative algorithm:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S_t\A\x=\S_t\b.
  \end{align*}
Then, the following convergence guarantee holds:
\begin{align}
  \E\big[\|\x^{t+1}-\x^*\|^2\big]
  \leq (1-\bar\kappa)\,\E\big[\|\x^{t}-\x^*\|^2\big],
  \quad\text{where}\quad
  \bar\kappa = \frac{\sigma_{\min}^2(\A)}{\sigma_{\min}^2(\A)+1/\gamma}.
\end{align}
\end{conjecture}
Similar but weaker guarantee can be obtained as a corollary of
Conjecture \ref{h:non-residual}. There, we still suffer from the
approximation parameter $\epsilon$, but the dependence is now
multiplicative instead of additive, which means that the result is non-vacuous
even when the smallest eigenvalue is very small.

\subsection{High-probability guarantees and the spectral norm
error}

In addition to analyzing the low-rank approximation error using the
Frobenius norm, a large body of literature has also considered other
norms, and particularly the spectral norm \cite{tropp2011structure}:
\begin{align*}
  \text{(spectral norm error)}\quad \|\A-\A\P\| = \|\A\P_{\perp}\|.
\end{align*}
In addition to analysis of the low-rank approximation error, the above
quantity also appears in the analysis of certain randomized iterative
algorithms, such as the one proposed by \cite{lacotte2019high}.

Note that while the Frobenius norm error depends linearly on the residual projection,
i.e., $\|\A\-\A\P\|_F^2 = \tr(\A^\top\A\P_{\perp})$, the spectral norm
depends on it in a non-linear fashion. As a result, the analysis of
expected residual projection does not 
apply directly. Instead, one could hope to obtain high-probability
guarantees on the residual projection, which can then be used to
estimate the spectral norm error.

\paragraph{Open problem.} Let $r$ be the stable rank of $\A$. Show
that for any fixed vector $\x\in\R^n$, we have:
\begin{align*}
  \x^\top\P_{\!\perp}\x\ \overset\epsilon\simeq\ \x^\top(\gamma\Sigmab+\I)^{-1}\x\quad\text{with probability $1-\ee^{-cr}$}.
  \end{align*}

\subsection{Deterministic equivalents for matrix resolvents via 
  stable rank conditions}
Standard random matrix theory analysis focuses on constructing
deterministic equivalents for the resolvent matrix, defined as
follows:
\begin{align*}
  \Q(z) = (\A^\top\S^\top\S\A + z\I)^{-1}\quad\text{for}\quad z\in\mathbb{C}.
\end{align*}
The residual projection matrix which we analyze can be obtained via
the following limit:
\begin{align*}
  \P_{\!\perp}= \I - (\S\A)^\dagger\S\A =  \lim_{z\rightarrow 0}\ z\,\Q(z).
\end{align*}
Many of our results and conjectures can be naturally extended so that
they are also formulated for the resolvent matrices. The obtained
statements should be of independent interest in this context, because
of our use of stable rank to control the spectral properties of the
matrix $\A$ (as opposed to using the condition number, or bounding $z$
away from zero). For example, the following conjecture should follow
similarly as our main result.
\begin{conjecture}
  Let $\S$ be a sketch of size $k$ with i.i.d.~mean-zero sub-Gaussian entries and let
  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.  If
$\rho = r/k>1$ is a constant, then for
  any $z\neq 0$
  \begin{align*}
    (1-\epsilon)\,\bar\Q(z)\preceq\E[\Q(z)]\preceq
    (1+\epsilon)\,\bar\Q(z)\quad\text{with}\quad \epsilon =
    O(\tfrac1{\sqrt r}),
  \end{align*}
  where $\Q(z) := (\A^\top\S^\top\S\A + z\I)^{-1}$ and $\bar\Q(z):=(\gamma\A^\top\A +
  z\I)^{-1}$.
\end{conjecture}

\noindent
This prompts the question whether or not the conjecture leads to
a novel variant of asymptotic spectral analysis for random matrices, for example, one where we do
not require the limiting spectral distribution to be bounded away from
zero. Also, similarly as in the case of residual projection, we can
ask whether or not the results can be extended beyond the stable rank.

Another motivation for analyzing resolvents is that they naturally
arise when analyzing the bias of sketching for $l_2$-regularized
second-order methods (such as the Newton Sketch), with potential applications to distributed averaging.

\bibliographystyle{alpha}
  \bibliography{../pap}

\newpage
\appendix

\section{Analysis for Section \ref{s:non-residual}}

\begin{theorem}%\label{t:main-tech}
Let $\P = \X^\dagger \X$ for $\X = \S\A$, where $\A \in \R^{m \times n}$ with $m \ge n$ and $\S \in \R^{k\times m}$ has i.i.d.~$K$-sub-Gaussian entries having zero mean and unit variance covariance. Define: 
\begin{align*}
\bar\P = \gamma \A^\top \A (\gamma \A^\top \A + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P = k.
\end{align*}
Let $\Sigmab = \A^\top \A \in \R^{n \times n}$ be invertible (possible with $m \ge n$) and $r =\| \A \|_F^2/\|\A\|^2 = \tr(\Sigmab)/ \| \Sigmab \|$ be the stable rank of $\A$ and fix $\rho=r/k > 1$ that is assumed to be sufficiently large and $\tau = n/r$. There exists a constant
$C_{\rho,\tau}>0$, depending only on $\rho$, $\tau$ and $K$, then there is an event $E$ such that:
\begin{align*}
  \Pr(E)\geq 1-\ee^{-\Theta(r)}\quad\text{and}\quad
  (1-\epsilon)\,\bar\P\preceq\E[\,\P\mid E\,]
  \preceq(1+\epsilon)\,\bar\P\quad\text{for}\quad
  \epsilon=O(\tfrac1{\sqrt r}),
\end{align*}
\end{theorem}

\begin{proof}
Recall $\Sigmab = \A^\top \A$ and $\P = \X^\dagger \X$ and $\bar\P= \gamma\Sigmab
(\gamma\Sigmab+\I)^{-1}$. We use $\E_E$ to denote $\E[\cdot\mid E]$
   \begin{align*}
     \bar\P^{-\frac12}\E_E[\P]\bar\P^{-\frac12} - \I
     &=\bar\P^{-\frac12}(\E_E[\P]\bar\P^{-1} - \I)\bar\P^{\frac12}
     \\
     &=\bar\P^{-\frac12}(\I-\E_E[\P_\perp]\bar\P_\perp^{-1})\gamma^{-1}\Sigmab^{-1}\bar\P^{\frac12}
     % \\
     % &=\bar\P^{-\frac12}(\bar\gamma\Sigmab)^{-\frac12}(\I-\E_E[\P]\bar\P^{-1})(\bar\gamma\Sigmab)^{-\frac12}\bar\P^{\frac12}
     % \\
     % &=\bar\P^{-\frac12}\Sigmab^{-\frac12}
     %   (\T_1 + \T_2 + \T_3) \Sigmab^{-\frac12}\bar\P^{\frac12},
   \end{align*}
   where we recall $\P_\perp = \I - \P = \I - \X^\dagger \X$ and $\bar \P_\perp = (\gamma \Sigmab + \I)^{-1}$. Note that for any matrices $\A, \B$ of the same dimension, we have that $\A\B$ and $\B \A$ share the same eigenvalue. As a consequence, 
  \begin{align*}
    \| \bar\P^{-\frac12}\E_E[\P]\bar\P^{-\frac12} - \I \| &= \lambda_{\max} (\bar\P^{-\frac12}\E_E[\P]\bar\P^{-\frac12} - \I) = \lambda_{\max} \left( \Sigmab^{-\frac12} \gamma^{-1}( \I - \E_E[\P_\perp] \bar \P_\perp^{-1}) \Sigmab^{-\frac12} \right) \\
    & \le \left\| \Sigmab^{-\frac12} \gamma^{-1}( \I - \E_E[\P_\perp] \bar \P_\perp^{-1}) \Sigmab^{-\frac12} \right\| = \left\| \Sigmab^{-\frac12} (\T_1 + \T_2 + \T_3) \Sigmab^{-\frac12} \right\|
  \end{align*}
  with
   \begin{align*}
     \T_1
     &= \E_E\bigg[(\bar s - \hat s)\cdot
            \frac{ ( \I - \P_{-k}) \x_k\x_k^\top}{\x_k^\top ( \I - \P_{-k}) \x_k}\bigg],
     \\
     \T_2 
     &=\E_E[ ( \I - \P_{-k}) \x_k\x_k^\top] - \E_E[\I - \P_{-k}]\Sigmab,
     \\
     \T_3
     &=\E_E[ (\I - \P_{-k}) - \P_\perp]\Sigmab = \E_E[ \P - \P_{-k}]\Sigmab.
   \end{align*}
   for $\x_k \in \R^n$ the $k$-th row of $\X$, i.e., $\x_k = \A^\top \s_k$ for $\s_k \in \R^m$ the $k$-th row of $\S$ and $\P_{-k} = \X_{-k}^\dagger \X_{-k}$ for $\X_{-k} \in \R^{(k-1) \times n}$ the matrix obtained by deleting the $k$-th row from $\X$.

   For the term with $\T_1$ we have, with $\x_k = \A^\top \s_k$ that
\begin{align*}
      \| \Sigmab^{-\frac12} \T_1 \Sigmab^{-\frac12} \| &=  \left\| \E_E\bigg[(\bar s - \hat s)\cdot
            \frac{ \Sigmab^{-\frac12} ( \I - \P_{-k}) \A^\top \s_k\s_k^\top \A \Sigmab^{-\frac12} }{\x_k^\top ( \I - \P_{-k}) \x_k}\bigg] \right\|\\
      &\leq
     \underbrace{\E_E\Big[\big\|\Sigmab^{-\frac12} (\I - \P_{-k}) \Sigmab^{\frac12}\big\|\Big]}_{T_{1,1}}
     \cdot
     \underbrace{\E_E\bigg[\Big\|\E_E\Big[(\bar s- \hat s)\cdot
     \frac{\Sigmab^{-\frac12} \A^\top \s_k\s_k^\top \A \Sigmab^{-\frac12} }{\x_k^\top (\I - \P_{-k}) \x_k}\mid
     \P_{-k}\Big]\Big\|\bigg]}_{T_{1,2}}.
\end{align*}
To bound the term $T_{1,2}$ we write 
\begin{align*}
  \Big\|\E_E\Big[(\bar s- \hat s)\cdot
     \frac{\Sigmab^{-\frac12} \A^\top \s_k\s_k^\top \A \Sigmab^{-\frac12} }{\x_k^\top (\I - \P_{-k}) \x_k}\mid
     \P_{-k}\Big]\Big\| &\leq \frac32 \frac1{r-k} \cdot \E_E\bigg[\sup_{\| \v \| = 1}\E\Big[
            |\bar s-\hat s|\cdot \v^\top \Sigmab^{-\frac12} \A^\top \s_k \s_k^\top \A \Sigmab^{-\frac12} \v \mid
            \P_{-k}\Big]\bigg]
  \\
  &\leq \frac{C_\rho}{r}\cdot \E_E\bigg[\underbrace{\sqrt{\E_E\big[(\bar s-\hat
    s)^2 \mid\P_{-k}\big]}}_{ T_{1,2,1} }\cdot
    \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E_E\big[(\v^\top (\A^\top \A)^{-\frac12} \A^\top \s_k )^4\big]}}_{ T_{1,2,2} }\bigg]
\end{align*}
and it remains to show that $T_{1,2,2}$ is bounded by a constant that is independent of the \emph{minimum} eigenvalue of $\Sigmab$. To this end, note that with the svd of $\A \in \R^{m \times n}$ as $\A = \U \begin{bmatrix} \D \\ 0\end{bmatrix} \V^\top$ for orthonormal $\U \in \R^{m \times m}$, $\V \in \R^{n \times n}$ and diagonal $\D \in \R^{n \times n}$ that is invertible. As a consequence, we have
\[
  (\A^\top \A)^{-\frac12} \A^\top = \V \begin{bmatrix} \I_n & 0\end{bmatrix} \U^\top.
\]
Since $\s_k$ has independent zero mean $K$-sub-Gaussian random variable as entries, $\s_k$ is a $CK$-sub-Gaussian random vector \cite[Lemma~3.4.2]{vershynin2018high} for some universal constant $C> 0$ and so is $(\A^\top \A)^{-\frac12} \A^\top \s_k$. We thus have $T_{1,2,2} \le c K^2$ for some $c > 0$.

\medskip

To bound the term $\T_{1,1}$, we write, with $\X = \Z \Sigmab^{\frac12}$ that
\begin{align*}
 \big\|\Sigmab^{-\frac12}( \I -\P_{-k}) \Sigmab^{\frac12}\big\|
 &= \big\|\I -
   \Sigmab^{-\frac12}\X_{-k}^\dagger\X_{-k}\Sigmab^{\frac12}\big\|
 \\
 &\leq 1 +
   \big\|\Sigmab^{-\frac12}\Sigmab^{\frac12}\Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k}\Sigmab\big\| = 1 + \big\|\Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k}\Sigmab\big\| \\ 
 &\le 1 + \big\|\Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k}\big\| = 1 + \lambda_{\max} \left( \Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k} \right) \\ 
 &= 1 + \lambda_{\max} \left( (\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k} \Z_{-k}^\top \right) \le 1 + \| (\X_{-k}\X_{-k}^\top)^\dagger \| \cdot \| \Z_{-k} \Z_{-k}^\top \|.
\end{align*}

We introduce the following lemma to bound the minimum and maximum eigenvalue of $\X \X^\top = \Z \Sigmab \Z^\top \in \R^{k \times k}$, for $\Z \in \R^{k \times n}$ having i.i.d.\@ sub-Gaussian rows with $k < n$.

\begin{lemma}[Two-sided eigenvalue bounds on sub-Gaussian Gram matrix]\label{lem:eig-bound-Gram}\label{lem:eig-bound}
For $\X = \Z \Sigmab^{\frac12} \in \R^{k \times n}$ with p.s.d.\@ $\Sigmab \in \R^{n \times n}$ and $\Z$ having i.i.d.\@ zero mean, unit variance sub-Gaussian random variables with sub-Gaussian norm $\| \Z_{ij} \|_{\psi_2} \le K$, then for every $t \ge 0$, with probability at least $1 - 2 e^{-t^2}$ we have
\begin{equation}\label{eq:lem-eig-Gram}
  \| r^{-1} \Z \Sigmab \Z^\top - \I \| \le K^2 \max(\delta, \delta^2), \quad \delta = C \left( \sqrt{ \frac{k}{r} } + \frac{t}{\sqrt r} \right)
\end{equation}
for some constants $C > 0$. % that only depends on $K$.
\end{lemma}

\begin{proof}[Proof of Lemma~\ref{lem:eig-bound}]
The idea follows from the fact that $\E[\Z \Sigmab \Z^\top] = \tr (\Sigmab) \I_k = r \I_k$ so that we wish to have the concentration $\| r^{-1} \Z \Sigmab \Z^\top - \I_k \| \le C$ with high probability.

Following the proof of \cite[Theorem~4.6.1]{vershynin2018high}, Lemma~\ref{lem:eig-bound-Gram} essentially needs to control $\| \X^\top \u \|$ for all vectors $\u $ on the unit sphere $S^{k}$, the proof of which  comes in three steps: (i) we discretize the sphere using a net $\mathit N$ (the approximation step), so as to (ii) establish a tight control of $\| \X^\top \u \|$ for every fixed $\u \in \mathit N$ with high probability (the concentration step) and (iii) finish by taking a union bound over all $\u$ in the net.

\paragraph{Approximation.} With \cite[Corollary~4.2.13]{vershynin2018high}, we can evaluation the operator norm in \eqref{eq:lem-eig-Gram} on a $1/4$-net $\mathit N$ of $S^{k-1} $ that has cardinality $|\mathit N| \le 9^k$ as
\[
  \| r^{-1} \Z \Sigmab \Z^\top - \I \| \le 2 \max_{\u \in \mathit N} \left| \u^\top (r^{-1} \Z \Sigmab \Z^\top - \I) \u \right| = 2 \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 |
\]
so it suffices to prove, with the required high probability, we have
\[
  \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 | \leq \frac{\varepsilon}2, \quad \varepsilon := C K^2  \max(\delta, \delta^2) %\frac{2 C K^2}{\sqrt r}  \max(\delta, \delta^2)
\]
for $\delta$ defined in \eqref{eq:lem-eig-Gram}.

\paragraph{Concentration.} For any fixed $\u \in \R^k$ such that $\| \u \| =1$, we have $\| \X^\top \u \|^2 = \u^\top \Z \Sigmab \Z^\top \u$. Since $\Z \in \R^{k \times n}$ has i.i.d.\@ zero mean, unit variance sub-Gaussian entries with $\| \Z_{ij} \|_{\psi_2} \le K$, $\Z^\top \u \in \R^n$ has independent sub-Gaussian entries of zero mean and unit variance such that $\| [\Z^\top \u]_i \|_{\psi_2} \le C K$ for some universal constant $C > 0$ \cite[Proposition~2.6.1]{vershynin2018high}. As a consequence, we can bound, with Hanson-Wright inequality that, for symmetric positive semi-definite $\Sigmab \in \R^{n \times n}$ and $t \ge 0$:
\[
  \Pr \left\{ \left| \u^\top \Z \Sigmab \Z^\top \u - r  \right| \ge t \right\} \le 2 \exp \left( - c\,\min \left\{ \frac{t^2}{C^4 K^4 \tr\Sigmab^2 },  \frac{t}{C^2 K^2 \| \Sigmab \| } \right\} \right)
\]
for some universal constants $c, C > 0$. Taking $t = \varepsilon\,r/2$ with $\tr \Sigmab^2 \le \| \Sigmab \| \tr \Sigmab = r$, we get
\begin{equation}
  \Pr \left\{ \left| r^{-1} \u^\top \Z \Sigmab \Z^\top \u - 1  \right| \ge \varepsilon/2 \right\} \le 2 \exp \left( - c r \min \left\{ \frac{ \varepsilon^2 }{C^2 K^4 },  \frac{\varepsilon}{C K^2 } \right\} \right).
\end{equation}

\paragraph{Union bound.} We apply a union bound for all $\u$ in the net $\mathit N$ that has cardinality bounded by $9^k$ and obtain
\begin{align*}
  \Pr \left\{ 2 \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 | \ge \varepsilon \right\} &\le 9^k \cdot 2 \exp \left( - c r \min \left\{ \frac{ \varepsilon^2 }{C^2 K^4 },  \frac{\varepsilon}{C K^2 } \right\} \right) \\
  &\le 9^k \cdot 2 \exp \left( - c r \min \{ (\max\{\delta, \delta^2\} )^2, \max\{\delta, \delta^2\} \} \right) = 9^k \cdot 2 \exp (- c r \delta^2).
\end{align*}
By definition of $\delta$ in \eqref{eq:lem-eig-Gram}, we have $r \delta^2 \ge C^2 (k +  t^2)$ so that 
\[
  \Pr \left\{ 2 \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 | \ge \varepsilon \right\} \le 9^k \cdot 2 \exp \left(- c C^2 ( k + t^2) \right) \le 2 \exp (-t^2)
\]
if we choose $C$ large enough (for example in such as way that $c C^2 \ge 3$). This concludes the proof of Lemma~\ref{lem:eig-bound}.
\end{proof}

As a consequence of Lemma~\ref{lem:eig-bound}, we have in particular, with probability at least $1- 2 e^{-k^2}$, for fixed $r/k = \rho > 1$ that
\[
  |\lambda_{\min} (\X \X^\top) - r| \le K^2 \max ( 2 C \rho^{-1/2}, 4 C^2 \rho^{-1}) \cdot r.
\]
Then, for large enough $\rho$ such that $K^2 \max ( 2 C \rho^{-1/2}, 4 C^2 \rho^{-1}) \le 1 - c_\rho$ for some $0 < c_\rho < 1$, we have
\[
  \lambda_{\min}(\X \X^\top) \ge c_\rho r.
\]
As a consequence, with probability at least $1- 2 e^{-(k-1)^2}$, we have
\[
  \| (\X_{-k}\X_{-k}^\top)^\dagger \| \le \frac{C_\rho}{r}
\]
for some constant $C_\rho > 1$ depending on the ratio $\rho = r/k$ that is assumed to be large enough.

Taking $\Sigmab = \I_n$ (so that $r = n$) in Lemma~\ref{lem:eig-bound}, we can similarly show that, with probability at least $1- 2 e^{-(k-1)^2}$, for fixed $n/r = \tau$ so that $n/k = \tau \rho$,
\[
  \| \Z_{-k} \Z_{-k}^\top \| \le n \left( 1 + K^2 \max( 2 C (\tau \rho)^{-1/2}, 4 C^2 (\tau \rho)^{-1}) \right) = c_{\rho, \tau} n
\]
for some constant $c_{\rho, \tau} > 1$ that depends on the ratio $\rho = r/k$ and $n/r = \tau$.

As a result, we reach $\T_{1,1} \le 1 + \C_{\rho, \tau} \cdot \frac{n}{r} = \C_{\rho, \tau} \cdot \tau$.
 \end{proof}

\begin{lemma}\label{lem:rank-one-update}
    For $\X \in \mathbb R^{k \times n}$ with $k<n$, denote $\P = \X^\dagger \X$ and $\P_{-k} = \X_{-k}^\dagger \X_{-k}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its i-th row $\x_i \in \mathbb R^n$. Then, conditioned on the event $E_k: \left\{ \left| \frac{\tr \Sigmab (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} - 1 \right| \le \frac12 \right\}$:
    \begin{align*}
      (\X^\top\X)^\dagger\x_k = \frac{(\I - \P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}, \quad \P - \P_{-k} = \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top (\I - \P_{-k}) \x_k}.
    \end{align*}
\end{lemma}

\clearpage

\begin{theorem}%\label{t:main-tech}
Let $\P = \X^\dagger \X$ for $\X = \Z\Sigmab^{\frac12}$,
where $\Z \in \mathbb R^{k\times n}$ has i.i.d.~$K$-sub-Gaussian entries having zero mean and unit variance covariance, and $\Sigmab$ is an $n\times n$ positive semi-definite matrix. Define: 
\begin{align*}
\bar\P = \gamma \Sigmab (\gamma\Sigmab + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P = k.
\end{align*}
Let $r =\tr(\Sigmab)/\|\Sigmab\|$
be the stable rank of
$\Sigmab^{\frac12}$ and fix $\rho=r/k > 1$ that is assumed to be sufficiently large and $\tau = n/r$. There exists a constant
$C_{\rho,\tau}>0$, depending only on $\rho$, $\tau$ and $K$, then there is an event $E$ such that:
\begin{align*}
  \Pr(E)\geq 1-\ee^{-\Theta(r)}\quad\text{and}\quad
  (1-\epsilon)\,\bar\P\preceq\E[\,\P\mid E\,]
  \preceq(1+\epsilon)\,\bar\P\quad\text{for}\quad
  \epsilon=O(\tfrac1{\sqrt r}),
\end{align*}
\end{theorem}

\begin{proof}
Recall $\P = \X^\dagger \X$ and $\bar\P= \gamma\Sigmab(\gamma\Sigmab+\I)^{-1}$. We use $\E_E$ to denote $\E[\cdot\mid E]$
   \begin{align*}
     \bar\P^{-\frac12}\E_E[\P]\bar\P^{-\frac12} - \I
     &=\bar\P^{-\frac12}(\E_E[\P]\bar\P^{-1} - \I)\bar\P^{\frac12}
     \\
     &=\bar\P^{-\frac12}(\I-\E_E[\P_\perp]\bar\P_\perp^{-1})\gamma^{-1}\Sigmab^{-1}\bar\P^{\frac12}
     % \\
     % &=\bar\P^{-\frac12}(\bar\gamma\Sigmab)^{-\frac12}(\I-\E_E[\P]\bar\P^{-1})(\bar\gamma\Sigmab)^{-\frac12}\bar\P^{\frac12}
     % \\
     % &=\bar\P^{-\frac12}\Sigmab^{-\frac12}
     %   (\T_1 + \T_2 + \T_3) \Sigmab^{-\frac12}\bar\P^{\frac12},
   \end{align*}
   where we recall $\P_\perp = \I - \P = \I - \X^\dagger \X$ and $\bar \P_\perp = (\gamma \Sigmab + \I)^{-1}$. Note that for any matrices $\A, \B$ of the same dimension, we have that $\A\B$ and $\B \A$ share the same eigenvalue. As a consequence, 
  \begin{align*}
    \| \bar\P^{-\frac12}\E_E[\P]\bar\P^{-\frac12} - \I \| &= \lambda_{\max} (\bar\P^{-\frac12}\E_E[\P]\bar\P^{-\frac12} - \I) = \lambda_{\max} \left( \Sigmab^{-\frac12} \gamma^{-1}( \I - \E_E[\P_\perp] \bar \P_\perp^{-1}) \Sigmab^{-\frac12} \right) \\
    & \le \left\| \Sigmab^{-\frac12} \gamma^{-1}( \I - \E_E[\P_\perp] \bar \P_\perp^{-1}) \Sigmab^{-\frac12} \right\| = \left\|  \Sigmab^{-\frac12} (\T_1 + \T_2 + \T_3) \Sigmab^{-\frac12} \right\|
  \end{align*}
  with
   \begin{align*}
     \T_1
     &= \E_E\bigg[(\bar s - \hat s)\cdot
            \frac{ ( \I - \P_{-k}) \x_k\x_k^\top}{\x_k^\top ( \I - \P_{-k}) \x_k}\bigg],
     \\
     \T_2 
     &=\E_E[ ( \I - \P_{-k}) \x_k\x_k^\top] - \E_E[\I - \P_{-k}]\Sigmab,
     \\
     \T_3
     &=\E_E[ (\I - \P_{-k}) - \P_\perp]\Sigmab = \E_E[ \P - \P_{-k}]\Sigmab.
   \end{align*}

For the term with $\Sigmab^{-\frac12}\T_1 \Sigmab^{-\frac12}$ we have, with $\x_k = \Sigmab^{\frac12} \z_k$ that
\begin{align*}
      \|\Sigmab^{-\frac12}\T_1 \Sigmab^{-\frac12}\| &=  \left\| \E_E\bigg[(\bar s - \hat s)\cdot
            \frac{ \Sigmab^{-\frac12} ( \I - \P_{-k}) \Sigmab^{\frac12} \z_k\z_k^\top}{\x_k^\top ( \I - \P_{-k}) \x_k}\bigg] \right\|\\
      &\leq
     \underbrace{\E_E\Big[\big\|\Sigmab^{-\frac12} (\I - \P_{-k}) \Sigmab^{\frac12}\big\|\Big]}_{\T_{1,1}}
     \cdot
     \underbrace{\E_E\bigg[\Big\|\E_E\Big[(\bar s- \hat s)\cdot
     \frac{\z_k\z_k^\top}{\x_k^\top (\I - \P_{-k}) \x_k}\mid
     \P_{-k}\Big]\Big\|\bigg]}_{\T_{1,2}}.
\end{align*}
To bound the term $\T_{1,1}$, we write, with $\X = \Z \Sigmab^{\frac12}$ that
\begin{align*}
 \big\|\Sigmab^{-\frac12}( \I -\P_{-k}) \Sigmab^{\frac12}\big\|
 &= \big\|\I -
   \Sigmab^{-\frac12}\X_{-k}^\dagger\X_{-k}\Sigmab^{\frac12}\big\|
 \\
 &\leq 1 +
   \big\|\Sigmab^{-\frac12}\Sigmab^{\frac12}\Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k}\Sigmab\big\| = 1 + \big\|\Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k}\Sigmab\big\| \\ 
 &\le 1 + \big\|\Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k}\big\| = 1 + \lambda_{\max} \left( \Z_{-k}^\top(\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k} \right) \\ 
 &= 1 + \lambda_{\max} \left( (\X_{-k}\X_{-k}^\top)^{\dagger}\Z_{-k} \Z_{-k}^\top \right) \le 1 + \| (\X_{-k}\X_{-k}^\top)^\dagger \| \cdot \| \Z_{-k} \Z_{-k}^\top \|.
\end{align*}

We introduce the following lemma to bound the minimum and maximum eigenvalue of $\X \X^\top = \Z \Sigmab \Z^\top \in \R^{k \times k}$, for $\Z \in \R^{k \times n}$ having i.i.d.\@ sub-Gaussian rows with $k < n$.

\begin{lemma}[Two-sided eigenvalue bounds on sub-Gaussian Gram matrix]\label{lem:eig-bound-Gram}\label{lem:eig-bound}
For $\X = \Z \Sigmab^{\frac12} \in \R^{k \times n}$ with p.s.d.\@ $\Sigmab \in \R^{n \times n}$ and $\Z$ having i.i.d.\@ zero mean, unit variance sub-Gaussian random variables with sub-Gaussian norm $\| \Z_{ij} \|_{\psi_2} \le K$, then for every $t \ge 0$, with probability at least $1 - 2 e^{-t^2}$ we have
\begin{equation}\label{eq:lem-eig-Gram}
  \| r^{-1} \Z \Sigmab \Z^\top - \I \| \le K^2 \max(\delta, \delta^2), \quad \delta = C \left( \sqrt{ \frac{k}{r} } + \frac{t}{\sqrt r} \right)
\end{equation}
for some constants $C > 0$. % that only depends on $K$.
\end{lemma}

\begin{proof}[Proof of Lemma~\ref{lem:eig-bound}]
The idea follows from the fact that $\E[\Z \Sigmab \Z^\top] = \tr (\Sigmab) \I_k = r \I_k$ so that we wish to have the concentration $\| r^{-1} \Z \Sigmab \Z^\top - \I_k \| \le C$ with high probability.

Following the proof of \cite[Theorem~4.6.1]{vershynin2018high}, Lemma~\ref{lem:eig-bound-Gram} essentially needs to control $\| \X^\top \u \|$ for all vectors $\u $ on the unit sphere $S^{k}$, the proof of which  comes in three steps: (i) we discretize the sphere using a net $\mathit N$ (the approximation step), so as to (ii) establish a tight control of $\| \X^\top \u \|$ for every fixed $\u \in \mathit N$ with high probability (the concentration step) and (iii) finish by taking a union bound over all $\u$ in the net.

\paragraph{Approximation.} With \cite[Corollary~4.2.13]{vershynin2018high}, we can evaluation the operator norm in \eqref{eq:lem-eig-Gram} on a $1/4$-net $\mathit N$ of $S^{k-1} $ that has cardinality $|\mathit N| \le 9^k$ as
\[
  \| r^{-1} \Z \Sigmab \Z^\top - \I \| \le 2 \max_{\u \in \mathit N} \left| \u^\top (r^{-1} \Z \Sigmab \Z^\top - \I) \u \right| = 2 \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 |
\]
so it suffices to prove, with the required high probability, we have
\[
  \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 | \leq \frac{\varepsilon}2, \quad \varepsilon := C K^2  \max(\delta, \delta^2) %\frac{2 C K^2}{\sqrt r}  \max(\delta, \delta^2)
\]
for $\delta$ defined in \eqref{eq:lem-eig-Gram}.

\paragraph{Concentration.} For any fixed $\u \in \R^k$ such that $\| \u \| =1$, we have $\| \X^\top \u \|^2 = \u^\top \Z \Sigmab \Z^\top \u$. Since $\Z \in \R^{k \times n}$ has i.i.d.\@ zero mean, unit variance sub-Gaussian entries with $\| \Z_{ij} \|_{\psi_2} \le K$, $\Z^\top \u \in \R^n$ has independent sub-Gaussian entries of zero mean and unit variance such that $\| [\Z^\top \u]_i \|_{\psi_2} \le C K$ for some universal constant $C > 0$ \cite[Proposition~2.6.1]{vershynin2018high}. As a consequence, we can bound, with Hanson-Wright inequality that, for symmetric positive semi-definite $\Sigmab \in \R^{n \times n}$ and $t \ge 0$:
\[
  \Pr \left\{ \left| \u^\top \Z \Sigmab \Z^\top \u - r  \right| \ge t \right\} \le 2 \exp \left( - c\,\min \left\{ \frac{t^2}{C^4 K^4 \tr\Sigmab^2 },  \frac{t}{C^2 K^2 \| \Sigmab \| } \right\} \right)
\]
for some universal constants $c, C > 0$. Taking $t = \varepsilon\,r/2$ with $\tr \Sigmab^2 \le \| \Sigmab \| \tr \Sigmab = r$, we get
\begin{equation}
  \Pr \left\{ \left| r^{-1} \u^\top \Z \Sigmab \Z^\top \u - 1  \right| \ge \varepsilon/2 \right\} \le 2 \exp \left( - c r \min \left\{ \frac{ \varepsilon^2 }{C^2 K^4 },  \frac{\varepsilon}{C K^2 } \right\} \right).
\end{equation}

\paragraph{Union bound.} We apply a union bound for all $\u$ in the net $\mathit N$ that has cardinality bounded by $9^k$ and obtain
\begin{align*}
  \Pr \left\{ 2 \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 | \ge \varepsilon \right\} &\le 9^k \cdot 2 \exp \left( - c r \min \left\{ \frac{ \varepsilon^2 }{C^2 K^4 },  \frac{\varepsilon}{C K^2 } \right\} \right) \\
  &\le 9^k \cdot 2 \exp \left( - c r \min \{ (\max\{\delta, \delta^2\} )^2, \max\{\delta, \delta^2\} \} \right) = 9^k \cdot 2 \exp (- c r \delta^2).
\end{align*}
By definition of $\delta$ in \eqref{eq:lem-eig-Gram}, we have $r \delta^2 \ge C^2 (k +  t^2)$ so that 
\[
  \Pr \left\{ 2 \max_{\u \in \mathit N} | r^{-1} \| \X^\top \u \|^2 - 1 | \ge \varepsilon \right\} \le 9^k \cdot 2 \exp \left(- c C^2 ( k + t^2) \right) \le 2 \exp (-t^2)
\]
if we choose $C$ large enough (for example in such as way that $c C^2 \ge 3$). This concludes the proof of Lemma~\ref{lem:eig-bound}.
\end{proof}

\textbf{** Comments: to extend to general sub-Gaussian random vector, we need to use sub-Gaussian concentration to bound the probability
\[
  \Pr \left\{ | \u^\top \Z \Sigmab \Z^\top \u - r| \ge t \right\}
\]
however, the vector $\Z^\top \u$ is in general not sub-Gaussian having identity covariance. I have tried also Hanson-Wright-type inequality for general sub-Gaussian random vectors such as \cite[Corollary 2.9]{zajkowski2018bounds}, but the rate is not good enough, in general we need $r = O(k^2)$ in that case. **}

As a consequence of Lemma~\ref{lem:eig-bound}, we have in particular, with probability at least $1- 2 e^{-k^2}$, for fixed $r/k = \rho > 1$ that
\[
  |\lambda_{\min} (\X \X^\top) - r| \le K^2 \max ( 2 C \rho^{-1/2}, 4 C^2 \rho^{-1}) \cdot r.
\]
Then, for large enough $\rho$ such that $K^2 \max ( 2 C \rho^{-1/2}, 4 C^2 \rho^{-1}) \le 1 - c_\rho$ for some $0 < c_\rho < 1$, we have
\[
  \lambda_{\min}(\X \X^\top) \ge c_\rho r.
\]
As a consequence, with probability at least $1- 2 e^{-(k-1)^2}$, we have
\[
  \| (\X_{-k}\X_{-k}^\top)^\dagger \| \le \frac{C_\rho}{r}
\]
for some constant $C_\rho > 1$ depending on the ratio $\rho = r/k$ that is assumed to be large enough.

Taking $\Sigmab = \I_n$ (so that $r = n$) in Lemma~\ref{lem:eig-bound}, we can similarly show that, with probability at least $1- 2 e^{-(k-1)^2}$, for fixed $n/r = \tau$ so that $n/k = \tau \rho$,
\[
  \| \Z_{-k} \Z_{-k}^\top \| \le n \left( 1 + K^2 \max( 2 C (\tau \rho)^{-1/2}, 4 C^2 (\tau \rho)^{-1}) \right) = c_{\rho, \tau} n
\]
for some constant $c_{\rho, \tau} > 1$ that depends on the ratio $\rho = r/k$ and $n/r = \tau$.

As a result, we reach $\T_{1,1} \le 1 + \C_{\rho, \tau} \cdot \frac{n}{r} = \C_{\rho, \tau} \cdot \tau$.
\end{proof}

\end{document}

### 2022-08-23 design/projections.tex
Commit original repos

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
  colorlinks,
  linkcolor={red!40!gray},
  citecolor={blue!40!gray},
  urlcolor={blue!70!gray}
}
\usepackage{cleveref}
\input{shortdefs}

\title{Exact expressions for random projections:
low-rank approximation, randomized second-order methods and implicit regularization}

% \setlength{\parindent}{0cm}
% \setlength{\parskip}{0mm}

\begin{document}
\maketitle

\section{Main result}
\paragraph{Main Theorem.}
Let $\X = \Z\Sigmab^{\frac12}$, where $\Z$ is an $n\times d$ random matrix with
entries jointly
$K$-sub-Gaussian (sub-Gaussian Orlicz norm
$\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$),
and $\Sigmab$ is a $d\times d$
positive definite matrix. Assume that $n<d$ and define:
\begin{align*}
\bar\P_{\Sigmab,n} = (\bar\gamma\Sigmab +
  \I)^{-1}\bar\gamma\Sigmab,\quad\text{so that}\quad\tr\,\bar\P=n. 
\end{align*}
Let $r=\frac{\tr(\Sigmab)}{\|\Sigmab\|}$ be the nuclear rank of
$\Sigmab$ and fix $\rho=\frac rn > 1$. There is $C_{\rho}$ such that if $r\geq C_\rho$, then
\begin{align}
\Big(1-\frac {C_\rho K^2}{\sqrt
  r}\Big)\cdot\bar\P_{\Sigmab,n}\ \preceq\ \E[\X^\dagger\X]\ \preceq\ 
  \Big(1+\frac {C_\rho K^2}{\sqrt r}\Big)\cdot\bar\P_{\Sigmab,n}.
\end{align}

\begin{corollary}[Low-rank approximation]
  Let $\A$ be an $m\times n$ matrix with stable rank
  $r=\frac{\|\A\|_F^2}{\|\A\|^2}$ and define
  $\Er_\A(\tilde\A)=\|\A-\P_{\!\tilde\A}\A\|_F^2$, where $\P_{\!\tilde\A}$
  is the projection onto the column-span of $\tilde\A$. If $\S$ is an
  $n\times k$ matrix with i.i.d.~Gaussian entries and $k<r$, then we have
  \begin{align*}
    \E\big[\Er_\A(\A\S)\big]  \overset{\epsilon}{\simeq}
    \tr\,\A\A^\top\big(\I-\bar\P_{\A\A^\top\!,k}\big).
  \end{align*}
\end{corollary}
\begin{corollary}[Iterative optimization]
  Given $m\times n$ matrix $\A$ with stable rank $r$ and vector
  $\b\in\R^m$, let $\x^*$ be the solution of $\A\x^*=b$ and consider
  the iterative algorithm:
  \begin{align*}
    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\text{subject to}\quad\S\A=\S\b,
  \end{align*}
  where $\S$ is a $k\times m$ sub-Gaussian sketching matrix with
  $k<r$. Then we have:
  \begin{align*}
    \E\big[\x^{t+1}-\x^*\big] &\overset{\epsilon}{=}
    \big(\I - \bar\P_{\A^\top\!\A,k}\big)\cdot\E\big[\x^t-\x^*\big],\\
    \E\big[\|\x^{t+1}-\x^*\|^2\big]&\leq
\Big(1-\lambda_{\min}\big((1-\epsilon)\bar\P_{\A^\top\!\A,k}\big)\Big)\cdot
    \E\big[\|\x^t-\x^*\|^2\big].
  \end{align*}
\end{corollary}
\begin{corollary}[Implicit regularization]
Let $f_\w(\x)=\x^\top\w$ and $(\x,y)\sim\mu$
  be s.t.~$\x\sim\Nc(0,\Sigmab)$ and $y=\x^\top \bar \w+\xi$,
  with $\E[\xi]=0$. Assume that $\Sigmab$ has nuclear rank
  $r>n$. Then,
  \begin{align*}
    \E_{\Dc\sim\mu^n} \Big[\argmin_{\w}\|\w\|^2\text{ s.t. }f_\w(\x)=y\,\forall_{(\x,y)\in \Dc}\Big]
   \ \overset{\epsilon}{\simeq}\
    \argmin_{\w}\Big\{\E_{(\x,y)\sim\mu}\big(f_\w(\x)-y\big)^2 + \lambda\|\w\|^2\Big\},
  \end{align*}
  where $\lambda$ is chosen so that $n=\tr\,\Sigmab(\Sigmab+\lambda\I)^{-1}$.
\end{corollary}

\noindent
% Note that this expression for the low-rank approximation error of a
% Gaussian sketch is identical to the expected error of DPP sampling in the Column
% Subset Selection Problem (CSSP) given in
% \cite{nystrom-multiple-descent}, where several bounds are given for
% the expression in terms of the best rank $k$ approximation.
Further research questions associated with the corollaries:
\begin{enumerate}
\item Analyze which sketches are sub-Gaussian and with what constant?
  \begin{enumerate}
    \item $z_{ij} = N(0,1)$, then $K = \|z_{ij}\|_{\psi_2} = \sqrt{8/3} < 1.64$
      (\cite{van1996weak} exercise 2.2.1).
  \item Simple Rademacher sketch (i.i.d.~$\pm1$ entries)
    \begin{align*}
      \|z_{ij}\|_{\psi_2} = 
      \inf \left\{ c > 0 : \E e^{x^2 / c^2} \leq 2\right\}
      = \log(2)^{-1/2} < 1.21
    \end{align*}
  \item Sparsified Rademacher sketch (randomly zeroed out i.i.d.~$\pm1$ entries),
  \item Column/row sampling (uniform, leverage score, etc). For uniform sampling
    with replacement, $\S^\top \in \R^{k \times n}$ has rows which
    are uniformly distributed over 
    $\sqrt{\frac{n}{k}} e_i$ where $e_i \in \R^n$ is the $i$th standard
    basis vector. Therefore,
    for $\|\a\| \leq 1$
    \begin{align*}
      \E e^{\a^\top \z}
      = \sum_{i=1}^n \frac{1}{n} e^{a_i \sqrt{\frac{n}{k}}}
      \leq e^{\frac{1}{\sqrt{k}}}
    \end{align*}
    with equality attained when $\a$ has entries identically 
    $\frac{1}{\sqrt{n}}$. As a consequence, the sub-Gaussian
    Orlicz norm $K^2 \leq C \frac{2}{\sqrt{k}}$
    for some universal constant $C$.

    Similar analysis can be extended to non-uniform discrete distributions
    over $[n]$ (like leverage scores) by suitably replacing $\frac{1}{n}$, but 
    the results may have additional dependence on $\Sigmab$.
  \item Subsampled Randomized Hadamard Transform (probably harder
    since it is non-i.i.d.).
  \end{enumerate}
  \item Can we show this for the spectral norm error, i.e.,
    $\|\A-\P_{\!\A\S}\A\|$,  instead of the Frobenius norm? Presumably this
    requires directly analyzing $\I-\X^\dagger\X$ instead of $\E[\I-\X^\dagger\X]$.
  \item Extend the result to $k>r$, possibly under the assumption of slow eigenvalue decay.
  \end{enumerate}

\begin{theorem}\label{t:main}
% Let $\X = \Z\Sigmab^{\frac12}$, where $\Z$ is an $n\times d$ random
% matrix with entries jointly $K$-sub-Gaussian (sub-Gaussian Orlicz norm
% $\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$), and $\Sigmab$ is a $d\times d$ positive definite matrix. Assume that $n<d$ and define:
% \begin{align*}
% \bar\P=\bar\P(\Sigmab,n) = (\bar\gamma\Sigmab + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P=d-n.
% \end{align*}
% Let $r=\frac{\tr(\Sigmab)}{\|\Sigmab\|}$ be the nuclear rank of
% $\Sigmab$ and fix $\rho=\frac rn > 1$. There is $C_{\rho}$ such that if $r\geq C_\rho$, then
% \begin{align}
% \Big(1-\frac {C_\rho K^2}{\sqrt
%   r}\Big)\cdot\bar\P\preceq\E[\I-\X^\dagger\X]\preceq
%   \Big(1+\frac {C_\rho K^2}{\sqrt r}\Big)\cdot\bar\P.
% \end{align}
Let $\X = \Z\Sigmab^{\frac12}$, where $\Z \in \mathbb R^{n\times d}$ having i.i.d.~rows that are $K$-sub-Gaussian random vectors (with sub-Gaussian Orlicz norm $\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$), and $\Sigmab$ is a $d\times d$ positive definite matrix. Assume that $n<d$ and define:
\begin{align*}
\bar\P=\bar\P(\Sigmab,n) = (\bar\gamma\Sigmab + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P=d-n.
\end{align*}
Let $r=\frac{\tr(\Sigmab)}{\|\Sigmab\|}$ be the nuclear rank of
$\Sigmab$ and fix $\rho=\frac rn > 1$. There is $C_{\rho}$ such that if $r\geq C_\rho$, then
\begin{align}
\Big(1-\frac {C_\rho K^2}{\sqrt
  r}\Big)\cdot\bar\P\preceq\E[\I-\X^\dagger\X]\preceq
  \Big(1+\frac {C_\rho K^2}{\sqrt r}\Big)\cdot\bar\P.
\end{align}
\end{theorem}
\begin{proof}
  Note that $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$
  for any $\alpha \in \R \setminus \{0\}$, so
  assume without loss of generality (after rescaling $\bar{\P}$ correspondingly)
  that $\|\Sigmab\| = 1$.

  Let $\P=\I-\X^\dagger\X$ and define $\P_{-n}=\I-\X_{-n}^\dagger\X_{-n}$,
  where $\X_{-n}$ denotes the matrix $\X$ without $n$th row. We will use the
  following.

\begin{lemma}\label{l:rank-one-update}
    For any $n<d$, we have, conditioned on the event $E_n: \left\{ \left| \frac{\tr \P_{-n} \Sigmab}{\x_n^\top \P_{-n} \x_n} - 1 \right| \le \frac12 \right\}$ that
    \begin{align*}
       \P_{-n} - \P = \frac{\P_{-n} \x_n \x_n^\top \P_{-n}}{\x_n^\top \P_{-n} \x_n}, \quad (\X^\top\X)^\dagger\x_n =
      \frac{\P_{-n}\x_n}{\x_n^\top\P_{-n}\x_n}.
    \end{align*}
\end{lemma}

% \begin{lemma}[Concentration of quadratic forms of sub-Gaussian random vectors, {\cite[Corollary 2.6]{zajkowski2018bounds}}] \label{l:concentration-sub-gaussian}
% For a $K$-sub-Gaussian random vector $\x \in \mathbb R^d$ (i.e., $\| \x \|_{\psi_2} \leq K$) and $\A \in \mathbb R^{d \times d}$, we have
% \begin{align*}
%   \Pr\left[
%     \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert
%     \geq t
%     \right] \leq 2 \exp\left(
%     -\min\left\{
%       \frac{t^2}{16 C^2  K^4 (\tr \A)^2},
%       \frac{t}{4 C K^2 (\tr \A)},
%     \right\}
%   \right)
% \end{align*}
% and 
% \begin{align*}
%   \E\left[ \left( \x^\top \A \x - \E[\x^\top \A \x] \right)^2 \right]
%   &\leq C~K^4~\tr(\A)^2
% \end{align*}
% for some universal constant $C$.
% \end{lemma}

\begin{lemma}[Concentration of quadratic forms of sub-Gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}] \label{l:concentration-sub-gaussian}
For a $K$-sub-Gaussian random vector $\x \in \mathbb R^d$ (i.e., $\| \x \|_{\psi_2} \leq K$) with $\E[\x] = 0$, $\E[\x \x^\top] = \I_d$ and symmetric and p.s.d.~$\A \in \mathbb R^{d \times d}$, we have
\begin{align*}
  \Pr\left[
    \lvert \x^\top \A \x - \tr \A \rvert
    \geq \frac12 \tr \A
    \right] \leq 2 \exp\left( - \left(  \frac{ \sqrt{  \tr \A^2 + \| \A \| \tr \A } - \sqrt{\tr \A^2} }{2 \| \A \|} \right)^2 \right) 
\end{align*}
and 
\begin{align*}
  \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
  &\leq C~K^4~\tr(\A)^2
\end{align*}
for some universal constant $C>0$. In particular, if we have $\| \A \| \le 1$ and $\tr \A^2 \le \tr \A \le r$, then
\[
  \frac{ \sqrt{  \tr \A^2 + \| \A \| \tr \A } - \sqrt{\tr \A^2} }{2 \| \A \|} = \frac12 \frac{ \tr \A }{\sqrt{  \tr \A^2 + \| \A \| \tr \A } + \sqrt{\tr \A^2} } \ge \frac{\tr \A}{C \sqrt r}
\]
so that
\[
  \Pr\left[
    \lvert \x^\top \A \x - \tr \A \rvert
    \geq \frac12 \tr \A
    \right] \leq 2 \exp\left( - \frac{(\tr \A)^2}{C r} \right)
\]
for some $C > 0$.
\end{lemma}

\textbf{**In fact \cite[Theorem~2.1]{hsu2012tail} can not be applied directly, due to the mismatch in definition of sub-Gaussian and the form of \cite[Theorem~2.1]{hsu2012tail}**: we cannot really assume $\sigma = 1$ there, because that imposes constraints on the sub-Gaussian norm.}

% \begin{lemma}\label{l:rank-one-projection}
%   For any $d\times d$ p.s.d.~matrix $\A$ and
%   any random vector $\z$ with jointly
%   $K$-sub-Gaussian (sub-Gaussian Orlicz norm
%   $\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$),
%   \begin{align*}
%     \Bigg\|\E\bigg[\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\bigg]\Bigg\|
% \leq
%     \sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\v\Big)^2\bigg]}
%     \leq C\cdot K^2 \cdot \frac{\|\A\|}{\tr\,\A}.
%   \end{align*}
%   for some universal constant $C$.
% \end{lemma}
%\textbf{** Lemma 3 seems no longer necessary?**}

\begin{lemma}\label{l:trace}
For $X=\tr\,\Sigmab(\P_{-n}-\E[\P_{-n}])$
and $\|\Sigmab\| = 1$, we have
\begin{align*}
  \E[X^2]\leq
  cn\quad\text{and}\quad\Pr\{|X|\geq t\}\leq 2 \ee^{-\frac{t^2}{cn}}.
  \end{align*}
  for some universal constant $C$.
\end{lemma}

\noindent
Defining $\hat s = \x_n^\top\P_{-n}\x_n$ and $\bar s=\tr\bar\P\Sigmab=n/\bar\gamma$,
using the lemmas and the fact that $\I-\P=(\X^\top\X)^\dagger\X^\top\X$, we obtain:
\begin{align*}
  \I - \E[\P]\bar\P^{-1}
  &= \E[\I -\P] - \E[\P\bar\gamma\Sigmab]\\
  &=\sum_{i=1}^n\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top\big] -
    \E[\P]\bar\gamma\Sigmab\\
  &=n\,\E\bigg[\frac{\P_{-n}\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\bigg] -
    \E[\P]\bar\gamma\Sigmab\\
  &=\bar\gamma\,
    \E\bigg[\bar s\cdot\frac{\P_{-n}\x_n\x_n^\top} {\x_n^\top\P_{-n}\x_n}\bigg]
    -\bar\gamma\,\E\bigg[\hat s
    \cdot\frac{\P_{-n}\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\bigg] +
    \bar\gamma\,\E[\P_{-n}\Sigmab]
    -\E[\P]\bar\gamma\Sigmab\\
  &=\underbrace{\bar\gamma\,\E\bigg[(\bar s - \hat s)
    \cdot\frac{\P_{-n}\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\bigg]}_{\T_1}
+    \underbrace{\E[\P_{-n}-\P]\bar\gamma\Sigmab}_{\T_2}.
\end{align*}
We bound $\T_1$ as follows:
\begin{align*}
  \bar\gamma^{-1}\|\T_1\|
  &\leq \E\Bigg[\bigg\|\E\Big[(\bar s-\hat
  s)\cdot\frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}\mid
  \P_{-n}\Big]\bigg\|\Bigg]\\
  &\leq \E\Bigg[\sqrt{\E\big[(\bar s - \hat
  s)^2\mid
    \P_{-n}\big]}\
    \cdot\sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\P_{-n}\x_n\x_n\P_{-n}}{\x_n^\top\P_{-n}\x_n}\v\Big)^2\mid\P_{-n}\bigg]}\,\Bigg]\\
  &\leq  C\,K^2\,\E\Bigg[\sqrt{\E\big[(\bar s - \hat
  s)^2\mid
    \P_{-n}\big]}\,
    \cdot \frac{\|\P_{-n}\Sigmab\|}{\tr\,\P_{-n}\Sigmab}\Bigg]
% \ \leq\ C\,\sqrt{\E\big[(\bar s - \hat s)^2\big]}\,\cdot
% \underbrace{\sqrt{\E\bigg[\frac{\|\P_{-n}\Sigmab\|^2}{(\tr\,\P_{-n}\Sigmab)^2}\bigg]}}_{\Delta}\\[-4mm]
    \  \leq\ C\sqrt{(\bar s - s)^2 + \E\big[(s - \hat s)^2\big]}\cdot \frac{K^2}{r-n},
\end{align*}
where we let $s=\E[\hat s]=\tr\,\E[\P_{-n}]\Sigmab$ and the last
inequality follows because:
\begin{align*}
    \frac{\tr\,\P_{-n}\Sigmab} {\|\P_{-n}\Sigmab\|}\geq
    \tr\,\P_{-n}\Sigmab =
    \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-n}^\dagger\X_{-n}\|_F^2\geq
\sum_{i\geq n}\lambda_i\geq r-n,
\end{align*}
with $1=\lambda_1\geq\lambda_2\geq...$ denoting the eigenvalues of $\Sigmab$.
Furthermore, we have:
\begin{align*}
  \E\big[(s - \hat s)^2\big] &=
\E\big[\big(\tr\,\Sigmab(\E[\P_{-n}]-\P_{-n})\big)^2\big] +
  \E\big[(\tr\,\P_{-n}\Sigmab-\x_n^\top\P_{-n}\x_n)^2\big]\\
  &\leq C\Big(n + \E\big[\tr\,(\P_{-n}\Sigmab)^2\big]\Big)\leq
    C\big(n + \tr\,\E[\P_{-n}]\Sigmab\big)\\
  &=C(n+s)\leq C\big(n+\bar s + |s- \bar s|\big).
\end{align*}
Moreover, using the fact that
$\bar\P\Sigmab\preceq\frac1{\bar\gamma+1}\I$, we obtain that:
\begin{align*}
  |\bar s - s|
  &= \big|\tr\,(\I-\E[\P]\bar\P^{-1})\bar\P\Sigmab\big|
    =\big|\tr\,(\T_1+\T_2)\bar\P\Sigmab\big|\\
  &\leq \bar\gamma\,\E\bigg[|\bar s-\hat s|\cdot
    \frac{\tr\,\P_{-n}\x_n\x_n^\top\bar\P\Sigmab}{\tr\,\P_{-n}\x_n\x_n^\top}\bigg]
    + \bar\gamma\,
    \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
    {\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}}\bigg]\\
  &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(\E\big[|\bar s-\hat
    s|\big] + 1\Big)\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar
    s-s| + \E\big[|s-\hat
    s|\big] + 1\Big)\\
  &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| +
     c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
\end{align*}
Solving for $|\bar s-s|$, we conclude that
\begin{align*}
  |\bar s-s|\leq
  \bar\gamma(c\sqrt{n+\bar s}+1) + c^2\bar\gamma^2\leq \bar\gamma
  (c+1)\sqrt{n+\bar s} + c^2\bar\gamma^2.
\end{align*}
We next bound $\T_2$. Note that $\E[\P_{-n}-\P] =
\E\big[\frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}\big]$, so
using Lemma \ref{l:rank-one-projection}, we have:
\begin{align*}
\bar\gamma^{-1}\|\T_2\|\leq  \bigg\|
  \E\bigg[\E\Big[\frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}
  \mid \P_{-n}\Big]\bigg]\bigg\|  \leq
C\,K^2\,\E\Big[\frac{\|\P_{-n}\Sigmab\|}{\tr\,\P_{-n}\Sigmab}\Big]   
\leq C\cdot \frac{K^2}{r-n}.
\end{align*}
% Finally, we bound $\Delta$ by using the concentration of
% $\tr\,\P_{-n}\Sigmab$ around its mean $s$, combined with 
% the fact that  $\|\P_{-n} \Sigmab\| \leq \min\{1,\tr\,\P_{-n}\Sigmab\}$:
% \begin{align}
% \Delta
%   \leq    \sqrt{\E\bigg[\frac{\|\P_{-n}\Sigmab\|^2}{(\tr\,\P_{-n}\Sigmab)^2}
%     \mid \tr\,\P_{-n}\Sigmab\geq s/2\bigg] +
% \Pr\big\{\tr\,\P_{-n}\Sigmab\leq s/2\big\}}
%   \leq \sqrt{\tfrac 4{s^2} + 2 \ee^{-\frac{s^2}{cn}}}.\label{eq:projected-rank}
% \end{align}
% The high probability bound used above is derived from Lemma \ref{l:trace}:
% \begin{align*}
%   \Pr\big\{\tr\,\P_{-n}\Sigmab\leq s/2\big\}
%   &= \Pr\big\{\tr\,(\P_{-n} - \E[\P_{-n}])\Sigmab\leq - s/2 \big\}
%   \leq 2 \ee^{-\frac{s^2}{c n}}
% \end{align*}
% To complete the proof, we must show that $s$ is large enough for the
% bound in \eqref{eq:projected-rank} to be meaningful.
Using the fact that $\bar\P\succeq \frac1{\bar\gamma+1}\I$,
we have:
\begin{align*}
n = d - \tr\,\bar\P = \tr\,\bar\P (\bar\gamma\Sigmab + \I) - \tr\,\bar\P
  = \tr\,\bar\P \bar\gamma\Sigmab\geq \frac{\bar\gamma}{\bar\gamma+1}\tr\,\Sigmab,
\end{align*}
so $r=\tr\,\Sigmab\leq \frac{\bar\gamma+1}{\bar\gamma}n=n + \frac
n{\bar\gamma}=n+\tr\,\bar\P\Sigmab$. Thus, $r\geq\bar s \geq r-n$ and also $\bar\gamma\leq
\frac{n}{r-n}=\frac1{\rho-1}$.
% Assume that $r\geq \frac{4(c+1)^2\rho^2}{(\rho-1)^4}$. Then, we have:
% \begin{align*}
% \bar s-n\geq r-2n=r\,\frac{\rho-2}\rho\ \geq
%   \frac c{\sqrt{\rho/2}}\sqrt{r/\rho}\geq \frac{c}{\rho-1}\sqrt n\geq \bar\gamma c\sqrt{n}.
% \end{align*}
% \begin{align*}
%   n+s
%   &\geq  n+ \bar s - \bar\gamma(c+1)\sqrt{n+\bar s} -c^2\bar\gamma^2
%   =
%   (n+\bar s)\cdot \Big(1 - \frac{\bar\gamma (c+1)}{\sqrt{n+\bar s}} -
%     \frac{c^2\bar\gamma^2}{n+\bar s}\Big)\\
%   &\geq r\cdot\Big(1-\frac{\bar\gamma (c+1)}{\sqrt r} -
% \frac{c^2\bar\gamma^2}{r}\Big)
%   \geq \rho n\cdot \Big(1 - \frac{\rho-1}{2\rho} -
%   \frac{(\rho-1)^2}{4\rho^2} \Big)\\
%   &\geq \rho n\cdot\Big(1 - \frac34\,\frac{\rho-1}{\rho}\Big) = n +
%     n\cdot \frac34(\rho-1),
% \end{align*}
% so we conclude that $s\geq n\cdot \frac
% 34(\rho-1)=r\cdot\frac34\frac{\rho-1}{\rho}$. Using the fact that,
% for any $\alpha>0$, function 
% $f(x)=x^2\ee^{-\alpha x}$ is bounded over $x>0$, we have
% \begin{align*}
% r^2\Big(\frac4{s^2}+2 \ee^{-\frac{s^2}{cn}}\Big) \leq \frac 43\frac\rho{\rho-1} + 2 r^2
%   \ee^{-r\cdot\frac34\frac{\rho-1}{c}}\leq c_\rho,
% \end{align*}
% thus concluding that $\Delta^2\leq \frac{c_\rho}{r^2}$.
Observing that $\frac1{r-n} = \frac1r\cdot \frac{\rho}{\rho-1}$, we
put everything together:
\begin{align*}
  \|\I-\E[\P]\bar\P^{-1}\|\leq \frac{C\,K^2\,\bar\gamma}{r-n}\cdot
  \big(1 + \bar\gamma\sqrt{r} + \bar\gamma^2 + \sqrt r\big)
  \leq \frac{C_\rho K^2}{\sqrt r}.
\end{align*}
% \begin{align*}
%   \|\I-\E[\P]\bar\P^{-1}\|\leq \bar\gamma C_1\Delta\sqrt r \Big(1 +
%   \bar\gamma+\bar\gamma^2\Big)
%   +\bar\gamma C_2\Delta\leq \frac{C_\rho}{\sqrt r},
% \end{align*}
It remains to observe that since $\I\succeq\bar\P\succeq
\frac1{\bar\gamma+1}\I\succeq \frac{\rho-1}{\rho}\I$, we have
\begin{align*}
  \|\I-\bar\P^{-\frac12}\E[\P]\bar\P^{-\frac12}\| =
  \|\bar\P^{-\frac12}(\I-\E[\P]\bar\P^{-1})\bar\P^{\frac12}\|\leq
\|\bar\P^{-\frac12}\|\cdot\frac{C_\rho K^2}{\sqrt r}\cdot \|\bar\P^{\frac12}\|\leq
\frac{C_\rho K^2}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
\end{align*}
Defining $\epsilon =\frac{C_\rho K^2}{\sqrt r}
\sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
of the p.s.d.~matrix $\bar\P^{-\frac12}\E[\P]\bar\P^{-\frac12}$ lie
in the interval $[1-\epsilon,1+\epsilon]$, so:
\begin{align*}
  (1-\epsilon)\I\preceq
  \bar\P^{-\frac12}\E[\P]\bar\P^{-\frac12}\preceq (1+\epsilon)\I.
\end{align*}
Note that for any p.s.d.~matrices $\A,\B,\Z$, if $\A\preceq \B$ then
$\Z\A\Z\preceq \Z\B\Z$, so letting $\Z=\bar\P^{\frac12}$ in the above,
we get:
\begin{align*}
  (1-\epsilon)\bar\P\preceq
  \E[\P]\preceq (1+\epsilon)\bar\P,
\end{align*}
which completes the proof.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{t:main}]

Before getting into the details of the proof of Theorem~\ref{t:main}, we first provide the following intuitive derivation of Theorem~\ref{t:main}, 
\[
  \E[\X^\dagger \X] \simeq \I - \bar \P  = \I -  (\bar \gamma \Sigmab + \I)^{-1} = \bar \gamma \bar \P \Sigmab
\]
with $\bar \gamma = \frac{n}{\tr \Sigmab \bar \P}$ such that $\tr \bar \P = d - n$. Note that for $\X^\dagger \X = (\X^\top \X)^\dagger \X^\top \X$ we have, with Lemma~\ref{l:rank-one-update} that
\[
  \E[\I - \P] = \E[\X^\dagger \X] = \sum_{i=1}^n \E[(\X^\top \X)^\dagger \x_i \x_i^\top]  = n \E \left[\frac{ \P_{-n} \x_n \x_n^\top }{ \x_n^\top \P_{-n} \x_n } \right]
\]
where we denote $\x_i \in \mathbb R^d$ the $i$-th row of $\X$, $\P_{-i} = \I - \X_{-i}^\dagger \X_{-n}$, with $\X_{-i} \in \mathbb R^{(n-1) \times d}$ the matrix $\X$ without its $i$-th row. Due to the sub-Gaussianity of $\x_n$, the quadratic form $\x_n^\top \P_{-n} \x_n$ in the denominator is expected to concentrate around its expectation (with respect to $\x_n$, with $\E[\x_n \x_n^\top] = \Sigmab$) and thus
\[
  \E[\X^\dagger \X] \simeq n \E \left[\frac{ \P_{-n} \x_n \x_n^\top }{ \tr \Sigmab \P_{-n} } \right] = n\E \left[\frac{ \P_{-n} \Sigmab }{ \tr \Sigmab \P_{-n} } \right].
\]
Further note that, with $\E[\P_{-n}] \simeq \E[\P]$ for $n$ large and $\frac1n \tr \Sigmab \P_{-n} \simeq \frac1n \tr \Sigmab \E[\P]$ similarly with a concentration argument, we shall have
\[
  \E[\X^\dagger \X] = \I - \E[\P] \simeq \frac{ \E[\P] \Sigmab }{ \frac1n \tr \Sigmab \E[\P] }
\]
so that
\[
  \left( \I +  \frac{\Sigmab}{ \frac1n \tr \Sigmab \E[\P] } \right) \E[\P] \simeq \I \Rightarrow \E[\P] \simeq \left( \left( \frac1n \tr \Sigmab \E[\P] \right)^{-1} \Sigmab + \I \right)^{-1}
\]
and thus $\E[\P] \simeq \bar \P$ for $\bar \P = (\bar \gamma \Sigmab + \I)^{-1}$ and $\bar \gamma^{-1} = \frac1n \tr \Sigmab \bar \P$. This leads to the (implicit) expression of $\bar \P $ and $ \bar \gamma$.

\bigskip

To make the above intuition rigorous, we present now the proof of Theorem~\ref{t:main}. First note that:
\begin{enumerate}
  \item  $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, so we can assume without loss of generality (after rescaling $\bar{\P}$ correspondingly) that $\|\Sigmab\| = 1$.
  \item according to the definition of $\bar \P$ and $\bar \gamma$, the following bounds hold
  \begin{equation}\label{eq:bound-bar-gamma-P}
    \frac1{\bar \gamma + 1} \I \preceq \bar \P \preceq \I, \quad \bar \gamma \le \frac{n}{r-n} = \frac1{\rho - 1}
  \end{equation}
  for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}{n} > 1$, where we used the fact that
  \[
    n = d - \tr\,\bar\P = \tr\,\bar\P (\bar\gamma\Sigmab + \I) - \tr\,\bar\P
  = \tr\,\bar\P \bar\gamma\Sigmab\geq \frac{\bar\gamma}{\bar\gamma+1}\tr\,\Sigmab,
  \]
  so that $r = \tr \Sigmab \le n\frac{\bar \gamma + 1}{\bar \gamma} = n + \tr \bar \P \Sigmab$;
  \item to obtain the lower and upper bound for $\E[\I - \X^\dagger \X]$ in the sense of symmetric matrix as in Theorem~\ref{t:main}, it suffices to bound the following spectral norm
  \begin{equation}\label{eq:proof-1}
    \| \I - \E[\I - \X^\dagger \X] \bar \P^{-1} \| \le \frac{C_\rho K^2}{\sqrt r}
  \end{equation}
  so that, with $\I\succeq\bar\P\succeq \frac{\rho-1}{\rho}\I$ from \eqref{eq:bound-bar-gamma-P}, we have
  \begin{align*}
    \|\I-\bar\P^{-\frac12}\E[\P]\bar\P^{-\frac12}\| =
    \|\bar\P^{-\frac12}(\I-\E[\P]\bar\P^{-1})\bar\P^{\frac12}\|\leq
  \|\bar\P^{-\frac12}\|\cdot\frac{C_\rho K^2}{\sqrt r}\cdot \|\bar\P^{\frac12}\|\leq
  \frac{C_\rho K^2}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
  \end{align*}
  Defining $\epsilon =\frac{C_\rho K^2}{\sqrt r}
  \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
  of the p.s.d.~matrix $\bar\P^{-\frac12}\E[\P]\bar\P^{-\frac12}$ lie
  in the interval $[1-\epsilon,1+\epsilon]$, so:
  \begin{align*}
    (1-\epsilon)\I\preceq
    \bar\P^{-\frac12}\E[\P]\bar\P^{-\frac12}\preceq (1+\epsilon)\I.
  \end{align*}
  Note that for p.s.d.~matrices $\A,\B,\Z$, if $\A\preceq \B$ then
  $\Z\A\Z\preceq \Z\B\Z$, so letting $\Z=\bar\P^{\frac12}$ in the above,
  we get:
  \begin{align*}
    (1-\epsilon)\bar\P\preceq
    \E[\P]\preceq (1+\epsilon)\bar\P
  \end{align*}
  as desired.
\end{enumerate}

\bigskip

As a consequence of the above observations, we only need to prove \eqref{eq:proof-1} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:

\begin{enumerate}
  \item Recall $\P_{-i} = \I - \X_{-i}^\dagger \X_{-n}$, with $\X_{-i} \in \mathbb R^{(n-1) \times d}$ the matrix $\X$ without its $i$-th row, we define, for all $i \in \{ 1, \ldots,n \}$, the following events
  \begin{equation}\label{eq:def-events-E-F}
    E_i: \left\{ \left| \frac{\tr \P_{-i} \Sigmab}{\x_i^\top \P_{-i} \x_i} - 1 \right| \le \frac12 \right\}, \quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\} %F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
  \end{equation}
  for which we can bound, with Lemma~\ref{l:concentration-sub-gaussian}, the probability of both $\neg E_i$ and $\neg F_i$, and consequently $\neg E$ for $E = \bigwedge_{i=1}^n(E_i \wedge F_i)$;
  \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P] \bar \P^{-1} \|$. Since 
  \begin{align*}
  \I&-\E[\P]\bar\P^{-1}
  = \E[\X^\dagger\X] - \bar\gamma\E[\P]\Sigmab\\
  &= \E[\X^\dagger\X\one_E] +
    \E[\X^\dagger\X \one_{\neg E}]
    -\bar\gamma\E[\P]\Sigmab\\
  &=n\,\E\bigg[\frac{\P_{-n}\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\one_E\bigg]
    -\bar\gamma\E[\P]\Sigmab \ +\     \E[\X^\dagger\X\one_{\neg E}]\\
  &=\bar\gamma\,\underbrace{\E\bigg[(\bar s-\hat
    s)\cdot\frac{\P_{-n}\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\one_E\bigg]}_{\T_1}
    + \bar\gamma \underbrace{\E[\P_{-n}\x_n\x_n^\top\one_E] - \E[\P_{-n}]\Sigmab}_{\T_2} + \bar\gamma \underbrace{\E[\P_{-n}-\P]\Sigmab}_{\T_3} + \underbrace{\E[\X^\dagger\X\one_{\neg E}]}_{\T_4}
\end{align*}
where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_n^\top\P_{-n}\x_n$, $\bar s=n/\bar\gamma$. It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ respectively to reach the conclusion.
\end{enumerate}

Another important relation that will be constantly used throughout the proof is
\begin{equation}\label{eq:bound-trace-P-Sigma}
    \tr\,\P_{-n}\Sigmab = \tr \Sigmab^{\frac12} \P_{-n} \P_{-n} \Sigmab^{\frac12} =
    \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-n}^\dagger\X_{-n}\|_F^2\geq
\sum_{i\geq n}\lambda_i (\Sigmab)\geq r-n.
\end{equation}
where we have, under Assumption~(general position?) that $\rank(\X_{-n}) = n-1$ and arrange the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_d(\Sigmab)$ in a non-increasing order.

\bigskip

For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian} and \eqref{eq:bound-trace-P-Sigma} that
\[
  \Pr (\neg E_i) \le 2 \ee^{ - \frac{ (\tr \P_{-n} \Sigmab)^2 }{c r} } \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
\]
for some constant $c_\rho$ depending on fixed $\rho \equiv \frac{r}n > 1$. Similarly we have
\[
  \Pr (\neg F_i) \le \ee^{-c_\rho r} \le \ee^{-c_\rho (r-n)}
\]
and with the union bound we obtain
\begin{equation}\label{eq:proba-E}
  \Pr (\neg E) \le 2 n \ee^{-c_\rho (r-n)} \le \frac{n}{(r-n)^2} \cdot 2 (r-n)^2 \ee^{-c_\rho (r-n)} \le \frac{C_\rho}{r - n} \le \frac{C_\rho}{r}
\end{equation}
where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ on $x > 0$.

\medskip

Note in particular that
\begin{enumerate}
  \item conditioned on the event $E$, we have for $i \in \{1, \ldots, n \}$
  \begin{equation}
    \frac12 \frac1{\tr \P_{-i} \Sigmab} \le \frac1{\x_i^\top \P_{-i} \x_i} \le \frac32 \frac1{\tr \P_{-i} \Sigmab}, \quad  \frac1{2r} = \frac12 \frac1{\tr \Sigmab} \le \frac1{\x_i^\top \x_i} \le \frac32 \frac1{\tr \Sigmab} = \frac3{2r} %\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
  \end{equation}
  \item with \eqref{eq:proba-E}, we have $\|\E[\X^\dagger\X\one_{\neg E}]\|\leq \frac{C_\rho}{r}$ and $\|\E[\x_n\x_n^\top\one_E\mid\P_{-n}]-\Sigmab\|\leq
  \frac{C_\rho}{r}$ and thus
  \begin{align*}
  \| \T_2 \| = \big\|\E[\P_{-n}]\Sigmab -
  \E[\P_{-n}\x_n\x_n^\top\one_E]\big\|
    =\Big\|\E\Big[\P_{-n}\big(\Sigmab-\E[\x_n\x_n^\top\one_E\mid\P_{-n}]\big)\Big]\Big\|\leq \frac{C_\rho}{\sqrt{r}}.
  \end{align*}
\end{enumerate}

As a consequence, we have $\| \T_2 \| + \| \T_4 \| \le \frac{C_\rho}{r}$. And it thus remains to handle the terms $\T_1$ and $\T_3$ to prove the second step.

To bound $\T_3$, note that  $\P_{-n} - \P = \frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}$, we have
\begin{align*}
\|\T_3\|\leq  \bigg\|
  \E\bigg[\E\Big[\frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}
  \mid \P_{-n}\Big]\bigg]\bigg\|  \leq
C\,K^2\,\E\Big[\frac{\|\P_{-n}\Sigmab\|}{\tr\,\P_{-n}\Sigmab}\Big]   
\leq C\cdot \frac{K^2}{r-n} =  C K^2 \frac{r}{r-n} \frac1r =  K^2 \frac{C_\rho}{r}
\end{align*}
where we used the fact that $\frac{\tr\,\P_{-n}\Sigmab} {\|\P_{-n}\Sigmab\|}\ge \tr\,\P_{-n}\Sigmab $, $\tr\,\P_{-n}\Sigmab \ge r -n $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv \frac{r}n > 1$.

For $\T_1$ we write
\begin{align*}
  \|\T_1\|    &\leq\E\bigg[\|\P_{-n}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
    \frac{\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\one_E\mid\P_{-n}\Big]
    \Big\|\bigg]
  \\
          &\leq \frac32 \frac1{r-n} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
            |\bar s-\hat s|\cdot \v^\top\x_n\x_n^\top\v\cdot\one_E\mid
            \P_{-n}\Big]\bigg]
  \\
  &\leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
    s)^2\cdot \one_E \mid\P_{-n}\big]}}_{ T_{1,1} }\cdot
    \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_n)^4\big]}}_{ T_{1,2} }\bigg]
\end{align*}
where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} and $\| \P_{-n} \| \le 1$ for the second inequality, and Cauchy–Schwarz for the third inequality.

To bound $T_{1,2}$, by definition of sub-Gaussian random vectors, we have for $\x_n$ a $K$-sub-Gaussian and $\| \v \| = 1$ that, $\v^\top \x_n$ is a sub-Gaussian random variable with $\|\v^\top \a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute constant $C > 0$, see for example \cite[Section~2.5.2]{vershynin2018high}.

For $T_{1,1}$ we have
\[
  \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-n}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
\]
where we denote $s=\E[\hat s]=\tr\,\E[\P_{-n}]\Sigmab$. Note that
\begin{align*}
  \E\big[(s - \hat s)^2 \cdot\one_E \mid \P_{-n}\big] &=
\E\big[\big(\tr\,\Sigmab(\E[\P_{-n}]-\P_{-n})\big)^2 \cdot\one_E \big] +
  \E\big[(\tr\,\P_{-n}\Sigmab-\x_n^\top\P_{-n}\x_n)^2 \cdot\one_E \big]\\
  &\leq C\Big(n + \E\big[\tr\,(\P_{-n}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
    C\big(n + \tr\,\E[\P_{-n}]\Sigmab \cdot\one_E\big)\\
  &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
\end{align*}
where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}.

% Moreover, using the fact that
% $\bar\P\Sigmab\preceq\frac1{\bar\gamma+1}\I$, we obtain that:
% \begin{align*}
%   |\bar s - s| \cdot\one_E &= |\tr (\bar \P - \E[\P_{-n}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-n}] \Sigmab| \\
%   &= \big|\tr\,(\I-\E[\P]\bar\P^{-1})\bar\P\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top \P_{-n} \x_n} \right] \Sigmab \\
%   %&= \bar \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P\Sigmab\big| \cdot\one_E + 1 \\
%   &\leq \bar\gamma\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,\P_{-n}\x_n\x_n^\top\bar\P\Sigmab}{\tr\,\P_{-n}\x_n\x_n^\top} \cdot\one_E\bigg]
%     + \bar\gamma\,
%     \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
%     {\tr\,\P_{-n}\x_n\x_n^\top}\bigg] + 1\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(\E\left[|\bar s-\hat
%     s| \cdot \frac{\x_n^\top \x_n}{\x_n^\top \P_{-n} \x_n} \cdot\one_E \right] + 1\Big)\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar
%     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
%     s|\big] + 1\Big)\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| +
%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
% \end{align*}

Note that $\I - \P$ is symmetric and $\I-\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X (\X^\top\X)^\dagger$, we obtain
\begin{align*}
  &\I - \E[\P]\bar\P^{-1} + \I - \bar\P^{-1} \E[\P]
  = 2\E[\I -\P] - \E[\bar\gamma\P\Sigmab] - \E[\bar\gamma\Sigmab\P]\\
  &= \sum_{i=1}^n\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \bar\gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
  &=n\,\E\bigg[\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top\P_{-n}\x_n}\bigg] - \bar\gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
  &=\bar\gamma\,
    \E\bigg[\bar s\cdot\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}} {\x_n^\top\P_{-n}\x_n}\bigg]
    -\bar\gamma\,\E\bigg[\hat s
    \cdot\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top\P_{-n}\x_n}\bigg] +
    \bar\gamma\,(\E[\P_{-n}\Sigmab] + \E[\Sigmab\P_{-n}]) \\ 
  &- \bar\gamma (\E[\P]\Sigmab + \Sigmab \E[\P]) \\
  &=\bar\gamma\,\E\bigg[(\bar s - \hat s)
    \cdot\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top\P_{-n}\x_n}\bigg]
+    \bar\gamma (\E[\P_{-n}-\P]\Sigmab + \Sigmab \E[\P_{-n}-\P])
\end{align*}
%where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
Moreover, using the fact that $\bar\P\Sigmab\preceq\frac1{\bar\gamma+1}\I$ and $\bar \P \Sigmab = \Sigmab \bar \P$, we obtain that:
\begin{align*}
  |\bar s - s|   &= |\tr (\bar \P - \E[\P_{-n}] )\Sigmab|   \le |\tr (\bar \P - \E[\P] )\Sigmab|   + |\tr \E[\P - \P_{-n}] \Sigmab| \\
  &= \frac12 \big|\tr\,(\I-\E[\P]\bar\P^{-1})\bar\P\Sigmab + \tr\,\bar\P (\I - \bar \P^{-1} \E[\P]) \Sigmab \big|   + \tr\,\E \left[ \frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top \P_{-n} \x_n} \right] \Sigmab \\
  &= \frac12 \big|\tr\,(\I-\E[\P]\bar\P^{-1} + \I - \bar \P^{-1} \E[\P])\bar\P\Sigmab \big|   + 1 \\
  &\leq \frac{\bar\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
    \frac{\tr\,(\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}) \bar\P\Sigmab}{\tr\,\P_{-n}\x_n\x_n^\top}  \bigg]
    + \bar\gamma\,
    \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
    {\tr\,\P_{-n}\x_n\x_n^\top}\bigg] + 1\\
  &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(\E\left[|\bar s-\hat
    s| \cdot \frac{\x_n^\top \P_{-n} \x_n}{\x_n^\top \P_{-n} \x_n}  \right] + 1\Big) + 1\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar
    s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1 \\
  &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| +
     c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 2\Big).
\end{align*}

% $\bar\P\Sigmab\preceq\frac1{\bar\gamma+1}\I$, we obtain that:
% \begin{align*}
%   |\bar s - s| \cdot\one_E &= |\tr (\bar \P - \E[\P_{-n}] )\Sigmab| \le |\tr (\bar \P - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-n}] \Sigmab| \\
%   &= \big|\tr\,(\I-\E[\P]\bar\P^{-1})\bar\P\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top \P_{-n} \x_n} \right] \Sigmab \\
%   &\leq \bar\gamma\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,\P_{-n}\x_n\x_n^\top\bar\P\Sigmab}{\tr\,\P_{-n}\x_n\x_n^\top} \cdot\one_E\bigg]
%     + \bar\gamma\,
%     \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
%     {\tr\,\P_{-n}\x_n\x_n^\top}\bigg] + 1\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(\E\left[|\bar s-\hat
%     s| \cdot \frac{\x_n^\top \x_n}{\x_n^\top \P_{-n} \x_n} \cdot\one_E \right] + 1\Big)\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar
%     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
%     s|\big] + 1\Big)\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| +
%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
% \end{align*}

% Moreover, since
% \begin{align*}
%   |\bar s - s| \cdot\one_E
%   &= \big|\tr\,(\bar \P - \E[\P])\Sigmab \big| \cdot\one_E \\
%     %=\big|\tr\,(\T_1+\T_2)\bar\P\Sigmab\big|\\
%   &\leq \bar\gamma\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,\P_{-n}\x_n\x_n^\top\bar\P\Sigmab}{\x_n^\top \P_{-n} \x_n} \cdot\one_E\bigg]
%     + \bar\gamma\,
%     \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
%     {\x_n^\top \P_{-n} \x_n }\bigg]\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{ \x_n^\top \x_n }{\x_n^\top \P_{-n} \x_n} \cdot\one_E\bigg]
%     + \frac{\bar\gamma}{\bar\gamma+1}\,
%     \E\bigg[\frac{\tr \P_{-n} \x_n \x_n^\top \P_{-n}}
%     {\x_n^\top \P_{-n} \x_n }\bigg]\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(\E\big[|\bar s-\hat
%     s|\big] + 1\Big)\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar
%     s-s| + \E\big[|s-\hat
%     s|\big] + 1\Big)\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| +
%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
% \end{align*}
% where we used, for the first inequality, the fact that
% \[
%   \E[\P \cdot\one_E] - \bar \P = \I - \bar \P - \E[\X^\dagger \X \cdot\one_E] = \bar \gamma \Sigmab \bar \P - \frac{n}2 \E \left[ \frac{ \P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top \P_{-n} \x_n } \cdot\one_E \right]
% \]
% so that
% \[
%   \bar \gamma \Sigmab \bar \P - \frac{3n}4 \E \left[ \frac{ \P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{ \tr \P_{-n} \Sigmab } \right] \preceq \E[\P \cdot\one_E] - \bar \P \preceq \bar \gamma \Sigmab \bar \P - \frac{n}4 \E \left[ \frac{ \P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{ \tr \P_{-n} \Sigmab } \right]
% \]
% or equivalently
% \[
%   \bar \gamma \Sigmab \bar \P - \frac{3n}4 \E \left[ \frac{ \P_{-n}\Sigmab + \Sigmab \P_{-n}}{ \tr \P_{-n} \Sigmab } \right] \preceq \E[\P \cdot\one_E] - \bar \P \preceq \bar \gamma \Sigmab \bar \P - \frac{n}4 \E \left[ \frac{ \P_{-n}\Sigmab + \Sigmab \P_{-n}}{ \tr \P_{-n} \Sigmab } \right]
% \]
% \begin{align*}
%    \\
%   &\bar \gamma \E \left[ (\bar s-\hat s)\cdot \frac{ \P_{-n}\x_n\x_n^\top}{\x_n^\top \P_{-n} \x_n } \right] + \bar \gamma \E[\P_{-n} - \P] \Sigmab \\ 
%   &= \bar \gamma \E \left[ (\bar s-\hat s)\cdot \frac{ \P_{-n}\x_n\x_n^\top}{\x_n^\top \P_{-n} \x_n} \right] + \bar \gamma \E \left[ \frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top \P_{-n} \x_n} \right] \Sigmab.
% \end{align*}

% For the second inequality, we use the fact that $\max\{ \| \Sigmab \bar \P \Sigmab \|, \| \bar\P\Sigmab \P_{-n} \} \| \le \| \bar\P\Sigmab \| \le \frac1{\bar\gamma+1}$ and for symmetric and p.s.d.~matrix $\A$ and arbitrary $\B$ of appropriate dimension we have $| \tr \A \B| \le \| \B \| \cdot \tr \A$. We proceed similarly to the bound of $\T_1$ to obtain the third inequality.

% Note that $\I - \P$ is symmetric and $\I-\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X (\X^\top\X)^\dagger$, we obtain
% \begin{align*}
%   &\I - \E[\P]\bar\P^{-1} + \I - \bar\P^{-1} \E[\P]
%   = 2\E[\I -\P] - \E[\bar\gamma\P\Sigmab] - \E[\bar\gamma\Sigmab\P]\\
%   &= \sum_{i=1}^n\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \bar\gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
%   &=n\,\E\bigg[\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top\P_{-n}\x_n}\bigg] - \bar\gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
%   &=\bar\gamma\,
%     \E\bigg[\bar s\cdot\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}} {\x_n^\top\P_{-n}\x_n}\bigg]
%     -\bar\gamma\,\E\bigg[\hat s
%     \cdot\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top\P_{-n}\x_n}\bigg] +
%     \bar\gamma\,(\E[\P_{-n}\Sigmab] + \E[\Sigmab\P_{-n}]) \\ 
%   &- \bar\gamma (\E[\P]\Sigmab + \Sigmab \E[\P]) \\
%   &=\bar\gamma\,\E\bigg[(\bar s - \hat s)
%     \cdot\frac{\P_{-n}\x_n\x_n^\top + \x_n\x_n^\top \P_{-n}}{\x_n^\top\P_{-n}\x_n}\bigg]
% +    \bar\gamma (\E[\P_{-n}-\P]\Sigmab + \Sigmab \E[\P_{-n}-\P]) 
% \end{align*}

% As a consequence
% \begin{align*}
%   |\bar s - s| &=  \big|\tr\,(\bar \P - \E[\P])\Sigmab \big| = \frac12 \big|\tr\,(\I-\E[\P]\bar\P^{-1})\bar\P\Sigmab + \tr\,(\I-\bar\P^{-1}\E[\P])\Sigmab \bar\P \big| \\
%     %=\big|\tr\,(\T_1+\T_2)\bar\P\Sigmab\big|\\
%   &\leq \frac{\bar\gamma}2\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,(\P_{-n}\x_n\x_n^\top\bar\P + \bar \P \x_n\x_n^\top \P_{-n})\Sigmab}{\x_n^\top \P_{-n} \x_n} \bigg]
%     + \bar\gamma\,
%     \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
%     {\x_n^\top \P_{-n} \x_n }\bigg]\\
%   &\leq \frac{\bar\gamma}2\,\E\bigg[|\bar s-\hat s|\cdot
%     \frac{\tr\,(\P_{-n}\x_n\x_n^\top\P_{-n} + \bar \P \x_n\x_n^\top \bar \P)\Sigmab}{\x_n^\top \P_{-n} \x_n} \bigg]
%     + \bar\gamma\,
%     \E\bigg[\frac{\tr\,\P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab\bar\P\Sigmab}
%     {\x_n^\top \P_{-n} \x_n }\bigg]\\
%   &\leq \frac{\bar\gamma}2\,\E\bigg[|\bar s-\hat s|\cdot \left(
%     1 +  \right)\bigg]
%     + \frac{\bar\gamma}{\bar\gamma+1}\,
%     \E\bigg[\frac{\tr \P_{-n} \x_n \x_n^\top \P_{-n}}
%     {\x_n^\top \P_{-n} \x_n }\bigg]\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(\E\big[|\bar s-\hat
%     s|\big] + 1\Big)\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar
%     s-s| + \E\big[|s-\hat
%     s|\big] + 1\Big)\\
%   &\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| +
%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
% \end{align*}
% where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).

% where we used, for the first inequality, the fact that
% \begin{align*}
%   \I - \E[\P] \bar \P^{-1} &= \E[\X^\dagger \X] - \bar \gamma \E[\P] \Sigmab  = \bar \gamma \E \left[ (\bar s-\hat s)\cdot \frac{ \P_{-n}\x_n\x_n^\top}{\x_n^\top \P_{-n} \x_n } \right] + \bar \gamma \E[\P_{-n} - \P] \Sigmab \\ 
%   &= \bar \gamma \E \left[ (\bar s-\hat s)\cdot \frac{ \P_{-n}\x_n\x_n^\top}{\x_n^\top \P_{-n} \x_n} \right] + \bar \gamma \E \left[ \frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top \P_{-n} \x_n} \right] \Sigmab.
% \end{align*}
% For the second inequality, we use the fact that $\max\{ \| \Sigmab \bar \P \Sigmab \|, \| \bar\P\Sigmab \P_{-n} \} \| \le \| \bar\P\Sigmab \| \le \frac1{\bar\gamma+1}$. For the third inequality we used \eqref{eq:bound-trace-P-Sigma} and recall $\rho = \frac{r}n$.
%and for symmetric and p.s.d.~matrix $\A$ and arbitrary $\B$ of appropriate dimension we have $| \tr \A \B| \le \| \B \| \cdot \tr \A$. We proceed similarly to the bound of $\T_1$ to obtain the third inequality.

Solving for $|\bar s-s|$, we conclude that
\begin{align*}
  |\bar s-s|\leq
  \bar\gamma(c\sqrt{n+\bar s}+2) + c^2\bar\gamma^2\leq \bar\gamma
  (c+1)\sqrt{n+\bar s} + c^2\bar\gamma^2.
\end{align*}
Since $\bar s = \tr \bar \P \Sigmab \le \tr \Sigmab =  r $, this allows us to conclude that $T_{1,1} \le \frac{C_\rho}{\sqrt r} $ and thus the conclusion.
\end{proof}

\subsection{Proof of Lemma \ref{l:rank-one-update}}
Recall $\P = \I - \X^\dagger \X = \I - (\X^\top \X)^\dagger \X^\top \X = \I - (\X^\top \X)^\dagger (\A + \x_n \x_n^\top)$ and $\P_{-n} = \I - \X_{-n}^\dagger \X_{-n} = \I - \A^\dagger \A$ for $\A = \X_{-n}^\top \X_{-n}$. We have, from \cite[Theorem 1]{10.2307/2099767} that
\begin{align*}
  (\X^\top \X)^\dag
  &= (\A + \x_n \x_n^\top)^\dag
  = \A^\dag - k u^\dag - v^\dag h + \beta v^\dag u^\dag
\end{align*}
where $k = \A^\dag \x_n$,
$h = \x_n^\top \A^\dag$,
$u = \P_{-n} \x_n$,
$v = \x_n^\top \P_{-n}$,
and $\beta = 1 + \x_n^\top \A^\dag \x_n$.
Noting that column vectors $u$ have
pseudo-inverse $u^\dag = \frac{u^\top}{\|u\|^2}$,
we have
\begin{align*}
  (\X^\top \X)^\dag
  &= \A^\dag
  - \frac{\A^\dag \x_n \x_n^\top \P_{-n}}{\x_n^\top \P_{-n} \x_n}
  - \frac{\P_{-n} \x_n \x_n^\top \A^\dag}{\x_n^\top \P_{-n} \x_n}
  + (1 + \x_n^\top \A^\dag \x_n) \frac{\P_{-n} \x_n \x_n \P_{-n}}{(\x_n^\top \P_{-n} \x_n)^2}
\end{align*}
where we have used
$\|\P_{-n} \x_n\|^2 = \x_n^\top \P_{-n}^\top \P_{-n} \x_n = \x_n^\top \P_{-n} \x_n$
because $\P_{-n}$ is a projection matrix hence idempotent.
As a consequence, we have 
\[
  \P_{-n} - \P = \X^\dagger \X - \X_{-n}^\dagger \X_{-n} = \frac{\P_{-n} \x_n \x_n^\top \P_{-n}}{\x_n^\top \P_{-n} \x_n}
\]
and by left multiplying by $\x_n$ and simplifying we get
\begin{align*}
  (\X^\top \X)^\dag \x_n
  &= - \frac{\P_{-n} \x_n \x_n^\top \A^\dag \x_n}{\x_n^\top \P_{-n} \x_n}
  + (1 + \x_n^\top \A^\dag \x_n) \frac{\P_{-n} \x_n}{\x_n^\top \P_{-n} \x_n}
  = \frac{\P_{-n} \x_n}{\x_n^\top \P_{-n} \x_n}.
\end{align*}

\subsection{Proof of Lemma~\ref{l:concentration-sub-gaussian}}

% Let $\x=\A^{\frac12}\z$. For the first inequality
% \begin{align*}
%   \left\|\E\left[
%     \frac{\x \x^\top}{\x^\top \x}
%   \right]\right\|
%   &= \sup_{\v : \|\v\| = 1} \E\left[
%     \frac{\v \x \x^\top \v}{\x^\top \x}
%   \right]
%   \leq \sup_{\v : \|\v\| = 1} \sqrt{
%     \E\left[
%       \left(\frac{\v \x \x^\top \v}{\x^\top \x}\right)^2
%     \right]
%   }
% \end{align*}

% Next, fix any $\v$ and let $E$ be the event that
% $\x^\top\x\geq \frac12\tr\,\A$.
% $E$ has small probability because $\z$ is $K$-sub-gaussian, so
% \begin{align*}
%   \hspace{-3mm}\E\bigg[\Big(\v^\top\frac{\x\x^\top}
%   {\x^\top\x}\v\Big)^2\bigg]
%       &\leq
%       \Big(\frac2{\tr\,\A}\Big)^2\E\big[(\v^\top\x)^4\mid
%       E\big]\Pr(E) +
%       \E\bigg[\Big(\v^\top\frac{\x\x^\top}
%         {\x^\top\x}\v\Big)^2 \mid
%       E\bigg]\Pr(\neg E)\\
%       &\leq
%       \Big(\frac2{\tr\,\A}\Big)^2\E\big[(\z^\top\A^{\frac12}\v\v^\top\A^{\frac12}\z)^2\big]
%       + 2\ee^{-c  r / K^4} \|\v\|^4\\
%       &\overset{(*)}{\leq}
%       \Big(\frac{2}{\tr\,\A}\Big)^2 \left(
%         C K^4 \tr\,(\A^{\frac12}\v\v^\top\A^{\frac12})^2
%         + (\tr \A^{\frac12}\v\v^\top\A^{\frac12} \E[\z \z^\top])^2
%       \right)
%       + 2\ee^{-c  r / K^4} \|\v\|^4 \\
%       &\leq
%       C \Big(\frac{\|\A\|}{\tr\,\A}\Big)^2 \left(
%         K^4 \tr (\v \v^\top)^2
%         + K^4 \tr (\v\v)^2
%       \right)
%       + 2\ee^{-c  r / K^4} \|\v\|^4 \\
%       &
%       = O(K^4/r^2)\cdot \|\v\|^4,
% \end{align*}
% where $(*)$ uses \Cref{l:concentration-sub-gaussian},
% the next line bounds 
% $\E[\z \z^\top] \preceq C K^2 \I$ for some universal constant $C$
% (\cite{pollard2015few} Theorem 16)
% \feynman{Here is where we could assume $\E[\z \z^\top] = \I$, but I think
% sub-Gaussian lets you bound the diagonal by $K^2 \I$}.

%\begin{proof}
  From \cite{zajkowski2018bounds} Corollary 2.6, we have the tail bound
  \begin{align*}
    \Pr\left[
      \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert
      \geq t
      \right] \leq 2 \exp\left(
      -\min\left\{
        \frac{t^2}{16 C^2  K^4 (\tr \A)^2},
        \frac{t}{4 C K^2 (\tr \A)},
      \right\}
    \right)
  \end{align*}
  Integrating this bound yields
  \begin{align*}
      \E\left[ \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert^2 \right]
      &= \int_0^\infty \Pr\left[
        \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert
        \geq \sqrt{t}
      \right]dt \\
      &\leq 2 \int_0^\infty \exp\left(
        -\min\left\{
          \frac{t}{C^2  (\tr \A)^2 K^4},
          \frac{\sqrt{t}}{C  (\tr \A) K^2},
        \right\}
      \right)dt \\
      &\leq 2 \int_0^\infty \left(\exp\left(
        - \frac{t}{C^2  (\tr \A)^2 K^4},
        \right) + \exp\left(
        \frac{\sqrt{t}}{C  (\tr \A) K^2},
      \right)\right)dt \\
      &\leq C~K^4~\tr(\A)^2.
  \end{align*}
%\end{proof}

\subsection{Proof of Lemma \ref{l:trace}}

\begin{proof}
To simplify notations, we work on $\P$ instead of $\P_{-n}$, the same line of argument applies to $\P_{-n}$ by changing the sample size $n$ to $n-1$.

First note that
\begin{align*}
  X
  &= \tr \Sigmab (\P - \E \P)
  = \E_n [\tr \Sigmab \P]  - \E_0[ \tr \Sigmab \P ]\\
  %%%
  &= \sum_{i=1}^n \left( \E_i[\tr \Sigmab \P] - \E_{i-1}[\tr \Sigmab \P]\right)
  = \sum_{i=1}^n (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})
\end{align*}
where we used the fact that $\E_i [\tr \Sigmab \P_{-i}] = \E_{i-1} [\tr \Sigmab \P_{-i}]$, for $\E_i[\cdot]$ the conditional expectation with respect to $\mathcal F_i$ the $\sigma$-field generating the rows $\x_1 \ldots, \x_i$ of $\X$
This forms a martingale difference sequence (it is a difference sequence of
the Doob martingale for $\tr \Sigmab (\P - \P_{-i})$ with respect to filtration $\mathcal{F}_i$)
hence it falls within the scope of the Burkholder inequality \cite{burkholder1973distribution}, recalled as follows.

\begin{lemma}\label{l:burkholder}
For $\{ x_i \}_{i=1}^n$ a real martingale difference sequence with respect to the increasing $\sigma$ field $\mathcal F_i$, we have, for $k > 1$, there exists $C_k > 0$ such that
\[
  \E \left[
    \left| \sum_{i=1}^n x_i \right|^k
  \right]
  \le C_k \E \left[
    \left( \sum_{i=1}^n |x_i|^2 \right)^{k/2}
  \right]
\]
\end{lemma}

Since $\P_{-i} - \P = \frac{\P_{-i} \x_i \x_i^\top \P_{-i} }{ \x_i \P_{-i} \x_i}$ that is positive semi-definite, we have $|\tr \Sigmab (\P_{-i} - \P)| \le \| \Sigmab \| = 1$ so that with Lemma~\ref{l:burkholder} we obtain with $x_i = (\E_i - \E_{i-1}) \tr \Sigmab (\P_{-i} - \P)$ that, for $k > 1$
\[
  \E |X|^k \le C_k n^{k/2}.
\]
In particular, for $k=2$, we have $\E |X|^2 \le c n$ for $c = (36 \sqrt{2})^2$
(details in proof of Theorem~3.2 in \cite{burkholder1973distribution}).

For the second result, since we have almost surely bounded martingale differences
($\lvert x_i \rvert \leq 2$), by the Azuma-Hoeffding inequality
\begin{align*}
  \Pr\{
    \lvert X \rvert
    \geq t
  \}
  \leq 2 \ee^{\frac{-t^2}{8 n}}
\end{align*}
as desired.

\end{proof}

\section{Technical results}

\subsection{Limiting behavior}

\subsubsection{Proof of Conjecture~1}

We consider the limiting behavior of $\tr (\X^\top \X)^\dagger = \tr (\X \X^\top)^\dagger$ as $n,d \to \infty$ with $d/n \to c \in (1,\infty)$, for $\X = \Z \Sigmab^{\frac12}$ with $\Z \in \mathbb R^{n \times d}$ having i.i.d.~zero mean, unit variance sub-Gaussian entries, i.e., the behavior of
\begin{equation}\label{eq:limit1}
  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left( \frac1n \X \X^\top + z \I_n \right)^{-1}
\end{equation}
by definition of the pseudoinverse.

The proof comes in three steps: 1) for fixed $z > 0$, consider the limiting behavior of $ \delta(z) \equiv \tr (\X \X^\top/n + z \I_n)^{-1}/n$ as $n,d \to \infty$ and state
\begin{equation}\label{eq:limit2}
  \lim_{n,d \to \infty} \delta(z)  \to m(z)
\end{equation}
almost surely; 2) show that both $\delta(z)$ and its derivate $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take the $z \to 0^+$ limit in \eqref{eq:limit2} and state
\begin{equation}\label{eq:limit3}
  \lim_{z \to 0^+} \lim_{n,d \to \infty} \delta(z)  \to \lim_{z \to 0^+} m(z)
\end{equation}
almost surely, given that the limit $\lim_{z \to 0^+} m(z) \equiv m(0)$ exists and eventually 3) exchange the two limits in \eqref{eq:limit3} with Moore-Osgood theorem, to reach
\[
  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left(\frac1n \X \X^\top + z \I_n \right)^{-1} \to m(0).
\]

Step 1) follows from \cite{silverstein1995empirical} that, we have, for $z > 0$ that
\[
  \delta(z) \equiv \frac1n \tr \left( \frac1n \X \X^\top  + z \I_n \right)^{-1}  \to m(z)
\]
almost surely as $n,d \to \infty$, for $m(z)$ the positive unique solution to 
\begin{equation}\label{eq:def-m}
  m(z) = \left( z + \frac1n \tr \Sigmab (\I_d + m(z) \Sigmab)^{-1} \right)^{-1}
\end{equation}
that admits a limit as $n,d \to \infty$ if we assume the weak convergence of the spectral measure of $\Sigmab$ as $d \to \infty$.

For the above step 2), \textbf{we assume that, there exists $\epsilon > 0$ such that $\lambda_{\min}( \Sigmab) > \epsilon$ for all $d,n$ large, i.e., as $d,n \to \infty$ the matrix $\Sigmab$ is not ``becoming'' more and more singular to avoid the invertibility issue.}

Since $\X = \Z \Sigmab^{\frac12}$, we have as $n, d \to \infty$
\[
  \lambda_{\min} (\X \X^\top/n) \ge \lambda_{\min} (\Z \Z^\top/n) \lambda_{\min} (\Sigmab) \ge \frac{\epsilon}2 (\sqrt c - 1 )^2
\]
almost surely, where we used and Bai-Yin theorem \cite{bai1993limit}, which states that the minimum eigenvalue of $\Z \Z^\top/n$ is almost surely larger that $( \sqrt c - 1 )^2/2$ for $n,d$ sufficiently large.

** This assumption seems necessary and is also demanded in (Hestie) **.

** We should compare the above assumption to the (General position) Assumption.**

Observe that
\[
  |\delta(z)| = \left| \frac1n \tr \left( \frac1n \X \X^\top  + z \I_n \right)^{-1} \right| \le \frac1{ \lambda_{\min} (\X \X^\top/n) }
\]
and similarly for its derivative, so that we are allowed to take the $z \to 0^+$ limit. Note that the existence of the $\lim_{z \to 0^+} m(z)$ for $m(z)$ defined in \eqref{eq:def-m} is well known, see for example \cite{ledoit2011eigenvectors}. Then, by Moore-Osgood theorem we finish step 3) and conclude that
\[
  \tr (\X^\top \X)^\dagger \to m(0)
\]
for $m(0)$ the unique positive solution to
\[
  m(0) = \left( \frac1n \tr \Sigmab (\I_d + m(0) \Sigmab)^{-1} \right)^{-1}.
\]

\subsubsection{Proof of Conjecture~2}

Since $\X^\dagger \X = \X^\top (\X \X^\top)^\dagger \X$, to prove Conjecture~2, we are interested in the limiting behavior of the following quadratic form 
\[
  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w
\]
for deterministic $\w \in \mathbb R^{d}$ of bounded Euclidean norm (i.e., $\| \w \| \le C$ as $n,d \to \infty$), as $n,d \to \infty$ with $d/n \to c \in (1,\infty)$. The limiting behavior of the above quadratic form, or more generally, bilinear form of the type $\frac1n \w_1^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w_2$ for $\w_1, \w_2 \in \mathbb R^{d}$ of bounded Euclidean norm are widely studied in random matrix literature, see for example \cite{hachem2013bilinear}.

For the proof of Conjecture~2 we follow the same protocol as that of Conjecture~1, namely: 1) we consider, for fixed $z > 0$, the limiting behavior of $\frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w$. Note that
\begin{align*}
  \delta(z) &\equiv \frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w = \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \frac1n \X^\top \X \w \\ 
  &= \| \w \|^2 - z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w
\end{align*}
and it remains to work on the second $z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w$ term.
It follows from \cite{hachem2013bilinear} that 
\[
   \delta(z) = \| \w \|^2 - z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w  \to \| \w \|^2 - \w^\top (\I_d + m(z) \Sigmab)^{-1} \w^\top 
\]
almost surely as $n,d \to \infty$, where we recall $m(z)$ is the unique solution to \eqref{eq:def-m} and implicitly assume that the right-hand side admits a limit as $n,d \to \infty$.

We move on to step 2), under the assumption that $C_{\min} \le \lambda_{\min} (\Sigmab) \le C_{\max}$ and $\| \w \| \le C$, we have
\[
  \lambda_{\max} \left( \frac1n \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \right) \le \frac{ \lambda_{\max} (\X \X^\top/n) }{ \lambda_{\min} (\X \X^\top/n) + z } \le \frac{ \lambda_{\max} (\Z \Z^\top) \lambda_{\max} (\Sigmab) }{ \lambda_{\min} (\Z \Z^\top) \lambda_{\min} (\Sigmab) } \le 4 \frac{ (\sqrt c + 1)^2 C_{\max} }{ (\sqrt c -1)^2 C_{\min} }
\]
so that $\delta(z)$ remains bounded and similarly for its derivative $\delta'(z)$, which, by Arzela-Ascoli theorem, yields uniform convergence and we are allowed to take the $z \to 0^+$ limit. Ultimately, in step 3) we exchange the two limits with Moore-Osgood theorem, and thus the conclusion.

\begin{corollary}\label{coro:projection}
Let $\X = \Z\Sigmab^{\frac12}$, where $\Z$ is an $n\times d$ random
matrix with
i.i.d.~sub-Gaussian entries and $\Sigmab$ is a $d\times d$
positive definite matrix. Assume that $n<d$ and define:
\begin{align*}
\bar\P = (\bar\gamma\Sigmab + \I)^{-1},\quad\text{such that}\quad\tr\,\bar\P=d-n.
\end{align*}
Let $r=\frac{\tr(\Sigmab)}{\|\Sigmab\|}$ be the nuclear rank of
$\Sigmab$ and fix $\rho=\frac rn > 1$. There is $C_{\rho}$ such that if $r\geq C_\rho$, then
\begin{align}
  \| \E[\X^\dagger\X] (\I - \bar \P)^{-1} - \I \| \le \frac {C_\rho}{\sqrt r}.
  % \big\|\I - \E[\I-\X^\dagger\X]\bar\P^{-1} \big\| \leq C\cdot
  % \frac{\sqrt n}r.
\end{align}
\end{corollary}

\begin{proof}[Proof of Corollary~\ref{coro:projection}]
The proof of Corollary~\ref{coro:projection} follows the same line of arguments as that of Theorem~\ref{t:main}, with some tiny modifications. It is recorded here mainly for self-consistency.

First note that
\[
  \E[\X^\dagger\X] (\I - \bar \P)^{-1} - \I = (\I - \E[\P] \bar \P^{-1}) \bar \gamma^{-1} \Sigmab^{-1}
\]
\[
  (\I - \bar \P)^{-\frac12} = \bar \gamma^{-\frac12} \bar \P^{-\frac12} \Sigmab^{-\frac12} = \bar \gamma^{-\frac12} \Sigmab^{-\frac12} \bar \P^{-\frac12}.
\]
so that
\[
  (\I - \bar \P)^{-\frac12} \E[\X^\dagger\X] (\I - \bar \P)^{-\frac12} - \I = \bar \gamma^{-1} \Sigmab^{-\frac12} (\I - \P) 
\]
so that a naive application of Theorem~\ref{t:main} leads to an error bound that involves the minimum eigenvalue of $\Sigmab$. To avoid this issue, we define, similar to the proof of Theorem~\ref{t:main}, $\hat s = \x_n^\top\P_{-n}\x_n$ and $\bar s=\tr\bar\P\Sigmab=n/\bar\gamma$, and use Lemma~\ref{l:rank-one-update} and the fact that $\I-\P=(\X^\top\X)^\dagger\X^\top\X$ to obtain
%%% original
\begin{align*}
  &(\I - \E[\P] \bar \P^{-1}) \bar \gamma^{-1} \Sigmab^{-1} = \E[\I -\P] \bar \gamma^{-1} \Sigmab^{-1} - \E[\P]\\
  &= \sum_{i=1}^n\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top\big] \bar \gamma^{-1} \Sigmab^{-1} -
    \E[\P]\\
  &=n\,\E\bigg[\frac{\P_{-n}\x_n\x_n^\top}{\x_n^\top\P_{-n}\x_n}\bigg] \bar \gamma^{-1} \Sigmab^{-1} -
    \E[\P]\\
  &=
    \E\bigg[\bar s\cdot\frac{\P_{-n}\x_n\x_n^\top }{\x_n^\top\P_{-n}\x_n}\bigg]\Sigmab^{-1} 
    - \E\bigg[\hat s
    \cdot\frac{ \P_{-n}\x_n\x_n^\top }{\x_n^\top\P_{-n}\x_n}\bigg] \Sigmab^{-1}  +
    \E[\P_{-n}]
    -\E[\P]\\
  &=\underbrace{ \E\bigg[(\bar s - \hat s)
    \cdot\frac{ \P_{-n}\x_n\x_n^\top }{\x_n^\top\P_{-n}\x_n}\bigg] \Sigmab^{-1}}_{\T_1}
+    \underbrace{\E[\P_{-n}-\P]}_{\T_2}.
\end{align*}
%%%
% where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which follows from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).

% We bound $\T_1$ as follows:
% \begin{align*}
%   \|\T_1\|
%   &\leq \E\Bigg[\bigg\|\E\Big[(\bar s-\hat
%   s)\cdot\frac{ \Sigmab^{-\frac12} \P_{-n}\x_n\x_n^\top \Sigmab^{-\frac12} }{\x_n^\top\P_{-n}\x_n}\mid
%   \P_{-n}\Big]\bigg\|\Bigg]\\
%   &\leq \E\Bigg[\bigg\|\E\Big[(\bar s-\hat
%   s)\cdot\frac{ \Sigmab^{-\frac12} \P_{-n}\x_n\x_n^\top \P_{-n} \Sigmab^{-\frac12} }{\x_n^\top\P_{-n}\x_n}\mid
%   \P_{-n}\Big] \bigg\| + \bigg\|\E\Big[(\bar s-\hat
%   s)\cdot\frac{ \Sigmab^{-\frac12} \x_n \x_n^\top  \Sigmab^{-\frac12} }{\x_n^\top\P_{-n}\x_n}\mid
%   \P_{-n}\Big] \bigg\| \Bigg]\\
%   &\leq \E\Bigg[\sqrt{\E\big[(\bar s - \hat
%   s)^2\mid
%     \P_{-n}\big]}\
%     \cdot \sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\Sigmab^{-\frac12} \P_{-n}\x_n\x_n^\top\P_{-n}\Sigmab^{-\frac12} }{\x_n^\top\P_{-n}\x_n}\v\Big)^2\mid\P_{-n}\bigg]}\,\Bigg]\\
%   &+\E\Bigg[\sqrt{\E\big[(\bar s - \hat
%   s)^2\mid
%     \P_{-n}\big]}\
%     \cdot \sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\Sigmab^{-\frac12}\x_n\x_n^\top\Sigmab^{-\frac12}}{\x_n^\top\P_{-n}\x_n}\v\Big)^2\mid\P_{-n}\bigg]}\,\Bigg]\\
%   &\leq  C\,\E\Bigg[\sqrt{\E\big[(\bar s - \hat
%   s)^2\mid
%     \P_{-n}\big]}\,
%     \cdot \frac{ 2 }{\tr\,\P_{-n}\Sigmab - 1}\Bigg]
% % \ \leq\ C\,\sqrt{\E\big[(\bar s - \hat s)^2\big]}\,\cdot
% % \underbrace{\sqrt{\E\bigg[\frac{\|\P_{-n}\Sigmab\|^2}{(\tr\,\P_{-n}\Sigmab)^2}\bigg]}}_{\Delta}\\[-4mm]
% \  \leq\ C\sqrt{(\bar s - s)^2 + \E\big[(s - \hat s)^2\big]}\cdot \frac2{r-n},
% \end{align*}

We bound $\T_1$ as follows:
\begin{align*}
  \|\T_1\|
  &\leq \E\Bigg[\E\bigg\|(\bar s-\hat
  s)\cdot\frac{\P_{-n}\x_n\x_n^\top \Sigmab^{-1} }{\x_n^\top\P_{-n}\x_n}\mid
  \P_{-n}\bigg\|\Bigg]\\
  &\leq \E\Bigg[\sqrt{\E\big[(\bar s - \hat
  s)^2\mid
    \P_{-n}\big]}\
    \cdot \sqrt{\E \left[\frac{\| \Sigmab^{-\frac12} \x_n \x_n^\top \Sigmab^{-\frac12} \|^2}{ (\x_n^\top \P_{-n} \x_n)^2 }\mid \P_{-n}\right]} \Bigg]\\
  &\leq  C\,K^2\,\E\Bigg[\sqrt{\E\big[(\bar s - \hat
  s)^2\mid
    \P_{-n}\big]}\,
    \cdot \frac1{\tr \Sigmab \P_{-n} } \Bigg] %\frac{\|\P_{-n}\Sigmab\|}{\tr\,\P_{-n}\Sigmab}
% \ \leq\ C\,\sqrt{\E\big[(\bar s - \hat s)^2\big]}\,\cdot
% \underbrace{\sqrt{\E\bigg[\frac{\|\P_{-n}\Sigmab\|^2}{(\tr\,\P_{-n}\Sigmab)^2}\bigg]}}_{\Delta}\\[-4mm]
    \  \leq\ C\sqrt{(\bar s - s)^2 + \E\big[(s - \hat s)^2\big]}\cdot \frac{K^2}{r-n},
\end{align*}
where we used the fact that $\| \P_{-n} \| \le 1$ (\textbf{here we need a bound of the type $\E \frac{\z \z^\top}{\z^\top \A \z} \preceq \frac{\I}{\tr \A}$}, which is so far only good for Gaussian) and let $s=\E[\hat s]=\tr\,\E[\P_{-n}]\Sigmab$ and the last
inequality follows because:
\begin{align*}
    \tr\,\P_{-n}\Sigmab =
    \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-n}^\dagger\X_{-n}\|_F^2\geq
\sum_{i\geq n}\lambda_i\geq r-n,
\end{align*}
with $1=\lambda_1\geq\lambda_2\geq...$ denoting the eigenvalues of $\Sigmab$. It follows from the proof of Theorem~\ref{t:main} that
\begin{align*}
  \E\big[(s - \hat s)^2\big] &\le C(n+s)\leq C\big(n+\bar s + |s- \bar s|\big) \\
  |\bar s - s|&\leq \frac{\bar\gamma}{\bar\gamma+1}\Big(|\bar s- s| + c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big)
\end{align*}
and thus again
\begin{align*}
  |\bar s-s|\leq
  \bar\gamma(c\sqrt{n+\bar s}+1) + c^2\bar\gamma^2\leq \bar\gamma
  (c+1)\sqrt{n+\bar s} + c^2\bar\gamma^2
\end{align*}
always hold.

We next treat $\T_2$, the bound for which remains the same as
\begin{align*}
\|\T_2\|\leq  \bigg\|
  \E\bigg[\E\Big[\frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}
  \mid \P_{-n}\Big]\bigg]\bigg\|  \leq C\,\E\Big[\frac{\|\P_{-n}\Sigmab\|}{\tr\,\P_{-n}\Sigmab}\Big]   
\leq C\cdot \frac1{r-n}.
\end{align*}
where we used $\E[\P_{-n}-\P] = \E\big[\frac{\P_{-n}\x_n\x_n^\top\P_{-n}}{\x_n^\top\P_{-n}\x_n}\big]$.

Recall from the proof of Theorem~\ref{t:main} that $r \le n + \frac{n}{\bar \gamma}$, $r\geq\bar s \geq r-n$ and $\bar\gamma\leq
\frac{n}{r-n}=\frac1{\rho-1}$. This allows us to conclude that
\[
  \| \E[\X^\dagger\X] (\I - \bar \P)^{-1} - \I \| \le \frac{C_\rho}{\sqrt r}.
\]
\end{proof}

\section{Euclidean norm case}

For given hypothesis class $f_\w$ of functions parameterized by $\w$ and training set $\{ \x_i, y_i \}_{i=1}^n$, consider the set of $\w$ that interpolates the training set, i.e., $f(\x_i) = y_i$ for $i \in \{0, \ldots, n\}$ and choose the following particular $\w_*$ from the set such that some capacity metric $g(\w)$ is minimized as
\[
  \w_* \equiv \argmin_\w g(\w), \quad \textmd{such that $f(\x_i) = y_i$ for all $i$}.
\]

Consider also the following (population level, for $(\x, y) \sim \mu$) minimizer
\[
  \w_\mu \equiv \argmin_\w \left\{ \E_{(\x,y) \sim \mu} \ell(f_\w(\x), y) + \lambda g(\w) \right\}
\]
we have the following theorem.

\begin{theorem}\label{t:main-Eulidean}
Let $f_\w(\x)=\x^\top\w$ and $(\x,y)\sim\mu$
  be s.t.~$\x\sim\Nc(0,\Sigmab)$ and $y=\x^\top \bar \w+\xi$,
  with $\E[\xi]=0$. Assume that $\Sigmab$ has nuclear rank
  $r>2n$. Then, for $g(\w)=\|\w\|_2^2$
  and $\ell(\hat y, y)=(\hat y-y)^2$, we have:
  \begin{align*}
    \E [\w_*]
   \ \overset{\epsilon}{\simeq}\ \w_\mu
\quad    \text{with}\quad\epsilon = O\big(\tfrac{\sqrt n}r\big),
  \end{align*}
  where $\w'\overset\epsilon\simeq\w$ iff
  $\|\w'-\w\|_2\leq \epsilon\,\|\w\|_2$, and we define $\lambda$ so that $n=\tr\,\Sigmab(\Sigmab+\lambda\I)^{-1}$.
\end{theorem}
\begin{proof}
In this case, we have $\w_*$ the minimum norm solution explicitly given by $\w_* = \X^\dagger \y$ so that
\[
  \E [\w_*]= \E[\X^\dagger \y] = \E[\X^\dagger \X] \bar \w
\]
as well as
\[
  \w_\mu = \argmin_\w \left\{ \E_{(\x,y) \sim \mu} (y - \w^\top \x)^2 + \lambda \| \w \|^2 \right\} = (\Sigmab + \lambda \I)^{-1} \Sigmab \bar \w.
\]
It then follows from Theorem~\ref{t:projection} that, for $\lambda$ the solution to $n = \tr \Sigmab (\Sigmab + \lambda \I)^{-1}$ we have $ (\Sigmab + \lambda \I)^{-1} \Sigmab = \I - \bar \P $ with
\[
  \bar \P = (\bar \gamma \Sigmab + \I)^{-1}, \quad \lambda = \bar \gamma^{-1}
\]
so that
\begin{align*}
  \E[\X^\dagger \X] \bar \w - (\Sigmab + \lambda \I)^{-1} \Sigmab \bar \w &= \left( \E[\X^\dagger \X] (\I - \bar \P)^{-1} - \I  \right)  \cdot (\Sigmab + \lambda \I)^{-1} \Sigmab \bar \w \\
  %%%
  &= \left( \E[\X^\dagger \X] (\I - \bar \P)^{-1} - \I  \right) \cdot \w_\mu.
\end{align*}
\end{proof}

\section{General Euclidean norm case}

We consider now the $\w$ that minimizes the following general Euclidean norm metric
\[
  g(\w) = \w^\top \A \w
\]
for a given positive definite $\A \in \mathbb R^{d \times d}$. For $f_\w(\x) = \w^\top \x$, this is equivalent to find $\tilde \w \equiv \A^{\frac12} \w$ such that $f_{\tilde \w} (\x) = \tilde \w^\top \A^{-\frac12} \x$ with $g(\tilde \w) = \| \tilde \w \|^2$ minimized, or, to find $\tilde \w$ such that $f_{\tilde \w} (\tilde \x) = \tilde \w^\top \tilde \x$ for $\tilde \x \equiv \A^{-\frac12} \x \sim \Nc (\zero, \A^{-\frac12} \Sigmab \A^{-\frac12} )$ with minimal Euclidean norm $\| \tilde \w \|^2$. The desired solution can then be retrieved as $\w = \A^{-\frac12} \tilde \w$. This is given in the following theorem.

\begin{theorem}\label{t:main-Mahalanobis}
Let $f_\w(\x)=\x^\top\w$ and $(\x,y)\sim\mu$
  be s.t.~$\x\sim\Nc(0,\Sigmab)$ and $y=\x^\top \bar \w+\xi$,
  with $\E[\xi]=0$. Then, for $g(\w)=\|\w^\top \A \w \|_2^2$
  and $\ell(\hat y, y)=(\hat y-y)^2$, assume $\Sigmab \A^{-1}$ has nuclear rank
  $r>2n$, we have:
  \begin{align*}
    \E [\w_*]
   \ \overset{\epsilon}{\simeq}\ \w_\mu
\quad    \text{with}\quad\epsilon = O\big(\tfrac{\sqrt n}r\big),
  \end{align*}
  where $\w'\overset\epsilon\simeq\w$ iff
  $\|\w'-\w\|_\A \leq \epsilon\,\|\w\|_\A$ for $\| \w \|_\A = \| \A^{\frac12} \w \| $ and we define $\lambda$ so that $n=\tr\,\Sigmab(\Sigmab+\lambda\A)^{-1}$.
\end{theorem}
\begin{proof}
Let us first focus on $f_{\tilde \w} (\tilde \x) = \tilde \w^\top \tilde \x$ for $\tilde \x \equiv \A^{-\frac12} \x \sim \Nc (\zero, \A^{-\frac12} \Sigmab \A^{-\frac12} )$ with minimal Euclidean norm $\| \tilde \w \|^2$. In this case, we have $\tilde \w_*$ the minimum norm solution explicitly given by $\tilde \w_* = \tilde \X^\dagger \y$ with $y = \tilde \x^\top \A^{\frac12} \bar \w + \xi$, so that
\[
  \E [\tilde \w_*]= \E[\tilde \X^\dagger \y] = \E[\tilde \X^\dagger \tilde \X] \A^{\frac12} \bar \w
\]
as well as
\[
  \tilde \w_\mu = \argmin_{\tilde \w} \left\{ \E_{(\x,y) \sim \mu} (y - \tilde \w^\top \A^{-\frac12} \x)^2 + \lambda \| \tilde \w \|^2 \right\} = (\A^{-\frac12} \Sigmab \A^{-\frac12} + \lambda \I)^{-1} \A^{-\frac12} \Sigmab \A^{-\frac12} \cdot \A^{\frac12} \bar \w.
\]

Denote the shortcut $\tilde \Sigmab \equiv \A^{-\frac12} \Sigmab \A^{-\frac12}$, it then follows from Theorem~\ref{t:projection} that, for $\lambda$ the solution to $n = \tr \tilde \Sigmab (\tilde \Sigmab + \lambda \I)^{-1}$ we have $ (\tilde \Sigmab + \lambda \I)^{-1} \tilde \Sigmab = \I - \bar \P $ with
\[
  \bar \P = (\bar \gamma \tilde \Sigmab + \I)^{-1}, \quad \lambda = \bar \gamma^{-1}
\]
so that
\begin{align*}
  \E[\tilde \X^\dagger \tilde \X] \A^{\frac12} \bar \w - (\tilde \Sigmab + \lambda \I)^{-1} \tilde \Sigmab \bar \A^{\frac12} \w &= \left( \E[\tilde \X^\dagger \tilde \X] - (\I - \bar \P) \right) (\I - \bar \P)^{-1} \cdot (\tilde \Sigmab + \lambda \I)^{-1} \tilde \Sigmab \A^{\frac12} \bar \w \\
  %%%
  &= \left( \E[\tilde \X^\dagger \tilde \X] (\I - \bar \P)^{-1} - \I  \right) \cdot \tilde \w_\mu.
\end{align*}
Since $\w = \A^{-\frac12} \tilde \w$, we have
\begin{align*}
  \A^{\frac12} ( \E[\w_*] - \w_\mu ) = \E[\tilde \w_*] - \tilde \w_\mu = \left( \E[\tilde \X^\dagger \tilde \X] (\I - \bar \P)^{-1} - \I  \right) \cdot \A^{\frac12} \w_\mu
\end{align*}
and thus
\[
  \| \E[\w_*] - \w_\mu \|_\A \le \left\| \left( \E[\tilde \X^\dagger \tilde \X] (\I - \bar \P)^{-1} - \I  \right) \right\| \cdot \| \w_\mu \|_\A.
\]
\end{proof}

\section{Generic output}

The idea is to evaluate the implicit ``regularization'' effect. For the minimum norm solution, it thus remains to assess
\begin{equation}\label{eq:implicit-regularization}
  \E[ \X^\dagger \y] = \begin{cases} (\Sigmab + \lambda \I_d)^{-1} \E[y(\x) \x], \quad n < d \\ \Sigmab^{-1} \E[y(\x) \x], \quad n \ge d \end{cases}
\end{equation}
with $\lambda$ the unique solution to $n = \tr \Sigmab (\Sigmab+ \lambda \I_d)^{-1}$.

%\section{Proof of \texorpdfstring{\eqref{eq:implicit-regularization}}{something} }

\begin{proof}[Proof of \eqref{eq:implicit-regularization}]
The idea is to show the following approximation in the $n < d$ case
\[
  %\| \E[\X^\dagger \y] - (\Sigmab + \lambda \I_d)^{-1} \E[y(\x) \x] \| \le C \frac{\sqrt n}{r} \cdot \left( \| \E[y(\x) \x] \| + \left( \E \| \x y(\x) \|^4 \right)^{\frac14} \right)
  \| \E[\X^\dagger \y] - (\Sigmab + \lambda \I_d)^{-1} \E[y(\x) \x] \| \le C \frac{\sqrt n}{r} \cdot \| \E[y(\x) \x] \| + \left( \E[y^4(\x)] \right)^{\frac14}
\]
where we recall $\X \in \mathbb R^{n \times d}$ having i.i.d.~rows of $\mathcal N(0, \Sigmab)$.

Since $\X^\dagger = (\X^\top \X)^\dagger \X^\top$, we have
\begin{align*}
  \E[ \X^\dagger \y] & = \E[ (\X^\top \X)^\dagger \X^\top \y] =  \E \left[ (\X^\top \X)^\dagger \left( \sum_{i=1}^n y(\x_i) \x_i \right) \right] \\
  %%%
  & = \sum_{i=1}^n \E [(\X^\top \X)^\dagger \x_i y(\x_i)] = n \cdot \E \frac{\P_{-n} \x_n y(\x_n)}{ \x_n^\top \P_{-n} \x_n }
\end{align*}
where we used Lemma~1 for the last equality. Since $\P_{-n}$ is independent of $\x_i$, we can write
\[
  n \cdot \E \frac{\P_{-n} \x_n y(\x_n)}{ \x_n^\top \P_{-n} \x_n } = n \cdot \E \frac{\P_{-n} \E [\x_n y(\x_n)] }{ \tr \Sigmab \P_{-n} } + n \cdot \E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} }
\]
with the second RHS term bounded (in a Frobenius norm sense) as
% \begin{align}
%    & \left\|\E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} } \right\| = \left\| \E \left[ (\X^\top \X)^\dagger \x_n y(\x_n) \cdot \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \tr \Sigmab \P_{-n} } \right] \right\| \nonumber \\
%    %%%
%    &\le \sqrt{ \E \| (\X^\top \X)^\dagger \x_n y(\x_n) \|^2} \cdot \sqrt{\E_{\P_{-n}} \E_{\x_n} \left[ \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \tr \Sigmab \P_{-n} } \right]^2 }. \label{eq:bound-dif-1}
% \end{align}
%%%
\begin{align}
   & n \cdot \left\|\E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} } \right\| = n \cdot \left\| \E \left[ \frac{ \P_{-n} \x_n y(\x_n) }{\x_n^\top \P_{-n} \x_n } \cdot \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \tr \Sigmab \P_{-n} } \right] \right\| \nonumber \\
   %%%
   &\le n \sqrt{ \E \left\| \frac{ \P_{-n} \x_n y(\x_n) }{\x_n^\top \P_{-n} \x_n } \right\|^2 } \cdot \sqrt{\E_{\P_{-n}} \E_{\x_n} \left[ \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \tr \Sigmab \P_{-n} } \right]^2 }. \label{eq:bound-dif-1}
\end{align}
%%%
% \begin{align}
%    & \left\|\E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} } \right\| = \left\| \E \left[ \frac{ \P_{-n} \x_n y(\x_n) }{ \tr \Sigmab \P_{-n} } \cdot \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \x_n^\top \P_{-n} \x_n } \right] \right\| \nonumber \\
%    %%%
%    &\le \sqrt{ \E \left\| \frac{ \P_{-n} \x_n y(\x_n) }{\tr \Sigmab \P_{-n}} \right\|^2 } \cdot \sqrt{\E \left[ \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \x_n^\top \P_{-n} \x_n } \right]^2 } \nonumber \\
%    %%%
%    & \le \sqrt{ \E \frac1{ (\tr \Sigmab \P_{-n})^2 } \cdot \E \| \x_n y(\x) \|^2 } \cdot \sqrt{\E \left[ \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \x_n^\top \P_{-n} \x_n } \right]^2 } \nonumber \\
%    %%%
%    & \le \frac{C}{ \sqrt{r-n} } \sqrt{ \E \| \x_n y(\x) \|^2 } \cdot \sqrt{\E \left[ \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \x_n^\top \P_{-n} \x_n } \right]^2 } \nonumber \\
%    %%%
%    &\le \frac{C}{ \sqrt{r-n} } \sqrt{ \E \| \x_n y(\x) \|^2 } \cdot \left(\E (\tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n)^4 \cdot \E \left[ \frac1{ \x_n^\top \P_{-n} \x_n } \right]^4 \right)^{\frac14} \label{eq:bound-dif-1}
% \end{align}
% with Cauchy–Schwarz inequality, $\| \P_{-n} \| \le 1$ and the fact that $\E \frac1{ (\tr \Sigmab \P_{-n}^2 } \le \frac{C}{s}$ for $s= \tr \Sigmab \E[\P_{-n}] \ge r - n$ from previous proof.

For the first RHS term, note that
\[
  \E \left\| \frac{ \P_{-n} \x_n y(\x_n) }{\x_n^\top \P_{-n} \x_n } \right\|^2 = \E \frac{ y^2(\x) \x_n^\top \P_{-n} \x_n }{ (\x_n^\top \P_{-n} \x_n )^2} = \E \frac{y^2(\x)}{ \x_n^\top \P_{-n} \x_n } \le \sqrt{ \E[y^4(\x)] \cdot \E \frac1{ (\x_n^\top \P_{-n} \x_n)^2 } }
\]
with Cauchy–Schwarz inequality, where we need the following lemmas.

\begin{lemma}\label{lem:inverse-moment-bound}
  Fix $k \geq 2$. There exists constants $c, C > 0$ (depending on $k$)
  such that for any $d \geq 1$,
  $\z\sim\Nc(\zero,\I_d)$,
  and any $d\times d$ symmetric positive semidefinite matrix
  $\A \in \mathbb R^{d \times d}$ satisfying $\tr \A \ge c \cdot \ln d$,
  we have
  \[
    %\left( \E \left[ \frac1{ (\z^\top \A \z)^k } \right] \right)^{1/k} \le C \cdot \frac1{\tr \A} 
    \E \left[ \frac1{ (\z^\top \A \z)^k } \right] \le C \cdot \frac1{ (\tr \A)^k} 
  \]
\end{lemma}

\begin{proof}
  Fix $a = 2 k + 1$.
  Let $Z = \z^\top \A \z 
  = \sum_{i=1}^d \lambda_i z_i^2 
  = Z_{\leq a} + Z_{> a}$
  where $Z_{\leq a} = \sum_{i \leq a} \lambda_i z_i^2$
  and $Z_{\geq a} = \sum_{i > a} \lambda_i z_i^2$.
  Splitting on the event $B = \{ Z_{> a} \geq \frac{1}{2} \E Z_{> a} \}$
  \begin{align*}
    \E\left[ Z^{-k} \right]
    &= \Pr[B] \E\left[ Z^{-k} \mid B\right] 
    + \Pr[B^c] \E\left[ Z^{-k} \mid B^c\right] \\
    &\leq \Pr[B] \E\left[ Z_{> a}^{-k} \mid B\right] 
    + \Pr[B^c] \E\left[ Z_{\leq a}^{-k} \mid B^c\right]
    = \Pr[B] \E\left[ Z_{> a}^{-k} \mid B\right] 
    + \Pr[B^c] \E\left[ Z_{\leq a}^{-k} \right] \\
    &\leq 1 \cdot \left(\frac{1}{2} \E Z_{>a} \right)^{-k}
    + \Pr[B^c] \E\left[ Z_{\leq a}^{-k}\right] \\
    &\leq \left(\frac{2}{r - a} \right)^{k}
    + \Pr[B^c] \lambda_a^{-k} \E\left[ (\sum_{i=1}^a z_i^2)^{-k} \right]
  \end{align*}
  where we have used independence of $Z_{> a}$ and $Z_{\leq a}$ to remove
  the conditioning on $B^c$,
  and $\|\A\| = 1 \implies \E Z_{> a} \geq r - a$.
  Recognizing the last expectation as
  the $k$th moment of an inverse Chi-square with $a$ degrees of freedom,
  we have
  \begin{align*}
    \E\left[ Z^{-k} \right]
    &\leq \left(\frac{2}{r - a} \right)^{k}
    + \Pr[B^c] \lambda_a^{-k} \frac{1}{(a - 2)!!}
    \leq \left(\frac{2}{r - a} \right)^{k}
    + \Pr[B^c] \left(\frac{d-a}{r-a}\right)^k \frac{1}{(a - 2)!!}
  \end{align*}
  where we have used $r - a \leq (d - a) \lambda_a$.
  To bound $\Pr[B^c]$, note that $Z$ is $(4 \tr (\A^2), 4 \|\A\|)$ sub-exponential,
  so
  \begin{align*}
    \label{eq:sub-exp-quad-form}
    \Pr[Z - \E z \leq -t] \leq \begin{cases}
      \exp\left(-\frac{t^2}{8 (\tr \A)^2}\right),
      &\text{ if } 0 \leq t \leq \frac{\tr( \A^2 )}{\|\A\|_2} \\
      \exp\left(-\frac{t}{8 \|\A\|_2}\right), &\text{ if }t > \frac{\tr \A}{\|\A\|_2}
    \end{cases}
  \end{align*}
  \feynman{Any significance that the Gaussian to exponential tail transition
  occurs when the deviation $t$ exceeds the soft rank of $\A$?}
  Hence, using $\|\A\| = 1$ to bound $\E Z_{> a} \geq r - a$, we have
  \begin{align*}
    \Pr[B^c]
    &= \Pr[Z_{> a} < \frac{1}{2} \E Z_{> a}]
    \leq \Pr[Z_{> a} - \E Z_{> a} < -\frac{1}{2} (r - a)]
    \leq \exp\left( - \frac{r - a}{16 \cancelto{1}{\|\A\|_2}} \right)
  \end{align*}
  Taking $c_0 = \frac{1}{16}$, we have altogether
  \begin{align*}
    \E\left[ Z^{-k} \right]
    &\leq \left(\frac{2}{r - a} \right)^{k}
    + e^{-c_0 (r - a)} \left(\frac{d-a}{r-a}\right)^k \frac{1}{(a - 2)!!}
  \end{align*}
  When $r - a \geq c_0^{-1} k \log(d)$ or
  equivalently $r \geq C \log(d)$ for some constant $C$ depending on $k$,
  \begin{align*}
    \E\left[ Z^{-k} \right]
    &\leq \left(\frac{2}{r - a} \right)^{k}
    + \left(\frac{1}{r-a}\right)^k \frac{1}{(a - 2)!!} \\
    &\leq \left(
        \frac{2}{r - a} + \frac{1}{r-a} ((a - 2)!!)^{-k}
    \right)^k\\
    &\leq \left(C \frac{1}{r}\right)^k
  \end{align*}
  for some constant $C$ independent of $r$ and $d$.
\end{proof}

Since
\[
  \E \frac1{ (\x_n^\top \P_{-n} \x_n)^2} = \E \left[ \E \frac1{ (\x_n^\top \P_{-n} \x_n)^2}~|~\P_{-n} \right] 
\]
so that by taking $\A = \Sigmab^{\frac12} \P_{-n} \Sigmab^{\frac12}$ and $k=2$ in Lemma~\ref{lem:inverse-moment-bound}, we have
\[
  \E \frac1{ (\x_n^\top \P_{-n} \x_n)^2} \le C \E \frac1{ (\tr \Sigmab \P_{-n})^2}
\]
and thus
\[
  \sqrt{\E \left\| \frac{ \P_{-n} \x_n y(\x_n) }{\x_n^\top \P_{-n} \x_n } \right\|^2} \le C \cdot \left( \E[y^4(\x)] \right)^{\frac14} \cdot \left( \E \frac1{ \tr (\Sigmab \P_{-n})^2 } \right)^{\frac14 } \le C \cdot \left( \E[y^4(\x)] \right)^{\frac14} \cdot \frac1{\sqrt{r - n}}
\]
where we used the fact that $\E \frac1{ (\tr \Sigmab \P_{-n})^2 } \le \frac{C}{s^2}$ for $s= \tr \Sigmab \E[\P_{-n}] \ge r - n$ from previous proof.

\begin{lemma}[Lemma B.26 in \cite{bai2010spectral}]
    \label{lem:bai-b26}
    Let $\x \in \mathbb R^d$ have i.i.d.~zero mean, unit variance entries and $\E[\lvert x_0 \rvert^L] \leq \nu_L$ for some $L \geq 4$. Then for $\A \in \mathbb R^{d \times d}$ of bounded norm and independent of $\x$ and $1 \leq \ell \leq L/2$
    \begin{align*}
        \E \left[\left| \x^\top \A \x - \tr \A \right|^\ell\right] \leq C_\ell \left[ \left( \nu_4 \tr(\A \A^\top)\right)^{\ell/2} + \nu_{2\ell} \tr(\A \A^\top)^{\ell/2} \right]
    \end{align*}
    for some $C_\ell > 0$.
\end{lemma}

To bound the second RHS in \eqref{eq:bound-dif-1}, we take then $\ell = 2$ in Lemma~\ref{lem:bai-b26}, we have
\begin{align*}
  &\E_{\x_n} \left[ \frac{ \x_n^\top \P_{-n} \x_n - \tr \Sigmab \P_{-n} }{ \tr \Sigmab \P_{-n} } \right]^2 = \frac{ \E_{\x_n} (\tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n)^2 }{ (\tr \Sigmab \P_{-n})^2 } \\
  %%%
  &\le C \frac{ \tr \Sigmab \P_{-n} \Sigmab }{ (\tr \Sigmab \P_{-n})^2 } \le \frac{C}{ \tr \Sigmab \P_{-n} }.
\end{align*}

As a consequence, we obtain
\[
   \sqrt{\E \left[ \frac{ \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n }{ \tr \Sigmab \P_{-n} } \right]^2 } \le C \cdot \frac1{\sqrt{r - n}}
\]
so that
\[
  n \cdot \left\|\E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} } \right\| \le C \cdot \left( \E[y^4(\x)] \right)^{\frac14} \cdot \frac{n}{r - n}.
\]
% Taking $\ell = 4$ in Lemma~\ref{lem:bai-b26}, we have
% \begin{align*}
%   &\E_{\x_n} (\x_n^\top \P_{-n} \x_n - \tr \Sigmab \P_{-n})^4 \le C (\tr(\Sigmab \P_{-n} \Sigmab) )^2
% \end{align*}
% for some constant $C > 0$. For the term $\E (\x_n^\top \P_{-n} \x_n)^{-4}$, note that
% \begin{align*}
%   \frac1{\x_n^\top \P_{-n} \x_n} =
% \end{align*}
% so that with $\E \X^\dagger \y = n \E (\X^\top \X)^\dagger \x_n y(\x_n)$, we can conclude from \eqref{eq:bound-dif-1} that
% \begin{align}
%    & n\left|\E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} } \right| \le n\sqrt{ \E \| (\X^\top \X)^\dagger \x_n y(\x_n) \|^2 \cdot C \E \left[ \frac1{ \tr \Sigmab \P_{-n} } \right] } \nonumber \\
%    %%%
%    &\le \frac{C}{\sqrt s} \sqrt{ \E \| (\X)^\dagger \y \|^2 } \le \frac{C}{\sqrt{r-n} } \sqrt{ \E \| (\X)^\dagger \y \|^2 } \label{eq:bound-dif-3}
% \end{align}

As a consequence,
\begin{align*}
  \left\| \E[\X^\dagger \y] - n \E \frac{\P_{-n} \E [\x_n y(\x_n)] }{ \tr \Sigmab \P_{-n} } \right\| \le n \left\| \E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\top \P_{-n} \x_n \right) }{ \x_n^\top \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} } \right\| \le C \cdot \left( \E[y^4(\x)] \right)^{\frac14} %\left( \E \| \x_n y(\x_n) \|^4 \right)^{\frac14} \cdot \frac{C}{\sqrt{r - n}}.
\end{align*}

Let us move on and focus on the expectation $\E \frac{\P_{-n} }{ \tr \Sigmab \P_{-n} }$, we have from previous proof that
\[
  n\E \frac{\P_{-n} }{ \tr \Sigmab \P_{-n} } = \E[ \tilde \gamma \P_{-n} ] =  \bar \gamma \E[\P_{-n}] + \E [(\tilde \gamma - \bar \gamma) \P_{-n}] = \bar \gamma \bar \P + \underbrace{\bar \gamma ( \E[\P_{-n}] - \bar \P)}_{T_3} + \underbrace{\E [(\tilde \gamma - \bar \gamma) \P_{-n}]}_{T_4}
\]
where we recall the definition $\tilde \gamma = n/\tr \P_{-n} \Sigmab$ and it thus remains to bound $T_3$ and $T_4$ respectively to reach
\begin{align*}
  \left\| n \E \frac{\P_{-n} }{ \tr \Sigmab \P_{-n} } - \gamma \bar \P \right\| \le \| T_3 \| + \| T_4 \|
\end{align*}
and thus the conclusion.

To this end, we write, similar to the bound of $T_1$ in previous proof that
\[
  \| T_4 \| \le \bar \gamma C \frac1s ( (\bar \gamma + 1) \sqrt{cn} + 2 ) \le \frac{ \bar \gamma C ( (\bar \gamma + 1) \sqrt{cn} + 2 ) }n
\]
as well as
\[
  \| T_3 \| \le \bar \gamma \| \E[\P_{-n} - \P] \| + \bar \gamma \| \E[\P] - \bar \P \| \le \bar \gamma C \frac1n + \bar \gamma C' \frac{\sqrt n}{r}
\]
and ultimately
\[
  \| \E[\X^\dagger \y] - (\Sigmab + \lambda \I_d)^{-1} \E[y(\x) \x] \| \le (\| T_3 \| + \| T_4 \|) \| \E [\x y(\x) ] \| + C \left( \E[y^4(\x)] \right)^{\frac14} %+ \left( \E \| \x y(\x) \|^4 \right)^{\frac14} \cdot c \cdot \frac{\sqrt n}{r}
\]
which concludes the proof.
\end{proof}

\subsection{New strategy}

Let $\xbt=(\Sigmab+\lambda\I)^{-1}\x$ and use $y$ as a shorthand for
$y(\x)$. This way, we have $\E[\xbt y ]=(\Sigmab+\lambda\I)^{-1}\E[\x y]$.
Note that $(\Sigmab+\lambda\I)^{-1} = \bar\gamma\bar\P$. Letting
$\gamma_n=n/(\x_n^\top\P_{-n}\x_n)$ and $y_n=y(\x_n)$, we have
\begin{align*}
  \E[\X^\dagger\y]
  & = \E[\gamma_n\P_{-n}\x_ny_n]\\
  &= \E[\gamma_n\P_{-n}\bar\gamma^{-1}\bar\P^{-1}\xbt_ny_n]\\
  &=\E[\bar\gamma\P_{-n}\bar\gamma^{-1}\bar\P^{-1}\xbt_n y_n] +
    \E\big[\P_{-n}\bar\P^{-1}\xbt_ny_n(\gamma_n-\bar\gamma)\bar\gamma^{-1}\big] \\
  &=\E[\P_{-n}]\bar\P^{-1}\E[\xbt_ny_n]
    + \E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n}(\bar\alpha-\alpha_n)\bigg],
\end{align*}
where $\bar\alpha=n/\bar\gamma$ and
$\alpha_n=\x_n^\top\P_{-n}\x_n$. So we have
\begin{align*}
  \big\|\E[\X^\dagger\y]-\E[\xbt y]\big\|
  &\leq \|\E[\P_{-n}]\bar\P^{-1}-\I\|\cdot\|\E[\xbt_ny_n]\| +
    \bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n}
    (\bar\alpha-\alpha_n)\bigg]\bigg\|.
\end{align*}

With Theorem~\ref{t:main}, the first term is bounded by $O(\frac{\sqrt n}{r})\cdot \|\E[\xbt_ny_n]\|$ and it remains to treat the second term.

Note that
\[
  \E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg] = \E\bigg[ n \frac{\P_{-n} \x_n y_n}{\x_n^\top\P_{-n}\x_n} \cdot \frac{\bar\alpha-\alpha_n}{\bar \alpha} \bigg]
\]
so that with Cauchy–Schwarz inequality
\begin{align*}
  &\bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg]\bigg\| \le \sqrt{ \frac{ \E (\bar \alpha - \alpha_n)^2 }{\bar \alpha^2 } \cdot \E \left[ \left\|  n \frac{\P_{-n} \x_n y_n}{\x_n^\top\P_{-n}\x_n} \right\|^2 \right]  } \\ 
  &= \frac{ \sqrt{ \E (\bar \alpha - \alpha_n)^2 } }{\bar \alpha }  \cdot \sqrt{ \E \| (\X)^\dagger \y \|^2 }.
\end{align*}

Recall $\bar s = \tr \bar \P \Sigmab = \bar \alpha$ and $s = \tr \E[\P_{-n}] \Sigmab$, we have
\begin{align*}
  &\E (\bar \alpha - \alpha_n)^2 = \E (\alpha_n - \tr \P_{-n} \Sigmab + \tr \P_{-n} \Sigmab - s + s - \bar s)^2 \\ 
  %%%
  &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^2 + \E (\tr \P_{-n} \Sigmab - s)^2 + (s - \bar s)^2  
  %%%
  %+ + 2 \E (\alpha_n - \tr \P_{-n} \Sigmab) (\tr \P_{-n} \Sigmab - s) + 2 \E (\alpha_n - \tr \P_{-n} \Sigmab) (s - \bar s) + 2 \E (\tr \P_{-n} \Sigmab - s) (s - \bar s)
\end{align*}
so that by taking $\A = \Sigmab^{\frac12} \P_{-n} \Sigmab^{\frac12}$ and $\ell = 2$ in Lemma~\ref{lem:bai-b26}, together with Lemma~\ref{l:rank-one-projection} and $|s - \bar s| \le \bar \gamma \sqrt{cn} + 2$, we deduce
\begin{align*}
  \E (\bar \alpha - \alpha_n)^2 &\le c_1 s + c_2 n + \bar \gamma^2 c_3 n + 4 \bar \gamma \sqrt{c_3 n} + 4 \le 4 + c_1 (\bar s + \bar \gamma \sqrt{c_3 n} + 2) + 4 \bar \gamma \sqrt{c_3 n} + (c_2 + \bar \gamma^2 c_3)n \\ 
  &= c_1 \bar s + (c_2 + \bar \gamma^2 c_3)n + (c_1 + 4)\bar \gamma \sqrt{c_3 n} + 4 + 2 c_1
\end{align*}
and thus
\begin{align*}
  \bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg]\bigg\| &\le \sqrt{ \E \| (\X)^\dagger \y \|^2 } \cdot \sqrt{ \frac{ c_1 \bar s + (c_2 + \bar \gamma^2 c_3)n + (c_1 + 4)\bar \gamma \sqrt{c_3 n} + 4 + 2 c_1 }{\bar s^2} } \\ 
  &\le \sqrt{ \E \| (\X)^\dagger \y \|^2 } \cdot \frac{C}{\sqrt n}
\end{align*}
for some constant $C> 0$, where we recall $\bar s \ge n$. Putting thing together, we get
\begin{equation}
  \big\|\E[\X^\dagger\y]-\E[\xbt y]\big\| \leq C \frac{\sqrt n}{r}
  \cdot\|\E[\xbt_ny_n]\| + C\frac{\sqrt n}{r} \cdot \sqrt{ \E \|
    (\X)^\dagger \y \|^2 }. 
\end{equation}

\clearpage

Note that with Jensen's inequality,
\begin{align*}
  &\bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg]\bigg\| \le \sqrt{ \E \left[ \left( \frac{\bar \alpha - \alpha_n}{\alpha_n} \right)^2 y_n^2 \cdot \tilde \x_n^\top \bar \P^{-1} \P_{-n} \bar \P^{-1} \tilde \x_n \right] } \\ 
  &\le \sqrt{ \E \left[ \left( \frac{\bar \alpha - \alpha_n}{\alpha_n} \right)^2 \cdot (\bar \gamma + 1)^2 \cdot \| \tilde \x_n y_n \|^2  \right] }
\end{align*}
where we used the fact that $\bar \P^{-1} = \bar \gamma \Sigmab + \I$ so that $\bar\P \preceq (\bar\gamma+1) \I$ and $\P^{-1} \P_{-n} \bar \P^{-1} \preceq (\bar\gamma+1)^2 \I$.

% As a consequence, with Cauchy–Schwarz inequality, we have
% \begin{align*}
%   \bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg]\bigg\| \le (\bar \gamma + 1) \cdot \left( \E[ \| \tilde \x_n y_n \|^4 ] \right)^{1/4} \cdot \left( \E \frac{  (\alpha_n - \bar \alpha)^4}{ \alpha_n^4} \right)^{1/4}.
% \end{align*}
% Let us first focus on the expectation (conditioned on $\P_{-n}$) $\E_{\x_n} \frac{  (\alpha_n - \bar \alpha)^4}{ \alpha_n^4}$, and write
% \begin{align*}
%   \E_{\x_n} \frac{  (\x_i^\top \P_{-n} \x_i - \bar \alpha)^4}{ (\x_i^\top \P_{-n} \x_i)^4 } \le \E_{\x_n} \left[ \frac{  (\x_i^\top \P_{-n} \x_i - \bar \alpha)^4 }{ (\x_i^\top \P_{-n} \x_i)^4 }~|~\x_n^\top \P_{-n} \x_n \ge \right]
% \end{align*}

As a consequence, with Cauchy–Schwarz inequality, we have
\begin{align*}
  \bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg]\bigg\| \le (\bar \gamma + 1) \cdot \left( \E[ \| \tilde \x_n y_n \|^4 ] \right)^{1/4} \cdot \left( \E (\bar \alpha - \alpha_n)^8 \right)^{1/8} \cdot \left( \E \frac1{ \alpha_n^8 } \right)^{1/8}.
\end{align*}

Note from Lemma~\ref{lem:inverse-moment-bound} that
\[
  \E \frac1{ \alpha_n^8 } = \E \frac1{ (\x_n^\top \P_{-n} \x_n)^8} \le C \cdot \E \frac1{ (\tr \Sigmab \P_{-n})^8 }.
\]
While we can easily bound the probability of the bad event $\tr \P_{-n} \Sigmab \ge \epsilon s$ for $s = \tr \E[\P_{-n}] \Sigmab$, we do not have an easy control of what happens in this worst case scenario, where the denominator can get arbitrarily close to zero. In the previous proof, we always work with $\frac{\| \Sigmab \P_{-n} \| }{ \tr \P_{-n} \Sigmab } \le 1$ and allow for a control, here we have only $\frac{ \P_{-n} }{ \tr \P_{-n} \Sigmab }$ and the bound will depend on $\lambda_{\min}(\Sigmab)$.

\bigskip

We recall $\bar s = \tr \bar \P \Sigmab = \bar \alpha$, $s = \tr \E[\P_{-n}] \Sigmab$ and $|s - \bar s| \le \bar \gamma \sqrt{cn} + 2$, we have
\begin{align}
  \E (\bar \alpha - \alpha_n)^8 &= \E (\alpha_n - s + s - \bar s)^8 = \sum_{k=0}^8 C_k \E(\alpha_n - s )^k (s-\bar s)^{8-k} \nonumber \\ 
  %%%
  &\le \sum_{k=0}^8 C_k \E(\alpha_n - s )^k (\bar \gamma \sqrt{cn} + 2)^{8-k}. \label{eq:diff}
\end{align}
%where we used the fact that $\E(\alpha_n - s )^k \le \E(\alpha_n - s )^8$ for $0 \le k \le 8$.

Similarly, we have $\E [\alpha_n - s] = 0$ and for $2 \le k \le 8$
\[
  \E(\alpha_n - s )^k = \E (\alpha_n - \tr \P_{-n} \Sigmab + \tr \P_{-n} \Sigmab - s)^k
\]
so that
\begin{align*}
  \E(\alpha_n - s )^2 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^2 + \E (\tr \P_{-n} \Sigmab - s)^2 \\ 
  %%%
  \E(\alpha_n - s )^3 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^3 + \E (\tr \P_{-n} \Sigmab - s)^3 \\ 
  %%%
  \E(\alpha_n - s )^4 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^4 + 6 \E(\alpha_n - \tr \P_{-n} \Sigmab)^2 \cdot \E (\tr \P_{-n} \Sigmab - s)^2 + \E (\tr \P_{-n} \Sigmab - s)^4 \\ 
  %%%
  \E(\alpha_n - s )^5 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^5 + 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^3 \cdot \E (\tr \P_{-n} \Sigmab - s)^2 \\ 
  &+ 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^2 \cdot \E (\tr \P_{-n} \Sigmab - s)^3 + \E (\tr \P_{-n} \Sigmab - s)^5 \\ 
  &\ldots
  %%%
  % \E(\alpha_n - s )^6 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^5 + 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^3 \cdot \E (\tr \P_{-n} \Sigmab - s)^2 \\ 
  % &+ 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^2 \cdot \E (\tr \P_{-n} \Sigmab - s)^3 + \E (\tr \P_{-n} \Sigmab - s)^5 \\
  % %%%
  % \E(\alpha_n - s )^7 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^5 + 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^3 \cdot \E (\tr \P_{-n} \Sigmab - s)^2 \\ 
  % &+ 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^2 \cdot \E (\tr \P_{-n} \Sigmab - s)^3 + \E (\tr \P_{-n} \Sigmab - s)^5 \\
  % %%%
  % \E(\alpha_n - s )^8 &= \E (\alpha_n - \tr \P_{-n} \Sigmab)^5 + 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^3 \cdot \E (\tr \P_{-n} \Sigmab - s)^2 \\ 
  % &+ 10 \E(\alpha_n - \tr \P_{-n} \Sigmab)^2 \cdot \E (\tr \P_{-n} \Sigmab - s)^3 + \E (\tr \P_{-n} \Sigmab - s)^5 
\end{align*}
Note with Lemma~\ref{lem:bai-b26} we have, for $k \ge 2$ that
\[
  \E (\alpha_n - \tr \P_{-n} \Sigmab)^k \le c \cdot \E (\tr \Sigmab \P_{-n} \Sigmab)^{k/2} \le c \cdot \E (\tr \Sigmab \P_{-n})^{k/2}.
\]
Consider then the event $\tr \P_{-n} \Sigmab \ge \frac32 s$, which, according to Lemma~\ref{l:trace}, happens with exponentially low probability as
\[
  \Pr(\tr \P_{-n} \Sigmab \le s/2) \le 2 e^{- \frac{c s^2}n}.
\]
As a consequence, we have
\[
  \E (\alpha_n - \tr \P_{-n} \Sigmab)^k \le c \cdot \E \left[ (\tr \Sigmab \P_{-n})^{k/2}~|~\tr \P_{-n} \Sigmab \le \frac32 s\right] + c \cdot (\tr \Sigmab)^{k/2} \cdot e^{-xx} \le c \cdot s^{k/2}
\]
for $s \tr \Sigmab $.

This allows us to conclude, together with Lemma~\ref{l:trace} (\textbf{To be done!}) that
\[
  \E (\alpha_n - s)^k \le c_k n^{k/2}
\]
holds for some constant $c_k$ that depends on $2 \le k \le 8$ but independent of $n$, as long as

Plugging back into \eqref{eq:diff}, we obtain
\[
  \E (\bar \alpha - \alpha_n)^8 \le C n^4.
\]
Further note from Lemma~\ref{lem:inverse-moment-bound} that
\[
  \E \frac1{ \alpha_n^8 } = \E \frac1{ (\x_n^\top \P_{-n} \x_n)^8} \le C \cdot \E \frac1{ (\tr \Sigmab \P_{-n})^8 }
\]
we conclude that
\[
  \bigg\|\E\bigg[\frac{\P_{-n}\bar\P^{-1}\xbt_ny_n}{\x_n^\top\P_{-n}\x_n} (\bar\alpha-\alpha_n)\bigg]\bigg\| \le (\bar \gamma + 1) \cdot \left( \E[ \| \tilde \x_n y_n \|^4 ] \right)^{1/4} \cdot \left( \E (\bar \alpha - \alpha_n)^8 \right)^{1/8} \cdot \left( \E \frac1{ \alpha_n^8 } \right)^{1/8}
\]

\subsection{Linear plus non-linear}
Suppose that $y = \x^\top\w + f + \xi$, where
$\xi\sim\Nc(0,\sigma^2)$ and $f=f(\x)$ is a non-linearity.
Let $\w_\mu = \Sigmab_\lambda^{-1}\E[\x y]$, where $\Sigmab_\lambda =
\Sigmab+\lambda\I$, and assume that  
$ \big(\E_\x[\|\x f\|^4]\big)^{\frac14}\leq M\cdot \|\w_\mu\|$. Next,
letting $\f = f(\X)$, suppose we have:
\begin{align*}
  \|\E[\X^\dagger\f] - (\Sigmab+\lambda\I)^{-1}\E[\x f]\|\leq \epsilon\cdot
 \big(\E_\x[\|\x f\|^4]\big)^{\frac14},
\end{align*}
for $\epsilon=O(\frac{\sqrt n}{r})$. Then it follows that:
\begin{align*}
  \|\E[\X^\dagger\y] - \w_\mu\|
  &= \|\E[\X^\dagger\X]\w + \E[\X^\dagger\f] -
    \Sigmab_\lambda^{-1}(\Sigmab\w + \E[\x f])\|\\
  &\leq \epsilon\cdot \|\Sigmab_\lambda^{-1}\Sigmab\w\| +
    \epsilon\cdot \big(\E[\|\x f\|^4]\big)^{\frac14}\\
  &\leq  \epsilon\cdot \|\w_\mu\| + \epsilon\cdot
    \|\Sigmab_\lambda^{-1}\E[\x f]\| +  \epsilon\cdot \big(\E[\|\x
    f\|^4]\big)^{\frac14}\\
  &\leq \epsilon\cdot \|\w_\mu\| + 2\epsilon\cdot \big(\E[\|\x
    f\|^4]\big)^{\frac14}\\
  &\leq O(\epsilon M)\cdot \|\w_\mu\|.
\end{align*}
\section{Computation of the expectation}

\subsection{Generalized linear model}

Consider the nonlinear model $\y = \sigma(\X \w) + \xi$, with $\y \in \mathbb R^n$, $\w \in \mathbb R^d$, $\X \in \mathbb R^{n \times d}$ and independent noise $\xi$ of zero mean. Here $\sigma: \mathbb R \mapsto \mathbb R$ is the \emph{link function}, popular choices of $\sigma(\cdot)$ are
\begin{enumerate}
  \item \emph{logit link function}: $\sigma(t) = \frac1{ 1 + e^{-t} }$;
  \item \emph{prohibit link function}: $\sigma(t) = \Phi(t)$ with $\Phi$ the cumulative distribution function of standard normal;
  \item \emph{cloglog link function}: $\sigma(t) = 1 - e^{-e^t}$;
  \item possible non-differentiable function such as the ReLU $\sigma(t) = \max(t,0)$ in a neural network context.
\end{enumerate}

For $\x \sim \Nc(0, \Sigmab)$ we denote $\x = \Sigmab^{\frac12} \z$ with $\z \sim \Nc(0, \I_d)$ standard normal. We can decompose $\z$ as follows
\[
  \z = \eta \frac{\Sigmab^{\frac12} \w}{ \sqrt{\w^\T \Sigmab \w} } + \z_\perp, \quad \eta = \frac{\w^\T \Sigmab^{\frac12} \z}{ \sqrt{\w^\T \Sigmab \w} }
\]
with $\Sigmab^{\frac12} \w/\sqrt{\w^\T \Sigmab \w}$ the unit vector in the direction of $\Sigmab^{\frac12} \w$ and $\z_\perp$ lying on the $(d-1)$-dimensional subspace orthogonal to $\Sigmab^{\frac12} \w$. Using the orthogonal invariance of the standard Gaussian we have  that both $\eta$ and $\z_\perp$ and they are \emph{independent}.

As such, we have
\[
  \E[ \sigma(\w^\T \x) \x ] = \E \left[ \sigma(\eta \sqrt{\w^\T \Sigmab \w} ) \left( \eta \frac{\Sigmab \w}{ \sqrt{\w^\T \Sigmab \w} } + \Sigmab^{\frac12} \z_\perp \right) \right] = \E \left[ \eta \sigma(\eta \sqrt{\w^\T \Sigmab \w} ) \right] \frac{\Sigmab \w}{ \sqrt{\w^\T \Sigmab \w} }.
\]
which is a constant scaling of $\Sigmab \w$, with the constant $\kappa$ depends on the link function $\sigma(\cdot)$ via
\[
  \kappa = \E \left[ \eta \sigma(\eta \sqrt{\w^\T \Sigmab \w} ) \right] = \frac1{\sqrt{2 \pi}} \int x \sigma(x \sqrt{\w^\T \Sigmab \w}) e^{-x^2/2} dx.
\]

\subsection{More generic models}

More generally, for $y(\x) \in \mathbb R$ an arbitrary function of $\x \in \mathbb R^d$, we have the following lemma to compute the expectation.

\begin{lemma}[Stein's lemma, from \cite{stein1981estimation}]\label{lem:stein}
Let $x \sim \Nc(0,1)$ and $y: \mathbb R \mapsto \mathbb R$ a continuously differentiable function such that $\E[y'(x)] < \infty$. Then,
\begin{equation}
  \E[x y(x)] = \E[y'(x)].
\end{equation}
Moreover, for $\x \sim \Nc(\zero, \Sigmab)$ with $\Sigmab \in \mathbb R^{d \times d}$ and $y: \mathbb R^d \mapsto \mathbb R$ a continuously differentiable function with derivatives having at most polynomial growth with respect to $d$,
\begin{equation}
  \E[[\x]_i y(\x)] = \sum_{j=1}^d \Sigmab_{ij} \E \left[ \frac{\partial y(\x)}{\partial [\x]_j} \right]
\end{equation}
where $\partial/\partial [\x]_i$ indicates differentiation with respect to the $i$-th entry of $\x$. Or in vector form
\[
  \E[\x y(\x)] = \Sigmab \E[\nabla y(\x)].
\]
\end{lemma}

% \clearpage
% Since we have the following concentration
% \[
%   \frac1n \x_n^\T \P_{-n} \x_n \simeq \bar \gamma^{-1}, \quad \| \E [\P_{-i} - \bar \P] \| \simeq 0
% \]
% therefore
% \[
%   \E[\X^\dagger \y] = \E \sum_{i=1}^n \frac{\P_{-i} \x_i y(\x_i)}{ \x_i^\T \P_{-i} \x_i } \simeq \bar \gamma \E [\P_{-i}] \E[\x_i y(\x_i)] \simeq \bar \gamma \bar \P \E[y(\x) \x] = ( \Sigmab + \bar \gamma^{-1} \I)^{-1} \E[y(\x) \x]
% \]
% note that by definition $\gamma^{-1} = \lambda$ and thus the conclusion.

% To rigorously prove the above result, we need to either find a way, similar to Lemma~2, to compute the expectation
% \[
%   \E \frac{\P_{-n} \x_n y(\x_n)}{ \x_n^\T \P_{-n} \x_n } = \frac{ \E[\P_{-n}] \E[\x_n y(\x_n)]}{ \tr \Sigmab \P_{-n} }
% \]
% for arbitrary $y(\x)$, or to bound the difference
% \[
%   \E \frac{\P_{-n} \x_n y(\x_n)}{ \x_n^\T \P_{-n} \x_n } - \frac{\P_{-n} \x_n y(\x_n) }{ \tr \Sigmab \P_{-n} } = \E \frac{\P_{-n} \x_n y(\x_n) \left( \tr \Sigmab \P_{-n} - \x_n^\T \P_{-n} \x_n \right) }{ \x_n^\T \P_{-n} \x_n \cdot \tr \Sigmab \P_{-n} }
% \]
% perhaps with Bai and Silverstein lemma or Hanson-wright.

% We need the following lemma, which follows from the integration by parts.

% \begin{Lemma}[Nash-Poincar\'e inequality, from \cite{pastur2005simple}]\label{lem:poincare}
% For $\x \sim \Nc(\zero, \Sigmab)$ with $\Sigmab \in \mathbb R^{d \times d}$ and $f: \mathbb R^d \mapsto \mathbb R$ continuously differentiable with derivatives having at most polynomial growth with respect to $d$,
% \begin{equation*}
%     \var[ f(\x) ] \le 2 \E[(\nabla f)^\T \Sigmab (\nabla f)] = 2 \sum_{i,j=1}^d \Sigmab_{ij} \E\left[ \frac{\partial f(\x)}{\partial [\x]_i}\frac{\partial f(\x)}{\partial [\x]_j}\right].
% \end{equation*}
% \end{Lemma}

\section{Beyond min-norm solution}

Another idea is to (always) work with the linear model $\y = \X^\T \w + \xi$, and consider the possible implicit regularization effect of enhancing different structures of the solution. Instead of choosing the min-norm solution, we can consider the solution to the following general optimization problem
\[
  \min_{\w \in \mathbb R^p} \ell(\w)
\]
for a convex function $\ell: \mathbb R^p \mapsto \mathbb R$ that enhances certain structure of the solution. For instance
\begin{enumerate}
  \item min-norm or minimal energy solution $\ell(\cdot) = \| \cdot \|_2$
  \item sparse solution $\ell(\cdot) = \| \cdot \|_1$
  \item binary response $\ell(\cdot) = \| \cdot \|_\infty$ so that $\w \in \{ \pm \}^p$.
\end{enumerate}

\section{Extension to other sketches}

We are interested in extending Lemma~\ref{l:rank-one-projection},
which currently is only applicable for Gaussian $\z$.

Let $z_i$ be independent, mean zero, variance 1, sub-Gaussian
with $\|z_i\|_{\psi_2} \leq K$.
Fix $\A \succ 0$ symmetric.

\begin{proposition}
  \label{prop:inv-moment-high-prob}
  For any $k \geq 1$ and $\epsilon \geq 0$,
  with probability $\geq 1 - e^{-\frac{r}{K^2} (1 - \frac{\epsilon}{\|\A\|})}$
  \begin{align*}
    (\z^\top \A \z)^{-k} < (\epsilon r)^{-k}
  \end{align*}

  In particular, 
  \begin{itemize}
    \item 
      Taking $\epsilon = 1/2$ gives
      $(\z^\top \A \z)^{-k} < (r / 2)^{-k}$.
      with probability $\geq 1 - e^{-\frac{r}{K^2}(1 - \frac{1}{2 \|\A\|})}$.
    \item 
      Taking $\epsilon = \|\A\|/2$ gives
      $(\z^\top \A \z)^{-k} < (\tr \A / 2)^{-k}$.
      with probability $\geq 1 - e^{-\frac{r}{2 K^2}}$.
  \end{itemize}
\end{proposition}

\feynman{We could alternatively bound by $(\epsilon \tr \A)^{-k}$ instead of $r$
with a probability not involving $\|\A\|$}

\begin{proof}
  Since $z_i$ are independent, $\E \z^\top \A \z = \tr \A$.
  By Hanson-Wright,
  the event $E = \{ \z^\top \A \z \geq (1 - \epsilon) \tr \A \}$
  satisfies the required high probability bound.
\end{proof}

\begin{lemma}
  \begin{align*}
    \left\| \E \left[
      \frac{\z \z^\top}{\z^\top \A \z}
      \mid E
    \right] - \frac{\I}{\tr \A} \right\|
    &\leq C \Pr[E]^{-1/2} (\tr \A)^{-1} r^{-1/2}
  \end{align*}
  for some event $E$ with $\Pr[E] \geq 1 - e^{-\frac{r}{2 K^2}}$.

  Alternatively,
  \begin{align*}
    \left\| \E \left[
      \frac{\z \z^\top}{\z^\top \A \z}
      \mid E
    \right] - \frac{\I}{\tr \A} \right\|
    &\leq C \Pr[E]^{-1/2} r^{-3/2}
  \end{align*}
  for some event $E$ with $\Pr[E] \geq 1 - e^{-\frac{r}{K^2} (1 - \frac{1}{2 \|\A\|})}$.
\end{lemma}

\begin{proof}
  By \Cref{prop:inv-moment-high-prob}, the event $E$
  (with $\epsilon = 1/2$ or $\epsilon = \|\A\| / 2$) satisfies the desired
  probability bound. Notice
  \begin{align*}
    \left\| \E \left[ 
      \frac{\z \z^\top}{\z^\top \A \z}
    \mid E \right] - \frac{\I}{\tr \A} \right\| 
    &= \left\| \E \left[ 
      \frac{\z \z^\top}{\z^\top \A \z} - \frac{\z \z^\top}{\tr \A} 
    \mid E \right] \right\| 
    = \left\| \E \left[ 
      \frac{\z \z^\top}{\z^\top \A \z} \cdot \frac{ \tr \A - \z^\top \A \z }{\tr \A} 
    \mid E \right] \right\| \\ 
    %%%
    &\le \sqrt{\E \left[
      \frac{ (\z^\top \z)^2 }{ (\z^\top \A \z)^2 } 
  \mid E \right] \cdot \frac{ \E[(\tr \A - \z^\top \A \z)^2 \mid E]}{(\tr \A)^2} } \\
  \end{align*}
  where we used Jensen and Cauchy–Schwarz.

  We next apply Lemma~\ref{lem:bai-b26} conditionally
  to simplify the second expectation
  \begin{align*}
    \E[(\tr \A - \z^\top \A \z)^2  \mid E]
    &\leq \E[(\tr \A - \z^\top \A \z)^2] \Pr[E]^{-1} \\
    &\leq C \Pr[E]^{-1} \tr(\A^2)
  \end{align*}
  Bounding $\tr(\A^2) = \|\A\|^2 \tr((\A / \|\A\|)^2) \leq \|\A\| \tr(\A)$
  and substituting back in gives
  \[
    \left\| \E \left[ 
      \frac{\z \z^\top}{\z^\top \A \z}
    \mid E \right] - \frac{\I}{\tr \A} \right\| 
    \le C \sqrt{\E \left[
      \frac{ (\z^\top \z)^2 }{ (\z^\top \A \z)^2 } 
    \mid E \right]} \Pr[E]^{-1/2} \sqrt{\frac{1}{r}}
  \]
  Applying \Cref{prop:inv-moment-high-prob}, in the $\epsilon = 1/2$ case
  we have
  \[
    \left\| \E \left[ 
      \frac{\z \z^\top}{\z^\top \A \z} \mid E
    \right] - \frac{\I}{\tr \A} \right\| 
    \le C \Pr[E]^{-1/2} \sqrt{\E[\|\z\|^4 \mid E]} r^{-3/2}
  \]
  The conclusion follows because sub-Gaussian implies bounded $4$th moments on
  $z_i$, so
  \begin{align*}
    \E[\|\z\|^4 \mid E]
    &= \E[\|\z\|^4 1_{E}] / \E[1_E]
    \leq \E[\|\z\|^4] / \E[1_E]
    = \Pr[E]^{-1} \sum_{i,j} \E[z_i^2 z_j^2]
    < \infty 
  \end{align*}

  The other result follows from using
  \Cref{prop:inv-moment-high-prob} with $\epsilon = \|\A\| / 2$.
\end{proof}

\bibliographystyle{alpha}
  \bibliography{pap}

\end{document}

### 2022-08-23 design/pseudoinverse.tex
Commit original repos

\documentclass[12pt]{sty/colt2019/colt2018-arxiv}
% \else
% \documentclass[anon,12pt]{sty/colt2019/colt2019} % Anonymized submission
% \documentclass[12pt]{colt2019} % Include author names
% \fi

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

% \title[Short Title]{Full Title of Article}
\title{Averaging subsampled pseudoinverses}
\usepackage{times}

\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{xfrac}
\usepackage{algorithmic}
\input{shortdefs}

\coltauthor{\Name{Micha{\l } Derezi\'{n}ski}
  \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics, UC Berkeley 
}

\begin{document}

\maketitle
For a square $n\times n$ matrix $\A$, we use $\adj(\A)$ to denote its adjugate, defined as an $n\times n$ matrix whose $(i,j)$th entry is $(-1)^{i+j}\det(\A_{-j,-i})$, where $\A_{-j,-i}$ denotes $\A$ without $j$th row and $i$th column. 
The adjugate matrix provides a key connection between the inverse and
the determinant because for any invertible matrix $\A$, we have
$\adj(\A) = \det(\A)\A^{-1}$. Also let $\pdet(\A)$ denote the pseudo-determinant.

Given a sequence of matrices $(\Z_i)_{i=1}^n$, we will use $\Z_{[S]}$ to
denote $\sum_{i\in S}\Z_i$ and let $\Z_{[n]}=\sum_{i=1}^n\Z_i$. 
Consider the following assumption, which is an
extension of the traditional \emph{general position} assumption.
\begin{assumption}
We say that a sequence of square rank-1 matrices $(\Z_i)_{i=1}^n$ is in
$r$-general position if for any subset $S\subseteq[n]$ of size at
least $r$ we have $\rank(\Z_{[S]}) = r$.
\end{assumption}

The fundamental result we use does not require the above assumption.
\begin{theorem}\label{t:adj}
  If $(\Z_i)_{i=1}^n$ are $d\times d$ matrices then for any $k\geq
  d-1$,
  \begin{align*}
    \sum_{S:|S|=k}\adj(\Z_{[S]})=   {n-d+1\choose k-d+1}\adj(\Z_{[n]}).
  \end{align*}
\end{theorem}

The following Theorem is a straightforward extension of the result
from \cite{unbiased-estimates-journal}, and follows as a corollary of
Theorem \ref{t:adj}.
   \begin{theorem}\label{t:sqinv}
Suppose that $(\Z_i)_{i=1}^n$ are $d\times d$ and in $d$-general
position. Then for any $k\geq d$, 
     \begin{align*}
\sum_{S:|S|=k} w_S \,\Z_{[S]}^{-1} =
       \tfrac{n-d+1}{k-d+1}\,\Z_{[n]}^{-1}\quad\text{and}\quad 
     \sum_{S:|S|=k}\!w_S = 1,\quad \text{for}\quad w_S=
\tfrac{\det(\Z_{[S]})}{{n-d\choose k-d}\det(\Z_{[n]})}.
     \end{align*}
   \end{theorem}
Note that we do not require that any matrices be psd, but we do
require invertibility. If the matrices are not psd, then some of the
weights $w_S$ will be negative, but they will still sum to 1.
Next, we relax the invertiblity assumption.

   \begin{theorem}\label{t:sqinv}
Suppose that $(\Z_i)_{i=1}^n$ are square and in $r$-general
position. Then for any $k\geq r$, 
     \begin{align*}
\sum_{S:|S|=k} w_S \,\tr(\Z_{[S]}^{\dagger}) =
       \tfrac{n-r+1}{k-r+1}\,\tr(\Z_{[n]}^{\dagger})\quad\text{and}\quad 
     \sum_{S:|S|=k}\!w_S = 1,\quad \text{for}\quad w_S=
\tfrac{\pdet(\Z_{[S]})}{{n-r\choose k-r}\pdet(\Z_{[n]})}.
     \end{align*}
   \end{theorem}
Note that trace could be replaced by any linear function of the eigenvalues
of the matrix. 
     \begin{proof}
Let $\X,\Y$ be $n\times d$ matrices such that
$\Z_{[S]}=\X_S^\top\Y_S$, where $\X_S$ selects out the rows of
$\X$. Note that $\L=\X\Y^\top$ has rank $r$, so there are $n\times r$
matrices $\A,\B$ such that $\Y\X^\top=\B\A^\top$. Moreover, since
       $\Y_S\X_S^\top=\L_{S,S}=\B_S\A_S^\top$ for any $S\subseteq [n]$, we have
       \begin{align*}
         \tr\big(\Z_{[S]}^\dagger\big) =
         \tr\big((\Y_S\X_S^\top)^\dagger\big) =
         \tr\big((\B_S\A_S^\top)^\dagger\big) = \tr\big((\A_S^\top\B_S)^{-1}\big).
       \end{align*}
       In particular, we also have $\tr\big((\X^\top\Y)^\dagger\big) =
       \tr\big((\A^\top\B)^{-1}\big)$. The result now follows by
       applying Theorem \ref{t:sqinv} to matrices $\tilde{\Z}_{[S]}=\A_S^\top\B_S$.
     \end{proof}

%\bibliographystyle{alpha}
\bibliography{pap}

\end{document}

### 2022-08-23 design/talk/Michal_Derezinski_Answers.tex
Commit original repos

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{fullpage}
% \usepackage[1in]{geometry}                		% See
%geometry.pdf to learn the layout options. There are lots.

%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{url}
%SetFonts

%SetFonts

\title{Improved guarantees and a multiple-descent curve for \\
  % the CSSP
  Column Subset Selection %Problem
  and the Nystr\"om method}
\author{Micha{\l} Derezi\'nski (speaker), Rajiv Khanna and Michael W. Mahoney}
 \date{}							% Activate to display a given date or no date

\begin{document}
\maketitle

\end{document}

### 2022-08-23 design/talk/Michal_Derezinski_Statement.tex
Commit original repos

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{fullpage}
% \usepackage[1in]{geometry}                		% See
%geometry.pdf to learn the layout options. There are lots.

%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{url}
%SetFonts

%SetFonts

\title{Improved guarantees and a multiple-descent curve for \\
  % the CSSP
  Column Subset Selection %Problem
  and the Nystr\"om method}
\author{Micha{\l} Derezi\'nski (speaker), Rajiv Khanna and Michael W. Mahoney}
 \date{}							% Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}

A key motivation for our work is interpretable data summarization,
which is a core challenge in reasoning 
about large datasets and machine learning models.  Data summarization
can be used, for example, to select a representative subset of gene
variants from a genetics dataset, or a collection of most informative
documents from a text database.  When data are represented
numerically, they are often described via matrices, in which case
linear algebra suggests a natural (and in a certain sense optimal) way
of performing data summarization: find the principal components
corresponding to the largest directions of variance.  These principal
components work well for black-box models that are evaluated only in
terms of prediction quality, but they are generally not interpretable
in terms of the domain from which the data are drawn. They do not
correspond to, say, a particular document or a gene variant, but
rather a complex mixture of them.  Neural networks, on which the
machine learning community increasingly relies, simply exacerbate this
problem.  A long-standing challenge has been to find summaries of data
which mimic the numerical properties of principal components and which
are also interpretable.

The cost of interpretability, when formulated this way, has strong
connections to the so-called Column Subset Selection Problem in
Randomized Numerical Linear Algebra (RandNLA). It
has been extensively studied in the literature, resulting in optimal
worst-case guarantees, developed for the first time nearly fifteen
years ago. These results are still relied upon as an important
technical tool in recent works, including the best paper at last
year’s International Conference on Machine Learning, which applied
them to Gaussian Process regression. The worst-case results suggest
that the cost of interpretability should become progressively higher
as the data summaries get larger.
To verify this, we carefully constructed a worst-case example for this problem; and we
showed that, for this example, the cost of interpretability does reach
the worse-case bound for one size of the summary, but then it rapidly
drops down for both larger and smaller sizes. As it turns out, in this and other
worst-case examples, the high cost of interpretability is only a
corner case.

% We empirically tested this claim on
% a worst-case example taken from the literature. In this example, the
% cost of interpretability does reach the bound for one size of the
% summary, but then it rapidly drops down for both larger and smaller
% sizes. So, even in the worst-case example, high cost is only a corner
% case.

To explain this, we go beyond worst-case analysis, and our results are
able to accurately capture the non-linear behavior of the cost of
interpretability. Specifically, we conclude that, except for certain
pathological examples which we can characterize, the cost of
interpretability is far smaller than suggested by prior work, and it
is often negligible for real-world problems. To construct our
interpretable summaries, we use a randomized subset selection
technique based on Determinantal Point Processes (DPPs), which provide
%technique called Determinantal Point Processes, which is 
a probabilistic model of diversity that has emerged across many scientific domains.
DPPs were first discovered in physics as a model of fermions in thermal
equilibrium, and they have since been used in random matrix theory, graph
theory and quantum mechanics. More recently, thanks to the emergence
of efficient algorithms for DPP sampling, they have also become
popular in RandNLA and Machine Learning.

Our analysis reveals that
the previously observed worst-case examples are part of a larger
phenomenon, which we call the multiple-descent curve in data
summarization.  This curve represents a family of phase transitions,
observed through a spike in the cost of interpretability, which occur
when the data exhibit an underlying hierarchical structure. The
multiple-descent curve can be observed not only in artificially
constructed pathological examples but also in real-world
problems. However, it can generally be avoided by tuning model
parameters.

This phenomenon is very much reminiscent of the so-called double
descent curve, which is exhibited by many machine learning models
including deep neural networks. In the context of double descent, we
distinguish two regimes of machine learning: the "classical" regime,
where we have an abundance of training data relative to the number of
model parameters we wish to learn; and the "modern" regime, which
includes deep learning architectures, where there is far more
parameters than training data. While each regime allows for
constructing machine learning models that perform well on unseen data,
the phase transition between the two regimes may lead to a spike in
the error rate, resulting in poor performance.

Both the double descent curve in machine learning, and the new
multiple-descent curve in data summarization, shown in our work, are
related to fundamental phase transitions observed in the behavior of
high-dimensional random matrices. Obtaining precise characterizations
of these phenomena is of crucial importance to understanding modern
machine learning as well as algorithmic-statistical tradeoffs that are
central to the foundations of data science.
\vspace{2mm}

\noindent
\textbf{Media Contact:}
Micha{\l} Derezi\'nski\\
\underline{Email:} \texttt{mderezin@berkeley.edu}\\
\underline{Phone:} 408-680-4652
%\vspace{5mm}
\subsection*{Author Bios} 
\begin{minipage}{.7\textwidth}
  \underline{Full name:} \textbf{Micha{\l} Derezi\'nski}\\
\underline{Title:} Postdoctoral Researcher\\
\underline{Organization:} University of California, Berkeley\\
\underline{Bio:} Michał Dereziński is a Postdoctoral Researcher in the
Department of Statistics at University of California,
Berkeley. Previously, he was a research fellow at the Simons Institute
for the Theory of Computing. He obtained his Ph.D. in Computer Science
at University of California, Santa Cruz, advised by professor Manfred
Warmuth. Michał's research is focused on developing scalable
randomized algorithms with robust statistical guarantees for machine
learning, data science and optimization. More information is available
at: \url{https://users.soe.ucsc.edu/~mderezin/}.\\
\underline{Headshot:} \url{https://users.soe.ucsc.edu/~mderezin/photo.jpg}
\end{minipage}\hfill
\begin{minipage}{.25\textwidth}
  \includegraphics[width=\textwidth]{figs/photo-michal.jpg}
\end{minipage}
\vspace{5mm}

\noindent
\begin{minipage}{.7\textwidth}
\underline{Full Name:} \textbf{Rajiv Khanna}\\
\underline{Title:} Postdoctoral Researcher\\
\underline{Organization:} University of California, Berkeley\\
\underline{Bio:} Rajiv Khanna is a Postdoctoral Researcher at the
Foundations of Data Analysis Institute at University of California,
Berkeley working with Michael Mahoney. He graduated with a PhD from UT
Austin advised by Professors Joydeep Ghosh, and Alex Dimakis.  His
research interests include theoretical aspects of optimization and
more recently generalization in neural networks.
\end{minipage}
\hfill
\begin{minipage}{.25\textwidth}
  \includegraphics[width=\textwidth]{figs/photo-rajiv.jpg}
\end{minipage}
\vspace{5mm}

\noindent
\begin{minipage}{.7\textwidth}
\underline{Full name:} \textbf{Michael W. Mahoney}\\
\underline{Title:} Associate Professor\\
\underline{Organization:} International Computer Science Institute and
University of California at Berkeley\\
\underline{Bio:} Michael W. Mahoney is at the University of California
at Berkeley in the Department of Statistics and at the International
Computer Science Institute (ICSI).  He works on algorithmic and
statistical aspects of modern large-scale data analysis.  Much of his
recent research has focused on large-scale machine learning, including
randomized matrix algorithms and randomized numerical linear algebra,
geometric network analysis tools for structure extraction in large
informatics graphs, scalable implicit regularization methods, and
applications in genetics, astronomy, medical imaging, social network
analysis, and internet data analysis.  He received him PhD from Yale
University with a dissertation in computational statistical mechanics,
and he has worked and taught at Yale University in the mathematics
department, at Yahoo Research, and at Stanford University in the
mathematics department.  Among other things, he is on the national
advisory committee of the Statistical and Applied Mathematical
Sciences Institute (SAMSI), he was on the National Research Council's
Committee on the Analysis of Massive Data, he co-organized the Simons
Institute's fall 2013 and 2018 programs on the foundations of data
science, and he runs the biennial MMDS Workshops on Algorithms for
Modern Massive Data Sets.  He is currently the Director of the
NSF/TRIPODS-funded FODA (Foundations of Data Analysis) Institute at UC
Berkeley.  He holds several patents for work done at Yahoo Research
and as Lead Data Scientist for Vieu Labs, Inc., a  startup reimagining
consumer video for billions of users.  More information is available
at: \url{https://www.stat.berkeley.edu/~mmahoney/}.\\
\underline{Headshot:} \url{https://www.stat.berkeley.edu/~mmahoney/misc/mwm18-big.png}
 \end{minipage}\hfill
\begin{minipage}[t]{.25\textwidth}
  \includegraphics[width=\textwidth]{figs/photo-mwm.png}
\end{minipage}

\end{document}

### 2022-08-23 design/talk/Michal_Derezinski_Statement_new.tex
Commit original repos

--- 2022-08-23 design/talk/Michal_Derezinski_Statement.tex
@@
 %\subsection{}
+\vspace{-5mm}
+
+Thank you \textbf{very} much. We are \textbf{really} excited to receive this award. My
+name is Micha{\l}~Derezi\'nski, I am a postdoctoral researcher at UC Berkeley.
 
@@
 linear algebra suggests a natural (and in a certain sense optimal) way
-of performing data summarization: find the principal components
+of performing data summarization: namely, finding the principal components
 corresponding to the largest directions of variance.  These principal
@@
 machine learning community increasingly relies, simply exacerbate this
-problem.  A long-standing challenge has been to find summaries of data
+problem of non-interpretability.  A long-standing challenge has been to find summaries of data
 which mimic the numerical properties of principal components and which
@@
 connections to the so-called Column Subset Selection Problem in
-Randomized Numerical Linear Algebra (RandNLA). It
+Randomized Numerical Linear Algebra. It
 has been extensively studied in the literature, resulting in optimal
@@
 To verify this, we carefully constructed a worst-case example for this problem; and we
-showed that, for this example, the cost of interpretability does reach
-the worse-case bound for one size of the summary, but then it rapidly
+showed that, for this example, the cost of interpretability \textbf{does} reach
+the worse-case bound for \textbf{one} size of the summary, but then it rapidly
 drops down for both larger and smaller sizes. As it turns out, in this and other
@@
 interpretable summaries, we use a randomized subset selection
-technique based on Determinantal Point Processes (DPPs), which provide
+technique based on Determinantal Point Processes (DPPs for short), which provide
 %technique called Determinantal Point Processes, which is 
@@
 of efficient algorithms for DPP sampling, they have also become
-popular in RandNLA and Machine Learning.
+popular in Randomized Numerical Linear Algebra and Machine~Learning.
 
@@
 central to the foundations of data science.
+\vfill
+\textbf{Summary}:
+Interpretable data summarization is a core challenge in reasoning
+about large datasets and machine learning models.  Data summarization
+can be used, e.g., to select a representative subset of gene variants
+from a genetics dataset, or a collection of most informative documents
+from a text database.  When data are represented numerically, they are
+often described via matrices, in which case linear algebra suggests a
+natural (and in certain ways optimal) way of performing data
+summarization: find the principal components corresponding to the
+largest directions of variance.  These principal components work well
+for black-box models that are evaluated only in terms of prediction
+quality, but they are generally not interpretable in terms of the
+domain from which the data are drawn. They do not correspond to, say,
+a particular document or a gene variant, but rather a complex mixture
+of them.  Neural networks, on which the machine learning community
+increasingly relies, simply exacerbate this problem.  A long-standing
+challenge has been to find summaries of data which mimic the numerical
+properties of principal components and which are also
+interpretable. The added cost of this interpretability can be
+significant, as shown by prior worst-case analysis.  In our work, we
+exploit the inherent structure of data to show that, except for
+pathological examples which we characterize, the cost of
+interpretability is far smaller than suggested by prior worst-case
+analysis, and it is often negligible for real-world problems.  Our
+analysis reveals an intriguing new phenomenon, which we call the
+multiple-descent curve in subset selection.  This is a phase
+transition, observed through a spike in the cost of interpretability,
+which occurs when the data exhibit an underlying hierarchical
+structure.  The multiple-descent phenomenon can be observed in
+practice, but it can also be avoided by tuning model parameters.
+
+\newpage
+\begin{itemize}
+\item Interpretability is important, but Modern ML models are
+  black boxes.
+  \begin{quote}
+    Obtaining interpretable data summaries is an
+    important challenge, especially in the context of modern ML and
+    Deep Learning models, which are often black boxes.
+  \end{quote}
+  \begin{quote}
+    Our work suggests that obtaining interpretable summaries can be
+    done effectively, if we ignore pathological problem instances,
+    which do not occur in the real-world.
+  \end{quote}
+\item Going beyond worst-case analysis to understand different phases
+  of learning.
+  \begin{quote}
+Theoretical work in ML and Data Science, particularly involving
+techniques from Randomized Numerical Linear Algebra, often focuses on
+the worst-case examples.
+\end{quote}
+\begin{quote}
+  Worst-case analysis does not capture the
+fundamental phase transitions associated with learning, such as
+double-descent and multiple-descent.
+\end{quote}
+\begin{quote}
+Using techniques from statistical physics and random matrix theory,
+such as Determinantal Point Processes, we can go beyond the worst-case
+analysis and obtain precise characterizations of these phenomena.
+\end{quote}
+\item The paper that “forced Timnit Gebru out of Google”
+  \begin{quote}
+    I don’t know the specifics of the situation with Timnit, so I
+    can’t comment on that. However, it's worth noting that Timnit's work underscores the
+    importance of interpretability in machine learning, particularly when
+    it comes to the implicit bias of machine learning models trained
+    on large datasets. Our work suggests that the cost of interpretability is small, even
+    negligible, for many real-world applications.
+  \end{quote}
+\item Connection to text summarization
+  \begin{quote}
+    While our work is foundational, and not specific to, say natural
+    language processing, the method we analyze in this paper, namely
+    Determinantal Point Processes, has been used in a number of
+    concrete tasks such as document summarization and video
+    summarization, and has proven empirically effective. Our
+    theoretical analysis provides a better understanding of these
+    empirical results.
+  \end{quote}
+  \end{itemize}
 \vspace{2mm}

### 2022-08-23 design/talk/asymptotic.tex
Commit original repos

\documentclass[10pt]{beamer}
%\beamertemplateshadingbackground{brown!70}{yellow!10}
\mode<presentation>
{
  %\usetheme{Warsaw}
  \usecolortheme{crane}
  % or ...

%  \setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}[frame number]{}
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=newest}
\usepackage[utf8]{inputenc}
\usetikzlibrary{patterns}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{colortbl}
%\usepackage{multicol}
\usepackage{cancel}
\usepackage{ulem}
\usepackage{multirow}
\usepackage{relsize}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{forloop}% http://ctan.org/pkg/forloop
\newcounter{loopcntr}
\newcommand{\rpt}[2][1]{%
  \forloop{loopcntr}{0}{\value{loopcntr}<#1}{#2}%
}
%\pagestyle{plain}
%\input{defs2}
\input{../shortdefs}

\edef\polishl{\l}
\setlength{\columnsep}{0.7em}
\setlength{\columnseprule}{0mm}
\setlength{\arrayrulewidth}{1pt}

\newcommand{\svr}[1]{{\textcolor{darkSilver}{#1}}}
\definecolor{brightyellow}{cmyk}{0,0,0.7,0.0}
\definecolor{lightyellow}{cmyk}{0,0,0.3,0.0}
\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
\definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}
\AtBeginSection[]
{
\begin{frame}<beamer>
\frametitle{Outline}
\tableofcontents[currentsection]
\end{frame}
}
\def\layersep{2.5cm}

%  \fboxsep=3pt
% %\fboxsep=0mm%padding thickness
% \fboxrule=2pt%border thickness

\setkeys{Gin}{width=0.7\textwidth}

\title[]{Strong asymptotic equivalence}

\author[]{Micha{\l} Derezi\'{n}ski\\
UC Berkeley}

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\linespread{1.3}

\begin{frame}
  \frametitle{Standard asymptotic equivalence}
  Matrix sequences $\A_n$, $\B_n$ (deterministic or random) of growing dimensions are
  \emph{equivalent}, denoted by
  \begin{align*}
    \A_n\asymp \B_n,
  \end{align*}
  if
  \begin{align*}
    \lim_{n\rightarrow\infty}\big|\tr\,\C_n(\A_n-\B_n)\big| =0,
  \end{align*}
  almost surely, for any sequence $\C_n$ of not necessarily symmetric matrices with bounded trace
  norm, i.e., such that
  \begin{align*}
    \lim\sup\|\C_n\|_{\tr}<\infty.
  \end{align*}
  Here, the trace norm is defined by $\|\M\|_{\tr} = \tr((\M^\top\M)^{\frac12})$.
\end{frame}

\begin{frame}
  \frametitle{\emph{New:} Strong asymptotic equivalence}
  Positive semi-definite matrix sequences $\A_n$, $\B_n$ (deterministic or random) of growing dimensions are
  \emph{strongly equivalent}, denoted by
  \begin{align*}
    \A_n\simeq \B_n,
  \end{align*}
  if
  \begin{align*}
    \A_n^{\dagger/2}\B_n\A_n^{\dagger/2} \asymp
    \B_n^{\dagger/2}\A_n\B_n^{\dagger/2}\asymp \I_n,
  \end{align*}
  where $\I_n$ is the sequence of identity matrices with matching
  dimensions, and $\M^{\dagger/2}=(\M^\dagger)^{\frac12}$ is the
  square-root of the Moore-Penrose inverse.\\[5mm]
  \emph{Alternative condition:} \\
  Instead of relying on standard equivalence, we can use the spectral norm:
  \begin{align*}
    \lim_{n\rightarrow\infty}\|\A_n^{\dagger/2}\B_n\A_n^{\dagger/2} -
    \I_n\|
    =
        \lim_{n\rightarrow\infty}\|
    \B_n^{\dagger/2}\A_n\B_n^{\dagger/2} - \I_n\|
    = 0.
  \end{align*}
\end{frame}

\begin{frame}
  \frametitle{Desired properties of strong equivalence}
  \begin{enumerate}
  \item Strong equivalence implies standard equivalence, i.e.:
    \begin{align*}
      \text{if}\quad \A_n\simeq\B_n \quad\text{then}\quad
      \A_n\asymp\B_n
      \end{align*}
  \item if $\A_n$ and $\B_n$ have uniformly bounded condition numbers,
    then equivalence implies strong equivalence\\[3mm]
  \item Some form of calculus of deterministic equivalents applies to
    the strong equivalence relation.
  \end{enumerate}
  \vspace{3mm}

  \emph{Note:} Strong equivalence differs from standard equivalence
  when sequences $\A_n$ and $\B_n$ have vanishing eigenvalues.
\end{frame}

\begin{frame}
  \frametitle{Some examples}

Let $\X_n\sim\Nc(\zero,\I_n\otimes\Sigmab_n)$. We can consider various sequences:
\begin{align*}
  \A_n &= \E[\X_n^\dagger\X_n],\\
  \A_n &= \E[(\tfrac1n\X_n^\top\X_n + \lambda_n\I)^{-1}],\quad\text{where}\quad
  \lambda_n\rightarrow 0,\\
  \A_n &= \E[(\tfrac1n\X_n^\top\X_n + \lambda_n\I)^{-p}],\quad\text{where}\quad
  \lambda_n\rightarrow 0
\end{align*}
\emph{Note:} Standard deterministic equivalents for $\A_n$ are $\B_n=(\gamma_n\Sigmab + \delta_n\I)^{-p}$, for the right choices
of $\gamma_n,\delta_n$\\[3mm]

However, they will not satisfy the \emph{strong} equivalence when, e.g.:
\begin{align*}
  \Sigmab_n =
  \begin{pmatrix}
    \I_n&\zero\\
    \zero & \frac1n\I_n
  \end{pmatrix}.
\end{align*}
\emph{Question:} How do standard and strong equivalences differ when
$\Sigmab_n$ exhibits some form of polynomial or exponential spectral decay?

\end{frame}

\end{document}

### 2022-08-23 design/talk/talk-nystrom-script.tex
Commit original repos

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[margin=.5in]{geometry}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[sc]{mathpazo}
\newcommand{\HRule}{\rule{\linewidth}{0.5mm}}
\newcommand{\Hrule}{\rule{\linewidth}{0.3mm}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\makeatletter% since there's an at-sign (@) in the command name
\renewcommand{\@maketitle}{%
  \parindent=0pt% don't indent paragraphs in the title block
  \centering
  {\Large \bfseries\textsc{\@title}}
  \HRule\par%
  \textbf{\@author \hfill \@date}
  \par
}
\def\b{{~\textbullet~}}
\makeatother% resets the meaning of the at-sign (@)

\title{Nystr\"om Talk Script}
\author{Micha{\l} Derezi\'{n}ski}
\date{\texttt{mderezin@berkeley.edu}}

\begin{document}
  \maketitle% prints the title block
\setlength{\parskip}{0.5\baselineskip}
\setlength{\parindent}{0pt}

Hi, my name is Micha{\l}, and I will be talking about an intriguing
phenomenon, which we refer to as the multiple-descent curve in subset
selection. This is joint work with Rajiv and Michael.

A core motivation for the tasks we are studying is interpretable
dimensionality reduction. As an example, consider a biological
application, with a large genetics dataset where each individual is
described by a sequence of genetic variations, called Single
Nucleotide Polymorphisms, SNPs for short. Prior work has shown that
such data can be effectively characterized by a subset of SNPs, chosen
so that it is correlated with the top Principal Components of the data
matrix. This subset of SNPs can then be used by domain experts to
extract insight from the data. Many other applications of
interpretable dimensionality reduction have been demonstrated in the
literature, including for example document clustering.

The two basic dimensionality reduction tasks we consider in this work
are Column Subset Selection and the
Nystrom method. Those two tasks are very closely related, even though they often
appear independently in the literature. The first one, on the left, is
most commonly described as feature selection from a data matrix A,
although we could just as well consider a setting where the columns of
A are the instances instead of features. For the Nystrom method, on
the right, the matrix in question is a square kernel matrix, and we
wish to represent it by a subset of entries. This task arises in the
context of kernel machines, Gaussian Process regression and
Independent Component Analysis, among others. As we demonstrate in the
paper, an equivalence between these tasks can be established, which
means that all of our results apply in both settings.

In low-rank approximation our goal is to project data onto a
low-dimensional subspace, while minimizing the projection
error. Without additional constraints, there is a unique choice to
minimize this error, by projecting onto the top k principal components
in the data. However, typically these principal component vectors will
not correspond to any interpretable feature or instance, which is why
we may want to\b choose a subspace spanned by a column subset of the
data matrix. Naturally, there are other notions of interpretability
than selecting subsets of the data matrix, however this model provides
a useful framework for analyzing, what can be described as, the cost
of interpretability. We measure this cost via the so-called
approximation factor, comparing the error of the best interpretable
representation, relative to the optimal non-interpretable choice given
by the principal components, for a given dimensionality parameter k.

A large body of literature has been dedicated to obtaining worst-case
bounds on the approximation factor that describes the cost of
interpretability. In particular, it was shown, for 
Column Subset Selection by Deshpande, Rademacher, Vempala and Wang,
and for the Nystrom method by Belabbas and Wolfe, that the cost of
interpretability in rank k approximation can always be bounded by a
factor of k+1. These results are still used as a technical tool
in recent works, including the best paper at the 2019 International
Conference on Machine Learning, which
applied them to sparse variational Gaussian Process regression. The
above bound is in fact worst-case optimal in that there are problem
instances for which factor k+1 cannot be improved. Nevertheless, these
results are still somewhat unsatisfying because they suggest that the
cost of interpretability gets progressively higher as the rank
parameter k grows, as seen in the plot below, which is not generally
observed in practice. To understand this,\b we empirically tested the
approximation factor for the worst-case example given by Deshpande et
al. As we can see, the factor does reach the worst-case bound for one
value of k, but then it rapidly drops down in either direction. So,
even for the worst-case example, a high approximation factor is only a
corner case. To explain this,\b we go beyond worst-case analysis, and
our bounds are able to accurately capture the non-linear behavior of
the approximation factor.

In particular, our new guarantees show that the worst-case example is
part of a larger phenomenon, which we refer to as the multiple-descent
curve in subset selection. This curve is illustrated in the
following plot, both empirically and through our theory. As we can
see, the approximation factor exhibits occasional sharp spikes, but is
otherwise very small.  Our new lower bound shows that more than one
spike can occur for one data matrix, which was not the case for the
previously demonstrated worst-case constructions. Moreover, we show
that the spikes are correlated with sharp drops in the data spectrum.

The multiple-descent curve in subset selection is very much
reminiscent of the so-called double descent curve, proposed by Belkin
et al, which is exhibited by the generalization error of many machine
learning models, when varying the ratio between the number of
parameters and the training data size. In this context, we can
distinguish two regimes: the Classical one, where we have more data
than parameters, and the Modern one, which includes deep learning
architectures, where there is more parameters than data and so the
model can interpolate the training set. The phase transition between
the two regimes exhibits a peak in the generalization
error. Naturally, there are important differences compared to our
setting: first, we consider a combinatorial optimization task of
subset selection, as opposed to a statistical learning one, and
second, we study the approximation factor instead of the
generalization error. However, several connections arise.\b Similarly as
in the case of multiple-descent in subset selection, the double
descent peak is correlated with the condition number of the data, and
moreover, recent works showed that the generalization error can also
exhibit multiple peaks. Finally, techniques similar to ours can be
used to obtain exact expressions for double descent in terms of the
mean squared error of the minimum norm interpolating linear model,
which is actually what we used to generate the plot shown on the
right.

So, what properties of the data affect the non-linear behavior of the
cost of interpretability? For example, 
do we expect the cost to be smaller if the data
is more or less uniformly distributed in the domain space, as in
Dataset 1, or if it has some low-dimensional structure, as in Dataset 2?
Well, first, we need to be able to easily characterize such properties of the data.\b For
this we can use the so-called ``spectral decay profile'', which looks at
the magnitude of the largest directions of variance in the data, represented
mathematically as the singular
values of the data matrix. To illustrate the structure of the dataset through spectral decay, we plot
the top singular values in decreasing order. For example, for the first dataset, we see a relatively
flat decay with multiple singular values of comparable
magnitude, because there are many directions with large
variance, whereas for the second dataset, there is one or two large singular values
followed by a rapid decay. This corresponds to the
low-dimensional structure we can see in the top right.

There are several types of spectral decays which are commonly
observed in real world datasets, such as a flat decay, a sharp
drop, and a  range of smooth decays with polynomial or
exponential rates. In fact, often we may be able to anticipate
the type of spectral decay even before we look at the data, for
example because of some prior knowledge coming from domain
experts about the underlying data distribution, and in some cases
we can actually control it through model selection, by choosing
an apprioriate similarity kernel.

In this work, we used the spectral decay profile of the data to
provide a complete characterization of the cost of interpretability,
showing that if the data exhibits sufficiently 
smooth spectral decay, then the spikes in the multiple-descent curve
can be avoided. In particular, this  
allows us to deduce that the spikes in the cost of
interpretability are directly correlated with sharp drops in the
spectrum of the data. This also implies that when the spectral decay is
sufficiently smooth, for example, when it exhibits a polynomial
rate which is commonly observed in practice, then we can prove
that there is no spikes and the cost of interpretability is
small for all values of the dimensionality parameter,
significantly improving on the prior work.

Of course, the practical performance of dimensionality reduction depends on
the subset selection method used. Here, we focus primarily on the method called
Determinantal Point Process sampling. % , although in the paper we
% empirically demonstrate that the multiple-descent curve is also
% exhibited by another commonly used method, called greedy subset
% selection.
A Determinantal Point Process is a non-i.i.d. randomized
method that uses negative correlation to induce diversity in the
sampled subset. DPPs have found many applications in machine
learning for the past 15 years or so, including in recommendation
systems, experimental design and stochastic optimization, and they can
be implemented very efficiently, with runtimes faster than singular
value decomposition. For a state-of-the-art implementation, check out
a paper by myself, Daniele Calandriello and Michal
Valko from the last year's Neurips conference. And to learn more about
related applications of DPPs, see a new 
survey by myself and Michael Mahoney, as well as an earlier monograph
by Kulesza and Taskar.

In our empirical results, we showed that the multiple-descent
phenomenon can be observed not only in worst-case 
examples but also in real-world subset selection tasks. Here, we
demonstrate this for the Nystrom method with the Gaussian RBF kernel,
studying the effect of varying the RBF parameter. On the top plots, we
show the Nystrom approximation factor for two Libsvm datasets whereas
at the bottom we compare it to decreasing singular values of the
data. We can clearly see that the spikes in approximation factor align
with big drops in the spectral decay, and that this behavior is
greatly influenced by the value of the RBF parameter. Decreasing this
parameter results in a smoother spectral decay, which in turn
eliminates the spikes, resulting in a uniformly small approximation
factor. The right choice of the RBF parameter ultimately depends on
the down-stream machine learning task, so depending on the structure
of the problem, we may end up in the regime where the
approximation factor exhibits a multiple-descent curve or in the
regime where it is relatively flat. Here, I showed the results when using a
Determinantal Point Process for subset selection, however, remarkably,
we can observe the same phenomenon even if we switch to a completely
different method, called Greedy Subset Selection.\b This suggests that
the multiple-descent curve is not specific to DPPs, but rather
it is a method-agnostic phenomenon.

So, to conclude, in this work we studied the cost of interpretability
in dimensionality reduction. We particularly focused on two methods of
dimensionality reduction, namely, Column Subset Selection and the
Nystrom method, both of which use data subsets to construct small
interpretable representations of large datasets. In our main results,
we overcome the limitations of prior worst-case analysis to accurately
capture the non-linear behavior of the cost of interpretability. Moreover, our
analysis reveals that the previously observed worst-case examples are
part of a larger phenomenon, which we call the multiple-descent curve
in subset selection. This multiple-descent curve can be observed not
only in artificially constructed pathological examples but also in
real-world problems. However, it can generally be avoided by tuning
model parameters.

Here is a complete list of references from the talk, and thank you
very much for your attention!

\newpage

\end{document}

### 2022-08-23 ftvi/ftvi_aistats/ftvi_aistats.tex
Commit original repos

--- 2022-08-23 ftvi/ftvi_icml/ftvi_icml.tex
@@
-%%%%%%%% ICML 2022 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
-\documentclass[nohyperref]{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
-\usepackage{graphicx}
-%\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2022} with \usepackage[nohyperref]{icml2022} above.
-\usepackage{hyperref}
-
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-% \usepackage{icml2022}
-
-% If accepted, instead use the following line for the camera-ready submission:
-\usepackage[accepted]{icml2022}
-
-% For theorems and such
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{mathtools}
-    \usepackage{amsthm}
-
-% if you use cleveref..
-\usepackage[capitalize,noabbrev]{cleveref}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% THEOREMS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
+\documentclass[twoside]{article}
+
+\usepackage{aistats2022}
+%\usepackage[accepted]{aistats2022}
+% If your paper is accepted, change the options for the package
+% aistats2022 as follows:
+%
+%\usepackage[accepted]{aistats2022}
+%
+% This option will print headings for the title of your paper and
+% headings for the authors names, plus a copyright note at the end of
+% the first column of the first page.
+
+% If you set papersize explicitly, activate the following three lines:
+%\special{papersize = 8.5in, 11in}
+%\setlength{\pdfpageheight}{11in}
+%\setlength{\pdfpagewidth}{8.5in}
+
+% If you use natbib package, activate the following three lines:
+\usepackage[round]{natbib}
+\renewcommand{\bibname}{References}
+\renewcommand{\bibsection}{\subsubsection*{\bibname}}
+
+% If you use BibTeX in apalike style, activate the following line:
+\bibliographystyle{apalike}
+
+\usepackage{amssymb,amsmath,amsthm}
+\newtheorem{theorem}{Theorem}
 \theoremstyle{definition}
@@
 \newtheorem{assumption}[theorem]{Assumption}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{claim}[theorem]{Claim}
+\newtheorem{example}[theorem]{Example}
+\newtheorem{corollary}[theorem]{Corollary}
 \theoremstyle{remark}
 \newtheorem{remark}[theorem]{Remark}
-
-% Todonotes is useful during development; simply uncomment the next line
-%    and comment out the line below the next line to turn off comments
-%\usepackage[disable,textsize=tiny]{todonotes}
-\usepackage[textsize=tiny]{todonotes}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Fat--Tailed VI with Anisotropic Tail Adaptive Flows}
-
-% custom stuff
 \usepackage[bold]{hhtensor}
+
+\usepackage{hyperref}
+\usepackage{cleveref}
 \usepackage{mathtools}
+\usepackage{booktabs}
 \usepackage{float}
@@
 \usepackage{enumitem}
+
+\crefname{assumption}{assumption}{assumptions}
+\Crefname{assumption}{Assumption}{Assumptions}
+\crefname{proposition}{proposition}{propositions}
+\Crefname{proposition}{Proposition}{Propositions}
+\crefname{lemma}{lemma}{lemmas}
+\Crefname{lemma}{Lemma}{Lemmas}
+\crefname{remark}{remark}{remarks}
+\Crefname{remark}{Remark}{Remarks}
+
 \input{../macros}
@@
 
+% If your paper is accepted and the title of your paper is very long,
+% the style will print as headings an error message. Use the following
+% command to supply a shorter title of your paper so that it can be
+% used as headings.
+%
+%\runningtitle{I use this title instead because the last one was very long}
+
+% If your paper is accepted and the number of authors is large, the
+% style will print as headings an error message. Use the following
+% command to supply a shorter version of the authors names so that
+% they can be used as headings (for example, use only the surnames)
+%
+%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
+\runningtitle{Fat-Tailed VI with Anisotropic Tail Adaptive Flows}
+
 \twocolumn[
-\icmltitle{Fat--Tailed Variational Inference with Anisotropic Tail Adaptive Flows}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2022
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Feynman Liang}{yyy,fb}
-\icmlauthor{Liam Hodgkinson}{yyy,zzz}
-\icmlauthor{Michael W. Mahoney}{yyy,zzz}
-% \icmlauthor{Firstname4 Lastname4}{sch}
-% \icmlauthor{Firstname5 Lastname5}{yyy}
-% \icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
-% \icmlauthor{Firstname7 Lastname7}{comp}
-% %\icmlauthor{}{sch}
-% \icmlauthor{Firstname8 Lastname8}{sch}
-% \icmlauthor{Firstname8 Lastname8}{yyy,comp}
-% %\icmlauthor{}{sch}
-% %\icmlauthor{}{sch}
-\end{icmlauthorlist}
-
-\icmlaffiliation{fb}{Meta, Menlo Park, CA}
-\icmlaffiliation{yyy}{Department of Statistics, University of California, Berkeley, CA}
-\icmlaffiliation{zzz}{International Computer Science Institute, Berkeley, CA}
-% \icmlaffiliation{comp}{Company Name, Location, Country}
-% \icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}
-
-\icmlcorrespondingauthor{Feynman Liang}{feynman@berkeley.edu}
-%\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+
+\aistatstitle{Fat--Tailed Variational Inference \\ with Anisotropic Tail Adaptive Flows}
+
+\aistatsauthor{ Feynman Liang \And Liam Hodgkinson \And Michael W. Mahoney }
+
+\aistatsaddress{ Department of Statistics\\UC Berkeley \And ICSI and Department of Statistics\\UC Berkeley \And ICSI and Department of Statistics\\UC Berkeley } ]
+
+% \maketitle
 
 \begin{abstract}
-    While fat-tailed densities commonly arise as posterior and marginal distributions in
-    robust models and scale mixtures, they present challenges when Gaussian-based 
-    variational inference fails to capture tail decay accurately. 
-    We first improve previous theory on tails of Lipschitz flows 
-    %\citep{jaini2020tails} 
-    by quantifying how the tails affect the \emph{rate} of tail decay 
-    and by expanding the theory to non-Lipschitz polynomial flows.
-    We then develop an alternative theory for multivariate tail parameters which is 
-    sensitive to tail-anisotropy. 
-    In doing so, we unveil a fundamental problem which plagues many existing flow-based 
-    methods: they can only model tail-isotropic distributions (i.e., distributions 
-    having the same tail parameter in every direction).
-    To mitigate this and enable modeling of tail-anisotropic targets, we propose 
-    anisotropic tail-adaptive flows (ATAF).
-    Experimental results on both synthetic and real-world targets confirm that ATAF 
+    While fat-tailed densities commonly arise as posterior and marginal
+    distributions in robust models and scale mixtures, they present a
+    problematic scenario where Gaussian-based variational inference fails
+    to accurately capture tail decay. We first improve
+    previous theory on tails of Lipschitz flows \citep{jaini2020tails} by
+    quantifying how they affect the \emph{rate} of tail decay and
+    expanding the theory to non-Lipschitz polynomial flows.
+    Next, we develop an alternative theory for multivariate tail parameters which is sensitive to tail-anisotropy. In doing so, we unveil a
+    fundamental problem which plagues many existing flow-based methods:
+    they can only model tail-isotropic distributions (i.e., distributions having the same tail parameter in every direction).
+    To mitigate this and enable modeling of
+    tail-anisotropic targets, we propose anisotropic tail-adaptive flows (ATAF).
+    Experimental results confirm ATAF on both synthetic and real-world
+    targets
     is competitive with prior work while also exhibiting appropriate tail-anisotropy.
@@
 
-Flow-based methods 
+Flow based methods 
 \citep{papamakarios2021normalizing}
@@
 These methods start with a random variable $X$ having a simple and tractable
-distribution $\mu$, and then apply a learnable transport map $f_\theta$ to build
+distribution $\mu$, and apply a learnable transport map $f_\theta$ to build
 another random variable $Y = f_\theta(X)$ with a more expressive \emph{pushforward}
 probability measure $(f_\theta)_\ast \mu$ \citep{papamakarios2021normalizing}.
-In contrast to the implicit distributions \citep{huszar2017variational} produced by generative adversarial networks (GANs), flow-based methods restrict the transport map $f_\theta$ to be invertible and to have efficiently-computable Jacobian determinants.
+In contrast to the implicit distributions \citep{huszar2017variational} produced by generative adversarial networks (GANs), flow based methods restrict the transport map $f_\theta$ to be invertible and to have efficiently-computable Jacobian determinants.
 As a result, probability density functions can be tractably computed
@@
         \left.\frac{d f_\theta^{-1}(z)}{dz} \right\vert_{z=y}
-      \right\rvert .
+      \right\rvert
 \end{align}
@@
 However, in \Cref{thm:distn_class_closed}, we show this choice results in significant
-restrictions on the expressivity of the model, limiting its utility for data that
+restrictions to the expressivity of the model, limiting its utility for data that
 exhibits fat-tailed (or heavy-tailed) structure.
 Prior work addressing heavy-tailed flows \citep{jaini2020tails}
-are limited to tail-isotropic base distributions.
-In \Cref{prop:isotropic-pushforward}, we prove flows built on these base distributions
-are unable to model accurately multivariate anisotropic fat-tailed structure.
+are limited to tail-isotropic base distributions ---
+in \Cref{prop:isotropic-pushforward}, we also prove flows built on these base distributions
+are unable to accurately model multivariate anisotropic fat-tailed structure.
 % \michael{I'm a little confused by this paragraph.
@@
   \includegraphics{../Figures/pancake.pdf}
-    \vspace{-5mm}
+    \vspace{-3mm}
   \caption{
@@
     In contrast, ADVI's (top right) Gaussian base distribution and TAF's (bottom left) tail-isotropic $\prod_{i=1}^2 \text{StudentT}(\nu)$
-    base distribution  can only model tail-isotropic distributions (\Cref{prop:isotropic-pushforward}),
+    base distribution  can only model tail-isotropic distributions (\Cref{prop:isotropic-pushforward})
     which erroneously imposes power-law tails with the same rate of decay along both the $x$ and $y$ axes.
@@
 we develop and apply theory for fat-tailed random variables and their transformations under Lipschitz-continuous functions.
-Our approach leverages the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration} to sharpen significantly and extend prior results
-\citep[Theorem 4]{jaini2019sum} by describing precisely the tail parameters of the pushforward distribution $(f_\theta)_\ast \mu$ under both Lipschitz-continuous (\Cref{thm:distn_class_closed}) and polynomial (\Cref{corr:closure_polynomials}) transport maps.
+Our approach leverages the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration} to significantly sharpen and extend prior results
+\citep[Theorem 4]{jaini2019sum} by precisely describing the tail parameters of the pushforward distribution $(f_\theta)_\ast \mu$ under both Lipschitz-continuous (\Cref{thm:distn_class_closed}) and polynomial (\Cref{corr:closure_polynomials}) transport maps.
 % \michael{We say ``characterize'' but what do we mean.
@@
 In the multivariate setting,
-we develop a theory of direction-dependent tail parameters (\Cref{def:mv-tail-param}), and we show that tail-isotropic base distributions yield tail-isotropic pushforward measures (\Cref{prop:isotropic-pushforward}).
+we develop a theory of direction-dependent tail parameters (\Cref{def:mv-tail-param}), and show that tail-isotropic base distributions yield tail-isotropic pushforward measures (\Cref{prop:isotropic-pushforward}).
 As a consequence of \Cref{prop:isotropic-pushforward}, prior methods \citep{jaini2020tails} are limited in that
@@
 This motivates the construction of \emph{anisotropic tail adaptive flows} (ATAF, \Cref{def:ataf}) as a means to
-alleviate this issue (\Cref{remark:anisotropic}) and to improve modeling of tail-anisotropic distributions.
+alleviate this issue (\Cref{remark:anisotropic}) and improve modeling of tail-anisotropic distributions.
 % \michael{What limitation is precisely being referred to, since in the previous sentence we say that we get several results in this paper, and then we say how our results are a limitation.
 Probably the best way to do this is to be more detailed, e.g., say that our \Cref{prop:isotropic-pushforward} means that the mapping can't introduce anisotropy.}
-Our experiments show that ATAF exhibits correct tail behaviour in synthetic target distributions exhibiting fat-tails (\Cref{fig:cauchy_normal_student} of \Cref{sec:cauchy_normal_student}) and tail-anisotropy (\Cref{fig:pancake}).
+Our experiments show ATAF exhibits correct tail behaviour in synthetic target distributions exhibiting fat-tails (\Cref{fig:cauchy_normal_student}) and tail-anisotropy (\Cref{fig:pancake}).
 On realistic targets,
@@
 \vspace{-1mm}
-\paragraph{Fat-Tails in Variational Inference.}
+\paragraph{Fat-tails in variational inference}
 
@@
 Recent work in variational autoencoders (VAEs) have considered relaxing Gaussian assumptions to heavier-tailed distributions \citep{mathieu2019disentangling,chen2019residual,boenninghoff2020variational,abiri2020variational}.
-In \citet{mathieu2019disentangling}, a StudentT prior distribution $p(z)$ is considered over the latent code
+In \citep{mathieu2019disentangling}, a StudentT prior distribution $p(z)$ is considered over the latent code
 $z$ in a VAE with Gaussian encoder $q(z \mid x)$. They argue
 %\footnote{\url{https://github.com/iffsid/disentangling-disentanglement/blob/3396d40f46c34dd928a1241f567a86276b0ff41b/src/main.py\#L52}},
-that the anisotropy of a StudentT product distribution leads to more disentangled representations, as compared to the standard choice of Normal distributions.
-A similar modification is performed in \citet{chen2020use} for
-a coupled VAE \citep{cao2019coupled}. 
-This result showed improvements in the marginal
+that the anisotropy of a StudentT product distribution leads to more disentangled representations as compared to the standard choice of Normal distributions.
+A similar modification is performed in \citet{chen2020use}, for
+a coupled VAE (see \citet{cao2019coupled}). It showed improvements in the marginal
 likelihoods of reconstructed images. In addition,
@@
 decoder model $p(x \mid z)$ and the prior $p(z)$. Our work differs from
-\citet{mathieu2019disentangling,chen2020use,boenninghoff2020variational}, in
+\citet{mathieu2019disentangling,chen2020use,boenninghoff2020variational} in
 that we consider fat-tailed variational approximations $q(z \mid x)$ rather
@@
 a StudentT approximate posterior, our work involves a more general
-variational family which uses normalizing flows.
+variational family which use normalizing flows.
 % Relaxation of priors to heavy-tailed distributions has numerous
@@
 \vspace{-1mm}
-\paragraph{Flow-Based Methods.}
-
-Normalizing flows and other flow-based methods have a rich history within variational
+\paragraph{Flow based methods}
+
+Normalizing flows and other flow based methods have a rich history within variational
 inference \citep{kingma2016improved,rezende2015variational,agrawal2020advances,webb2019improving}.
 Consistent with our experience (\Cref{fig:blr-anisotropic}), \citet{webb2019improving}
-documents normalizing flows can offer improvements over ADVI and NUTS across thirteen different
+reports that normalizing flows can offer improvements over ADVI and NUTS across thirteen different
 Bayesian linear regression models from \citet{gelman2006data}.
@@
 \vspace{-2mm}
-\section{Flow-Based Methods for Fat-Tailed Variational Inference}
+\section{Flow-based methods for fat-tailed variational inference}
 
@@
 \vspace{-2mm}
-\subsection{Flow-Based VI Methods}
+\subsection{Flow-based VI methods}
 
@@
 While alternatives exist \citep{li2016variational,wang2018variational}, VI typically
-seeks to find $q_\phi$ ``close'' to $\pi$, as measured by Kullback-Leibler divergence $\infdiv{q_\phi}{\pi}$.
+seeks to find $q_\phi$ ``close'' to $\pi$ as measured by Kullback-Leibler divergence $\infdiv{q_\phi}{\pi}$.
 To ensure tractability without sacrificing generality, in practice \citep{wingate2013automated,ranganath2014black}
@@
   x_i \simiid q_\phi,\;
-  \bar{\pi} \propto \pi. 
+  \bar{\pi} \propto \pi
 \end{align*}
@@
 $\pi(x) = p(x \mid y) = \frac{p(x, y)}{p(y)}$ is typically intractable but
-$\bar\pi(x) = p(x,y)$ is computable (i.e., represented by the probabilistic program's
+$\bar\pi(x) = p(x,y)$ is computable (i.e. represented by the probabilistic program's
 generative / forward execution).
@@
     \toprule
-    Model                                  & Autoregressive transform                                                                              & Suff. cond. for Lipschitz-continuity        \\
+    Model                                  & Autoregressive transform                                                                              & Suff. conditions for Lipschitz         \\
     \midrule
@@
   and sufficient conditions conditions for Lipschitz-continuity. A subset of this table was first presented
-  in \citet{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.
-  \vspace{-5mm}
-  }
+  in \citet{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.}
   \label{tab:flows}
@@
 While it is possible to construct a variational family $\cQ$ tailored to a specific task, we are interested in VI methods which are more broadly applicable and convenient to use: $\cQ$ should be automatically constructed from introspection of a given probabilistic model/program.
-Automatic differentiation variational inference (ADVI, \citet{kucukelbir2017automatic}) is an early implementation of automatic VI and it is still the default in certain probabilistic programming languages \citep{carpenter2017stan}.
+Automatic differentiation variational inference (ADVI) \citep{kucukelbir2017automatic} is an early implementation of automatic VI and it is still the default in certain probabilistic programming languages \citep{carpenter2017stan}.
 ADVI uses a Gaussian base distribution $\mu$ and a transport map $f_\theta = f \circ \Phi_\text{Affine}$ comprised of an invertible affine transform composed with a deterministic transformation $f$ from $\RR$ to the target distribution's support (e.g., $\exp : \RR \to \RR_{\geq 0}$, $\text{sigmoid} : \RR \to [0,1]$).
@@
     (f~\circ~\Phi_\text{Flow})_\ast \mu
-    \}$, 
+    \}$ 
   where $\mu = \text{Normal}(0_d, I_d)$,
@@
 To address this issue, tail-adaptive flows \citep{jaini2020tails} use a
-base distribution $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$,
-where a single degrees-of-freedom $\nu \in \RR$ is used across all $d$ dimensions. Here is a more precise definition.
+base distribution $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$
+where a single degrees-of-freedom $\nu \in \RR$ is used across all $d$ dimensions. More precisely,
 \begin{definition}
@@
     (f \circ \Phi_\text{Flow})_\ast \mu_\nu
-    \}$,
+    \}$
   where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$ with $\nu$ shared across all $d$ dimensions,
@@
 \vspace{-2mm}
-\subsection{Fat-Tailed Variational Inference}
+\subsection{Fat-tailed variational inference}
 
@@
 \citep[Section 3]{jaini2020tails}, here we propose a more natural and finer-grained classification based upon
-the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration}, which is sensitive to
+the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration} which is sensitive to
 the rate of tail decay.
@@
 \vspace{-2mm}
-\section{Tail Behavior of Lipschitz Flows}
+\section{Tail behavior of Lipschitz flows}
 
@@
 We sharpen previous impossibility results approximating fat-tailed targets
-using light-tailed base distributions \citep[Theorem 4]{jaini2020tails}
+using thin-tailed base distributions \citep[Theorem 4]{jaini2020tails}
 by characterizing the effects of Lipschitz-continuous transport maps on not only the tail class
@@
 Most of our results are developed within the context of Lipschitz-continuous transport maps $f_\theta$.
-In practice, many flow-based methods exhibit Lipschitz-continuity in their transport map, either by design \citep{grathwohl2018ffjord,chen2019residual}, or as a consequence of choice of architecture and activation function (\Cref{tab:flows}).
+In practice, many flow-based methods exhibit Lipschitz-continuity in their transport map either by design \citep{grathwohl2018ffjord,chen2019residual}, or as a consequence of choice of architecture and activation function (\Cref{tab:flows}).
 % or by being enforced stemming from the Banach fixed point theorem
@@
 \end{assumption}
-It is worth noting that domains other than $\mathbb{R}^d$ may require an additional bijection between supports (e.g. $\exp : \mathbb{R} \to \mathbb{R}_+$) which
-could violate \cref{assump:lipschitz}.
 
 \vspace{-2mm}
-\subsection{Closure of Tail Classes}
+\subsection{Closure of tail classes}
 \label{ssec:failure}
@@
 \vspace{-1mm}
-Our first set of results pertains to the closure of the tail classes in \Cref{def:tail-classification}
+Our first set of results pertain to closure of the tail classes in \Cref{def:tail-classification}
 under Lipschitz-continuous transport maps. While earlier work \citep{jaini2020tails} demonstrated
-closure of exponential-type distributions $\cup_{p > 0} \overline{\cE^p}$ under flows satisfying \Cref{assump:lipschitz}, our results in Theorem \ref{thm:distn_class_closed} and Corollaries \ref{corr:heavy_to_light} and \ref{corr:closure_polynomials} sharpen these observations, showing that: (1) Lipschitz transport maps cannot decrease the class index $p$ for exponential-type random variables, but they can alter the tail parameter $\alpha$; and
-(2) under additional assumptions, they cannot change either class index $p$ or the tail parameter $\alpha$ for logarithmic-type random variables.
+closure of exponential-type distributions $\cup_{p > 0} \overline{\cE^p}$ under flows satisfying \Cref{assump:lipschitz}, our results in Theorem \ref{thm:distn_class_closed}, and Corollaries \ref{corr:heavy_to_light} and \ref{corr:closure_polynomials} sharpen these observations, showing that (1) Lipschitz transport maps cannot decrease the class index $p$ for exponential-type random variables, but can alter the tail parameter $\alpha$; and
+(2) under additional assumptions, cannot change either class index $p$ or the tail parameter $\alpha$ for logarithmic-type random variables.
 %         tail parameter $\alpha$; 
@@
 by itself does not preclude transformations of fat-tailed base distributions to light-tailed targets.
-Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushfoward,
+Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushfoward
 hence heavy-to-light transformations are impossible. Note here there is no ascending union over
@@
 \vspace{-2mm}
-\subsection{Multivariate Fat-Tails and Anisotropic Tail Adaptive Flows}
+\subsection{Multivariate fat-tails and anisotropic tail adaptive flows}
 
 \vspace{-1mm}
-Next, we restrict attention to power-law tails $\cL^1_\alpha$, and we develop a multivariate fat-tailed theory and notions of isotropic/anisotropic tail indices.
+Next, we restrict attention to power-law tails $\cL^1_\alpha$ and develop a multivariate fat-tailed theory and notions of isotropic/anisotropic tail indices.
 Using our theory, we prove that both ADVI and TAF are fundamentally limited because they
 are only capable of fitting tail-isotropic target measures (\Cref{prop:isotropic-pushforward}).
-We consider anisotropic tail adaptive flows (ATAF): a density
+We propose anisotropic tail adaptive flows (ATAF) as a density
 modeling method which can represent tail-anisotropic distributions (\Cref{remark:anisotropic}).
@@
   %\input{../Figures/radial-fat-tail.pgf}
-  \vspace{-7mm}
+  \vspace{-3mm}
   \caption{
-    Illustration of the direction-dependent tail-parameter function (right) on a tail-anisotropic distribution (left)
+    Illustration of the direction-dependent tail-parameter function (bottom) on a tail-anisotropic distribution (top)
     with PDF $\dd P(r,\theta) = r^{-\alpha(\theta)} r \dd r \dd\theta$ and tail parameter $\alpha(\theta) = 2 + \cos(2\theta)$.
@@
     capture \emph{tail-anisotropy}.
-    \vspace{-5mm}
   }
@@
 % has index $\max_i \nu_i = \nu$.
+
 Of course, one can construct pathological densities where this definition is not effective
-(see \Cref{eg:spiral}), but it will suffice for our purposes. 
-It is illustrative to contrast with the theory presented for TAF \citep{jaini2020tails},
+(see \Cref{eg:spiral}), but we will find it to suffice for our purposes. It is illustrative to contrast to the theory presented for TAF \citep{jaini2020tails}
 where only the tail exponent of $\|X\|_2$ is considered.
@@
   = \max_{1 \leq i \leq d} \nu_i
-  = \max_{0 \leq i \leq d-1} \alpha_X(e_i) .
+  = \max_{0 \leq i \leq d-1} \alpha_X(e_i)
 \end{align*}
@@
   where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu_i)$, each $\nu_i$ is \emph{distinct}, and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
-  Analogous to \citet{jaini2020tails}, ATAF's implementation
+  Analogous to \cite{jaini2020tails}, ATAF's implementation
   treats $\nu_i$ identically to the other parameters in the flow and jointly optimizes over them.
@@
 
-Naturally, there are other parameterizations of the tail parameters $\nu_i$ that may be more effective depending on the application.
-For example, in high dimensions, one might prefer not to allow for $d$ unique indices, but perhaps only fewer.
-On the other hand, by using only $d$ tail parameters, an approximation error will necessarily be incurred when more than $d$ different tail parameters are present.
-\Cref{fig:radial-fat-tail} presents a worst-case scenario where the target distribution has a continuum of tail parameters.
-In theory, this density could itself be used as an underlying base distribution, although we have not found this to be a good option in practice.
-The key takeaway is that to capture several different tails in the target density, one must consider a base distribution that incorporates sufficiently many \emph{distinct} tail parameters.
-
-Concerning the choice of StudentT families, we remark that since $\text{StudentT}(\nu) \Rightarrow \cN(0,1)$ 
-as $\nu \to \infty$, ATAF should still provide reasonably good approximations to target 
-distributions in $\overline{\cE^2}$ by taking $\nu$ sufficiently large. This can be seen in practice in \Cref{sec:normal-normal-location-mixture}.
-%whereas the tail-parameter function may take on more than $d$ values. \Cref{fig:radial-fat-tail}
-%presents a worst-case scenario.
-%As ATAF is only capable of representing $d$ different tail parameters in its pushforward
-%(\Cref{remark:anisotropic}), it will necessarily incur an approximation error when $> d$ different
-%tail parameters are present.
-
 % Compared to \citep{jaini2020tails}, the degrees of freedom $\nu$ is no longer
@@
 
-%\vspace{-2mm}
-%\subsection{Discussion of limitations}
-%\label{ssec:limitations}
-
-%\vspace{-1mm}
-%By only considering projections $\braket{X,v}$ along rays through the origin
-%$v \in \cS^{d-1}$, the limit in $\alpha_X(\cdot)$ may not be defined
-%(see \cref{eg:spiral}).
-
-%Furthermore, 
+\vspace{-2mm}
+\subsection{Discussion of limitations}
+\label{ssec:limitations}
+
+\vspace{-1mm}
+
+Additionally, ATAF only uses a finite collection $\{\nu_i\}_{i=1}^d$ of tail parameters
+whereas the tail-parameter function may take on more than $d$ values. \Cref{fig:radial-fat-tail}
+presents a worst-case scenario, where the target distribution has a continuum of tail parameters.
+As ATAF is only capable of representing $d$ different tail parameters in its pushforward
+(\Cref{remark:anisotropic}), it will necessarily incur an approximation error when $> d$ different
+tail parameters are present.
+Furthermore, distributions supported on domains other than Euclidian $\mathbb{R}^d$ may require an additional bijection between supports (e.g. $\exp : \mathbb{R} \to \mathbb{R}_+$) which
+may violate \cref{assump:lipschitz}.
+Finally, note that $\text{StudentT}(\nu) \Rightarrow \cN(0,1)$ 
+as $\nu \to \infty$ so we should expect ATAF to provide reasonably good approximations to target 
+distributions in $\overline{\cE^2}$ by taking $\nu$ sufficiently large. We observed this occurs in 
+practice empirically in \Cref{sec:normal-normal-location-mixture}.
 % We found that applying ATAF when the target is light tailed results in minimal
@@
 density modelling when fat tails are considered, and
-our experiments are complementary by evaluating TAFs and ATAFs for variational inference tasks as well as by demonstrating the effect of tail-anisotropy for modelling real-world financial returns and insurance claims datasets.
+our experiments are complementary by evaluating TAFs and ATAFs for variational inference tasks as well as demonstrating the necessity of tail-anisotropy for modelling real-world financial returns and insurance claims datasets.
 We implement using the 
-\texttt{beanmachine} probabilistic programming language \citep{tehrani2020bean}
-%\texttt{REDACTED FOR REVIEW} probabilistic programming language[REDACTED]
+%\texttt{beanmachine} probabilistic programming language\citep{tehrani2020bean}
+\texttt{REDACTED FOR REVIEW} probabilistic programming language[REDACTED]
 and the
-\texttt{flowtorch} library for normalizing flows \citep{flowtorchai},
-%\texttt{REDACTED FOR REVIEW} library for normalizing flows[REDACTED],
+%\texttt{flowtorch} library for normalizing flows\citep{flowtorchai},
+\texttt{REDACTED FOR REVIEW} library for normalizing flows[REDACTED],
 %and we have open-sourced the code\citep{ghlic}
-and we have open-sourced code for reproducing experiments in Supplementary Materials. 
-% TODO: add facebookresearch reference 
+and we have open-sourced code for reproducing experiments[REDACTED, reviewers please see supplementary materials for code]. 
 Additional details for the experiments are detailed in \Cref{sec:additional-exp-details}.
@@
   \centering
-  \vspace{-0.2cm}
+  \vspace{-0.5cm}
   \includegraphics{../Figures/blr_aniso.pdf}
@@
     \Cref{prop:isotropic-pushforward} implies
-    Gaussian (Gaussian, second row) or isotropic StudentT product (TAF, third row) base distributions
-    yield Gaussian or power-law tails, respectively, for \emph{both} $\sigma$ and $\beta$.
-    In contrast, ATAF (bottom row) illustrates \Cref{remark:anisotropic} by
-    modeling simultaneously a power-law tail on $\sigma$ and Gaussian tail on $\beta$.
+    Gaussian (Gaussian, second row) or isotropic StudentT product (TAF, third row) base distribution
+    yields Gaussian or power-law tails respectively for \emph{both} $\sigma$ and $\beta$.
+    In contrast, ATAF (bottom row) illustrates \Cref{remark:anisotropic} by simultaneously
+    modeling a power-law tail on $\sigma$ and Gaussian tail on $\beta$.
   }
@@
             \midrule
-            ADVI      & $\mathbf{2873.90} \pm 6.95$    & $2969.73 \pm 1.73$ \\
+            ADVI      & $2873.90 \pm 6.95$    & $2969.73 \pm 1.73$ \\
             TAF       & $2839.64 \pm 9.10$    & $2973.85 \pm 0.87$ \\
-            ATAF      & $2842.75 \pm 8.83$    & $\mathbf{2976.75} \pm 0.66$ \\
-            \hline
+            ATAF      & $2842.75 \pm 8.83$    & $2976.75 \pm 0.66$ \\
             NUTS      & n/a                  & $3724.59 \pm 0.036$ \\
@@
             TAF       & $-64.64 \pm 4.88$    & $-52.51 \pm 4.41$ \\
-            ATAF      & $\mathbf{-58.63} \pm 4.75$    & $\mathbf{-51.01} \pm 3.71$ \\
-            \hline
+            ATAF      & $-58.63 \pm 4.75$    & $-51.01 \pm 3.71$ \\
             NUTS      & n/a                  & $-47.78 \pm 0.093$ \\
@@
     \end{subfigure}
-    \vspace{-2mm}
+
         \caption{Monte-Carlo ELBO and importance weighted Monte-Carlo marginal likelihood
@@
         estimates from VI on real-world datasets.
-        To understand the variational approximation gap, we include marginal likelihoods based on ``golden samples'' from \texttt{posteriordb} \citep{ghposteriordb} computed using No-U-Turn-Sampling (NUTS, \citet{hoffman2014no,carpenter2017stan}).
+        For comparison, we include results computed using ``golden samples'' from \texttt{posteriordb} \citep{ghposteriordb} (NUTS).
+        These golden samples are obtained using No-U-Turn-Sampling \citep{hoffman2014no,carpenter2017stan},
+        a state-of-the-art MCMC method which does not optimize ELBO, is more computationally expensive than
+        VI, and only produces samples rather than parametric probability density functions.
     }
@@
         TAF       & $-4.703 \pm 0.023$    & $-1.659 \pm 0.004$ \\
-        ATAF      & $\mathbf{-4.699} \pm 0.024$    & $\mathbf{-1.603} \pm 0.034$ \\
+        ATAF      & $-4.699 \pm 0.024$    & $-1.603 \pm 0.034$ \\
         \bottomrule
     \end{tabular}
-    \vspace{-2mm}
     \caption{
-        Log-likelihoods (higher is better, $\pm$ standard errors) achieved on
-        density modeling tasks involving financial returns \citep{fama2015five} and insurance claims \citep{cms} data. 
-        \vspace{-5mm}
+        Log-likelihoods (higher is better, $\pm$ standard errors, $10$ trials) achieved on
+        density modeling tasks involving real-world financial returns \citep{fama2015five} and insurance claims \citep{cms} data. 
     }
@@
 
-\subsection{Bayesian Linear Regression}
+\subsection{Bayesian linear regression}
 
@@
     \beta \mid \sigma^2 \sim \cN(0, \sigma^2),\qquad
-    y \mid X, \beta, \sigma \sim \cN(X \beta, \sigma^2) ,
+    y \mid X, \beta, \sigma \sim \cN(X \beta, \sigma^2) 
 \end{gather*}
@@
     a_0 + \frac{n}{2}, 
-    b_0 + \frac{1}{2}(y^\top y - \mu_n^\top \Sigma_n \mu_n)\bigg). % \\
+    b_0 + \frac{1}{2}(y^\top y - \mu_n^\top \Sigma_n \mu_n)\bigg)% \\
     \]
@@
 can produce a tail-anisotropic pushforward. This intuition is confirmed in \Cref{fig:blr-anisotropic},
-where we see that only ATAF is the only method capable of modeling the tail-anisotropy present in the data.
-
-Conducting Bayesian linear regression is among the standard tasks requested of a probabilistic programming language, yet it still displays tail-anisotropy.
-To accurately capture large quantiles, this tail-anisotropy  should not be ignored, necessitating a method such as ATAF.
-
-\subsection{Diamond Price Prediction Using Non-Conjugate Bayesian Regression}
+where we see that only ATAF is the only method capable of modeling the tail-anisotropy present.
+
+\subsection{Diamond price prediction using non-conjugate Bayesian regression}
 \label{ssec:diamonds}
@@
 diamonds each with $24$ features as well as an outcome variable $y \in \RR^{5000}$ representing each diamond's price.
-The probabilistic model for this inference task is specified in Stan code provided by \citet{ghposteriordb} and is reproduced here
-for convenience:
+The probabilistic model for this inference task is specified in Stan code provided by \citep{ghposteriordb} and is reproduced
+for convenience
 \begin{gather*}
@@
     \beta \sim \cN(0, \mI_{24}),\qquad
-    y \sim \cN(\alpha + X \beta, \sigma) .
+    y \sim \cN(\alpha + X \beta, \sigma)
 \end{gather*}
@@
 
-\subsection{Eight Schools SAT Score Modelling with Fat-tailed Scale Mixtures}
+\subsection{Eight schools SAT score modelling with fat-tailed scale mixtures}
 \label{ssec:eight_schools}
@@
 test scores and coaching programs in place at eight schools.
-A variation using half Cauchy non-informative priors \citep{gelman2006prior} provides
+A variation utilizing half Cauchy non-informative priors \citep{gelman2006prior} provides
 a real-world inference problem involving fat-tailed distributions, and is formally specified
@@
     \theta \sim \cN(\mu, \tau),\qquad
-    y \sim \cN(\theta, \sigma) .
+    y \sim \cN(\theta, \sigma)
 \end{gather*}
@@
 posterior distribution over treatment effects $\theta_1,\ldots,\theta_d$. The experimental
-parameters are identical to \Cref{ssec:diamonds}, and results are reported in \Cref{tab:eight_schools}.
-
-\subsection{Financial and Actuarial Applications}
-
-To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021 \citep{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010 \citep{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events.
+parameters are identical to \Cref{ssec:diamonds} and results are reported in \Cref{tab:eight_schools}.
+
+\subsection{Tail-anisotropic modeling of real-world data}
+
+To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021, \cite{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010, \cite{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events.
 Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}.
@@
 
-In this work, we have sharpened existing theory for approximating fat-tailed distributions with normalizing flows, and we formalized tail-(an)isotropy through a direction-dependent tail parameter.
-With this, we have shown that many prior flow-based methods are inherently limited by tail-isotropy.
-With this in mind, we proposed a simple flow-based method capable of modeling tail-anisotropic targets.
-As we have seen, anisotropic FTVI is already applicable in fairly elementary examples such as Bayesian linear regression;
-and ATAFs provide one of the first methods for using the representational capacity of flow-based methods,
-while simultaneously producing tail-anisotropic distributions.
-A number of open problems still remain, including the study of other parameterizations of the tail behaviour of the base distribution.
-Even so, going forward, it seems prudent that density estimators, especially those used in black-box settings, consider accounting for tail-anisotropy using a method such as~ATAF.% expect the scientific impact of ATAFs to be
-%fairly easily realized, as it can be applied wherever a parametric density estimator is used.
-%While we do not expect ATAF to directly have any negative societal impacts,
-%its broad applicability as a parametric density estimator which accommodates fat-tails means that it may be
-%used to improve applications which do cause negative societal impact.
+We sharpened existing theory for approximating fat-tailed distributions with normalizing flows, formalized  tail-(an)isotropy through a direction-dependent tail parameter, proved many prior flow-based methods are limited by tail-isotropy, and proposed a new flow-based method called
+ATAF which is capable of modeling tail-anisotropic targets.
+As we have seen, anisotropic FTVI is already applicable in fairly elementary examples such as Bayesian linear regression
+and ATAFs provide one of the first methods for using the representational capacity of flow-based methods
+while simultaneously producing tail-anisotropic distributions. We expect the scientific impact of ATAFs to be
+fairly easily realized, as it can be applied wherever a parametric density estimator is used.
+While we do not expect ATAF to directly have any negative societal impacts,
+its broad applicability as a parametric density estimator which accommodates fat-tails means that it may be
+used to improve applications which do cause negative societal impact.
 
@@
 
-\paragraph{Acknowledgements}
-L.H. and M.M's contributions were supported in part by DARPA, NSF, and ONR.
-F.L.’s research while at the University of California, Berkeley was supported in part by a GFSD fellowship.
+\emph{Acknowledgements}
+%   Feynman Liang is supported by funding from NPSC and Facebook.
+%   Michael Mahoney acknowledges funding from ARO, DARPA, NSF, and ONR.
+Redacted
 
 \bibliography{refs_ftvi}
-\bibliographystyle{icml2022}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
+\clearpage
 \appendix
-\onecolumn
-
-\section{Experiments Performing VI Against a Fat-tailed Cauchy Target}
-\label{sec:cauchy_normal_student}
-The motivation for the fat-tailed variational families used in TAF/ATAF
-is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
-As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
-appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
-imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
-As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
-test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
-than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.
-
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics{../Figures/fat_tail_ks.pdf}
-  \vspace{-6mm}
-  \caption{
-    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
-    the use of a Gaussian variational family (ADVI, solid blue line) can incur
-    exponentially bad tail approximations (middle panel) compared to
-    methods such as ATAF which permit heavier tails (orange dashed line).
-    As a consequence, ADVI samples (blue, right panel) are rejected by
-    the Kolmogorov-Smirnov test more often than ATAF samples (orange, right panel).
-    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.
-It might be good to show ATAF is much better but still not perfect.
-Also, what is the X axis on each figure.
-}
-  }
-  \label{fig:cauchy_normal_student}
-\end{figure*}
-
-\section{Proofs of Our Main Theoretical Results}
+
+{\onecolumn
+\section{Proofs}
 \label{sec:proofs}
@@
   Its concentration function
-  (Equation 1.6 \citet{ledoux2001concentration}
+  \cite[Equation 1.6]{ledoux2001concentration}
   is given by
@@
     \coloneqq \sup \{ \mu\{x : d(x,A) \geq r\}; A \subset \text{supp}~X, \mu(A) \geq 1/2\}
-    = \PP(\lvert X - m_X \rvert \geq r) .
+    = \PP(\lvert X - m_X \rvert \geq r)
   \]
   Under Assumption 1, $f_\theta$ is Lipschitz (say with Lipschitz
-  constant $L$) so by Proposition 1.3 of \citet{ledoux2001concentration},
+  constant $L$) so by \cite[Proposition 1.3]{ledoux2001concentration},
   \[
@@
     &= \cO(\PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)) \nonumber\\
-    &= \cO(\alpha_X(r/L)) ,
-    \label{eq:pushforward-conc-fn}
+    &= \cO(\alpha_X(r/L)) \label{eq:pushforward-conc-fn}
   \end{align}
@@
     \[
-    \frac{d}{dx}(f_\theta)^{-1}(x) = \frac{1}{(f_\theta)'((f_\theta)^{-1}(x))} \leq \frac{1}{M} .
+    \frac{d}{dx}(f_\theta)^{-1}(x) = \frac{1}{(f_\theta)'((f_\theta)^{-1}(x))} \leq \frac{1}{M}
     \]
@@
 It will suffice to show that (I) $\mathbb{P}(|X+Y|\geq r)=\mathcal{O}(r^{-\gamma})$, and (II) $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\gamma})$.
-Since $(X,Y)\mapsto|X+Y|$ is a 1-Lipschitz function on $\mathbb{R}^{2}$ and $\mathbb{P}(|X|\geq r)+\mathbb{P}(|Y|\geq r)=\mathcal{O}(r^{-\gamma})$, (I) follows directly from the hypotheses and Proposition 1.11 of \citet{ledoux2001concentration}.
+Since $(X,Y)\mapsto|X+Y|$ is a 1-Lipschitz function on $\mathbb{R}^{2}$ and $\mathbb{P}(|X|\geq r)+\mathbb{P}(|Y|\geq r)=\mathcal{O}(r^{-\gamma})$, (I) follows directly from the hypotheses and \cite[Proposition 1.11]{ledoux2001concentration}.
 To show (II), note that for any $M>0$, conditioning on the event $|Y|\leq M$,\[
@@
 \]
-From Theorem C.2.1 of \citet{buraczewski2016stochastic}, since $\nu \not \in \mathbb{Z}$, there exists a non-zero measure $\mu$ such that
+From \cite[Theorem C.2.1]{buraczewski2016stochastic}, since $\nu \not \in \mathbb{Z}$, there exists a non-zero measure $\mu$ such that
 \[
@@
 for any Borel set $E$. Consequently, $\mu$ is regularly varying, and so 
-by the spectral representation of regularly varying random vectors (see p. 281 \citet{buraczewski2016stochastic}), there exists a measure $P$ such that
+by the spectral representation of regularly varying random vectors (see \cite[pg. 281]{buraczewski2016stochastic}), there exists a measure $P$ such that
 \[
@@
 &\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|X\| > x(m(1-\cos\theta))^{-1}, X/\|X\| \in F_v)}{\|X\| > x / M} \\
-&\geq P(F_v) \left(\frac{M}{m(1-\cos\theta)}\right)^{-\nu} > 0,
-yaB\end{align*}
+&\geq P(F_v) \left(\frac{M}{m(1-\cos\theta)}\right)^{-\nu} > 0.
+\end{align*}
 where $P(F_v) > 0$ follows from the bilipschitz condition for $f$.
@@
 
-\section{Example of Non-existence of Tail Parameter Due to Oscillations}
-\label{eg:spiral}
-
-Consider $\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
-using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
-oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.
+\section{Experiments performing VI against a fat-tailed Cauchy target}
+\label{sec:cauchy_normal_student}
+The motivation for the fat-tailed variational families used in TAF/ATAF
+is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
+As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
+appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
+imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
+As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
+test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
+than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.
 
@@
   \centering
-    \vspace{-.5in}
-    \includegraphics[scale=0.8]{Figures/spiral.pdf}
-    % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\
-    % \includegraphics[trim={5cm 0 4.9cm 0},clip]{Figures/spiral.pdf}\\
-    % \includegraphics[trim={9.3cm 0 0cm 0},clip]{Figures/spiral.pdf}\\
-    \caption{Taking a tail-anisotropic distribution (left) and ``spinning'' it (middle) results in
-        one-dimensional projections which oscillate between tail parameters (as seen in
-        $\log p(\braket{X,e_0})$ in right panel) and result in an ill-defined
-        direction-dependent tail parameter function $\alpha_X(\cdot)$ due to a
-        divergent limit.
+  \includegraphics{../Figures/fat_tail_ks.pdf}
+  \vspace{-6mm}
+  \caption{
+    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
+    the use of a sub-Gaussian variational family (ADVI, solid blue line) can incur
+    exponentially bad tail approximations (middle panel) compared to
+    methods which permit heavier tails (ATAF, green dashed line) and results in
+    samples which are more similar to the target (as measured by $1000$ repeats of
+    1-sample $N=100$ Kolmogorov-Smirnov $p$-values, right panel).
+    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.
+It might be good to show ATAF is much better but still not perfect.
+Also, what is the X axis on each figure.
 }
-    \label{fig:spiral}
+  }
+  \label{fig:cauchy_normal_student}
 \end{figure*}
 
-\section{Normal-normal Conjugate Model}
+\section{Normal-normal location mixture}
 \label{sec:normal-normal-location-mixture}
@@
   \includegraphics[width=0.6\textwidth]{../Figures/normal_normal_posterior.png}
-  \caption{Variational inference against a light tailed Normal posterior. Both light and heavy tail
-  variational families yield similar results.}
+  \caption{VI against a Normal posterior}
   \label{fig:normal_normal}
@@
 
-\section{Additional Details For Experiments}
+\section{Example of non-existence of tail parameter due to oscillations}
+\label{eg:spiral}
+
+Consider $\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
+using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
+oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.
+
+\begin{figure*}[htbp]
+    \centering
+    \includegraphics[scale=0.8]{Figures/spiral.pdf}
+    % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\
+    % \includegraphics[trim={5cm 0 4.9cm 0},clip]{Figures/spiral.pdf}\\
+    % \includegraphics[trim={9.3cm 0 0cm 0},clip]{Figures/spiral.pdf}\\
+    \caption{Taking a tail-anisotropic distribution (top) and ``spinning'' it (middle) results in
+        one-dimensional projections which oscillate between tail parameters (as seen in
+        $\log p(\braket{X,e_0})$ in bottom panel) and cause $\alpha_X(\cdot)$ to be not well defined.
+    }
+    \label{fig:spiral}
+\end{figure*}
+
+\section{Additional details for experiments}
 \label{sec:additional-exp-details}
@@
 running PyTorch 1.9.0 / Python 3.8.5 / CUDA 11.2 / Ubuntu Linux 20.04 via Windows Subsystem for Linux.
-For all flow-transforms $\Phi_{\text{Flow}}$, we used inverse autoregressive flows \citep{kingma2016improved} with a
+For all flow-transforms $\Phi_{\text{Flow}}$ we used inverse autoregressive flows \citep{kingma2016improved} with a
 dense autoregressive conditioner consisting of two layers of either 32 or 256 hidden units depending on problem (see code for details) and
 ELU activation functions.
-As described in \citet{jaini2020tails}, TAF is trained by including $\nu$ within the Adam optimizer alongside other flow parameters. For ATAF, we include all $\nu_i$ within the optimizer.
+As described in \cite{jaini2020tails}, TAF is trained by including $\nu$ within the Adam optimizer alongside other flow parameters. For ATAF, we include all $\nu_i$ within the optimizer.
 Models were trained using the Adam optimizer with $10^{-3}$ learning rate
@@
 For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
-is comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
+are comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
 using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
-used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empkkirically observed
+used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empirically observed
 this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
@@
 the ``Target'' row. A similar progressive training scheme used for \cref{tab:diamonds} was also used here, with
-the TAF flow transform $\Phi_{\text{Flow}}$ initialized from the result of ADVI and ATAF additionally initialized
+the TAF flow transform $\Phi_{\text{Flow}}$ was initialized from the result of ADVI and ATAF additionally initialized
 all tail parameters $\nu_i$ based on the final shared tail parameter obtained from TAF training. Tails are computed
-along the $\beta = 1$ or $\sigma = 1$ axes because the posterior is identically zero for $\sigma = 0$, hence it reveals
+along the $\beta = 1$ or $\sigma = 1$ axes because the posterior is identically zero for $\sigma = 0$ hence it reveals
 no information about the tails.
-
-% \section{Electronic Submission}
-% \label{submission}
-
-% Submission to ICML 2022 will be entirely electronic, via a web site
-% (not email). Information about the submission process and \LaTeX\ templates
-% are available on the conference web site at:
+}
+\end{document}
+
+% \begin{abstract}
+%   The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
+%   both left and right-hand margins. Use 10~point type, with a vertical
+%   spacing of 11~points. The \textbf{Abstract} heading must be centered,
+%   bold, and in point size 12. Two line spaces precede the
+%   Abstract. The Abstract must be limited to one paragraph.
+% \end{abstract}
+
+% \section{GENERAL FORMATTING INSTRUCTIONS}
+
+% The camera-ready versions of the accepted papers are 8 pages,
+% plus any additional pages needed for references.
+
+% Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
+% Each column is 3.25~inches wide (19.5~picas).  The space
+% between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
+% Use 10~point type with a vertical spacing of
+% 11~points. Please use US Letter size paper instead of A4.
+
+% Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
+% Top rule is 4~points thick and bottom rule is 1~point thick.
+% Allow 1/4~inch space above and below title to rules.
+
+% Author descriptions are center-justified, initial caps.  The lead
+% author is to be listed first (left-most), and the Co-authors are set
+% to follow.  If up to three authors, use a single row of author
+% descriptions, each one center-justified, and all set side by side;
+% with more authors or unusually long names or institutions, use more
+% rows.
+
+% Use one-half line space between paragraphs, with no indent.
+
+% \section{FIRST LEVEL HEADINGS}
+
+% First level headings are all caps, flush left, bold, and in point size
+% 12. Use one line space before the first level heading and one-half line space
+% after the first level heading.
+
+% \subsection{Second Level Heading}
+
+% Second level headings are initial caps, flush left, bold, and in point
+% size 10. Use one line space before the second level heading and one-half line
+% space after the second level heading.
+
+% \subsubsection{Third Level Heading}
+
+% Third level headings are flush left, initial caps, bold, and in point
+% size 10. Use one line space before the third level heading and one-half line
+% space after the third level heading.
+
+% \paragraph{Fourth Level Heading}
+
+% Fourth level headings must be flush left, initial caps, bold, and
+% Roman type.  Use one line space before the fourth level heading, and
+% place the section text immediately after the heading with no line
+% break, but an 11 point horizontal space.
+
+% %%%
+% \subsection{Citations, Figure, References}
+
+% \subsubsection{Citations in Text}
+
+% Citations within the text should include the author's last name and
+% year, e.g., (Cheesman, 1985). 
+% %Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
+% Be sure that the sentence reads
+% correctly if the citation is deleted: e.g., instead of ``As described
+% by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
+% described by Cheesman (1985), we first frobulate the widgets.''
+
+% The references listed at the end of the paper can follow any style as long as it is used consistently.
+
+% %Be sure to avoid
+% %accidentally disclosing author identities through citations.
+
+% \subsubsection{Footnotes}
+
+% Indicate footnotes with a number\footnote{Sample of the first
+%   footnote.} in the text. Use 8 point type for footnotes. Place the
+% footnotes at the bottom of the column in which their markers appear,
+% continuing to the next column if required. Precede the footnote
+% section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
+% long.\footnote{Sample of the second footnote.}
+
+% \subsubsection{Figures}
+
+% All artwork must be centered, neat, clean, and legible.  All lines
+% should be very dark for purposes of reproduction, and art work should
+% not be hand-drawn.  Figures may appear at the top of a column, at the
+% top of a page spanning multiple columns, inline within a column, or
+% with text wrapped around them, but the figure number and caption
+% always appear immediately below the figure.  Leave 2 line spaces
+% between the figure and the caption. The figure caption is initial caps
+% and each figure should be numbered consecutively.
+
+% Make sure that the figure caption does not get separated from the
+% figure. Leave extra white space at the bottom of the page rather than
+% splitting the figure and figure caption.
+% \begin{figure}[h]
+% \vspace{.3in}
+% \centerline{\fbox{This figure intentionally left non-blank}}
+% \vspace{.3in}
+% \caption{Sample Figure Caption}
+% \end{figure}
+
+% \subsubsection{Tables}
+
+% All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
+% Table number and title always appear above the table.
+% See Table~\ref{sample-table}.
+
+% Use one line space before the table title, one line space after the table title,
+% and one line space after the table. The table title must be
+% initial caps and each table numbered consecutively.
+
+% \begin{table}[h]
+% \caption{Sample Table Title} \label{sample-table}
 % \begin{center}
-% \textbf{\texttt{http://icml.cc/}}
+% \begin{tabular}{ll}
+% \textbf{PART}  &\textbf{DESCRIPTION} \\
+% \hline \\
+% Dendrite         &Input terminal \\
+% Axon             &Output terminal \\
+% Soma             &Cell body (contains cell nucleus) \\
+% \end{tabular}
 % \end{center}
-
-% The guidelines below will be enforced for initial submissions and
-% camera-ready copies. Here is a brief summary:
-% \begin{itemize}
-% \item Submissions must be in PDF\@. 
-% \item \textbf{New to this year}: If your paper has appendices, submit the appendix together with the main body and the references \textbf{as a single file}.
-Reviewers will not look for appendices as a separate PDF file.
-So if you submit such an extra file, reviewers will very likely miss it.
-% \item Page limit: The main body of the paper has to be fitted to 8 pages, excluding references and appendices; the space for the latter two is not limited.
-For the final version of the paper, authors can add one extra page to the main body.
-% \item \textbf{Do not include author information or acknowledgements} in your
-%     initial submission.
-% \item Your paper should be in \textbf{10 point Times font}.
-% \item Make sure your PDF file only uses Type-1 fonts.
-% \item Place figure captions \emph{under} the figure (and omit titles from inside
-%     the graphic file itself). Place table captions \emph{over} the table.
-% \item References must include page numbers whenever possible and be as complete
-%     as possible. Place multiple citations in chronological order.
-% \item Do not alter the style template; in particular, do not compress the paper
-%     format by reducing the vertical spaces.
-% \item Keep your abstract brief and self-contained, one paragraph and roughly
-%     4--6 sentences. Gross violations will require correction at the
-%     camera-ready phase. The title should have content words capitalized.
-% \end{itemize}
-
-% \subsection{Submitting Papers}
-
-% \textbf{Paper Deadline:} The deadline for paper submission that is
-% advertised on the conference website is strict. If your full,
-% anonymized, submission does not reach us on time, it will not be
-% considered for publication.
-
-% \textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
-% author information may appear on the title page or in the paper
-% itself. \cref{author info} gives further details.
-
-% \textbf{Simultaneous Submission:} ICML will not accept any paper which,
-% at the time of submission, is under review for another conference or
-% has already been published. This policy also applies to papers that
-% overlap substantially in technical content with conference papers
-% under review or previously published. ICML submissions must not be
-% submitted to other conferences and journals during ICML's review
-% period.
-% %Authors may submit to ICML substantially different versions of journal papers
-% %that are currently under review by the journal, but not yet accepted
-% %at the time of submission.
-% Informal publications, such as technical
-% reports or papers in workshop proceedings which do not appear in
-% print, do not fall under these restrictions.
-
-% \medskip
-
-% Authors must provide their manuscripts in \textbf{PDF} format.
-% Furthermore, please make sure that files contain only embedded Type-1 fonts
-% (e.g.,~using the program \texttt{pdffonts} in linux or using
-% File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
-% might come from graphics files imported into the document.
-
-% Authors using \textbf{Word} must convert their document to PDF\@. Most
-% of the latest versions of Word have the facility to do this
-% automatically. Submissions will not be accepted in Word format or any
-% format other than PDF\@. Really. We're not joking. Don't send Word.
-
-% Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
-% Those using \texttt{latex} and \texttt{dvips} may need the following
-% two commands:
-
-% {\footnotesize
-% \begin{verbatim}
-% dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
-% ps2pdf paper.ps
-% \end{verbatim}}
-% It is a zero following the ``-G'', which tells dvips to use
-% the config.pdf file. Newer \TeX\ distributions don't always need this
-% option.
-
-% Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
-% results. This program avoids the Type-3 font problem, and supports more
-% advanced features in the \texttt{microtype} package.
-
-% \textbf{Graphics files} should be a reasonable size, and included from
-% an appropriate format. Use vector formats (.eps/.pdf) for plots,
-% lossless bitmap formats (.png) for raster graphics with sharp lines, and
-% jpeg for photo-like images.
-
-% The style file uses the \texttt{hyperref} package to make clickable
-% links in documents. If this causes problems for you, add
-% \texttt{nohyperref} as one of the options to the \texttt{icml2022}
-% usepackage statement.
-
-% \subsection{Submitting Final Camera-Ready Copy}
-
-% The final versions of papers accepted for publication should follow the
-% same format and naming convention as initial submissions, except that
-% author information (names and affiliations) should be given. See
-% \cref{final author} for formatting instructions.
-
-% The footnote, ``Preliminary work. Under review by the International
-% Conference on Machine Learning (ICML). Do not distribute.'' must be
-% modified to ``\textit{Proceedings of the
-% $\mathit{39}^{th}$ International Conference on Machine Learning},
-% Baltimore, Maryland, USA, PMLR 162, 2022.
-% Copyright 2022 by the author(s).''
-
-% For those using the \textbf{\LaTeX} style file, this change (and others) is
-% handled automatically by simply changing
-% $\mathtt{\backslash usepackage\{icml2022\}}$ to
-% $$\mathtt{\backslash usepackage[accepted]\{icml2022\}}$$
-% Authors using \textbf{Word} must edit the
-% footnote on the first page of the document themselves.
-
-% Camera-ready copies should have the title of the paper as running head
-% on each page except the first one. The running title consists of a
-% single line centered above a horizontal rule which is $1$~point thick.
-% The running head should be centered, bold and in $9$~point type. The
-% rule should be $10$~points above the main text. For those using the
-% \textbf{\LaTeX} style file, the original title is automatically set as running
-% head using the \texttt{fancyhdr} package which is included in the ICML
-% 2022 style file package. In case that the original title exceeds the
-% size restrictions, a shorter form can be supplied by using
-
-% \verb|\icmltitlerunning{...}|
-
-% just before $\mathtt{\backslash begin\{document\}}$.
-% Authors using \textbf{Word} must edit the header of the document themselves.
-
-% \section{Format of the Paper}
-
-% All submissions must follow the specified format.
-
-% \subsection{Dimensions}
-
-% The text of the paper should be formatted in two columns, with an
-% overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
-% between the columns. The left margin should be 0.75~inches and the top
-% margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
-% whether you print on US letter or A4 paper, but all final versions
-% must be produced for US letter size.
-% Do not write anything on the margins.
-
-% The paper body should be set in 10~point type with a vertical spacing
-% of 11~points. Please use Times typeface throughout the text.
-
-% \subsection{Title}
-
-% The paper title should be set in 14~point bold type and centered
-% between two horizontal rules that are 1~point thick, with 1.0~inch
-% between the top rule and the top edge of the page. Capitalize the
-% first letter of content words and put the rest of the title in lower
-% case.
-
-% \subsection{Author Information for Submission}
-% \label{author info}
-
-% ICML uses double-blind review, so author information must not appear. If
-% you are using \LaTeX\/ and the \texttt{icml2022.sty} file, use
-% \verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
-% will not be printed unless \texttt{accepted} is passed as an argument to the
-% style file.
-% Submissions that include the author information will not
-% be reviewed.
-
-% \subsubsection{Self-Citations}
-
-% If you are citing published papers for which you are an author, refer
-% to yourself in the third person. In particular, do not use phrases
-% that reveal your identity (e.g., ``in previous work \cite{langley00}, we
-% have shown \ldots'').
-
-% Do not anonymize citations in the reference section. The only exception are manuscripts that are
-% not yet published (e.g., under submission). If you choose to refer to
-% such unpublished manuscripts \cite{anonymous}, anonymized copies have
-% to be submitted
-% as Supplementary Material via CMT\@. However, keep in mind that an ICML
-% paper should be self contained and should contain sufficient detail
-% for the reviewers to evaluate the work. In particular, reviewers are
-% not required to look at the Supplementary Material when writing their
-% review (they are not required to look at more than the first $8$ pages of the submitted document).
-
-% \subsubsection{Camera-Ready Author Information}
-% \label{final author}
-
-% If a paper is accepted, a final camera-ready copy must be prepared.
-% %
-% For camera-ready papers, author information should start 0.3~inches below the
-% bottom rule surrounding the title. The authors' names should appear in 10~point
-% bold type, in a row, separated by white space, and centered. Author names should
-% not be broken across lines. Unbolded superscripted numbers, starting 1, should
-% be used to refer to affiliations.
-
-% Affiliations should be numbered in the order of appearance. A single footnote
-% block of text should be used to list all the affiliations. (Academic
-% affiliations should list Department, University, City, State/Region, Country.
-% Similarly for industrial affiliations.)
-
-% Each distinct affiliations should be listed once. If an author has multiple
-% affiliations, multiple superscripts should be placed after the name, separated
-% by thin spaces. If the authors would like to highlight equal contribution by
-% multiple first authors, those authors should have an asterisk placed after their
-% name in superscript, and the term ``\textsuperscript{*}Equal contribution"
-% should be placed in the footnote block ahead of the list of affiliations. A
-% list of corresponding authors and their emails (in the format Full Name
-% \textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
-% Ideally only one or two names should be listed.
-
-% A sample file with author names is included in the ICML2022 style file
-% package. Turn on the \texttt{[accepted]} option to the stylefile to
-% see the names rendered. All of the guidelines above are implemented
-% by the \LaTeX\ style file.
-
-% \subsection{Abstract}
-
-% The paper abstract should begin in the left column, 0.4~inches below the final
-% address. The heading `Abstract' should be centered, bold, and in 11~point type.
-% The abstract body should use 10~point type, with a vertical spacing of
-% 11~points, and should be indented 0.25~inches more than normal on left-hand and
-% right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
-% abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
-% sentences. Gross violations will require correction at the camera-ready phase.
-
-% \subsection{Partitioning the Text}
-
-% You should organize your paper into sections and paragraphs to help
-% readers place a structure on the material and understand its
-% contributions.
-
-% \subsubsection{Sections and Subsections}
-
-% Section headings should be numbered, flush left, and set in 11~pt bold
-% type with the content words capitalized. Leave 0.25~inches of space
-% before the heading and 0.15~inches after the heading.
-
-% Similarly, subsection headings should be numbered, flush left, and set
-% in 10~pt bold type with the content words capitalized. Leave
-% 0.2~inches of space before the heading and 0.13~inches afterward.
-
-% Finally, subsubsection headings should be numbered, flush left, and
-% set in 10~pt small caps with the content words capitalized. Leave
-% 0.18~inches of space before the heading and 0.1~inches after the
-% heading.
-
-% Please use no more than three levels of headings.
-
-% \subsubsection{Paragraphs and Footnotes}
-
-% Within each section or subsection, you should further partition the
-% paper into paragraphs. Do not indent the first line of a given
-% paragraph, but insert a blank line between succeeding ones.
-
-% You can use footnotes\footnote{Footnotes
-% should be complete sentences.} to provide readers with additional
-% information about a topic without interrupting the flow of the paper.
-% Indicate footnotes with a number in the text where the point is most
-% relevant. Place the footnote in 9~point type at the bottom of the
-% column in which it appears. Precede the first footnote in a column
-% with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
-% appear in each column, in the same order as they appear in the text,
-% but spread them across columns and pages if possible.}
-
-% \begin{figure}[ht]
-% \vskip 0.2in
-% \begin{center}
-% \centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
-% \caption{Historical locations and number of accepted papers for International
-% Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
-% Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
-% produced, the number of accepted papers for ICML 2008 was unknown and instead
-% estimated.}
-% \label{icml-historical}
-% \end{center}
-% \vskip -0.2in
-% \end{figure}
-
-% \subsection{Figures}
-
-% You may want to include figures in the paper to illustrate
-% your approach and results. Such artwork should be centered,
-% legible, and separated from the text. Lines should be dark and at
-% least 0.5~points thick for purposes of reproduction, and text should
-% not appear on a gray background.
-
-% Label all distinct components of each figure. If the figure takes the
-% form of a graph, then give a name for each axis and include a legend
-% that briefly describes each curve. Do not include a title inside the
-% figure; instead, the caption should serve this function.
-
-% Number figures sequentially, placing the figure number and caption
-% \emph{after} the graphics, with at least 0.1~inches of space before
-% the caption and 0.1~inches after it, as in
-% \cref{icml-historical}. The figure caption should be set in
-% 9~point type and centered unless it runs two or more lines, in which
-% case it should be flush left. You may float figures to the top or
-% bottom of a column, and you may set wide figures across both columns
-% (use the environment \texttt{figure*} in \LaTeX). Always place
-% two-column figures at the top or bottom of the page.
-
-% \subsection{Algorithms}
-
-% If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
-% environments to format pseudocode. These require
-% the corresponding stylefiles, algorithm.sty and
-% algorithmic.sty, which are supplied with this package.
-% \cref{alg:example} shows an example.
-
-% \begin{algorithm}[tb]
-%   \caption{Bubble Sort}
-%   \label{alg:example}
-% \begin{algorithmic}
-%   \STATE {\bfseries Input:} data $x_i$, size $m$
-%   \REPEAT
-%   \STATE Initialize $noChange = true$.
-%   \FOR{$i=1$ {\bfseries to} $m-1$}
-%   \IF{$x_i > x_{i+1}$}
-%   \STATE Swap $x_i$ and $x_{i+1}$
-%   \STATE $noChange = false$
-%   \ENDIF
-%   \ENDFOR
-%   \UNTIL{$noChange$ is $true$}
-% \end{algorithmic}
-% \end{algorithm}
-
-% \subsection{Tables}
-
-% You may also want to include tables that summarize material. Like
-% figures, these should be centered, legible, and numbered consecutively.
-% However, place the title \emph{above} the table with at least
-% 0.1~inches of space before the title and the same after it, as in
-% \cref{sample-table}. The table title should be set in 9~point
-% type and centered unless it runs two or more lines, in which case it
-% should be flush left.
-
-% % Note use of \abovespace and \belowspace to get reasonable spacing
-% % above and below tabular lines.
-
-% \begin{table}[t]
-% \caption{Classification accuracies for naive Bayes and flexible
-% Bayes on various data sets.}
-% \label{sample-table}
-% \vskip 0.15in
-% \begin{center}
-% \begin{small}
-% \begin{sc}
-% \begin{tabular}{lcccr}
-% \toprule
-% Data set & Naive & Flexible & Better? \\
-% \midrule
-% Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
-% Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
-% Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
-% Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
-% Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
-% Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
-% Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
-% Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
-% \bottomrule
-% \end{tabular}
-% \end{sc}
-% \end{small}
-% \end{center}
-% \vskip -0.1in
 % \end{table}
 
-% Tables contain textual material, whereas figures contain graphical material.
-% Specify the contents of each row and column in the table's topmost
-% row. Again, you may float tables to a column's top or bottom, and set
-% wide tables across both columns. Place two-column tables at the
-% top or bottom of the page.
-
-% \subsection{Theorems and such}
-% The preferred way is to number definitions, propositions, lemmas, etc. consecutively, within sections, as shown below.
-% \begin{definition}
-% \label{def:inj}
-% A function $f:X \to Y$ is injective if for any $x,y\in X$ different, $f(x)\ne f(y)$.
-% \end{definition}
-% Using \cref{def:inj} we immediate get the following result:
-% \begin{proposition}
-% If $f$ is injective mapping a set $X$ to another set $Y$, 
-% the cardinality of $Y$ is at least as large as that of $X$
-% \end{proposition}
-% \begin{proof} 
-% Left as an exercise to the reader. 
-% \end{proof}
-% \cref{lem:usefullemma} stated next will prove to be useful.
-% \begin{lemma}
-% \label{lem:usefullemma}
-% For any $f:X \to Y$ and $g:Y\to Z$ injective functions, $f \circ g$ is injective.
-% \end{lemma}
-% \begin{theorem}
-% \label{thm:bigtheorem}
-% If $f:X\to Y$ is bijective, the cardinality of $X$ and $Y$ are the same.
-% \end{theorem}
-% An easy corollary of \cref{thm:bigtheorem} is the following:
-% \begin{corollary}
-% If $f:X\to Y$ is bijective, 
-% the cardinality of $X$ is at least as large as that of $Y$.
-% \end{corollary}
-% \begin{assumption}
-% The set $X$ is finite.
-% \label{ass:xfinite}
-% \end{assumption}
-% \begin{remark}
-% According to some, it is only the finite case (cf. \cref{ass:xfinite}) that is interesting.
-% \end{remark}
-%restatable
-
-% \subsection{Citations and References}
-
-% Please use APA reference format regardless of your formatter
-% or word processor. If you rely on the \LaTeX\/ bibliographic
-% facility, use \texttt{natbib.sty} and \texttt{icml2022.bst}
-% included in the style-file package to obtain this format.
-
-% Citations within the text should include the authors' last names and
-% year. If the authors' names are included in the sentence, place only
-% the year in parentheses, for example when referencing Arthur Samuel's
-% pioneering work \yrcite{Samuel59}. Otherwise place the entire
-% reference in parentheses with the authors and year separated by a
-% comma \cite{Samuel59}. List multiple references separated by
-% semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
-% construct only for citations with three or more authors or after
-% listing all authors to a publication in an earlier reference \cite{MachineLearningI}.
-
-% Authors should cite their own work in the third person
-% in the initial version of their paper submitted for blind review.
-% Please refer to \cref{author info} for detailed instructions on how to
-% cite your own papers.
-
-% Use an unnumbered first-level section heading for the references, and use a
-% hanging indent style, with the first line of the reference flush against the
-% left margin and subsequent lines indented by 10 points. The references at the
-% end of this document give examples for journal articles \cite{Samuel59},
-% conference publications \cite{langley00}, book chapters \cite{Newell81}, books
-% \cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
-% \cite{mitchell80}, and dissertations \cite{kearns89}.
-
-% Alphabetize references by the surnames of the first authors, with
-% single author entries preceding multiple author entries. Order
-% references for the same authors by year of publication, with the
-% earliest first. Make sure that each reference includes all relevant
-% information (e.g., page numbers).
-
-% Please put some effort into making references complete, presentable, and
-% consistent, e.g. use the actual current name of authors.
-% If using bibtex, please protect capital letters of names and
-% abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
-% in your .bib file.
-
-% \section*{Accessibility}
-% Authors are kindly asked to make their submissions as accessible as possible for everyone including people with disabilities and sensory or neurological differences.
-% Tips of how to achieve this and what to pay attention to will be provided on the conference website \url{http://icml.cc/}.
-
-% \section*{Software and Data}
-
-% If a paper is accepted, we strongly encourage the publication of software and data with the
-% camera-ready version of the paper whenever appropriate. This can be
-% done by including a URL in the camera-ready copy. However, \textbf{do not}
-% include URLs that reveal your institution or identity in your
-% submission for review. Instead, provide an anonymous URL or upload
-% the material as ``Supplementary Material'' into the CMT reviewing
-% system. Note that reviewers are not required to look at this material
-% when writing their review.
-
-% % Acknowledgements should only appear in the accepted version.
-% \section*{Acknowledgements}
-
-% \textbf{Do not} include acknowledgements in the initial version of
-% the paper submitted for blind review.
-
-% If a paper is accepted, the final camera-ready version can (and
-% probably should) include acknowledgements. In this case, please
-% place such acknowledgements in an unnumbered section at the
-% end of the paper. Typically, this will include thanks to reviewers
-% who gave useful comments, to colleagues who contributed to the ideas,
-% and to funding agencies and corporate sponsors that provided financial
-% support.
-
-% % In the unusual situation where you want a paper to appear in the
-% % references without citing it in the main text, use \nocite
-% % \nocite{langley00}
-
-% % \bibliography{example_paper}
-% % \bibliographystyle{icml2022}
-
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % APPENDIX
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \newpage
-% \appendix
-% \onecolumn
-% \section{You \emph{can} have an appendix here.}
-
-% You can have as much text here as you want. The main body must be at most $8$ pages long.
-% For the final version, one more page can be added.
-% If you want, you can use an appendix like this one, even using the one-column format.
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{document}
-
-% % This document was modified from the file originally made available by
-% % Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% % by Iain Murray in 2018, and modified by Alexandre Bouchard in
-% % 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022. 
-% % Previous contributors include Dan Roy, Lise Getoor and Tobias
-% % Scheffer, which was slightly modified from the 2010 version by
-% % Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% % 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% % slightly modified from Prasad Tadepalli's 2007 version which is a
-% % lightly changed version of the previous year's version by Andrew
-% % Moore, which was in turn edited from those of Kristian Kersting and
-% % Codrina Lauth. Alex Smola contributed to the algorithmic style files.
+% \section{SUPPLEMENTARY MATERIAL}
+
+% If you need to include additional appendices during submission, you can include them in the supplementary material file.
+% You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
+% Note that reviewers are under no obligation to examine your supplementary material. 
+% If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.
+
+% You must use \texttt{aistats2022.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
+% The only difference is that it must be in a \emph{single-column} format.
+% You can use \texttt{supplement.tex} in our starter pack as a starting point.
+% Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.
+
+% \section{SUBMISSION INSTRUCTIONS}
+
+% To submit your paper to AISTATS 2022, please follow these instructions.
+
+% \begin{enumerate}
+%     \item Download \texttt{aistats2022.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
+%     Please, do not modify the style files as this might result in a formatting violation.
+
+%     \item Use \texttt{sample\_paper.tex} as a starting point.
+%     \item Begin your document with
+%     \begin{flushleft}
+%     \texttt{\textbackslash documentclass[twoside]\{article\}}\\
+%     \texttt{\textbackslash usepackage\{aistats2022\}}
+%     \end{flushleft}
+%     The \texttt{twoside} option for the class article allows the
+%     package \texttt{fancyhdr.sty} to include headings for even and odd
+%     numbered pages.
+%     \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
+%     \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
+%     \item Upload the PDF file along with other supplementary material files to the CMT website.
+% \end{enumerate}
+
+% \subsection{Camera-ready Papers}
+
+% %For the camera-ready paper, if you are using \LaTeX, please make sure
+% %that you follow these instructions.  
+% % (If you are not using \LaTeX,
+% %please make sure to achieve the same effect using your chosen
+% %typesetting package.)
+
+% If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
+% \begin{enumerate}
+%     %\item Download \texttt{fancyhdr.sty} -- the
+%     %\texttt{aistats2022.sty} file will make use of it.
+%     \item Change the beginning of your document to
+%     \begin{flushleft}
+%     \texttt{\textbackslash documentclass[twoside]\{article\}}\\
+%     \texttt{\textbackslash usepackage[accepted]\{aistats2022\}}
+%     \end{flushleft}
+%     The option \texttt{accepted} for the package
+%     \texttt{aistats2022.sty} will write a copyright notice at the end of
+%     the first column of the first page. This option will also print
+%     headings for the paper.  For the \emph{even} pages, the title of
+%     the paper will be used as heading and for \emph{odd} pages the
+%     author names will be used as heading.  If the title of the paper
+%     is too long or the number of authors is too large, the style will
+%     print a warning message as heading. If this happens additional
+%     commands can be used to place as headings shorter versions of the
+%     title and the author names. This is explained in the next point.
+%     \item  If you get warning messages as described above, then
+%     immediately after $\texttt{\textbackslash
+%     begin\{document\}}$, write
+%     \begin{flushleft}
+%     \texttt{\textbackslash runningtitle\{Provide here an alternative
+%     shorter version of the title of your paper\}}\\
+%     \texttt{\textbackslash runningauthor\{Provide here the surnames of
+%     the authors of your paper, all separated by commas\}}
+%     \end{flushleft}
+%     Note that the text that appears as argument in \texttt{\textbackslash
+%       runningtitle} will be printed as a heading in the \emph{even}
+%     pages. The text that appears as argument in \texttt{\textbackslash
+%       runningauthor} will be printed as a heading in the \emph{odd}
+%     pages.  If even the author surnames do not fit, it is acceptable
+%     to give a subset of author names followed by ``et al.''
+
+%     %\item Use the file sample\_paper.tex as an example.
+
+%     \item The camera-ready versions of the accepted papers are 8
+%       pages, plus any additional pages needed for references.
+
+%     \item If you need to include additional appendices,
+%       you can include them in the supplementary
+%       material file.
+
+%     \item Please, do not change the layout given by the above
+%       instructions and by the style file.
+
+% \end{enumerate}
+
+% \subsubsection*{Acknowledgements}
+% All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
+% To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
+
+% \subsubsection*{References}
+
+% References follow the acknowledgements.  Use an unnumbered third level
+% heading for the references section.  Please use the same font
+% size for references as for the body of the paper---remember that
+% references do not count against your page length total.
+
+% \begin{thebibliography}{}
+% \setlength{\itemindent}{-\leftmargin}
+% \makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
+% \bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
+%     \newblock Performance of a stochastic learning microchip.
+%     \newblock In D. S. Touretzky (ed.),
+%     \textit{Advances in Neural Information Processing Systems 1}, 748--760.
+%     San Mateo, Calif.: Morgan Kaufmann.
+
+% \bibitem{} F.~Rosenblatt (1962).
+%     \newblock \textit{Principles of Neurodynamics.}
+%     \newblock Washington, D.C.: Spartan Books.
+
+% \bibitem{} G.~Tesauro (1989).
+%     \newblock Neurogammon wins computer Olympiad.
+%     \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
+% \end{thebibliography}
+
+% \end{document}

### 2022-08-23 ftvi/ftvi_aistats/supplement.tex
Commit original repos

--- 2022-08-23 ftvi/ftvi_aistats/ftvi_aistats.tex
@@
 \usepackage{aistats2022}
-%\usepackage[accepted]{aistats2022}
-% If your paper is accepted, change the options for the package
-% aistats2022 as follows:
-%
-%\usepackage[accepted]{aistats2022}
-%
-% This option will print headings for the title of your paper and
-% headings for the authors names, plus a copyright note at the end of
-% the first column of the first page.
 
-% If you set papersize explicitly, activate the following three lines:
-%\special{papersize = 8.5in, 11in}
-%\setlength{\pdfpageheight}{11in}
-%\setlength{\pdfpagewidth}{8.5in}
-
-% If you use natbib package, activate the following three lines:
 \usepackage[round]{natbib}
@@
 
+\usepackage{xr}
 \usepackage{amssymb,amsmath,amsthm}
@@
 
+\makeatletter
+\newcommand*{\addFileDependency}[1]{% argument=file name and extension
+  \typeout{(#1)}
+  \@addtofilelist{#1}
+  \IfFileExists{#1}{}{\typeout{No file #1.}}
+}
+\makeatother
+
+\newcommand*{\myexternaldocument}[1]{%
+    \externaldocument{#1}%
+    \addFileDependency{#1.tex}%
+    \addFileDependency{#1.aux}%
+}
+
+\myexternaldocument{ftvi_aistats}
+
+% If your paper is accepted, change the options for the package
+% aistats2022 as follows:
+%
+%\usepackage[accepted]{aistats2022}
+%
+% This option will print headings for the title of your paper and
+% headings for the authors names, plus a copyright note at the end of
+% the first column of the first page.
+
+% If you set papersize explicitly, activate the following three lines:
+%\special{papersize = 8.5in, 11in}
+%\setlength{\pdfpageheight}{11in}
+%\setlength{\pdfpagewidth}{8.5in}
+
+% If you use natbib package, activate the following three lines:
+%\usepackage[round]{natbib}
+%\renewcommand{\bibname}{References}
+%\renewcommand{\bibsection}{\subsubsection*{\bibname}}
+
+% If you use BibTeX in apalike style, activate the following line:
+%\bibliographystyle{apalike}
+
 \begin{document}
@@
 %\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
-\runningtitle{Fat-Tailed VI with Anisotropic Tail Adaptive Flows}
 
-\twocolumn[
+% Supplementary material: To improve readability, you must use a single-column format for the supplementary material.
+\onecolumn
+\aistatstitle{
+Supplementary Material for ``Fat--Tailed Variational Inference with Anisotropic Tail Adaptive Flows''}
 
-\aistatstitle{Fat--Tailed Variational Inference \\ with Anisotropic Tail Adaptive Flows}
-
-\aistatsauthor{ Feynman Liang \And Liam Hodgkinson \And Michael W. Mahoney }
-
-\aistatsaddress{ Department of Statistics\\UC Berkeley \And ICSI and Department of Statistics\\UC Berkeley \And ICSI and Department of Statistics\\UC Berkeley } ]
-
-% \maketitle
-
-\begin{abstract}
-    While fat-tailed densities commonly arise as posterior and marginal
-    distributions in robust models and scale mixtures, they present a
-    problematic scenario where Gaussian-based variational inference fails
-    to accurately capture tail decay. We first improve
-    previous theory on tails of Lipschitz flows \citep{jaini2020tails} by
-    quantifying how they affect the \emph{rate} of tail decay and
-    expanding the theory to non-Lipschitz polynomial flows.
-    Next, we develop an alternative theory for multivariate tail parameters which is sensitive to tail-anisotropy. In doing so, we unveil a
-    fundamental problem which plagues many existing flow-based methods:
-    they can only model tail-isotropic distributions (i.e., distributions having the same tail parameter in every direction).
-    To mitigate this and enable modeling of
-    tail-anisotropic targets, we propose anisotropic tail-adaptive flows (ATAF).
-    Experimental results confirm ATAF on both synthetic and real-world
-    targets
-    is competitive with prior work while also exhibiting appropriate tail-anisotropy.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-
-%\michael{Example}
-%\liam{Example}
-
-Flow based methods 
-\citep{papamakarios2021normalizing}
-have proven to be effective techniques to model complex
-probability densities. They compete with the state of the art on
-density estimation \citep{huang2018neural,durkan2019neural,jaini2020tails},
-generative modeling \citep{chen2019residual,kingma2018glow}, and variational inference \citep{kingma2016improved,agrawal2020advances} tasks.
-These methods start with a random variable $X$ having a simple and tractable
-distribution $\mu$, and apply a learnable transport map $f_\theta$ to build
-another random variable $Y = f_\theta(X)$ with a more expressive \emph{pushforward}
-probability measure $(f_\theta)_\ast \mu$ \citep{papamakarios2021normalizing}.
-In contrast to the implicit distributions \citep{huszar2017variational} produced by generative adversarial networks (GANs), flow based methods restrict the transport map $f_\theta$ to be invertible and to have efficiently-computable Jacobian determinants.
-As a result, probability density functions can be tractably computed
-through direct application of a change of variables
-\begin{align}
-    \label{eq:change-of-variable}
-    p_{Y}(y)
-      = p_{X}(f_\theta^{-1}(y)) \left\lvert \det
-        \left.\frac{d f_\theta^{-1}(z)}{dz} \right\vert_{z=y}
-      \right\rvert
-\end{align}
-
-While recent developments \citep{chen2019residual,huang2018neural,durkan2019neural} have focused primarily
-on the transport map $f_\theta$, the base distribution $\mu$ has received comparatively less investigation. 
-%PREVIOUS%We believe this asymmetric focus is detrimental to the research community because the sensible default choice of Gaussian base distribution $\mu = \cN(0,\mI)$ may result in significant limitations to the expressivity of the model (\Cref{thm:distn_class_closed}).
-%PREVIOUS%\michael{Clarify what does ``asymmetric'' mean in that sentence, does it mean that the focus is more on the transport map than the base distribution, or something else.}
-The most common choice for the base distribution is standard Gaussian $\mu = \cN(0,\mI)$.
-However, in \Cref{thm:distn_class_closed}, we show this choice results in significant
-restrictions to the expressivity of the model, limiting its utility for data that
-exhibits fat-tailed (or heavy-tailed) structure.
-Prior work addressing heavy-tailed flows \citep{jaini2020tails}
-are limited to tail-isotropic base distributions ---
-in \Cref{prop:isotropic-pushforward}, we also prove flows built on these base distributions
-are unable to accurately model multivariate anisotropic fat-tailed structure.
-% \michael{I'm a little confused by this paragraph.
-We cite two results of ours in this paper, but it sounds like prior work, since we then say that we will address those issues.
-Are we saying that prior work did such-and-such, and one of our contributions is to make that explicit and then improve that.
-Or was that known.
-In either case, we should reword here.}
-%PREVIOUS%Addressing this issue was an important aim for recent work \citep{jaini2020tails}, and our work here represents an additional advancement towards this goal.
-
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics{../Figures/pancake.pdf}
-    \vspace{-3mm}
-  \caption{
-    Variational inference against a tail-anisotropic target distribution $\cN(0,1) \otimes \text{StudentT}(\nu=1)$ (top left).
-    Only ATAF (bottom right) is able to correctly reproduce the tail-anisotropy (fat-tailed along $x$-axis,
-    Gaussian along $y$-axis).
-    In contrast, ADVI's (top right) Gaussian base distribution and TAF's (bottom left) tail-isotropic $\prod_{i=1}^2 \text{StudentT}(\nu)$
-    base distribution  can only model tail-isotropic distributions (\Cref{prop:isotropic-pushforward})
-    which erroneously imposes power-law tails with the same rate of decay along both the $x$ and $y$ axes.
-    \vspace{-5mm}
-  }
-  \label{fig:pancake}
-\end{figure*}
-
-Our work here aims to identify and address these deficiencies.
-To understand the impact of the base distribution $\mu$ in flow-based models,
-we develop and apply theory for fat-tailed random variables and their transformations under Lipschitz-continuous functions.
-Our approach leverages the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration} to significantly sharpen and extend prior results
-\citep[Theorem 4]{jaini2019sum} by precisely describing the tail parameters of the pushforward distribution $(f_\theta)_\ast \mu$ under both Lipschitz-continuous (\Cref{thm:distn_class_closed}) and polynomial (\Cref{corr:closure_polynomials}) transport maps.
-% \michael{We say ``characterize'' but what do we mean.
-Probably this is the place to say that it is particularly important to be flexible to have different tail indices in different directions.
-Then some of the later sentences will make sense.}
-%In the multivariate setting, we observe that it is important to be flexible and allow for a multivariate distribution
-%to have different tail parameters in different directions. %Accordingly,
-In the multivariate setting,
-we develop a theory of direction-dependent tail parameters (\Cref{def:mv-tail-param}), and show that tail-isotropic base distributions yield tail-isotropic pushforward measures (\Cref{prop:isotropic-pushforward}).
-As a consequence of \Cref{prop:isotropic-pushforward}, prior methods \citep{jaini2020tails} are limited in that
-they are unable to capture \emph{tail-anisotropy}.
-This motivates the construction of \emph{anisotropic tail adaptive flows} (ATAF, \Cref{def:ataf}) as a means to
-alleviate this issue (\Cref{remark:anisotropic}) and improve modeling of tail-anisotropic distributions.
-% \michael{What limitation is precisely being referred to, since in the previous sentence we say that we get several results in this paper, and then we say how our results are a limitation.
-Probably the best way to do this is to be more detailed, e.g., say that our \Cref{prop:isotropic-pushforward} means that the mapping can't introduce anisotropy.}
-Our experiments show ATAF exhibits correct tail behaviour in synthetic target distributions exhibiting fat-tails (\Cref{fig:cauchy_normal_student}) and tail-anisotropy (\Cref{fig:pancake}).
-On realistic targets,
-% prior work \citep[Table 2]{jaini2020tails} showed tail-adaptive flows (TAF, a subset of ATAF)
-% can improve density modeling and 
-we find that ATAF can yield improvements in variational inference (VI) by capturing potential tail-anisotropy (\Cref{sec:experiments}).
-% \michael{That sentence sounds like we are a footnote on them.
-I would say that we get results for a ragne of cases, both for isotropic and anisotropic cases, thereby improving prior work that considered Gaussian things as well as prior work from \cite{jaini2020tails} that considered isotropic HT.}
-\vspace{-3mm}
-\subsection*{Related Work}
-\vspace{-1mm}
-\paragraph{Fat-tails in variational inference}
-
-% The bulk of related work focuses on fat-tails arising from relaxing priors.
-Recent work in variational autoencoders (VAEs) have considered relaxing Gaussian assumptions to heavier-tailed distributions \citep{mathieu2019disentangling,chen2019residual,boenninghoff2020variational,abiri2020variational}.
-In \citep{mathieu2019disentangling}, a StudentT prior distribution $p(z)$ is considered over the latent code
-$z$ in a VAE with Gaussian encoder $q(z \mid x)$. They argue
-%\footnote{\url{https://github.com/iffsid/disentangling-disentanglement/blob/3396d40f46c34dd928a1241f567a86276b0ff41b/src/main.py\#L52}},
-that the anisotropy of a StudentT product distribution leads to more disentangled representations as compared to the standard choice of Normal distributions.
-A similar modification is performed in \citet{chen2020use}, for
-a coupled VAE (see \citet{cao2019coupled}). It showed improvements in the marginal
-likelihoods of reconstructed images. In addition,
-\citet{boenninghoff2020variational} consider a mixture of StudentTs for the
-prior $p(z)$. To position
-our work in context, note that the encoder $q(z \mid x)$ may be viewed
-as a variational approximation to the posterior $p(z \mid x)$ defined by the
-decoder model $p(x \mid z)$ and the prior $p(z)$. Our work differs from
-\citet{mathieu2019disentangling,chen2020use,boenninghoff2020variational} in
-that we consider fat-tailed variational approximations $q(z \mid x)$ rather
-than priors $p(z)$. Although \citet{abiri2020variational} also considers
-a StudentT approximate posterior, our work involves a more general
-variational family which use normalizing flows.
-% Relaxation of priors to heavy-tailed distributions has numerous
-% applications beyond VAEs. In \cite{silnova2018fast}, the authors perform
-% inference in heavy-tailed probabilistic linear discriminant analysis
-% using Gaussian mean-field variational inference and show improved
-% accuracy in speaker identification. Our work is complementary to these approaches;
-% whereas they consider heavy-tailed priors $p(z)$ we consider heavy-tailed
-% variational families $q(z \mid x)$.
-Similarly, although \citet{wang2018variational} also deals with fat-tails in variational inference,
-their goal is to improve $\alpha$-divergence VI by controlling the moments of importance
-sampling ratios (which may be heavy-tailed). Our work here adopts
-Kullback-Leibler divergence and is concerned with enriching the variational family
-to include anisotropic fat-tailed distributions.
-More directly comparable recent work \citep{ding2011t,futami2017expectation} studies the $t$-exponential family
-variational approximation which includes StudentTs and other
-heavier-tailed densities. Critically, the selection of their parameter $t$ (directly related to the
-StudentT's degrees of freedom $\nu$), and the issue of tail anisotropy, are not discussed. %Other differences include their derivation of
-%expectation-propagation update equations while 
-% We directly backpropagate a noisy
-% ELBO estimate, and introduce a richer variational family used by ATAF.
-
-\vspace{-1mm}
-\paragraph{Flow based methods}
-
-Normalizing flows and other flow based methods have a rich history within variational
-inference \citep{kingma2016improved,rezende2015variational,agrawal2020advances,webb2019improving}.
-Consistent with our experience (\Cref{fig:blr-anisotropic}), \citet{webb2019improving}
-reports that normalizing flows can offer improvements over ADVI and NUTS across thirteen different
-Bayesian linear regression models from \citet{gelman2006data}.
-\citet{agrawal2020advances} shows that normalizing flows compose nicely with other
-advances in black-box VI (e.g., stick the landing, importance weighting).
-However, none of these works treat the issue of fat-tailed targets and inappropriate tail
-decay.
-To our knowledge, only TAFs \citep{jaini2020tails} explicitly consider flows with tails
-heavier than Gaussians. Our work here can be viewed as a direct improvement of \citet{jaini2020tails},
-and we make extensive comparison to this work throughout the body of this paper. At
-a high level, we provide a theory for fat-tails which is sensitive to the rate of
-tail decay and develop a framework to characterize and address the tail-isotropic limitations plaguing
-TAFs.
-
-\vspace{-2mm}
-\section{Flow-based methods for fat-tailed variational inference}
-
-%Continuing from the notation introduced in \Cref{sec:intro}, in this section we
-%finish establishing the VI setting and definitions/notation
-%\michael{I'm not sure what ``Continuing ...'' means.  Can we say ``In this section we do such-and such ...".  Maybe something like:
-%``In this section, we provide background on flow-vased methods for VI, and we describe some of the issues that arise when dealing with fat-tailed data and fat-tailed VI that will be important for understanding our main results.
-%}
-
-\vspace{-2mm}
-\subsection{Flow-based VI methods}
-
-The objective of VI is to approximate a target distribution $\pi(x)$ by searching over
-a \emph{variational family} $\cQ = \{q_\phi : \phi \in \Phi\}$ of probability distributions $q_\phi$.
-While alternatives exist \citep{li2016variational,wang2018variational}, VI typically
-seeks to find $q_\phi$ ``close'' to $\pi$ as measured by Kullback-Leibler divergence $\infdiv{q_\phi}{\pi}$.
-To ensure tractability without sacrificing generality, in practice \citep{wingate2013automated,ranganath2014black}
-a Monte-Carlo approximation of the evidence lower bound (ELBO) is maximized:
-\begin{align*}
-  %-\infdiv{q_\phi}{\pi}
-  %\propto 
-  \text{ELBO}(\phi)
-  &= \int q_\phi(x) \log \frac{\bar\pi(x)}{q_\phi(x)} dx\\
-  & \approx \frac{1}{n} \sum_{i=1}^n \log \frac{\bar\pi(x_i)}{q_\phi(x_i)},\;
-  x_i \simiid q_\phi,\;
-  \bar{\pi} \propto \pi
-\end{align*}
-To summarize, this procedure enables tractable black-box VI
-by replacing $\pi$ with $\bar\pi \propto \pi$ and approximating expectations with respect to $q_\phi$ (which are tractable only in simple variational families) through Monte-Carlo approximation.
-In Bayesian inference and probabilistic programming applications, the target posterior
-$\pi(x) = p(x \mid y) = \frac{p(x, y)}{p(y)}$ is typically intractable but
-$\bar\pi(x) = p(x,y)$ is computable (i.e. represented by the probabilistic program's
-generative / forward execution).
-
-\begin{table*}
-  \centering
-  \begin{tabular}{ccc}
-    \toprule
-    Model                                  & Autoregressive transform                                                                              & Suff. conditions for Lipschitz         \\
-    \midrule
-    NICE\citep{dinh2014nice}               & $z_j + \mu_j \cdot \mathds{1}_{k \not \in [j]}$                                                       & $\mu_j$ Lipschitz                      \\
-    MAF\citep{papamakarios2017masked}      & $\sigma_j z_j + (1 - \sigma_j) \mu_j$                                                                 & $\sigma_j$ bounded                     \\
-    IAF\citep{kingma2016improved}          & $z_j \cdot \exp(\lambda_j) + \mu_j$                                                                   & $\lambda_j$ bounded, $\mu_j$ Lipschitz \\
-    Real-NVP\citep{dinh2016density}        & $\exp(\lambda_j \cdot \mathds{1}_{k \not\in[j]}) \cdot z_j + \mu_j \cdot \mathds{1}_{k \not \in [j]}$ & $\lambda_j$ bounded, $\mu_j$ Lipschitz \\
-    Glow\citep{kingma2018glow}             & $\sigma_j \cdot z_j + \mu_j \cdot \mathds{1}_{k \not\in [j]}$                                         & $\sigma_j$ bounded, $\mu_j$ Lipschitz  \\
-    NAF\citep{huang2018neural}             & $\sigma^{-1}(w^\top \cdot \sigma(\sigma_j z_j + \mu_j))$                                              & Always \par (logistic mixture CDF)          \\
-    NSF\citep{durkan2019neural}            & $z_j \mathds{1}_{z_j \not\in [-B,B]} + M_j(z_j;z_{<j}) \mathds{1}_{x_j \in [-B,B]}$         & Always \par (linear outside $[-B,B]$)     \\
-    FFJORD\citep{grathwohl2018ffjord} & n/a (not autoregressive)                                                                              & Always \par (required for invertibility)    \\
-    ResFlow\citep{chen2019residual} & n/a (not autoregressive)                                                                              & Always \par (required for invertibility)    \\
-    \bottomrule
-  \end{tabular}
-  \caption{Some popular / recently developed flows, the autoregressive transform used in the flow (if applicable),
-  and sufficient conditions conditions for Lipschitz-continuity. A subset of this table was first presented
-  in \citet{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.}
-  \label{tab:flows}
-\end{table*}
-
-While it is possible to construct a variational family $\cQ$ tailored to a specific task, we are interested in VI methods which are more broadly applicable and convenient to use: $\cQ$ should be automatically constructed from introspection of a given probabilistic model/program.
-Automatic differentiation variational inference (ADVI) \citep{kucukelbir2017automatic} is an early implementation of automatic VI and it is still the default in certain probabilistic programming languages \citep{carpenter2017stan}.
-ADVI uses a Gaussian base distribution $\mu$ and a transport map $f_\theta = f \circ \Phi_\text{Affine}$ comprised of an invertible affine transform composed with a deterministic transformation $f$ from $\RR$ to the target distribution's support (e.g., $\exp : \RR \to \RR_{\geq 0}$, $\text{sigmoid} : \RR \to [0,1]$).
-As Gaussians are closed under affine transformations, ADVI's representational capacity is limited to deterministic transformations of Gaussians.
-Hence it cannot represent complex multi-modal distributions.
-To address this, more recent work \citep{kingma2016improved,webb2019improving} replaces the affine map $\Phi_\text{Affine}$ with a flow $\Phi_{\text{Flow}}$ typically parameterized by an invertible neural network:
-\begin{definition}
-    \label[definition]{def:advi}
-  ADVI (with normalizing flows) comprise the variational family
-  $\cQ_\text{ADVI}~\coloneqq~\{
-    (f~\circ~\Phi_\text{Flow})_\ast \mu
-    \}$ 
-  where $\mu = \text{Normal}(0_d, I_d)$,
-  $\Phi_\text{Flow}$ is an invertible flow transform (e.g., \Cref{tab:flows})
-  and $f$ is a deterministic bijection between constrained supports \citep{kucukelbir2017automatic}.
-\end{definition}
-
-As first noted in \citet{jaini2020tails}, the pushforward of a light-tailed Gaussian base distribution under a Lipschitz-continuous flow will remain light-tailed and provide poor approximation to fat-tailed targets.
-Despite this, 
-%at the time of writing this paper 
-many major probabilistic programming packages still make a default choice of Gaussian base distribution (\texttt{AutoNormalizingFlow}/\texttt{AutoIAFNormal} in Pyro \citep{bingham2019pyro}, \texttt{method=variational} in Stan \citep{carpenter2017stan}, \texttt{NormalizingFlowGroup} in PyMC \citep{patil2010pymc}).
-%\michael{What is the verb in the following sentence}
-To address this issue, tail-adaptive flows \citep{jaini2020tails} use a
-base distribution $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$
-where a single degrees-of-freedom $\nu \in \RR$ is used across all $d$ dimensions. More precisely,
-\begin{definition}
-  \label[definition]{def:taf}
-  Tail adaptive flows (TAF) comprise the variational family
-  $\cQ_\text{TAF}
-    \coloneqq \{
-    (f \circ \Phi_\text{Flow})_\ast \mu_\nu
-    \}$
-  where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$ with $\nu$ shared across all $d$ dimensions,
-  $\Phi_{\text{Flow}}$ is an invertible flow,
-  and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
-  During training, the shared degrees of freedom $\nu$ is treated as an additional variational parameter.
-\end{definition}
-
-\vspace{-2mm}
-\subsection{Fat-tailed variational inference}
-
-\vspace{-1mm}
-Fat-tailed variational inference (FTVI)
-% \michael{I'm inclined to say that we should have ``(FTVI)'' and then refer to that acronym everywhere; then we can say: VI is important, FTVI is very important, ADVI did a good job with FTVI, we do a great job with FTVI, and you the reader should follow up on us and do FTVI.
-It's the sort of acronym that is nice to remember.}
-% will do, and add this discussion to future directions in conclusion
-considers the setting where the target $\pi(x)$ is fat-tailed. 
-Such distributions commonly arise during a standard
-``robustification'' approach where light-tailed noise distributions are
-replaced with fat-tailed ones \citep{tipping2005variational}. They also
-appear when weakly informative prior distributions are used in Bayesian
-hierarchical models \citep{gelman2006prior}.
-
-To formalize these notions of fat-tailed versus light-tailed distributions, a quantitative classification for tails is required.
-While prior work classified distribution tails according to quantiles and the existence of moment generating functions
-\citep[Section 3]{jaini2020tails}, here we propose a more natural and finer-grained classification based upon
-the theory of concentration functions \citep[Chapter 1.2]{ledoux2001concentration} which is sensitive to
-the rate of tail decay.
-
-\begin{definition}[Classification of tails]
-    \label[definition]{def:tail-classification}
-    For each $\alpha,p > 0$, we let 
-    \vspace{-1mm}
-    \begin{itemize}[leftmargin=*]
-        \item $\mathcal{E}_\alpha^p$ denote the set of \emph{exponential-type} random variables $X$ with $\mathbb{P}(|X| \geq x) = \Theta(e^{-\alpha x^p})$;
-    \vspace{-1mm}
-        \item $\mathcal{L}_\alpha^p$ denote the set of \emph{logarithmic-type} random variables $X$ with $\mathbb{P}(|X| \geq x) = \Theta(e^{-\alpha(\log x)^p})$.
-    \end{itemize}
-    \vspace{-1mm}
-    % A random variable $X$ is of %\textbf{exponential-type}, denoted $X \in \cE_\alpha^p$, w
-    % \begin{description}
-    %     \item[exponential-type], denoted $X \in \cE_\alpha^p$,
-    %   whenever $\PP(\lvert X \rvert \geq x) = \Theta(e^{-\alpha x^p})$, and
-    %     \item[logarithmic-type], denoted $X \in \cL^p_\alpha$, whenever $\PP(\lvert X \rvert \geq x) = \Theta(e^{-\alpha (\log x)^p})$.
-    % \end{description}
-    In both cases, we call $p$ the \emph{class index} and $\alpha$ the \emph{tail parameter} for $X$.
-    Note that every $\cE_\alpha^p$ and $\cL_\beta^q$ are disjoint, that is,
-    $\cE_\alpha^p \cap \cL_\beta^q = \emptyset$ for all $\alpha,\beta,p,q > 0$.
-    For brevity, we define the ascending families
-    $\overline{\cE_\alpha^p}$ and $\overline{\cL_\alpha^p}$
-    analogously as before except with $\Theta(\cdot)$ replaced
-    by $\cO(\cdot)$. Similarly, we denote the class of distributions with exponential-type
-    tails with class index at least $p$
-    by $\overline{\cE^p} = \cup_{\alpha \in \RR_+} \overline{\cE_\alpha^p}$, and similarly for $\overline{\cL^p}$.
-\end{definition}
-
-For example, $\overline{\cE_\alpha^2}$ corresponds to $\alpha^{-1/2}$-sub-Gaussian random variables,
-    $\overline{\cE_\alpha^1}$ corresponds to sub-exponentials, and (of particular relevance to this paper) $\cL^1_\alpha$ corresponds to the class of power-law distributions.
-
-\vspace{-2mm}
-\section{Tail behavior of Lipschitz flows}
-
-\vspace{-1mm}
-This section states our main theoretical contributions; proofs are deferred to \Cref{sec:proofs}.
-We sharpen previous impossibility results approximating fat-tailed targets
-using thin-tailed base distributions \citep[Theorem 4]{jaini2020tails}
-by characterizing the effects of Lipschitz-continuous transport maps on not only the tail class
-but also the class index and tail parameter (\Cref{def:tail-classification}). Furthermore, we extend the theory
-to include polynomial flows \citep{jaini2019sum}. For the multivariate setting,
-we define the tail-parameter function (\Cref{def:mv-tail-param}) to help formalize the notion
-of tail-isotropic distributions and prove a fundamental limitation that tail-isotropic
-pushforwards remain tail-isotropic (\Cref{prop:isotropic-pushforward}).
-
-Most of our results are developed within the context of Lipschitz-continuous transport maps $f_\theta$.
-In practice, many flow-based methods exhibit Lipschitz-continuity in their transport map either by design \citep{grathwohl2018ffjord,chen2019residual}, or as a consequence of choice of architecture and activation function (\Cref{tab:flows}).
-% or by being enforced stemming from the Banach fixed point theorem
-%
-%or .
-The following assumption encapsulates this premise.
-\begin{assumption}\label[assumption]{assump:lipschitz}
-    $f_\theta$ is invertible, and both $f_\theta$ and $f^{-1}_\theta$
-    are $L$-Lipschitz continuous (e.g., sufficient conditions in \Cref{tab:flows} are satisfied).
-\end{assumption}
-
-\vspace{-2mm}
-\subsection{Closure of tail classes}
-\label{ssec:failure}
-
-\vspace{-1mm}
-Our first set of results pertain to closure of the tail classes in \Cref{def:tail-classification}
-under Lipschitz-continuous transport maps. While earlier work \citep{jaini2020tails} demonstrated
-closure of exponential-type distributions $\cup_{p > 0} \overline{\cE^p}$ under flows satisfying \Cref{assump:lipschitz}, our results in Theorem \ref{thm:distn_class_closed}, and Corollaries \ref{corr:heavy_to_light} and \ref{corr:closure_polynomials} sharpen these observations, showing that (1) Lipschitz transport maps cannot decrease the class index $p$ for exponential-type random variables, but can alter the tail parameter $\alpha$; and
-(2) under additional assumptions, cannot change either class index $p$ or the tail parameter $\alpha$ for logarithmic-type random variables.
-%         tail parameter $\alpha$; 
-% \begin{itemize}[leftmargin=*]
-%     \item cannot decrease the class index $p$ for exponential-type random variables, but can alter the
-%         tail parameter $\alpha$; and
-%     \item under additional assumptions, cannot change either class index $p$ or the tail
-%         parameter $\alpha$ for logarithmic-type random variables.
-% \end{itemize}
-
-\begin{theorem}[Lipschitz maps of tail classes]
-  \label{thm:distn_class_closed}
-  Under \Cref{assump:lipschitz},
-  the distribution classes $\overline{\cE^p}$
-  and $\overline{\cL^p_\alpha}$ (with $p,\alpha > 0$) are closed
-  under every flow transformation in \Cref{tab:flows}.
-\end{theorem}
-
-Informally, \Cref{thm:distn_class_closed} asserts that light-tailed base distributions cannot be transformed
-via Lipschitz transport maps into fat-tailed target distributions.
-Note this does not violate universality theorems for certain flows \citep{huang2018neural}
-as these results only apply in the infinite-dimensional limit.
-Indeed, certain exponential-type families (such as Gaussian mixtures) are dense in the class of \emph{all} distributions, including those that are fat-tailed.
-
-Note that $\overline{\cL^p_\alpha} \supset \cE^q_\beta$ for all $p,q,\alpha,\beta$, so \Cref{thm:distn_class_closed}
-by itself does not preclude transformations of fat-tailed base distributions to light-tailed targets.
-Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushfoward
-hence heavy-to-light transformations are impossible. Note here there is no ascending union over
-tail parameters (i.e., $\cL^p_\alpha$ instead of $\overline{\cL^p_\alpha}$).
-
-\begin{corollary}[Closure of $\cL^p_\alpha$]
-  \label[corollary]{corr:heavy_to_light}
-  If in addition $f_\theta$ is smooth
-  with no critical points on the interior or boundary of
-  its domain, then $\cL_\alpha^p$ is closed.
-\end{corollary}
-
-This implies that simply fixing a fat-tailed base
-distribution \emph{a priori} is insufficient; the tail-parameter(s) of the base distribution must be explicitly optimized alongside
-the other variational parameters during training.
-While these additional assumptions may seem restrictive, note that many flow transforms
-explicitly enforce smoothness and monotonicity \citep{wehenkel2019unconstrained,huang2018neural,durkan2019neural}
-and hence satisfy the premises. In fact, we can show a version of \Cref{thm:distn_class_closed} ensuring closure of exponential-type
-distributions under polynomial transport maps which do not satisfy \Cref{assump:lipschitz}.
-This is significant because it extends the closure results to
-include polynomial flows such as sum-of-squares flows \citep{jaini2019sum}.
-
-\begin{corollary}[Closure under polynomial maps]
-    \label[corollary]{corr:closure_polynomials}
-  For any $\alpha, \beta, p, q \in \RR_+$, there does not exist a
-  finite-degree polynomial map from $\cE_\alpha^p$ into $\cL_\beta^q$.
-\end{corollary}
-
-% Suffices to show for k \in \RR_{> 0}, this covers inverse powers
-
-% \begin{remark}
-%   There does not exist a inverse polynomial map (e.g., sqrt) from $\cL_\alpha$ to $\cE$.
-% \end{remark}
-
-\vspace{-2mm}
-\subsection{Multivariate fat-tails and anisotropic tail adaptive flows}
-
-\vspace{-1mm}
-Next, we restrict attention to power-law tails $\cL^1_\alpha$ and develop a multivariate fat-tailed theory and notions of isotropic/anisotropic tail indices.
-Using our theory, we prove that both ADVI and TAF are fundamentally limited because they
-are only capable of fitting tail-isotropic target measures (\Cref{prop:isotropic-pushforward}).
-We propose anisotropic tail adaptive flows (ATAF) as a density
-modeling method which can represent tail-anisotropic distributions (\Cref{remark:anisotropic}).
-
-% Karamata's theorem enables recovery of a regularly varying random variable's
-% index: $X \in \cL^1_\alpha$ means that $\lim_{x \to \infty} \frac{x \PP(X \geq
-%     x)}{\int_x^\infty \PP(X \geq t) dt} = \alpha$ for $\alpha < -1$.
-% Motivated by Karamata's theorem, we introduce
-% the following definition to aid in describing multivariate fat-tailed random variables:
-
-For example, consider the target distribution shown earlier in \Cref{fig:pancake} formed as the product of $\mathcal{N}(0,1)$ and $\text{StudentT}(\nu=1)$ distributions.
-The marginal/conditional distribution along a horizontal slice (e.g., the distribution of $\braket{X,e_0}$)
-is fat-tailed, while along a vertical slice (e.g., $\braket{X,e_1}$) it is Gaussian.
-Another extreme example of tail-anisotropy where the tail parameter for
-$\braket{X,v}$ is different in every direction $v \in \cS^{1}$
-is given in \Cref{fig:radial-fat-tail}. Here $\mathcal{S}^{d-1}$ denotes the $(d-1)$-sphere in $d$ dimensions. 
-Noting that the tail parameter depends on the choice of direction, we are motivated to consider
-the following direction-dependent definition of multivariate tail parameters.
-
-\begin{definition}
-  \label[definition]{def:mv-tail-param}
-  For a $d$-dimensional random vector $X$,
-  its \emph{tail parameter function} $\alpha_X : \cS^{d-1} \to \bar{\RR}_+$
-  is defined as
-  % $\alpha_X(v) = \lim_{x \to \infty} \frac{x \PP(\braket{v,X} \geq x)}{\int_x^\infty \PP(\braket{v,X} \geq t) dt}$.
-  $\alpha_X(v) = -\lim_{x \to \infty} \log \PP(\braket{v,X} \geq x) / \log x$ when the limit exists, and $\alpha_X(v) = +\infty$ otherwise.
-  In other words, $\alpha_X(v)$ maps directions $v$ into the tail parameter of the corresponding one-dimensional projection $\braket{v,X}$.
-The random vector $X$ is \emph{tail-isotropic} if $\alpha_X(v) \equiv c$ is constant and
-  \emph{tail-anisotropic} if $\alpha_X(v)$ is not constant but bounded.
-\end{definition}
-
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics{../Figures/radial-fat-tail.pdf}%
-%   \includegraphics[trim={0 0 7cm 0},clip]{../Figures/radial-fat-tail.pdf}\\%
-%   \vspace{-8mm}
-%   \includegraphics[trim={7.2cm 0 0 0},clip]{../Figures/radial-fat-tail.pdf}%
-  %\includegraphics[width=\textwidth]{../Figures/radial-fat-tail.png}%
-  %\input{../Figures/radial-fat-tail.pgf}
-  \vspace{-3mm}
-  \caption{
-    Illustration of the direction-dependent tail-parameter function (bottom) on a tail-anisotropic distribution (top)
-    with PDF $\dd P(r,\theta) = r^{-\alpha(\theta)} r \dd r \dd\theta$ and tail parameter $\alpha(\theta) = 2 + \cos(2\theta)$.
-    While prior fat-tailed theory based on $\|X\|_2 = \sup_{\|v\|_2 = 1} \braket{X,v}$
-    is only sensitive to the largest tail parameter $\max_{\theta \in [0, 2\pi]} \alpha(\theta) = 3.0$,
-    our direction-dependent tail parameter function (bottom, red line)
-    and its values along the standard basis axes ($\alpha(0)$ and $\alpha(\pi/2)$)
-    capture \emph{tail-anisotropy}.
-  }
-  \label{fig:radial-fat-tail}
-\end{figure*}
-% \begin{definition}
-%   \label[definition]{def:isotropic}
-%   Multivarate random variable $X$ is tail-isotropic if $\alpha_X(v) \equiv c$ is constant.
-%   Otherwise, it is tail-anisotropic.
-% \end{definition}
-
-% \begin{align*}
-%   \PP[\braket{X,u} \geq x]
-%   = \PP[\braket{\mu,u} + R \braket{AU,u} \geq x]
-%   = \PP[R \geq \frac{x - \braket{\mu,u}}{\braket{u,AU}}]
-%   \sim C_1 \left(\frac{x - \braket{\mu,u}}{\braket{u,AU}}\right)^{-\alpha}
-%   \sim C_2 x^{-\alpha}
-% \end{align*}
-% and
-% theory around tail parameters is developed for elliptically
-% contoured multivariate distributions $X = \mu + R A U$
-% for some heavy-tailed random variable $R$, fixed vectors $\mu \in \RR^d$
-% and $U \in S^{d-1}$, and $A$ a (Cholesky factor)
-% defining the ellipsoid axes.
-
-% While elliptically contoured multivariate distributions admit a
-% straightforward generalization from the scalar case, they are severely
-% limited in practical applications.  One fundamental limitation of
-% elliptically contoured $X$ is that the tail parameter is the same for every
-% 1-dimension projection.  If $R$ has tail index $\alpha$ then for all for $u
-%   \in S^{d-1}$
-% \begin{align*}
-%   \PP[\braket{X,u} \geq x]
-%   = \PP[\braket{\mu,u} + R \braket{AU,u} \geq x]
-%   = \PP[R \geq \frac{x - \braket{\mu,u}}{\braket{u,AU}}]
-%   \sim C_1 \left(\frac{x - \braket{\mu,u}}{\braket{u,AU}}\right)^{-\alpha}
-%   \sim C_2 x^{-\alpha}
-% \end{align*}
-
-% Furthermore, the assumption of elliptical distributions are
-% easily violated in many common and useful applications.  For example, in
-% probabilistic programming collections of random variables are oftentimes
-% grouped together into a single multivariate random variable (e.g.,  blocked
-% Gibbs, Hamiltonian Monte Carlo).  In fact, $Q_{TAF}$ (used in the experiments
-% in \citep{jaini2020tails}) uses a StudentT product base distribution which
-% is not elliptically symmetric.
-
-% \begin{figure}[htbp]
-%   \centering
-%   \includegraphics[width=0.4\textwidth]{../Figures/studentt-product.png}
-%   \caption{
-%     Illustrating multivariate fat-tails using a 2-dimensional StudentT.
-%     The multivariate theory from \citep{jaini2020tails} is for elliptically
-%     contoured distributions and not applicable to the StudentT product base distributions
-%     used in their experiments.
-%   }
-%   \feynman{Consider just using words here to delineate from our contributions, this is not a super important fact}
-%   \label{fig:studentt-product}
-% \end{figure}
-
-% Elliptical distributions are \emph{tail isotropic} i.e., $\alpha(v) \equiv c$ is constant.
-% The base distribution for \citep{jaini2020tails}, $\prod_1^d \text{StudentT}(\nu_i)$ with $\nu_i \equiv \nu$,
-% is also tail isotropic because
-% $\braket{v,X}$ is a sum of StudentT and by tail index algebra
-% has index $\max_i \nu_i = \nu$.
-
-Of course, one can construct pathological densities where this definition is not effective
-(see \Cref{eg:spiral}), but we will find it to suffice for our purposes. It is illustrative to contrast to the theory presented for TAF \citep{jaini2020tails}
-where only the tail exponent of $\|X\|_2$ is considered.
-For $X = (X_1, \ldots, X_d)$ with $X_i \in \cL^1_{\alpha_i}$, 
-by Fatou-Lebesgue and \Cref{lem:sum-rule}
-\begin{align*}
-  &\PP[\|X\|_2 \geq t]
-  = \PP\left[\sup_{z \in \cS^{d-1}} \braket{X,z} \geq t\right]\\
-  &\quad \geq \sup_{z \in \cS^{d-1}} \PP[\braket{X,z} \geq t]
-  = \max_{1 \leq i \leq d} \nu_i
-  = \max_{0 \leq i \leq d-1} \alpha_X(e_i)
-\end{align*}
-Therefore, considering only the tail exponent of $\|X\|_2$ is equivalent to summarizing $\alpha_X(\cdot)$ by an upper bound.
-Given the absence of the tail parameters for other directions (i.e., $\alpha_X(v) \neq \sup_{\|v\|=1} \alpha_X(v)$)
-in the theory for TAF \citep{jaini2020tails}, it should be unsurprising that both their multivariate 
-theory as well as their experiments only consider tail-isotropic distributions obtained either
-as an elliptically-contoured distribution with fat-tailed radial distribution or 
-$\prod_{i=1}^d \text{StudentT}(\nu)$ (tail-isotropic by \Cref{lem:sum-rule}). 
-Our next proposition shows that this presents a significant limitation when the target distribution is
-tail-anisotropic.
-
-\begin{proposition}[Pushforwards of tail-isotropic distributions]
-  \label[proposition]{prop:isotropic-pushforward}
-  Let $\mu$ be tail isotropic with non-integer parameter $\nu$
-  and suppose $f_\theta$ satisfies \Cref{assump:lipschitz}.
-  Then $(f_\theta)_\ast \mu$ is tail isotropic with parameter $\nu$.
-\end{proposition}
-
-% However, $\alpha(v)$ is non-trivial to work with; it is an asymptotic quantity and is defined
-% for uncountably many $v \in S^{d-1}$. In this work, we propose approximating
-% $\alpha(v)$ using the standard basis vectors:
-
-% \begin{definition}
-%   The \emph{standard basis tail parameters} of a fat-tailed $X \in \RR^d$
-%   is $\{\alpha(e_i) : i \in [d]\}$ where $\alpha$ is defined in
-%   \Cref{def:mv-tail-param} and $e_i$ is the $i$th standard basis vector.
-% \end{definition}
-
-% The standard basis vectors provide a natural choice of projections for
-% multivariate product distributions (as commonly encountered during blocking /
-% grouping of random variables). Going back to our previous example,
-% we still have that $\alpha_{\|X\|} = \max_i \alpha(e_i)$ but now the tail
-% indices of the remaining $e_j$ need not be equal. Admittedly, the standard basis
-% is less suitable for correlated multivariate distributions. For example,
-% let $t_i \sim \text{StudentT}(\nu_i)$ and consider the rotated random variable
-% $X = R_{\pi/4} [t_1; t_2]$. Its standard basis tail parameters are
-% \begin{align*}
-%   \Pr[\braket{X,e_1} > x]
-%   = \Pr[\braket{X,e_2} > x]
-%   = \Pr[0.5 t_1 + 0.5 t_2 > x]
-%   \leq \max_i \nu_i
-% \end{align*}
-% This example illustrates that the standard basis tail parameters
-% provide multivariate tail parameters which are no worse than
-% previous work.
-
-% \subsection{Limitations of our theory}
-
-% \begin{itemize}
-%   \item Standard basis parameters not general, only works for axis-aligned tails
-%   i.e., independent product distributions
-%   \item Still fails on the spiral, but it should do better than TAF
-%   \item Theory not fully general, only considers rays from origin rather than
-%   path integrals
-% \end{itemize}
-
-To work around this limitation without relaxing \Cref{assump:lipschitz}, it is evident
-that tail-anisotropic base distributions $\mu$ must be considered.
-Perhaps the most straightforward modification to incorporate a tail-anisotropic base distribution replaces TAF's isotropic base distribution $\prod_{i=1}^d \text{StudentT}(\nu)$
-with $\prod_{i=1}^d \text{StudentT}(\nu_i)$. Note that $\nu$ is no longer shared across dimensions,
-enabling $d$ different tail parameters to be represented:
-
-\begin{definition}\label[definition]{def:ataf}
-  Anisotropic Tail-Adaptive Flows (ATAF) comprise the variational family
-  $\cQ_\text{ATAF}~\coloneqq~\{
-    (f \circ \Phi_\text{Flow})_\ast \mu_\nu
-    \},$
-  where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu_i)$, each $\nu_i$ is \emph{distinct}, and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
-  Analogous to \cite{jaini2020tails}, ATAF's implementation
-  treats $\nu_i$ identically to the other parameters in the flow and jointly optimizes over them.
-\end{definition}
-
-\begin{remark}\label[remark]{remark:anisotropic}
-  Anisotropic tail-adaptive flows can represent tail-anisotropic distributions with up to $d$ different
-  tail parameters while simultaneously satisfying \Cref{assump:lipschitz}.
-  For example, if $\Phi_\text{Flow} = \text{Identity}$ and $\mu_\nu = \prod_{i=1}^d \text{StudentT}(i)$
-  then the pushforward $(\Phi_\text{Flow})_\ast \mu_\nu = \mu_\nu$ is tail-anisotropic.
-  % Moreover, its standard basis tail parameters are equal
-  % (up to permutation) to those for $\mu$.
-\end{remark}
-
-% Compared to \citep{jaini2020tails}, the degrees of freedom $\nu$ is no longer
-% shared across all $d$ dimensions so the variational approximations may exhibit
-% varying degrees of fat-tailedness across different dimensions. As seen in
-% \Cref{fig:pancake}, such anisotropy is particularly important in situations
-% such as the ``heavy-tailed pancake.'' This situation is particularly relevant in
-% probabilistic programming, where multiple latent variariables (potentially of different
-% tail index) are blocked together for joint sampling / approximation.
-
-\vspace{-2mm}
-\subsection{Discussion of limitations}
-\label{ssec:limitations}
-
-\vspace{-1mm}
-
-Additionally, ATAF only uses a finite collection $\{\nu_i\}_{i=1}^d$ of tail parameters
-whereas the tail-parameter function may take on more than $d$ values. \Cref{fig:radial-fat-tail}
-presents a worst-case scenario, where the target distribution has a continuum of tail parameters.
-As ATAF is only capable of representing $d$ different tail parameters in its pushforward
-(\Cref{remark:anisotropic}), it will necessarily incur an approximation error when $> d$ different
-tail parameters are present.
-Furthermore, distributions supported on domains other than Euclidian $\mathbb{R}^d$ may require an additional bijection between supports (e.g. $\exp : \mathbb{R} \to \mathbb{R}_+$) which
-may violate \cref{assump:lipschitz}.
-Finally, note that $\text{StudentT}(\nu) \Rightarrow \cN(0,1)$ 
-as $\nu \to \infty$ so we should expect ATAF to provide reasonably good approximations to target 
-distributions in $\overline{\cE^2}$ by taking $\nu$ sufficiently large. We observed this occurs in 
-practice empirically in \Cref{sec:normal-normal-location-mixture}.
-% We found that applying ATAF when the target is light tailed results in minimal
-% error (\Cref{sec:normal-normal-location-mixture}); the result agrees with
-% intuition because $\text{StudentT}(\nu) \to N(0,1)$ as $\nu \to \infty$ so
-% it is reasonable to expect ATAF to learn reasonable approximations.
-
-% \subsection{Non-parametric tail index estimation}
-% \label{ssec:non-param-index}
-
-% To perform variational inference with $\cQ_{ATAF}(\nu,\Phi)$, note that if
-% $\Phi$ is Lipschitz then the tail index of any element of
-% $\cQ_{ATAF}(\nu,\Phi)$ depends only on $\nu$. This motivates a two step
-% procedure where (1) $\hat\nu$ estimates the degrees of freedom and
-% (2) standard normalizing-flow variational inference $\max_\Phi \cQ_{ATAF}(\hat\nu,\Phi)$ is performed. The key advantage here is that both steps are standard,
-% so existing techniques for tail index estimation may be applied.
-
-% Let $X$ be a random variable with fat-tailed density $\pi(x)$
-% with tail index $\alpha$ which we wish to estimate.
-% When samples $X_i \simiid \pi$ are tractable
-% (e.g., no \texttt{observe} or \texttt{condition} statements), traditional
-% tail-index estimation techniques may be applied:
-
-% \begin{figure}[H]
-%   \centering
-%   \includegraphics[width=0.6\textwidth]{../Figures/hill.png}
-%   \caption{Hill estimator against $\nu=1.5$ StudentT}
-%   \label{fig:hill}
-% \end{figure}
-
-% Oftentimes $\pi$ is the posterior distribution of latent variable $X$
-% after conditioning on observations $Y=y$, in which case
-% $\pi(x) = p(X=x \mid Y=y) = \frac{p(x,y)}{p(y)}$
-% and sampling from $\pi$ is non-trivial.
-% To circumvent this, we propose the following asymptotically correct
-% estimator. Note for $a, b$ large:
-% \begin{align*}
-%   \Pr[X > a]                 & \sim b^{-\alpha}                                           \\
-%   \pi(a)                     & \sim \alpha a^{-\alpha - 1}                                \\
-%   \log \pi(a)                & \sim \log \alpha - (\alpha + 1) \log a                     \\
-%   \log \frac{p(a,y)}{p(b,y)} & = \log \frac{\pi(a)}{\pi(b)}
-%   \sim (\alpha + 1) \log \frac{b}{a}                                                      \\
-%   \alpha                     & \sim \frac{\log p(a,y) - \log p(b,y)}{\log b - \log a} - 1
-% \end{align*}
-% The joint density $p(x,y)$ is tractably computed by running the
-% probabilistic program forwards.
-% For example. setting $a=10$ and $b=20$ yields an estimate of $1.4799$ for a
-% $\nu=1.5$ StudentT.
-
-% \subsubsection{Approximating 1D-marginals}
-
-% This is an example of a two-point log-linear extrapolation at points $a$ and $b$.
-% \feynman{Consider Richardson extrapolation to accelerate this limit. Rate of convergece in limit must be a power}
-
-% In general, PPs may include several latent variables.
-% The 1-dimensional marginal $p(x, y) = \int_{\setminus x} p(x,z,y) dz$
-% is required to apply the above interpolation method.
-% We propose approximating this integral using discrete samples.
-% If $z$ is a descendent of $y$, then
-% we can compute $p(x,y)$ by executing the probabilistic program
-% (possibly terminating early).
-% If $z$ is an ancestor of $x$ (hence also an ancestor of $y$), then
-% \begin{align*}
-%   p(x,y) \approx \frac{1}{N} \sum_{i=1}^N p(y, x \mid z_i)
-%   \qquad z_i \simiid p(z)
-% \end{align*}
-% In the last case where $x \rightarrow z \rightarrow y$,
-% we have
-% \begin{align*}
-%   p(x,y) \approx p(x) \frac{1}{N} \sum_{i=1}^n p(y \mid x, z_i)
-%   \qquad z_i \simiid p(z_i \mid x)
-% \end{align*}
-
-% \begin{figure}[H]
-%   \centering
-%   \includegraphics[width=0.9\textwidth]{../Figures/our-estimator.png}
-%   \caption{Tail index estimation on location-scale mixture representation
-%     for StudentT, where the mixture is discretized to $n$ components.}
-%   \label{fig:approx_marg}
-% \end{figure}
-
-\vspace{-2mm}
-\section{Experiments}
-\label{sec:experiments}
-
-\vspace{-1mm}
-Here we validate ATAF's ability to improve
-a range of probabilistic modeling tasks.
-Prior work \citep{jaini2020tails} demonstrated improved
-density modelling when fat tails are considered, and
-our experiments are complementary by evaluating TAFs and ATAFs for variational inference tasks as well as demonstrating the necessity of tail-anisotropy for modelling real-world financial returns and insurance claims datasets.
-We implement using the 
-%\texttt{beanmachine} probabilistic programming language\citep{tehrani2020bean}
-\texttt{REDACTED FOR REVIEW} probabilistic programming language[REDACTED]
-and the
-%\texttt{flowtorch} library for normalizing flows\citep{flowtorchai},
-\texttt{REDACTED FOR REVIEW} library for normalizing flows[REDACTED],
-%and we have open-sourced the code\citep{ghlic}
-and we have open-sourced code for reproducing experiments[REDACTED, reviewers please see supplementary materials for code]. 
-Additional details for the experiments are detailed in \Cref{sec:additional-exp-details}.
-
-% These experiments investigate the behavior of neural density estimators with
-% \emph{heavy-tailed base distribution}.
-% Specifically, we consider a masked autoregressive flow \cite{papamakarios2017masked}
-% transform of a generalized Student's t distribution as a density estimator $q_\theta(X)$
-% in a variational inference framework. To fit $q_\theta$ to a target distribution
-% $\pi$, the ELBO gradient is reparameterized and Monte-Carlo approximated
-% \begin{align*}
-%   \nabla_\theta \EE_{q_\theta} \log \frac{\pi(X)}{q_\theta(X)}
-%   & = \nabla_\theta \EE_p \log \frac{
-%     \pi(X)
-%   }{p_\theta(f_\theta^{-1}(X))
-%   \left\lvert \det \nabla f_\theta^{-1}(X) \right\rvert}    \\
-%   & = \EE_p \nabla_\theta \log \frac{
-%     \pi(X)
-%   }{p_\theta(f_\theta^{-1}(X))
-%   \left\lvert \det \nabla f_\theta^{-1}(X) \right\rvert}    \\
-%   & \approx \frac{1}{n} \sum_i^n \nabla_\theta \log \frac{
-%     \pi(x_i)
-%   }{p_\theta(f_\theta^{-1}(x_i))
-%     \left\lvert \det \nabla f_\theta^{-1}(x_i) \right\rvert}
-% \end{align*}
-
-\begin{figure*}[htbp]
-  \centering
-  \vspace{-0.5cm}
-  \includegraphics{../Figures/blr_aniso.pdf}
-  \vspace{-0.3cm}
-  \caption{
-    Bayesian linear regression's tail-anisotropic posterior
-    (top left) exhibits a fat-tailed conditional in $\sigma$ (as evidenced by
-    the convex power-law decay in the top middle panel) and a Gaussian conditional in $\beta$ (concave graph in top right panel).
-    While all methods appear to provide a good approximation of the bulk (left column),
-    \Cref{prop:isotropic-pushforward} implies
-    Gaussian (Gaussian, second row) or isotropic StudentT product (TAF, third row) base distribution
-    yields Gaussian or power-law tails respectively for \emph{both} $\sigma$ and $\beta$.
-    In contrast, ATAF (bottom row) illustrates \Cref{remark:anisotropic} by simultaneously
-    modeling a power-law tail on $\sigma$ and Gaussian tail on $\beta$.
-  }
-  \label{fig:blr-anisotropic}
-\end{figure*}
-
-\begin{table*}[htbp]
-  \centering
-%   \includegraphics[width=1.0\textwidth]{../Figures/eight_schools.png}
-    \begin{subfigure}[t]{0.49\textwidth}
-        \centering
-        \begin{tabular}{rcc}
-            \toprule
-                      & ELBO                & $\log p(y)$       \\
-            \midrule
-            ADVI      & $2873.90 \pm 6.95$    & $2969.73 \pm 1.73$ \\
-            TAF       & $2839.64 \pm 9.10$    & $2973.85 \pm 0.87$ \\
-            ATAF      & $2842.75 \pm 8.83$    & $2976.75 \pm 0.66$ \\
-            NUTS      & n/a                  & $3724.59 \pm 0.036$ \\
-            \bottomrule
-        \end{tabular}
-        \caption{diamonds}
-        \label{tab:diamonds}
-    \end{subfigure}
-    \begin{subfigure}[t]{0.49\textwidth}
-        \centering
-        \begin{tabular}{rcc}
-            \toprule
-                      & ELBO                & $\log p(y)$       \\
-            \midrule
-            ADVI      & $-72.13 \pm 6.89$    & $-53.25 \pm 3.44$ \\
-            TAF       & $-64.64 \pm 4.88$    & $-52.51 \pm 4.41$ \\
-            ATAF      & $-58.63 \pm 4.75$    & $-51.01 \pm 3.71$ \\
-            NUTS      & n/a                  & $-47.78 \pm 0.093$ \\
-            \bottomrule
-        \end{tabular}
-        \caption{Eight schools}
-        \label{tab:eight_schools}
-    \end{subfigure}
-
-        \caption{Monte-Carlo ELBO and importance weighted Monte-Carlo marginal likelihood
-        $p(y) = \EE_{x \sim q_\theta} \frac{p(x,y)}{q_\theta(x)}$ (higher is better, $\pm$ standard errors) 
-        estimates from VI on real-world datasets.
-        For comparison, we include results computed using ``golden samples'' from \texttt{posteriordb} \citep{ghposteriordb} (NUTS).
-        These golden samples are obtained using No-U-Turn-Sampling \citep{hoffman2014no,carpenter2017stan},
-        a state-of-the-art MCMC method which does not optimize ELBO, is more computationally expensive than
-        VI, and only produces samples rather than parametric probability density functions.
-    }
-  \label{fig:eight_schools}
-  \vspace{-4mm}
-\end{table*}
-
-\begin{table*}[htbp]
-    \centering
-    \begin{tabular}{rcc}
-        \toprule
-                  & Fama-French 5 Industry Daily & CMS 2008-2010 DE-SynPUF       \\
-        \midrule
-        ADVI      & $-5.018 \pm 0.056$    & $-1.883 \pm 0.012$ \\
-        TAF       & $-4.703 \pm 0.023$    & $-1.659 \pm 0.004$ \\
-        ATAF      & $-4.699 \pm 0.024$    & $-1.603 \pm 0.034$ \\
-        \bottomrule
-    \end{tabular}
-    \caption{
-        Log-likelihoods (higher is better, $\pm$ standard errors, $10$ trials) achieved on
-        density modeling tasks involving real-world financial returns \citep{fama2015five} and insurance claims \citep{cms} data. 
-    }
-    \label{tab:density-estimation}
-\end{table*}
-
-\subsection{Bayesian linear regression}
-
-Consider one-dimensional Bayesian linear regression (BLR)
-with conjugate priors, defined by priors and likelihood
-\begin{gather*}
-    \sigma^2 \sim \text{Inv-Gamma}(a_0, b_0)\\
-    \beta \mid \sigma^2 \sim \cN(0, \sigma^2),\qquad
-    y \mid X, \beta, \sigma \sim \cN(X \beta, \sigma^2) 
-\end{gather*}
-where $a_0$, $b_0$ are hyperparameters and the task is to approximate the posterior
-distribution $p(\beta,\sigma^2 \mid X, y)$. Owing to conjugacy,
-the posterior distribution can be explicitly computed.
-Indeed, $p(\beta,\sigma^2 \mid X, y) = \rho(\sigma^2)\rho(\beta \mid \sigma)$ where $\rho(\beta \mid \sigma) = \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 \Sigma_n)$, $\Sigma_n = (X^\top X + \sigma^{-2})^{-1}$, $\hat\beta = (X^\top X)^{-1} X^\top y$, and
-%\begin{align*}
-    %p(\beta,\sigma^2 \mid X, y) &= \rho(\sigma^2) \rho(\beta \mid \sigma) \\
-    \[
-    \rho(\sigma^2) = \text{Inv-Gamma}\bigg(
-    %\underbrace{a_0 + \frac{n}{2}}_{\eqqcolon a_n}, 
-    a_0 + \frac{n}{2}, 
-    b_0 + \frac{1}{2}(y^\top y - \mu_n^\top \Sigma_n \mu_n)\bigg)% \\
-    \]
-    %\rho(\beta \mid \sigma) &= \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 (X^\top X + \sigma^{-2} I)^{-1})
-%\end{align*}
-This calculation reveals that the posterior distribution is tail-anisotropic:
-for fixed $c$ we have that $p(\sigma^2, \beta=c \mid X, y) \propto \rho(\sigma^2) \in \cL^1_{\alpha_n}$
-as a function of $\sigma$ (with $\alpha_n$ a function of $n$)
-and $p(\sigma^2=c, \beta \mid X, y) \propto \rho(\beta \mid c) \in \overline{\cE^2}$
-as a function of $\beta$.
-As a result of \Cref{prop:isotropic-pushforward}, we expect ADVI and TAF to erroneously impose
-Gaussian and power-law tails respectively for both $\beta$ and $\sigma^2$ as neither method
-can produce a tail-anisotropic pushforward. This intuition is confirmed in \Cref{fig:blr-anisotropic},
-where we see that only ATAF is the only method capable of modeling the tail-anisotropy present.
-
-\subsection{Diamond price prediction using non-conjugate Bayesian regression}
-\label{ssec:diamonds}
-
-Without conjugacy, the BLR posterior is intractable and there is no reason \emph{a priori} to expect tail-anisotropy.
-Regardless, this presents a realistic and practical scenario for evaluating ATAF's ability to improve VI.
-For this experiment, we consider BLR on the \texttt{diamonds} dataset \citep{wickham2011ggplot2} included in
-\texttt{posteriordb} \citep{ghposteriordb}.
-This dataset contains a covariate matrix $X \in \RR^{5000 \times 24}$ consisting of $5000$
-diamonds each with $24$ features as well as an outcome variable $y \in \RR^{5000}$ representing each diamond's price.
-The probabilistic model for this inference task is specified in Stan code provided by \citep{ghposteriordb} and is reproduced
-for convenience
-\begin{gather*}
-    \alpha \sim \text{StudentT}(\nu=3, \text{loc}=8, \text{scale}=10)\\
-    \sigma \sim \text{HalfStudentT}(\nu=3, \text{loc}=0, \text{scale}=10)\\
-    \beta \sim \cN(0, \mI_{24}),\qquad
-    y \sim \cN(\alpha + X \beta, \sigma)
-\end{gather*}
-
-For each VI method, we performed 100 trials each consisting of 5000 descent steps
-on the Monte-Carlo ELBO estimated using 1000 samples and report the results in
-\Cref{tab:diamonds}. We report both the final Monte-Carlo ELBO
-as well as a Monte-Carlo importance-weighted approximation to
-the log marginal likelihood $\log p(y) = \log \EE_{x \sim q_\theta} \frac{p(x,y)}{q_\theta(y)}$
-both estimated using 1000 samples.
-
-% \subsection{Bayesian Robust Linear Regression}
-
-% $n = 100$, $d = 10$.
-
-% $X_{ij} \simiid N(0,1)$ for $i \in [n]$, $j \in [d]$.
-
-% $y_i \simiid \text{StudentT}(\text{loc}=X \beta, df=5)$
-
-% Improper ``flat'' prior on $\beta$ to ensure heavy-tailed posterior.
-
-% \begin{figure}[H]
-%   \centering
-%   \includegraphics[width=0.9\textwidth]{../Figures/blr.png}
-%   \caption{VI of Bayesian robust linear regression}
-%   \label{fig:blr}
-% \end{figure}
-
-\subsection{Eight schools SAT score modelling with fat-tailed scale mixtures}
-\label{ssec:eight_schools}
-
-The eight-schools model \citep{rubin1981estimation,gelman2013bayesian} is a classical
-Bayesian hierarchical model used originally to consider the relationship between standardized
-test scores and coaching programs in place at eight schools.
-A variation utilizing half Cauchy non-informative priors \citep{gelman2006prior} provides
-a real-world inference problem involving fat-tailed distributions, and is formally specified
-by the probabilistic model
-\begin{gather*}
-    \tau \sim \text{HalfCauchy}(\text{loc}=0, \text{scale}=5)\\
-    \mu \sim \cN(0, 5),\qquad
-    \theta \sim \cN(\mu, \tau),\qquad
-    y \sim \cN(\theta, \sigma)
-\end{gather*}
-Given test scores and standard errors $\{(y_i, \sigma_i)\}_{i=1}^8$, we are interested in the
-posterior distribution over treatment effects $\theta_1,\ldots,\theta_d$. The experimental
-parameters are identical to \Cref{ssec:diamonds} and results are reported in \Cref{tab:eight_schools}.
-
-\subsection{Tail-anisotropic modeling of real-world data}
-
-To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021, \cite{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010, \cite{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events.
-Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}.
-Both datasets exhibited superior fits after allowing for heavier tails, with a further improved fit using ATAF for the CMS claims dataset.
-%Using identical flow architectures and optimizers, the log likelihoods in \Cref{tab:density-estimation} suggest that models permitting tail-anisotropy (ATAF) yield significantly superior fits.
-
-% \feynman{TODO}
-
-% \subsection{Bayesian regression analysis on diamonds}
-% \feynman{Remove? These are taking too long and don't look good}
-
-% \begin{table}[htbp]
-%   \caption{Final ELBO and (MC estimate of) log marginal $\log P(y)$ after 10000 steps on \texttt{diamonds}}
-%   \label{fig:blr_diamonds}
-%   \centering
-%   \begin{tabular}{rcc}
-%     \toprule
-%               & ELBO   & $\log P(y)$ \\
-%     \midrule
-%     ADVI      & -374.4 & -6912.0           \\
-%     ADVI-HT   & -246.3 & -6876.6     \\
-%     MAF       & -211.1 & -6894.8     \\
-%     MAF-HT    & -208.5 & -7442.0     \\
-%     MAF-2L    & -197.9 & -6532.4     \\
-%     MAF-2L-HT & -191.9 & -6839.9     \\
-%     MAF-3L    & -194.1 & -7027.3     \\
-%     MAF-3L-HT & -209.8 & -7128.1     \\
-%     \bottomrule
-%   \end{tabular}
-% \end{table}
-
-% \feynman{
-%   % \subsection{Importance weights}
-
-%   When the importance sampling density is more peaked than the target density.
-
-%   \cite{wang2018variational} example 3.1: Let $p = N(0, \sigma_p^2)$, $q = N(0, \sigma_q^2)$,
-%   and for $x \sim q$ let $w(x) = \frac{p(x)}{q(x)}$. If $\sigma_p > \sigma_q$,
-%   then $w$ has tail index $\frac{\sigma_p^2}{\sigma_p^2 - \sigma_q^2}$.
-%   Otherwise, $w$ is not fat-tailed.
-% }
-
-%\section{Related Work}
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We sharpened existing theory for approximating fat-tailed distributions with normalizing flows, formalized  tail-(an)isotropy through a direction-dependent tail parameter, proved many prior flow-based methods are limited by tail-isotropy, and proposed a new flow-based method called
-ATAF which is capable of modeling tail-anisotropic targets.
-As we have seen, anisotropic FTVI is already applicable in fairly elementary examples such as Bayesian linear regression
-and ATAFs provide one of the first methods for using the representational capacity of flow-based methods
-while simultaneously producing tail-anisotropic distributions. We expect the scientific impact of ATAFs to be
-fairly easily realized, as it can be applied wherever a parametric density estimator is used.
-While we do not expect ATAF to directly have any negative societal impacts,
-its broad applicability as a parametric density estimator which accommodates fat-tails means that it may be
-used to improve applications which do cause negative societal impact.
-
-% VI is important, FTVI is very important, ADVI did a good job with FTVI, we do a great job with FTVI, and you the reader should follow up on us and do FTVI.
-
-\emph{Acknowledgements}
-%   Feynman Liang is supported by funding from NPSC and Facebook.
-%   Michael Mahoney acknowledges funding from ARO, DARPA, NSF, and ONR.
-Redacted
-
-\bibliography{refs_ftvi}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section*{Checklist}
-
-% %%% BEGIN INSTRUCTIONS %%%
-% % The checklist follows the references.  Please
-% % read the checklist guidelines carefully for information on how to answer these
-% % questions.  For each question, change the default \answerTODO{} to \answerYes{},
-% % \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
-% %     justification to your answer}, either by referencing the appropriate section of
-% % your paper or providing a brief inline description.  For example:
-% % \begin{itemize}
-% %   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
-% %   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
-% %   \item Did you include the license to the code and datasets? \answerNA{}
-% % \end{itemize}
-% % Please do not modify the questions and only use the provided macros for your
-% % answers.  Note that the Checklist section does not count towards the page
-% % limit.  In your paper, please delete this instructions block and only keep the
-% % Checklist section heading above along with the questions/answers below.
-% %%% END INSTRUCTIONS %%%
-
-% \begin{enumerate}
-
-%   \item For all authors...
-%         \begin{enumerate}
-%           \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
-%                 \answerYes{}
-%           \item Did you describe the limitations of your work?
-%                 \answerYes{See \Cref{ssec:limitations}}
-%           \item Did you discuss any potential negative societal impacts of your work?
-%                 \answerYes{See \Cref{sec:conclusion}}
-%           \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
-%                 \answerYes{}
-%         \end{enumerate}
-
-%   \item If you are including theoretical results...
-%         \begin{enumerate}
-%           \item Did you state the full set of assumptions of all theoretical results?
-%                 \answerYes{}
-%           \item Did you include complete proofs of all theoretical results?
-%                 \answerYes{Deferred to \Cref{sec:proofs}}
-%         \end{enumerate}
-
-%   \item If you ran experiments...
-%         \begin{enumerate}
-%           \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
-%                 \answerYes{See \Cref{sec:experiments}}
-%           \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
-%                 \answerYes{See \Cref{sec:experiments}, \Cref{sec:additional-exp-details}, and open-sourced code}
-%           \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
-%                 \answerYes{Tables/figures report standard errors}
-%           \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
-%                 \answerYes{See \Cref{sec:additional-exp-details}}
-%         \end{enumerate}
-
-%   \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
-%         \begin{enumerate}
-%           \item If your work uses existing assets, did you cite the creators?
-%                 \answerYes{}
-%           \item Did you mention the license of the assets?
-%                 \answerYes{See \texttt{LICENSE} and \texttt{README.md} file on Github repository.}
-%           \item Did you include any new assets either in the supplemental material or as a URL?
-%                 \answerYes{Experiments use newly developed code within \texttt{beanmachine} which is released in \Cref{sec:experiments}.}
-%           \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
-%                 \answerNo{Datasets used are either synthetic or released as part of MIT/BSD-3 licensed code.}
-%           \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
-%                 \answerNo{We did not screen datasets for PII / offensive content. All experiments perform probabilistic modeling on top of
-%                 existing data, so we do not expect any PII / offensive content to appear in the body of the paper. Please notify us if something
-%                 could potentially make a reader uncomfortable and we will be happy to fix it!
-%                 }
-%         \end{enumerate}
-
-%   \item If you used crowdsourcing or conducted research with human subjects...
-%         \begin{enumerate}
-%           \item Did you include the full text of instructions given to participants and screenshots, if applicable?
-%                 \answerNA{}
-%           \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
-%                 \answerNA{}
-%           \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
-%                 \answerNA{}
-%         \end{enumerate}
-
-% \end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\appendix
-
-{\onecolumn
 \section{Proofs}
@@
 for any Borel set $E$. Consequently, $\mu$ is regularly varying, and so
-by the spectral representation of regularly varying random vectors (see \cite[pg. 281]{buraczewski2016stochastic}), there exists a measure $P$ such that
+by the spectral representation of regularly varying random vectors (see \citet[pg. 281]{buraczewski2016stochastic}), there exists a measure $P$ such that
 \[
@@
 
-\begin{figure}[H]
+\begin{figure}
   \centering
@@
 no information about the tails.
-}
+
+\bibliography{refs_ftvi}
+
 \end{document}
-
-% \begin{abstract}
-%   The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
-%   both left and right-hand margins. Use 10~point type, with a vertical
-%   spacing of 11~points. The \textbf{Abstract} heading must be centered,
-%   bold, and in point size 12. Two line spaces precede the
-%   Abstract. The Abstract must be limited to one paragraph.
-% \end{abstract}
-
-% \section{GENERAL FORMATTING INSTRUCTIONS}
-
-% The camera-ready versions of the accepted papers are 8 pages,
-% plus any additional pages needed for references.
-
-% Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
-% Each column is 3.25~inches wide (19.5~picas).  The space
-% between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
-% Use 10~point type with a vertical spacing of
-% 11~points. Please use US Letter size paper instead of A4.
-
-% Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
-% Top rule is 4~points thick and bottom rule is 1~point thick.
-% Allow 1/4~inch space above and below title to rules.
-
-% Author descriptions are center-justified, initial caps.  The lead
-% author is to be listed first (left-most), and the Co-authors are set
-% to follow.  If up to three authors, use a single row of author
-% descriptions, each one center-justified, and all set side by side;
-% with more authors or unusually long names or institutions, use more
-% rows.
-
-% Use one-half line space between paragraphs, with no indent.
-
-% \section{FIRST LEVEL HEADINGS}
-
-% First level headings are all caps, flush left, bold, and in point size
-% 12. Use one line space before the first level heading and one-half line space
-% after the first level heading.
-
-% \subsection{Second Level Heading}
-
-% Second level headings are initial caps, flush left, bold, and in point
-% size 10. Use one line space before the second level heading and one-half line
-% space after the second level heading.
-
-% \subsubsection{Third Level Heading}
-
-% Third level headings are flush left, initial caps, bold, and in point
-% size 10. Use one line space before the third level heading and one-half line
-% space after the third level heading.
-
-% \paragraph{Fourth Level Heading}
-
-% Fourth level headings must be flush left, initial caps, bold, and
-% Roman type.  Use one line space before the fourth level heading, and
-% place the section text immediately after the heading with no line
-% break, but an 11 point horizontal space.
-
-% %%%
-% \subsection{Citations, Figure, References}
-
-% \subsubsection{Citations in Text}
-
-% Citations within the text should include the author's last name and
-% year, e.g., (Cheesman, 1985). 
-% %Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
-% Be sure that the sentence reads
-% correctly if the citation is deleted: e.g., instead of ``As described
-% by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
-% described by Cheesman (1985), we first frobulate the widgets.''
-
-% The references listed at the end of the paper can follow any style as long as it is used consistently.
-
-% %Be sure to avoid
-% %accidentally disclosing author identities through citations.
-
-% \subsubsection{Footnotes}
-
-% Indicate footnotes with a number\footnote{Sample of the first
-%   footnote.} in the text. Use 8 point type for footnotes. Place the
-% footnotes at the bottom of the column in which their markers appear,
-% continuing to the next column if required. Precede the footnote
-% section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
-% long.\footnote{Sample of the second footnote.}
-
-% \subsubsection{Figures}
-
-% All artwork must be centered, neat, clean, and legible.  All lines
-% should be very dark for purposes of reproduction, and art work should
-% not be hand-drawn.  Figures may appear at the top of a column, at the
-% top of a page spanning multiple columns, inline within a column, or
-% with text wrapped around them, but the figure number and caption
-% always appear immediately below the figure.  Leave 2 line spaces
-% between the figure and the caption. The figure caption is initial caps
-% and each figure should be numbered consecutively.
-
-% Make sure that the figure caption does not get separated from the
-% figure. Leave extra white space at the bottom of the page rather than
-% splitting the figure and figure caption.
-% \begin{figure}[h]
-% \vspace{.3in}
-% \centerline{\fbox{This figure intentionally left non-blank}}
-% \vspace{.3in}
-% \caption{Sample Figure Caption}
-% \end{figure}
-
-% \subsubsection{Tables}
-
-% All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
-% Table number and title always appear above the table.
-% See Table~\ref{sample-table}.
-
-% Use one line space before the table title, one line space after the table title,
-% and one line space after the table. The table title must be
-% initial caps and each table numbered consecutively.
-
-% \begin{table}[h]
-% \caption{Sample Table Title} \label{sample-table}
-% \begin{center}
-% \begin{tabular}{ll}
-% \textbf{PART}  &\textbf{DESCRIPTION} \\
-% \hline \\
-% Dendrite         &Input terminal \\
-% Axon             &Output terminal \\
-% Soma             &Cell body (contains cell nucleus) \\
-% \end{tabular}
-% \end{center}
-% \end{table}
-
-% \section{SUPPLEMENTARY MATERIAL}
-
-% If you need to include additional appendices during submission, you can include them in the supplementary material file.
-% You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
-% Note that reviewers are under no obligation to examine your supplementary material. 
-% If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.
-
-% You must use \texttt{aistats2022.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
-% The only difference is that it must be in a \emph{single-column} format.
-% You can use \texttt{supplement.tex} in our starter pack as a starting point.
-% Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.
-
-% \section{SUBMISSION INSTRUCTIONS}
-
-% To submit your paper to AISTATS 2022, please follow these instructions.
-
-% \begin{enumerate}
-%     \item Download \texttt{aistats2022.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
-%     Please, do not modify the style files as this might result in a formatting violation.
-
-%     \item Use \texttt{sample\_paper.tex} as a starting point.
-%     \item Begin your document with
-%     \begin{flushleft}
-%     \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-%     \texttt{\textbackslash usepackage\{aistats2022\}}
-%     \end{flushleft}
-%     The \texttt{twoside} option for the class article allows the
-%     package \texttt{fancyhdr.sty} to include headings for even and odd
-%     numbered pages.
-%     \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
-%     \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
-%     \item Upload the PDF file along with other supplementary material files to the CMT website.
-% \end{enumerate}
-
-% \subsection{Camera-ready Papers}
-
-% %For the camera-ready paper, if you are using \LaTeX, please make sure
-% %that you follow these instructions.  
-% % (If you are not using \LaTeX,
-% %please make sure to achieve the same effect using your chosen
-% %typesetting package.)
-
-% If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
-% \begin{enumerate}
-%     %\item Download \texttt{fancyhdr.sty} -- the
-%     %\texttt{aistats2022.sty} file will make use of it.
-%     \item Change the beginning of your document to
-%     \begin{flushleft}
-%     \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-%     \texttt{\textbackslash usepackage[accepted]\{aistats2022\}}
-%     \end{flushleft}
-%     The option \texttt{accepted} for the package
-%     \texttt{aistats2022.sty} will write a copyright notice at the end of
-%     the first column of the first page. This option will also print
-%     headings for the paper.  For the \emph{even} pages, the title of
-%     the paper will be used as heading and for \emph{odd} pages the
-%     author names will be used as heading.  If the title of the paper
-%     is too long or the number of authors is too large, the style will
-%     print a warning message as heading. If this happens additional
-%     commands can be used to place as headings shorter versions of the
-%     title and the author names. This is explained in the next point.
-%     \item  If you get warning messages as described above, then
-%     immediately after $\texttt{\textbackslash
-%     begin\{document\}}$, write
-%     \begin{flushleft}
-%     \texttt{\textbackslash runningtitle\{Provide here an alternative
-%     shorter version of the title of your paper\}}\\
-%     \texttt{\textbackslash runningauthor\{Provide here the surnames of
-%     the authors of your paper, all separated by commas\}}
-%     \end{flushleft}
-%     Note that the text that appears as argument in \texttt{\textbackslash
-%       runningtitle} will be printed as a heading in the \emph{even}
-%     pages. The text that appears as argument in \texttt{\textbackslash
-%       runningauthor} will be printed as a heading in the \emph{odd}
-%     pages.  If even the author surnames do not fit, it is acceptable
-%     to give a subset of author names followed by ``et al.''
-
-%     %\item Use the file sample\_paper.tex as an example.
-
-%     \item The camera-ready versions of the accepted papers are 8
-%       pages, plus any additional pages needed for references.
-
-%     \item If you need to include additional appendices,
-%       you can include them in the supplementary
-%       material file.
-
-%     \item Please, do not change the layout given by the above
-%       instructions and by the style file.
-
-% \end{enumerate}
-
-% \subsubsection*{Acknowledgements}
-% All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
-% To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
-
-% \subsubsection*{References}
-
-% References follow the acknowledgements.  Use an unnumbered third level
-% heading for the references section.  Please use the same font
-% size for references as for the body of the paper---remember that
-% references do not count against your page length total.
-
-% \begin{thebibliography}{}
-% \setlength{\itemindent}{-\leftmargin}
-% \makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
-% \bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
-%     \newblock Performance of a stochastic learning microchip.
-%     \newblock In D. S. Touretzky (ed.),
-%     \textit{Advances in Neural Information Processing Systems 1}, 748--760.
-%     San Mateo, Calif.: Morgan Kaufmann.
-
-% \bibitem{} F.~Rosenblatt (1962).
-%     \newblock \textit{Principles of Neurodynamics.}
-%     \newblock Washington, D.C.: Spartan Books.
-
-% \bibitem{} G.~Tesauro (1989).
-%     \newblock Neurogammon wins computer Olympiad.
-%     \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
-% \end{thebibliography}
-
-% \end{document}

### 2022-08-23 ftvi/ftvi_neurips/ftvi_neurips.tex
Commit original repos

--- 2022-08-23 ftvi/ftvi_aistats/ftvi_aistats.tex
@@
-\documentclass[twoside]{article}
-
-\usepackage{aistats2022}
-%\usepackage[accepted]{aistats2022}
-% If your paper is accepted, change the options for the package
-% aistats2022 as follows:
+\documentclass{article}
+
+% if you need to pass options to natbib, use, e.g.:
+\PassOptionsToPackage{numbers, compress, sort}{natbib}
+% before loading neurips_2021
+
+% ready for submission
+%\usepackage{neurips_2021}
+
+% to compile a preprint version, e.g., for submission to arXiv, add the
+% [preprint] option:
+%\usepackage[preprint]{neurips_2021}
+\usepackage{neurips_2021}
+
+% to compile a camera-ready version, add the [final] option, e.g.:
+%     \usepackage[final]{neurips_2021}
+
+% to avoid loading the natbib package, add option nonatbib:
+%    \usepackage[nonatbib]{neurips_2021}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{xcolor}         % colors
+\usepackage{enumitem}
+
+\title{Fat--Tailed Variational Inference with \\ Anisotropic Tail Adaptive Flows}
+
+% The \author macro works with any number of authors. There are two commands
+% used to separate the names and addresses of multiple authors: \And and \AND.
 %
-%\usepackage[accepted]{aistats2022}
-%
-% This option will print headings for the title of your paper and
-% headings for the authors names, plus a copyright note at the end of
-% the first column of the first page.
-
-% If you set papersize explicitly, activate the following three lines:
-%\special{papersize = 8.5in, 11in}
-%\setlength{\pdfpageheight}{11in}
-%\setlength{\pdfpagewidth}{8.5in}
-
-% If you use natbib package, activate the following three lines:
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
-
-% If you use BibTeX in apalike style, activate the following line:
-\bibliographystyle{apalike}
+% Using \And between authors leaves it to LaTeX to determine where to break the
+% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
+% authors names on the first line, and the last on the second line, try using
+% \AND instead of \And before the third author name.
+
+\author{%
+  Feynman Liang\\
+  Department of Statistics\\
+  University of California, Berkeley\\
+  \texttt{feynman@berkeley.edu} \\
+  \And
+  Liam Hodgkinson\\
+  ICSI and Department of Statistics\\
+  University of California, Berkeley\\
+  \texttt{liam.hodgkinson@berkeley.edu} \\
+  \And
+  Michael W. Mahoney\\
+  ICSI and Department of Statistics\\
+  University of California, Berkeley\\
+  \texttt{mmahoney@stat.berkeley.edu} \\
+}
 
@@
 \newtheorem{corollary}[theorem]{Corollary}
+
 \theoremstyle{remark}
 \newtheorem{remark}[theorem]{Remark}
+
 \usepackage[bold]{hhtensor}
 
-\usepackage{hyperref}
+% \usepackage{natbib}
+\bibliographystyle{abbrv}
+% \usepackage{hyperref}
 \usepackage{cleveref}
@@
 \usepackage{subcaption}
-\usepackage{enumitem}
 
@@
 
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-%\runningtitle{I use this title instead because the last one was very long}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
-\runningtitle{Fat-Tailed VI with Anisotropic Tail Adaptive Flows}
-
-\twocolumn[
-
-\aistatstitle{Fat--Tailed Variational Inference \\ with Anisotropic Tail Adaptive Flows}
-
-\aistatsauthor{ Feynman Liang \And Liam Hodgkinson \And Michael W. Mahoney }
-
-\aistatsaddress{ Department of Statistics\\UC Berkeley \And ICSI and Department of Statistics\\UC Berkeley \And ICSI and Department of Statistics\\UC Berkeley } ]
-
-% \maketitle
+\maketitle
 
@@
     to accurately capture tail decay. We first improve
-    previous theory on tails of Lipschitz flows \citep{jaini2020tails} by
+    previous theory on tails of Lipschitz flows\citep{jaini2020tails} by
     quantifying how they affect the \emph{rate} of tail decay and
@@
     tail-anisotropic targets, we propose anisotropic tail-adaptive flows (ATAF).
-    Experimental results confirm ATAF on both synthetic and real-world
-    targets
-    is competitive with prior work while also exhibiting appropriate tail-anisotropy.
+    Experimental results confirm using ATAF improves
+    variational inference (VI) against both synthetic and real-world
+    targets which exhibit fat-tails and tail-anisotropy.
 \end{abstract}
@@
 
-\begin{figure*}[htbp]
+\begin{figure}[htbp]
   \centering
@@
   \label{fig:pancake}
-\end{figure*}
-
-Our work here aims to identify and address these deficiencies.
+\end{figure}
+
+Our work here aims to address these deficiencies.
 To understand the impact of the base distribution $\mu$ in flow-based models,
@@
 they are unable to capture \emph{tail-anisotropy}.
-This motivates the construction of \emph{anisotropic tail adaptive flows} (ATAF, \Cref{def:ataf}) as a means to
+This motivates our development of \emph{anisotropic tail adaptive flows} (ATAF, \Cref{def:ataf}) as a means to
 alleviate this issue (\Cref{remark:anisotropic}) and improve modeling of tail-anisotropic distributions.
@@
 Probably the best way to do this is to be more detailed, e.g., say that our \Cref{prop:isotropic-pushforward} means that the mapping can't introduce anisotropy.}
-Our experiments show ATAF exhibits correct tail behaviour in synthetic target distributions exhibiting fat-tails (\Cref{fig:cauchy_normal_student}) and tail-anisotropy (\Cref{fig:pancake}).
+Our experiments show ATAF yields significant improvements in synthetic target distributions exhibiting fat-tails (\Cref{fig:cauchy_normal_student}) and tail-anisotropy (\Cref{fig:pancake}).
 On realistic targets,
@@
 % can improve density modeling and 
-we find that ATAF can yield improvements in variational inference (VI) by capturing potential tail-anisotropy (\Cref{sec:experiments}).
+We find (\Cref{sec:experiments}) that ATAF can yield significant improvements when applied for variational inference (VI).
 % \michael{That sentence sounds like we are a footnote on them.
@@
 In \citep{mathieu2019disentangling}, a StudentT prior distribution $p(z)$ is considered over the latent code
-$z$ in a VAE with Gaussian encoder $q(z \mid x)$. They argue
+$z$ in a VAE with Gaussian encoder $q(z \mid x)$, arguing
 %\footnote{\url{https://github.com/iffsid/disentangling-disentanglement/blob/3396d40f46c34dd928a1241f567a86276b0ff41b/src/main.py\#L52}},
 that the anisotropy of a StudentT product distribution leads to more disentangled representations as compared to the standard choice of Normal distributions.
-A similar modification is performed in \citet{chen2020use}, for
-a coupled VAE (see \citet{cao2019coupled}). It showed improvements in the marginal
+A similar modification is performed in \cite{chen2020use}, for
+a coupled VAE (see \cite{cao2019coupled}). It showed improvements in the marginal
 likelihoods of reconstructed images. In addition,
-\citet{boenninghoff2020variational} consider a mixture of StudentTs for the
+\cite{boenninghoff2020variational} consider a mixture of StudentTs for the
 prior $p(z)$. To position
@@
 decoder model $p(x \mid z)$ and the prior $p(z)$. Our work differs from
-\citet{mathieu2019disentangling,chen2020use,boenninghoff2020variational} in
+\citep{mathieu2019disentangling,chen2020use,boenninghoff2020variational} in
 that we consider fat-tailed variational approximations $q(z \mid x)$ rather
-than priors $p(z)$. Although \citet{abiri2020variational} also considers
+than priors $p(z)$. Although \citep{abiri2020variational} also considers
 a StudentT approximate posterior, our work involves a more general
@@
 % variational families $q(z \mid x)$.
-Similarly, although \citet{wang2018variational} also deals with fat-tails in variational inference,
+Similarly, although \citep{wang2018variational} also deals with fat-tails in variational inference,
 their goal is to improve $\alpha$-divergence VI by controlling the moments of importance
@@
 to include anisotropic fat-tailed distributions.
-More directly comparable recent work \citep{ding2011t,futami2017expectation} studies the $t$-exponential family
+More directly comparable recent work \cite{ding2011t,futami2017expectation} studies the $t$-exponential family
 variational approximation which includes StudentTs and other
-heavier-tailed densities. Critically, the selection of their parameter $t$ (directly related to the
-StudentT's degrees of freedom $\nu$), and the issue of tail anisotropy, are not discussed. %Other differences include their derivation of
+heavier-tailed densities. Critically, these do
+not discuss selection of the parameter $t$ (which is deterministically to the
+Student-T's DoF $\nu$), nor do they address the issue of tail anisotropy. %Other differences include their derivation of
 %expectation-propagation update equations while 
@@
 inference \citep{kingma2016improved,rezende2015variational,agrawal2020advances,webb2019improving}.
-Consistent with our experience (\Cref{fig:blr-anisotropic}), \citet{webb2019improving}
-reports that normalizing flows can offer improvements over ADVI and NUTS across thirteen different
-Bayesian linear regression models from \citet{gelman2006data}.
-\citet{agrawal2020advances} shows that normalizing flows compose nicely with other
+Consistent with our experience (\Cref{fig:blr-anisotropic}), \citep{webb2019improving}
+documents normalizing flows can offer improvements over ADVI and NUTS across thirteen different
+Bayesian linear regression models from \citep{gelman2006data}.
+\citep{agrawal2020advances} shows that normalizing flows compose nicely with other
 advances in black-box VI (e.g., stick the landing, importance weighting).
@@
 To our knowledge, only TAFs \citep{jaini2020tails} explicitly consider flows with tails
-heavier than Gaussians. Our work here can be viewed as a direct improvement of \citet{jaini2020tails},
+heavier than Gaussians. Our work here can be viewed as a direct improvement of \citep{jaini2020tails},
 and we make extensive comparison to this work throughout the body of this paper. At
@@
   \text{ELBO}(\phi)
-  &= \int q_\phi(x) \log \frac{\bar\pi(x)}{q_\phi(x)} dx\\
-  & \approx \frac{1}{n} \sum_{i=1}^n \log \frac{\bar\pi(x_i)}{q_\phi(x_i)},\;
-  x_i \simiid q_\phi,\;
-  \bar{\pi} \propto \pi
+  = \int q_\phi(x) \log \frac{\bar\pi(x)}{q_\phi(x)} dx
+   & \approx \frac{1}{n} \sum_{i=1}^n \log \frac{\bar\pi(x_i)}{q_\phi(x_i)},
+   && x_i \simiid q_\phi,
+   && \bar{\pi} \propto \pi  .
 \end{align*}
@@
 
-\begin{table*}
+\begin{table}
   \centering
@@
   and sufficient conditions conditions for Lipschitz-continuity. A subset of this table was first presented
-  in \citet{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.}
+  in \citep{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.}
   \label{tab:flows}
-\end{table*}
+\end{table}
 
@@
 
-As first noted in \citet{jaini2020tails}, the pushforward of a light-tailed Gaussian base distribution under a Lipschitz-continuous flow will remain light-tailed and provide poor approximation to fat-tailed targets.
+As first noted in \citep{jaini2020tails}, the pushforward of a light-tailed Gaussian base distribution under a Lipschitz-continuous flow will remain light-tailed and provide poor approximation to fat-tailed targets.
 Despite this, 
@@
     \}$
-  where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$ with $\nu$ shared across all $d$ dimensions,
-  $\Phi_{\text{Flow}}$ is an invertible flow,
+  where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu)$ with $\nu$ shared across all $d$ dimensions
   and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
@@
 ``robustification'' approach where light-tailed noise distributions are
-replaced with fat-tailed ones \citep{tipping2005variational}. They also
+replaced with fat-tailed ones \cite{tipping2005variational}. They also
 appear when weakly informative prior distributions are used in Bayesian
-hierarchical models \citep{gelman2006prior}.
+hierarchical models \cite{gelman2006prior}.
 
@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics{../Figures/fat_tail_ks.pdf}
+  \vspace{-6mm}
+  \caption{
+    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
+    the use of a sub-Gaussian variational family (ADVI, solid blue line) can incur
+    exponentially bad tail approximations (middle panel) compared to
+    methods which permit heavier tails (ATAF, green dashed line) and results in
+    samples which are more similar to the target (as measured by $1000$ repeats of
+    1-sample $N=100$ Kolmogorov-Smirnov $p$-values, right panel).
+    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.
+It might be good to show ATAF is much better but still not perfect.
+Also, what is the X axis on each figure.
+}
+    % \feynman{updated the figure, hope it addresses the comment!}
+  }
+  \label{fig:cauchy_normal_student}
+\end{figure}
+
 For example, $\overline{\cE_\alpha^2}$ corresponds to $\alpha^{-1/2}$-sub-Gaussian random variables,
     $\overline{\cE_\alpha^1}$ corresponds to sub-exponentials, and (of particular relevance to this paper) $\cL^1_\alpha$ corresponds to the class of power-law distributions.
+The motivation for the fat-tailed variational families used in TAF/ATAF
+is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
+As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
+appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
+imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
+As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
+test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
+than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.
 
@@
 but also the class index and tail parameter (\Cref{def:tail-classification}). Furthermore, we extend the theory
-to include polynomial flows \citep{jaini2019sum}. For the multivariate setting,
+to include polynomial flows \cite{jaini2019sum}. For the multivariate setting,
 we define the tail-parameter function (\Cref{def:mv-tail-param}) to help formalize the notion
@@
 \begin{assumption}\label[assumption]{assump:lipschitz}
-    $f_\theta$ is invertible, and both $f_\theta$ and $f^{-1}_\theta$
-    are $L$-Lipschitz continuous (e.g., sufficient conditions in \Cref{tab:flows} are satisfied).
+    $f_\theta$ is Lipschitz continuous (e.g., sufficient conditions in \Cref{tab:flows} are satisfied) and invertible.
 \end{assumption}
@@
 via Lipschitz transport maps into fat-tailed target distributions.
-Note this does not violate universality theorems for certain flows \citep{huang2018neural}
+Note this does not violate universality theorems for certain flows \cite{huang2018neural}
 as these results only apply in the infinite-dimensional limit.
@@
 \vspace{-1mm}
-Next, we restrict attention to power-law tails $\cL^1_\alpha$ and develop a multivariate fat-tailed theory and notions of isotropic/anisotropic tail indices.
+Next, we develop theory for fat-tailed random elements in the multivariate setting which leads to formalized notions of isotropic/anisotropic tail indices.
 Using our theory, we prove that both ADVI and TAF are fundamentally limited because they
 are only capable of fitting tail-isotropic target measures (\Cref{prop:isotropic-pushforward}).
-We propose anisotropic tail adaptive flows (ATAF) as a density
+To address this limitation, we propose anisotropic tail adaptive flows (ATAF) as a density
 modeling method which can represent tail-anisotropic distributions (\Cref{remark:anisotropic}).
@@
   % $\alpha_X(v) = \lim_{x \to \infty} \frac{x \PP(\braket{v,X} \geq x)}{\int_x^\infty \PP(\braket{v,X} \geq t) dt}$.
-  $\alpha_X(v) = -\lim_{x \to \infty} \log \PP(\braket{v,X} \geq x) / \log x$ when the limit exists, and $\alpha_X(v) = +\infty$ otherwise.
-  In other words, $\alpha_X(v)$ maps directions $v$ into the tail parameter of the corresponding one-dimensional projection $\braket{v,X}$.
-The random vector $X$ is \emph{tail-isotropic} if $\alpha_X(v) \equiv c$ is constant and
-  \emph{tail-anisotropic} if $\alpha_X(v)$ is not constant but bounded.
+  $\alpha_X(v) = \lim_{x \to \infty} \frac{\log \PP(\braket{v,X} \geq x)}{\log x}$ when the limit exists, and $\alpha_X(v) = +\infty$ otherwise.
+  In other words, $\alpha_X(v)$ maps directions $v$ into the tail index of the corresponding one-dimensional projection $\braket{v,X}$.
+The random vector $X$ is \emph{tail-isotropic} if $\alpha_X(v) \equiv c$ is constant.
+Otherwise, it is \emph{tail-anisotropic}.
 \end{definition}
 
-\begin{figure*}[htbp]
+\begin{figure}[htbp]
   \centering
   \includegraphics{../Figures/radial-fat-tail.pdf}%
-%   \includegraphics[trim={0 0 7cm 0},clip]{../Figures/radial-fat-tail.pdf}\\%
-%   \vspace{-8mm}
-%   \includegraphics[trim={7.2cm 0 0 0},clip]{../Figures/radial-fat-tail.pdf}%
   %\includegraphics[width=\textwidth]{../Figures/radial-fat-tail.png}%
@@
   \caption{
-    Illustration of the direction-dependent tail-parameter function (bottom) on a tail-anisotropic distribution (top)
+    Illustration of the direction-dependent tail-parameter function (right) on a tail-anisotropic distribution (left)
     with PDF $\dd P(r,\theta) = r^{-\alpha(\theta)} r \dd r \dd\theta$ and tail parameter $\alpha(\theta) = 2 + \cos(2\theta)$.
@@
     is only sensitive to the largest tail parameter $\max_{\theta \in [0, 2\pi]} \alpha(\theta) = 3.0$,
-    our direction-dependent tail parameter function (bottom, red line)
+    our direction-dependent tail parameter function (right, red line)
     and its values along the standard basis axes ($\alpha(0)$ and $\alpha(\pi/2)$)
@@
   \label{fig:radial-fat-tail}
-\end{figure*}
+\end{figure}
 % \begin{definition}
@@
 
-Of course, one can construct pathological densities where this definition is not effective
-(see \Cref{eg:spiral}), but we will find it to suffice for our purposes. It is illustrative to contrast to the theory presented for TAF \citep{jaini2020tails}
+It is illustrative to contrast this to the theory presented for TAF \citep{jaini2020tails}
 where only the tail exponent of $\|X\|_2$ is considered.
 For $X = (X_1, \ldots, X_d)$ with $X_i \in \cL^1_{\alpha_i}$, 
-by Fatou-Lebesgue and \Cref{lem:sum-rule}
 \begin{align*}
-  &\PP[\|X\|_2 \geq t]
-  = \PP\left[\sup_{z \in \cS^{d-1}} \braket{X,z} \geq t\right]\\
-  &\quad \geq \sup_{z \in \cS^{d-1}} \PP[\braket{X,z} \geq t]
+  \PP[\|X\|_2 \geq t]
+  = \PP\left[\sup_{v \in \cS^{d-1}} \braket{X,v} \geq t\right]
+  \geq \sup_{v \in \cS^{d-1}} \PP[\braket{X,v} \geq t]
   = \max_{1 \leq i \leq d} \nu_i
-  = \max_{0 \leq i \leq d-1} \alpha_X(e_i)
+  = \max_{0 \leq i \leq d-1} \alpha_X(e_i).
 \end{align*}
@@
   \label[proposition]{prop:isotropic-pushforward}
-  Let $\mu$ be tail isotropic with non-integer parameter $\nu$
+  Let $\mu$ be tail isotropic with index $\nu$.
   and suppose $f_\theta$ satisfies \Cref{assump:lipschitz}.
-  Then $(f_\theta)_\ast \mu$ is tail isotropic with parameter $\nu$.
+  Then $(f_\theta)_\ast \mu$ is tail isotropic with index $\nu$.
 \end{proposition}
@@
 To work around this limitation without relaxing \Cref{assump:lipschitz}, it is evident
-that tail-anisotropic base distributions $\mu$ must be considered.
-Perhaps the most straightforward modification to incorporate a tail-anisotropic base distribution replaces TAF's isotropic base distribution $\prod_{i=1}^d \text{StudentT}(\nu)$
+that tail-anisotropic base distributions $\mu$ must be considered. In this work, we propose
+the following model which replaces TAF's isotropic base distribution $\prod_{i=1}^d \text{StudentT}(\nu)$
 with $\prod_{i=1}^d \text{StudentT}(\nu_i)$. Note that $\nu$ is no longer shared across dimensions,
@@
   where $\mu_\nu = \prod_{i=1}^d \text{StudentT}(\nu_i)$, each $\nu_i$ is \emph{distinct}, and $f$ is a bijection between constrained supports \citep{kucukelbir2017automatic}.
-  Analogous to \cite{jaini2020tails}, ATAF's implementation
-  treats $\nu_i$ identically to the other parameters in the flow and jointly optimizes over them.
 \end{definition}
@@
 \vspace{-1mm}
-
-Additionally, ATAF only uses a finite collection $\{\nu_i\}_{i=1}^d$ of tail parameters
+Whereas the tail-parameter function $\alpha_X(\cdot)$ provides finer-grained characterization
+of multivariate tails, it still possesses some notable limitations.
+By only considering projections $\braket{X,v}$ along rays through the origin
+$v \in \cS^{d-1}$ precludes situations where the limit in $\alpha_X(\cdot)$ is not well defined.
+For example, consider $\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
+using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
+oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics{Figures/spiral.pdf}
+    \caption{Taking a tail-anisotropic distribution (left) and ``spinning'' it (middle) results in
+        one-dimensional projections which oscillate between tail parameters (as seen in
+        $\log p(\braket{X,e_0})$ in right panel) and cause $\alpha_X(\cdot)$ to be not well defined.
+    }
+    \label{fig:spiral}
+\end{figure}
+
+Additionally, note that ATAF only uses a finite collection $\{\nu_i\}_{i=1}^d$ of tail parameters
 whereas the tail-parameter function may take on more than $d$ values. \Cref{fig:radial-fat-tail}
@@
 tail parameters are present.
-Furthermore, distributions supported on domains other than Euclidian $\mathbb{R}^d$ may require an additional bijection between supports (e.g. $\exp : \mathbb{R} \to \mathbb{R}_+$) which
-may violate \cref{assump:lipschitz}.
-Finally, note that $\text{StudentT}(\nu) \Rightarrow \cN(0,1)$ 
-as $\nu \to \infty$ so we should expect ATAF to provide reasonably good approximations to target 
-distributions in $\overline{\cE^2}$ by taking $\nu$ sufficiently large. We observed this occurs in 
-practice empirically in \Cref{sec:normal-normal-location-mixture}.
+
+While it is true that ATAF imposes logarithmic-type tails for all projections $\braket{X,v}$,
+note that $\text{StudentT}(\nu) \Rightarrow \cN(0,1)$ as $\nu \to \infty$ so we should expect
+ATAF to provide reasonably good approximations to target distributions in $\overline{\cE^2}$
+by taking $\nu$ sufficiently large. We observed this occurs in practice empirically
+in \Cref{sec:normal-normal-location-mixture}.
 % We found that applying ATAF when the target is light tailed results in minimal
@@
 \vspace{-1mm}
-Here we validate ATAF's ability to improve
-a range of probabilistic modeling tasks.
-Prior work \citep{jaini2020tails} demonstrated improved
-density modelling when fat tails are considered, and
-our experiments are complementary by evaluating TAFs and ATAFs for variational inference tasks as well as demonstrating the necessity of tail-anisotropy for modelling real-world financial returns and insurance claims datasets.
-We implement using the 
+While previous sections demonstrated the improvements of ATAF on synthetic problems
+exhibiting tail-anisotropy, in this section we validate ATAF's ability to improve
+a range of probabilistic modeling tasks. As improvements in density
+modeling using TAFs (a subset of ATAF) have already been shown in prior work \citep{jaini2020tails},
+our experiments complement these results by focusing on VI tasks and demonstrating the
+efficacy of ATAF.
+Our experiments are implemented using the 
 %\texttt{beanmachine} probabilistic programming language\citep{tehrani2020bean}
@@
 %and we have open-sourced the code\citep{ghlic}
-and we have open-sourced code for reproducing experiments[REDACTED, reviewers please see supplementary materials for code]. 
-Additional details for the experiments are detailed in \Cref{sec:additional-exp-details}.
+and we have open-sourced the code[REDACTED, reviewers please see supplementary materials for code]
+under a CC-BY-NC license. Additional details for the experiments are detailed in \Cref{sec:additional-exp-details}.
 
@@
 
-\begin{figure*}[htbp]
+\subsection{Bayesian linear regression with conjugate priors}
+
+For our first experiment, we consider one-dimensional Bayesian linear regression (BLR)
+with conjugate priors. This model is defined by priors and likelihood
+\begin{align*}
+    \sigma^2 \sim \text{Inv-Gamma}(a_0, b_0), \qquad
+    \beta \mid \sigma^2 \sim \cN(0, \sigma^2),\qquad 
+    y \mid X, \beta, \sigma \sim \cN(X \beta, \sigma^2) 
+\end{align*}
+where $a_0$, $b_0$ are hyperparameters and the task is to approximate the posterior
+distribution $p(\beta,\sigma^2 \mid X, y)$. Owing to conjugacy,
+the posterior distribution can be explicitly computed.
+Indeed, $p(\beta,\sigma^2 \mid X, y) = \rho(\sigma^2)\rho(\beta \mid \sigma)$ where $\rho(\beta \mid \sigma) = \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 \Sigma_n)$, $\Sigma_n = (X^\top X + \sigma^{-2})^{-1}$, $\hat\beta = (X^\top X)^{-1} X^\top y$, and
+%\begin{align*}
+    %p(\beta,\sigma^2 \mid X, y) &= \rho(\sigma^2) \rho(\beta \mid \sigma) \\
+    \[
+    \rho(\sigma^2) = \text{Inv-Gamma}\bigg(
+    %\underbrace{a_0 + \frac{n}{2}}_{\eqqcolon a_n}, 
+    a_0 + \frac{n}{2}, 
+    b_0 + \frac{1}{2}(y^\top y - \mu_n^\top \Sigma_n \mu_n)\bigg)% \\
+    \]
+    %\rho(\beta \mid \sigma) &= \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 (X^\top X + \sigma^{-2} I)^{-1})
+%\end{align*}
+This calculation reveals that the posterior distribution is tail-anisotropic:
+for fixed $c$ we have that $p(\sigma^2, \beta=c \mid X, y) \propto \rho(\sigma^2) \in \cL^1_{\alpha_n}$
+as a function of $\sigma$
+and $p(\sigma^2=c, \beta \mid X, y) \propto \rho(\beta \mid c) \in \overline{\cE^2}$
+as a function of $\beta$.
+As a result of \Cref{prop:isotropic-pushforward}, we expect ADVI and TAF to erroneously impose
+Gaussian and power-law tails respectively for both $\beta$ and $\sigma^2$ as neither method
+can produce a tail-anisotropic pushforward. This intuition is confirmed in \Cref{fig:blr-anisotropic},
+where we see that only ATAF is the only method capable of modeling the tail-anisotropy present.
+
+\begin{figure}[htbp]
   \centering
@@
   \includegraphics{../Figures/blr_aniso.pdf}
-  \vspace{-0.3cm}
+  \vspace{-0.5cm}
   \caption{
@@
   \label{fig:blr-anisotropic}
-\end{figure*}
-
-\begin{table*}[htbp]
+\end{figure}
+
+\begin{table}[htbp]
   \centering
@@
         \caption{Monte-Carlo ELBO and importance weighted Monte-Carlo marginal likelihood
-        $p(y) = \EE_{x \sim q_\theta} \frac{p(x,y)}{q_\theta(x)}$ (higher is better, $\pm$ standard errors) 
+        $p(y) = \EE_{x \sim q_\theta} \frac{p(x,y)}{q_\theta(x)}$ (higher is better, $\pm$ indicates standard errors) 
         estimates from VI on real-world datasets.
@@
   \vspace{-4mm}
-\end{table*}
-
-\begin{table*}[htbp]
-    \centering
-    \begin{tabular}{rcc}
-        \toprule
-                  & Fama-French 5 Industry Daily & CMS 2008-2010 DE-SynPUF       \\
-        \midrule
-        ADVI      & $-5.018 \pm 0.056$    & $-1.883 \pm 0.012$ \\
-        TAF       & $-4.703 \pm 0.023$    & $-1.659 \pm 0.004$ \\
-        ATAF      & $-4.699 \pm 0.024$    & $-1.603 \pm 0.034$ \\
-        \bottomrule
-    \end{tabular}
-    \caption{
-        Log-likelihoods (higher is better, $\pm$ standard errors, $10$ trials) achieved on
-        density modeling tasks involving real-world financial returns \citep{fama2015five} and insurance claims \citep{cms} data. 
-    }
-    \label{tab:density-estimation}
-\end{table*}
-
-\subsection{Bayesian linear regression}
-
-Consider one-dimensional Bayesian linear regression (BLR)
-with conjugate priors, defined by priors and likelihood
-\begin{gather*}
-    \sigma^2 \sim \text{Inv-Gamma}(a_0, b_0)\\
-    \beta \mid \sigma^2 \sim \cN(0, \sigma^2),\qquad
-    y \mid X, \beta, \sigma \sim \cN(X \beta, \sigma^2) 
-\end{gather*}
-where $a_0$, $b_0$ are hyperparameters and the task is to approximate the posterior
-distribution $p(\beta,\sigma^2 \mid X, y)$. Owing to conjugacy,
-the posterior distribution can be explicitly computed.
-Indeed, $p(\beta,\sigma^2 \mid X, y) = \rho(\sigma^2)\rho(\beta \mid \sigma)$ where $\rho(\beta \mid \sigma) = \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 \Sigma_n)$, $\Sigma_n = (X^\top X + \sigma^{-2})^{-1}$, $\hat\beta = (X^\top X)^{-1} X^\top y$, and
-%\begin{align*}
-    %p(\beta,\sigma^2 \mid X, y) &= \rho(\sigma^2) \rho(\beta \mid \sigma) \\
-    \[
-    \rho(\sigma^2) = \text{Inv-Gamma}\bigg(
-    %\underbrace{a_0 + \frac{n}{2}}_{\eqqcolon a_n}, 
-    a_0 + \frac{n}{2}, 
-    b_0 + \frac{1}{2}(y^\top y - \mu_n^\top \Sigma_n \mu_n)\bigg)% \\
-    \]
-    %\rho(\beta \mid \sigma) &= \cN(\Sigma_n(X^\top X \hat\beta), \sigma^2 (X^\top X + \sigma^{-2} I)^{-1})
-%\end{align*}
-This calculation reveals that the posterior distribution is tail-anisotropic:
-for fixed $c$ we have that $p(\sigma^2, \beta=c \mid X, y) \propto \rho(\sigma^2) \in \cL^1_{\alpha_n}$
-as a function of $\sigma$ (with $\alpha_n$ a function of $n$)
-and $p(\sigma^2=c, \beta \mid X, y) \propto \rho(\beta \mid c) \in \overline{\cE^2}$
-as a function of $\beta$.
-As a result of \Cref{prop:isotropic-pushforward}, we expect ADVI and TAF to erroneously impose
-Gaussian and power-law tails respectively for both $\beta$ and $\sigma^2$ as neither method
-can produce a tail-anisotropic pushforward. This intuition is confirmed in \Cref{fig:blr-anisotropic},
-where we see that only ATAF is the only method capable of modeling the tail-anisotropy present.
-
-\subsection{Diamond price prediction using non-conjugate Bayesian regression}
+\end{table}
+
+\subsection{Non-conjugate BLR on diamonds}
 \label{ssec:diamonds}
@@
 \begin{gather*}
-    \alpha \sim \text{StudentT}(\nu=3, \text{loc}=8, \text{scale}=10)\\
-    \sigma \sim \text{HalfStudentT}(\nu=3, \text{loc}=0, \text{scale}=10)\\
-    \beta \sim \cN(0, \mI_{24}),\qquad
-    y \sim \cN(\alpha + X \beta, \sigma)
+    \beta \sim \cN(0, \mI_{24}),\quad
+    \alpha \sim \text{StudentT}(\nu=3, \text{loc}=8, \text{scale}=10),\quad\\
+    \sigma \sim \text{HalfStudentT}(\nu=3, \text{loc}=0, \text{scale}=10),\quad
+    y \sim \cN(\alpha + X \beta, \sigma).
 \end{gather*}
@@
 
-\subsection{Eight schools SAT score modelling with fat-tailed scale mixtures}
+\subsection{Eight schools}
 \label{ssec:eight_schools}
@@
 by the probabilistic model
-\begin{gather*}
-    \tau \sim \text{HalfCauchy}(\text{loc}=0, \text{scale}=5)\\
-    \mu \sim \cN(0, 5),\qquad
-    \theta \sim \cN(\mu, \tau),\qquad
-    y \sim \cN(\theta, \sigma)
-\end{gather*}
+\begin{align*}
+    \mu \sim \cN(0, 5),\quad
+    \tau \sim \text{HalfCauchy}(\text{loc}=0, \text{scale}=5),\quad
+    \theta \sim \cN(\mu, \tau),\quad
+    y \sim \cN(\theta, \sigma).
+\end{align*}
 Given test scores and standard errors $\{(y_i, \sigma_i)\}_{i=1}^8$, we are interested in the
@@
 parameters are identical to \Cref{ssec:diamonds} and results are reported in \Cref{tab:eight_schools}.
-
-\subsection{Tail-anisotropic modeling of real-world data}
-
-To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021, \cite{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010, \cite{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events.
-Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}.
-Both datasets exhibited superior fits after allowing for heavier tails, with a further improved fit using ATAF for the CMS claims dataset.
-%Using identical flow architectures and optimizers, the log likelihoods in \Cref{tab:density-estimation} suggest that models permitting tail-anisotropy (ATAF) yield significantly superior fits.
 
@@
 
-We sharpened existing theory for approximating fat-tailed distributions with normalizing flows, formalized  tail-(an)isotropy through a direction-dependent tail parameter, proved many prior flow-based methods are limited by tail-isotropy, and proposed a new flow-based method called
+We further developed theory on the capability of flow-based methods to represent fat-tailed
+target distributions, proposed a direction-dependent tail parameter for the multivariate setting
+in order to formalize the idea of tail-(an)isotropy, proved many existing flow-based methods are limited to only represent tail-isotropic distributions, and proposed a new flow-based method called
 ATAF which is capable of modeling tail-anisotropic targets.
@@
 
-\emph{Acknowledgements}
+% \begin{ack}
 %   Feynman Liang is supported by funding from NPSC and Facebook.
 %   Michael Mahoney acknowledges funding from ARO, DARPA, NSF, and ONR.
+% \end{ack}
+\begin{ack}
   Redacted
+\end{ack}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section*{Checklist}
-
-% %%% BEGIN INSTRUCTIONS %%%
-% % The checklist follows the references.  Please
-% % read the checklist guidelines carefully for information on how to answer these
-% % questions.  For each question, change the default \answerTODO{} to \answerYes{},
-% % \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
-% %     justification to your answer}, either by referencing the appropriate section of
-% % your paper or providing a brief inline description.  For example:
-% % \begin{itemize}
-% %   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
-% %   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
-% %   \item Did you include the license to the code and datasets? \answerNA{}
-% % \end{itemize}
-% % Please do not modify the questions and only use the provided macros for your
-% % answers.  Note that the Checklist section does not count towards the page
-% % limit.  In your paper, please delete this instructions block and only keep the
-% % Checklist section heading above along with the questions/answers below.
-% %%% END INSTRUCTIONS %%%
-
-% \begin{enumerate}
-
-%   \item For all authors...
-%         \begin{enumerate}
-%           \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
-%                 \answerYes{}
-%           \item Did you describe the limitations of your work?
-%                 \answerYes{See \Cref{ssec:limitations}}
-%           \item Did you discuss any potential negative societal impacts of your work?
-%                 \answerYes{See \Cref{sec:conclusion}}
-%           \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
-%                 \answerYes{}
-%         \end{enumerate}
-
-%   \item If you are including theoretical results...
-%         \begin{enumerate}
-%           \item Did you state the full set of assumptions of all theoretical results?
-%                 \answerYes{}
-%           \item Did you include complete proofs of all theoretical results?
-%                 \answerYes{Deferred to \Cref{sec:proofs}}
-%         \end{enumerate}
-
-%   \item If you ran experiments...
-%         \begin{enumerate}
-%           \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
-%                 \answerYes{See \Cref{sec:experiments}}
-%           \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
-%                 \answerYes{See \Cref{sec:experiments}, \Cref{sec:additional-exp-details}, and open-sourced code}
-%           \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
-%                 \answerYes{Tables/figures report standard errors}
-%           \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
-%                 \answerYes{See \Cref{sec:additional-exp-details}}
-%         \end{enumerate}
-
-%   \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
-%         \begin{enumerate}
-%           \item If your work uses existing assets, did you cite the creators?
-%                 \answerYes{}
-%           \item Did you mention the license of the assets?
-%                 \answerYes{See \texttt{LICENSE} and \texttt{README.md} file on Github repository.}
-%           \item Did you include any new assets either in the supplemental material or as a URL?
-%                 \answerYes{Experiments use newly developed code within \texttt{beanmachine} which is released in \Cref{sec:experiments}.}
-%           \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
-%                 \answerNo{Datasets used are either synthetic or released as part of MIT/BSD-3 licensed code.}
-%           \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
-%                 \answerNo{We did not screen datasets for PII / offensive content. All experiments perform probabilistic modeling on top of
-%                 existing data, so we do not expect any PII / offensive content to appear in the body of the paper. Please notify us if something
-%                 could potentially make a reader uncomfortable and we will be happy to fix it!
-%                 }
-%         \end{enumerate}
-
-%   \item If you used crowdsourcing or conducted research with human subjects...
-%         \begin{enumerate}
-%           \item Did you include the full text of instructions given to participants and screenshots, if applicable?
-%                 \answerNA{}
-%           \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
-%                 \answerNA{}
-%           \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
-%                 \answerNA{}
-%         \end{enumerate}
-
-% \end{enumerate}
+\section*{Checklist}
+
+%%% BEGIN INSTRUCTIONS %%%
+% The checklist follows the references.  Please
+% read the checklist guidelines carefully for information on how to answer these
+% questions.  For each question, change the default \answerTODO{} to \answerYes{},
+% \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
+%     justification to your answer}, either by referencing the appropriate section of
+% your paper or providing a brief inline description.  For example:
+% \begin{itemize}
+%   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
+%   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
+%   \item Did you include the license to the code and datasets? \answerNA{}
+% \end{itemize}
+% Please do not modify the questions and only use the provided macros for your
+% answers.  Note that the Checklist section does not count towards the page
+% limit.  In your paper, please delete this instructions block and only keep the
+% Checklist section heading above along with the questions/answers below.
+%%% END INSTRUCTIONS %%%
+
+\begin{enumerate}
+
+  \item For all authors...
+        \begin{enumerate}
+          \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
+                \answerYes{}
+          \item Did you describe the limitations of your work?
+                \answerYes{See \Cref{ssec:limitations}}
+          \item Did you discuss any potential negative societal impacts of your work?
+                \answerYes{See \Cref{sec:conclusion}}
+          \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
+                \answerYes{}
+        \end{enumerate}
+
+  \item If you are including theoretical results...
+        \begin{enumerate}
+          \item Did you state the full set of assumptions of all theoretical results?
+                \answerYes{}
+          \item Did you include complete proofs of all theoretical results?
+                \answerYes{Deferred to \Cref{sec:proofs}}
+        \end{enumerate}
+
+  \item If you ran experiments...
+        \begin{enumerate}
+          \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
+                \answerYes{See \Cref{sec:experiments}}
+          \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
+                \answerYes{See \Cref{sec:experiments}, \Cref{sec:additional-exp-details}, and open-sourced code}
+          \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
+                \answerYes{Tables/figures report standard errors}
+          \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
+                \answerYes{See \Cref{sec:additional-exp-details}}
+        \end{enumerate}
+
+  \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
+        \begin{enumerate}
+          \item If your work uses existing assets, did you cite the creators?
+                \answerYes{}
+          \item Did you mention the license of the assets?
+                \answerYes{See \texttt{LICENSE} and \texttt{README.md} file on Github repository.}
+          \item Did you include any new assets either in the supplemental material or as a URL?
+                \answerYes{Experiments use newly developed code within \texttt{beanmachine} which is released in \Cref{sec:experiments}.}
+          \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
+                \answerNo{Datasets used are either synthetic or released as part of MIT/BSD-3 licensed code.}
+          \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
+                \answerNo{We did not screen datasets for PII / offensive content. All experiments perform probabilistic modeling on top of
+                existing data, so we do not expect any PII / offensive content to appear in the body of the paper. Please notify us if something
+                could potentially make a reader uncomfortable and we will be happy to fix it!
+                }
+        \end{enumerate}
+
+  \item If you used crowdsourcing or conducted research with human subjects...
+        \begin{enumerate}
+          \item Did you include the full text of instructions given to participants and screenshots, if applicable?
+                \answerNA{}
+          \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
+                \answerNA{}
+          \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
+                \answerNA{}
+        \end{enumerate}
+
+\end{enumerate}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
+\newpage
 \appendix
 
-{\onecolumn
 \section{Proofs}
@@
     \PP(\lvert f_\theta(X) \rvert \geq r)
-    &= \cO(e^{-\alpha (\log (r/L))^p})
-    % &= \cO(e^{-\alpha (\log r)^p} e^{-\alpha (-\log L)^p}) \\
-    = \cO(e^{-\alpha (\log r)^p}),
+    &= \cO(e^{-\alpha (\log (r/L))^p}) \\
+    &= \cO(e^{-\alpha (\log r)^p} e^{-\alpha (-\log L)^p}) \\
+    &= \cO(e^{-\alpha (\log r)^p}),
   \end{align*}
@@
   Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
-  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$. 
+  so $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$. 
 \end{proof}
@@
     Suppose $X \in \cL^1_\alpha$ and $Y \in \cL^1_\beta$.
-    Then $X + Y \in \cL^1_{\min\{\alpha,\beta\}}$.
+    Then $X + Y \in \cL^1_{\min(\alpha,\beta)}$.
 \end{lemma}
@@
 \begin{proof}
-First, let $\gamma=\min\{\alpha,\beta\}$.
+First, let $\gamma=\min\{a,b\}$.
 It will suffice to show that (I) $\mathbb{P}(|X+Y|\geq r)=\mathcal{O}(r^{-\gamma})$, and (II) $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\gamma})$.
@@
 
-To show Proposition \ref{prop:isotropic-pushforward}, we will require a few extra assumptions to rule out pathological cases.
-The full content of Proposition \ref{prop:isotropic-pushforward} is contained in the following theorem.
-
-\begin{theorem}
-Suppose there exists $\nu > 0$ such that $C:\mathcal{S}^{d-1}\to(0,\infty)$ satisfies $C(v) \coloneqq \lim_{x \to \infty}x^{\nu}\mathbb{P}(|\langle v, X\rangle| > x)$ for all $v \in \mathcal{S}^{d-1}$.
-If $\nu$ is not an integer and $f$ is a bilipschitz function,
-%such that $f(X)/\|f(X)\|$ has full support on $\mathcal{S}^{d-1}$
-then $f(X)$ is tail-isotropic with tail index $\nu$. 
-\end{theorem}
-\begin{proof}
-Since $x \mapsto \langle v, f(x)\rangle$ is Lipschitz continuous for any $v \in \mathcal{S}^{d-1}$, Theorem \ref{thm:distn_class_closed} implies $\langle v, f(X)\rangle \in \overline{\mathcal{L}_{\nu}^{1}}$.
-Let $\theta \in (0,\pi/2)$ (say, $\theta = \pi / 4$), and let $S_v = \{x\, : \, \cos^{-1}(\langle x/\|x\|, v\rangle)\leq\theta\}$ for each $v \in \mathcal{S}^{d-1}$.
-Then
-\[
-H_v \coloneqq \{x\,:\,\langle v, x \rangle > 1\} \supset \{x\,:\,\|x\| > (1-\cos\theta)^{-1}\} \cap S_v.
-\]
-From \cite[Theorem C.2.1]{buraczewski2016stochastic}, since $\nu \not \in \mathbb{Z}$, there exists a non-zero measure $\mu$ such that
-\[
-\mu(E) = \lim_{x \to \infty} \frac{\mathbb{P}(x^{-1}X \in E)}{\mathbb{P}(\|X\| > x)},
-\]
-for any Borel set $E$. Consequently, $\mu$ is regularly varying, and so 
-by the spectral representation of regularly varying random vectors (see \cite[pg. 281]{buraczewski2016stochastic}), there exists a measure $P$ such that
-\[
-\lim_{x \to \infty}\frac{\mathbb{P}(\|X\|>tx, X/\|X\| \in E)}{\mathbb{P}(\|X\| > x)} = t^{-\nu} P(E),
-\]
-for any Borel set $E$ on $\mathcal{S}^{d-1}$ and any $t > 0$.
-Letting $F_v = \{y / \|y\|\,:\, f(y) \in S_v\} \subset \mathcal{S}^{d-1}$ (noting that $P(F_v) > 0$ by assumption), since $m\|x - y\| \leq \|f(x) - f(y)\| \leq M\|x - y\|$ for all $x,y$,
-\begin{align*}
-\liminf_{x \to \infty}
-\frac{\mathbb{P}(f(X) \in x H_v)}{\mathbb{P}(\|f(X)\| > x)} 
-&\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|f(X)\| > x(1-\cos\theta)^{-1}, f(X) \in S_v)}{\mathbb{P}(\|f(X)\| > x)} \\
-&\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|X\| > x(m(1-\cos\theta))^{-1}, X/\|X\| \in F_v)}{\|X\| > x / M} \\
-&\geq P(F_v) \left(\frac{M}{m(1-\cos\theta)}\right)^{-\nu} > 0.
-\end{align*}
-where $P(F_v) > 0$ follows from the bilipschitz condition for $f$.
-Therefore, we have shown that $\mathbb{P}(\langle v, f(X)\rangle > x) = \Theta(\mathbb{P}(\|f(X)\| > x))$ for every $v \in \mathcal{S}^{d-1}$.
-Since $\mathbb{P}(\|f(X)\| > x)$ obeys a power law with exponent $\nu$ by Corollary \ref{corr:heavy_to_light}, $f(X)$ is tail-isotropic with exponent $\nu$.
+\begin{proof}[Proof of \Cref{prop:isotropic-pushforward}]
+  Suppose $\mu$ is isotropic, so $\braket{X,v}$ has index $\nu$ for all $v \in \cS^{d-1}$.
+  Then for any affine autoregressive flow,
+  $\braket{v,f(X)}$ is an affine combination of random variables with tail index $\nu$.
+  Note tail parameters are invariant under scalar multiplication as a consequence of
+  asymptotic notation, so combined with repeated applications of \Cref{lem:sum-rule} we
+  have that the linear combination $\braket{v,f(X)}$ has index $\nu$.
 \end{proof}
-
-%\begin{proof}[Proof of \Cref{prop:isotropic-pushforward}]
-%  Since $f_\theta$ satisfies \cref{assump:lipschitz}, we have $\max(\|f_\theta\|_{\text{Lip}},\|f_\theta^{-1}\|_{\text{Lip}} \leq L < \infty$
-
-%   Fix $z \in \cS^{d-1}$.
-%   Note tail parameters are invariant under scalar multiplication as a consequence of
-%   asymptotic notation, so combined with repeated applications of \Cref{lem:sum-rule} we
-%   have that the linear combination $\braket{f_\theta(X), z}$
-%   of $\{f_\theta(X)_i\}_{i \in [d]}$ is tail-isotropic provided
-%   every $f_\theta(X)_i \in \cL_\nu^1$.
-%   Let $i \in [d]$ be arbitrary.
-%   Notice that the projection map $\pi_i = x \mapsto x_i$
-%   is $1$-Lipschitz hence $\pi_i \circ f_\theta = x \mapsto f_\theta(x)_i$
-%   is $L$-Lipschitz. In particular, $g_{x_{-i}} = x_i \mapsto (\pi \circ_i f_\theta)(x_{-i}, x_i)$ is $L$-Lipschitz for every $x_{-i} \in \mathbb{R}^{d-1}$.
-
-%   Consider the $d$ times iterated conditional expectation
-%   \begin{align*}
-%       \mathbb{P}[f_\theta(X)_i \geq t]
-%       &= \EE[\EE[\cdots\EE[1(f(X)_i \geq t) \mid X_1, \ldots, X_{i-1},X_{i+1},\ldots,X_d] \cdots \mid X_1 ]] \\
-%       &= \EE[\EE[\cdots\mathbb{P}[g_{X_{-i}}(X_i) \geq t \mid X_{-i} ] \cdots \mid X_1 ]]
-%   \end{align*}
-%   Since $X \sim \mu$ is $\nu$-isotropic, we have that
-%   $X_i = \braket{X, e_i} \in \cL_\nu^1$.
-%   \feynman{To apply \cref{corr:heavy_to_light}, we need $x_i \mapsto f(x_{-i}, x_i)_i$ to be bi-Lipschitz for every $i \in [d]$ i.e.
-separately bilipschitz, see component Lipschitz constants in \url{http://www.optimization-online.org/DB_FILE/2014/12/4679.pdf}}
-%\end{proof}
-
-\section{Experiments performing VI against a fat-tailed Cauchy target}
-\label{sec:cauchy_normal_student}
-The motivation for the fat-tailed variational families used in TAF/ATAF
-is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
-As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
-appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
-imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
-As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
-test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
-than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.
-
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics{../Figures/fat_tail_ks.pdf}
-  \vspace{-6mm}
-  \caption{
-    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
-    the use of a sub-Gaussian variational family (ADVI, solid blue line) can incur
-    exponentially bad tail approximations (middle panel) compared to
-    methods which permit heavier tails (ATAF, green dashed line) and results in
-    samples which are more similar to the target (as measured by $1000$ repeats of
-    1-sample $N=100$ Kolmogorov-Smirnov $p$-values, right panel).
-    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.
-It might be good to show ATAF is much better but still not perfect.
-Also, what is the X axis on each figure.
-}
-  }
-  \label{fig:cauchy_normal_student}
-\end{figure*}
 
@@
 
-\section{Example of non-existence of tail parameter due to oscillations}
-\label{eg:spiral}
-
-Consider $\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
-using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
-oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.
-
-\begin{figure*}[htbp]
-    \centering
-    \includegraphics[scale=0.8]{Figures/spiral.pdf}
-    % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\
-    % \includegraphics[trim={5cm 0 4.9cm 0},clip]{Figures/spiral.pdf}\\
-    % \includegraphics[trim={9.3cm 0 0cm 0},clip]{Figures/spiral.pdf}\\
-    \caption{Taking a tail-anisotropic distribution (top) and ``spinning'' it (middle) results in
-        one-dimensional projections which oscillate between tail parameters (as seen in
-        $\log p(\braket{X,e_0})$ in bottom panel) and cause $\alpha_X(\cdot)$ to be not well defined.
-    }
-    \label{fig:spiral}
-\end{figure*}
-
 \section{Additional details for experiments}
@@
 For all flow-transforms $\Phi_{\text{Flow}}$ we used inverse autoregressive flows \citep{kingma2016improved} with a
-dense autoregressive conditioner consisting of two layers of either 32 or 256 hidden units depending on problem (see code for details) and
-ELU activation functions.
-As described in \cite{jaini2020tails}, TAF is trained by including $\nu$ within the Adam optimizer alongside other flow parameters. For ATAF, we include all $\nu_i$ within the optimizer.
-Models were trained using the Adam optimizer with $10^{-3}$ learning rate
+dense autoregressive conditioner consisting of varying numbers of hidden layers each with 32 hidden units and
+ELU activation functions. Models were trained using the Adam optimizer with $10^{-3}$ learning rate
 for 10000 iterations, which we found empirically in all our experiments to result in negligible change in ELBO
@@
 For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
-are comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
-using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
+are comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters
+were trained first for ADVI, then used as initialization for TAF. Additionally, the tail parameters $\nu_i$
 used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empirically observed
@@
 no information about the tails.
-}
+
 \end{document}
-
-% \begin{abstract}
-%   The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
-%   both left and right-hand margins. Use 10~point type, with a vertical
-%   spacing of 11~points. The \textbf{Abstract} heading must be centered,
-%   bold, and in point size 12. Two line spaces precede the
-%   Abstract. The Abstract must be limited to one paragraph.
-% \end{abstract}
-
-% \section{GENERAL FORMATTING INSTRUCTIONS}
-
-% The camera-ready versions of the accepted papers are 8 pages,
-% plus any additional pages needed for references.
-
-% Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
-% Each column is 3.25~inches wide (19.5~picas).  The space
-% between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
-% Use 10~point type with a vertical spacing of
-% 11~points. Please use US Letter size paper instead of A4.
-
-% Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
-% Top rule is 4~points thick and bottom rule is 1~point thick.
-% Allow 1/4~inch space above and below title to rules.
-
-% Author descriptions are center-justified, initial caps.  The lead
-% author is to be listed first (left-most), and the Co-authors are set
-% to follow.  If up to three authors, use a single row of author
-% descriptions, each one center-justified, and all set side by side;
-% with more authors or unusually long names or institutions, use more
-% rows.
-
-% Use one-half line space between paragraphs, with no indent.
-
-% \section{FIRST LEVEL HEADINGS}
-
-% First level headings are all caps, flush left, bold, and in point size
-% 12. Use one line space before the first level heading and one-half line space
-% after the first level heading.
-
-% \subsection{Second Level Heading}
-
-% Second level headings are initial caps, flush left, bold, and in point
-% size 10. Use one line space before the second level heading and one-half line
-% space after the second level heading.
-
-% \subsubsection{Third Level Heading}
-
-% Third level headings are flush left, initial caps, bold, and in point
-% size 10. Use one line space before the third level heading and one-half line
-% space after the third level heading.
-
-% \paragraph{Fourth Level Heading}
-
-% Fourth level headings must be flush left, initial caps, bold, and
-% Roman type.  Use one line space before the fourth level heading, and
-% place the section text immediately after the heading with no line
-% break, but an 11 point horizontal space.
-
-% %%%
-% \subsection{Citations, Figure, References}
-
-% \subsubsection{Citations in Text}
-
-% Citations within the text should include the author's last name and
-% year, e.g., (Cheesman, 1985). 
-% %Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
-% Be sure that the sentence reads
-% correctly if the citation is deleted: e.g., instead of ``As described
-% by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
-% described by Cheesman (1985), we first frobulate the widgets.''
-
-% The references listed at the end of the paper can follow any style as long as it is used consistently.
-
-% %Be sure to avoid
-% %accidentally disclosing author identities through citations.
-
-% \subsubsection{Footnotes}
-
-% Indicate footnotes with a number\footnote{Sample of the first
-%   footnote.} in the text. Use 8 point type for footnotes. Place the
-% footnotes at the bottom of the column in which their markers appear,
-% continuing to the next column if required. Precede the footnote
-% section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
-% long.\footnote{Sample of the second footnote.}
-
-% \subsubsection{Figures}
-
-% All artwork must be centered, neat, clean, and legible.  All lines
-% should be very dark for purposes of reproduction, and art work should
-% not be hand-drawn.  Figures may appear at the top of a column, at the
-% top of a page spanning multiple columns, inline within a column, or
-% with text wrapped around them, but the figure number and caption
-% always appear immediately below the figure.  Leave 2 line spaces
-% between the figure and the caption. The figure caption is initial caps
-% and each figure should be numbered consecutively.
-
-% Make sure that the figure caption does not get separated from the
-% figure. Leave extra white space at the bottom of the page rather than
-% splitting the figure and figure caption.
-% \begin{figure}[h]
-% \vspace{.3in}
-% \centerline{\fbox{This figure intentionally left non-blank}}
-% \vspace{.3in}
-% \caption{Sample Figure Caption}
-% \end{figure}
-
-% \subsubsection{Tables}
-
-% All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
-% Table number and title always appear above the table.
-% See Table~\ref{sample-table}.
-
-% Use one line space before the table title, one line space after the table title,
-% and one line space after the table. The table title must be
-% initial caps and each table numbered consecutively.
-
-% \begin{table}[h]
-% \caption{Sample Table Title} \label{sample-table}
-% \begin{center}
-% \begin{tabular}{ll}
-% \textbf{PART}  &\textbf{DESCRIPTION} \\
-% \hline \\
-% Dendrite         &Input terminal \\
-% Axon             &Output terminal \\
-% Soma             &Cell body (contains cell nucleus) \\
-% \end{tabular}
-% \end{center}
-% \end{table}
-
-% \section{SUPPLEMENTARY MATERIAL}
-
-% If you need to include additional appendices during submission, you can include them in the supplementary material file.
-% You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
-% Note that reviewers are under no obligation to examine your supplementary material. 
-% If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.
-
-% You must use \texttt{aistats2022.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
-% The only difference is that it must be in a \emph{single-column} format.
-% You can use \texttt{supplement.tex} in our starter pack as a starting point.
-% Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.
-
-% \section{SUBMISSION INSTRUCTIONS}
-
-% To submit your paper to AISTATS 2022, please follow these instructions.
-
-% \begin{enumerate}
-%     \item Download \texttt{aistats2022.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
-%     Please, do not modify the style files as this might result in a formatting violation.
-
-%     \item Use \texttt{sample\_paper.tex} as a starting point.
-%     \item Begin your document with
-%     \begin{flushleft}
-%     \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-%     \texttt{\textbackslash usepackage\{aistats2022\}}
-%     \end{flushleft}
-%     The \texttt{twoside} option for the class article allows the
-%     package \texttt{fancyhdr.sty} to include headings for even and odd
-%     numbered pages.
-%     \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
-%     \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
-%     \item Upload the PDF file along with other supplementary material files to the CMT website.
-% \end{enumerate}
-
-% \subsection{Camera-ready Papers}
-
-% %For the camera-ready paper, if you are using \LaTeX, please make sure
-% %that you follow these instructions.  
-% % (If you are not using \LaTeX,
-% %please make sure to achieve the same effect using your chosen
-% %typesetting package.)
-
-% If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
-% \begin{enumerate}
-%     %\item Download \texttt{fancyhdr.sty} -- the
-%     %\texttt{aistats2022.sty} file will make use of it.
-%     \item Change the beginning of your document to
-%     \begin{flushleft}
-%     \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-%     \texttt{\textbackslash usepackage[accepted]\{aistats2022\}}
-%     \end{flushleft}
-%     The option \texttt{accepted} for the package
-%     \texttt{aistats2022.sty} will write a copyright notice at the end of
-%     the first column of the first page. This option will also print
-%     headings for the paper.  For the \emph{even} pages, the title of
-%     the paper will be used as heading and for \emph{odd} pages the
-%     author names will be used as heading.  If the title of the paper
-%     is too long or the number of authors is too large, the style will
-%     print a warning message as heading. If this happens additional
-%     commands can be used to place as headings shorter versions of the
-%     title and the author names. This is explained in the next point.
-%     \item  If you get warning messages as described above, then
-%     immediately after $\texttt{\textbackslash
-%     begin\{document\}}$, write
-%     \begin{flushleft}
-%     \texttt{\textbackslash runningtitle\{Provide here an alternative
-%     shorter version of the title of your paper\}}\\
-%     \texttt{\textbackslash runningauthor\{Provide here the surnames of
-%     the authors of your paper, all separated by commas\}}
-%     \end{flushleft}
-%     Note that the text that appears as argument in \texttt{\textbackslash
-%       runningtitle} will be printed as a heading in the \emph{even}
-%     pages. The text that appears as argument in \texttt{\textbackslash
-%       runningauthor} will be printed as a heading in the \emph{odd}
-%     pages.  If even the author surnames do not fit, it is acceptable
-%     to give a subset of author names followed by ``et al.''
-
-%     %\item Use the file sample\_paper.tex as an example.
-
-%     \item The camera-ready versions of the accepted papers are 8
-%       pages, plus any additional pages needed for references.
-
-%     \item If you need to include additional appendices,
-%       you can include them in the supplementary
-%       material file.
-
-%     \item Please, do not change the layout given by the above
-%       instructions and by the style file.
-
-% \end{enumerate}
-
-% \subsubsection*{Acknowledgements}
-% All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
-% To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
-
-% \subsubsection*{References}
-
-% References follow the acknowledgements.  Use an unnumbered third level
-% heading for the references section.  Please use the same font
-% size for references as for the body of the paper---remember that
-% references do not count against your page length total.
-
-% \begin{thebibliography}{}
-% \setlength{\itemindent}{-\leftmargin}
-% \makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
-% \bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
-%     \newblock Performance of a stochastic learning microchip.
-%     \newblock In D. S. Touretzky (ed.),
-%     \textit{Advances in Neural Information Processing Systems 1}, 748--760.
-%     San Mateo, Calif.: Morgan Kaufmann.
-
-% \bibitem{} F.~Rosenblatt (1962).
-%     \newblock \textit{Principles of Neurodynamics.}
-%     \newblock Washington, D.C.: Spartan Books.
-
-% \bibitem{} G.~Tesauro (1989).
-%     \newblock Neurogammon wins computer Olympiad.
-%     \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
-% \end{thebibliography}
-
-% \end{document}

### 2022-08-23 ic/aistats2021/aistats2021_author_response.tex
Commit original repos

\documentclass{article}

\usepackage{aistats2021_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % define colors in text
\usepackage{xspace}         % fix spacing around commands
\usepackage{mathtools,natbib}

\setlength{\bibsep}{0.0pt}

\bibliographystyle{abbrv}

\begin{document}

We thank our reviewers and will fix all typos/grammar errors. Any
reproducability issues will be addressed when we release all implementation
(reviewer 1) and experiment code in early 2021. We will also add additional
details on $n$-schools notation and motivation (reviewers 2, 6).

\textbf{Reviewer 2}:

We respectfully point out that reviewer 2 may have entirely misunderstood our work as
posterior adaptation rather than amortized inference and we strongly recommend
reconsideration of the claims in this review.
While reviewer 2 summarizes the ``first step'' of LIC as running
MCMC to sample $p(x \mid y)$, in fact LIC (see equation 2) and other methods
for amortized inference \cite{le2017inference} sample the joint distribution
$p(x,y)$ by running the probabilistic program forwards. This misunderstanding
is further evidenced by reviewer 2's major concern on lack of novelty
compared against three prior works on posterior adaptation
\cite{de2001variational,neklyudov2018metropolis,titsias2019gradient}, when in
fact our work is on amortized inference (see section 1.2). As amortized
inference must provide speedups (i.e. amortize the upfront compilation
cost) across many different observations $y$ and not just a single target
posterior $p(x \mid y)$, the prior works in reviewer 2's first major concern
are not comparable to our work. Prior work in IC \cite{le2017inference} also
does not make direct comparison against posterior adaptation methods, and we have
followed suit.
Additionally, reviewer 2's second major concern related to being ``safe from
mode-collapse'' are likely a consequence of misunderstanding the ``first
step'' as performing MCMC sampling of the posterior. Because no Markov chains
are actually involved during learning for LIC, there is no concern of any chains getting stuck
/ collapsing.

\textbf{Reviewer 3}:

The major criticism from the reviewer in ``Appropriateness of Graph Neural Networks in theory''
is that averaging together nodes in a Markov blanket (e.g. the mean and variance) may
not yield meaningful signal. While this is true for a 1D-embedding, we point out that
the vector space embeddings (``embedding net''s in Figure 2) yield multi-dimensional
embeddings where different dimensions handle how nodes in the Markov blanket ``play
different roles from one another.'' For the mean/variance example, the
embedding networks could potentially learn to embed the mean to dimension 1
and variance to dimension 2 so the resulting sum yields a 2-dimensional
[mean, variance] vector.

Regarding weakness of empirical validation, we attempted to conduct
the first experiment of the two ``missing'' experiments
and found LIC to improve over IC \cite{le2017inference} in terms of 
compilation/inference times, model size, and ESS/PLL on Bayesian logistic
regression and $n$-schools. 
However, we opted to not include these results because IC's reference
implementation (\texttt{pyprob}, \cite{le2017inference}) lacks support for
vector-valued random variables (see e.g.
\texttt{pyprob.distributions.mixture.Mixture.event\_shape}) so we were forced
to unvectorize our \texttt{pyprob} models. If there is interest, we can
include these results in the appendix.

Regarding relation to prior work, we concede that handling the open-universe
case is not novel and will remove this claim from our paper. However,
although neural Gibbs proposers are also used in \cite{wang2018meta}, a key
distinction is that they ``focus on hand-identified common structures'' to
construct proposers whereas LIC (and related work in IC) automate proposal
construction. While their framework includes Markov blanket parameterized
proposers, they raise the issue that ``Markov blankets... might not be
consistent across all instantiations'' whereas we use permutation-invariance
to motivate graph neural networks as a solution for summarizing dynamic
Markov blankets. We will include this discussion in related works.

With respect to additional questions:
    Infinite-domain discrete random variables -- yes, and we are
    actively looking at $q(\cdot; \phi) \sim \text{Poisson}(\phi)$ and
    negative-binomial.
    ``Is $x_i$ actually part of the input?'' -- yes, see Figure 2.
    Imperative vs declarative -- agree, will fix in camera-ready.
    On reproduceability -- GMM components is on page 7.
\bibliography{response}

\end{document}

### 2022-08-23 ic/aistats2021/paper.tex
Commit original repos

\documentclass[twoside]{article}

%\usepackage{aistats2021}
\usepackage[accepted]{aistats2021}
% If your paper is accepted, change the options for the package
% aistats2021 as follows:
%
%\usepackage[accepted]{aistats2021}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
\special{papersize = 8.5in, 11in}
\setlength{\pdfpageheight}{11in}
\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

% If you use BibTeX in apalike style, activate the following line:
\bibliographystyle{apalike}

\input{macros.tex}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

  \aistatstitle{Accelerating Metropolis-Hastings with Lightweight Inference Compilation}

  \aistatsauthor{ Feynman Liang \And Nimar Arora \And Nazanin Tehrani \\ \\
    \AND Yucen Li \And Michael Tingley \And Erik Meijer }
  \vspace{-0.60in}
  \aistatsaddress{ UC Berkeley \\ feynman@berkeley.edu
    \And Facebook \\ nimarora@fb.com
    \And Facebook \\ nazaninkt@fb.com
    \AND Facebook \\ yucenli@fb.com
    \And Facebook \\ tingley@fb.com
    \And Facebook \\ erikm@fb.com}
]

\runningauthor{Liang, Arora, Tehrani, Li, Tingley, Meijer}
\begin{abstract}
  In order to construct accurate proposers for Metropolis-Hastings Markov
  Chain Monte Carlo, we integrate ideas from probabilistic graphical models
  and neural networks in a framework we call Lightweight
  Inference Compilation (LIC). LIC implements amortized inference within an
  open-universe declarative probabilistic programming language (PPL).
  Graph neural networks are used to parameterize proposal distributions
  as functions of Markov blankets, which during ``compilation'' are optimized
  to approximate single-site Gibbs sampling distributions. Unlike prior work
  in inference compilation (IC), LIC forgoes importance sampling of linear
  execution traces in favor of operating directly on Bayesian networks.
  Through using a declarative PPL, the Markov blankets of nodes (which may be
  non-static) are queried at inference-time to produce proposers. Experimental
  results show LIC can produce proposers which have less parameters, greater
  robustness to nuisance random variables, and improved posterior sampling in
  a Bayesian logistic regression and $n$-schools inference application.
\end{abstract}

\section{Background}

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{Figures/intuition.pdf}
  \caption{
    Intuition for Lightweight Inference Compilation (LIC).
    LIC uses samples $(x_i, y_i) \overset{\text{iid}}{\sim} p$ (blue ``x'' in left)
    drawn from the joint density $p(x, y)$ to approximate the expected inclusive KL-divergence
    $\bE_{p(y)} D_{\text{KL}}\infdivx{p(x \mid y)}{q(x; \phi(y))}$
    between the posterior $p(x \mid y)$ and the LIC proposal distribution
    $q(x ; \phi(y))$. For an observation $y=0.25$ (dashed red line in left),
    the posterior $p(x \mid y=0.25)$ (dashed red line in right) is ``approximated''
    by samples ``close'' to $y$ (blue ``x'' in right) to form an empirical
    inclusive KL-divergence minimized by LIC. As inclusive KL-divergence
    encourages a mass-covering / mean-seeking fit, the resulting proposal
    distribution $q(x;\phi(y=0.25), K=1)$ (green dotted line in right)
    when using a single ($K=1$) Gaussian proposal density covers both modes
    and can successfully propose moves which cross between the two mixture
    components. Using a 2-component ($K=2$) GMM proposal density results in
    $q(x;\phi(y=0.25), K=2)$ (purple solid line in right) which captures both the
    bi-modality of the posterior as well as the low probability region between
    the two modes.
    As a result of sampling the generative model, LIC can discover
    both posterior modes and their appropriate mixture weights
    (whereas other state of the art MCMC samplers fail, see \cref{fig:gmm_mode_escape}).
  }\label{fig:intuition}
\end{figure*}

Deriving and implementing samplers has traditionally been a high-effort and
application-specific endeavour \citep{porteous2008fast,murray2010elliptical},
motivating the development of general-purpose probabilistic programming
languages (PPLs) where a non-expert can specify a generative model (i.e.
joint distribution) $p(\vx, \vy)$ and the software automatically performs
inference to sample latent variables $\vx$ from the posterior $p(\vx \mid
  \vy)$ conditioned on observations $\vy$. While exceptions exist,
modern general-purpose PPLs typically implement variational inference
\citep{bingham2019pyro}, importance sampling
\citep{wood2014new,le2017inference}, or Monte Carlo Markov Chain (MCMC,
\cite{wingate2011lightweight,tehrani2020beanmachine}).

Our work focuses on MCMC. More specifically, we target lightweight
Metropolis-Hastings (LMH, \cite{wingate2011lightweight}) within a recently developed
declarative PPL called \texttt{beanmachine} \citep{tehrani2020beanmachine}.
The performance of Metropolis-Hastings critically depends on the quality of
the proposal distribution used, which is the primary goal of LIC.
Broadly speaking, LIC amortizes MCMC by constructing function approximators
(parameterized by graph neural networks) from Markov blankets to proposal
distribution parameters which are learned via forward simulation and training
to match the full conditionals (equivalently, to minimize the inclusive KL
divergence). In doing so, LIC makes the following contributions:
\begin{enumerate}
  \item We present a novel implementation of inference compilation (IC) within an
        open-universe declarative PPL which combines Markov blanket structure with
        graph neural network architectures. Our primary innovation is the use of a
        graph neural network to aggregate Markov blankets, which enables proposers
        to ignore irrelevant variables by construction.
  \item We demonstrate LIC's ability to escape local modes, improved
        robustness to nuisance random variables, and improvements over
        state-of-the-art methods across a number of metrics in two
        industry-relevant applications.
\end{enumerate}

\subsection{Declarative Probabilistic Programming}

To make Bayesian inference accessible to non-experts, PPLs provide
user-friendly primitives in high-level programming languages for abstraction
and composition in model representation
\citep{goodman2013principles,ghahramani2015probabilistic}. Existing PPLs
can be broadly classified based on the representation inference is performed
over, with declarative PPLs \citep{lunn2000winbugs,plummer2003jags,milch20071,tehrani2020beanmachine}
performing inference over Bayesian graphical networks and imperative PPLs
conducting importance sampling \citep{wood2014new} or MCMC
\citep{wingate2011lightweight} on linearized execution traces.
Because an execution trace is a topological sort of an instantiated
Bayesian network, declarative PPLs naturally preserve additional model
structure such as Markov blanket relationships while imperative PPLs require
additional dependency tracking instrumentation \citep{mansinghka2014venture}
or analysis tooling \citep{gorinova2020conditional,cusumano2019gen}
to achieve similar functionality.

\begin{definition}
  The \emph{Markov Blanket} $\MB(x_i)$ of a node $x_i$
  is the minimal set of random variables such that
  \begin{align}
    p(x_i \mid \vx_{-i}, \vy)
     & = p(x_i \mid \MB(x_i))
    \label{eq:gibbs-mb}
  \end{align}
  In a Bayesian network, $\MB(x_i)$ consists of the parents, children, and
  children's parents of $x_i$ \citep{pearl1987evidential}.
\end{definition}

% Universal PPLs \citep{milch20071,wood2014new} \todo{cite beanmachine}
% permit an unknown (potentially unbounded) number of objects and Turing-complete control flow.

\subsection{Inference Compilation}

Amortized inference \citep{gershman2014amortized} refers to the re-use of
initial up-front learning to improve future queries. In context of
Bayesian inference \citep{marino2018iterative,zhang2018advances} and IC
\citep{paige2016inference,weilbach2019efficient,harvey2019attention},
this means using acceleration performing multiple inferences over
different observations $\vy$ to amortize a one-time
``compilation time.'' While compilation in both trace-based IC
\citep{paige2016inference,le2017inference,harvey2019attention} and LIC
consists of drawing forward samples from the generative model $p(\vx, \vy)$ and
training neural networks to minimize inclusive KL-divergence,
trace-based IC uses the resulting neural network artifacts to parameterize
proposal distributions for importance sampling while LIC uses them for MCMC
proposers.

\subsection{Lightweight Metropolis Hastings}

Lightweight Metropolis Hastings (LMH,
\cite{wingate2011lightweight,ritchie2016c3}) is a simple but general method
of implementing a PPL. In LMH, random variables are assigned string
identifiers and their values and likelihoods are stored in a database
(\texttt{World} in our implementation). MCMC is performed in a
Metropolis-within-Gibbs manner where: (1) a single random variable is
modified according to a proposal distribution while all others remain fixed,
(2) the computations dependant upon the modified random variable (e.g.\ the
continuation in a continuation-passing-style implementation) is re-executed
to generate the remaining trace (re-using the database values for all other
random variables) and the new trace's likelihood, and finally (3) a
Metropolis-Hastings accept/reject correction is performed.

While a number of choices for proposal distribution exist, the single-site
Gibbs sampler which proposes from \cref{eq:gibbs-mb} enjoys a 100\%
acceptance probability \citep{pearl1987evidential} and provides a good choice
when available \citep{lunn2000winbugs,plummer2003jags}. Unfortunately,
outside of discrete models they are oftentimes intractable to directly sample
so another proposal distribution must be used. LIC seeks to approximate these
single-site Gibbs distributions using tractable neural network
approximations.

\subsection{Related Works}

\begin{figure*}
  \centering
  \includegraphics[width=0.90\linewidth]{Figures/lic_overview.pdf}
  \caption{The Markov blankets in the Bayesian network for \cref{eq:conj-normal}
    (left, expressed in plate notation) are available in a declarative PPL,
    and are used as inputs to LIC. The LIC proposer for node $x_1$ (right)
    is obtained by first performing neural network embedding of $x_1$
    and every node in its Markov blanket, followed by a graph convolutional
    network aggregation over the Markov blanket of $x_1$.
    The resulting vectors are then combined to yield a parameter vector $\phi(x_1)$
    for a proposal distribution $q(\cdot; \phi(x_1))$ which is then sampled
    for proposing an update within Metropolis-Hastings.
  }\label{fig:example-ic-arch}
\end{figure*}

Prior work on IC in imperative PPLs can be broadly classified based on the
order in which nodes are sampled. ``Backwards'' methods approximate an inverse
factorization, starting at observations and using IC artifacts to propose
propose parent random variables. Along these lines, \cite{paige2016inference}
use neural autoregressive density estimators but heuristically invert the model
by expanding parent sets. \cite{webb2018faithful} proposes a more principled
approach utilizing minimal I-maps and demonstrate that minimality of inputs to
IC neural networks can be beneficial; an insight also exploited through LIC's
usage of Markov blankets. Unfortunately, model inversion is not possible
in universal PPLs \citep{le2017inference}.

The other group of ``forwards'' methods operate in the same direction as the
probabilistic model's dependency graph. Starting at root nodes, these methods
produce inference compilation artifacts which map an execution trace's prefix
to a proposal distribution. In \cite{ritchie2016deep}, a user-specified
model-specific guide program parameterizes the proposer's architecture and
results in more interpretable IC artifacts at the expense of increased user
effort. \cite{le2017inference} automates this by using a recurrent neural
network (RNN) to summarize the execution prefix while constructing a node's
proposal distribution. This approach suffers from well-documented RNN
limitations learning long distance dependencies
\citep{hochreiter1998vanishing}, requiring mechanisms like attention
\citep{harvey2019attention} to avoid degradation in the presence of long
execution trace prefixes (e.g. when nuisance random variables are present).

With respect to prior work, LIC is most similar to the attention-based
extension \citep{harvey2019attention} of \cite{le2017inference}. Both methods
minimize inclusive KL-divergence empirically approximated by samples from the
generative model $p(\vx, \vy)$, and both methods use neural networks to
produce a parametric proposal distribution from a set of inputs sufficient
for determining a node's posterior distribution. However, important distinctions
include (1) LIC's use of a declarative PPL implies Markov blanket
relationships are directly queryable and ameliorates the need for also
learning an attention mechanism, and (2) LIC uses a graph neural network to
summarize the Markov blanket rather than a RNN over the execution trace
prefix.
\cite{wang2017meta} is also closely related to LIC as both are methods for
amortizing MCMC by learning Markov blanket parameterized neural Gibbs
proposers, but a key difference is that LIC exploits permutation-invariance
of graph neural networks to address the issue where ``Markov blankets...
might not be consistent across all instantiations'' whereas
\cite{wang2017meta} restricts ``focus on hand-identified common structur[al
    motifs]'' for constructing proposers.

% \todo{Separate adaptation into separate publication?}
% \citep{marino2018iterative} iteratively optimizes approximate posterior gradients of ELBO
% analogous to our adaptation scheme (except we do inclusive KL).

\section{Lightweight Inference Compilation}

\subsection{Architecture}
\label{ssec:architecture}

\Cref{fig:example-ic-arch} shows a sketch of LIC's architecture.
For every latent node $x_i$, LIC constructs a mapping $(x_i, \MB(x_i)) \mapsto \phi(x_i)$
parameterized by feedforward and graph neural networks to produce a parameter
vector $\phi(x_i)$ for a parametric density estimator $q(\cdot; \phi(x_i))$.
Every node $x_i$ has feedforward ``node embedding network'' used
to map the value of the underlying random variable into a vector space of
common dimensionality. The set of nodes in the Markov blanket are then
summarized to a fixed-length vector following \cref{ssec:dynamic-mb}, and
a feedforward neural network ultimately maps the concatenation of the
node's embedding with its Markov blanket summary to proposal
distribution parameters $\phi(x_i)$.

\subsubsection{Dynamic Markov Blanket embeddings}
\label{ssec:dynamic-mb}

% SEE https://cs.mcgill.ca/~wlh/comp766/files/chapter4_draft_mar29.pdf

Because a node's Markov blanket may vary in both its size and elements (e.g.
in a GMM, a data point's component membership may change during MCMC),
$\MB(x_i)$ is a non-static set of vectors (albeit all of the same dimension
after node embeddings are applied) and a feed-forward network with fixed
input dimension is unsuitable for computing a fixed-length proposal parameter
vector $\phi(x_i)$. Furthermore, Markov blankets (unlike execution trace
prefixes) are unordered sets and lack a natural ordering hence use of a
RNN as done in \cite{le2017inference} is inappropriate.
Instead, LIC draws motivation from graph neural networks
\citep{scarselli2008graph,dai2016discriminative} which have demonstrated
superior results in representation learning \cite{bruna2013spectral} and
performs summarization of Markov Blankets following \cite{kipf2016semi} by
defining
\begin{align*}
  \phi(x_i) = \sigma\left(
  \matr{W}
  \underset{x_j \in \text{MB}(x_i)}{\square}
  \frac{1}{\sqrt{\lvert \text{MB}(x_i) \rvert \lvert \text{MB}(x_j) \rvert}}
  f_j(x_j)
  \right)
\end{align*}
where $f_j(x_j)$ denotes the output of the node embedding network for node
$x_j$ when provided its current value as an input, $\square$ is any
differentiable permutation-invariant function (summation in LIC's case), and
$\sigma$ is an activation function. This technique is analogous to the ``Deep
sets trick'' \citep{zaheer2017deep}, and we expect it to perform well when
elements of the Markov blanket are conditionally exchangeable. However,
we note this assumption of permutation invariance may not always hold
hence additional investigation into more sophisticated aggregation
schemes (e.g. identifying exchangeable elements, using permutation-dependent
aggregators like RNNs) is important future work.

\subsubsection{Parameterized density estimation}

The resulting parameter vectors $\phi(x_i)$ of LIC are ultimately used to
parameterize proposal distributions $q(x_i; \phi(\MB(x_i)))$ for MCMC sampling.
For discrete $x_i$, LIC directly estimates logit scores over the
support. For continuous $x_i$, LIC transforms continuous $x_i$ to
unconstrained space following \cite{carpenter2017stan} and models
the density using a Gaussian mixture model (GMM). Note that although more
sophisticated density estimators such as masked autoregressive flows
\citep{kingma2016improved,papamakarios2017masked} can equally be used.

\subsection{Objective Function}

To ``compile'' LIC, parameters are optimized to minimize the inclusive
KL-divergence between the posterior distributions and inference compilation
proposers: $D_{\text{KL}}\infdivx{p(\vx \mid \vy)}{q(\vx \mid \vy; \phi)}$.
Consistent with \cite{le2017inference}, observations $\vy$
are sampled from the marginal distribution $p(\vy)$, but note that this may
not be representative of observations $\vy$ encountered at inference. The
resulting objective function is given by
\begin{align}
   & \bE_{p(\vy)}\left[
    D_{\text{KL}}\infdivx{p(\vx \mid \vy)}{q(\vx \mid \vy; \phi)}
  \right]                                               \label{eq:e-kl-inc} \\
   & = \bE_{p(\vx, \vy)}\left[
    \log \frac{p(\vx \mid \vy)}{q(\vx \mid \vy; \phi)}
  \right]                                               \nonumber           \\
   & \propto \bE_{p(\vx, \vy)}\left[
    -\log q(\vx \mid \vy; \phi)
  \right]                                               \nonumber           \\
   & \approx \sum_{i=1}^N  -\log q(\vx \mid \vy; \phi),
  \qquad (\vx,\vy) \simiid p(\vx,\vy)                   \nonumber           \\
   & \eqqcolon \mathcal{L}(\phi) \nonumber
\end{align}
where we have neglected a conditional entropy term independent of $\phi$
and performed Monte-Carlo approximation to an expectation.
The intuition for this objective is shown in \Cref{fig:intuition}, which shows
how samples from $p(\vx, \vy)$ (left) form an empirical joint approximation where
``slices'' (at $y=0.25$ in \cref{fig:intuition}) yield posterior approximations
which the objective is computed over (right).

% \subsubsection{Open-universe support / novel data at runtime}

% Because the number of random variables in an open-universe probabilistic can
% be unbounded, it is impossible for any finite-time training procedure which relies
% on sampling the generative model $p(\vx, \vy)$ to encounter all possible random variables.
% As a consequence, any implementation of IC within a universal PPL must address the issue
% of novel data encountered at inference time. This issue is not explicitly addressed
% in prior work \citep{le2017inference,harvey2019attention}. In LIC's case,
% if a node $x_i$ does not have an existing LIC artifact $x_i \mapsto \phi(x_i)$
% then we fall back to proposing from the parent-conditional prior $q(x_i;
%   \MB(x_i), \phi_i) = p(x_i \mid \Pa(x_i))$ (i.e. ancestral sampling).

\section{Experiments}

To validate LIC's competitiveness, we conducted experiments benchmarking a variety
of desired behaviors and relevant applications. In particular:
\begin{itemize}
  \item Training on samples from the joint distribution $p(\vx, \vy)$ should enable
        discovery of distant modes, so LIC samplers should be less likely to
        to get ``stuck'' in a local mode. We validate this in \cref{ssec:gmm}
        using a GMM mode escape experiment, where we see LIC escape not only
        escape a local mode but also yield accurate mixture component
        weights.
  \item When there is no approximation error (i.e. the true posterior density is
        within the family of parametric densities representable by LIC), we expect
        LIC to closely approximate the posterior at least for the range of
        observations $\vy$ sampled during compilation (\cref{eq:e-kl-inc})
        with high probability under the prior $p(\vy)$. \Cref{ssec:conj-gaussian}
        shows this is indeed the case in a conjugate Gaussian-Gaussian model
        where a closed form expression for the posterior is available.
  \item Because Markov blankets can be explicitly queried, we expect LIC's
        performance to be unaffected by the presence of nuisance random variables
        (i.e. random variables which extend the execution trace but are statistically
        independent from the observations and queried latent variables). This is
        confirmed in \cref{ssec:nuisance} using the probabilistic program from
        \cite{harvey2019attention}, where we see
        trace-based IC suffering an order of
        magnitude increase in model parameters and compilation time while
        yielding an effective sample size almost $5\times$ smaller
        (\Cref{fig:nuisance}).
  \item To verify LIC yields competitive performance in applications of
        interest, we benchmark LIC against other state-of-the-art MCMC methods
        on a Bayesian logistic regression problem
        (\cref{ssec:blr}) and on a generalization of the classical eight schools
        problem \citep{rubin1981estimation} called $n$-schools
        (\cref{ssec:nschools}) which is used in production at a large internet
        company for Bayesian meta-analysis \cite{sutton2001bayesian}.
        We find that LIC exceeds the performance of adaptive random walk
        Metropolis-Hastings \citep{garthwaite2016adaptive}
        and Newtonian Monte Carlo \citep{arora2020newtonian} and
        yields comparable performance to NUTS \citep{hoffman2014no} despite
        being implemented in an interpreted (Python) versus compiled (C\texttt{++})
        language.
\end{itemize}

\begin{figure*}
  \includegraphics[width=\linewidth]{Figures/mode_escape.pdf}
  \caption{
    When sampling the bi-modal posterior density from \cref{fig:intuition},
    only inference compilation (IC, this paper) and adaptive step-size random walk
    Metropolis-Hastings (Adaptive RWMH, \cite{garthwaite2016adaptive}) are
    able to recover both posterior modes. Whereas RWMH's posterior samples
    erroneously assign approximately equal probability to both modes,
    IC's samples faithfully reproduce the ground truth and yields
    higher probability for the mode at $x=10$ than the mode at $x=0$.
  }\label{fig:gmm_mode_escape}
\end{figure*}

A reference implementation for LIC and code to reproduce our experiments
have been made publically available \citep{lic2020}.

\subsection{GMM mode escape}
\label{ssec:gmm}

Consider the multi-modal posterior resulting from conditioning on $y=0.25$ in
the 2-dimensional GMM in \cref{fig:intuition}, which is comprised of two
Gaussian components with greater mixture probability on the right-hand
component and a large energy barrier separating the two components. Because
LIC is compiled by training on samples from the joint distribution $p(x, y)$,
it is reasonable to expect LIC's proposers to assign high probability to
values for the latent variable $x$ from both modes. In contrast, uncompiled
methods such as random walk Metropolis-Hastings (RWMH) and NUTS may encounter
difficulty crossing the low-probability energy barrier and be unable to escape
the basin of attraction of the mode closest to their initialization.

This intuition is confirmed in \cref{fig:gmm_mode_escape}, which illustrates
kernel density estimates of 1,000 posterior samples obtained by a variety of
MCMC algorithms as well as ground truth samples. HMC with adaptive step size
\citep{hoffman2014no}, NMC, and NUTS with the default settings as implemented
in Stan \citep{carpenter2017stan} are all unable to escape the mode they are
initialized in. While both LIC and RWMH with adaptive step size escape the
local mode, RWMH's samples erroneously imply equal mixture component
probabilities whereas LIC's samples faithfully reproduce a higher component
probability for the right-hand mode.

\subsection{Conjugate Gaussian-Gaussian Model}
\label{ssec:conj-gaussian}

We next consider a Gaussian likelihood with a Gaussian mean prior, a
conjugate model with closed-form posterior given by:
\begin{align}
  x                                 \sim \mathcal{N}(0, \sigma_x),
                                    & \qquad y \sim \mathcal{N}(x, \sigma_y) \label{eq:conj-normal} \\
  \Pr[x \mid y, \sigma_x, \sigma_y] & \sim \mathcal{N}\left(
  \frac{\sigma_y^{-2}}{\sigma_x^{-2} + \sigma_{y}^{-2}} y,
  \frac{1}{\sigma_x^{-2} + \sigma_{y}^{-2}}
  \right) \nonumber
\end{align}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{Figures/normal_normal_mean_type1.pdf}
  \caption{In a conjugate normal-normal model, LIC's proposal distribution
    mean (dashed orange line) closely follows the closed-form posterior mean
    (blue solid line) across a wide range of observed values $y$.}\label{fig:normal-normal}
\end{figure}

There is minimal approximation error because the posterior density is in the
same family as LIC's GMM proposal distributions and the relationship between
the Markov blanket $\MB(x) = \{y\}$ and the posterior mean is a linear
function easily approximated (locally) by neural networks. As a result,
we expect LIC's proposal distribution to provide a good approximation to the
true posterior and LIC to approximately implement a direct posterior sampler.

To confirm this behavior, we trained LIC with a $K=1$ component GMM proposal
density on 1,000 samples and show the resulting LIC proposer's mean as the
observed value $y$ varies in \cref{fig:normal-normal}. Here $\sigma_x = 2$
and $\sigma_y = 0.1$, so the marginal distribution of $y$ (i.e. the
observations sampled during compilation in \cref{eq:e-kl-inc}) is Gaussian
with mean $0$ and standard deviation $\sqrt{\sigma_x^2 + \sigma_y^2} \approx
  2.0025$. Consistent with our expectations, LIC provides a good approximation
to the true posterior for observed values $y$ well-represented during
training (i.e. with high probability under the marginal $p(y)$). While LIC
also provides a reasonable proposer by extrapolation to less represented
observed values $y$, it is clear from \cref{fig:normal-normal} that the
approximation is less accurate. This motivates future work into modifying the
forward sampling distribution used to approximate \cref{eq:e-kl-inc} (e.g.
``inflating'' the prior) as well as adapting LIC towards the distribution of
observations $y$ used at inference time.

\subsection{Robustness to Nuisance Variables}
\label{ssec:nuisance}

An important innovation of LIC is its use of a declarative PPL
and ability to query for Markov blanket relationships so that only statistically
relevant inputs are utilized when constructing proposal distributions.
To validate this yields significant improvement over prior work in IC,
we reproduced an experiment from \cite{harvey2019attention} where
nuisance random variables (i.e. random variables which are statistically independent
from the queried latent variables/observations whose only purpose is to extend the
execution trace) are introduced and the impact on system performance is measured.
As trace-based inference compilation utilizes the execution trace prefix to summarize
program state, extending the trace of the program with nuisance random variables
typically results in degradation of performance due to difficulties encountered
by RNN in capturing long range dependencies as well as the
production of irrelevant neural network embedding artifacts.

We reproduce trace-based IC as described in \cite{le2017inference} using the
author-provided software package \citep{pyprob2020}, and implement Program 1
from \cite{harvey2019attention} with the source code illustrated in
\cref{lst:pyprob} where 100 nuisance random variables are added. Note that
although \texttt{nuisance} has no relationship to the remainder of the
program, the line number where they are instantiated has a dramatic impact on
performance. By extending the trace between where \texttt{x} and \texttt{y}
are defined, trace-based IC's RNNs degrade due to difficulty learning a
long-range dependency\citep{hochreiter1998vanishing} between the two
variables. For LIC, the equivalent program expressed in the
\texttt{beanmachine} declarative PPL \citep{tehrani2020beanmachine} is shown
in \cref{lst:bm}. In this case, the order in which random variable
declarations appear is irrelevant as all permutations describe the same
probabilistic graphical model.

\begin{lstlisting}[language=Python,caption={A version of Program 1 from \cite{harvey2019attention} to illustrate nuisance random variables},label={lst:pyprob}]
def magnitude(obs):
  x = sample(Normal(0, 10))
  for _ in range(100):
    nuisance = sample(Normal(0, 10))
  y = sample(Normal(0, 10))
  observe(
    obs**2,
    likelihood=Normal(
      x**2 + y**2, 
      0.1))
  return x
\end{lstlisting}

\begin{lstlisting}[language=Python,caption={The equivalent program in beanmachine, where independencies are explicit in program specification},label={lst:bm}]
class NuisanceModel:
  @random_variable
  def x(self):
      return dist.Normal(0, 10)
  @random_variable
  def nuisance(self, i):
      return dist.Normal(0, 10)
  @random_variable
  def y(self):
      return dist.Normal(0, 10)
  @random_variable
  def noisy_sq_length(self):
      return dist.Normal(
        self.x()**2 + self.y()**2, 
        0.1)
\end{lstlisting}

\Cref{fig:nuisance} compares the results between LIC and trace-based IC
\citep{le2017inference} for this nuisance variable model. Both \texttt{pyprob}'s
defaults (1 layer 4 dimension sample embedding, 64 dimension address
embedding, 8 dimension distribution type embedding, 10 component GMM
proposer, 1 layer 512 dimension LSTM) and LIC's defaults (used for all
experiments in this paper, 1 layer 4 dimension node embedding, 3 layer 8
dimension Markov blanket embedding, 1 layer node proposal network) with a 10
component GMM proposer are trained on 10,000 samples and subsequently used to
draw 100 posterior samples. Although model size is not directly comparable due
differences in model architecture, \texttt{pyprob}'s resulting models were over $7\times$
larger than those of LIC. Furthermore, despite requiring more than $10\times$ longer
time to train, the resulting sampler produced by \texttt{pyprob} yields
an effective sample size almost $5\times$ smaller than that produced by LIC.
\begin{figure}
  \centering
  \begin{tabular}{lccc}
    \toprule
           & \# params & compile time & ESS   \\
    \midrule
    LIC    & 3,358     & 44 sec.      & 49.75 \\
    PyProb & 21,952    & 472 sec.     & 10.99 \\
    \bottomrule
  \end{tabular}
  \caption{Number of parameters, compilation time (10,000 samples), and effective sample size (100 samples)
    for inference compilation in LIC (this work)
    versus \citep{pyprob2020}
  }\label{fig:nuisance}
\end{figure}

\subsection{Bayesian Logistic Regression}
\label{ssec:blr}

Consider a Bayesian logistic regression model over $d$ covariates
with prior
$\vec{\beta} \sim \cN_{d+1}(\vec{0}_{d+1}, \diag(10, 2.5\vec{1}_{d}))$
and likelihood $y_i \mid \vx_i \simiid \text{Bernoulli}(\sigma(\vec{\beta}^\top \vx_i))$
where $\sigma(t) = (1 + e^{-t})^{-1}$ is the logistic function.
This model is appropriate in a wide range of classification problems where prior
knowledge about the regression coefficients $\vec{\beta}$ are available.

\Cref{fig:blr} shows the results of performing inference using LIC compared
against other MCMC inference methods. Results for existing IC approaches
\citep{le2017inference} are omitted because they are not comparable due to
lack of support for vector-valued random variables in the publically
available reference implementation \citep{pyprob2020}. All methods yield
similar predictive log-likelihoods on held-out test data, but LIC and NUTS
yield significantly higher ESS and $\widehat{R}$s closer to $1.0$ suggesting
better mixing and more effective sampling.

\begin{figure*}
  \centering
  \caption{
    Results drawing 100 samples across 10 chains (after 1,000 burn-in / adaptation / compilation samples)
    on two Bayesian inference problems.
    The top row of each subfigure aggregates over the 10 chains the (left to
    right) compilation time (time to instantiate the Python class, includes
    neural network training for LIC and Stan's C\texttt{++} codegen and compilation
    for NUTS), inference time (time from invoking \texttt{infer()} to when
    sampling terminates), and the predictive log-likelihood (PLL) on held-out
    test data.
    The bottom row aggregates over both the 10 chains as well as over all
    latent variables the (left to right) estimated expected sample size (ESS,
    higher is better, \cite{geyer2011introduction}) and the rank normalized
    $\widehat{R}$ diagnostic (Rhat, closer to $1$ is better, \cite{vehtari2020rank}).
  }\label{fig:blr_nschools}
  \begin{subfigure}[b]{\textwidth}
    \centering
    \caption{Bayesian logistic regression ($2000$ rows, $10$ features).
      Both LIC and Stan (NUTS) amortize the upfront compilation cost with accelerated inference times.
      LIC achieves comparable PLL to NUTS \cite{hoffman2014no} and other methods,
      and yields ESS comparable to NUTS and higher than any other method.
      The $\widehat{R}$ of LIC is close to $1$ (similar to NUTS and lower than
      all other methods).
    }\label{fig:blr}
    \includegraphics[width=\linewidth]{Figures/blr_pll_type1.pdf}
    \includegraphics[width=0.6\linewidth]{Figures/blr_ess_rhat_type1.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{\textwidth}
    \centering
    \caption{n-schools (1000 schools, 8 states, 5 districts, 5 types).
      Again, increased compilation times are offset by accelerated inference times.
      In this case, LIC achieves comparable PLL to NUTS while simultaneously producing
      higher ESS and lower $\widehat{R}$, suggesting that the resulting posterior samples
      are less autocorrelated and provide a more accurate posterior approximation.
    }\label{fig:nschools}
    \includegraphics[width=\linewidth]{Figures/nschools_pll_type1.pdf}
    \includegraphics[width=0.66\linewidth]{Figures/nschools_ess_rhat_type1.pdf}
  \end{subfigure}
\end{figure*}

\subsection{n-Schools}
\label{ssec:nschools}

The eight schools model \citep{rubin1981estimation} is a Bayesian
hierarchical model originally used to model the effectiveness of schools at
improving SAT scores. $n$-schools is a generalization of this model from $8$
to $n$ possible treatments, and is used at a large internet company for
performing Bayesian meta-analysis \cite{sutton2001bayesian} to estimate
(fixed) effect sizes.
Let $K$ denote the total number of schools, $n_j$ the number of
districts/states/types, and $j_k$ the district/state/type of school $k$.
\begin{align*}
  \beta_0     & \sim \text{StudentT}(3, 0, 10)                                                                         \\
  \tau_{i}    & \sim \text{HalfCauchy}(\sigma_i) \qquad \text{for}~i \in [\text{district}, \text{state}, \text{type}]  \\
  \beta_{i,j} & \sim \cN(0, \tau_i)  \qquad \text{for}~i \in [\text{district}, \text{state}, \text{type}], j \in [n_i] \\
  y_k         & \sim \cN(\beta_0 + \sum_i \beta_{i,j_k}, \sigma_k)
\end{align*}
The treatment effects $y_k$ and standard errors $\sigma_i$ and $\sigma_k$ are observed.

Intuitively, each ``school'' corresponds to a set of treatment parameters
(here a tuple of district, state, and type) and $\beta_{i,j}$ measures the
average change in response $y$ when treatment parameter $i$ is set equal to
$j$ (e.g. $\beta_{\text{state},\text{CA}}$ measures the change in SAT scores
when a school is located in California).

\Cref{fig:nschools} presents results in a format analogous to
\cref{ssec:blr}. Here, we see that while both LIC and NUTS yield higher PLLs
(with NUTS outperforming LIC in this case), LICs ESS is significantly higher
than other compared methods. Additionally, the $\widehat{R}$ of NUTS is also
larger than the other methods which suggests that even after 1,000 burn-in samples
NUTS has still not properly mixed.

\section{Conclusion}

We introduced Lightweight Inference Compilation (LIC) for building high
quality single-site proposal distributions to accelerate Metropolis-Hastings
MCMC. LIC utilizes declarative probabilistic programming to retain graphical
model structure and graph neural networks to approximate single-site Gibbs
sampling distributions. To our knowledge, LIC is the first proposed method
for inference compilation within an open-universe declarative probabilistic
programming language and an open-source implementation will be released in
early 2021. Compared to prior work, LIC's use of Markov blankets resolves the
need for attention to handle nuisance random variances and yields posterior
sampling comparable to state-of-the-art MCMC samplers such as NUTS and
adaptive RWMH.

% TODO: include in camera ready
\subsubsection*{Acknowledgements}
Feynman Liang is supported by a NPSC Graduate Fellowship.
This work was completed during an internship at Facebook.

\bibliography{refs.bib}

\end{document}

### 2022-08-23 ic/aistats2021/poster/main.tex
Commit original repos

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Jacobs Landscape Poster
% LaTeX Template
% Version 1.1 (14/06/14)
%
% Created by:
% Computational Physics and Biophysics Group, Jacobs University
% https://teamwork.jacobs-university.de:8443/confluence/display/CoPandBiG/LaTeX+Poster
% 
% Further modified by:
% Nathaniel Johnston (nathaniel@njohnston.ca)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[final]{beamer}

\usepackage[scale=1.24]{beamerposter} % Use the beamerposter package for laying out the poster

\usetheme{confposter} % Use the confposter theme supplied with this template

\setbeamercolor{block title}{fg=ngreen,bg=white} % Colors of the block titles
\setbeamercolor{block body}{fg=black,bg=white} % Colors of the body of blocks
\setbeamercolor{block alerted title}{fg=white,bg=dblue!70} % Colors of the highlighted block titles
\setbeamercolor{block alerted body}{fg=black,bg=dblue!10} % Colors of the body of highlighted blocks
% Many more colors are available for use in beamerthemeconfposter.sty

%-----------------------------------------------------------
% Define the column widths and overall poster size
% To set effective sepwid, onecolwid and twocolwid values, first choose how many columns you want and how much separation you want between columns
% In this template, the separation width chosen is 0.024 of the paper width and a 4-column layout
% onecolwid should therefore be (1-(# of columns+1)*sepwid)/# of columns e.g. (1-(4+1)*0.024)/4 = 0.22
% Set twocolwid to be (2*onecolwid)+sepwid = 0.464
% Set threecolwid to be (3*onecolwid)+2*sepwid = 0.708

\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\paperwidth}{48in} % A0 width: 46.8in
\setlength{\paperheight}{36in} % A0 height: 33.1in
\setlength{\sepwid}{0.024\paperwidth} % Separation width (white space) between columns
\setlength{\onecolwid}{0.22\paperwidth} % Width of one column
\setlength{\twocolwid}{0.464\paperwidth} % Width of two columns
\setlength{\threecolwid}{0.708\paperwidth} % Width of three columns
\setlength{\topmargin}{-0.5in} % Reduce the top margin size
%-----------------------------------------------------------

\usepackage{graphicx}  % Required for including images

\usepackage{booktabs} % Top and bottom rules for tables

\usepackage{setspace}
\usepackage{listings}
\setbeamertemplate{bibliography entry article}{}
\setbeamertemplate{bibliography entry title}{}
\setbeamertemplate{bibliography entry location}{}
\setbeamertemplate{bibliography entry note}{}
\setbeamertemplate{bibliography entry howpublished}{}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\vx}{\vec{x}}
\newcommand{\simiid}{\overset{\text{iid}}{\sim}}
\DeclareMathOperator{\diag}{diag}

%----------------------------------------------------------------------------------------
%	TITLE SECTION 
%----------------------------------------------------------------------------------------

\title{Accelerating Metropolis-Hastings with Lightweight Inference Compilation} % Poster title

\author[shortname]{Feynman Liang\inst{1}, 
Nimar Arora\inst{2},
Nazanin Tehrani\inst{2},
Yucen Li\inst{2},
Michael Tingley\inst{2},
Erik Meijer\inst{2}
}

\institute[shortinst]{\inst{1} UC Berkeley,
  \inst{2} Facebook}

\setbeamertemplate{headline}{
 \leavevmode
  \begin{columns}[T]
    \begin{column}{.1\linewidth}
        \vskip1cm
        \hskip1cm
        \includegraphics[width=.8\linewidth]{Figures/UCBseal.png}
    \end{column}
        \begin{column}{.8\linewidth}
         \vskip2cm
         \centering
         \usebeamercolor{title in headline}{\color{jblue}\huge{\textbf{\inserttitle}}\\[0.5ex]}
         \usebeamercolor{author in headline}{\color{fg}\Large{\insertauthor}\\[1ex]}
         \usebeamercolor{institute in headline}{\color{fg}\large{\insertinstitute}\\[1ex]}
         \vskip1cm
        \end{column}
    \begin{column}{.1\linewidth}
        \vskip1cm
        \includegraphics[width=.9\linewidth]{Figures/FBseal.png}
        \hskip1cm
    \end{column}        
   \vspace{1cm}
  \end{columns}
 \vspace{0.5in}
 \hspace{0.5in}\begin{beamercolorbox}[wd=47in,colsep=0.15cm]{cboxb}\end{beamercolorbox}
 \vspace{0.1in}
}

%----------------------------------------------------------------------------------------

\begin{document}

\addtobeamertemplate{block end}{}{\vspace*{2ex}} % White space under blocks
\addtobeamertemplate{block alerted end}{}{\vspace*{2ex}} % White space under highlighted (alert) blocks

\setlength{\belowcaptionskip}{2ex} % White space under figures
\setlength\belowdisplayshortskip{2ex} % White space under equations

\begin{frame}[t,containsverbatim] % The whole poster is enclosed in one beamer frame

  \begin{columns}[t] % The whole poster consists of three major columns, the second of which is split into two columns twice - the [t] option aligns each column's content to the top

    \begin{column}{\sepwid}\end{column} % Empty spacer column

    \begin{column}{\onecolwid} % The first column

      %----------------------------------------------------------------------------------------
      %	OBJECTIVES
      %----------------------------------------------------------------------------------------

      \begin{alertblock}{Summary}
        Accelerate lightweight Metropolis-Hastings \cite{wingate2011lightweight}
        by using neural network approximations to Gibbs sampling distributions.
        Unlike prior work \cite{le2017inference}, lightweight inference compilation
        (LIC) leverages Markov blanket structure provided by its host
        probabilistic programming language (PPL) to inform its neural network
        architectures. As a result, LIC's proposers have less parameters,
        greater robustness to nuisance random variables, and improved
        posterior sampling in a Bayesian logistic regression and $n$-schools
        inference application
      \end{alertblock}

      %----------------------------------------------------------------------------------------
      %	Data Representation
      %----------------------------------------------------------------------------------------

      \begin{block}{Intuition for inference compilation}
        Leverage generative sampling (i.e. running the probabilistic program)
        for amortized proposers $q(x;\phi(x,y))$
        \begin{figure}
          \includegraphics[width=\textwidth,trim=0 20 305 0,clip]{../Figures/intuition.pdf}
          \includegraphics[width=\textwidth,trim=305 10 0 10,clip]{../Figures/intuition.pdf}
          \caption{
            Generative samples $(x,y) \sim p(x,y)$ yield information about the posterior
            $p(x \mid y)$ for multiple values of $y$, enabling amortized inference
          }
        \end{figure}
      \end{block}

      %----------------------------------------------------------------------------------------

      \hrule~\\
      \begin{singlespace}
        \rmfamily{\footnotesize{
            Presented at the 24\textsuperscript{th}\,International Conference
            on Artificial Intelligence and Statistics (AISTATS) 2021, San
            Diego\@. Copyright 2021\/ by the author(s).
          }}
      \end{singlespace}

    \end{column} % End of the first column

    \begin{column}{\sepwid}\end{column} % Empty spacer column

    \begin{column}{\twocolwid} % Begin a column which is two columns wide (column 2)

      %----------------------------------------------------------------------------------------
      %	IMPORTANT RESULT
      %----------------------------------------------------------------------------------------

      % \begin{alertblock}{High Level Model}

      %   Analysis tools:
      %     Semantic interpretation of specificity of LSTM hidden states via plotting gate activations for inputs
      %     What would Bach do? MAP estimate of next note
      %     Fixing LSTM weights and optimizing the inputs

      %   Training, unrolling, BPTT
      %   Sampling

      %   Implementation: keras, tensorflow

      % \end{alertblock}

      %----------------------------------------------------------------------------------------

      %----------------------------------------------------------------------------------------
      %	One-pass polyphonic
      %----------------------------------------------------------------------------------------

      % \begin{block}{Learning a conjugate posterior}
      %   \begin{itemize}
      %     \item Given initial seed, generate entire chorale
      %     \item Baseline: n/a, subjective evaluation
      %     \item Experiments:
      %           \begin{itemize}
      %             \item Bi-axial and grid architectures
      %             \item Convolutional vs recurrent dimensions
      %           \end{itemize}
      %   \end{itemize}

      %   \begin{figure}
      %     \includegraphics[width=0.95\linewidth]{../Figures/normal_normal_mean_type1.pdf}
      %     \caption{}
      %   \end{figure}
      % \end{block}

      % \begin{block}{Bi-modal GMM}
      %   \begin{itemize}
      %     \item MAP : what would Bach do?
      %     \item Interpreting hidden state activations
      %     \item Enhance ``Bach-ness'' of input
      %   \end{itemize}

      %   \begin{figure}
      %     \includegraphics[width=0.95\linewidth]{../Figures/mode_escape.pdf}
      %     \caption{}
      %   \end{figure}
      % \end{block}

      %----------------------------------------------------------------------------------------

      \begin{block}{LIC's architecture and training}
        \begin{itemize}
          \item Run probabilistic program forwards to generate $n$ joint samples $(x,y) \sim p(x,y)$
          \item Minimize Monte-Carlo estimate of inclusive KL-divergence
                \[
                  \mathbb{E}_{y \sim p(y)} KL(p(x \mid y), q(x))
                  \approx \frac{1}{n} \sum_{i=1}^n \log \frac{p(x_i \mid y_i)}{q(x_i;\phi)}
                  \propto \frac{1}{n} \sum_{i=1}^n \log \frac{p(x_i, y_i)}{q(x_i;\phi)}
                \]
        \end{itemize}
        \begin{figure}
          \includegraphics[width=0.75\linewidth]{../Figures/lic_overview.pdf}
          \caption{LIC's architecture for a proposal distribution over
            latent $x_i$ conditioned on $y_i$ and with parent $\sigma$}
        \end{figure}
      \end{block}
      \vspace{-2cm}

      \begin{block}{Bayesian logistic regression and $n$-schools}
        \begin{columns}[t]
          \begin{column}{0.5\textwidth}
            Bayesian logistic regression (BLR)
            \newline
            \begin{align*}
              \vec{\beta} & \sim \cN_{d+1}(\vec{0}_{d+1}, \diag(10, 2.5\vec{1}_{d}))            \\
              y_i         & \mid \vx_i \simiid \text{Bernoulli}(\sigma(\vec{\beta}^\top \vx_i)) \\
              \sigma(t)   & = (1 + e^{-t})^{-1}
            \end{align*}
          \end{column}
          \begin{column}{0.5\textwidth}
            $n$-schools, a generalization of $8$-schools \cite{rubin1981estimation}
            used for Bayesian meta-analysis at a large internet company
            \begin{align*}
              \beta_0     & \sim \text{StudentT}(3, 0, 10)                                                                         \\
              \tau_{i}    & \sim \text{HalfCauchy}(\sigma_i) \qquad \text{for}~i \in [\text{district}, \text{state}, \text{type}]  \\
              \beta_{i,j} & \sim \cN(0, \tau_i)  \qquad \text{for}~i \in [\text{district}, \text{state}, \text{type}], j \in [n_i] \\
              y_k         & \sim \cN(\beta_0 + \sum_i \beta_{i,j_k}, \sigma_k)
            \end{align*}
          \end{column}
        \end{columns}

        \begin{figure}
          \includegraphics[width=0.475\linewidth,trim=17 0 205 0,clip]{../Figures/blr_pll_type1.pdf}
          \includegraphics[width=0.485\linewidth,trim=17 0 205 0,clip]{../Figures/nschools_pll_type1.pdf}
          \includegraphics[width=0.475\linewidth,trim=17 0 0 0,clip]{../Figures/blr_ess_rhat_type1.pdf}
          \includegraphics[width=0.485\linewidth,trim=27 0 0 0,clip]{../Figures/nschools_ess_rhat_type1.pdf}
          \caption{Empirical results for Bayesian logistic regression (left) and $n$-schools (right),
            compile time (seconds, lower is better, only applicable to LIC and NUTS), inference time (seconds, lower is better), effective sample size
            (higher is better), and $\hat{R}$ \cite{brooks1998general} (lower is better)}
        \end{figure}
      \end{block}

      %----------------------------------------------------------------------------------------

    \end{column} % End of the second column

    \begin{column}{\sepwid}\end{column} % Empty spacer column

    \begin{column}{\onecolwid} % The third column

      %----------------------------------------------------------------------------------------
      %	Status
      %----------------------------------------------------------------------------------------

      \begin{block}{Robustness to nuisance parameters}
        A version of Program 1 from \cite{harvey2019attention} illustrates
        LIC's improved robustness to nuisance parameters compared
        to \cite{pyprob2020}:
        \begin{lstlisting}[language=Python]
x = sample(Normal(0, 10))
for _ in range(100):
  nuisance = sample(Normal(0, 10))
y = sample(Normal(0, 10))
observe(obs**2, 
  likelihood=Normal( x**2 + y**2, 0.1))
        \end{lstlisting}
        \vspace{1cm}
        \begin{figure}
          \centering
          \begin{tabular}{lccc}
            \toprule
                   & \# params & compile time & ESS   \\
            \midrule
            LIC    & 3,358     & 44 sec.      & 49.75 \\
            PyProb & 21,952    & 472 sec.     & 10.99 \\
            \bottomrule
          \end{tabular}
        \end{figure}
      \end{block}

      %----------------------------------------------------------------------------------------
      %	REFERENCES
      %----------------------------------------------------------------------------------------

      \begin{block}{References}

        \nocite{*} % Insert publications even if they are not cited in the poster
        \footnotesize{
          \bibliographystyle{unsrt}
          \bibliography{./refs.bib}
        }

      \end{block}

      %----------------------------------------------------------------------------------------
      %	ACKNOWLEDGEMENTS
      %----------------------------------------------------------------------------------------

      \setbeamercolor{block title}{fg=red,bg=white} % Change the block title color

      \begin{block}{Acknowledgements}

        \small{\rmfamily{Feynman Liang is supported by a NPSC Graduate Fellowship.
            This work was done during an internship at Facebook.}} \\

      \end{block}

      %----------------------------------------------------------------------------------------
      %	CONTACT INFORMATION
      %----------------------------------------------------------------------------------------

      \setbeamercolor{block alerted title}{fg=black,bg=norange} % Change the alert block title colors
      \setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors

      \begin{alertblock}{Contact Information}

        \begin{itemize}
          \item \href{https://github.com/feynmanliang/lightweight-inference-compilation}{github.com/feynmanliang/lightweight-inference-compilation}
          \item \href{mailto:feynman@berkeley.edu}{feynman@berkeley.edu}
        \end{itemize}

      \end{alertblock}

      % \begin{center}
      %   \includegraphics[width=0.5\linewidth]{Figures/UCBwordmark.png}
      % \end{center}

      %----------------------------------------------------------------------------------------

    \end{column} % End of the third column

  \end{columns} % End of all the columns in the poster

\end{frame} % End of the enclosing frame

\end{document}

### 2022-08-23 ic/aistats2021/sample_paper.tex
Commit original repos

\documentclass[twoside]{article}

\usepackage{aistats2021}
% If your paper is accepted, change the options for the package
% aistats2021 as follows:
%
%\usepackage[accepted]{aistats2021}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
%\usepackage[round]{natbib}
%\renewcommand{\bibname}{References}
%\renewcommand{\bibsection}{\subsubsection*{\bibname}}

% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for Paper Submissions to AISTATS 2021}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The \textbf{Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

%%%
\subsection{Citations, Figure, References}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). 
%Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
Be sure that the sentence reads
correctly if the citation is deleted: e.g., instead of ``As described
by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
described by Cheesman (1985), we first frobulate the widgets.''

The references listed at the end of the paper can follow any style as long as it is used consistently.

%Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title,
and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
\textbf{PART}  &\textbf{DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you can include them in the supplementary material file.
You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
Note that reviewers are under no obligation to examine your supplementary material. 
If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.

You must use \texttt{aistats2021.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
The only difference is that it must be in a \emph{single-column} format.
You can use \texttt{supplement.tex} in our starter pack as a starting point.

\section{SUBMISSION INSTRUCTIONS}

To submit your paper to AISTATS 2021, please follow these instructions.

\begin{enumerate}
    \item Download \texttt{aistats2021.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
    Please, do not modify the style files as this might result in a formatting violation.

    \item Use \texttt{sample\_paper.tex} as a starting point.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage\{aistats2021\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages.
    \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
    \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
    \item Upload the PDF file along with other supplementary material files to the CMT website.
\end{enumerate}

\subsection{Camera-ready Papers}

%For the camera-ready paper, if you are using \LaTeX, please make sure
%that you follow these instructions.  
% (If you are not using \LaTeX,
%please make sure to achieve the same effect using your chosen
%typesetting package.)

If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
\begin{enumerate}
    %\item Download \texttt{fancyhdr.sty} -- the
    %\texttt{aistats2021.sty} file will make use of it.
    \item Change the beginning of your document to
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2021\}}
    \end{flushleft}
    The option \texttt{accepted} for the package
    \texttt{aistats2021.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative
    shorter version of the title of your paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of
    the authors of your paper, all separated by commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    %\item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, do not change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}
All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support. 
To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.

\subsubsection*{References}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

\begin{thebibliography}{}
\setlength{\itemindent}{-\leftmargin}
\makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
\bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
    \newblock Performance of a stochastic learning microchip.
    \newblock In D. S. Touretzky (ed.),
    \textit{Advances in Neural Information Processing Systems 1}, 748--760.
    San Mateo, Calif.: Morgan Kaufmann.

\bibitem{} F.~Rosenblatt (1962).
    \newblock \textit{Principles of Neurodynamics.}
    \newblock Washington, D.C.: Spartan Books.

\bibitem{} G.~Tesauro (1989).
    \newblock Neurogammon wins computer Olympiad.
    \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
\end{thebibliography}

\end{document}

### 2022-08-23 design/code/asymptotics.m

@@ -1,0 +1,58 @@
+d_list = floor(logspace(1,2,20));
+alpha_list = [0.5,0.75,0.9];
+w_list = [1,1,1];
+d_num = length(d_list);
+alpha_num = length(alpha_list);
+T = 100000;
+mse_error = zeros(d_num,alpha_num);
+mse_std = zeros(d_num,alpha_num);
+sg = 0; %1
+
+for i=1:d_num
+    d = d_list(i)
+    Sgm = diag(logspace(-2,0,d));%diag((d:-1:1).^(-2));%eye(d); Sgm(1,1)=0.1;% eye(d);%
+    B = sqrt(Sgm);
+    lbds = eig(Sgm);
+    %w = zeros(d,1); w(ceil(0.25*d))=1;%[zeros(d-1,1);1]; %[1;zeros(d-1,1)]; %ones(d,1);%zeros(d,1);%  
+    
+    lbd = 1e10;
+    for j=1:alpha_num
+        w_ind = max(1,ceil(w_list(j)*d));
+        w = zeros(d,1); w(w_ind)=1;
+        w_squared = norm(w)^2;        
+        alpha = alpha_list(j);
+        n = floor(alpha*d)
+        while sum(lbds./(lbds+lbd)) < n, lbd = lbd/(1+1e-3); end
+        while sum(lbds./(lbds+lbd)) > n, lbd = lbd*(1+1e-6); end
+        while sum(lbds./(lbds+lbd)) < n, lbd = lbd/(1+1e-9); end
+        sum(lbds./(lbds+lbd))
+        gamma = det((Sgm+lbd*eye(d))\Sgm);
+        theory = sg^2*sum(1./(lbds+lbd))*(1-gamma)/(d-n) + (d-n)*w'*inv(Sgm+lbd*eye(d))*w / sum(1./(lbds+lbd));
+        mse_est = zeros(T,1);
+        for t=1:T
+            X = randn(n,d)*B;
+            Xinv = pinv(X);
+            mse_est(t) = sg^2*trace(Xinv'*Xinv) + w_squared - w'*Xinv*X*w;
+        end
+        mse_std(i,j) = std(mse_est) / (sqrt(T) * theory);
+        mse_error(i,j) = (mean(mse_est) - theory) / theory;
+    end
+end
+    
+set(0,'DefaultAxesFontSize', 15);
+set(0,'DefaultTextFontSize', 15);
[…]

### 2022-08-23 design/code/bothbounds.m

@@ -1,0 +1,38 @@
+
+r = 30; 
+
+all_eps =  0.01:0.001:10;
+num_eps = length(all_eps);
+all_ub = zeros(num_eps,1); 
+all_lb = zeros(num_eps,1);
+all_k = zeros(num_eps,1);
+
+
+for i = 1:num_eps
+    eps = all_eps(i)
+    k = eps/(1+eps) * r;
+    all_k(i) = k; 
+    all_ub(i) = 1 + 3*eps;
+    all_lb(i) = 1 + eps * (k-1)/ (2*k*k);
+end   
+sel_k = all_k > 1;
+set(0,'DefaultAxesFontSize', 15);
+set(0,'DefaultTextFontSize', 15);
+figure('Color','white','Position', [500, 500, 500, 400])
+
+
+%    ph_list(end+1) = semilogx(ks,ratios,'LineWidth',1,'Color',colors{delta});
+%    legend_list{end+1} = sprintf('\\delta=%.3f',delta_list(delta));
+    hold on
+    %semilogx(1:d,theory_ratios,'--','LineWidth',1,'Color',colors{delta});    
+    %scatter(sr,0,20,colors{delta},'LineWidth',2,'MarkerEdgeAlpha',0.5);
+    plot (all_k(sel_k), all_lb(sel_k), '--r', all_k(sel_k), all_ub(sel_k), '--b');
+
+    x2 = [all_k(sel_k), fliplr(all_k(sel_k))];
+inBetween = [all_lb(sel_k), fliplr( all_ub(sel_k))];
+%patch(x2, inBetween, 'g');
+
+%legend(ph_list,legend_list,'Location','best');
+xlim([0,r])
+%title(decay_list{decay});
+hold off

### 2022-08-23 design/code/bound.m

@@ -1,0 +1,12 @@
+function val = bound( lbd , c)
+    if nargin == 1, c = sqrt(3); end
+    lbd = abs(lbd);
+    lbd = sort(lbd,'descend');
+    d = length(lbd);
+    r = sum(lbd) / lbd(1);
+    Er = sum(lbd(1:d>=r/c));
+    %if ceil(r/c) >= 2, Er = Er + (ceil(r/c)-(r/c))*lbd(ceil(r/c)-1); end
+    alpha = c^2 * Er / r;
+    k = sum(lbd./(lbd+alpha));
+    val = -k*c/r;
+    %val = - alpha*k/(c*Er); 

### 2022-08-23 design/code/convexity.m

@@ -1,0 +1,62 @@
+% nystrom symmetric polynomials
+
+n=51;
+T = 10;
+R = 7;
+sparsity = 1;
+dataset = 5;
+lower = [];
+
+if dataset == 1
+    X = randn(n,n);
+elseif dataset == 2
+    [U,~,V] = svd(randn(n));
+    l = ones(n,1);
+    l(floor(n/2)+1:n) = 0.01;
+    X = U*diag(l)*V';
+elseif dataset == 3
+    [U,~,V] = svd(randn(n));
+    l = ones(n,1);
+    l(21:n) = 0.01;
+    l(40:n) = 0.001;
+    X = U*diag(l)*V';   
+    %lower = [19,0.8*21;59,0.8*51]; 
+elseif dataset == 4
+    [U,~,V] = svd(randn(n));
+    l = 1./(1:n).^2;
+%    l(ceil(n/2):n) = 1/ceil(n/2);
+    X = U*diag(l)*V';    
+elseif dataset==5
+    l=[ones(10,1);.1*ones(2,1)];
+    l = l/sum(l);
+    X = diag(sqrt(l));
+    [n,~]=size(X);
+end
+
+k_list = 1:n-1;
+k_num = length(k_list);
+L = decompose_kernel(X*X');
+lbds = sort(L.D,'descend');
+theory = ones(k_num,1);
+poly = elem_sympoly(lbds,n);
+
+
+for i=k_list
+    k = k_list(i);
+    if k>=n, break; end
+    %Ek = sum(lbds(k+1:n));
+    theory(i) = ((k+1)/n)*poly(k+2,n+1)/poly(k+1,n+1);
+end
+
+
+% set(0,'DefaultAxesFontSize', 13);
+% set(0,'DefaultTextFontSize', 13);
+% figure('Color','white','Position', [1000, 1000, 500, 400])
+
+ph_list = [];
+legend_list = {};
+%colors = {[1.0, 0.5, 0], [1, 0, 0], [0, 0.5, 0],[0,0,0.5]};
+%types = {'-','--','-.',':'};
+
+ph_list(end+1) = plot(k_list,theory,'-','LineWidth',2,'Color',[1,0,0,0.5]);
+legend_list{end+1} = 'Expected error';

### 2022-08-23 design/code/descent.m

@@ -1,0 +1,114 @@
+n=100;
+
+d_list = unique(ceil(logspace(1,3,20)'));
+d_list = d_list(abs(d_list-n)>5);
+d_max = max(d_list);
+d_min = min(d_list);
+d_list_big = (d_min:d_max)';
+d_num = length(d_list);
+T = 20;
+show_experiments = false;
+show_norm = true;
+
+% B = randn(d,d);
+% [U,~,V] = svd(B,'econ');
+%Sgm = eye(d); %Sgm(1,1) = 0.0001; %diag(logspace(-5,0,d));%diag((1:d).^(-2)); %
+%B = sqrt(Sgm);
+%lbds = eig(Sgm);
+%w = [1;zeros(d-1,1)];
+%w_squared = norm(w)^2;
+
+
+
+num_opts = 3;
+sg = 1;
+w_norm = 1;
+logkappas = [0,2,5];
+mse_mean = zeros(d_num,num_opts);
+mse_error = zeros(d_num,num_opts);
+mse_count = 1;
+theory = zeros(length(d_list_big),num_opts);
+for i=1:length(d_list_big)
+    d = d_list_big(i);
+    for opt=1:num_opts
+        if opt <= 5
+            Sgm = diag(logspace(0,-logkappas(opt),d));
+        else
+            Sgm = diag(logspace(0,-1,d));
+        end
+        Sgm = Sgm * trace(inv(Sgm)) / d;
+        B = sqrt(Sgm);
+        lbds = eig(Sgm);
+        if opt>5
+            w = w_norm(opt-5)*ones(d,1)/sqrt(d); %w_norm * [zeros(d-1,1);1]; %w = w_norm * ones(d,1)/sqrt(d);
+        else
+            w = w_norm(1)*ones(d,1)/sqrt(d);
+        end
+%         ws(1,1) = w_norm(1);
+%         ws(ceil(d/2),2) = w_norm(2);
+%         ws(d,3) = w_norm(3);
+        if d <= n
+            theory(i,opt) = sg^2*sum(1./lbds);
+            if d < n, theory(i,opt) = theory(i,opt) * (1-exp(d-n))/(n-d); end
+        else
+            lbd = 1e10;
+            while sum(lbds./(lbds+lbd)) < n, lbd = lbd/1.1; end
+            while sum(lbds./(lbds+lbd)) > n, lbd = lbd*1.001; end
[…]

### 2022-08-23 design/code/dpp/examples/plane_samples.m

@@ -1,0 +1,33 @@
+%% compare dpp/poisson samples of points in the plane
+
+% config
+n = 60;      % grid dimension, N = n^2
+sigma = 0.1; % kernel width
+
+% choose a grid of points
+[x, y] = meshgrid((1:n)/n);
+
+% gaussian kernel
+L = exp(- (bsxfun(@minus,x(:),x(:)').^2 + ...
+           bsxfun(@minus,y(:),y(:)').^2) / sigma^2);
+
+% sample
+dpp_sample = sample_dpp(decompose_kernel(L));
+ind_sample = randsample(n*n,length(dpp_sample));
+  
+% plot
+subplot(1,2,1);
+plot(x(dpp_sample),y(dpp_sample),'b.');
+axis([0 1.02 0 1.02]);
+axis square;
+set(gca,'YTick',[]);
+set(gca,'XTick',[]);
+xlabel('DPP');
+
+subplot(1,2,2);
+plot(x(ind_sample),y(ind_sample),'r.');
+axis([0 1.02 0 1.02]);
+axis square;
+set(gca,'YTick',[]);
+set(gca,'XTick',[]);
+xlabel('Independent');

### 2022-08-23 design/code/dppsample.m

@@ -1,0 +1,4 @@
+n = 100;   %size
+m=50;  % number of dense
+
+% construct the matrix

### 2022-08-23 design/code/expectations.m

@@ -1,0 +1,37 @@
+d = 5;
+n = 5;
+X = randn(n,d);
+[~,D,V] = svd(X,'econ');
+D = diag(1:d);%eye(d);%diag(logspace(-1,0,d));
+X = U*D*V;
+
+Einv1 = 0;
+Einv2 = 0;
+Ek = 0;
+Zval = 0;
+
+Z = det(eye(d) + X'*X);
+
+for k=0:d
+    S = nchoosek(1:n,k);
+    for i=1:nchoosek(n,k)
+        XS = X(S(i,:),:);
+        PS = det(XS*XS')/Z;
+        Einv1 = Einv1 + PS * trace(pinv(XS'*XS)); 
+        Einv2 = Einv2 + PS * trace(X*pinv(XS'*XS)*X');
+        Ek = Ek + PS * k;
+        Zval = Zval + PS;
+    end
+end
+
+Zval
+
+Ek
+Thk = trace(X'*X*inv(eye(d)+X'*X))
+Diffk = Ek - Thk
+
+Einv1
+Th1 = trace(inv(eye(n)+X*X')) - (n-d)*det(X'*X)/Z
+Diff1 = Einv1 - Th1
+
+Einv2

### 2022-08-23 design/code/gaussian.m

@@ -1,0 +1,73 @@
+d=100;
+
+n_list = 1:1:d-2;
+n_num = length(n_list);
+r = 1;
+T = 1000;
+
+% B = randn(d,d);
+% [U,~,V] = svd(B,'econ');
+% D = diag((1:d).^(-2));%diag(logspace(-2,0,d));
+% B = U*D*V;
+% Sgm = B'*B;
+% lbds = eig(Sgm);
+% w = 10*randn(d,1);
+
+sg = 1;
+set(0,'DefaultAxesFontSize', 15);
+set(0,'DefaultTextFontSize', 15);
+figure('Color','white','Position', [500, 500, 500, 400])
+ph_list = [];
+legend_list = {};
+colors = [1,0.8,0];
+
+mse_std = zeros(n_num,1);
+theory = zeros(n_num,1);
+lbd = 1e10;
+for i=1:n_num
+    n = n_list(i);
+    while sum(lbds./(lbds+lbd)) < n, lbd = lbd/2; end
+    while sum(lbds./(lbds+lbd)) > n, lbd = lbd*1.01; end
+    gamma = det((Sgm+lbd*eye(d))\Sgm);
+    theory(i) = sg^2*sum(1./(lbds+lbd))/(d-n-1) + (d-n)*w'*inv(Sgm+lbd*eye(d))*w / sum(1./(lbds+lbd));
+    mse_list = zeros(T,1);
+    for t=1:T
+        X = randn(n,d)*B;
+        w_est = pinv(X)*(X*w+sg*randn(n,1));
+        mse_list(t) = norm(w_est-w)^2;
+    end
+    mse_mean(i) = mean(mse_list);
+    mse_std(i) = std(mse_list)/sqrt(T);
+%     rectangle('Position',[n_list(i)-1.5,theory(i)-mse_std(i),1,2*mse_std(i)],...
+%         'Curvature',[0.3 0.3],'FaceColor',[.9,.9,.9,0.3],'EdgeColor',[1,1,1,0]);
+%     if n>80
+%         mid_x = (ridge_x(i-1)+ridge_x(i))/2;
+%         mid_y = (ridge_y(i-1)+ridge_y(i))/2;
+%         rectangle('Position',[mid_x-mse_std(i),mid_y-mse_std(i),2*mse_std(i),2*mse_std(i)],...
+%             'Curvature',[0.3 0.3],'FaceColor',[.9,.9,.9,0.3],'EdgeColor',[1,1,1,0]);
+%         if n>90
[…]

### 2022-08-23 design/code/intro.m

@@ -1,0 +1,239 @@
+d=100;
+
+n_list = unique(ceil(linspace(5,2*d,25)'));
+n_list = n_list(abs(n_list-d)>5);
+n_max = max(n_list);
+n_min = min(n_list);
+n_list_big = (1:n_max)';
+n_num = length(n_list);
+T =50;
+%R=5;
+
+% B = randn(d,d);
+% [U,~,V] = svd(B,'econ');
+%Sgm = eye(d); %Sgm(1,1) = 0.0001; %diag(logspace(-5,0,d));%diag((1:d).^(-2)); %
+%B = sqrt(Sgm);
+%lbds = eig(Sgm);
+%w = [1;zeros(d-1,1)];
+%w_squared = norm(w)^2;
+
+rng(2);
+
+num_opts = 3;
+show = [true,true,true];
+show_norm = true;
+plot_type = 'mse';
+show_experiments = true;
+show_iid = false;
+sg = .4;
+w_norm = 1;
+logkappas = [0,2,5];
+mse_mean = zeros(n_num,num_opts);
+bias_mean = zeros(n_num,num_opts);
+shrinkage_mean = zeros(n_num,num_opts);
+mse_error = zeros(n_num,num_opts);
+bias_error = zeros(n_num,num_opts);
+shrinkage_error = zeros(n_num,num_opts);
+mse_count = 1;
+theory = zeros(length(n_list_big),num_opts);
+theory2 = zeros(length(n_list_big),num_opts);
+bias_theory = zeros(length(n_list_big),num_opts);
+shrinkage_theory = zeros(length(n_list_big),num_opts);
+lambda_theory = zeros(length(n_list_big),num_opts);
+for i=1:length(n_list_big)
+    n = n_list_big(i);
+    for opt=1:num_opts
+        Sgm = diag(logspace(0,-logkappas(opt),d));
+        Sgm = Sgm * trace(inv(Sgm)) / d;
+        lbds = eig(Sgm);
+        B = sqrt(Sgm);
+        w = w_norm(1)*ones(d,1)/sqrt(d);
+        if d <= n
+            theory(i,opt) = sg^2*sum(1./lbds);
+            bias_theory(i,opt) = 0;
+            shrinkage_theory(i,opt) = w_norm;
+            lambda_theory(i,opt) = 0;
+            if d < n-1
+                theory2(i,opt) = sg^2*sum(1./lbds)/(n-d-1); 
+            else
[…]

### 2022-08-23 design/code/lb.m

@@ -1,0 +1,76 @@
+d=100;
+m=20;
+decay = 3; %
+decay_list = {'exponential decay', 'one-step drop','mixed','linear decay','polynomial'};
+
+alpha_list = logspace(-10,1,100)';
+alpha_num = length(alpha_list);
+ratios = zeros(alpha_num,1);
+ks = zeros(alpha_num,1);
+delta_list = [0.01];
+theory_ratios = zeros(d,1);
+
+
+set(0,'DefaultAxesFontSize', 15);
+set(0,'DefaultTextFontSize', 15);
+figure('Color','white','Position', [500, 500, 500, 400])
+ph_list = [];
+legend_list = {};
+colors = {[0 0 1],[1 0 0],[0 0.8 0]};
+for delta=1:length(delta_list)
+    if decay==1        
+        lbd = 0.9.^(1:d)';
+    elseif decay==2
+        lbd = [ones(1,m),delta_list(delta)*ones(1,d-m)]';
+    elseif decay==3
+        lbd = [ones(1,m),0.9.^(m+1:2*m),0.1*0.9^(2*m)*ones(1,d-2*m)]';
+    elseif decay==4
+        m=20;
+        lbd = [ones(1,m),1e-1*ones(1,m),1e-2*ones(1,m),5e-4*ones(1,2*m)]';
+    elseif decay==5
+        lbd = (1:d).^(-1/4);
+    end
+    sr_list = [];
+    next = 1;
+    while next < d        
+        sr_list(end+1) = next-1 + sum(lbd(next:end))/lbd(next);
+        next = ceil(sr_list(end));
+    end
+    for i=1:alpha_num
+        alpha  = alpha_list(i);
+        ks(i) = sum(lbd./(lbd+alpha));
+        kceil = ceil(ks(i));
+        Ek = sum(lbd(kceil:end));
+        if kceil > 1
+             Ek = Ek + (kceil-ks(i))*lbd(kceil-1);
+        end        
+        ratios(i) = alpha*ks(i)/Ek;
+    end
+    for k=1:floor(sr)-1
+        theory_ratios(k) = 1 + k/(sr-k);
+    end
+    for k=floor(sr):d
+        theory_ratios(k)=k+1;
+    end
+
+    ph_list(end+1) = plot(ks,ratios,'LineWidth',1,'Color',colors{delta});
+    yylim = max(ratios);
[…]

### 2022-08-23 design/code/mse.m

@@ -1,0 +1,60 @@
+d=50;
+
+n_list = ceil(0.9*d);
+n_num = length(n_list);
+T = 10000;
+
+% B = randn(d,d);
+% [U,~,V] = svd(B,'econ');
+Sgm = eye(d); Sgm(1,1) = 0.0001; %diag(logspace(-5,0,d));%diag((1:d).^(-2)); %
+B = sqrt(Sgm);
+lbds = eig(Sgm);
+w = [1;zeros(d-1,1)];
+w_squared = norm(w)^2;
+sg = 0;
+
+set(0,'DefaultAxesFontSize', 15);
+set(0,'DefaultTextFontSize', 15);
+figure('Color','white','Position', [500, 500, 500, 400])
+ph_list = [];
+legend_list = {};
+colors = {[1.0, 0.5, 0], [1, 0, 0], [0, 0, 0]};
+
+mse_mean = zeros(n_num,T);
+theory = zeros(n_num,1);
+lbd = 1e10;
+for i=1:n_num
+    n = n_list(i)
+    while sum(lbds./(lbds+lbd)) < n, lbd = lbd/1.1; end
+    while sum(lbds./(lbds+lbd)) > n, lbd = lbd*1.001; end
+    while sum(lbds./(lbds+lbd)) < n, lbd = lbd/1.00001; end
+    sum(lbds./(lbds+lbd))-n
+    gamma = det((Sgm+lbd*eye(d))\Sgm);
+    theory(i) = sg^2*sum(1./(lbds+lbd))*(1-gamma)/(d-n) + (d-n)*w'*inv(Sgm+lbd*eye(d))*w / sum(1./(lbds+lbd));
+    mse_list = zeros(T,1);
+    error_list = zeros(T,1);
+    for t=1:T
+        X = randn(n,d)*B;
+        Xinv = pinv(X);
+        mse_list(t) = sg^2*trace(Xinv'*Xinv) + w_squared - w'*Xinv*X*w;
+%         w_est = pinv(X)*(X*w+sg*randn(n,1));
+%         mse_list(t) = norm(w_est-w)^2;
+        if t==1 
+            mse_mean(i,t) = mse_list(t); 
+        else
+            mse_mean(i,t) = mse_mean(i,t-1) + mse_list(t);
+        end
+        error_list(t) = abs(mse_mean(i,t)/t - theory(i))/theory(i);
+    end
+    
[…]

### 2022-08-23 design/code/nystrom.m

@@ -1,0 +1,223 @@
+% Nystrom plots
+
+% bodyfat: sgm = 2000,1000,500
+% eunite2001: sgm = 10,5,2.5,1
+n=200;
+T = 100;
+R = 7;
+sparsity = 1;
+experiments = false;
+from_file = false;
+dataset = 2;
+data_dir = '../../tail/data'; %'../../Microsoft/big_data'; %
+expand_dims=0;
+kernelize = 0;
+data_files = {'bodyfat','eunite2001','mg','triazines','housing','space_ga','abalone','mpg',...
+              'MSD','a9a/a9a.t','covtype/bet_stages/covtype.t',...
+              'w8a/w8a.t','mushrooms.txt','phishing.txt',...
+              'madelon.t','splice_scale.txt','sonar_scale.txt'};%'rcv1/rcv1_train.binary',
+% data_files = {'cadata','MSD',...
+%     'housing','space_ga','cpusmall','cpusmall_scale','abalone',...
+%     'abalone_scale','bodyfat','bodyfat_scale','mg','mg_scale',...
+%     'housing_scale','triazines','pyrim','mpg','mpg_scale'};
+lower = [];
+
+if from_file
+    [~, X] = libsvmread(sprintf('%s/%s',data_dir,data_files{dataset}));
+    [n,d] = size(X);
+    if expand_dims        
+        Xe = zeros(d^2,n);
+        for i=1:n
+            A = X(i,:)'*X(i,:);
+            Xe(:,i) = A(:);
+        end
+        X = Xe'./svds(Xe,1);
+    end
+    [n,d] = size(X);
+    if kernelize
+        sgm = 250;
+        C = 3*sgm;
+        eta = ceil((d+1)/2);
+        K=zeros(n,n);
+        for i=1:n
+            for j=1:n
+                ll = norm(X(i,:)-X(j,:));
+                K(i,j) =  exp(-(ll/sgm)^2);%max((1-ll/C)^eta,0) *                 
+            end
+        end
+        X = chol(K)';
+    end
+
+else
+    if dataset == 1
+        X = randn(n,n);
+    elseif dataset == 2
+        [U,~,V] = svd(randn(n));
+        l = ones(n,1);
+        l(30:n) = 0.01;
+        X = U*diag(l)*V';
+    elseif dataset == 3
+        [U,~,V] = svd(randn(n));
+        l = ones(n,1);
[…]

### 2022-08-23 design/code/optimize.m

@@ -1,0 +1,11 @@
+d=10;
+options = optimset('PlotFcns',@optimplotfval,'MaxIter',5000);
+eps = 0.01;
+
+lbd = sort(abs(randn(1,d)))
+
+for i=1:2
+    [lbd,val] = fminsearch(@bound,lbd,options);
+    i, -val %, sort(abs(lbd),'descend')
+    lbd = lbd + eps*randn(1,d);
+end

### 2022-08-23 design/code/projectors.m

@@ -1,0 +1,90 @@
+d_list = [20,25,30,35,40];
+alpha_list = 0.2;
+d_num = length(d_list);
+alpha_num = length(alpha_list);
+T = 40000;
+R = 4;
+proj_error = zeros(d_num,alpha_num);
+
+proj_std = zeros(d_num,alpha_num);
+
+
+for i=1:d_num
+    d = d_list(i);
+    Sgm = eye(d); Sgm(1,1)=100000000;%diag((d:-1:1).^(-2));% eye(d);%=diag(logspace(-2,0,d));%
+    B = sqrt(Sgm);
+    lbds = eig(Sgm);
+    w = zeros(d,1);
+    w(1) = 1;
+    
+    lbd = 1e10;
+    for j=1:alpha_num
+        alpha = alpha_list(j);
+        n = floor(alpha*d);
+        fprintf('d=%d, n=%d\n',d,n);
+        while sum(lbds./(lbds+lbd)) < n, lbd = lbd/(1+1e-3); end
+        while sum(lbds./(lbds+lbd)) > n, lbd = lbd*(1+1e-6); end
+        while sum(lbds./(lbds+lbd)) < n, lbd = lbd/(1+1e-9); end
+        %sum(lbds./(lbds+lbd))
+        SqSgmlbd = sqrt(eye(d) + Sgm/lbd);    
+        theory = w'*inv(eye(d) + Sgm/lbd)*w;
+        proj_global = 0;
+        %proj_global = zeros(d,d);
+        cnt = 0;
+        TT = T;
+        for r=1:R
+            fprintf('...sampling T=%d...\n',TT);
+            btstrp_l = randsample(TT,TT,true); 
+            btstrp_c = zeros(TT,1);
+            for t=1:TT
+                btstrp_c(btstrp_l(t)) = btstrp_c(btstrp_l(t)) + 1;
+            end
+            proj_mean = 0;
+            btstrp_mean = 0;
+            %proj_mean = zeros(d,d);
+            %btstrp_mean = zeros(d,d);
+            for t=1:TT
+                cnt = cnt + 1;
+                X = randn(n,d)*B;                
+                M = eye(d) - pinv(X)*X;
+                Bias = w'*M*w;
+                proj_mean = proj_mean + Bias; %M
+                proj_global = proj_global + Bias; %M
+                btstrp_mean = btstrp_mean + btstrp_c(t) * Bias; %M
+            end
+            btstrp_mean = btstrp_mean / TT;
[…]

### 2022-08-23 design/code/rbfexperiment.m

@@ -1,0 +1,151 @@
+% Nystrom plots
+
+% toy: sgm = 10,5,1
+% bodyfat: sgm = 2000,1000,200
+% eunite2001: sgm = 10,5,1 or 8,4,1 ??
+T = 1000;
+R = 2;
+sparsity = 1;
+dataset = 2;
+method = 1;
+data_dir = '../../tail/data'; %'../../Microsoft/big_data'; %
+kernelize = 1;
+data_files = {'bodyfat','eunite2001','mg','triazines','housing','space_ga','abalone','mpg',...
+              'MSD','a9a/a9a.t','covtype/bet_stages/covtype.t',...
+              'w8a/w8a.t','mushrooms.txt','phishing.txt',...
+              'madelon.t','splice_scale.txt','sonar_scale.txt'};
+method_names = {'k-DPP','greedy'};
+k_max = 40;
+k_list = 1:k_max;
+k_num = length(k_list);
+sgm_list =  [8,4,1]; %[1400,500,200];%
+sgm_num = length(sgm_list);
+ratio_mean = ones(sgm_num,k_num);
+ratio_std = zeros(sgm_num,k_num);
+lbd_lists = zeros(sgm_num,k_num);
+
+for sgm_i = 1:sgm_num
+    sgm = sgm_list(sgm_i);
+    if dataset==0
+        n=50;
+        d=50;
+        [U,~,V] = svd(randn(n,d));
+        l = ones(n,1);
+        l(21:n) = 1/sgm;
+        X = U*diag(l)*V';
+    else
+        [~, X] = libsvmread(sprintf('%s/%s',data_dir,data_files{dataset}));
+        [n,d] = size(X);
+    end
+    if kernelize
+        eta = ceil((d+1)/2);
+        K=zeros(n,n);
+        for i=1:n
+            for j=1:n
+                ll = norm(X(i,:)-X(j,:));
+                K(i,j) =  exp(-ll^2/(2*sgm^2));
+            end
+        end
+        X = chol(K)';
+    end
+
+    L = decompose_kernel(full(X*X'));
+    lbds = sort(L.D,'descend');
+    lbd_lists(sgm_i,:) = lbds(k_list);
+
+    for i=k_list
+        k = k_list(i);
+        if k>=n, break; end
+        Ek = sum(lbds(k+1:n));
+        if method == 1
+            for r=1:R
+                [k,r]
+                T_temp = T*2^r;
+                ratio_list = zeros(T_temp,1);
[…]

### 2022-08-23 design/code/ridge.m

@@ -1,0 +1,89 @@
+d=100;
+
+n_list = 1:1:d;
+n_num = length(n_list);
+r = 1;
+T = 1000;
+
+% B = randn(d,d);
+% [U,~,V] = svd(B,'econ');
+% D = diag(logspace(-2,0,d));
+% B = U*D*V;
+% Sgm = B'*B;
+% lbds = eig(Sgm);
+% w = randn(d,1);
+
+set(0,'DefaultAxesFontSize', 15);
+set(0,'DefaultTextFontSize', 15);
+figure('Color','white','Position', [500, 500, 500, 400])
+ph_list = [];
+legend_list = {};
+
+x_max = 0;
+w_mean = zeros(2,n_num);
+w_std = zeros(n_num,1);
+ridge_x = zeros(n_num,1);
+ridge_y = zeros(n_num,1);
+colors = [1,0.8,0];
+lbd = 1e10;
+for i=1:n_num
+    n = n_list(i);
+    while sum(lbds./(lbds+lbd)) < n, lbd = lbd/2; end
+    while sum(lbds./(lbds+lbd)) > n, lbd = lbd*1.01; end
+    w_ridge = (Sgm+lbd*eye(d))\Sgm * w;
+
+    for j=1:r   
+        x_list = zeros(T,1);
+        y_list = zeros(T,1);
+        for t=1:T
+            X = randn(n,d)*B;
+            w_est = pinv(X)*X*w;
+            x_list(t) = w_est(1);
+            y_list(t) = w_est(2);
+            %w_min(:,j) = w_min(:,j) + pinv(X)*X*w;           
+        end
+        w_mean(1,i) = mean(x_list);
+        w_mean(2,i) = mean(y_list);
+        w_std(i) = (std(x_list)+std(y_list))/sqrt(T);
+        %w_min(:,j) = w_min(:,j)/T;
+        x_max = max(x_max, abs(w_mean(1,i)));
+        x_max = max(x_max, abs(w_mean(2,i)));
+    end
+    n_eff = sum(lbds./(lbds+lbd));
+    ridge_x(i) = w_ridge(1);
+    ridge_y(i) = w_ridge(2);
+    %ph_list(end+1) = plot(w_ridge(1),w_ridge(2),40,colors*n_eff/d,'filled');
+    %legend_list{end+1} = sprintf('n=%d',round(n_eff));
+    %hold on
[…]

### 2022-08-23 design/equivalents_code/aws/build_and_push.sh

@@ -1,0 +1,39 @@
+#!/bin/bash
+
+#TAG=$(git rev-parse HEAD)
+TAG=latest
+
+REPO=equivalents
+
+# docker build -t $REPO:$TAG .
+# docker tag $REPO:$TAG 366424314563.dkr.ecr.us-east-1.amazonaws.com/$REPO:$TAG
+# docker push 366424314563.dkr.ecr.us-east-1.amazonaws.com/$REPO:$TAG
+
+CONTAINER_PROPERTIES=`cat << EOF
+{
+	"image": "366424314563.dkr.ecr.us-east-1.amazonaws.com/$REPO:$TAG",
+	"vcpus": 1,
+	"memory": 1024,
+	"environment": [
+		{
+			"name": "D_MAX",
+			"value": "100"
+		}
+	],
+	"jobRoleArn": "arn:aws:iam::366424314563:role/double-descent-batch-s3"
+}
+EOF`
+CMD="aws batch register-job-definition \
+	--job-definition-name $REPO \
+	--type container \
+	--container-properties '$CONTAINER_PROPERTIES'"
+echo $CMD
+RESULT=$(eval $CMD)
+JOB_ARN=$(echo $RESULT | jq '.jobDefinitionArn' -r)
+
+SIZE=$(python run_experiment.py)
+aws batch submit-job \
+	--job-name $REPO \
+	--job-queue default \
+	--array-properties size=$SIZE \
+	--job-definition $JOB_ARN

### 2022-08-23 design/equivalents_code/aws/run_experiment.py

@@ -1,0 +1,78 @@
+import boto3
+import collections
+from datetime import datetime
+import itertools
+import numpy as np
+import os
+import pandas as pd
+import pickle
+import scipy.linalg
+import scipy.optimize
+import scipy.stats
+import sys
+
+BUCKET_NAME = 'equivalents'
+
+def _solve_lambda_n(Sigma, n, tol=1e-10):
+    if np.isclose(n, np.linalg.matrix_rank(Sigma)):
+        return 0
+    else:
+        _, S, _ = scipy.linalg.svd(Sigma, full_matrices=False)
+        exp_subset_size_func = lambda lam: np.sum(S / (S + lam))
+
+        opt_result = scipy.optimize.minimize_scalar(
+            lambda lam: (exp_subset_size_func(lam) - n)**2,
+            method='bounded',
+            bounds=(0, d * max(S) / n),
+            tol=tol,
+        )
+        return opt_result.x
+
+def run_experiment(param):
+    return {
+        'param': param,
+    }
+
+def get_current_results():
+    s3 = boto3.resource('s3')
+    data = []
+    for obj in s3.Bucket(BUCKET_NAME).objects.all():
+        result = pickle.loads(obj.get()['Body'].read())
+        data.append(result)
+    return pd.DataFrame(data)
+
+if __name__ == '__main__':
+    start = datetime.now()
+    rng = np.random.RandomState(42)
+
+    param_grid = [0, 1, 2, 3]
+
+    if os.environ.get('AWS_BATCH_JOB_ARRAY_INDEX') is None:
+        # local, exit early printing size, see `build_and_push.sh`
+        print(len(param_grid))
+        sys.exit(0)
+    print(os.environ.get('AWS_BATCH_JOB_ARRAY_INDEX'))
+
+    param = param_grid[int(os.environ.get('AWS_BATCH_JOB_ARRAY_INDEX'))]
+
+    current_results = get_current_results()
+    try:
+        previous_matching_runs = (
+            (current_results['param'] == param)
+        ).sum()
+        if previous_matching_runs != 0:
+            print("Previous run found, exiting")
+    except:
+        pass
+
+    print("No previous runs, executing")
+
+    result = run_experiment(param)
+    print(result)
+
+    result_filename = f"result-param={param}.pkl"
+    with open(result_filename, "wb") as f:
+        pickle.dump(result, f)
+
+    s3 = boto3.client('s3')
+    s3.upload_file(result_filename, BUCKET_NAME, result_filename)

### 2022-08-23 ic/gaussian_unknown_mean.py

@@ -1,0 +1,47 @@
+import pyprob
+from pyprob import Model
+from pyprob.distributions import Normal
+
+import torch
+import numpy as np
+import math
+import matplotlib.pyplot as plt
+
+class GaussianUnknownMean(Model):
+    def __init__(self):
+        super().__init__(name='Gaussian with unknown mean') # give the model a name
+        self.prior_mean = 1
+        self.prior_std = math.sqrt(5)
+        self.likelihood_std = math.sqrt(2)
+
+    def forward(self): # Needed to specifcy how the generative model is run forward
+        # sample the (latent) mean variable to be inferred:
+        mu = pyprob.sample(Normal(self.prior_mean, self.prior_std)) # NOTE: sample -> denotes latent variables
+
+        # define the likelihood
+        likelihood = Normal(mu, self.likelihood_std)
+
+        # Lets add two observed variables
+        # -> the 'name' argument is used later to assignment values:
+        pyprob.observe(likelihood, name='obs0') # NOTE: observe -> denotes observable variables
+        pyprob.observe(likelihood, name='obs1')
+
+        # return the latent quantity of interest
+        return mu
+
+model = GaussianUnknownMean()
+model.learn_inference_network(num_traces=20000,
+        observe_embeddings={'obs0' : {'dim' : 32},
+            'obs1': {'dim' : 32}},
+        inference_network=pyprob.InferenceNetwork.LSTM)
+
+observed_list = [12, 10] # Observations
+# sample from posterior (200 samples)
+posterior = model.posterior_results(
+        num_traces=200, # the number of samples estimating the posterior
+        inference_engine=pyprob.InferenceEngine.IMPORTANCE_SAMPLING_WITH_INFERENCE_NETWORK, # specify which inference engine to use
+        observe={
+            'obs0': observed_list[0],
+            'obs1': observed_list[1]} # assign values to the observed values
+        )
+print(posterior.expectation(lambda x: x))

### 2022-08-23 ic/gmm_model.py

@@ -1,0 +1,64 @@
+import pyprob
+from pyprob import Model
+from pyprob.distributions import Normal
+
+import numpy as np
+import torch
+from pyprob.distributions import Gamma, Mixture
+
+class GMM(Model):
+    def __init__(self, K=2):
+        super().__init__(name='Gaussian mixture model')
+        self.d = 2
+        self.K = K
+
+    def forward(self):
+        muxs = list()
+        muys = list()
+        probs = [1./self.K for _ in range(self.K)]
+        for _ in range(self.K):
+            muxs.append(pyprob.sample(Normal(0, 1)))
+            muys.append(pyprob.sample(Normal(0, 1)))
+
+        # define the likelihood factorized over coordinates
+        likelihoodx = Mixture(
+                [Normal(muxs[i], 0.1) for i in range(self.K)],
+                probs=probs)
+        likelihoody = Mixture(
+                [Normal(muys[i], 0.1) for i in range(self.K)],
+                probs=probs)
+
+        pyprob.observe(likelihoodx, name='obs0x')
+        pyprob.observe(likelihoody, name='obs0y')
+        pyprob.observe(likelihoodx, name='obs1x')
+        pyprob.observe(likelihoody, name='obs1y')
+
+        return torch.tensor([muxs, muys]).T
+
+
+if __name__ == '__main__':
+    model = GMM()
+    # prior = model.prior_results(num_traces=10)
+    # print(prior.get_values())
+    model.learn_inference_network(
+            num_traces=20000,
+            observe_embeddings={
+                'obs0x' : {'dim' : 16},
+                'obs0y' : {'dim' : 16},
+                'obs1x' : {'dim' : 16},
+                'obs1y' : {'dim' : 16},
+            },
+            inference_network=pyprob.InferenceNetwork.LSTM)
+
+    observed_list = [100, 100, -100, -100] # Observations
+    # sample from posterior (200 samples)
+    posterior = model.posterior_results(
+            num_traces=200, # the number of samples estimating the posterior
+            inference_engine=pyprob.InferenceEngine.IMPORTANCE_SAMPLING_WITH_INFERENCE_NETWORK, # specify which inference engine to use
+            observe={
+                'obs0x': observed_list[0],
+                'obs0y': observed_list[1],
+                'obs1x': observed_list[2],
+                'obs1y': observed_list[3]} # assign values to the observed values
+            )
+    print(posterior.expectation(lambda x: x))

### 2022-08-23 design/aistats/bayesian.tex
Point chapter links

@@
 \begin{document}
-\ifisarxiv
-  \title{Bayesian experimental design using regularized \\
-    determinantal point processes}
-  \author{%
-          Micha{\l } Derezi\'{n}ski \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mderezin@berkeley.edu}\\
-  \And
-  Feynman Liang \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{feynman@berkeley.edu}
-  \And
-   Michael W. Mahoney\\
-  ICSI and Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mmahoney@stat.berkeley.edu}
-    % examples of more authors
+% \ifisarxiv
+%   \title{Bayesian experimental design using regularized \\
+%     determinantal point processes}
+%   \author{%
+%           Micha{\l } Derezi\'{n}ski \\
+%   Department of Statistics\\
+%   University of California, Berkeley\\
+%   \texttt{mderezin@berkeley.edu}\\
 %   \And
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
-    % \AND
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
+%   Feynman Liang \\
+%   Department of Statistics\\
+%   University of California, Berkeley\\
+%   \texttt{feynman@berkeley.edu}
 %   \And
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
+%   Michael W. Mahoney\\
+%   ICSI and Department of Statistics\\
+%   University of California, Berkeley\\
+%   \texttt{mmahoney@stat.berkeley.edu}
+%     % examples of more authors
+%     % \And
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%     % \AND
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%     % \And
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%     % \And
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%   }
+%   \else
+
+%   \runningtitle{Bayesian experimental design using regularized DPPs}
+%   \twocolumn[
+%   \aistatstitle{Bayesian experimental design \\using regularized 
+%     determinantal point processes}
+%   \aistatsauthor{%
+%     Micha{\l } Derezi\'{n}ski
 %     \And
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
-  }
-  \else
-
-  \runningtitle{Bayesian experimental design using regularized DPPs}
-  \twocolumn[
-  \aistatstitle{Bayesian experimental design \\using regularized 
-    determinantal point processes}
-  \aistatsauthor{%
-    Micha{\l } Derezi\'{n}ski
-    \And
-    Feynman Liang
-    \And
-    Michael W. Mahoney
-  }
-  \aistatsaddress {
-    Department of Statistics\\
-    University of California, Berkeley\\
-    \texttt{mderezin@berkeley.edu}
-    \And
-    Department of Statistics\\
-    University of California, Berkeley\\
-    \texttt{feynman@berkeley.edu}
-    \And
-    ICSI and Department of Statistics\\
-    University of California, Berkeley\\
-    \texttt{mmahoney@stat.berkeley.edu}
-  } 
-% \aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
-% \aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }
-  ]
-\fi
+%     Feynman Liang
+%     \And
+%     Michael W. Mahoney
+%   }
+%   \aistatsaddress {
+%     Department of Statistics\\
+%     University of California, Berkeley\\
+%     \texttt{mderezin@berkeley.edu}
+%     \And
+%     Department of Statistics\\
+%     University of California, Berkeley\\
+%     \texttt{feynman@berkeley.edu}
+%     \And
+%     ICSI and Department of Statistics\\
+%     University of California, Berkeley\\
+%     \texttt{mmahoney@stat.berkeley.edu}
+%   } 
+% % \aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+% % \aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }
+%   ]
+% \fi
 

### 2022-08-23 ic/supplement.tex
Point chapter links

--- 2022-08-23 ic/aistats2021/sample_paper.tex
@@
 
-\twocolumn[
+% Supplementary material: To improve readability, you must use a single-column format for the supplementary material.
+\onecolumn
+\aistatstitle{Instructions for Paper Submissions to AISTATS 2021: \\ 
+Supplementary Materials}
 
-\aistatstitle{Instructions for Paper Submissions to AISTATS 2021}
+\section{FORMATTING INSTRUCTIONS}
 
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
-
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]
-
-\begin{abstract}
-  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
-  both left and right-hand margins. Use 10~point type, with a vertical
-  spacing of 11~points. The \textbf{Abstract} heading must be centered,
-  bold, and in point size 12. Two line spaces precede the
-  Abstract. The Abstract must be limited to one paragraph.
-\end{abstract}
-
-\section{GENERAL FORMATTING INSTRUCTIONS}
-
-The camera-ready versions of the accepted papers are 8 pages,
-plus any additional pages needed for references.
-
-Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
-Each column is 3.25~inches wide (19.5~picas).  The space
-between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
-Use 10~point type with a vertical spacing of
-11~points. Please use US Letter size paper instead of A4.
-
-Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
-Top rule is 4~points thick and bottom rule is 1~point thick.
-Allow 1/4~inch space above and below title to rules.
-
-Author descriptions are center-justified, initial caps.  The lead
-author is to be listed first (left-most), and the Co-authors are set
-to follow.  If up to three authors, use a single row of author
-descriptions, each one center-justified, and all set side by side;
-with more authors or unusually long names or institutions, use more
-rows.
-
-Use one-half line space between paragraphs, with no indent.
-
-\section{FIRST LEVEL HEADINGS}
-
-First level headings are all caps, flush left, bold, and in point size
-12. Use one line space before the first level heading and one-half line space
-after the first level heading.
-
-\subsection{Second Level Heading}
-
-Second level headings are initial caps, flush left, bold, and in point
-size 10. Use one line space before the second level heading and one-half line
-space after the second level heading.
-
-\subsubsection{Third Level Heading}
-
-Third level headings are flush left, initial caps, bold, and in point
-size 10. Use one line space before the third level heading and one-half line
-space after the third level heading.
-
-\paragraph{Fourth Level Heading}
-
-Fourth level headings must be flush left, initial caps, bold, and
-Roman type.  Use one line space before the fourth level heading, and
-place the section text immediately after the heading with no line
-break, but an 11 point horizontal space.
-
-%%%
-\subsection{Citations, Figure, References}
-
-\subsubsection{Citations in Text}
-
-Citations within the text should include the author's last name and
-year, e.g., (Cheesman, 1985). 
-%Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
-Be sure that the sentence reads
-correctly if the citation is deleted: e.g., instead of ``As described
-by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
-described by Cheesman (1985), we first frobulate the widgets.''
-
-The references listed at the end of the paper can follow any style as long as it is used consistently.
-
-%Be sure to avoid
-%accidentally disclosing author identities through citations.
-
-\subsubsection{Footnotes}
-
-Indicate footnotes with a number\footnote{Sample of the first
-  footnote.} in the text. Use 8 point type for footnotes. Place the
-footnotes at the bottom of the column in which their markers appear,
-continuing to the next column if required. Precede the footnote
-section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
-long.\footnote{Sample of the second footnote.}
-
-\subsubsection{Figures}
-
-All artwork must be centered, neat, clean, and legible.  All lines
-should be very dark for purposes of reproduction, and art work should
-not be hand-drawn.  Figures may appear at the top of a column, at the
-top of a page spanning multiple columns, inline within a column, or
-with text wrapped around them, but the figure number and caption
-always appear immediately below the figure.  Leave 2 line spaces
-between the figure and the caption. The figure caption is initial caps
-and each figure should be numbered consecutively.
-
-Make sure that the figure caption does not get separated from the
-figure. Leave extra white space at the bottom of the page rather than
-splitting the figure and figure caption.
-\begin{figure}[h]
-\vspace{.3in}
-\centerline{\fbox{This figure intentionally left non-blank}}
-\vspace{.3in}
-\caption{Sample Figure Caption}
-\end{figure}
-
-\subsubsection{Tables}
-
-All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
-Table number and title always appear above the table.
-See Table~\ref{sample-table}.
-
-Use one line space before the table title, one line space after the table title,
-and one line space after the table. The table title must be
-initial caps and each table numbered consecutively.
-
-\begin{table}[h]
-\caption{Sample Table Title} \label{sample-table}
-\begin{center}
-\begin{tabular}{ll}
-\textbf{PART}  &\textbf{DESCRIPTION} \\
-\hline \\
-Dendrite         &Input terminal \\
-Axon             &Output terminal \\
-Soma             &Cell body (contains cell nucleus) \\
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{SUPPLEMENTARY MATERIAL}
-
-If you need to include additional appendices during submission, you can include them in the supplementary material file.
-You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
-Note that reviewers are under no obligation to examine your supplementary material. 
-If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.
-
-You must use \texttt{aistats2021.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
-The only difference is that it must be in a \emph{single-column} format.
+To prepare a supplementary pdf file, we ask the authors to use \texttt{aistats2021.sty} as a style file and to follow the same formatting instructions as in the main paper. 
+The only difference is that the supplementary material must be in a \emph{single-column} format.
 You can use \texttt{supplement.tex} in our starter pack as a starting point.
 
-\section{SUBMISSION INSTRUCTIONS}
+Note that reviewers are under no obligation to examine your supplementary material.
 
-To submit your paper to AISTATS 2021, please follow these instructions.
+\section{MISSING PROOFS}
 
-\begin{enumerate}
-    \item Download \texttt{aistats2021.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
-    Please, do not modify the style files as this might result in a formatting violation.
+The supplementary materials may contain detailed proofs of the results that are missing in the main paper.
 
-    \item Use \texttt{sample\_paper.tex} as a starting point.
-    \item Begin your document with
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage\{aistats2021\}}
-    \end{flushleft}
-    The \texttt{twoside} option for the class article allows the
-    package \texttt{fancyhdr.sty} to include headings for even and odd
-    numbered pages.
-    \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
-    \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
-    \item Upload the PDF file along with other supplementary material files to the CMT website.
-\end{enumerate}
+\subsection{Proof of Lemma 3}
 
-\subsection{Camera-ready Papers}
+\textit{In this section, we present the detailed proof of Lemma 3 and then [ ... ]}
 
-%For the camera-ready paper, if you are using \LaTeX, please make sure
-%that you follow these instructions.  
-% (If you are not using \LaTeX,
-%please make sure to achieve the same effect using your chosen
-%typesetting package.)
+\section{ADDITIONAL EXPERIMENTS}
 
-If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
-\begin{enumerate}
-    %\item Download \texttt{fancyhdr.sty} -- the
-    %\texttt{aistats2021.sty} file will make use of it.
-    \item Change the beginning of your document to
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage[accepted]\{aistats2021\}}
-    \end{flushleft}
-    The option \texttt{accepted} for the package
-    \texttt{aistats2021.sty} will write a copyright notice at the end of
-    the first column of the first page. This option will also print
-    headings for the paper.  For the \emph{even} pages, the title of
-    the paper will be used as heading and for \emph{odd} pages the
-    author names will be used as heading.  If the title of the paper
-    is too long or the number of authors is too large, the style will
-    print a warning message as heading. If this happens additional
-    commands can be used to place as headings shorter versions of the
-    title and the author names. This is explained in the next point.
-    \item  If you get warning messages as described above, then
-    immediately after $\texttt{\textbackslash
-    begin\{document\}}$, write
-    \begin{flushleft}
-    \texttt{\textbackslash runningtitle\{Provide here an alternative
-    shorter version of the title of your paper\}}\\
-    \texttt{\textbackslash runningauthor\{Provide here the surnames of
-    the authors of your paper, all separated by commas\}}
-    \end{flushleft}
-    Note that the text that appears as argument in \texttt{\textbackslash
-      runningtitle} will be printed as a heading in the \emph{even}
-    pages. The text that appears as argument in \texttt{\textbackslash
-      runningauthor} will be printed as a heading in the \emph{odd}
-    pages.  If even the author surnames do not fit, it is acceptable
-    to give a subset of author names followed by ``et al.''
+If you have additional experimental results, you may include them in the supplementary materials.
 
-    %\item Use the file sample\_paper.tex as an example.
+\subsection{The Effect of Regularization Parameter}
 
-    \item The camera-ready versions of the accepted papers are 8
-      pages, plus any additional pages needed for references.
+\textit{Our algorithm depends on the regularization parameter $\lambda$. Figure 1 below illustrates the effect of this parameter on the performance of our algorithm. As we can see, [ ... ]}
 
-    \item If you need to include additional appendices,
-      you can include them in the supplementary
-      material file.
-
-    \item Please, do not change the layout given by the above
-      instructions and by the style file.
-
-\end{enumerate}
-
-\subsubsection*{Acknowledgements}
-All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
-To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
-
-\subsubsection*{References}
-
-References follow the acknowledgements.  Use an unnumbered third level
-heading for the references section.  Please use the same font
-size for references as for the body of the paper---remember that
-references do not count against your page length total.
-
-\begin{thebibliography}{}
-\setlength{\itemindent}{-\leftmargin}
-\makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
-\bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
-    \newblock Performance of a stochastic learning microchip.
-    \newblock In D. S. Touretzky (ed.),
-    \textit{Advances in Neural Information Processing Systems 1}, 748--760.
-    San Mateo, Calif.: Morgan Kaufmann.
-
-\bibitem{} F.~Rosenblatt (1962).
-    \newblock \textit{Principles of Neurodynamics.}
-    \newblock Washington, D.C.: Spartan Books.
-
-\bibitem{} G.~Tesauro (1989).
-    \newblock Neurogammon wins computer Olympiad.
-    \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
-\end{thebibliography}
+\vfill
 

### 2022-08-23 ic/gaussian_unknown_mean.py

@@ -1,47 +1,0 @@
-import pyprob
-from pyprob import Model
-from pyprob.distributions import Normal
-
-import torch
-import numpy as np
-import math
-import matplotlib.pyplot as plt
-
-class GaussianUnknownMean(Model):
-    def __init__(self):
-        super().__init__(name='Gaussian with unknown mean') # give the model a name
-        self.prior_mean = 1
-        self.prior_std = math.sqrt(5)
-        self.likelihood_std = math.sqrt(2)
-
-    def forward(self): # Needed to specifcy how the generative model is run forward
-        # sample the (latent) mean variable to be inferred:
-        mu = pyprob.sample(Normal(self.prior_mean, self.prior_std)) # NOTE: sample -> denotes latent variables
-
-        # define the likelihood
-        likelihood = Normal(mu, self.likelihood_std)
-
-        # Lets add two observed variables
-        # -> the 'name' argument is used later to assignment values:
-        pyprob.observe(likelihood, name='obs0') # NOTE: observe -> denotes observable variables
-        pyprob.observe(likelihood, name='obs1')
-
-        # return the latent quantity of interest
-        return mu
-
-model = GaussianUnknownMean()
-model.learn_inference_network(num_traces=20000,
-        observe_embeddings={'obs0' : {'dim' : 32},
-            'obs1': {'dim' : 32}},
-        inference_network=pyprob.InferenceNetwork.LSTM)
-
-observed_list = [12, 10] # Observations
-# sample from posterior (200 samples)
-posterior = model.posterior_results(
-        num_traces=200, # the number of samples estimating the posterior
-        inference_engine=pyprob.InferenceEngine.IMPORTANCE_SAMPLING_WITH_INFERENCE_NETWORK, # specify which inference engine to use
-        observe={
-            'obs0': observed_list[0],
-            'obs1': observed_list[1]} # assign values to the observed values
-        )
-print(posterior.expectation(lambda x: x))

### 2022-08-23 ic/gmm_model.py

@@ -1,64 +1,0 @@
-import pyprob
-from pyprob import Model
-from pyprob.distributions import Normal
-
-import numpy as np
-import torch
-from pyprob.distributions import Gamma, Mixture
-
-class GMM(Model):
-    def __init__(self, K=2):
-        super().__init__(name='Gaussian mixture model')
-        self.d = 2
-        self.K = K
-
-    def forward(self):
-        muxs = list()
-        muys = list()
-        probs = [1./self.K for _ in range(self.K)]
-        for _ in range(self.K):
-            muxs.append(pyprob.sample(Normal(0, 1)))
-            muys.append(pyprob.sample(Normal(0, 1)))
-
-        # define the likelihood factorized over coordinates
-        likelihoodx = Mixture(
-                [Normal(muxs[i], 0.1) for i in range(self.K)],
-                probs=probs)
-        likelihoody = Mixture(
-                [Normal(muys[i], 0.1) for i in range(self.K)],
-                probs=probs)
-
-        pyprob.observe(likelihoodx, name='obs0x')
-        pyprob.observe(likelihoody, name='obs0y')
-        pyprob.observe(likelihoodx, name='obs1x')
-        pyprob.observe(likelihoody, name='obs1y')
-
-        return torch.tensor([muxs, muys]).T
-
-
-if __name__ == '__main__':
-    model = GMM()
-    # prior = model.prior_results(num_traces=10)
-    # print(prior.get_values())
-    model.learn_inference_network(
-            num_traces=20000,
-            observe_embeddings={
-                'obs0x' : {'dim' : 16},
-                'obs0y' : {'dim' : 16},
-                'obs1x' : {'dim' : 16},
-                'obs1y' : {'dim' : 16},
-            },
-            inference_network=pyprob.InferenceNetwork.LSTM)
-
-    observed_list = [100, 100, -100, -100] # Observations
-    # sample from posterior (200 samples)
-    posterior = model.posterior_results(
-            num_traces=200, # the number of samples estimating the posterior
-            inference_engine=pyprob.InferenceEngine.IMPORTANCE_SAMPLING_WITH_INFERENCE_NETWORK, # specify which inference engine to use
-            observe={
-                'obs0x': observed_list[0],
-                'obs0y': observed_list[1],
-                'obs1x': observed_list[2],
-                'obs1y': observed_list[3]} # assign values to the observed values
-            )
-    print(posterior.expectation(lambda x: x))

### 2022-08-23 design/aistats/bayesian.tex
Updates from Overleaf

--- 2022-08-23 design/aistats/bayesian.tex
@@
 \newif\ifisarxiv
-\isarxivfalse
+\isarxivtrue
 
 % ready for submission
-\ifisarxiv
-  \documentclass[11pt]{article}
-  \usepackage{fullpage}
-\else
-  \documentclass[twoside]{article}
-  \usepackage[accepted]{aistats2020}
-
-  % \special{papersize = 8.5in, 11in}
-  % \setlength{\pdfpageheight}{11in}
-  % \setlength{\pdfpagewidth}{8.5in}
-
-  \usepackage{natbib}
-  \renewcommand{\bibname}{References}
-  \renewcommand{\bibsection}{\subsubsection*{\bibname}}
-\fi
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{xfrac}
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{wrapfig}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage{caption}
-\hypersetup{
-	colorlinks,
-	linkcolor={red!40!gray},
-	citecolor={blue!40!gray},
-	urlcolor={blue!70!gray}
-}
-
-\input{../shortdefs}
-\newcommand{\mub}{\bm{\mu}}
-
-\usepackage{grffile}
+% \ifisarxiv
+%   \documentclass[11pt]{article}
+%   \usepackage{fullpage}
+% \else
+%   \documentclass[twoside]{article}
+%   \usepackage[accepted]{aistats2020}
+
+%   % \special{papersize = 8.5in, 11in}
+%   % \setlength{\pdfpageheight}{11in}
+%   % \setlength{\pdfpagewidth}{8.5in}
+
+%   \usepackage{natbib}
+%   \renewcommand{\bibname}{References}
+%   \renewcommand{\bibsection}{\subsubsection*{\bibname}}
+% \fi
+\documentclass[thesis.tex]{subfiles}
 
@@
 
-\ifisarxiv
-  \date{}
-  \def\And{\and}
-\fi
-
-\ifisarxiv
-  \maketitle
-\fi
-
-\begin{abstract}
+% \ifisarxiv
+%   \date{}
+%   \def\And{\and}
+% \fi
+
+% \ifisarxiv
+%   \maketitle
+% \fi
+
+% \begin{abstract}
 We establish a fundamental connection between Bayesian
@@
 with comparable guarantees. 
-\end{abstract}
+% \end{abstract}
 
@@
 
-\bibliographystyle{apalike}
-\bibliography{../pap}
-
-\newpage
-\appendix
-
-\onecolumn
+% \bibliographystyle{apalike}
+% \bibliography{../pap}
+
+% \newpage
+% \appendix
+
+% \onecolumn
 

### 2022-08-23 design/bayesian.tex
Commit original repos
Updates from Overleaf

--- 2022-08-23 bayesian/bayesian.tex
@@
 % ready for submission
-\ifisarxiv
-  \documentclass[11pt]{article}
-  \usepackage{fullpage}
-
-  \usepackage{natbib}
-\else
-  \documentclass[twoside]{article}
-  \usepackage{aistats2020}
-
-  \usepackage{natbib}
-  \renewcommand{\bibname}{References}
-  \renewcommand{\bibsection}{\subsubsection*{\bibname}}
-\fi
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{xfrac}
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{wrapfig}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage{caption}
-
-\input{shortdefs}
-\newcommand{\mub}{\bm{\mu}}
-
-\usepackage{grffile}
+% \ifisarxiv
+%   \documentclass[11pt]{article}
+%   \usepackage{fullpage}
+
+%   \usepackage{natbib}
+% \else
+%   \documentclass[twoside]{article}
+%   \usepackage{aistats2020}
+
+%   \usepackage{natbib}
+%   \renewcommand{\bibname}{References}
+%   \renewcommand{\bibsection}{\subsubsection*{\bibname}}
+% \fi
+
+% \usepackage[utf8]{inputenc} % allow utf-8 input
+% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+% \usepackage{hyperref}       % hyperlinks
+% \usepackage{url}            % simple URL typesetting
+% \usepackage{booktabs}       % professional-quality tables
+% \usepackage{amsfonts}       % blackboard math symbols
+% \usepackage{amsmath}
+% \usepackage{bm}
+% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
+% \usepackage{xfrac}
+% \usepackage{microtype}      % microtypography
+% \usepackage{color}
+% \usepackage{graphicx}
+% \usepackage{tikz}
+% \usepackage{wrapfig}
+% \usepackage{algorithm}
+% \usepackage{algorithmic}
+% \usepackage{caption}
+
+% \input{shortdefs}
+% \newcommand{\mub}{\bm{\mu}}
+
+% \usepackage{grffile}
+
+\documentclass[thesis.tex]{subfiles}
 
 \begin{document}
-\ifisarxiv
-  \title{Bayesian experimental design using regularized \\
-    determinantal point processes}
-  \author{%
-          Micha{\l } Derezi\'{n}ski \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mderezin@berkeley.edu}\\
-  \And
-  Feynman Liang \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{feynman@berkeley.edu}
-  \And
-   Michael W. Mahoney\\
-  ICSI and Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mmahoney@stat.berkeley.edu}
-    % examples of more authors
-    % \And
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
-    % \AND
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
-    % \And
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
-    % \And
-    % Coauthor \\
-    % Affiliation \\
-    % Address \\
-    % \texttt{email} \\
-  }
-\else
-  \twocolumn[
-  \aistatstitle{Bayesian experimental design \\using regularized 
-    determinantal point processes}
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
-}
-  % \aistatsauthor{%
-  %   Micha{\l } Derezi\'{n}ski
-  %   \And
-  %   Feynman Liang
-  %   \And
-  %   Michael W. Mahoney
-  % }
-  % \aistatsaddress {
+% \ifisarxiv
+%   \title{Bayesian experimental design using regularized \\
+%     determinantal point processes}
+%   \author{%
+%           Micha{\l } Derezi\'{n}ski \\
 %   Department of Statistics\\
 %   University of California, Berkeley\\
-  %   \texttt{mderezin@berkeley.edu}
+%   \texttt{mderezin@berkeley.edu}\\
 %   \And
+%   Feynman Liang \\
 %   Department of Statistics\\
@@
 %   \And
+%   Michael W. Mahoney\\
 %   ICSI and Department of Statistics\\
@@
 %   \texttt{mmahoney@stat.berkeley.edu}
+%     % examples of more authors
+%     % \And
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%     % \AND
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%     % \And
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
+%     % \And
+%     % Coauthor \\
+%     % Affiliation \\
+%     % Address \\
+%     % \texttt{email} \\
 %   }
-  ]
-\fi
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\ifisarxiv
-  \date{}
-  \def\And{\and}
-\fi
-
-\ifisarxiv
-  \maketitle
-\fi
-
-\begin{abstract}
+% \else
+%   \twocolumn[
+%   \aistatstitle{Bayesian experimental design \\using regularized 
+%     determinantal point processes}
+% \aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+% \aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
+% }
+%   % \aistatsauthor{%
+%   %   Micha{\l } Derezi\'{n}ski
+%   %   \And
+%   %   Feynman Liang
+%   %   \And
+%   %   Michael W. Mahoney
+%   % }
+%   % \aistatsaddress {
+%   %   Department of Statistics\\
+%   %   University of California, Berkeley\\
+%   %   \texttt{mderezin@berkeley.edu}
+%   %   \And
+%   %   Department of Statistics\\
+%   %   University of California, Berkeley\\
+%   %   \texttt{feynman@berkeley.edu}
+%   %   \And
+%   %   ICSI and Department of Statistics\\
+%   %   University of California, Berkeley\\
+%   %   \texttt{mmahoney@stat.berkeley.edu}
+%   % } 
+%   ]
+% \fi
+
+% % The \author macro works with any number of authors. There are two commands
+% % used to separate the names and addresses of multiple authors: \And and \AND.
+% %
+% % Using \And between authors leaves it to LaTeX to determine where to break the
+% % lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
+% % authors names on the first line, and the last on the second line, try using
+% % \AND instead of \And before the third author name.
+
+% \ifisarxiv
+%   \date{}
+%   \def\And{\and}
+% \fi
+
+% \ifisarxiv
+%   \maketitle
+% \fi
+
+% \begin{abstract}
 We establish a fundamental connection between Bayesian
@@
 
-\bibliographystyle{apalike}
-\bibliography{pap}
+% \bibliographystyle{apalike}
+% \bibliography{pap}
 

### 2022-08-23 design/continuous_nips/continuous.tex
Updates from Overleaf

@@
 
- \ifisarxiv
- \documentclass[11pt]{article}
- \usepackage[round]{natbib}
- \usepackage{fullpage}
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsthm}
- \usepackage{mathabx}
- \usepackage{color}
- \usepackage{cancel}
- \usepackage{wrapfig}
- \usepackage{subfigure}
- \usepackage{caption}
- \usepackage{xfrac}
- \usepackage{algorithm}
- \usepackage{algorithmic}
- \usepackage{mdframed}
- \usepackage{hyperref}
- \usepackage{xcolor}
- \hypersetup{
- 	colorlinks,
- 	linkcolor={red!40!gray},
- 	citecolor={blue!40!gray},
- 	urlcolor={blue!70!gray}
- }
- \usepackage{cleveref}
+%  \ifisarxiv
+ \documentclass[thesis.tex]{subfiles}
+%  \usepackage[round]{natbib}
+%  \usepackage{fullpage}
+%  \usepackage{amsmath}
+%  \usepackage{amssymb}
+%  \usepackage{amsthm}
+%  \usepackage{mathabx}
+%  \usepackage{color}
+%  \usepackage{cancel}
+%  \usepackage{wrapfig}
+%  \usepackage{subfigure}
+%  \usepackage{caption}
+%  \usepackage{xfrac}
+%  \usepackage{algorithm}
+%  \usepackage{algorithmic}
+%  \usepackage{mdframed}
+%  \usepackage{hyperref}
+%  \usepackage{xcolor}
+%  \hypersetup{
+%  	colorlinks,
+%  	linkcolor={red!40!gray},
+%  	citecolor={blue!40!gray},
+%  	urlcolor={blue!70!gray}
+%  }
+%  \usepackage{cleveref}
 
@@
 
-\else
-
-\documentclass{article}
-\usepackage[final]{neurips_2020}
-
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-
-\usepackage{amsmath}
-\usepackage{cancel}
-\usepackage{cleveref}
-\usepackage{subfigure}
-\usepackage{todonotes}
- \usepackage{wrapfig}
-\usepackage{xcolor}
-\usepackage{xfrac}
-\hypersetup{
-	colorlinks,
-	linkcolor={red!40!gray},
-	citecolor={blue!40!gray},
-	urlcolor={blue!70!gray}
-}
-
-\title{Exact expressions for double descent and implicit regularization via
-surrogate random design}
-  \author{%
-Micha{\l } Derezi\'{n}ski \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mderezin@berkeley.edu}\\
-  \And
-Feynman Liang \\
-  Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{feynman@berkeley.edu}
-  \And
-Michael W. Mahoney\\
-  ICSI and Department of Statistics\\
-  University of California, Berkeley\\
-  \texttt{mmahoney@stat.berkeley.edu}
-}
-% \usepackage{times}
-
-% \coltauthor{%
-%   \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
-%   \addr Department of Statistics\\
+% \else
+
+% \documentclass{article}
+% \usepackage[final]{neurips_2020}
+
+% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+% \usepackage{hyperref}       % hyperlinks
+% \usepackage{url}            % simple URL typesetting
+% \usepackage{booktabs}       % professional-quality tables
+% \usepackage{amsfonts}       % blackboard math symbols
+% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
+% \usepackage{microtype}      % microtypography
+
+% \usepackage{amsmath}
+% \usepackage{cancel}
+% \usepackage{cleveref}
+% \usepackage{subfigure}
+% \usepackage{todonotes}
+%  \usepackage{wrapfig}
+% \usepackage{xcolor}
+% \usepackage{xfrac}
+% \hypersetup{
+% 	colorlinks,
+% 	linkcolor={red!40!gray},
+% 	citecolor={blue!40!gray},
+% 	urlcolor={blue!70!gray}
+% }
+
+% \title{Exact expressions for double descent and implicit regularization via
+% surrogate random design}
+%   \author{%
+% Micha{\l } Derezi\'{n}ski \\
+%   Department of Statistics\\
 %   University of California, Berkeley\\
-%   \AND
-%   \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
-%   \addr Department of Statistics\\
+%   \texttt{mderezin@berkeley.edu}\\
+%   \And
+% Feynman Liang \\
+%   Department of Statistics\\
 %   University of California, Berkeley\\
-%   \AND
-%   \Name{Zhenyu Liao} \Email{TODO}\\
-%   \addr Department of Statistics\\
+%   \texttt{feynman@berkeley.edu}
+%   \And
+% Michael W. Mahoney\\
+%   ICSI and Department of Statistics\\
 %   University of California, Berkeley\\
-%   \AND
-%   \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
-%   \addr ICSI and Department of Statistics\\
-%   University of California, Berkeley
+%   \texttt{mmahoney@stat.berkeley.edu}
 % }
-
-% \usepackage{float}
-% \usepackage{wrapfig}
-
-\fi
-
-\input{../shortdefs}
-
-\newcommand{\RED}{\color[rgb]{0.70,0,0}}
+% % \usepackage{times}
+
+% % \coltauthor{%
+% %   \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
+% %   \addr Department of Statistics\\
+% %   University of California, Berkeley\\
+% %   \AND
+% %   \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
+% %   \addr Department of Statistics\\
+% %   University of California, Berkeley\\
+% %   \AND
+% %   \Name{Zhenyu Liao} \Email{TODO}\\
+% %   \addr Department of Statistics\\
+% %   University of California, Berkeley\\
+% %   \AND
+% %   \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
+% %   \addr ICSI and Department of Statistics\\
+% %   University of California, Berkeley
+% % }
+
+% % \usepackage{float}
+% % \usepackage{wrapfig}
+
+% \fi
+
+% \input{../shortdefs}
+
+% \newcommand{\RED}{\color[rgb]{0.70,0,0}}
 
 \begin{document}
-\maketitle
-
-\begin{abstract}
+% \maketitle
+
+% \begin{abstract}
   Double descent refers to the phase transition that is exhibited by
@@
   determinant commutes with expectation.
-\end{abstract}
+% \end{abstract}
 
@@
   \centering
-  \subfigure[%
-    Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
+  \begin{subfigure}{0.49\textwidth}%
+    \subcaption{Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
     numerical estimates even for non-isotropic
     features. Eigenvalue decay leads to a steeper
-    descent curve in the under-determined regime ($n<d$).]{%
-    \includegraphics[width=0.48\textwidth]{figs/descent-intro}
-  }
+    descent curve in the under-determined regime ($n<d$).}
+    \includegraphics[width=0.48\textwidth]{../figs/descent-intro}
+  \end{subfigure}
   \hfill
-  \subfigure[%
-    The mean of the estimator $\X^\dagger\y$ exhibits
+  \begin{subfigure}{0.49\textwidth}
+    \subcaption{The mean of the estimator $\X^\dagger\y$ exhibits
     shrinkage which closely matches the shrinkage of a
     ridge-regularized least squares optimum (theory lines), as characterized by
-    Theorem \ref{t:unbiased}.]{%
-    \includegraphics[width=0.48\textwidth]{figs/descent-shrinkage}
-  }
+    Theorem \ref{t:unbiased}.}
+    \includegraphics[width=0.48\textwidth]{../figs/descent-shrinkage}
+  \end{subfigure}
   \caption{Illustration of the main results for $d=100$ and
@@
   \centering
-  \includegraphics[width=.43\textwidth]{figs/descent-model-log}
+  \includegraphics[width=.43\textwidth]{../figs/descent-model-log}
   \ifisarxiv\else \vspace{-3mm}\fi
@@
 
-\ifisarxiv
-  \bibliographystyle{plainnat}
-\else
-  \bibliographystyle{alpha}
-\fi
-
-\bibliography{../pap}
-
-\ifisarxiv\else\newpage\fi
-
-\appendix
+% \ifisarxiv
+%   \bibliographystyle{plainnat}
+% \else
+%   \bibliographystyle{alpha}
+% \fi
+
+% \bibliography{../pap}
+
+% \ifisarxiv\else\newpage\fi
+
+% \appendix
 

### 2022-08-23 design/equivalents_nips/neurips_2020.tex
Updates from Overleaf

@@
 
-\documentclass{article}
+% \documentclass{article}
+\documentclass[thesis.tex]{subfiles}
 
@@
 \ifisarxiv
-\usepackage{fullpage}
-\usepackage{natbib}
-\else
-\usepackage[final]{neurips_2020}
-\fi
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
- \usepackage{subfigure}
-\usepackage{amsmath,amsthm}
-\usepackage{cancel}
-\usepackage{cleveref}
-\usepackage{mathtools}
-\usepackage{enumitem}
-\usepackage{todonotes}
-\usepackage{xcolor}
-\usepackage{xfrac}
-\usepackage{caption}
-\hypersetup{
-	colorlinks,
-	linkcolor={red!40!gray},
-	citecolor={blue!40!gray},
-	urlcolor={blue!70!gray}
-}
-\newcommand{\BLUE}{\color[rgb]{0,0,0.69}}
-\input{../shortdefs}
+% \usepackage{fullpage}
+% \usepackage{natbib}
+% \else
+% \usepackage[final]{neurips_2020}
+% \fi
+% \usepackage[utf8]{inputenc} % allow utf-8 input
+% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+% \usepackage{hyperref}       % hyperlinks
+% \usepackage{url}            % simple URL typesetting
+% \usepackage{booktabs}       % professional-quality tables
+% \usepackage{amsfonts}       % blackboard math symbols
+% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
+% \usepackage{microtype}      % microtypography
+%  \usepackage{subfigure}
+% \usepackage{amsmath,amsthm}
+% \usepackage{cancel}
+% \usepackage{cleveref}
+% \usepackage{mathtools}
+% \usepackage{enumitem}
+% \usepackage{todonotes}
+% \usepackage{xcolor}
+% \usepackage{xfrac}
+% \usepackage{caption}
+% \hypersetup{
+% 	colorlinks,
+% 	linkcolor={red!40!gray},
+% 	citecolor={blue!40!gray},
+% 	urlcolor={blue!70!gray}
+% }
+% \newcommand{\BLUE}{\color[rgb]{0,0,0.69}}
+% \input{../shortdefs}
 
@@
 \begin{document}
-\maketitle
-
-\begin{abstract}
+% \maketitle
+
+% \begin{abstract}
 It is often desirable to reduce the dimensionality of a large dataset
@@
 methods, down to lower-order effects and even constant factors. 
-\end{abstract}
+% \end{abstract}
 
@@
 \centering
-\subfigure[%
-Singular values are given by $\sigma_i^2=C\cdot \alpha^{i-1}$.]{%
+    \begin{subfigure}{0.49\textwidth}%
+\subcaption{Singular values are given by $\sigma_i^2=C\cdot \alpha^{i-1}$.}
     \includegraphics[width=0.48\textwidth]{explicit_exp}
-  }
+    \end{subfigure}
 \hfill
-\subfigure[%
-Singular values are given by $\sigma_i^2=C\cdot i^{-\beta}$.
-]{%
+    \begin{subfigure}{0.49\textwidth}%
+\subcaption{Singular values are given by $\sigma_i^2=C\cdot i^{-\beta}$.}
       \includegraphics[width=0.48\textwidth]{explicit_poly}
-    }
+    \end{subfigure}
 \caption{
@@
 
-\bibliographystyle{alpha}
-\bibliography{../pap}
+% \bibliographystyle{alpha}
+% \bibliography{../pap}
 
@@
 
-\ifisarxiv\else\newpage\fi
-\appendix
+% \ifisarxiv\else\newpage\fi
+% \appendix
 

### 2022-08-23 ftvi/ftvi_icml/ftvi_icml.tex
Updates from Overleaf

@@
 
-\documentclass[nohyperref]{article}
+% \documentclass[nohyperref]{article}
+\documentclass[thesis.tex]{subfiles}
 
 % Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
-\usepackage{graphicx}
-%\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2022} with \usepackage[nohyperref]{icml2022} above.
-\usepackage{hyperref}
-
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-% \usepackage{icml2022}
-
-% If accepted, instead use the following line for the camera-ready submission:
-\usepackage[accepted]{icml2022}
-
-% For theorems and such
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{mathtools}
-    \usepackage{amsthm}
-
-% if you use cleveref..
-\usepackage[capitalize,noabbrev]{cleveref}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% THEOREMS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\theoremstyle{remark}
-\newtheorem{remark}[theorem]{Remark}
-
-% Todonotes is useful during development; simply uncomment the next line
-%    and comment out the line below the next line to turn off comments
-%\usepackage[disable,textsize=tiny]{todonotes}
-\usepackage[textsize=tiny]{todonotes}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Fat--Tailed VI with Anisotropic Tail Adaptive Flows}
-
-% custom stuff
-\usepackage[bold]{hhtensor}
-\usepackage{mathtools}
-\usepackage{float}
-\usepackage{braket}
-\usepackage{dsfont}
-\usepackage{graphbox}
-\usepackage{pgf}
-\usepackage{comment}
-\usepackage{subcaption}
-\usepackage{enumitem}
-\input{../macros}
+% \usepackage{microtype}
+% \usepackage{graphicx}
+% %\usepackage{subfigure}
+% \usepackage{booktabs} % for professional tables
+
+% % hyperref makes hyperlinks in the resulting PDF.
+% % If your build breaks (sometimes temporarily if a hyperlink spans a page)
+% % please comment out the following usepackage line and replace
+% % \usepackage{icml2022} with \usepackage[nohyperref]{icml2022} above.
+% \usepackage{hyperref}
+
+% % Attempt to make hyperref and algorithmic work together better:
+% \newcommand{\theHalgorithm}{\arabic{algorithm}}
+
+% % Use the following line for the initial blind version submitted for review:
+% % \usepackage{icml2022}
+
+% % If accepted, instead use the following line for the camera-ready submission:
+% \usepackage[accepted]{icml2022}
+
+% % For theorems and such
+% \usepackage{amsmath}
+% \usepackage{amssymb}
+% \usepackage{mathtools}
+%     \usepackage{amsthm}
+
+% % if you use cleveref..
+% \usepackage[capitalize,noabbrev]{cleveref}
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % THEOREMS
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \theoremstyle{plain}
+% \newtheorem{theorem}{Theorem}[section]
+% \newtheorem{proposition}[theorem]{Proposition}
+% \newtheorem{lemma}[theorem]{Lemma}
+% \newtheorem{corollary}[theorem]{Corollary}
+% \theoremstyle{definition}
+% \newtheorem{definition}[theorem]{Definition}
+% \newtheorem{assumption}[theorem]{Assumption}
+% \theoremstyle{remark}
+% \newtheorem{remark}[theorem]{Remark}
+
+% % Todonotes is useful during development; simply uncomment the next line
+% %    and comment out the line below the next line to turn off comments
+% %\usepackage[disable,textsize=tiny]{todonotes}
+% \usepackage[textsize=tiny]{todonotes}
+
+% % The \icmltitle you define below is probably too long as a header.
+% % Therefore, a short form for the running title is supplied here:
+% \icmltitlerunning{Fat--Tailed VI with Anisotropic Tail Adaptive Flows}
+
+% % custom stuff
+% \usepackage[bold]{hhtensor}
+% \usepackage{mathtools}
+% \usepackage{float}
+% \usepackage{braket}
+% \usepackage{dsfont}
+% \usepackage{graphbox}
+% \usepackage{pgf}
+% \usepackage{comment}
+% \usepackage{subcaption}
+% \usepackage{enumitem}
+% \input{../macros}
 
@@
 
-\twocolumn[
-\icmltitle{Fat--Tailed Variational Inference with Anisotropic Tail Adaptive Flows}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2022
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Feynman Liang}{yyy,fb}
-\icmlauthor{Liam Hodgkinson}{yyy,zzz}
-\icmlauthor{Michael W. Mahoney}{yyy,zzz}
-% \icmlauthor{Firstname4 Lastname4}{sch}
-% \icmlauthor{Firstname5 Lastname5}{yyy}
-% \icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
-% \icmlauthor{Firstname7 Lastname7}{comp}
-% %\icmlauthor{}{sch}
-% \icmlauthor{Firstname8 Lastname8}{sch}
-% \icmlauthor{Firstname8 Lastname8}{yyy,comp}
-% %\icmlauthor{}{sch}
-% %\icmlauthor{}{sch}
-\end{icmlauthorlist}
-
-\icmlaffiliation{fb}{Meta, Menlo Park, CA}
-\icmlaffiliation{yyy}{Department of Statistics, University of California, Berkeley, CA}
-\icmlaffiliation{zzz}{International Computer Science Institute, Berkeley, CA}
-% \icmlaffiliation{comp}{Company Name, Location, Country}
-% \icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}
-
-\icmlcorrespondingauthor{Feynman Liang}{feynman@berkeley.edu}
-%\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
+% \twocolumn[
+% \icmltitle{Fat--Tailed Variational Inference with Anisotropic Tail Adaptive Flows}
+
+% % It is OKAY to include author information, even for blind
+% % submissions: the style file will automatically remove it for you
+% % unless you've provided the [accepted] option to the icml2022
+% % package.
+
+% % List of affiliations: The first argument should be a (short)
+% % identifier you will use later to specify author affiliations
+% % Academic affiliations should list Department, University, City, Region, Country
+% % Industry affiliations should list Company, City, Region, Country
+
+% % You can specify symbols, otherwise they are numbered in order.
+% % Ideally, you should not use this facility. Affiliations will be numbered
+% % in order of appearance and this is the preferred way.
+% \icmlsetsymbol{equal}{*}
+
+% \begin{icmlauthorlist}
+% \icmlauthor{Feynman Liang}{yyy,fb}
+% \icmlauthor{Liam Hodgkinson}{yyy,zzz}
+% \icmlauthor{Michael W. Mahoney}{yyy,zzz}
+% % \icmlauthor{Firstname4 Lastname4}{sch}
+% % \icmlauthor{Firstname5 Lastname5}{yyy}
+% % \icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
+% % \icmlauthor{Firstname7 Lastname7}{comp}
+% % %\icmlauthor{}{sch}
+% % \icmlauthor{Firstname8 Lastname8}{sch}
+% % \icmlauthor{Firstname8 Lastname8}{yyy,comp}
+% % %\icmlauthor{}{sch}
+% % %\icmlauthor{}{sch}
+% \end{icmlauthorlist}
+
+% \icmlaffiliation{fb}{Meta, Menlo Park, CA}
+% \icmlaffiliation{yyy}{Department of Statistics, University of California, Berkeley, CA}
+% \icmlaffiliation{zzz}{International Computer Science Institute, Berkeley, CA}
+% % \icmlaffiliation{comp}{Company Name, Location, Country}
+% % \icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}
+
+% \icmlcorrespondingauthor{Feynman Liang}{feynman@berkeley.edu}
+% %\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}
+
+% % You may provide any keywords that you
+% % find helpful for describing your paper; these are used to populate
+% % the "keywords" metadata in the PDF but will not be shown in the document
+% \icmlkeywords{Machine Learning, ICML}
+
+% \vskip 0.3in
+% ]
+
+% % this must go after the closing bracket ] following \twocolumn[ ...
+
+% % This command actually creates the footnote in the first column
+% % listing the affiliations and the copyright notice.
+% % The command takes one argument, which is text to display at the start of the footnote.
+% % The \icmlEqualContribution command is standard text for equal contribution.
+% % Remove it (just {}) if you do not need this facility.
+
+% \printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+% %\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+
+% \begin{abstract}
     While fat-tailed densities commonly arise as posterior and marginal distributions in
@@
     is competitive with prior work while also exhibiting appropriate tail-anisotropy.
-\end{abstract}
+% \end{abstract}
 
@@
 
-\bibliography{refs_ftvi}
-\bibliographystyle{icml2022}
+% \bibliography{refs_ftvi}
+% \bibliographystyle{icml2022}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\appendix
-\onecolumn
+% \newpage
+% \appendix
+% \onecolumn
 
@@
     \vspace{-.5in}
-    \includegraphics[scale=0.8]{Figures/spiral.pdf}
+    \includegraphics[scale=0.8]{../Figures/spiral.pdf}
     % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\

### 2022-08-23 hta/neurips_2022.tex
Updates from Overleaf

@@
-\documentclass{article}
+\documentclass[thesis.tex]{subfiles}
+% \documentclass{article}
 \csname remappendix\endcsname
@@
 % ready for submission
-\usepackage{neurips_2022}
+% \usepackage{neurips_2022}
 
@@
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{xcolor}         % colors
-\usepackage{longtable}
-\usepackage{wrapfig}
-\usepackage{array}
-\usepackage{enumerate}
-\usepackage{enumitem}
-\usepackage{hhline}
-\usepackage{multirow}
-
-\usepackage{amsmath}
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-\usepackage{todonotes}
-\title{Static Analysis of Tail Behaviour with a \\ Generalized Gamma Algebra}
-
-\newcommand{\feynman}[1]{\todo{(Feynman): #1}}
-\newcommand{\dd}{\mathrm{d}}
-\usepackage{tikz}
-\usetikzlibrary{bayesnet}
-\usepackage{cleveref}
-\usepackage{listings}
-\usepackage{multirow}
-\usepackage{siunitx}
-\usepackage{etoolbox}
-\robustify\bfseries
-\usepackage[ruled]{algorithm2e}
-\usepackage{float}
-
-\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
-\newcolumntype{M}[1]{>{\centering\arraybackslash}m{#1}}
+% \usepackage[utf8]{inputenc} % allow utf-8 input
+% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+% \usepackage{hyperref}       % hyperlinks
+% \usepackage{url}            % simple URL typesetting
+% \usepackage{booktabs}       % professional-quality tables
+% \usepackage{amsfonts}       % blackboard math symbols
+% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
+% \usepackage{microtype}      % microtypography
+% \usepackage{xcolor}         % colors
+% \usepackage{longtable}
+% \usepackage{wrapfig}
+% \usepackage{array}
+% \usepackage{enumerate}
+% \usepackage{enumitem}
+% \usepackage{hhline}
+% \usepackage{multirow}
+
+% \usepackage{amsmath}
+% \usepackage{amsthm}
+% \newtheorem{theorem}{Theorem}
+% \newtheorem{proposition}{Proposition}
+% \newtheorem{lemma}[theorem]{Lemma}
+% \theoremstyle{definition}
+% \newtheorem{definition}{Definition}
+% \newtheorem{example}{Example}
+% \usepackage{todonotes}
+% \title{Static Analysis of Tail Behaviour with a \\ Generalized Gamma Algebra}
+
+% \newcommand{\feynman}[1]{\todo{(Feynman): #1}}
+% \newcommand{\dd}{\mathrm{d}}
+% \usepackage{tikz}
+% \usetikzlibrary{bayesnet}
+% \usepackage{cleveref}
+% \usepackage{listings}
+% \usepackage{multirow}
+% \usepackage{siunitx}
+% \usepackage{etoolbox}
+% \robustify\bfseries
+% \usepackage[ruled]{algorithm2e}
+% \usepackage{float}
+
+% \newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}
+% \newcolumntype{M}[1]{>{\centering\arraybackslash}m{#1}}
 
@@
 
-\maketitle
-
-\begin{abstract}
+% \maketitle
+
+% \begin{abstract}
 	Despite the successes of probabilistic models based on passing noise through
@@
 	inference tasks.
-\end{abstract}
+% \end{abstract}
 
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{figures/schematic.pdf}
+	\includegraphics[width=\textwidth]{./figures/schematic.pdf}
 	\caption{Our overall approach for density approximations with calibrated tails.
@@
 %% The argument will be surrounded by /* ... */
-\SetKwComment{Comment}{/* }{ */}
+% \SetKwComment{Comment}{/* }{ */}
 
@@
 	\caption{Pseudocode for a GGA tails static analysis pass}\label{alg:bfs_typecheck}
-	\KwData{Abstract syntax tree for a PPL program}
-	\KwResult{GGA parameter estimates for all random variables}
-	frontier $\gets$ [rv : Parents(rv) = $\emptyset$]\;
-	GGAs $\gets \{\}$\;
-	\While{\text{frontier} $\neq \emptyset$}{
-		next $\gets$ frontier.popLeft()\;
-		GGAs[next] $\gets$ computeGGA(next.op, next.parent)\;
-		frontier $\gets$ frontier + next.children()\;
-	}
-	\Return{GGAs}
+	\begin{algorithmic}
+	\REQUIRE{Abstract syntax tree for a PPL program}
+
+	\STATE{frontier $\gets$ [rv : Parents(rv) = $\emptyset$]}\\
+	\STATE{GGAs $\gets \{\}$}
+
+	\WHILE{\text{frontier} $\neq \emptyset$} 
+	\STATE{next $\gets$ frontier.popLeft()}\\
+	\STATE{GGAs[next] $\gets$ computeGGA(next.op, next.parent)}\\
+	\STATE{frontier $\gets$ frontier + next.children()}
+	\ENDWHILE
+
+	\RETURN{GGA parameter estimates for all random variables}
+	\end{algorithmic}
 \end{algorithm}
@@
 		\centering
-		\input{experiments/normal-power-sum.pgf}
-		\input{experiments/lip-nf.pgf}
+		\input{./experiments/normal-power-sum.pgf}
+		\input{./experiments/lip-nf.pgf}
 		\caption{(Top) $5000$ samples of $\lvert x^\top A y \rvert$ vs the
@@
 
-\begin{ack}
-	Use unnumbered first level headings for the acknowledgments. All acknowledgments
-	go at the end of the paper before the list of references. Moreover, you are required to declare
-	funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
-	More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2021/PaperInformation/FundingDisclosure}.
-
-	Do {\bf not} include this section in the anonymized submission, only in the final paper.
+% \begin{ack}
+% 	Use unnumbered first level headings for the acknowledgments. All acknowledgments
+% 	go at the end of the paper before the list of references. Moreover, you are required to declare
+% 	funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
+% 	More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2021/PaperInformation/FundingDisclosure}.
+
+% 	Do {\bf not} include this section in the anonymized submission, only in the final paper.
 You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
-\end{ack}
+% \end{ack}
 
@@
 
-{
-\small
-\bibliographystyle{unsrtnat}
-\bibliography{refs}
-}
+% {
+% \small
+% \bibliographystyle{unsrtnat}
+% \bibliography{refs}
+% }
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Checklist}
+% \section*{Checklist}
 
@@
 
-\begin{enumerate}
-
-\item For all authors...
-\begin{enumerate}
-  \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
-    \answerYes{}
-  \item Did you describe the limitations of your work?
-    \answerYes{See \cref{sec:conclusion}}
-  \item Did you discuss any potential negative societal impacts of your work?
-    \answerYes{See \cref{sec:conclusion}}
-  \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
-    \answerYes{}
-\end{enumerate}
-
-\item If you are including theoretical results...
-\begin{enumerate}
-  \item Did you state the full set of assumptions of all theoretical results?
-    \answerYes{Stated at start of \cref{sec:gga}}
-	\item Did you include complete proofs of all theoretical results?
-    \answerYes{Most are deferred to supplementary materials}
-\end{enumerate}
-
-\item If you ran experiments...
-\begin{enumerate}
-  \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
-    \answerYes{See supplemental, a URL for code will be provided witn camera ready}
-  \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
-    \answerYes{}
-	\item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
-    \answerYes{Standard errors reported in parenthesis on \cref{tab:de} and \cref{tab:vi}}
-	\item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
-    \answerYes{See \cref{sec:experiments}}
-\end{enumerate}
-
-\item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
-\begin{enumerate}
-  \item If your work uses existing assets, did you cite the creators?
-    \answerYes{See citations of \citep{tehrani2020bean}}
-  \item Did you mention the license of the assets?
-    \answerYes{Licence included in code}
-  \item Did you include any new assets either in the supplemental material or as a URL?
-    \answerYes{See supplemental material}
-  \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
-    \answerNA{}
-  \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
-    \answerNA{}
-\end{enumerate}
-
-\item If you used crowdsourcing or conducted research with human subjects...
-\begin{enumerate}
-  \item Did you include the full text of instructions given to participants and screenshots, if applicable?
-    \answerNA{}
-  \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
-    \answerNA{}
-  \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
-    \answerNA{}
-\end{enumerate}
-
-\end{enumerate}
+% \begin{enumerate}
+
+% \item For all authors...
+% \begin{enumerate}
+%   \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
+%     \answerYes{}
+%   \item Did you describe the limitations of your work?
+%     \answerYes{See \cref{sec:conclusion}}
+%   \item Did you discuss any potential negative societal impacts of your work?
+%     \answerYes{See \cref{sec:conclusion}}
+%   \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
+%     \answerYes{}
+% \end{enumerate}
+
+% \item If you are including theoretical results...
+% \begin{enumerate}
+%   \item Did you state the full set of assumptions of all theoretical results?
+%     \answerYes{Stated at start of \cref{sec:gga}}
+% 	\item Did you include complete proofs of all theoretical results?
+%     \answerYes{Most are deferred to supplementary materials}
+% \end{enumerate}
+
+% \item If you ran experiments...
+% \begin{enumerate}
+%   \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
+%     \answerYes{See supplemental, a URL for code will be provided witn camera ready}
+%   \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
+%     \answerYes{}
+% 	\item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
+%     \answerYes{Standard errors reported in parenthesis on \cref{tab:de} and \cref{tab:vi}}
+% 	\item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
+%     \answerYes{See \cref{sec:experiments}}
+% \end{enumerate}
+
+% \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
+% \begin{enumerate}
+%   \item If your work uses existing assets, did you cite the creators?
+%     \answerYes{See citations of \citep{tehrani2020bean}}
+%   \item Did you mention the license of the assets?
+%     \answerYes{Licence included in code}
+%   \item Did you include any new assets either in the supplemental material or as a URL?
+%     \answerYes{See supplemental material}
+%   \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
+%     \answerNA{}
+%   \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
+%     \answerNA{}
+% \end{enumerate}
+
+% \item If you used crowdsourcing or conducted research with human subjects...
+% \begin{enumerate}
+%   \item Did you include the full text of instructions given to participants and screenshots, if applicable?
+%     \answerNA{}
+%   \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
+%     \answerNA{}
+%   \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
+%     \answerNA{}
+% \end{enumerate}
+
+% \end{enumerate}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\ifdefined\remappendix
-\newpage
-
-\appendix
+% \ifdefined\remappendix
+% \newpage
+
+% \appendix
 
@@
 
-\fi
+% \fi
 \end{document}

### 2022-08-23 ic/aistats2021/paper.tex
Updates from Overleaf

@@
-\documentclass[twoside]{article}
+% \documentclass[twoside]{article}
+\documentclass[thesis.tex]{subfiles}
 
 %\usepackage{aistats2021}
-\usepackage[accepted]{aistats2021}
+% \usepackage[accepted]{aistats2021}
 % If your paper is accepted, change the options for the package
@@
 % If you set papersize explicitly, activate the following three lines:
-\special{papersize = 8.5in, 11in}
-\setlength{\pdfpageheight}{11in}
-\setlength{\pdfpagewidth}{8.5in}
+% \special{papersize = 8.5in, 11in}
+% \setlength{\pdfpageheight}{11in}
+% \setlength{\pdfpagewidth}{8.5in}
 
 % If you use natbib package, activate the following three lines:
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
+% \usepackage[round]{natbib}
+% \renewcommand{\bibname}{References}
+% \renewcommand{\bibsection}{\subsubsection*{\bibname}}
 
 % If you use BibTeX in apalike style, activate the following line:
-\bibliographystyle{apalike}
+% \bibliographystyle{apalike}
 
-\input{macros.tex}
+% \input{macros.tex}
 
@@
 
-\twocolumn[
+% \twocolumn[
 
-  \aistatstitle{Accelerating Metropolis-Hastings with Lightweight Inference Compilation}
+%   \aistatstitle{Accelerating Metropolis-Hastings with Lightweight Inference Compilation}
 
-  \aistatsauthor{ Feynman Liang \And Nimar Arora \And Nazanin Tehrani \\ \\
-    \AND Yucen Li \And Michael Tingley \And Erik Meijer }
-  \vspace{-0.60in}
-  \aistatsaddress{ UC Berkeley \\ feynman@berkeley.edu
-    \And Facebook \\ nimarora@fb.com
-    \And Facebook \\ nazaninkt@fb.com
-    \AND Facebook \\ yucenli@fb.com
-    \And Facebook \\ tingley@fb.com
-    \And Facebook \\ erikm@fb.com}
-]
+%   \aistatsauthor{ Feynman Liang \And Nimar Arora \And Nazanin Tehrani \\ \\
+%     \AND Yucen Li \And Michael Tingley \And Erik Meijer }
+%   \vspace{-0.60in}
+%   \aistatsaddress{ UC Berkeley \\ feynman@berkeley.edu
+%     \And Facebook \\ nimarora@fb.com
+%     \And Facebook \\ nazaninkt@fb.com
+%     \AND Facebook \\ yucenli@fb.com
+%     \And Facebook \\ tingley@fb.com
+%     \And Facebook \\ erikm@fb.com}
+% ]
 
-\runningauthor{Liang, Arora, Tehrani, Li, Tingley, Meijer}
-\begin{abstract}
+% \runningauthor{Liang, Arora, Tehrani, Li, Tingley, Meijer}
+% \begin{abstract}
   In order to construct accurate proposers for Metropolis-Hastings Markov
@@
   a Bayesian logistic regression and $n$-schools inference application.
-\end{abstract}
+% \end{abstract}
 
@@
 
-\bibliography{refs.bib}
+% \bibliography{refs.bib}
 

### 2022-11-07 design/continuous_nips/continuous.tex
Updates from Overleaf

@@
- \newif\ifisarxiv
- \isarxivfalse
-
-%  \ifisarxiv
   \documentclass[thesis.tex]{subfiles}
@@
 
- \title{Exact expressions for double descent
- and implicit regularization
- \\
- via surrogate random design}
-   \author{%
-           \textbf{Micha{\l } Derezi\'{n}ski} \\
-   Department of Statistics\\
-   University of California, Berkeley\\
-   \texttt{mderezin@berkeley.edu}\\
-   \and
-   \textbf{Feynman Liang} \\
-   Department of Statistics\\
-   University of California, Berkeley\\
-   \texttt{feynman@berkeley.edu}
-   \and
-    \textbf{Michael W. Mahoney}\\
-   ICSI and Department of Statistics\\
-   University of California, Berkeley\\
-   \texttt{mmahoney@stat.berkeley.edu}
- }
-
- %  \newcommand{\jmlrBlackBox}{\rule{1.5ex}{1.5ex}}
-   \providecommand{\BlackBox}{\rule{1.5ex}{1.5ex}}
- %  \newcommand{\jmlrQED}{\hfill\jmlrBlackBox\par\bigskip}
-   \renewenvironment{proof}%
-   {%
-    \par\noindent{\bfseries\upshape Proof\ }%
-   }%
- {\hfill\BlackBox\par\bigskip}
+%  \title{Exact expressions for double descent
+%  and implicit regularization
+%  \\
+%  via surrogate random design}
+%   \author{%
+%           \textbf{Micha{\l } Derezi\'{n}ski} \\
+%   Department of Statistics\\
+%   University of California, Berkeley\\
+%   \texttt{mderezin@berkeley.edu}\\
+%   \and
+%   \textbf{Feynman Liang} \\
+%   Department of Statistics\\
+%   University of California, Berkeley\\
+%   \texttt{feynman@berkeley.edu}
+%   \and
+%     \textbf{Michael W. Mahoney}\\
+%   ICSI and Department of Statistics\\
+%   University of California, Berkeley\\
+%   \texttt{mmahoney@stat.berkeley.edu}
+%  }
+
+%  %  \newcommand{\jmlrBlackBox}{\rule{1.5ex}{1.5ex}}
+%   \providecommand{\BlackBox}{\rule{1.5ex}{1.5ex}}
+%  %  \newcommand{\jmlrQED}{\hfill\jmlrBlackBox\par\bigskip}
+%   \renewenvironment{proof}%
+%   {%
+%     \par\noindent{\bfseries\upshape Proof\ }%
+%   }%
+%  {\hfill\BlackBox\par\bigskip}
 
@@
     descent curve in the under-determined regime ($n<d$).}
-    \includegraphics[width=0.48\textwidth]{../figs/descent-intro}
+    \includegraphics[width=\textwidth]{../figs/descent-intro}
   \end{subfigure}
@@
     Theorem \ref{t:unbiased}.}
-    \includegraphics[width=0.48\textwidth]{../figs/descent-shrinkage}
+    \includegraphics[width=\textwidth]{../figs/descent-shrinkage}
   \end{subfigure}
@@
 
-\begin{wrapfigure}{r}{\ifisarxiv 0.5\else 0.44\fi\textwidth}
-  \ifisarxiv\else \vspace{-.2cm}\fi
+\begin{figure}[h]
   \centering
-  \includegraphics[width=.43\textwidth]{../figs/descent-model-log}
-  \ifisarxiv\else \vspace{-3mm}\fi
+  \includegraphics[width=.5\textwidth]{../figs/descent-model-log}
   \caption{
@@
   \label{f:model}
-  \ifisarxiv\else  \vspace{-.5cm}\fi
-\end{wrapfigure}
+\end{figure}
 
@@
 % Gaussian analysis.
-\ifisarxiv\else
-  \section*{Broader Impact}
-
-  While the double descent phenomenon has been empirically observed in a variety
-  of applications, mathematical descriptions of it are oftentimes complex and
-  inaccesible to non-experts. In contrast, our surrogate design and the
-  accompanying closed-form expressions provide an easily computable rule of thumb
-  for estimating the generalization error. One important application is
-  the high-dimensional (i.e. underdetermined, $n < d$) regime
-  experienced by modern machine learning systems where the number of
-  parameters vastly exceeds the quantity of
-  available data. Our research can be applied here to provide a theoretical
-  understanding of the surprising phenomenon where without proper
-  regularization \citep{nakkiran2020optimal} more data (i.e. increasing $n$)
-  may lead to worse generalization performance.
-
-  Better theoretical understanding of generalization error in the small data
-  regime has important societal impact. Through theoretically modeling a system's
-  performance, we can build safer systems by better understanding how badly an
-  estimator fails and accounting for these failure modes in the system's design.
-  Furthermore, the improved understanding of minimum norm solutions performing
-  worse with more data offers an appealing trade-off where certain systems can
-  both improve their performance and respect the privacy of its users by
-  collecting less data.
-\fi
-
-\paragraph{Acknowledgements.}
-We would like to acknowledge ARO, DARPA, NSF, ONR, and GFSD for providing
-partial support of this work. We also thank Zhenyu Liao for pointing
-out fruitful connections between our results and the asymptotic
-analysis of random matrix resolvents.
+% \ifisarxiv\else
+%   \section*{Broader Impact}
+
+%   While the double descent phenomenon has been empirically observed in a variety
+%   of applications, mathematical descriptions of it are oftentimes complex and
+%   inaccesible to non-experts. In contrast, our surrogate design and the
+%   accompanying closed-form expressions provide an easily computable rule of thumb
+%   for estimating the generalization error. One important application is
+%   the high-dimensional (i.e. underdetermined, $n < d$) regime
+%   experienced by modern machine learning systems where the number of
+%   parameters vastly exceeds the quantity of
+%   available data. Our research can be applied here to provide a theoretical
+%   understanding of the surprising phenomenon where without proper
+%   regularization \citep{nakkiran2020optimal} more data (i.e. increasing $n$)
+%   may lead to worse generalization performance.
+
+%   Better theoretical understanding of generalization error in the small data
+%   regime has important societal impact. Through theoretically modeling a system's
+%   performance, we can build safer systems by better understanding how badly an
+%   estimator fails and accounting for these failure modes in the system's design.
+%   Furthermore, the improved understanding of minimum norm solutions performing
+%   worse with more data offers an appealing trade-off where certain systems can
+%   both improve their performance and respect the privacy of its users by
+%   collecting less data.
+% \fi
+
+% \paragraph{Acknowledgements.}
+% We would like to acknowledge ARO, DARPA, NSF, ONR, and GFSD for providing
+% partial support of this work. We also thank Zhenyu Liao for pointing
+% out fruitful connections between our results and the asymptotic
+% analysis of random matrix resolvents.
 

### 2022-11-07 hta/neurips_2022.tex
Updates from Overleaf

@@
 % \documentclass{article}
-\csname remappendix\endcsname
+% \csname remappendix\endcsname
 
@@
 
-\author{%
-  David S.~Hippocampus\thanks{Use footnote for providing further information
-    about author (webpage, alternative address)---\emph{not} for acknowledging
-    funding agencies.} \\
-  Department of Computer Science\\
-  Cranberry-Lemon University\\
-  Pittsburgh, PA 15213 \\
-  \texttt{hippo@cs.cranberry-lemon.edu} \\
-  % examples of more authors
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \AND
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-}
-
-\sisetup{detect-all = true}
+% \author{%
+%   David S.~Hippocampus\thanks{Use footnote for providing further information
+%     about author (webpage, alternative address)---\emph{not} for acknowledging
+%     funding agencies.} \\
+%   Department of Computer Science\\
+%   Cranberry-Lemon University\\
+%   Pittsburgh, PA 15213 \\
+%   \texttt{hippo@cs.cranberry-lemon.edu} \\
+%   % examples of more authors
+%   % \And
+%   % Coauthor \\
+%   % Affiliation \\
+%   % Address \\
+%   % \texttt{email} \\
+%   % \AND
+%   % Coauthor \\
+%   % Affiliation \\
+%   % Address \\
+%   % \texttt{email} \\
+%   % \And
+%   % Coauthor \\
+%   % Affiliation \\
+%   % Address \\
+%   % \texttt{email} \\
+%   % \And
+%   % Coauthor \\
+%   % Affiliation \\
+%   % Address \\
+%   % \texttt{email} \\
+% }
+
+% \sisetup{detect-all = true}
 
@@
     \bgroup
-\def\arraystretch{2.5}
-\tolerance=1
-\emergencystretch=\maxdimen
-\hyphenpenalty=10000
-\hbadness=10000
+{\small
 \begin{tabular}{|M{0.15\linewidth} | p{0.75\linewidth}|}
@@
     \end{tabular}
+}
     \vspace{.25cm}
@@
 
-\textbf{Variational Inference} The optimization objective is the ELBO $E_q \log \frac{p(X)}{q(X)} \approx \frac{1}{N} \sum_{i=1}^N \log \frac{p(x_i)}{q(x_i)},\quad x_i \sim q$.
+\textbf{Variational Inference} The optimization objective is the ELBO
+\[
+    E_q \log \frac{p(X)}{q(X)} \approx \frac{1}{N} \sum_{i=1}^N \log \frac{p(x_i)}{q(x_i)},\quad x_i \sim q
+\]
 Here, the density $p$ must also be evaluated so for simplicity
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{figures/LogNormal_Bulk.png}
-	\includegraphics[width=0.48\textwidth]{figures/LogNormal_Tail.png}
+	\includegraphics[width=0.4\textwidth]{figures/LogNormal_Bulk.png}
+	\includegraphics[width=0.4\textwidth]{figures/LogNormal_Tail.png}
 	\caption{Estimation of the log-normal density by tail algebra applied to $V_k$.}
@@
 where we also see that GGA achieves similar approximation quality.
-This is unsurprising as the GGA approximation in \cref{tab:dist_list}
+This is unsurprising as the GGA approximation in \Cref{tab:dist_list}
 is also a Normal distribution.
 
-\begin{figure}[H]
+\begin{figure}
 	\centering
-	\includegraphics[width=.6\textwidth]{figures/de_normal_sum.png}
-	\includegraphics[width=.6\textwidth]{figures/vi_normal_sum.png}
+	\includegraphics[width=1.0\textwidth]{figures/de_normal_sum.png}
+	\includegraphics[width=.4\textwidth]{figures/vi_normal_sum.png}
 	\caption{Density estimation and VI against a known normal target}
@@
 It is important here to recognize that the power operation $X \mapsto X^2$ is not equivalent to the multiplication operation $X \mapsto X \otimes X$, as multiplication assumes independence.
-\begin{figure}[H]
+\begin{figure}
 	\centering
@@
 
+{\small
 \bgroup
@@
 \hline 
-Generalized hyperbolic distribution & $(-\infty,\infty)$ & $\propto e^{\beta(x-\mu)}\frac{K_{\lambda-1/2}(\alpha\sqrt{\delta^{2}+(x-\mu)^{2}})}{(\delta^{2}+(x-\mu)^{2})^{1/4-\lambda/2}}$ & $(\lambda-1,\alpha-\beta,1)$\tabularnewline
-\hline 
-Generalized Normal distribution & $(-\infty,\infty)$ & $\frac{\beta}{2\alpha\Gamma(1/\beta)}\exp\left(-\left(\frac{|x-\mu|}{\alpha}\right)^{\beta}\right)$ & $(0,\alpha^{-\beta},\beta)$\tabularnewline
+Gen.
+hyperbolic distribution & $(-\infty,\infty)$ & $\propto e^{\beta(x-\mu)}\frac{K_{\lambda-1/2}(\alpha\sqrt{\delta^{2}+(x-\mu)^{2}})}{(\delta^{2}+(x-\mu)^{2})^{1/4-\lambda/2}}$ & $(\lambda-1,\alpha-\beta,1)$\tabularnewline
+\hline 
+Gen. Normal distribution & $(-\infty,\infty)$ & $\frac{\beta}{2\alpha\Gamma(1/\beta)}\exp\left(-\left(\frac{|x-\mu|}{\alpha}\right)^{\beta}\right)$ & $(0,\alpha^{-\beta},\beta)$\tabularnewline
 \hline 
@@
 \hline 
-Inverse chi-squared distribution & $(0,\infty)$ & $\frac{2^{-k/2}}{\Gamma(k/2)}x^{-k/2-1}e^{-1/(2x)}$ & $(-\frac{k}{2}-1,\frac{1}{2},-1)$\tabularnewline
-\hline 
-Inverse gamma distribution & $(0,\infty)$ & $\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{-\alpha-1}e^{-\beta/x}$ & $(-\alpha-1,\beta,-1)$\tabularnewline
+Inv. chi-squared distribution & $(0,\infty)$ & $\frac{2^{-k/2}}{\Gamma(k/2)}x^{-k/2-1}e^{-1/(2x)}$ & $(-\frac{k}{2}-1,\frac{1}{2},-1)$\tabularnewline
+\hline 
+Inv. gamma distribution & $(0,\infty)$ & $\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{-\alpha-1}e^{-\beta/x}$ & $(-\alpha-1,\beta,-1)$\tabularnewline
 \hline 
@@
 \egroup
+}
 

### 2022-11-07 ic/aistats2021/paper.tex
Updates from Overleaf

@@
   \centering
-  \includegraphics[width=0.8\linewidth]{Figures/normal_normal_mean_type1.pdf}
+  \includegraphics[width=0.6\linewidth]{Figures/normal_normal_mean_type1.pdf}
   \caption{In a conjugate normal-normal model, LIC's proposal distribution
@@
 Consider a Bayesian logistic regression model over $d$ covariates
-with prior
+with prior distribution 
 $\vec{\beta} \sim \cN_{d+1}(\vec{0}_{d+1}, \diag(10, 2.5\vec{1}_{d}))$
@@
 \begin{figure*}
-  \centering
   \caption{
-    Results drawing 100 samples across 10 chains (after 1,000 burn-in / adaptation / compilation samples)
-    on two Bayesian inference problems.
-    The top row of each subfigure aggregates over the 10 chains the (left to
-    right) compilation time (time to instantiate the Python class, includes
-    neural network training for LIC and Stan's C\texttt{++} codegen and compilation
-    for NUTS), inference time (time from invoking \texttt{infer()} to when
-    sampling terminates), and the predictive log-likelihood (PLL) on held-out
-    test data.
-    The bottom row aggregates over both the 10 chains as well as over all
-    latent variables the (left to right) estimated expected sample size (ESS,
+    Results of MCMC on two Bayesian inference tasks where we compare
+    the compilation time (neural network training for LIC and Stan's C\texttt{++} codegen / compilation
+    for NUTS), inference time, predictive log-likelihood (PLL) on
+    hold-out data, expected sample size (ESS,
     higher is better, \cite{geyer2011introduction}) and the rank normalized
@@
 % TODO: include in camera ready
-\subsubsection*{Acknowledgements}
-Feynman Liang is supported by a NPSC Graduate Fellowship.
-This work was completed during an internship at Facebook.
+% \subsubsection*{Acknowledgements}
+% Feynman Liang is supported by a NPSC Graduate Fellowship.
+% This work was completed during an internship at Facebook.
 

### 2022-11-07 intro.tex
Updates from Overleaf

\chapter{Introduction}

In \todo{Chapter design}, we consider approximately optimal Bayesian experimental design
using an adaptive row sampling algorithm based on DPPs and show that it provides
good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
applied here, in \todo{Chapter double descent} an extension of the DPP-based design is analyzed in
closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
linear regression which closely matches empirical experiments.
In \todo{Chapter random projections} we isolate the part of the proof involving concentration
of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
expected projections $X_S^\dag X_S$ when $X_S = S X$ is obtained by sub-Gaussian sketching matrix $S$.

The second three chapters concern the setting of probabilistic programming, a computational tool
for specifying probabilistic models and automating inference through MCMC and VI. In particular, we  
study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
importance sampling and variational inference.
In \todo{Chapter LIC}, we consider parameterizing $q(x)$ by graph neural networks which condition each
node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
\cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
the length of execution traces.
When the target $p(x)$ is both multivariate and heavy tailed, \todo{Chapter FTVI}
considers the problem of tail anisotropy through both a theoretical and practical perspective.
We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
an anisotropic approximation (fat-tailed variational inference, FTVI) where
an anisotropic product base measures is pushed forwards through a bijective neural network,
and confirm that in practice FTVI improves both density estimation as well as variational inference.
However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
the tail parameter. In \todo{Chapter HTA}, we consider addressing this issue during static analysis of a
probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
such as sums and products. This enables a priori computation of tail parameters, which we show
improves the stability and convergence of a number of inference tasks.

### 2022-11-07 abstract.tex
Adds devcontainer and magic comments

--- 2022-11-07 intro.tex
@@
-\chapter{Introduction}
+% !TeX root = thesis.tex
+% (This file is included by thesis.tex; you do not latex it by itself.)
 
-In \todo{Chapter design}, we consider approximately optimal Bayesian experimental design
+\begin{abstract}
+
+    % The text of the abstract goes here.  If you need to use a \section
+    % command you will need to use \section*, \subsection*, etc. so that
+    % you don't get any numbering.  You probably won't be using any of
+    % these commands in the abstract anyway.
+
+    In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
     using an adaptive row sampling algorithm based on DPPs and show that it provides
     good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
-applied here, in \todo{Chapter double descent} an extension of the DPP-based design is analyzed in
+    applied here, in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
     closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
     linear regression which closely matches empirical experiments.
-In \todo{Chapter random projections} we isolate the part of the proof involving concentration
+    In \cref{ch:projections} we isolate the part of the proof involving concentration
     of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
@@
     importance sampling and variational inference.
-In \todo{Chapter LIC}, we consider parameterizing $q(x)$ by graph neural networks which condition each
+    In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
     node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
@@
     the length of execution traces.
-When the target $p(x)$ is both multivariate and heavy tailed, \todo{Chapter FTVI}
+    When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
     considers the problem of tail anisotropy through both a theoretical and practical perspective.
@@
     However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
-the tail parameter. In \todo{Chapter HTA}, we consider addressing this issue during static analysis of a
+    the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
     probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
@@
     improves the stability and convergence of a number of inference tasks.
+
+\end{abstract}

### 2022-11-07 intro.tex
Adds devcontainer and magic comments

@@
+% !TeX root = thesis.tex
 \chapter{Introduction}
 
-In \todo{Chapter design}, we consider approximately optimal Bayesian experimental design
+In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
 using an adaptive row sampling algorithm based on DPPs and show that it provides
 good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
-applied here, in \todo{Chapter double descent} an extension of the DPP-based design is analyzed in
+applied here, in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
 closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
 linear regression which closely matches empirical experiments.
-In \todo{Chapter random projections} we isolate the part of the proof involving concentration
+In \cref{ch:projections} we isolate the part of the proof involving concentration
 of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
@@
 importance sampling and variational inference.
-In \todo{Chapter LIC}, we consider parameterizing $q(x)$ by graph neural networks which condition each
+In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
 node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
@@
 the length of execution traces.
-When the target $p(x)$ is both multivariate and heavy tailed, \todo{Chapter FTVI}
+When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
 considers the problem of tail anisotropy through both a theoretical and practical perspective.
@@
 However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
-the tail parameter. In \todo{Chapter HTA}, we consider addressing this issue during static analysis of a
+the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
 probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of

### 2022-11-23 abstract.tex
Working compile

@@
 
-    In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
+    This dissertation considers a range of problems at the intersection of randomized methods
+    and statistics. Initially in the first three chapters, we develop novel analyses to recent problems
+    in experimental design, double descent, and random projections by using determinantal
+    point processes (DPP).
+    \Cref{ch:design} consider approximately optimal Bayesian experimental design
     using an adaptive row sampling algorithm based on DPPs and show that it provides
-    good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
-    applied here, in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
+    good approximations. In \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
     closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
@@
 
-    The second three chapters concern the setting of probabilistic programming, a computational tool
-    for specifying probabilistic models and automating inference through MCMC and VI. In particular, we
-    study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
-    importance sampling and variational inference.
-    In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
+    In the later chapters we focus on probabilistic programming and developing theory and tools
+    to automate statistical inference using randomized algorithms based on
+    Monte Carlo Markoc chain (MCMC) and variational inference (VI).
+    \Cref{ch:lic} considers parameterizing $q(x)$ by graph neural networks which condition each
     node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
@@
     considers the problem of tail anisotropy through both a theoretical and practical perspective.
-    We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
-    an anisotropic approximation (fat-tailed variational inference, FTVI) where
-    an anisotropic product base measures is pushed forwards through a bijective neural network,
-    and confirm that in practice FTVI improves both density estimation as well as variational inference.
     However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
     the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
-    probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
-    elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
-    such as sums and products. This enables a priori computation of tail parameters, which we show
-    improves the stability and convergence of a number of inference tasks.
+    probabilistic program's source code.
 

### 2022-11-23 design/aistats/bayesian.tex
Updates from Overleaf
Working compile

@@
 % \fi
-\documentclass[thesis.tex]{subfiles}
+\documentclass[../../thesis.tex]{subfiles}
 
@@
 % \begin{abstract}
-We establish a fundamental connection between Bayesian
+
+\chapter{Bayesian experimental design with regularized determinantal point processes}
+\label{ch:design}
+
+In this chapter, we establish a fundamental connection between Bayesian
 experimental design and determinantal point processes
@@
 of our algorithm significantly improves on existing approaches
-with comparable guarantees. 
+with comparable guarantees. Some of the results here were initially
+published in \fullcite{bayesian-experimental-design}.
 % \end{abstract}
@@
 \begin{align*}
-  \w\mid \y_S \ &\sim\ \Nc(\mub, \Sigmab), \\
-  \text{where } \mub &= (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
-  \Sigmab &= \sigma^2(\X_S^\top\X_S+\A)^{-1}.
+  \w\mid \y_S \      & \sim\ \Nc(\mub, \Sigmab),                 \\
+  \text{where } \mub & = (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
+  \Sigmab            & = \sigma^2(\X_S^\top\X_S+\A)^{-1}.
 \end{align*}
@@
 
-\textbf{Question 1:} \quad
+\begin{question}\label{q:bayesian:q1}
   Given $\X$, $f_{\A}$ and $k$, can we efficiently find a $(1+\epsilon)$-approximation for
   $\opt$?
-
-\textbf{Question 2:} \quad
+\end{question}
+
+\begin{question}\label{q:bayesian:q2}
   Given only $\Sigmab_{\X}$, $f_{\A}$ and $k$,
   what is the upper bound on $\opt$?
+\end{question}
 
@@
 
-  \paragraph{Answer to Question 1.}
+\paragraph{Answer to \Cref{q:bayesian:q1}}
 We propose an efficient $(1+\epsilon)$-approximation algorithm for
@@
 number of different methods can be used to solve the SDP relaxation (see
-Section \ref{s:experiments}). For
+Section \ref{s:bayesian:experiments}). For
 example, \cite{near-optimal-design} suggest using an iterative
@@
 \begin{figure*}[htbp]
-  \vspace{-2mm}
   \centering
-  \renewcommand{\arraystretch}{1.5}
   \begin{tabular}{r||c|c|l|l}
-  &Criteria&Bayesian& $k$ %& dep.~on $d$ & dep.~on $1/\epsilon$
-  & Cost after SDP\\%of rounding\\
+                              & Criteria             & Bayesian & $k$            %& dep.~on $d$ & dep.~on $1/\epsilon$
+                              & Cost after SDP                                   \\%of rounding\\
     \hline\hline
     \small  \cite{tractable-experimental-design}
-  &\small A,V&\xmark  &$d^2/\epsilon$
+                              & \small A,V           & \xmark   & $d^2/\epsilon$
     % &\Red{quadratic}& \Orange{linear}
-  &$n^2\cdot d$\\
+                              & $n^2\cdot d$                                     \\
     \small  \cite{near-optimal-design}
-  &\small A,C,D,E,G,V&\cmark
-&$ d/\epsilon^2$
+                              & \small A,C,D,E,G,V   & \cmark
+                              & $ d/\epsilon^2$
     % &\Orange{linear}& \Red{quadratic}
-  &$n\cdot kd^2$\\ 
+                              & $n\cdot kd^2$                                    \\
     \small  \cite{proportional-volume-sampling}
-  &\small A,D&\xmark &$d/\epsilon$
+                              & \small A,D           & \xmark   & $d/\epsilon$
     % & \Orange{linear}& \Orange{linear}
-  & $n^4\cdot k^2d$\\ 
+                              & $n^4\cdot k^2d$                                  \\
     \hline
-  \small\textbf{this paper}&\small A,C,D,V&\cmark
+    \small\textbf{this paper} & \small A,C,D,V       & \cmark
                               & $d_{\A}/\epsilon$
@@
   \end{tabular}
-%\vspace{2mm}
   \captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
@@
   \label{tab:comp}
-\vspace{-3mm}
 \end{figure*}
 
-\paragraph{Answer to Question 2.}
+\paragraph{Answer to \Cref{q:bayesian:q2}}
 %Formalizing the above intuitions,
@@
 Section \ref{s:related-work} for a~comparison). In Section
-  \ref{s:experiments}, we we provide numerical evidence suggesting
+\ref{s:bayesian:experiments}, we we provide numerical evidence suggesting
 that for many real datasets the quantity $f_{\A}(\frac
@@
 factor of 2.
+
+\subsection{Comparison of different effective dimensions}
 
@@
 Note that $d_{\frac nk\!\A}\leq d_{\A}$ and this gap can be very large for some
-problems (see discussion in Appendix~\ref{a:deff}).
-
-\subsection{Technical contributions}
-  To establish Theorems \ref{t:q2} and \ref{t:q1}, 
-  we develop a theoretical framework for a new sampling distribution
-  % $\DPPreg{p}(\X,\A)$, where 
-  % $p=(p_1,...,p_n)\in[0,1]^n$ is a vector of weights. 
-  which can be seen as a \emph{regularized} variant of a determinantal point
-  process (DPP). DPPs are a well-studied family of distributions
-with numerous applications in sampling diverse subsets
-  of negatively correlated elements \citep[see][]{dpp-ml}.
-
-  Given a psd matrix $\A$ and a weight vector $p=(p_1,...,p_n)\in[0,1]^n$, we
-define $\DPPreg{p}(\X,\A)$ as a distribution over subsets $S\subseteq[n]$ (of all
-sizes) such that (see Definition~\ref{d:r-dpp}):
-\begin{align*}
- \Pr(S)\propto \det(
-  \X_S^\top\X_S+\A)\ \cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in
-  S}(1-p_i).
-\end{align*}
-A number of regularized DPPs have been proposed recently
-\citep{dpp-intermediate,regularized-volume-sampling}, mostly within the context
-of Randomized Numerical Linear Algebra (RandNLA)~
-\citep{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR}.
-To our knowledge, ours is the first such definition that strictly falls
-under the umbrella of traditional DPPs \citep{dpp-ml}.
-We show this in Section \ref{s:r-dpp}, where we also prove that
-regularized DPPs can be decomposed into a low-rank DPP plus
-i.i.d. Bernoulli sampling (Theorem \ref{t:algorithm}). This decomposition reduces
-the sampling cost from $O(n^3)$ to $O(nd^2)$,
-and involves a more general result about DPPs defined via a
-correlation kernel (Lemma~\ref{l:decomposition}), which is of
-independent interest.
-
-In Section \ref{s:guarantees} we demonstrate a fundamental connection between an
-$\A$-regularized DPP and Bayesian experimental design with precision
-matrix $\A$. For simplicity of exposition, let the weight vector $p$ be uniformly equal $(\frac kn,...,\frac kn)$. If
-$S\sim\DPPreg{p}(\X,\A)$ and $f_{\A}$ is any one of the
-A/C/D/V-optimality criteria, then:
- \begin{align*}
-   \E\big[f_{\A}(\X_S^\top\X_S)\big]
-   \leq f_{\A}\big(\tfrac kn\Sigmab_{\X}\big)
-   \quad\text{and}\quad \E\big[|S|\big]
-   \leq d_{\frac nk\!\A}+k. 
- \end{align*}
-The proof of Theorem \ref{t:q1} relies on these two inequalities and a
-concentration bound for the subset size $|S|$,
-whereas to obtain Theorem
-\ref{t:q2} we additionally use the SDP relaxation to find the optimal
-weight vector~$p$. When $\A=\zero$, then $\DPPreg{p}(\X,\A)$ bears a
-lot of similarity to 
-\emph{proportional volume sampling} which is an (unregularized) determinantal
-distribution proposed by \cite{proportional-volume-sampling}.
-Our algorithm not only extends it to the Bayesian setting but also
-offers a drastic time complexity improvement from 
-the $O(n^4dk^2\log k)$ required by \cite{proportional-volume-sampling}
-down to the $O(nd^2)$ required for sampling from $\DPPreg{p}(\X,\A)$,
-and recent advances in RandNLA for DPP sampling
-\citep{leveraged-volume-sampling,correcting-bias,dpp-intermediate} suggest that
-$O(nd\log n + \poly(d))$ time is also possible.
-
-\section{Related work}
-\label{s:related-work}
-
-A number of works proposed $(1+\epsilon)$-approximation algorithms
-for experimental design which start with solving a convex
-relaxation of the problem, and then use some rounding strategy to
-obtain a discrete solution (see Table \ref{tab:comp} for comparison).
-In this line of work we wish to find the smallest $k$ for which a
-polynomial time approximation algorithm is possible. For example,
-\cite{tractable-experimental-design} gave an approximation algorithm
-for classical A/V-optimality with $k=\Omega(\frac{d^2}{\epsilon})$, where the
-rounding is done in a greedy fashion, and some randomized rounding
-strategies are also discussed. \cite{proportional-volume-sampling}
-suggested \emph{proportional volume sampling} for the rounding step
-and obtained approximation algorithms for classical A/D-optimality with
-$k=\Omega(\frac d\epsilon+\frac {\log1/\epsilon}{\epsilon^2})$. Their
-approach is particularly similar to ours (when $\A=\zero$). However,
-as discussed earlier, while their algorithms run in polynomial time,
-they scale very poorly with the number of experiments $n$ (see Table 
-\ref{tab:comp}). \cite{near-optimal-design} proposed  
-an efficient algorithm with a $(1+\epsilon)$-approximation guarantee for a wide
-range of optimality criteria, including A/C/D/E/V/G-optimality, both classical and Bayesian, when
-$k=\Omega(\frac d{\epsilon^2})$.  Our results (in Theorem \ref{t:q2}) improve on this work in
-two important ways: 
-\begin{itemize}
-  \item In terms of the dependence on $\epsilon$ for A/C/D/V-optimality,
-  \item In terms of the dependence on the dimension 
-  (by replacing $d$ with $d_{\A}$) in the Bayesian setting.
-\end{itemize}
-A lower bound shown by \cite{proportional-volume-sampling} implies that our
-Theorem \ref{t:q2} cannot be
-directly extended to E-optimality, but a similar lower bound does not
-exist for G-optimality. We remark that the approximation approaches
-relying on a convex relaxation can generally be converted to
-an upper bound on $\opt$ akin to our Theorem \ref{t:q1},
-however, unlike our bound, none of them apply to the regime of $k\leq d$.
-
-Non-trivial bounds for the \emph{classical} A-optimality criterion
-(i.e., $\opt$
-with $\A=\zero$) were first given by \cite{avron-boutsidis13}, where
-they show that
-for any $k\geq d$, $\opt \leq (1+ \frac{d-1}{k-d+1})\cdot f_{\zero}(\frac kn\Sigmab_\X)$
-and the subset $S$ attaining the bound can be found in polynomial time.
-The result was later extended \citep{unbiased-estimates,
-regularized-volume-sampling,unbiased-estimates-journal} to the case
-where $\A=\lambda\I$, proving that for any $k\geq d_{\lambda\I}$,
-we have
-$\opt\leq (1 +\frac{d_{\lambda\I}-1}{k-d_{\lambda\I}+1})\cdot f_{\frac
-  kn\!\lambda\I}(\frac kn\Sigmab_\X)$,
-and also a faster $O(nd^2)$ time algorithm was provided.
-In comparison, our results (in Theorem \ref{t:q1}) offer the following
-improvements for upper bounding $\opt$:
-\begin{itemize}
-  \item We cover a wider range of subset sizes, because
-    $d_{\frac nk\!\lambda\I}\leq d_{\lambda\I}$,
-  \item Our upper bound can be much tighter because
-    $f_{\lambda\I}(\frac kn\Sigmab_\X)\leq f_{\frac kn\!\lambda\I}(\frac
-    kn\Sigmab_\X)$. 
-\end{itemize}
-Additionally, \cite{minimax-experimental-design} propose a new notion of
-\emph{minimax} experimental design, which is related to A/V-optimality.  They
-also use a determinantal distribution for subset selection, however, due to
-different assumptions, their bounds are incomparable.
-
-Purely greedy approximation algorithms have been shown to provide
-guarantees in a number of special cases for experimental design. One
-example is classical D-optimality criterion, which can be converted to
-a submodular function \citep{submodularity-optimal-design}. Also, greedy algorithms
-for Bayesian A/V-optimality criteria have been considered by
-\cite{greedy-supermodular} and \cite{greedy-graph-sampling}. These
-methods can only provide a constant factor approximation guarantee
-(as opposed to $1+\epsilon$), and the factor is generally problem
-dependent (which means it could be arbitrarily large).
-Finally, a number of
-heuristics with good empirical performance have been proposed, such as Fedorov's
-exchange method \citep{cook1980comparison}. However, in this work
-we focus on methods that provide theoretical approximation guarantees.
-% In Section \ref{s:experiments} we experimentally show that
-% our regularized DPP sampling (without solving an SDP) performs
-% better than iid sampling methods like uniform and predictive length
-% sampling \cite{zhu2015optimal} and that by using the SDP solution
-% we can improve it to match the performance of
-% greedy algorithms for A-optimal design \cite{chamon2017approximate,greedy-supermodular}
-% which empirically have been found to perform well on real world data.
-
-\section{A new regularized determinantal point process}
-\label{s:r-dpp}
-  In this section we develop the theory for a novel regularized extension of
-  determinantal point processes (DPP) which we use as the sampling distribution
-  for obtaining guarantees in Bayesian experimental design.
-  DPPs form a family of distributions which are used to model repulsion between
-  elements in a random set, with many applications in machine learning
-  \citep{dpp-ml}. Here, we focus on the setting where we are sampling out of all
-  $2^n$ subsets $S\subseteq[n]$. Traditionally, a DPP is defined by a
-  correlation kernel, which is an $n\times n$ psd matrix $\K$ with eigenvalues
-  between 0 and 1, i.e., such that $\zero\preceq\K\preceq \I$. Given a
-  correlation kernel $\K$, the corresponding DPP is
-  defined as
-  \begin{align*}
-    S\sim\DPPcor(\K)\quad\text{iff}\quad
-    \Pr(T\subseteq S) = \det(\K_{T,T})\ \  \forall_{T\in[n]},
-  \end{align*}
-where $\K_{T,T}$ is the submatrix of $\K$ with rows and columns
-indexed by $T$. Another way of defining a DPP, popular in the machine
-learning community, is via an ensemble kernel $\L$. Any psd matrix
-$\L$ is an ensemble kernel of a DPP defined as:
-  \begin{align*}
-    S\sim\DPPens(\L)\quad\text{iff}\quad \Pr(S) \propto \det(\L_{S,S}).\quad~
-  \end{align*}
-Crucially, every $\DPPens$ is also a $\DPPcor$, but not the other way
-around. Specifically, $\DPPens(\L)=\DPPcor(\K)$ when:
-\begin{align*}
-  &\text{(a)}\ \  \L=\K(\I-\K)^{-1},\qquad
-  &\text{(b)}\ \  \K = \I -  (\I+\L)^{-1},
-\end{align*}
-but (a) requires that $\I-\K$ be invertible which is not true
-for some DPPs.
-(This will be important in our analysis.)
-The classical algorithm for sampling from a DPP requires the
-eigendecomposition of either matrix $\K$ or $\L$, which in general
-costs $O(n^3)$, followed by a sampling procedure which costs
-$O(n\,|S|^2)$ \cite{dpp-independence,dpp-ml}.
-
-We now define our regularized DPP and describe its
-connection with correlation and ensemble~DPPs.
-\begin{definition}\label{d:r-dpp}
-Given matrix $\X\in\R^{n\times d}$, a sequence  $p=(p_1,\dots,p_n)\in[0,1]^n$
-and a psd matrix $\A\in\R^{d\times d}$ such that
-$\sum_ip_i\x_i\x_i^\top\!+\A$ is
-full rank, let
-$\DPPreg{p}(\X,\A)$ be a distribution over $S\subseteq [n]$:
-  \begin{align}
-  \Pr(S) = \frac{\det(
-  \X_S^\top\X_S+\A)}{\det\!\big(\!
- \sum_ip_i\x_i\x_i^\top\!+ \A\big)}
-\cdot \!\prod_{i\in S}p_{i}\cdot\!\prod_{i\not\in S}(1\!-\!p_i).\label{eq:poisson-prob}
-\end{align}
-%   \Pr(\pi)\ \propto\ \det\!\big(\A +
-%   \X_{\pi}^\top\X_{\pi}\big)
-%   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
-% \end{align*}
-\end{definition}
-The fact that this is a proper distribution (i.e., that it sums to
-one) can be restated as a determinantal expectation formula: if
-$b_i\sim\mathrm{Bernoulli}(p_i)$ are independent Bernoulli random
-variables, then
-\begin{align*}
-  &\sum_{S\subseteq [n]}\!\det(
-  \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
-  &\!=
-  \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
-  \det\!\Big(\sum_i\E[b_i]\x_i\x_i^\top\!+\A\Big),
-\end{align*}
-where $(*)$ follows  from Lemma 7 of \citet{determinantal-averaging}.
-% follows from a recently shown determinantal expectation formula.
-% \begin{lemma}[\cite{determinantal-averaging}]\label{l:cb}
-% For $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$
-% and a random set $S\subseteq [n]$ drawn with independent Bernoullis,
-% \begin{align*}
-%   \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)\Big]
-%   = \det\!\Big( \E\big[\X_S^\top\X_S\big]+\A\Big).
-% \end{align*}
-% \end{lemma}
-
-  The main theoretical contribution in this section is the
-  following efficient algorithm for $\DPPreg{p}(\X,\A)$ which reduces it to
-  sampling from a correlation DPP and unioning with i.i.d.~Bernoulli samples:
-\begin{theorem}\label{t:algorithm}
-For any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
-s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
-full rank, let
-\begin{align*}
-  T&\sim\DPPcor\big(\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big),
-  \quad\\
-  &\quad\text{where}\quad\D_p=\diag(p).
-\end{align*}
-If $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random variables, then
-$T\cup \{i:b_i\!=\!1\}\sim \DPPreg{p}(\X,\A)$.
-\end{theorem}
-
-\begin{remark}
-    Figure~\ref{fig:algorithm-pseudocode} illustrates how to exploit this result
-    to build an efficient sampling algorithm.
-  Since the correlation kernel matrix has rank at most $d$, the preprocessing
-  cost of eigendecomposition is $O(nd^2)$. Then, each sample costs only $O(n\,|T|^2)$.
-\end{remark}
-
-We prove the theorem in three steps. 
-First, we express $\DPPreg{p}(\X,\A)$ as an ensemble DPP, which requires some
-additional assumptions on $\A$ and $p$ to be possible.
-Then, we convert the ensemble to a correlation kernel (eliminating the extra
-assumptions),
-and finally show that this kernel can be decomposed into a rank $d$ kernel plus
-Bernoulli sampling.
-In the process, we establish several novel theoretical properties
-regarding the representation, decomposition, and closure properties
-of regularized DPPs which may be of independent interest.
-
-\renewcommand{\thealgorithm}{}
-\floatname{algorithm}{}
-\vspace{-3mm}
-\begin{figure}[htbp]
-\begin{center}
-  \begin{algorithm}[H]
-    {\fontsize{9}{9}\selectfont
-      \caption{\bf \small Sampling \ $S\sim\DPPreg{p}(\X,\A)$}
-      \begin{algorithmic}[0]
-        \STATE \textbf{Input:} $\X\!\in\!\R^{n\times d}\!,\text{ psd } \A\!\in\!\R^{d\times
-        d}\!, p\!\in\![0,1]^n$\\[1mm]
-        \STATE Compute $\Z \leftarrow \A+\X^\top\D_p\X$\\[1mm]
-        \STATE Compute \textsc{SVD} of  $\B=\D_p^{\sfrac12}\X\Z^{-\sfrac12}$\\[1mm]
-        \STATE Sample $T\sim\DPPcor(\B\B^\top)$ \hfill \citep{dpp-independence}\\[1mm]
-        \STATE Sample $b_i\sim\mathrm{Bernoulli}(p_i)$ for $i\in[n]$\\[1mm]
-        \RETURN $S=T\cup\{i:b_i=1\}$
-      \end{algorithmic}
-    }
-  \end{algorithm}
-\end{center}
-\vspace{-3mm}
-\caption{Algorithm which exploits
-  Theorem~\ref{t:algorithm} to sample $S \sim \DPPreg{p}(\X, \A)$
-  in $O( nd^2 )$ time.}
-\label{fig:algorithm-pseudocode}
-\end{figure}
-
-\vspace{2mm}
-\begin{lemma}\label{t:reduction}
-  Given $\X$, $\A$ and $\D_p$ as in Theorem \ref{t:algorithm}, assume that $\A$ and $\I-\D_p$ are
-  invertible. Then,
-  \begin{align*}
-    \DPPreg{p}(\X,\A)&=\DPPens\big(\Dbt+
-    \Dbt^{\sfrac12}\X\A^{-1}\X^\top
-    \Dbt^{\sfrac12}\big),\quad\\
-    \text{where}
-    \quad\Dbt &= \D_p(\I-\D_p)^{-1}.
-  \end{align*}
-\end{lemma}
-\begin{proof}
-  Let $S\sim \DPPreg{p}(\X,\A)$. By Definition \ref{d:r-dpp} and
-  the fact that $\det(\A\B+\I)=\det(\B\A+\I)$,
-  \begin{align*}
-    \Pr(S) &\propto \det(\X_S^\top\X_S+\A)\cdot
-\prod_{i\in S}p_i\cdot\prod_{i\not\in S}(1-p_i) \\
-&= \det(\X_S^\top\X_S+\A)\cdot
-\prod_{i\in S}\frac{p_i}{1-p_i}\cdot\prod_{i=1}^n(1-p_i)
-\\ &\propto\det\!\big(\A(\A^{-1}\X_S^\top\X_S+\I)\big) \det(\Dbt_{S,S}) \\
-     &=\det(\A)\det(\A^{-1}\X_S^\top\X_S+\I) \det(\Dbt_{S,S})
-\\ &\propto \det(\X_S \A^{-1}\X_S^\top+\I) \det(\Dbt_{S,S}) \\
-&=\det\!\Big(\big[\Dbt^{\sfrac12}\X \A^{-1}\X^\top\Dbt^{\sfrac12}+\Dbt\big]_{S,S}\Big),
-  \end{align*}
-which matches the definition of the L-ensemble DPP.
-\end{proof}
- At this point, to sample from  $\DPPreg{p}(\X,\A)$, we could simply
-invoke any algorithm for sampling from
-an ensemble DPP. However, this would only work for invertible
-$\A$, which in particular excludes the important case of
-$\A=\zero$ corresponding to classical experimental
-design. Moreover, the standard algorithm would require computing the
-eigendecomposition of the ensemble kernel, which (at
-least if done na\"ively) costs $O(n^3)$. Even after this is done, the
-sampling cost would still be $O(n\,|S|^2)$ which can be considerably
-more than $O(nd^2)$. We first address the issue of invertibility of matrix
-$\A$ by expressing our distribution via a correlation DPP.
-\begin{lemma}\label{l:correlation}
-  Given $\X$, $\A$, and $\D_p$ as in Theorem \ref{t:algorithm} (without
-  any additional assumptions), we have
-  \begin{align*}
-    &\DPPreg{p}(\X,\A) 
-    = \DPPcor\big(\D_p + \\
-    &\ (\I\!-\!\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A\!+\!\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}(\I\!-\!\D_p)^{\sfrac12}\big).
-    \end{align*}
-  \end{lemma}
-  When $\A$ and $\I-\D_p$ are invertible, then the proof (given in
-  Appendix \ref{a:proofs}) is a straightforward calculation.
-Then, we use a limit argument with $p_\epsilon=(1-\epsilon)p$ and
-$\A_\epsilon=\A+\epsilon\I$, where $\epsilon\rightarrow 0$.
-
-Finally, we show that the correlation DPP arrived at in Lemma
-\ref{l:correlation} can be decomposed into a smaller DPP plus
-Bernoulli sampling. In fact, in the following lemma we obtain a more
-general recipe for combining DPPs with Bernoulli sampling, which may
-be of independent interest. Note that if $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random
-variables then $\{i:b_i\!=1\!\}\sim\DPPcor(\D_p)$.
-\begin{lemma}\label{l:decomposition}
-  Let $\K$ and $\D$ be $n\times n$ psd matrices with eigenvalues between
-0 and 1, and assume that $\D$ is diagonal. If\, $T\!\sim\DPPcor(\K)$ and
-$R\sim\DPPcor(\D)$, then
-\begin{align*}T\cup R\sim \DPPcor\big(\D+(\I-\D)^{\sfrac12}\K
-  (\I-\D)^{\sfrac12}\big).
-  \end{align*}
-\end{lemma}
-The lemma is proven in Appendix \ref{a:proofs}. Theorem
-\ref{t:algorithm} now follows by combining Lemmas \ref{l:correlation} and \ref{l:decomposition}.
-
-\section{Guarantees for Bayesian experimental design}
-\label{s:guarantees}
-In this section we prove our main results regarding Bayesian
-experimental design (Theorems \ref{t:q2} and \ref{t:q1}). First, we
-establish certain properties of the regularized DPP distribution that
-make it effective in this setting. Even though the size of the
-sampled subset $S\sim\DPPreg{p}(\X,\A)$ is random and can be as large
-as $n$, it is also highly concentrated around its expectation, which
-can be bounded in terms of the $\A$-effective dimension. This is
-crucial, since both of our main results require a subset of
-deterministically bounded size. Recall that the effective dimension is
-defined as a function $d_{\A}(\Sigmab) =
-\tr\big(\Sigmab(\A+\Sigmab)^{-1}\big)$. The omitted proofs are in
-Appendix \ref{a:proofs}.
-\begin{lemma}\label{l:size}
-  Given any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
-s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
-full rank, let $S=T\cup \{i:b_i=1\}\sim \DPPreg{p}(\X,\A)$ be defined
-as in Theorem \ref{t:algorithm}. Then
-\begin{align*}
-  \E\big[|S|\big] &\leq \E\big[|T|\big] + \E\Big[\sum_i b_i\Big]  \\
-                  & = d_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big) + \sum_i p_i.
-\end{align*}
-\end{lemma}
-Next, we show two expectation inequalities for the matrix inverse and
-matrix determinant, which hold for the regularized DPP.
-We use them to bound the Bayesian optimality criteria in expectation.
-% These are
-% immediately followed by a corollary applying them to bound
-% Bayesian optimality criteria.
-\begin{lemma}\label{t:expectations}
-Whenever $S\sim \DPPreg{p}(\X,\A)$ is a well-defined distribution it holds that
-\begin{align}
-  \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &\preceq \Big(\!
-    \sum_ip_i\x_i\x_i^\top \!+\A \Big)^{-1}\!\!,\label{eq:sqinv}
-  \\
-\E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &\leq\det\!\Big(\!
-    \sum_ip_i\x_i\x_i^\top\!+\A \Big)^{\!-1}\!\!.\!\!\label{eq:det}
-\end{align}
-\end{lemma}
-\begin{corollary}\label{c:expectations}
-Let $f_{\A}$ be A/C/D/V-optimality. Whenever $S\sim\DPPreg{p}(\X,\A)$ is
-well-defined,
-\begin{align*}
-  \E\big[f_{\A}(\X_S^\top\X_S)\big] \leq f_{\A}\Big(\sum_i p_i\x_i\x_i^\top\Big).
-\end{align*}
-\end{corollary}
-\begin{proof}
-  In the case of A-, C-, and V-optimality, the function $f_{\A}$ is a
-  linear transformation of the matrix $(\X_S^\top\X_S+\A)^{-1}$ so the
-  bound follows from \eqref{eq:sqinv}. For D-optimality, we apply
-  \eqref{eq:det} as follows:
-  \begin{align*}
-    \E\big[f_{\A}(\X_S^\top\X_S)\big]
-    & = \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-\sfrac1d}\Big] \\
-    &\leq
-     \E\bigg[\Big(\det\!\big(\X_S^\top\X_S+\A\big)^{-\sfrac1d}\Big)^d\bigg]^{\sfrac1d} \\
-    \\ &=  \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]^{\sfrac1d}\\
-       &\leq
-         \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{-\sfrac1d},
-  \end{align*}
-  which completes the proof.
-\end{proof}
-Finally, we present the key lemma that puts everything together. This
-result is essentially a generalization of Theorem \ref{t:q1} from
-which also follows Theorem \ref{t:q2}.
-\begin{lemma}\label{l:guarantees}
-  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be $n\times
-  d$.  For some $w=(w_1,\dots,w_n)\in[0,1]^n$, let
-  $\Sigmab_w=\sum_iw_i\x_i\x_i^\top$ and assume that
-  $\sum_i w_i=k\in[n]$. If $k\geq 4\,d_{\A}(\Sigmab_w)$, then a
-subset $S\subseteq[n]$ of size $k$ can be found in $O(ndk+k^2d^2)$ time that satisfies
-  \begin{align*}
-    &f_{\A}\big(\X_S^\top\X_S\big)  \\
-    &\leq \bigg(\ 1+8\,\frac{d_{\A}(\Sigmab_w)}{k} + 8\sqrtshort{\frac{\ln (k/d_{\A}(\Sigmab_w))}{k}}\ \bigg)\cdot f_{\A}\big(\Sigmab_w\big).
-  \end{align*}
-  \end{lemma}
-\begin{proof}
-  Let $p=(p_1,\dots,p_n)$ be defined so that $p_i =
-  \frac{w_i}{1+\epsilon}$, and suppose that $S\sim
-  \DPPreg{p}(\X,\A)$. Then, using Corollary \ref{c:expectations}, we have
-  \begin{align*}
-\Pr&\big(|S|\leq k\big)\,\E\Big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\Big]\\
-&\leq
-\E\big[f_{\A}(\X_S^\top\X_S)\big] \\
-&\leq
-f_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big)
-\\ &\leq (1+\epsilon)\cdot f_{\A}\Big(\sum_iw_i\x_i\x_i^\top\Big).
-  \end{align*}
-  Using Lemma \ref{l:size} we can bound the expected size of $S$ as
-follows:
-\begin{align*}
-  \E\big[|S|\big]
-  &\leq d_{\A}(\Sigmab_w) + \sum_i p_i \\
-  &=d_{\A}(\Sigmab_w)+\frac{k}{1+\epsilon}\\
-  &=k\cdot \Big(1+\frac {d_{\A}(\Sigmab_w) }k - \frac{\epsilon}{1+\epsilon}\Big).
-\end{align*}
-Let $d_w = d_{\A}(\Sigmab_w)$ and $\alpha=1+\frac {d_w} k -\frac{\epsilon}{1+\epsilon}$.
-If $1\geq\epsilon\geq \frac{4d_w}{k}$, then $\alpha\leq 1 +
-\frac\epsilon4-\frac\epsilon2=1-\frac\epsilon4$. Since $\DPPreg{p}(\X,\A)$
-is a  determinantal point process, $|S|$ is a
-Poisson binomial r.v.~so for $\epsilon\geq 6\sqrt{\frac{\ln(k/d_w)}{k}}$,
-\begin{align*}
-  \Pr(|S|>k) \leq \ee^{-\frac{(k-\alpha k)^2}{2k}} = \ee^{-\frac
-  k2(1-\alpha)^2}\leq \ee^{-\frac{k\epsilon^2}{32}}\leq \frac{d_w}k.
-\end{align*}
-% If $k\geq\frac{4 d_{\A}(\Sigmab_w)}\epsilon$,
-% then $\alpha \leq 1 + \frac\epsilon4 -\frac\epsilon2=1-\frac\epsilon4$.
-%  Since $\Vol_p(\X,\A)$ is a determinantal point process, Lemma
-% \ref{l:concent} implies that for $\epsilon'=\frac\epsilon4$ and $k\geq
-%  48\frac{\ln 1/\epsilon'}{\epsilon'^2}$ we have
-% \begin{align*}
-% \Pr(|S|> k)\leq  \Pr\big(|S|\ \geq\ \alpha k + \epsilon' k\big)\leq
-%   3\ee^{-\frac{\epsilon'^2k^2}{16(\epsilon'k+2\alpha k)}} \leq
-%   3\ee^{-\frac{\epsilon'^2k}{48}}\leq 3\epsilon'.
-% \end{align*}
-For any $\epsilon\geq
-4\,\frac{d_w}k + 6\sqrt{\frac{\ln(k/d_w)}{k}}$, we have
-\begin{align*}
-\E&\big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\big] \\
-&\leq \frac{1+\epsilon}{1-\frac{d_w}k}\cdot f_{\A}(\Sigmab_w)\\
-  &\leq
-  \bigg(1+\frac{\epsilon+\frac{d_w}k}{1-\frac{d_w}k}\bigg)\cdot
-f_{\A}(\Sigmab_w)
-\\ &\leq \bigg(1+7\,\frac{d_w}{k} +
-  8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\cdot f_{\A}(\Sigmab_w).
-\end{align*}
-Denoting $\E\big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\big]$ as $F_k$,
-Markov's inequality implies that
-\[\Pr\Big(f_{\A}(\X_S^\top\X_S) \geq
-  (1+\delta) F_k\ \mid\ |S|\leq k\Big)\leq \frac1{1+\delta}.\]
-Also, we showed
-that $\Pr(|S|\leq k)\geq 1-\frac{d_w}{k}\geq \frac34$. Setting $\delta=\frac{d_w}{C
-  k}$ for sufficiently large $C$ we obtain that with probability
-$\Omega(\frac{d_w}{k})$, the random set $S$ has size at most $k$ and
-\begin{align*}
-f_{\A}&(\X_S^\top\X_S) \\
-&\leq \bigg(1+\frac{d_w}{C k}\bigg)\!\cdot\! \bigg(1+7\,\frac{d_w}{k} +
-  8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\!\cdot\! f_{\A}(\Sigmab_w)\\
-&\leq
-\bigg(1+8\,\frac{d_w}{k} +
-  8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\cdot f_{\A}(\Sigmab_w).
-\end{align*}
-We can sample from $\DPPreg{p}(\X,\A)$ conditioned on
-$|S|\leq k$ and $f_{\A}(\X_S^\top\X_S)$ bounded as above by rejection
-sampling. When $|S|<k$,  the set is completed to
-$k$ with arbitrary indices. On average, $O(\frac k{d_w})$ samples from
-$\DPPreg{p}(\X,\A)$ are needed, so the cost is $O(nd^2)$ for the
-eigendecomposition, $O(\frac{k}{d_w}\cdot nd_w^2)=O(nd_wk)$ for
-sampling and $O(\frac k{d_w}\cdot kd^2)$ for recomputing $f_{\A}(\X_S^\top\X_S)$.
-\end{proof}
-To prove the main results, we use Lemma \ref{l:guarantees} with
-appropriately chosen weights $w$.
-\begin{proofof}{Theorem}{\ref{t:q2}}
-  As discussed by \cite{near-optimal-design} and
-  \cite{boyd2004convex}, the following convex 
-relaxation of experimental design can be written as a semi-definite
-program and solved using standard SDP solvers:
-\begin{align}
-  &w^* \ = \  \argmin_w\ \
-  f_{\A}\Big(\sum_{i=1}^nw_i\x_i\x_i^\top\Big),\quad\\
-  &\text{subject to}\quad
-  \forall_i\ \ 0\leq w_i\leq 1,\quad \sum_i w_i=k.\label{eq:sdp}
-\end{align}
-The solution $w^*$ satisfies $f_{\A}\big(\Sigmab_{w^*}\big)\leq
-\opt$. If we use $w^*$ in Lemma \ref{l:guarantees}, then observing
-that $d_{\A}(\Sigmab_{w^*})\leq d_{\A}$, and setting $k\geq C(\frac
-{d_{\A}}{\epsilon} + \frac{\log1/\epsilon}{\epsilon^2})$ for
-sufficiently large $C$, the algorithm in the lemma finds subset $S$
-such that
-\[f_{\A}(\X_S^\top\X_S)\leq (1+\epsilon)\cdot
-f_{\A}(\Sigmab_{w^*})\leq
-(1+\epsilon)\cdot\opt.\]
-% \cite{tractable-experimental-design} show that
-% the support of $w^*$ has size at most $k+d^2$, which reduces the cost of
-% our sampling to $O(k^2d^2)$.
-Note that we did not need to
-solve the SDP exactly, so approximate solvers could be used instead.
-\end{proofof}
-\begin{proofof}{Theorem}{\ref{t:q1}}
-Let $w=(\frac kn,...,\frac kn)$ in Lemma~\ref{l:guarantees}. Then, we have
-$\Sigmab_w=\frac kn\Sigmab_{\X}$ and also $d_{\A}(\Sigmab_w) =
-d_{\frac nk\!\A}$. Since for any set $S$ of size $k$, we
-have $\opt\leq f_{\A}(\X_S^\top\X_S)$, the result follows.
-\end{proofof}
-
-\input{bayesian_experiments.tex}
-
-\section{Conclusions}
-
-We proposed a new algorithm for finding
-$(1+\epsilon)$-approximate Bayesian experimental designs by leveraging 
-a fundamental connection with determinantal point processes. Compared
-to the state-of-the-art approaches, our
-method provides stronger theoretical guarantees in terms of the allowed
-range of subset sizes, as well as
-offering significantly better time complexity guarantees. At the same
-time, our experiments show that on the task of A-optimal design the
-proposed algorithm performs as well as or better than several methods that
-are used in practice.
-
-\subsubsection*{Acknowledgements}
-
-MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
-  support of this work. Also, MWM and MD thank the NSF TRIPODS program
-    and FL thanks the NPSC for funding.  The authors thank Uthaipon
-  T.~Tantipongpipat for valuable discussions.
-
-% \bibliographystyle{apalike}
-% \bibliography{../pap}
-
-% \newpage
-% \appendix
-
-% \onecolumn
-
-\section{Properties of regularized DPPs}
-\label{a:proofs}
-In this section we provide proofs omitted from Sections \ref{s:r-dpp}
-and \ref{s:guarantees}. We start with showing the fact that the
-regularized DPP distribution $\DPPreg{p}(\X,\A)$ is a correlation DPP.
-\begin{lemma}[restated Lemma \ref{l:correlation}]
-  Given $\X$, $\A$, and $\D_p$ as in Theorem \ref{t:algorithm}, we have
-  \begin{align*}
-    \DPPreg{p}(\X,\A)= \DPPcor\big(\D_p +
-    (\I-\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}(\I-\D_p)^{\sfrac12}\big).
-    \end{align*}
-  \end{lemma}
-  \begin{proof}
-    First, we show this under the invertibility assumptions of Lemma
-    \ref{t:reduction}, i.e., given that $\A$ and $\I-\D_p$ are
-    invertible. In this case $\DPPreg{p}(\X,\A)=\DPPens(\L)$, where
-    \begin{align}\L=\Dbt+
-    \Dbt^{\sfrac12}\X\A^{-1}\X^\top
-      \Dbt^{\sfrac12}\quad\text{ and }\quad\Dbt=\D_p(\I-\D_p)^{-1}.\label{eq:ensemble}
-\end{align}
-Converting this to
-    a correlation kernel $\K$ and denoting $\Xt=\D_p^{\sfrac12}\X$, we obtain
-    \begin{align*}
-      \K&=\I-(\I+\L)^{-1}
-\\      &=\I - \big(\I + (\I-\D_p)^{-1}\D_p +
-(\I-\D_p)^{-\sfrac12}\Xt\A^{-1}\Xt^\top(\I-\D_p)^{-\sfrac12}\big)^{-1}
-\\ &= \I - \big((\I-\D_p)^{-1} +
-     (\I-\D_p)^{-\sfrac12}\Xt\A^{-1}\Xt^\top(\I-\D_p)^{-\sfrac12}\big)^{-1}
-\\ &=\I -
-     (\I-\D_p)^{\sfrac12}(\I+\Xt\A^{-1}\Xt^\top)^{-1}(\I-\D_p)^{\sfrac12}
-      \\
-      &\overset{(*)}{=}\I -
-     (\I-\D_p)^{\sfrac12}\big(\I-\Xt\A^{-\sfrac12}
-     (\I+\A^{-\sfrac12}\Xt^\top\Xt\A^{-\sfrac12})^{-1}
-     \A^{-\sfrac12}\Xt^\top\big)(\I-\D_p)^{\sfrac12}
-      \\
-      &=\I - (\I-\D_p) +
-     (\I-\D_p)^{\sfrac12}\Xt(\A+\Xt^\top\Xt)^{-1}\Xt^\top(\I-\D_p)^{\sfrac12}
-\\ &=\D_p + (\I-\D_p)^{\sfrac12}\Xt(\A+\Xt^\top\Xt)^{-1}\Xt^\top(\I-\D_p)^{\sfrac12},
-    \end{align*}
-    where $(*)$ follows from Fact 2.16.19 in
-    \cite{matrix-mathematics}. Note that converting from $\L$ to $\K$
-    got rid of the inverses $\A^{-1}$  and $(\I-\D_p)^{-1}$ appearing
-    in \eqref{eq:ensemble}. The intuition 
-    is that when $\A$ or $\I-\D_p$ is non-invertible, then
-    $\DPPreg{p}(\X,\A)$ is not an L-ensemble but it is still a
-    correlation DPP. To show this, we use a limit argument. For
-    $\epsilon\in[0,1]$, let $p_\epsilon=(1-\epsilon)p$ and
-    $\A_\epsilon=\A+\epsilon\I$. Observe that if $\epsilon>0$ then $\A_\epsilon$ and
-    $\I-\D_{p_{\epsilon}}$ are always invertible even if $\A$ and
-    $\I-\D_p$ are not. Denote $\K_\epsilon$ as the
-    above correlation kernel with $p$ replaced by $p_{\epsilon}$ and
-    $\A$ replaced by $\A_{\epsilon}$. Note that all matrix operations
-    defining kernel $\K_\epsilon$ are continuous w.r.t. $\epsilon\in[0,1]$, including the inverse, since
-    $\A+\Xt^\top\Xt$ is assumed to be invertible. Therefore, the
-    following equalities hold (with limits taken point-wise and $\epsilon>0$):
-    \begin{align*}
-      \DPPreg{p}(\X,\A) = \lim_{\epsilon\rightarrow
-      0}\DPPreg{p_\epsilon}(\X,\A_{\epsilon}) =
-      \lim_{\epsilon\rightarrow 0}\DPPcor(\K_\epsilon) = \DPPcor(\K),
-    \end{align*}
-where we did not have to assume invertibility of $\A$ or $\I-\D_p$.
-\end{proof}
-We now prove a lemma about combining a determinantal point process
-with Bernoulli sampling, which itself is a DPP with a diagonal correlation
-kernel.
-\begin{lemma}[restated Lemma \ref{l:decomposition}]
-  Let $\K$ and $\D$ be $n\times n$ psd matrices with eigenvalues between
-0 and 1, and assume that $\D$ is diagonal. If\, $T\!\sim\DPPcor(\K)$ and
-$R\sim\DPPcor(\D)$, then
-\begin{align*}T\cup R\sim \DPPcor\big(\D+(\I-\D)^{\sfrac12}\K
-  (\I-\D)^{\sfrac12}\big).
-  \end{align*}
-\end{lemma}
-\begin{proof}
-For this proof we will use the shorthand $\K_A$ for $\K_{A,A}$. If
-$\D$ has no zeros on the diagonal then $\det(\D_A)>0$ for all
-$A\subseteq[n]$ and
-  \begin{align*}
-    \Pr(A \subset T \cup R)
-    &=\sum_{B\subset A}\Pr(R\cap A=A\setminus B) \ \Pr(B\subseteq T)\\
-    &=\sum_{B\subset A}\det(\D_{A\setminus
-      B})\det\!\big([\I-\D]_B\big)\,\det(\K_B) \\
-    &=\sum_{B\subset A}\det(\D_{A\setminus
-      B})\det\!\Big(\big[(\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\big]_B\Big) \\
-    &=\det(\D_A) \sum_{B\subset A}\det\!\Big(\big[\D^{-\sfrac12}
-      (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\D^{-\sfrac12}\big]_{B}\Big)\\[-1mm]
-    &\overset{(*)}{=}\det(\D_A)\det\!\Big(\I + \big[\D^{-\sfrac12}
-      (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\D^{-\sfrac12}\big]_A\Big)\\
-    &=\det\!\Big(\big[\D+ (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\big]_A\Big),
-  \end{align*}
-  where $(*)$ follows from a standard determinantal identity used to
-  compute the L-ensemble partition function
-  \cite[Theorem~2.1]{dpp-ml}. If $\D$ has zeros on the diagonal, a
-  similar limit argument as in Lemma \ref{l:correlation} with
-  $\D_\epsilon=\D+\epsilon\,\I$ holds.
-\end{proof}
-Next, we give a bound on the expected size of a regularized DPP.
-\begin{lemma}[restated Lemma \ref{l:size}]
-  Given any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
-  s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
-  full rank, let $S=T\cup \{i:b_i=1\}\sim \DPPreg{p}(\X,\A)$ be defined
-as in Theorem \ref{t:algorithm}. Then
-\begin{align*}
-\E\big[|S|\big]\leq \E\big[|T|\big] + \E\Big[\sum_i b_i\Big] =
-  d_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big) + \sum_i  p_i.
-\end{align*}
-\end{lemma}
-\begin{proof}
-For correlation kernels it is known that the expected size of $\DPPcor(\K)$
-is $\tr(\K)$. Thus, using $\D_p=\diag(p)$, we can invoke Lemma \ref{l:correlation} to obtain
-\begin{align*}
-  \E\big[|S|\big]
-  &= \tr\big(\D_p +
-  (\I-\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}
-                    \X^\top\D_p^{\sfrac12}(\I-\D_p)^{\sfrac12}\big)
-  \\ &\leq \tr(\D_p) +\tr\big(\D_p^{\sfrac12}
-       \X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big)
-\\ &=\tr(\D_p) +\tr\big(\X^\top\D_p\X (\A+\X^\top\D_p\X)^{-1}\big) =
-     \tr(\D_p) + d_{\A}(\X^\top\D_p\X),
-\end{align*}
-from which the claim follows.
-\end{proof}
-Next, we show two expectation inequalities for the matrix inverse and
-matrix determinant.
-\begin{lemma}[restated Lemma \ref{t:expectations}]
-Whenever $S\sim \DPPreg{p}(\X,\A)$ is a well-defined distribution it holds that
-\begin{align}
-  \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &\preceq \Big(
-    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1},\label{eq:sqinv2}
-  \\
-\E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &\leq\det\!\Big(
-    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1}.\label{eq:det2}
-\end{align}
-\end{lemma}
-\begin{proof}
-For a square matrix $\M$, define its adjugate, denoted $\adj(\M)$, as a matrix
-whose $i,j$-th entry is $(-1)^{i+j}\det(\M_{-j,-i})$, where
-$\M_{-j,-i}$ is the matrix $\M$ without $j$th row and $i$th column. If
-$\M$ is invertible, then $\adj(\M) = \det(\M)\M^{-1}$. Now, let
-$b_i\sim\mathrm{Bernoulli}(p_i)$ be independent random variables. As seen
-in previous section, the identity
-$\E[\det(\sum_ib_i\x_i\x_i^\top+\A)]=\det(\sum_ip_i\x_i\x_i^\top+\A)$ gives
-us the normalization constant for $\DPPreg{p}(\X,\A)$. Moreover, as
-noted in a different context by \cite{determinantal-averaging}, when
-applied entrywise to the adjugate matrix, this identity implies that
-$\E[\adj(\sum_ib_i\x_i\x_i^\top+\A)]=\adj(\sum_ip_i\x_i\x_i^\top+\A)$. Let
-$\Ic$ denote the set of all subsets $S\subseteq [n]$ such that
-$\X_S^\top\X_S+\A$ is invertible. We have
-\begin{align*}
-  \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &= \sum_{S\in\Ic}\big(\X_S^\top\X_S+\A\big)^{-1}
-\frac{\det(\X_S^\top\X_S+\A)}{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
-    S}p_i\prod_{i\not\in S}(1-p_i)
-\\ &=\sum_{S\in\Ic}\frac{\adj(\X_S^\top\X_S+\A\big)}
-{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
-     S}p_i\prod_{i\not\in S}(1-p_i)
-\\ &\preceq \sum_{S\subseteq[n]}\frac{\adj(\X_S^\top\X_S+\A\big)}
-{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
-     S}p_i\prod_{i\not\in S}(1-p_i)
-\\ &=\frac{\E\big[\!\adj(\sum_i
-     b_i\x_i\x_i^\top+\A)\big]}{\det(\sum_ip_i\x_i\x_i^\top+\A)}
-  \\
-  &=\frac{\adj(\sum_ip_i\x_i\x_i^\top+\A)}{\det(\sum_ip_i\x_i\x_i^\top+\A)}
-    = \Big(\sum_ip_i\x_i\x_i^\top+\A\Big)^{-1}.
-\end{align*}
-Note that if $\Ic$ contains all subsets of $[n]$, for example when
-$\A\succ\zero$, then the inequality turns into equality. Thus, we
-showed \eqref{eq:sqinv2}, and \eqref{eq:det2} follows even more easily:
-\begin{align*}
-  \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
-  &= \sum_{S\in\Ic}
-\frac{1}{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
-    S}p_i\prod_{i\not\in S}(1-p_i)
-\leq \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{\!-1}\!,
-\end{align*}
-where the equality holds if $\Ic$ consists of all subsets of $[n]$.
-\end{proof}
-
-\section{Comparison of different effective dimensions}
-\label{a:deff}
-In this section we compare the two notions of effective dimension for
-Bayesian experimental design considered in this work. Here, we let
-$\X$ be the full $n\times d$ design matrix and use $k$ to denote the
-desired subset size. Recall that the effective dimension is  defined
-as a function of the data covariance matrix $\Sigmab_\X=\X^\top\X$ and
-the prior precision matrix $\A$:
-It is given by $d_{\A} =
-\tr\big(\Sigmab_\X(\A+\Sigmab_\X)^{-1}\big)$. In
-\cite{regularized-volume-sampling} it was suggested that $d_{\A}$
-should also be used as the effective dimension for the experimental
-design problem. Theorem \ref{t:q1} suggests it may not reflect the true degrees
-of freedom of the problem because it does not scale with subset size
-$k$. Instead we propose to use the \emph{scaled effective dimension}
-$d_{\frac nk\!\A}$. Thus, the two definitions we are comparing can be
-summarized as follows:
+problems.
+
+Consider the two definitions we are comparing:
 \begin{description}
@@
   \centering
-      \includegraphics[width=0.5\textwidth]{../figs/deff_identity}\nobreak\includegraphics[width=0.5\textwidth]{../figs/deff_lowrank}
+  \includegraphics[width=0.5\textwidth]{design/figs/deff_identity}\nobreak\includegraphics[width=0.5\textwidth]{design/figs/deff_lowrank}
   \caption{Scaled effective dimension compared to the full effective
@@
 
-\section{Additional details for the experiments}
-\label{a:experiments}
-\input{bayesian_experiments-appendix.tex}
+\subsection{Technical contributions}
+To establish Theorems \ref{t:q2} and \ref{t:q1},
+we develop a theoretical framework for a new sampling distribution
+% $\DPPreg{p}(\X,\A)$, where 
+% $p=(p_1,...,p_n)\in[0,1]^n$ is a vector of weights. 
+which can be seen as a \emph{regularized} variant of a determinantal point
+process (DPP). DPPs are a well-studied family of distributions
+with numerous applications in sampling diverse subsets
+of negatively correlated elements \citep[see][]{dpp-ml}.
+
+Given a psd matrix $\A$ and a weight vector $p=(p_1,...,p_n)\in[0,1]^n$, we
+define $\DPPreg{p}(\X,\A)$ as a distribution over subsets $S\subseteq[n]$ (of all
+sizes) such that (see Definition~\ref{d:r-dpp}):
+\begin{align*}
+  \Pr(S)\propto \det(
+  \X_S^\top\X_S+\A)\ \cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in
+    S}(1-p_i).
+\end{align*}
+A number of regularized DPPs have been proposed recently
+\citep{dpp-intermediate,regularized-volume-sampling}, mostly within the context
+of Randomized Numerical Linear Algebra (RandNLA)~
+\citep{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR}.
+To our knowledge, ours is the first such definition that strictly falls
+under the umbrella of traditional DPPs \citep{dpp-ml}.
+We show this in Section \ref{s:r-dpp}, where we also prove that
+regularized DPPs can be decomposed into a low-rank DPP plus
+i.i.d. Bernoulli sampling (Theorem \ref{t:algorithm}). This decomposition reduces
+the sampling cost from $O(n^3)$ to $O(nd^2)$,
+and involves a more general result about DPPs defined via a
+correlation kernel (Lemma~\ref{l:decomp_corr_dpp}), which is of
+independent interest.
+
+In Section \ref{s:guarantees} we demonstrate a fundamental connection between an
+$\A$-regularized DPP and Bayesian experimental design with precision
+matrix $\A$. For simplicity of exposition, let the weight vector $p$ be uniformly equal $(\frac kn,...,\frac kn)$. If
+$S\sim\DPPreg{p}(\X,\A)$ and $f_{\A}$ is any one of the
+A/C/D/V-optimality criteria, then:
+\begin{align*}
+  \E\big[f_{\A}(\X_S^\top\X_S)\big]
+  \leq f_{\A}\big(\tfrac kn\Sigmab_{\X}\big)
+  \quad\text{and}\quad \E\big[|S|\big]
+  \leq d_{\frac nk\!\A}+k.
+\end{align*}
+The proof of Theorem \ref{t:q1} relies on these two inequalities and a
+concentration bound for the subset size $|S|$,
+whereas to obtain Theorem
+\ref{t:q2} we additionally use the SDP relaxation to find the optimal
+weight vector~$p$. When $\A=\zero$, then $\DPPreg{p}(\X,\A)$ bears a
+lot of similarity to
+\emph{proportional volume sampling} which is an (unregularized) determinantal
+distribution proposed by \cite{proportional-volume-sampling}.
+Our algorithm not only extends it to the Bayesian setting but also
+offers a drastic time complexity improvement from
+the $O(n^4dk^2\log k)$ required by \cite{proportional-volume-sampling}
+down to the $O(nd^2)$ required for sampling from $\DPPreg{p}(\X,\A)$,
+and recent advances in RandNLA for DPP sampling
+\citep{leveraged-volume-sampling,correcting-bias,dpp-intermediate} suggest that
+$O(nd\log n + \poly(d))$ time is also possible.
+
+\section{Related work}
+\label{s:related-work}
+
+A number of works proposed $(1+\epsilon)$-approximation algorithms
+for experimental design which start with solving a convex
+relaxation of the problem, and then use some rounding strategy to
+obtain a discrete solution (see Table \ref{tab:comp} for comparison).
+In this line of work we wish to find the smallest $k$ for which a
+polynomial time approximation algorithm is possible. For example,
+\cite{tractable-experimental-design} gave an approximation algorithm
+for classical A/V-optimality with $k=\Omega(\frac{d^2}{\epsilon})$, where the
+rounding is done in a greedy fashion, and some randomized rounding
+strategies are also discussed. \cite{proportional-volume-sampling}
+suggested \emph{proportional volume sampling} for the rounding step
+and obtained approximation algorithms for classical A/D-optimality with
+$k=\Omega(\frac d\epsilon+\frac {\log1/\epsilon}{\epsilon^2})$. Their
+approach is particularly similar to ours (when $\A=\zero$). However,
+as discussed earlier, while their algorithms run in polynomial time,
+they scale very poorly with the number of experiments $n$ (see Table
+\ref{tab:comp}). \cite{near-optimal-design} proposed
+an efficient algorithm with a $(1+\epsilon)$-approximation guarantee for a wide
+range of optimality criteria, including A/C/D/E/V/G-optimality, both classical and Bayesian, when
+$k=\Omega(\frac d{\epsilon^2})$.  Our results (in Theorem \ref{t:q2}) improve on this work in
+two important ways:
+\begin{itemize}
+  \item In terms of the dependence on $\epsilon$ for A/C/D/V-optimality,
+  \item In terms of the dependence on the dimension
+        (by replacing $d$ with $d_{\A}$) in the Bayesian setting.
+\end{itemize}
+A lower bound shown by \cite{proportional-volume-sampling} implies that our
+Theorem \ref{t:q2} cannot be
+directly extended to E-optimality, but a similar lower bound does not
+exist for G-optimality. We remark that the approximation approaches
+relying on a convex relaxation can generally be converted to
+an upper bound on $\opt$ akin to our Theorem \ref{t:q1},
+however, unlike our bound, none of them apply to the regime of $k\leq d$.
+
+Non-trivial bounds for the \emph{classical} A-optimality criterion
+(i.e., $\opt$
+with $\A=\zero$) were first given by \cite{avron-boutsidis13}, where
+they show that
+for any $k\geq d$, $\opt \leq (1+ \frac{d-1}{k-d+1})\cdot f_{\zero}(\frac kn\Sigmab_\X)$
+and the subset $S$ attaining the bound can be found in polynomial time.
+The result was later extended \citep{unbiased-estimates,
+  regularized-volume-sampling,unbiased-estimates-journal} to the case
+where $\A=\lambda\I$, proving that for any $k\geq d_{\lambda\I}$,
+we have
+$\opt\leq (1 +\frac{d_{\lambda\I}-1}{k-d_{\lambda\I}+1})\cdot f_{\frac
+      kn\!\lambda\I}(\frac kn\Sigmab_\X)$,
+and also a faster $O(nd^2)$ time algorithm was provided.
+In comparison, our results (in Theorem \ref{t:q1}) offer the following
+improvements for upper bounding $\opt$:
+\begin{itemize}
+  \item We cover a wider range of subset sizes, because
+        $d_{\frac nk\!\lambda\I}\leq d_{\lambda\I}$,
+  \item Our upper bound can be much tighter because
+        $f_{\lambda\I}(\frac kn\Sigmab_\X)\leq f_{\frac kn\!\lambda\I}(\frac
+          kn\Sigmab_\X)$.
+\end{itemize}
+Additionally, \cite{minimax-experimental-design} propose a new notion of
+\emph{minimax} experimental design, which is related to A/V-optimality.  They
+also use a determinantal distribution for subset selection, however, due to
+different assumptions, their bounds are incomparable.
+
+Purely greedy approximation algorithms have been shown to provide
+guarantees in a number of special cases for experimental design. One
+example is classical D-optimality criterion, which can be converted to
+a submodular function \citep{submodularity-optimal-design}. Also, greedy algorithms
+for Bayesian A/V-optimality criteria have been considered by
+\cite{greedy-supermodular} and \cite{greedy-graph-sampling}. These
+methods can only provide a constant factor approximation guarantee
+(as opposed to $1+\epsilon$), and the factor is generally problem
+dependent (which means it could be arbitrarily large).
+Finally, a number of
+heuristics with good empirical performance have been proposed, such as Fedorov's
+exchange method \citep{cook1980comparison}. However, in this work
+we focus on methods that provide theoretical approximation guarantees.
+% In Section \ref{s:bayesian:experiments} we experimentally show that
+% our regularized DPP sampling (without solving an SDP) performs
+% better than iid sampling methods like uniform and predictive length
+% sampling \cite{zhu2015optimal} and that by using the SDP solution
+% we can improve it to match the performance of
+% greedy algorithms for A-optimal design \cite{chamon2017approximate,greedy-supermodular}
+% which empirically have been found to perform well on real world data.
+
+\section{A new regularized determinantal point process}
+\label{s:r-dpp}
+In this section we develop the theory for a novel regularized extension of
+determinantal point processes (DPP) which we use as the sampling distribution
+for obtaining guarantees in Bayesian experimental design.
+DPPs form a family of distributions which are used to model repulsion between
+elements in a random set, with many applications in machine learning
+\citep{dpp-ml}. Here, we focus on the setting where we are sampling out of all
+$2^n$ subsets $S\subseteq[n]$. Traditionally, a DPP is defined by a
+correlation kernel, which is an $n\times n$ psd matrix $\K$ with eigenvalues
+between 0 and 1, i.e., such that $\zero\preceq\K\preceq \I$. Given a
+correlation kernel $\K$, the corresponding DPP is
+defined as
+\begin{align*}
+  S\sim\DPPcor(\K)\quad\text{iff}\quad
+  \Pr(T\subseteq S) = \det(\K_{T,T})\ \  \forall_{T\in[n]},
+\end{align*}
+where $\K_{T,T}$ is the submatrix of $\K$ with rows and columns
+indexed by $T$. Another way of defining a DPP, popular in the machine
+learning community, is via an ensemble kernel $\L$. Any psd matrix
+$\L$ is an ensemble kernel of a DPP defined as:
+\begin{align*}
+  S\sim\DPPens(\L)\quad\text{iff}\quad \Pr(S) \propto \det(\L_{S,S}).\quad~
+\end{align*}
+Crucially, every $\DPPens$ is also a $\DPPcor$, but not the other way
+around. Specifically, $\DPPens(\L)=\DPPcor(\K)$ when:
+\begin{align*}
+   & \text{(a)}\ \  \L=\K(\I-\K)^{-1},\qquad
+   & \text{(b)}\ \  \K = \I -  (\I+\L)^{-1},
+\end{align*}
+but (a) requires that $\I-\K$ be invertible which is not true
+for some DPPs.
+(This will be important in our analysis.)
+The classical algorithm for sampling from a DPP requires the
+eigendecomposition of either matrix $\K$ or $\L$, which in general
+costs $O(n^3)$, followed by a sampling procedure which costs
+$O(n\,|S|^2)$ \cite{dpp-independence,dpp-ml}.
+
+We now define our regularized DPP and describe its
+connection with correlation and ensemble~DPPs.
+\begin{definition}\label{d:r-dpp}
+  Given matrix $\X\in\R^{n\times d}$, a sequence  $p=(p_1,\dots,p_n)\in[0,1]^n$
+  and a psd matrix $\A\in\R^{d\times d}$ such that
+  $\sum_ip_i\x_i\x_i^\top\!+\A$ is
+  full rank, let
+  $\DPPreg{p}(\X,\A)$ be a distribution over $S\subseteq [n]$:
+  \begin{align}
+    \Pr(S) = \frac{\det(
+      \X_S^\top\X_S+\A)}{\det\!\big(\!
+      \sum_ip_i\x_i\x_i^\top\!+ \A\big)}
+    \cdot \!\prod_{i\in S}p_{i}\cdot\!\prod_{i\not\in S}(1\!-\!p_i).\label{eq:poisson-prob}
+  \end{align}
+  %   \Pr(\pi)\ \propto\ \det\!\big(\A +
+  %   \X_{\pi}^\top\X_{\pi}\big)
+  %   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
+  % \end{align*}
+\end{definition}
+The fact that this is a proper distribution (i.e., that it sums to
+one) can be restated as a determinantal expectation formula: if
+$b_i\sim\mathrm{Bernoulli}(p_i)$ are independent Bernoulli random
+variables, then
+\begin{align*}
+   & \sum_{S\subseteq [n]}\!\det(
+  \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
+   & \!=
+  \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
+  \det\!\Big(\sum_i\E[b_i]\x_i\x_i^\top\!+\A\Big),
+\end{align*}
+where $(*)$ follows  from Lemma 7 of \citet{determinantal-averaging}.
+% follows from a recently shown determinantal expectation formula.
+% \begin{lemma}[\cite{determinantal-averaging}]\label{l:cb}
+% For $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$
+% and a random set $S\subseteq [n]$ drawn with independent Bernoullis,
+% \begin{align*}
+%   \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)\Big]
+%   = \det\!\Big( \E\big[\X_S^\top\X_S\big]+\A\Big).
+% \end{align*}
+% \end{lemma}
+
+The main theoretical contribution in this section is the
+following efficient algorithm for $\DPPreg{p}(\X,\A)$ which reduces it to
+sampling from a correlation DPP and unioning with i.i.d.~Bernoulli samples:
+\begin{theorem}\label{t:algorithm}
+  For any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
+  s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
+  full rank, let
+  \begin{align*}
+    T & \sim\DPPcor\big(\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big),
+    \quad                                                                                    \\
+      & \quad\text{where}\quad\D_p=\diag(p).
+  \end{align*}
+  If $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random variables, then
+  $T\cup \{i:b_i\!=\!1\}\sim \DPPreg{p}(\X,\A)$.
+\end{theorem}
+
+\begin{remark}
+  Figure~\ref{fig:algorithm-pseudocode} illustrates how to exploit this result
+  to build an efficient sampling algorithm.
+  Since the correlation kernel matrix has rank at most $d$, the preprocessing
+  cost of eigendecomposition is $O(nd^2)$. Then, each sample costs only $O(n\,|T|^2)$.
+\end{remark}
+
+We prove the theorem in three steps.
+First, we express $\DPPreg{p}(\X,\A)$ as an ensemble DPP, which requires some
+additional assumptions on $\A$ and $p$ to be possible.
+Then, we convert the ensemble to a correlation kernel (eliminating the extra
+assumptions),
+and finally show that this kernel can be decomposed into a rank $d$ kernel plus
+Bernoulli sampling.
+In the process, we establish several novel theoretical properties
+regarding the representation, decomposition, and closure properties
+of regularized DPPs which may be of independent interest.
+
+\renewcommand{\thealgorithm}{}
+\floatname{algorithm}{}
+\vspace{-3mm}
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{algorithm}[H]
+      {\fontsize{9}{9}\selectfont
+        \caption{\bf \small Sampling \ $S\sim\DPPreg{p}(\X,\A)$}
+        \begin{algorithmic}[0]
+          \STATE \textbf{Input:} $\X\!\in\!\R^{n\times d}\!,\text{ psd } \A\!\in\!\R^{d\times
+              d}\!, p\!\in\![0,1]^n$\\[1mm]
+          \STATE Compute $\Z \leftarrow \A+\X^\top\D_p\X$\\[1mm]
+          \STATE Compute \textsc{SVD} of  $\B=\D_p^{\sfrac12}\X\Z^{-\sfrac12}$\\[1mm]
+          \STATE Sample $T\sim\DPPcor(\B\B^\top)$ \hfill \citep{dpp-independence}\\[1mm]
+          \STATE Sample $b_i\sim\mathrm{Bernoulli}(p_i)$ for $i\in[n]$\\[1mm]
+          \RETURN $S=T\cup\{i:b_i=1\}$
+        \end{algorithmic}
+      }
+    \end{algorithm}
+  \end{center}
+  \vspace{-3mm}
+  \caption{Algorithm which exploits
+    Theorem~\ref{t:algorithm} to sample $S \sim \DPPreg{p}(\X, \A)$
+    in $O( nd^2 )$ time.}
+  \label{fig:algorithm-pseudocode}
+\end{figure}
+
+\vspace{2mm}
+\begin{lemma}\label{t:reduction}
+  Given $\X$, $\A$ and $\D_p$ as in Theorem \ref{t:algorithm}, assume that $\A$ and $\I-\D_p$ are
+  invertible. Then,
+  \begin{align*}
+    \DPPreg{p}(\X,\A) & =\DPPens\big(\Dbt+
+    \Dbt^{\sfrac12}\X\A^{-1}\X^\top
+    \Dbt^{\sfrac12}\big),\quad                \\
+    \text{where}
+    \quad\Dbt         & = \D_p(\I-\D_p)^{-1}.
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  Let $S\sim \DPPreg{p}(\X,\A)$. By Definition \ref{d:r-dpp} and
+  the fact that $\det(\A\B+\I)=\det(\B\A+\I)$,
+  \begin{align*}
+    \Pr(S) & \propto \det(\X_S^\top\X_S+\A)\cdot
+    \prod_{i\in S}p_i\cdot\prod_{i\not\in S}(1-p_i)                                                 \\
+           & = \det(\X_S^\top\X_S+\A)\cdot
+    \prod_{i\in S}\frac{p_i}{1-p_i}\cdot\prod_{i=1}^n(1-p_i)
+    \\ &\propto\det\!\big(\A(\A^{-1}\X_S^\top\X_S+\I)\big) \det(\Dbt_{S,S}) \\
+           & =\det(\A)\det(\A^{-1}\X_S^\top\X_S+\I) \det(\Dbt_{S,S})
+    \\ &\propto \det(\X_S \A^{-1}\X_S^\top+\I) \det(\Dbt_{S,S}) \\
+           & =\det\!\Big(\big[\Dbt^{\sfrac12}\X \A^{-1}\X^\top\Dbt^{\sfrac12}+\Dbt\big]_{S,S}\Big),
+  \end{align*}
+  which matches the definition of the L-ensemble DPP.
+\end{proof}
+At this point, to sample from  $\DPPreg{p}(\X,\A)$, we could simply
+invoke any algorithm for sampling from
+an ensemble DPP. However, this would only work for invertible
+$\A$, which in particular excludes the important case of
+$\A=\zero$ corresponding to classical experimental
+design. Moreover, the standard algorithm would require computing the
+eigendecomposition of the ensemble kernel, which (at
+least if done na\"ively) costs $O(n^3)$. Even after this is done, the
+sampling cost would still be $O(n\,|S|^2)$ which can be considerably
+more than $O(nd^2)$. We first address the issue of invertibility of matrix
+$\A$ by expressing our distribution via a correlation DPP.
+\begin{lemma}\label{l:correlation}
+  Given $\X$, $\A$, and $\D_p$ as in Theorem \ref{t:algorithm} (without
+  any additional assumptions), we have
+  \begin{align*}
+     & \DPPreg{p}(\X,\A)
+    = \DPPcor\big(\D_p +                                                                                                            \\
+     & \ (\I\!-\!\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A\!+\!\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}(\I\!-\!\D_p)^{\sfrac12}\big).
+  \end{align*}
+\end{lemma}
+When $\A$ and $\I-\D_p$ are invertible, then the proof is a
+straightforward calculation.
+Then, we use a limit argument with $p_\epsilon=(1-\epsilon)p$ and
+$\A_\epsilon=\A+\epsilon\I$, where $\epsilon\rightarrow 0$.
+
+\begin{proof}
+  First, we show this under the invertibility assumptions of Lemma
+  \ref{t:reduction}, i.e., given that $\A$ and $\I-\D_p$ are
+  invertible. In this case $\DPPreg{p}(\X,\A)=\DPPens(\L)$, where
+  \begin{align}\L=\Dbt+
+    \Dbt^{\sfrac12}\X\A^{-1}\X^\top
+    \Dbt^{\sfrac12}\quad\text{ and }\quad\Dbt=\D_p(\I-\D_p)^{-1}.\label{eq:ensemble}
+  \end{align}
+  Converting this to
+  a correlation kernel $\K$ and denoting $\Xt=\D_p^{\sfrac12}\X$, we obtain
+  \begin{align*}
+    \K & =\I-(\I+\L)^{-1}
+    \\      &=\I - \big(\I + (\I-\D_p)^{-1}\D_p +
+    (\I-\D_p)^{-\sfrac12}\Xt\A^{-1}\Xt^\top(\I-\D_p)^{-\sfrac12}\big)^{-1}
+    \\ &= \I - \big((\I-\D_p)^{-1} +
+    (\I-\D_p)^{-\sfrac12}\Xt\A^{-1}\Xt^\top(\I-\D_p)^{-\sfrac12}\big)^{-1}
+    \\ &=\I -
+    (\I-\D_p)^{\sfrac12}(\I+\Xt\A^{-1}\Xt^\top)^{-1}(\I-\D_p)^{\sfrac12}
+    \\
+       & \overset{(*)}{=}\I -
+    (\I-\D_p)^{\sfrac12}\big(\I-\Xt\A^{-\sfrac12}
+    (\I+\A^{-\sfrac12}\Xt^\top\Xt\A^{-\sfrac12})^{-1}
+    \A^{-\sfrac12}\Xt^\top\big)(\I-\D_p)^{\sfrac12}
+    \\
+       & =\I - (\I-\D_p) +
+    (\I-\D_p)^{\sfrac12}\Xt(\A+\Xt^\top\Xt)^{-1}\Xt^\top(\I-\D_p)^{\sfrac12}
+    \\ &=\D_p + (\I-\D_p)^{\sfrac12}\Xt(\A+\Xt^\top\Xt)^{-1}\Xt^\top(\I-\D_p)^{\sfrac12},
+  \end{align*}
+  where $(*)$ follows from Fact 2.16.19 in
+  \cite{matrix-mathematics}. Note that converting from $\L$ to $\K$
+  got rid of the inverses $\A^{-1}$  and $(\I-\D_p)^{-1}$ appearing
+  in \eqref{eq:ensemble}. The intuition
+  is that when $\A$ or $\I-\D_p$ is non-invertible, then
+  $\DPPreg{p}(\X,\A)$ is not an L-ensemble but it is still a
+  correlation DPP. To show this, we use a limit argument. For
+  $\epsilon\in[0,1]$, let $p_\epsilon=(1-\epsilon)p$ and
+  $\A_\epsilon=\A+\epsilon\I$. Observe that if $\epsilon>0$ then $\A_\epsilon$ and
+  $\I-\D_{p_{\epsilon}}$ are always invertible even if $\A$ and
+  $\I-\D_p$ are not. Denote $\K_\epsilon$ as the
+  above correlation kernel with $p$ replaced by $p_{\epsilon}$ and
+  $\A$ replaced by $\A_{\epsilon}$. Note that all matrix operations
+  defining kernel $\K_\epsilon$ are continuous w.r.t. $\epsilon\in[0,1]$, including the inverse, since
+  $\A+\Xt^\top\Xt$ is assumed to be invertible. Therefore, the
+  following equalities hold (with limits taken point-wise and $\epsilon>0$):
+  \begin{align*}
+    \DPPreg{p}(\X,\A) = \lim_{\epsilon\rightarrow
+      0}\DPPreg{p_\epsilon}(\X,\A_{\epsilon}) =
+    \lim_{\epsilon\rightarrow 0}\DPPcor(\K_\epsilon) = \DPPcor(\K),
+  \end{align*}
+  where we did not have to assume invertibility of $\A$ or $\I-\D_p$.
+\end{proof}
+
+Finally, we show that the correlation DPP arrived at in Lemma
+\ref{l:correlation} can be decomposed into a smaller DPP plus
+Bernoulli sampling. In fact, in the following lemma we obtain a more
+general recipe for combining DPPs with Bernoulli sampling, which may
+be of independent interest. Note that if $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random
+variables then $\{i:b_i\!=1\!\}\sim\DPPcor(\D_p)$.
+\begin{lemma}\label{l:decomp_corr_dpp}
+  Let $\K$ and $\D$ be $n\times n$ psd matrices with eigenvalues between
+  0 and 1, and assume that $\D$ is diagonal. If\, $T\!\sim\DPPcor(\K)$ and
+  $R\sim\DPPcor(\D)$, then
+  \begin{align*}T\cup R\sim \DPPcor\big(\D+(\I-\D)^{\sfrac12}\K
+      (\I-\D)^{\sfrac12}\big).
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  For this proof we will use the shorthand $\K_A$ for $\K_{A,A}$. If
+  $\D$ has no zeros on the diagonal then $\det(\D_A)>0$ for all
+  $A\subseteq[n]$ and
+  \begin{align*}
+    \Pr(A \subset T \cup R)
+     & =\sum_{B\subset A}\Pr(R\cap A=A\setminus B) \ \Pr(B\subseteq T)          \\
+     & =\sum_{B\subset A}\det(\D_{A\setminus
+    B})\det\!\big([\I-\D]_B\big)\,\det(\K_B)                                    \\
+     & =\sum_{B\subset A}\det(\D_{A\setminus
+    B})\det\!\Big(\big[(\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\big]_B\Big)       \\
+     & =\det(\D_A) \sum_{B\subset A}\det\!\Big(\big[\D^{-\sfrac12}
+    (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\D^{-\sfrac12}\big]_{B}\Big)          \\[-1mm]
+     & \overset{(*)}{=}\det(\D_A)\det\!\Big(\I + \big[\D^{-\sfrac12}
+    (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\D^{-\sfrac12}\big]_A\Big)            \\
+     & =\det\!\Big(\big[\D+ (\I-\D)^{\sfrac12}\K(\I-\D)^{\sfrac12}\big]_A\Big),
+  \end{align*}
+  where $(*)$ follows from a standard determinantal identity used to
+  compute the L-ensemble partition function
+  \cite[Theorem~2.1]{dpp-ml}. If $\D$ has zeros on the diagonal, a
+  similar limit argument as in Lemma \ref{l:correlation} with
+  $\D_\epsilon=\D+\epsilon\,\I$ holds.
+\end{proof}
+
+Theorem
+\ref{t:algorithm} now follows by combining Lemmas \ref{l:correlation} and \ref{l:decomp_corr_dpp}.
+
+\section{Guarantees for Bayesian experimental design}
+\label{s:guarantees}
+In this section we prove our main results regarding Bayesian
+experimental design (Theorems \ref{t:q2} and \ref{t:q1}). First, we
+establish certain properties of the regularized DPP distribution that
+make it effective in this setting. Even though the size of the
+sampled subset $S\sim\DPPreg{p}(\X,\A)$ is random and can be as large
+as $n$, it is also highly concentrated around its expectation, which
+can be bounded in terms of the $\A$-effective dimension. This is
+crucial, since both of our main results require a subset of
+deterministically bounded size. Recall that the effective dimension is
+defined as a function $d_{\A}(\Sigmab) =
+  \tr\big(\Sigmab(\A+\Sigmab)^{-1}\big)$.
+\begin{lemma}\label{l:size}
+  Given any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
+  s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
+  full rank, let $S=T\cup \{i:b_i=1\}\sim \DPPreg{p}(\X,\A)$ be defined
+  as in Theorem \ref{t:algorithm}. Then
+  \begin{align*}
+    \E\big[|S|\big] & \leq \E\big[|T|\big] + \E\Big[\sum_i b_i\Big]          \\
+                    & = d_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big) + \sum_i p_i.
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  For correlation kernels it is known that the expected size of $\DPPcor(\K)$
+  is $\tr(\K)$. Thus, using $\D_p=\diag(p)$, we can invoke Lemma \ref{l:correlation} to obtain
+  \begin{align*}
+    \E\big[|S|\big]
+     & = \tr\big(\D_p +
+    (\I-\D_p)^{\sfrac12}\,\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}
+    \X^\top\D_p^{\sfrac12}(\I-\D_p)^{\sfrac12}\big)
+    \\ &\leq \tr(\D_p) +\tr\big(\D_p^{\sfrac12}
+    \X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big)
+    \\ &=\tr(\D_p) +\tr\big(\X^\top\D_p\X (\A+\X^\top\D_p\X)^{-1}\big) =
+    \tr(\D_p) + d_{\A}(\X^\top\D_p\X),
+  \end{align*}
+  from which the claim follows.
+\end{proof}
+
+Next, we show two expectation inequalities for the matrix inverse and
+matrix determinant, which hold for the regularized DPP.
+We use them to bound the Bayesian optimality criteria in expectation.
+% These are
+% immediately followed by a corollary applying them to bound
+% Bayesian optimality criteria.
+\begin{lemma}\label{t:expectations}
+  Whenever $S\sim \DPPreg{p}(\X,\A)$ is a well-defined distribution it holds that
+  \begin{align}
+    \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
+     & \preceq \Big(\!
+    \sum_ip_i\x_i\x_i^\top \!+\A \Big)^{-1}\!\!,\label{eq:sqinv}
+    \\
+    \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
+     & \leq\det\!\Big(\!
+    \sum_ip_i\x_i\x_i^\top\!+\A \Big)^{\!-1}\!\!.\!\!\label{eq:det}
+  \end{align}
+\end{lemma}
+
+\begin{proof}
+  For a square matrix $\M$, define its adjugate, denoted $\adj(\M)$, as a matrix
+  whose $i,j$-th entry is $(-1)^{i+j}\det(\M_{-j,-i})$, where
+  $\M_{-j,-i}$ is the matrix $\M$ without $j$th row and $i$th column. If
+  $\M$ is invertible, then $\adj(\M) = \det(\M)\M^{-1}$. Now, let
+  $b_i\sim\mathrm{Bernoulli}(p_i)$ be independent random variables. As seen
+  in previous section, the identity
+  $\E[\det(\sum_ib_i\x_i\x_i^\top+\A)]=\det(\sum_ip_i\x_i\x_i^\top+\A)$ gives
+  us the normalization constant for $\DPPreg{p}(\X,\A)$. Moreover, as
+  noted in a different context by \cite{determinantal-averaging}, when
+  applied entrywise to the adjugate matrix, this identity implies that
+  $\E[\adj(\sum_ib_i\x_i\x_i^\top+\A)]=\adj(\sum_ip_i\x_i\x_i^\top+\A)$. Let
+  $\Ic$ denote the set of all subsets $S\subseteq [n]$ such that
+  $\X_S^\top\X_S+\A$ is invertible. We have
+  \begin{align*}
+    \E\Big[\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
+     & = \sum_{S\in\Ic}\big(\X_S^\top\X_S+\A\big)^{-1}
+    \frac{\det(\X_S^\top\X_S+\A)}{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
+      S}p_i\prod_{i\not\in S}(1-p_i)
+    \\ &=\sum_{S\in\Ic}\frac{\adj(\X_S^\top\X_S+\A\big)}
+    {\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
+      S}p_i\prod_{i\not\in S}(1-p_i)
+    \\ &\preceq \sum_{S\subseteq[n]}\frac{\adj(\X_S^\top\X_S+\A\big)}
+    {\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
+      S}p_i\prod_{i\not\in S}(1-p_i)
+    \\ &=\frac{\E\big[\!\adj(\sum_i
+        b_i\x_i\x_i^\top+\A)\big]}{\det(\sum_ip_i\x_i\x_i^\top+\A)}
+    \\
+     & =\frac{\adj(\sum_ip_i\x_i\x_i^\top+\A)}{\det(\sum_ip_i\x_i\x_i^\top+\A)}
+    = \Big(\sum_ip_i\x_i\x_i^\top+\A\Big)^{-1}.
+  \end{align*}
+  Note that if $\Ic$ contains all subsets of $[n]$, for example when
+  $\A\succ\zero$, then the inequality turns into equality. Thus, we
+  showed \eqref{eq:sqinv}, and \eqref{eq:det} follows even more easily:
+  \begin{align*}
+    \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]
+     & = \sum_{S\in\Ic}
+    \frac{1}{\det(\sum_ip_i\x_i\x_i^\top+\A)}\ \prod_{i\in
+      S}p_i\prod_{i\not\in S}(1-p_i)
+    \leq \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{\!-1}\!,
+  \end{align*}
+  where the equality holds if $\Ic$ consists of all subsets of $[n]$.
+\end{proof}
+
+\begin{corollary}\label{c:expectations}
+  Let $f_{\A}$ be A/C/D/V-optimality. Whenever $S\sim\DPPreg{p}(\X,\A)$ is
+  well-defined,
+  \begin{align*}
+    \E\big[f_{\A}(\X_S^\top\X_S)\big] \leq f_{\A}\Big(\sum_i p_i\x_i\x_i^\top\Big).
+  \end{align*}
+\end{corollary}
+\begin{proof}
+  In the case of A-, C-, and V-optimality, the function $f_{\A}$ is a
+  linear transformation of the matrix $(\X_S^\top\X_S+\A)^{-1}$ so the
+  bound follows from \eqref{eq:sqinv}. For D-optimality, we apply
+  \eqref{eq:det} as follows:
+  \begin{align*}
+    \E\big[f_{\A}(\X_S^\top\X_S)\big]
+     & = \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)^{-\sfrac1d}\Big]                     \\
+     & \leq
+    \E\bigg[\Big(\det\!\big(\X_S^\top\X_S+\A\big)^{-\sfrac1d}\Big)^d\bigg]^{\sfrac1d} \\
+    \\ &=  \E\Big[\!\det\!\big(\X_S^\top\X_S+\A\big)^{-1}\Big]^{\sfrac1d}\\
+     & \leq
+    \det\!\Big(\sum_i p_i\x_i\x_i^\top\Big)^{-\sfrac1d},
+  \end{align*}
+  which completes the proof.
+\end{proof}
+Finally, we present the key lemma that puts everything together. This
+result is essentially a generalization of Theorem \ref{t:q1} from
+which also follows Theorem \ref{t:q2}.
+\begin{lemma}\label{l:guarantees}
+  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be $n\times
+    d$.  For some $w=(w_1,\dots,w_n)\in[0,1]^n$, let
+  $\Sigmab_w=\sum_iw_i\x_i\x_i^\top$ and assume that
+  $\sum_i w_i=k\in[n]$. If $k\geq 4\,d_{\A}(\Sigmab_w)$, then a
+  subset $S\subseteq[n]$ of size $k$ can be found in $O(ndk+k^2d^2)$ time that satisfies
+  \begin{align*}
+     & f_{\A}\big(\X_S^\top\X_S\big)                                                                                                            \\
+     & \leq \bigg(\ 1+8\,\frac{d_{\A}(\Sigmab_w)}{k} + 8\sqrtshort{\frac{\ln (k/d_{\A}(\Sigmab_w))}{k}}\ \bigg)\cdot f_{\A}\big(\Sigmab_w\big).
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  Let $p=(p_1,\dots,p_n)$ be defined so that $p_i =
+    \frac{w_i}{1+\epsilon}$, and suppose that $S\sim
+    \DPPreg{p}(\X,\A)$. Then, using Corollary \ref{c:expectations}, we have
+  \begin{align*}
+    \Pr & \big(|S|\leq k\big)\,\E\Big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\Big] \\
+        & \leq
+    \E\big[f_{\A}(\X_S^\top\X_S)\big]                                          \\
+        & \leq
+    f_{\A}\Big(\sum_ip_i\x_i\x_i^\top\Big)
+    \\ &\leq (1+\epsilon)\cdot f_{\A}\Big(\sum_iw_i\x_i\x_i^\top\Big).
+  \end{align*}
+  Using Lemma \ref{l:size} we can bound the expected size of $S$ as
+  follows:
+  \begin{align*}
+    \E\big[|S|\big]
+     & \leq d_{\A}(\Sigmab_w) + \sum_i p_i                                            \\
+     & =d_{\A}(\Sigmab_w)+\frac{k}{1+\epsilon}                                        \\
+     & =k\cdot \Big(1+\frac {d_{\A}(\Sigmab_w) }k - \frac{\epsilon}{1+\epsilon}\Big).
+  \end{align*}
+  Let $d_w = d_{\A}(\Sigmab_w)$ and $\alpha=1+\frac {d_w} k -\frac{\epsilon}{1+\epsilon}$.
+  If $1\geq\epsilon\geq \frac{4d_w}{k}$, then $\alpha\leq 1 +
+    \frac\epsilon4-\frac\epsilon2=1-\frac\epsilon4$. Since $\DPPreg{p}(\X,\A)$
+  is a  determinantal point process, $|S|$ is a
+  Poisson binomial r.v.~so for $\epsilon\geq 6\sqrt{\frac{\ln(k/d_w)}{k}}$,
+  \begin{align*}
+    \Pr(|S|>k) \leq \ee^{-\frac{(k-\alpha k)^2}{2k}} = \ee^{-\frac
+      k2(1-\alpha)^2}\leq \ee^{-\frac{k\epsilon^2}{32}}\leq \frac{d_w}k.
+  \end{align*}
+  % If $k\geq\frac{4 d_{\A}(\Sigmab_w)}\epsilon$,
+  % then $\alpha \leq 1 + \frac\epsilon4 -\frac\epsilon2=1-\frac\epsilon4$.
+  %  Since $\Vol_p(\X,\A)$ is a determinantal point process, Lemma
+  % \ref{l:concent} implies that for $\epsilon'=\frac\epsilon4$ and $k\geq
+  %  48\frac{\ln 1/\epsilon'}{\epsilon'^2}$ we have
+  % \begin{align*}
+  % \Pr(|S|> k)\leq  \Pr\big(|S|\ \geq\ \alpha k + \epsilon' k\big)\leq
+  %   3\ee^{-\frac{\epsilon'^2k^2}{16(\epsilon'k+2\alpha k)}} \leq
+  %   3\ee^{-\frac{\epsilon'^2k}{48}}\leq 3\epsilon'.
+  % \end{align*}
+  For any $\epsilon\geq
+    4\,\frac{d_w}k + 6\sqrt{\frac{\ln(k/d_w)}{k}}$, we have
+  \begin{align*}
+    \E & \big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\big]                \\
+       & \leq \frac{1+\epsilon}{1-\frac{d_w}k}\cdot f_{\A}(\Sigmab_w) \\
+       & \leq
+    \bigg(1+\frac{\epsilon+\frac{d_w}k}{1-\frac{d_w}k}\bigg)\cdot
+    f_{\A}(\Sigmab_w)
+    \\ &\leq \bigg(1+7\,\frac{d_w}{k} +
+    8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\cdot f_{\A}(\Sigmab_w).
+  \end{align*}
+  Denoting $\E\big[f_{\A}(\X_S^\top\X_S)\mid |S|\leq k\big]$ as $F_k$,
+  Markov's inequality implies that
+  \[\Pr\Big(f_{\A}(\X_S^\top\X_S) \geq
+    (1+\delta) F_k\ \mid\ |S|\leq k\Big)\leq \frac1{1+\delta}.\]
+  Also, we showed
+  that $\Pr(|S|\leq k)\geq 1-\frac{d_w}{k}\geq \frac34$. Setting $\delta=\frac{d_w}{C
+      k}$ for sufficiently large $C$ we obtain that with probability
+  $\Omega(\frac{d_w}{k})$, the random set $S$ has size at most $k$ and
+  \begin{align*}
+    f_{\A} & (\X_S^\top\X_S)                                                        \\
+           & \leq \bigg(1+\frac{d_w}{C k}\bigg)\!\cdot\! \bigg(1+7\,\frac{d_w}{k} +
+    8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\!\cdot\! f_{\A}(\Sigmab_w)              \\
+           & \leq
+    \bigg(1+8\,\frac{d_w}{k} +
+    8\sqrtshort{\frac{\ln(k/d_w)}{k}}\bigg)\cdot f_{\A}(\Sigmab_w).
+  \end{align*}
+  We can sample from $\DPPreg{p}(\X,\A)$ conditioned on
+  $|S|\leq k$ and $f_{\A}(\X_S^\top\X_S)$ bounded as above by rejection
+  sampling. When $|S|<k$,  the set is completed to
+  $k$ with arbitrary indices. On average, $O(\frac k{d_w})$ samples from
+  $\DPPreg{p}(\X,\A)$ are needed, so the cost is $O(nd^2)$ for the
+  eigendecomposition, $O(\frac{k}{d_w}\cdot nd_w^2)=O(nd_wk)$ for
+  sampling and $O(\frac k{d_w}\cdot kd^2)$ for recomputing $f_{\A}(\X_S^\top\X_S)$.
+\end{proof}
+To prove the main results, we use Lemma \ref{l:guarantees} with
+appropriately chosen weights $w$.
+\begin{proofof}{Theorem}{\ref{t:q2}}
+  As discussed by \cite{near-optimal-design} and
+  \cite{boyd2004convex}, the following convex
+  relaxation of experimental design can be written as a semi-definite
+  program and solved using standard SDP solvers:
+  \begin{align}
+     & w^* \ = \  \argmin_w\ \
+    f_{\A}\Big(\sum_{i=1}^nw_i\x_i\x_i^\top\Big),\quad \\
+     & \text{subject to}\quad
+    \forall_i\ \ 0\leq w_i\leq 1,\quad \sum_i w_i=k.\label{eq:sdp}
+  \end{align}
+  The solution $w^*$ satisfies $f_{\A}\big(\Sigmab_{w^*}\big)\leq
+    \opt$. If we use $w^*$ in Lemma \ref{l:guarantees}, then observing
+  that $d_{\A}(\Sigmab_{w^*})\leq d_{\A}$, and setting $k\geq C(\frac
+    {d_{\A}}{\epsilon} + \frac{\log1/\epsilon}{\epsilon^2})$ for
+  sufficiently large $C$, the algorithm in the lemma finds subset $S$
+  such that
+  \[f_{\A}(\X_S^\top\X_S)\leq (1+\epsilon)\cdot
+    f_{\A}(\Sigmab_{w^*})\leq
+    (1+\epsilon)\cdot\opt.\]
+  % \cite{tractable-experimental-design} show that
+  % the support of $w^*$ has size at most $k+d^2$, which reduces the cost of
+  % our sampling to $O(k^2d^2)$.
+  Note that we did not need to
+  solve the SDP exactly, so approximate solvers could be used instead.
+\end{proofof}
+\begin{proofof}{Theorem}{\ref{t:q1}}
+  Let $w=(\frac kn,...,\frac kn)$ in Lemma~\ref{l:guarantees}. Then, we have
+  $\Sigmab_w=\frac kn\Sigmab_{\X}$ and also $d_{\A}(\Sigmab_w) =
+    d_{\frac nk\!\A}$. Since for any set $S$ of size $k$, we
+  have $\opt\leq f_{\A}(\X_S^\top\X_S)$, the result follows.
+\end{proofof}
+
+\input{bayesian_experiments.tex}
+
+\section{Conclusions}
+
+We proposed a new algorithm for finding
+$(1+\epsilon)$-approximate Bayesian experimental designs by leveraging
+a fundamental connection with determinantal point processes. Compared
+to the state-of-the-art approaches, our
+method provides stronger theoretical guarantees in terms of the allowed
+range of subset sizes, as well as
+offering significantly better time complexity guarantees. At the same
+time, our experiments show that on the task of A-optimal design the
+proposed algorithm performs as well as or better than several methods that
+are used in practice.
+
+% \subsubsection*{Acknowledgements}
+
+% MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
+% support of this work. Also, MWM and MD thank the NSF TRIPODS program
+% and FL thanks the NPSC for funding.  The authors thank Uthaipon
+% T.~Tantipongpipat for valuable discussions.
+
+% \bibliographystyle{apalike}
+% \bibliography{../pap}
+
+% \newpage
+% \appendix
+
+% \onecolumn
 

### 2022-11-23 design/aistats/bayesian_experiments-appendix.tex
Working compile

@@
-This section presents additional details and experimental results omitted from
-the main body of the paper. In addition to the \texttt{mg\_scale} dataset presented in
+% This section presents additional details and experimental results omitted from
+% the main body of the paper. 
+In addition to the \texttt{mg\_scale} dataset presented in
 Section~\ref{s:experiments}, we also benchmarked on three other data sets
@@
     $n$ & 1385               & 252                     & 392                 & 506                     \\
-        $d$ & 6 & 14 & 7 & 13\\
+    $d$ & 6                  & 14                      & 7                   & 13                      \\
     \bottomrule
@@
   \centering
-    \includegraphics[width=\textwidth]{../bayesian_figures/obj_grid.pdf}
+  \includegraphics[width=\textwidth]{design/bayesian_figures/obj_grid.pdf}
   \caption{A-optimality values achieved by the methods compared. In all cases
@@
   \centering
-    \includegraphics[width=\textwidth]{../bayesian_figures/runtime_grid.pdf}
+  \includegraphics[width=\textwidth]{design/bayesian_figures/runtime_grid.pdf}
   \caption{Runtimes of the methods compared. Our method (without SDP) is
@@
   \centering
-    \includegraphics[width=\textwidth]{../bayesian_figures/ratios_grid.pdf}
+  \includegraphics[width=\textwidth]{design/bayesian_figures/ratios_grid.pdf}
   \caption{The ratio controlled by Lemma~\ref{l:guarantees}. This ratio converges

### 2022-11-23 design/continuous_nips/continuous-empirical.tex
Updates from Overleaf
Working compile

@@
 \begin{figure}[t]
-  \includegraphics[width=\textwidth]{continuous_figures/decays.pdf}
+  \includegraphics[width=\textwidth]{design/continuous_figures/decays.pdf}
   \vspace{-1cm}
@@
 \begin{figure} %[ht]
-  \includegraphics[width=\textwidth]{continuous_figures/wishart_var.pdf}
-  \includegraphics[width=\textwidth]{continuous_figures/proj_bias.pdf}
+  \includegraphics[width=\textwidth]{design/continuous_figures/wishart_var.pdf}
+  \includegraphics[width=\textwidth]{design/continuous_figures/proj_bias.pdf}
   \vspace{-.8cm}
@@
 samples (up to two million) required for this experiment made the computations
-much more expensive (more details in Appendix \ref{a:empirical}). Based on the
+much more expensive (more details in \Cref{a:empirical}). Based on the
 above empirical results, we conclude with a conjecture.
@@
 % computations much more expensive.
+
+\subsection{Additional details for empirical evaluation}
+\label{a:empirical}
+
+Our empirical investigation of the rate of asymptotic convergence in \Cref{t:asymptotic} (and, more
+specifically, the variance and bias discrepancies defined in
+Section~\ref{sec:asymp-conj-details}), in the context of
+Gaussian random matrices, is related to open problems which have been
+extensively studied in the literature. Note that when
+$\X=\Z\Sigmab^{1/2}$ were $\Z$ has i.i.d.~Gaussian entries (as in
+Section \ref{sec:asymp-conj-details}), then $\W=\X^\top\X$ is known as the
+pseudo-Wishart distribution (also called the singular Wishart),
+denoted as $\W \sim \Pc\Wc(\Sigmab, n)$, and the variance
+term from the MSE can be written as $\sigma^2\E[\tr(\W^\dagger)]$.
+\cite{srivastava2003} first derived the probability density function of the
+pseudo-Wishart distribution, and
+\cite{cook2011} computed the first and second moments of generalized
+inverses. However, for the Moore-Penrose inverse and arbitrary
+covariance $\Sigmab$, \cite{cook2011} claims that the quantities required to
+express the mean ``do not have tractable closed-form representation.''
+The bias term, $\w^{*\top}\E[\I-\X^\dagger\X]\w^*$, has connections to
+directional statistics.  Using the SVD,
+we have the equivalent representation $\X^\dagger \X = \V \V^\top$ where $\V$
+is an element of the Stiefel manifold $V_{n,d}$ (i.e., orthonormal $n$-frames
+in $\R^d$).  The distribution of $\V$ is known as the matrix angular central
+Gaussian (MACG) distribution \citep{chikuse1990matrix}. While prior work has
+considered high dimensional limit theorems \citep{CHIKUSE1991145} as well as
+density estimation and hypothesis testing \citep{CHIKUSE1998188} on $V_{n,d}$,
+they only analyzed the invariant measure (which corresponds in our setting to
+$\Sigmab = \I$), and to our knowledge a closed form expression of
+$\E[\V\V^\top]$ where $\V$ is distributed according to MACG with arbitrary
+$\Sigmab$ remains an open question.
+
+For analyzing the rate of decay of variance and bias discrepancies (as
+defined in Section \ref{sec:asymp-conj-details}), it suffices to only consider diagonal
+covariance matrices $\Sigmab$.  This is because if $\Sigmab = \Q \D \Q^\top$ is
+its eigendecomposition and $\X\sim\Nc_{n,d}(\zero, \I_n \otimes \Q\D\Q^\top)$,
+then we have for $\W \sim \Pc\Wc(\Sigmab, n)$ that $\W \overset{d}{=} \X^\top
+  \X$ and hence, defining $\Xt\sim\Nc_{n,d}(\zero,\I_n \otimes \D)$, by linearity
+and unitary invariance of trace,
+\begin{align*}
+  \E[\tr(\W^\dagger)]
+   & = \tr\big( \E[(\X^\top\X)^\dagger] \big)
+  = \tr\Big( \Q\E\big[(\Xt^\top\Xt)^\dagger\big]\Q^\top \Big)
+  = \tr\Big( \E\big[(\Xt^\top\Xt)^\dagger\big] \Big)
+  = \E\left[\tr \big((\Xt^\top\Xt)^\dagger\big) \right].
+\end{align*}
+Similarly, we have that $\E[\X^\dagger\X]=\Q\E\big[\Xt^\dagger\Xt\big]\Q^\top$,
+and a simple calculation shows that the bias discrepancy is
+also independent of the choice of matrix $\Q$.
+
+In our experiments, we increase $d$ while keeping the aspect ratio $n/d$
+fixed and examining the rate of decay of the discrepancies.
+We estimate $\E\big[\tr(\W^\dagger)\big]$ (for the variance) and
+$\E[\I-\X^\dagger\X]$ (for the bias) through Monte Carlo sampling.
+Confidence intervals are constructed using ordinary bootstrapping for
+the variance. We rewrite the supremum over $\w$ in bias discrepancy as
+a spectral norm:
+\[\big\|\Bc(\Sigmab,n)^{-\frac12}\E[\I-\X^\dagger\X]\Bc(\Sigmab,n)^{-\frac12} -
+  \I\big\|,\]
+and apply existing methods for constructing bootstrapped operator
+norm confidence intervals described in \cite{lopes2019bootstrapping}.  To
+ensure that estimation noise is sufficiently small, we continually increase the
+number of Monte Carlo samples until the bootstrap confidence intervals are
+within $\pm 12.5\%$ of the measured discrepancies.  We found that while
+variance discrepancy required a relatively small number of trials (up
+to one thousand), estimation noise was much larger for the bias
+discrepancy, and it necessitated over two million trials to obtain
+good estimates near $d=100$.
+
+\subsection{Eigenvalue decay profiles}
+\label{sec:eig-decay-details}
+
+Letting $\lambda_i(\Sigmab)$ be the $i$th largest eigenvalue of
+$\Sigmab$, we consider the
+following eigenvalue profiles (visualized in Figure~\ref{fig:eig-decays}):
+\begin{itemize}
+  \item \texttt{diag\_linear}: linear decay, $\lambda_i(\Sigmab)= b-a i$;
+  \item \texttt{diag\_exp}: exponential decay, $\lambda_i(\Sigmab) = b\,10^{- a i} $;
+  \item \texttt{diag\_poly}: fixed-degree polynomial decay, $\lambda_i(\Sigmab) = (b-a i)^2$;
+  \item \texttt{diag\_poly\_2}: variable-degree polynomial decay, $\lambda_i(\Sigmab) = b i^{-a}$.
+\end{itemize}
+The constants $a$ and $b$ are chosen to ensure $\lambda_{\text{max}}(\Sigmab) = 1$ and
+$\lambda_{\text{min}}(\Sigmab) = 10^{-4}$ (i.e., the condition number
+$\kappa(\Sigmab) = 10^{4}$ remains constant).

### 2022-11-23 design/continuous_nips/continuous.tex
Working compile

@@
-  \documentclass[thesis.tex]{subfiles}
+\documentclass[../../thesis.tex]{subfiles}
+
 %  \usepackage[round]{natbib}
@@
 % \begin{abstract}
+
+\chapter{Exact expressions for double descent in determinantal random designs}
+\label{ch:double_descent}
+
+Building on the theory established in \cref{ch:design}, in this chapter
+we analyze a determinantal ``surrogate'' random design to develop new
+non-asymptotic theory on double descent in overparameterized linear models.
 Double descent refers to the phase transition that is exhibited by
@@
 determinant commutes with expectation.
+Some of the results presented in this chapter were first published
+in \fullcite{surrogate-design}.
+
 % \end{abstract}
@@
 % design to capture its main properties while being ``nicer'' in some useful way.
-% In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
+% In Theorem~\ref{t:asymptotic} and \Cref{sec:asymp-conj-details} we show, both
 % theoretically and empirically, that our surrogate design accurately preserves
@@
       descent curve in the under-determined regime ($n<d$).}
-    \includegraphics[width=\textwidth]{../figs/descent-intro}
+    \includegraphics[width=\textwidth]{design/figs/descent-intro}
   \end{subfigure}
@@
       Theorem \ref{t:unbiased}.}
-    \includegraphics[width=\textwidth]{../figs/descent-shrinkage}
+    \includegraphics[width=\textwidth]{design/figs/descent-shrinkage}
   \end{subfigure}
@@
 
-\noindent
-Proof of Theorem \ref{t:mse} is given in Appendix \ref{a:mse-proof}.
+Proof of Theorem \ref{t:mse} is given in \Cref{a:mse-proof}.
 For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
@@
 does \emph{not} require the responses to follow any linear model as in
-Assumption~\ref{a:linear} (proof in Appendix~\ref{s:unbiased-proof}).
+Assumption~\ref{a:linear} (proof in \Cref{s:unbiased-proof}).
 \begin{theorem}[Implicit regularization of Moore-Penrose estimator]
@@
 \section{Related work}
-\label{s:related-work}
 
@@
   \centering
-  \includegraphics[width=.5\textwidth]{../figs/descent-model-log}
+  \includegraphics[width=.5\textwidth]{design/figs/descent-model-log}
   \caption{
@@
 
-The first non-trivial property of the surrogate design $S_\mu^n$ is
-that the expected sample size is in fact always equal to $n$, which we
-prove in Appendix \ref{appx: proof-of-l-size}.
-\begin{lemma} \label{l:size}
+We first record an important property of the design $S_\mu^d$
+which can be used to construct an over-determined design for any $n>d$. A similar
+version of this result was also previously shown by
+\cite{correcting-bias-journal} for a different determinantal design.
+
+\begin{lemma}\label{l:decomposition}
+  Let $\Xb\sim S_\mu^d$ and $\X\sim \mu^K$, where
+  $K\sim\Poisson(\gamma)$. Then the matrix composed of a random
+  permutation of the rows from $\Xb$ and $\X$ is distributed according to
+  $S_\mu^{d+\gamma}$.
+\end{lemma}
+
+\begin{proof}
+  Let $\Xt$ denote the matrix constructed from the permuted rows of
+  $\Xb$ and $\X$.  Letting $\Z\sim\mu^{K+d}$, we derive the probability
+  $\Pr\big\{\Xt\!\in\! E\big\}$ by summing over the possible index subsets  $S\subseteq
+    [K+d]$ that correspond to the rows coming from $\Xb$:
+  \begin{align*}
+    \Pr\big\{\Xt\in E\big\} & = \E\bigg[\frac{1}{\binom{K+d}{d}}
+      \sum_{S:\,|S|=d}\frac{\E[\det(\Z_{S,*})^2\one_{[\Z\in E]}\mid
+    K]}{d!\det(\Sigmab_\mu)}\bigg]                               \\
+                            & =\sum_{k=0}^\infty
+    \frac{\gamma^k\ee^{-\gamma}}{k!}\,\frac{\gamma^dk!}{(k+d)!}\,
+    \frac{\E\big[\sum_{S:\,|S|=d}\det(\Z_{S,*})^2\one_{[\Z\in E]}\mid
+    K=k\big]}{\det(\gamma\Sigmab_\mu)}                           \\
+                            & \overset{(*)}{=} \sum_{k=0}^\infty
+    \frac{\gamma^{k+d}\ee^{-\gamma}}{(k+d)!}
+    \,\frac{\E[\det(\Z^\top\Z)\one_{[\Z\in E]}\mid K=k]}{\det(\gamma\Sigmab_\mu)},
+  \end{align*}
+  where $(*)$ uses the Cauchy-Binet formula to sum over all subsets $S$
+  of size $d$. Finally, since the sum shifts from $k$
+  to $k+d$, the last expression can be rewritten as
+  $\E[\det(\X^\top\X)\one_{[\X\in E]}]/\det(\gamma\Sigmab_\mu)$, where recall that
+  $\X\sim\mu^K$ and $K\sim\Poisson(\gamma)$, matching the definition of $S_\mu^{d+\gamma}$.
+\end{proof}
+
+Another non-trivial property of the surrogate design $S_\mu^n$ is
+that the expected sample size is in fact always equal to $n$.
+\begin{lemma} \label{l:expected_size_n}
   Let $\Xb\sim S_\mu^n$ for any $n>0$.
@@
 
-\noindent
+\begin{proofof}{Lemma}{\ref{l:expected_size_n}}
+  The result is obvious when $n=d$, whereas
+  for $n>d$ it is an immediate consequence
+  of Lemma \ref{l:decomposition}.
+  Finally, for $n<d$ the expected sample
+  size follows as a corollary of Lemma \ref{l:proj}, which states that
+  \begin{align*}
+    \text{(Lemma \ref{l:proj})} \qquad\E\big[\I - \Xb^\dagger\Xb\big] =
+    (\gamma_n\Sigmab_\mu + \I)^{-1},
+  \end{align*}
+  where $\Xb^\dagger\Xb$ is the orthogonal projection onto
+  the subspace spanned by the rows of $\Xb$. Since the rank of this
+  subspace is equal to the number of the rows, we have
+  $\#(\Xb)=\tr(\Xb^\dagger\Xb)$, so
+  \begin{align*}
+    \E\big[\#(\Xb)\big] = d - \tr\big((\gamma_n\Sigmab_\mu +
+    \I)^{-1}\big) =
+    \tr\big(\gamma_n\Sigmab_\mu(\gamma_n\Sigmab_\mu+\I)^{-1}\big) = n,
+  \end{align*}
+  which completes the proof.
+\end{proofof}
+
 Our general template for computing expectations under
@@
 the complement of the row-span of $\Xb$, is proven in
-Appendix \ref{s:unbiased-proof}.
+\Cref{s:unbiased-proof}.
 \begin{lemma}\label{l:proj}
@@
 No such expectation formula is known for i.i.d.~designs, except when
-$\mu$ is an isotropic Gaussian. In Appendix \ref{s:unbiased-proof}, we
+$\mu$ is an isotropic Gaussian. In \Cref{s:unbiased-proof}, we
 also prove a generalization of Lemma \ref{l:proj} which is then used
@@
 for the trace of the Moore-Penrose inverse of the covariance
-matrix for a surrogate design (proof in Appendix \ref{a:mse-proof}).
+matrix for a surrogate design (proof in \Cref{a:mse-proof}).
 \begin{lemma}\label{l:sqinv-all}
@@
 
-\noindent
 We next give a few simple examples to provide some intuition. First, note
@@
 
-\noindent
 In fact, it can be shown that all random matrices with independent entries
@@
 
-\noindent
+We use $\adj(\A)$ to denote the adjugate of $\A$, defined as follows: the
+$(i,j)$th entry of $\adj(\A)$ is
+$(-1)^{i+j}\det(\A_{[n]\backslash\{j\},[n]\backslash\{i\}})$.
+We will use two useful identities related to the adjugate: (1)
+$\adj(\A)=\det(\A)\A^{-1}$ for invertible $\A$, and (2)
+$\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$
+\citep[see Fact 2.14.2 in][]{matrix-mathematics}.
+Note that from the definition of an adjugate matrix it immediately follows that if $\A$ is
+determinant preserving then adjugate commutes with expectation for this matrix:
+\begin{align}
+  \E\big[\big(\!\adj(\A)\big)_{i,j}\big] & =
+  \E\big[(-1)^{i+j}\det(\A_{[d]\backslash\{j\},[d]\backslash\{i\}})\big]\nonumber
+  \\
+                                         & =(-1)^{i+j}\det\!\big(\E[\A_{[d]\backslash\{j\},[d]\backslash\{i\}}]\big)
+  \\
+                                         & = \big(\!\adj(\E[\A])\big)_{i,j}.\label{eq:adj}
+\end{align}
+The adjugate is useful in our analysis because it connects the
+determinant and the inverse via the formula
+$\adj(\A)=\det(\A)\A^{-1}$, which holds for any invertible $\A$.
+
 To construct more complex examples, we show that determinant preserving random matrices are
@@
 \cite{determinantal-averaging} in the proof of Lemma~7, for computing
-the expected determinant of the sum of rank-1 random matrices (proof in Appendix \ref{a:dp}).
+the expected determinant of the sum of rank-1 random matrices.
 \begin{lemma}[Closure properties]\label{t:ring}
@@
 
-\noindent
+\begin{proofof}{Lemma}{\ref{t:ring}} \
+  First, we show that $\A+\u\v^\top$ is d.p.~for any fixed
+  $\u,\v\in\R^d$. Below, we use the identity for a rank one
+  update of a determinant:
+  $\det(\A+\u\v^\top)=\det(\A)+\v^\top\!\adj(\A)\u$. It follows that
+  for any $\Ic$ and $\Jc$ of the same size,
+  \begin{align*}
+    \E\big[\!\det(\A_{\Ic,\Jc}\!+\u_{\Ic}\v_{\Jc}^\top)\big] & =
+    \E\big[\!\det(\A_{\Ic,\Jc}) +
+    \v_{\Jc}^\top\adj(\A_{\Ic,\Jc}) \u_{\Ic}\big]                                                                           \\
+                                                             & \overset{(*)}{=}\det\!\big(\E[\A_{\Ic,\Jc}]\big) +
+    \v_{\Jc}^\top\adj\!\big(\E[\A_{\Ic,\Jc}]\big) \u_{\Ic}                                                                  \\
+                                                             & =\det\!\big(\E[\A_{\Ic,\Jc} \!+ \u_{\Ic}\v_{\Jc}^\top]\big),
+  \end{align*}
+  where $(*)$ used \eqref{eq:adj}, i.e., the fact that for d.p.~matrices, adjugate commutes
+  with expectation. Crucially, through the definition of an adjugate
+  this step implicitly relies on the assumption that all the square
+  submatrices of $\A_{\Ic,\Jc}$ are also  determinant preserving.
+  Iterating this, we get that $\A+\Z$ is d.p.~for any fixed
+  $\Z$. We now show the same for $\A+\B$:
+  \begin{align*}
+    \E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\big]
+     & =
+    \E\Big[\E\big[\!\det(\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc})\mid\B\big]\Big]                \\
+     & \overset{(*)}{=}\E\Big[\!\det\!\big(\E[\A_{\Ic,\Jc}]\!+\B_{\Ic,\Jc}\big)\Big] \\
+     & = \det\!\big(\E[\A_{\Ic,\Jc}\!+\B_{\Ic,\Jc}]\big),
+  \end{align*}
+  where $(*)$  uses the fact that after conditioning on $\B$ we can
+  treat it as a fixed matrix. Next, we show that $\A\B$ is determinant preserving via the Cauchy-Binet formula:
+  \begin{align*}
+    \E\big[\!\det\!\big((\A\B)_{\Ic,\Jc}\big)\big]
+     & = \E\big[\!\det(\A_{\Ic,*}\B_{*,\Jc})\big]                  \\
+     & =\E\bigg[\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\A_{\Ic,S}\big)
+    \det\!\big(\B_{S,\Jc}\big)\bigg]                               \\
+     & =\!\!\sum_{S:\,|S|=|\Ic|}\!\!\det\!\big(\E[\A]_{\Ic,S}\big)
+    \det\!\big(\E[\B]_{S,\Jc}\big)                                 \\
+     & =\det\!\big(\E[\A]_{\Ic,*}\, \E[\B]_{*,\Jc}\big)            \\
+     & = \det\!\big(\E[\A\B]_{\Ic,\Jc}\big),
+  \end{align*}
+  where recall that $\A_{\Ic,*}$ denotes the submatrix of $\A$
+  consisting of its (entire) rows indexed by $\Ic$.
+\end{proofof}
+
 Next, we introduce another important class of d.p.~matrices:
@@
 extension of an expectation formula given by \cite{dpp-intermediate}
-for sampling from discrete distributions (proof in Appendix \ref{a:dp}).
+for sampling from discrete distributions.
 \begin{lemma}\label{l:poisson}
@@
 
+To prove Lemma \ref{l:poisson}, we will use the following
+lemma, many variants of which appeared in the literature
+\cite[e.g.,][]{expected-generalized-variance}. We use the one given by
+\cite{correcting-bias}.
+\begin{lemma}[\cite{correcting-bias}]\label{l:cb}
+  If the rows of random $k\times d$ matrices $\A,\B$
+  are sampled as an i.i.d.~sequence of $k\geq d$ pairs of joint random vectors, then
+  \begin{align}
+    k^d\,\E \big[\det(\A^\top\B)\big]
+     & = \ktd\,\det\!\big(\E[\A^\top\B]\big).
+  \end{align}
+\end{lemma}
+
 \noindent
+Here, we use the following standard shorthand: $\ktd =
+  \frac{k!}{(k-d)!} = k\,(k-1)\dotsm(k-d+1)$. Note that the above result
+almost looks like we are claiming that the matrix $\A^\top\B$ is d.p.,
+but in fact it is not because $k^d\neq \ktd$. The difference
+in those factors is precisely what we are going to correct with the
+Poisson random variable. We now present the proof of Lemma
+\ref{l:poisson}.
+\begin{proofof}{Lemma}{\ref{l:poisson}}
+  Without loss of generality, it suffices to check Definition \ref{d:main} with both $\Ic$ and
+  $\Jc$ equal $[d]$. We first expand the expectation by
+  conditioning on the value of $K$ and letting $\gamma=\E[K]$:
+  \begin{align*}
+    \E\big[\!\det(\A^\top\B)\big]
+          & = \sum_{k=0}^\infty
+    \E\big[\det(\A^\top\B)\mid K\!=\!k\big]\
+    \Pr(K\!=\!k)                     \\
+    \text{(Lemma \ref{l:cb})}
+    \quad & =
+    \sum_{k=d}^\infty\frac{k! k^{-d}}{(k-d)!}\det\!\big(\E[\A^\top\B\mid
+      K\!=\!k]\big)
+    \frac{\gamma^k\ee^{-\gamma}}{k!} \\
+          & =\sum_{k=d}^\infty
+    \Big(\frac\gamma k\Big)^d\det\!\big(\E[\A^\top\B\mid K\!=\!k]\big)
+    \frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}.
+    %  \\
+    %  &=\sum_{k=0}^\infty \det\!\big(\E[\A^\top\B]\big)\,\Pr(K\!=\!k)
+    % \ =\ \det\!\big(\E[\A^\top\B]\big).
+  \end{align*}
+  Note that $\frac\gamma k\,\E[\A^\top\B\mid K\!=\!k]=\E[\A^\top\B]$,
+  which is independent of $k$. Thus we can rewrite the above
+  expression as:
+  \begin{align*}
+    \det\!\big(\E[\A^\top\B]\big)\sum_{k=d}^\infty\frac{\gamma^{k-d}\ee^{-\gamma}}{(k-d)!}
+    =
+    \det\!\big(\E[\A^\top\B]\big)\sum_{k=0}^\infty
+    \frac{\gamma^{k}\ee^{-\gamma}}{k!}=\det\!\big(\E[\A^\top\B]\big),
+  \end{align*}
+  which concludes the proof.
+\end{proofof}
+
 Finally, we show the expectation formula needed for obtaining the
@@
 \ref{t:mse} and~\ref{t:unbiased}. The proof uses
-Lemmas \ref{t:ring} and \ref{l:poisson} (see Appendix \ref{a:dp}).
+Lemmas \ref{t:ring} and \ref{l:poisson}.
 \begin{lemma}\label{l:normalization}
@@
 \end{lemma}
+
+To prove Lemma \ref{l:normalization}, we use the following standard
+determinantal formula which is used to derive the normalization
+constant of a discrete determinantal point process.
+\begin{lemma}[\cite{dpp-ml}]\label{l:det-standard}
+  For any $k\times d$ matrices $\A,\B$ we have
+  \[\det(\I+\A\B^\top)=\sum_{S\subseteq[k]}\det(\A_{S,*}\B_{S,*}^\top).\]
+\end{lemma}
+
+\begin{proofof}{Lemma}{\ref{l:normalization}}
+  By Lemma \ref{l:poisson}, the matrix $\B^\top\A$ is determinant
+  preserving. Applying Lemma \ref{t:ring} we conclude that
+  $\I+\B^\top\A$ is also d.p., so
+  \begin{align*}
+    \det\!\big(\I+\E[\B^\top\A]\big) = \E\big[\det(\I+\B^\top\A)\big] =\E\big[\det(\I+\A\B^\top)\big],
+  \end{align*}
+  where the second equality is known as Sylvester's Theorem.
+  We rewrite the expectation of $\det(\I+\A\B^\top)$ by applying Lemma
+  \ref{l:det-standard}.  Letting $\gamma=\E[K]$, we obtain:
+  \begin{align*}
+    \E\big[\det(\I+\A\B^\top)\big] & =\E\bigg[\sum_{S\subseteq [K]}\E\big[\det(\A_{S,*}\B_{S,*}^\top)\mid
+    K\big]\bigg]                                                                                          \\
+                                   & \overset{(*)}{=}\sum_{k=0}^\infty\frac{\gamma^k\ee^{-\gamma}}{k!}
+    \sum_{i=0}^k\binom{k}{i} \E\big[\det(\A\B^\top)\mid K=i\big]                                          \\
+                                   & =\sum_{i=0}^\infty \E\big[\det(\A\B^\top)\mid K=i\big]
+    \sum_{k\geq i}^\infty \binom{k}{i}
+    \frac{\gamma^k\ee^{-\gamma}}{k!}                                                                      \\
+                                   & =\sum_{i=0}^\infty
+    \frac{\gamma^i\ee^{-\gamma}}{i!}\E\big[\det(\A\B^\top)\mid K=i\big]
+    \sum_{k\geq i}^\infty\frac{\gamma^{k-i}}{(k-i)!} = \E\big[\det(\A\B^\top)\big]\cdot\ee^\gamma,
+  \end{align*}
+  where $(*)$ follows from the exchangeability of the rows of $\A$ and
+  $\B$, which implies that the distribution of $\A_{S,*}\B_{S,*}^\top$ is the
+  same for all subsets $S$ of a fixed size $k$.
+\end{proofof}
+
+\section{Proof of Theorem \ref{t:mse}}
+\label{a:mse-proof}
+In this section we use $Z_\mu^n$ to denote the normalization
+constant that appears in \eqref{eq:cases} when computing an expectation for surrogate design
+$S_\mu^n$.
+We first prove Lemma \ref{l:sqinv-all}. % by splitting the under- and
+% over-determined cases and start with proving the former.
+% Note that those
+% results require $\mu$ to satisfy general position (Assumption \ref{a:general-position}),
+% which implies that if $\X\sim\mu^k$ for $k\leq d$ then $\rank(\X)=k$.
+\begin{lemma}[restated Lemma \ref{l:sqinv-all}]\label{l:sqinv-under}
+  If  $\Xb\sim S_\mu^n$ for $n<d$, then we have
+  \begin{align*}
+    \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
+     & ={\gamma_n}\big(1- \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big).
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  Let $\X\sim\mu^K$ for $K\sim\Poisson({\gamma_n})$. Note that if
+  $\det(\X\X^\top)>0$ then using the fact that
+  $\det(\A)\A^{-1}=\adj(\A)$ for any invertible matrix $\A$, we can write:
+  \begin{align*}
+    \det(\X\X^\top)\tr\big((\X^\top\X)^{\dagger}\big)
+     & = \det(\X\X^\top)\tr\big((\X\X^\top)^{-1}\big) \\
+     & = \tr(\adj(\X\X^\top))                         \\[-1mm]
+     & = \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top),
+  \end{align*}
+  where $\X_{-i}$ is a shorthand for $\X_{[K]\backslash\{i\},*}$.
+  Assumption \ref{a:general-position} ensures that
+  $\Pr\big\{\det(\X\X^\top)>0\big\}=1$, which allows us to write:
+  \begin{align*}
+    Z_\mu^n\cdot \E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
+     & =\E\bigg[
+      \sum_{i=1}^K\det(\X_{-i}\X_{-i}^\top)\ \big|\
+    \det(\X\X^\top)>0\bigg]\cdot\overbrace{\Pr\big\{\det(\X\X^\top)>0\big\}}^{1}   \\
+     & =\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\E\Big[
+    \sum_{i=1}^k\det(\X_{-i}\X_{-i}^\top)\ \big|\  K=k\Big]                        \\
+     & =\sum_{k=0}^d\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}\, k\
+    \E\big[\det(\X\X^\top)\mid K=k-1\big]                                          \\
+     & =\gamma_n\sum_{k=0}^{d-1}\frac{\gamma_n^{k}\ee^{-\gamma_n}}{k!}
+    \E\big[\det(\X\X^\top)\mid K=k\big]                                            \\
+     & =\gamma_n\Big(\E\big[\det(\X\X^\top)\big]\ -\
+    \frac{\gamma_n^{d}\ee^{-\gamma_n}}{d!}\E\big[\det(\X)^2\mid K=d\big]
+    \Big)                                                                          \\
+     & \overset{(*)}{=}\gamma_n\big(\ee^{-\gamma_n}\det(\I +\gamma_n\Sigmab_\mu) -
+    \ee^{-\gamma_n}\det(\gamma_n\Sigmab_\mu)\big),
+  \end{align*}
+  where $(*)$ uses Lemma \ref{l:normalization} for the first term and
+  Lemma \ref{l:cb} for the second term. We obtain the desired result by
+  dividing both sides by
+  $Z_\mu^n=\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)$.
+\end{proof}
+In the over-determined regime, a more general matrix expectation
+formula can be shown (omitting the trace). The following result is
+related to an expectation formula derived by
+\cite{correcting-bias-journal}, however they use a slightly
+different determinantal design so the results are incomparable.
+\begin{lemma}\label{l:sqinv-over}
+  If $\Xb\sim S_\mu^n$ and $n>d$, then we
+  have
+  \begin{align*}
+    \E\big[ (\Xb^\top\Xb)^{\dagger}\big] =
+    \Sigmab_\mu^{-1}\cdot \frac{1-\ee^{-\gamma_n}}{\gamma_n}.
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$. Assumption
+  \ref{a:general-position} implies that for $K\neq d-1$ we have
+  \begin{align}
+    \det(\X^\top\X)(\X^\top\X)^\dagger=\adj(\X^\top\X),\label{eq:adj-over}
+  \end{align}
+  however when $k=d-1$ then \eqref{eq:adj-over} does not hold because
+  $\det(\X^\top\X)=0$ while $\adj(\X^\top\X)$ may be non-zero. It
+  follows that:
+  \begin{align*}
+    Z_\mu^n\cdot
+    \E\big[ (\Xb^\top\Xb)^{\dagger}\big]
+     & =\E\big[\det(\X^\top\X)(\X^\top\X)^\dagger\big]                                      \\
+     & =\E\big[\adj(\X^\top\X)\big]-
+    \frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)!}
+    \E\big[\adj(\X^\top\X)\mid K=d-1\big]                                                   \\
+     & \overset{(*)}{=}\adj\!\big(\E[\X^\top\X]\big) -
+    \frac{\gamma_n^{d-1}\ee^{-\gamma_n}}{(d-1)^{d-1}}
+    \adj\!\big(\E[\X^\top\X\mid K=d-1]\big)                                                 \\
+     & =\adj(\gamma_n\Sigmab_\mu) - \ee^{-\gamma_n}\adj(\gamma_n\Sigmab_\mu)                \\
+     & =\det(\gamma_n\Sigmab_\mu)\,(\gamma_n\Sigmab_\mu)^{-1}(1-\ee^{-\gamma_n})            \\
+     & =\det(\gamma_n\Sigmab_\mu)\,\Sigmab_\mu^{-1}\cdot\frac{1-\ee^{-\gamma_n}}{\gamma_n},
+  \end{align*}
+  where the first term in $(*)$ follows from Lemma
+  \ref{l:normalization} and \eqref{eq:adj}, whereas the second term comes
+  from Lemma 2.3 of \cite{correcting-bias-journal}.
+  Dividing both sides by $Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ completes the proof.
+\end{proof}
+
+Applying the closed form expressions from Lemmas
+\ref{l:proj}, \ref{l:sqinv-all} and \ref{l:sqinv-over}, we derive
+the formula for the MSE and prove Theorem \ref{t:mse} (we defer the
+proof of Lemma \ref{l:proj} to \Cref{s:unbiased-proof}).
+\begin{proofof}{Theorem}{\ref{t:mse}}
+  First, assume that $n<d$, in which case we have
+  $\gamma_n=\frac1{\lambda_n}$ and moreover
+  \begin{align*}
+    n & = \tr\big(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1}\big)                          \\
+      & =\tr\big((\Sigmab_\mu+\lambda_n\I-\lambda_n\I)(\Sigmab_\mu+\lambda_n\I)^{-1}\big) \\
+      & =d - \lambda_n\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big),
+    %\frac{\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)}{d-n},
+  \end{align*}
+  so we can write $\lambda_n$ as $(d-n)/\tr((\Sigmab_\mu+\lambda_n\I)^{-1})$.
+  From this and Lemmas \ref{l:proj} and \ref{l:sqinv-under}, we
+  obtain the desired expression, where recall
+  that $\alpha_n = \det\!\big(\Sigmab_\mu (\Sigmab_\mu+\frac1{\gamma_n})^{-1}\big)$:
+  \begin{align*}
+    \MSE{\Xb^\dagger\ybb} & = \sigma^2\,\gamma_n(1-\alpha_n) +
+    \tfrac1{\gamma_n} \,\w^{*\top}(\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\w^*
+    \\
+                          & \overset{(a)}{=}\sigma^2\,\frac{1-\alpha_n}{\lambda_n} +
+    \lambda_n\,\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*                                                           \\
+                          & \overset{(b)}{=}\sigma^2\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\frac{1-\alpha_n}{d-n}
+    +
+    (d-n)\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
+    {\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)}.
+  \end{align*}
+  While the expression given after $(a)$ is simpler than the one
+  after $(b)$, the latter better illustrates how the MSE depends on
+  the sample size $n$ and the dimension $d$.
+  Now, assume that $n>d$. In this case, we have $\gamma_n=n-d$ and apply Lemma
+  \ref{l:sqinv-over}:
+  \begin{align*}
+    \MSE{\Xb^\dagger\ybb}
+     & = \sigma^2\,\tr(\Sigmab_\mu^{-1})\,
+    \frac{1-\ee^{-\gamma_n}}{\gamma_n}
+    =\sigma^2\,\tr(\Sigmab_\mu^{-1})
+    \,\frac{1-\beta_n}{n-d}.
+  \end{align*}
+  The case of $n=d$ was shown in Theorem~2.12 of \cite{correcting-bias-journal}.
+  This concludes the proof.
+\end{proofof}
+
+\section{Proof of Theorem \ref{t:unbiased}}
+\label{s:unbiased-proof}
+
+As in the previous section, we use $Z_\mu^n$ to denote the normalization
+constant that appears in \eqref{eq:cases} when computing an expectation
+for surrogate design $S_\mu^n$.
+Recall that our goal is to compute the expected value of
+$\Xb^\dagger\ybb$ under the surrogate design $S_\mu^n$. Similarly as for Theorem
+\ref{t:mse}, the case of $n=d$ was shown in Theorem 2.10 of
+\cite{correcting-bias-journal}. We break the rest down into the
+under-determined case $(n<d)$ and the over-determined case ($n>d$),
+starting with the former. Recall that we do \emph{not} require any
+modeling assumptions on the responses.
+\begin{lemma}\label{l:ridge-under}
+  If $\Xb\sim S_\mu^n$ and $n<d$, then for any $y(\cdot)$
+  such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
+  denoting $\yb_i$ as $y(\xbb_i)$, we have
+  \begin{align*}
+    \E\big[\Xb^\dagger \ybb\big]
+     & =
+    \big(\Sigmab_\mu+\tfrac1{\gamma_n}\I\big)^{-1}\E_{\mu,y}[y(\x)\,\x].
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
+  $y(\x_i)$ as $y_i$.
+  Note that when $\det(\X\X^\top)>0$, then
+  the $j$th entry of $\X^\dagger\y$ equals
+  $\f_j^\top(\X\X^\top)^{-1}\y$, where $\f_j$ is the $j$th
+  column of $\X$, so:
+  \begin{align*}
+    \det(\X\X^\top)\,(\X^\dagger\y)_j
+     & = \det(\X\X^\top)\, \f_j^\top(\X\X^\top)^{-1}\y \\
+     & =
+    \det(\X\X^\top+\y\f_j^\top) - \det(\X\X^\top).
+  \end{align*}
+  If $\det(\X\X^\top)=0$, then also
+  $\det(\X\X^\top+\y\f_j^\top)=0$, so we can write:
+  \begin{align*}
+    Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
+     & =  \E\big[\det(\X\X^\top)(\X^\dagger\y)_j\big]                                   \\
+     & = \E\big[\det(\X\X^\top+\y\f_j^\top)-\det(\X\X^\top)\big]                        \\
+     & =\E\big[\det\!\big([\X,\y][\X,\f_j]^\top\big)\big] - \E\big[\det(\X\X^\top)\big] \\
+     & \overset{(a)}{=}\ee^{-\gamma_n}\det\!\bigg(\I +
+    \gamma_n\,\E_{\mu,y}\bigg[\begin{pmatrix}\x\x^\top &
+               \!\!\x\, y(\x) \\ x_j\,\x^\top&\!\!
+               x_j\,y(\x)\end{pmatrix}\bigg]\bigg)
+    -\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)                                        \\
+     & \overset{(b)}{=}\ee^{-\gamma_n}\det(\I+\gamma_n\Sigmab_\mu)                      \\
+     & \qquad \times \Big(\E_{\mu,y}\big[\gamma_n
+        x_j\,y(\x)\big] - \E_{\mu}\big[\gamma_n
+        x_j\,\x^\top\big](\I+\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[\gamma_n\x\,
+        y(\x)\big]\Big),
+  \end{align*}
+  where $(a)$ uses Lemma \ref{l:normalization} twice, with the first
+  application involving two different matrices $\A=[\X,\y]$ and
+  $\B=[\X,\f_j]$, whereas $(b)$ is a standard determinantal identity
+  \cite[see Fact 2.14.2 in][]{matrix-mathematics}.
+  Dividing both sides by $Z_\mu^n$ and letting $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$, we obtain that:
+  \begin{align*}
+    \E\big[\Xb^\dagger\ybb\big]
+     & = \gamma_n\v_{\mu,y} - \gamma_n^2\Sigmab_\mu(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y} \\
+     & =\gamma_n\big(\I - \gamma_n\Sigmab_\mu (\I+\gamma_n\Sigmab_\mu)^{-1}\big)\v_{\mu,y}
+    =\gamma_n(\I+\gamma_n\Sigmab_\mu)^{-1}\v_{\mu,y},
+  \end{align*}
+  which completes the proof.
+\end{proof}
+We return to Lemma \ref{l:proj}, regarding the expected orthogonal
+projection onto the complement of the row-span of $\Xb$, i.e.,
+$\E[\I-\Xb^\dagger\Xb]$, which follows as a corollary of Lemma~\ref{l:ridge-under}.
+\begin{proofof}{Lemma}{\ref{l:proj}}
+  We let $y(\x)=x_j$ where $j\in[d]$ and apply Lemma
+  \ref{l:ridge-under} for each $j$, obtaining:
+  \begin{align*}
+    \I - \E\big[\Xb^\dagger\Xb] = \I -
+    (\Sigmab_\mu+\tfrac1{\gamma_n}\I)^{-1}\Sigmab_\mu,
+  \end{align*}
+  from which the result follows by simple algebraic manipulation.
+\end{proofof}
+
+We move on to the over-determined case, where the ridge regularization
+of adding the identity to $\Sigmab_\mu$ vanishes. Recall that we
+assume throughout the paper that $\Sigmab_\mu$ is invertible.
+\begin{lemma}\label{l:ridge-over}
+  If $\Xb\sim S_\mu^n$ and $n>d$, then for any real-valued random function $y(\cdot)$
+  such that $\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
+  denoting $\yb_i$ as $y(\xbb_i)$, we have
+  \begin{align*}
+    \E\big[\Xb^\dagger \ybb\big]
+     & =\Sigmab_\mu^{-1}\E_{\mu,y}\big[y(\x)\,\x\big].
+  \end{align*}
+\end{lemma}
+\begin{proof}
+  Let $\X\sim\mu^K$ for $K\sim\Poisson(\gamma_n)$ and denote
+  $y_i=y(\x_i)$. Similarly as in the proof of
+  Lemma~\ref{l:ridge-under}, we note that when $\det(\X^\top\X)>0$,
+  then
+  the $j$th entry of $\X^\dagger\y$ equals
+  $\e_j^\top(\X^\top\X)^{-1}\X^\top\y$, where $\e_j$ is the $j$th
+  standard basis vector, so:
+  \begin{align*}
+    \det(\X^\top\X)\,(\X^\dagger\y)_j =
+    \det(\X^\top\X)\, \e_j^\top(\X^\top\X)^{-1}\X^\top\y =
+    \det(\X^\top\X+\X^\top\y\e_j^\top) - \det(\X^\top\X).
+  \end{align*}
+  If $\det(\X^\top\X)=0$, then also
+  $\det(\X^\top\X+\X^\top\y\e_j^\top)=0$. We proceed to compute the
+  expectation:
+  \begin{align*}
+    Z_\mu^n\cdot\E\big[(\Xb^\dagger\ybb)_j\big]
+     & =  \E\big[\det(\X^\top\X)(\X^\dagger\y)_j\big]                                     \\
+     & = \E\big[\det(\X^\top\X+\X^\top\y\e_j^\top)-\det(\X^\top\X)\big]                   \\
+     & =\E\big[\det\!\big(\X^\top(\X+\y\e_j^\top)\big)\big] - \E\big[\det(\X^\top\X)\big] \\
+     & \overset{(*)}{=}\det\!\Big(
+    \gamma_n\,\E_{\mu,y}\big[\x(\x+ y(\x)\e_j)^\top\big]\Big)
+    -\det(\gamma_n\Sigmab_\mu)                                                            \\
+     & =\det\!\big(\gamma_n\Sigmab_\mu + \gamma_n\E_{\mu,y}[\x\,y(\x)]\e_j^\top\big)
+    -\det(\gamma_n\Sigmab_\mu)                                                            \\
+     & =\det(\gamma_n\Sigmab_\mu)\cdot
+    \gamma_n\e_j^\top(\gamma_n\Sigmab_\mu)^{-1}\E_{\mu,y}\big[y(\x)\,\x\big],
+  \end{align*}
+  where $(*)$ uses Lemma \ref{l:poisson} twice (the first time, with
+  $\A=\X$ and $\B=\X+\y\e_j^\top$). Dividing both sides by
+  $Z_\mu^n=\det(\gamma_n\Sigmab_\mu)$ concludes the proof.
+\end{proof}
+
+\noindent
+We combine Lemmas \ref{l:ridge-under} and \ref{l:ridge-over} to obtain
+the proof of Theorem \ref{t:unbiased}.
+\begin{proofof}{Theorem}{\ref{t:unbiased}}
+  The case of $n=d$ follows directly from Theorem~2.10 of
+  \cite{correcting-bias}.
+  Assume that $n<d$. Then we have
+  $\gamma_n=\frac1{\lambda_n}$, so the result follows
+  from Lemma \ref{l:ridge-under}.
+  % , noting that unlike in the lemma,
+  % Theorem \ref{t:unbiased} allows for the response model $y(\cdot)$ to be randomized:
+  % \begin{align*}
+  %   \E[\Xb^\dagger\ybb] = \E\big[\E[\Xb^\dagger\ybb\mid\ybb]\big] =
+  % % \w_{\lambda_n}^* = \argmin_\w\E_\mu\big[(\x^\top\w-y(\x))^2\big] +
+  % %   \lambda_n\|\w\|^2 =
+  % %   \big(\Sigmab_\mu+\lambda_n\I\big)^{-1}
+  % %   \E_{\mu}\big[y(\x)\,\x\big].
+  % \end{align*}
+  If $n>d$, then the result follows from Lemma \ref{l:ridge-over}.
+\end{proofof}
+
+\section{Proof of Theorem \ref{t:asymptotic}}
+\label{sec:proof-of-t-asymptotic}
+
+The proof of Theorem \ref{t:asymptotic} follows the standard decomposition of
+MSE in Equation~\ref{eq:mse-derivation}, and in the process,
+establishes consistency of the variance and bias terms
+independently. To this end, we introduce the following two useful
+lemmas that capture the limiting behavior of the
+variance and bias terms, respectively.
+
+\begin{lemma}\label{c:wishart}
+  Under the setting of Theorem~\ref{t:asymptotic}, we have, as $n,d \to \infty$
+  with $n/d \to \bar c \in (0, \infty) \setminus \{ 1 \}$ that
+  \begin{align}
+    %\bigg|\frac{\E\big[\tr((\X^\top \X)^\dagger)\big]}{\Vc(\Sigmab,n)} -1\bigg| \to 0 \qquad\text{for}\quad\Vc(\Sigmab,n)= \frac{1-\alpha_n}{\lambda_n}
+    \begin{cases}
+      \E\big[\tr((\X^\top \X)^\dagger)\big] - (1-\alpha_n) \lambda_n^{-1} \to 0,                    & \text{for }\bar c < 1, \\
+      \E\big[\tr((\X^\top \X)^\dagger)\big] - \frac{1-\beta_n}{n - d} \cdot \tr \Sigmab^{-1} \to 0, & \text{for }\bar c > 1
+    \end{cases}
+    \label{eq:wishart}
+  \end{align}
+  where $\lambda_n\geq 0$ is the unique solution to
+  $n=\tr(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$,
+  $\alpha_n=\det(\Sigmab(\Sigmab+\lambda_n\I)^{-1})$,
+  and $\beta_n = e^{d-n}$.
+\end{lemma}
+
+\noindent
+
+The second term in the MSE derivation \eqref{eq:mse-derivation}, $\E[\I -
+    \X^\dag \X]$, involves the expectation of a projection onto the orthogonal
+complement of a sub-Gaussian general position sample $\X$.
+This term is zero when $n > d$, and for $n < d$
+we prove in \cref{sec:proof-of-c-projection} that the surrogate design's
+bias $\mathcal{B}(\Sigmab, n)$ provides an asymptotically consistent approximation
+to all of the eigenvectors and eigenvalues:
+
+\begin{lemma}\label{c:projection}
+  Under the setting of Theorem~\ref{t:asymptotic}, for $\w \in \R^d$ of bounded Euclidean norm (i.e., $\| \w \| \le C'$ for all $d$), we have, as $n,d \to \infty$ with $n/d \to \bar c \in (0,1)$ that
+  \begin{align}
+    %\sup_{\w\in\R^d\backslash\{\zero\}}\bigg|\frac{\w^\top\E[\I-\X^\dagger\X]\w}{\w^\top \Bc(\Sigmab,n)\w} - 1\bigg| \to 0 \qquad\text{for}\quad\Bc(\Sigmab,n) = \lambda_n(\Sigmab+\lambda_n\I)^{-1}
+    \w^\top\E[\I-\X^\dagger\X]\w - \lambda_n \w^\top (\Sigmab + \lambda_n \I)^{-1}\w \to 0
+    \label{eq:projection}
+  \end{align}
+  while $\I - \X^\dagger \X = 0$ for $\bar c > 1$.
+\end{lemma}
+
+\subsection{Proof of \cref{c:wishart}}
+\label{sec:proof-of-c-wishart}
+
+\subsubsection{The $\bar c \in (0,1)$ case}
+
+For $n < d$, we first establish (1) $\liminf_n \lambda_n > 0$ and (2) $\alpha_n \to 0$.
+To prove (1), by hypothesis $\Sigmab \succeq c \I$ for all $d$. Since $\frac{n}{d} < 1$,
+we have (by definition of $\lambda_n$) for some $\delta > 0$
+\begin{align*}
+  1 - \delta
+  > \frac{n}{d}
+  = \frac{1}{d} \tr(\Sigmab(\Sigmab + \lambda_n \I)^{-1})
+  > \frac{c}{c + \lambda_n}
+\end{align*}
+Rearranging, we have $\lambda_n > \frac{\delta c}{1 - \delta} > 0$.
+For (2), let $(\tau_i)_{i \in [d]}$ denote the eigenvalues of $\Sigmab$.
+Since $1 - x \leq e^{-x}$ and $C \I \succeq \Sigmab \succeq c \I$ for all $d$,
+\begin{align*}
+  \alpha_n
+  = \prod_{i=1}^d \frac{\tau_i}{\tau_i + \lambda_n}
+  \leq \left(\frac{C}{C + \lambda_n}\right)^d
+  = \left( 1 - \frac{\lambda_n}{C + \lambda_n}\right)^d
+  \leq \exp\left(-d \frac{\lambda_n}{C + \lambda_n}\right)
+\end{align*}
+and since $\lambda_n > 0$ eventually as $d \to \infty$ we have
+$\alpha_n \to 0$ so that $(1-\alpha_n) \lambda_n^{-1} - \lambda_n^{-1} \to 0$.
+
+% (1) and (2) together imply $\mathcal{V}(\Sigmab, n) \to \lambda_n^{-1}$, and
+% by Slutsky's theorem
+% \begin{align*}
+%   \lim_{n,d} \left\lvert
+%   \frac{\E[\tr((\X^\top \X)^\dag)] - \mathcal{V}(\Sigmab, n)}{\mathcal{V}(\Sigmab, n)}
+%   \right\rvert
+%   = \lim_{n,d} \left\lvert
+%   \frac{\E[\tr((\X^\top \X)^\dag)] - \lambda_n^{-1}}{\lambda_n^{-1}}
+%   \right\rvert
+% \end{align*}
+% Since $\liminf_n \lambda_n > 0$ was shown in (1),
+As a consequence of (2) and Slutsky's theorem, it suffices to show
+$\tr(\X^\top \X)^\dag - \lambda_n^{-1} \overset{d}{\to} 0$ as $n,d \to \infty$.
+To do this, we consider the limiting behavior of $\tr
+  (\X^\top \X)^\dagger/n = \tr (\X \X^\top)^\dagger/n$ as $n/d \to \bar c \in
+  (0,1)$, for $\X = \Z \Sigmab^{\frac12}$ with $\Z \in \mathbb R^{n \times d}$
+having i.i.d.~zero mean, unit variance sub-Gaussian entries, i.e., the behavior
+of
+\begin{equation}\label{eq:limit1}
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left( \frac1n \X \X^\top + z \I_n \right)^{-1}
+\end{equation}
+by definition of the pseudo-inverse.
+
+The proof comes in three steps: (i) for fixed $z > 0$, consider the limiting behavior of $ \delta(z) \equiv \tr (\X \X^\top/n + z \I_n)^{-1}/n$ as $n,d \to \infty$ and state
+\begin{equation}\label{eq:limit2}
+  \lim_{n,d \to \infty} \delta(z) - m(z) \to 0
+\end{equation}
+almost surely for some $m(z)$ to be defined; (ii) show that both $\delta(z)$ and its derivate $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take $z \to 0^+$ in \eqref{eq:limit2} and state
+\begin{equation}\label{eq:limit3}
+  \lim_{z \to 0^+} \lim_{n,d \to \infty} \delta(z) - \lim_{z \to 0^+} m(z) \to 0
+\end{equation}
+almost surely, given that the limit $\lim_{z \to 0^+} m(z) \equiv m(0)$ exists and eventually (iii) exchange the two limits in \eqref{eq:limit3} with Moore-Osgood theorem, to reach
+\[
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left(\frac1n \X \X^\top + z \I_n \right)^{-1} - m(0) \to 0.
+\]
+
+Step (i) follows from \cite{silverstein1995empirical} that, we have, for $z > 0$ that
+\[
+  \delta(z) \equiv \frac1n \tr \left( \frac1n \X \X^\top  + z \I_n \right)^{-1}  - m(z) \to 0
+\]
+almost surely as $n,d \to \infty$, for $m(z)$ the unique positive solution to
+\begin{equation}\label{eq:def-m}
+  m(z) = \left( z + \frac1n \tr \Sigmab (\I + m(z) \Sigmab)^{-1} \right)^{-1}.
+\end{equation}
+
+For the above step (ii), we use the assumption $\Sigmab \succeq c \I \succ 0$
+for all $d$ large, so that with $\X = \Z \Sigmab^{\frac12}$, we have for large enough $n,d$ that
+\[
+  \lambda_{\min} (\X \X^\top/n) \ge \lambda_{\min} (\Z \Z^\top/n) \lambda_{\min} (\Sigmab) \ge \frac{c}2 (\sqrt{\bar c} - 1 )^2
+\]
+almost surely, where we used Bai-Yin theorem \cite{bai1993limit}, which states that the minimum eigenvalue of $\Z \Z^\top/n$ is almost surely larger than $( \sqrt{\bar c} - 1 )^2/2$ for $n<d$ sufficiently large.
+Note that here the case $\bar c = 1$ is excluded.
+
+Observe that
+\[
+  |\delta(z)| = \left| \frac1n \tr \left( \frac1n \X \X^\top  + z \I_n \right)^{-1} \right| \le \frac1{ \lambda_{\min} (\X \X^\top/n) }
+\]
+and similarly for its derivative, so that we are allowed to take the $z \to 0^+$ limit.
+Note that the existence of the $\lim_{z \to 0^+} m(z)$ for $m(z)$ defined in \eqref{eq:def-m} is well known, see for example \cite{ledoit2011eigenvectors}.
+Then, by Moore-Osgood theorem we finish step (iii) and by concluding that
+\[
+  \tr (\X^\top \X)^\dagger - m(0) \to 0
+\]
+for $m(0) = \lambda_n^{-1}$ the unique solution to $\lambda_n^{-1} = \left( \frac1n \tr \Sigmab (\I + \lambda_n^{-1} \Sigmab)^{-1} \right)^{-1}$, or equivalently, to
+\[
+  n = \tr \Sigmab (\Sigmab + \lambda_n \I)^{-1}
+\]
+as desired.
+
+\subsubsection{The $\bar c \in (1, \infty)$ case}
+
+First note that as $n, d \to \infty$ with $n > d$, we have $\beta_n = e^{d-n} \to 0$ and it
+it suffices to show
+\[
+  \tr (\X^\top \X)^\dagger - \frac1{n-d} \tr \Sigmab^{-1} \to 0
+\]
+almost surely to conclude the proof.
+
+In the $\bar c \in (1, \infty)$ case, it is more convenient to work on the following co-resolvent
+\[
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left( \frac1n \X^\top \X + z \I_d \right)^{-1}
+\]
+where we recall $\X^\top \X = \Sigmab^{\frac12} \Z^\top \Z \Sigmab^{\frac12} \in \R^{d \times d}$ and following the same three-step procedure as in the $\bar c < 1$ case above.
+The only difference is in step (i) we need to assess the asymptotic behavior of $\delta \equiv \tr (\X^\top \X/n + z \I_d)^{-1}/n$.
+This was established in \cite{bai1998no} where it was shown that, for $z > 0$ we have
+\[
+  \frac1n \tr (\X^\top \X/n + z \I_d)^{-1} - \frac{d}n m(z) \to 0
+\]
+almost surely as $n,d \to \infty$, for $m(z)$ the unique solution to
+\[
+  m(z) = \frac1d \tr \left( \left(1- \frac{d}n - \frac{d}n z m(z) \right) \Sigmab - z \I_d \right)^{-1}
+\]
+so that for $d < n$ by taking $z = 0$ we have
+\[
+  m(0) = \frac{n}d \frac1{n-d} \tr \Sigmab^{-1}.
+\]
+The steps (ii) and (iii) follow exactly the same line of arguments as the $\bar c < 1$ case and are thus omitted.
+
+% {\RED TODO: proof for $n>d$ sub-Gaussians. We can't appeal
+% to inverse Wishart mean anymore}
+% Now consider $n > d$ with $n/d \to \bar c \in (1, \infty)$.  Firstly notice
+% $\beta_n = e^{d - n} \to 0$, so $\lim \mathcal{V}(\Sigmab, n) = \lim
+% \frac{1}{n-d} \tr(\Sigmab^{-1})$. Noting
+% $\lim \frac{1}{n-d} \tr(\Sigmab^{-1}) = \int \tau^{-1} dH(\tau)$ (where $H$ is the
+% limiting spectral distribution of $\Sigmab$) is a constant, by Slutsky's theorem
+% it suffices to show
+% \[
+%   \tr (\X^\top \X)^\dagger - \frac1{n-d} \tr \Sigmab^{-1} \to 0
+% \]
+% Recognizing $(\X^\top \X)^\dag = (\X^\top\X)^{-1}$ as an inverse-Wishart
+% matrix when $n > d$, we have that the left-hand side is identically zero.
+
+\subsection{Proof of \cref{c:projection}}
+\label{sec:proof-of-c-projection}
+
+Since $\X^\dagger \X = \X^\top (\X \X^\top)^\dagger \X$, to prove
+\cref{c:projection}, we are interested in the limiting behavior of the
+following quadratic form
+\[
+  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w
+\]
+for deterministic $\w \in \mathbb R^{d}$ of bounded Euclidean norm (i.e., $\| \w \| \le C'$ as $n,d \to \infty$), as $n,d \to \infty$ with $n/d \to \bar c \in (0,1)$.
+The limiting behavior of the above quadratic form, or more generally, bilinear form of the type $\frac1n \w_1^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w_2$ for $\w_1, \w_2 \in \mathbb R^{d}$ of bounded Euclidean norm are widely studied in random matrix literature, see for example \cite{hachem2013bilinear}.
+
+For the proof of \Cref{c:projection} we follow the same protocol as that of
+\Cref{c:wishart}, namely: (i) we consider, for fixed $z > 0$, the limiting
+behavior of $\frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n
+  \right)^{-1} \X \w$. Note that
+\begin{align*}
+  \delta(z) & \equiv \frac1n \w^\top \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \w = \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \frac1n \X^\top \X \w \\
+            & = \| \w \|^2 - z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w
+\end{align*}
+and it remains to work on the second $z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w$ term.
+It follows from \cite{hachem2013bilinear} that
+\[
+  %\delta(z) = \| \w \|^2 - z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w  \to \| \w \|^2 - \w^\top (\I_d + m(z) \Sigmab)^{-1} \w^\top
+  z \w^\top \left( \frac1n \X^\top \X + z \I_d \right)^{-1} \w - \w^\top (\I_d + m(z) \Sigmab)^{-1} \w^\top \to 0
+\]
+almost surely as $n,d \to \infty$, where we recall $m(z)$ is the unique solution to \eqref{eq:def-m}.
+
+We move on to step (ii), under the assumption that $c \le \lambda_{\min} (\Sigmab) \le \lambda_{\max} (\Sigmab) \le C$ and $\| \w \| \le C'$, we have
+\begin{align*}
+  \lambda_{\max} \left( \frac1n \X^\top \left( \frac1n \X \X^\top + z \I_n \right)^{-1} \X \right) & \le \frac{ \lambda_{\max} (\X \X^\top/n) }{ \lambda_{\min} (\X \X^\top/n) + z } \le \frac{ \lambda_{\max} (\Z \Z^\top/n) \lambda_{\max} (\Sigmab) }{ \lambda_{\min} (\Z \Z^\top/n) \lambda_{\min} (\Sigmab) } \\
+                                                                                                   & \le 4 \frac{ (\sqrt{\bar c} + 1)^2 C }{ (\sqrt{\bar c} -1)^2 c }
+\end{align*}
+so that $\delta(z)$ remains bounded and similarly for its derivative
+$\delta'(z)$, which, by Arzela-Ascoli theorem, yields uniform
+convergence and we are allowed to take the $z \to 0^+$
+limit. Ultimately, in step (iii) we exchange the two limits with
+Moore-Osgood theorem, concluding the proof.
+
+\subsection{Finishing the proof of Theorem~\ref{t:asymptotic}}
+
+To finish the proof of Theorem~\ref{t:asymptotic}, it remains to write
+\[
+  \MSE{\X^\dagger\y} = \sigma^2\E\big[\tr\big((\X^\top\X)^{\dagger}\big)\big] +
+  \w^{*\top}\E\big[\I-\X^\dagger\X\big]\w^*
+\]
+Since $\lambda_n = \frac{d-n}{ \tr (\Sigmab + \lambda_n \I)^{-1} }$,
+by \Cref{c:wishart} and \Cref{c:projection} we have
+$\MSE{\X^\dagger\y} - \Mc(\Sigmab, \w^*, \sigma^2, n) \to 0$ as $n,d
+  \to \infty$ with $n/d \to \bar c \in (0,\infty) \setminus \{ 1 \}$,
+which concludes the proof of Theorem~\ref{t:asymptotic}.
+% \begin{align*}
+%   &\MSE{\X^\dagger\y} - \Mc(\Sigmab, \w^*, \sigma^2, n) \\
+%   &=
+%     \underbrace{\left(
+%       \sigma^2\E\big[\tr\big((\X^\top\X)^{\dagger}\big)\big]
+%       - \frac{\sigma^2(1 - \alpha_n)}{\lambda_n}
+%     \right)}_{\to 0} + \underbrace{\left(
+%       \w^{*\top}\E\big[\I-\X^\dagger\X\big]\w^*
+%       - \lambda_n \w^{*\top} (\Sigmab + \lambda_n \I)^{-1} \w^*
+%     \right)}_{\to 0}
+% \end{align*}
+%as $n,d \to \infty$ with $n/d \to \bar c \in (0,1)$.
+%This allows us to conclude the proof of
+
+% For
+% the case $n <d$, note that $\X^\dagger \X = \I$ and
+% \[
+%   \MSE{\X^\dagger\y} - \frac1{n-d} \tr (\Sigmab^{-1}) \to 0.
+% \]
 
@@
 
-% \bibliography{../pap}
+% \bibliography{design/pap}
 
@@
 
-\input{appendix}
+% \input{appendix}
 

### 2022-11-23 design/equivalents_nips/neurips_2020.tex
Updates from Overleaf
Working compile

@@
 % \documentclass{article}
-\documentclass[thesis.tex]{subfiles}
+\documentclass[../../thesis.tex]{subfiles}
 
@@
 % \begin{abstract}
+
+\chapter{Exact expectation expressions for sub-Gaussian random projections}
+\label{ch:projections}
+
 It is often desirable to reduce the dimensionality of a large dataset
@@
 guarantees are typically very different from what is observed in
-practice.  We exploit recent developments in the spectral analysis of
-random matrices to develop novel techniques that provide provably
-accurate expressions for the expected value of random projection
-matrices obtained via sketching.  These expressions can be used to
-characterize the performance of dimensionality reduction in a variety
+practice. Building on the Stieltjes transform methods employed
+in \cref{sec:proof-of-c-projection} while establishing asymptotic
+consistency of surrogate design's MSE, this chapter develops novel
+techniques
+that provide provably accurate expressions for the expected value of
+random projection matrices obtained via sketching.  These expressions
+can be used to characterize the performance of dimensionality reduction
+in a variety
 of common machine learning tasks, ranging from low-rank approximation
@@
 methods, down to lower-order effects and even constant factors.
+Some of the results here were originally published in
+\fullcite{precise-expressions}.
+
 % \end{abstract}
@@
 coming from the literature on Randomized Numerical Linear Algebra
-(RandNLA), consists of data-oblivious sketches \cite{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}. 
+(RandNLA), consists of data-oblivious sketches \citep{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}.
 Consider a large $m\times n$
@@
 the trace norm error of a sketched Nystr\"om approximation
-\cite{Williams01Nystrom,revisiting-nystrom}.
+\citep{Williams01Nystrom,revisiting-nystrom}.
 Among others, this approximation error
 appears in the analysis of sketched kernel
-ridge regression \cite{fanuel2020diversity} and Gaussian process
-regression \cite{sparse-variational-gp}. Furthermore, a variety of
+ridge regression \citep{fanuel2020diversity} and Gaussian process
+regression \citep{sparse-variational-gp}. Furthermore, a variety of
 iterative algorithms, such as randomized second-order methods for
-convex optimization \cite{Qu2015Feb,Qu2016,Gower2019,jacsketch}
+convex optimization \citep{Qu2015Feb,Qu2016,Gower2019,jacsketch}
 and linear system solvers based on
-the generalized Kaczmarz method \cite{generalized-kaczmarz},
+the generalized Kaczmarz method \citep{generalized-kaczmarz},
 have convergence guarantees which depend on the extreme eigenvalues of
@@
 regularization of the interpolating solutions in over-parameterized
-linear models \cite{surrogate-design,BLLT19_TR}.
+linear models \citep{surrogate-design,BLLT19_TR}.
 
@@
 associated matrix resolvents and Stieltjes transforms
-\cite{hachem2007deterministic,bai2010spectral}. However, our analysis
+\citep{hachem2007deterministic,bai2010spectral}. However, our analysis
 is novel in several respects:
@@
         taking $z\rightarrow 0$. Prior work (e.g.,
-      \cite{HMRT19_TR}) combined this with an exchange-of-limits argument to
+        \citep{HMRT19_TR}) combined this with an exchange-of-limits argument to
         analyze the asymptotic behavior of the residual projection.
@@
 embeddings (e.g.,
-\cite{sarlos-sketching,cw-sparse,nn-sparse,projection-cost-preserving,optimal-matrix-product}),
+\citep{sarlos-sketching,cw-sparse,nn-sparse,projection-cost-preserving,optimal-matrix-product}),
 in the sense that here our object of interest is not to obtain
@@
 Rademacher sketches where each entry $s_{ij}$ is a $\pm1$ Rademacher
-random variable \cite{achlioptas2003database}. We start by analyzing the Frobenius
+random variable \citep{achlioptas2003database}. We start by analyzing the Frobenius
 norm error $\|\A-\A\P\|_F^2=\tr\,\A^\top\A\,\P_{\!\perp}$ of
@@
 extended variant of the
-Nystr\"om method \cite{revisiting-nystrom}. A Nystr\"om approximation
+Nystr\"om method \citep{revisiting-nystrom}. A Nystr\"om approximation
 constructed from a sketching matrix $\S$ is defined as $\tilde\K =
@@
 regression, kernel machines and Independent
-Component Analysis \cite{sparse-variational-gp,Williams01Nystrom,Bach2003}. By setting
-$\A=\K^{\frac12}$, it is easy to see \cite{nystrom-multiple-descent} that
+Component Analysis \citep{sparse-variational-gp,Williams01Nystrom,Bach2003}. By setting
+$\A=\K^{\frac12}$, it is easy to see \citep{nystrom-multiple-descent} that
 the trace norm error $\|\K-\tilde\K\|_*$ is identical to the squared
@@
 Basis Function (RBF) or the Mat\'ern kernel induce a well-understood eigenvalue
-decay \cite{Santa97Gaussianregression,RasmussenWilliams06}.
+decay \citep{Santa97gaussianregression,RasmussenWilliams06}.
 
 Metrics other than the aforementioned Frobenius norm error,
-such as the spectral norm error \cite{tropp2011structure}, are also of significant
+such as the spectral norm error \citep{tropp2011structure}, are also of significant
 interest in the low-rank approximation literature. We leave these
@@
 properties determine the convergence behavior of the algorithms
-\cite{generalized-kaczmarz}. With
+\citep{generalized-kaczmarz}. With
 our new results, we can precisely characterize not only the rate of
@@
   \item \emph{Generalized Kaczmarz method}
-    \cite{generalized-kaczmarz} for approximately solving a linear system $\A\x=\b$;
-  \item \emph{Randomized Subspace Newton} \cite{Gower2019}, a second order
+        \citep{generalized-kaczmarz} for approximately solving a linear system $\A\x=\b$;
+  \item \emph{Randomized Subspace Newton} \citep{Gower2019}, a second order
         method, where we sketch the Hessian matrix.
-  \item \emph{Jacobian Sketching} \cite{jacsketch}, a
+  \item \emph{Jacobian Sketching} \citep{jacsketch}, a
         class of first order methods which use additional information via a
         weight matrix $\W$ that is sketched at every iteration.
-    % \item \emph{Iterative Adaptive Sketching} \cite{lacotte2019high},
+        % \item \emph{Iterative Adaptive Sketching} \citep{lacotte2019high},
         %   a second order method where the sketching matrices are
@@
         % \footnote{While
-      %   \cite{lacotte2019high} also use the residual projection in
+        %   \citep{lacotte2019high} also use the residual projection in
         %   their analysis, adapting our techniques is less
@@
 extensions of our techniques will apply to other algorithms,
-such as that of \cite{lacotte2019high}.
+such as that of \citep{lacotte2019high}.
 
@@
 give a result in the context of linear systems
-for the generalized Kaczmarz method \cite{generalized-kaczmarz}, but a similar
+for the generalized Kaczmarz method \citep{generalized-kaczmarz}, but a similar
 convergence analysis is given for the methods of
-\cite{Gower2019,jacsketch} in Appendix~\ref{a:newton}.
+\citep{Gower2019,jacsketch} in \Cref{a:newton}.
 \begin{corollary}\label{c:kaczmarz}
@@
 The corollary follows from Theorem \ref{t:main} combined with Theorem
-4.1 in \cite{generalized-kaczmarz}. 
+4.1 in \citep{generalized-kaczmarz}.
 Note that when $\A^\top\A$ is positive definite then
@@
 context, Theorem \ref{t:main} provides new evidence for the implicit
-  regularization conjecture posed by \cite{surrogate-design} (see
+regularization conjecture posed by \citep{surrogate-design} (see
 their Theorem 2 and associated discussion), with the amount of
@@
 \end{align*}
-While implicit regularization has received attention recently in the context of SGD algorithms for overparameterized machine learning models, it was originally discussed in the context of approximation algorithms more generally~\cite{Mah12}.
-Recent work has made precise this notion in the context of RandNLA \cite{surrogate-design}, and our results here can be viewed in terms of implicit regularization of scalable RandNLA methods.
+While implicit regularization has received attention recently in the context of SGD algorithms for overparameterized machine learning models, it was originally discussed in the context of approximation algorithms more generally~\citep{Mah12}.
+Recent work has made precise this notion in the context of RandNLA \citep{surrogate-design}, and our results here can be viewed in terms of implicit regularization of scalable RandNLA methods.
 
@@
 particularly in the context of
-RandNLA \cite{DM16_CACM,RandNLA_PCMIchapter_chapter}. This line of work includes
-i.i.d.~row sampling methods \cite{BoutsidisMD08,ridge-leverage-scores} which 
+RandNLA \citep{DM16_CACM,RandNLA_PCMIchapter_chapter}. This line of work includes
+i.i.d.~row sampling methods \citep{BoutsidisMD08,ridge-leverage-scores} which
 preserve the structure of the data, and data-oblivious methods such as
-Gaussian and Rademacher sketches \cite{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}. However,
+Gaussian and Rademacher sketches \citep{Mah-mat-rev_JRNL,tropp2011structure,woodruff2014sketching}. However,
 all of these results focus on worst-case
@@
 line of works on non-i.i.d.~row sampling with Determinantal Point Processes
-(DPP, \cite{dpps-in-randnla}). In this case, exact analysis of the low-rank approximation
-error \cite{nystrom-multiple-descent}, as well as precise convergence analysis
-of stochastic second order methods \cite{randomized-newton}, have been
+(DPP, \citep{dpps-in-randnla}). In this case, exact analysis of the low-rank approximation
+error \citep{nystrom-multiple-descent}, as well as precise convergence analysis
+of stochastic second order methods \citep{randomized-newton}, have been
 obtained. Remarkably, the expressions they obtain are analogous to
@@
 between DPPs and Gaussian sketches was recently explored by
-\cite{surrogate-design} in the context of analyzing the implicit
+\citep{surrogate-design} in the context of analyzing the implicit
 regularization effect of choosing a minimum norm solution in
@@
 distribution. Similar observations were made by
-\cite{debiasing-second-order} in the context of sketching for
+\citep{debiasing-second-order} in the context of sketching for
 regularized least squares and second order optimization. While
@@
 results directly applicable. A related family of methods uses
-the so-called Iterative Hessian Sketch (IHS) approach \cite{pilanci2016iterative,lacotte2019faster}. The
+the so-called Iterative Hessian Sketch (IHS) approach \citep{pilanci2016iterative,lacotte2019faster}. The
 similarities between IHS and the Subspace Newton-type methods (see
-\cite{Qu2015Feb} for a comparison) suggest that our techniques could be
+\citep{Qu2015Feb} for a comparison) suggest that our techniques could be
 extended to provide precise convergence guarantees also to the IHS.
 Finally, yet another family of Hessian sketching methods has been studied by
-\cite{roosta2019sub,sketched-ridge-regression,XRM17_theory_TR,YXRM18_TR,fred_newtonMR_TR,distributed-newton,determinantal-averaging}.
+\citep{roosta2019sub,sketched-ridge-regression,XRM17_theory_TR,YXRM18_TR,fred_newtonMR_TR,distributed-newton,determinantal-averaging}.
 These methods  preserve the  rank of the
@@
 guarantees do not rely on the residual projection.
+
+\section{Convergence analysis of randomized iterative methods}
+\label{a:newton}
+
+Here, we discuss how our surrogate expressions for the expected
+residual projection can be used to perform convergence analysis for
+several randomized iterative optimization methods discussed in
+Section~\ref{s:newton}.
+
+\subsection{Generalized Kaczmarz method}
+Generalized Kaczmarz \citep{generalized-kaczmarz} is an iterative
+method for solving an $m\times n$ linear system
+$\A\x=\b$, which uses a $k\times m$ sketching matrix $\S_t$ to reduce
+the linear system and update an iterate $\x^t$ as follows:
+\begin{align*}
+  \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S_t\A\x=\S_t\b.
+\end{align*}
+Assume that $\x^*$ is the unique solution to the linear system $\A\x=\b$. In Theorems 4.1 and 4.6, \citep{generalized-kaczmarz} show that the
+expected trajectory of the generalized Kaczmarz iterates, as
+they converge to $\x^*$, is controlled
+by the projection matrix $\P=(\S_t\A)^\dagger\S_t\A$ as follows:
+\begin{align*}
+  \text{(\citep{generalized-kaczmarz}, Theorem 4.1)}\qquad\quad \ \ \,
+  \E[\x^{t+1}-\x^*]
+   & = \big(\I-\E[\P]\big)\,\E[\x^t-\x^*],         \\
+  \text{(\citep{generalized-kaczmarz}, Theorem 4.6)}\qquad
+  \E\big[\|\x^{t+1}-\x^*\|^2\big]
+   & \leq (1-\kappa)\,\E\big[\|\x^t-\x^*\|^2\big],
+  \text{ where }\kappa=\lambda_{\min}\big(\E[\P]\big).
+\end{align*}
+Both of these results depend on the expected projection
+$\E[\P]$. The first one describes the expected trajectory of the
+iterate, whereas the second one gives the worst-case convergence
+rate in terms of the so-called \emph{stochastic condition number}
+$\kappa$. We next demonstrate how Theorem \ref{t:main} can be used  in
+combination with the above results to
+obtain convergence analysis for generalized Kaczmarz which is
+formulated in terms of the spectral properties of $\A$. This includes
+precise expressions for both the expected trajectory and $\kappa$. The following
+result is a more detailed version of Corollary \ref{c:kaczmarz} from
+Section \ref{s:newton}.
+\begin{corollary}\label{c:kaczmarz2}
+  Let $\sigma_i$ denote the singular values of $\A$, and let $k$ denote
+  the size of sketch $\S_t$. Define:
+  \begin{align*}
+    \Delta_t=\x^t-\x^*\quad\text{and}\quad  \bar\Delta_{t+1} =
+    (\gamma\A^\top\A+\I)^{-1}\E[\Delta_t]\quad\text{s.t.}\quad
+    \sum_i\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1}=k.
+  \end{align*}
+  Suppose that $\S_t$ has i.i.d.~mean-zero sub-gaussian entries and let
+  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$.
+  Assume that $\rho = r/k$ is a constant larger than $1$. Then, the
+  expected trajectory satisfies:
+  \begin{align}
+    \big\|\E[\Delta_{t+1}] - \bar\Delta_{t+1}\big\|
+    \leq \epsilon\cdot \|\bar\Delta_{t+1}\|,
+    \quad\text{for}\quad\epsilon=O\big(\tfrac1{\sqrt
+      r}\big).\label{eq:trajectory}
+  \end{align}
+  Moreover, we obtain the following worst-case convergence guarantee:
+  \begin{align}
+    \E\big[\|\Delta_{t+1}\|^2\big]
+    \leq \big(1-
+    (\bar\kappa-\epsilon)\big)\,\E\big[\|\Delta_t\|^2\big],
+    \quad\text{where}\quad
+    \bar\kappa = \frac{\sigma_{\min}^2}{\sigma_{\min}^2+1/\gamma}.\label{eq:worst-case}
+  \end{align}
+\end{corollary}
+\begin{remark}
+  Our worst-case convergence guarantee \eqref{eq:worst-case}
+  requires the matrix $\A$ to be sufficiently well-conditioned so
+  that $\bar\kappa-\epsilon>0$. However, we believe that our surrogate expression
+  $\bar\kappa$ for the stochastic condition number is far more
+  accurate than suggested by the current analysis.
+\end{remark}
+
+\begin{proofof}{Corollary}{\ref{c:kaczmarz2}}
+  Using Theorem \ref{t:main}, for $\bar\P_{\perp}$ as defined in \eqref{eq:surrogate}, we have
+  \begin{align*}
+    (1-\epsilon)\bar\P_{\perp}\preceq \I-\E[\P] =
+    \E[\P_{\perp}]\preceq(1+\epsilon)\bar\P_{\perp},
+    \quad\text{where}\quad\epsilon=O\big(\tfrac 1{\sqrt r}\big).
+  \end{align*}
+  In particular, this implies that
+  $\|\bar\P_{\perp}^{-\frac12}(\E[\P_{\perp}]-\bar\P_{\perp})\bar\P_{\perp}^{-\frac12}\|\leq\epsilon$.
+  Moreover, in the proof of Theorem \ref{t:main-tech} we showed that
+  $\frac{\rho-1}\rho\I\preceq\bar\P_{\perp}\preceq\I$, see
+  \eqref{eq:bound-bar-gamma-P}, so it follows that:
+  \begin{align*}
+    \bar\P_{\perp}^{-1}(\E[\P_{\perp}]-\bar\P_{\perp})^2\bar\P_{\perp}^{-1}\preceq
+    \frac{\rho}{\rho-1}\big(\bar\P_{\perp}^{-\frac12}(\E[\P_{\perp}]-\bar\P_{\perp})\bar\P_{\perp}^{-\frac12}\big)^2
+    \preceq     \frac{\rho}{\rho-1}\,\epsilon^2\cdot \I,
+  \end{align*}
+  where note that  $\frac{\rho}{\rho-1}\,\epsilon^2=O(1/r)$, since
+  $\rho$ is treated as a constant. Thus we conclude that:
+  \begin{align*}
+    \|\E[\Delta_{t+1}]-\bar\Delta_{t+1}\|^2
+     & =\E[\Delta_t]^\top(\E[\P_{\perp}]-\bar\P_{\perp})^2\E[\Delta_t]
+    \\
+     & \leq O(1/r)\cdot \E[\Delta_t]^\top\bar\P_{\perp}^2\E[\Delta_t] =
+    O(1/r)\cdot \|\bar\Delta_{t+1}\|^2,
+  \end{align*}
+  which completes the proof of \eqref{eq:trajectory}. To show
+  \eqref{eq:worst-case}, it suffices to observe that
+  \begin{align*}
+    \lambda_{\min}(\E[\P])
+    = 1- \lambda_{\max}(\E[\P_{\perp}])
+    \geq 1-(1+\epsilon)\lambda_{\max}(\bar\P_{\perp})
+    \geq \lambda_{\min}(\I-\bar\P_{\perp}) - \epsilon,
+  \end{align*}
+  which completes the proof since
+  $\I-\bar\P_{\perp}=\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1}$.
+\end{proofof}
+Corollaries \ref{c:rsn} and \ref{c:jacsketch} follow analogously from
+Theorem \ref{t:main}.
+
+\subsection{Randomized Subspace Newton}
+Randomized Subspace Newton (RSN, \citep{Gower2019})  is a randomized
+Newton-type method  for minimizing a
+smooth, convex and twice differentiable function
+$f:\R^d\times \R$. The iterative
+update for this algorithm is defined as follows:
+\begin{align*}
+  \x^{t+1} = \x^t - \frac1L\S_t^\top(\S_t\H(\x^t)\S_t^\top)^\dagger\S_t\g(\x^t),
+\end{align*}
+where $\H(\x^t)$ and $\g(\x^t)$ are the Hessian and gradient of $f$ at
+$\x^t$, respectively, whereas $\S_t$ is a $k\times d$ sketching
+matrix (with $k\ll d$) which is refreshed at every iteration. Here,
+$L$ denotes the \emph{relative smoothness} constant defined by
+\citep{Gower2019} in Assumption 1, which also defines relative strong
+convexity, denoted by $\mu$. In Theorem 2, they prove the
+following convergence guarantee for RSN:
+\begin{align*}
+  \E[f(\x^t)] - f(\x^*) \leq \Big(1-\kappa\frac{\mu}{L}\Big)^t(f(\x^0)-f(\x^*)),
+\end{align*}
+where $\kappa =\min_{\x}\kappa(\x)$ and
+$\kappa(\x)=\lambda_{\min}^+(\E[\P(\x)])$ is the smallest positive
+eigenvalue of the expectation of the projection matrix
+$\P(\x)=
+  \H^{\frac12}(\x)\S_t^\top(\S_t\H(\x)\S_t^\top)^\dagger\S_t\H^{\frac12}(\x)$. Our
+results lead to the following surrogate expression for this expected
+projection when the sketch is sub-gaussian:
+\begin{align*}
+  \E[\P(\x)] \simeq \H(\x)\big(\H(\x) +
+  \tfrac1{\gamma(\x)}\I\big)^{-1}\quad\text{for}\quad\gamma(\x)>0
+  \quad\text{s.t.}\quad\tr\,\H(\x)\big(\H(\x) +
+  \tfrac1{\gamma(\x)}\I\big)^{-1} = k.
+\end{align*}
+Thus, the condition number $\kappa$ of RSN can be estimated using the
+following surrogate expression:
+\begin{align*}
+  \kappa\simeq \bar\kappa :=
+  \min_\x\frac{\lambda_{\min}^+(\H(\x))}{\lambda_{\min}^+(\H(\x)) +
+    1/\gamma(\x)}.
+\end{align*}
+Just as in Corollary \ref{c:kaczmarz2}, an approximation
+of the form $|\bar\kappa-\kappa|\leq\epsilon$ can be shown
+from Theorem \ref{t:main}.
+\begin{corollary}\label{c:rsn}
+  Suppose that sketch $\S_t$ has size $k$ and i.i.d.~mean-zero sub-gaussian entries. Let
+  $r=\min_\x\tr\,\H(\x)/\|\H(\x)\|$ be the (minimum) stable rank of
+  the (square root) Hessian and assume that $\rho = r/k$ is a constant larger than
+  $1$. Then,
+  \begin{align*}
+    |\kappa-\bar\kappa| \leq O\big(\tfrac1{\sqrt r}\big).
+  \end{align*}
+\end{corollary}
+
+\subsection{Jacobian Sketching}
+Jacobian Sketching (JacSketch, \citep{jacsketch}) defines an $n\times n$ positive semi-definite weight matrix
+$\W$, and combines it with an $k\times n$ sketching matrix $\S$
+(which is refreshed at every iteration of the algorithm), to
+implicitly construct the following projection matrix:
+\begin{align*}
+  \Pi_\S = \S^\top(\S\W\S^\top)^\dagger\S\W,
+\end{align*}
+which is used to sketch the Jacobian at the current iterate (for the
+complete method, we refer to their Algorithm 1).
+The convergence rate guarantee given in their Theorem 3.6 for
+JacSketch is given in terms of the Lyapunov function:
+\begin{align*}
+  \Psi^t = \|\x^t-\x^*\|^2 + \frac{\alpha}{2\mathcal{L}_2}\|\mathbf{J}^t-\nabla F(\x^*)\|_{\W^{-1}}^2,
+\end{align*}
+where $\alpha$ is the step size used by the algorithm. Under
+appropriate choice of the step-size, Theorem~3.6 states that:
+\begin{align*}
+  \E[\Psi^t] \leq \bigg(1 - \mu\,\min\Big\{\frac{1}{4\mathcal{L}_1},\frac{\kappa}{4\mathcal{L}_2\rho/n^2+\mu}\Big\}\bigg)^t\cdot\Psi^0,
+\end{align*}
+where $\kappa=\lambda_{\min}(\E[\Pi_\S])$ is the
+\emph{stochastic condition number} analogous to the one defined for
+the Generalized Kaczmarz method, $n$ is the data size and
+parameters $\rho$, $\mathcal{L}_1$, $\mathcal{L}_2$ and $\mu$ are
+problem dependent constants defined in Theorem~3.6. Similarly as
+before, we can use our surrogate expressions
+for the expected residual projection to obtain a precise estimate for
+the stochastic condition number
+$\kappa$ under sub-gaussian sketching:
+\begin{align*}
+  \kappa\simeq\bar\kappa :=\frac{\lambda_{\min}(\W)}{\lambda_{\min}(\W) +
+    1/\gamma}
+  \quad\text{for}\quad\gamma>0\quad\text{s.t.}\quad\tr\,\W(\W +
+  \tfrac1\gamma\I)^{-1} = k.
+\end{align*}
+\begin{corollary}\label{c:jacsketch}
+  Suppose $\S_t$ has size $k$ and i.i.d.~mean-zero sub-gaussian entries. Let
+  $r=\tr\,\W/\|\W\|$ be the stable rank of $\W^{\frac12}$ and assume
+  that $\rho = r/k$ is a constant larger than
+  $1$. Then,
+  \begin{align*}
+    |\kappa-\bar\kappa| \leq O\big(\tfrac1{\sqrt r}\big).
+  \end{align*}
+\end{corollary}
 
@@
   \I-\E[\P_\perp]\bar\P_\perp^{-1}
-    &= \E[\P] - \gamma\E[\P_\perp]\Sigmab = \E[\P \cdot \one_E] +
+   & = \E[\P] - \gamma\E[\P_\perp]\Sigmab = \E[\P \cdot \one_E] +
   \E[\P  \cdot \one_{\neg E}]
-    -\gamma\E[\P_\perp]\Sigmab\\
+  -\gamma\E[\P_\perp]\Sigmab                                                                                   \\
   %&=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg] -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
-  &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
+   & =\gamma\,\underbrace{\E\bigg[(\bar s-\hat
       s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
   - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2}                           \\
-    &\quad+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
+   & \quad+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
   %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
@@
 bounded with a careful application of  the Hanson-Wright type
-\cite{rudelson2013hanson} sub-gaussian concentration 
+\citep{rudelson2013hanson} sub-gaussian concentration
 inequalities (Lemmas~\ref{l:concentration-sub-gaussian}~and~\ref{l:trace} in the
@@
 
+\section{Proof of Theorem~\ref{t:main-tech}}
+\label{sec:proof-of-theo-main-tech}
+
+We first introduce the following technical lemmas.
+
+\begin{lemma}\label{l:rank-one-update}
+  For $\X \in \mathbb R^{k \times n}$ with $k<n$, denote $\P = \X^\dagger \X$ and $\P_{-k} = \X_{-k}^\dagger \X_{-k}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its i-th row $\x_i \in \mathbb R^n$.
+Then, conditioned on the event $E_k: \left\{ \left| \frac{\tr \Sigmab (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} - 1 \right| \le \frac12 \right\}$:
+  \begin{align*}
+    (\X^\top\X)^\dagger\x_k = \frac{(\I - \P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}, \quad \P - \P_{-k} = \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top (\I - \P_{-k}) \x_k}.
+  \end{align*}
+\end{lemma}
+
+\begin{proof}
+  % Recall $\P = \I - \X^\dagger \X = \I - (\X^\top \X)^\dagger \X^\top \X = \I - (\X^\top \X)^\dagger (\A + \x_k \x_k^\top)$ and $\P_{-k} = \I - \X_{-k}^\dagger \X_{-k} = \I - \A^\dagger \A$ 
+  Since conditioned on $E_k$ we have $\x_k^\top (\I - \P_{-k}) \x_k \neq 0$, from \cite[Theorem 1]{10.2307/2099767} we deduce
+  \begin{align*}
+    (\X^\top \X)^\dag = \left( \A + \x_k \x_k^\top \right)^\dagger
+     & = \A^\dag
+    - \frac{\A^\dag \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top ( \I - \P_{-k}) \x_k}
+    - \frac{( \I -\P_{-k}) \x_k \x_k^\top \A^\dag}{\x_k^\top ( \I - \P_{-k}) \x_k}                                           \\
+     & + (1 + \x_k^\top \A^\dag \x_k) \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k})}{(\x_k^\top (\I - \P_{-k}) \x_k)^2}
+  \end{align*}
+  for $\A = \X_{-k}^\top \X_{-k}$ so that $\I - \P_{-k} = \I - \A^\dagger \A$, where we used the fact that $\I - \P_{-k}$ is a projection matrix so that $(\I - \P_{-k})^2 = \I - \P_{-k}$.
+As a consequence, multiplying by $\x_k$ and simplifying we get
+  \[
+    (\X^\top \X)^\dag \x_k = \frac{ ( \I -\P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}.
+  \]
+  By definition of the pseudoinverse, $\P = \X^\dagger \X = (\X^\top \X)^\dagger \X^\top \X$ so that
+  \[
+    \P - \P_{-k} = \X^\dagger \X - \X_{-k}^\dagger \X_{-k} = \frac{( \I - \P_{-k}) \x_k \x_k^\top ( \I - \P_{-k})}{\x_k^\top ( \I - \P_{-k}) \x_k}
+  \]
+  where we used $\A (\I - \P_{-k}) = \A - \A \A^\dagger \A = 0$ and thus the conclusion.
+\end{proof}
+
+% \begin{lemma}[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Corollary 2.6]{zajkowski2018bounds}}] \label{l:concentration-sub-gaussian}
+% For a $K$-sub-gaussian random vector $\x \in \mathbb R^d$ (i.e., $\| \x \|_{\psi_2} \leq K$) and $\A \in \mathbb R^{d \times d}$, we have
+% \begin{align*}
+%   \Pr\left[
+%     \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert
+%     \geq t
+%     \right] \leq 2 \exp\left(
+%     -\min\left\{
+%       \frac{t^2}{16 C^2  K^4 (\tr \A)^2},
+%       \frac{t}{4 C K^2 (\tr \A)},
+%     \right\}
+%   \right)
+% \end{align*}
+% and 
+% \begin{align*}
+%   \E\left[ \left( \x^\top \A \x - \E[\x^\top \A \x] \right)^2 \right]
+%   &\leq C~K^4~\tr(\A)^2
+% \end{align*}
+% for some universal constant $C$.
+% \end{lemma}
+
+% \begin{lemma}
+% %[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}]
+% \label{l:concentration-sub-gaussian}
+% For a $K$-sub-gaussian random vector $\x \in \mathbb R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and positive semi-definite matrix $\A \in \mathbb R^{n \times n}$ such thats $\| \A \| \le 1 $, we have
+% \[
+%   \Pr\left[
+%     \lvert \x^\top \A \x - \tr \A \rvert
+%     \geq \frac12 \tr \A
+%     \right] \leq 2 \exp\left( - \min \left\{ \frac{r_\A}{4 C^2 K^4}, \frac{\sqrt r_\A}{2 C K^2}  \right\} \right)
+% \]
+% with $r_\A = \frac{\tr \A}{\| \A \|}$ the nuclear rank of $\A$, and 
+% \begin{align*}
+%   \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
+%   &\leq cK^4\tr\A^2
+% \end{align*}
+% for some $C,c > 0$ independent of $k$.
+% \end{lemma}
+
+% \begin{proof}
+% % From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \Sigmab$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
+% % \[
+% %   \Pr \left\{ \left| \x^\top \A \x - \tr \A \Sigmab \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
+% % \]
+% % for some universal constant $C > 0$. Taking $t = \frac12 \tr \A \Sigmab$ we obtain
+% From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
+% \[
+%   \Pr \left\{ \left| \x^\top \A \x - \tr \A  \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
+% \]
+% for some universal constant $C > 0$. Taking $t = \frac12 \tr \A$ we have
+% \[
+%   \frac{t^2}{C^2 K^4 \tr\A^2 } = \frac{ (\tr \A)^2 }{4 C^2 K^4 \tr\A^2 } \ge \frac{ \tr \A }{4 C^2 K^4 \| \A \| } = \frac{r_\A}{4 C^2 K^4}, \quad \frac{t}{C K^2 \sqrt{\tr\A^2}  } \ge \frac{\sqrt r_\A}{2 C K^2}
+% \]
+% where we use the fact that $\tr \A^2 \le \| \A \| \tr \A$ and denote $r_\A = \frac{\A}{\| \A \|}$ the nuclear rank of $\A$.
+
+% Integrating this bound yields, we get
+% \[
+%   \E\left[ (\x^\top \A \x - \tr \A )^2 \right] \le c~K^4~\tr \A^2
+% \]
+% and thus the conclusion.
+% \end{proof}
+%
+%
+%
+\begin{lemma}
+  %[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}]
+  \label{l:concentration-sub-gaussian}
+  For a $K$-sub-gaussian random vector $\x \in \mathbb R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and positive semi-definite matrix $\A \in \mathbb R^{n \times n}$, we have
+  \[
+    \Pr\left[
+      \lvert \x^\top \A \x - \tr \A \rvert
+      \geq \frac13 \tr \A
+      \right] \leq 2 \exp\left( - \min \left\{ \frac{r_{\!\A}}{9 C^2 K^4}, \frac{\sqrt{r_{\!\A}}}{3 C K^2}  \right\} \right)
+  \]
+  with $r_{\!\A} = \tr \A/\| \A \|$ the stable rank of $\A$, and
+  \begin{align*}
+    \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
+     & \leq c~K^4~\tr \A^2
+  \end{align*}
+  for some $C,c > 0$ independent of $K$.
+\end{lemma}
+
+\begin{proof}
+  % From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \Sigmab$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
+  % \[
+  %   \Pr \left\{ \left| \x^\top \A \x - \tr \A \Sigmab \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
+  % \]
+  % for some universal constant $C > 0$. Taking $t = \frac12 \tr \A \Sigmab$ we obtain
+  This follows from a Hanson-Wright type \citep{rudelson2013hanson} sub-gaussian concentration inequality. More precisely,
+  from \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
+  \[
+    \Pr \left\{ \left| \x^\top \A \x - \tr \A  \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
+  \]
+  for some universal constant $C > 0$. Taking $t = \frac13 \tr \A$ we have
+  \[
+    \frac{t^2}{C^2 K^4 \tr\A^2 } = \frac{ (\tr \A)^2 }{9 C^2 K^4 \tr\A^2 } \ge \frac{ \tr \A }{9 C^2 K^4 \| \A \| } = \frac{r_{\!\A}}{9 C^2 K^4}, \quad \frac{t}{C K^2 \sqrt{\tr\A^2}  } \ge \frac{\sqrt{r_{\!\A}}}{3 C K^2}
+  \]
+  where we use the fact that $\tr \A^2 \le \| \A \| \tr \A$.
+
+  Integrating this bound yields:
+  \[
+    \E\left[ (\x^\top \A \x - \tr \A )^2 \right] \le c~K^4~\tr \A^2
+  \]
+  and thus the conclusion.
+\end{proof}
+
+% \begin{lemma}\label{l:rank-one-projection}
+%   For any $d\times d$ p.s.d.~matrix $\A$ and
+%   any random vector $\z$ with jointly
+%   $K$-sub-gaussian (sub-gaussian Orlicz norm
+%   $\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$),
+%   \begin{align*}
+%     \Bigg\|\E\bigg[\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\bigg]\Bigg\|
+% \leq
+%     \sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\v\Big)^2\bigg]}
+%     \leq C\cdot K^2 \cdot \frac{\|\A\|}{\tr\,\A}.
+%   \end{align*}
+%   for some universal constant $C$.
+% \end{lemma}
+%\textbf{** Lemma 3 seems no longer necessary?**}
+
+\begin{lemma}\label{l:trace}
+  With the notations of Lemma~\ref{l:rank-one-update}, for $X=\tr\,\Sigmab(\P_{-k}-\E[\P_{-k}])$
+  and $\|\Sigmab\| = 1$, we have
+  \begin{align*}
+    \E[X^2]\leq
+    Ck\quad\text{and}\quad\Pr\{|X|\geq t\}\leq 2 \ee^{-\frac{t^2}{ck}}.
+  \end{align*}
+  for some universal constant $C,c > 0$.
+\end{lemma}
+\begin{proof}
+  To simplify notations, we work on $\P$ instead of $\P_{-k}$, the same line of argument applies to $\P_{-k}$ by changing the sample size $k$ to $k-1$.
+
+  First note that
+  \begin{align*}
+    X
+     & = \tr \Sigmab (\P - \E \P)
+    = \E_k [\tr \Sigmab \P]  - \E_0[ \tr \Sigmab \P ]                               \\
+    %%%
+     & = \sum_{i=1}^k \left( \E_i[\tr \Sigmab \P] - \E_{i-1}[\tr \Sigmab \P]\right)
+    = \sum_{i=1}^k (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})
+  \end{align*}
+  where we used the fact that $\E_i [\tr \Sigmab \P_{-i}] = \E_{i-1} [\tr \Sigmab \P_{-i}]$, for $\E_i[\cdot]$ the conditional expectation with respect to $\mathcal F_i$ the $\sigma$-field generating the rows $\x_1 \ldots, \x_i$ of $\X$.
+This forms a martingale difference sequence (it is a difference sequence of
+  the Doob martingale for $\tr \Sigmab (\P - \P_{-i})$ with respect to filtration $\mathcal{F}_i$)
+  hence it falls within the scope of the Burkholder inequality \citep{burkholder1973distribution}, recalled as follows.
+
+  \begin{lemma}\label{l:burkholder}
+    For $\{ x_i \}_{i=1}^k$ a real martingale difference sequence with respect to the increasing $\sigma$ field $\mathcal F_i$, we have, for $L > 1$, there exists $C_L > 0$ such that
+    \[
+      \E \bigg[
+        \Big| \sum_{i=1}^k x_i \Big|^L
+        \bigg]
+      \le C_L \E \bigg[
+        \Big( \sum_{i=1}^k |x_i|^2 \Big)^{L/2}
+        \bigg].
+    \]
+  \end{lemma}
+
+  From Lemma~\ref{l:rank-one-update}, $\P - \P_{-i} = \frac{(\I - \P_{-i}) \x_i \x_i^\top (\I - \P_{-i}) }{ \x_i^\top (\I - \P_{-i}) \x_i}$ is positive semi-definite, we have $\tr \Sigmab (\P - \P_{-i}) \le \| \Sigmab \| = 1$ so that with Lemma~\ref{l:burkholder} we obtain with $x_i = (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})$ that, for $L > 1$
+  \[
+    \E |X|^L \le C_L k^{L/2}.
+  \]
+  In particular, for $L=2$, we obtain $\E |X|^2 \le C k$.
+  %for $c = (36 \sqrt{2})^2$ (details in proof of Theorem~3.2 in \citep{burkholder1973distribution}).
+
+  For the second result, since we have almost surely bounded martingale differences
+  ($\lvert x_i \rvert \leq 2$), by the Azuma-Hoeffding inequality
+  \begin{align*}
+    \Pr\{
+    \lvert X \rvert
+    \geq t
+    \}
+    \leq 2 \ee^{\frac{-t^2}{8 k}}
+  \end{align*}
+  as desired.
+\end{proof}
+
+% \subsection{Complete proof of Theorem~\ref{t:main-tech}}
+
+% Equipped with the lemmas above, we are ready to prove Theorem~\ref{t:main-tech}. First note that:
+% \begin{enumerate}[leftmargin=*]
+%   \item  Since $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, we can assume without loss of generality (after rescaling $\bar\P_\perp$ correspondingly) that $\|\Sigmab\| = 1$.
+%   \item According to the definition of $\bar \P_\perp$ and $\gamma$, the following bounds hold
+%   \begin{equation}\label{eq:bound-bar-gamma-P}
+%     \frac1{\gamma + 1} \I \preceq \bar \P_\perp \preceq \I, \quad \gamma \le \frac{k}{r-k} = \frac1{\rho - 1}
+%   \end{equation}
+%   for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}k > 1$, where we used the fact that
+%   \[
+%     k = n - \tr\,\bar\P_\perp = \tr\,\bar\P_\perp (\gamma\Sigmab + \I) - \tr\,\bar\P_\perp
+%   = \gamma \tr\,\bar\P_\perp \Sigmab\geq \frac{\gamma}{\gamma+1}\tr\,\Sigmab,
+%   \]
+%   so that $r = \tr \Sigmab \le k \cdot \frac{\gamma + 1}{\gamma}$.%= k + \tr \bar \P_\perp \Sigmab
+%   \item As already discussed in Section~\ref{subsec:proof-sketch}, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the following spectral norm
+%   \begin{equation}\label{eq:proof-1-SM}
+%     \| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}
+%   \end{equation}
+%   so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ from \eqref{eq:bound-bar-gamma-P}, we have
+%   \begin{align*}
+%     \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
+%     \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp^{\frac12}\|\leq
+%   \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
+%   \end{align*}
+%   Defining $\epsilon =\frac{C_\rho}{\sqrt r}
+%   \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
+%   of the p.s.d.~matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
+%   in the interval $[1-\epsilon,1+\epsilon]$, and
+%   \begin{align*}
+%     (1-\epsilon)\I\preceq
+%     \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.
+%   \end{align*}
+%   so that by multiplying $\bar \P_{\perp}^{\frac12}$ on both sides, we obtain the
+%   desired bound.
+% \end{enumerate}
+
+% \bigskip
+
+% As a consequence of the above observations, we only need to prove \eqref{eq:proof-1-SM} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:
+
+% \begin{enumerate}[leftmargin=*]
+%   \item For $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its $i$-th row, we define, for $i \in \{ 1, \ldots, k \}$, the following events
+%   \begin{equation}\label{eq:def-events-E-F}
+%     E_i: \left\{ \left| \frac{\tr (\I - \P_{-i}) \Sigmab}{\x_i^\top ( \I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\}, \quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\}.
+%F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
+%   \end{equation}
+%   where we recall $\x_i \in \R^n$ is the $i$-th row of $\X$ so that $\E[\x_i] = 0$ and $\E[\x_i \x_i^\top] = \Sigmab$.
+With Lemma~\ref{l:concentration-sub-gaussian}, we can bound the probability of $\neg E_i$ and $\neg F_i$, and consequently that of $\neg E$ for $E = \bigwedge_{i=1}^k(E_i \wedge F_i)$;
+%   \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \|$. More precisely, since
+% %   \begin{align*}
+% %   \I&-\E[\P]\bar\P_\perp^{-1}
+% %   = \E[\X^\dagger\X] - \gamma\E[\P]\Sigmab\\
+% %   &= \E[\X^\dagger\X\one_E] +
+% %     \E[\X^\dagger\X \one_{\neg E}]
+% %     -\gamma\E[\P]\Sigmab\\
+% %   &=n\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top}{\x_k^\top\P_{-k}\x_k}\one_E\bigg]
+% %     -\gamma\E[\P]\Sigmab \ +\     \E[\X^\dagger\X\one_{\neg E}]\\
+% %   &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
+% %     s)\cdot\frac{\P_{-k}\x_k\x_k^\top}{\x_k^\top\P_{-k}\x_k}\one_E\bigg]}_{\T_1}
+% %     + \gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2} + \gamma \underbrace{\E[\P_{-k}-\P]\Sigmab}_{\T_3} + \underbrace{\E[\X^\dagger\X\one_{\neg E}]}_{\T_4}
+% % \end{align*}
+% \begin{align*}
+%   \I-\E[\P_\perp]\bar\P_\perp^{-1}
+%     &= \E[\P] - \gamma\E[\P_\perp]\Sigmab
+%     \\
+%     &= \E[\P \cdot \one_E] +
+%     \E[\P  \cdot \one_{\neg E}]
+%     -\gamma\E[\P_\perp]\Sigmab\\
+%   &=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg]
+%     -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
+%   &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
+%     s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
+%     - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2} \\ 
+%     &+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
+%     %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
+% \end{align*}
+% where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_k^\top ( \I - \P_{-k}) \x_k$ as well as $\bar s = \tr \bar \P_\perp \Sigmab =k/\gamma$.
+It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ to reach the conclusion.
+% \end{enumerate}
+
+% Another important relation that will be constantly used throughout the proof is
+% \begin{equation}\label{eq:bound-trace-P-Sigma}
+%     \tr (\I - \P_{-k}) \Sigmab = \tr \Sigmab^{\frac12} (\I - \P_{-k})^2 \Sigmab^{\frac12} =
+%     \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-k}^\dagger\X_{-k}\|_F^2\geq
+% \sum_{i\geq k}\lambda_i (\Sigmab)\geq r-k
+% \end{equation}
+% where we used the fact that $\rank(\X_{-k}^\dagger \X_{-k}) \le \rank(\X_{-k}) \le k-1$ and arranged the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_n(\Sigmab)$ in a non-increasing order.
+As a consequence, we also have
+% \begin{equation}\label{eq:bound-norm-P-Sigma}
+%   \frac{\tr (\I - \P_{-k}) \Sigmab}{ \|  (\I - \P_{-k}) \Sigmab \| } \ge \tr (\I - \P_{-k}) \Sigmab \geq r-k.
+% \end{equation}
+
+% \bigskip
+
+% For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian}~and~\eqref{eq:bound-norm-P-Sigma} that
+% \begin{align*}
+%   \Pr (\neg E_i) &\le \Pr \left\{ \lvert \x_i^\top (\I - \P_{-i}) \x_i - \tr \Sigmab (\I - \P_{-i})  \rvert
+%     \geq \frac13 \tr \Sigmab (\I - \P_{-i}) \right\} \\ 
+%     &\le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }. % \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
+% \end{align*}
+% %for some constant $c_\rho$ depending on fixed $\rho \equiv \frac{r}n > 1$. 
+% Similarly we have
+% \[
+%   \Pr (\neg F_i) \le 2 \ee^{ - \min \left\{ \frac{ r }{9 C^2 K^4}, \frac{\sqrt{r} }{3 C K^2} \right\} } \le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }
+% \]
+% and with the union bound we obtain
+% \begin{equation}
+%   \Pr (\neg E) \le 4 k \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{k}{(r-k)^2} \cdot 4 (r-k)^2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{C_\rho}{r - k}
+% \end{equation}
+% where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ and  $x^4 \ee^{-\alpha x} \le \frac{256 \ee^{-4}}{\alpha^4}$ on $x > 0$.
+Also, denote $c_\rho = \frac{r - k}r = \frac{\rho - 1}{\rho} > 0$, we have
+% \begin{equation}\label{eq:proba-E}
+%   \Pr(\neg E) \le \frac{C_\rho}{r - k} = \frac{C_\rho}{c_\rho r} = \frac{C'_\rho}{r} 
+% \end{equation}
+% for some $C'_\rho > 0$ that depends on $\rho = r/k > 1$ and the sub-gaussian norm $K$.
+
+% \medskip
+
+% At this point, note that, conditioned on the event $E$, we have for $i \in \{1, \ldots, k \}$
+% \begin{equation}
+%     \frac12 \frac1{\tr (\I - \P_{-i}) \Sigmab} \le \frac1{\x_i^\top (\I - \P_{-i}) \x_i} \le \frac32 \frac1{\tr (\I - \P_{-i}) \Sigmab}, \quad  \frac1{2r} \le \frac1{\x_i^\top \x_i} \le \frac3{2r}.
+%\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
+% \end{equation}
+% Also, with \eqref{eq:proba-E} and the fact that $\| \I - \P_{-k} \|
+% \le 1$ and $\| \P \| \le 1$, we have $\| \T_2 \| + \| \T_4 \| \le
+% \frac{C_\rho}{r}$ for some $\C_\rho > 0$ that depends on $\rho$ and $K$. And it thus remains to handle the terms $\T_1$ and $\T_3$ to obtain a bound on $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| $.
+
+% To bound $\T_3$, with $\P - \P_{-k} = \frac{(\I - \P_{-k})\x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}$ in Lemma~\ref{l:rank-one-update}, we have
+% \begin{align*}
+% \|\T_3\| &\leq  \bigg\|
+%   \E\bigg[\frac{ (\I - \P_{-k}) \x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}
+%   \cdot \one_E \bigg]\bigg\|  + \| \E[ (\P - \P_{-k}) \cdot \one_{\neg E} ] \| \\ 
+%   %%%
+%   &\le \frac32 \E \left[ \frac1{ \tr (\I - \P_{-k} ) \Sigmab } \right] + \frac{c_\rho}{r - k} \leq \frac{C_\rho}{r-k} = \frac{C'_\rho}{r}
+% \end{align*}
+% where we used the fact that $\tr\,(\I - \P_{-k}) \Sigmab \ge r -k $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv r/k > 1$.
+
+% For $\T_1$ we write
+% \begin{align*}
+%   \|\T_1\|    &\leq\E\bigg[\|\I - \P_{-k}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
+%     \frac{\x_k\x_k^\top}{\x_k^\top( \I - \P_{-k}) \x_k} \cdot \one_E\mid\P_{-k}\Big]
+%     \Big\|\bigg]
+%   \\
+%           &\leq \frac32 \frac1{r-k} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
+%             |\bar s-\hat s|\cdot \v^\top\x_k\x_k^\top\v\cdot\one_E\mid
+%             \P_{-k}\Big]\bigg]
+%   \\
+%   &\leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
+%     s)^2\cdot \one_E \mid\P_{-k}\big]}}_{ T_{1,1} }\cdot
+%     \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_k)^4\big]}}_{ T_{1,2} }\bigg]
+% \end{align*}
+% where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} for the second inequality, and Cauchy–Schwarz for the third inequality.
+
+% We first bound $T_{1,2}$ by definition of sub-gaussian random
+% vectors. We have for $\x_k$ a $K$-sub-gaussian and $\| \v \| = 1$
+% that, $\v^\top \x_k$ is a sub-gaussian random variable with $\|\v^\top
+% \a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute
+% constant $C > 0$, see for example
+% \cite[Section~2.5.2]{vershynin2018high}.
+
+% For $T_{1,1}$ we have
+% \[
+%   \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-k}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
+% \]
+% where we denote $s=\E[\hat s]=\tr\,\E[\I - \P_{-k}]\Sigmab$. Note that
+% % \begin{align*}
+% %   \E\big[(s - \hat s)^2 \cdot\one_E \big] &=
+% % \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2 \cdot\one_E \big] +
+% %   \E\big[(\tr\,\P_{-k}\Sigmab-\x_k^\top\P_{-k}\x_k)^2 \cdot\one_E \big]\\
+% %   &\leq C\Big(n + \E\big[\tr\,(\P_{-k}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
+% %     C\big(n + \tr\,\E[\P_{-k}]\Sigmab \cdot\one_E\big)\\
+% %   &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
+% % \end{align*}
+% \begin{align*}
+%   \E\big[(s - \hat s)^2 \big] &=
+% \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2  \big] +
+%   \E\big[(\tr\,(\I - \P_{-k})\Sigmab-\x_k^\top(\I - \P_{-k})\x_k)^2  \big]\\
+%   &\leq C_1 k +  C_2 \E\big[\tr\,(\Sigmab - \P_{-k}\Sigmab)^2 \big] \le C(k+s)\leq C\big(k+\bar s + |s- \bar s| \big)
+% \end{align*}
+% where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}. Recall that $\bar s = \tr \bar \P_\perp \Sigmab \le \tr \Sigmab = r$ and $k < r$, we have
+% \begin{equation}\label{eq:bound-T11}
+%   T_{1,1} \le \sqrt{ (\bar s - s)^2 + C (|\bar s - s| + 2 r) }
+% \end{equation}
+
+% % Moreover, using the fact that
+% % $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$, we obtain that:
+% % \begin{align*}
+% %   |\bar s - s| \cdot\one_E &= |\tr (\bar \P_\perp - \E[\P_{-k}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P_\perp - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-k}] \Sigmab| \\
+% %   &= \big|\tr\,(\I-\E[\P]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-k}\x_k\x_k^\top\P_{-k}}{\x_k^\top \P_{-k} \x_k} \right] \Sigmab \\
+% %   %&= \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P_\perp\Sigmab\big| \cdot\one_E + 1 \\
+% %   &\leq \gamma\,\E\bigg[|\bar s-\hat s|\cdot
+% %     \frac{\tr\,\P_{-k}\x_k\x_k^\top\bar\P_\perp\Sigmab}{\tr\,\P_{-k}\x_k\x_k^\top} \cdot\one_E\bigg]
+% %     + \gamma\,
+% %     \E\bigg[\frac{\tr\,\P_{-k}\x_k\x_k^\top\P_{-k}\Sigmab\bar\P_\perp\Sigmab}
+% %     {\tr\,\P_{-k}\x_k\x_k^\top}\bigg] + 1\\
+% %   &\leq \frac{\gamma}{\gamma+1}\Big(\E\left[|\bar s-\hat
+% %     s| \cdot \frac{\x_k^\top \x_k}{\x_k^\top \P_{-k} \x_k} \cdot\one_E \right] + 1\Big)\leq \frac{\gamma}{\gamma+1}\Big(|\bar
+% %     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
+% %     s|\big] + 1\Big)\\
+% %   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
+% %      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
+% % \end{align*}
+
+% It remains to bound $|\bar s - s|$. Note that $\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X
+% (\X^\top\X)^\dagger$ and is symmetric, so
+% \begin{align*}
+%   &\I - \E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar\P_\perp^{-1} \E[\P_\perp]
+%   = 2\E[\P] - \E[\gamma\P_\perp\Sigmab] - \E[\gamma\Sigmab\P_\perp]\\
+%   &= \sum_{i=1}^k\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp])\\
+%   %&=k\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top + \x_k\x_k^\top \P_{-k}}{\x_k^\top\P_{-k}\x_k}\bigg] - \gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
+%   &=\gamma\,
+%     \E\bigg[\bar s\cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})} {\x_k^\top(\I - \P_{-k}) \x_k}\bigg]
+%     -\gamma\,\E\bigg[\hat s
+%     \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg] \\ 
+%   & +\gamma\,(\E[(\I - \P_{-k})\Sigmab] + \E[\Sigmab ( \I -\P_{-k}) ])- \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp]) \\
+%   &=\gamma\,\E\bigg[(\bar s - \hat s)
+%     \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]
+% +    \gamma (\E[\P-\P_{-k}]\Sigmab + \Sigmab \E[\P-\P_{-k}]). 
+% \end{align*}
+% %where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
+% Moreover, using the fact that $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$ and $\bar \P_\perp \Sigmab = \Sigmab \bar \P_\perp$, we obtain that
+% \begin{align*}
+%   &|\bar s - s|  = |\tr (\bar \P_\perp - \E[\I - \P_{-k}] )\Sigmab|   \le |\tr (\bar \P_\perp - \E[\P_\perp] )\Sigmab|   + |\tr \E[\P - \P_{-k}] \Sigmab| \\
+%   &= \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab + \tr\,\bar\P_\perp (\I - \bar \P_\perp^{-1} \E[\P_\perp]) \Sigmab \big|   + \tr\,\E \left[ \frac{(\I - \P_{-k})\x_k\x_k^\top(\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} \right] \Sigmab \\
+%   &\leq \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar \P_\perp^{-1} \E[\P_\perp])\bar\P_\perp\Sigmab \big|   + 1 \\
+%   &\leq \frac{\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
+%     \frac{\tr\,( (\I -\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I -\P_{-k})) \bar\P_\perp\Sigmab}{\tr\,(\I -\P_{-k})\x_k\x_k^\top}  \bigg] \\
+%   &+ \gamma\, \E\bigg[\frac{\tr\,(\I-\P_{-k})\x_k\x_k^\top(\I-\P_{-k})\bar\P_\perp\Sigmab}
+%     {\tr\,(\I-\P_{-k})\x_k\x_k^\top}\bigg] + 1\\
+%   &\leq \frac{\gamma}{\gamma+1}\Bigg(\E\bigg[|\bar s-\hat
+%     s| \cdot \frac{\x_k^\top (\I -\P_{-k}) \x_k}{\x_k^\top (\I -\P_{-k}) \x_k}  \bigg] + 1\Bigg) + 1\leq \frac{\gamma}{\gamma+1}\Big(|\bar
+%     s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1 \\
+%   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
+%      C\sqrt{|\bar s- s|} + C\sqrt{2 r} + 1\Big) + 1.
+% \end{align*}
+
+% Solving for $|\bar s-s|$, we deduce that
+% \begin{align*}
+%   |\bar s-s| \leq C_1 \sqrt{r} + C_2,
+%   %\gamma(c\sqrt{k+\bar s}+2) + c^2\gamma^2\leq \gamma (c+1)\sqrt{k+\bar s} + c^2\gamma^2.
+% \end{align*}
+% so plugging back to \eqref{eq:bound-T11} we get $T_{1,1} \le C
+% \sqrt r$ and $\| \T_1\| \le \frac{C_\rho}{\sqrt r} $, thus completing
+% the proof.
+
+\subsection{Complete proof of Theorem~\ref{t:main-tech}}
+
+Equipped with the lemmas above, we are ready to prove Theorem~\ref{t:main-tech}. First note that:
+\begin{enumerate}[leftmargin=*]
+  \item  Since $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, we can assume without loss of generality (after rescaling $\bar\P_\perp$ correspondingly) that $\|\Sigmab\| = 1$.
+  \item According to the definition of $\bar \P_\perp$ and $\gamma$, the following bounds hold
+        \begin{equation}\label{eq:bound-bar-gamma-P}
+          \frac1{\gamma + 1} \I \preceq \bar \P_\perp \preceq \I, \quad \gamma \le \frac{k}{r-k} = \frac1{\rho - 1}
+        \end{equation}
+        for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}k > 1$, where we used the fact that
+        \[
+          k = n - \tr\,\bar\P_\perp = \tr\,\bar\P_\perp (\gamma\Sigmab + \I) - \tr\,\bar\P_\perp
+          = \gamma \tr\,\bar\P_\perp \Sigmab\geq \frac{\gamma}{\gamma+1}\tr\,\Sigmab,
+        \]
+        so that $r = \tr \Sigmab \le k \cdot \frac{\gamma + 1}{\gamma}$.%= k + \tr \bar \P_\perp \Sigmab
+  \item As already discussed in Section~\ref{subsec:proof-sketch}, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the following spectral norm
+        \begin{equation}\label{eq:proof-1-SM}
+          \| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}, %\quad~\text{or}~\| \I - \bar \P_\perp^{-1} \E[\P_\perp] \| \le \frac{C_\rho}{\sqrt r}
+        \end{equation}
+        so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ from \eqref{eq:bound-bar-gamma-P}, we have
+        \begin{align*}
+          \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
+          %\|\bar\P_\perp^{\frac12}(\I-\bar\P_\perp^{-1}\E[\P_\perp])\bar\P_\perp^{-\frac12}\| =
+          \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp^{\frac12}\|\leq
+          \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
+        \end{align*}
+        Defining $\epsilon =\frac{C_\rho}{\sqrt r}
+          \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues
+        of the p.s.d.~matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
+        in the interval $[1-\epsilon,1+\epsilon]$, and
+        \begin{align*}
+          (1-\epsilon)\I\preceq
+          \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.
+        \end{align*}
+        so that by multiplying $\bar \P_{\perp}^{\frac12}$ on both sides, we obtain the
+        desired bound.
+\end{enumerate}
+
+\bigskip
+
+As a consequence of the above observations, we only need to prove \eqref{eq:proof-1-SM} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:
+
+\begin{enumerate}[leftmargin=*]
+  \item For $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its $i$-th row, we define, for $i \in \{ 1, \ldots, k \}$, the following events
+        \begin{equation}\label{eq:def-events-E-F}
+          E_i: \left\{ \left| \frac{\tr (\I - \P_{-i}) \Sigmab}{\x_i^\top ( \I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\},
+          %\quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\}. %F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
+        \end{equation}
+        where we recall $\x_i \in \R^n$ is the $i$-th row of $\X$ so that $\E[\x_i] = 0$ and $\E[\x_i \x_i^\top] = \Sigmab$.
+With Lemma~\ref{l:concentration-sub-gaussian}, we can bound the probability of $\neg E_i$, and consequently that of $\neg E$ for $E = \bigwedge_{i=1}^k E_i$;
+  \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \|$. More precisely, since
+        \begin{align*}
+          \I-\E[\P_\perp]\bar\P_\perp^{-1}
+           & = \E[\P] - \gamma\E[\P_\perp]\Sigmab
+          \\
+           & = \E[\P \cdot \one_E] +
+          \E[\P  \cdot \one_{\neg E}]
+          -\gamma\E[\P_\perp]\Sigmab                                                                              \\
+           & =k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg]
+          -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]                                          \\
+           & =\gamma\,\underbrace{\E\bigg[(\bar s-\hat
+              s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
+          - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2}                      \\
+           & + \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
+          %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
+        \end{align*}
+        where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_k^\top ( \I - \P_{-k}) \x_k$ as well as $\bar s = \tr \bar \P_\perp \Sigmab =k/\gamma$.
+It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ to reach the conclusion.
+\end{enumerate}
+
+% Note that, $\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I- \bar\P_\perp^{-1}\E[\P_\perp]$ is symmetric and 
+% \begin{equation}
+%   \I-\E[\P_\perp]\bar\P_\perp^{-1} + \I- \bar\P_\perp^{-1}\E[\P_\perp] = \gamma (\T_1 + \T_1^\top) - \gamma (\T_2 + \T_2^\top) + 2 \gamma \T_3 + 2 \T_4
+% \end{equation}
+
+Another important relation that will be constantly used throughout the proof is
+\begin{equation}\label{eq:bound-trace-P-Sigma}
+  \tr (\I - \P_{-k}) \Sigmab = \tr \Sigmab^{\frac12} (\I - \P_{-k})^2 \Sigmab^{\frac12} =
+  \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-k}^\dagger\X_{-k}\|_F^2\geq
+  \sum_{i\geq k}\lambda_i (\Sigmab)\geq r-k
+\end{equation}
+where we used the fact that $\rank(\X_{-k}^\dagger \X_{-k}) \le \rank(\X_{-k}) \le k-1$ and arranged the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_n(\Sigmab)$ in a non-increasing order.
+As a consequence, we also have
+\begin{equation}\label{eq:bound-norm-P-Sigma}
+  \frac{\tr (\I - \P_{-k}) \Sigmab}{ \|  (\I - \P_{-k}) \Sigmab \| } \ge \tr (\I - \P_{-k}) \Sigmab \geq r-k.
+\end{equation}
+
+\bigskip
+
+For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian}~and~\eqref{eq:bound-norm-P-Sigma} that
+\begin{align*}
+  \Pr (\neg E_i) & \le \Pr \left\{ \lvert \x_i^\top (\I - \P_{-i}) \x_i - \tr \Sigmab (\I - \P_{-i})  \rvert
+  \geq \frac13 \tr \Sigmab (\I - \P_{-i}) \right\}                                                               \\
+                 & \le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }. % \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
+\end{align*}
+so that with the union bound we obtain
+\begin{equation}\label{eq:union-bound-neg-E}
+  \Pr (\neg E) \le 2k \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{k}{(r-k)^2} \cdot 2 (r-k)^2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{C_\rho}{r - k}
+\end{equation}
+where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ and  $x^4 \ee^{-\alpha x} \le \frac{256 \ee^{-4}}{\alpha^4}$ on $x > 0$.
+Also, denote $c_\rho = \frac{r - k}r = \frac{\rho - 1}{\rho} > 0$, we have
+\begin{equation}\label{eq:proba-E}
+  \Pr(\neg E) \le \frac{C_\rho}{r - k} = \frac{C_\rho}{c_\rho r} = \frac{C'_\rho}{r}
+\end{equation}
+for some $C'_\rho > 0$ that depends on $\rho = r/k > 1$ and the sub-gaussian norm $K$.
+
+\medskip
+
+At this point, note that, conditioned on the event $E$, we have for $i \in \{1, \ldots, k \}$
+\begin{equation}
+  \frac12 \frac1{\tr (\I - \P_{-i}) \Sigmab} \le \frac1{\x_i^\top (\I - \P_{-i}) \x_i} \le \frac32 \frac1{\tr (\I - \P_{-i}) \Sigmab}, %\quad  \frac1{2r} \le \frac1{\x_i^\top \x_i} \le \frac3{2r}.
+%\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
+\end{equation}
+Also, with \eqref{eq:proba-E} and the fact that $\| \P \| \le 1$, we have $ \| \T_4 \| \le
+  \frac{C_\rho}{r}$ for some $C_\rho > 0$ that depends on $\rho$ and $K$. To handle non-symmetric matrix $\T_2$, note that $\T_2 + \T_2^\top$ is symmetric and
+\begin{equation}
+  - \E[ (\I - \P_{-k}) \cdot \one_{\neg E}] - \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}] \preceq \T_2 + \T_2^\top \preceq \E[ (\I - \P_{-k}) \cdot \one_{\neg E}] + \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}]
+\end{equation}
+with $-(\A \A^\top + \B \B^\top) \preceq \A \B^\top + \B \A^\top \preceq \A \A^\top + \B \B^\top$.
+To obtain an upper bound for operator norm of $\E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}]$, note that
+\begin{align*}
+  \| \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}] \| & \le \E[ (\x_k^\top \x_k)^2 \cdot \one_{\neg E}] = \int_0^\infty \Pr ( \x^\top \x \cdot \one_{\neg E} \ge \sqrt t ) dt                          \\
+                                                                & \le \int_0^X \Pr ( \x^\top \x \cdot \one_{\neg E} \ge \sqrt t ) dt + \int_X^\infty \Pr ( \x^\top \x \ge \sqrt t ) dt                           \\
+                                                                & \le X \cdot \Pr(\neg E) + \int_X^\infty \ee^{- \min \left\{ \frac{t}{C^2 K^4 r}, \frac{\sqrt t}{CK^2 \sqrt r} \right\} } dt \le \frac{c_\rho}r
+\end{align*}
+where we recall $\E[\x^\top \x] = \tr \Sigmab = r$ and take $X \ge C^2 K^4 r$, the third line follows from the proof of Lemma~\ref{l:concentration-sub-gaussian} and the forth line from the same argument as in \eqref{eq:union-bound-neg-E}.
+Moreover, since $\| \T_2 \| \le \| \T_2 + \T_2^\top \|$ (see for example \cite[Proposition~5.11]{serre2010matrices}), we conclude that $\| \T_2 \| \le \frac{C_\rho}r$.
+
+And it thus remains to handle the terms $\T_1$ and $\T_3$ to obtain a bound on $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| $.
+
+To bound $\T_3$, with $\P - \P_{-k} = \frac{(\I - \P_{-k})\x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}$ in Lemma~\ref{l:rank-one-update}, we have
+\begin{align*}
+  \|\T_3\| & \leq  \bigg\|
+  \E\bigg[\frac{ (\I - \P_{-k}) \x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}
+  \cdot \one_E \bigg]\bigg\|  + \| \E[ (\P - \P_{-k}) \cdot \one_{\neg E} ] \|                                                                      \\
+  %%%
+           & \le \frac32 \E \left[ \frac1{ \tr (\I - \P_{-k} ) \Sigmab } \right] + \frac{c_\rho}{r - k} \leq \frac{C_\rho}{r-k} = \frac{C'_\rho}{r}
+\end{align*}
+where we used the fact that $\tr\,(\I - \P_{-k}) \Sigmab \ge r -k $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv r/k > 1$.
+
+For $\T_1$ we write
+\begin{align*}
+  \|\T_1\| & \leq\E\bigg[\|\I - \P_{-k}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
+      \frac{\x_k\x_k^\top}{\x_k^\top( \I - \P_{-k}) \x_k} \cdot \one_E\mid\P_{-k}\Big]
+    \Big\|\bigg]
+  \\
+           & \leq \frac32 \frac1{r-k} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
+      |\bar s-\hat s|\cdot \v^\top\x_k\x_k^\top\v\cdot\one_E\mid
+      \P_{-k}\Big]\bigg]
+  \\
+           & \leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
+        s)^2\cdot \one_E \mid\P_{-k}\big]}}_{ T_{1,1} }\cdot
+  \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_k)^4\big]}}_{ T_{1,2} }\bigg]
+\end{align*}
+where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} for the second inequality, and Cauchy–Schwarz for the third inequality.
+
+We first bound $T_{1,2}$ by definition of sub-gaussian random
+vectors. We have for $\x_k$ a $K$-sub-gaussian and $\| \v \| = 1$
+that, $\v^\top \x_k$ is a sub-gaussian random variable with $\|\v^\top
+  \a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute
+constant $C > 0$, see for example
+\cite[Section~2.5.2]{vershynin2018high}.
+
+For $T_{1,1}$ we have
+\[
+  \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-k}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
+\]
+where we denote $s=\E[\hat s]=\tr\,\E[\I - \P_{-k}]\Sigmab$. Note that
+% \begin{align*}
+%   \E\big[(s - \hat s)^2 \cdot\one_E \big] &=
+% \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2 \cdot\one_E \big] +
+%   \E\big[(\tr\,\P_{-k}\Sigmab-\x_k^\top\P_{-k}\x_k)^2 \cdot\one_E \big]\\
+%   &\leq C\Big(n + \E\big[\tr\,(\P_{-k}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
+%     C\big(n + \tr\,\E[\P_{-k}]\Sigmab \cdot\one_E\big)\\
+%   &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
+% \end{align*}
+\begin{align*}
+  \E\big[(s - \hat s)^2 \big] & =
+  \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2  \big] +
+  \E\big[(\tr\,(\I - \P_{-k})\Sigmab-\x_k^\top(\I - \P_{-k})\x_k)^2  \big]                                                                         \\
+                              & \leq C_1 k +  C_2 \E\big[\tr\,(\Sigmab - \P_{-k}\Sigmab)^2 \big] \le C(k+s)\leq C\big(k+\bar s + |s- \bar s| \big)
+\end{align*}
+where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}. Recall that $\bar s = \tr \bar \P_\perp \Sigmab \le \tr \Sigmab = r$ and $k < r$, we have
+\begin{equation}\label{eq:bound-T11}
+  T_{1,1} \le \sqrt{ (\bar s - s)^2 + C (|\bar s - s| + 2 r) }
+\end{equation}
+
+% Moreover, using the fact that
+% $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$, we obtain that:
+% \begin{align*}
+%   |\bar s - s| \cdot\one_E &= |\tr (\bar \P_\perp - \E[\P_{-k}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P_\perp - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-k}] \Sigmab| \\
+%   &= \big|\tr\,(\I-\E[\P]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-k}\x_k\x_k^\top\P_{-k}}{\x_k^\top \P_{-k} \x_k} \right] \Sigmab \\
+%   %&= \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P_\perp\Sigmab\big| \cdot\one_E + 1 \\
+%   &\leq \gamma\,\E\bigg[|\bar s-\hat s|\cdot
+%     \frac{\tr\,\P_{-k}\x_k\x_k^\top\bar\P_\perp\Sigmab}{\tr\,\P_{-k}\x_k\x_k^\top} \cdot\one_E\bigg]
+%     + \gamma\,
+%     \E\bigg[\frac{\tr\,\P_{-k}\x_k\x_k^\top\P_{-k}\Sigmab\bar\P_\perp\Sigmab}
+%     {\tr\,\P_{-k}\x_k\x_k^\top}\bigg] + 1\\
+%   &\leq \frac{\gamma}{\gamma+1}\Big(\E\left[|\bar s-\hat
+%     s| \cdot \frac{\x_k^\top \x_k}{\x_k^\top \P_{-k} \x_k} \cdot\one_E \right] + 1\Big)\leq \frac{\gamma}{\gamma+1}\Big(|\bar
+%     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
+%     s|\big] + 1\Big)\\
+%   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
+%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
+% \end{align*}
+
+It remains to bound $|\bar s - s|$. Note that $\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X
+  (\X^\top\X)^\dagger$ and is symmetric, so
+\begin{align*}
+   & \I - \E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar\P_\perp^{-1} \E[\P_\perp]
+  = 2\E[\P] - \E[\gamma\P_\perp\Sigmab] - \E[\gamma\Sigmab\P_\perp]                                                                                        \\
+   & = \sum_{i=1}^k\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp]) \\
+  %&=k\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top + \x_k\x_k^\top \P_{-k}}{\x_k^\top\P_{-k}\x_k}\bigg] - \gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
+   & =\gamma\,
+  \E\bigg[\bar s\cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})} {\x_k^\top(\I - \P_{-k}) \x_k}\bigg]
+  -\gamma\,\E\bigg[\hat s
+  \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]                                                      \\
+   & +\gamma\,(\E[(\I - \P_{-k})\Sigmab] + \E[\Sigmab ( \I -\P_{-k}) ])- \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp])                               \\
+   & =\gamma\,\E\bigg[(\bar s - \hat s)
+    \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]
+  +    \gamma (\E[\P-\P_{-k}]\Sigmab + \Sigmab \E[\P-\P_{-k}]).
+\end{align*}
+%where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
+Moreover, using the fact that $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$ and $\bar \P_\perp \Sigmab = \Sigmab \bar \P_\perp$, we obtain that
+\begin{align*}
+   & |\bar s - s|  = |\tr (\bar \P_\perp - \E[\I - \P_{-k}] )\Sigmab|   \le |\tr (\bar \P_\perp - \E[\P_\perp] )\Sigmab|   + |\tr \E[\P - \P_{-k}] \Sigmab|                                                                                                                \\
+   & = \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab + \tr\,\bar\P_\perp (\I - \bar \P_\perp^{-1} \E[\P_\perp]) \Sigmab \big|   + \tr\,\E \left[ \frac{(\I - \P_{-k})\x_k\x_k^\top(\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} \right] \Sigmab \\
+   & \leq \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar \P_\perp^{-1} \E[\P_\perp])\bar\P_\perp\Sigmab \big|   + 1                                                                                                                                       \\
+   & \leq \frac{\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
+  \frac{\tr\,( (\I -\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I -\P_{-k})) \bar\P_\perp\Sigmab}{\tr\,(\I -\P_{-k})\x_k\x_k^\top}  \bigg]                                                                                                                                     \\
+   & + \gamma\, \E\bigg[\frac{\tr\,(\I-\P_{-k})\x_k\x_k^\top(\I-\P_{-k})\bar\P_\perp\Sigmab}
+  {\tr\,(\I-\P_{-k})\x_k\x_k^\top}\bigg] + 1                                                                                                                                                                                                                               \\
+   & \leq \frac{\gamma}{\gamma+1}\Bigg(\E\bigg[|\bar s-\hat
+    s| \cdot \frac{\x_k^\top (\I -\P_{-k}) \x_k}{\x_k^\top (\I -\P_{-k}) \x_k}  \bigg] + 1\Bigg) + 1\leq \frac{\gamma}{\gamma+1}\Big(|\bar
+  s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1                                                                                                                                                                                                                               \\
+   & \leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
+  C\sqrt{|\bar s- s|} + C\sqrt{2 r} + 1\Big) + 1.
+\end{align*}
+
+Solving for $|\bar s-s|$, we deduce that
+\begin{align*}
+  |\bar s-s| \leq C_1 \sqrt{r} + C_2,
+  %\gamma(c\sqrt{k+\bar s}+2) + c^2\gamma^2\leq \gamma (c+1)\sqrt{k+\bar s} + c^2\gamma^2.
+\end{align*}
+so plugging back to \eqref{eq:bound-T11} we get $T_{1,1} \le C
+  \sqrt r$ and $\| \T_1\| \le \frac{C_\rho}{\sqrt r} $, thus completing
+the proof.
+
 \section{Explicit formulas under known spectral decay}
@@
 exponential or polynomial rate, arise in many machine
-learning problems \cite{randomized-newton}. Such behavior can be naturally
+learning problems \citep{randomized-newton}. Such behavior can be naturally
 occurring in data, or it can be induced by feature expansion using, say, the
-RBF kernel (for exponential decay) \cite{Santa97Gaussianregression} or the
-Mat\'ern kernel (for polynomial decay) \cite{RasmussenWilliams06}. Understanding these
+RBF kernel (for exponential decay) or
+Mat\'ern (for polynomial decay) kernels \citep{Santa97gaussianregression,RasmussenWilliams06}. Understanding these
 two classes of decay plays an important role in distinguishing the
@@
 matrix $\A$, and study the sketched Nystr\"om method
-\cite{revisiting-nystrom} for low-rank approximation. To handle 
+\citep{revisiting-nystrom} for low-rank approximation. To handle
 the kernel setting in our analysis, it suffices to replace the squared singular
@@
     \subcaption{Singular values are given by $\sigma_i^2=C\cdot \alpha^{i-1}$.}
-    \includegraphics[width=0.48\textwidth]{explicit_exp}
+    \includegraphics[width=\textwidth]{explicit_exp}
   \end{subfigure}
@@
     \subcaption{Singular values are given by $\sigma_i^2=C\cdot i^{-\beta}$.}
-      \includegraphics[width=0.48\textwidth]{explicit_poly}
+    \includegraphics[width=\textwidth]{explicit_poly}
   \end{subfigure}
@@
 
-\ifisarxiv
-\else
-\begin{minipage}{.48\textwidth}
-\hspace{-4.5mm}\includegraphics[width=.57\textwidth]{abalone-nystrom}\nobreak\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{eunite-nystrom}\vspace{-1mm}
-  \captionof{figure}{Theoretical predictions versus approximation error for the
-    sketched Nystr\"om with the RBF kernel
-    (spectral decay shown at the bottom).}\label{f:nystrom}
-\end{minipage}\hspace{4mm}
-\begin{minipage}{.48\textwidth}
-\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{covtype-linear}\nobreak\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{splice-linear}\vspace{-1mm}
-  \captionof{figure}{Theoretical predictions versus approximation error for
-the Gaussian and Rademacher sketches
-  (spectral decay shown at the bottom).}\label{f:linear}
-\end{minipage}
-\fi
+% \ifisarxiv
+% \else
+%   \begin{minipage}{.48\textwidth}
+%     \hspace{-4.5mm}\includegraphics[width=.57\textwidth]{abalone-nystrom}\nobreak\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{eunite-nystrom}\vspace{-1mm}
+%     \captionof{figure}{Theoretical predictions versus approximation error for the
+%       sketched Nystr\"om with the RBF kernel
+%       (spectral decay shown at the bottom).}\label{f:nystrom}
+%   \end{minipage}\hspace{4mm}
+%   \begin{minipage}{.48\textwidth}
+%     \hspace{-2.5mm}\includegraphics[width=.57\textwidth]{covtype-linear}\nobreak\hspace{-2.5mm}\includegraphics[width=.57\textwidth]{splice-linear}\vspace{-1mm}
+%     \captionof{figure}{Theoretical predictions versus approximation error for
+%       the Gaussian and Rademacher sketches
+%       (spectral decay shown at the bottom).}\label{f:linear}
+%   \end{minipage}
+% \fi
 
@@
 benchmark datasets from the libsvm repository
-\cite{libsvm} (further numerical results are in
+\citep{libsvm} (further numerical results are in
 Appendix~\ref{a:experiments}). We repeated every experiment 10 times,
@@
 
-\ifisarxiv
+% \ifisarxiv
 \begin{figure}
@@
 \end{figure}
-\fi
-
-\ifisarxiv
+% \fi
+
+% \ifisarxiv
 \begin{figure}
@@
 \end{figure}
-\fi
+% \fi
 
@@
 % cross-validation.
-Following \cite{revisiting-nystrom}, we use the RBF kernel because
+Following \citep{revisiting-nystrom}, we use the RBF kernel because
 varying the scale parameter $\sigma$
@@
 for two datasets and a range of sketching sizes, as well as in other
-empirical results shown in Appendix \ref{a:experiments}.
+empirical results shown in \Cref{a:experiments}.
+
+\subsection{Additional empirical results on libsvm datasets}
+\label{a:experiments}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=.47\textwidth]{bodyfat-supp}\nobreak\includegraphics[width=.47\textwidth]{housing-supp}
+  \includegraphics[width=.47\textwidth]{mg-supp}\nobreak\includegraphics[width=.47\textwidth]{triazines-supp}
+  \caption{Theoretical predictions versus approximation error for the
+    sketched Nystr\"om with the RBF kernel, using Gaussian and
+    Rademacher sketches (spectral decay shown at the bottom).}\label{f:nystrom2}
+\end{figure}
+
+We complement the results of Section \ref{s:experiments} with
+empirical results on four additional libsvm datasets \citep{libsvm} (bringing
+the total number of benchmark datasets to eight), which further
+establish the accuracy of our surrogate expressions for the low-rank approximation
+error. Similary as in Figure \ref{f:nystrom}, we use the sketched
+Nystr\"om method \citep{revisiting-nystrom} with the RBF kernel
+$k(\a_i,\a_j)=\exp(-\|\a_i-\a_j\|^2/(2\sigma^2))$, for several values
+of the parameter $\sigma$. The values of $\sigma$ were chosen so as to
+demonstrate the effectiveness of our theoretical predictions both when the
+stable rank is moderately large and when it is very small.
+
+In Figure \ref{f:nystrom2} we show the results for both Gaussian and
+Rademacher sketches. These results reinforce the conclusions we made in
+Section \ref{s:experiments}: our theoretical estimates are very
+accurate in all cases, for both sketching methods, and even when the
+stable rank is close to 1 (a regime that is not supported by the
+current theory).
 
@@
 
-\ifisarxiv\else
-\section*{Broader Impact}
-In this paper, we investigate the spectral properties of residual
-(random) projection matrices, commonly appearing in various
-sketching-based methods. The precise theoretical description given in
-this paper provides performance guarantees for popular algorithms
-such as low-rank approximation and many randomized (iterative)
-optimization methods, and contributes to the development of more
-robust and reliable large-scale learning systems. The
-theoretical framework developed in this work presents no foreseeable
-negative societal consequence. 
-\fi
-
-\paragraph{Acknowledgments.}
-We would like to acknowledge DARPA, NSF, and ONR for providing partial support of this work.
+% \ifisarxiv\else
+%   \section*{Broader Impact}
+%   In this paper, we investigate the spectral properties of residual
+%   (random) projection matrices, commonly appearing in various
+%   sketching-based methods. The precise theoretical description given in
+%   this paper provides performance guarantees for popular algorithms
+%   such as low-rank approximation and many randomized (iterative)
+%   optimization methods, and contributes to the development of more
+%   robust and reliable large-scale learning systems. The
+%   theoretical framework developed in this work presents no foreseeable
+%   negative societal consequence.
+% \fi
+
+% \paragraph{Acknowledgments.}
+% We would like to acknowledge DARPA, NSF, and ONR for providing partial support of this work.
 
@@
 % We construct bootstrapped operator norm confidence intervals using
-% \cite{lopes2019bootstrapping}.
+% \citep{lopes2019bootstrapping}.
 
@@
 
-\section{Proof of Theorem~\ref{t:main-tech}}
-\label{sec:proof-of-theo-main-tech}
-
-We first introduce the following technical lemmas.
-
-\begin{lemma}\label{l:rank-one-update}
-    For $\X \in \mathbb R^{k \times n}$ with $k<n$, denote $\P = \X^\dagger \X$ and $\P_{-k} = \X_{-k}^\dagger \X_{-k}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its i-th row $\x_i \in \mathbb R^n$.
-Then, conditioned on the event $E_k: \left\{ \left| \frac{\tr \Sigmab (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} - 1 \right| \le \frac12 \right\}$:
-    \begin{align*}
-      (\X^\top\X)^\dagger\x_k = \frac{(\I - \P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}, \quad \P - \P_{-k} = \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top (\I - \P_{-k}) \x_k}.
-    \end{align*}
-\end{lemma}
-
-\begin{proof}
-% Recall $\P = \I - \X^\dagger \X = \I - (\X^\top \X)^\dagger \X^\top \X = \I - (\X^\top \X)^\dagger (\A + \x_k \x_k^\top)$ and $\P_{-k} = \I - \X_{-k}^\dagger \X_{-k} = \I - \A^\dagger \A$ 
-Since conditioned on $E_k$ we have $\x_k^\top (\I - \P_{-k}) \x_k \neq 0$, from \cite[Theorem 1]{10.2307/2099767} we deduce
-\begin{align*}
-  (\X^\top \X)^\dag = \left( \A + \x_k \x_k^\top \right)^\dagger
-  &= \A^\dag
-  - \frac{\A^\dag \x_k \x_k^\top (\I - \P_{-k}) }{\x_k^\top ( \I - \P_{-k}) \x_k}
-  - \frac{( \I -\P_{-k}) \x_k \x_k^\top \A^\dag}{\x_k^\top ( \I - \P_{-k}) \x_k} \\ 
-  &+ (1 + \x_k^\top \A^\dag \x_k) \frac{(\I - \P_{-k}) \x_k \x_k^\top (\I - \P_{-k})}{(\x_k^\top (\I - \P_{-k}) \x_k)^2}
-\end{align*}
-for $\A = \X_{-k}^\top \X_{-k}$ so that $\I - \P_{-k} = \I - \A^\dagger \A$, where we used the fact that $\I - \P_{-k}$ is a projection matrix so that $(\I - \P_{-k})^2 = \I - \P_{-k}$.
-As a consequence, multiplying by $\x_k$ and simplifying we get
-\[
-  (\X^\top \X)^\dag \x_k = \frac{ ( \I -\P_{-k}) \x_k}{\x_k^\top (\I - \P_{-k}) \x_k}.
-\]
-By definition of the pseudoinverse, $\P = \X^\dagger \X = (\X^\top \X)^\dagger \X^\top \X$ so that 
-\[
-  \P - \P_{-k} = \X^\dagger \X - \X_{-k}^\dagger \X_{-k} = \frac{( \I - \P_{-k}) \x_k \x_k^\top ( \I - \P_{-k})}{\x_k^\top ( \I - \P_{-k}) \x_k}
-\]
-where we used $\A (\I - \P_{-k}) = \A - \A \A^\dagger \A = 0$ and thus the conclusion.
-\end{proof}
-
-% \begin{lemma}[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Corollary 2.6]{zajkowski2018bounds}}] \label{l:concentration-sub-gaussian}
-% For a $K$-sub-gaussian random vector $\x \in \mathbb R^d$ (i.e., $\| \x \|_{\psi_2} \leq K$) and $\A \in \mathbb R^{d \times d}$, we have
-% \begin{align*}
-%   \Pr\left[
-%     \lvert \x^\top \A \x - \E[\x^\top \A \x] \rvert
-%     \geq t
-%     \right] \leq 2 \exp\left(
-%     -\min\left\{
-%       \frac{t^2}{16 C^2  K^4 (\tr \A)^2},
-%       \frac{t}{4 C K^2 (\tr \A)},
-%     \right\}
-%   \right)
-% \end{align*}
-% and 
-% \begin{align*}
-%   \E\left[ \left( \x^\top \A \x - \E[\x^\top \A \x] \right)^2 \right]
-%   &\leq C~K^4~\tr(\A)^2
-% \end{align*}
-% for some universal constant $C$.
-% \end{lemma}
-
-% \begin{lemma}
-% %[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}]
-% \label{l:concentration-sub-gaussian}
-% For a $K$-sub-gaussian random vector $\x \in \mathbb R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and positive semi-definite matrix $\A \in \mathbb R^{n \times n}$ such thats $\| \A \| \le 1 $, we have
-% \[
-%   \Pr\left[
-%     \lvert \x^\top \A \x - \tr \A \rvert
-%     \geq \frac12 \tr \A
-%     \right] \leq 2 \exp\left( - \min \left\{ \frac{r_\A}{4 C^2 K^4}, \frac{\sqrt r_\A}{2 C K^2}  \right\} \right)
-% \]
-% with $r_\A = \frac{\tr \A}{\| \A \|}$ the nuclear rank of $\A$, and 
-% \begin{align*}
-%   \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
-%   &\leq cK^4\tr\A^2
-% \end{align*}
-% for some $C,c > 0$ independent of $k$.
-% \end{lemma}
-
-% \begin{proof}
-% % From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \Sigmab$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
-% % \[
-% %   \Pr \left\{ \left| \x^\top \A \x - \tr \A \Sigmab \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
-% % \]
-% % for some universal constant $C > 0$. Taking $t = \frac12 \tr \A \Sigmab$ we obtain
-% From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
-% \[
-%   \Pr \left\{ \left| \x^\top \A \x - \tr \A  \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
-% \]
-% for some universal constant $C > 0$. Taking $t = \frac12 \tr \A$ we have
-% \[
-%   \frac{t^2}{C^2 K^4 \tr\A^2 } = \frac{ (\tr \A)^2 }{4 C^2 K^4 \tr\A^2 } \ge \frac{ \tr \A }{4 C^2 K^4 \| \A \| } = \frac{r_\A}{4 C^2 K^4}, \quad \frac{t}{C K^2 \sqrt{\tr\A^2}  } \ge \frac{\sqrt r_\A}{2 C K^2}
-% \]
-% where we use the fact that $\tr \A^2 \le \| \A \| \tr \A$ and denote $r_\A = \frac{\A}{\| \A \|}$ the nuclear rank of $\A$.
-
-% Integrating this bound yields, we get
-% \[
-%   \E\left[ (\x^\top \A \x - \tr \A )^2 \right] \le c~K^4~\tr \A^2
-% \]
-% and thus the conclusion.
-% \end{proof}
-%
-%
-%
-\begin{lemma}
-%[Concentration of quadratic forms of sub-gaussian random vectors, {\cite[Theorem~2.1]{hsu2012tail}}]
-\label{l:concentration-sub-gaussian}
-For a $K$-sub-gaussian random vector $\x \in \mathbb R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and positive semi-definite matrix $\A \in \mathbb R^{n \times n}$, we have
-\[
-  \Pr\left[
-    \lvert \x^\top \A \x - \tr \A \rvert
-    \geq \frac13 \tr \A
-    \right] \leq 2 \exp\left( - \min \left\{ \frac{r_{\!\A}}{9 C^2 K^4}, \frac{\sqrt{r_{\!\A}}}{3 C K^2}  \right\} \right)
-\]
-with $r_{\!\A} = \tr \A/\| \A \|$ the stable rank of $\A$, and 
-\begin{align*}
-  \E\left[ \left( \x^\top \A \x - \tr \A \right)^2 \right]
-  &\leq c~K^4~\tr \A^2
-\end{align*}
-for some $C,c > 0$ independent of $K$.
-\end{lemma}
-
-\begin{proof}
-% From \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \Sigmab$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
-% \[
-%   \Pr \left\{ \left| \x^\top \A \x - \tr \A \Sigmab \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
-% \]
-% for some universal constant $C > 0$. Taking $t = \frac12 \tr \A \Sigmab$ we obtain
-This follows from a Hanson-Wright type \cite{rudelson2013hanson} sub-gaussian concentration inequality. More precisely,
-from \cite[Corollary 2.9]{zajkowski2018bounds} we have, for $K$-sub-gaussian $\x \in \R^n$ with $\E[\x] = 0$, $\E[\x \x^\top] = \I_n$ and symmetric positive semi-definite $\A \in \R^{n \times n}$ that
-\[
-  \Pr \left\{ \left| \x^\top \A \x - \tr \A  \right| \ge t \right\} \le 2 \exp \left( - \min \left\{ \frac{t^2}{C^2 K^4 \tr\A^2 },  \frac{t}{C K^2 \sqrt{\tr\A^2}  } \right\} \right)
-\]
-for some universal constant $C > 0$. Taking $t = \frac13 \tr \A$ we have
-\[
-  \frac{t^2}{C^2 K^4 \tr\A^2 } = \frac{ (\tr \A)^2 }{9 C^2 K^4 \tr\A^2 } \ge \frac{ \tr \A }{9 C^2 K^4 \| \A \| } = \frac{r_{\!\A}}{9 C^2 K^4}, \quad \frac{t}{C K^2 \sqrt{\tr\A^2}  } \ge \frac{\sqrt{r_{\!\A}}}{3 C K^2}
-\]
-where we use the fact that $\tr \A^2 \le \| \A \| \tr \A$.
-
-Integrating this bound yields:
-\[
-  \E\left[ (\x^\top \A \x - \tr \A )^2 \right] \le c~K^4~\tr \A^2
-\]
-and thus the conclusion.
-\end{proof}
-
-% \begin{lemma}\label{l:rank-one-projection}
-%   For any $d\times d$ p.s.d.~matrix $\A$ and
-%   any random vector $\z$ with jointly
-%   $K$-sub-gaussian (sub-gaussian Orlicz norm
-%   $\|\z^\top \a \|_{\psi_2} \leq K$ for all $\|\a\| \leq 1$),
-%   \begin{align*}
-%     \Bigg\|\E\bigg[\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\bigg]\Bigg\|
-% \leq
-%     \sup_{\v:\|\v\|=1}\sqrt{\E\bigg[\Big(\v^\top\frac{\A^{\sfrac12}\z\z^\top\A^{\sfrac12}}{\z^\top\A\z}\v\Big)^2\bigg]}
-%     \leq C\cdot K^2 \cdot \frac{\|\A\|}{\tr\,\A}.
-%   \end{align*}
-%   for some universal constant $C$.
-% \end{lemma}
-%\textbf{** Lemma 3 seems no longer necessary?**}
-
-\begin{lemma}\label{l:trace}
-With the notations of Lemma~\ref{l:rank-one-update}, for $X=\tr\,\Sigmab(\P_{-k}-\E[\P_{-k}])$
-and $\|\Sigmab\| = 1$, we have
-\begin{align*}
-  \E[X^2]\leq
-  Ck\quad\text{and}\quad\Pr\{|X|\geq t\}\leq 2 \ee^{-\frac{t^2}{ck}}.
-  \end{align*}
-  for some universal constant $C,c > 0$.
-\end{lemma}
-\begin{proof}
-To simplify notations, we work on $\P$ instead of $\P_{-k}$, the same line of argument applies to $\P_{-k}$ by changing the sample size $k$ to $k-1$.
-
-First note that
-\begin{align*}
-  X
-  &= \tr \Sigmab (\P - \E \P)
-  = \E_k [\tr \Sigmab \P]  - \E_0[ \tr \Sigmab \P ]\\
-  %%%
-  &= \sum_{i=1}^k \left( \E_i[\tr \Sigmab \P] - \E_{i-1}[\tr \Sigmab \P]\right)
-  = \sum_{i=1}^k (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})
-\end{align*}
-where we used the fact that $\E_i [\tr \Sigmab \P_{-i}] = \E_{i-1} [\tr \Sigmab \P_{-i}]$, for $\E_i[\cdot]$ the conditional expectation with respect to $\mathcal F_i$ the $\sigma$-field generating the rows $\x_1 \ldots, \x_i$ of $\X$.
-This forms a martingale difference sequence (it is a difference sequence of
-the Doob martingale for $\tr \Sigmab (\P - \P_{-i})$ with respect to filtration $\mathcal{F}_i$)
-hence it falls within the scope of the Burkholder inequality \cite{burkholder1973distribution}, recalled as follows.
-
-\begin{lemma}\label{l:burkholder}
-For $\{ x_i \}_{i=1}^k$ a real martingale difference sequence with respect to the increasing $\sigma$ field $\mathcal F_i$, we have, for $L > 1$, there exists $C_L > 0$ such that
-\[
-  \E \bigg[
-    \Big| \sum_{i=1}^k x_i \Big|^L
-  \bigg]
-  \le C_L \E \bigg[
-    \Big( \sum_{i=1}^k |x_i|^2 \Big)^{L/2}
-  \bigg].
-\]
-\end{lemma}
-
-From Lemma~\ref{l:rank-one-update}, $\P - \P_{-i} = \frac{(\I - \P_{-i}) \x_i \x_i^\top (\I - \P_{-i}) }{ \x_i^\top (\I - \P_{-i}) \x_i}$ is positive semi-definite, we have $\tr \Sigmab (\P - \P_{-i}) \le \| \Sigmab \| = 1$ so that with Lemma~\ref{l:burkholder} we obtain with $x_i = (\E_i - \E_{i-1}) \tr \Sigmab (\P - \P_{-i})$ that, for $L > 1$
-\[
-  \E |X|^L \le C_L k^{L/2}.
-\]
-In particular, for $L=2$, we obtain $\E |X|^2 \le C k$.
-%for $c = (36 \sqrt{2})^2$ (details in proof of Theorem~3.2 in \cite{burkholder1973distribution}).
-
-For the second result, since we have almost surely bounded martingale differences
-($\lvert x_i \rvert \leq 2$), by the Azuma-Hoeffding inequality
-\begin{align*}
-  \Pr\{
-    \lvert X \rvert
-    \geq t
-  \}
-  \leq 2 \ee^{\frac{-t^2}{8 k}}
-\end{align*}
-as desired.
-\end{proof}
-
-% \subsection{Complete proof of Theorem~\ref{t:main-tech}}
-
-% Equipped with the lemmas above, we are ready to prove Theorem~\ref{t:main-tech}. First note that:
-% \begin{enumerate}[leftmargin=*]
-%   \item  Since $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, we can assume without loss of generality (after rescaling $\bar\P_\perp$ correspondingly) that $\|\Sigmab\| = 1$.
-%   \item According to the definition of $\bar \P_\perp$ and $\gamma$, the following bounds hold
-%   \begin{equation}\label{eq:bound-bar-gamma-P}
-%     \frac1{\gamma + 1} \I \preceq \bar \P_\perp \preceq \I, \quad \gamma \le \frac{k}{r-k} = \frac1{\rho - 1}
-%   \end{equation}
-%   for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}k > 1$, where we used the fact that
-%   \[
-%     k = n - \tr\,\bar\P_\perp = \tr\,\bar\P_\perp (\gamma\Sigmab + \I) - \tr\,\bar\P_\perp
-%   = \gamma \tr\,\bar\P_\perp \Sigmab\geq \frac{\gamma}{\gamma+1}\tr\,\Sigmab,
-%   \]
-%   so that $r = \tr \Sigmab \le k \cdot \frac{\gamma + 1}{\gamma}$.%= k + \tr \bar \P_\perp \Sigmab
-%   \item As already discussed in Section~\ref{subsec:proof-sketch}, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the following spectral norm
-%   \begin{equation}\label{eq:proof-1-SM}
-%     \| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}
-%   \end{equation}
-%   so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ from \eqref{eq:bound-bar-gamma-P}, we have
-%   \begin{align*}
-%     \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
-%     \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp^{\frac12}\|\leq
-%   \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
-%   \end{align*}
-%   Defining $\epsilon =\frac{C_\rho}{\sqrt r}
-%   \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
-%   of the p.s.d.~matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
-%   in the interval $[1-\epsilon,1+\epsilon]$, and
-%   \begin{align*}
-%     (1-\epsilon)\I\preceq
-%     \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.
-%   \end{align*}
-%   so that by multiplying $\bar \P_{\perp}^{\frac12}$ on both sides, we obtain the
-%   desired bound.
-% \end{enumerate}
-
-% \bigskip
-
-% As a consequence of the above observations, we only need to prove \eqref{eq:proof-1-SM} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:
-
-% \begin{enumerate}[leftmargin=*]
-%   \item For $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its $i$-th row, we define, for $i \in \{ 1, \ldots, k \}$, the following events
-%   \begin{equation}\label{eq:def-events-E-F}
-%     E_i: \left\{ \left| \frac{\tr (\I - \P_{-i}) \Sigmab}{\x_i^\top ( \I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\}, \quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\}.
-%F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
-%   \end{equation}
-%   where we recall $\x_i \in \R^n$ is the $i$-th row of $\X$ so that $\E[\x_i] = 0$ and $\E[\x_i \x_i^\top] = \Sigmab$.
-With Lemma~\ref{l:concentration-sub-gaussian}, we can bound the probability of $\neg E_i$ and $\neg F_i$, and consequently that of $\neg E$ for $E = \bigwedge_{i=1}^k(E_i \wedge F_i)$;
-%   \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \|$. More precisely, since
-% %   \begin{align*}
-% %   \I&-\E[\P]\bar\P_\perp^{-1}
-% %   = \E[\X^\dagger\X] - \gamma\E[\P]\Sigmab\\
-% %   &= \E[\X^\dagger\X\one_E] +
-% %     \E[\X^\dagger\X \one_{\neg E}]
-% %     -\gamma\E[\P]\Sigmab\\
-% %   &=n\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top}{\x_k^\top\P_{-k}\x_k}\one_E\bigg]
-% %     -\gamma\E[\P]\Sigmab \ +\     \E[\X^\dagger\X\one_{\neg E}]\\
-% %   &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
-% %     s)\cdot\frac{\P_{-k}\x_k\x_k^\top}{\x_k^\top\P_{-k}\x_k}\one_E\bigg]}_{\T_1}
-% %     + \gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2} + \gamma \underbrace{\E[\P_{-k}-\P]\Sigmab}_{\T_3} + \underbrace{\E[\X^\dagger\X\one_{\neg E}]}_{\T_4}
-% % \end{align*}
-% \begin{align*}
-%   \I-\E[\P_\perp]\bar\P_\perp^{-1}
-%     &= \E[\P] - \gamma\E[\P_\perp]\Sigmab
-%     \\
-%     &= \E[\P \cdot \one_E] +
-%     \E[\P  \cdot \one_{\neg E}]
-%     -\gamma\E[\P_\perp]\Sigmab\\
-%   &=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg]
-%     -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
-%   &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
-%     s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
-%     - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2} \\ 
-%     &+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
-%     %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
-% \end{align*}
-% where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_k^\top ( \I - \P_{-k}) \x_k$ as well as $\bar s = \tr \bar \P_\perp \Sigmab =k/\gamma$.
-It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ to reach the conclusion.
-% \end{enumerate}
-
-% Another important relation that will be constantly used throughout the proof is
-% \begin{equation}\label{eq:bound-trace-P-Sigma}
-%     \tr (\I - \P_{-k}) \Sigmab = \tr \Sigmab^{\frac12} (\I - \P_{-k})^2 \Sigmab^{\frac12} =
-%     \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-k}^\dagger\X_{-k}\|_F^2\geq
-% \sum_{i\geq k}\lambda_i (\Sigmab)\geq r-k
-% \end{equation}
-% where we used the fact that $\rank(\X_{-k}^\dagger \X_{-k}) \le \rank(\X_{-k}) \le k-1$ and arranged the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_n(\Sigmab)$ in a non-increasing order.
-As a consequence, we also have
-% \begin{equation}\label{eq:bound-norm-P-Sigma}
-%   \frac{\tr (\I - \P_{-k}) \Sigmab}{ \|  (\I - \P_{-k}) \Sigmab \| } \ge \tr (\I - \P_{-k}) \Sigmab \geq r-k.
-% \end{equation}
-
-% \bigskip
-
-% For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian}~and~\eqref{eq:bound-norm-P-Sigma} that
-% \begin{align*}
-%   \Pr (\neg E_i) &\le \Pr \left\{ \lvert \x_i^\top (\I - \P_{-i}) \x_i - \tr \Sigmab (\I - \P_{-i})  \rvert
-%     \geq \frac13 \tr \Sigmab (\I - \P_{-i}) \right\} \\ 
-%     &\le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }. % \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
-% \end{align*}
-% %for some constant $c_\rho$ depending on fixed $\rho \equiv \frac{r}n > 1$. 
-% Similarly we have
-% \[
-%   \Pr (\neg F_i) \le 2 \ee^{ - \min \left\{ \frac{ r }{9 C^2 K^4}, \frac{\sqrt{r} }{3 C K^2} \right\} } \le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }
-% \]
-% and with the union bound we obtain
-% \begin{equation}
-%   \Pr (\neg E) \le 4 k \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{k}{(r-k)^2} \cdot 4 (r-k)^2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{C_\rho}{r - k}
-% \end{equation}
-% where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ and  $x^4 \ee^{-\alpha x} \le \frac{256 \ee^{-4}}{\alpha^4}$ on $x > 0$.
-Also, denote $c_\rho = \frac{r - k}r = \frac{\rho - 1}{\rho} > 0$, we have
-% \begin{equation}\label{eq:proba-E}
-%   \Pr(\neg E) \le \frac{C_\rho}{r - k} = \frac{C_\rho}{c_\rho r} = \frac{C'_\rho}{r} 
-% \end{equation}
-% for some $C'_\rho > 0$ that depends on $\rho = r/k > 1$ and the sub-gaussian norm $K$.
-
-% \medskip
-
-% At this point, note that, conditioned on the event $E$, we have for $i \in \{1, \ldots, k \}$
-% \begin{equation}
-%     \frac12 \frac1{\tr (\I - \P_{-i}) \Sigmab} \le \frac1{\x_i^\top (\I - \P_{-i}) \x_i} \le \frac32 \frac1{\tr (\I - \P_{-i}) \Sigmab}, \quad  \frac1{2r} \le \frac1{\x_i^\top \x_i} \le \frac3{2r}.
-%\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
-% \end{equation}
-% Also, with \eqref{eq:proba-E} and the fact that $\| \I - \P_{-k} \|
-% \le 1$ and $\| \P \| \le 1$, we have $\| \T_2 \| + \| \T_4 \| \le
-% \frac{C_\rho}{r}$ for some $\C_\rho > 0$ that depends on $\rho$ and $K$. And it thus remains to handle the terms $\T_1$ and $\T_3$ to obtain a bound on $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| $.
-
-% To bound $\T_3$, with $\P - \P_{-k} = \frac{(\I - \P_{-k})\x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}$ in Lemma~\ref{l:rank-one-update}, we have
-% \begin{align*}
-% \|\T_3\| &\leq  \bigg\|
-%   \E\bigg[\frac{ (\I - \P_{-k}) \x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}
-%   \cdot \one_E \bigg]\bigg\|  + \| \E[ (\P - \P_{-k}) \cdot \one_{\neg E} ] \| \\ 
-%   %%%
-%   &\le \frac32 \E \left[ \frac1{ \tr (\I - \P_{-k} ) \Sigmab } \right] + \frac{c_\rho}{r - k} \leq \frac{C_\rho}{r-k} = \frac{C'_\rho}{r}
-% \end{align*}
-% where we used the fact that $\tr\,(\I - \P_{-k}) \Sigmab \ge r -k $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv r/k > 1$.
-
-% For $\T_1$ we write
-% \begin{align*}
-%   \|\T_1\|    &\leq\E\bigg[\|\I - \P_{-k}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
-%     \frac{\x_k\x_k^\top}{\x_k^\top( \I - \P_{-k}) \x_k} \cdot \one_E\mid\P_{-k}\Big]
-%     \Big\|\bigg]
-%   \\
-%           &\leq \frac32 \frac1{r-k} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
-%             |\bar s-\hat s|\cdot \v^\top\x_k\x_k^\top\v\cdot\one_E\mid
-%             \P_{-k}\Big]\bigg]
-%   \\
-%   &\leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
-%     s)^2\cdot \one_E \mid\P_{-k}\big]}}_{ T_{1,1} }\cdot
-%     \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_k)^4\big]}}_{ T_{1,2} }\bigg]
-% \end{align*}
-% where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} for the second inequality, and Cauchy–Schwarz for the third inequality.
-
-% We first bound $T_{1,2}$ by definition of sub-gaussian random
-% vectors. We have for $\x_k$ a $K$-sub-gaussian and $\| \v \| = 1$
-% that, $\v^\top \x_k$ is a sub-gaussian random variable with $\|\v^\top
-% \a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute
-% constant $C > 0$, see for example
-% \cite[Section~2.5.2]{vershynin2018high}.
-
-% For $T_{1,1}$ we have
-% \[
-%   \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-k}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
-% \]
-% where we denote $s=\E[\hat s]=\tr\,\E[\I - \P_{-k}]\Sigmab$. Note that
-% % \begin{align*}
-% %   \E\big[(s - \hat s)^2 \cdot\one_E \big] &=
-% % \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2 \cdot\one_E \big] +
-% %   \E\big[(\tr\,\P_{-k}\Sigmab-\x_k^\top\P_{-k}\x_k)^2 \cdot\one_E \big]\\
-% %   &\leq C\Big(n + \E\big[\tr\,(\P_{-k}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
-% %     C\big(n + \tr\,\E[\P_{-k}]\Sigmab \cdot\one_E\big)\\
-% %   &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
-% % \end{align*}
-% \begin{align*}
-%   \E\big[(s - \hat s)^2 \big] &=
-% \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2  \big] +
-%   \E\big[(\tr\,(\I - \P_{-k})\Sigmab-\x_k^\top(\I - \P_{-k})\x_k)^2  \big]\\
-%   &\leq C_1 k +  C_2 \E\big[\tr\,(\Sigmab - \P_{-k}\Sigmab)^2 \big] \le C(k+s)\leq C\big(k+\bar s + |s- \bar s| \big)
-% \end{align*}
-% where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}. Recall that $\bar s = \tr \bar \P_\perp \Sigmab \le \tr \Sigmab = r$ and $k < r$, we have
-% \begin{equation}\label{eq:bound-T11}
-%   T_{1,1} \le \sqrt{ (\bar s - s)^2 + C (|\bar s - s| + 2 r) }
-% \end{equation}
-
-% % Moreover, using the fact that
-% % $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$, we obtain that:
-% % \begin{align*}
-% %   |\bar s - s| \cdot\one_E &= |\tr (\bar \P_\perp - \E[\P_{-k}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P_\perp - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-k}] \Sigmab| \\
-% %   &= \big|\tr\,(\I-\E[\P]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-k}\x_k\x_k^\top\P_{-k}}{\x_k^\top \P_{-k} \x_k} \right] \Sigmab \\
-% %   %&= \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P_\perp\Sigmab\big| \cdot\one_E + 1 \\
-% %   &\leq \gamma\,\E\bigg[|\bar s-\hat s|\cdot
-% %     \frac{\tr\,\P_{-k}\x_k\x_k^\top\bar\P_\perp\Sigmab}{\tr\,\P_{-k}\x_k\x_k^\top} \cdot\one_E\bigg]
-% %     + \gamma\,
-% %     \E\bigg[\frac{\tr\,\P_{-k}\x_k\x_k^\top\P_{-k}\Sigmab\bar\P_\perp\Sigmab}
-% %     {\tr\,\P_{-k}\x_k\x_k^\top}\bigg] + 1\\
-% %   &\leq \frac{\gamma}{\gamma+1}\Big(\E\left[|\bar s-\hat
-% %     s| \cdot \frac{\x_k^\top \x_k}{\x_k^\top \P_{-k} \x_k} \cdot\one_E \right] + 1\Big)\leq \frac{\gamma}{\gamma+1}\Big(|\bar
-% %     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
-% %     s|\big] + 1\Big)\\
-% %   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
-% %      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
-% % \end{align*}
-
-% It remains to bound $|\bar s - s|$. Note that $\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X
-% (\X^\top\X)^\dagger$ and is symmetric, so
-% \begin{align*}
-%   &\I - \E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar\P_\perp^{-1} \E[\P_\perp]
-%   = 2\E[\P] - \E[\gamma\P_\perp\Sigmab] - \E[\gamma\Sigmab\P_\perp]\\
-%   &= \sum_{i=1}^k\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp])\\
-%   %&=k\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top + \x_k\x_k^\top \P_{-k}}{\x_k^\top\P_{-k}\x_k}\bigg] - \gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
-%   &=\gamma\,
-%     \E\bigg[\bar s\cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})} {\x_k^\top(\I - \P_{-k}) \x_k}\bigg]
-%     -\gamma\,\E\bigg[\hat s
-%     \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg] \\ 
-%   & +\gamma\,(\E[(\I - \P_{-k})\Sigmab] + \E[\Sigmab ( \I -\P_{-k}) ])- \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp]) \\
-%   &=\gamma\,\E\bigg[(\bar s - \hat s)
-%     \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]
-% +    \gamma (\E[\P-\P_{-k}]\Sigmab + \Sigmab \E[\P-\P_{-k}]). 
-% \end{align*}
-% %where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
-% Moreover, using the fact that $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$ and $\bar \P_\perp \Sigmab = \Sigmab \bar \P_\perp$, we obtain that
-% \begin{align*}
-%   &|\bar s - s|  = |\tr (\bar \P_\perp - \E[\I - \P_{-k}] )\Sigmab|   \le |\tr (\bar \P_\perp - \E[\P_\perp] )\Sigmab|   + |\tr \E[\P - \P_{-k}] \Sigmab| \\
-%   &= \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab + \tr\,\bar\P_\perp (\I - \bar \P_\perp^{-1} \E[\P_\perp]) \Sigmab \big|   + \tr\,\E \left[ \frac{(\I - \P_{-k})\x_k\x_k^\top(\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} \right] \Sigmab \\
-%   &\leq \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar \P_\perp^{-1} \E[\P_\perp])\bar\P_\perp\Sigmab \big|   + 1 \\
-%   &\leq \frac{\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
-%     \frac{\tr\,( (\I -\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I -\P_{-k})) \bar\P_\perp\Sigmab}{\tr\,(\I -\P_{-k})\x_k\x_k^\top}  \bigg] \\
-%   &+ \gamma\, \E\bigg[\frac{\tr\,(\I-\P_{-k})\x_k\x_k^\top(\I-\P_{-k})\bar\P_\perp\Sigmab}
-%     {\tr\,(\I-\P_{-k})\x_k\x_k^\top}\bigg] + 1\\
-%   &\leq \frac{\gamma}{\gamma+1}\Bigg(\E\bigg[|\bar s-\hat
-%     s| \cdot \frac{\x_k^\top (\I -\P_{-k}) \x_k}{\x_k^\top (\I -\P_{-k}) \x_k}  \bigg] + 1\Bigg) + 1\leq \frac{\gamma}{\gamma+1}\Big(|\bar
-%     s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1 \\
-%   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
-%      C\sqrt{|\bar s- s|} + C\sqrt{2 r} + 1\Big) + 1.
-% \end{align*}
-
-% Solving for $|\bar s-s|$, we deduce that
-% \begin{align*}
-%   |\bar s-s| \leq C_1 \sqrt{r} + C_2,
-%   %\gamma(c\sqrt{k+\bar s}+2) + c^2\gamma^2\leq \gamma (c+1)\sqrt{k+\bar s} + c^2\gamma^2.
-% \end{align*}
-% so plugging back to \eqref{eq:bound-T11} we get $T_{1,1} \le C
-% \sqrt r$ and $\| \T_1\| \le \frac{C_\rho}{\sqrt r} $, thus completing
-% the proof.
-
-\subsection{Complete proof of Theorem~\ref{t:main-tech}}
-
-Equipped with the lemmas above, we are ready to prove Theorem~\ref{t:main-tech}. First note that:
-\begin{enumerate}[leftmargin=*]
-  \item  Since $\X^\dag \X \overset{d}{=} (\alpha \X)^\dag (\alpha \X)$ for any $\alpha \in \R \setminus \{0\}$, we can assume without loss of generality (after rescaling $\bar\P_\perp$ correspondingly) that $\|\Sigmab\| = 1$.
-  \item According to the definition of $\bar \P_\perp$ and $\gamma$, the following bounds hold
-  \begin{equation}\label{eq:bound-bar-gamma-P}
-    \frac1{\gamma + 1} \I \preceq \bar \P_\perp \preceq \I, \quad \gamma \le \frac{k}{r-k} = \frac1{\rho - 1}
-  \end{equation}
-  for $r \equiv \frac{\tr \Sigmab}{ \| \Sigmab \| } = \tr \Sigmab$ and $\rho \equiv \frac{r}k > 1$, where we used the fact that
-  \[
-    k = n - \tr\,\bar\P_\perp = \tr\,\bar\P_\perp (\gamma\Sigmab + \I) - \tr\,\bar\P_\perp
-  = \gamma \tr\,\bar\P_\perp \Sigmab\geq \frac{\gamma}{\gamma+1}\tr\,\Sigmab,
-  \]
-  so that $r = \tr \Sigmab \le k \cdot \frac{\gamma + 1}{\gamma}$.%= k + \tr \bar \P_\perp \Sigmab
-  \item As already discussed in Section~\ref{subsec:proof-sketch}, to obtain the lower and upper bound for $\E[\P_\perp]$ in the sense of symmetric matrix as in Theorem~\ref{t:main-tech}, it suffices to bound the following spectral norm
-  \begin{equation}\label{eq:proof-1-SM}
-    \| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| \le \frac{C_\rho}{\sqrt r}, %\quad~\text{or}~\| \I - \bar \P_\perp^{-1} \E[\P_\perp] \| \le \frac{C_\rho}{\sqrt r}
-  \end{equation}
-  so that, with $\frac{\rho-1}{\rho}\I \preceq \bar\P_\perp \preceq \I$ from \eqref{eq:bound-bar-gamma-P}, we have
-  \begin{align*}
-    \|\I-\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\| =
-    %\|\bar\P_\perp^{\frac12}(\I-\bar\P_\perp^{-1}\E[\P_\perp])\bar\P_\perp^{-\frac12}\| =
-    \|\bar\P_\perp^{-\frac12}(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp^{\frac12}\|\leq
-  \frac{C_\rho}{\sqrt r} \sqrt{\frac{\rho}{\rho-1}}.
-  \end{align*}
-  Defining $\epsilon =\frac{C_\rho}{\sqrt r}
-  \sqrt{\frac{\rho}{\rho-1}}$, this means that all eigenvalues 
-  of the p.s.d.~matrix $\bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}$ lie
-  in the interval $[1-\epsilon,1+\epsilon]$, and
-  \begin{align*}
-    (1-\epsilon)\I\preceq
-    \bar\P_\perp^{-\frac12}\E[\P_\perp]\bar\P_\perp^{-\frac12}\preceq (1+\epsilon)\I.
-  \end{align*}
-  so that by multiplying $\bar \P_{\perp}^{\frac12}$ on both sides, we obtain the
-  desired bound.
-\end{enumerate}
-
-\bigskip
-
-As a consequence of the above observations, we only need to prove \eqref{eq:proof-1-SM} under the setting $\| \Sigmab \| = 1$. The proof comes in the following two steps:
-
-\begin{enumerate}[leftmargin=*]
-  \item For $\P_{-i} = \X_{-i}^\dagger \X_{-i}$, with $\X_{-i} \in \mathbb R^{(k-1) \times n}$ the matrix $\X$ without its $i$-th row, we define, for $i \in \{ 1, \ldots, k \}$, the following events
-  \begin{equation}\label{eq:def-events-E-F}
-    E_i: \left\{ \left| \frac{\tr (\I - \P_{-i}) \Sigmab}{\x_i^\top ( \I - \P_{-i}) \x_i} - 1 \right| \le \frac12 \right\}, 
-    %\quad F_i : \left\{ \left|\frac{\tr \Sigmab}{\x_i^\top \x_i} - 1 \right| \le \frac12  \right\}. %F_i : \left\{ \frac{\tr \Sigmab}{\x_i^\top \x_i} \le 2 \right\}
-  \end{equation}
-  where we recall $\x_i \in \R^n$ is the $i$-th row of $\X$ so that $\E[\x_i] = 0$ and $\E[\x_i \x_i^\top] = \Sigmab$.
-With Lemma~\ref{l:concentration-sub-gaussian}, we can bound the probability of $\neg E_i$, and consequently that of $\neg E$ for $E = \bigwedge_{i=1}^k E_i$;
-  \item We then bound, conditioned on $E$ and $\neg E$ respectively, the spectral norm $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \|$. More precisely, since
-\begin{align*}
-  \I-\E[\P_\perp]\bar\P_\perp^{-1}
-    &= \E[\P] - \gamma\E[\P_\perp]\Sigmab
-    \\
-    &= \E[\P \cdot \one_E] +
-    \E[\P  \cdot \one_{\neg E}]
-    -\gamma\E[\P_\perp]\Sigmab\\
-  &=k\,\E\bigg[\frac{(\I - \P_{-k})\x_k\x_k^\top}{\x_k^\top(\I - \P_{-k})\x_k} \cdot \one_E\bigg]
-    -\gamma\E[\P_\perp]\Sigmab \ +\     \E[\P \cdot \one_{\neg E}]\\
-  &=\gamma\,\underbrace{\E\bigg[(\bar s-\hat
-    s)\cdot\frac{(\I -\P_{-k})\x_k\x_k^\top}{\x_k^\top(\I -\P_{-k})\x_k} \cdot \one_E\bigg]}_{\T_1}
-    - \gamma \underbrace{\E[ (\I - \P_{-k}) \x_k\x_k^\top \cdot \one_{\neg E}]}_{\T_2} \\ 
-    &+ \gamma \underbrace{\E[\P-\P_{-k}]\Sigmab}_{\T_3} + \underbrace{\E[\P \cdot \one_{\neg E}]}_{\T_4},
-    %\gamma \underbrace{\E[\P_{-k}\x_k\x_k^\top\one_E] - \E[\P_{-k}]\Sigmab}_{\T_2}
-\end{align*}
-where we used Lemma~\ref{l:rank-one-update} for the third equality and denote $\hat s = \x_k^\top ( \I - \P_{-k}) \x_k$ as well as $\bar s = \tr \bar \P_\perp \Sigmab =k/\gamma$.
-It then remains to bound the spectral norms of $ \T_1, \T_2, \T_3 ,\T_4$ to reach the conclusion.
-\end{enumerate}
-
-% Note that, $\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I- \bar\P_\perp^{-1}\E[\P_\perp]$ is symmetric and 
-% \begin{equation}
-%   \I-\E[\P_\perp]\bar\P_\perp^{-1} + \I- \bar\P_\perp^{-1}\E[\P_\perp] = \gamma (\T_1 + \T_1^\top) - \gamma (\T_2 + \T_2^\top) + 2 \gamma \T_3 + 2 \T_4
-% \end{equation}
-
-Another important relation that will be constantly used throughout the proof is
-\begin{equation}\label{eq:bound-trace-P-Sigma}
-    \tr (\I - \P_{-k}) \Sigmab = \tr \Sigmab^{\frac12} (\I - \P_{-k})^2 \Sigmab^{\frac12} =
-    \|\Sigmab^{\frac12}-\Sigmab^{\frac12}\X_{-k}^\dagger\X_{-k}\|_F^2\geq
-\sum_{i\geq k}\lambda_i (\Sigmab)\geq r-k
-\end{equation}
-where we used the fact that $\rank(\X_{-k}^\dagger \X_{-k}) \le \rank(\X_{-k}) \le k-1$ and arranged the eigenvalues $ 1 = \lambda_1(\Sigmab) \ge \ldots \ge \lambda_n(\Sigmab)$ in a non-increasing order.
-As a consequence, we also have
-\begin{equation}\label{eq:bound-norm-P-Sigma}
-  \frac{\tr (\I - \P_{-k}) \Sigmab}{ \|  (\I - \P_{-k}) \Sigmab \| } \ge \tr (\I - \P_{-k}) \Sigmab \geq r-k.
-\end{equation}
-
-\bigskip
-
-For the first step, we have, with Lemma~\ref{l:concentration-sub-gaussian}~and~\eqref{eq:bound-norm-P-Sigma} that
-\begin{align*}
-  \Pr (\neg E_i) &\le \Pr \left\{ \lvert \x_i^\top (\I - \P_{-i}) \x_i - \tr \Sigmab (\I - \P_{-i})  \rvert
-    \geq \frac13 \tr \Sigmab (\I - \P_{-i}) \right\} \\ 
-    &\le 2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} }. % \le 2 \ee^{ - \frac{ r -n }{c r} (r-n) } = 2 \ee^{c_\rho (r - n)}
-\end{align*}
-so that with the union bound we obtain
-\begin{equation}\label{eq:union-bound-neg-E}
-  \Pr (\neg E) \le 2k \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{k}{(r-k)^2} \cdot 2 (r-k)^2 \ee^{ - \min \left\{ \frac{ r -k }{9 C^2 K^4}, \frac{\sqrt{r -k} }{3 C K^2} \right\} } \le \frac{C_\rho}{r - k}
-\end{equation}
-where we used the fact that, for $\alpha >0$, $x^2 \ee^{-\alpha x} \le \frac{4 \ee^{-2}}{\alpha^2}$ and  $x^4 \ee^{-\alpha x} \le \frac{256 \ee^{-4}}{\alpha^4}$ on $x > 0$.
-Also, denote $c_\rho = \frac{r - k}r = \frac{\rho - 1}{\rho} > 0$, we have
-\begin{equation}\label{eq:proba-E}
-  \Pr(\neg E) \le \frac{C_\rho}{r - k} = \frac{C_\rho}{c_\rho r} = \frac{C'_\rho}{r} 
-\end{equation}
-for some $C'_\rho > 0$ that depends on $\rho = r/k > 1$ and the sub-gaussian norm $K$.
-
-\medskip
-
-At this point, note that, conditioned on the event $E$, we have for $i \in \{1, \ldots, k \}$
-\begin{equation}
-    \frac12 \frac1{\tr (\I - \P_{-i}) \Sigmab} \le \frac1{\x_i^\top (\I - \P_{-i}) \x_i} \le \frac32 \frac1{\tr (\I - \P_{-i}) \Sigmab}, %\quad  \frac1{2r} \le \frac1{\x_i^\top \x_i} \le \frac3{2r}.
-%\frac1{\x_i^\top \x_i} \le \frac2{\tr \Sigmab} = \frac2{r}
-\end{equation}
-Also, with \eqref{eq:proba-E} and the fact that $\| \P \| \le 1$, we have $ \| \T_4 \| \le
-\frac{C_\rho}{r}$ for some $C_\rho > 0$ that depends on $\rho$ and $K$. To handle non-symmetric matrix $\T_2$, note that $\T_2 + \T_2^\top$ is symmetric and
-\begin{equation}
-  - \E[ (\I - \P_{-k}) \cdot \one_{\neg E}] - \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}] \preceq \T_2 + \T_2^\top \preceq \E[ (\I - \P_{-k}) \cdot \one_{\neg E}] + \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}]
-\end{equation}
-with $-(\A \A^\top + \B \B^\top) \preceq \A \B^\top + \B \A^\top \preceq \A \A^\top + \B \B^\top$.
-To obtain an upper bound for operator norm of $\E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}]$, note that
-\begin{align*}
-  \| \E[ (\x_k^\top \x_k) \x_k\x_k^\top \cdot \one_{\neg E}] \| &\le \E[ (\x_k^\top \x_k)^2 \cdot \one_{\neg E}] = \int_0^\infty \Pr ( \x^\top \x \cdot \one_{\neg E} \ge \sqrt t ) dt \\ 
-  &\le \int_0^X \Pr ( \x^\top \x \cdot \one_{\neg E} \ge \sqrt t ) dt + \int_X^\infty \Pr ( \x^\top \x \ge \sqrt t ) dt  \\ 
-  &\le X \cdot \Pr(\neg E) + \int_X^\infty \ee^{- \min \left\{ \frac{t}{C^2 K^4 r}, \frac{\sqrt t}{CK^2 \sqrt r} \right\} } dt \le \frac{c_\rho}r
-\end{align*}
-where we recall $\E[\x^\top \x] = \tr \Sigmab = r$ and take $X \ge C^2 K^4 r$, the third line follows from the proof of Lemma~\ref{l:concentration-sub-gaussian} and the forth line from the same argument as in \eqref{eq:union-bound-neg-E}.
-Moreover, since $\| \T_2 \| \le \| \T_2 + \T_2^\top \|$ (see for example \cite[Proposition~5.11]{serre2010matrices}), we conclude that $\| \T_2 \| \le \frac{C_\rho}r$.
-
-And it thus remains to handle the terms $\T_1$ and $\T_3$ to obtain a bound on $\| \I - \E[\P_\perp] \bar \P_\perp^{-1} \| $.
-
-To bound $\T_3$, with $\P - \P_{-k} = \frac{(\I - \P_{-k})\x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}$ in Lemma~\ref{l:rank-one-update}, we have
-\begin{align*}
-\|\T_3\| &\leq  \bigg\|
-  \E\bigg[\frac{ (\I - \P_{-k}) \x_k\x_k^\top (\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k}
-  \cdot \one_E \bigg]\bigg\|  + \| \E[ (\P - \P_{-k}) \cdot \one_{\neg E} ] \| \\ 
-  %%%
-  &\le \frac32 \E \left[ \frac1{ \tr (\I - \P_{-k} ) \Sigmab } \right] + \frac{c_\rho}{r - k} \leq \frac{C_\rho}{r-k} = \frac{C'_\rho}{r}
-\end{align*}
-where we used the fact that $\tr\,(\I - \P_{-k}) \Sigmab \ge r -k $ from \eqref{eq:bound-trace-P-Sigma} and recall $\rho \equiv r/k > 1$.
-
-For $\T_1$ we write
-\begin{align*}
-  \|\T_1\|    &\leq\E\bigg[\|\I - \P_{-k}\|\cdot \Big\|\E\Big[|\bar s-\hat s|\cdot
-    \frac{\x_k\x_k^\top}{\x_k^\top( \I - \P_{-k}) \x_k} \cdot \one_E\mid\P_{-k}\Big]
-    \Big\|\bigg]
-  \\
-          &\leq \frac32 \frac1{r-k} \cdot \E\bigg[\sup_{\| \v \| = 1}\E\Big[
-            |\bar s-\hat s|\cdot \v^\top\x_k\x_k^\top\v\cdot\one_E\mid
-            \P_{-k}\Big]\bigg]
-  \\
-  &\leq \frac{C_\rho}{r}\cdot \E\bigg[\underbrace{\sqrt{\E\big[(\bar s-\hat
-    s)^2\cdot \one_E \mid\P_{-k}\big]}}_{ T_{1,1} }\cdot
-    \underbrace{\sup_{ \| \v \| = 1 }\sqrt{\E\big[(\v^\top\x_k)^4\big]}}_{ T_{1,2} }\bigg]
-\end{align*}
-where we used Jensen's inequality for the first inequality, the relation in \eqref{eq:bound-trace-P-Sigma} for the second inequality, and Cauchy–Schwarz for the third inequality.
-
-We first bound $T_{1,2}$ by definition of sub-gaussian random
-vectors. We have for $\x_k$ a $K$-sub-gaussian and $\| \v \| = 1$
-that, $\v^\top \x_k$ is a sub-gaussian random variable with $\|\v^\top
-\a \|_{\psi_2} \leq K$. As such, $T_{1,2} \le C K^2$ for some absolute
-constant $C > 0$, see for example
-\cite[Section~2.5.2]{vershynin2018high}.
-
-For $T_{1,1}$ we have
-\[
-  \sqrt{ \E [(\bar s-\hat s)^2 \cdot\one_E \mid\P_{-k}] } = \sqrt{ (\bar s - s)^2 + \E[(s - \hat s)^2 \cdot\one_E] }
-\]
-where we denote $s=\E[\hat s]=\tr\,\E[\I - \P_{-k}]\Sigmab$. Note that
-% \begin{align*}
-%   \E\big[(s - \hat s)^2 \cdot\one_E \big] &=
-% \E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2 \cdot\one_E \big] +
-%   \E\big[(\tr\,\P_{-k}\Sigmab-\x_k^\top\P_{-k}\x_k)^2 \cdot\one_E \big]\\
-%   &\leq C\Big(n + \E\big[\tr\,(\P_{-k}\Sigmab)^2 \cdot\one_E \big]\Big)\leq
-%     C\big(n + \tr\,\E[\P_{-k}]\Sigmab \cdot\one_E\big)\\
-%   &=C(n+s \cdot\one_E)\leq C\big(n+\bar s + |s- \bar s| \cdot\one_E\big)
-% \end{align*}
-\begin{align*}
-  \E\big[(s - \hat s)^2 \big] &=
-\E\big[\big(\tr\,\Sigmab(\P_{-k} - \E[\P_{-k}])\big)^2  \big] +
-  \E\big[(\tr\,(\I - \P_{-k})\Sigmab-\x_k^\top(\I - \P_{-k})\x_k)^2  \big]\\
-  &\leq C_1 k +  C_2 \E\big[\tr\,(\Sigmab - \P_{-k}\Sigmab)^2 \big] \le C(k+s)\leq C\big(k+\bar s + |s- \bar s| \big)
-\end{align*}
-where we used Lemma~\ref{l:trace} and Lemma~\ref{l:concentration-sub-gaussian}. Recall that $\bar s = \tr \bar \P_\perp \Sigmab \le \tr \Sigmab = r$ and $k < r$, we have
-\begin{equation}\label{eq:bound-T11}
-  T_{1,1} \le \sqrt{ (\bar s - s)^2 + C (|\bar s - s| + 2 r) }
-\end{equation}
-
-% Moreover, using the fact that
-% $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$, we obtain that:
-% \begin{align*}
-%   |\bar s - s| \cdot\one_E &= |\tr (\bar \P_\perp - \E[\P_{-k}] )\Sigmab| \cdot\one_E \le |\tr (\bar \P_\perp - \E[\P] )\Sigmab| \cdot\one_E + |\tr \E[\P - \P_{-k}] \Sigmab| \\
-%   &= \big|\tr\,(\I-\E[\P]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab\big| \cdot\one_E + \tr\,\E \left[ \frac{\P_{-k}\x_k\x_k^\top\P_{-k}}{\x_k^\top \P_{-k} \x_k} \right] \Sigmab \\
-%   %&= \gamma \big|\tr\,( \T_1 + \T_2 + \T_3 )\bar\P_\perp\Sigmab\big| \cdot\one_E + 1 \\
-%   &\leq \gamma\,\E\bigg[|\bar s-\hat s|\cdot
-%     \frac{\tr\,\P_{-k}\x_k\x_k^\top\bar\P_\perp\Sigmab}{\tr\,\P_{-k}\x_k\x_k^\top} \cdot\one_E\bigg]
-%     + \gamma\,
-%     \E\bigg[\frac{\tr\,\P_{-k}\x_k\x_k^\top\P_{-k}\Sigmab\bar\P_\perp\Sigmab}
-%     {\tr\,\P_{-k}\x_k\x_k^\top}\bigg] + 1\\
-%   &\leq \frac{\gamma}{\gamma+1}\Big(\E\left[|\bar s-\hat
-%     s| \cdot \frac{\x_k^\top \x_k}{\x_k^\top \P_{-k} \x_k} \cdot\one_E \right] + 1\Big)\leq \frac{\gamma}{\gamma+1}\Big(|\bar
-%     s-s| \frac{3\rho}{\rho - 1}  + \E\big[|s-\hat
-%     s|\big] + 1\Big)\\
-%   &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
-%      c\sqrt{|\bar s- s|} + c\sqrt{n+\bar s} + 1\Big).
-% \end{align*}
-
-It remains to bound $|\bar s - s|$. Note that $\P=(\X^\top\X)^\dagger\X^\top\X = \X^\top\X
-(\X^\top\X)^\dagger$ and is symmetric, so
-\begin{align*}
-  &\I - \E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar\P_\perp^{-1} \E[\P_\perp]
-  = 2\E[\P] - \E[\gamma\P_\perp\Sigmab] - \E[\gamma\Sigmab\P_\perp]\\
-  &= \sum_{i=1}^k\E\big[(\X^\top\X)^\dagger\x_i\x_i^\top + \x_i\x_i^\top (\X^\top\X)^\dagger \big] - \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp])\\
-  %&=k\,\E\bigg[\frac{\P_{-k}\x_k\x_k^\top + \x_k\x_k^\top \P_{-k}}{\x_k^\top\P_{-k}\x_k}\bigg] - \gamma (\E[\P]\Sigmab + \Sigmab \E[\P])\\
-  &=\gamma\,
-    \E\bigg[\bar s\cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})} {\x_k^\top(\I - \P_{-k}) \x_k}\bigg]
-    -\gamma\,\E\bigg[\hat s
-    \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg] \\ 
-  & +\gamma\,(\E[(\I - \P_{-k})\Sigmab] + \E[\Sigmab ( \I -\P_{-k}) ])- \gamma (\E[\P_\perp]\Sigmab + \Sigmab \E[\P_\perp]) \\
-  &=\gamma\,\E\bigg[(\bar s - \hat s)
-    \cdot\frac{(\I-\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I-\P_{-k})}{\x_k^\top(\I-\P_{-k})\x_k}\bigg]
-+    \gamma (\E[\P-\P_{-k}]\Sigmab + \Sigmab \E[\P-\P_{-k}]). 
-\end{align*}
-%where we used the fact that for $\mathbf{a}, \mathbf{b} \in \mathbb R^d$, we have $\mathbf{a} \mathbf{b}^\top + \mathbf{b} \mathbf{a}^\top \preceq \mathbf{a} \mathbf{a}^\top + \mathbf{b} \mathbf{b}^\top$ (which unfolds from $(\mathbf{a} - \mathbf{b}) (\mathbf{a} - \mathbf{b})^\top \succeq 0$).
-Moreover, using the fact that $\bar\P_\perp\Sigmab\preceq\frac1{\gamma+1}\I$ and $\bar \P_\perp \Sigmab = \Sigmab \bar \P_\perp$, we obtain that
-\begin{align*}
-  &|\bar s - s|  = |\tr (\bar \P_\perp - \E[\I - \P_{-k}] )\Sigmab|   \le |\tr (\bar \P_\perp - \E[\P_\perp] )\Sigmab|   + |\tr \E[\P - \P_{-k}] \Sigmab| \\
-  &= \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1})\bar\P_\perp\Sigmab + \tr\,\bar\P_\perp (\I - \bar \P_\perp^{-1} \E[\P_\perp]) \Sigmab \big|   + \tr\,\E \left[ \frac{(\I - \P_{-k})\x_k\x_k^\top(\I - \P_{-k})}{\x_k^\top (\I - \P_{-k}) \x_k} \right] \Sigmab \\
-  &\leq \frac12 \big|\tr\,(\I-\E[\P_\perp]\bar\P_\perp^{-1} + \I - \bar \P_\perp^{-1} \E[\P_\perp])\bar\P_\perp\Sigmab \big|   + 1 \\
-  &\leq \frac{\gamma}2 \,\E\bigg[|\bar s-\hat s|\cdot
-    \frac{\tr\,( (\I -\P_{-k})\x_k\x_k^\top + \x_k\x_k^\top (\I -\P_{-k})) \bar\P_\perp\Sigmab}{\tr\,(\I -\P_{-k})\x_k\x_k^\top}  \bigg] \\
-  &+ \gamma\, \E\bigg[\frac{\tr\,(\I-\P_{-k})\x_k\x_k^\top(\I-\P_{-k})\bar\P_\perp\Sigmab}
-    {\tr\,(\I-\P_{-k})\x_k\x_k^\top}\bigg] + 1\\
-  &\leq \frac{\gamma}{\gamma+1}\Bigg(\E\bigg[|\bar s-\hat
-    s| \cdot \frac{\x_k^\top (\I -\P_{-k}) \x_k}{\x_k^\top (\I -\P_{-k}) \x_k}  \bigg] + 1\Bigg) + 1\leq \frac{\gamma}{\gamma+1}\Big(|\bar
-    s-s| + \E\big[|s-\hat s|\big] + 1\Big) + 1 \\
-  &\leq \frac{\gamma}{\gamma+1}\Big(|\bar s- s| +
-     C\sqrt{|\bar s- s|} + C\sqrt{2 r} + 1\Big) + 1.
-\end{align*}
-
-Solving for $|\bar s-s|$, we deduce that
-\begin{align*}
-  |\bar s-s| \leq C_1 \sqrt{r} + C_2,
-  %\gamma(c\sqrt{k+\bar s}+2) + c^2\gamma^2\leq \gamma (c+1)\sqrt{k+\bar s} + c^2\gamma^2.
-\end{align*}
-so plugging back to \eqref{eq:bound-T11} we get $T_{1,1} \le C
-\sqrt r$ and $\| \T_1\| \le \frac{C_\rho}{\sqrt r} $, thus completing
-the proof.
-\section{Convergence analysis of randomized iterative methods}
-\label{a:newton}
-
-Here, we discuss how our surrogate expressions for the expected
-residual projection can be used to perform convergence analysis for
-several randomized iterative optimization methods discussed in
-Section~\ref{s:newton}.
-
-\subsection{Generalized Kaczmarz method}
-Generalized Kaczmarz \cite{generalized-kaczmarz} is an iterative
-method for solving an $m\times n$ linear system
-$\A\x=\b$, which uses a $k\times m$ sketching matrix $\S_t$ to reduce
-the linear system and update an iterate $\x^t$ as follows:
-  \begin{align*}
-    \x^{t+1} = \argmin_\x\|\x-\x^t\|^2\quad\textnormal{subject to}\quad\S_t\A\x=\S_t\b.
-  \end{align*}
-Assume that $\x^*$ is the unique solution to the linear system $\A\x=\b$. In Theorems 4.1 and 4.6, \cite{generalized-kaczmarz} show that the
-expected trajectory of the generalized Kaczmarz iterates, as
-they converge to $\x^*$, is controlled
-by the projection matrix $\P=(\S_t\A)^\dagger\S_t\A$ as follows:
-\begin{align*}
-  \text{(\cite{generalized-kaczmarz}, Theorem 4.1)}\qquad\quad \ \ \,
-  \E[\x^{t+1}-\x^*]
-  &= \big(\I-\E[\P]\big)\,\E[\x^t-\x^*],\\
-  \text{(\cite{generalized-kaczmarz}, Theorem 4.6)}\qquad
-  \E\big[\|\x^{t+1}-\x^*\|^2\big]
-&\leq (1-\kappa)\,\E\big[\|\x^t-\x^*\|^2\big],
-\text{ where }\kappa=\lambda_{\min}\big(\E[\P]\big).
-\end{align*}
-Both of these results depend on the expected projection
-$\E[\P]$. The first one describes the expected trajectory of the
-iterate, whereas the second one gives the worst-case convergence
-rate in terms of the so-called \emph{stochastic condition number}
-$\kappa$. We next demonstrate how Theorem \ref{t:main} can be used  in 
-combination with the above results to
-obtain convergence analysis for generalized Kaczmarz which is
-formulated in terms of the spectral properties of $\A$. This includes
-precise expressions for both the expected trajectory and $\kappa$. The following
-result is a more detailed version of Corollary \ref{c:kaczmarz} from
-Section \ref{s:newton}.
-\begin{corollary}\label{c:kaczmarz2}
-Let $\sigma_i$ denote the singular values of $\A$, and let $k$ denote
-the size of sketch $\S_t$. Define:
-\begin{align*}
-\Delta_t=\x^t-\x^*\quad\text{and}\quad  \bar\Delta_{t+1} =
-  (\gamma\A^\top\A+\I)^{-1}\E[\Delta_t]\quad\text{s.t.}\quad
-  \sum_i\frac{\gamma\sigma_i^2}{\gamma\sigma_i^2+1}=k.
-\end{align*}
-  Suppose that $\S_t$ has i.i.d.~mean-zero sub-gaussian entries and let
-  $r=\|\A\|_F^2/\|\A\|^2$ be the stable rank of $\A$. 
-Assume that $\rho = r/k$ is a constant larger than $1$. Then, the
-expected trajectory satisfies:
-\begin{align}
-  \big\|\E[\Delta_{t+1}] - \bar\Delta_{t+1}\big\|
-  \leq \epsilon\cdot \|\bar\Delta_{t+1}\|,
-  \quad\text{for}\quad\epsilon=O\big(\tfrac1{\sqrt
-  r}\big).\label{eq:trajectory}
-\end{align}
-Moreover, we obtain the following worst-case convergence guarantee:
-\begin{align}
-  \E\big[\|\Delta_{t+1}\|^2\big]
-  \leq \big(1-
-  (\bar\kappa-\epsilon)\big)\,\E\big[\|\Delta_t\|^2\big],
-  \quad\text{where}\quad
-  \bar\kappa = \frac{\sigma_{\min}^2}{\sigma_{\min}^2+1/\gamma}.\label{eq:worst-case}
-\end{align}
-\end{corollary}
-\begin{remark}
-Our worst-case convergence guarantee \eqref{eq:worst-case}
-  requires the matrix $\A$ to be sufficiently well-conditioned so
-  that $\bar\kappa-\epsilon>0$. However, we believe that our surrogate expression
-  $\bar\kappa$ for the stochastic condition number is far more
-  accurate than suggested by the current analysis. 
-\end{remark}
-
-\subsection{Randomized Subspace Newton}
-Randomized Subspace Newton (RSN, \cite{Gower2019})  is a randomized
-Newton-type method  for minimizing a
-smooth, convex and twice differentiable function
-$f:\R^d\times \R$. The iterative 
-update for this algorithm is defined as follows:
-\begin{align*}
-  \x^{t+1} = \x^t - \frac1L\S_t^\top(\S_t\H(\x^t)\S_t^\top)^\dagger\S_t\g(\x^t),
-\end{align*}
-where $\H(\x^t)$ and $\g(\x^t)$ are the Hessian and gradient of $f$ at
-$\x^t$, respectively, whereas $\S_t$ is a $k\times d$ sketching
-matrix (with $k\ll d$) which is refreshed at every iteration. Here,
-$L$ denotes the \emph{relative smoothness} constant defined by
-\cite{Gower2019} in Assumption 1, which also defines relative strong
-convexity, denoted by $\mu$. In Theorem 2, they prove the
-following convergence guarantee for RSN:
-\begin{align*}
-  \E[f(\x^t)] - f(\x^*) \leq \Big(1-\kappa\frac{\mu}{L}\Big)^t(f(\x^0)-f(\x^*)),
-\end{align*}
-where $\kappa =\min_{\x}\kappa(\x)$ and
-$\kappa(\x)=\lambda_{\min}^+(\E[\P(\x)])$ is the smallest positive
-eigenvalue of the expectation of the projection matrix
-$\P(\x)=
-\H^{\frac12}(\x)\S_t^\top(\S_t\H(\x)\S_t^\top)^\dagger\S_t\H^{\frac12}(\x)$. Our
-results lead to the following surrogate expression for this expected
-projection when the sketch is sub-gaussian:
-\begin{align*}
-  \E[\P(\x)] \simeq \H(\x)\big(\H(\x) +
-  \tfrac1{\gamma(\x)}\I\big)^{-1}\quad\text{for}\quad\gamma(\x)>0
-  \quad\text{s.t.}\quad\tr\,\H(\x)\big(\H(\x) +
-  \tfrac1{\gamma(\x)}\I\big)^{-1} = k.
-\end{align*}
-Thus, the condition number $\kappa$ of RSN can be estimated using the
-following surrogate expression:
-\begin{align*}
-\kappa\simeq \bar\kappa :=
-  \min_\x\frac{\lambda_{\min}^+(\H(\x))}{\lambda_{\min}^+(\H(\x)) +
-  1/\gamma(\x)}. 
-\end{align*}
-Just as in Corollary \ref{c:kaczmarz2}, an approximation
-of the form $|\bar\kappa-\kappa|\leq\epsilon$ can be shown
-from Theorem \ref{t:main}.
-\begin{corollary}\label{c:rsn}
-  Suppose that sketch $\S_t$ has size $k$ and i.i.d.~mean-zero sub-gaussian entries. Let
-  $r=\min_\x\tr\,\H(\x)/\|\H(\x)\|$ be the (minimum) stable rank of
-  the (square root) Hessian and assume that $\rho = r/k$ is a constant larger than
-  $1$. Then,
-  \begin{align*}
-    |\kappa-\bar\kappa| \leq O\big(\tfrac1{\sqrt r}\big).
-  \end{align*}
-\end{corollary}
-
-\subsection{Jacobian Sketching}
-Jacobian Sketching (JacSketch, \cite{jacsketch}) defines an $n\times n$ positive semi-definite weight matrix
-$\W$, and combines it with an $k\times n$ sketching matrix $\S$
-(which is refreshed at every iteration of the algorithm), to
-implicitly construct the following projection matrix:
-\begin{align*}
-  \Pi_\S = \S^\top(\S\W\S^\top)^\dagger\S\W,
-\end{align*}
-which is used to sketch the Jacobian at the current iterate (for the
-complete method, we refer to their Algorithm 1).
-The convergence rate guarantee given in their Theorem 3.6 for
-JacSketch is given in terms of the Lyapunov function:
-\begin{align*}
- \Psi^t = \|\x^t-\x^*\|^2 + \frac{\alpha}{2\mathcal{L}_2}\|\mathbf{J}^t-\nabla F(\x^*)\|_{\W^{-1}}^2,
-\end{align*}
-where $\alpha$ is the step size used by the algorithm. Under
-appropriate choice of the step-size, Theorem~3.6 states that:
-\begin{align*}
-\E[\Psi^t] \leq \bigg(1 - \mu\,\min\Big\{\frac{1}{4\mathcal{L}_1},\frac{\kappa}{4\mathcal{L}_2\rho/n^2+\mu}\Big\}\bigg)^t\cdot\Psi^0,
-\end{align*}
-where $\kappa=\lambda_{\min}(\E[\Pi_\S])$ is the 
-\emph{stochastic condition number} analogous to the one defined for
-the Generalized Kaczmarz method, $n$ is the data size and
-parameters $\rho$, $\mathcal{L}_1$, $\mathcal{L}_2$ and $\mu$ are
-problem dependent constants defined in Theorem~3.6. Similarly as
-before, we can use our surrogate expressions 
-for the expected residual projection to obtain a precise estimate for
-the stochastic condition number 
-$\kappa$ under sub-gaussian sketching:
-\begin{align*}
-  \kappa\simeq\bar\kappa :=\frac{\lambda_{\min}(\W)}{\lambda_{\min}(\W) +
-  1/\gamma}
-  \quad\text{for}\quad\gamma>0\quad\text{s.t.}\quad\tr\,\W(\W +
-  \tfrac1\gamma\I)^{-1} = k.
-\end{align*}
-\begin{corollary}\label{c:jacsketch}
-  Suppose $\S_t$ has size $k$ and i.i.d.~mean-zero sub-gaussian entries. Let
-  $r=\tr\,\W/\|\W\|$ be the stable rank of $\W^{\frac12}$ and assume
-  that $\rho = r/k$ is a constant larger than 
-  $1$. Then,
-  \begin{align*}
-    |\kappa-\bar\kappa| \leq O\big(\tfrac1{\sqrt r}\big).
-  \end{align*}
-\end{corollary}
-
-\subsection{Omitted proofs}
-\begin{proofof}{Corollary}{\ref{c:kaczmarz2}}
-Using Theorem \ref{t:main}, for $\bar\P_{\perp}$ as defined in \eqref{eq:surrogate}, we have
-  \begin{align*}
-    (1-\epsilon)\bar\P_{\perp}\preceq \I-\E[\P] =
-    \E[\P_{\perp}]\preceq(1+\epsilon)\bar\P_{\perp},
-    \quad\text{where}\quad\epsilon=O\big(\tfrac 1{\sqrt r}\big).
-  \end{align*}
-In particular, this implies that
-$\|\bar\P_{\perp}^{-\frac12}(\E[\P_{\perp}]-\bar\P_{\perp})\bar\P_{\perp}^{-\frac12}\|\leq\epsilon$.
-Moreover, in the proof of Theorem \ref{t:main-tech} we showed that
-  $\frac{\rho-1}\rho\I\preceq\bar\P_{\perp}\preceq\I$, see
-  \eqref{eq:bound-bar-gamma-P}, so it follows that:
-  \begin{align*}
-    \bar\P_{\perp}^{-1}(\E[\P_{\perp}]-\bar\P_{\perp})^2\bar\P_{\perp}^{-1}\preceq
-    \frac{\rho}{\rho-1}\big(\bar\P_{\perp}^{-\frac12}(\E[\P_{\perp}]-\bar\P_{\perp})\bar\P_{\perp}^{-\frac12}\big)^2
-    \preceq     \frac{\rho}{\rho-1}\,\epsilon^2\cdot \I,
-  \end{align*}
-  where note that  $\frac{\rho}{\rho-1}\,\epsilon^2=O(1/r)$, since
-  $\rho$ is treated as a constant. Thus we conclude that:
-  \begin{align*}
-    \|\E[\Delta_{t+1}]-\bar\Delta_{t+1}\|^2
-    &=\E[\Delta_t]^\top(\E[\P_{\perp}]-\bar\P_{\perp})^2\E[\Delta_t]
-    \\
-    &\leq O(1/r)\cdot \E[\Delta_t]^\top\bar\P_{\perp}^2\E[\Delta_t] =
-      O(1/r)\cdot \|\bar\Delta_{t+1}\|^2,
-  \end{align*}
-  which completes the proof of \eqref{eq:trajectory}. To show
-  \eqref{eq:worst-case}, it suffices to observe that
-  \begin{align*}
-    \lambda_{\min}(\E[\P])
-    = 1- \lambda_{\max}(\E[\P_{\perp}])
-    \geq 1-(1+\epsilon)\lambda_{\max}(\bar\P_{\perp})
-    \geq \lambda_{\min}(\I-\bar\P_{\perp}) - \epsilon,
-  \end{align*}
-  which completes the proof since
-  $\I-\bar\P_{\perp}=\gamma\A^\top\A(\gamma\A^\top\A+\I)^{-1}$. 
-\end{proofof}
-
-Corollaries \ref{c:rsn} and \ref{c:jacsketch} follow analogously from
-Theorem \ref{t:main}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.47\textwidth]{bodyfat-supp}\nobreak\includegraphics[width=.47\textwidth]{housing-supp}
-  \includegraphics[width=.47\textwidth]{mg-supp}\nobreak\includegraphics[width=.47\textwidth]{triazines-supp} 
-  \caption{Theoretical predictions versus approximation error for the
-    sketched Nystr\"om with the RBF kernel, using Gaussian and
-    Rademacher sketches (spectral decay shown at the bottom).}\label{f:nystrom2}
-\end{figure}
-
-\section{Additional empirical results}
-\label{a:experiments}
-
-We complement the results of Section \ref{s:experiments} with
-empirical results on four additional libsvm datasets \cite{libsvm} (bringing
-the total number of benchmark datasets to eight), which further
-establish the accuracy of our surrogate expressions for the low-rank approximation
-error. Similary as in Figure \ref{f:nystrom}, we use the sketched
-Nystr\"om method \cite{revisiting-nystrom} with the RBF kernel
-$k(\a_i,\a_j)=\exp(-\|\a_i-\a_j\|^2/(2\sigma^2))$, for several values
-of the parameter $\sigma$. The values of $\sigma$ were chosen so as to
-demonstrate the effectiveness of our theoretical predictions both when the
-stable rank is moderately large and when it is very small.
-
-In Figure \ref{f:nystrom2} we show the results for both Gaussian and
-Rademacher sketches. These results reinforce the conclusions we made in
-Section \ref{s:experiments}: our theoretical estimates are very
-accurate in all cases, for both sketching methods, and even when the
-stable rank is close to 1 (a regime that is not supported by the
-current theory).
-
 \end{document}

### 2022-11-23 ftvi/ftvi_icml/ftvi_icml.tex
Updates from Overleaf
Working compile

@@
 % \documentclass[nohyperref]{article}
-\documentclass[thesis.tex]{subfiles}
+\documentclass[../../thesis.tex]{subfiles}
 
@@
 % \begin{abstract}
-    While fat-tailed densities commonly arise as posterior and marginal distributions in
-    robust models and scale mixtures, they present challenges when Gaussian-based 
-    variational inference fails to capture tail decay accurately. 
-    We first improve previous theory on tails of Lipschitz flows 
+
+\chapter{Fat-tailed variational inference}
+\label{ch:ftvi}
+
+In some applications of probabilistic modeling,
+researchers may desire to explicitly model large deviation ``tail''  or ``black-swan''
+events. While fat-tailed densities commonly arise as posterior and marginal distributions in
+robust models and scale mixtures, they present challenges when
+prior methods (including \cref{ch:lic})
+fails to capture tail decay accurately.
+In this chapter, we first improve previous theory on tails
+of Lipschitz flows
 %\citep{jaini2020tails} 
@@
 is competitive with prior work while also exhibiting appropriate tail-anisotropy.
+Parts of this chapter were first presented in \fullcite{ftvi}.
 % \end{abstract}
@@
   \centering
-  \includegraphics{../Figures/pancake.pdf}
-    \vspace{-5mm}
+  \includegraphics{ftvi/Figures/pancake.pdf}
   \caption{
@@
     which erroneously imposes power-law tails with the same rate of decay along both the $x$ and $y$ axes.
-    \vspace{-5mm}
   }
@@
 I would say that we get results for a ragne of cases, both for isotropic and anisotropic cases, thereby improving prior work that considered Gaussian things as well as prior work from \cite{jaini2020tails} that considered isotropic HT.}
-\vspace{-3mm}
 \subsection*{Related Work}
-\vspace{-1mm}
 \paragraph{Fat-Tails in Variational Inference.}
@@
 
-\vspace{-1mm}
 \paragraph{Flow-Based Methods.}
@@
 
-\vspace{-2mm}
 \section{Flow-Based Methods for Fat-Tailed Variational Inference}
@@
 
-\vspace{-2mm}
 \subsection{Flow-Based VI Methods}
@@
   \text{ELBO}(\phi)
-  &= \int q_\phi(x) \log \frac{\bar\pi(x)}{q_\phi(x)} dx\\
+   & = \int q_\phi(x) \log \frac{\bar\pi(x)}{q_\phi(x)} dx                     \\
    & \approx \frac{1}{n} \sum_{i=1}^n \log \frac{\bar\pi(x_i)}{q_\phi(x_i)},\;
@@
   \centering
+  {\tiny
     \begin{tabular}{ccc}
@@
     \end{tabular}
+  }
   \caption{Some popular / recently developed flows, the autoregressive transform used in the flow (if applicable),
@@
     in \citet{jaini2020tails}. $M(\cdot)$ denotes monotonic rational quadratic splines \citep{durkan2019neural}.
-  \vspace{-5mm}
   }
@@
   ADVI (with normalizing flows) comprise the variational family
-  $\cQ_\text{ADVI}~\coloneqq~\{
+  \[
+    \cQ_\text{ADVI}~\coloneqq~\{
     (f~\circ~\Phi_\text{Flow})_\ast \mu
-    \}$, 
+    \}
+  \]
   where $\mu = \text{Normal}(0_d, I_d)$,
@@
 
-\vspace{-2mm}
 \subsection{Fat-Tailed Variational Inference}
 
-\vspace{-1mm}
 Fat-tailed variational inference (FTVI)
@@
   For each $\alpha,p > 0$, we let
-    \vspace{-1mm}
   \begin{itemize}[leftmargin=*]
     \item $\mathcal{E}_\alpha^p$ denote the set of \emph{exponential-type} random variables $X$ with $\mathbb{P}(|X| \geq x) = \Theta(e^{-\alpha x^p})$;
-    \vspace{-1mm}
     \item $\mathcal{L}_\alpha^p$ denote the set of \emph{logarithmic-type} random variables $X$ with $\mathbb{P}(|X| \geq x) = \Theta(e^{-\alpha(\log x)^p})$.
   \end{itemize}
-    \vspace{-1mm}
   % A random variable $X$ is of %\textbf{exponential-type}, denoted $X \in \cE_\alpha^p$, w
@@
 
-\vspace{-2mm}
 \section{Tail Behavior of Lipschitz Flows}
 
-\vspace{-1mm}
-This section states our main theoretical contributions; proofs are deferred to \Cref{sec:proofs}.
+This section contains our main theoretical contributions and proofs.
 We sharpen previous impossibility results approximating fat-tailed targets
@@
 
-Most of our results are developed within the context of Lipschitz-continuous transport maps $f_\theta$.
+As an initial but nevertheless crucial result, we first bound the tail
+parameters of the sum of two power law random variables.
+
+\begin{lemma}
+  \label[lemma]{lem:sum-rule}
+  Suppose $X \in \cL^1_\alpha$ and $Y \in \cL^1_\beta$.
+  Then $X + Y \in \cL^1_{\min\{\alpha,\beta\}}$.
+\end{lemma}
+
+\begin{proof}
+  First, let $\gamma=\min\{\alpha,\beta\}$.
+It will suffice to show that (I) $\mathbb{P}(|X+Y|\geq r)=\mathcal{O}(r^{-\gamma})$, and (II) $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\gamma})$.
+Since $(X,Y)\mapsto|X+Y|$ is a 1-Lipschitz function on $\mathbb{R}^{2}$ and $\mathbb{P}(|X|\geq r)+\mathbb{P}(|Y|\geq r)=\mathcal{O}(r^{-\gamma})$, (I) follows directly from the hypotheses and Proposition 1.11 of \citet{ledoux2001concentration}.
+To show (II), note that for any $M>0$, conditioning on the event $|Y|\leq M$,\[
+    \mathbb{P}\left(\left|X\right|+|Y|\geq r\,\vert\,|Y|\leq M\right)\geq\mathbb{P}\left(\left|X\right|\geq r-M\right).
+  \]
+  Therefore, by taking $M$ to be sufficiently large so that $\mathbb{P}(|Y|\leq M)\geq\frac{1}{2}$,
+  \begin{align*}
+    \mathbb{P}\left(|X+Y|\geq r\right) & \geq\mathbb{P}\left(|X|+|Y|\geq r\right)                                                                         \\
+                                       & \geq\mathbb{P}\left(\left|X\right|+|Y|\geq r\,\vert\,|Y|\leq M\right)\mathbb{P}\left(\left|Y\right|\leq M\right) \\
+                                       & \geq\frac{1}{2}\mathbb{P}\left(\left|X\right|\geq r-M\right)=\Theta(r^{-\alpha}).
+  \end{align*}
+  The same process with $X$ and $Y$ reversed implies $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\beta})$ as well. Both (II) and the claim follow.
+\end{proof}
+
+Most of our following results are developed within the context of Lipschitz-continuous transport maps $f_\theta$.
 In practice, many flow-based methods exhibit Lipschitz-continuity in their transport map, either by design \citep{grathwohl2018ffjord,chen2019residual}, or as a consequence of choice of architecture and activation function (\Cref{tab:flows}).
@@
 
-\vspace{-2mm}
 \subsection{Closure of Tail Classes}
@@
 
-\vspace{-1mm}
 Our first set of results pertains to the closure of the tail classes in \Cref{def:tail-classification}
@@
 
+\begin{proof}[Proof of \Cref{thm:distn_class_closed}]
+  \label{proof:distn_class_closed}
+  Let $X$ be a random variable from either $\cE_\alpha^p$
+  or $\cL_\alpha^p$.
+  Its concentration function
+  (Equation 1.6 \citet{ledoux2001concentration}
+  is given by
+  \[
+    \alpha_X(r)
+    \coloneqq \sup \{ \mu\{x : d(x,A) \geq r\}; A \subset \text{supp}~X, \mu(A) \geq 1/2\}
+    = \PP(\lvert X - m_X \rvert \geq r) .
+  \]
+  Under Assumption 1, $f_\theta$ is Lipschitz (say with Lipschitz
+  constant $L$) so by Proposition 1.3 of \citet{ledoux2001concentration},
+  \[
+    \PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)
+    \leq 2 \alpha_X(r/L)
+    = \cO(\alpha_X(r/L)),
+  \]
+  where $m_{f_\theta(X)}$ is a median of $f_\theta(X)$.
+  Furthermore, by the triangle inequality
+  \begin{align}
+    \PP(\lvert f_\theta(X) \rvert \geq r)
+     & = \PP(\lvert f_\theta(X) - m_{f_\theta(X)} + m_{f_\theta(X)} \rvert \geq r) \nonumber                 \\
+     & \leq \PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r - \lvert m_{f_\theta(X)}\rvert ) \nonumber \\
+     & = \cO(\PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)) \nonumber                               \\
+     & = \cO(\alpha_X(r/L)) ,
+    \label{eq:pushforward-conc-fn}
+  \end{align}
+  where the asymptotic equivalence holds because $\lvert m_{f_\theta(X)} \rvert$ is independent of $r$.
+  When $X \in \cE_\alpha^p$, \Cref{eq:pushforward-conc-fn} implies
+  \[
+    \PP(\lvert f_\theta(X) \rvert \geq r)
+    = \cO(e^{-\frac{\alpha}{L} r^p}) \implies f_\theta(X) \in \overline{\cE}_{\alpha/L}^p,
+  \]
+  from whence we find that the Lipschitz transform of exponential-type
+  tails continues to possess exponential-type tails with the same
+  class index $p$, although the tail parameter may have changed. Hence,
+  $\overline{\cE^p}$ is closed under Lipschitz maps for each $p \in \RR_{>0}$.
+  On the other hand, when $X \in \cL_\alpha^p$, \Cref{eq:pushforward-conc-fn} also implies that
+  \begin{align*}
+    \PP(\lvert f_\theta(X) \rvert \geq r)
+     & = \cO(e^{-\alpha (\log (r/L))^p})
+    % &= \cO(e^{-\alpha (\log r)^p} e^{-\alpha (-\log L)^p}) \\
+    = \cO(e^{-\alpha (\log r)^p}),
+  \end{align*}
+  and therefore, $f_\theta(X) \in \overline{\cL_\alpha^p}$.
+  Unlike exponential-type tails, Lipschitz transforms of
+  logarithmic-type tails not only remain logarithmic, but
+  their tails decay no slower than a logarithmic-type tail
+  of the same class index with the \emph{same} tail parameter $\alpha$.
+  This upper bound suffices to show closure under Lipschitz maps for the
+  ascending family $\overline{\cL_\alpha^p}$.
+\end{proof}
+
 Note that $\overline{\cL^p_\alpha} \supset \cE^q_\beta$ for all $p,q,\alpha,\beta$, so \Cref{thm:distn_class_closed}
@@
 explicitly enforce smoothness and monotonicity \citep{wehenkel2019unconstrained,huang2018neural,durkan2019neural}
-and hence satisfy the premises. In fact, we can show a version of \Cref{thm:distn_class_closed} ensuring closure of exponential-type
+and hence satisfy the premises.
+
+\begin{proof}[Proof of \Cref{corr:heavy_to_light}]
+  Let $f_\theta$ be as before with the additional assumptions.
+  Since $f_\theta$ is a smooth continuous bijection, it is a diffeomorphism.
+  Furthermore, by assumption $f_\theta$ has invertible Jacobian on the closure of its
+  domain hence $\sup_{x \in \text{dom}~f_\theta} \lvert (f_\theta)'(x) \rvert \geq M > 0$.
+  By the inverse function theorem, $(f_\theta)^{-1}$ exists and is
+  a diffeomorphism with
+  \[
+    \frac{d}{dx}(f_\theta)^{-1}(x) = \frac{1}{(f_\theta)'((f_\theta)^{-1}(x))} \leq \frac{1}{M} .
+  \]
+  Therefore, $(f_\theta)^{-1}$ is $M^{-1}$-Lipschitz and we may apply
+  \Cref{thm:distn_class_closed} to conclude the desired result.
+  %\footnote{\url{https://math.stackexchange.com/questions/394908/diffeomorphism-from-inverse-function-theorem}}
+\end{proof}
+
+In fact, we can show a version of \Cref{thm:distn_class_closed} ensuring closure of exponential-type
 distributions under polynomial transport maps which do not satisfy \Cref{assump:lipschitz}.
@@
 
+\begin{proof}[Proof of \Cref{corr:closure_polynomials}]
+  Let $X \in \cE^p_\alpha$.
+  By considering sufficiently large $X$ such that leading powers dominate, it suffices to consider monomials $Y = X^k$.
+  Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
+  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$.
+\end{proof}
+
 % Suffices to show for k \in \RR_{> 0}, this covers inverse powers
@@
 
-\vspace{-2mm}
 \subsection{Multivariate Fat-Tails and Anisotropic Tail Adaptive Flows}
 
-\vspace{-1mm}
 Next, we restrict attention to power-law tails $\cL^1_\alpha$, and we develop a multivariate fat-tailed theory and notions of isotropic/anisotropic tail indices.
@@
   \centering
-  \includegraphics{../Figures/radial-fat-tail.pdf}%
-%   \includegraphics[trim={0 0 7cm 0},clip]{../Figures/radial-fat-tail.pdf}\\%
-%   \vspace{-8mm}
-%   \includegraphics[trim={7.2cm 0 0 0},clip]{../Figures/radial-fat-tail.pdf}%
-  %\includegraphics[width=\textwidth]{../Figures/radial-fat-tail.png}%
-  %\input{../Figures/radial-fat-tail.pgf}
-  \vspace{-7mm}
+  \includegraphics{ftvi/Figures/radial-fat-tail.pdf}%
+  %   \includegraphics[trim={0 0 7cm 0},clip]{ftvi/Figures/radial-fat-tail.pdf}\\%
+  %   \includegraphics[trim={7.2cm 0 0 0},clip]{ftvi/Figures/radial-fat-tail.pdf}%
+  %\includegraphics[width=\textwidth]{ftvi/Figures/radial-fat-tail.png}%
+  %\input{ftvi/Figures/radial-fat-tail.pgf}
   \caption{
@@
 %   \centering
-%   \includegraphics[width=0.4\textwidth]{../Figures/studentt-product.png}
+%   \includegraphics[width=0.4\textwidth]{ftvi/Figures/studentt-product.png}
 %   \caption{
@@
 % has index $\max_i \nu_i = \nu$.
-Of course, one can construct pathological densities where this definition is not effective
-(see \Cref{eg:spiral}), but it will suffice for our purposes. 
+
+\subsubsection{Example of Non-existence of Tail Parameter Due to Oscillations}
+\label{eg:spiral}
+Of course, one can construct pathological densities where this definition is not effective.
+As a degenerate example, consider
+$\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
+using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
+oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.
+
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[scale=0.8]{ftvi/Figures/spiral.pdf}
+  % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\
+  % \includegraphics[trim={5cm 0 4.9cm 0},clip]{Figures/spiral.pdf}\\
+  % \includegraphics[trim={9.3cm 0 0cm 0},clip]{Figures/spiral.pdf}\\
+  \caption{Taking a tail-anisotropic distribution (left) and ``spinning'' it (middle) results in
+    one-dimensional projections which oscillate between tail parameters (as seen in
+    $\log p(\braket{X,e_0})$ in right panel) and result in an ill-defined
+    direction-dependent tail parameter function $\alpha_X(\cdot)$ due to a
+    divergent limit.
+  }
+  \label{fig:spiral}
+\end{figure*}
+
+\subsubsection{Comparison versus TAF}
+
 It is illustrative to contrast with the theory presented for TAF \citep{jaini2020tails},
@@
 \begin{align*}
-  &\PP[\|X\|_2 \geq t]
-  = \PP\left[\sup_{z \in \cS^{d-1}} \braket{X,z} \geq t\right]\\
-  &\quad \geq \sup_{z \in \cS^{d-1}} \PP[\braket{X,z} \geq t]
+   & \PP[\|X\|_2 \geq t]
+  = \PP\left[\sup_{z \in \cS^{d-1}} \braket{X,z} \geq t\right]  \\
+   & \quad \geq \sup_{z \in \cS^{d-1}} \PP[\braket{X,z} \geq t]
   = \max_{1 \leq i \leq d} \nu_i
@@
 \end{proposition}
+
+To show Proposition \ref{prop:isotropic-pushforward}, we will require a few extra assumptions to rule out pathological cases.
+The full content of Proposition \ref{prop:isotropic-pushforward} is contained in the following theorem.
+
+\begin{theorem}
+  Suppose there exists $\nu > 0$ such that $C:\mathcal{S}^{d-1}\to(0,\infty)$ satisfies $C(v) \coloneqq \lim_{x \to \infty}x^{\nu}\mathbb{P}(|\langle v, X\rangle| > x)$ for all $v \in \mathcal{S}^{d-1}$.
+If $\nu$ is not an integer and $f$ is a bilipschitz function,
+  %such that $f(X)/\|f(X)\|$ has full support on $\mathcal{S}^{d-1}$
+  then $f(X)$ is tail-isotropic with tail index $\nu$.
+\end{theorem}
+\begin{proof}
+  Since $x \mapsto \langle v, f(x)\rangle$ is Lipschitz continuous for any $v \in \mathcal{S}^{d-1}$, Theorem \ref{thm:distn_class_closed} implies $\langle v, f(X)\rangle \in \overline{\mathcal{L}_{\nu}^{1}}$.
+Let $\theta \in (0,\pi/2)$ (say, $\theta = \pi / 4$), and let $S_v = \{x\, : \, \cos^{-1}(\langle x/\|x\|, v\rangle)\leq\theta\}$ for each $v \in \mathcal{S}^{d-1}$.
+Then
+  \[
+    H_v \coloneqq \{x\,:\,\langle v, x \rangle > 1\} \supset \{x\,:\,\|x\| > (1-\cos\theta)^{-1}\} \cap S_v.
+  \]
+  From Theorem C.2.1 of \citet{buraczewski2016stochastic}, since $\nu \not \in \mathbb{Z}$, there exists a non-zero measure $\mu$ such that
+  \[
+    \mu(E) = \lim_{x \to \infty} \frac{\mathbb{P}(x^{-1}X \in E)}{\mathbb{P}(\|X\| > x)},
+  \]
+  for any Borel set $E$. Consequently, $\mu$ is regularly varying, and so
+  by the spectral representation of regularly varying random vectors (see p. 281 \citet{buraczewski2016stochastic}), there exists a measure $P$ such that
+  \[
+    \lim_{x \to \infty}\frac{\mathbb{P}(\|X\|>tx, X/\|X\| \in E)}{\mathbb{P}(\|X\| > x)} = t^{-\nu} P(E),
+  \]
+  for any Borel set $E$ on $\mathcal{S}^{d-1}$ and any $t > 0$.
+Letting $F_v = \{y / \|y\|\,:\, f(y) \in S_v\} \subset \mathcal{S}^{d-1}$ (noting that $P(F_v) > 0$ by assumption), since $m\|x - y\| \leq \|f(x) - f(y)\| \leq M\|x - y\|$ for all $x,y$,
+  \begin{align*}
+    \liminf_{x \to \infty}
+    \frac{\mathbb{P}(f(X) \in x H_v)}{\mathbb{P}(\|f(X)\| > x)}
+     & \geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|f(X)\| > x(1-\cos\theta)^{-1}, f(X) \in S_v)}{\mathbb{P}(\|f(X)\| > x)} \\
+     & \geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|X\| > x(m(1-\cos\theta))^{-1}, X/\|X\| \in F_v)}{\|X\| > x / M}         \\
+     & \geq P(F_v) \left(\frac{M}{m(1-\cos\theta)}\right)^{-\nu} > 0,
+    yaB\end{align*}
+  where $P(F_v) > 0$ follows from the bilipschitz condition for $f$.
+Therefore, we have shown that $\mathbb{P}(\langle v, f(X)\rangle > x) = \Theta(\mathbb{P}(\|f(X)\| > x))$ for every $v \in \mathcal{S}^{d-1}$.
+  Since $\mathbb{P}(\|f(X)\| > x)$ obeys a power law with exponent $\nu$ by Corollary \ref{corr:heavy_to_light}, $f(X)$ is tail-isotropic with exponent $\nu$.
+\end{proof}
 
@@
 
+\subsubsection{Anisotropic tail adaptive flows}
+
 To work around this limitation without relaxing \Cref{assump:lipschitz}, it is evident
@@
 
-%\vspace{-2mm}
 %\subsection{Discussion of limitations}
@@
 
-%\vspace{-1mm}
 %By only considering projections $\braket{X,v}$ along rays through the origin
@@
 %   \centering
-%   \includegraphics[width=0.6\textwidth]{../Figures/hill.png}
+%   \includegraphics[width=0.6\textwidth]{ftvi/Figures/hill.png}
 %   \caption{Hill estimator against $\nu=1.5$ StudentT}
@@
 %   \centering
-%   \includegraphics[width=0.9\textwidth]{../Figures/our-estimator.png}
+%   \includegraphics[width=0.9\textwidth]{ftvi/Figures/our-estimator.png}
 %   \caption{Tail index estimation on location-scale mixture representation
@@
 
-\vspace{-2mm}
 \section{Experiments}
@@
 
-\vspace{-1mm}
 Here we validate ATAF's ability to improve
@@
 % TODO: add facebookresearch reference
-Additional details for the experiments are detailed in \Cref{sec:additional-exp-details}.
+
+All experiments were performed on an Intel i8700K with 32GB RAM and a NVIDIA GTX 1080
+running PyTorch 1.9.0 / Python 3.8.5 / CUDA 11.2 / Ubuntu Linux 20.04 via Windows Subsystem for Linux.
+
+\subsection{Toy Examples}
+\subsubsection{Normal-normal Conjugate Model}
+\label{sec:normal-normal-location-mixture}
+
+We consider a Normal-Normal conjugate inference problem where the posterior
+is known to be a Normal distribution as well. Here, we aim to show that ATAF
+performs no worse than ADVI because $\text{StudentT}(\nu) \to N(0, 1)$ as $\nu \to \infty$.
+\Cref{fig:normal_normal} shows the resulting density approximation, which can
+be seen to be reasonable for both a Normal base distribution (the ``correct'' one)
+and a StudentT base distribution. This suggests that mis-specification (i.e., heavier
+tails in the base distribution than the target) may not be too problematic.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\textwidth]{ftvi/Figures/normal_normal_posterior.png}
+  \caption{Variational inference against a light tailed Normal posterior. Both light and heavy tail
+    variational families yield similar results.}
+  \label{fig:normal_normal}
+\end{figure}
+
+\subsubsection{Experiments Performing VI Against a Fat-tailed Cauchy Target}
+\label{sec:cauchy_normal_student}
+The motivation for the fat-tailed variational families used in TAF/ATAF
+is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
+As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
+appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
+imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
+As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
+test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
+than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.
+
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics{ftvi/Figures/fat_tail_ks.pdf}
+  \caption{
+    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
+    the use of a Gaussian variational family (ADVI, solid blue line) can incur
+    exponentially bad tail approximations (middle panel) compared to
+    methods such as ATAF which permit heavier tails (orange dashed line).
+    As a consequence, ADVI samples (blue, right panel) are rejected by
+    the Kolmogorov-Smirnov test more often than ATAF samples (orange, right panel).
+    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.
+It might be good to show ATAF is much better but still not perfect.
+Also, what is the X axis on each figure.
+}
+  }
+  \label{fig:cauchy_normal_student}
+\end{figure*}
+
+\subsection{Real-World Datasets}
+
+For all flow-transforms $\Phi_{\text{Flow}}$, we used inverse autoregressive flows \citep{kingma2016improved} with a
+dense autoregressive conditioner consisting of two layers of either 32 or 256 hidden units depending on problem (see code for details) and
+ELU activation functions.
+As described in \citet{jaini2020tails}, TAF is trained by including $\nu$ within the Adam optimizer alongside other flow parameters. For ATAF, we include all $\nu_i$ within the optimizer.
+Models were trained using the Adam optimizer with $10^{-3}$ learning rate
+for 10000 iterations, which we found empirically in all our experiments to result in negligible change in ELBO
+at the end of training.
+
+For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
+is comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
+using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
+used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empkkirically observed
+this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
+the absence of outliers when using a Gaussian base distribution resulting in more stable ELBO gradients. This suggests
+other techniques for handling outliers such as winsorization may also be helpful, and we leave further investigation
+for future work.
+
+For \cref{fig:blr-anisotropic}, the closed-form posterior was computed over a finite element grid to produce
+the ``Target'' row. A similar progressive training scheme used for \cref{tab:diamonds} was also used here, with
+the TAF flow transform $\Phi_{\text{Flow}}$ initialized from the result of ADVI and ATAF additionally initialized
+all tail parameters $\nu_i$ based on the final shared tail parameter obtained from TAF training. Tails are computed
+along the $\beta = 1$ or $\sigma = 1$ axes because the posterior is identically zero for $\sigma = 0$, hence it reveals
+no information about the tails.
 
@@
   \centering
-  \vspace{-0.2cm}
-  \includegraphics{../Figures/blr_aniso.pdf}
-  \vspace{-0.3cm}
+  \includegraphics{ftvi/Figures/blr_aniso.pdf}
   \caption{
@@
   \centering
-%   \includegraphics[width=1.0\textwidth]{../Figures/eight_schools.png}
+  %   \includegraphics[width=1.0\textwidth]{ftvi/Figures/eight_schools.png}
   \begin{subfigure}[t]{0.49\textwidth}
@@
   \end{subfigure}
-    \vspace{-2mm}
   \caption{Monte-Carlo ELBO and importance weighted Monte-Carlo marginal likelihood
@@
   \label{fig:eight_schools}
-  \vspace{-4mm}
 \end{table*}
@@
   \end{tabular}
-    \vspace{-2mm}
   \caption{
@@
     density modeling tasks involving financial returns \citep{fama2015five} and insurance claims \citep{cms} data.
-        \vspace{-5mm}
   }
@@
 %   \centering
-%   \includegraphics[width=0.9\textwidth]{../Figures/blr.png}
+%   \includegraphics[width=0.9\textwidth]{ftvi/Figures/blr.png}
 %   \caption{VI of Bayesian robust linear regression}
@@
 
-\paragraph{Acknowledgements}
-L.H. and M.M's contributions were supported in part by DARPA, NSF, and ONR.
-F.L.’s research while at the University of California, Berkeley was supported in part by a GFSD fellowship.
+% \paragraph{Acknowledgements}
+% L.H. and M.M's contributions were supported in part by DARPA, NSF, and ONR.
+% F.L.’s research while at the University of California, Berkeley was supported in part by a GFSD fellowship.
 
@@
 
-\section{Experiments Performing VI Against a Fat-tailed Cauchy Target}
-\label{sec:cauchy_normal_student}
-The motivation for the fat-tailed variational families used in TAF/ATAF
-is easily illustrated on a toy example consisting of $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1) \in \cL^1_1$.
-As seen in \Cref{fig:cauchy_normal_student}, while ADVI with normalizing flows \citep{kingma2016improved,webb2019improving}
-appears to provide a reasonable fit to the bulk of the target distribution (left panel), the improper
-imposition of sub-Gaussian tails results in an exponentially bad tail approximation (middle panel).
-As a result, samples drawn from the variational approximation fail a Kolmogorov-Smirnov goodness-of-fit
-test against the true target distribution much more often (right panel, smaller $p$-values imply more rejections)
-than a variational approximation which permits fat-tails. This example is a special case of \Cref{thm:distn_class_closed}.
-
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics{../Figures/fat_tail_ks.pdf}
-  \vspace{-6mm}
-  \caption{
-    When performing FTVI to approximate a $X \sim \text{Cauchy}(x_0 = 0, \gamma = 1)$ target (left panel, green dotted line),
-    the use of a Gaussian variational family (ADVI, solid blue line) can incur
-    exponentially bad tail approximations (middle panel) compared to
-    methods such as ATAF which permit heavier tails (orange dashed line).
-    As a consequence, ADVI samples (blue, right panel) are rejected by
-    the Kolmogorov-Smirnov test more often than ATAF samples (orange, right panel).
-    % \michael{Why does (b) look so good; are the tails really perfect, of if we went out to 15 or 30 on the X axis.
-It might be good to show ATAF is much better but still not perfect.
-Also, what is the X axis on each figure.
-}
-  }
-  \label{fig:cauchy_normal_student}
-\end{figure*}
-
-\section{Proofs of Our Main Theoretical Results}
-\label{sec:proofs}
-
-\begin{proof}[Proof of \Cref{thm:distn_class_closed}]
-  \label{proof:distn_class_closed}
-  Let $X$ be a random variable from either $\cE_\alpha^p$
-  or $\cL_\alpha^p$.
-  Its concentration function
-  (Equation 1.6 \citet{ledoux2001concentration}
-  is given by
-  \[
-    \alpha_X(r)
-    \coloneqq \sup \{ \mu\{x : d(x,A) \geq r\}; A \subset \text{supp}~X, \mu(A) \geq 1/2\}
-    = \PP(\lvert X - m_X \rvert \geq r) .
-  \]
-  Under Assumption 1, $f_\theta$ is Lipschitz (say with Lipschitz
-  constant $L$) so by Proposition 1.3 of \citet{ledoux2001concentration},
-  \[
-    \PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)
-    \leq 2 \alpha_X(r/L)
-    = \cO(\alpha_X(r/L)),
-  \]
-  where $m_{f_\theta(X)}$ is a median of $f_\theta(X)$.
-  Furthermore, by the triangle inequality
-  \begin{align}
-    \PP(\lvert f_\theta(X) \rvert \geq r)
-    &= \PP(\lvert f_\theta(X) - m_{f_\theta(X)} + m_{f_\theta(X)} \rvert \geq r) \nonumber\\
-    &\leq \PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r - \lvert m_{f_\theta(X)}\rvert ) \nonumber\\
-    &= \cO(\PP(\lvert f_\theta(X) - m_{f_\theta(X)}\rvert \geq r)) \nonumber\\
-    &= \cO(\alpha_X(r/L)) ,
-    \label{eq:pushforward-conc-fn}
-  \end{align}
-  where the asymptotic equivalence holds because $\lvert m_{f_\theta(X)} \rvert$ is independent of $r$.
-  When $X \in \cE_\alpha^p$, \Cref{eq:pushforward-conc-fn} implies
-  \[
-    \PP(\lvert f_\theta(X) \rvert \geq r)
-    = \cO(e^{-\frac{\alpha}{L} r^p}) \implies f_\theta(X) \in \overline{\cE}_{\alpha/L}^p,
-  \]
-  from whence we find that the Lipschitz transform of exponential-type
-  tails continues to possess exponential-type tails with the same
-  class index $p$, although the tail parameter may have changed. Hence,
-  $\overline{\cE^p}$ is closed under Lipschitz maps for each $p \in \RR_{>0}$.
-  On the other hand, when $X \in \cL_\alpha^p$, \Cref{eq:pushforward-conc-fn} also implies that
-  \begin{align*}
-    \PP(\lvert f_\theta(X) \rvert \geq r)
-    &= \cO(e^{-\alpha (\log (r/L))^p})
-    % &= \cO(e^{-\alpha (\log r)^p} e^{-\alpha (-\log L)^p}) \\
-    = \cO(e^{-\alpha (\log r)^p}),
-  \end{align*}
-  and therefore, $f_\theta(X) \in \overline{\cL_\alpha^p}$.
-  Unlike exponential-type tails, Lipschitz transforms of
-  logarithmic-type tails not only remain logarithmic, but
-  their tails decay no slower than a logarithmic-type tail
-  of the same class index with the \emph{same} tail parameter $\alpha$.
-  This upper bound suffices to show closure under Lipschitz maps for the
-  ascending family $\overline{\cL_\alpha^p}$.
-\end{proof}
-
-\begin{proof}[Proof of \Cref{corr:heavy_to_light}]
-    Let $f_\theta$ be as before with the additional assumptions.
-    Since $f_\theta$ is a smooth continuous bijection, it is a diffeomorphism.
-    Furthermore, by assumption $f_\theta$ has invertible Jacobian on the closure of its
-    domain hence $\sup_{x \in \text{dom}~f_\theta} \lvert (f_\theta)'(x) \rvert \geq M > 0$.
-    By the inverse function theorem, $(f_\theta)^{-1}$ exists and is
-    a diffeomorphism with
-    \[
-    \frac{d}{dx}(f_\theta)^{-1}(x) = \frac{1}{(f_\theta)'((f_\theta)^{-1}(x))} \leq \frac{1}{M} .
-    \]
-    Therefore, $(f_\theta)^{-1}$ is $M^{-1}$-Lipschitz and we may apply
-    \Cref{thm:distn_class_closed} to conclude the desired result.
-    %\footnote{\url{https://math.stackexchange.com/questions/394908/diffeomorphism-from-inverse-function-theorem}}
-\end{proof}
-
-\begin{proof}[Proof of \Cref{corr:closure_polynomials}]
-  Let $X \in \cE^p_\alpha$.
-  By considering sufficiently large $X$ such that leading powers dominate, it suffices to consider monomials $Y = X^k$.
-  Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
-  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$. 
-\end{proof}
-
-\begin{lemma}
-    \label[lemma]{lem:sum-rule}
-    Suppose $X \in \cL^1_\alpha$ and $Y \in \cL^1_\beta$.
-    Then $X + Y \in \cL^1_{\min\{\alpha,\beta\}}$.
-\end{lemma}
-
-\begin{proof}
-First, let $\gamma=\min\{\alpha,\beta\}$.
-It will suffice to show that (I) $\mathbb{P}(|X+Y|\geq r)=\mathcal{O}(r^{-\gamma})$, and (II) $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\gamma})$.
-Since $(X,Y)\mapsto|X+Y|$ is a 1-Lipschitz function on $\mathbb{R}^{2}$ and $\mathbb{P}(|X|\geq r)+\mathbb{P}(|Y|\geq r)=\mathcal{O}(r^{-\gamma})$, (I) follows directly from the hypotheses and Proposition 1.11 of \citet{ledoux2001concentration}.
-To show (II), note that for any $M>0$, conditioning on the event $|Y|\leq M$,\[
-\mathbb{P}\left(\left|X\right|+|Y|\geq r\,\vert\,|Y|\leq M\right)\geq\mathbb{P}\left(\left|X\right|\geq r-M\right).
-\]
-Therefore, by taking $M$ to be sufficiently large so that $\mathbb{P}(|Y|\leq M)\geq\frac{1}{2}$,
-\begin{align*}
-\mathbb{P}\left(|X+Y|\geq r\right)&\geq\mathbb{P}\left(|X|+|Y|\geq r\right)\\
-&\geq\mathbb{P}\left(\left|X\right|+|Y|\geq r\,\vert\,|Y|\leq M\right)\mathbb{P}\left(\left|Y\right|\leq M\right)\\
-&\geq\frac{1}{2}\mathbb{P}\left(\left|X\right|\geq r-M\right)=\Theta(r^{-\alpha}).
-\end{align*}
-The same process with $X$ and $Y$ reversed implies $\mathbb{P}(|X+Y|\geq r)\geq\Theta(r^{-\beta})$ as well. Both (II) and the claim follow.
-\end{proof}
-
-To show Proposition \ref{prop:isotropic-pushforward}, we will require a few extra assumptions to rule out pathological cases.
-The full content of Proposition \ref{prop:isotropic-pushforward} is contained in the following theorem.
-
-\begin{theorem}
-Suppose there exists $\nu > 0$ such that $C:\mathcal{S}^{d-1}\to(0,\infty)$ satisfies $C(v) \coloneqq \lim_{x \to \infty}x^{\nu}\mathbb{P}(|\langle v, X\rangle| > x)$ for all $v \in \mathcal{S}^{d-1}$.
-If $\nu$ is not an integer and $f$ is a bilipschitz function,
-%such that $f(X)/\|f(X)\|$ has full support on $\mathcal{S}^{d-1}$
-then $f(X)$ is tail-isotropic with tail index $\nu$. 
-\end{theorem}
-\begin{proof}
-Since $x \mapsto \langle v, f(x)\rangle$ is Lipschitz continuous for any $v \in \mathcal{S}^{d-1}$, Theorem \ref{thm:distn_class_closed} implies $\langle v, f(X)\rangle \in \overline{\mathcal{L}_{\nu}^{1}}$.
-Let $\theta \in (0,\pi/2)$ (say, $\theta = \pi / 4$), and let $S_v = \{x\, : \, \cos^{-1}(\langle x/\|x\|, v\rangle)\leq\theta\}$ for each $v \in \mathcal{S}^{d-1}$.
-Then
-\[
-H_v \coloneqq \{x\,:\,\langle v, x \rangle > 1\} \supset \{x\,:\,\|x\| > (1-\cos\theta)^{-1}\} \cap S_v.
-\]
-From Theorem C.2.1 of \citet{buraczewski2016stochastic}, since $\nu \not \in \mathbb{Z}$, there exists a non-zero measure $\mu$ such that
-\[
-\mu(E) = \lim_{x \to \infty} \frac{\mathbb{P}(x^{-1}X \in E)}{\mathbb{P}(\|X\| > x)},
-\]
-for any Borel set $E$. Consequently, $\mu$ is regularly varying, and so 
-by the spectral representation of regularly varying random vectors (see p. 281 \citet{buraczewski2016stochastic}), there exists a measure $P$ such that
-\[
-\lim_{x \to \infty}\frac{\mathbb{P}(\|X\|>tx, X/\|X\| \in E)}{\mathbb{P}(\|X\| > x)} = t^{-\nu} P(E),
-\]
-for any Borel set $E$ on $\mathcal{S}^{d-1}$ and any $t > 0$.
-Letting $F_v = \{y / \|y\|\,:\, f(y) \in S_v\} \subset \mathcal{S}^{d-1}$ (noting that $P(F_v) > 0$ by assumption), since $m\|x - y\| \leq \|f(x) - f(y)\| \leq M\|x - y\|$ for all $x,y$,
-\begin{align*}
-\liminf_{x \to \infty}
-\frac{\mathbb{P}(f(X) \in x H_v)}{\mathbb{P}(\|f(X)\| > x)} 
-&\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|f(X)\| > x(1-\cos\theta)^{-1}, f(X) \in S_v)}{\mathbb{P}(\|f(X)\| > x)} \\
-&\geq \liminf_{x \to \infty}\frac{\mathbb{P}(\|X\| > x(m(1-\cos\theta))^{-1}, X/\|X\| \in F_v)}{\|X\| > x / M} \\
-&\geq P(F_v) \left(\frac{M}{m(1-\cos\theta)}\right)^{-\nu} > 0,
-yaB\end{align*}
-where $P(F_v) > 0$ follows from the bilipschitz condition for $f$.
-Therefore, we have shown that $\mathbb{P}(\langle v, f(X)\rangle > x) = \Theta(\mathbb{P}(\|f(X)\| > x))$ for every $v \in \mathcal{S}^{d-1}$.
-Since $\mathbb{P}(\|f(X)\| > x)$ obeys a power law with exponent $\nu$ by Corollary \ref{corr:heavy_to_light}, $f(X)$ is tail-isotropic with exponent $\nu$.
-\end{proof}
-
 %\begin{proof}[Proof of \Cref{prop:isotropic-pushforward}]
@@
 %\end{proof}
-
-\section{Example of Non-existence of Tail Parameter Due to Oscillations}
-\label{eg:spiral}
-
-Consider $\text{StudentT}(\nu=1) \otimes \text{StudentT}(\nu=2)$ and ``spin'' it
-using the radial transformation $(r,\theta) \mapsto (r,r+\theta)$ (\Cref{fig:spiral}). Due to
-oscillations, $\alpha_X(v)$ is not well defined for all $v \in \cS^{1}$.
-
-\begin{figure*}[htbp]
-    \centering
-    \vspace{-.5in}
-    \includegraphics[scale=0.8]{../Figures/spiral.pdf}
-    % \includegraphics[trim={0 0 9cm 0},clip]{Figures/spiral.pdf}\\
-    % \includegraphics[trim={5cm 0 4.9cm 0},clip]{Figures/spiral.pdf}\\
-    % \includegraphics[trim={9.3cm 0 0cm 0},clip]{Figures/spiral.pdf}\\
-    \caption{Taking a tail-anisotropic distribution (left) and ``spinning'' it (middle) results in
-        one-dimensional projections which oscillate between tail parameters (as seen in
-        $\log p(\braket{X,e_0})$ in right panel) and result in an ill-defined
-        direction-dependent tail parameter function $\alpha_X(\cdot)$ due to a
-        divergent limit.
-    }
-    \label{fig:spiral}
-\end{figure*}
-
-\section{Normal-normal Conjugate Model}
-\label{sec:normal-normal-location-mixture}
-
-We consider a Normal-Normal conjugate inference problem where the posterior
-is known to be a Normal distribution as well. Here, we aim to show that ATAF
-performs no worse than ADVI because $\text{StudentT}(\nu) \to N(0, 1)$ as $\nu \to \infty$.
-\Cref{fig:normal_normal} shows the resulting density approximation, which can
-be seen to be reasonable for both a Normal base distribution (the ``correct'' one)
-and a StudentT base distribution. This suggests that mis-specification (i.e., heavier
-tails in the base distribution than the target) may not be too problematic.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.6\textwidth]{../Figures/normal_normal_posterior.png}
-  \caption{Variational inference against a light tailed Normal posterior. Both light and heavy tail
-  variational families yield similar results.}
-  \label{fig:normal_normal}
-\end{figure}
-
-\section{Additional Details For Experiments}
-\label{sec:additional-exp-details}
-
-All experiments were performed on an Intel i8700K with 32GB RAM and a NVIDIA GTX 1080
-running PyTorch 1.9.0 / Python 3.8.5 / CUDA 11.2 / Ubuntu Linux 20.04 via Windows Subsystem for Linux.
-For all flow-transforms $\Phi_{\text{Flow}}$, we used inverse autoregressive flows \citep{kingma2016improved} with a
-dense autoregressive conditioner consisting of two layers of either 32 or 256 hidden units depending on problem (see code for details) and
-ELU activation functions.
-As described in \citet{jaini2020tails}, TAF is trained by including $\nu$ within the Adam optimizer alongside other flow parameters. For ATAF, we include all $\nu_i$ within the optimizer.
-Models were trained using the Adam optimizer with $10^{-3}$ learning rate
-for 10000 iterations, which we found empirically in all our experiments to result in negligible change in ELBO
-at the end of training.
-
-For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
-is comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
-using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
-used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empkkirically observed
-this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
-the absence of outliers when using a Gaussian base distribution resulting in more stable ELBO gradients. This suggests
-other techniques for handling outliers such as winsorization may also be helpful, and we leave further investigation
-for future work.
-
-For \cref{fig:blr-anisotropic}, the closed-form posterior was computed over a finite element grid to produce
-the ``Target'' row. A similar progressive training scheme used for \cref{tab:diamonds} was also used here, with
-the TAF flow transform $\Phi_{\text{Flow}}$ initialized from the result of ADVI and ATAF additionally initialized
-all tail parameters $\nu_i$ based on the final shared tail parameter obtained from TAF training. Tails are computed
-along the $\beta = 1$ or $\sigma = 1$ axes because the posterior is identically zero for $\sigma = 0$, hence it reveals
-no information about the tails.
 

### 2022-11-23 hta/neurips_2022.tex
Working compile

@@
-\documentclass[thesis.tex]{subfiles}
+\documentclass[../thesis.tex]{subfiles}
 % \documentclass{article}
@@
 
+\chapter{The generalized gamma tail algebra}
+\label{ch:hta}
+
 % \maketitle
@@
 % \begin{abstract}
-	Despite the successes of probabilistic models based on passing noise through
-	neural networks, recent work has identified that such methods often fail to
-	capture tail behavior accurately unless the tails of the base distribution
-	are appropriately calibrated. To overcome this deficiency, we propose a
+Whereas previous chapters considered adaptive methods which learn an
+approximation's bulk (\cref{ch:lic}) and tail (\cref{ch:ftvi}) from
+samples, in this chapter we develop a
 systematic approach for analyzing the tails of random variables during the
@@
 superior performance across a number of density modeling and variational
-	inference tasks.
+inference tasks. Parts of this chapter have been submitted for peer review
+as \fullcite{gga}.
 % \end{abstract}
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{./figures/schematic.pdf}
+  \includegraphics[width=\textwidth]{hta/figures/schematic.pdf}
   \caption{Our overall approach for density approximations with calibrated tails.
@@
 \end{itemize}
+
+\section{Related Work}
+
+\textbf{Heavy tails and probabilistic machine learning}
+
+For studying heavy tails, methods based on subexponential distributions
+\citep{goldie1998subexponential} and generalized Pareto distributions (GPD) or
+equivalently regularly varying distributions \citep{tajvidi2003confidence}
+have recieved attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
+Heavy tails can impact probabilistic machine learning methods in a number of ways.
+The observation that density ratios $\frac{p(x)}{q(x)}$ tend to be heavy tailed
+has resulted in new methods for smoothing importance sampling \citep{vehtari2015pareto},
+adaptively modifying divergences \citep{wang2018variational}, and
+diagnosing variational inference through the Pareto $\hat{k}$ diagnostic \citep{yao2018yes}.
+These works are complementary to our paper and our reported results include $\hat{k}$
+diagnostics for VI and $\hat\alpha$ tail index estimates based on GPD.
+
+Our work considers heavy-tailed targets $p(x)$ which is the same setting as
+\citet{jaini2020tails,ftvi}. Whereas those respective works
+lump the tail parameter in as another variational parameter and may be more generally applicable, the GGA may be applied before samples are drawn and leads to perfectly calibrated tails when applicable.
+
+\textbf{Probabilistic programming}
+
+PPLs can be characterized by the primary use case optimized for, whether that's
+Gibbs sampling over Bayes nets \citep{spiegelhalter1996bugs,de2017programming},
+stochastic control flow \citep{goodman2012church,wingate2011lightweight},
+deep stochastic variational inference \citep{tran2018simple,bingham2019pyro}, or
+Hamiltonian Monte-Carlo \citep{carpenter2017stan,xu2020advancedhmc}. Our implementation target
+\texttt{beanmachine} \citep{tehrani2020bean} is a declarative PPL selected
+due to availability of a PPL compiler and support for static analysis plugins.
+Similar to \citet{bingham2019pyro,siddharth2017learning}, it uses PyTorch \citep{paszke2019pytorch} for GPU tensors and automatic differentiation.
+Synthesizing an approximating distribution during PPL compilation (\Cref{sec:impl}) is
+also performed in the Stan language by \citet{kucukelbir2017automatic} and normalizing
+flow extensions in \citet{webb2019aml}. We compare directly against these related density approximators in \Cref{sec:hta:experiments}.
+
+\textbf{Static analysis}
+
+There is a long history of formal methods and probabilistic programming
+\cite{kozen1979semantics,jones1989probabilistic}. While much of the research
+\cite{claret2013bayesian}
+is concerned with defining formal semantics
+and establishing invariants \cite{wang2018pmaf}
+See \cite{bernstein2019static} for a recent review.
+Static analysis utilizes the abstract syntax tree (AST)
+representation of a program in order to compute invariants
+(e.g. the return type of a function, the number of classes implementing a trait)
+without executing the underlying program.
+As dynamic analysis in PPs is less reliable due to non-determinism, static analysis
+methods for PPs become increasingly important.
+
+Within PPLs, static analysis has traditionally been applied in the context of formalizing
+semantics \citep{kozen1979semantics} and has been used to verify probabilistic
+programs by ensuring termination, bounding random values values
+\citep{sankaranarayanan2013static}. \cite{lee2019towards} proposes a static analyzer for the
+Pyro PPL \citep{bingham2019pyro} to verify
+distribution supports and avoid $-\texttt{Inf}$ log probabilities.
+
+More relevant to our work are applications of static analysis to improve inference.
+\citet{nori2014r2} statically analyzes a probabilistic program and computes pre-images of observations in order to better adapt MCMC proposal distributions.
+While we also perform static analysis over
+abstract syntax tree (AST) representations of a probabilistic program,
+applying GGA yields an upper bound on the tails of all random variables
+so that calibrated tails can be imposed on distribution estimates.
 
@@
                                                            & $\equiv \begin{cases}
-\max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\} & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1\\
-\left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right) & \text{ if }\rho_{1}=\rho_{2}=1\\
+                                                                         \max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\}                                              & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1 \\
+                                                                         \left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right)                                                   & \text{ if }\rho_{1}=\rho_{2}=1                                \\
                                                                          (\nu_{1}+\nu_{2}+\frac{2-\rho}{2},(\sigma_{1}^{-\frac{1}{\rho-1}}+\sigma_{2}^{-\frac{1}{\rho-1}})^{1-\rho},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}>1.
@@
                                                                                               \end{cases}$                                                                                          \\ \hline
-\centering \textbf{Scalar \linebreak Multiplication} & $c(\nu,\sigma,\rho) \equiv (\nu,\sigma/|c|^\rho,\rho)$\\\hline
+      \centering \textbf{Scalar \linebreak Multiplication} & $c(\nu,\sigma,\rho) \equiv (\nu,\sigma/|c|^\rho,\rho)$                                                                                                                                                                \\\hline
       \multirow{3}{*}{\centering \textbf{Multiplication}}  & $(\nu_{1},\sigma_{1},\rho_{1})\otimes(\nu_{2},\sigma_{2},\rho_{2})$                                                                                                                                                   \\
-&$\qquad \equiv\begin{cases}
-\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0\\
-\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}>0\\
+                                                           & $\qquad \equiv\begin{cases}
+                                                                               \left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0      \\
+                                                                               \left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)      & \text{ if }\rho_{1},\rho_{2}>0      \\
                                                                                \mathcal{R}_{|\nu_1|}                                                                                                          & \mbox{ if }\rho_{1}\leq0,\rho_{2}>0 \\
@@
                                                                              \end{cases}$                                    \\
-&where $\mu=\frac{1}{|\rho_{1}|}+\frac{1}{|\rho_{2}|}=\frac{|\rho_{1}|+|\rho_{2}|}{|\rho_{1}\rho_{2}|}$, $\sigma=\mu(\sigma_{1}|\rho_{1}|)^{\frac{1}{\mu|\rho_{1}|}}(\sigma_{2}|\rho_{2}|)^{\frac{1}{\mu|\rho_{2}|}}$.\\ \hline
+                                                           & where $\mu=\frac{1}{|\rho_{1}|}+\frac{1}{|\rho_{2}|}=\frac{|\rho_{1}|+|\rho_{2}|}{|\rho_{1}\rho_{2}|}$, $\sigma=\mu(\sigma_{1}|\rho_{1}|)^{\frac{1}{\mu|\rho_{1}|}}(\sigma_{2}|\rho_{2}|)^{\frac{1}{\mu|\rho_{2}|}}$.
+\\ \hline
       \centering \textbf{Product of Densities}             & $(\nu_1,\sigma_1,\rho_1)\&(\nu_2,\sigma_2,\rho_2)
       \equiv \begin{cases}
-(\nu_{1}+\nu_{2},\sigma_{1},\rho_{1}) & \text{ if }\rho_{1}<\rho_{2}\\
-(\nu_{1}+\nu_{2},\sigma_{1}+\sigma_{2},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}\\
+                 (\nu_{1}+\nu_{2},\sigma_{1},\rho_{1})        & \text{ if }\rho_{1}<\rho_{2}      \\
+                 (\nu_{1}+\nu_{2},\sigma_{1}+\sigma_{2},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2} \\
                  (\nu_{1}+\nu_{2},\sigma_{2},\rho_{2})        & \text{ otherwise.}
@@
 All operations in the GGA can be proven to exhibit identical behaviour with their corresponding operations on random variables, with the sole exception of reciprocals (marked by asterisk), where additional assumptions are required.
-For further details, refer to the Supplementary Material.
 
@@
 
-\section{Implementation}\label{sec:impl}
-
-\subsection{Compile-time static analysis}
-
-To illustrate an implementation of GGA for static analysis, we sketch the operation of the PPL compiler at a high-level
-and defer to the supplementary code for details.
-A probabilistic program is first inspected using Python's built-in \texttt{ast} module
-and transformed to static single assignment (SSA) form \citep{rosen1988global}.
-Next, standard compiler optimizations (e.g. dead code elimination, constant propagation)
-are applied and an execution of the optimized program is traced \citep{wingate2011lightweight,bingham2019pyro}
-and accumulated in a directed acyclic graph representation.
-A breadth-first type checking pass, as seen in Algorithm~\ref{alg:bfs_typecheck}, completes in linear time, and GGA results may be applied to implement \texttt{computeGGA()} using the following steps:
-\begin{itemize}[leftmargin=*]
-	\item If a node has no parents, then it is an atomic distribution and its tail parameters are known (\Cref{tab:dist_list})
- 	\item Otherwise, the node is an operation taking its potentially stochastic inputs (parents) to its output. Consult \Cref{tab:gga_operations} for the output GGA tails.
-\end{itemize}
-
-%% This declares a command \Comment
-%% The argument will be surrounded by /* ... */
-% \SetKwComment{Comment}{/* }{ */}
-
-\begin{algorithm}
-	\caption{Pseudocode for a GGA tails static analysis pass}\label{alg:bfs_typecheck}
-	\begin{algorithmic}
-	\REQUIRE{Abstract syntax tree for a PPL program}
-
-	\STATE{frontier $\gets$ [rv : Parents(rv) = $\emptyset$]}\\
-	\STATE{GGAs $\gets \{\}$}
-
-	\WHILE{\text{frontier} $\neq \emptyset$} 
-	\STATE{next $\gets$ frontier.popLeft()}\\
-	\STATE{GGAs[next] $\gets$ computeGGA(next.op, next.parent)}\\
-	\STATE{frontier $\gets$ frontier + next.children()}
-	\ENDWHILE
-
-	\RETURN{GGA parameter estimates for all random variables}
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Representative distributions}\label{ssec:repr_dist}
-
-For each $(\nu,\sigma,\rho)$ we make a carefully defined choice of $p$ on $\mathbb{R}$ such that if $X \sim p$, then $X \equiv (\nu,\sigma,\rho)$.
-This way, any random variable $f(X)$, where $f$ is $1$-Lipschitz, will exhibit the correct tail, and so approximations of this form may be used for variational inference or density estimation.
-Let $X \equiv (\nu,\sigma,\rho)$ and $0 < \epsilon \ll 1$ denote a small parameter such that tails $e^{-x^\epsilon}$ are deemed to be ``very heavy'' (we chose $\epsilon = 0.1$).
-%We also let $X\equiv(\nu,\sigma,\rho)$ and $\zeta$ denote an independent Rademacher random variable with $\mathbb{P}(\zeta=1)=\mathbb{P}(\zeta=-1)=\frac12$.
-
-\begin{itemize}
-    \item[($\rho \leq 0$)] If $\rho \leq -1$, then $p_X(x) \sim c x^{-|\nu|}$.
-A prominent distribution on $\mathbb{R}$ with power law tails is the \emph{Student $t$ distribution}, in this case, with $|\nu|-1$ degrees of freedom if $\nu < -1$ (generate $X \sim t_{|\nu|-1})$.
-    \item[($\rho > \epsilon$)] For moderately sized $\rho > 0$, we consider a symmetrized variant of the generalized Gamma density (\Cref{eq:GenGammaDensity}). 
-    \item[($\rho \leq \epsilon$)] If $X \equiv (\nu,\sigma,\rho)$ where $\rho$ is small, then $X$ will exhibit much heavier tails, and the generalized Gamma distribution in Case 1 will become challenging to sample from.
-In these cases, we expect that the tail of $X$ should be well represented by a power law.
-    %Letting $\alpha = \sigma \rho - \nu - 1$, using an approximation to the logarithm, $p_{|X}(x) \approx c x^{-\alpha - 1}$ (see the Supplementary Material for more details).
-Therefore, we represent tails of this form by the Student $t$ distribution with $\alpha$ degrees
-    The generalized Gamma density (\Cref{eq:GenGammaDensity}) satisfies $\mathbb{E}X^r = \sigma^{-r/\rho} \Gamma(\frac{\nu+1+r}{\rho})/\Gamma(\frac{\nu+1}{\rho})$ for $r > 0$.
-Let $\alpha > 0$ be such that $\mathbb{E}X^\alpha = 2$.
-By Markov's inequality, the tail of $X$ satisfies $\mathbb{P}(X>x)\leq 2 x^{-\alpha}$.
-Therefore, we can represent tails of this form by the Student $t$ distribution with $\alpha+1$ degrees of freedom (generate $X \sim t_{\alpha}$).
-\end{itemize}
-
-\subsection{Bulk correction by Lipschitz mapping}
-
-While a representative distribution will exhibit the desired tails,
-the target distribution's bulk may be very different from a generalized Gamma and result in poor distributional approximation.
-To address this, we propose splicing together the tails from a generalized Gamma with a flexible density approximation for the bulk.
-While many combinations are possible, in this work we rely on \Cref{lem:lipschitz} and post-compose neural
-spline flows \citep{durkan2019neural} (which are identity functions outside of a bounded interval) after
-properly initialized generalized Gamma distributions.
-Optimizing the parameters of the flow results in good bulk approximation while simultaneously
-preserving the tail correctness guarantees attained by the GGA.
-
-\begin{example}
-	Let $A \in \mathbb{R}^{k \times k}$, $x,y\in\mathbb{R}^k$, with $x_i,y_i,A_{ij} \overset{\text{iid}}{\sim} \mathcal{N}(-1,1)$.
-	The distribution of $x^\top A y = \sum_{i,j} x_i A_{ij} y_j$ is convolution of normal-powers \citep{gupta2008analyzing} and has no convenient
-	closed form expression.
-	Using GGA's closure theorems (\cref{tab:gga_operations}), one can compute
-	its tail parameters to be $(\frac{k}{2}-1,\frac{3}{2}, \frac{2}{3})$.
-
-	\begin{figure}[h]
-		\centering
-		\input{./experiments/normal-power-sum.pgf}
-		\input{./experiments/lip-nf.pgf}
-		\caption{(Top) $5000$ samples of $\lvert x^\top A y \rvert$ vs the
-		calibrated GGA density $q(x)$. While calibrated tails are provably guarantees,
-		the target distribution's bulk differs from the assumed generalized Gamma representative distribution (\cref{ssec:repr_dist}) for all $k$.
-		To fix the bulk approximation, a normalizing flow is composed with
-		the GGA representative to form $\text{flow}(q)(x)$. The bulk approximation
-		is improved (bottom) while the tails continue to exhibit the same
-		behavior (bottom right).
-		}
-		\label{fig:power_normal}
-	\end{figure}
-
-	The GGA representative is a gamma distribution with the correct tails, but there is non-negligible error in the bulk where $x$ is small. To address this, a learnable bijector can be
-	optimized as in \Cref{fig:power_normal} bottom left to correct the bulk approximation. Guaranteed by \Cref{lem:lipschitz} and
-	visualized in \Cref{fig:power_normal} bottom right, the tails of the overall composition remain calibrated.
-\end{example}
-
-\section{Experiments}\label{sec:experiments}
-
-In this section we demonstrate that GGA-based density estimation yields improvements across a variety of metrics.
-We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushfowards of Normal distributions. To understand the individual
-effect of using a GGA base distribution over standard normals versus more expressive pushforward maps \citep{durkan2019neural}, we also report ablation results where
-normalizing flows are replaced by affine transforms as originally
-proposed in \citep{kucukelbir2017automatic}. All experiments are repeated for 50 trials, trained to convergence using the Adam optimizer with manually tuned learning rate,
-and conducted on i7-8700K CPU and GTX 1080 GPU hardware.
-
-All target distributions in this section are expressed as generative PPL programs:
-Cauchy using a reciprocal normal, Chi2 using a sum of squared normals, IG (Inverse Gamma)
-using a reciprocal exponential, Normal using a sum of normals, and StudentT using a
-normal and Cauchy ratio. Doing so tasks the static analyzer to infer the target's tails and makes the analysis non-trivial.
-See supplementary for full details.
-
-Our results in the following tables share a consistent narrative where a GGA base
-distribution rarely hurts and can significantly help with heavy tailed targets.
-Except for when targets are truly light tailed ($\alpha=\infty$ in Chi2 and Normal),
-GGA-based approximations are the only ones to reproduce appropriate GPD tail index $\hat\alpha$ in density estimation and achieve a passing Pareto $\hat{k}$ diagnostic
-\citep{yao2018yes} below $0.2$ in variational inference. When viewed through traditional
-evaluation metrics such as negative cross-entropy $H(p,q) = E_p \log p(X)$,
-ELBO $E_q \log \frac{q(X)}{p(X)}$,
-and importance-weighted autoencoder bound \citep{burda2015importance} $E_q \log \sum_i^{1000} \frac{p(X)}{q(X)}$, GGA-based approximations remain favorable on almost all heavy-tailed
-targets and have negligible difference for light tailed targets.
-Less surprising is the result that adding a flow improved approximation metrics, as we expect the additional representation flexibility to be beneficial.
-
-%\feynman{explain the targets, generative and marginalized, reference appendix}
-
-\textbf{Density Estimation} We minimize a Monte-Carlo estimate of the cross entropy $H(p,q) = -E_p[\log q(X)] \approx -\frac{1}{N} \sum_{i=1}^N \log q(x_i),\quad x_i \sim p$.
-The results are shown in \Cref{tab:de} along with power-law tail index estimates \citep{clauset2009power} $\hat\alpha$.
-Overall, we see that GGA performs better (lower NLL, $\hat\alpha$ closer to target) when the target
-has heavier tails (lower $\hat\alpha$ target/theory) and that the difference is smaller but still non-negligible
-for distributions such as Chi1 which possess tails heavier than Gaussian.
-
-\begin{table}[h]
-	\centering
-	\caption{Density estimation metrics attained (mean, standard deviation in parenthesis)
-	on targets of varying tail index (smaller $\alpha =$ heavier tails).
-	Higher negative cross entropy $-H(p,q) = E_p \log q(X)$ implies a better overall approximation (row maxes bolded)
-	while close agreement between the target Pareto tail index $\alpha$ \citep{clauset2009power}
-	and its estimate $\hat\alpha$ in $q(x)$ suggest calibrated tails (closest in row bolded).
-	}
-	\label{tab:de}
-	\input{experiments/de_table.tex}
-\end{table}
-
-\textbf{Variational Inference} The optimization objective is the ELBO
-\[
-    E_q \log \frac{p(X)}{q(X)} \approx \frac{1}{N} \sum_{i=1}^N \log \frac{p(x_i)}{q(x_i)},\quad x_i \sim q
-\]
-Here, the density $p$ must also be evaluated so for simplicity
-experiments in \cref{tab:vi} use closed-form marginalized densities for targets.
-The overall trends also show that GGA yields consistent improvements as measured by both ELBO and importance-weighted
-estimates of marginal likelihood and that the difference was greater when the tails of $p(z)$ were heavier.
-The $\hat{k}$ diagnostics \citep{yao2018yes} corroborate our findings that variational inference succeeds
-($\hat{k} < -1.2$) when a GGA with appropriately matched tails is used and fails ($\hat{k} > 1$) when
-Gaussian tails are erroneously imposed.
-
-\begin{table}[h]
-	\centering
-	\caption{Variational inference metrics (mean, standard deviation in parenthesis)
-	on targets of varying tail index (smaller $\alpha =$ heavier tails).
-Both the IWAE bound $E_q \log \sum_i^K \frac{p(X_i)}{q(X_i)}$ and the ELBO ($K=1$) measure (a lower bound) on the marginal likelihood where larger is better (row maxes bolded).
-In \citet{yao2018yes},
-	a Pareto $\hat{k}$ diagnostic $> 0.2$ is interpreted as potentially problematic so only values below are bolded.}
-	\label{tab:vi}
-	\small\input{experiments/vi_table.tex}
-\end{table}
-
-\subsection{SGD for least-squares linear regression}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.32\textwidth]{figures/Kesten_0.40_1.5.png}
-\includegraphics[width=0.32\textwidth]{figures/Kesten_0.50_1.5.png}
-\includegraphics[width=0.32\textwidth]{figures/Kesten_0.50_2.png}
-\caption{\label{fig:SGD}Density of iterates of SGD vs. predicted tail behaviour}
-\end{figure}
-
-For inputs $X$ and labels $Y$ from a dataset $\mathcal{D}$, the least squares estimator for linear regression satisfies $\beta = \min_\beta \tfrac12 \mathbb{E}_{X,Y\sim\mathcal{D}}(Y - X\beta)^2$.
-To solve for this estimator, one can apply stochastic gradient descent (SGD) sampling over independent $X_k,Y_k\sim \mathcal{D}$ to obtain the sequence of iterations
-\[
-\beta_{k+1} = (I - \delta X_k X_k^\top) \beta_k + \delta Y_k X_k
-\]
-for a step size $\delta > 0$.
-For large $\delta$, the iterates $\beta_k$ typically exhibit heavy-tailed fluctuations; in this regard, this sequence of iterates has been used as a simple model for more general stochastic optimization dynamics \citep{gurbuzbalaban2021heavy,hodgkinson2021multiplicative}.
-In particular, generalization performance has been tied to the heaviness of the tails in the iterates \cite{simsekli2019tail}.
-Here we use our algebra to predict the tail behaviour in a simple one-dimensional setting where $X_k \sim \mathcal{N}(0,\sigma^2)$ and $Y_k \sim \mathcal{N}(0,1)$.
-From classical theory \citep{buraczewski2016stochastic}, it is known that $X_k$ converges in distribution to a power law with tail exponent $\alpha > 0$ satisfying $\mathbb{E}|1 - \delta X_k^2|^\alpha = 1$.
-In \cref{fig:SGD}, we plot the density of the representative obtained using our algebra after $10^4$ iterations against a kernel density estimate of the first $10^6$ iterates when $\sigma \in \{0.4,0.5\}$ and $\delta \in \{1.5,2.0\}$.
-In all cases, the density obtained from the algebra provides a surprisingly close fit.
-
-\section{Related Work}
-
-\textbf{Heavy tails and probabilistic machine learning}
-
-For studying heavy tails, methods based on subexponential distributions
-\citep{goldie1998subexponential} and generalized Pareto distributions (GPD) or
-equivalently regularly varying distributions \citep{tajvidi2003confidence}
-have recieved attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
-Heavy tails can impact probabilistic machine learning methods in a number of ways.
-The observation that density ratios $\frac{p(x)}{q(x)}$ tend to be heavy tailed
-has resulted in new methods for smoothing importance sampling \citep{vehtari2015pareto},
-adaptively modifying divergences \citep{wang2018variational}, and
-diagnosing variational inference through the Pareto $\hat{k}$ diagnostic \citep{yao2018yes}.
-These works are complementary to our paper and our reported results include $\hat{k}$
-diagnostics for VI and $\hat\alpha$ tail index estimates based on GPD.
-
-Our work considers heavy-tailed targets $p(x)$ which is the same setting as
-\citet{jaini2020tails,ftvi}. Whereas those respective works
-lump the tail parameter in as another variational parameter and may be more generally applicable, the GGA may be applied before samples are drawn and leads to perfectly calibrated tails when applicable.
-
-\textbf{Probabilistic programming}
-
-PPLs can be characterized by the primary use case optimized for, whether that's
-Gibbs sampling over Bayes nets \citep{spiegelhalter1996bugs,de2017programming},
-stochastic control flow \citep{goodman2012church,wingate2011lightweight},
-deep stochastic variational inference \citep{tran2018simple,bingham2019pyro}, or
-Hamiltonian Monte-Carlo \citep{carpenter2017stan,xu2020advancedhmc}. Our implementation target
-\texttt{beanmachine} \citep{tehrani2020bean} is a declarative PPL selected
-due to availability of a PPL compiler and support for static analysis plugins.
-Similar to \citet{bingham2019pyro,siddharth2017learning}, it uses PyTorch \citep{paszke2019pytorch} for GPU tensors and automatic differentiation.
-Synthesizing an approximating distribution during PPL compilation (\Cref{sec:impl}) is
-also performed in the Stan language by \citet{kucukelbir2017automatic} and normalizing
-flow extensions in \citet{webb2019aml}. We compare directly against these related density approximators in \Cref{sec:experiments}.
-
-\textbf{Static analysis}
-
-There is a long history of formal methods and probabilistic programming
-\cite{kozen1979semantics,jones1989probabilistic}. While much of the research
-\cite{claret2013bayesian}
-is concerned with defining formal semantics
-and establishing invariants \cite{wang2018pmaf}
-See \cite{bernstein2019static} for a recent review.
-Static analysis utilizes the abstract syntax tree (AST)
-representation of a program in order to compute invariants
-(e.g. the return type of a function, the number of classes implementing a trait)
-without executing the underlying program.
-As dynamic analysis in PPs is less reliable due to non-determinism, static analysis
-methods for PPs become increasingly important.
-
-Within PPLs, static analysis has traditionally been applied in the context of formalizing
-semantics \citep{kozen1979semantics} and has been used to verify probabilistic
-programs by ensuring termination, bounding random values values
-\citep{sankaranarayanan2013static}. \cite{lee2019towards} proposes a static analyzer for the
-Pyro PPL \citep{bingham2019pyro} to verify
-distribution supports and avoid $-\texttt{Inf}$ log probabilities.
-
-More relevant to our work are applications of static analysis to improve inference.
-\citet{nori2014r2} statically analyzes a probabilistic program and computes pre-images of observations in order to better adapt MCMC proposal distributions.
-While we also perform static analysis over
-abstract syntax tree (AST) representations of a probabilistic program,
-applying GGA yields an upper bound on the tails of all random variables
-so that calibrated tails can be imposed on distribution estimates.
-
-\section{Conclusion}\label{sec:conclusion}
-
-In this work, we have proposed a novel systematic approach for conducting tail inferential static analysis by implementing a three-parameter generalized Gamma algebra into a PPL compiler.
-Initial results are promising, showing that improved inference with simpler approximation families is possible when combined with tail metadata.
-While already useful, the generalized Gamma algebra and its implementation currently has
-some notable limitations:
-\begin{itemize}[leftmargin=*]
-\item Since the algebra assumes independence, handling of dependencies between defined random variables must be conducted externally.
-This will inevitably require interoperability with a symbolic package to decompose complex expressions into operations on independent random variables.
-\item The GGA is formulated for univariate distributions only. Suitably defining multivariate tails is an open problem with interesting alternatives \citep{jaini2020tails,ftvi} all of which
-could extend GGA to higher dimensions. 
-\item Conditioning is arguably the most important feature of a PPL and what distinguishes it from a glorified simulator.
-Exact marginalization in general is NP-hard \citep{koller2009probabilistic}, so treatment of conditional distributions using symbolic manipulations is a significant open problem, with some basic developments \cite{shan2017exact,cho2019disintegration}.
-Since only the tails are required in our setup, it may be possible to construct a dual algebra for operations under conditioning; this is left for future work.
-\item Compile-time static analysis only applicable to fixed model structure. While out of scope for our current work, open-universe models \citep{milch2010extending}
-and PPLs to support them \citep{bingham2019pyro} are an important research direction.
-\item The most significant omission to the algebra itself is classification of log-normal tails; while addition may be treated using \cite{gulisashvili2016tail} for example, multiplicative convolution with log-normal tails remains elusive.
-\item At present, reciprocals are approximated by assuming behaviour near zero. Reciprocals may be better treated by covering near-zero asymptotics separately. 
-\end{itemize}
-
-The GGA provides a necessary first step into the static analysis of tails in
-a probabilistic program.
-As the above limitations are improved in future work and GGA becomes more broadly applicable,
-we are excited to see how improved tail modelling will improve downstream PPL applications
-as well as
-other researchers will utilize GGA metadata to develop
-novel PPL applications.
-
-% \begin{ack}
-% 	Use unnumbered first level headings for the acknowledgments. All acknowledgments
-% 	go at the end of the paper before the list of references. Moreover, you are required to declare
-% 	funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
-% 	More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2021/PaperInformation/FundingDisclosure}.
-
-% 	Do {\bf not} include this section in the anonymized submission, only in the final paper.
-You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
-% \end{ack}
-
-% \subsection{Limitations}
-
-% \todo{Liam: Log-Normal insensitivity, edge case (see example)}
-
-% The GGA is formulated for univariate distributions only.
-
-% {
-% \small
-% \bibliographystyle{unsrtnat}
-% \bibliography{refs}
-% }
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section*{Checklist}
-
-% %%% BEGIN INSTRUCTIONS %%%
-% The checklist follows the references.  Please
-% read the checklist guidelines carefully for information on how to answer these
-% questions.  For each question, change the default \answerTODO{} to \answerYes{},
-% \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
-% justification to your answer}, either by referencing the appropriate section of
-% your paper or providing a brief inline description.  For example:
-% \begin{itemize}
-%   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
-%   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
-%   \item Did you include the license to the code and datasets? \answerNA{}
-% \end{itemize}
-% Please do not modify the questions and only use the provided macros for your
-% answers.  Note that the Checklist section does not count towards the page
-% limit.  In your paper, please delete this instructions block and only keep the
-% Checklist section heading above along with the questions/answers below.
-% %%% END INSTRUCTIONS %%%
-
-% \begin{enumerate}
-
-% \item For all authors...
-% \begin{enumerate}
-%   \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
-%     \answerYes{}
-%   \item Did you describe the limitations of your work?
-%     \answerYes{See \cref{sec:conclusion}}
-%   \item Did you discuss any potential negative societal impacts of your work?
-%     \answerYes{See \cref{sec:conclusion}}
-%   \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
-%     \answerYes{}
-% \end{enumerate}
-
-% \item If you are including theoretical results...
-% \begin{enumerate}
-%   \item Did you state the full set of assumptions of all theoretical results?
-%     \answerYes{Stated at start of \cref{sec:gga}}
-% 	\item Did you include complete proofs of all theoretical results?
-%     \answerYes{Most are deferred to supplementary materials}
-% \end{enumerate}
-
-% \item If you ran experiments...
-% \begin{enumerate}
-%   \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
-%     \answerYes{See supplemental, a URL for code will be provided witn camera ready}
-%   \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
-%     \answerYes{}
-% 	\item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
-%     \answerYes{Standard errors reported in parenthesis on \cref{tab:de} and \cref{tab:vi}}
-% 	\item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
-%     \answerYes{See \cref{sec:experiments}}
-% \end{enumerate}
-
-% \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
-% \begin{enumerate}
-%   \item If your work uses existing assets, did you cite the creators?
-%     \answerYes{See citations of \citep{tehrani2020bean}}
-%   \item Did you mention the license of the assets?
-%     \answerYes{Licence included in code}
-%   \item Did you include any new assets either in the supplemental material or as a URL?
-%     \answerYes{See supplemental material}
-%   \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
-%     \answerNA{}
-%   \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
-%     \answerNA{}
-% \end{enumerate}
-
-% \item If you used crowdsourcing or conducted research with human subjects...
-% \begin{enumerate}
-%   \item Did you include the full text of instructions given to participants and screenshots, if applicable?
-%     \answerNA{}
-%   \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
-%     \answerNA{}
-%   \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
-%     \answerNA{}
-% \end{enumerate}
-
-% \end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \ifdefined\remappendix
-% \newpage
-
-% \appendix
-
-\section{Additional examples}\label{sec:addtl_eg}
+\subsection{Illustrative examples}\label{sec:addtl_eg}
 
@@
 
-\section{Additional details for experiments}
-
-The targets in \Cref{tab:de} and \Cref{tab:vi} are analyzed using
-the GGA in \Cref{sec:addtl_eg}. Note that Inverse Gamma (``IG'') corresponds
-to the inverse exponential. We selected closed form targets so that
-the Pareto tail index $\alpha$ is known analytically and the quality
-of theoretical predictions as well as empirical results can be evaluated
-against. All experiments are repeated for $100$ trials and $1,000$ samples
-from the model (as well as the approximation in VI) were used to compute
-each gradient estimate. Losses were trained until convergence, which 
-all occurred in under $10,000$ iterations at a $0.05$ learning rate and
-the Adam \citep{kingma2014adam} optimizer.
-
-\subsection{Normal target}
-
-Consider the toy example of a Normal target. This case is trivial for
-Gaussian based methods and is oftentimes the initialization.
-This lack of approximation gap in ADVI is seen in \Cref{fig:normal},
-where we also see that GGA achieves similar approximation quality.
-This is unsurprising as the GGA approximation in \Cref{tab:dist_list}
-is also a Normal distribution.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\textwidth]{figures/de_normal_sum.png}
-	\includegraphics[width=.4\textwidth]{figures/vi_normal_sum.png}
-	\caption{Density estimation and VI against a known normal target}
-	\label{fig:normal}
-\end{figure}
-
-\subsection{Chi-square}
-
-Now let $X_{ij} \sim N(0,1)$ and consider $tr X^\top X$.
-Such quantities arise in the analysis of random projections.
-It is important here to recognize that the power operation $X \mapsto X^2$ is not equivalent to the multiplication operation $X \mapsto X \otimes X$, as multiplication assumes independence.
-\begin{figure}
-	\centering
-	\includegraphics[width=\textwidth]{figures/jl-chi2.png}
-	\caption{$5000$ samples of JL matrix trace (blue) vs GGA prediction (yellow)}
-	\label{fig:jl_chi2}
-\end{figure}
-
-\section{Operations in the Generalized Gamma Algebra}
+\subsection{Operations in the Generalized Gamma Algebra}
 
@@
     \\\equiv \begin{cases}
-\max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\} & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1\\
-\left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right) & \text{ if }\rho_{1}=\rho_{2}=1\\
+      \max\{(\nu_{1},\sigma_{1},\rho_{1}),(\nu_{2},\sigma_{2},\rho_{2})\}                                              & \text{ if }\rho_{1}\neq\rho_{2}\text{ or }\rho_{1},\rho_{2}<1 \\
+      \left(\nu_{1}+\nu_{2}+1,\min\{\sigma_{1},\sigma_{2}\},1\right)                                                   & \text{ if }\rho_{1}=\rho_{2}=1                                \\
       (\nu_{1}+\nu_{2}+1-\frac{\rho}{2},(\sigma_{1}^{-\frac{1}{\rho-1}}+\sigma_{2}^{-\frac{1}{\rho-1}})^{1-\rho},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}>1.
@@
     \equiv\begin{cases}
-\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0\\
-\left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}>0\\
+      \left(\frac{1}{\mu}\left(\frac{\nu_{1}}{|\rho_{1}|}+\frac{\nu_{2}}{|\rho_{2}|}+\frac{1}{2}\right),\sigma,-\frac{1}{\mu}\right) & \text{ if }\rho_{1},\rho_{2}<0      \\
+      \left(\frac{1}{\mu}\left(\frac{\nu_{1}}{\rho_{1}}+\frac{\nu_{2}}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)      & \text{ if }\rho_{1},\rho_{2}>0      \\
       \mathcal{R}_{|\nu_1|}                                                                                                          & \mbox{ if }\rho_{1}\leq0,\rho_{2}>0 \\
@@
   \equiv \begin{cases}
-(\nu_{1}+\nu_{2},\sigma_{1},\rho_{1}) & \text{ if }\rho_{1}<\rho_{2}\\
-(\nu_{1}+\nu_{2},\sigma_{1}+\sigma_{2},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2}\\
+    (\nu_{1}+\nu_{2},\sigma_{1},\rho_{1})        & \text{ if }\rho_{1}<\rho_{2}      \\
+    (\nu_{1}+\nu_{2},\sigma_{1}+\sigma_{2},\rho) & \text{ if }\rho=\rho_{1}=\rho_{2} \\
     (\nu_{1}+\nu_{2},\sigma_{2},\rho_{2})        & \text{ otherwise.}
@@
 
-\section{List of univariate distributions}
+\subsection{List of univariate distributions}
+
+Here we provide an enumeration of common parametric distributions and their
+corresponding GGA parameterizations.
 
@@
     \begin{longtable}{|c|c|c|c|}
-\caption{List of univariate distributions}\label{tab:dist_list}\\
+      \caption{List of univariate distributions}\label{tab:dist_list}                                                                                                                                                                                                                                                                         \\
       \hline
@@
       \hline
-Gen. Normal distribution & $(-\infty,\infty)$ & $\frac{\beta}{2\alpha\Gamma(1/\beta)}\exp\left(-\left(\frac{|x-\mu|}{\alpha}\right)^{\beta}\right)$ & $(0,\alpha^{-\beta},\beta)$\tabularnewline
+      Gen.
+Normal distribution                                              & $(-\infty,\infty)$                                      & $\frac{\beta}{2\alpha\Gamma(1/\beta)}\exp\left(-\left(\frac{|x-\mu|}{\alpha}\right)^{\beta}\right)$                             & $(0,\alpha^{-\beta},\beta)$\tabularnewline
       \hline
@@
       \hline
-Inv. chi-squared distribution & $(0,\infty)$ & $\frac{2^{-k/2}}{\Gamma(k/2)}x^{-k/2-1}e^{-1/(2x)}$ & $(-\frac{k}{2}-1,\frac{1}{2},-1)$\tabularnewline
-\hline 
-Inv. gamma distribution & $(0,\infty)$ & $\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{-\alpha-1}e^{-\beta/x}$ & $(-\alpha-1,\beta,-1)$\tabularnewline
+      Inv.
+chi-squared distribution                                         & $(0,\infty)$                                            & $\frac{2^{-k/2}}{\Gamma(k/2)}x^{-k/2-1}e^{-1/(2x)}$                                                                             & $(-\frac{k}{2}-1,\frac{1}{2},-1)$\tabularnewline
+      \hline
+      Inv.
+gamma distribution                                               & $(0,\infty)$                                            & $\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{-\alpha-1}e^{-\beta/x}$                                                                & $(-\alpha-1,\beta,-1)$\tabularnewline
       \hline
@@
 
-\section{Proofs}\label{sec:proofs}
+\subsection{Proofs of new results}
+\label{sec:proofs}
 
@@
   \begin{align*}
-p_Z(z) &= \frac{1}{2\pi}\int_{\mathbb{R}}\left|x\right|^{-1}e^{-\frac{1}{2}x^{2}}e^{-\frac{1}{2}z^{2}/x^{2}}\dd x\\
-&=\frac{1}{\pi}\int_{0}^{\infty}x^{-1}e^{-\frac{1}{2}x^{2}-\frac{1}{2}z^{2}/x^{2}}\dd x\\
-&=\frac{1}{\pi}\int_{0}^{\infty}x^{-1}e^{-\frac{1}{2}x^{2}-\frac{1}{2}z^{2}/x^{2}}\dd x.
+    p_Z(z) & = \frac{1}{2\pi}\int_{\mathbb{R}}\left|x\right|^{-1}e^{-\frac{1}{2}x^{2}}e^{-\frac{1}{2}z^{2}/x^{2}}\dd x \\
+           & =\frac{1}{\pi}\int_{0}^{\infty}x^{-1}e^{-\frac{1}{2}x^{2}-\frac{1}{2}z^{2}/x^{2}}\dd x                    \\
+           & =\frac{1}{\pi}\int_{0}^{\infty}x^{-1}e^{-\frac{1}{2}x^{2}-\frac{1}{2}z^{2}/x^{2}}\dd x.
   \end{align*}
@@
   \begin{align*}
-p_{Z}(z)&=\int_{\mathbb{R}}\frac{1}{\sqrt{2\pi}x^{2}}e^{-\frac{1}{2x^{2}}}\frac{x^{2}}{\sqrt{2\pi}z^{2}}e^{-\frac{x^{2}}{2z^{2}}}\frac{1}{\left|x\right|}\dd x\\
-&=\frac{1}{2\pi z^{2}}\int_{\mathbb{R}}e^{-\frac{1}{2x^{2}}-\frac{x^{2}}{2z^{2}}}\frac{1}{\left|x\right|}\dd x\\
-&=\frac{1}{\pi z^{2}}\int_{0}^{\infty}e^{-\frac{1}{2x^{2}}-\frac{x^{2}}{2z^{2}}}\frac{1}{x}\dd x\\
-&=\frac{1}{\pi z^{2}}K_{0}(|z|^{-1}) \sim \sqrt{\frac{1}{2\pi}}|z|^{-3/2}e^{-|z|^{-1}},
+    p_{Z}(z) & =\int_{\mathbb{R}}\frac{1}{\sqrt{2\pi}x^{2}}e^{-\frac{1}{2x^{2}}}\frac{x^{2}}{\sqrt{2\pi}z^{2}}e^{-\frac{x^{2}}{2z^{2}}}\frac{1}{\left|x\right|}\dd x \\
+             & =\frac{1}{2\pi z^{2}}\int_{\mathbb{R}}e^{-\frac{1}{2x^{2}}-\frac{x^{2}}{2z^{2}}}\frac{1}{\left|x\right|}\dd x                                         \\
+             & =\frac{1}{\pi z^{2}}\int_{0}^{\infty}e^{-\frac{1}{2x^{2}}-\frac{x^{2}}{2z^{2}}}\frac{1}{x}\dd x                                                       \\
+             & =\frac{1}{\pi z^{2}}K_{0}(|z|^{-1}) \sim \sqrt{\frac{1}{2\pi}}|z|^{-3/2}e^{-|z|^{-1}},
   \end{align*}
@@
     \mathbb{P}(XY>x,X\notin[x^{a_{-}},x^{a_{+}}]) & \leq\mathbb{P}(X>x^{a_{+}})+\mathbb{P}(Y>x^{1-a_{-}})                                                                                                 \\
-& \sim c_{1}x^{\nu_{1}a_{+}}e^{-\sigma_{1}x^{\rho_{1}a_{+}}}+c_{2}x^{\nu_{2}(1-a_{-})}e^{-\sigma_{2}x^{\rho_{2}(1-a_{-})}}\\
+                                                  & \sim c_{1}x^{\nu_{1}a_{+}}e^{-\sigma_{1}x^{\rho_{1}a_{+}}}+c_{2}x^{\nu_{2}(1-a_{-})}e^{-\sigma_{2}x^{\rho_{2}(1-a_{-})}}                              \\
                                                   & \leq\left(c_{1}x^{\nu_{1}a_{+}}+c_{2}x^{\nu_{2}(1-a_{-})}\right)e^{-\min\{\sigma_{1},\sigma_{2}\}x^{(1+\epsilon)\rho}} = o(x^\nu e^{-\sigma x^\rho}),
@@
   \begin{align*}
-(\nu_{1},\sigma_{1},-\rho_{1})^{-1}\otimes(\nu_{2},\sigma_{2},-\rho_{2})^{-1}	&\equiv\left(-\nu_{1}-2,\sigma_{1},\rho_{1}\right)\otimes\left(-\nu_{2}-2,\sigma_{2},\rho_{2}\right)\\
-&\equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}-2}{\rho_{1}}+\frac{-\nu_{2}-2}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)\\
-&\equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}}{\rho_{1}}+\frac{-\nu_{2}}{\rho_{2}}-2\mu-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)\\
-&\equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}}{\rho_{1}}+\frac{-\nu_{2}}{\rho_{2}}-\frac{1}{2}\right)-2,\sigma,\frac{1}{\mu}\right),
+    (\nu_{1},\sigma_{1},-\rho_{1})^{-1}\otimes(\nu_{2},\sigma_{2},-\rho_{2})^{-1} & \equiv\left(-\nu_{1}-2,\sigma_{1},\rho_{1}\right)\otimes\left(-\nu_{2}-2,\sigma_{2},\rho_{2}\right)                                    \\
+                                                                                  & \equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}-2}{\rho_{1}}+\frac{-\nu_{2}-2}{\rho_{2}}-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right)  \\
+                                                                                  & \equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}}{\rho_{1}}+\frac{-\nu_{2}}{\rho_{2}}-2\mu-\frac{1}{2}\right),\sigma,\frac{1}{\mu}\right) \\
+                                                                                  & \equiv\left(\frac{1}{\mu}\left(\frac{-\nu_{1}}{\rho_{1}}+\frac{-\nu_{2}}{\rho_{2}}-\frac{1}{2}\right)-2,\sigma,\frac{1}{\mu}\right),
   \end{align*}
@@
 
+\section{Implementation}\label{sec:impl}
+
+\subsection{Compile-time static analysis}
+
+To illustrate an implementation of GGA for static analysis, we sketch the operation of the PPL compiler at a high-level
+and defer to the supplementary code for details.
+A probabilistic program is first inspected using Python's built-in \texttt{ast} module
+and transformed to static single assignment (SSA) form \citep{rosen1988global}.
+Next, standard compiler optimizations (e.g. dead code elimination, constant propagation)
+are applied and an execution of the optimized program is traced \citep{wingate2011lightweight,bingham2019pyro}
+and accumulated in a directed acyclic graph representation.
+A breadth-first type checking pass, as seen in Algorithm~\ref{alg:bfs_typecheck}, completes in linear time, and GGA results may be applied to implement \texttt{computeGGA()} using the following steps:
+\begin{itemize}[leftmargin=*]
+  \item If a node has no parents, then it is an atomic distribution and its tail parameters are known (\Cref{tab:dist_list})
+  \item Otherwise, the node is an operation taking its potentially stochastic inputs (parents) to its output. Consult \Cref{tab:gga_operations} for the output GGA tails.
+\end{itemize}
+
+%% This declares a command \Comment
+%% The argument will be surrounded by /* ... */
+% \SetKwComment{Comment}{/* }{ */}
+
+\begin{algorithm}
+  \caption{Pseudocode for a GGA tails static analysis pass}\label{alg:bfs_typecheck}
+  \begin{algorithmic}
+    \REQUIRE{Abstract syntax tree for a PPL program}
+
+    \STATE{frontier $\gets$ [rv : Parents(rv) = $\emptyset$]}\\
+    \STATE{GGAs $\gets \{\}$}
+
+    \WHILE{\text{frontier} $\neq \emptyset$}
+    \STATE{next $\gets$ frontier.popLeft()}\\
+    \STATE{GGAs[next] $\gets$ computeGGA(next.op, next.parent)}\\
+    \STATE{frontier $\gets$ frontier + next.children()}
+    \ENDWHILE
+
+    \RETURN{GGA parameter estimates for all random variables}
+  \end{algorithmic}
+\end{algorithm}
+
+\subsection{Representative distributions}\label{ssec:repr_dist}
+
+For each $(\nu,\sigma,\rho)$ we make a carefully defined choice of $p$ on $\mathbb{R}$ such that if $X \sim p$, then $X \equiv (\nu,\sigma,\rho)$.
+This way, any random variable $f(X)$, where $f$ is $1$-Lipschitz, will exhibit the correct tail, and so approximations of this form may be used for variational inference or density estimation.
+Let $X \equiv (\nu,\sigma,\rho)$ and $0 < \epsilon \ll 1$ denote a small parameter such that tails $e^{-x^\epsilon}$ are deemed to be ``very heavy'' (we chose $\epsilon = 0.1$).
+%We also let $X\equiv(\nu,\sigma,\rho)$ and $\zeta$ denote an independent Rademacher random variable with $\mathbb{P}(\zeta=1)=\mathbb{P}(\zeta=-1)=\frac12$.
+
+\begin{itemize}
+  \item[($\rho \leq 0$)] If $\rho \leq -1$, then $p_X(x) \sim c x^{-|\nu|}$.
+A prominent distribution on $\mathbb{R}$ with power law tails is the \emph{Student $t$ distribution}, in this case, with $|\nu|-1$ degrees of freedom if $\nu < -1$ (generate $X \sim t_{|\nu|-1})$.
+  \item[($\rho > \epsilon$)] For moderately sized $\rho > 0$, we consider a symmetrized variant of the generalized Gamma density (\Cref{eq:GenGammaDensity}).
+  \item[($\rho \leq \epsilon$)] If $X \equiv (\nu,\sigma,\rho)$ where $\rho$ is small, then $X$ will exhibit much heavier tails, and the generalized Gamma distribution in Case 1 will become challenging to sample from.
+In these cases, we expect that the tail of $X$ should be well represented by a power law.
+    %Letting $\alpha = \sigma \rho - \nu - 1$, using an approximation to the logarithm, $p_{|X}(x) \approx c x^{-\alpha - 1}$ (see the Supplementary Material for more details).
+Therefore, we represent tails of this form by the Student $t$ distribution with $\alpha$ degrees
+    The generalized Gamma density (\Cref{eq:GenGammaDensity}) satisfies $\mathbb{E}X^r = \sigma^{-r/\rho} \Gamma(\frac{\nu+1+r}{\rho})/\Gamma(\frac{\nu+1}{\rho})$ for $r > 0$.
+Let $\alpha > 0$ be such that $\mathbb{E}X^\alpha = 2$.
+By Markov's inequality, the tail of $X$ satisfies $\mathbb{P}(X>x)\leq 2 x^{-\alpha}$.
+Therefore, we can represent tails of this form by the Student $t$ distribution with $\alpha+1$ degrees of freedom (generate $X \sim t_{\alpha}$).
+\end{itemize}
+
+\subsection{Bulk correction by Lipschitz mapping}
+
+While a representative distribution will exhibit the desired tails,
+the target distribution's bulk may be very different from a generalized Gamma and result in poor distributional approximation.
+To address this, we propose splicing together the tails from a generalized Gamma with a flexible density approximation for the bulk.
+While many combinations are possible, in this work we rely on \Cref{lem:lipschitz} and post-compose neural
+spline flows \citep{durkan2019neural} (which are identity functions outside of a bounded interval) after
+properly initialized generalized Gamma distributions.
+Optimizing the parameters of the flow results in good bulk approximation while simultaneously
+preserving the tail correctness guarantees attained by the GGA.
+
+\begin{example}
+  Let $A \in \mathbb{R}^{k \times k}$, $x,y\in\mathbb{R}^k$, with $x_i,y_i,A_{ij} \overset{\text{iid}}{\sim} \mathcal{N}(-1,1)$.
+  The distribution of $x^\top A y = \sum_{i,j} x_i A_{ij} y_j$ is convolution of normal-powers \citep{gupta2008analyzing} and has no convenient
+  closed form expression.
+  Using GGA's closure theorems (\cref{tab:gga_operations}), one can compute
+  its tail parameters to be $(\frac{k}{2}-1,\frac{3}{2}, \frac{2}{3})$.
+
+  \begin{figure}[h]
+    \centering
+    \input{hta/experiments/normal-power-sum.pgf}
+    \input{hta/experiments/lip-nf.pgf}
+    \caption{(Top) $5000$ samples of $\lvert x^\top A y \rvert$ vs the
+      calibrated GGA density $q(x)$. While calibrated tails are provably guarantees,
+      the target distribution's bulk differs from the assumed generalized Gamma representative distribution (\cref{ssec:repr_dist}) for all $k$.
+      To fix the bulk approximation, a normalizing flow is composed with
+      the GGA representative to form $\text{flow}(q)(x)$. The bulk approximation
+      is improved (bottom) while the tails continue to exhibit the same
+      behavior (bottom right).
+    }
+    \label{fig:power_normal}
+  \end{figure}
+
+  The GGA representative is a gamma distribution with the correct tails, but there is non-negligible error in the bulk where $x$ is small. To address this, a learnable bijector can be
+  optimized as in \Cref{fig:power_normal} bottom left to correct the bulk approximation. Guaranteed by \Cref{lem:lipschitz} and
+  visualized in \Cref{fig:power_normal} bottom right, the tails of the overall composition remain calibrated.
+\end{example}
+
+\section{Experiments}\label{sec:hta:experiments}
+
+In this section we demonstrate that GGA-based density estimation yields improvements across a variety of metrics.
+We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushfowards of Normal distributions. To understand the individual
+effect of using a GGA base distribution over standard normals versus more expressive pushforward maps \citep{durkan2019neural}, we also report ablation results where
+normalizing flows are replaced by affine transforms as originally
+proposed in \citep{kucukelbir2017automatic}. All experiments are repeated for 50 trials, trained to convergence using the Adam optimizer with manually tuned learning rate,
+and conducted on i7-8700K CPU and GTX 1080 GPU hardware.
+
+All target distributions in this section are expressed as generative PPL programs:
+Cauchy using a reciprocal normal, Chi2 using a sum of squared normals, IG (Inverse Gamma)
+using a reciprocal exponential, Normal using a sum of normals, and StudentT using a
+normal and Cauchy ratio. Doing so tasks the static analyzer to infer the target's tails and makes the analysis non-trivial.
+See supplementary for full details.
+
+Our results in the following tables share a consistent narrative where a GGA base
+distribution rarely hurts and can significantly help with heavy tailed targets.
+Except for when targets are truly light tailed ($\alpha=\infty$ in Chi2 and Normal),
+GGA-based approximations are the only ones to reproduce appropriate GPD tail index $\hat\alpha$ in density estimation and achieve a passing Pareto $\hat{k}$ diagnostic
+\citep{yao2018yes} below $0.2$ in variational inference. When viewed through traditional
+evaluation metrics such as negative cross-entropy $H(p,q) = E_p \log p(X)$,
+ELBO $E_q \log \frac{q(X)}{p(X)}$,
+and importance-weighted autoencoder bound \citep{burda2015importance} $E_q \log \sum_i^{1000} \frac{p(X)}{q(X)}$, GGA-based approximations remain favorable on almost all heavy-tailed
+targets and have negligible difference for light tailed targets.
+Less surprising is the result that adding a flow improved approximation metrics, as we expect the additional representation flexibility to be beneficial.
+
+%\feynman{explain the targets, generative and marginalized, reference appendix}
+
+\textbf{Density Estimation} We minimize a Monte-Carlo estimate of the cross entropy $H(p,q) = -E_p[\log q(X)] \approx -\frac{1}{N} \sum_{i=1}^N \log q(x_i),\quad x_i \sim p$.
+The results are shown in \Cref{tab:de} along with power-law tail index estimates \citep{clauset2009power} $\hat\alpha$.
+Overall, we see that GGA performs better (lower NLL, $\hat\alpha$ closer to target) when the target
+has heavier tails (lower $\hat\alpha$ target/theory) and that the difference is smaller but still non-negligible
+for distributions such as Chi1 which possess tails heavier than Gaussian.
+
+\begin{table}[h]
+  \centering
+  \caption{Density estimation metrics attained (mean, standard deviation in parenthesis)
+    on targets of varying tail index (smaller $\alpha =$ heavier tails).
+    Higher negative cross entropy $-H(p,q) = E_p \log q(X)$ implies a better overall approximation (row maxes bolded)
+    while close agreement between the target Pareto tail index $\alpha$ \citep{clauset2009power}
+    and its estimate $\hat\alpha$ in $q(x)$ suggest calibrated tails (closest in row bolded).
+  }
+  \label{tab:de}
+  \input{experiments/de_table.tex}
+\end{table}
+
+\textbf{Variational Inference} The optimization objective is the ELBO
+\[
+  E_q \log \frac{p(X)}{q(X)} \approx \frac{1}{N} \sum_{i=1}^N \log \frac{p(x_i)}{q(x_i)},\quad x_i \sim q
+\]
+Here, the density $p$ must also be evaluated so for simplicity
+experiments in \cref{tab:vi} use closed-form marginalized densities for targets.
+The overall trends also show that GGA yields consistent improvements as measured by both ELBO and importance-weighted
+estimates of marginal likelihood and that the difference was greater when the tails of $p(z)$ were heavier.
+The $\hat{k}$ diagnostics \citep{yao2018yes} corroborate our findings that variational inference succeeds
+($\hat{k} < -1.2$) when a GGA with appropriately matched tails is used and fails ($\hat{k} > 1$) when
+Gaussian tails are erroneously imposed.
+
+\begin{table}[h]
+  \centering
+  \caption{Variational inference metrics (mean, standard deviation in parenthesis)
+    on targets of varying tail index (smaller $\alpha =$ heavier tails).
+Both the IWAE bound $E_q \log \sum_i^K \frac{p(X_i)}{q(X_i)}$ and the ELBO ($K=1$) measure (a lower bound) on the marginal likelihood where larger is better (row maxes bolded).
+In \citet{yao2018yes},
+    a Pareto $\hat{k}$ diagnostic $> 0.2$ is interpreted as potentially problematic so only values below are bolded.}
+  \label{tab:vi}
+  \small\input{experiments/vi_table.tex}
+\end{table}
+
+The targets in \Cref{tab:de} and \Cref{tab:vi} are analyzed using
+the GGA. Note that Inverse Gamma (``IG'') corresponds
+to the inverse exponential. We selected closed form targets so that
+the Pareto tail index $\alpha$ is known analytically and the quality
+of theoretical predictions as well as empirical results can be evaluated
+against. All experiments are repeated for $100$ trials and $1,000$ samples
+from the model (as well as the approximation in VI) were used to compute
+each gradient estimate. Losses were trained until convergence, which
+all occurred in under $10,000$ iterations at a $0.05$ learning rate and
+the Adam \citep{kingma2014adam} optimizer.
+
+\subsection{SGD for least-squares linear regression}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.32\textwidth]{figures/Kesten_0.40_1.5.png}
+  \includegraphics[width=0.32\textwidth]{figures/Kesten_0.50_1.5.png}
+  \includegraphics[width=0.32\textwidth]{figures/Kesten_0.50_2.png}
+  \caption{\label{fig:SGD}Density of iterates of SGD vs. predicted tail behaviour}
+\end{figure}
+
+For inputs $X$ and labels $Y$ from a dataset $\mathcal{D}$, the least squares estimator for linear regression satisfies $\beta = \min_\beta \tfrac12 \mathbb{E}_{X,Y\sim\mathcal{D}}(Y - X\beta)^2$.
+To solve for this estimator, one can apply stochastic gradient descent (SGD) sampling over independent $X_k,Y_k\sim \mathcal{D}$ to obtain the sequence of iterations
+\[
+  \beta_{k+1} = (I - \delta X_k X_k^\top) \beta_k + \delta Y_k X_k
+\]
+for a step size $\delta > 0$.
+For large $\delta$, the iterates $\beta_k$ typically exhibit heavy-tailed fluctuations; in this regard, this sequence of iterates has been used as a simple model for more general stochastic optimization dynamics \citep{gurbuzbalaban2021heavy,hodgkinson2021multiplicative}.
+In particular, generalization performance has been tied to the heaviness of the tails in the iterates \cite{simsekli2019tail}.
+Here we use our algebra to predict the tail behaviour in a simple one-dimensional setting where $X_k \sim \mathcal{N}(0,\sigma^2)$ and $Y_k \sim \mathcal{N}(0,1)$.
+From classical theory \citep{buraczewski2016stochastic}, it is known that $X_k$ converges in distribution to a power law with tail exponent $\alpha > 0$ satisfying $\mathbb{E}|1 - \delta X_k^2|^\alpha = 1$.
+In \cref{fig:SGD}, we plot the density of the representative obtained using our algebra after $10^4$ iterations against a kernel density estimate of the first $10^6$ iterates when $\sigma \in \{0.4,0.5\}$ and $\delta \in \{1.5,2.0\}$.
+In all cases, the density obtained from the algebra provides a surprisingly close fit.
+
+\subsection{Normal target}
+
+Consider the toy example of a Normal target. This case is trivial for
+Gaussian based methods and is oftentimes the initialization.
+This lack of approximation gap in ADVI is seen in \Cref{fig:normal},
+where we also see that GGA achieves similar approximation quality.
+This is unsurprising as the GGA approximation in \Cref{tab:dist_list}
+is also a Normal distribution.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\textwidth]{figures/de_normal_sum.png}
+  \includegraphics[width=.4\textwidth]{figures/vi_normal_sum.png}
+  \caption{Density estimation and VI against a known normal target}
+  \label{fig:normal}
+\end{figure}
+
+\subsection{Chi-square}
+
+Now let $X_{ij} \sim N(0,1)$ and consider $tr X^\top X$.
+Such quantities arise in the analysis of random projections.
+It is important here to recognize that the power operation $X \mapsto X^2$ is not equivalent to the multiplication operation $X \mapsto X \otimes X$, as multiplication assumes independence.
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{figures/jl-chi2.png}
+  \caption{$5000$ samples of JL matrix trace (blue) vs GGA prediction (yellow)}
+  \label{fig:jl_chi2}
+\end{figure}
+
+\section{Conclusion}\label{sec:hta:conclusion}
+
+In this work, we have proposed a novel systematic approach for conducting tail inferential static analysis by implementing a three-parameter generalized Gamma algebra into a PPL compiler.
+Initial results are promising, showing that improved inference with simpler approximation families is possible when combined with tail metadata.
+While already useful, the generalized Gamma algebra and its implementation currently has
+some notable limitations:
+\begin{itemize}[leftmargin=*]
+  \item Since the algebra assumes independence, handling of dependencies between defined random variables must be conducted externally.
+This will inevitably require interoperability with a symbolic package to decompose complex expressions into operations on independent random variables.
+  \item The GGA is formulated for univariate distributions only. Suitably defining multivariate tails is an open problem with interesting alternatives \citep{jaini2020tails,ftvi} all of which
+        could extend GGA to higher dimensions.
+  \item Conditioning is arguably the most important feature of a PPL and what distinguishes it from a glorified simulator.
+Exact marginalization in general is NP-hard \citep{koller2009probabilistic}, so treatment of conditional distributions using symbolic manipulations is a significant open problem, with some basic developments \cite{shan2017exact,cho2019disintegration}.
+Since only the tails are required in our setup, it may be possible to construct a dual algebra for operations under conditioning; this is left for future work.
+  \item Compile-time static analysis only applicable to fixed model structure. While out of scope for our current work, open-universe models \citep{milch2010extending}
+        and PPLs to support them \citep{bingham2019pyro} are an important research direction.
+  \item The most significant omission to the algebra itself is classification of log-normal tails; while addition may be treated using \cite{gulisashvili2016tail} for example, multiplicative convolution with log-normal tails remains elusive.
+  \item At present, reciprocals are approximated by assuming behaviour near zero. Reciprocals may be better treated by covering near-zero asymptotics separately.
+\end{itemize}
+
+The GGA provides a necessary first step into the static analysis of tails in
+a probabilistic program.
+As the above limitations are improved in future work and GGA becomes more broadly applicable,
+we are excited to see how improved tail modelling will improve downstream PPL applications
+as well as
+other researchers will utilize GGA metadata to develop
+novel PPL applications.
+
+% \begin{ack}
+% 	Use unnumbered first level headings for the acknowledgments. All acknowledgments
+% 	go at the end of the paper before the list of references. Moreover, you are required to declare
+% 	funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
+% 	More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2021/PaperInformation/FundingDisclosure}.
+
+% 	Do {\bf not} include this section in the anonymized submission, only in the final paper.
+You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
+% \end{ack}
+
+% \subsection{Limitations}
+
+% \todo{Liam: Log-Normal insensitivity, edge case (see example)}
+
+% The GGA is formulated for univariate distributions only.
+
+% {
+% \small
+% \bibliographystyle{unsrtnat}
+% \bibliography{refs}
+% }
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section*{Checklist}
+
+% %%% BEGIN INSTRUCTIONS %%%
+% The checklist follows the references.  Please
+% read the checklist guidelines carefully for information on how to answer these
+% questions.  For each question, change the default \answerTODO{} to \answerYes{},
+% \answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
+% justification to your answer}, either by referencing the appropriate section of
+% your paper or providing a brief inline description.  For example:
+% \begin{itemize}
+%   \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
+%   \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
+%   \item Did you include the license to the code and datasets? \answerNA{}
+% \end{itemize}
+% Please do not modify the questions and only use the provided macros for your
+% answers.  Note that the Checklist section does not count towards the page
+% limit.  In your paper, please delete this instructions block and only keep the
+% Checklist section heading above along with the questions/answers below.
+% %%% END INSTRUCTIONS %%%
+
+% \begin{enumerate}
+
+% \item For all authors...
+% \begin{enumerate}
+%   \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
+%     \answerYes{}
+%   \item Did you describe the limitations of your work?
+%     \answerYes{See \cref{sec:hta:conclusion}}
+%   \item Did you discuss any potential negative societal impacts of your work?
+%     \answerYes{See \cref{sec:hta:conclusion}}
+%   \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
+%     \answerYes{}
+% \end{enumerate}
+
+% \item If you are including theoretical results...
+% \begin{enumerate}
+%   \item Did you state the full set of assumptions of all theoretical results?
+%     \answerYes{Stated at start of \cref{sec:gga}}
+% 	\item Did you include complete proofs of all theoretical results?
+%     \answerYes{Most are deferred to supplementary materials}
+% \end{enumerate}
+
+% \item If you ran experiments...
+% \begin{enumerate}
+%   \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
+%     \answerYes{See supplemental, a URL for code will be provided witn camera ready}
+%   \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
+%     \answerYes{}
+% 	\item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
+%     \answerYes{Standard errors reported in parenthesis on \cref{tab:de} and \cref{tab:vi}}
+% 	\item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
+%     \answerYes{See \cref{sec:hta:experiments}}
+% \end{enumerate}
+
+% \item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
+% \begin{enumerate}
+%   \item If your work uses existing assets, did you cite the creators?
+%     \answerYes{See citations of \citep{tehrani2020bean}}
+%   \item Did you mention the license of the assets?
+%     \answerYes{Licence included in code}
+%   \item Did you include any new assets either in the supplemental material or as a URL?
+%     \answerYes{See supplemental material}
+%   \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
+%     \answerNA{}
+%   \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
+%     \answerNA{}
+% \end{enumerate}
+
+% \item If you used crowdsourcing or conducted research with human subjects...
+% \begin{enumerate}
+%   \item Did you include the full text of instructions given to participants and screenshots, if applicable?
+%     \answerNA{}
+%   \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
+%     \answerNA{}
+%   \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
+%     \answerNA{}
+% \end{enumerate}
+
+% \end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \ifdefined\remappendix
+% \newpage
+
+% \appendix
+
 % \fi

### 2022-11-23 ic/aistats2021/paper.tex
Working compile

@@
 % \documentclass[twoside]{article}
-\documentclass[thesis.tex]{subfiles}
+\documentclass[../../thesis.tex]{subfiles}
 
@@
 % \begin{abstract}
+
+\chapter{Accelerating Metropolis-hastings with lightweight inference compilation}
+\label{ch:lic}
+
+While the subsequent chapters depart from a previously common theme of DPPs,
+they continue our study of statistical applications of randomized methods.
+In this chapter,
+we are concerned with the problem of sampling intractable posteriors of
+Bayesian graphical models.
 In order to construct accurate proposers for Metropolis-Hastings Markov
@@
 a Bayesian logistic regression and $n$-schools inference application.
+Parts of this chapter were originally published in
+\fullcite{lic}.
 % \end{abstract}
@@
 A reference implementation for LIC and code to reproduce our experiments
-have been made publically available \citep{lic2020}.
+have been made publically available\footnote{\url{https://github.com/facebookresearch/lightweight-inference-compilation}}.
 
@@
 an effective sample size almost $5\times$ smaller than that produced by LIC.
-\begin{figure}
+\begin{table}
   \centering
@@
   }\label{fig:nuisance}
-\end{figure}
+\end{table}
 

### 2022-11-23 intro.tex
Working compile

@@
-% !TeX root = thesis.tex
+\documentclass[thesis.tex]{subfiles}
+
+\begin{document}
+
 \chapter{Introduction}
+\label{ch:intro}
 
+When faced with difficult problems, randomized approximations are a tool
+commonly employed by quantitative scientists which can offer algernative
+algorithms and analyses. In this dissertation, we consider a range of problems
+at the intersection of randomized methods and statistics.
+Throughout our work, randomized methods are a unifying theme used in the first section to
+construct tractable approximations and later as tools for automating computational Bayesian statistics.
+
+Initially in the first three chapters, we develop novel analyses to recent problems
+in experimental design, double descent, and random projections by using determinantal
+point processes (DPP).
 In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
 using an adaptive row sampling algorithm based on DPPs and show that it provides
-good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
-applied here, in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
+good approximations. Through generalizing the previous chapter's proof techniques,
+in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
 closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
@@
 expected projections $X_S^\dag X_S$ when $X_S = S X$ is obtained by sub-Gaussian sketching matrix $S$.
+These chapters correspond to the following publications:
+\begin{itemize}
+    \item \fullcite{bayesian-experimental-design}
+    \item \fullcite{surrogate-design}
+    \item \fullcite{precise-expressions}
+\end{itemize}
 
+In the later chapters we focus on probabilistic programming and developing theory and tools
+to automate statistical inference using randomized algorithms based on
+Monte Carlo Markoc chain (MCMC) and variational inference (VI).
 The second three chapters concern the setting of probabilistic programming, a computational tool
@@
 improves the stability and convergence of a number of inference tasks.
+These chapters correspond to the following publications:
+\begin{itemize}
+    \item \fullcite{lic}
+    \item \fullcite{ftvi}
+    \item \fullcite{gga}
+\end{itemize}
+
+\end{document}

### 2022-12-09 abstract.tex
Submitted for alan review

@@
     to automate statistical inference using randomized algorithms based on
-    Monte Carlo Markoc chain (MCMC) and variational inference (VI).
-    \Cref{ch:lic} considers parameterizing $q(x)$ by graph neural networks which condition each
+    Monte Carlo Markov chain (MCMC) and variational inference (VI).
+    \Cref{ch:lic} considers parameterize $q(x)$ by graph neural networks which condition each
     node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over

### 2022-12-09 design/continuous_nips/continuous.tex
Submitted for alan review

@@
 performance of learning models which occurs at an \emph{interpolation
-  thershold}, i.e., a point where training error goes to zero (as one varies the
+  threshold}, i.e., a point where training error goes to zero (as one varies the
 ratio between the model complexity and the sample size). Moving away from this
@@
 that every $1\times 1$ random matrix is determinant preserving simply
-because taking a determinant is an identity transfomation in one
+because taking a determinant is an identity transformation in one
 dimension. Similarly, every fixed matrix is determinant preserving because
@@
 \end{equation}
-almost surely for some $m(z)$ to be defined; (ii) show that both $\delta(z)$ and its derivate $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take $z \to 0^+$ in \eqref{eq:limit2} and state
+almost surely for some $m(z)$ to be defined; (ii) show that both $\delta(z)$ and its derivative $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take $z \to 0^+$ in \eqref{eq:limit2} and state
 \begin{equation}\label{eq:limit3}

### 2022-12-09 design/equivalents_nips/neurips_2020.tex
Submitted for alan review

@@
 Subspace Newton and the Generalized Kaczmarz method, relies on
-projecting the Hessian downto a low-dimensional subspace, which makes our
+projecting the Hessian down to a low-dimensional subspace, which makes our
 results directly applicable. A related family of methods uses
@@
 establish the accuracy of our surrogate expressions for the low-rank approximation
-error. Similary as in Figure \ref{f:nystrom}, we use the sketched
+error. Similarly as in Figure \ref{f:nystrom}, we use the sketched
 Nystr\"om method \citep{revisiting-nystrom} with the RBF kernel

### 2022-12-09 ftvi/ftvi_icml/ftvi_icml.tex
Submitted for alan review

@@
 by itself does not preclude transformations of fat-tailed base distributions to light-tailed targets.
-Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushfoward,
+Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushforward,
 hence heavy-to-light transformations are impossible. Note here there is no ascending union over
@@
   Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
-  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$.
+  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointedness of $\mathcal{E}$ and $\mathcal{L}$.
 \end{proof}
@@
 using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
-used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empkkirically observed
+used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empirically observed
 this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
@@
 
-To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021 \citep{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010 \citep{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events.
+To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021 \citep{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicaid (CMS) claims during 2008--2010 \citep{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events.
 Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}.

### 2022-12-09 hta/neurips_2022.tex
Submitted for alan review

@@
 For example, importance sampling estimators can exhibit infinite variance if the tail of the approximating density is lighter than the target.
-Most prominant black-box variational inference methods are incapable of changing their tail behaviour from an initial proposal distribution \citep{jaini2020tails,ftvi}.
+Most prominent black-box variational inference methods are incapable of changing their tail behaviour from an initial proposal distribution \citep{jaini2020tails,ftvi}.
 MCMC algorithms may also lose ergodicity when the tail of the target density falls outside of a particular family \citep{roberts1996exponential}.
@@
 equivalently regularly varying distributions \citep{tajvidi2003confidence}
-have recieved attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
+have received attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
 Heavy tails can impact probabilistic machine learning methods in a number of ways.
@@
 without executing the underlying program.
-As dynamic analysis in PPs is less reliable due to non-determinism, static analysis
-methods for PPs become increasingly important.
+As dynamic analysis in PPLs is less reliable due to non-determinism, static analysis
+methods for PPLs become increasingly important.
 
@@
 In this section we demonstrate that GGA-based density estimation yields improvements across a variety of metrics.
-We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushfowards of Normal distributions. To understand the individual
+We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushforwards of Normal distributions. To understand the individual
 effect of using a GGA base distribution over standard normals versus more expressive pushforward maps \citep{durkan2019neural}, we also report ablation results where

### 2022-12-09 ic/aistats2021/paper.tex
Submitted for alan review

@@
 
-\chapter{Accelerating Metropolis-hastings with lightweight inference compilation}
+\chapter{Accelerating Metropolis-Hastings with lightweight inference compilation}
 \label{ch:lic}
@@
 A reference implementation for LIC and code to reproduce our experiments
-have been made publically available\footnote{\url{https://github.com/facebookresearch/lightweight-inference-compilation}}.
+have been made publicly available\footnote{\url{https://github.com/facebookresearch/lightweight-inference-compilation}}.
 
@@
 \citep{le2017inference} are omitted because they are not comparable due to
-lack of support for vector-valued random variables in the publically
+lack of support for vector-valued random variables in the publicly
 available reference implementation \citep{pyprob2020}. All methods yield

### 2022-12-09 intro.tex
Submitted for alan review

@@
 When faced with difficult problems, randomized approximations are a tool
-commonly employed by quantitative scientists which can offer algernative
+commonly employed by quantitative scientists which can offer alternative
 algorithms and analyses. In this dissertation, we consider a range of problems
@@
 to automate statistical inference using randomized algorithms based on
-Monte Carlo Markoc chain (MCMC) and variational inference (VI).
+Monte Carlo Markov chain (MCMC) and variational inference (VI).
 The second three chapters concern the setting of probabilistic programming, a computational tool

### 2022-12-09 check_spelling.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+
+Files=(
+    thesis.tex \
+    intro.tex \
+    design/aistats/bayesian.tex \
+    design/continuous_nips/continuous.tex \
+    design/equivalents_nips/neurips_2020.tex \
+    ic/aistats2021/paper.tex \
+    ftvi/ftvi_icml/ftvi_icml.tex \
+    hta/neurips_2022.tex
+)
+for f in ${Files[@]}; do
+    aspell \
+        --lang=en \
+        --mode=tex \
+        --conf=./tex_conf \
+        --add-extra-dicts=./aspelling_dict \
+        check \
+        $f
+done;
